### 2013-07-18 2013specconstr/Main.tex
specconstr paper

\documentclass{tmr}
\usepackage{style/utils}
\usepackage{style/code}

\title{Constructor specialisation in GHC}
\author{Amos Robinson\email{amosr@cse.unsw.edu.au}}

\begin{document}
\makeatactive

\begin{introduction}
In Haskell, loops, even those of higher-order combinators such as @map@, are expressed as recursive functions. 
With boxed objects such as @Int@, each iteration ends up allocating new objects in memory, only to be deconstructed immediately in the next iteration.
For high-performance code, these spurious allocations turn out to have serious penalties.

This problem has been solved specifically for boxed objects such as @Int@, @Float@ and others in the worker/wrapper transform~\CITE.
However, in the case of code produced by stream fusion~\CITE, other types such as @Either@ and tuples are used as loop state.
Constructor specialisation, or SpecConstr, is a generalisation of worker/wrapper for arbitrary types.
This is able to remove such allocations in stream-fusion code, and in some circumstances, produce object code that competes with hand-written C.
\end{introduction}

\section{}

SpecConstr

ForceSpecConstr, termination

And not specialising ``too many times'' on recursive types

Code blowup

Seeding specialisation of non-exported functions

\end{document}

### 2013-07-19 2013specconstr/Main.tex
ok

@@
 \begin{introduction}
-In Haskell, loops, even those of higher-order combinators such as @map@, are expressed as recursive functions. 
-With boxed objects such as @Int@, each iteration ends up allocating new objects in memory, only to be deconstructed immediately in the next iteration.
+In Haskell, loops are expressed as recursive functions. 
+When loop variables are updated, each iteration ends up allocating new objects in memory, only to be deconstructed immediately in the next iteration.
 For high-performance code, these spurious allocations turn out to have serious penalties.
@@
 This problem has been solved specifically for boxed objects such as @Int@, @Float@ and others in the worker/wrapper transform~\CITE.
-However, in the case of code produced by stream fusion~\CITE, other types such as @Either@ and tuples are used as loop state.
+However, in the case of code produced by stream fusion~\CITE, other types such as @Either@ and tuples are used as loop variables.
 Constructor specialisation, or SpecConstr, is a generalisation of worker/wrapper for arbitrary types.
@@
 
-\section{}
+\section{Background}
+\subsection{Core}
+Core is GHC's intermediate language, an explicitly typed System-F style language.
+After the source program has been parsed, its types checked and inferred, it is converted to core.
+The majority of optimisations are then performed on the core code,
+before eventually being converted to object code.
+
+To an experienced Haskell programmer, learning to read core should take little effort, and is a valuable skill for writing high-performance code.
+Where clauses are turned into @let@s, or @letrec@s for recursive expressions.
+Pattern matching and guards are converted to explicit @case@ expressions.
+
+To illustrate this, there follows an example implementation of @foldl@, and its core equivalent.
+\begin{code}
+foldl :: (a -> b -> a) -> a -> [b] -> a
+foldl k z xs = go xs z
+ where
+  go []     acc = acc
+  go (x:xs) acc = go xs (k acc x)
+\end{code}
+
+The core is quite a bit noisier.
+
+\begin{code}
+foldl = \@a @b k z xs ->
+ letrec {
+     go = \xs' acc ->
+       case xs' of {
+         []     -> acc;
+         : x xs -> go xs (k acc x)
+       }
+ } in go xs z
+\end{code}
+
+\subsection{Unboxed types}
+In GHC, the @Int@ type is defined as a wrapper around an unboxed int,
+and unboxed int literals and operators have @#@ appended to their name.
+
+\section{Motivation}
+Consider a simple example of summing a list of @Int@s.
+\begin{code}
+sum :: [Int] -> Int
+sum = foldl (+) 0
+\end{code}
+
+After inlining @foldl@, we get a recursive worker function.
+\begin{code}
+data Int = I# Int#
+
+sum xs = go xs (I# 0#)
+ where
+  go []          (I# acc) = I# acc
+  go ((I# x):xs) (I# acc) = go xs (I# (acc +# x))
+\end{code}
+
+\begin{code}
+dotp :: Vector Double -> Vector Double -> Double
+dotp as bs = go (Nothing, 0) 0
+ where
+  go (_, i) acc
+   | i > V.length as
+   = acc
+  go (Nothing, i) acc
+   = go (Just (as!i), i) acc
+  go (Just a, i) acc
+   = go (Nothing, i + 1) (acc + (a * bs!i))
+\end{code}
+
+\section{Constructor specialisation}
 

### 2013-07-24 2013specconstr/Main.tex
ok

@@
 
-The core is quite a bit noisier.
+The core is quite a bit noisier,
+but it is easy to see that the idea is the same.
 
@@
 
+\subsection{Simplifier}
+The simplifier is the workhorse of GHC's optimisations.
+It includes many different local transforms such as inlining, case-of-constructor, etc.
+
+\subsection{Rewrite rules}
+
 \subsection{Unboxed types}
@@
 and unboxed int literals and operators have @#@ appended to their name.
+
+\begin{code}
+data Int = I# Int#
+
+(+) :: Int -> Int -> Int
+(+) (I# x) (I# y) = I# (x +# y)
+\end{code}
+
+With sufficient inlining and case-of-constructor simplification,
+an arithmetic expression can be reduced to as few allocations as possible.
+
+In the following sections, ints will be boxed and unboxed explicitly to make reductions in allocations more obvious.
 
@@
 \begin{code}
-data Int = I# Int#
-
 sum xs = go xs (I# 0#)
@@
 
+\subsection{Code blowup}
+\subsection{Termination}
+\subsection{ForceSpecConstr}
+
 ForceSpecConstr, termination

### 2014-03-15 2014betterfusionforfilters/Main.tex
initial commit of fusion paper

\documentclass[preprint]{sigplanconf}
\usepackage{version}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathptmx}
\usepackage{hyperref}
\usepackage{alltt}
\usepackage{url}
\usepackage{float}
\usepackage{style/utils}
\usepackage{style/code}
\usepackage{style/proof}

% -----------------------------------------------------------------------------
\begin{document}
\title{Better fusion for filters}

\authorinfo{ 
  Amos Robinson$^\dagger$ 
  \and Ben Lippmeier$^\dagger$
  \and Manuel M. T. Chakravarty$^\dagger$
  \and Gabriele Keller$^\dagger$ 
}{
  \vspace{5pt}
  \shortstack{
    $^\dagger$Computer Science and Engineering \\
    University of New South Wales, Australia \\[2pt]
    \textsf{\{amosr,benl,chak,keller\}@cse.unsw.edu.au}
  }
}

\maketitle
\makeatactive

\begin{abstract}
Our recent work on flow fusion shows how to extract a single imperative loop from a set of combinators, but if multiple loops are required, finding the optimal clustering is non-trivial.
The optimal clustering must maximise cache locality, minimise memory traffic, and minimise loop overhead.
Given the sheer number of possible clusterings an exhaustive search is not feasible, and greedy algorithms tend to only find an approximate solution.
Instead, we show how to formulate the problem as an \emph{integer linear programming} problem, and use an external library to find the optimal solution.
\end{abstract}

\category
	{D.3.3}
	{Programming Languages}
	{Language Constructs and Features---Concurrent programming structures; Control structures; Abstract data types}

\terms
	Languages, Performance

\keywords
	Arrays, Fusion, Haskell

\input{section/01-Introduction.tex}

\section{Acceptable functions}
We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.
This form is a list of combinator bindings, where variables are split into scalar and array variables.
The main restriction is that worker functions may only reference scalar variables, and thus do not perform array combinators.

The only combinators we fuse are @map@, @filter@, @fold@ and @gather@.
Most of these are standard combinators except for @gather@, which is equivalent to @gather xs ys = map (index xs) ys@.
However, as we support no @index@ operation, @gather@ is implemented as a primitive.

Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@. This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
The @external@ also signifies variables produced by a non-primitive combinators.
Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points. They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing.

It is important to note, however, that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
\input{figures/Program.tex}

\section{Size inference}

\section{Integer linear program formulation}

\subsection{Cost function}
We argue that our cost function makes intuitive sense, by reducing the number of manifest arrays and loops.
However, due to the nature of our heuristic cost function we have no strong guarantees that the clustering with the lowest cost will actually be the clustering with the shortest runtime.
In general, since we do not know the expected sizes of the arrays, this is impossible to guarantee correctly.
It is possible that for certain restricted subsets of the language, a more sophisticated cost function may be able to guarantee optimality.
This is left to future work.

\section{Related work}

\subsection{Haskell short-cut fusion}
Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.

Consider the following @filterMax@ function, where each element in the input array is incremented, then the maximum is found, and the array is filtered to those greater than zero.
In this case, the result of the map @vs'@ cannot be inlined into both occurences without duplicating work, so no fusion can be performed.
This has the effect of performing three loops instead of one, with two arrays instead of one.

\begin{code}
filterMax vs =
 let vs' = map    (+1)  vs
     m   = fold   max 0 vs'
     flt = filter (>0)  vs'
\end{code}

\subsection{Integer linear programming in imperative languages}
The idea of using integer linear programming to find optimal fusion clusterings is not new, and has been discussed for imperative languages before.
These methods first construct a \emph{loop dependence graph} (LDG) from a given program, and then use this graph to create the integer linear program.
The LDG has nodes for each loop in the program, and edges between loops are dependencies.
Edges may be fusible, or fusion-preventing, in which case the two nodes may not be merged together.

Talk about the simple formulation by Darte\cite{darte2002new}.
Has an integer variable for each node, denoting the number of the cluster it's in.
Also includes a binary variable for each node, which is whether the node is fused with all its successors - in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes. Explain why.

The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ - that is, no fusion is possible.
This alone is not enough to guarantee a valid clustering.
To constrain the solution to acyclic and precedence preserving clusterings, a variable $\pi_i$ is added for each node $i$.
Constraints are added that require two nodes $i,j$ to have $\pi_i = \pi_j$ if $x_{ij} = 0$, and otherwise $\pi_i > \pi_j$ if $i$ is after $j$.
For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.

The difference to our combinator-based approach is that with combinators we retain more information about the meaning of the program.

\section{Conclusion}

\bibliographystyle{plain}
\bibliography{Main}

\end{document}

### 2014-03-15 2014betterfusionforfilters/section/01-Introduction.tex
initial commit of fusion paper

\section{Introduction}
Combinators are a natural and composable way of expressing array programs, with many correctness benefits over imperative programs.
If compiled naively, however, executing the combinator program will often require many more loops and intermediate arrays than an imperative program written by hand.
These intermediate arrays can often be removed by merging several combinators into a single loop.
This merging is known as \emph{fusion}.

Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
However, for scientific applications where efficiency is important, we cannot afford these limitations.
We have instead implemented our fusion system as a compiler plugin, which gives us total control of when fusion will occur.

Our recent work on flow fusion\cite{lippmeier2013data} generates a single imperative loop for a set of combinators, but when multiple loops are required, deciding which combinators to group together is difficult.
The optimal clustering must maximise cache locality, minimise memory traffic, and minimise loop overhead.
For most functions, there are so many possible clusterings that finding the optimal by exhaustive search is simply not feasible.

We must also decide which clustering is the `best' or most optimal. 
One obvious criterion for this is the minimum number of loops, but there may even be multiple clusterings with the minimum number of loops.
In this case, the number of required manifest arrays must also be taken into account.
Consider the @normaliseInc@ function below; it requires at least two loops, since @sum@ must be computed from the entire array before any of the output array @norm@ can be filled.
Scheduling @incs@ and @sum@ into a single loop, then @norm@ into another loop only requires two loops, but requires @incs@ to be stored in a manifest array before being read again by @norm@. If we were to schedule @sum@ in its own loop and @incs@ and @norm@ together, we would only require one array to be filled - the output array.

\begin{code}
normaliseInc (us : Vector Int) =
 let incs = map  (+1)    us
     sum  = fold (+) 0   us
     norm = map  (/sum)  incs
 in  norm
\end{code}

### 2014-03-15 2014betterfusionforfilters/Main.tex
reorganise

@@
 \input{section/01-Introduction.tex}
-
-\section{Acceptable functions}
-We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.
-This form is a list of combinator bindings, where variables are split into scalar and array variables.
-The main restriction is that worker functions may only reference scalar variables, and thus do not perform array combinators.
-
-The only combinators we fuse are @map@, @filter@, @fold@ and @gather@.
-Most of these are standard combinators except for @gather@, which is equivalent to @gather xs ys = map (index xs) ys@.
-However, as we support no @index@ operation, @gather@ is implemented as a primitive.
-
-Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@.
-This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
-The @external@ also signifies variables produced by a non-primitive combinators.
-Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points.
-They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing.
-
-It is important to note, however, that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
-None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
-\input{figures/Program.tex}
-
-\section{Size inference}
-
-\section{Integer linear program formulation}
-
-\subsection{Cost function}
-We argue that our cost function makes intuitive sense, by reducing the number of manifest arrays and loops.
-However, due to the nature of our heuristic cost function we have no strong guarantees that the clustering with the lowest cost will actually be the clustering with the shortest runtime.
-In general, since we do not know the expected sizes of the arrays, this is impossible to guarantee correctly.
-It is possible that for certain restricted subsets of the language, a more sophisticated cost function may be able to guarantee optimality.
-This is left to future work.
-
-\section{Related work}
-
-\subsection{Haskell short-cut fusion}
-Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
-This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
-
-Consider the following @filterMax@ function, where each element in the input array is incremented, then the maximum is found, and the array is filtered to those greater than zero.
-In this case, the result of the map @vs'@ cannot be inlined into both occurences without duplicating work, so no fusion can be performed.
-This has the effect of performing three loops instead of one, with two arrays instead of one.
-
-\begin{code}
-filterMax vs =
- let vs' = map    (+1)  vs
-     m   = fold   max 0 vs'
-     flt = filter (>0)  vs'
-\end{code}
-
-\subsection{Integer linear programming in imperative languages}
-The idea of using integer linear programming to find optimal fusion clusterings is not new, and has been discussed for imperative languages before.
-These methods first construct a \emph{loop dependence graph} (LDG) from a given program, and then use this graph to create the integer linear program.
-The LDG has nodes for each loop in the program, and edges between loops are dependencies.
-Edges may be fusible, or fusion-preventing, in which case the two nodes may not be merged together.
-
-Talk about the simple formulation by Darte\cite{darte2002new}.
-Has an integer variable for each node, denoting the number of the cluster it's in.
-Also includes a binary variable for each node, which is whether the node is fused with all its successors - in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
-The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
-It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
-As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
-The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes. Explain why.
-
-The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
-For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
-Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
-If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ - that is, no fusion is possible.
-This alone is not enough to guarantee a valid clustering.
-To constrain the solution to acyclic and precedence preserving clusterings, a variable $\pi_i$ is added for each node $i$.
-Constraints are added that require two nodes $i,j$ to have $\pi_i = \pi_j$ if $x_{ij} = 0$, and otherwise $\pi_i > \pi_j$ if $i$ is after $j$.
-For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.
-
-The difference to our combinator-based approach is that with combinators we retain more information about the meaning of the program.
-
+\input{section/02-CNF.tex}
+\input{section/03-SizeInference.tex}
+\input{section/04-ILP.tex}
+\input{section/05-Related.tex}
 \section{Conclusion}

### 2014-03-15 2014betterfusionforfilters/section/02-CNF.tex
reorganise

\section{Combinator normal form}
We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.
This form is a list of combinator bindings, where variables are split into scalar and array variables.
The main restriction is that worker functions may only reference scalar variables, and thus do not perform array combinators.

The only combinators we fuse are @map@, @filter@, @fold@ and @gather@.
Most of these are standard combinators except for @gather@, which is equivalent to @gather xs ys = map (index xs) ys@.
However, as we support no @index@ operation, @gather@ is implemented as a primitive.

Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@. This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
The @external@ also signifies variables produced by a non-primitive combinators.
Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points. They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing.

It is important to note, however, that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
\input{figures/Program.tex}

### 2014-03-15 2014betterfusionforfilters/section/04-ILP.tex
reorganise

\section{Integer linear program formulation}

\subsection{Cost function}
We argue that our cost function makes intuitive sense, by reducing the number of manifest arrays and loops.
However, due to the nature of our heuristic cost function we have no strong guarantees that the clustering with the lowest cost will actually be the clustering with the shortest runtime.
In general, since we do not know the expected sizes of the arrays, this is impossible to guarantee correctly.
It is possible that for certain restricted subsets of the language, a more sophisticated cost function may be able to guarantee optimality.
This is left to future work.

\subsection{Proof}
To prove correctness of our linear program formulation, we need to prove two different things.
Firstly, the formulation's constraints must always be satisfiable; that is, there must exist a variable assignment that satisfies all constraints.
This is rather simple to show, but guarantees that the linear program will always give an answer.
The next thing to show is that any produced clustering is legal: if a variable assignment satisfies the constraints, then it is a valid and legal clustering.
This means that, not only do we get \emph{an} answer, we also get the \emph{right} answer.

\subsubsection{Satisfiability}
For any program $p$, there exists a trivial clustering with no fusion at all.
We can use this as the variable assignment of $ilp(p)$.
For each pair of nodes $m,n \in p$, $x_{mn} = 1$ - no fusion is possible.
For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.

Now, prove that this assignment actually satisfies the constraints.

\subsubsection{Legality}
For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_ij$ in $v$ is legal.

For a clustering to be legal, it must satisfy three constraints:
\begin{description}
\item[Acyclic]
after merging nodes of same cluster together, the resulting graph must be a dag
\item[Precedence preserving]
if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
\item[Type constraint]
if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
    \\
    \TODO This is wrong.
\end{description}

### 2014-03-15 2014betterfusionforfilters/section/05-Related.tex
reorganise

\section{Related work}

\subsection{Haskell short-cut fusion}
Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.

Consider the following @filterMax@ function, where each element in the input array is incremented, then the maximum is found, and the array is filtered to those greater than zero.
In this case, the result of the map @vs'@ cannot be inlined into both occurences without duplicating work, so no fusion can be performed.
This has the effect of performing three loops instead of one, with two arrays instead of one.

\begin{code}
filterMax vs =
 let vs' = map    (+1)  vs
     m   = fold   max 0 vs'
     flt = filter (>0)  vs'
\end{code}

\subsection{Integer linear programming in imperative languages}
The idea of using integer linear programming to find optimal fusion clusterings is not new, and has been discussed for imperative languages before.
These methods first construct a \emph{loop dependence graph} (LDG) from a given program, and then use this graph to create the integer linear program.
The LDG has nodes for each loop in the program, and edges between loops are dependencies.
Edges may be fusible, or fusion-preventing, in which case the two nodes may not be merged together.

Talk about the simple formulation by Darte\cite{darte2002new}.
Has an integer variable for each node, denoting the number of the cluster it's in.
Also includes a binary variable for each node, which is whether the node is fused with all its successors - in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes. Explain why.

The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ - that is, no fusion is possible.
This alone is not enough to guarantee a valid clustering.
To constrain the solution to acyclic and precedence preserving clusterings, a variable $\pi_i$ is added for each node $i$.
Constraints are added that require two nodes $i,j$ to have $\pi_i = \pi_j$ if $x_{ij} = 0$, and otherwise $\pi_i > \pi_j$ if $i$ is after $j$.
For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.

The difference to our combinator-based approach is that with combinators we retain more information about the meaning of the program.

### 2014-03-21 2014betterfusionforfilters/section/03-SizeInference.tex
ok

\section{Size inference}
Knowing the relative sizes of loops is important for fusion.
While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times.
For simplicity, we do away with this added complexity and only support fusion of equal sized loops.
\emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.

It is important to emphasise that we are only interested in relative sizes of arrays; which arrays are equal, and which may be smaller.
The $@map@_n$ combinators require all input arrays to be the same size, and their output is the same.
Since @fold@s do not produce arrays, they have no constraints.
@gather@ takes two arrays; the data and the indices. Its output size is the size of the indices.
The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size.

Talk about the relation to Chatterjee\cite{chatterjee1991size}.

\begin{tabbing}
MMMMM       \= MM \= MMMM \= MM \= \kill
$constraint$\> @::=@ \> $array$  \> $=_c$ \> $array$ \\
            \> $~|$  \> $array$  \> $\le_c$ \> $array$ \\
\\
$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{constraint\}$ \\
MMMMMMMMMMMM                 \= MM  \= \kill
$constrs(o = @fold@~f~n)  $ \> $=$ \> $\{\}$     \\
$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o =_c n~|~n \in ns\}$     \\
$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o \le_c b\}$              \\
$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o =_c i\}$ \\
\end{tabbing}

Then, for some program $\lambda\ldots~@let@~binds~@in@~\ldots$, find the union of all constraints:
$\bigcup_{b \in binds} constrs(b)$.
Since equality is transitive, the transitive closure of the constraints is useful.
For example, if $a =_c b$ and $b =_c c$, then $a =_c c$. 
After finding the transitive closure the constraints are partitioned into equivalence sets.
All combinators whose constraints are in the same equivalence set may be fused together.

\subsection{Disallowed filters}

Filtered constraints of the form $a \le_c b$ are special.
The actual size of the filtered output is unknown, and so cannot be constrained.
For example, consider the constraints generated by this program:
\begin{code}
test vs    =
 let flt   = filter p vs
     wrong = map2   f flt vs
 in  wrong
\end{code}
With the resulting constraints being $\{flt \le_c vs, wrong =_c flt, wrong =_c vs\}$.
By transitivity, we find that $flt \le_c flt$ - the size of the filtered output is upper-bounded by the size of the filtered output! 
Why is this bad?
Actually implementing this would require to buffer the entire contents of @flt@, so for simplicity we disallow it. Talk more about this.

Similarly, this example:
\begin{code}
test vs    =
 let flt   = filter p  vs
     flt2  = filter p' vs
     wrong = map2   f  flt flt2
 in  wrong
\end{code}
Generates the constraints $\{flt \le_c vs, flt2 \le_c vs, flt =_c flt2\}$.
Here we are constraining two different filtered outputs to be the same size.
We disallow this.

Not all constraints on filtered data are disallowed, of course;
\begin{code}
test vs    =
 let flt   = filter p  vs
     xs    = map    f  flt
     ys    = map    g  flt
     sm    = map2   f  xs ys
 in  sm
\end{code}
This generates the constraints $\{flt \le_c vs, xs =_c flt, ys =_c flt, sm =_c xs, sm =_c ys\}$,
but this is perfectly valid, since the constraints are all actually based on @flt@.

\begin{tabbing}
MMM       \= MM \= MMMMMM \= MM \= \kill
$valid$   \> @::@  \> $\{constraint\}$  \> $\to$ \> $\mathbb{B}$ \\
MMMMMMMMMMMM                 \= MM  \= \kill
$valid(\{             \})$   \> $=$ \> $\top$  \\
$valid(\{a =_c b, rest\})$   \> $=$ \> $valid(subst~a:=b~in~rest)$  \\
$valid(\{a \le_c a, rest\})$ \> $=$ \> $\bot$                        \\
$valid(\{a \le_c b, rest\})$ \> $=$ \> $a \le_c \_ \not\in closure(rest) \wedge valid(rest)$                        \\
\\
$closure(\{\}, out)$                \> $=$ \> $out$ \\
$closure(\{a =_c b, rest\}, out)$   \> $=$ \> $closure(subst~a:=b~in~rest, subst~a:=b~in~out)$  \\
$closure(\{a \le_c b, rest\}, out)$ \> $=$ \> $closure(rest, \{a \le_c b\} \cup out)$ \\
\end{tabbing}

\subsection{Generators}

\begin{tabbing}
MMM       \= MM \= MMMMMM \= MM \= MMM \= MM \= \kill
$gen$   \> @::@  \> $\{constraint\}$  \> $\to$ \> $array$ \> $\to$ \> $\{array\}$ \\
MMMMMMM                 \= M  \= MMMMMMMM \= MM \= \kill
$gen(constrs, a)$ \> $|$ \> $a \le_c b \in constrs$ \> $=$ \> $\{b\}$                        \\
$gen(constrs, a)$ \> $|$ \> $a =_c b \in constrs$   \> $=$ \> $\bigcup gen(constrs, b)$                        \\
$gen(constrs, a)$ \> $|$ \> $otherwise$             \> $=$ \> $\emptyset$                        \\
\end{tabbing}

%\begin{lemma}
\textbf{Lemma: unique generation of filters}
For some bindings $bs$ and associated constraints $cs$, then
\[
valid(cs) \implies \forall x. (\exists y \in bs. gen(cs, x) = \{y\}) \vee gen(cs, x) = \emptyset
\]
%\end{lemma}

### 2014-03-21 2014betterfusionforfilters/section/04-ILP.tex
ok

@@
 \section{Integer linear program formulation}
+
+\subsection{Function $\to$ graph}
+Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
+Each binding becomes a node in the graph.
+When a binding references other arrays or scalars, there is an edge between those two nodes.
+Edges may be either \emph{fusible} or \emph{fusion-preventing}.
+Fusion-preventing edges mean that the entire input node must completely finish its execution before the output node can start executing.
+For example, @fold@s consume their entire input array before producing a single result, so any references to folds must be fusion-preventing.
+Conversely, @map@s produce output data for every input element, so may be fused.
+
+The @gather@ operation is interesting; it takes an indices array and a data array, and for each element in indices returns that element in data.
+Thus, @gather@ requires random-access in its data array, and is not fusible, but consumes indices linearly and may fuse.
+
+\subsection{Graph $\to$ ILP}
+
+ILP variables:
+\begin{tabbing}
+M   \= M \= MMMMMMM \= MM \= \kill
+$x$ \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
+$\pi$ \> @::@  \> $node$ \> $\to$ \> $\mathbb{N}$ \\
+\end{tabbing}
+
+Talk about how $x_{ij} = 0$ if and only if $i$ and $j$ are fused together, otherwise $x_{ij} = 1$. This may seem reversed, but makes certain constraints easier to formulate.
+
+Use the lemma:
+a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{N}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.
+
+\begin{tabbing}
+MMMMM   \= MMMMMMMMMMM \= \kill
+Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$       \> ($W_{ij}$ is constant weight)      \\
+Subject to \> $-N x_{ij} \le \pi_j - \pi_i \le N x_{ij}$ \> (no edge from i to j)            \\
+           \> $              \pi_j - \pi_i \le N x_{ij}$ \> (an edge from i to j)            \\
+\end{tabbing}
+
+\subsection{Better than stream fusion}
+Abdallah suggests something like:
+\[
+\forall g.\ \exists c.\ ilp(g, c) \le_c streamfusion(g)
+\]
+ie, for any graph there exists a criterion to optimise against, where we are better than or equal to stream fusion.
 
@@
 We can use this as the variable assignment of $ilp(p)$.
-For each pair of nodes $m,n \in p$, $x_{mn} = 1$ - no fusion is possible.
+For each pair of nodes $m,n \in p$, $x_{mn} = 1$ --- no fusion is possible.
 For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.
@@
 
-\subsubsection{Legality}
-For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_ij$ in $v$ is legal.
+\subsubsection{Soundness}
+For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_{ij}$ in $v$ is legal.
 
@@
     \\
-    \TODO This is wrong.
+    \TODO Actually, let us say $x_{ij} = 0 \implies check_{ij}$
+
 \end{description}
+where
+\begin{tabbing}
+MMMMM      \= M \= MMMMMMM \= MM \= \kill
+$check$ \> @::@  \> $array \times array$ \> $\to$ \> $\mathbb{B}$ \\
+MMMMM      \= M \= MMMMMM \= MM \= \kill
+$check(i, j)$     \> $|$ \> $tau_i = tau_j$ \> $=$ \> $x_{i,j} = 0$                        \\
+$check(i, j)$     \> $|$ \> $i' \in gen(i) $ \> $=$ \> $x_{i',j} = 0 \wedge x_{i,i'} = 0 \wedge check(i', j)$                        \\
+$check(i, j)$     \> $|$ \> $j' \in gen(j) $ \> $=$ \> $x_{i,j'} = 0 \wedge x_{j,j'} = 0 \wedge check(i, j')$                        \\
+$check(i, j)$     \> $|$ \> $tau_i \not= tau_j$\> $=$ \> $\bot$
+\end{tabbing}

### 2014-03-21 2014betterfusionforfilters/section/05-Related.tex
ok

@@
 \subsection{Haskell short-cut fusion}
-Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
+Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, mainland2013haskell}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules\cite{jones2001playingby}, to fuse combinators without having to modify the compiler itself.
 This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
@@
 Has an integer variable for each node, denoting the number of the cluster it's in.
-Also includes a binary variable for each node, which is whether the node is fused with all its successors - in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
+Also includes a binary variable for each node, which is whether the node is fused with all its successors --- in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
 The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
@@
 As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
-The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes. Explain why.
+The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes.
+\TODO{Explain why. Perhaps go into more detail about how arrays are contracted, which is different from Megiddo.}
 
@@
 Slightly awkwardly, but for simplicity of other constraints, $x_{ij} = 0$ if the two nodes are fused together.
-If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ - that is, no fusion is possible.
+If there is a fusion preventing edge between $i$ and $j$, then $x_{ij}$ is constrained to be $1$ --- that is, no fusion is possible.
 This alone is not enough to guarantee a valid clustering.
@@
 The difference to our combinator-based approach is that with combinators we retain more information about the meaning of the program.
+
+\subsection{Non-optimal heuristics}
+
+\TODO{Gao\cite{gao1993collective}} Collective loop fusion for array contraction.
+I think this is a good introduction to the min/max edge following algorithm that is central to most of these.
+Finds minimal number of clusters for single types. Despite the name, I don't think it actually minimises clustering for array contraction.
+
+\CITEME{kennedy1993typed} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
+Typed fusion --- choose an ordering of types, then find clustering of first type, fuse those together, then cluster second type, and so on.
+It ain't optimal, even if you try all orderings (see Darte below)
+
+\TODO{Chatterjee\cite{chatterjee1991size}}
+Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
+Like typed fusion, they choose clustering for different types rather arbitrarily, which is sub-par.
+
+\CITEME{darte1998fusion} Alain Darte --- On the complexity of loop fusion.
+Shows that loop fusion, minimising certain things (ie manifest arrays and number of loops) is NP-hard.
+
+\TODO{Song\cite{song2004improving}} Improving data locality by array contraction.
+All about \emph{controlled SFC}, ie shifting, fusion and contraction.
+Formalises cost of memory references depending on distance / how many other references in between. Then only fuses loops if it doesn't raise the distance too much.
+
+\CITEME{ashby2006iterative} T.J. Ashby --- Iterative collective loop fusion.
+Executes the programs to decide which clustering is best, apparently. I didn't read it thoroughly, but didn't understand where they actually get the test data from. Unless it's a kind of JIT thing.
+I honestly don't think much of it, but it \emph{is} more recent than the other stuff, so I suspect it'd be good to mention just so we're not ignoring ``modern fusion''.
+
+\CITEME{kennedy2000fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.
+
+\TODO{Sarkar\cite{sarkar1991optimization}} Optimization of array accesses by collective loop transformations.
+This is probably not at all relevant, but it's \emph{very cute}. Use a two-colouring algorithm to decide when to reverse loops, to get best fusion. Not our scene.
+
+\TODO{I don't really have a story on polyhedral analysis or unimodular.}
+Now, here's my understanding of unimodular (and I'm just guessing this applies to polyhedral too):
+if the dependency matrix for a loop nest or set of loops forms a \emph{unimodular matrix} (an integral matrix whose inverse is also integral), only then can it be dealt with by unimodular transforms.

### 2014-03-23 2014betterfusionforfilters/section/04-ILP.tex
formalise graph construction
add @cross@ combinator. todo fix up constraint checking

@@
 Thus, @gather@ requires random-access in its data array, and is not fusible, but consumes indices linearly and may fuse.
+
+\begin{tabbing}
+MMMM        \= MM   \= \kill
+$nodes$     \> @::@ \> $program \to V$          \\
+$edges$     \> @::@ \> $program \to E$          \\
+$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
+\\
+MMMMM       \= \kill
+$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
+\\
+$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
+\\
+MMMM             \= M \= \kill
+$edge(bs, out = @fold@~f~in)$ \\
+                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
+                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+$edge(bs, out = @map@~f~in)$  \\
+                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
+                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+$edge(bs, out = @filter@~f~in)$  \\
+                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
+                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+$edge(bs, out = @gather@~data~indices)$  \\
+                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
+                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
+$edge(bs, out = @cross@~a~b)$            \\
+                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
+                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
+$edge(bs, outs = @external@~ins)$  \\
+                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
+\\
+$inedge(bs,to,from)$ \\
+                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
+                     \> $=$ \> $(to, from, fusion~preventing)$  \\
+                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
+                     \> $=$ \> $(to, from, fusion~preventing)$  \\
+                     \> $|$ \> $otherwise$                      \\
+                     \> $=$ \> $(to, from, fusible)$            \\
+\end{tabbing}
 
@@
 \end{tabbing}
+
+\subsection{Subtype optimisations}
+For two clusters $i, j$, if $x_{ij} = 0 \vee \tau_i \not= \tau_j$, then $i$'s generator must be fused with $j$ and vice versa, recursively ($x_{ij} = 0 \implies check(i,j)$).
+Problem: actually generating constraints for \emph{all} pairs of nodes with unequal types is excessive.
+I see two solutions to this:
+\begin{itemize}
+\item
+Generate fewer constraints and rely on transitivity.
+This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
+This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.
+\item
+Alternatively:
+we can say, only generate subtype constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
+Here, we generate fewer constraints than originally, but still more than above.
+We also must preprocess the graph to find such blocking paths.
+But we still get the minimisation to count non-connected nodes.
+\end{itemize}
 

### 2014-03-25 2014betterfusionforfilters/Main.tex
ok
Copy edits to abstract, add note about "optimal"

@@
 \documentclass[preprint]{sigplanconf}
-\usepackage{version}
 \usepackage{amssymb}
@@
 \begin{abstract}
-Our recent work on flow fusion shows how to extract a single imperative loop from a set of combinators, but if multiple loops are required, finding the optimal clustering is non-trivial.
-The optimal clustering must maximise cache locality, minimise memory traffic, and minimise loop overhead.
-Given the sheer number of possible clusterings an exhaustive search is not feasible, and greedy algorithms tend to only find an approximate solution.
-Instead, we show how to formulate the problem as an \emph{integer linear programming} problem, and use an external library to find the optimal solution.
+Recent work on array fusion shows how to compile a specific class of data flow programs into single imperative loops.
+However, until now there has been no automated way to take the operator graph expressing a complete program and split it into subgraphs that can be individually fused.
+Finding a satisfactory clustering of operators is non-trivial, as we must find some balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
+As real programs contain tens or hundreds of individual operators, performing an exhaustive search for an optimal clustering is not feasible, and greedy algorithms tend to produce poor solutions.
+Instead, we show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find an good solution.
+
+% BL: I've removed uses of the word "optimal" here because we don't have a proper
+% language cost model, and our solution is ultimately based on heuristics. ILP can find 
+% an "optimal" solution for a heuristic cost metric, but this isn't the thing we're
+% really trying to optimize.
+
 \end{abstract}
@@
 \input{section/01-Introduction.tex}
+\input{section/XX-Programs.tex}
 \input{section/02-CNF.tex}

### 2014-03-25 2014betterfusionforfilters/Main.tex
Rework introduction

@@
 \begin{abstract}
-Recent work on array fusion shows how to compile a specific class of data flow programs into single imperative loops.
-However, until now there has been no automated way to take the operator graph expressing a complete program and split it into subgraphs that can be individually fused.
-Finding a satisfactory clustering of operators is non-trivial, as we must find some balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
-As real programs contain tens or hundreds of individual operators, performing an exhaustive search for an optimal clustering is not feasible, and greedy algorithms tend to produce poor solutions.
-Instead, we show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find an good solution.
+Recent work on array fusion demonstrates how to compile a specific class of data flow programs into single imperative loops.
+However, until now there has been no satisfactory way to take the operator graph expressing a complete program and split it into fusible subgraphs.
+Finding good clusterings of operators is non-trivial, as we must achieve a balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
+We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find an good solution.
+\end{abstract}
 
@@
 % really trying to optimize.
-
-\end{abstract}
 
@@
 \input{section/01-Introduction.tex}
-\input{section/XX-Programs.tex}
 \input{section/02-CNF.tex}
@@
 \input{section/05-Related.tex}
+\input{section/XX-Programs.tex}
 \section{Conclusion}

### 2014-03-25 2014betterfusionforfilters/section/01-Introduction.tex
Rework introduction

@@
 \section{Introduction}
-Combinators are a natural and composable way of expressing array programs, with many correctness benefits over imperative programs.
+
+Recent work on data flow fusion~\cite{lippmeier2013data} shows how to compile a specific class of data flow programs into \emph{single}, efficient imperative loops.
+This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays \cite{repa}.
+The key benefits of data flow fusion over prior work are that 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion is guaranteed for all programs that fall into the specific class that can be fused.
+
+Ultimately, though, we cannot expect \emph{all} programs to be fused into single imperative loops, as most contain fundamental data dependencies that prevent complete fusion. For example:
+\begin{code}
+  normalize :: Array Int -> Array Int
+  normalize xs = let sum = fold (+) 0 xs
+                 in  map (/ sum) xs
+\end{code}
+
+If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using two loops: one to determine the sum, and one to divide the elements.
+The application of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
+
+However, many programs \emph{do} contain opportunities for fusion, if we only new which opportunities to take.
+The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion:
+\begin{code}
+ normalizeInc :: Array Int -> Array Int
+ normalizeInc xs
+  = let incs = map  (+1)    us      (A1) (B2)
+        sum  = fold (+) 0   us      (A1) (B1)
+        norm = map  (/ sum) incs    (A2) (B2)
+    in  norm
+\end{code}
+
+For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
+The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
+For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+
+Our contributions are as follows:
+
+\begin{itemize}
+\item   We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
+We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering.
+\TODO{ref}
+
+\item   Our constraint system also expresses heuristics on the cost of various clusterings.
+For example, we encode the fact that intermediate arrays are more expensive to create than intermediate integers, which ensures the second clustering for @normalizeInt@ will be chosen over the first.
+\TODO{ref}
+
+\item   We present benchmarks of our algorithm applied to several common programming patterns, and to several pathological examples.
+Our algorithm is complete and yields good results in practice, though an optimal solution is uncomputable in general.
+\TODO{ref}
+\end{itemize}
+
+\TODO{Briefly mention related work. Other approaches to typed fusion}
+
+% We must also decide which clustering is the `best' or most optimal.
+One obvious criterion for this is the minimum number of loops, but there may even be multiple clusterings with the minimum number of loops.
+In this case, the number of required manifest arrays must also be taken into account.
+
+% Consider the @normaliseInc@ function below; it requires at least two loops, since @sum@ must be computed from the entire array before any of the output array @norm@ can be filled.
+Scheduling @incs@ and @sum@ into a single loop, then @norm@ into another loop only requires two loops, but requires @incs@ to be stored in a manifest array before being read again by @norm@.
+If we were to schedule @sum@ in its own loop and @incs@ and @norm@ together, we would only require one array to be filled - the output array.
+
+% Combinators are a natural and composable way of expressing array programs, with many correctness benefits over imperative programs.
 If compiled naively, however, executing the combinator program will often require many more loops and intermediate arrays than an imperative program written by hand.
@@
 
-Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
+% Existing fusion systems for Haskell such as stream fusion~\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
 This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
@@
 
-Our recent work on flow fusion\cite{lippmeier2013data} generates a single imperative loop for a set of combinators, but when multiple loops are required, deciding which combinators to group together is difficult.
-The optimal clustering must maximise cache locality, minimise memory traffic, and minimise loop overhead.
-For most functions, there are so many possible clusterings that finding the optimal by exhaustive search is simply not feasible.
-
-We must also decide which clustering is the `best' or most optimal. 
-One obvious criterion for this is the minimum number of loops, but there may even be multiple clusterings with the minimum number of loops.
-In this case, the number of required manifest arrays must also be taken into account.
-Consider the @normaliseInc@ function below; it requires at least two loops, since @sum@ must be computed from the entire array before any of the output array @norm@ can be filled.
-Scheduling @incs@ and @sum@ into a single loop, then @norm@ into another loop only requires two loops, but requires @incs@ to be stored in a manifest array before being read again by @norm@.
-If we were to schedule @sum@ in its own loop and @incs@ and @norm@ together, we would only require one array to be filled - the output array.
-
-\begin{code}
-normaliseInc (us : Vector Int) =
- let incs = map  (+1)    us
-     sum  = fold (+) 0   us
-     norm = map  (/sum)  incs
- in  norm
-\end{code}
+% As real programs contain tens or hundreds of individual operators, performing an exhaustive search for an optimal clustering is not feasible, and greedy algorithms tend to produce poor solutions.

### 2014-03-27 2014betterfusionforfilters/section/04-ILP.tex
add fusion preventing constraint
mention quadtree
start filling in ILP formulation and weights

@@
 $x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
-$\pi$ \> @::@  \> $node$ \> $\to$ \> $\mathbb{N}$ \\
+      \>       \> ($x_{ij} = 0$ iff $i$ and $j$ are fused together) \\
+$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$ \\
+      \>       \> ($\pi_i$ is used to ensure an acyclic clustering) \\
+$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$ \\
+      \>       \> ($c_i = 0$ iff $\forall (i,j) \in E. x_{ij} = 0$) \\
+      \>       \> (i.e. no intermediate array necessary)            \\
 \end{tabbing}
@@
 Use the lemma:
-a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{N}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.
+a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{R}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.
 
 \begin{tabbing}
-MMMMM   \= MMMMMMMMMMM \= \kill
-Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$       \> ($W_{ij}$ is constant weight)      \\
-Subject to \> $-N x_{ij} \le \pi_j - \pi_i \le N x_{ij}$ \> (no edge from i to j)            \\
-           \> $              \pi_j - \pi_i \le N x_{ij}$ \> (an edge from i to j)            \\
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \> \ldots \\
+Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (no edge from i to j)            \\
+           \>             \>       \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (an edge from i to j)            \\
+\\
+           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \> (for all edges from i to j)      \\
+\\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (fusion-preventing edges from i to j)      \\
+\\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
+           \> $x_{ij}$    \> $\le$ \> $x_{i'j}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
+           \> $x_{ij}$    \> $\le$ \> $x_{ij'}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+\end{tabbing}
+
+\subsubsection{The benefits of fusion}
+
+Fusing loops can have three main benefits:
+\begin{itemize}
+\item
+reducing memory traffic, such as multiple loops reading from the same array;
+\item
+removing intermediate arrays, thus reducing the amount of memory required;
+\item
+and reducing loop overhead, such as when two loops operate on different arrays of the same size.
+\end{itemize}
+However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
+When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
+An excessive number of intermediate arrays can also cause issues if all are live in memory at once, potentially leading to \emph{thrashing}.
+The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
+
+This total ordering can be encoded into an ILP objective function as weights.
+If the program graph contains $n$ combinators, then there are at most $n$ opportunities for fusion.
+The encoding of loop overhead is weight $1$, removing intermediate arrays is weight $n$, and reducing memory traffic is weight $n^2$.
+This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array,
+and likewise no number of removed intermediate arrays can outweigh a reduction in memory traffic.
+Although, it is worth noting that reducing memory traffic does \emph{tend} to remove intermediate arrays, and vice versa.
+
+\begin{tabbing}
+MMMMM   \= MMMM \= M \= \kill
+Minimise   \> \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
+Where      \>                                       \\
+           \> $W_{ij} = N^2$ \> $|$ \> $(i,j) \in E $         \\
+           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
+           \> $W_{ij} = N^2$ \> $|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
+           \> \> \> ($i$ and $j$ share an input array)                                         \\
+           \> $W_{ij} = 1$   \> $|$ \> $@otherwise@$                                                  \\
+           \> \> \> (the only benefit is loop overhead)                                        \\
+           \\
+           \> $N = |V|$
 \end{tabbing}
@@
 \end{itemize}
+
+\begin{tabbing}
+MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
+$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
+MMMM        \= M    \= \kill
+$split(vs,es)$ \\
+    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
+$clusterable(v,vs,es)$  \\
+    \>$=$       \>$\{v' | v' \in vs$                                \\
+    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
+    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
+\end{tabbing}
 
@@
 if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
+\item[Fusion preventing]
+likewise, if there is a fusion-preventing edge between two nodes $i$ and $j$, then we require $\pi_j > \pi_i$, which implies that they are not merged together
 \item[Type constraint]

### 2014-03-28 2014betterfusionforfilters/section/04-ILP.tex
minor fix

@@
 Minimise   \> \ldots \\
-Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (no edge from i to j)            \\
-           \>             \>       \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \> (an edge from i to j)            \\
+Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (no edge from i to j)            \\
+           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (an edge from i to j)            \\
 \\
-           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \> (for all edges from i to j)      \\
+           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
+           \> (for all edges from i to j; thus $c_i=0$ only if all $x{ij} = 0$)      \\
 \\
-           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (fusion-preventing edges from i to j)      \\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (fusion-preventing edges from i to j)      \\
 \\
-           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
-           \> $x_{ij}$    \> $\le$ \> $x_{i'j}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
-           \> $x_{ij}$    \> $\le$ \> $x_{ij'}$       \>       \>            \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
+           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
+           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
 \end{tabbing}
+It is worth noting that for the edge cases, $x_{ij} = 1 \implies \pi_i < \pi_j$, meaning if $i$ and $j$ are not fused together, $j$ must be scheduled after $i$.
 
@@
 This total ordering can be encoded into an ILP objective function as weights.
-If the program graph contains $n$ combinators, then there are at most $n$ opportunities for fusion.
-The encoding of loop overhead is weight $1$, removing intermediate arrays is weight $n$, and reducing memory traffic is weight $n^2$.
+If the program graph contains $N$ combinators, then there are at most $N$ opportunities for fusion.
+The encoding of loop overhead is weight $1$, removing intermediate arrays is weight $N$, and reducing memory traffic is weight $N^2$.
 This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array,
@@
 Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+           \> \> \> (memory traffic and loop overhead)         \\
            \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
+           \> \> \> (removing intermediate arrays)         \\
+Subject to \> \ldots                                \\
 Where      \>                                       \\
-           \> $W_{ij} = N^2$ \> $|$ \> $(i,j) \in E $         \\
+           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
            \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
-           \> $W_{ij} = N^2$ \> $|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
+           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
            \> \> \> ($i$ and $j$ share an input array)                                         \\
-           \> $W_{ij} = 1$   \> $|$ \> $@otherwise@$                                                  \\
+           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
            \> \> \> (the only benefit is loop overhead)                                        \\

### 2014-03-28 2014betterfusionforfilters/section/04-ILP.tex
minor fix
simplified ILP constraints

@@
 
+\subsubsection{The simple version}
+\TODO{Explain this simple version}
 \begin{tabbing}
@@
 Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (no edge from i to j)            \\
+           \>             (no edge from i to j, but there is a fusion benefit)            \\
            \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
@@
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (for all edges from i to j; thus $c_i=0$ only if all $x{ij} = 0$)      \\
+           \> (for all edges from i; thus $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$)      \\
 \\
@@
 It is worth noting that for the edge cases, $x_{ij} = 1 \implies \pi_i < \pi_j$, meaning if $i$ and $j$ are not fused together, $j$ must be scheduled after $i$.
+
+\subsubsection{A slightly smarter version}
+We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
+Likewise with clusters of different, incompatible types.
+The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.
+
+\begin{tabbing}
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \> \ldots \\
+Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (no edge from i to j, but there is a fusion benefit)            \\
+           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (an edge from i to j)            \\
+           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
+           \>             (a \emph{fusion-preventing} edge from i to j)            \\
+\\
+           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
+           \> (for all edges from i with no fusion-preventing outputs)      \\
+\\
+           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
+           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+\end{tabbing}
 
@@
 Alternatively:
-we can say, only generate subtype constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
+we can say, only generate constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
 Here, we generate fewer constraints than originally, but still more than above.

### 2014-04-01 2014betterfusionforfilters/section/04-ILP.tex
ilp intro

@@
 \section{Integer linear program formulation}
+While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
+To simplify creation of the integer linear program formulation, we first convert a function to a dependency graph.
+The dependency graph has a node for each combinator, and edges between two combinators when one uses another's output.
+After the graph is created, it is converted to an integer linear program with an objective of the least memory traffic and loops, solved, and the solution converted to a clustering.
+
+When creating the graph, it is important to note that not all loops can be fused together; firstly, loops of different size cannot be fused.
+Secondly, two loops cannot be fused together if an iteration in one loop relies on the output of a later iteration in another loop.
+For example, a @map@ may not be fused with a @fold@ if the @map@'s worker function uses the result of the @fold@.
+This fusion restriction is encoded as a \emph{fusion-preventing} edge between two combinators.
+Other edges are \emph{fusible}, and may be fused together.
+If they are not fused together for whatever reason, the \emph{from} combinator must be scheduled before the \emph{to} combinator.
+
+The dependency graph is translated to an integer linear program.
+The integer linear program has some integral, boolean and real variables, an objective function to minimise, and a set of constraints that the variables must conform to.
+Finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, but existing ILP solvers tend to be sufficient for realistic problem sizes.
+For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still gives the exact answer for small problems.
 

### 2014-04-01 2014betterfusionforfilters/section/01-Introduction.tex
Add note to intro
simplified ILP constraints
Adding citations

@@
 
-Recent work on data flow fusion~\cite{lippmeier2013data} shows how to compile a specific class of data flow programs into \emph{single}, efficient imperative loops.
-This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays \cite{repa}.
+Recent work on data flow fusion~\cite{lippmeier2013flow} shows how to compile a specific class of data flow programs into \emph{single}, efficient imperative loops.
+This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays \cite{keller2010repa}.
 The key benefits of data flow fusion over prior work are that 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion is guaranteed for all programs that fall into the specific class that can be fused.
+
+\ben{reword last sentence, reads like a tautology}
 
@@
 
-However, many programs \emph{do} contain opportunities for fusion, if we only new which opportunities to take.
+However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
 The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion:
+
+\TODO{If the main contribution is fusion for *filters* then we need to use a motivating example with a filter}
+
 \begin{code}
@@
 
-\TODO{Briefly mention related work. Other approaches to typed fusion}
+The reduction of the clustering problem to integer linear programming was previously described by \cite{megiddo1998weighted}, though they do not consider filtering operations.
 

### 2014-04-01 2014betterfusionforfilters/section/01-Introduction.tex
Add .bib file and fix citations

@@
 
-The reduction of the clustering problem to integer linear programming was previously described by \cite{megiddo1998weighted}, though they do not consider filtering operations.
+The reduction of the clustering problem to integer linear programming was previously described by \cite{megiddo1998optimal}, though they do not consider filtering operations.
 

### 2014-04-01 2014betterfusionforfilters/section/05-Related.tex
Add .bib file and fix citations

@@
-\section{Related work}
+f\section{Related work}
 
 \subsection{Haskell short-cut fusion}
-Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, mainland2013haskell}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules\cite{jones2001playingby}, to fuse combinators without having to modify the compiler itself.
+Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, mainland2013exploiting}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules\cite{peytonjones2001rules}, to fuse combinators without having to modify the compiler itself.
 This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
@@
 
-Talk about the simple formulation by Darte\cite{darte2002new}.
-Has an integer variable for each node, denoting the number of the cluster it's in.
+Talk about the simple formulation by Darte\cite{darte2002contraction}. Has an integer variable for each node, denoting the number of the cluster it's in.
 Also includes a binary variable for each node, which is whether the node is fused with all its successors --- in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
@@
 
-The formulation by Megiddo\cite{megiddo1997optimal} supports different sized loops, and is therefore more relevant for our purposes.
+The formulation by Megiddo\cite{megiddo1998optimal} supports different sized loops, and is therefore more relevant for our purposes.
 For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
@@
 
-\TODO{Gao\cite{gao1993collective}} Collective loop fusion for array contraction.
+\cite{gao1993collective} Collective loop fusion for array contraction.
 I think this is a good introduction to the min/max edge following algorithm that is central to most of these.
@@
 
-\CITEME{kennedy1993typed} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
+\cite{kennedy1993typedfusion} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
 Typed fusion --- choose an ordering of types, then find clustering of first type, fuse those together, then cluster second type, and so on.
@@
 
-\TODO{Chatterjee\cite{chatterjee1991size}}
-Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
+\cite{chatterjee1991size} Chatterjee -- Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
 Like typed fusion, they choose clustering for different types rather arbitrarily, which is sub-par.
 
-\CITEME{darte1998fusion} Alain Darte --- On the complexity of loop fusion.
+\cite{darte1999complexity} Alain Darte --- On the complexity of loop fusion.
 Shows that loop fusion, minimising certain things (ie manifest arrays and number of loops) is NP-hard.
 
-\TODO{Song\cite{song2004improving}} Improving data locality by array contraction.
+\cite{song2004improving} Improving data locality by array contraction.
 All about \emph{controlled SFC}, ie shifting, fusion and contraction.
@@
 
-\CITEME{ashby2006iterative} T.J. Ashby --- Iterative collective loop fusion.
+\cite{ashby2006iterative} T.J. Ashby --- Iterative collective loop fusion.
 Executes the programs to decide which clustering is best, apparently. I didn't read it thoroughly, but didn't understand where they actually get the test data from. Unless it's a kind of JIT thing.
@@
 
-\CITEME{kennedy2000fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.
+\cite{kennedy2001fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.
 
-\TODO{Sarkar\cite{sarkar1991optimization}} Optimization of array accesses by collective loop transformations.
-This is probably not at all relevant, but it's \emph{very cute}. Use a two-colouring algorithm to decide when to reverse loops, to get best fusion. Not our scene.
+\cite{sarkar1991optimization} Optimization of array accesses by collective loop transformations.
+This is probably not at all relevant, but it's \emph{very cute}.
+Use a two-colouring algorithm to decide when to reverse loops, to get best fusion.
+Not our scene.
 

### 2014-04-02 2014betterfusionforfilters/section/04-ILP.tex
objective function
explain ILP variables

@@
 $nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
-\\
+\end{tabbing}
+Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
+Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.
+
+\begin{tabbing}
+MMMMM       \= \kill
 $edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
@@
 \subsection{Graph $\to$ ILP}
+With the dependency graph now constructed, it must be converted to an integer linear program.
+The integer linear program will contain some variables, an objective function, and constraints that the variables must conform to.
+An external solver will find and return a variable assignment that minimises the objective function.
+With the variable assignment, the graph's nodes are partitioned into a clustering, and then flow fusion\cite{lippmeier2013flow} is used to extract imperative loops for each cluster.
 
-ILP variables:
+The variables in the ILP formulation can be split into three groups:
+
 \begin{tabbing}
-M   \= M \= MMMMMMM \= MM \= \kill
-$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$ \\
-      \>       \> ($x_{ij} = 0$ iff $i$ and $j$ are fused together) \\
-$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$ \\
-      \>       \> ($\pi_i$ is used to ensure an acyclic clustering) \\
-$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$ \\
-      \>       \> ($c_i = 0$ iff $\forall (i,j) \in E. x_{ij} = 0$) \\
-      \>       \> (i.e. no intermediate array necessary)            \\
+M   \= MM \= MMMMMMM \= MM \= \kill
+$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
 \end{tabbing}
+The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused.
+If $x_{ij} = 0$, then $i$ and $j$ are fused together.
+While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler.
+The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
 
-Talk about how $x_{ij} = 0$ if and only if $i$ and $j$ are fused together, otherwise $x_{ij} = 1$. This may seem reversed, but makes certain constraints easier to formulate.
+\begin{tabbing}
+M   \= MM \= MMMMMMM \= MM \= \kill
+$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$
+\end{tabbing}
+The second group of variables is used simply to ensure that the resulting clustering is acyclic.
+For some node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ has a value $\pi_j > \pi_i$.
+This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor.
+If two nodes are to be fused together, their $\pi$ values must be the same.
 
-Use the lemma:
-a graph $G$ is acyclic if and only if there exists some mapping $\pi :: node \to \mathbb{R}$ such that for every edge in $G$ from $i$ to $j$, $\pi_i < \pi_j$.
+\begin{tabbing}
+M   \= MM \= MMMMMMM \= MM \= \kill
+$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$
+\end{tabbing}
+The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
+They are not required for correctness of the clustering, but are used in the objective function.
+A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together:
+$c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$.
 
@@
 
-\subsubsection{The benefits of fusion}
+\subsubsection{The objective function}
 
-Fusing loops can have three main benefits:
+To find the objective function, note that fusing loops can have three main benefits:
 \begin{itemize}

### 2014-04-02 2014betterfusionforfilters/section/04-ILP.tex
explain ILP variables
move simplification/opt talk, expand explanation of ILP formulation

@@
 \subsubsection{The simple version}
-\TODO{Explain this simple version}
+\paragraph{Acyclic and precedence-preserving}
+
 \begin{tabbing}
@@
 Minimise   \> \ldots \\
-Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (no edge from i to j, but there is a fusion benefit)            \\
+Subject to \> \ldots \\
            \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (an edge from i to j)            \\
-\\
-           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (for all edges from i; thus $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$)      \\
-\\
+           \>             (an edge from $i$ to $j$)            \\
+           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (no edge from $i$ to $j$)            \\
+           \> \ldots
+\end{tabbing}
+Here, we look at every pair of nodes $i$ and $j$.
+
+Whether there is an edge between $i$ and $j$ or not, if the two are fused together, then $x_{ij} = 0$.
+If $x_{ij} = 0$, then both cases may be simplified to $0 \le \pi_j - \pi_i \le 0$, which is equivalent to $\pi_i = \pi_j$.
+
+Otherwise, if the two nodes are not fused together, then $x_{ij} = 1$.
+If there is an edge between $i$ and $j$, the constraint simplifies to $1 \le \pi_j - \pi_i \le N$, where $N$ is the number of nodes in the graph.
+This constraint means that the difference between the two $\pi$s must be at least 1, and less than $N$.
+Since there are $N$ nodes, the maximum difference between any two $\pi$s would be at most $N$, so the upper bound of $N$ is large enough to be safely ignored.
+This means the constraint can roughly be translated to $\pi_i < \pi_j$, which enforces the acyclic constraint.
+
+If there is no edge between $i$ and $j$, the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively means there is no constraint on the $\pi$ values.
+Note, however, that other edges may still constrain $i$ or $j$ by transitivity.
+
+\paragraph{Fusion-preventing edges}
+\begin{tabbing}
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \> \ldots \\
+Subject to \> \ldots \\
            \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
-           \> (fusion-preventing edges from i to j)      \\
-\\
+           \> (fusion-preventing edges from $i$ to $j$)      \\
+           \> \ldots
+\end{tabbing}
+For every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
+This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.
+
+\paragraph{Fusion between different types}
+\begin{tabbing}
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \> \ldots \\
+Subject to \> \ldots \\
            \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
@@
            \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+           \> \ldots
 \end{tabbing}
-It is worth noting that for the edge cases, $x_{ij} = 1 \implies \pi_i < \pi_j$, meaning if $i$ and $j$ are not fused together, $j$ must be scheduled after $i$.
+Nodes with different loop sizes cannot be fused together, generally.
+If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators.
+\TODO{Expand}.
 
-\subsubsection{A slightly smarter version}
-We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
-Likewise with clusters of different, incompatible types.
-The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.
-
+\paragraph{Array contraction}
 \begin{tabbing}
@@
 Minimise   \> \ldots \\
-Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (no edge from i to j, but there is a fusion benefit)            \\
-           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (an edge from i to j)            \\
-           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
-           \>             (a \emph{fusion-preventing} edge from i to j)            \\
-\\
+Subject to \> \ldots \\
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (for all edges from i with no fusion-preventing outputs)      \\
-\\
-           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
-           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+           \> (for all edges from i)            \\
+           \> \ldots \\
 \end{tabbing}
+An array is fully contracted if all its output edges are fused with it.
+Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$.
+By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.
 
@@
 
+\subsubsection{An optimised version}
+We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
+Likewise with clusters of different, incompatible types.
+The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.
+
+\begin{tabbing}
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \> \ldots \\
+Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (no edge from i to j, but there is a fusion benefit)            \\
+           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \>             (an edge from i to j)            \\
+           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
+           \>             (a \emph{fusion-preventing} edge from i to j)            \\
+\\
+           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
+           \> (for all edges from i with no fusion-preventing outputs)      \\
+\\
+           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
+           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+\end{tabbing}
+
 \subsection{Better than stream fusion}

### 2014-04-04 2014betterfusionforfilters/section/01-Introduction.tex
explicitly state similarities and differences to existing ILP formulations

@@
 \begin{itemize}
-\item   We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
-We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering.
+\item   
+We use existing integer linear programming clustering formulations such as Megiddo\cite{megiddo1998optimal} and Darte\cite{darte2002contraction},
+and modify them to express constraints on fusing loops of different sizes if they can be fused into generator loops of the same size.
 \TODO{ref}
+% We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
+% We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering. \TODO{ref}
+
+\item
+We present a simplification to constraint generation that would also be applicable to existing integer linear programming formulations,
+where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two. \TODO{ref}
 

### 2014-04-04 2014betterfusionforfilters/section/04-ILP.tex
explicitly state similarities and differences to existing ILP formulations

@@
 \end{tabbing}
-Here, we look at every pair of nodes $i$ and $j$.
+As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.
 
@@
 \end{tabbing}
-For every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
+As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
 This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.
@@
 \end{tabbing}
-Nodes with different loop sizes cannot be fused together, generally.
+The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
 If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators.
@@
 \end{tabbing}
+As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
 An array is fully contracted if all its output edges are fused with it.

### 2014-04-04 2014betterfusionforfilters/section/01-Introduction.tex
suggest normalise2 for better motivating program?

@@
 
+\TODO{This program has filters and other interesting stuff, but doesn't have ``multiple unique best solutions''}
+\begin{code}
+ normalize2 :: Array Int -> Array Int
+ normalize2 us
+  = let sum1 = fold   (+) 0 us      (A1) (S1) (I1)
+        gts  = filter (>0)  us      (A1) (S2) (I1)
+        sum2 = fold   (+) 0 gts     (A1) (S2) (I2)
+        nor1 = map  (/sum1) us      (A2) (S3) (I3)
+        nor2 = map  (/sum2) us      (A2) (S4) (I3)
+    in (nor1, nor2)
+\end{code}
+Our approach, @A1-A2@, requires only two loops for this particular case.
+Stream fusion (@S1-S4@) is unable to fuse @sum1@ and @gts@ together, as there is nothing to be inlined between them, so requires four whole traversals of the input array, @us@.
+Similarly, existing imperative solutions (@I1-I3@) would be unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.
+
 Our contributions are as follows:
@@
 
-The reduction of the clustering problem to integer linear programming was previously described by \cite{megiddo1998optimal}, though they do not consider filtering operations.
+The reduction of the clustering problem to integer linear programming was previously described by\cite{megiddo1998optimal}, though they do not consider filtering operations.
 

### 2014-04-07 2014betterfusionforfilters/section/02-CNF.tex
ok
Rework introduction
suggest normalise2 for better motivating program?
Add diagrams for normalize2 example

@@
+%!TEX root = ../Main.tex
+%\eject
+
+\begin{figure*}[ht!]
+\begin{center}
+\includegraphics[scale=0.5]{figures/ex1-compare.pdf}
+\end{center}
+
+\caption{Clusterings for normalize2 example: with stream fusion; our system; best imperative system}
+\label{f:normalize2-cluterings}
+\end{figure*}
+
 \section{Combinator normal form}
@@
 
-It is important to note, however, that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
+It is important to note that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
 None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.

### 2014-04-07 2014betterfusionforfilters/section/01-Introduction.tex
Add diagrams for normalize2 example
Shift normalizeInc example to a later section

@@
+%!TEX root = ../Main.tex
 \section{Introduction}
@@
 However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
-The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion:
-
-\TODO{If the main contribution is fusion for *filters* then we need to use a motivating example with a filter}
-
-\begin{code}
- normalizeInc :: Array Int -> Array Int
- normalizeInc xs
-  = let incs = map  (+1)    us      (A1) (B2)
-        sum  = fold (+) 0   us      (A1) (B1)
-        norm = map  (/ sum) incs    (A2) (B2)
-    in  norm
-\end{code}
-
-For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
-The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
-For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion.
+A diagram of the data flow graph for this example is on the next page.
 
 \TODO{This program has filters and other interesting stuff, but doesn't have ``multiple unique best solutions''}
+
 \begin{code}
@@
 
+% NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
 \begin{itemize}

### 2014-04-07 2014betterfusionforfilters/section/04-ILP.tex
Add diagrams for normalize2 example
Shift normalizeInc example to a later section

@@
+%!TEX root = ../Main.tex
 \section{Integer linear program formulation}
+
+\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering.
+Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.}
+\begin{code}
+ normalizeInc :: Array Int -> Array Int
+ normalizeInc xs
+  = let incs = map  (+1)    us      (A1) (B2)
+        sum  = fold (+) 0   us      (A1) (B1)
+        norm = map  (/ sum) incs    (A2) (B2)
+    in  norm
+\end{code}
+
+For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
+The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
+For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+
 While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
@@
 
+%!TEX root = ../Main.tex
 \subsubsection{Satisfiability}

### 2014-04-08 2014betterfusionforfilters/section/01-Introduction.tex
intro fixes

@@
 Recent work on data flow fusion~\cite{lippmeier2013flow} shows how to compile a specific class of data flow programs into \emph{single}, efficient imperative loops.
-This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays \cite{keller2010repa}.
-The key benefits of data flow fusion over prior work are that 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion is guaranteed for all programs that fall into the specific class that can be fused.
-
-\ben{reword last sentence, reads like a tautology}
+This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
+The key benefits of data flow fusion over prior work are that 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between combinators.
 
@@
 However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
-The following function is similar to the one from before, except that it offers two unique \emph{but mutually exclusive} approaches to fusion.
+The following function is similar to the one from before, except that it offers several unique \emph{but mutually exclusive} approaches to fusion.
 A diagram of the data flow graph for this example is on the next page.
-
-\TODO{This program has filters and other interesting stuff, but doesn't have ``multiple unique best solutions''}
 
@@
 Stream fusion (@S1-S4@) is unable to fuse @sum1@ and @gts@ together, as there is nothing to be inlined between them, so requires four whole traversals of the input array, @us@.
-Similarly, existing imperative solutions (@I1-I3@) would be unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.
+Similarly, existing imperative solutions (@I1-I3@) are unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.
 
@@
 \item
-We present a simplification to constraint generation that would also be applicable to existing integer linear programming formulations,
+We present a simplification to constraint generation that is also applicable to some existing integer linear programming formulations such as Megiddo's,
 where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two. \TODO{ref}
 
-\item   Our constraint system also expresses heuristics on the cost of various clusterings.
-For example, we encode the fact that intermediate arrays are more expensive to create than intermediate integers, which ensures the second clustering for @normalizeInt@ will be chosen over the first.
+\item
+Our constraint system also encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program.
+For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced.
 \TODO{ref}
 
-\item   We present benchmarks of our algorithm applied to several common programming patterns, and to several pathological examples.
-Our algorithm is complete and yields good results in practice, though an optimal solution is uncomputable in general.
-\TODO{ref}
+\item
+We present benchmarks of our algorithm applied to several common programming patterns, and to several pathological examples.
+Our algorithm is complete and yields good results in practice, though if array sizes are unknown, an optimal solution is uncomputable in general. \TODO{ref}
 \end{itemize}

### 2014-04-08 2014betterfusionforfilters/section/01-Introduction.tex
fix some citations

@@
 \item   
-We use existing integer linear programming clustering formulations such as Megiddo\cite{megiddo1998optimal} and Darte\cite{darte2002contraction},
+We use existing integer linear programming clustering formulations such as Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction},
 and modify them to express constraints on fusing loops of different sizes if they can be fused into generator loops of the same size.
@@
 
-The reduction of the clustering problem to integer linear programming was previously described by\cite{megiddo1998optimal}, though they do not consider filtering operations.
+The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider filtering operations.
 

### 2014-04-08 2014betterfusionforfilters/section/02-CNF.tex
update CNF

@@
 This form is a list of combinator bindings, where variables are split into scalar and array variables.
-The main restriction is that worker functions may only reference scalar variables, and thus do not perform array combinators.
+The main restriction is that worker functions may only reference scalar variables, nor may they include any array combinators.
 
-The only combinators we fuse are @map@, @filter@, @fold@ and @gather@.
-Most of these are standard combinators except for @gather@, which is equivalent to @gather xs ys = map (index xs) ys@.
-However, as we support no @index@ operation, @gather@ is implemented as a primitive.
+We support the following combinators: $@map@_n$, @filter@, @fold@, @gather@ and @cross@.
+Most of these are standard combinators except for @gather@ and @cross@, however $@map@_n$ differs slightly from usual @map@ in that it takes $n$ arrays of the same length, and applies the worker function to all elements of the same index.
+The @gather@ combinator is equivalent to @gather xs ys = map (index xs) ys@.
+However, as we support no @index@ operation, and worker functions may not refer to arrays, @gather@ is implemented as a primitive.
+The @cross@ combinator takes two arrays, and returns the cartesian product of the two.
 
@@
 This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
-The @external@ also signifies variables produced by a non-primitive combinators.
+The @external@ also lists the variables produced by the external computation.
 Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points.
 They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing.
+This also allows us to support combinators other than those here, but as they must be treated conservatively will likely be less optimal than if they were implemented as primitives.
 
-It is important to note that because of the purity of Haskell, we are free to take certain liberties when reordering the program.
+It is important to note that because of the purity of Haskell, we are free to take certain liberties when fusing the program.
 None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
+This means that any reordering of the program is allowed, as long as mentioned variables are bound beforehand.
 \input{figures/Program.tex}

### 2014-04-11 2014betterfusionforfilters/section/03-SizeInference.tex
some example motivating programs. sadly, closest points is 3 clusters for us but only 4 clusters for stream fusion
add @cross@ combinator. todo fix up constraint checking
Add diagrams for normalize2 example
size inference update

@@
+%!TEX root = ../Main.tex
 \section{Size inference}
+
 Knowing the relative sizes of loops is important for fusion.
@@
 \emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.
+
+After size inference, the goal is to have a function that only requires runtime size checks at the start of the function. This outlaws certain filter constraints.
 
@@
 The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size.
+Similarly, the size of @external@ outputs is not known at all, and thus cannot be constrained.
 
@@
 \begin{tabbing}
+MMMMM       \= MM \= MMMMMM \= \kill
+$rate$      \> @::=@ \> $Filtered_n$        \> (output of @filter@s) \\
+            \> $~|$  \> $External_n$        \> (output of @external@s) \\
+            \> $~|$  \> $rate \times rate$  \> (cross product) \\
+            \> $~|$  \> $k_n$               \> (rate variable)\\
+\end{tabbing}
+
+Finding the relative sizes of arrays is very similar to type inference.
+A rate can be either some rate variable, a cross product of two rates, or an unknown, but fixed, filtered or external output.
+For each binding, a set of equality constraints are created.
+
+\begin{tabbing}
 MMMMM       \= MM \= MMMM \= MM \= \kill
-$constraint$\> @::=@ \> $array$  \> $=_c$ \> $array$ \\
-            \> $~|$  \> $array$  \> $\le_c$ \> $array$ \\
-\\
-$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{constraint\}$ \\
-MMMMMMMMMMMM                 \= MM  \= \kill
+$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{var = rate\}$ \\
+MMMMMMMMMMMMMM               \= MM  \= \kill
 $constrs(o = @fold@~f~n)  $ \> $=$ \> $\{\}$     \\
-$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o =_c n~|~n \in ns\}$     \\
-$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o \le_c b\}$              \\
-$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o =_c i\}$ \\
+$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o = n~|~n \in ns\}$     \\
+$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_b\}$              \\
+$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o = i\}$ \\
+$constrs(o = @cross@~a~b) $ \> $=$ \> $\{o = a \times b\}$ \\
+$constrs(outs = @external@~ins) $ \> $=$ \> $\{o = External_o~|~o \in outs\}$ \\
 \end{tabbing}
@@
 $\bigcup_{b \in binds} constrs(b)$.
-Since equality is transitive, the transitive closure of the constraints is useful.
-For example, if $a =_c b$ and $b =_c c$, then $a =_c c$. 
-After finding the transitive closure the constraints are partitioned into equivalence sets.
-All combinators whose constraints are in the same equivalence set may be fused together.
+The generated equalities are grouped into equivalence classes.
+
+All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.
+These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.
+In other words, only programs that can be statically determined to require only array length checks at the start of execution.
+
+\TODO{Well\ldots} note that $Filtered_n$, $External_n$ are \emph{existential} constructors, bound at $n$.
+This basically translates into them not being able to be referred to \emph{before} their bindings.
+Also note that there can be at most one distinct constructor per equivalence class.
+Go through the examples below.
+
+\begin{code}
+ normalize2 :: Array Int -> Array Int
+ normalize2 us                   {us   = k_us}
+  = let sum1 = fold   (+) 0 us   
+        gts  = filter (>0)  us   {gts  = Filtered_us}
+        sum2 = fold   (+) 0 gts   
+        nor1 = map  (/sum1) us   {nor1 = k_us}
+        nor2 = map  (/sum2) us   {nor2 = k_us}
+    in (nor1, nor2)
+\end{code}
+
+\begin{code}
+ { us   : us = nor1 = nor2
+ , gts  : Filtered_us }
+\end{code}
 
@@
 
-\begin{tabbing}
-MMM       \= MM \= MMMMMM \= MM \= \kill
-$valid$   \> @::@  \> $\{constraint\}$  \> $\to$ \> $\mathbb{B}$ \\
-MMMMMMMMMMMM                 \= MM  \= \kill
-$valid(\{             \})$   \> $=$ \> $\top$  \\
-$valid(\{a =_c b, rest\})$   \> $=$ \> $valid(subst~a:=b~in~rest)$  \\
-$valid(\{a \le_c a, rest\})$ \> $=$ \> $\bot$                        \\
-$valid(\{a \le_c b, rest\})$ \> $=$ \> $a \le_c \_ \not\in closure(rest) \wedge valid(rest)$                        \\
-\\
-$closure(\{\}, out)$                \> $=$ \> $out$ \\
-$closure(\{a =_c b, rest\}, out)$   \> $=$ \> $closure(subst~a:=b~in~rest, subst~a:=b~in~out)$  \\
-$closure(\{a \le_c b, rest\}, out)$ \> $=$ \> $closure(rest, \{a \le_c b\} \cup out)$ \\
-\end{tabbing}
-
 \subsection{Generators}

### 2014-04-12 2014betterfusionforfilters/Main.tex
Wibble
explicitly state similarities and differences to existing ILP formulations
Add diagrams for normalize2 example
size inference

@@
 \title{Better fusion for filters}
-
 \authorinfo{ 
@@
 Finding good clusterings of operators is non-trivial, as we must achieve a balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
-We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find an good solution.
+We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find a good solution. 
 \end{abstract}
 
-% BL: I've removed uses of the word "optimal" here because we don't have a proper
-% language cost model, and our solution is ultimately based on heuristics. ILP can find 
-% an "optimal" solution for a heuristic cost metric, but this isn't the thing we're
-% really trying to optimize.
+% BL: I've removed uses of the word "optimal" here because we don't have a proper language cost model, and our solution is ultimately based on heuristics.
+ILP can find  an "optimal" solution for a heuristic cost metric, but this isn't the thing we're really trying to optimize.
 
@@
 
+\section*{Acknowledgements}
+Many thanks are due to
+Robert Clifton-Everest,
+Kai Engelhardt,
+Bill Kroon,
+Frederik Madsen,
+Abdallah Saffidine,
+and Jingling Xue
+for enlightening discussions relating to this work.
+
 \bibliographystyle{plain}

### 2014-04-12 2014betterfusionforfilters/section/03-SizeInference.tex
size inference

@@
 
-After size inference, the goal is to have a function that only requires runtime size checks at the start of the function. This outlaws certain filter constraints.
-
 It is important to emphasise that we are only interested in relative sizes of arrays; which arrays are equal, and which may be smaller.
@@
 
-Talk about the relation to Chatterjee\cite{chatterjee1991size}.
+Size inference has been explored before in the context of fusion by Chatterjee~\cite{chatterjee1991size}, but the only constraints they support are equality.
+Their formulation has no filtering operations, and all array operations in their case generate the same size as at least one of the operation's inputs.
+A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
+Since we have fewer and simpler primitives, our size inference does not need to be as complicated as Jay's shape inference.
 
@@
 $constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o = n~|~n \in ns\}$     \\
-$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_b\}$              \\
+$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_o\}$              \\
 $constrs(o = @gather@~i~d)$ \> $=$ \> $\{o = i\}$ \\
@@
 $\bigcup_{b \in binds} constrs(b)$.
-The generated equalities are grouped into equivalence classes.
+The constraints are all equalities, so the generated equalities are easily grouped into equivalence classes.
 
@@
 These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.
-In other words, only programs that can be statically determined to require only array length checks at the start of execution.
-
-\TODO{Well\ldots} note that $Filtered_n$, $External_n$ are \emph{existential} constructors, bound at $n$.
-This basically translates into them not being able to be referred to \emph{before} their bindings.
-Also note that there can be at most one distinct constructor per equivalence class.
-Go through the examples below.
 
@@
   = let sum1 = fold   (+) 0 us   
-        gts  = filter (>0)  us   {gts  = Filtered_us}
+        gts  = filter (>0)  us   {gts  = Filtered_gts}
         sum2 = fold   (+) 0 gts   
@@
 
-\begin{code}
- { us   : us = nor1 = nor2
- , gts  : Filtered_us }
-\end{code}
+\subsection{Runtime checks on array lengths}
+The $@map@_n$ combinator requires all its input arrays to be the same size.
+We do not, however, want to perform runtime checks on the size of arrays before every combinator.
+We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.
+Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.
 
-\subsection{Disallowed filters}
+We use the constraints to rule out any programs that require runtime checks on array lengths during the program.
+Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.
+Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid.
+Finally, any initial equality constraints, such as two input arrays being the same length, or one input array being the length of the cross product of two others, are dealt with by inserting runtime checks only at the start of the function.
 
-Filtered constraints of the form $a \le_c b$ are special.
-The actual size of the filtered output is unknown, and so cannot be constrained.
-For example, consider the constraints generated by this program:
+The constraints $Filtered_n$, $External_n$ are treated as \emph{existential} constructors, bound at $n$.
+This means that such constraints are not able to be referred to before their bindings.
+Intuitively, we know that the output of a @filter@ or an @external@ has \emph{some} length, but we don't know what its value is until after they have executed.
+Thus, constraining the input of something to be some exact length, before we know what the length is, is impossible.
+
+Similarly, these constructors cannot be constrained to equal any other constructor. 
+For example, $Filtered_n = External_n$ is invalid, as is $a \times b = Filtered_n$.
+This corresponds to each equivalence class having \emph{at most} one distinct constructor associated with it.
+Any other elements of the equivalence class must be either the same constructor, or simply rate variables.
+
+\subsubsection{Examples}
+Following are a few examples of disallowed programs.
+Consider the constraints generated by this program:
 \begin{code}
 test vs    =
- let flt   = filter p vs
-     wrong = map2   f flt vs
+ let flt   = filter p vs        { flt   = Filtered_flt }
+     wrong = map2   f flt vs    { wrong = flt = vs     }
  in  wrong
 \end{code}
-With the resulting constraints being $\{flt \le_c vs, wrong =_c flt, wrong =_c vs\}$.
-By transitivity, we find that $flt \le_c flt$ - the size of the filtered output is upper-bounded by the size of the filtered output! 
-Why is this bad?
-Actually implementing this would require to buffer the entire contents of @flt@, so for simplicity we disallow it. Talk more about this.
+By transitivity, we find that $vs = Filtered_{flt}$, but $vs$ is bound before @flt@.
+Operationally, actually implementing this would require to buffer the entire contents of @flt@, then check the length of @flt@ against the length of @vs@.
 
@@
 test vs    =
- let flt   = filter p  vs
-     flt2  = filter p' vs
-     wrong = map2   f  flt flt2
+ let flt   = filter p  vs       { flt   = Filtered_flt  }
+     flt2  = filter p' vs       { flt2  = Filtered_flt2 }
+     wrong = map2   f  flt flt2 { wrong = flt = flt2    }
  in  wrong
 \end{code}
-Generates the constraints $\{flt \le_c vs, flt2 \le_c vs, flt =_c flt2\}$.
-Here we are constraining two different filtered outputs to be the same size.
-We disallow this.
-
-Not all constraints on filtered data are disallowed, of course;
-\begin{code}
-test vs    =
- let flt   = filter p  vs
-     xs    = map    f  flt
-     ys    = map    g  flt
-     sm    = map2   f  xs ys
- in  sm
-\end{code}
-This generates the constraints $\{flt \le_c vs, xs =_c flt, ys =_c flt, sm =_c xs, sm =_c ys\}$,
-but this is perfectly valid, since the constraints are all actually based on @flt@.
+Again, transitivity here requires that $Filtered_{flt} = Filtered_{flt2}$, but since these are both concrete constructors, we cannot have such a constraint.
+Basically, we are constraining two different filtered outputs to be the same size.
+Operationally, implementing this would require to buffer both @flt@ and @flt2@, before checking the two lengths.
 

### 2014-04-12 2014betterfusionforfilters/Main.tex
Rewrite abstract to highlight that our main contribution is how to handle filters

@@
 \begin{document}
-\title{Better fusion for filters}
+
+\title{Fusing Filters with Integer Linear Programming}
+% Better title?
+% \title{Better fusion for filters}
+
 \authorinfo{ 
@@
 \begin{abstract}
-Recent work on array fusion demonstrates how to compile a specific class of data flow programs into single imperative loops.
-However, until now there has been no satisfactory way to take the operator graph expressing a complete program and split it into fusible subgraphs.
-Finding good clusterings of operators is non-trivial, as we must achieve a balance between minimizing memory traffic and loop overhead, while maximizing cache locality.
-We show how to formulate operator clustering as an \emph{integer linear programming} problem, and use an external library to find a good solution. 
+Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
+However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
+Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle filter-like operators which produce arrays of a different size than the input.
+We extend the existing ILP approach with support for filters, using an external ILP solver to find good clusterings.
 \end{abstract}
-
-% BL: I've removed uses of the word "optimal" here because we don't have a proper language cost model, and our solution is ultimately based on heuristics.
-ILP can find  an "optimal" solution for a heuristic cost metric, but this isn't the thing we're really trying to optimize.
 

### 2014-04-12 2014betterfusionforfilters/section/01-Introduction.tex
Rewrite abstract to highlight that our main contribution is how to handle filters

@@
 
-Recent work on data flow fusion~\cite{lippmeier2013flow} shows how to compile a specific class of data flow programs into \emph{single}, efficient imperative loops.
+Data flow fusion~\cite{lippmeier2013flow} is a technique to compile a specific class of data flow programs into single, efficient imperative loops.
 This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.

### 2014-04-12 2014betterfusionforfilters/section/01-Introduction.tex
More work on intro, suggested title change

@@
 This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
-The key benefits of data flow fusion over prior work are that 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between combinators.
+The key benefits of data flow fusion over prior work are: 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
 
-Ultimately, though, we cannot expect \emph{all} programs to be fused into single imperative loops, as most contain fundamental data dependencies that prevent complete fusion. For example:
+Fusion-preventing dependencies express the fact that some operators simply must wait for others to complete before they can produce their own output. For example, in the following:
 \begin{code}
@@
 If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using two loops: one to determine the sum, and one to divide the elements.
-The application of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
+The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
 
 However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
-The following function is similar to the one from before, except that it offers several unique \emph{but mutually exclusive} approaches to fusion.
-A diagram of the data flow graph for this example is on the next page.
-
+The following example offers \emph{several} unique, but mutually exclusive approaches to fusion.
+Figure~\ref{f:normalize2-cluterings} on the next page show some of the possibilities.
 \begin{code}
@@
  normalize2 us
-  = let sum1 = fold   (+) 0 us      (A1) (S1) (I1)
-        gts  = filter (>0)  us      (A1) (S2) (I1)
-        sum2 = fold   (+) 0 gts     (A1) (S2) (I2)
-        nor1 = map  (/sum1) us      (A2) (S3) (I3)
-        nor2 = map  (/sum2) us      (A2) (S4) (I3)
+  = let sum1 = fold   (+)  0   us
+        gts  = filter (> 0)    us
+        sum2 = fold   (+)  0   gts
+        nor1 = map    (/ sum1) us
+        nor2 = map    (/ sum2) us
     in (nor1, nor2)
 \end{code}
-Our approach, @A1-A2@, requires only two loops for this particular case.
-Stream fusion (@S1-S4@) is unable to fuse @sum1@ and @gts@ together, as there is nothing to be inlined between them, so requires four whole traversals of the input array, @us@.
-Similarly, existing imperative solutions (@I1-I3@) are unable to fuse @gts@ and @sum2@ together, as the iteration size is completely different.
 
-Our contributions are as follows:
+In Figure~\ref{f:normalize2-cluterings}, the dotted lines show possible clusterings of operators.
+Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers.
+The best existing ILP approach will chose the solution on the right as it cannot cluster operators that consume arrays of different lengths.
+Our system choses the solution in the middle, which is also optimal for this example.
 
 % NOTE: This set of bullets needs to fit on the first page, without spilling to the second.
+Our contributions are as follows:
 \begin{itemize}
 \item   
-We use existing integer linear programming clustering formulations such as Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction},
-and modify them to express constraints on fusing loops of different sizes if they can be fused into generator loops of the same size.
-\TODO{ref}
-% We present an algorithm to cluster the operators of a data flow graph into sub-graphs that can be fused into individual imperative loops.
-% We express the constraints on fusion as an integer linear program, and use an off-the-shelf solver to compute a valid clustering. \TODO{ref}
+We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for filtering operators.
+Filters can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
 
@@
 
-% Consider the @normaliseInc@ function below; it requires at least two loops, since @sum@ must be computed from the entire array before any of the output array @norm@ can be filled.
-Scheduling @incs@ and @sum@ into a single loop, then @norm@ into another loop only requires two loops, but requires @incs@ to be stored in a manifest array before being read again by @norm@.
-If we were to schedule @sum@ in its own loop and @incs@ and @norm@ together, we would only require one array to be filled - the output array.
-
-% Combinators are a natural and composable way of expressing array programs, with many correctness benefits over imperative programs.
-If compiled naively, however, executing the combinator program will often require many more loops and intermediate arrays than an imperative program written by hand.
-These intermediate arrays can often be removed by merging several combinators into a single loop.
-This merging is known as \emph{fusion}.
-
-% Existing fusion systems for Haskell such as stream fusion~\cite{coutts2007streamfusion}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules, to fuse combinators without having to modify the compiler itself.
-This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
-However, for scientific applications where efficiency is important, we cannot afford these limitations.
-We have instead implemented our fusion system as a compiler plugin, which gives us total control of when fusion will occur.
-
 % As real programs contain tens or hundreds of individual operators, performing an exhaustive search for an optimal clustering is not feasible, and greedy algorithms tend to produce poor solutions.

### 2014-04-12 2014betterfusionforfilters/section/01-Introduction.tex
Add notes and move sections around

@@
 We present a simplification to constraint generation that is also applicable to some existing integer linear programming formulations such as Megiddo's,
-where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two. \TODO{ref}
+where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two (\S\ref{s:OptimisedConstraints}).
 
@@
 Our constraint system also encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program.
-For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced.
-\TODO{ref}
+For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced (\S\ref{s:ObjectiveFunction}).
 

### 2014-04-12 2014betterfusionforfilters/section/03-SizeInference.tex
More work on intro, suggested title change
Add notes and move sections around

@@
 We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.
-Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.
+
+\TODO{We assign different rates to both @ys1@ and @ys2@ in this example.}
+\begin{code}
+ diff xs = let ys1 = filter p xs
+               ys2 = filter p xs
+           in  (ys1, ys2)
+\end{code}
+
+% \ben{This point is only relavant in the context of our specific implementation, which is no the focus of this paper. }
+% Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.
 
@@
 Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.
+
+\ben{Refactor presentation to use the actual $\exists k. $ quantifier, as in ``The essence of ML type inference'', from ATAPL}
+
 Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid.
@@
 
-The constraints $Filtered_n$, $External_n$ are treated as \emph{existential} constructors, bound at $n$.
-This means that such constraints are not able to be referred to before their bindings.
+The constraints $Filtered_n$, $External_n$ are treated as \emph{existential} constructors, bound at $n$. This means that such constraints are not able to be referred to before their bindings.
 Intuitively, we know that the output of a @filter@ or an @external@ has \emph{some} length, but we don't know what its value is until after they have executed.

### 2014-04-12 2014betterfusionforfilters/section/04-ILP.tex
More work on intro, suggested title change
Add notes and move sections around

@@
 \section{Integer linear program formulation}
+\label{s:ILP}
 
@@
 For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+
+\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
 
@@
 \subsubsection{The simple version}
+\TODO{Why is this ``the simple version''?
+Why present a simple version before the real one?} \TODO{We need to show a complete constraint system at the start.
+Maybe use a simple objective function to start with, then refine it in a later section.}
+
 \paragraph{Acyclic and precedence-preserving}
@@
 Note, however, that other edges may still constrain $i$ or $j$ by transitivity.
+\ben{If there is ``no constraint'' on the $\pi$ values then why bother adding the constraint at all?}
 
@@
 \paragraph{Fusion between different types}
+
+\ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}
+
 \begin{tabbing}
@@
 If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators.
+\ben{What is a subrate?
+How do we determine the subrates of each rate?}
 \TODO{Expand}.
@@
 \subsubsection{The objective function}
+\label{s:ObjectiveFunction}
 
@@
 The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
+\TODO{Refer to example code that demonstrates these different opportunities.}
 
@@
 \subsubsection{An optimised version}
+\label{s:OptimisedConstraints}
 We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.

### 2014-04-12 2014betterfusionforfilters/section/05-Benchmarks.tex
Add notes and move sections around

%!TEX root = ../Main.tex
\section{Benchmarks}
\label{s:Benchmarks}
\ben{Use the larger programs as benchmarks, or as running examples. We only need small, simple programs to demonstrate how the algorithm works.}

\subsection{Quickselect}
We can't quite do @quickselect@, sadly; it requires @head@.
\begin{code}
select :: Vec a -> Int -> a
select xs k
 | length xs > 0
 = let p    = head xs 
       xslt = filter  (<p) xs
       xsge = filter (>=p) xs
   in  if     length xslt > k
       then   select xsge (k - length xslt)
       else   select xslt  k
 | length xs == 1
 = head xs
 | length xs == 0
 = error "Empty array!"

median :: Vec a -> a
median xs
 = select xs (length xs `div` 2)
\end{code}

\subsection{Closest pairs}
Closest pairs makes use of fast-ish @median@ to ensure balanced division of work.
\begin{code}
closest :: Vec Pt -> (Pt,Pt)
closest pts
 | length pts < 250
 = naive pts
 | otherwise
 = divide pts

divide :: Vec Pt -> (Pt,Pt)
divide pts
 = let p      = median pts
       aboves = filter (above p) pts
       belows = filter (below p) pts
       above' = closest aboves
       below' = closest belows

       border = min (distance above') (distance below')

       aboveB = filter (above (p - border)) pts
       belowB = filter (below (p + border)) pts

       cs     = cross aboveB belowB
       bord   = minBy distance cs
   in  above' `minDist` below' `minDist` bord

naive :: Vec Pt -> (Pt,Pt)
naive pts
 = let c = cross pts pts
   in  minBy distance c

minBy :: Ord b => (a -> b) -> Vec a -> a
minBy f xs
 = fold (min . f...) ... xs
\end{code}

Let's translate @divide@ to a program in CNF.
\begin{code}
divide :: Vec Pt -> (Pt,Pt)
divide pts
 = let p      = external pts

       aboves = filter (... p) pts          -- A
       belows = filter (... p) pts          -- A

       above' = external aboves
       below' = external belows
       border = external above' below'

       aboveB = filter (... p border) pts   -- C
       belowB = filter (... p border) pts   -- B

       cs     = cross  aboveB belowB        -- C
       bord   = fold   (...) cs             -- C

       min'   = external above' below' bord
   in  min'
\end{code}
where @A@, @B@ and @C@ are distinct clusters.

What happens if we do this using stream fusion?
\begin{code}
divide :: Vec Pt -> (Pt,Pt)
divide pts
 = let p      = external pts

       aboves = filter (... p) pts          -- A
       belows = filter (... p) pts          -- B

       above' = external aboves
       below' = external belows
       border = external above' below'

       aboveB = filter (... p border) pts   -- D
       belowB = filter (... p border) pts   -- C

       cs     = cross  aboveB belowB        -- D
       bord   = fold   (...) cs             -- D

       min'   = external above' below' bord
   in  min'
\end{code}
So, we have four clusters instead of three, and the same number of manifest arrays. Not particularly impressive.

\subsection{Quickhull}
Is Quickhull any better? Seems like it's just one cluster; @filterMax@. And we don't have append (@++@), anyway.
\begin{code}
quickhull :: Vec Pt -> Vec Pt
quickhull pts
 = let top  = fold getTop pts
       bot  = fold getBot pts
       tops = hull (top,bot) pts
       bots = hull (bot,top) pts
   in  tops ++ bots

hull :: (Pt,Pt) -> Vec Pt -> Vec Pt
hull line@(l,r) pts
 = let pts' = filter (above   line) pts
       ma   = fold   (maxFrom line) pts'
       hl   = hull   (l, ma)        as
       hr   = hull   (ma, r)        as
   in  hl  ++ hr
\end{code}
Yep. @hull@ is pretty boring.

Something with a \emph{fold}, and then filtering or mapping based on that fold would be good.
Or something with
\begin{code}
let xs' = filter f    xs
    i   = fold   g    xs'
    xs''= map   (h i) xs'
\end{code}
would be really good.

FFT - not really.

\subsection{QuadTree}
\begin{code}
bounds pts
 = let x1 = fold minX pts
       y1 = fold minY pts
       x2 = fold maxX pts
       y2 = fold maxY pts
   in ((x1,y1), (x2,y2))

%splitbox ((x1,y1), (x2,y2))
% = let xm = mid x1 x2
%       ym = mid y1 y2
%   in ( ((x1, y1), (xm, ym))
%      , ((x1, ym), (xm, y2))
%      , ((xm, y1), (x2, ym))
%      , ((xm, ym), (x2, y2)))
%
quadtree pts
 = go (bounds pts)
 where
  go bounds pts
   | length pts > 0
   = let (b1,b2,b3,b4) = splitbox bounds
         pts1  = filter (inbox b1) pts
         pts2  = filter (inbox b2) pts
         pts3  = filter (inbox b3) pts
         pts4  = filter (inbox b4) pts
         tree1 = go b1 pts1
         tree2 = go b2 pts2
         tree3 = go b3 pts3
         tree4 = go b4 pts4
      in Tree tree1 tree2 tree3 tree4
   | otherwise
   = Empty
\end{code}
It's easy to see that @bounds@ should only require one loop, but stream fusion requires four, as there is no inlining that can occur.
The same is true of @go@ in @quadtree@. Our implementation only requires one loop for each of these.

### 2014-04-12 2014betterfusionforfilters/section/02-CNF.tex
Add notes concerning loop communication graphs

@@
 \section{Combinator normal form}
+\TODO{Mention that in the data flow graph we draw materialized vectors with a box around them.}
+\TODO{Say a program in this form is just a textual representation of its data-flow graph.
+This is similar to Loop Communication Graphs (LCG) from \cite{sarkar1991optimization}, except that we avoid needing ``array access vectors'' that track which direction an array is accessed in.
+All our operators access elements from front to back, and we don't consider reversals.
+},
 We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.

### 2014-04-13 2014betterfusionforfilters/section/02-CNF.tex
Update Fig 1 to represent true clustering produced by related work

@@
 This is similar to Loop Communication Graphs (LCG) from \cite{sarkar1991optimization}, except that we avoid needing ``array access vectors'' that track which direction an array is accessed in.
-All our operators access elements from front to back, and we don't consider reversals.
-},
+Our operators are directionless, operating in a bulk manner with no implied order of processing for the individual array elements.}.
+
 We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.

### 2014-04-13 2014betterfusionforfilters/section/04-ILP.tex
Add data flow diagram of normalizeInc

@@
 
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics[scale=0.5]{figures/ex2-normalizeInc.pdf}
+\end{center}
+\caption{Possible clusterings for \texttt{normalizeInc}}
+\end{figure}
+
 \TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering.
-Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.}
+Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.
+Compare with prior work on typed fusion that only tries to minimise the number of clusters.}
 \begin{code}

### 2014-04-15 2014betterfusionforfilters/section/03-SizeInference.tex
attempt to turn size inference into natural deduction proofs. ugly. why can't I center them?

@@
 MMMMM       \= MM \= MMMMMM \= \kill
-$rate$      \> @::=@ \> $Filtered_n$        \> (output of @filter@s) \\
-            \> $~|$  \> $External_n$        \> (output of @external@s) \\
+$rate$      \> @::=@ \> $k_n$               \> (rate variable)\\
             \> $~|$  \> $rate \times rate$  \> (cross product) \\
-            \> $~|$  \> $k_n$               \> (rate variable)\\
+%            \> $~|$  \> $Filtered_n$        \> (output of @filter@s) \\
+%            \> $~|$  \> $External_n$        \> (output of @external@s) \\
+\\
+$constraint$ \> @::=@ \> $rate = rate$                      \\
+             \> $~|$  \> $rate \le rate$                    \\
+             \> $~|$  \> $constraint \wedge constraint$     \\
+             \> $~|$  \> $\forall k_n. constraint$          \\
+             \> $~|$  \> $\exists k_n. constraint$          \\
 \end{tabbing}
@@
 
-\begin{tabbing}
-MMMMM       \= MM \= MMMM \= MM \= \kill
-$constrs$  \> @::@  \> $bind$  \> $\to$ \> $\{var = rate\}$ \\
-MMMMMMMMMMMMMM               \= MM  \= \kill
-$constrs(o = @fold@~f~n)  $ \> $=$ \> $\{\}$     \\
-$constrs(o = @map@_n~f~ns)$ \> $=$ \> $\{o = n~|~n \in ns\}$     \\
-$constrs(o = @filter@~f~b)$ \> $=$ \> $\{o = Filtered_o\}$              \\
-$constrs(o = @gather@~i~d)$ \> $=$ \> $\{o = i\}$ \\
-$constrs(o = @cross@~a~b) $ \> $=$ \> $\{o = a \times b\}$ \\
-$constrs(outs = @external@~ins) $ \> $=$ \> $\{o = External_o~|~o \in outs\}$ \\
-\end{tabbing}
+\vspace{1em}
 
-Then, for some program $\lambda\ldots~@let@~binds~@in@~\ldots$, find the union of all constraints:
-$\bigcup_{b \in binds} constrs(b)$.
-The constraints are all equalities, so the generated equalities are easily grouped into equivalence classes.
+\infer{\lambda scalars~arrays \to @let @ binds : \forall_{a \in arrays} k_a. C}
+      { \{k_a | a \in arrays\} \vdash binds : C}
 
+\vspace{1em}
+
+\infer{\Gamma \vdash o = @fold@~f~n; binds : C}
+      {\Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+\infer{\Gamma \vdash o = @map@_n~f~ns; binds : \forall k_o. \bigwedge_{n \in ns}\{k_o = k_n\} \wedge C}
+      {k_o, \Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+\infer{\Gamma \vdash o = @filter@~f~n; binds : \exists k_o. k_o \le k_n \wedge C}
+      {k_o, \Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+\infer{\Gamma \vdash o = @gather@~i~d; binds : \forall k_o. k_o = k_i \wedge C}
+      {k_o, \Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+\infer{\Gamma \vdash o = @cross@~a~b; binds : \forall k_o. k_o = k_a \times k_b \wedge C}
+      {k_o, \Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+\infer{\Gamma \vdash outs = @external@~ins; binds : \exists_{o \in outs} k_o. C}
+      {\{k_o | o \in outs\} \cup \Gamma \vdash binds                 : C}
+
+\vspace{1em}
+
+The generated equalities are easily grouped into equivalence classes.
 All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.

### 2014-04-15 2014betterfusionforfilters/section/03-SizeInference.tex
working on size inference

@@
 \begin{tabbing}
-MMMMM       \= MM \= MMMMMM \= \kill
+MMMMM       \= MM \= MMMMMMMMMMM \= \kill
 $rate$      \> @::=@ \> $k_n$               \> (rate variable)\\
@@
 \\
-$constraint$ \> @::=@ \> $rate = rate$                      \\
-             \> $~|$  \> $rate \le rate$                    \\
-             \> $~|$  \> $constraint \wedge constraint$     \\
-             \> $~|$  \> $\forall k_n. constraint$          \\
-             \> $~|$  \> $\exists k_n. constraint$          \\
+$constraint$ \> @::=@ \> $rate = rate$                          \> (equality constraint) \\
+             \> $~|$  \> $constraint \wedge constraint$         \> (conjunction) \\
+             \> $~|$  \> $\forall k_n.\ constraint$              \> (forall quantification) \\
+             \> $~|$  \> $\exists k_n.\ constraint$              \> (exists with no bounds)\\
+             \> $~|$  \> $\exists k_n \le rate.\ constraint$     \> (exists with upper bound)    \\
+\\
+$\tau$       \> @::=@ \> $rate$                                  \> (loop size of some rate) \\
+             \> $~|$  \> @external@                              \> (external and unfusible) \\
 \end{tabbing}
+
+\[ binds \vDash constraint : (\tau_{name} \Rightarrow \tau) \]
 
@@
 
-\infer{\lambda scalars~arrays \to @let @ binds : \forall_{a \in arrays} k_a. C}
-      { \{k_a | a \in arrays\} \vdash binds : C}
+\infer{\lambda scalars~arrays \to @let @ binds : \forall_{a \in arrays} k_a. C : T}
+      { binds \vDash C : T}
 
@@
 
-\infer{\Gamma \vdash o = @fold@~f~n; binds : C}
-      {\Gamma \vdash binds                 : C}
+\infer{o = @fold@~f~n; binds \vDash C : \tau_o = k_n, T}
+      {binds                 \vDash C : T}
 
@@
 
-\infer{\Gamma \vdash o = @map@_n~f~ns; binds : \forall k_o. \bigwedge_{n \in ns}\{k_o = k_n\} \wedge C}
-      {k_o, \Gamma \vdash binds                 : C}
+\infer{o = @map@_n~f~ns; binds \vDash \forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge C : \tau_o = k_o, T}
+      {binds                   \vDash C : T}
 
@@
 
-\infer{\Gamma \vdash o = @filter@~f~n; binds : \exists k_o. k_o \le k_n \wedge C}
-      {k_o, \Gamma \vdash binds                 : C}
+\infer{o = @filter@~f~n; binds \vDash \exists k_o \le k_n.\ C : \tau_o = k_n, T}
+      {binds                   \vDash C : T}
+Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
+This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.
 
@@
 
-\infer{\Gamma \vdash o = @gather@~i~d; binds : \forall k_o. k_o = k_i \wedge C}
-      {k_o, \Gamma \vdash binds                 : C}
+\infer{o = @gather@~i~d; binds \vDash \forall k_o.\ k_o = k_i \wedge C : \tau_o = k_i, T}
+      {binds                   \vDash C : T}
 
@@
 
-\infer{\Gamma \vdash o = @cross@~a~b; binds : \forall k_o. k_o = k_a \times k_b \wedge C}
-      {k_o, \Gamma \vdash binds                 : C}
+\infer{o = @cross@~a~b; binds \vDash \forall k_o.\ k_o = k_a \times k_b \wedge C : \tau_o = k_a \times k_b, T}
+      {binds                  \vDash C : T}
 
@@
 
-\infer{\Gamma \vdash outs = @external@~ins; binds : \exists_{o \in outs} k_o. C}
-      {\{k_o | o \in outs\} \cup \Gamma \vdash binds                 : C}
+\infer{outs = @external@~ins; binds \vDash \exists_{o \in outs} k_o.\ C : \tau_{outs} = @external@, T}
+      {binds                        \vDash C : T}
 
@@
 
-The generated equalities are easily grouped into equivalence classes.
+Solving the generated equality constraints allows each rate to be grouped into equivalence classes.
 All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.

### 2014-04-17 2014betterfusionforfilters/section/03-SizeInference.tex
minor cleanup on judgments

@@
 
-\[ binds \vDash constraint : (\tau_{name} \Rightarrow \tau) \]
+\newcommand{\constraint}[3]{#1 \vDash #2 : #3}
+\newcommand{\inferC}[2]{
+    $$\begin{array}{c}
+        \infer{#1}{#2}
+    \end{array}$$}
+\newcommand{\ruleIC}[3]{
+    \inferC
+        {\constraint{#1; binds}{#2 C}{#3 ~ T}}
+        {\constraint{binds}{C}{T}}}
+
+\[ \constraint{binds}{constraint}{(\tau_{name} \Rightarrow \tau)} \]
 
@@
 
-\infer{\lambda scalars~arrays \to @let @ binds : \forall_{a \in arrays} k_a. C : T}
-      { binds \vDash C : T}
+\inferC{\constraint{\lambda scalars~arrays \to @let @ binds}{\forall_{a \in arrays} k_a. C}{T}}
+       {\constraint{binds}{C}{T}}
 
@@
 
-\infer{o = @fold@~f~n; binds \vDash C : \tau_o = k_n, T}
-      {binds                 \vDash C : T}
+\ruleIC{o = @fold@~f~n}{}{\tau_o = k_n, }
 
@@
 
-\infer{o = @map@_n~f~ns; binds \vDash \forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge C : \tau_o = k_o, T}
-      {binds                   \vDash C : T}
+\ruleIC{o = @map@_n~f~ns}{\forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge}{\tau_o = k_o, }
 
@@
 
-\infer{o = @filter@~f~n; binds \vDash \exists k_o \le k_n.\ C : \tau_o = k_n, T}
-      {binds                   \vDash C : T}
+\ruleIC{o = @filter@~f~n}{\exists k_o \le k_n.\ }{\tau_o = k_n, }
+
 Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
@@
 
-\infer{o = @gather@~i~d; binds \vDash \forall k_o.\ k_o = k_i \wedge C : \tau_o = k_i, T}
-      {binds                   \vDash C : T}
+\ruleIC{o = @gather@~i~d}{\forall k_o.\ k_o = k_i \wedge}{\tau_o = k_i, }
 
@@
 
-\infer{o = @cross@~a~b; binds \vDash \forall k_o.\ k_o = k_a \times k_b \wedge C : \tau_o = k_a \times k_b, T}
-      {binds                  \vDash C : T}
+\ruleIC{o = @cross@~a~b}{\forall k_o.\ k_o = k_a \times k_b \wedge C}{\tau_o = k_a \times k_b, }
 
@@
 
-\infer{outs = @external@~ins; binds \vDash \exists_{o \in outs} k_o.\ C : \tau_{outs} = @external@, T}
-      {binds                        \vDash C : T}
+\ruleIC{outs = @external@~ins}{\exists_{o \in outs} k_o.\ }{\tau_{outs} = @external@, }
 

### 2014-04-18 2014betterfusionforfilters/Main.tex
Add notes and move sections around
into: stress filter is just one example of a length changing operation

@@
 However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
-Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle filter-like operators which produce arrays of a different size than the input.
-We extend the existing ILP approach with support for filters, using an external ILP solver to find good clusterings.
+Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different length than their input.
+We extend the existing ILP approach with support for length changing operators, using an external ILP solver to find good clusterings.
 \end{abstract}
@@
 \input{section/04-ILP.tex}
-\input{section/05-Related.tex}
-\input{section/XX-Programs.tex}
+\input{section/05-Benchmarks.tex}
+\input{section/06-Related.tex}
 \section{Conclusion}

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
into: stress filter is just one example of a length changing operation

@@
 This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
-The key benefits of data flow fusion over prior work are: 1) it can fuse programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
+The key benefits of data flow fusion over this prior work are: 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
 
@@
 
-If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using two loops: one to determine the sum, and one to divide the elements.
+If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using at least two loops: one to determine the sum, and one to divide the elements.
 The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
@@
 The following example offers \emph{several} unique, but mutually exclusive approaches to fusion.
-Figure~\ref{f:normalize2-cluterings} on the next page show some of the possibilities.
+Figure~\ref{f:normalize2-cluterings} on the next page shows some of the possibilities.
 \begin{code}
@@
 \item   
-We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for filtering operators.
-Filters can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
+We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for length changing operators.
+Length changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
 
@@
 
-The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider filtering operations.
+The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider length changing operators.
 

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
intro: update var names in normalize2 example to match diagram

@@
  normalize2 :: Array Int -> Array Int
- normalize2 us
-  = let sum1 = fold   (+)  0   us
-        gts  = filter (> 0)    us
+ normalize2 xs
+  = let sum1 = fold   (+)  0   xs
+        gts  = filter (> 0)    xs
         sum2 = fold   (+)  0   gts
-        nor1 = map    (/ sum1) us
-        nor2 = map    (/ sum2) us
-    in (nor1, nor2)
+        ys1  = map    (/ sum1) xs
+        ys2  = map    (/ sum2) xs
+    in (ys1, ys2)
 \end{code}

### 2014-04-18 2014betterfusionforfilters/Main.tex
intro: s/length/size/ so it matches "Size inference"

@@
 However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
-Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different length than their input.
-We extend the existing ILP approach with support for length changing operators, using an external ILP solver to find good clusterings.
+Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input.
+We extend the existing ILP approach with support for size changing operators, using an external ILP solver to find good clusterings.
 \end{abstract}

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
intro: s/length/size/ so it matches "Size inference"

@@
 Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers.
-The best existing ILP approach will chose the solution on the right as it cannot cluster operators that consume arrays of different lengths.
+The best existing ILP approach will chose the solution on the right as it cannot cluster operators that consume arrays of different sizes.
 Our system choses the solution in the middle, which is also optimal for this example.
@@
 \item   
-We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for length changing operators.
-Length changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
+We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for size changing operators.
+Size changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
 

### 2014-04-18 2014betterfusionforfilters/section/02-CNF.tex
cnf: describe format of data flow graphs

--- 2014-04-08 2014betterfusionforfilters/section/02-CNF.tex
@@
 %!TEX root = ../Main.tex
-%\eject
 
@@
 \end{center}
-
 \caption{Clusterings for normalize2 example: with stream fusion; our system; best imperative system}
-\label{f:normalize2-cluterings}
+\label{f:normalize2-clusterings}
 \end{figure*}
 
-\section{Combinator normal form}
-We can accept functions written in \emph{combinator normal form}, which is a specialised form of first-order array programs detailed in Figure~\ref{f:CombinatorNormalForm}.
-This form is a list of combinator bindings, where variables are split into scalar and array variables.
+% -----------------------------------------------------------------------------
+\section{Combinator Normal Form}
+Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph.
+The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}.
+Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}.
+Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion \CITE.
+We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}).
+
+Once we have determined how the operators in a data flow graph should be clustered, we highlight materialized array variables by drawing them in boxes.
+In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program.
+However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused.
+
+In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
+
+\TODO{blend this in.}
 The main restriction is that worker functions may only reference scalar variables, nor may they include any array combinators.
@@
 
-Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@.
-This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
+Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@.This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
 The @external@ also lists the variables produced by the external computation.

### 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
cnf: describe format of data flow graphs
cnf: copy edits, add types for array combinators, say the CNF program is in a host program

@@
 The following example offers \emph{several} unique, but mutually exclusive approaches to fusion.
-Figure~\ref{f:normalize2-cluterings} on the next page shows some of the possibilities.
+Figure~\ref{f:normalize2-clusterings} on the next page shows some of the possibilities.
 \begin{code}
@@
 
-In Figure~\ref{f:normalize2-cluterings}, the dotted lines show possible clusterings of operators.
+In Figure~\ref{f:normalize2-clusterings}, the dotted lines show possible clusterings of operators.
 Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers.

### 2014-04-18 2014betterfusionforfilters/section/02-CNF.tex
cnf: copy edits, add types for array combinators, say the CNF program is in a host program

@@
 We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}).
+In data flow graphs we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.
 
@@
 
-\TODO{blend this in.}
-The main restriction is that worker functions may only reference scalar variables, nor may they include any array combinators.
+Most of our array combinators are standard, and suggestive types are given at the bottom of Figure~\ref{f:CombinatorNormalForm}.
+The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index.
+As such it is similar to Haskell's @zipWith@, except for the length restriction on the argument arrays.
+The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index.
+The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that were at each index.
+In Haskell this would be @gather arr ixs = map (index arr) ixs@.
+The @cross@ combinator returns the cartesian product of two arrays.
 
-We support the following combinators: $@map@_n$, @filter@, @fold@, @gather@ and @cross@.
-Most of these are standard combinators except for @gather@ and @cross@, however $@map@_n$ differs slightly from usual @map@ in that it takes $n$ arrays of the same length, and applies the worker function to all elements of the same index.
-The @gather@ combinator is equivalent to @gather xs ys = map (index xs) ys@.
-However, as we support no @index@ operation, and worker functions may not refer to arrays, @gather@ is implemented as a primitive.
-The @cross@ combinator takes two arrays, and returns the cartesian product of the two.
+The exact form of the worker functions is left unspecified, as it is not important for the discussion.
+We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment.
+We assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like.
+Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program.
+All access to locally bound array variables is via the formal parameters of array combinators, which ensures that all data dependencies we need to consider for fusion are explicit in the data flow graph.
 
-Since it is unlikely that an entire function will be comprised of these few combinators, we support one additional binding type: @external@.This signifies that the referenced variables are used by a computation that is not a primitive combinator, and must be materialised fully in memory at this stage.
-The @external@ also lists the variables produced by the external computation.
-Without knowing the nature of the computation expressed by @external@, we must naturally take a conservative view, and allow no fusion to occur between at these points.
-They are, in effect, fusion barriers, forcing arrays and scalars to be fully computed before continuing.
-This also allows us to support combinators other than those here, but as they must be treated conservatively will likely be less optimal than if they were implemented as primitives.
+The @external@ binding invokes a host library function that can produce and consume arrays, but not be fused with other combinators.
+All arrays passed to, and returned from host functions are full materialised.
+External bindings are explicit \emph{fusion barriers}, which force arrays and scalars to be fully computed before continuing.
 
-It is important to note that because of the purity of Haskell, we are free to take certain liberties when fusing the program.
-None of the worker functions, nor any @external@ computations may produce visible side-effects; the only observable effect must be to produce their output.
-This means that any reordering of the program is allowed, as long as mentioned variables are bound beforehand.
+% When the function computed is the identity, they can also be used to force a particular intermediate array to be materialized, which provides the programmer fine-grained control over exactly what clustering is produced by our algorithm.
+
 \input{figures/Program.tex}

### 2014-04-20 2014betterfusionforfilters/code/fortran/dumb.f90
very simple fortran example. but gfortran doesn't fuse?

@@ -1,0 +1,21 @@
+ function dumb(A) result (ou)
+     integer, intent(in), dimension(:) :: A
+     integer :: ou
+
+     integer :: i1, i2, N, s1, s2
+     N  = size(A)
+     s1 = 0
+     s2 = 0
+
+     do i1 = 1, N
+         s1 = s1 + A(i1)
+     enddo
+    
+     do i2 = 1, N
+         s2 = s2 - A(i2)
+     enddo
+
+     ou = s1 * s2
+        
+ end function dumb
+

### 2014-04-20 2014betterfusionforfilters/section/03-SizeInference.tex
minor size inf
constraint generation fixes
change constraint generation to array, somewhat prettier
more size inference

@@
 A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
-Since we have fewer and simpler primitives, our size inference does not need to be as complicated as Jay's shape inference.
+Since we have fewer primitives and simpler goals, our size inference does not need to be as complicated as Jay's shape inference.
+
+\newcommand{\constr}[1]{\llbracket #1 \rrbracket}
+\newcommand{\hole}[0]{[]}
+\newcommand{\fillhole}[2]{#1\Left[#2\right]}
 
@@
 \\
-$constraint$ \> @::=@ \> $rate = rate$                          \> (equality constraint) \\
-             \> $~|$  \> $constraint \wedge constraint$         \> (conjunction) \\
-             \> $~|$  \> $\forall k_n.\ constraint$              \> (forall quantification) \\
-             \> $~|$  \> $\exists k_n.\ constraint$              \> (exists with no bounds)\\
-             \> $~|$  \> $\exists k_n \le rate.\ constraint$     \> (exists with upper bound)    \\
+$C$          \> @::=@ \> $true$                                 \> (trivially true) \\
+             \> $~|$  \> $\hole$                                \> (constraint hole) \\
+             \> $~|$  \> $rate = rate$                          \> (equality constraint) \\
+             \> $~|$  \> $C \wedge C$                           \> (conjunction) \\
+             \> $~|$  \> $\forall k_n.\ C$                      \> (forall quantification) \\
+             \> $~|$  \> $\exists k_n.\ C$                      \> (exists with no bounds)\\
+             \> $~|$  \> $\exists k_n \le rate.\ C$             \> (exists with upper bound)    \\
 \\
@@
 
-\newcommand{\constraint}[3]{#1 \vDash #2 : #3}
-\newcommand{\inferC}[2]{
-    $$\begin{array}{c}
-        \infer{#1}{#2}
-    \end{array}$$}
-\newcommand{\ruleIC}[3]{
-    \inferC
-        {\constraint{#1; binds}{#2 C}{#3 ~ T}}
-        {\constraint{binds}{C}{T}}}
-
-\[ \constraint{binds}{constraint}{(\tau_{name} \Rightarrow \tau)} \]
-
 Finding the relative sizes of arrays is very similar to type inference.
@@
 
-\vspace{1em}
+$$\begin{array}{lcrcl}
+&&\constr{\_} & :: & binds \rightarrow C \\
+\\
+% MMMMMMMMMMMMMMMMMM \= MM \= \kill
+% $\constr{\lambda scalars~arrays \to @let @binds@ in @x}$ \> $=$ \> $\forall_{a \in arrays} k_a. \constr{binds}$ \\
 
-\inferC{\constraint{\lambda scalars~arrays \to @let @ binds}{\forall_{a \in arrays} k_a. C}{T}}
-       {\constraint{binds}{C}{T}}
+\constr{o &=& @fold @f@ @n}       &  = &  \hole \\
+\constr{o &=& @map@_n~f~ns}       &  = &  \forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
+\constr{o &=& @filter@~f~n}       &  = &  \exists k_o \le k_n.\ \hole \\
+\constr{o &=& @gather@~i~d}       &  = &  \forall k_o.\ k_o = k_i \wedge \hole \\
+\constr{o &=& @cross@~a~b}        &  = &  \forall k_o.\ k_o = k_a \times k_b \wedge \hole \\
+\constr{outs &=& @external@~ins}  &  = &  \exists_{o \in outs} k_o.\ \hole \\
+\\
+&&\constr{b;~bs}  &  = &  \fillhole{\constr{b}}{\constr{bs}}       \\
+&&\constr{nil}    &  = &  true                          \\
+\end{array}$$
 
-\vspace{1em}
-
-\ruleIC{o = @fold@~f~n}{}{\tau_o = k_n, }
-
-\vspace{1em}
-
-\ruleIC{o = @map@_n~f~ns}{\forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge}{\tau_o = k_o, }
-
-\vspace{1em}
-
-\ruleIC{o = @filter@~f~n}{\exists k_o \le k_n.\ }{\tau_o = k_n, }
+\begin{tabbing}
+MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
+$\tau$  \>$::$\> $binds \rightarrow name \rightarrow \tau$ \\
+\\
+$\tau_{bs,o}$    
+            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $k_n$ \\
+            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $k_o$ \\
+            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $k_n$ \\
+            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $k_i$ \\
+            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $k_a \times k_b$ \\
+            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $@external@$ \\
+\end{tabbing}
 
@@
 This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.
-
-\vspace{1em}
-
-\ruleIC{o = @gather@~i~d}{\forall k_o.\ k_o = k_i \wedge}{\tau_o = k_i, }
-
-\vspace{1em}
-
-\ruleIC{o = @cross@~a~b}{\forall k_o.\ k_o = k_a \times k_b \wedge C}{\tau_o = k_a \times k_b, }
-
-\vspace{1em}
-
-\ruleIC{outs = @external@~ins}{\exists_{o \in outs} k_o.\ }{\tau_{outs} = @external@, }
-
-\vspace{1em}
 

### 2014-04-20 2014betterfusionforfilters/section/03-SizeInference.tex
oops
size inf

@@
 
+Our formulation of size inference is very similar to type inference, a la \CITE.
+First, constraints are generated for the bindings, such as equality among two rates, the conjunction of two constraints, and existentials.
+If these constraints can be solved, the equality constraints are used to group the rates into equivalence classes.
+Otherwise if the constraints are unable to be solved, the program cannot be assured to require no runtime size checks and will not be fused by our system.
+
 \newcommand{\constr}[1]{\llbracket #1 \rrbracket}
 \newcommand{\hole}[0]{[]}
-\newcommand{\fillhole}[2]{#1\Left[#2\right]}
+\newcommand{\fillhole}[2]{#1\left[#2\right]}
+
+Rates may either be some rate variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
 
@@
 %            \> $~|$  \> $External_n$        \> (output of @external@s) \\
-\\
+\end{tabbing}
+
+The most important constraint is equality on two rates.
+Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
+Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
+Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every rate.
+Existential quantifiers simply state that there must exist a rate that satisfies the constraint.
+
+\begin{tabbing}
+MMMMM       \= MM \= MMMMMMMMMMM \= \kill
 $C$          \> @::=@ \> $true$                                 \> (trivially true) \\
@@
              \> $~|$  \> $\forall k_n.\ C$                      \> (forall quantification) \\
-             \> $~|$  \> $\exists k_n.\ C$                      \> (exists with no bounds)\\
-             \> $~|$  \> $\exists k_n \le rate.\ C$             \> (exists with upper bound)    \\
-\\
+             \> $~|$  \> $\forall k_n \le rate.\ C$             \> (forall with upper bound) \\
+             \> $~|$  \> $\exists k_n.\ C$                      \> (existential quantification)\\
+%             \> $~|$  \> $\exists k_n \le rate.\ C$             \> (exists with upper bound)    \\
+\end{tabbing}
+
+After the constraints are solved, each combinator is assigned a $rate$ as an iteration size -- the size of the loop required to generate the output.
+It is important to note that for filters, the iteration size is different to the output size.
+The iteration sizes, $tau_n$, are used to check whether two loops may be fused.
+Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
+External computations are treated separately, as they cannot be fused with any other nodes.
+
+\begin{tabbing}
+MMMMM       \= MM \= MMMMMMMMMMM \= \kill
 $\tau$       \> @::=@ \> $rate$                                  \> (loop size of some rate) \\
@@
 
-Finding the relative sizes of arrays is very similar to type inference.
-A rate can be either some rate variable, a cross product of two rates, or an unknown, but fixed, filtered or external output.
-For each binding, a set of equality constraints are created.
+Constraints are generated for a list of bindings.
+As folds produce no constraints, they simply return a hole to be filled later.
+Maps require their inputs and output to all have the same rate, so if there exists some rate that is equal to all its inputs, then that is the rate of the output.
+As the output rate of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible rates less than the upper bound, the subconstraint must be satisfiable.
+Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all rates.
+
+The final clauses show how the constraint holes are filled by constraints of subsequent bindings.
+If there no bindings, the constraint is trivially true.
 
@@
 \constr{o &=& @fold @f@ @n}       &  = &  \hole \\
-\constr{o &=& @map@_n~f~ns}       &  = &  \forall k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
-\constr{o &=& @filter@~f~n}       &  = &  \exists k_o \le k_n.\ \hole \\
-\constr{o &=& @gather@~i~d}       &  = &  \forall k_o.\ k_o = k_i \wedge \hole \\
-\constr{o &=& @cross@~a~b}        &  = &  \forall k_o.\ k_o = k_a \times k_b \wedge \hole \\
-\constr{outs &=& @external@~ins}  &  = &  \exists_{o \in outs} k_o.\ \hole \\
+\constr{o &=& @map@_n~f~ns}       &  = &  \exists k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
+\constr{o &=& @filter@~f~n}       &  = &  \forall k_o \le k_n.\ \hole \\
+\constr{o &=& @gather@~i~d}       &  = &  \exists k_o.\ k_o = k_i \wedge \hole \\
+\constr{o &=& @cross@~a~b}        &  = &  \exists k_o.\ k_o = k_a \times k_b \wedge \hole \\
+\constr{outs &=& @external@~ins}  &  = &  \forall_{o \in outs} k_o.\ \hole \\
 \\
@@
 \end{array}$$
+
+As an example of constraint generation, here is the @normalize2@ program from earlier.
+\begin{tabbing}
+@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+@normalize2 us@                     \> $\exists k_{us}.$      \\
+@ = let sum1 = fold   (+) 0 us@     \>                      \\
+@       gts  = filter (>0)  us@     \> $\forall k_{gts} \le k_{us}.$ \\
+@       sum2 = fold   (+) 0 gts@    \> \\
+@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
+@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
+@   in (nor1, nor2)@                \> $true$ \\
+\end{tabbing}
+The entire constraint ends up being $\exists k_{us}.\ \forall k_{gts} \le k_{us}.\ \exists k_{nor1}.\ k_{nor1} = k_{us} \wedge \exists k_{nor2}.\ k_{nor2} = k_{us} \wedge true$.
+
+Once the constraints are solved, known to be satisfiable, and sorted into equivalence classes, each combinator is assigned a rate.
+Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
+This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.
 
@@
 
-Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
-This is because, in order to produce the filtered output, all elements of the input $n$ must be considered.
-
 Solving the generated equality constraints allows each rate to be grouped into equivalence classes.
@@
 These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.
-
-\begin{code}
- normalize2 :: Array Int -> Array Int
- normalize2 us                   {us   = k_us}
-  = let sum1 = fold   (+) 0 us   
-        gts  = filter (>0)  us   {gts  = Filtered_gts}
-        sum2 = fold   (+) 0 gts   
-        nor1 = map  (/sum1) us   {nor1 = k_us}
-        nor2 = map  (/sum2) us   {nor2 = k_us}
-    in (nor1, nor2)
-\end{code}
 

### 2014-04-21 2014betterfusionforfilters/section/06-Related.tex
Add notes and move sections around
Cite papers on the polyhedral model

--- 2014-04-01 2014betterfusionforfilters/section/05-Related.tex
@@
-f\section{Related work}
+%!TEX root = ../Main.tex
+\section{Related work}
 
@@
 
-\TODO{I don't really have a story on polyhedral analysis or unimodular.}
-Now, here's my understanding of unimodular (and I'm just guessing this applies to polyhedral too):
-if the dependency matrix for a loop nest or set of loops forms a \emph{unimodular matrix} (an integral matrix whose inverse is also integral), only then can it be dealt with by unimodular transforms.
+The polyhedral model is an algebraic representation of imperative loop nests and transformations on them, including fusion transformations.
+Polyhedral systems \cite{pouchet2011polyhedral} are able to express \emph{all possible} distinct loop transformations where the array indices, conditionals and loop bounds are affine functions of the surrounding loop indices.
+However, the polyhedral model is not applicable to (or intended for) one dimentional filter-like operations where the size of the result array depends on the source data.
+Recent work extends the polyhedral model to support arbitrary indexing~\cite{venkat2014polyhedral}, as well as conditional control flow that is predicated on arbitrary (ie, non-affine) functions of the loop indices~\cite{benabderrahmane2010polyhedral}.
+However, the indices used to write into the destination array must still be computed with affine functions.
+
+\TODO{Is unimodular the same as the polyhedral model?. 
+Now, here's my understanding of unimodular (and I'm just guessing this applies to polyhedral too): if the dependency matrix for a loop nest or set of loops forms a \emph{unimodular matrix} (an integral matrix whose inverse is also integral), only then can it be dealt with by unimodular transforms.}

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
reorg size inf

@@
 Otherwise if the constraints are unable to be solved, the program cannot be assured to require no runtime size checks and will not be fused by our system.
+After the constraints are generated and solved, each combinator is given an iteration size.
+Two loops with the same iteration size can be fused together.
 
@@
 
+\subsection{Constraint generation}
 Rates may either be some rate variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
@@
 
-After the constraints are solved, each combinator is assigned a $rate$ as an iteration size -- the size of the loop required to generate the output.
-It is important to note that for filters, the iteration size is different to the output size.
-The iteration sizes, $tau_n$, are used to check whether two loops may be fused.
-Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
-External computations are treated separately, as they cannot be fused with any other nodes.
-
-\begin{tabbing}
-MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$\tau$       \> @::=@ \> $rate$                                  \> (loop size of some rate) \\
-             \> $~|$  \> @external@                              \> (external and unfusible) \\
-\end{tabbing}
-
 Constraints are generated for a list of bindings.
@@
 
+\subsubsection{Examples}
+
 As an example of constraint generation, here is the @normalize2@ program from earlier.
@@
 \end{tabbing}
-The entire constraint ends up being $\exists k_{us}.\ \forall k_{gts} \le k_{us}.\ \exists k_{nor1}.\ k_{nor1} = k_{us} \wedge \exists k_{nor2}.\ k_{nor2} = k_{us} \wedge true$.
+This constraint is satisfiable, with the equivalence classes being:
+\newcommand{\eqclasses}[1]{
+    \begin{tabbing}
+        MM \= M \= \kill
+        #1
+    \end{tabbing}}
+\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
+\eqclasses{
+    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
+    \eqclass{k_{gts}}{k_{gts}}
+}
+
+The next example involves two filters using the same predicate.
+Despite using the same predicate and input data, we produce different output rates for each filter.
+\begin{tabbing}
+@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+@diff xs@                           \> $\exists k_{xs}.$ \\
+@ = let ys1 = filter p xs@          \> $\forall k_{ys1} \le k_{xs}.$       \\
+@       ys2 = filter p xs@          \> $\forall k_{ys2} \le k_{xs}.$       \\
+@   in (ys1, ys2)@                  \> $true$                   \\
+\end{tabbing}
+
+This constraint is satisfiable, with the equivalence classes being:
+\eqclasses{
+    \eqclass{k_{xs}}    {k_{xs}}
+    \eqclass{k_{ys1}}   {k_{ys1}}
+    \eqclass{k_{ys2}}   {k_{ys2}}
+}
+
+This example is disallowed, as it would require a runtime check on the size of the @flt@ array.
+\begin{tabbing}
+@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+@bad1 vs@                           \> $\exists k_{vs}.$ \\
+@ = let flt   = filter p vs@        \> $\forall k_{flt} \le k_{vs}.$ \\
+@       wrong = map2   f flt vs@    \> $\exists k_{wrong}.\ k_{wrong} = k_{flt}$ \\
+                                    \> $\wedge k_{wrong} = k_{vs} \wedge$ \\
+@   in  wrong@                      \> $true$
+\end{tabbing}
+This constraint is unsatisfiable, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
+Obviously, there exists no sole size such that all sizes are equal to it.
+As a result, no equivalence classes are generated for this example, and no fusion is performed.
+
+Similarly, this example:
+\begin{tabbing}
+@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+@bad2 vs@                           \> $\exists k_{vs}.$ \\
+@ = let flt  = filter p  vs@        \> $\forall k_{flt}.$ \\
+@       flt2 = filter p' vs@        \> $\forall k_{flt2}.$ \\
+@       mix  = map2   f  flt flt2@  \> $\exists k_{mix}.\ k_{mix} = k_{flt}$ \\
+                                    \> $\wedge k_{mix} = k_{flt2} \wedge$ \\
+@   in  mix@                        \> $true$                               \\
+\end{tabbing}
+Again, transitivity here requires that $Filtered_{flt} = Filtered_{flt2}$, but since these are both concrete constructors, we cannot have such a constraint.
+Basically, we are constraining two different filtered outputs to be the same size.
+Operationally, implementing this would require to buffer both @flt@ and @flt2@, before checking the two lengths.
+
+\subsection{Iteration size}
+After the constraints are solved, each combinator is assigned a $rate$ as an iteration size -- the size of the loop required to generate the output.
+It is important to note that for filters, the iteration size is different to the output size.
+The iteration sizes, $tau_n$, are used to check whether two loops may be fused.
+Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
+External computations are treated separately, as they cannot be fused with any other nodes.
+
+\begin{tabbing}
+MMMMM       \= MM \= MMMMMMMMMMM \= \kill
+$\tau$       \> @::=@ \> $rate$                                  \> (loop size of some rate) \\
+             \> $~|$  \> @external@                              \> (external and unfusible) \\
+\end{tabbing}
 
@@
 
-\subsubsection{Examples}
-Following are a few examples of disallowed programs.
-Consider the constraints generated by this program:
-\begin{code}
-test vs    =
- let flt   = filter p vs        { flt   = Filtered_flt }
-     wrong = map2   f flt vs    { wrong = flt = vs     }
- in  wrong
-\end{code}
-By transitivity, we find that $vs = Filtered_{flt}$, but $vs$ is bound before @flt@.
-Operationally, actually implementing this would require to buffer the entire contents of @flt@, then check the length of @flt@ against the length of @vs@.
-
-Similarly, this example:
-\begin{code}
-test vs    =
- let flt   = filter p  vs       { flt   = Filtered_flt  }
-     flt2  = filter p' vs       { flt2  = Filtered_flt2 }
-     wrong = map2   f  flt flt2 { wrong = flt = flt2    }
- in  wrong
-\end{code}
-Again, transitivity here requires that $Filtered_{flt} = Filtered_{flt2}$, but since these are both concrete constructors, we cannot have such a constraint.
-Basically, we are constraining two different filtered outputs to be the same size.
-Operationally, implementing this would require to buffer both @flt@ and @flt2@, before checking the two lengths.
-
 \subsection{Generators}

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
remove forall with upper bound from size inference; it actually made unsatisfiable constraints satisfiable, *and* was more complicated

@@
              \> $~|$  \> $\forall k_n.\ C$                      \> (forall quantification) \\
-             \> $~|$  \> $\forall k_n \le rate.\ C$             \> (forall with upper bound) \\
              \> $~|$  \> $\exists k_n.\ C$                      \> (existential quantification)\\
@@
 Maps require their inputs and output to all have the same rate, so if there exists some rate that is equal to all its inputs, then that is the rate of the output.
-As the output rate of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible rates less than the upper bound, the subconstraint must be satisfiable.
+As the output rate of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible rates, the subconstraint must be satisfiable.
 Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all rates.
@@
 \constr{o &=& @map@_n~f~ns}       &  = &  \exists k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
-\constr{o &=& @filter@~f~n}       &  = &  \forall k_o \le k_n.\ \hole \\
+\constr{o &=& @filter@~f~n}       &  = &  \forall k_o.\ \hole \\
 \constr{o &=& @gather@~i~d}       &  = &  \exists k_o.\ k_o = k_i \wedge \hole \\
@@
 @ = let sum1 = fold   (+) 0 us@     \>                      \\
-@       gts  = filter (>0)  us@     \> $\forall k_{gts} \le k_{us}.$ \\
+@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
 @       sum2 = fold   (+) 0 gts@    \> \\
@@
 @diff xs@                           \> $\exists k_{xs}.$ \\
-@ = let ys1 = filter p xs@          \> $\forall k_{ys1} \le k_{xs}.$       \\
-@       ys2 = filter p xs@          \> $\forall k_{ys2} \le k_{xs}.$       \\
+@ = let ys1 = filter p xs@          \> $\forall k_{ys1}.$       \\
+@       ys2 = filter p xs@          \> $\forall k_{ys2}.$       \\
 @   in (ys1, ys2)@                  \> $true$                   \\
@@
 @bad1 vs@                           \> $\exists k_{vs}.$ \\
-@ = let flt   = filter p vs@        \> $\forall k_{flt} \le k_{vs}.$ \\
+@ = let flt   = filter p vs@        \> $\forall k_{flt}.$ \\
 @       wrong = map2   f flt vs@    \> $\exists k_{wrong}.\ k_{wrong} = k_{flt}$ \\

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
size inf

@@
 
-Similarly, this example:
+Similarly, this example is not allowed, as it would require a runtime check to determine that the output of two filters @flt@ and @flt2@ are the same size.
 \begin{tabbing}
@@
 \end{tabbing}
-Again, transitivity here requires that $Filtered_{flt} = Filtered_{flt2}$, but since these are both concrete constructors, we cannot have such a constraint.
-Basically, we are constraining two different filtered outputs to be the same size.
-Operationally, implementing this would require to buffer both @flt@ and @flt2@, before checking the two lengths.
+Again, by transitivity of equality this constraint can be simplified to $\forall k_{flt}.\ \forall k_{flt2}.\ k_{flt} = k_{flt2}$.
+It is easy to see that this constraint is unsatisfiable.
 

### 2014-04-22 2014betterfusionforfilters/code/fortran/dumb.f90
normalise2 in fortran. surprisingly, intel fortran doesn't do any fusion in here. it does fuse dumb, though.

@@ -8,11 +8,11 @@
      s2 = 0
 
      do i1 = 1, N
-         s1 = s1 + A(i1)
+         s1 = s1 + A(i1) * 1111
      enddo
     
      do i2 = 1, N
-         s2 = s2 - A(i2)
+         s2 = s2 - A(i2) * 2222
      enddo
 
      ou = s1 * s2

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
size inf

@@
 After the constraints are solved, each combinator is assigned a $rate$ as an iteration size -- the size of the loop required to generate the output.
-It is important to note that for filters, the iteration size is different to the output size.
-The iteration sizes, $tau_n$, are used to check whether two loops may be fused.
+It is important to note that for filters, the iteration size is not the output size, but is instead the size of the input.
+The output size of a filter is, however, a \emph{subsize} of the input's size, as not only is it known to be less than or equal to its input in size, it is also generated depending on the input.
+The iteration sizes, $\tau_n$, are used to check whether two loops may be fused.
 Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
+The difference from previous work is that loops of different iteration sizes \emph{can} be fused, if one is a subsize of the other, and the subsize's \emph{generator} is fused together it as well.
+Basically, operations on filtered data can be fused with operations on the original data, if it is fused with the filter as well.
 External computations are treated separately, as they cannot be fused with any other nodes.

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
Refactor rules to generate constraints  Rename rate -> size consistently.  Use u^k as a size variable generator.  Make the main interpretation function take a list of bindings directly.  Add an interpretation for the top level function form.

@@
 \section{Size inference}
-
 Knowing the relative sizes of loops is important for fusion.
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Constraint generation}
-Rates may either be some rate variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
-
+Sizes may either be some variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
 \begin{tabbing}
-MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$rate$      \> @::=@ \> $k_n$               \> (rate variable)\\
-            \> $~|$  \> $rate \times rate$  \> (cross product) \\
-%            \> $~|$  \> $Filtered_n$        \> (output of @filter@s) \\
-%            \> $~|$  \> $External_n$        \> (output of @external@s) \\
+MMMMM   \= MM \= MMMMMMMMM \= \kill
+$size$  \> @::=@ \> $k_n$               \> (size variable) \\
+        \> $~|$  \> $size \times size$  \> (cross product)
 \end{tabbing}
 
-The most important constraint is equality on two rates.
+The most important constraint is equality on two sizes.
 Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
@@
 Existential quantifiers simply state that there must exist a rate that satisfies the constraint.
-
 \begin{tabbing}
-MMMMM       \= MM \= MMMMMMMMMMM \= \kill
+MMMMM   \= MM \= MMMMMMMMM \= \kill
 $C$     \> @::=@ \> $true$              \> (trivially true)             \\
-             \> $~|$  \> $\hole$                                \> (constraint hole) \\
-             \> $~|$  \> $rate = rate$                          \> (equality constraint) \\
+%       \> $~|$  \> $\hole$             \> (constraint hole)            \\
+        \> $~|$  \> $size = size$       \> (equality constraint)        \\
         \> $~|$  \> $C \wedge C$        \> (conjunction)                \\
         \> $~|$  \> $\forall k_n.\ C$   \> (forall quantification)      \\
-             \> $~|$  \> $\exists k_n.\ C$                      \> (existential quantification)\\
-%             \> $~|$  \> $\exists k_n \le rate.\ C$             \> (exists with upper bound)    \\
+        \> $~|$  \> $\exists k_n.\ C$   \> (existential quantification)
 \end{tabbing}
@@
 
-The final clauses show how the constraint holes are filled by constraints of subsequent bindings.
-If there no bindings, the constraint is trivially true.
+The final clauses show how the constraint holes are filled by constraints of subsequent bindings. If there no bindings, the constraint is trivially true.
 
 $$\begin{array}{lcrcl}
-&&\constr{\_} & :: & binds \rightarrow C \\
-\\
-% MMMMMMMMMMMMMMMMMM \= MM \= \kill
-% $\constr{\lambda scalars~arrays \to @let @binds@ in @x}$ \> $=$ \> $\forall_{a \in arrays} k_a. \constr{binds}$ \\
+&&\constr{\_} & :: & \textrm{function} \rightarrow C
+\\[1ex]
 
-\constr{o &=& @fold @f@ @n}       &  = &  \hole \\
-\constr{o &=& @map@_n~f~ns}       &  = &  \exists k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
-\constr{o &=& @filter@~f~n}       &  = &  \forall k_o.\ \hole \\
-\constr{o &=& @gather@~i~d}       &  = &  \exists k_o.\ k_o = k_i \wedge \hole \\
-\constr{o &=& @cross@~a~b}        &  = &  \exists k_o.\ k_o = k_a \times k_b \wedge \hole \\
-\constr{outs &=& @external@~ins}  &  = &  \forall_{o \in outs} k_o.\ \hole \\
-\\
-&&\constr{b;~bs}  &  = &  \fillhole{\constr{b}}{\constr{bs}}       \\
-&&\constr{nil}    &  = &  true                          \\
+\multicolumn{3}{l}{
+    \constr{
+       f~ \{ x_i \}~ \{ u_j \}^{\; j \gets 1..m}~ 
+         =  \textrm{let}~ bs~ \textrm{in}~ exp
+    }
+}
+        & = & \{ \exists u_j^k. \}^{j \gets 1..m}
+           ~\constr{bs}
+
+\\[2ex]
+
+&&\constr{\_} & :: & \overline{bind} \rightarrow C 
+\\[1ex]
+
+&&\constr{~ \cdot ~}    
+        & = &   true                          
+\\[0.5ex]
+
+\constr{u_o &=& @fold@~ f~ u_1;~ bs}
+        & = &   \constr{bs}
+\\[0.5ex]
+
+\constr{u_o &=& @map@_n~ f~ \{u_i\}^{\;i \gets 1..n};~ bs }       
+        & = &   \exists u_o^k.\ \bigwedge_{i \gets 1..n}\{u_o^k = u_i^k\} 
+                \wedge \constr{bs}
+\\[0.5ex]
+
+\constr{u_o &=& @filter@~ f~ u_1;~ bs}
+        & = &   \forall u_o^k.\ \constr{bs}
+\\[0.5ex]
+
+\constr{u_o &=& @gather@~ u_{ix}~ u_{data};~ bs}
+        & = &   \exists u_o^k.\ u_o^k = u_{ix}^k
+                ~\wedge~ \constr{bs}
+\\[0.5ex]
+
+\constr{u_o &=& @cross@~ u_a~ u_b;~ bs}
+        & = &   \exists u_o^k.\ u_o^k = u_a^k \times u_b^k 
+                ~\wedge~ \constr{bs}
+\\[0.5ex]
+
+\multicolumn{5}{l}{
+\constr{x,~ \{u_i\}^{i \gets 1..n} = @external@~ \{v_j\}^{j \gets 1..m};~ bs}
+          =     \{ \forall u_i^k. \}^{\; i \gets 1..n}
+                ~\constr{bs}
+}
+\\[0.5ex]
 \end{array}$$
 
+% -----------------------------------------------------------------------------
 \subsubsection{Examples}
-
 As an example of constraint generation, here is the @normalize2@ program from earlier.

### 2014-04-23 2014betterfusionforfilters/section/03-SizeInference.tex
reorganise size inference section; remove runtime checks subsec; clean up

@@
 Our formulation of size inference is very similar to type inference, a la \CITE.
-First, constraints are generated for the bindings, such as equality among two rates, the conjunction of two constraints, and existentials.
-If these constraints can be solved, the equality constraints are used to group the rates into equivalence classes.
+First, constraints are generated for the bindings, such as equality among two sizes, the conjunction of two constraints, and existentials.
+If these constraints can be solved, the equality constraints are used to group the sizes into equivalence classes.
 Otherwise if the constraints are unable to be solved, the program cannot be assured to require no runtime size checks and will not be fused by our system.
@@
 \subsection{Constraint generation}
-Sizes may either be some variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
+Sizes may either be some variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two sizes.
 \begin{tabbing}
@@
 Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
-Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every rate.
-Existential quantifiers simply state that there must exist a rate that satisfies the constraint.
+Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every size.
+Existential quantifiers simply state that there must exist a size that satisfies the constraint.
 \begin{tabbing}
@@
 As folds produce no constraints, they simply return a hole to be filled later.
-Maps require their inputs and output to all have the same rate, so if there exists some rate that is equal to all its inputs, then that is the rate of the output.
-As the output rate of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible rates, the subconstraint must be satisfiable.
-Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all rates.
+Maps require their inputs and output to all have the same size, so if there exists some size that is equal to all its inputs, then that is the size of the output.
+As the output size of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible sizes, the subconstraint must be satisfiable.
+Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all sizes.
 
@@
 The next example involves two filters using the same predicate.
-Despite using the same predicate and input data, we produce different output rates for each filter.
+Despite using the same predicate and input data, we produce different output sizes for each filter.
 \begin{tabbing}
@@
 \subsection{Iteration size}
-After the constraints are solved, each combinator is assigned a $rate$ as an iteration size -- the size of the loop required to generate the output.
+After the constraints are solved, each combinator is assigned a $size$ as an iteration size -- the number of iterations in the loop required to generate the output.
 It is important to note that for filters, the iteration size is not the output size, but is instead the size of the input.
@@
 MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$\tau$       \> @::=@ \> $rate$                                  \> (loop size of some rate) \\
+$\tau$       \> @::=@ \> $size$                                  \> (known size) \\
              \> $~|$  \> @external@                              \> (external and unfusible) \\
@@
 
-Once the constraints are solved, known to be satisfiable, and sorted into equivalence classes, each combinator is assigned a rate.
+Once the constraints are solved, known to be satisfiable, and sorted into equivalence classes, each combinator is assigned a size.
 Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
@@
 
-Solving the generated equality constraints allows each rate to be grouped into equivalence classes.
-All combinators whose constraints are in the same equivalence class have the same iteration size, and may be fused together.
-These constraints are also used to rule out any programs that require runtime checks on array lengths during the program.
+After the generated equality constraints are solved, and sizes are grouped into equivalence classes, the combinators with iteration sizes in the same equivalence class can be fused together.
 
-\subsection{Runtime checks on array lengths}
-The $@map@_n$ combinator requires all its input arrays to be the same size.
-We do not, however, want to perform runtime checks on the size of arrays before every combinator.
-We have decided to only fuse programs that can be statically determined to require only array length checks at the start of execution.
-
-\TODO{We assign different rates to both @ys1@ and @ys2@ in this example.}
-\begin{code}
- diff xs = let ys1 = filter p xs
-               ys2 = filter p xs
-           in  (ys1, ys2)
-\end{code}
-
-% \ben{This point is only relavant in the context of our specific implementation, which is no the focus of this paper. }
-% Any other programs are simply left alone by our fusion system, potentially falling back to stream fusion.
-
-We use the constraints to rule out any programs that require runtime checks on array lengths during the program.
-Firstly, filter and external constraints are only allowed to be mentioned after their binding site, as it is only at that stage that the lengths are actually known.
-
-\ben{Refactor presentation to use the actual $\exists k. $ quantifier, as in ``The essence of ML type inference'', from ATAPL}
-
-Secondly, it is not possible to constrain one constructor to equal another: $Filtered_n = External_n$ is invalid.
-Finally, any initial equality constraints, such as two input arrays being the same length, or one input array being the length of the cross product of two others, are dealt with by inserting runtime checks only at the start of the function.
-
-The constraints $Filtered_n$, $External_n$ are treated as \emph{existential} constructors, bound at $n$. This means that such constraints are not able to be referred to before their bindings.
-Intuitively, we know that the output of a @filter@ or an @external@ has \emph{some} length, but we don't know what its value is until after they have executed.
-Thus, constraining the input of something to be some exact length, before we know what the length is, is impossible.
-
-Similarly, these constructors cannot be constrained to equal any other constructor. 
-For example, $Filtered_n = External_n$ is invalid, as is $a \times b = Filtered_n$.
-This corresponds to each equivalence class having \emph{at most} one distinct constructor associated with it.
-Any other elements of the equivalence class must be either the same constructor, or simply rate variables.
-
-\subsection{Generators}
+\subsection{Subsizes and generators}
+Unlike previous work, we do allow combinators with different iteration sizes to be fused together.
+For example, an operation on filtered data may be fused with the filter operation that generates the data, even though the iteration sizes are different.
 

### 2014-05-02 2014betterfusionforfilters/section/03-SizeInference.tex
constraint validity. herbrand unification

@@
 
+\subsection{Constraint validity}
+Checking whether a constraint is valid can be done using Herbrand unification, similar to HM(X)~\cite{odersky1999type}.
+The difference, however, is that our constraints include both universal and existential quantifiers, whereas HM(X) only includes existential quantifiers.
+Herbrand unification is still applicable on constraints with both quantifiers~\cite{buss1995herbrand}.
+This is achieved by performing \emph{Herbrandization}, the dual of Skolemization, to remove the universal quantifiers and replace them with Herbrand functions.
+After the universal quantifiers have been removed, the constraint can easily be put into prenex form, and validated using the same unification method as HM(X).
+
+For example, a simple, but invalid constraint such as:
+\[ \exists x.\ \forall y.\ x = y \]
+will have its universal quantifier removed, and all references to $y$ changed to $f_y(x)$, with $x$ as an argument to $f$ as $x$ is in scope of the quantifier.
+The result being:
+\[ \exists x.\ x = f_y(x) \]
+This resulting constraint, while satisfiable for certain values of $f_y$, is not satisfiable for all values of $f_y$, and so is not valid.
+
+After the constraint has been validated, equivalence classes are easily constructed.
+The transitive closure of the equality constraints, ignoring quantifiers, is generated.
+As there are no disjunction, negation or implication all equality constraints will be applicable, regardless of the quantifiers used.
+
 \subsection{Iteration size}
-After the constraints are solved, each combinator is assigned a $size$ as an iteration size -- the number of iterations in the loop required to generate the output.
+After the constraints are validated and equivalence classes generated, each combinator is assigned a $size$ as an iteration size -- the number of iterations in the loop required to generate the output.
 It is important to note that for filters, the iteration size is not the output size, but is instead the size of the input.

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
fix citation
size inf

@@
 
-Our formulation of size inference is very similar to type inference, a la \CITE.
+Our formulation of size inference is very similar to type inference, particularly the exposition of HM(X)~\cite{odersky1999type}.
 First, constraints are generated for the bindings, such as equality among two sizes, the conjunction of two constraints, and existentials.
@@
 Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
-Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every size.
+Forall quantifiers mean the constraint must be satisfiable for every size.
 Existential quantifiers simply state that there must exist a size that satisfies the constraint.

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
size inf: constraints: change 'satisfiable' to 'valid'

@@
 Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
-Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
+Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be valid.
 Forall quantifiers mean the constraint must be satisfiable for every size.
@@
 \end{tabbing}
-This constraint is satisfiable, with the equivalence classes being:
+This constraint is valid, with the equivalence classes being:
 \newcommand{\eqclasses}[1]{
@@
 
-This constraint is satisfiable, with the equivalence classes being:
+This constraint is valid, with the equivalence classes being:
 \eqclasses{
@@
 \end{tabbing}
-This constraint is unsatisfiable, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
+This constraint is invalid, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
 Obviously, there exists no sole size such that all sizes are equal to it.
@@
 Again, by transitivity of equality this constraint can be simplified to $\forall k_{flt}.\ \forall k_{flt2}.\ k_{flt} = k_{flt2}$.
-It is easy to see that this constraint is unsatisfiable.
+It is easy to see that this constraint is invalid.
 
@@
 
-Once the constraints are solved, known to be satisfiable, and sorted into equivalence classes, each combinator is assigned a size.
+Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size.
 Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
size inf: constraints: remove ref to holes

@@
 \newcommand{\constr}[1]{\llbracket #1 \rrbracket}
-\newcommand{\hole}[0]{[]}
-\newcommand{\fillhole}[2]{#1\left[#2\right]}
 
@@
 The most important constraint is equality on two sizes.
-Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
 Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be valid.
@@
 $C$     \> @::=@ \> $true$              \> (trivially true)             \\
-%       \> $~|$  \> $\hole$             \> (constraint hole)            \\
         \> $~|$  \> $size = size$       \> (equality constraint)        \\
@@
 Constraints are generated for a list of bindings.
-As folds produce no constraints, they simply return a hole to be filled later.
+As folds produce no constraints, they simply return the constraints of the remaining bindings.
 Maps require their inputs and output to all have the same size, so if there exists some size that is equal to all its inputs, then that is the size of the output.

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
Shift size inference constraints to new file
Another go at size inference constraint generation
add generate to constrs

@@
 
-The final clauses show how the constraint holes are filled by constraints of subsequent bindings. If there no bindings, the constraint is trivially true.
+The final clauses show how the constraint holes are filled by constraints of subsequent bindings. If there are no bindings, the constraint is trivially true.
 
@@
 \constr{u_o &=& @filter@~ f~ u_1;~ bs}
+        & = &   \forall u_o^k.\ \constr{bs}
+\\[0.5ex]
+
+\constr{u_o &=& @generate@~ s~ f;~ bs}
         & = &   \forall u_o^k.\ \constr{bs}
@@
 MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$\tau$       \> @::=@ \> $size$                                  \> (known size) \\
+$T$          \> @::=@ \> $size$                                  \> (known size) \\
              \> $~|$  \> @external@                              \> (external and unfusible) \\
@@
 MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
-$\tau$  \>$::$\> $binds \rightarrow name \rightarrow \tau$ \\
+$\tau$  \>$::$\> $binds \rightarrow name \rightarrow T$ \\
 \\
@@
             \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $k_n$ \\
+            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $k_o$ \\
             \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $k_i$ \\
@@
 \begin{tabbing}
-MMM       \= MM \= MMMMMM \= MM \= MMM \= MM \= \kill
-$gen$   \> @::@  \> $\{constraint\}$  \> $\to$ \> $array$ \> $\to$ \> $\{array\}$ \\
-MMMMMMM                 \= M  \= MMMMMMMM \= MM \= \kill
-$gen(constrs, a)$ \> $|$ \> $a \le_c b \in constrs$ \> $=$ \> $\{b\}$                        \\
-$gen(constrs, a)$ \> $|$ \> $a =_c b \in constrs$   \> $=$ \> $\bigcup gen(constrs, b)$                        \\
-$gen(constrs, a)$ \> $|$ \> $otherwise$             \> $=$ \> $\emptyset$                        \\
+MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
+$gen$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
+$gen(bs,o)$    \\
+            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $\emptyset$ \\
+            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $\bigcup_{n \in ns} gen(bs, n)$ \\
+            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $\{o\}$       \\
+            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
+            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $gen(bs,i)$ \\
+            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $\emptyset$ \\
+            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
 \end{tabbing}
@@
 %\begin{lemma}
-\textbf{Lemma: unique generation of filters}
-For some bindings $bs$ and associated constraints $cs$, then
+\textbf{Lemma: unique generators}
+For some bindings $bs$, then
 \[
-valid(cs) \implies \forall x. (\exists y \in bs. gen(cs, x) = \{y\}) \vee gen(cs, x) = \emptyset
+valid(\constr{bs}) \implies \forall x.\ |gen(bs, x)| \le 1
 \]

### 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
Add sketchy notes about new size inference rules

@@
 
-$$\begin{array}{lcrcl}
-&&\constr{\_} & :: & \textrm{function} \rightarrow C
-\\[1ex]
+\begin{figure*}
+\input{figures/Size-env.tex}
+\end{figure*}
 
-\multicolumn{3}{l}{
-    \constr{
-       f~ \{ x_i \}~ \{ u_j \}^{\; j \gets 1..m}~ 
-         =  \textrm{let}~ bs~ \textrm{in}~ exp
-    }
-}
-        & = & \{ \exists u_j^k. \}^{j \gets 1..m}
-           ~\constr{bs}
+\TODO{Notes about new constraint generation rules. These need to be fleshed out}
+\begin{itemize}
 
-\\[2ex]
+\item   There are three sorts of size variable: sizes of program variables, existential sizes, and unification variables.
+Sizes of program variables are named after the program variable (like $k_{xs}$), and the mapping appears in the environment like $xs : k_{xs}$.
+Existential variables are marked like $\exists k'$ in the environment.
+Unification variables like $k'$ appear in the environment with no $\exists$ qualifier.
 
-&&\constr{\_} & :: & \overline{bind} \rightarrow C 
-\\[1ex]
+\item   When generating contrains, we introduce fresh existential sizes, or unfication variables to constrain the program size variables appropriately.
+Program size variables always appear on the left of atomic equality constriants, while existential sizes and unification variables appear on the right.
 
-&&\constr{~ \cdot ~}    
-        & = &   true                          
-\\[0.5ex]
+\item   Generating the constraints like this makes the easy to solve.
+The constraint set is in solved form when each program size variable appears only once, and on the left of atomic equality constraints.
+Additionally, existentials and unification variables can appear many times, but only on the right of atomic constraints.
+To solve the constraints, keep finding pairs of atomic equality constraints where the same program size variable appears on the left, and perform an appropriate substitution into the whole constraint set.
 
-\constr{u_o &=& @fold@~ f~ u_1;~ bs}
-        & = &   \constr{bs}
-\\[0.5ex]
-
-\constr{u_o &=& @map@_n~ f~ \{u_i\}^{\;i \gets 1..n};~ bs }       
-        & = &   \exists u_o^k.\ \bigwedge_{i \gets 1..n}\{u_o^k = u_i^k\} 
-                \wedge \constr{bs}
-\\[0.5ex]
-
-\constr{u_o &=& @filter@~ f~ u_1;~ bs}
-        & = &   \forall u_o^k.\ \constr{bs}
-\\[0.5ex]
-
-\constr{u_o &=& @generate@~ s~ f;~ bs}
-        & = &   \forall u_o^k.\ \constr{bs}
-\\[0.5ex]
-
-\constr{u_o &=& @gather@~ u_{ix}~ u_{data};~ bs}
-        & = &   \exists u_o^k.\ u_o^k = u_{ix}^k
-                ~\wedge~ \constr{bs}
-\\[0.5ex]
-
-\constr{u_o &=& @cross@~ u_a~ u_b;~ bs}
-        & = &   \exists u_o^k.\ u_o^k = u_a^k \times u_b^k 
-                ~\wedge~ \constr{bs}
-\\[0.5ex]
-
-\multicolumn{5}{l}{
-\constr{x,~ \{u_i\}^{i \gets 1..n} = @external@~ \{v_j\}^{j \gets 1..m};~ bs}
-          =     \{ \forall u_i^k. \}^{\; i \gets 1..n}
-                ~\constr{bs}
-}
-\\[0.5ex]
-\end{array}$$
+\item   Once the constraints have been put into solved form, we use the generalisation rule to produce a type scheme for the whole graph.
+The type scheme contains both universal and existential quantifiers.
+The side conditions on the generalization rule reject the @bad1@ example: we must generalize the sizes of all input arrays, but the fact that an @vs@ in this example is given an existential size prevents this from happening.
+\end{itemize}
 

### 2014-05-05 2014betterfusionforfilters/section/02-CNF.tex
cnf: add todo for filter vs general size changing operators
Edits to Section 2

@@
 
-Once we have determined how the operators in a data flow graph should be clustered, we highlight materialized array variables by drawing them in boxes.
+Clusters of operators that are fused into single imperative loops are indicated by dotted lines, and we highlight materialized arrays by drawing them in boxes.
 In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program.
 However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused.
-
 In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
@@
 The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index.
-As such it is similar to Haskell's @zipWith@, except for the length restriction on the argument arrays.
+As such it is similar to Haskell's @zipWith@, with an added length restriction on the argument arrays.
 The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index.
@@
 The @external@ binding invokes a host library function that can produce and consume arrays, but not be fused with other combinators.
-All arrays passed to, and returned from host functions are full materialised.
+All arrays passed to, and returned from host functions are fully materialised.
 External bindings are explicit \emph{fusion barriers}, which force arrays and scalars to be fully computed before continuing.
 
-% When the function computed is the identity, they can also be used to force a particular intermediate array to be materialized, which provides the programmer fine-grained control over exactly what clustering is produced by our algorithm.
+Finally, note that @filter@ is only one representative size changing operator.
+We can handle more complex functions such as @unfold@ in our framework, but we stick with simple filtering to aid the discussion.
 

### 2014-05-05 2014betterfusionforfilters/section/03-SizeInference.tex
CNF: fix type of cross
Cut back intro to size inference

@@
 \section{Size inference}
-Knowing the relative sizes of loops is important for fusion.
-While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times.
-For simplicity, we do away with this added complexity and only support fusion of equal sized loops.
-\emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.
+Before performing fusion proper, we must infer the relative sizes of each array in the program.
+We achieve this with a simple constraint based inference algorithm, which we discuss in this section.
+Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}, but this system does not support size changing functions such as filter.
 
-It is important to emphasise that we are only interested in relative sizes of arrays; which arrays are equal, and which may be smaller.
-The $@map@_n$ combinators require all input arrays to be the same size, and their output is the same.
-Since @fold@s do not produce arrays, they have no constraints.
-@gather@ takes two arrays; the data and the indices. Its output size is the size of the indices.
-The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size.
-Similarly, the size of @external@ outputs is not known at all, and thus cannot be constrained.
+Although our constraint based formulation of size inference is reminiscent of type inference for HM(X)~\cite{odersky1999type}, there are important differences.
+Firstly, our type schemes include existential quantifiers, which represent the fact that the size of an arrays produced by filter operations are unknown in general.
+This is also the case for @generate@, where the result size is data dependent.
+HM(X) style type inferences makes use of an $\exists$ quantifier to bind local type variables in constraints, but they do not appear in type schemes.
+Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments.
+Provided we generate the contraints in the correct form, solving them is straightforward.
 
-Size inference has been explored before in the context of fusion by Chatterjee~\cite{chatterjee1991size}, but the only constraints they support are equality.
-Their formulation has no filtering operations, and all array operations in their case generate the same size as at least one of the operation's inputs.
-A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
-Since we have fewer primitives and simpler goals, our size inference does not need to be as complicated as Jay's shape inference.
-
-Our formulation of size inference is very similar to type inference, particularly the exposition of HM(X)~\cite{odersky1999type}.
+% Our formulation of size inference is very similar to type inference, particularly the exposition of HM(X).
 First, constraints are generated for the bindings, such as equality among two sizes, the conjunction of two constraints, and existentials.
@@
 Two loops with the same iteration size can be fused together.
+
+% While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times.
+For simplicity, we do away with this added complexity and only support fusion of equal sized loops.
+\emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.
+
+% A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
+Since we have fewer primitives and simpler goals, our size inference does not need to be as complicated as Jay's shape inference.
+
+% The $@map@_n$ combinators require all input arrays to be the same size, and their output is the same.
+Since @fold@s do not produce arrays, they have no constraints.
+@gather@ takes two arrays; the data and the indices.
+Its output size is the size of the indices.
+The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size.
+Similarly, the size of @external@ outputs is not known at all, and thus cannot be constrained.
 

### 2014-05-05 2014betterfusionforfilters/section/03-SizeInference.tex
Wibble

@@
 This is also the case for @generate@, where the result size is data dependent.
-HM(X) style type inferences makes use of an $\exists$ quantifier to bind local type variables in constraints, but they do not appear in type schemes.
+HM(X) style type inferences uses the $\exists$ quantifier to bind local type variables in constraints, existential quantifiers do not appear in type schemes.
 Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments.
-Provided we generate the contraints in the correct form, solving them is straightforward.
+Provided we generate the constraints in the correct form, solving them is straightforward.
 

### 2014-05-05 2014betterfusionforfilters/section/03-SizeInference.tex
transducer. proof sketch.

--- 2014-05-04 2014betterfusionforfilters/section/03-SizeInference.tex
@@
 
-\subsection{Subsizes and generators}
+\subsection{Transducers}
 Unlike previous work, we do allow combinators with different iteration sizes to be fused together.
 For example, an operation on filtered data may be fused with the filter operation that generates the data, even though the iteration sizes are different.
+More generally, if the output size of a combinator is different to its iteration size, it is a \emph{transducer} from the iteration size to the output size.
+As usual, a transducer may fuse with other nodes of the same iteration size,
+but transducers may also fused with nodes with iteration size the same as the transducer's output size.
+For our set of combinators, the only transducer is @filter@.
 
 \begin{tabbing}
-MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
-$gen$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
-$gen(bs,o)$    \\
+MMMM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
+$trans$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
+$trans(bs,o)$    \\
             \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $\emptyset$ \\
-            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $\bigcup_{n \in ns} gen(bs, n)$ \\
+            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $\bigcup_{x \in ns} trans(bs, x)$ \\
             \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $\{o\}$       \\
             \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
-            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $gen(bs,i)$ \\
+            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $trans(bs,i)$ \\
             \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $\emptyset$ \\
@@
 %\begin{lemma}
-\textbf{Lemma: unique generators}
-For some bindings $bs$, then
+\textbf{Lemma: unique transducers}.
+For some bindings $bs$, each name $n$ will have at most one transducer.
 \[
-valid(\constr{bs}) \implies \forall x.\ |gen(bs, x)| \le 1
+valid(\constr{bs}) \implies \forall n.\ |trans(bs, n)| \le 1
 \]
+\textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
+then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
+As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
+if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
+The other cases are trivial.
 %\end{lemma}

### 2014-05-05 2014betterfusionforfilters/section/03-SizeInference.tex
Add section on size types, constraints and schemes

@@
 Although our constraint based formulation of size inference is reminiscent of type inference for HM(X)~\cite{odersky1999type}, there are important differences.
-Firstly, our type schemes include existential quantifiers, which represent the fact that the size of an arrays produced by filter operations are unknown in general.
+Firstly, our type schemes include existential quantifiers, which represent the fact that the sizes of arrays produced by filter operations are unknown in general.
 This is also the case for @generate@, where the result size is data dependent.
-HM(X) style type inferences uses the $\exists$ quantifier to bind local type variables in constraints, existential quantifiers do not appear in type schemes.
+HM(X) style type inferences uses the $\exists$ quantifier to bind local type variables in constraints, and existential quantifiers do not appear in type schemes.
 Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments.
@@
 
+% -----------------------------------------------------------------------------
+\begin{figure}
+\begin{tabbing}
+MMMMMMMM \= MM  \= MM \= MMMMMM \= \kill
+\textbf{Size Type}
+\> $\tau$   \> @::=@  \> $k$                  \> (size variable)       \\
+\>          \> $~|$   \> $\tau \times \tau$   \> (cross product)
+\end{tabbing}
+
+\begin{tabbing}
+MMMMMMMM \= MM  \= MM \= MMMMMM \= \kill
+\textbf{Size Constraint}
+\> $C$      \> @::=@  \> $true$               \> (trivially true)      \\
+\>          \> $~|$   \> $\tau = \tau$        \> (equality constraint) \\
+\>          \> $~|$   \> $C \wedge C$         \> (conjunction)
+\end{tabbing}
+
+\begin{tabbing}
+MMMMMMMM \= MM  \= MM \= MMMMMM \= \kill
+\textbf{Size Scheme}
+\> $\sigma$ \> @::=@  
+        \> $\forall \ov{k}.~ \exists \ov{k}.~ (\ov{ x : \tau }) \to (\ov{x : \tau})$
+\end{tabbing}
+
+\caption{Sizes, Constraints and Schemes}
+\label{f:constraints}
+\end{figure}
+
 \newcommand{\constr}[1]{\llbracket #1 \rrbracket}
@@
 % -----------------------------------------------------------------------------
-\subsection{Constraint generation}
-Sizes may either be some variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two sizes.
-\begin{tabbing}
-MMMMM   \= MM \= MMMMMMMMM \= \kill
-$size$  \> @::=@ \> $k_n$               \> (size variable) \\
-        \> $~|$  \> $size \times size$  \> (cross product)
-\end{tabbing}
+\subsection{Size types, constraints and schemes}
+Figure~\ref{f:constraints} shows the grammar for size types, constraints and schemes.
+A size scheme is like a type constraint from Hindley-Milner type systems, except that it only mentions the size of each input array, instead of the element types as well.
 
-The most important constraint is equality on two sizes.
-Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be valid.
-Forall quantifiers mean the constraint must be satisfiable for every size.
-Existential quantifiers simply state that there must exist a size that satisfies the constraint.
-\begin{tabbing}
-MMMMM   \= MM \= MMMMMMMMM \= \kill
-$C$     \> @::=@ \> $true$              \> (trivially true)             \\
-        \> $~|$  \> $size = size$       \> (equality constraint)        \\
-        \> $~|$  \> $C \wedge C$        \> (conjunction)                \\
-        \> $~|$  \> $\forall k_n.\ C$   \> (forall quantification)      \\
-        \> $~|$  \> $\exists k_n.\ C$   \> (existential quantification)
-\end{tabbing}
+A size may either be a variable $k$, or a cross product of two sizes.
+We use the latter to represent the result size of the @cross@ operator discussed in the previous section.
+Constraints maybe either be trivially $true$, an equality $\tau = \tau$, or a conjunction of two constraints $C \wedge C$.
+We refer to the trivially true and equality constraints as \emph{atomic constraints}.
+Size schemes relate the sizes of each input and output array.
+For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
+$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+$$
+
+When giving the size scheme of an operator graph we use $:_s$ to indicate that this is a size scheme rather than a type scheme.
+
+The existential quantifier appears in size schemes when the array produced by a filter or similar operator appears in the result. For example:
+\begin{alltt}
+   filterLeft \(:\sb{s}\,\forall\,k\sb{1}.\,\exists\,k\sb{2}.\,(xs\,:\,k\sb{1})\;\to\;(ys\sb{1}\,:\,k\sb{1},\,ys\sb{2}\,:\,k\sb{2})\)
+   filterLeft xs
+     = let ys1 = map (+ 1)   xs
+           ys2 = filter even xs
+       in (ys1, ys2)
+\end{alltt}
+
+The size scheme of @filterLeft@ shows that it works for input arrays of all sizes. The first result array has the same size as the input, and the second has some unrelated size.
+
+Finally, note that size schemes form but one ``layer'' of the type information that would be expressible in a full dependently typed language.
+For example, in Coq or Agda~\CITE we could write something like:
+\begin{alltt}
+filterLeft 
+ : \(\forall\,k\sb{1}.\,\exists\,k\sb{2}.\) Array \(k\sb{1}\) Float
+              \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
+\end{alltt}
+
+However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process.
+In our situation we need automatic type generalization, but for a first order language only.
+
+\subsection{Constraint Generation}
 

### 2014-05-05 2014betterfusionforfilters/section/03-SizeInference.tex
Add section on Constraint Generation

@@
 
+% -----------------------------------------------------------------------------
 \subsection{Constraint Generation}
+The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}.
+The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a new environment $\Gamma_2$ and size constraints $C$''.
+The other judgment forms are similar.
+The environment $\Gamma$ has the following grammar:
+$$
+\Gamma @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
+$$
 
+The $\cdot$ and $\Gamma,~ \Gamma$ are the empty environment and environment concatenation as usual.
+The element $zs : k$ records the size $k$ of some array variable $zs$.
+A plain $k$ indicates that $k$ can be unified with other size types when solving constraints, whereas $\exists k$ indicates a ``rigid'' size variable that cannot be unified with other sizes.
+We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.
+
+Note that the constraints are generated in a specific form to make the constraint solving process easy.
+For each array variable in the program we generate a new size variable, like size $k_{zs}$ for array variable $zs$.
+These new size variables always appear on the \emph{left} of atomic equality constraints.
+For each array binding we may also introduce unification or rigid variables, and these appear on the \emph{right} of atomic equality constraints.
+
+For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
+$$
+\begin{array}{ll}
+       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
+\\
+\vdash & true 
+        ~\wedge~  k_{gts} = k_1
+        ~\wedge~  true
+\\     &~~~~~~~~ 
+          \wedge~  k_{xs}  ~= k_2
+        ~ \wedge~  k_{ys1}  = k_2 
+        ~ \wedge~  k_{xs}   = k_3
+        ~ \wedge~  k_{ys2}  = k_3
+\end{array}
+$$
+
+% -------------------------------------------------------------------
+\subsection{Constraint Solving and Generalization}
 Constraints are generated for a list of bindings.
@@
 
-The final clauses show how the constraint holes are filled by constraints of subsequent bindings. If there are no bindings, the constraint is trivially true.
+\input{figures/Size-env.tex}
 
-\begin{figure*}
-\input{figures/Size-env.tex}
-\end{figure*}
-
-\TODO{Notes about new constraint generation rules. These need to be fleshed out}
 \begin{itemize}
-
-\item   There are three sorts of size variable: sizes of program variables, existential sizes, and unification variables.
-Sizes of program variables are named after the program variable (like $k_{xs}$), and the mapping appears in the environment like $xs : k_{xs}$.
-Existential variables are marked like $\exists k'$ in the environment.
-Unification variables like $k'$ appear in the environment with no $\exists$ qualifier.
-
-\item   When generating contrains, we introduce fresh existential sizes, or unfication variables to constrain the program size variables appropriately.
-Program size variables always appear on the left of atomic equality constriants, while existential sizes and unification variables appear on the right.
-
-\item   Generating the constraints like this makes the easy to solve.
-The constraint set is in solved form when each program size variable appears only once, and on the left of atomic equality constraints.
-Additionally, existentials and unification variables can appear many times, but only on the right of atomic constraints.
-To solve the constraints, keep finding pairs of atomic equality constraints where the same program size variable appears on the left, and perform an appropriate substitution into the whole constraint set.
+\item    To solve the constraints, keep finding pairs of atomic equality constraints where the same program size variable appears on the left, and perform an appropriate substitution into the whole constraint set.
 

### 2014-05-07 2014betterfusionforfilters/section/04-ILP.tex
ilp

@@
 
-If there is no edge between $i$ and $j$, the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively means there is no constraint on the $\pi$ values.
-Note, however, that other edges may still constrain $i$ or $j$ by transitivity.
-\ben{If there is ``no constraint'' on the $\pi$ values then why bother adding the constraint at all?}
+If there is no edge between $i$ and $j$ and the two are not fused together, then $x_{ij} = 1$ and the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively puts no constraint on the $\pi$ values.
+Note, however, that other edges may still constrain $i$ or $j$.
 

### 2014-05-07 2014betterfusionforfilters/section/04-ILP.tex
ilp

@@
 
-\subsubsection{The simple version}
-\TODO{Why is this ``the simple version''?
-Why present a simple version before the real one?} \TODO{We need to show a complete constraint system at the start.
-Maybe use a simple objective function to start with, then refine it in a later section.}
+\subsubsection{Unoptimised version}
+Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
+The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent.
 
@@
            \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_i=\emptyset \wedge gen_j=\emptyset$)                 \\
+           \> (if $\tau_i \not= \tau_j \wedge trans_i=\emptyset \wedge trans_j=\emptyset$)                 \\
            \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
+           \> (if $\tau_i \not= \tau_j \wedge trans_i=\{i'\}$) \\
            \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+           \> (if $\tau_i \not= \tau_j \wedge trans_j=\{j'\}$) \\
            \> \ldots
@@
 
-\subsection{Subtype optimisations}
-For two clusters $i, j$, if $x_{ij} = 0 \vee \tau_i \not= \tau_j$, then $i$'s generator must be fused with $j$ and vice versa, recursively ($x_{ij} = 0 \implies check(i,j)$).
-Problem: actually generating constraints for \emph{all} pairs of nodes with unequal types is excessive.
-I see two solutions to this:
-\begin{itemize}
-\item
-Generate fewer constraints and rely on transitivity.
+\subsection{A note on transitivity}
+It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
 This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
 This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.
-\item
+
+\subsubsection{Fusion-preventing path optimisation}
+\label{s:OptimisedConstraints}
 Alternatively:
@@
 But we still get the minimisation to count non-connected nodes.
-\end{itemize}
 
@@
 
-\subsubsection{An optimised version}
-\label{s:OptimisedConstraints}
 We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
@@
 
-\subsection{Better than stream fusion}
-Abdallah suggests something like:
-\[
-\forall g.\ \exists c.\ ilp(g, c) \le_c streamfusion(g)
-\]
-ie, for any graph there exists a criterion to optimise against, where we are better than or equal to stream fusion.
-
-\subsection{Cost function}
-We argue that our cost function makes intuitive sense, by reducing the number of manifest arrays and loops.
-However, due to the nature of our heuristic cost function we have no strong guarantees that the clustering with the lowest cost will actually be the clustering with the shortest runtime.
-In general, since we do not know the expected sizes of the arrays, this is impossible to guarantee correctly.
-It is possible that for certain restricted subsets of the language, a more sophisticated cost function may be able to guarantee optimality.
-This is left to future work.
-
 \subsection{Proof}
@@
 
-Now, prove that this assignment actually satisfies the constraints.
+\TODO{Now, prove that this assignment actually satisfies the constraints.}
 

### 2014-05-07 2014betterfusionforfilters/Main.tex
constraint generation fixes
Another go at size inference constraint generation
Re-written the abstract to make the contribution a bit more explicit

@@
 \usepackage{mathptmx}
+\usepackage{stmaryrd}
 \usepackage{hyperref}
@@
 \usepackage{style/proof}
+\usepackage{style/judgements}
 
@@
 \begin{abstract}
+In this paper, we present an array fusion technique for functional programs, which is able to handle multi-loop programs containing size altering operations like filter.
 Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
+However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs requiremore.
+Other work demonstrated how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input.
+The technique we introduce in this paper combines the advantages of both, by  extending the existing ILP approach with support for size changing operators, using an external ILP solver to find good clusterings.
+
+%The original abstract:
+%Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
 However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.

### 2014-05-08 2014betterfusionforfilters/section/01-Introduction.tex
Add section on Constraint Generation
Rewritten intro, typo in the abstract

--- 2014-04-18 2014betterfusionforfilters/section/01-Introduction.tex
@@
 \section{Introduction}
+\label{s:Introduction}
 
-Data flow fusion~\cite{lippmeier2013flow} is a technique to compile a specific class of data flow programs into single, efficient imperative loops.
-This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
-The key benefits of data flow fusion over this prior work are: 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
+To compile collection-oriented array functional array programs, which play a particular important part in parallel programming, into efficient code, it is essential that we can compile the array operations to a small number of imperative loops.
+Data flow fusion~\cite{lippmeier2013flow} addressed this problem by presenting a technique to compile a specific class of data flow programs into single, efficient imperative loops.
+It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+However, it can only fuse code fragments into one loop that produce a single output array.
+Approaches based on Integer Linear Programming (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}) do not suffer from this shortcoming.
+On the other hand, they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.
+The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.
 
-Fusion-preventing dependencies express the fact that some operators simply must wait for others to complete before they can produce their own output. For example, in the following:
+To demonstrate the effect of the three approaches, consider the following code snippet:
 \begin{code}
  normalize :: Array Int -> Array Int
-  normalize xs = let sum = fold (+) 0 xs
-                 in  map (/ sum) xs
-\end{code}
-
-If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using at least two loops: one to determine the sum, and one to divide the elements.
-The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
-
-However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
-The following example offers \emph{several} unique, but mutually exclusive approaches to fusion.
-Figure~\ref{f:normalize2-cluterings} on the next page shows some of the possibilities.
-\begin{code}
- normalize2 :: Array Int -> Array Int
- normalize2 xs
+ normalize xs
   = let sum1 = fold   (+)  0   xs
@@
 \end{code}
+Two sums are computed, one of all the elements of @xs@, the other of all the ones greater than zero.
+Since we need to fully compute the sums to proceed with the maps, it is clear that we need at least two separate loops.
+These are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations.
+Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@ and the effect of all three fusion techniques, where the dotted lines represent the loops clusters.
+Applying data-flow fusion (left-most diagram) result in four distinct loops: one for the fold for the calculation of @sum1@, one for @sum2@, combining the fold and filter, and two for each map.
+The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output.
+Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops.
+The first to calculate both sums, the second to calculate the maps.
 
-In Figure~\ref{f:normalize2-cluterings}, the dotted lines show possible clusterings of operators.
+% Data flow fusion~\cite{lippmeier2013flow} is a technique to compile a specific class of data flow programs into single, efficient imperative loops.
+This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
+The key benefits of data flow fusion over this prior work are: 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
+This process of ``compilation'' is equivalent to performing array fusion on a combinator based functional array program, as per related work on stream fusion~\cite{coutts2007streamfusion} and delayed arrays~\cite{keller2010repa}.
+The key benefits of data flow fusion over this prior work are: 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) complete fusion into a single loop is guaranteed for all programs that operate on the same size input data, and contain no fusion-preventing dependencies between operators.
+
+% Fusion-preventing dependencies express the fact that some operators simply must wait for others to complete before they can produce their own output. For example, in the following:
+% \begin{code}
+%   normalize :: Array Int -> Array Int
+%   normalize xs = let sum = fold (+) 0 xs
+%                  in  map (/ sum) xs
+% \end{code}
+
+% If we wish to divide every element of an array by the sum of all elements, then it seems we are forever destined to compute the result using at least two loops: one to determine the sum, and one to divide the elements.
+The evaluation of @fold@ demands all elements of its source array, and we cannot produce any elements of the result array until we know the value of @sum@.
+
+% However, many programs \emph{do} contain opportunities for fusion, if we only knew which opportunities to take.
+The following example offers \emph{several} unique, but mutually exclusive approaches to fusion.
+Figure~\ref{f:normalize2-clusterings} on the next page shows some of the possibilities.
+% \begin{code}
+%  normalize2 :: Array Int -> Array Int
+%  normalize2 xs
+%   = let sum1 = fold   (+)  0   xs
+%         gts  = filter (> 0)    xs
+%         sum2 = fold   (+)  0   gts
+%         ys1  = map    (/ sum1) xs
+%         ys2  = map    (/ sum2) xs
+%     in (ys1, ys2)
+% \end{code}
+
+% In Figure~\ref{f:normalize2-clusterings}, the dotted lines show possible clusterings of operators.
 Stream fusion implicitly choses the solution on the left as its compilation process cannot fuse a produced array into multiple consumers.

### 2014-05-08 2014betterfusionforfilters/section/02-CNF.tex
Fix citation in section 2
Minor changes

@@
 Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}.
-Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion \CITE.
+Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion~\cite{gao1993collective}.
 We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}).
@@
 However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused.
-In Figure~\ref{f:CombinatorNormalForm} note that the bindings have been split into the ones that produce scalar values ($sbind$), and the ones that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
+In Figure~\ref{f:CombinatorNormalForm}, note that the bindings have been split those that produce scalar values ($sbind$), and those that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
 
@@
 We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment.
-We assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like.
+We also assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like.
 Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program.

### 2014-05-08 2014betterfusionforfilters/section/03-SizeInference.tex
Add Nat types to dep type version of signature
Add note for clock inference
Minor stuff

@@
 We achieve this with a simple constraint based inference algorithm, which we discuss in this section.
-Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}, but this system does not support size changing functions such as filter.
+Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}.
+In constrast to our algorithm, \cite{chatterjee1991size} does not support size changing functions such as filter.
 
@@
 
-When giving the size scheme of an operator graph we use $:_s$ to indicate that this is a size scheme rather than a type scheme.
+We write $:_s$ to distinguish size schemata from type schemata.
 
@@
 \begin{alltt}
-filterLeft 
- : \(\forall\,k\sb{1}.\,\exists\,k\sb{2}.\) Array \(k\sb{1}\) Float
-              \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
+filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
+  Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
 \end{alltt}
@@
 $$
-\Gamma @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
+\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
 $$
 
-The $\cdot$ and $\Gamma,~ \Gamma$ are the empty environment and environment concatenation as usual.
+As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$ environment concatenation.
 The element $zs : k$ records the size $k$ of some array variable $zs$.
@@
 \begin{array}{ll}
-       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
+       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
 \\
@@
 \end{itemize}
+
+\TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
 

### 2014-05-08 2014betterfusionforfilters/section/03-SizeInference.tex
Describe constraint solving process

@@
 The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}.
+The top level judgment ~~$\SizeF{function}{\sigma}$~~ assigns a size scheme to a function.
+It does this by extracting size constraints and them solving them.
+This rule, along with the constraint solving process is discussed in the next section.
 The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a new environment $\Gamma_2$ and size constraints $C$''.
-The other judgment forms are similar.
+The remaining judgment that extracts constraints from a list of bindings is similar.
 The environment $\Gamma$ has the following grammar:
@@
 \subsection{Constraint Solving and Generalization}
-Constraints are generated for a list of bindings.
-As folds produce no constraints, they simply return the constraints of the remaining bindings.
-Maps require their inputs and output to all have the same size, so if there exists some size that is equal to all its inputs, then that is the size of the output.
-As the output size of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible sizes, the subconstraint must be satisfiable.
-Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all sizes.
+The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result.
+In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes and environment and constraint, producing a solved environment and constraint.
+As the constraint solving process is both standard and straightforward, we only describe it informally.
+
+Recall from the previous section that in our generated constraints all the size variables named after program variables are on the left of atomic equality constraints, while all the unification and existential variables are on the right.
+To solve the constraints we keep finding pairs of atomic equality constraints where the same variable appears on the left, unify the right of both of these constraints, and apply the resulting substitution to both the environment and original constraints.
+When there are no more pairs of constraints with the same variable on the left then the constraints are in solved form and we are finished.
+
+During constraint solving, all unification variables mentioned in the environment can have other sizes substituted for them.
+In contrast, the rigid variables marked by the $\exists$ symbol cannot.
+For example, if we consider the constraints for @normalize2@ mentioned before:
+$$
+\begin{array}{ll}
+       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
+\\
+\vdash & true 
+        ~\wedge~  k_{gts} = k_1
+        ~\wedge~  true
+\\     &~~~~~~~~ 
+          \wedge~  k_{xs}  ~= k_2
+        ~ \wedge~  k_{ys1}  = k_2 
+        ~ \wedge~  k_{xs}   = k_3
+        ~ \wedge~  k_{ys2}  = k_3
+\end{array}
+$$
+
+Note that $k_xs$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $true$ terms then yields:
+$$
+\begin{array}{ll}
+       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
+\\
+\vdash & k_{gts} = k_1
+        ~\wedge~  k_{xs}  ~= k_2
+        ~\wedge~  k_{ys1}  = k_2 
+        ~\wedge~  k_{ys2}  = k_2
+\end{array}
+$$
+
+To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately.
+In this case no rigid size variables appear in the result, so we can universally quantify all of them:
+$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+$$
 
@@
 
-\begin{itemize}
-\item    To solve the constraints, keep finding pairs of atomic equality constraints where the same program size variable appears on the left, and perform an appropriate substitution into the whole constraint set.
-
-\item   Once the constraints have been put into solved form, we use the generalisation rule to produce a type scheme for the whole graph.
-The type scheme contains both universal and existential quantifiers.
-The side conditions on the generalization rule reject the @bad1@ example: we must generalize the sizes of all input arrays, but the fact that an @vs@ in this example is given an existential size prevents this from happening.
-\end{itemize}
-
-\TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
-
 % -----------------------------------------------------------------------------
-\subsubsection{Examples}
-As an example of constraint generation, here is the @normalize2@ program from earlier.
-\begin{tabbing}
-@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
-@normalize2 us@                     \> $\exists k_{us}.$      \\
-@ = let sum1 = fold   (+) 0 us@     \>                      \\
-@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
-@       sum2 = fold   (+) 0 gts@    \> \\
-@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
-@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
-@   in (nor1, nor2)@                \> $true$ \\
-\end{tabbing}
-This constraint is valid, with the equivalence classes being:
-\newcommand{\eqclasses}[1]{
-    \begin{tabbing}
-        MM \= M \= \kill
-        #1
-    \end{tabbing}}
-\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
-\eqclasses{
-    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
-    \eqclass{k_{gts}}{k_{gts}}
-}
-
-The next example involves two filters using the same predicate.
-Despite using the same predicate and input data, we produce different output sizes for each filter.
-\begin{tabbing}
-@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
-@diff xs@                           \> $\exists k_{xs}.$ \\
-@ = let ys1 = filter p xs@          \> $\forall k_{ys1}.$       \\
-@       ys2 = filter p xs@          \> $\forall k_{ys2}.$       \\
-@   in (ys1, ys2)@                  \> $true$                   \\
-\end{tabbing}
-
-This constraint is valid, with the equivalence classes being:
-\eqclasses{
-    \eqclass{k_{xs}}    {k_{xs}}
-    \eqclass{k_{ys1}}   {k_{ys1}}
-    \eqclass{k_{ys2}}   {k_{ys2}}
-}
+\subsection{Existentially quantified sizes}
+\TODO{Describe a well typed example that does a filter, and where the filtered result is returned. Then discuss the two mistyped examples and how our generalization rule rejects them.}
 
@@
 
+\TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
+
+\newcommand{\eqclasses}[1]{
+    \begin{tabbing}
+        MM \= M \= \kill
+        #1
+    \end{tabbing}}
+
+\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
+
+% As an example of constraint generation, here is the @normalize2@ program from earlier.
+% \begin{tabbing}
+% @MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+% @normalize2 us@                     \> $\exists k_{us}.$      \\
+% @ = let sum1 = fold   (+) 0 us@     \>                      \\
+% @       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
+% @       sum2 = fold   (+) 0 gts@    \> \\
+% @       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
+% @       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
+% @   in (nor1, nor2)@                \> $true$ \\
+% \end{tabbing}
+% This constraint is valid, with the equivalence classes being:
+% \eqclasses{
+%     \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
+%     \eqclass{k_{gts}}{k_{gts}}
+% }
+
+% The next example involves two filters using the same predicate.
+% Despite using the same predicate and input data, we produce different output sizes for each filter.
+% \begin{tabbing}
+% @MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
+% @diff xs@                           \> $\exists k_{xs}.$ \\
+% @ = let ys1 = filter p xs@          \> $\forall k_{ys1}.$       \\
+% @       ys2 = filter p xs@          \> $\forall k_{ys2}.$       \\
+% @   in (ys1, ys2)@                  \> $true$                   \\
+% \end{tabbing}
+% 
+% This constraint is valid, with the equivalence classes being:
+% \eqclasses{
+%     \eqclass{k_{xs}}    {k_{xs}}
+%     \eqclass{k_{ys1}}   {k_{ys1}}
+%     \eqclass{k_{ys2}}   {k_{ys2}}
+% }
+
+% -----------------------------------------------------------------------------
 \subsection{Constraint validity}
+\ben{I don't think this is needed with the new constraint solving approach}
+
 Checking whether a constraint is valid can be done using Herbrand unification, similar to HM(X)~\cite{odersky1999type}.
@@
 
+% As folds produce no constraints, they simply return the constraints of the remaining bindings.
+% Maps require their inputs and output to all have the same size, so if there exists some size that is equal to all its inputs, then that is the size of the output.
+As the output size of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible sizes, the subconstraint must be satisfiable.
+Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all sizes.
+
+% -----------------------------------------------------------------------------
 \subsection{Iteration size}

### 2014-05-08 2014betterfusionforfilters/section/03-SizeInference.tex
remove old constraint validity subsection

@@
 % -----------------------------------------------------------------------------
-\subsection{Constraint validity}
-\ben{I don't think this is needed with the new constraint solving approach}
-
-Checking whether a constraint is valid can be done using Herbrand unification, similar to HM(X)~\cite{odersky1999type}.
-The difference, however, is that our constraints include both universal and existential quantifiers, whereas HM(X) only includes existential quantifiers.
-Herbrand unification is still applicable on constraints with both quantifiers~\cite{buss1995herbrand}.
-This is achieved by performing \emph{Herbrandization}, the dual of Skolemization, to remove the universal quantifiers and replace them with Herbrand functions.
-After the universal quantifiers have been removed, the constraint can easily be put into prenex form, and validated using the same unification method as HM(X).
-
-For example, a simple, but invalid constraint such as:
-\[ \exists x.\ \forall y.\ x = y \]
-will have its universal quantifier removed, and all references to $y$ changed to $f_y(x)$, with $x$ as an argument to $f$ as $x$ is in scope of the quantifier.
-The result being:
-\[ \exists x.\ x = f_y(x) \]
-This resulting constraint, while satisfiable for certain values of $f_y$, is not satisfiable for all values of $f_y$, and so is not valid.
-
-After the constraint has been validated, equivalence classes are easily constructed.
-The transitive closure of the equality constraints, ignoring quantifiers, is generated.
-As there are no disjunction, negation or implication all equality constraints will be applicable, regardless of the quantifiers used.
-
-% As folds produce no constraints, they simply return the constraints of the remaining bindings.
-% Maps require their inputs and output to all have the same size, so if there exists some size that is equal to all its inputs, then that is the size of the output.
-As the output size of a filter is unknown and cannot be constrained except for the upper bound, it is a forall: for all possible sizes, the subconstraint must be satisfiable.
-Similarly, external must work for any possible output, so the subconstraint must be satisfiable for all sizes.
-
-% -----------------------------------------------------------------------------
 \subsection{Iteration size}

### 2014-05-09 2014betterfusionforfilters/section/05-Benchmarks.tex
benchmarks. allegedly

@@
 
-\subsection{Quickselect}
-We can't quite do @quickselect@, sadly; it requires @head@.
-\begin{code}
-select :: Vec a -> Int -> a
-select xs k
- | length xs > 0
- = let p    = head xs 
-       xslt = filter  (<p) xs
-       xsge = filter (>=p) xs
-   in  if     length xslt > k
-       then   select xsge (k - length xslt)
-       else   select xslt  k
- | length xs == 1
- = head xs
- | length xs == 0
- = error "Empty array!"
+\begin{tabular}{lrrr}
+                &   Unfused         & Stream fusion   & \textbf{ ILP               }\\
+Normalise2      &   14.0s           & 12.3s           & \textbf{ 11.4s             }\\
+Closest pairs   &   25.6s           & 25.1s           & \textbf{ 22.4s             }\\
+QuadTree        &   x               & x               & \textbf{ x                 }\\
+\end{tabular}
 
-median :: Vec a -> a
-median xs
- = select xs (length xs `div` 2)
-\end{code}
+Notes:
+\begin{itemize}
+\item
+Normalise2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
+\item
+ClosestPoints only tested with 80mb data.
+\item
+ILP solutions were created by hand-fusing based on the clustering of the implementation. In the future, this will be integrated to use data flow fusion.
+\item
+Benchmark programs are available at \url{https://github.com/amosr/clustering/tree/master/benches/code}
+\end{itemize}
 

### 2014-05-10 2014betterfusionforfilters/section/02-CNF.tex
Fix up section on rigid sizes

@@
 \section{Combinator Normal Form}
+\label{s:CombinatorNormalForm}
 Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph.

### 2014-05-10 2014betterfusionforfilters/section/03-SizeInference.tex
fix transducer
Fix up section on rigid sizes

@@
 \subsection{Size types, constraints and schemes}
+\label{s:SizeTypes}
 Figure~\ref{f:constraints} shows the grammar for size types, constraints and schemes.
@@
 Finally, note that size schemes form but one ``layer'' of the type information that would be expressible in a full dependently typed language.
-For example, in Coq or Agda~\CITE we could write something like:
+For example, in Coq or Agda we could write something like:
 \begin{alltt}
@@
 This rule, along with the constraint solving process is discussed in the next section.
-The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a new environment $\Gamma_2$ and size constraints $C$''.
+The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a result environment $\Gamma_2$ and size constraints $C$''.
 The remaining judgment that extracts constraints from a list of bindings is similar.
@@
 To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately.
-In this case no rigid size variables appear in the result, so we can universally quantify all of them:
+This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
 $$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
@@
 % -----------------------------------------------------------------------------
-\subsection{Existentially quantified sizes}
-\TODO{Describe a well typed example that does a filter, and where the filtered result is returned. Then discuss the two mistyped examples and how our generalization rule rejects them.}
+\subsection{Rigid Sizes}
+When the environment of our size constraints contains rigid variables (indicate by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme.
+Consider the @filterLeft@ function from S~\ref{s:SizeTypes}
+\begin{code}
+      filterLeft xs
+        = let ys1 = map (+ 1)   xs
+              ys2 = filter even xs
+          in (ys1, ys2)
+\end{code}
+The size constraints for this function are as follows, and they are already in solved form.
+$$
+\begin{array}{ll}
+       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
+\\
+\vdash &          k_{ys_1} = k_1
+        ~\wedge~  k_{ys_2} = k_2
+        ~\wedge~  k_{xs}   = k_2
+\end{array}
+$$
 
-This example is disallowed, as it would require a runtime check on the size of the @flt@ array.
-\begin{tabbing}
-@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
-@bad1 vs@                           \> $\exists k_{vs}.$ \\
-@ = let flt   = filter p vs@        \> $\forall k_{flt}.$ \\
-@       wrong = map2   f flt vs@    \> $\exists k_{wrong}.\ k_{wrong} = k_{flt}$ \\
-                                    \> $\wedge k_{wrong} = k_{vs} \wedge$ \\
-@   in  wrong@                      \> $true$
-\end{tabbing}
-This constraint is invalid, as by transitivity it can be simplified to $\exists k_{vs}.\ \forall k_{flt}.\ k_{vs} = k_{flt}$.
-Obviously, there exists no sole size such that all sizes are equal to it.
-As a result, no equivalence classes are generated for this example, and no fusion is performed.
+As variable $k_2$ is marked as rigid, we introduce an existential quantifier for it, producing the size scheme stated earlier:
 
-Similarly, this example is not allowed, as it would require a runtime check to determine that the output of two filters @flt@ and @flt2@ are the same size.
-\begin{tabbing}
-@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
-@bad2 vs@                           \> $\exists k_{vs}.$ \\
-@ = let flt  = filter p  vs@        \> $\forall k_{flt}.$ \\
-@       flt2 = filter p' vs@        \> $\forall k_{flt2}.$ \\
-@       mix  = map2   f  flt flt2@  \> $\exists k_{mix}.\ k_{mix} = k_{flt}$ \\
-                                    \> $\wedge k_{mix} = k_{flt2} \wedge$ \\
-@   in  mix@                        \> $true$                               \\
-\end{tabbing}
-Again, by transitivity of equality this constraint can be simplified to $\forall k_{flt}.\ \forall k_{flt2}.\ k_{flt} = k_{flt2}$.
-It is easy to see that this constraint is invalid.
+\begin{alltt}
+   filterLeft \(:\sb{s}\,\forall\,k\sb{1}.\,\exists\,k\sb{2}.\,(xs\,:\,k\sb{1})\;\to\;(ys\sb{1}\,:\,k\sb{1},\,ys\sb{2}\,:\,k\sb{2})\)
+\end{alltt}
 
-\TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
+Note that although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a ``generalisation'' process, there is no corresponding instantiation rule.
+The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another.
+To say this another way, all subgraphs are ``fully inlined''.
+Recall from \S\ref{s:CombinatorNormalForm} that we assume our operator graphs are embedded in a larger host program.
+We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.
 
-\newcommand{\eqclasses}[1]{
-    \begin{tabbing}
-        MM \= M \= \kill
-        #1
-    \end{tabbing}}
+When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes.
+This restriction is necessary to reject programs that we cannot statically guarantee will be well sized.
+For example:
+\begin{code}
+    bad1 xs  = let flt   = filter p xs
+                   ys    = map2   f flt xs
+               in  ys
+\end{code}
 
-\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
+The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array.
+As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true.
+The size constraints are as follows:
+$$
+\begin{array}{ll}
+       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
+\\
+\vdash &          k_{flt}  = k_1
+        ~\wedge~  k_{flt}  = k_2
+        ~\wedge~  k_{xs}   = k_2
+        ~\wedge~  k_{ys}   = k_2
+\end{array}
+$$
+
+\noindent
+Solving this then yields:
+$$
+\begin{array}{ll}
+       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_1
+\\
+\vdash &          k_{flt}  = k_1
+        ~\wedge~  k_{xs}   = k_1
+        ~\wedge~  k_{ys}   = k_1
+\end{array}
+$$
+
+In this case Rule~(SFun) does not apply because the parameter variable $xs$ has size $k_1$, but this is marked as rigid in the environment (with $\exists k_1$).
+
+As a final example, the following function ill-sized because the two filter operators are not guaranteed to produce the same number of elements.
+\begin{code}
+     bad2 xs = let as  = filter p1 xs
+                   bs  = filter p2 xs
+                   ys  = map2   f  as bs
+               in  ys
+\end{code}
+
+The initial size constraints for this function are:
+$$
+\begin{array}{ll}
+       & xs : k_{xs},~ as : k_{as},~ \exists k_1,~ bs : k_{bs},~ \exists k_2,~ ys : k_{ys},~ k_3
+\\
+\vdash &          k_{as}   = k_1
+        ~\wedge~  k_{bs}   = k_2
+        ~\wedge~  k_{as}   = k_3
+        ~\wedge~  k_{bs}   = k_3
+        ~\wedge~  k_{ys}   = k_3
+\end{array}
+$$
+
+To solve these, we note that $k_{as}$ is used twice on the left of an atomic equality constraint, so substitute $k_1$ for $k_3$:
+$$
+\begin{array}{ll}
+       & xs : k_{xs},~ as : k_{as},~ \exists k_1,~ bs : k_{bs},~ \exists k_2,~ ys : k_{ys},~ k_1
+\\
+\vdash &          k_{as}   = k_1
+        ~\wedge~  k_{bs}   = k_2
+        ~\wedge~  k_{bs}   = k_1
+        ~\wedge~  k_{ys}   = k_1
+\end{array}
+$$
+
+At this stage we are stuck because the constraints are not yet in solved form, and we cannot simplify them further.
+Both $k_1$ and $k_2$ are marked as rigid, so we cannot substitute one for the other and produce a single atomic constraint for $k_{bs}$.
+
+% \newcommand{\eqclasses}[1]{
+%     \begin{tabbing}
+%         MM \= M \= \kill
+%         #1
+%     \end{tabbing}}
+% 
+% \newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
 
@@
 
+\TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
+
 \begin{tabbing}
@@
 $trans(bs,o)$    \\
+            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $trans'(bs,n)$ \\
+            \> $|$ \> otherwise             \>          \> $=$ \> $trans'(bs,o)$ \\
+\\
+$trans'(bs,o)$    \\
             \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $\emptyset$ \\

### 2014-05-11 2014betterfusionforfilters/Main.tex
Rewritten intro, typo in the abstract
bench results

@@
 Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
-However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs requiremore.
+However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
 Other work demonstrated how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input.
@@
 Abdallah Saffidine,
+Carter Schonwald,
 and Jingling Xue

### 2014-05-11 2014betterfusionforfilters/section/05-Benchmarks.tex
bench results

@@
 
-\begin{tabular}{lrrr}
-                &   Unfused         & Stream fusion   & \textbf{ ILP               }\\
-Normalise2      &   14.0s           & 12.3s           & \textbf{ 11.4s             }\\
-Closest pairs   &   25.6s           & 25.1s           & \textbf{ 22.4s             }\\
-QuadTree        &   x               & x               & \textbf{ x                 }\\
+\begin{figure*}
+$$\begin{array}{c}
+
+\begin{tabular}{lrrrrrrrr}
+                &   \multicolumn{2}{c}{Unfused}         & \multicolumn{2}{c}{Stream}
+                & \multicolumn{2}{c}{Megiddo} &\multicolumn{2}{c}{\textbf{Ours}} \\
+                & Time & Loops   & Time & Loops      & Time & Loops & Time & Loops   \\
+\hline
+Normalise2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
+Closest pair    & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
+QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
 \end{tabular}
+
+\end{array}$$
+\caption{Benchmark results}
+\label{f:BenchResults}
+\end{figure*}
 
@@
 \begin{itemize}
+% \item
+% Normalise2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
+% \item
+% ClosestPoints only tested with 80mb data.
 \item
-Normalise2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
+In some cases, the clusterings were the same for different methods. In this case, the results are the same. 
 \item
-ClosestPoints only tested with 80mb data.
+The stream fusion benchmarks use the clustering that stream fusion \emph{would} use, but hand-fused and written in C.
+\item
+Programs were run five times with the same input data, and the fastest run was used.
 \item
@@
 \item
-Benchmark programs are available at \url{https://github.com/amosr/clustering/tree/master/benches/code}
+Benchmark programs are available at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}
 \end{itemize}

### 2014-05-11 2014betterfusionforfilters/section/04-ILP.tex
Add notes to ILP section

@@
 Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.
-Compare with prior work on typed fusion that only tries to minimise the number of clusters.}
+Compare with prior work on typed fusion that only tries to minimise the number of clusters.} \TODO{Also give a quick rundown of what ILP actually is.}
 \begin{code}
@@
 For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
-
-\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Function $\to$ graph}
+\ben{This is a different sort of graph than before.
+This is a dependency graph where the nodes are intermediate variables, verses the previous ``program graph'' where the nodes can also be operators.
+We need to contrast these properly and add a diagram of a dependency graph.}
+\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
+
 Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Graph $\to$ ILP}
@@
 M   \= MM \= MMMMMMM \= MM \= \kill
-$x$   \> @::@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
+$x$   \> @:@  \> $node \times node$ \> $\to$ \> $\mathbb{B}$
 \end{tabbing}
@@
 The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
+\ben{When we're talking about the dependency graph, when we say ``nodes can be fused'', it means that the operators that produce these values can be fused -- or that the values are produced by the same cluster of operators.}
 
@@
 M   \= MM \= MMMMMMM \= MM \= \kill
-$\pi$ \> @::@  \> $node$             \> $\to$ \> $\mathbb{R}$
+$\pi$ \> @:@  \> $node$             \> $\to$ \> $\mathbb{R}$
 \end{tabbing}
-The second group of variables is used simply to ensure that the resulting clustering is acyclic.
-For some node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ has a value $\pi_j > \pi_i$.
+The second group of variables is used to ensure that the clustering is acyclic.
+For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$.
 This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor.
 If two nodes are to be fused together, their $\pi$ values must be the same.
+\ben{I don't understand the ``acyclic'' part.
+If we don't add these constraints, then how might the clustering turn out to be ``cyclic''?}
 
@@
 M   \= MM \= MMMMMMM \= MM \= \kill
-$c$   \> @::@  \> $node$             \> $\to$ \> $\mathbb{B}$
+$c$   \> @:@  \> $node$             \> $\to$ \> $\mathbb{B}$
 \end{tabbing}
@@
 They are not required for correctness of the clustering, but are used in the objective function.
-A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together:
-$c_i = 0 \iff \forall (i,j) \in E. x_{ij} = 0$.
+A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i,j) \in E.
+x_{ij} = 0$.
+\ben{Using $\forall$ like this implies that the quantified $i$ isn't the same as the one attached to $c_i$.
+Write this with a list comprehension instead.}
 
+% -----------------------------------------------------------------------------
 \subsubsection{Unoptimised version}
@@
 The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent.
+\ben{What does ``equivalent'' mean when the constraints are different?} \ben{Say what ``optimised'' means in this case.} \ben{Give some hints about why the unoptimised version has more constraints.
+}
 
@@
 If $x_{ij} = 0$, then both cases may be simplified to $0 \le \pi_j - \pi_i \le 0$, which is equivalent to $\pi_i = \pi_j$.
+\ben{Highlight that ``fused together'' means ``in the same cluster'', and that operators can be fused even if there are no direct edges between them.}
 
@@
 \paragraph{Fusion between different types}
-
 \ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}
@@
 \ben{What is a subrate?
-How do we determine the subrates of each rate?}
-\TODO{Expand}.
+How do we determine the subrates of each rate?} \ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.} \ben{Explain how we are using the idea of transducers}.
 
@@
 An array is fully contracted if all its output edges are fused with it.
-Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$.
+Thus, $c_i=0$ only if $\forall j | (i,j) \in E.
+x_{ij} = 0$.
 By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.
+\ben{Say what array contraction is, and why we want this constraint.}
 

### 2014-05-12 2014betterfusionforfilters/Main.tex
Rewrote the abstract

@@
 \begin{abstract}
-In this paper, we present an array fusion technique for functional programs, which is able to handle multi-loop programs containing size altering operations like filter.
-Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
-However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
-Other work demonstrated how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input.
-The technique we introduce in this paper combines the advantages of both, by  extending the existing ILP approach with support for size changing operators, using an external ILP solver to find good clusterings.
+Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code.
+To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation, we also have to also \emph{cluster} separate traversals of the same array into one single loop.
+Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
+However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array.
+The technique we present in this paper extends the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.
 

### 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
mention github impl
rename normalize and normalise to normaliZe2

@@
 \label{s:Introduction}
+A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.
+It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance.
+This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{some fusion refs}, to name just a few.
+Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
+As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering.
+Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.
+The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.
 
-To compile collection-oriented array functional array programs, which play a particular important part in parallel programming, into efficient code, it is essential that we can compile the array operations to a small number of imperative loops.
+To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
+It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+
+% To compile collection-oriented array functional array programs, which play a particular important role in parallel programming, into efficient code, it is essential that we can compile the array operations to a small number of imperative loops.
 Data flow fusion~\cite{lippmeier2013flow} addressed this problem by presenting a technique to compile a specific class of data flow programs into single, efficient imperative loops.
@@
 
-To demonstrate the effect of the three approaches, consider the following code snippet:
+To demonstrate the effect of clustering and fusion, consider the following code snippet:
 \begin{code}
- normalize :: Array Int -> Array Int
- normalize xs
+ normalize2 :: Array Int -> Array Int
+ normalize2 xs
   = let sum1 = fold   (+)  0   xs
@@
 \end{code}
-Two sums are computed, one of all the elements of @xs@, the other of all the ones greater than zero.
-Since we need to fully compute the sums to proceed with the maps, it is clear that we need at least two separate loops.
-These are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations.
-Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@ and the effect of all three fusion techniques, where the dotted lines represent the loops clusters.
-Applying data-flow fusion (left-most diagram) result in four distinct loops: one for the fold for the calculation of @sum1@, one for @sum2@, combining the fold and filter, and two for each map.
-The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output.
+The function @normalize@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero.
+Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops.
+These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations.
+Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@.
+In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold.
+The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations.
 Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops.
@@
 
+An implementation of the clustering algorithm is available at \url{https://github.com/amosr/clustering}.
+
 % We must also decide which clustering is the `best' or most optimal.

### 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
benchmark ref

@@
 We present benchmarks of our algorithm applied to several common programming patterns, and to several pathological examples.
-Our algorithm is complete and yields good results in practice, though if array sizes are unknown, an optimal solution is uncomputable in general. \TODO{ref}
+Our algorithm is complete and yields good results in practice, though if array sizes are unknown, an optimal solution is uncomputable in general (\S\ref{s:Benchmarks}).
 \end{itemize}

### 2014-05-12 2014betterfusionforfilters/Main.tex
Rewrote the intro

@@
 Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
-However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array.
-The technique we present in this paper extends the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.
+However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
+The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.
 

### 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
Rewrote the intro

--- 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
@@
 \begin{code}
- normalize2 :: Array Int -> Array Int
- normalize2 xs
+ normalize :: Array Int -> Array Int
+ normalize xs
   = let sum1 = fold   (+)  0   xs

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
Add notes to ILP section
transducers

@@
 
+% -----------------------------------------------------------------------------
 \subsection{Transducers}
@@
 For example, an operation on filtered data may be fused with the filter operation that generates the data, even though the iteration sizes are different.
-More generally, if the output size of a combinator is different to its iteration size, it is a \emph{transducer} from the iteration size to the output size.
-As usual, a transducer may fuse with other nodes of the same iteration size,
-but transducers may also fused with nodes with iteration size the same as the transducer's output size.
+
+We define the concept of \emph{transducers} as combinators having a different output size to their iteration size.
+As usual, a transducer may fuse with other nodes of the same iteration size, but transducers may also fuse with nodes having iteration size the same as the transducer's output size.
 For our set of combinators, the only transducer is @filter@.
 
-\TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
+Looking back at the @normalize2@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
+The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$.
+Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
+The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
+Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
+\begin{code}
+ normalize2 xs
+  = let sum1 = fold   (+)  0   xs
+        gts  = filter (> 0)    xs
+        sum2 = fold   (+)  0   gts
+        ...
+\end{code}
+
+% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
+% No,
+We now define a function $trans$, to find the parent transducer of a given combinator.
 

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
Typos, clarifications

--- 2014-05-10 2014betterfusionforfilters/section/03-SizeInference.tex
@@
-%!TEX root = ../Main.tex
+`%!TEX root = ../Main.tex
 \section{Size inference}
@@
 Size schemes relate the sizes of each input and output array.
-For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
-$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize-clusterings} is as follows:
+$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
 $$
@@
 
-Finally, note that size schemes form but one ``layer'' of the type information that would be expressible in a full dependently typed language.
-For example, in Coq or Agda we could write something like:
+Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
 \begin{alltt}
@@
 However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process.
-In our situation we need automatic type generalization, but for a first order language only.
+In our situation, we need automatic type generalization, but for a first order language only.
 
@@
 The top level judgment ~~$\SizeF{function}{\sigma}$~~ assigns a size scheme to a function.
-It does this by extracting size constraints and them solving them.
+It does this by extracting size constraints and then solving them.
 This rule, along with the constraint solving process is discussed in the next section.
@@
 
-As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$ environment concatenation.
-The element $zs : k$ records the size $k$ of some array variable $zs$.
-A plain $k$ indicates that $k$ can be unified with other size types when solving constraints, whereas $\exists k$ indicates a ``rigid'' size variable that cannot be unified with other sizes.
+As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$
+environment concatenation. The element $zs : k$ records the size $k$ of some
+array variable $zs$. A plain $k$ indicates that $k$ can be unified with other
+size types when solving constraints, whereas $\exists k$ indicates a  \emph{rigid} size variable that cannot be unified with other sizes.
 We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.
 
-Note that the constraints are generated in a specific form to make the constraint solving process easy.
+Note that the constraints are generated in a specific form, to make the constraint solving process easy.
 For each array variable in the program we generate a new size variable, like size $k_{zs}$ for array variable $zs$.
@@
 
-For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
+For example, the final environment and constraints generated for the @normalize@ example from Section~\ref{s:Introduction} are as follows:
 $$
@@
 In contrast, the rigid variables marked by the $\exists$ symbol cannot.
-For example, if we consider the constraints for @normalize2@ mentioned before:
+For example, if we consider the constraints for @normalize@ mentioned before:
 $$
@@
 To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately.
-This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
-$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.
+In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
+$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
 $$
@@
 \end{code}
-The size constraints for this function are as follows, and they are already in solved form.
+The size constraints for this function, already in solved form, are as follows.
 $$
@@
 
-Note that although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a ``generalisation'' process, there is no corresponding instantiation rule.
+Note that although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule.
 The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another.
-To say this another way, all subgraphs are ``fully inlined''.
+To say this another way, all subgraphs are fully inlined.
 Recall from \S\ref{s:CombinatorNormalForm} that we assume our operator graphs are embedded in a larger host program.
@@
 
-In this case Rule~(SFun) does not apply because the parameter variable $xs$ has size $k_1$, but this is marked as rigid in the environment (with $\exists k_1$).
+In this case Rule~(SFun) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).
 
-As a final example, the following function ill-sized because the two filter operators are not guaranteed to produce the same number of elements.
+As a final example, the following function is ill-sized, because the two filter operators are not guaranteed to produce the same number of elements.
 \begin{code}
@@
 More generally, if the output size of a combinator is different to its iteration size, it is a \emph{transducer} from the iteration size to the output size.
-As usual, a transducer may fuse with other nodes of the same iteration size,
-but transducers may also fused with nodes with iteration size the same as the transducer's output size.
+As usual, a transducer may fuse with other nodes of the same iteration size, but transducers may also be fused with nodes with iteration size the same as the transducer's output size.
 For our set of combinators, the only transducer is @filter@.
-
-\TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
+\ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent.
+so $trans(trans(n)) \not= n$}
 

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex

@@
-`%!TEX root = ../Main.tex
+%!TEX root = ../Main.tex
 \section{Size inference}
@@
 For example, an operation on filtered data may be fused with the filter operation that generates the data, even though the iteration sizes are different.
-More generally, if the output size of a combinator is different to its iteration size, it is a \emph{transducer} from the iteration size to the output size.
-As usual, a transducer may fuse with other nodes of the same iteration size, but transducers may also be fused with nodes with iteration size the same as the transducer's output size.
+
+We define the concept of \emph{transducers} as combinators having a different output size to their iteration size.
+As usual, a transducer may fuse with other nodes of the same iteration size, but transducers may also fuse with nodes having iteration size the same as the transducer's output size.
 For our set of combinators, the only transducer is @filter@.
-\ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent.
-so $trans(trans(n)) \not= n$}
+
+Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
+The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$.
+Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
+The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
+Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
+\begin{code}
+ normalize2 xs
+  = let sum1 = fold   (+)  0   xs
+        gts  = filter (> 0)    xs
+        sum2 = fold   (+)  0   gts
+        ...
+\end{code}
+
+% \ben{Do other systems use the concept of transducers?} \TODO{Explain that a transducer of filter is the filter's parent. so $trans(trans(n)) \not= n$}
+% No,
+We now define a function $trans$, to find the parent transducer of a given combinator.
 

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
Minor fix
transducer parents

@@
 MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$T$          \> @::=@ \> $size$                                  \> (known size) \\
+$T$          \> @::=@ \> $\tau$                                  \> (known size) \\
              \> $~|$  \> @external@                              \> (external and unfusible) \\
@@
 
-After the generated equality constraints are solved, and sizes are grouped into equivalence classes, the combinators with iteration sizes in the same equivalence class can be fused together.
+After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.
 
@@
 Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
-The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$.
+%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
+%\gabi{gts is no combinator, so can't be a transducer
+The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
 Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
@@
 We now define a function $trans$, to find the parent transducer of a given combinator.
+The parent transducer $trans(bs, n)$ of a combinator has the same output size as $n$'s iteration size, but the two have different iteration sizes.
 
@@
 
-%\begin{lemma}
-\textbf{Lemma: unique transducers}.
+To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
+\begin{tabbing}
+MMMM \= M \= MMMMMMM \= M \= \kill
+$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
+$parents(a,b)$ \\
+        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
+        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
+        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
+\end{tabbing}
+
+\textbf{Lemma: sole transducers}.
 For some bindings $bs$, each name $n$ will have at most one transducer.
@@
 The other cases are trivial.
-%\end{lemma}
+
+\textbf{Lemma: transducers change types}.
+For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
+\[
+valid(\constr{bs}) \wedge \{m\} = trans(bs, n) \implies \tau(n) \not= \tau(m)
+\]
+
+\textbf{Lemma: sole parents}.
+For some bindings $bs$, each pair of names $a$ and $b$ will have at most one set of parents $parents(a,b)$.
+\[
+valid(\constr{bs}) \implies \forall a,b.\ |parents(bs, n)| \le 1
+\]
+\textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

### 2014-05-12 2014betterfusionforfilters/section/04-ILP.tex
start fixing ILP different types

@@
 
-\paragraph{Fusion between different types}
-\ben{We haven't said what a ``type'' is in the context of clustering. Do we mean ``rate''? If so, can we use $k$ for the variables instead of $\tau$?}
+\paragraph{Fusion between different iteration sizes}
 
@@
            \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge trans_i=\emptyset \wedge trans_j=\emptyset$)                 \\
+           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \emptyset$)  \\
+           \\
            \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge trans_i=\{i'\}$) \\
            \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge trans_j=\{j'\}$) \\
+           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
            \> \ldots
 \end{tabbing}
+If two nodes have different iteration sizes, they may not necessarily be fused.
+However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.
+
 The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
@@
 How do we determine the subrates of each rate?} \ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.} \ben{Explain how we are using the idea of transducers}.
+
+As a degenerate example, consider fusing an operation on filtered data with its generating filter:
+\begin{code}
+gts  = filter (>0) us
+sum2 = fold (+) 0  gts
+\end{code}
+Here, $parents(gts,sum2) = \{gts, gts\}$.
+This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
+While these constraints are unnecessary in this case, they are harmless.
 

### 2014-05-12 2014betterfusionforfilters/section/04-ILP.tex
note sum1 sum2 example in ilp different types

@@
 \begin{code}
+sum1 = fold (+) 0  us
 gts  = filter (>0) us
@@
 While these constraints are unnecessary in this case, they are harmless.
+
+In the same example, $sum1$ and $sum2$ also have different iteration sizes.
+We have that $parents(sum1,sum2) = \{sum1, gts\}$.
+This means that $sum1$ and $sum2$ may only be fused if $sum1$ is fused with $sum1$ (trivial), $sum2$ is fused with $gts$, and $sum1$ is fused with $gts$.
+While it may seem like these constraints should be implied by transitivity of clustering, all clustering variables are used for the objective function.
 

### 2014-05-12 2014betterfusionforfilters/section/04-ILP.tex
ilp: different sizes: why fusing different sized arrays if important

@@
 
-The main difference to existing integer linear programming solutions is that, while nodes with different loop sizes cannot generally be fused together, we do support some fusion of filtered sizes.
-If two nodes have different sizes or rates, but are both sub-rates of another rate, they \emph{can} be fused together if they are both fused with their generators.
-\ben{What is a subrate?
-How do we determine the subrates of each rate?} \ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.} \ben{Explain how we are using the idea of transducers}.
+This is the main difference to existing integer linear programming solutions: we allow nodes with different iteration sizes to be fused together, when their parent transducers are fused together.
+This is an important feature for combinator-based programming, where size-changing operations such as @filter@ are used frequently.
+\ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.}
 

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
remove proofs of transducer lemmas

@@
 \]
-\textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
-then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
-As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
-if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
-The other cases are trivial.
+% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
+% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
+% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
+% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
+% The other cases are trivial.
 
@@
 \]
-\textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.
+% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.

### 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
size inference: drop superceeded text
pl. of scheme is schemes, wheraas pl. of schema is schemata. scheme and schema are different things (I think)
Redo iteration size section, use \bot to indicate an unknown size

@@
 
-% Our formulation of size inference is very similar to type inference, particularly the exposition of HM(X).
-First, constraints are generated for the bindings, such as equality among two sizes, the conjunction of two constraints, and existentials.
-If these constraints can be solved, the equality constraints are used to group the sizes into equivalence classes.
-Otherwise if the constraints are unable to be solved, the program cannot be assured to require no runtime size checks and will not be fused by our system.
-After the constraints are generated and solved, each combinator is given an iteration size.
-Two loops with the same iteration size can be fused together.
-
-% While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times.
-For simplicity, we do away with this added complexity and only support fusion of equal sized loops.
-\emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.
-
-% A more in-depth analysis has also been explored by Jay~\cite{jay1996shape} in the form of shape inference, which is built from primitives such as @cons@ and @nil@.
-Since we have fewer primitives and simpler goals, our size inference does not need to be as complicated as Jay's shape inference.
-
-% The $@map@_n$ combinators require all input arrays to be the same size, and their output is the same.
-Since @fold@s do not produce arrays, they have no constraints.
-@gather@ takes two arrays; the data and the indices.
-Its output size is the size of the indices.
-The size of a @filter@'s output is most interesting; the exact size is not known until after it has been executed, only that it is less than or equal to the input size.
-Similarly, the size of @external@ outputs is not known at all, and thus cannot be constrained.
-
 % -----------------------------------------------------------------------------
@@
 \> $C$      \> @::=@  \> $true$               \> (trivially true)      \\
-\>          \> $~|$   \> $\tau = \tau$        \> (equality constraint) \\
+\>          \> $~|$   \> $k = \tau$           \> (equality constraint) \\
 \>          \> $~|$   \> $C \wedge C$         \> (conjunction)
@@
 Size schemes relate the sizes of each input and output array.
-For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize-clusterings} is as follows:
+For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
 $$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
@@
 
-We write $:_s$ to distinguish size schemata from type schemata.
+We write $:_s$ to distinguish size schemes from type schemes.
 
@@
 
-% As an example of constraint generation, here is the @normalize2@ program from earlier.
-% \begin{tabbing}
-% @MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
-% @normalize2 us@                     \> $\exists k_{us}.$      \\
-% @ = let sum1 = fold   (+) 0 us@     \>                      \\
-% @       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
-% @       sum2 = fold   (+) 0 gts@    \> \\
-% @       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
-% @       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
-% @   in (nor1, nor2)@                \> $true$ \\
-% \end{tabbing}
-% This constraint is valid, with the equivalence classes being:
-% \eqclasses{
-%     \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
-%     \eqclass{k_{gts}}{k_{gts}}
-% }
-
 % The next example involves two filters using the same predicate.
@@
 \subsection{Iteration size}
-After the constraints are validated and equivalence classes generated, each combinator is assigned a $size$ as an iteration size -- the number of iterations in the loop required to generate the output.
-It is important to note that for filters, the iteration size is not the output size, but is instead the size of the input.
-The output size of a filter is, however, a \emph{subsize} of the input's size, as not only is it known to be less than or equal to its input in size, it is also generated depending on the input.
-The iteration sizes, $\tau_n$, are used to check whether two loops may be fused.
-Any two iteration sizes in the same equivalence class are the same size, and so are fusible.
-The difference from previous work is that loops of different iteration sizes \emph{can} be fused, if one is a subsize of the other, and the subsize's \emph{generator} is fused together it as well.
-Basically, operations on filtered data can be fused with operations on the original data, if it is fused with the filter as well.
-External computations are treated separately, as they cannot be fused with any other nodes.
+After the inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator.
+For @filter@ and other size changing operators, the iteration and result sizes are necessarily different.
+For such an operator, we say that the result size is a \emph{descendant} of the iteration size.
+Conversely, the iteration size is a \emph{parent} of the result size.
+
+This descendant--parent size relation is transitive, so if we filter an array, then filter it again, the size of the result is a descendant of the iteration size of the initial filter.
+This relation arises naturally from Data Flow Fusion~\cite{lippmeier2013flow}, as such an operation would be compiled into a single loop --- with an iteration size identical to the size of the input array, and containing two nested if-expressions to perform the two layers of filtering.
+
+Iteration sizes are used to decide which operators can be fused with each other.
+As in prior work, operators with the same iteration size can be fused.
+However, in our system we also allow operators of different iteration sizes to be fused, provided those sizes are descendants of the same parent size.
+
+We use $T$ to range over iteration sizes, and write $\bot$ for the case where the iteration size is unknown.
+The $\bot$ size is needed to handle the @external@ operator, as in general we cannot know its true iteration size, and it cannot be fused with any other operator.
 
 \begin{tabbing}
-MMMMM       \= MM \= MMMMMMMMMMM \= \kill
-$T$          \> @::=@ \> $\tau$                                  \> (known size) \\
-             \> $~|$  \> @external@                              \> (external and unfusible) \\
+MMMMMMMM \= MM       \= MM \= MMMM \= \kill
+\textbf{Iteration Size}
+ \> $T$         \> @::=@  \> $\tau$        \> (known size) \\
+ \>             \> $~|$   \> $\bot$     \> (unknown size) \\
 \end{tabbing}
 
-Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size.
+Once the size constraints have been solved, we can use the following function to compute the iteration size of each binding.
+In the definition we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$.
+Similarly, we use the syntax $C(k)$ to find the corresponding $k = \tau$ constraint in $C$ and return the associated size type $\tau$.
+
+\begin{tabbing}
+MMMMM \= M \= MMMMMMMMMM \= MM \= \kill
+$\iiter_{\Gamma,C}$  
+        \>$:$\> $bind \rightarrow T$ 
+\\[1ex]
+$\iiter_{\Gamma,C}$
+        \> $|$  \> $(z~ = @fold@~ f~xs)$     
+                \> $=$ \> $C(\Gamma(xs))$ 
+\\
+        \> $|$  \> $(ys = @map@_n~f~\overline{xs})$
+                \> $=$ \> $C(\Gamma(ys))$ 
+\\
+        \> $|$  \> $(ys = @filter@~f~xs)$    
+                \> $=$ \> $C(\Gamma(xs))$ 
+\\
+        \> $|$  \> $(ys = @generate@~s~f)$  
+                \> $=$ \> $C(\Gamma(ys))$ 
+\\
+        \> $|$  \> $(ys = @gather@~is~xs)$    
+                \> $=$ \> $C(\Gamma(is))$ 
+\\
+        \> $|$  \> $(ys = @cross@~as~bs)$     
+                \> $=$ \> $C(\Gamma(as)) \times C(\Gamma(bs))$ 
+\\
+        \> $|$  \> $(ys = @external@~\overline{xs})$  
+                \> $=$ \> $\bot$ 
+\\
+\end{tabbing}
+
+% Once the constraints are solved, known to be valid, and sorted into equivalence classes, each combinator is assigned a size.
 Note that for a filter, the size of the output array $k_o$ is some existential that is less than or equal to $k_n$, but the actual loop size of the \emph{combinator} is equal to $k_n$.
@@
 
-\begin{tabbing}
-MM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
-$\tau$  \>$::$\> $binds \rightarrow name \rightarrow T$ \\
-\\
-$\tau_{bs,o}$    
-            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $k_n$ \\
-            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $k_o$ \\
-            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $k_n$ \\
-            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $k_o$ \\
-            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $k_i$ \\
-            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $k_a \times k_b$ \\
-            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $@external@$ \\
-\end{tabbing}
-
-After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.
+% After the generated equality constraints are solved, and sizes are grouped into equivalence classes, combinators with iteration sizes in the same equivalence class may be fused together if there are no fusion-preventing dependencies between them.
 

### 2014-05-13 2014betterfusionforfilters/section/02-CNF.tex
Caption of Figure 1

@@
 \end{center}
-\caption{Clusterings for normalize2 example: with stream fusion; our system; best imperative system}
+\caption{Clusterings for normalize example: with stream fusion; our system; best imperative system}
 \label{f:normalize2-clusterings}
@@
 \label{s:CombinatorNormalForm}
-Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph.
+Input programs is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph.
 The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}.
-Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}.
+The @normalize@  examples on the previous page is in CNF, and the matching data flow graph for @normalize@ is in Figure~\ref{f:normalize2-clusterings}.
 Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion~\cite{gao1993collective}.

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
ILP: np-hardness and ILP solvers

@@
 
-\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering.
-Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.
-Compare with prior work on typed fusion that only tries to minimise the number of clusters.} \TODO{Also give a quick rundown of what ILP actually is.}
+The problem of finding the ``best'' clustering according to some criteria is known to be NP-hard~\cite{darte1999complexity}.
+For example, in the special case of single-typed fusion, where all loops have the same iteration size, finding a clustering with the minimal number of loops can be done in polynomial time.
+However, to find the minimal number of manifest arrays, each node may need to be considered in several clusters, and the problem becomes NP-hard.
+Similarly, after adding multiple iteration sizes that cannot be fused together, the problem becomes NP-hard.
+
+The clusterings of a given program are not unique; there are many valid clusterings, far too many to feasibly enumerate all of them.
+There may even be multiple clusterings with the minimal number of loops. For example, consider the following program:
+
 \begin{code}
@@
 For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+
+Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
+The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.
+
+In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
+Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
+The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
+This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
+Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.
 

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
ilp: example of cyclic clustering

@@
 \end{tabbing}
-The second group of variables is used to ensure that the clustering is acyclic.
+The second group of variables is used to ensure that the clustering is acyclic -- that is, there must be an ordering to execute each clusters' dependency clusters before executing the cluster itself.
 For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$.
@@
 If two nodes are to be fused together, their $\pi$ values must be the same.
-\ben{I don't understand the ``acyclic'' part.
-If we don't add these constraints, then how might the clustering turn out to be ``cyclic''?}
+
+As an example of a cyclic clustering, consider:
+\begin{code}
+cycle us
+ = let xs  = map (+1) us          (C1)
+       sum = fold xs              (C2)
+       zs  = map (+sum) xs        (C1)
+\end{code}
+There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
+If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
 

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
move proof of ILP and comment out

@@
 
-\subsection{Proof}
-To prove correctness of our linear program formulation, we need to prove two different things.
-Firstly, the formulation's constraints must always be satisfiable; that is, there must exist a variable assignment that satisfies all constraints.
-This is rather simple to show, but guarantees that the linear program will always give an answer.
-The next thing to show is that any produced clustering is legal: if a variable assignment satisfies the constraints, then it is a valid and legal clustering.
-This means that, not only do we get \emph{an} answer, we also get the \emph{right} answer.
-
-%!TEX root = ../Main.tex
-\subsubsection{Satisfiability}
-For any program $p$, there exists a trivial clustering with no fusion at all.
-We can use this as the variable assignment of $ilp(p)$.
-For each pair of nodes $m,n \in p$, $x_{mn} = 1$ --- no fusion is possible.
-For the $\pi$ variables, we must find a topographical ordering of the nodes in $p$, which is simple since we are assured it is a dag.
-
-\TODO{Now, prove that this assignment actually satisfies the constraints.}
-
-\subsubsection{Soundness}
-For any program $p$ and variable asignment $v$, if $v$ satisfies the constraints for $ilp(p)$, the clustering denoted by $x_{ij}$ in $v$ is legal.
-
-For a clustering to be legal, it must satisfy three constraints:
-\begin{description}
-\item[Acyclic]
-after merging nodes of same cluster together, the resulting graph must be a dag
-\item[Precedence preserving]
-if there is an edge between two nodes $i$ and $j$, and they are not merged together, then we require $\pi_j > \pi_i$
-\item[Fusion preventing]
-likewise, if there is a fusion-preventing edge between two nodes $i$ and $j$, then we require $\pi_j > \pi_i$, which implies that they are not merged together
-\item[Type constraint]
-if two nodes $i$ and $j$ are in the same cluster, then $\tau_i = \tau_j$, or if $\tau_i$ is a subtype of $\tau_j$ (or $\tau_j$ is a subtype of $\tau_i$), then the \emph{generator} for $\tau_i$ (or $\tau_j$) must also be in the same cluster as $i$ and $j$.
-    \\
-    \TODO Actually, let us say $x_{ij} = 0 \implies check_{ij}$
-
-\end{description}
-where
-\begin{tabbing}
-MMMMM      \= M \= MMMMMMM \= MM \= \kill
-$check$ \> @::@  \> $array \times array$ \> $\to$ \> $\mathbb{B}$ \\
-MMMMM      \= M \= MMMMMM \= MM \= \kill
-$check(i, j)$     \> $|$ \> $tau_i = tau_j$ \> $=$ \> $x_{i,j} = 0$                        \\
-$check(i, j)$     \> $|$ \> $i' \in gen(i) $ \> $=$ \> $x_{i',j} = 0 \wedge x_{i,i'} = 0 \wedge check(i', j)$                        \\
-$check(i, j)$     \> $|$ \> $j' \in gen(j) $ \> $=$ \> $x_{i,j'} = 0 \wedge x_{j,j'} = 0 \wedge check(i, j')$                        \\
-$check(i, j)$     \> $|$ \> $tau_i \not= tau_j$\> $=$ \> $\bot$
-\end{tabbing}
+% \input{section/04-ILP-Proof.tex}

### 2014-05-13 2014betterfusionforfilters/section/03-SizeInference.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?

@@
 The existential quantifier appears in size schemes when the array produced by a filter or similar operator appears in the result. For example:
+\begin{figure}[H]
 \begin{alltt}
@@
 \end{alltt}
+\end{figure}
 
@@
 Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
+\begin{figure}[H]
 \begin{alltt}
@@
 \end{alltt}
-
+\end{figure}
 However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process.
@@
 Consider the @filterLeft@ function from S~\ref{s:SizeTypes}
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
+\end{figure}
 The size constraints for this function, already in solved form, are as follows.
@@
 For example:
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-
+\end{figure}
 The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array.
@@
 
+\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
+\end{figure}
 
@@
 Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
+\end{figure}
 
@@
 \end{tabbing}
+\end{figure}
 
 To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
+\begin{figure}[H]
 \begin{tabbing}

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?

@@
 
+\begin{figure}[H]
 \begin{code}
@@
 \end{tabbing}
+\end{figure}
 Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
@@
 As an example of a cyclic clustering, consider:
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
+\end{figure}
 There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
@@
 \end{tabbing}
+\end{figure}
 As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.
@@
 \paragraph{Fusion-preventing edges}
+\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
+\end{figure}
 As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
@@
 \end{tabbing}
+\end{figure}
 If two nodes have different iteration sizes, they may not necessarily be fused.
@@
 As a degenerate example, consider fusing an operation on filtered data with its generating filter:
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
+\end{figure}
 Here, $parents(gts,sum2) = \{gts, gts\}$.
@@
 \paragraph{Array contraction}
+\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
+\end{figure}
 As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
ILP section
changed the order of fusion options to match with figure, added some
text

--- 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
@@
 \caption{Possible clusterings for \texttt{normalizeInc}}
+\label{f:normInc}
 \end{figure}
 
-The problem of finding the ``best'' clustering according to some criteria is known to be NP-hard~\cite{darte1999complexity}.
-For example, in the special case of single-typed fusion, where all loops have the same iteration size, finding a clustering with the minimal number of loops can be done in polynomial time.
-However, to find the minimal number of manifest arrays, each node may need to be considered in several clusters, and the problem becomes NP-hard.
-Similarly, after adding multiple iteration sizes that cannot be fused together, the problem becomes NP-hard.
-
-The clusterings of a given program are not unique; there are many valid clusterings, far too many to feasibly enumerate all of them.
-There may even be multiple clusterings with the minimal number of loops. For example, consider the following program:
-
+\TODO{More justification of why we want to use a heavyweight technology like ILP to do the clustering.
+Use normalizeInc or one like it to argue that plenty of real programs will admit multiple valid clusterings, so we really want to do this via constraints and an objective function.
+Compare with prior work on typed fusion that only tries to minimise the number of clusters.} \TODO{Also give a quick rundown of what ILP actually is.}
 \begin{code}
@@
  normalizeInc xs
-  = let incs = map  (+1)    us      (A1) (B2)
+  = let incs = map  (+1)    us      (A2) (B1)
         sum  = fold (+) 0   us      (A1) (B1)
@@
 \end{code}
+% swapped A and B, so that the first option matches with the left graph, the second with the right graph in the diagram
+For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}.
+The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@).
+For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@.
+For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
 
-For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
+The @normalizeInc@ example demonstrates why it is, in general, not sufficient, to simply minimize the number of loops in a program: we have to assign weights to different types of loops and take those into account when choosing a possible clustering.
+Ordered typed fusion \cite{kennedy1993typedfusion} addresses this problem in the context of fusing imperative parallel and sequential loops, but also only in the absence of filter-like operations.
+Our problem can be directly mapped to an integer linear program with some integral, boolean and real variables, an objective function to minimise, and a set of contraints the variables must conform to.
+While finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, we can build on exisiting work here and use the best ILP-solver available, which tend to be sufficient for realistic problem sizes.
+For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still results in a very efficient, if not necessarily optimal program.
+
+%For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
 The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
 For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
-
-Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
-The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.
-
-In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
-Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
-The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
-This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
-Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.
 
@@
 
-The dependency graph is translated to an integer linear program.
+% The dependency graph is translated to an integer linear program.
 The integer linear program has some integral, boolean and real variables, an objective function to minimise, and a set of constraints that the variables must conform to.
@@
 \end{tabbing}
-The second group of variables is used to ensure that the clustering is acyclic -- that is, there must be an ordering to execute each clusters' dependency clusters before executing the cluster itself.
+The second group of variables is used to ensure that the clustering is acyclic.
 For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$.
@@
 If two nodes are to be fused together, their $\pi$ values must be the same.
-
-As an example of a cyclic clustering, consider:
-\begin{code}
-cycle us
- = let xs  = map (+1) us          (C1)
-       sum = fold xs              (C2)
-       zs  = map (+sum) xs        (C1)
-\end{code}
-There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
-If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
+\ben{I don't understand the ``acyclic'' part.
+If we don't add these constraints, then how might the clustering turn out to be ``cyclic''?}
 

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex

--- 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
@@
 \caption{Possible clusterings for \texttt{normalizeInc}}
+\label{f:normInc}
 \end{figure}
@@
 \end{code}
-\end{figure}
+% swapped A and B, so that the first option matches with the left graph, the second with the right graph in the diagram
+For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}.
+The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@).
+For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@.
+For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
 
-For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
+The @normalizeInc@ example demonstrates why it is, in general, not sufficient, to simply minimize the number of loops in a program: we have to assign weights to different types of loops and take those into account when choosing a possible clustering.
+Ordered typed fusion \cite{kennedy1993typedfusion} addresses this problem in the context of fusing imperative parallel and sequential loops, but also only in the absence of filter-like operations.
+Our problem can be directly mapped to an integer linear program with some integral, boolean and real variables, an objective function to minimise, and a set of contraints the variables must conform to.
+While finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, we can build on exisiting work here and use the best ILP-solver available, which tend to be sufficient for realistic problem sizes.
+For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still results in a very efficient, if not necessarily optimal program.
+
+%For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
 The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
 For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
-
-Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
-The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.
-
-In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
-Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
-The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
-This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
-Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.
 

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
Merge cleanup

--- 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
@@
 \end{figure}
-
-For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@).
-The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@).
-For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
+For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}.
+The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@).
+For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@.
+For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
 

### 2014-05-13 2014betterfusionforfilters/section/01-Introduction.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?
fill in some fusion refs in intro

--- 2014-05-12 2014betterfusionforfilters/section/01-Introduction.tex
@@
 It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance.
-This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{some fusion refs}, to name just a few.
+This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{coutts2007streamfusion, gao1993collective, kennedy1993typedfusion, kennedy2001fastgreedy, keller2010repa}, to name just a few.
 Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
@@
 To demonstrate the effect of clustering and fusion, consider the following code snippet:
+\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-The function @normalize@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero.
+\end{figure}
+The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero.
 Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops.

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
varied fixes

@@
 MMMMM       \= \kill
-$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
+$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
 \end{tabbing}
@@
 MMMMM       \= \kill
-$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
+$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
 \\
-MMMM             \= M \= \kill
+MM             \= M \= \kill
 $edge(bs, out = @fold@~f~in)$ \\
-                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
-                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
 $edge(bs, out = @map@~f~in)$  \\
-                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
-                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
 $edge(bs, out = @filter@~f~in)$  \\
-                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
-                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
 $edge(bs, out = @gather@~data~indices)$  \\
-                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
-                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
+    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
 $edge(bs, out = @cross@~a~b)$            \\
-                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
-                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
+    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
 $edge(bs, outs = @external@~ins)$  \\
-                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
+    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
 \\
@@
     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
-                     \> $=$ \> $(to, from, fusion~preventing)$  \\
-                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
-                     \> $=$ \> $(to, from, fusion~preventing)$  \\
+    \> $=$ \> $(to, from, \fusionpreventing)$  \\
+    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge from \in outs$     \\
+    \> $=$ \> $(to, outs, \fusionpreventing)$  \\
     \> $|$ \> $otherwise$                      \\
-                     \> $=$ \> $(to, from, fusible)$            \\
+    \> $=$ \> $(to, from, \fusible)$            \\
 \end{tabbing}
@@
 The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
-\ben{When we're talking about the dependency graph, when we say ``nodes can be fused'', it means that the operators that produce these values can be fused -- or that the values are produced by the same cluster of operators.}
+If two nodes are ``fused together'', it means that their combinators will be in the same cluster.
 
@@
 They are not required for correctness of the clustering, but are used in the objective function.
-A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i,j) \in E.
-x_{ij} = 0$.
-\ben{Using $\forall$ like this implies that the quantified $i$ isn't the same as the one attached to $c_i$.
-Write this with a list comprehension instead.}
+A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
 
@@
 Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
-The only difference is that fewer constraints and variables are required in the optimised version, but both versions are equivalent.
-\ben{What does ``equivalent'' mean when the constraints are different?} \ben{Say what ``optimised'' means in this case.} \ben{Give some hints about why the unoptimised version has more constraints.
-}
+The only difference is that fewer constraints and variables are required in the optimised version, but both versions give the same clustering.
+This unoptimised version generates clustering constraints and variables for every pair of nodes, regardless of whether they may, in fact, be fused together.
+Later, we show that certain constraints and variables can be removed when there is a fusion-preventing edge between two nodes.
 
@@
            \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \emptyset$)  \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
+           \\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
            \\
@@
 If two nodes have different iteration sizes, they may not necessarily be fused.
+If either of the nodes have no iteration size --- in which case, they are @external@ computations --- they may not be fused at all.
 However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.
@@
 An array is fully contracted if all its output edges are fused with it.
-Thus, $c_i=0$ only if $\forall j | (i,j) \in E.
-x_{ij} = 0$.
+Thus, $c_i=0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
 By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.
-\ben{Say what array contraction is, and why we want this constraint.}
+
+If an array is fully contracted, it means that the array does not need to be materialised in memory, and can simply be replaced by a scalar variable for the current element.
+Contracting arrays reduces memory traffic and can reduce overall memory usage.
 
@@
 
-\subsection{A note on transitivity}
+\subsubsection{A note on transitivity}
 It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
@@
 \label{s:OptimisedConstraints}
-Alternatively:
-we can say, only generate constraints if there is no \emph{fusion preventing path} between $i$ and $j$.
-Here, we generate fewer constraints than originally, but still more than above.
-We also must preprocess the graph to find such blocking paths.
-But we still get the minimisation to count non-connected nodes.
+We can remove certain constraints and variables, by only checking nodes if there is no \emph{fusion-preventing path} between $i$ and $j$.
+If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
+We can simplify the integer linear program, and make the job of the ILP solver easier, by ommitting these constraints.
 
@@
     \>$=$       \>$\{v' | v' \in vs$                                \\
-    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
-    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
+    \>$\wedge$  \>$\forall p \in path(v,v') \cup path(v',v).\ \fusionpreventing \not\in p\}$\\
 \end{tabbing}
 \end{figure}
-
-We can simplify the version above and remove many constraints and variables by noting that if there is a fusion-preventing edge between $i$ and $j$, $x_{ij} = 1$, so the precedence constraint can be simplified to just $\pi_i < \pi_j$, and the $x_{ij}$ variable removed.
-Likewise with clusters of different, incompatible types.
-The contraction variable $c_i$ can be removed, if $i$ has any fusion-preventing output edges: any such edges use the array and cannot be fused, thus there is no possibility of the array being contracted away.
 
@@
 \\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
+           \\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
+           \\
            \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
            \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
+           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
+           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
 \end{tabbing}

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
fix up opt version

@@
            \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$) \\
            \> \ldots
@@
 If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
-We can simplify the integer linear program, and make the job of the ILP solver easier, by ommitting these constraints.
+We can simplify the integer linear program and make the job of the ILP solver easier, by omitting these constraints.
+
+For the graph, we define a function $split(a,b)$ to check whether there exists a fusion-preventing path between two nodes, $a$ and $b$.
 
@@
 \begin{tabbing}
-MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
-$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
-MMMM        \= M    \= \kill
-$split(vs,es)$ \\
-    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
-$clusterable(v,vs,es)$  \\
-    \>$=$       \>$\{v' | v' \in vs$                                \\
-    \>$\wedge$  \>$\forall p \in path(v,v') \cup path(v',v).\ \fusionpreventing \not\in p\}$\\
+MMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
+$split$     \> @::@ \> $name \times name \to \mathbb{B}$      \\
+MM        \= M    \= \kill
+$split(a,b)$ \\
+    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$\\
 \end{tabbing}
 \end{figure}
+
+With $split$ defined, we can refine our formulation to only generate constraints between two nodes if they may be fused together.
 
@@
 Minimise   \> \ldots \\
-Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (no edge from i to j, but there is a fusion benefit)            \\
+Subject to \\
+           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
+           \> (if $split(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
+\\
            \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (an edge from i to j)            \\
+           \> (if $split(i,j) \wedge (i,j,\fusible) \in E$)     \\
+\\
            \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
-           \>             (a \emph{fusion-preventing} edge from i to j)            \\
+           \> (if $split(i,j) \wedge (i,j,\fusionpreventing) \in E$)    \\
+\\
 \\
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (for all edges from i with no fusion-preventing outputs)      \\
+           \> (if $split(i,j)$) \\
+           \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
+           \> (if $\neg(split(i,j))$) \\
 \\
@@
            \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\} \wedge$ \\
+
+           \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
 \end{tabbing}

### 2014-05-14 2014betterfusionforfilters/section/03-SizeInference.tex
transducer changes
varied fixes
remove those damn figure[H] because they were reordering things strangely

--- 2014-05-12 2014betterfusionforfilters/section/03-SizeInference.tex
@@
 \end{alltt}
-
 However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process.
@@
 \end{code}
-
 The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array.
@@
 \begin{tabbing}
-MMMM \= M \= MMMMMMM \= M \= \kill
-$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
-$parents(a,b)$ \\
-        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
-        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
-        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
+MMMM \= M \= M \= M \= \kill
+$parents$ \> $::$ \> $binds \to name \to name \to \{name \times name\}$ \\
+$parents(bs, a, b)$ \\
+        \> $|$ \> $\iiter_{\Gamma,C}(bs(a)) == \iiter_{\Gamma,C}(bs(b))$ \\
+        \>     \>                      \> $=$ \> $\{(a, b)\}$ \\
+        \> $|$ \> otherwise            \\
+        \>     \>                      \> $=$    \> $\{ parents(bs, a', b) ~|~ a' \in trans(bs, a) \} $      \\
+        \>     \>                      \> $\cup$ \> $\{ parents(bs, a, b') ~|~ b' \in trans(bs, b) \} $  \\
 \end{tabbing}
@@
 \[
-valid(\constr{bs}) \implies \forall n.\ |trans(bs, n)| \le 1
+\forall bs, n.\ |trans(bs, n)| \le 1
 \]
@@
 
-\textbf{Lemma: transducers change types}.
-For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
-\[
-valid(\constr{bs}) \wedge \{m\} = trans(bs, n) \implies \tau(n) \not= \tau(m)
-\]
+% this is only useful for proving sole parents
+% \textbf{Lemma: transducers change types}.
+% For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
+% \[
+% \forall bs, n, m.\ trans(bs, n) = \{m\} \implies \tau(n) \not= \tau(m)
+% \]
 
@@
 \[
-valid(\constr{bs}) \implies \forall a,b.\ |parents(bs, n)| \le 1
+\forall bs, a, b.\ |parents(bs, a, b)| \le 1
 \]

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
remove those damn figure[H] because they were reordering things strangely

@@
 
-\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-\end{figure}
 For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}.
@@
 
-\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
-\end{figure}
 Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
@@
 \end{tabbing}
-\end{figure}
 
@@
 As an example of a cyclic clustering, consider:
-\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-\end{figure}
 There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
@@
 \end{tabbing}
-\end{figure}
 As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.
@@
 \paragraph{Fusion-preventing edges}
-\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
-\end{figure}
 As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
@@
 \end{tabbing}
-\end{figure}
 If two nodes have different iteration sizes, they may not necessarily be fused.
@@
 As a degenerate example, consider fusing an operation on filtered data with its generating filter:
-\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-\end{figure}
 Here, $parents(gts,sum2) = \{gts, gts\}$.
@@
 \paragraph{Array contraction}
-\begin{figure}[H]
 \begin{tabbing}
@@
 \end{tabbing}
-\end{figure}
 As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
add obj function to final, optimised version

@@
 If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
-We can simplify the integer linear program and make the job of the ILP solver easier, by omitting these constraints.
+With our specific knowledge of the problem domain, we can simplify the integer linear program and make the job of the ILP solver easier by omitting these constraints.
+Creating fewer constraints and variables should make it faster for the ILP solver to find a solution.
 
@@
 
-With $split$ defined, we can refine our formulation to only generate constraints between two nodes if they may be fused together.
+With $split$ defined, we can refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
 
@@
 MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> \ldots \\
+Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+           \> (if $split(i,j)$)         \\
+           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
 Subject to \\
@@
            \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
+Where      \>                                       \\
+           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
+           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
+           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
+           \> \> \> ($i$ and $j$ share an input array)                                         \\
+           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
+           \> \> \> (the only benefit is loop overhead)                                        \\
+           \\
+           \> $N = |V|$
 \end{tabbing}

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
rename normalize and normalise to normaliZe2
entire formulation

@@
 \hline
-Normalise2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
+Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
 Closest pair    & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
@@
 % \item
-% Normalise2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
+% Normalize2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
 % \item
@@
 \item
-In some cases, the clusterings were the same for different methods. In this case, the results are the same. 
+In some cases, the clusterings generated by the different methods were the same. In this case, the results are also the same. 
 \item
@@
 
-\subsection{Closest pairs}
-Closest pairs makes use of fast-ish @median@ to ensure balanced division of work.
+\subsection{Example}
+To take the normalize2 example, and convert it to combinator normal form:
 \begin{code}
-closest :: Vec Pt -> (Pt,Pt)
-closest pts
- | length pts < 250
- = naive pts
- | otherwise
- = divide pts
-
-divide :: Vec Pt -> (Pt,Pt)
-divide pts
- = let p      = median pts
-       aboves = filter (above p) pts
-       belows = filter (below p) pts
-       above' = closest aboves
-       below' = closest belows
-
-       border = min (distance above') (distance below')
-
-       aboveB = filter (above (p - border)) pts
-       belowB = filter (below (p + border)) pts
-
-       cs     = cross aboveB belowB
-       bord   = minBy distance cs
-   in  above' `minDist` below' `minDist` bord
-
-naive :: Vec Pt -> (Pt,Pt)
-naive pts
- = let c = cross pts pts
-   in  minBy distance c
-
-minBy :: Ord b => (a -> b) -> Vec a -> a
-minBy f xs
- = fold (min . f...) ... xs
+ normalize2 :: Array Int -> Array Int
+ normalize2 xs
+  = let sum1 = fold   (+)  0   xs
+        gts  = filter (> 0)    xs
+        sum2 = fold   (+)  0   gts
+        ys1  = map    (/ sum1) xs
+        ys2  = map    (/ sum2) xs
+    in (ys1, ys2)
 \end{code}
 
-Let's translate @divide@ to a program in CNF.
+Split:
 \begin{code}
-divide :: Vec Pt -> (Pt,Pt)
-divide pts
- = let p      = external pts
+split(sum1, gts)    = T
+split(sum1, sum2)   = T
+split(sum1, ys2)    = T
 
-       aboves = filter (... p) pts          -- A
-       belows = filter (... p) pts          -- A
+split(gts, sum2)    = T
+split(gts, ys1)     = T
 
-       above' = external aboves
-       below' = external belows
-       border = external above' below'
+split(sum2, ys1)    = T
 
-       aboveB = filter (... p border) pts   -- C
-       belowB = filter (... p border) pts   -- B
+split(ys1, ys2)     = T
+\end{code}
 
-       cs     = cross  aboveB belowB        -- C
-       bord   = fold   (...) cs             -- C
-
-       min'   = external above' below' bord
-   in  min'
-\end{code}
-where @A@, @B@ and @C@ are distinct clusters.
-
-What happens if we do this using stream fusion?
-\begin{code}
-divide :: Vec Pt -> (Pt,Pt)
-divide pts
- = let p      = external pts
-
-       aboves = filter (... p) pts          -- A
-       belows = filter (... p) pts          -- B
-
-       above' = external aboves
-       below' = external belows
-       border = external above' below'
-
-       aboveB = filter (... p border) pts   -- D
-       belowB = filter (... p border) pts   -- C
-
-       cs     = cross  aboveB belowB        -- D
-       bord   = fold   (...) cs             -- D
-
-       min'   = external above' below' bord
-   in  min'
-\end{code}
-So, we have four clusters instead of three, and the same number of manifest arrays. Not particularly impressive.
+\begin{tabbing}
+MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+           \> (if $split(i,j)$)         \\
+           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
+Subject to \\
+    \> $-N x_{sum1, gts}$
+\end{tabbing}
 
@@
 \end{code}
-Yep. @hull@ is pretty boring.
-
-Something with a \emph{fold}, and then filtering or mapping based on that fold would be good.
-Or something with
-\begin{code}
-let xs' = filter f    xs
-    i   = fold   g    xs'
-    xs''= map   (h i) xs'
-\end{code}
-would be really good.
-
-FFT - not really.
-
-\subsection{QuadTree}
-\begin{code}
-bounds pts
- = let x1 = fold minX pts
-       y1 = fold minY pts
-       x2 = fold maxX pts
-       y2 = fold maxY pts
-   in ((x1,y1), (x2,y2))
-
-%splitbox ((x1,y1), (x2,y2))
-% = let xm = mid x1 x2
-%       ym = mid y1 y2
-%   in ( ((x1, y1), (xm, ym))
-%      , ((x1, ym), (xm, y2))
-%      , ((xm, y1), (x2, ym))
-%      , ((xm, ym), (x2, y2)))
-%
-quadtree pts
- = go (bounds pts)
- where
-  go bounds pts
-   | length pts > 0
-   = let (b1,b2,b3,b4) = splitbox bounds
-         pts1  = filter (inbox b1) pts
-         pts2  = filter (inbox b2) pts
-         pts3  = filter (inbox b3) pts
-         pts4  = filter (inbox b4) pts
-         tree1 = go b1 pts1
-         tree2 = go b2 pts2
-         tree3 = go b3 pts3
-         tree4 = go b4 pts4
-      in Tree tree1 tree2 tree3 tree4
-   | otherwise
-   = Empty
-\end{code}
-It's easy to see that @bounds@ should only require one loop, but stream fusion requires four, as there is no inlining that can occur.
-The same is true of @go@ in @quadtree@. Our implementation only requires one loop for each of these.

### 2014-05-14 2014betterfusionforfilters/section/04-ILP.tex
entire formulation
example normalize2 ilp program

@@
 With $split$ defined, we can refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
+The entire formulation of the integer linear program follows.
 
@@
 MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij} + \Sigma_{i \in V} N c_i$  \\
            \> (if $split(i,j)$)         \\
-           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
 Subject to \\
@@
            \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
-           \> (if $split(i,j) \wedge (i,j,\fusionpreventing) \in E$)    \\
+           \> (if $(i,j,\fusionpreventing) \in E$)    \\
 \\
@@
            \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
+MMMMM   \= MMMM \= M \= \kill
 Where      \>                                       \\

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
example normalize2 ilp program

@@
 \subsection{Example}
-To take the normalize2 example, and convert it to combinator normal form:
+The normalize2 example is already in combinator normal form.
 \begin{code}
@@
 
-Split:
+First, we must calculate $split$ -- that is, those nodes which have no fusion-preventing path between them.
 \begin{code}
@@
 split(sum1, ys2)    = T
-
 split(gts,  sum2)   = T
 split(gts,  ys1)    = T
-
 split(sum2, ys1)    = T
-
 split(ys1,  ys2)    = T
 \end{code}
+We only need to create cluster variables for these.
+
+Note that, in the objective function, the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1.
+This is because they do not share any input arrays.
 
 \begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
-           \> (if $split(i,j)$)         \\
-           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
+MMMMM   \= MMMMMM \= M \= MMMMMM \= M \= MMMMMM \= \kill
+Minimise   \> $25 x_{sum1, gts} + 1 x_{sum1,sum2} + 25 x_{sum1, ys2} + 25 x_{gts, sum2} +$ \\
+           \> $25 x_{gts, ys1} + 1 x_{sum2, ys1} + 25 x_{ys1, ys2} + $ \\
+           \> $ 5 c_{gts} + 5 c_{ys1} + 5 c_{ys2} $\\
 Subject to \\
-    \> $-N x_{sum1, gts}$
+    \> $-5 x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 x_{sum1, gts}$  \\
+    \> $-5 x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 x_{sum1, sum2}$ \\
+    \> $-5 x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 x_{sum1, ys2 }$ \\
+    \> $-5 x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 x_{gts, ys1  }$ \\
+    \> $-5 x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 x_{sum2, ys1 }$ \\
+    \> $-5 x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 x_{ys1, ys2  }$ \\
+\\
+    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 x_{gts, sum2 }$ \\
+\\
+    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
+    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$ \\
+\\
+    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$ \\
+\\
+    \> $x_{sum1,sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
+    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$ \\
+    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$ \\
 \end{tabbing}
+$parents(sum1,sum2)  = \{(sum1, gts)\}$
+
+A minimal solution to this is
+\begin{tabbing}
+MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
+$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}$
+    \> $=$ \> $0$ \\
+$x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
+    \> $=$ \> $1$ \\
+\\
+$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
+    \> $=$ \> $0$ \\
+$\pi_{ys1 }, \pi_{ys2 }$
+    \> $=$ \> $1$ \\
+\\
+$c_{gts}, c_{ys1}, c_{ys2}$           
+    \> $=$ \> $0$ \\
+\end{tabbing}
+Looking at just the non-zero variables in the objective function,
+$25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
+It is worth noting that, for example, the objective could be reduced by fusing $x_{sum1,ys2}$ together.
+However, this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, the constraint requires that $\pi_{sum1} = \pi_{sum2}$, and another constraint requires $\pi_{sum2} < \pi_{ys2}$.
+These constraints may not all hold, so a clustering that fused sum1 and ys2 together would not allow sum1 and sum2 to be fused together.
 

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
talk about stream fusion clustering in eg of normalize2 program

@@
 
-Note that, in the objective function, the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1.
-This is because they do not share any input arrays.
+Note that, in the objective function, the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.
 
@@
 MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}$
+$x_{sum1, gts}, x_{sum1, sum2}, x_{gts, sum2}, x_{ys1,  ys2}$
     \> $=$ \> $0$ \\
@@
 \end{tabbing}
-Looking at just the non-zero variables in the objective function,
+Looking at just the non-zero variables in the objective function, the value is
 $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
@@
 These constraints may not all hold, so a clustering that fused sum1 and ys2 together would not allow sum1 and sum2 to be fused together.
+
+We can also calculate the objective function for the clustering used by stream fusion, to contrast and show that the constraints are valid.
+To do this, the values of $x_{ij}$ variables are:
+\begin{tabbing}
+MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
+$x_{gts, sum2}$
+    \> $=$ \> $0$ \\
+$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
+    \> $=$ \> $1$ \\
+\end{tabbing}
+If we add these extra equality constraints to the integer linear program and solve it, one possible solution is:
+\begin{tabbing}
+MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
+$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
+    \> $=$ \> $0$ \\
+$\pi_{ys1 }, \pi_{ys2 }$
+    \> $=$ \> $1$ \\
+\\
+$c_{gts}, c_{ys1}, c_{ys2}$           
+    \> $=$ \> $0$ \\
+\end{tabbing}
+It is interesting to note that two nodes having equal $\pi_i$ values does not imply that the two nodes are fused together.
+Conversely, however, different $\pi_i$ values do imply that two nodes are not fused together.
+
+The objective function for this solution is
+$25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
 

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
talk about stream fusion clustering in eg of normalize2 program

@@
 MMMMM   \= MMMMMM \= M \= MMMMMM \= M \= MMMMMM \= \kill
-Minimise   \> $25 x_{sum1, gts} + 1 x_{sum1,sum2} + 25 x_{sum1, ys2} + 25 x_{gts, sum2} +$ \\
-           \> $25 x_{gts, ys1} + 1 x_{sum2, ys1} + 25 x_{ys1, ys2} + $ \\
-           \> $ 5 c_{gts} + 5 c_{ys1} + 5 c_{ys2} $\\
+Minimise   \> $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, sum2} +$ \\
+           \> $25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} + $ \\
+           \> $ 5 \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
 Subject to \\

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
add dots between coefficients. just slightly easier to read

--- 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
@@
 \begin{tabbing}
-MMMMM   \= MMMMMM \= M \= MMMMMM \= M \= MMMMMM \= \kill
-Minimise   \> $25 x_{sum1, gts} + 1 x_{sum1,sum2} + 25 x_{sum1, ys2} + 25 x_{gts, sum2} +$ \\
-           \> $25 x_{gts, ys1} + 1 x_{sum2, ys1} + 25 x_{ys1, ys2} + $ \\
-           \> $ 5 c_{gts} + 5 c_{ys1} + 5 c_{ys2} $\\
+MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
+Minimise   \> $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
+           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
+           \> $ 25 \cdot x_{ys1, ys2} + 5 \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
 Subject to \\
-    \> $-5 x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 x_{sum1, gts}$  \\
-    \> $-5 x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 x_{sum1, sum2}$ \\
-    \> $-5 x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 x_{sum1, ys2 }$ \\
-    \> $-5 x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 x_{gts, ys1  }$ \\
-    \> $-5 x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 x_{sum2, ys1 }$ \\
-    \> $-5 x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 x_{ys1, ys2  }$ \\
+    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
+    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
+    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
+    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
+    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
+    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ \\
 \\
-    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 x_{gts, sum2 }$ \\
+    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ \\
 \\

### 2014-05-14 2014betterfusionforfilters/section/05-Benchmarks.tex
explanation of quickhull clustering

@@
 
-\subsection{Quickhull}
-Is Quickhull any better? Seems like it's just one cluster; @filterMax@. And we don't have append (@++@), anyway.
+\subsection{QuickHull}
+The core of the QuickHull algorithm is very similar to @hull@, below.
+Given a line and an array of points, the points are filtered to those above the line,
+and the farthest point above the line is also found.
+
 \begin{code}
-quickhull :: Vec Pt -> Vec Pt
-quickhull pts
- = let top  = fold getTop pts
-       bot  = fold getBot pts
-       tops = hull (top,bot) pts
-       bots = hull (bot,top) pts
-   in  tops ++ bots
-
-hull :: (Pt,Pt) -> Vec Pt -> Vec Pt
+hull :: (Point,Point) -> Array Point -> Array Point
 hull line@(l,r) pts
@@
        ma   = fold   (maxFrom line) pts'
-       hl   = hull   (l, ma)        as
-       hr   = hull   (ma, r)        as
+       hl   = hull   (l, ma)        pts'
+       hr   = hull   (ma, r)        pts'
    in  hl  ++ hr
 \end{code}
+
+Stream fusion is unable to fuse the @pts'@ and @ma@ together, as @pts'@ is used multiple times and cannot be inlined into @ma@.
+Similarly, Megiddo~\cite{megiddo1998optimal} is unable to fuse them together, as their iteration sizes are different.
+Our formulation is able to fuse these together.
+
+If the @fold@ is changed to operate over @pts@ instead of @pts'@, Megiddo's formulation \emph{is} able to fuse the two.
+Interestingly, a programmer unaware of fusion would expect the original version to be faster, because @pts'@ is smaller than @pts@.
+
+\begin{code}
+ = let pts' = filter (above   line) pts
+       ma   = fold   (maxFrom line) pts
+\end{code}

### 2014-05-15 2014betterfusionforfilters/Main.tex
remove manuel as author; add to acknowledgements

@@
   \and Ben Lippmeier$^\dagger$
-  \and Manuel M. T. Chakravarty$^\dagger$
   \and Gabriele Keller$^\dagger$ 
@@
     University of New South Wales, Australia \\[2pt]
-    \textsf{\{amosr,benl,chak,keller\}@cse.unsw.edu.au}
+    \textsf{\{amosr,benl,keller\}@cse.unsw.edu.au}
   }
@@
 Many thanks are due to
+Manuel Chakravarty,
 Robert Clifton-Everest,
@@
 Bill Kroon,
-Frederik Madsen,
+Frederik M.\ Madsen,
 Abdallah Saffidine,

### 2014-05-15 2014betterfusionforfilters/section/02-CNF.tex
Change caption back
Typos, grammatical errors, clarifications

--- 2014-05-10 2014betterfusionforfilters/section/02-CNF.tex
@@
 \label{s:CombinatorNormalForm}
-Each input program is expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of its data flow graph.
+Input programs are expressed in \emph{Combinator Normal Form} (CNF), which is a textual description of the data flow graph.
 The grammar for CNF is in Figure~\ref{f:CombinatorNormalForm}.
-Both the @normalize@ and @normalize2@ examples on the previous page are in CNF, and the matching data flow graph for @normalize2@ is in Figure~\ref{f:normalize2-clusterings}.
+The @normalize2@  example on the previous page is in CNF,  as is the matching data flow graph for @normalize2@ in Figure~\ref{f:normalize2-clusterings}.
 Our data flow graphs are similar to Loop Communication Graphs (LCGs) from related work in imperative array fusion~\cite{gao1993collective}.
 We name edges after the corresponding variable from the CNF form, and edges which are fusion preventing are drawn with a dash through them (as per the edge labeled @sum1@ in Figure~\ref{f:normalize2-clusterings}).
-In data flow graphs we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.
+In data flow graphs, we tend to elide the worker functions to combinators when they are not important to the discussion --- so we don't show the @(+)@ operator on each use of @fold@.
 
 Clusters of operators that are fused into single imperative loops are indicated by dotted lines, and we highlight materialized arrays by drawing them in boxes.
-In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes as these are the material input and output arrays of the program.
-However, in the graph on the far right hand side @gts@ has also been materialized because in this version the producing and consuming operators (@filter@ and @fold@) have not been fused.
-In Figure~\ref{f:CombinatorNormalForm}, note that the bindings have been split those that produce scalar values ($sbind$), and those that produce array values ($abind$), and these groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
+In Figure~\ref{f:normalize2-clusterings}, the variables @xs@, @ys1@ and @ys2@ are always in boxes, as these are the material input and output arrays of the program.
+However, in the graph on the far right hand side, @gts@ has also been materialized because in this version, the producing and consuming operators (@filter@ and @fold@) have not been fused.
+In Figure~\ref{f:CombinatorNormalForm}, note that the bindings have been split into those that produce scalar values ($sbind$), and those that produce array values ($abind$).
+These groupings are represented as open and closed arrow-heads in Figure~\ref{f:normalize2-clusterings}.
 
+%What's a 'suggestive' type???
 Most of our array combinators are standard, and suggestive types are given at the bottom of Figure~\ref{f:CombinatorNormalForm}.
 The $@map@_n$ combinator takes a worker function, $n$ arrays of the same length, and applies the worker function to all elements at the same index.
-As such it is similar to Haskell's @zipWith@, with an added length restriction on the argument arrays.
+As such, it is similar to Haskell's @zipWith@, with an added length restriction on the argument arrays.
 The @generate@ combinator takes an array length and a worker function, and creates a new array by applying the worker to each index.
-The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that were at each index.
-In Haskell this would be @gather arr ixs = map (index arr) ixs@.
+The @gather@ combinator takes an array of elements, an array of indices, and produces the array of elements that are positioned at each index.
+In Haskell, this would be @gather arr ixs = map (index arr) ixs@.
 The @cross@ combinator returns the cartesian product of two arrays.
@@
 We assume workers are pure, can at least compute arithmetic functions of their scalar arguments, and index into arrays in the environment.
-We also assume that each CNF program considered for fusion is embedded in a larger ``host program'' which handles file IO and the like.
+We also assume that each CNF program considered for fusion is embedded in a larger host program which handles file IO and the like.
 Workers are additionally restricted so they can only directly reference the \emph{scalar} variables bound by the local CNF program, though they may reference array variables bound by the host program.

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
Wording. Replaced ref to normalise with normalize2

@@
 
-A size may either be a variable $k$, or a cross product of two sizes.
+A size may either be a variable $k$ or a cross product of two sizes.
 We use the latter to represent the result size of the @cross@ operator discussed in the previous section.
-Constraints maybe either be trivially $true$, an equality $\tau = \tau$, or a conjunction of two constraints $C \wedge C$.
+Constraints maybe either be trivially $true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$.
 We refer to the trivially true and equality constraints as \emph{atomic constraints}.
 Size schemes relate the sizes of each input and output array.
-For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
-$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
+$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
 $$
@@
 Note that the constraints are generated in a specific form, to make the constraint solving process easy.
-For each array variable in the program we generate a new size variable, like size $k_{zs}$ for array variable $zs$.
+For each array variable in the program, we generate a new size variable, like size $k_{zs}$ for array variable $zs$.
 These new size variables always appear on the \emph{left} of atomic equality constraints.
@@
 
-For example, the final environment and constraints generated for the @normalize@ example from Section~\ref{s:Introduction} are as follows:
+For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
 $$
@@
 In contrast, the rigid variables marked by the $\exists$ symbol cannot.
-For example, if we consider the constraints for @normalize@ mentioned before:
+For example, if we consider the constraints for @normalize2@ mentioned before:
 $$
@@
 In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
-$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
+$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
 $$
@@
 
-Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
+Looking back at the @normalize2@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
 %The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 

### 2014-05-15 2014betterfusionforfilters/Main.tex
abstract wording

@@
 Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code.
-To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation, we also have to also \emph{cluster} separate traversals of the same array into one single loop.
+To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation; we also have to \emph{cluster} separate traversals of the same array into one single loop.
 Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
-However, these approaches have one serious restrictions: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
-The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP server to find good clusterings.
+However, these approaches have one serious restriction: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
+The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
 

### 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
remove those damn figure[H] because they were reordering things strangely
minor change to contribution

@@
 To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
-It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) it fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
 
@@
 To demonstrate the effect of clustering and fusion, consider the following code snippet:
-\begin{figure}[H]
 \begin{code}
@@
 \end{code}
-\end{figure}
 The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero.
@@
 These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations.
-Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@.
+Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize2@.
 In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold.
 The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations.
-Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops.
+Our approach (diagram in the middle) produces, in this case, the optimal solution: two loops.
 The first to calculate both sums, the second to calculate the maps.
@@
 We extend prior work by Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, with support for size changing operators.
-Size changing operators can be clustered with the operators that generate their source arrays, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
+Size changing operators can be clustered with operations on both their source array and output array, and compiled naturally with data-flow fusion (\S\ref{s:ILP}).
 

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
Size inference clarifications

@@
 \subsection{Rigid Sizes}
-When the environment of our size constraints contains rigid variables (indicate by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme.
-Consider the @filterLeft@ function from S~\ref{s:SizeTypes}
+When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme.
+Consider the @filterLeft@ function from Section~\ref{s:SizeTypes}
 \begin{code}
@@
 
-Note that although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule.
+Note that, although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule.
 The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another.
@@
 \end{code}
-The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array.
+The above function filters its input array, and then applies @map2@ to the filtered version as well as the original array.
 As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true.
@@
 
-In this case Rule~(SFun) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).
+In this case, Rule~(SFun) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).
 
@@
 
-To solve these, we note that $k_{as}$ is used twice on the left of an atomic equality constraint, so substitute $k_1$ for $k_3$:
+To solve these, we note that $k_{as}$ is used twice on the left of an atomic equality constraint, so we substitute $k_1$ for $k_3$:
 $$
@@
 
-At this stage we are stuck because the constraints are not yet in solved form, and we cannot simplify them further.
+At this stage we are stuck, because the constraints are not yet in solved form, and we cannot simplify them further.
 Both $k_1$ and $k_2$ are marked as rigid, so we cannot substitute one for the other and produce a single atomic constraint for $k_{bs}$.
@@
 \subsection{Iteration size}
-After the inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator.
-For @filter@ and other size changing operators, the iteration and result sizes are necessarily different.
+After inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator.
+For @filter@ and other size changing operators, the iteration and result sizes are in general different.
 For such an operator, we say that the result size is a \emph{descendant} of the iteration size.
@@
 \subsection{Transducers}
-Unlike previous work, we do allow combinators with different iteration sizes to be fused together.
+%     we've said that several times before
+% Unlike previous work, we do allow combinators with different iteration sizes to be fused together.
 For example, an operation on filtered data may be fused with the filter operation that generates the data, even though the iteration sizes are different.
@@
 The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
-Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
+Even though in general $k_{gts}$ is not equal to $k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
 The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
@@
 We now define a function $trans$, to find the parent transducer of a given combinator.
-The parent transducer $trans(bs, n)$ of a combinator has the same output size as $n$'s iteration size, but the two have different iteration sizes.
+The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.
 
@@
 \textbf{Lemma: sole transducers}.
-For some bindings $bs$, each name $n$ will have at most one transducer.
+For any set of bindings $bs$, each name $n$ will have at most one transducer.
 \[
@@
 \textbf{Lemma: sole parents}.
-For some bindings $bs$, each pair of names $a$ and $b$ will have at most one set of parents $parents(a,b)$.
+% I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
+For any set of bindings $bs$, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
 \[

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
conflict resolve

@@
 
+Two combinators $a$ and $b$ may be fused together only if they have parents $(c, d) \in parents(a,b)$, and the combinators and their parents are also fused together.
+That is, in order for $a$ and $b$ to be fused together, $c$ and $d$ must be fused, $a$ and $c$ must be fused, and $d$ and $b$ must be fused.
+
 \textbf{Lemma: sole transducers}.
-For any set of bindings $bs$, each name $n$ will have at most one transducer.
+For some function $f$ with valid constraints, each name $n$ will have at most one transducer.
 \[
-\forall bs, n.\ |trans(bs, n)| \le 1
+\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
 \]
@@
 % I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
-For any set of bindings $bs$, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
+For some bindings $f$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
 \[
-\forall bs, a, b.\ |parents(bs, a, b)| \le 1
+\forall f, \sigma, a, b.\ f :_s \sigma \implies |parents(binds(f), a, b)| \le 1
 \]
 % \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.
+
+These two lemmas are used in the integer linear programming formulation, when constructing the program.
+When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
transducers

@@
 We define the concept of \emph{transducers} as combinators having a different output size to their iteration size.
-As usual, a transducer may fuse with other nodes of the same iteration size, but transducers may also fuse with nodes having iteration size the same as the transducer's output size.
+As with any other combinator, a transducer may fuse with other nodes of the same iteration size, but transducers may also fuse with nodes having iteration size the same as the transducer's output size.
 For our set of combinators, the only transducer is @filter@.
@@
 The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
-Even though in general $k_{gts}$ is not equal to $k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
+Even though $k_{gts}$ is not equal to $k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
 The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
@@
 
-Two combinators $a$ and $b$ may be fused together only if they have parents $(c, d) \in parents(a,b)$, and the combinators and their parents are also fused together.
+Two combinators $a$ and $b$ of different size may be fused together only if they have parents $(c, d) \in parents(a,b)$, and the combinators and their parents are also fused together.
 That is, in order for $a$ and $b$ to be fused together, $c$ and $d$ must be fused, $a$ and $c$ must be fused, and $d$ and $b$ must be fused.
+In the previous example, @sum1@ and @sum2@ have different iteration size, and their parents are $parents(@sum1@, @sum2@) = \{(@sum1@, @gts@)\})$.
+In order for @sum1@ and @sum2@ to be fused together, @sum1@ and @gts@ must be fused, @sum1@ and @sum1@ must be fused, and @gts@ and @sum2@ must be fused.
 

### 2014-05-15 2014betterfusionforfilters/section/05-Benchmarks.tex
benchmark

@@
 \label{s:Benchmarks}
-\ben{Use the larger programs as benchmarks, or as running examples. We only need small, simple programs to demonstrate how the algorithm works.}
 
@@
 
-Notes:
-\begin{itemize}
-% \item
-% Normalize2 tested with around 1gb input data. This should be small enough that all intermediates fit in memory, but large enough not to fit in cache.
-% \item
-% ClosestPoints only tested with 80mb data.
-\item
-In some cases, the clusterings generated by the different methods were the same. In this case, the results are also the same. 
-\item
-The stream fusion benchmarks use the clustering that stream fusion \emph{would} use, but hand-fused and written in C.
-\item
-Programs were run five times with the same input data, and the fastest run was used.
-\item
-ILP solutions were created by hand-fusing based on the clustering of the implementation. In the future, this will be integrated to use data flow fusion.
-\item
+We have run three benchmarks to constrast our clustering algorithm, Megiddo~\cite{megiddo1998optimal}'s clustering, and the clustering produced by stream fusion\cite{coutts2007streamfusion}.
+The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree.
+The benchmark programs are all hand-written, hand-fused C code based on the clustering.
+Each program was ran with the same input five times, and the minimum runtime was used.
+
 Benchmark programs are available at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}
-\end{itemize}
 
-\subsection{Example}
+\subsection{Normalize2}
 The normalize2 example is already in combinator normal form.

### 2014-05-15 2014betterfusionforfilters/section/06-Related.tex
Fix up section on rigid sizes
Cut back related work section

@@
 \section{Related work}
+The idea of using integer linear programming to cluster an operator graph for array fusion was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}~(1999).
+This algorithm was then extended to support multiple iteration sizes by Darte and Huard~\cite{darte2002contraction}.
+Both algorithms were developed in the context of imperative languages (Fortran) and are based around a Loop Dependence Graph (LDG).
+In a LDG the nodes represent imperative loops, and the edges indicate which loops may or may not be fused.
+Although this work was developed in a context of imperative programming, the conceptual framework and algorithms are language agnostic.
+In earlier work, Chatterjee~\cite{chatterjee1993nested} (1991) mentioned that ILP can be used to schedule a data flow graph, though did not give a complete formulation.
+Our system extends the prior ILP approaches with support for size changing operators such as @filter@.
 
-\subsection{Haskell short-cut fusion}
-Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, mainland2013exploiting}, tend to cleverly reuse compiler optimisations such as inlining and rewrite rules\cite{peytonjones2001rules}, to fuse combinators without having to modify the compiler itself.
-This approach has the advantage of simplicity, but is inherently limited in the amount of fusion it can perform.
+In the loop fusion literature, the ILP approach is considered ``optimal'' because it can find the clustering that minimises a global cost metric.
+In our case the metric is defined by the objective function of~\S\ref{s:ObjectiveFunction}.
+Besides optimal algorithms, there are also heuristic approaches.
+For example, Gao, Olsen and Sarkar~\cite{gao1993collective} use the maxflow-mincut algorithm to try to maximise the number of fused edges in the LDG.
+Kennedy~\cite{kennedy2001fastgreedy} describes another greedy approach which tries to maximize the reuse of intermediate arrays, and Song~\cite{song2004improving} tries to reduce memory references.
 
-Consider the following @filterMax@ function, where each element in the input array is incremented, then the maximum is found, and the array is filtered to those greater than zero.
-In this case, the result of the map @vs'@ cannot be inlined into both occurences without duplicating work, so no fusion can be performed.
-This has the effect of performing three loops instead of one, with two arrays instead of one.
+Darte~\cite{darte1999complexity} formalises the algorithmic complexity of various loop fusion problems and shows that globally minimising most useful cost metrics is NP-complete.
+Our ILP formulation itself is NP-hard, though in practice we have not yet found this to be a problem.
 
-\begin{code}
-filterMax vs =
- let vs' = map    (+1)  vs
-     m   = fold   max 0 vs'
-     flt = filter (>0)  vs'
-\end{code}
+Recent literature on array fusion for imperative languages largely focuses on the polyhedral model.
+This is an algebraic representation imperative loop nests and transformations on them, including fusion transformations.
+Polyhedral systems \cite{pouchet2011polyhedral} are able to express \emph{all possible} distinct loop transformations where the array indices, conditionals and loop bounds are affine functions of the surrounding loop indices.
+However, the polyhedral model is not applicable to (or intended for) one dimentional filter-like operations where the size of the result array depends on the source data.
+Recent work extends the polyhedral model to support arbitrary indexing~\cite{venkat2014polyhedral}, as well as conditional control flow that is predicated on arbitrary (ie, non-affine) functions of the loop indices~\cite{benabderrahmane2010polyhedral}.
+However, the indices used to write into the destination array must still be computed with affine functions.
 
-\subsection{Integer linear programming in imperative languages}
-The idea of using integer linear programming to find optimal fusion clusterings is not new, and has been discussed for imperative languages before.
+Ultimately, the job of an array fusion system is to make the program go as fast as possible on the available hardware.
+Although the cost metrics of ``optimal'' fusion systems try to model the performance behavior of this hardware, it is not practical to encode the intricacies of all available hardware in a single compiler implementation.
+Iterative compilation approaches such as~\cite{ashby2006iterative} instead enumerate many possible clusterings, use a the cost metric to rank them, and perform benchmark runs to identify which clustering actually performs the best.
+An ILP formulation like ours naturally supports this model, as the integer constraints define the available clusterings, and the objective function can be used to rank them.
+
+% -----------------------------------------------------------------------------
+% \subsection{Haskell short-cut fusion}
+% Existing fusion systems for Haskell such as stream fusion\cite{coutts2007streamfusion, % mainland2013exploiting}, tend to cleverly reuse compiler optimisations such as inlining%  and % rewrite rules\cite{peytonjones2001rules}, to fuse combinators without having to modify % the % compiler itself.
+This approach has the advantage of simplicity, but is inherently limit% ed i% n the % amount of fusion it can perform.%
+% 
+% Consider the following @filterMax@ function, where each element in the input array is % incremented, then the maximum is found, and the array is filtered to those greater tha% n zero.
+% In this case, the result of the map @vs'@ cannot be inlined into both occurences witho% ut % duplicating work, so no fusion can be performed.
+This has the effect of performing thr% ee % loops % instead of one, with two arrays instead of one.%
+% 
+% \begin{code}
+% filterMax vs =
+%  let vs' = map    (+1)  vs
+%      m   = fold   max 0 vs'
+%      flt = filter (>0)  vs'
+% \end{code}
+
+% -----------------------------------------------------------------------------
+% \subsection{Integer linear programming in imperative languages}
+% The idea of using integer linear programming to find optimal fusion clusterings is not new, and has been discussed for imperative languages before.
 These methods first construct a \emph{loop dependence graph} (LDG) from a given program, and then use this graph to create the integer linear program.
@@
 
-Talk about the simple formulation by Darte\cite{darte2002contraction}. Has an integer variable for each node, denoting the number of the cluster it's in.
+% Talk about the simple formulation by Darte\cite{darte2002contraction}.
+Has an integer variable for each node, denoting the number of the cluster it's in.
 Also includes a binary variable for each node, which is whether the node is fused with all its successors --- in which case no array would be required, and an integer variable which is the maximum of all cluster numbers.
-The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.
-It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
+The objective function is to maximise the number of nodes that are completely fused, requiring no arrays, and minimise the maximum cluster number, which in turn minimises the number of clusters.It doesn't require many constraints and is easy to implement, but doesn't work as well when there are loops of different sizes.
 As loops of different sizes cannot be fused together, a simple method is to introduce an ordering on the sizes, and then extract loops of the same cluster number in order of size.
 The problem here is that the objective function uses the maximum cluster number to minimise the number of loops, but this alone is no longer sufficient when there are multiple sizes.
-\TODO{Explain why. Perhaps go into more detail about how arrays are contracted, which is different from Megiddo.}
+\TODO{Explain why.
+Perhaps go into more detail about how arrays are contracted, which is different from Megiddo.}
 
-The formulation by Megiddo\cite{megiddo1998optimal} supports different sized loops, and is therefore more relevant for our purposes.
+% The formulation by Megiddo\cite{megiddo1998optimal} supports different sized loops, and is therefore more relevant for our purposes.
 For every pair of nodes $i,j$ in the LDG, a variable $x_{ij}$ is created, which denotes whether $i$ and $j$ are fused together.
@@
 For each pair of nodes, a weight constant $w_{ij}$ is given, and the objective function is to minimise $\Sigma_{ij} w_{ij} x_{ij}$, which has the effect of maximally fusing nodes, according to their weights.
-
 The difference to our combinator-based approach is that with combinators we retain more information about the meaning of the program.
 
-\subsection{Non-optimal heuristics}
+% \cite{gao1993collective} Collective loop fusion for array contraction.
+I think this is a good introduction to the min/max edge following algorithm that is central to most of these.
+Finds minimal number of clusters for single types.
+Despite the name, I don't think it actually minimises clustering for array contraction.
 
-\cite{gao1993collective} Collective loop fusion for array contraction.
-I think this is a good introduction to the min/max edge following algorithm that is central to most of these.
-Finds minimal number of clusters for single types. Despite the name, I don't think it actually minimises clustering for array contraction.
-
-\cite{kennedy1993typedfusion} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
+% \cite{kennedy1993typedfusion} Ken Kennedy --- Typed Fusion with Applications to Parallel and Sequential Code Generation.
 Typed fusion --- choose an ordering of types, then find clustering of first type, fuse those together, then cluster second type, and so on.
@@
 
-\cite{chatterjee1991size} Chatterjee -- Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
+% \cite{chatterjee1991size} Chatterjee -- Not optimal, heuristic based, etc, but does actually use ILP to reduce array crossing clusterings.
 Like typed fusion, they choose clustering for different types rather arbitrarily, which is sub-par.
 
-\cite{darte1999complexity} Alain Darte --- On the complexity of loop fusion.
-Shows that loop fusion, minimising certain things (ie manifest arrays and number of loops) is NP-hard.
+% \cite{darte1999complexity} Alain Darte --- On the complexity of loop fusion. Shows that loop fusion, minimising certain things (ie manifest arrays and number of loops) is NP-hard.
 
-\cite{song2004improving} Improving data locality by array contraction.
+% \cite{kennedy2001fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.
+
+% \cite{song2004improving} Improving data locality by array contraction.
 All about \emph{controlled SFC}, ie shifting, fusion and contraction.
-Formalises cost of memory references depending on distance / how many other references in between. Then only fuses loops if it doesn't raise the distance too much.
+Formalises cost of memory references depending on distance / how many other references in betweenThen only fuses loops if it doesn't raise the distance too much.
 
-\cite{ashby2006iterative} T.J. Ashby --- Iterative collective loop fusion.
-Executes the programs to decide which clustering is best, apparently. I didn't read it thoroughly, but didn't understand where they actually get the test data from. Unless it's a kind of JIT thing.
+% \cite{ashby2006iterative} T.J.
+Ashby --- Iterative collective loop fusion.
+Executes the programs to decide which clustering is best, apparently.
+I didn't read it thoroughly, but didn't understand where they actually get the test data from.
+Unless it's a kind of JIT thing.
 I honestly don't think much of it, but it \emph{is} more recent than the other stuff, so I suspect it'd be good to mention just so we're not ignoring ``modern fusion''.
 
-\cite{kennedy2001fastgreedy} Ken Kennedy --- Fast Greedy Weighted Fusion.
-
-\cite{sarkar1991optimization} Optimization of array accesses by collective loop transformations.
+% \cite{sarkar1991optimization} Optimization of array accesses by collective loop transformations.
 This is probably not at all relevant, but it's \emph{very cute}.
@@
 
-The polyhedral model is an algebraic representation of imperative loop nests and transformations on them, including fusion transformations.
-Polyhedral systems \cite{pouchet2011polyhedral} are able to express \emph{all possible} distinct loop transformations where the array indices, conditionals and loop bounds are affine functions of the surrounding loop indices.
-However, the polyhedral model is not applicable to (or intended for) one dimentional filter-like operations where the size of the result array depends on the source data.
-Recent work extends the polyhedral model to support arbitrary indexing~\cite{venkat2014polyhedral}, as well as conditional control flow that is predicated on arbitrary (ie, non-affine) functions of the loop indices~\cite{benabderrahmane2010polyhedral}.
-However, the indices used to write into the destination array must still be computed with affine functions.
-
-\TODO{Is unimodular the same as the polyhedral model?. 
+% \TODO{Is unimodular the same as the polyhedral model?.
 Now, here's my understanding of unimodular (and I'm just guessing this applies to polyhedral too): if the dependency matrix for a loop nest or set of loops forms a \emph{unimodular matrix} (an integral matrix whose inverse is also integral), only then can it be dealt with by unimodular transforms.}

### 2014-05-15 2014betterfusionforfilters/section/06-Related.tex
Typos

@@
 
-In the loop fusion literature, the ILP approach is considered ``optimal'' because it can find the clustering that minimises a global cost metric.
+In the loop fusion literature, the ILP approach is considered ``optimal'' because it can find the clustering that minimizes a global cost metric.
 In our case the metric is defined by the objective function of~\S\ref{s:ObjectiveFunction}.
 Besides optimal algorithms, there are also heuristic approaches.
-For example, Gao, Olsen and Sarkar~\cite{gao1993collective} use the maxflow-mincut algorithm to try to maximise the number of fused edges in the LDG.
+For example, Gao, Olsen and Sarkar~\cite{gao1993collective} use the maxflow-mincut algorithm to try to maximize the number of fused edges in the LDG.
 Kennedy~\cite{kennedy2001fastgreedy} describes another greedy approach which tries to maximize the reuse of intermediate arrays, and Song~\cite{song2004improving} tries to reduce memory references.
 
-Darte~\cite{darte1999complexity} formalises the algorithmic complexity of various loop fusion problems and shows that globally minimising most useful cost metrics is NP-complete.
+Darte~\cite{darte1999complexity} formalizes the algorithmic complexity of various loop fusion problems and shows that globally minimizing most useful cost metrics is NP-complete.
 Our ILP formulation itself is NP-hard, though in practice we have not yet found this to be a problem.
@@
 Polyhedral systems \cite{pouchet2011polyhedral} are able to express \emph{all possible} distinct loop transformations where the array indices, conditionals and loop bounds are affine functions of the surrounding loop indices.
-However, the polyhedral model is not applicable to (or intended for) one dimentional filter-like operations where the size of the result array depends on the source data.
+However, the polyhedral model is not applicable to (or intended for) one dimensional filter-like operations where the size of the result array depends on the source data.
 Recent work extends the polyhedral model to support arbitrary indexing~\cite{venkat2014polyhedral}, as well as conditional control flow that is predicated on arbitrary (ie, non-affine) functions of the loop indices~\cite{benabderrahmane2010polyhedral}.

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
Clarification of combinator vs name
in binding of combinator application

@@
 
-Looking back at the @normalize2@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
+Looking back at the @normalize2@ example, the iteration sizes of the combinators of @gts@ and @sum1@ are both $k_{xs}$.
 %The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
@@
 % No,
-We now define a function $trans$, to find the parent transducer of a given combinator.
+We now define a function $trans$, to find the parent transducer a combinator application.
+Since each name is bound to at most one combinator, we abuse terminology here slightly and write ``combinator $n$'' when refering to the combinator occuring in the binding of the name $n$.
 The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.

### 2014-05-15 2014betterfusionforfilters/section/06-Related.tex
Typo

@@
 Although the cost metrics of ``optimal'' fusion systems try to model the performance behavior of this hardware, it is not practical to encode the intricacies of all available hardware in a single compiler implementation.
-Iterative compilation approaches such as~\cite{ashby2006iterative} instead enumerate many possible clusterings, use a the cost metric to rank them, and perform benchmark runs to identify which clustering actually performs the best.
+Iterative compilation approaches such as~\cite{ashby2006iterative} instead enumerate many possible clusterings, use a cost metric to rank them, and perform benchmark runs to identify which clustering actually performs the best.
 An ILP formulation like ours naturally supports this model, as the integer constraints define the available clusterings, and the objective function can be used to rank them.

### 2014-05-15 2014betterfusionforfilters/section/05-Benchmarks.tex
benchmarks

@@
 
-We have run three benchmarks to constrast our clustering algorithm, Megiddo~\cite{megiddo1998optimal}'s clustering, and the clustering produced by stream fusion\cite{coutts2007streamfusion}.
+We have run three benchmarks to constrast our clustering algorithm, Megiddo~\cite{megiddo1998optimal}'s clustering, the clustering produced by stream fusion~\cite{coutts2007streamfusion}, and no fusion at all.
 The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree.
 The benchmark programs are all hand-written, hand-fused C code based on the clustering.
-Each program was ran with the same input five times, and the minimum runtime was used.
+Each program was run with the same input five times, and the minimum runtime was used.
+Runtimes and the number of loops for each clustering are shown in Figure~\ref{f:BenchResults}.
+In all cases, our clustering performs better than or as good as Megiddo's, and better than stream fusion and unfused.
+Interestingly, stream fusion's clustering for @normalize2@ performs better than Megiddo's, despite having more loops, as stream fusion is able to remove the intermediate array.
 
-Benchmark programs are available at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}
+Benchmark programs are available at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}.
 
 \subsection{Normalize2}
-The normalize2 example is already in combinator normal form.
+To demonstrate the ILP formulation, the clustering of @normalize2@ is derived as follows.
 \begin{code}
@@
 
+To make the program shorter and easier to work with, we use the optimised version.
 First, we must calculate $split$ -- that is, those nodes which have no fusion-preventing path between them.
-\begin{code}
-split(sum1, gts)    = T
-split(sum1, sum2)   = T
-split(sum1, ys2)    = T
-split(gts,  sum2)   = T
-split(gts,  ys1)    = T
-split(sum2, ys1)    = T
-split(ys1,  ys2)    = T
-\end{code}
-We only need to create cluster variables for these.
+Any elements in the following sets have no fusion-preventing paths.
+\[ \{ \{sum1, gts, sum2\}
+ , \{sum1, ys2\}
+ , \{gts, sum2, ys1\}
+ , \{ys1, ys2\} \} \]
 
@@
 \end{tabbing}
-$parents(sum1,sum2)  = \{(sum1, gts)\}$
 
-A minimal solution to this is
+One minimal solution to this is:
 \begin{tabbing}
@@
 \end{tabbing}
+This minimal solution is not unique, however in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.
+
 Looking at just the non-zero variables in the objective function, the value is
@@
 If the @fold@ is changed to operate over @pts@ instead of @pts'@, Megiddo's formulation \emph{is} able to fuse the two.
-Interestingly, a programmer unaware of fusion would expect the original version to be faster, because @pts'@ is smaller than @pts@.
-
-\begin{code}
- = let pts' = filter (above   line) pts
-       ma   = fold   (maxFrom line) pts
-\end{code}
+However, a programmer unaware of fusion would expect the original version to be faster, because @pts'@ is smaller than @pts@.

### 2014-05-15 2014betterfusionforfilters/Main.tex
Add TODO to intro
Cut back related work section
condense abstract

@@
 \usepackage{style/judgements}
+
+% TODO: Fix up writing to use the words combinator and operator consistently
+% We seem to be using "combinator" for the functions in the original program, 
+% but "operator" for to name the abstract operation that combinator performs.
+
+% \TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
 
@@
 Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code.
-To achieve this, it is not sufficient to just fuse subsequent array operations into a single, more complex computation; we also have to \emph{cluster} separate traversals of the same array into one single loop.
+It is not sufficient to simply fuse subsequent array operations into a single computation; we also have to cluster \emph{separate} traversals of the same array into a single loop.
 Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
-However, these approaches have one serious restriction: they can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
-The technique we present in this paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
-
-%The original abstract:
-%Recent work on array fusion shows how to extract data-flow graphs from functional programs and compile them into efficient imperative loops.
-However, the compilation process only handles graphs that can be compiled into \emph{single} loops, and most programs require more.
-Prior work shows how use Integer Linear Programming (ILP) to \emph{cluster} the operators in a general data-flow graph into subgraphs that can be individually fused, but until now this approach did not handle operations like filter, which produce arrays of a different size than their input.
-We extend the existing ILP approach with support for size changing operators, using an external ILP solver to find good clusterings.
+However, these approaches can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
+This paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
 \end{abstract}
@@
 \input{section/06-Related.tex}
-\section{Conclusion}
 

### 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
condense abstract

@@
 Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
-As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering.
-Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.
+As shown in (Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering.
+Unfortunately, they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.
 The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.
@@
 To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
-It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa}) as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
 

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
Cut back related work section
Cut back the ILP intro

@@
 %!TEX root = ../Main.tex
-\section{Integer linear program formulation}
-\label{s:ILP}
 
-\begin{figure}[ht!]
+\begin{figure}
 \begin{center}
@@
 \caption{Possible clusterings for \texttt{normalizeInc}}
-\label{f:normInc}
+\label{f:normalizeInc}
 \end{figure}
 
-The problem of finding the ``best'' clustering according to some criteria is known to be NP-hard~\cite{darte1999complexity}.
-For example, in the special case of single-typed fusion, where all loops have the same iteration size, finding a clustering with the minimal number of loops can be done in polynomial time.
-However, to find the minimal number of manifest arrays, each node may need to be considered in several clusters, and the problem becomes NP-hard.
-Similarly, after adding multiple iteration sizes that cannot be fused together, the problem becomes NP-hard.
-
-The clusterings of a given program are not unique; there are many valid clusterings, far too many to feasibly enumerate all of them.
-There may even be multiple clusterings with the minimal number of loops. For example, consider the following program:
-
+% -----------------------------------------------------------------------------
+\section{Integer Linear Programming}
+\label{s:ILP}
+It is usually possible to cluster a program graph in multiple ways. For example, consider the following simple function:
 \begin{code}
@@
  normalizeInc xs
-  = let incs = map  (+1)    us      (A2) (B1)
-        sum  = fold (+) 0   us      (A1) (B1)
-        norm = map  (/ sum) incs    (A2) (B2)
-    in  norm
+  = let incs = map  (+1)    us
+        sum1 = fold (+) 0   us
+        ys   = map  (/ sum) incs
+    in  ys
 \end{code}
-For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}.
-The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@).
-For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@.
-For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
 
-Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
-The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.
+Two possible clusterings are shown in Figure~\ref{f:normalizeInc}.
+One option is to compute @sum1@ first and fuse the computation of @incs@ and @ys@.
+Another option is to fuse the computation of @incs@ and @sum1@ into a single loop, then compute @ys@ separately.
+A third option (not shown) is to compute all results separately, and not perform any fusion.
 
-In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
-Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
+Which option is better?
+On current hardware we generally expect the cost of memory access to dominate runtime.
+The first clustering in Figure~\ref{f:normalizeInc} requires two reads from array @xs@ and one write to array @ys@.
+The second requires a single fused read from @xs@, one write to @incs@, a read back from @incs@ and a final write to @ys@.
+From the size constraints of the program we know that all intermediate arrays have the same size, so we expect the first clustering will peform better as it only needs three array accesses instead of four.
+
+For small programs such as @normalizeInc@ it is possible to naively enumerate all possible clusterings, select just those that are \emph{valid} with respect to fusion preventing edges, and chose the one that maximises a cost metric such as the number of array accesses needed.
+However, as the program size increases the number of possible clusterings becomes too large to naively enumerate.
+For example, Pouchet et al~\cite{pouchet2010combined} present a fusion system using the polyhedral model~\cite{pouchet2011polyhedral} and report that some simple numeric programs have over 40,000 possible clusterings, with one particular example having $10^{12}$.
+
+To deal with the combinatorial explosion in the number of potential clusterings, we instead use an Integer Linear Programming (ILP) formulation.
+ILP problems are defined as a set of variables, an objective linear function and a set of linear constraints.
 The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
-This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
-Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.
+For the clustering problem we express our constraints regarding fusion preventing edges as linear constraints on the ILP variables, then use the objective function to encode our cost metric.
+This general approach was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}, and our main contribution is to extend it to work with size changing operators such as @filter@.
 
-While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
+% -----------------------------------------------------------------------------
+\subsection{Function $\to$ graph}
+% \ben{This is a different sort of graph than before.
+This is a dependency graph where the nodes are intermediate variables, verses the previous ``program graph'' where the nodes can also be operators.
+We need to contrast these properly and add a diagram of a dependency graph.}
+
+% While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
 To simplify creation of the integer linear program formulation, we first convert a function to a dependency graph.
@@
 After the graph is created, it is converted to an integer linear program with an objective of the least memory traffic and loops, solved, and the solution converted to a clustering.
-
-When creating the graph, it is important to note that not all loops can be fused together; firstly, loops of different size cannot be fused.
-Secondly, two loops cannot be fused together if an iteration in one loop relies on the output of a later iteration in another loop.
-For example, a @map@ may not be fused with a @fold@ if the @map@'s worker function uses the result of the @fold@.
-This fusion restriction is encoded as a \emph{fusion-preventing} edge between two combinators.
-Other edges are \emph{fusible}, and may be fused together.
-If they are not fused together for whatever reason, the \emph{from} combinator must be scheduled before the \emph{to} combinator.
-
-% The dependency graph is translated to an integer linear program.
-The integer linear program has some integral, boolean and real variables, an objective function to minimise, and a set of constraints that the variables must conform to.
-Finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, but existing ILP solvers tend to be sufficient for realistic problem sizes.
-For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still gives the exact answer for small problems.
-
-% -----------------------------------------------------------------------------
-\subsection{Function $\to$ graph}
-\ben{This is a different sort of graph than before.
-This is a dependency graph where the nodes are intermediate variables, verses the previous ``program graph'' where the nodes can also be operators.
-We need to contrast these properly and add a diagram of a dependency graph.}
-\TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
 
@@
 
+% -----------------------------------------------------------------------------
 \subsubsection{The objective function}

### 2014-05-15 2014betterfusionforfilters/Main.tex
Cut back the ILP intro
change sigplan category; I think compilers/optimization more fitting

@@
 % \TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
+
+% \TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
 
@@
 \begin{abstract}
-Minimising memory traffic is key to compiling functional, collection oriented array programs into efficient code.
+Minimising memory traffic is key to compiling functional, collection-oriented array programs into efficient code.
 It is not sufficient to simply fuse subsequent array operations into a single computation; we also have to cluster \emph{separate} traversals of the same array into a single loop.
@@
 \category
-	{D.3.3}
+	{D.3.4}
 	{Programming Languages}
-	{Language Constructs and Features---Concurrent programming structures; Control structures; Abstract data types}
+	{Processors---Compilers; Optimization}
 

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
Work on dependency graphs

@@
 % -----------------------------------------------------------------------------
-\subsection{Function $\to$ graph}
-% \ben{This is a different sort of graph than before.
-This is a dependency graph where the nodes are intermediate variables, verses the previous ``program graph'' where the nodes can also be operators.
-We need to contrast these properly and add a diagram of a dependency graph.}
+\subsection{Dependency Graphs}
+A dependency graph represents the data dependencies of the program to be fused, and we use it as an intermediate stage when producing linear constraints for the ILP problem.
+The dependency graph contains enough information to determine the possible clusterings of the input program, while abstracting away from the exact operators used to compute each intermediate array.
+The rules for producing a dependency graphs are in Figure~\ref{f:DependencyGraph}.
 
-% While the different combinators are all implemented differently, certain aspects are shared between all; they operate on input arrays, use scalar variables in their worker functions, and produce output.
-To simplify creation of the integer linear program formulation, we first convert a function to a dependency graph.
-The dependency graph has a node for each combinator, and edges between two combinators when one uses another's output.
-After the graph is created, it is converted to an integer linear program with an objective of the least memory traffic and loops, solved, and the solution converted to a clustering.
+Each binding in the source program becomes a node in the dependency graph.
+For each intermediate variable, we add a directed edge from the binding that produces a value to all bindings that consume it.
+Each edge is also marked as either \emph{fusible} or \emph{fusion preventing}.
+Fusion preventing edges are used when the producer must finish its execution before the consumer node can start.
+For example, a @fold@ operation must complete execution before it can produce the scalar value needed by its consumers.
+Conversely, the @map@ operation produces an output value for each value it consumes, so is marked as fusible.
 
-Converting a function in \emph{combinator normal form} to a graph --- in fact, a DAG --- is quite simple.
-Each binding becomes a node in the graph.
-When a binding references other arrays or scalars, there is an edge between those two nodes.
-Edges may be either \emph{fusible} or \emph{fusion-preventing}.
-Fusion-preventing edges mean that the entire input node must completely finish its execution before the output node can start executing.
-For example, @fold@s consume their entire input array before producing a single result, so any references to folds must be fusion-preventing.
-Conversely, @map@s produce output data for every input element, so may be fused.
+The @gather@ operation is a hybrid: it takes an indices array and an elements array, and for each element in the indices array returns the corresponding data element.
+This means that gather can be fused with the operation that produces its indices, but not the operation that produces its elements --- because those are accessed in a random-access manner.
 
-The @gather@ operation is interesting; it takes an indices array and a data array, and for each element in indices returns that element in data.
-Thus, @gather@ requires random-access in its data array, and is not fusible, but consumes indices linearly and may fuse.
+% Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
+Creating edges between these nodes is simply when one binding references an earlier one.
+The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.
 
+\begin{figure}
 \begin{tabbing}
-MMMM        \= MM   \= \kill
-$nodes$     \> @::@ \> $program \to V$          \\
-$edges$     \> @::@ \> $program \to E$          \\
-$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
-\\
-MMMMM       \= \kill
-$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
-\end{tabbing}
-Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
-Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.
+MMMMM       \= M  \= \kill
+$nodes$     \> @:@ \> $program \to V$              \\
+$edges$     \> @:@ \> $program \to E$              \\
+$edge$      \> @:@ \> $\{bind\} \times bind \to E$ 
+\\[1ex]
+$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       
+\\[2ex]
 
-\begin{tabbing}
-MMMMM       \= \kill
-$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
-\\
+$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    
+\\[1ex]
 MM             \= M \= \kill
@@
 $edge(bs, outs = @external@~ins)$  \\
-    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
-\\
+    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ 
+\\[1ex]
 $inedge(bs,to,from)$ \\
@@
     \> $|$ \> $otherwise$                      \\
-    \> $=$ \> $(to, from, \fusible)$            \\
+    \> $=$ \> $(to, from, \fusible)$
 \end{tabbing}
+
+\caption{Dependency Graphs from Programs}
+\label{f:DependencyGraph}
+\end{figure}
 

### 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
cut back intro to fit on single page. rather brutal. revert if you dont think it makes sense

--- 2014-05-15 2014betterfusionforfilters/section/01-Introduction.tex
@@
 A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.
-It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance.
-This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{coutts2007streamfusion, gao1993collective, kennedy1993typedfusion, kennedy2001fastgreedy, keller2010repa}, to name just a few.
-Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
-As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering.
-Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.
+It is essential, however, that the compiler combines these operations and compiles them into a small number of efficient loops, as a naive translation leads to high memory traffic and poor performance.
+This is a well known problem, and techniques to fuse subsequent array operations together have been presented in~\cite{gill1993short, coutts2007streamfusion, keller2010repa}, to name just a few.
+This type of fusion alone is not generally sufficient to minimise memory traffic.
+% : first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
+As shown in (Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find good clusterings.
+Unfortunately, they cannot handle operations like filter, where the output size differs from the input size.
 The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.
 
-To compile subgraphs which are the result of our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
-It improved on existing related array fusion approaches (~\cite{coutts2007streamfusion}, ~\cite{keller2010repa} as it 1) fuses programs that use branching data flows where a produced array is consumed by several consumers, and 2) guarantees complete fusion into a single loop for all programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+To compile the clusters found by our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
+It improved on existing array fusion approaches~\cite{coutts2007streamfusion, keller2010repa} as it guarantees complete fusion into a single loop for programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
 
@@
 
-To demonstrate the effect of clustering and fusion, consider the following code snippet:
+To see the effect of clustering, consider the following program:
 \begin{code}
@@
 \end{code}
-The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero.
+
+The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of only elements greater than zero.
 Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops.
@@
 In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold.
-The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations.
-Our approach (diagram in the middle) produces, in this case, the optimal solution: two loops.
-The first to calculate both sums, the second to calculate the maps.
+The best existing ILP approach results in the rightmost graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations.
+Our approach, in the middle, produces the optimal solution in this case: one loop for the sums, another for the maps.
 
@@
 \item
-We present a simplification to constraint generation that is also applicable to some existing integer linear programming formulations such as Megiddo's,
-where constraints between two nodes need not be generated if there exists a fusion-preventing path between the two (\S\ref{s:OptimisedConstraints}).
+We present a simplification to constraint generation that is also applicable to some ILP formulations such as Megiddo's:
+constraints between two nodes need not be generated if there is a fusion-preventing path between the two (\S\ref{s:OptimisedConstraints}).
 
 \item
-Our constraint system also encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program.
-For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, the memory traffic will be reduced (\S\ref{s:ObjectiveFunction}).
+Our constraint system encodes a total ordering on the cost of clusterings, expressed using weights on the integer linear program.
+For example, we encode that memory traffic is more expensive than loop overheads, so given a choice between the two, memory traffic will be reduced (\S\ref{s:ObjectiveFunction}).
 
 \item
-We present benchmarks of our algorithm applied to several common programming patterns, and to several pathological examples.
-Our algorithm is complete and yields good results in practice, though if array sizes are unknown, an optimal solution is uncomputable in general (\S\ref{s:Benchmarks}).
+We present benchmarks of our algorithm applied to several common programming patterns.
+Our algorithm is complete and yields good results in practice, though an optimal solution is uncomputable in general (\S\ref{s:Benchmarks}).
 \end{itemize}
 
-The reduction of the clustering problem to integer linear programming was previously described by~\cite{megiddo1998optimal}, though they do not consider length changing operators.
-
-An implementation of the clustering algorithm is available at \url{https://github.com/amosr/clustering}.
+An implementation of our clustering algorithm is available at \url{https://github.com/amosr/clustering}.
 

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
Elaborate on decription of ILP variables

@@
 % -----------------------------------------------------------------------------
-\subsection{Graph $\to$ ILP}
-With the dependency graph now constructed, it must be converted to an integer linear program.
-The integer linear program will contain some variables, an objective function, and constraints that the variables must conform to.
-An external solver will find and return a variable assignment that minimises the objective function.
-With the variable assignment, the graph's nodes are partitioned into a clustering, and then flow fusion\cite{lippmeier2013flow} is used to extract imperative loops for each cluster.
-
-The variables in the ILP formulation can be split into three groups:
+\subsection{ILP Variables}
+The next step is to produce a set of linear constraints from the dependency graph defined in the previous section. The variables involved in these linear constraints are split into three groups:
 
@@
 \end{tabbing}
-The first and most important group is created for every pair of nodes, and has a boolean value indicating whether the two nodes can be fused.
-If $x_{ij} = 0$, then $i$ and $j$ are fused together.
-While it may seem counterintuitive for 0 to mean fused and 1 unfused, it means the objective function can simply minimise a weighted $x_{ij}$, and also makes later precedence constraints simpler.
-The values of these variables are used to construct the partitioning at the end, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
-If two nodes are ``fused together'', it means that their combinators will be in the same cluster.
+For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{ij}$ which indicates whether those two nodes are fused.
+We use $x_{ij} = 0$ when the nodes are fused and $x_{ij} = 1$ when they are not.
+Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{ij}$ variables, and minimizing it tends to increase the number of nodes that are fused.
+The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
 
@@
 \end{tabbing}
-The second group of variables is used to ensure that the clustering is acyclic -- that is, there must be an ordering to execute each clusters' dependency clusters before executing the cluster itself.
-For each node $i$, we associate a real $\pi_i$ such that every node $j$ after $i$ we have $\pi_j > \pi_i$.
-This has the effect of requiring an acyclic clustering, as if one node is after another its $\pi$ must be strictly greater than its predecessor, and the predecessor must be strictly less than the successor.
-If two nodes are to be fused together, their $\pi$ values must be the same.
+The second group of variables is used to ensure that the clustering is acyclic.
+This means that for each node in the graph, the dependencies of that node can be executed before the node itself.
+For each node $i$, we associate a real $\pi_i$ such that every node $j$ that depends on $i$ we have $\pi_j > \pi_i$.
+Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value.
 
@@
 \end{code}
-There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
-If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
+There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@.
+If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
 
@@
 \end{tabbing}
-The third, and final, group of variables are purely for minimisation, indicating whether a node's array is \emph{fully contracted}.
-They are not required for correctness of the clustering, but are used in the objective function.
-A node's array will be fully contracted --- that is, it will be replaced by a scalar --- if all nodes that use this array are fused together: $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
+The final group of variables is used to help define the cost model encoded by the objective function.
+Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}.
+When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
+In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.
 
 % -----------------------------------------------------------------------------
-\subsubsection{Unoptimised version}
+\subsection{Linear Constraints}
 Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
@@
 % -----------------------------------------------------------------------------
-\subsubsection{The objective function}
+\subsection{The objective function}
 \label{s:ObjectiveFunction}
@@
 
-\subsubsection{A note on transitivity}
+% -----------------------------------------------------------------------------
+\subsection{A note on transitivity}
 It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
@@
 
-\subsubsection{Fusion-preventing path optimisation}
+% -----------------------------------------------------------------------------
+\subsection{Fusion-preventing path optimisation}
 \label{s:OptimisedConstraints}

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
Copy edits to linear constraints

@@
 \subsection{Linear Constraints}
-Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
+The constraints we place on the ILP variables are split into four groups: constraints that ensure the clustering is acyclic; constraints that encode fusion preventing edges; constraints on nodes with different iteration sizes, and constraints involving array contraction.
+
+% Before showing the optimised version with certain constraints removed (\S\ref{s:OptimisedConstraints}), this simpler, unoptimised version is shown.
 The only difference is that fewer constraints and variables are required in the optimised version, but both versions give the same clustering.
@@
 
-\paragraph{Acyclic and precedence-preserving}
+\paragraph{Acyclic and precedence-preserving} The first group of constraints ensures that the clustering is acyclic:
+\begin{tabbing}
+MM  \= MMM \= M \= MMM \= M \= MMMM \= \kill
+    \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ 
+    \>             (with an edge from $i$ to $j$)            \\
+    \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ 
+    \>             (with no edge from $i$ to $j$)
+\end{tabbing}
+As per Megiddo~\cite{megiddo1998optimal} the form of these constraints is determined by whether there is an dependency between nodes $i$ and $j$.
+The $N$ value is set to the total number of nodes in the graph.
 
-\begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> \ldots \\
-Subject to \> \ldots \\
-           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (an edge from $i$ to $j$)            \\
-           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \>             (no edge from $i$ to $j$)            \\
-           \> \ldots
-\end{tabbing}
-As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.
+If there is an edge from node $i$ to $j$ we use the first constraint form shown above.
+If the two nodes are fused into the same cluster then we have $x_{ij} = 0$.
+In this case the constraint simplifies to $0 \le \pi_j - \pi_i \le 0$, which forces $\pi_i = \pi_j$.
+If the two nodes are in \emph{different} clusters then the constraint instead simplifies to $1 \le \pi_j - \pi_i \le N$.
+This means that the difference between the two $\pi$s must be at least 1, and less than $N$.
+Since there are $N$ nodes, the maximum difference between any two $\pi$s would be at most $N$, so the upper bound of $N$ is large enough to be safely ignored.
+This means the constraint can roughly be translated to $\pi_i < \pi_j$, which enforces the acyclicity constraint.
 
-Whether there is an edge between $i$ and $j$ or not, if the two are fused together, then $x_{ij} = 0$.
-If $x_{ij} = 0$, then both cases may be simplified to $0 \le \pi_j - \pi_i \le 0$, which is equivalent to $\pi_i = \pi_j$.
-\ben{Highlight that ``fused together'' means ``in the same cluster'', and that operators can be fused even if there are no direct edges between them.}
-
-Otherwise, if the two nodes are not fused together, then $x_{ij} = 1$.
-If there is an edge between $i$ and $j$, the constraint simplifies to $1 \le \pi_j - \pi_i \le N$, where $N$ is the number of nodes in the graph.
-This constraint means that the difference between the two $\pi$s must be at least 1, and less than $N$.
-Since there are $N$ nodes, the maximum difference between any two $\pi$s would be at most $N$, so the upper bound of $N$ is large enough to be safely ignored.
-This means the constraint can roughly be translated to $\pi_i < \pi_j$, which enforces the acyclic constraint.
-
-If there is no edge between $i$ and $j$ and the two are not fused together, then $x_{ij} = 1$ and the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively puts no constraint on the $\pi$ values.
-Note, however, that other edges may still constrain $i$ or $j$.
+If instead there is no edge from node $i$ to $j$ then we use the second constraint form above.
+As before, if the two nodes are fused into the same cluster then we have $x_{ij} = 0$, which forces $\pi_i = \pi_j$.
+If the nodes are in different clusters then the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively puts no constraint on the $\pi$ values.
 

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
More on linear constraints

@@
 
+% -------------------------------------
 \paragraph{Acyclic and precedence-preserving} The first group of constraints ensures that the clustering is acyclic:
@@
 
-\paragraph{Fusion-preventing edges}
+% -------------------------------------
+\paragraph{Fusion-preventing edges} As per Megiddo\cite{megiddo1998optimal}, if there is a fusion preventing edge between two nodes we add a constraint to ensure the nodes will be placed in different clusters.
 \begin{tabbing}
 MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> \ldots \\
-Subject to \> \ldots \\
            \> $x_{ij}$    \> $=$   \> $1$ \>       \>       \\
-           \> (fusion-preventing edges from $i$ to $j$)      \\
-           \> \ldots
+           \> (for fusion-preventing edges from $i$ to $j$) 
 \end{tabbing}
-As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur.
-This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.
+When combined with the precedence-preserving constraints earlier, setting $x_{ij} = 1$ also forces $\pi_i < \pi_j$.
 
+% -------------------------------------
 \paragraph{Fusion between different iteration sizes}

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
Squash onto 10 pages

@@
 % -------------------------------------
-\paragraph{Fusion-preventing edges} As per Megiddo\cite{megiddo1998optimal}, if there is a fusion preventing edge between two nodes we add a constraint to ensure the nodes will be placed in different clusters.
+\paragraph{Fusion-preventing edges} As per Megiddo~\cite{megiddo1998optimal}, if there is a fusion preventing edge between two nodes we add a constraint to ensure the nodes will be placed in different clusters.
 \begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
         \> $x_{ij}$    \> $=$   \> $1$ \>       \>       \\
@@
 \begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> \ldots \\
-Subject to \> \ldots \\
+MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
         \> $x_{ij}$    \> $=$   \> $1$          \>       \>     \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
+        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$
+\\
 \\
@@
         \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$    \>       \>     \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$) \\
+        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$)
+\\
         \> \ldots
@@
 MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> \ldots \\
-Subject to \> \ldots \\
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
@@
 \\
-\\
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
kerning in math mode

@@
 \textbf{Size Constraint}
-\> $C$      \> @::=@  \> $true$               \> (trivially true)      \\
+\> $C$      \> @::=@  \> $\true$               \> (trivially true)      \\
 \>          \> $~|$   \> $k = \tau$           \> (equality constraint) \\
@@
 We use the latter to represent the result size of the @cross@ operator discussed in the previous section.
-Constraints maybe either be trivially $true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$.
+Constraints maybe either be trivially $\true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$.
 We refer to the trivially true and equality constraints as \emph{atomic constraints}.
@@
 \\
-\vdash & true 
+\vdash & \true 
         ~\wedge~  k_{gts} = k_1
-        ~\wedge~  true
+        ~\wedge~  \true
 \\     &~~~~~~~~ 
@@
 
-Note that $k_xs$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $true$ terms then yields:
+Note that $k_xs$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $\true$ terms then yields:
 $$

### 2014-05-15 2014betterfusionforfilters/section/04-ILP.tex
kerning in math mode
Copy edits to linear constraints / fusion between different sizes

@@
 \\[1ex]
-$inedge(bs,to,from)$ \\
-    \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
-    \> $=$ \> $(to, from, \fusionpreventing)$  \\
-    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge from \in outs$     \\
+$inedge(bs,to,\from)$ \\
+    \> $|$ \> $(\from = @fold@~f~s) \in bs$     \\
+    \> $=$ \> $(to, \from, \fusionpreventing)$  \\
+    \> $|$ \> $(outs = @external@ \ldots) \in bs     \wedge \from \in outs$     \\
     \> $=$ \> $(to, outs, \fusionpreventing)$  \\
     \> $|$ \> $otherwise$                      \\
-    \> $=$ \> $(to, from, \fusible)$
+    \> $=$ \> $(to, \from, \fusible)$
 \end{tabbing}
@@
 
+\eject
 % -----------------------------------------------------------------------------
 \subsection{ILP Variables}
-The next step is to produce a set of linear constraints from the dependency graph defined in the previous section. The variables involved in these linear constraints are split into three groups:
-
+After generating the dependency graph, the next step is to produce a set of linear constraints from this graph. The variables involved in these constraints are split into three groups:
 \begin{tabbing}
@@
 The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
-
 \begin{tabbing}
@@
 
-As an example of a cyclic clustering, consider:
+Here is an example of a cyclic clustering:
 \begin{code}
-  cycle us
-   = let xs  = map (+1) us          (C1)
-         sum = fold xs              (C2)
-         zs  = map (+sum) xs        (C1)
+  cycle xs  = let ys  = map (+1) xs     (C1)
+                  sum = fold ys         (C2)
+                  zs  = map (+sum) ys   (C1)
+              in  zs
 \end{code}
@@
 If the @xs@ and @zs@ bindings were in the same cluster @C1@ and @sum@ was in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.
-
 \begin{tabbing}
@@
 \begin{tabbing}
-MM  \= MMM \= M \= MMM \= M \= MMMM \= \kill
-    \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ 
+MM  \= MMMx \= M \= MMM \= M \= MMMM \= \kill
+    \> ~~~~~~~ $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N.\; x_{i,j}$ 
     \>             (with an edge from $i$ to $j$)            \\
-    \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ 
+    \> $-N.\;  x_{i,j}$  \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N.\; x_{i,j}$ 
     \>             (with no edge from $i$ to $j$)
@@
 If there is an edge from node $i$ to $j$ we use the first constraint form shown above.
-If the two nodes are fused into the same cluster then we have $x_{ij} = 0$.
+If the two nodes are fused into the same cluster then we have $x_{i,j} = 0$.
 In this case the constraint simplifies to $0 \le \pi_j - \pi_i \le 0$, which forces $\pi_i = \pi_j$.
@@
 If instead there is no edge from node $i$ to $j$ then we use the second constraint form above.
-As before, if the two nodes are fused into the same cluster then we have $x_{ij} = 0$, which forces $\pi_i = \pi_j$.
+As before, if the two nodes are fused into the same cluster then we have $x_{i,j} = 0$, which forces $\pi_i = \pi_j$.
 If the nodes are in different clusters then the constraint simplifies to $-N \le \pi_j - \pi_i \le N$, which effectively puts no constraint on the $\pi$ values.
@@
 MMM     \= MMM \= M  \= MMM \= M \= MMM \= \kill
-        \> $x_{ij}$    \> $=$   \> $1$ \>       \>       \\
+        \> $x_{i,j}$ \> $=$ \> $1$ \>   \> \\
         \> (for fusion-preventing edges from $i$ to $j$) 
 \end{tabbing}
-When combined with the precedence-preserving constraints earlier, setting $x_{ij} = 1$ also forces $\pi_i < \pi_j$.
+When combined with the precedence-preserving constraints earlier, setting $x_{i,j} = 1$ also forces $\pi_i < \pi_j$.
 
 % -------------------------------------
-\paragraph{Fusion between different iteration sizes}
-
+\paragraph{Fusion between different iteration sizes} This group of constraints restricts which nodes can be placed in the same cluster based on their iteration size. The group has three parts. 
+Firstly, either of the two nodes connected by an edge have an unknown ($\bot$) iteration size then they cannot be fused and we set $x_{i,j} = 1$:
 \begin{tabbing}
 MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
-        \> $x_{ij}$    \> $=$   \> $1$          \>       \>     \\
-        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$
-\\
-\\
-        \> $x_{ij}$    \> $=$   \> $1$          \>       \>     \\
-        \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
-\\
-        \> $x_{i'j}$   \> $\le$ \> $x_{ij}$     \>       \>     \\
-        \> $x_{ij'}$   \> $\le$ \> $x_{ij}$     \>       \>     \\
-        \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$    \>       \>     \\
-        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\}$)
-\\
-        \> \ldots
+        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
+        \> (if $\iiter_{\Gamma,C}(i) = \bot 
+                ~\vee~ \iiter_{\Gamma,C}(j) = \bot$)
 \end{tabbing}
-If two nodes have different iteration sizes, they may not necessarily be fused.
-If either of the nodes have no iteration size --- in which case, they are @external@ computations --- they may not be fused at all.
-However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.
+Secondly, if the two nodes have different iteration sizes and no common parent then they also cannot be fused and we set $x_{i,j} = 1$:
+\begin{tabbing}
+MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
+        \> $x_{i,j}$   \> $=$   \> $1$          \>       \>     \\
+        \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) 
+                ~\wedge~ parents(i,j) = \emptyset$)
+\end{tabbing}
+Finally, if the two nodes had different iteration sizes but \emph{do} have parent transducers of the same size, then the two nodes can be fused if they are fused with their respective parents, and the parents themselves are fused:
+\begin{tabbing}
+MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
+        \> $x_{a,A}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
+        \> $x_{b,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
+        \> $x_{A,B}$   \> $\le$ \> $x_{a,b}$    \>       \>     \\
+        \> (if $\iiter_{\Gamma,C}(a) \not= \iiter_{\Gamma,C}(b) 
+                ~\wedge~ parents(a,b) = \{(A,B)\}$)
+\end{tabbing}
+This last part is the main difference to existing ILP solutions: we allow nodes with different iteration sizes to be fused when their parent transducers are fused.
+The actual constraints encode a ``no more fused than'' relationship.
+For example $x_{a,A} \le x_{a,b}$ means that nodes $a$ and $b$ can be no more fused than nodes $a$ and $A$.
 
-This is the main difference to existing integer linear programming solutions: we allow nodes with different iteration sizes to be fused together, when their parent transducers are fused together.
-This is an important feature for combinator-based programming, where size-changing operations such as @filter@ are used frequently.
-\ben{Highlight why we want to allow this sort of fusion, and why it is specific to data flow fusion.}
-
-As a degenerate example, consider fusing an operation on filtered data with its generating filter:
+As a simple example, consider fusing an operation on filtered data with its generating filter:
 \begin{code}
- sum1 = fold (+) 0  us
- gts  = filter (>0) us
+    sum1 = fold (+) 0  xs
+    gts  = filter (>0) xs
     sum2 = fold (+) 0  gts
 \end{code}
-Here, $parents(gts,sum2) = \{gts, gts\}$.
+Here $sum1$ and $sum2$ have different iteration sizes and we have that $parents(sum1, sum2) = \{(sum1, gts)\}$.
+This means that $sum1$ and $sum2$ may only be fused if $sum1$ is fused with $sum1$ (trivial), $sum2$ is fused with $gts$, and $sum1$ is fused with $gts$.
+
+% Here, $parents(gts,sum2) = \{gts, gts\}$.
 This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
@@
 
-In the same example, $sum1$ and $sum2$ also have different iteration sizes.
-We have that $parents(sum1,sum2) = \{sum1, gts\}$.
-This means that $sum1$ and $sum2$ may only be fused if $sum1$ is fused with $sum1$ (trivial), $sum2$ is fused with $gts$, and $sum1$ is fused with $gts$.
-While it may seem like these constraints should be implied by transitivity of clustering, all clustering variables are used for the objective function.
+% While it may seem like these constraints should be implied by transitivity of clustering, all clustering variables are used for the objective function.
 
-\paragraph{Array contraction}
+% -------------------------------------
+\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
 \begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
+MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
         \> $x_{ij}$    \> $\le$ \> $c_i$  \> \> \\
-           \> (for all edges from i)            \\
-           \> \ldots \\
+        \> (for all edges from i)
 \end{tabbing}
-As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
-An array is fully contracted if all its output edges are fused with it.
-Thus, $c_i=0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
-By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays.
-
-If an array is fully contracted, it means that the array does not need to be materialised in memory, and can simply be replaced by a scalar variable for the current element.
-Contracting arrays reduces memory traffic and can reduce overall memory usage.
+Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory.
+As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
+By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.
 

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
note on transitivity. make consistent section and subsection capitalisation

@@
 %!TEX root = ../Main.tex
-\section{Size inference}
+\section{Size Inference}
 Before performing fusion proper, we must infer the relative sizes of each array in the program.
@@
 % -----------------------------------------------------------------------------
-\subsection{Size types, constraints and schemes}
+\subsection{Size Types, Constraints and Schemes}
 \label{s:SizeTypes}
@@
 The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}.
-The top level judgment ~~$\SizeF{function}{\sigma}$~~ assigns a size scheme to a function.
+The top level judgment ~~$\SizeF{\function}{\sigma}$~~ assigns a size scheme to a function.
 It does this by extracting size constraints and then solving them.
@@
 % -----------------------------------------------------------------------------
-\subsection{Iteration size}
+\subsection{Iteration Size}
 After inferring the size of each array variable, each operator is assigned an \emph{iteration size}, which is the number of iterations needed in the loop which evaluates that operator.

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
note on transitivity. make consistent section and subsection capitalisation

@@
 % -----------------------------------------------------------------------------
-\subsection{The objective function}
+\subsection{The Objective Function}
 \label{s:ObjectiveFunction}
@@
 % -----------------------------------------------------------------------------
-\subsection{A note on transitivity}
+\subsection{Transitivity of Clustering}
 It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
-This means that for each pair $i,j$ we wouldn't generate an $x_{ij}$ and thus not generate those constraints, unless $i$ or $j$ is a @filter@, or arc between $i$ and $j$.
-This would help, but also means we can't \emph{minimise} on these removed $x_{ij}$, so the solution won't count the (rather smaller) benefits of fusing two non-connected nodes.
+This means that for each pair $i,j$ we would only generate an $x_{ij}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
+This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{ij}$.
+The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.
 
 % -----------------------------------------------------------------------------
-\subsection{Fusion-preventing path optimisation}
+\subsection{Fusion-preventing Path Optimisation}
 \label{s:OptimisedConstraints}

### 2014-05-16 2014betterfusionforfilters/section/06-Related.tex
note on transitivity. make consistent section and subsection capitalisation

@@
 %!TEX root = ../Main.tex
-\section{Related work}
+\section{Related Work}
 The idea of using integer linear programming to cluster an operator graph for array fusion was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}~(1999).

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
Sec 4 (sole transducer as def. for val. constr)

@@
 Although our constraint based formulation of size inference is reminiscent of type inference for HM(X)~\cite{odersky1999type}, there are important differences.
-Firstly, our type schemes include existential quantifiers, which represent the fact that the sizes of arrays produced by filter operations are unknown in general.
+Firstly, our type schemes include existential quantifiers, which express the fact that the sizes of arrays produced by filter operations are unknown in general.
 This is also the case for @generate@, where the result size is data dependent.
-HM(X) style type inferences uses the $\exists$ quantifier to bind local type variables in constraints, and existential quantifiers do not appear in type schemes.
+HM(X) style type inferences use the $\exists$ quantifier to bind local type variables in constraints, and existential quantifiers do not appear in type schemes.
 Secondly, our types are first order only, as program graphs cannot take other program graphs as arguments.
@@
 
-Note that the constraints are generated in a specific form, to make the constraint solving process easy.
+Note that the constraints are generated in a specific form, to facilitate the constraint solving process.
 For each array variable in the program, we generate a new size variable, like size $k_{zs}$ for array variable $zs$.
@@
 The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result.
-In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes and environment and constraint, producing a solved environment and constraint.
+In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes an environment and a constraint set, and produces a solved environment and constraint set.
 As the constraint solving process is both standard and straightforward, we only describe it informally.
@@
 Recall from the previous section that in our generated constraints all the size variables named after program variables are on the left of atomic equality constraints, while all the unification and existential variables are on the right.
-To solve the constraints we keep finding pairs of atomic equality constraints where the same variable appears on the left, unify the right of both of these constraints, and apply the resulting substitution to both the environment and original constraints.
-When there are no more pairs of constraints with the same variable on the left then the constraints are in solved form and we are finished.
+To solve the constraints, we keep finding pairs of atomic equality constraints where the same variable appears on the left, unify the right of both of these constraints, and apply the resulting substitution to both the environment and original constraints.
+When there are no more pairs of constraints with the same variable on the left, the constraints are in solved form and we are finished.
 
-During constraint solving, all unification variables mentioned in the environment can have other sizes substituted for them.
+During constraint solving, all unification variables occuring in the environment can have other sizes substituted for them.
 In contrast, the rigid variables marked by the $\exists$ symbol cannot.
-For example, if we consider the constraints for @normalize2@ mentioned before:
+For example, consider the constraints for @normalize2@ mentioned before:
 $$
@@
 
-To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately.
+To produce the final size scheme, we lookup the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately.
 This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.
-In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
+In this case, no rigid size variables appear in the result, so we can universally quantify all size variables.
 $$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
@@
 To say this another way, all subgraphs are fully inlined.
-Recall from \S\ref{s:CombinatorNormalForm} that we assume our operator graphs are embedded in a larger host program.
+Recall from \S\ref{s:CombinatorNormalForm}, that we assume our operator graphs are embedded in a larger host program.
 We use size information to guide the clustering process, and although the host program can certainly call the operator graph, static size information does not flow across this boundary.
@@
 We use $T$ to range over iteration sizes, and write $\bot$ for the case where the iteration size is unknown.
-The $\bot$ size is needed to handle the @external@ operator, as in general we cannot know its true iteration size, and it cannot be fused with any other operator.
+The $\bot$ size is needed to handle the @external@ operator, as we cannot statically infer its true iteration size, and it cannot be fused with any other operator.
 
@@
 Once the size constraints have been solved, we can use the following function to compute the iteration size of each binding.
-In the definition we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$.
+In the definition, we use the syntax $\Gamma(xs)$ to find the $xs : k$ element in the environment $\Gamma$ and return the associated size $k$.
 Similarly, we use the syntax $C(k)$ to find the corresponding $k = \tau$ constraint in $C$ and return the associated size type $\tau$.
@@
 We now define a function $trans$, to find the parent transducer a combinator application.
-Since each name is bound to at most one combinator, we abuse terminology here slightly and write ``combinator $n$'' when refering to the combinator occuring in the binding of the name $n$.
+Since each name is bound to at most one combinator, we abuse terminology here slightly and write \emph{combinator $n$} when refering to the combinator occuring in the binding of the name $n$.
 The parent transducer $trans(bs, n)$ of a combinator $n$ has the same output size as $n$'s iteration size, but the two have different iteration sizes.
@@
 In order for @sum1@ and @sum2@ to be fused together, @sum1@ and @gts@ must be fused, @sum1@ and @sum1@ must be fused, and @gts@ and @sum2@ must be fused.
+Now we can express the restriction on programs we view as valid for our transformation more formally:
 
-\textbf{Lemma: sole transducers}.
-For some function $f$ with valid constraints, each name $n$ will have at most one transducer.
+%\textbf{Lemma: sole transducers}.
+\TODO{Double check: is this really a sufficient property, or only necessary?}
+\textbf{Definition: valid constraints}.
+The constraints of a function $f$ are \emph{valid}, if and only if each name $n$ in its bindings will have at most one transducer:
 \[
@@
 
-These two lemmas are used in the integer linear programming formulation, when constructing the program.
+These two properties are used in the integer linear programming formulation, when constructing the program.
 When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
rewrite lemmas sole parents

--- 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
@@
 
-\textbf{Lemma: sole transducers}.
-For some function $f$ with valid constraints, each name $n$ will have at most one transducer.
-\[
-\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
-\]
-% \textbf{Proof:} by induction on $bs$. If $n = @map@_n~f~ns$,
-% then $trans(bs,n) = \bigcup_{x \in ns} trans(bs,x)$.
-% As @map@ requires its arguments to have the same size, and @filter@ introduces a fresh size for its output,
-% if any of the $trans(bs,x)$ are non-empty, they will refer to the same @filter@.
-% The other cases are trivial.
-
-% this is only useful for proving sole parents
-% \textbf{Lemma: transducers change types}.
-% For some bindings $bs$, if $n$ has a transducer, the iteration size is not the same.
-% \[
-% \forall bs, n, m.\ trans(bs, n) = \{m\} \implies \tau(n) \not= \tau(m)
-% \]
-
-\textbf{Lemma: sole parents}.
-% I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
-For some bindings $f$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
-\[
-\forall f, \sigma, a, b.\ f :_s \sigma \implies |parents(binds(f), a, b)| \le 1
-\]
-% \textbf{Proof:} since each binding only has at most one transducer, and transducers' iteration sizes are not the same, there will be at most one pair of transducers up the chain that have equal types.
-
-These two lemmas are used in the integer linear programming formulation, when constructing the program.
-When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.
+The size inference constraints reject programs with multiple transducers per node, so
+for valid programs, each name will have at most one transducer, and each pair of names will have at most one pair of parents.
+This simplifies the formulation, as it only needs to check one pair of parents.

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex

--- 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
@@
 $$
+Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma. \SizeF{\function}{\sigma}$.
 
@@
 
-%\textbf{Lemma: sole transducers}.
-\TODO{Double check: is this really a sufficient property, or only necessary?}
-\textbf{Definition: valid constraints}.
-The constraints of a function $f$ are \emph{valid}, if and only if each name $n$ in its bindings will have at most one transducer:
+\textbf{Lemma: sole transducers}.
+If a function $f$ is \emph{valid}, then its bindings will have at most one transducer:
 \[
@@
 
-These two properties are used in the integer linear programming formulation, when constructing the program.
+These two lemmas are used in the integer linear programming formulation, when constructing the program.
 When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

### 2014-05-16 2014betterfusionforfilters/Main.tex
erik suggestion on abstract

@@
 \begin{abstract}
-Minimising memory traffic is key to compiling functional, collection-oriented array programs into efficient code.
-It is not sufficient to simply fuse subsequent array operations into a single computation; we also have to cluster \emph{separate} traversals of the same array into a single loop.
+The key to compiling functional, collection oriented array programs into efficient code is to minimise memory trafffic.
+Simply fusing subsequent array operations into a single computation is not sufficient; we also need to cluster \emph{separate} traversals of the same array into a single traversal.
 Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
erik suggestion on abstract
rename split to possible - there is a possibility they may be fused

@@
 MMMMM       \= M  \= \kill
-$nodes$     \> @:@ \> $program \to V$              \\
-$edges$     \> @:@ \> $program \to E$              \\
+$nodes$     \> @:@ \> $\function \to V$              \\
+$edges$     \> @:@ \> $\function \to E$              \\
 $edge$      \> @:@ \> $\{bind\} \times bind \to E$ 
@@
 MM  \= MMMx \= M \= MMM \= M \= MMMM \= \kill
-    \> ~~~~~~~ $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N.\; x_{i,j}$ 
+    \> ~~~~~~~ $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
     \>             (with an edge from $i$ to $j$)            \\
-    \> $-N.\;  x_{i,j}$  \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N.\; x_{i,j}$ 
+    \> $-N \cdot  x_{i,j}$  \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ 
     \>             (with no edge from $i$ to $j$)
@@
 % -----------------------------------------------------------------------------
-\subsection{The Objective Function}
+\subsection{Objective Function}
 \label{s:ObjectiveFunction}
@@
 MMMMM   \= MMMM \= M \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
+Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij}$   \\
            \> \> \> (memory traffic and loop overhead)         \\
-           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
+           \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$  \\
            \> \> \> (removing intermediate arrays)         \\
@@
 
-For the graph, we define a function $split(a,b)$ to check whether there exists a fusion-preventing path between two nodes, $a$ and $b$.
+For the graph, we define a function $possible(a,b)$ to check whether there exists no fusion-preventing path between two nodes, $a$ and $b$.
+If there is no fusion-preventing path, then fusion between the two nodes is possible.
+We also define a function $possible'(a,b)$, to check whether the pair has parents that may also be fused.
 
 \begin{tabbing}
-MMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
-$split$     \> @::@ \> $name \times name \to \mathbb{B}$      \\
+MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
+$possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
 MM        \= M    \= \kill
-$split(a,b)$ \\
+$possible(a,b)$ \\
     \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$\\
+\\
+MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
+$possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
+MM        \= M    \= \kill
+$possible'(a,b)$ \\
+    \>$=$  \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
+    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
 \end{tabbing}
 
-With $split$ defined, we can refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
+With $possible$ and $possible'$ defined, we can refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
 The entire formulation of the integer linear program follows.
@@
 MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij} + \Sigma_{i \in V} N c_i$  \\
-           \> (if $split(i,j)$)         \\
+Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij} + \Sigma_{i \in V} N \cdot c_i$  \\
+           \> (if $possible(i,j)$)         \\
 Subject to \\
-           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \> (if $split(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
+           \> $-N \cdot x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
+           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
 \\
-           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
-           \> (if $split(i,j) \wedge (i,j,\fusible) \in E$)     \\
+           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
+           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     \\
 \\
@@
            \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (if $split(i,j)$) \\
+           \> (if $possible(i,j)$) \\
            \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
-           \> (if $\neg(split(i,j))$) \\
-\\
-           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \emptyset$)  \\
+           \> (if $\neg possible(i,j)$) \\
            \\
@@
            \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge parents(i,j) = \{i',j'\} \wedge$ \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$) \\
 
-           \> \> $ split(i,j) \wedge split(i',j) \wedge split(i,j') \wedge split(i',j') $) \\
+\\
+           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\
 MMMMM   \= MMMM \= M \= \kill

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
remove normalize2 code from transducer 3.6, get trans defn to fit without break

@@
 Even though $k_{gts}$ is not equal to $k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
-The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
-Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
-\begin{code}
- normalize2 xs
-  = let sum1 = fold   (+)  0   xs
-        gts  = filter (> 0)    xs
-        sum2 = fold   (+)  0   gts
-        ...
-\end{code}
+% The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
+% Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
 
@@
 MMMM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
-$trans$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
+$trans$  \>$:$\> $binds \rightarrow name \rightarrow \{name\}$ \\
 $trans(bs,o)$    \\
@@
 To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
+
 \begin{tabbing}
 MMMM \= M \= M \= M \= \kill
-$parents$ \> $::$ \> $binds \to name \to name \to \{name \times name\}$ \\
+$parents$ \> $:$ \> $binds \to name \to name \to \{name \times name\}$ \\
 $parents(bs, a, b)$ \\

### 2014-05-16 2014betterfusionforfilters/section/03-SizeInference.tex
mention size inf failure
size inf fail
spacing and typo

@@
 In constrast to our algorithm, \cite{chatterjee1991size} does not support size changing functions such as filter.
+If size inference fails, the programs may still be compiled, but fusion is not performed.
 
@@
 $$
-Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma. \SizeF{\function}{\sigma}$.
+Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\function}{\sigma}$.
 
@@
 
-To produce the final size scheme, we lookup the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately.
+To produce the final size scheme, we look up the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately.
 This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
fix possible
remove normalize2 code from transducer 3.6, get trans defn to fit without break
dep graph
consistent commas in x_{i,j}

@@
 $edges$     \> @:@ \> $\function \to E$              \\
-$edge$      \> @:@ \> $\{bind\} \times bind \to E$ 
-\\[1ex]
-$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       
-\\[2ex]
-
-$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    
-\\[1ex]
+$edge$      \> @:@ \> $\{bind\} \times bind \to E$   \\
+$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$   \\
+\\
+$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
+\\
+$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
+\\
 MM             \= M \= \kill
@@
 $edge(bs, outs = @external@~ins)$  \\
-    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ 
-\\[1ex]
+    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
+\\
 $inedge(bs,to,\from)$ \\
@@
 \end{tabbing}
-For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{ij}$ which indicates whether those two nodes are fused.
-We use $x_{ij} = 0$ when the nodes are fused and $x_{ij} = 1$ when they are not.
-Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{ij}$ variables, and minimizing it tends to increase the number of nodes that are fused.
-The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{ij} = 0 \iff cluster(i) = cluster(j)$.
+For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{i,j}$ which indicates whether those two nodes are fused.
+We use $x_{i,j} = 0$ when the nodes are fused and $x_{i,j} = 1$ when they are not.
+Using $0$ for the fused case means that the objective function can be a weighted function of the $x_{i,j}$ variables, and minimizing it tends to increase the number of nodes that are fused.
+The values of these variables are used to construct the final clustering, such that $\forall i,j.\ x_{i,j} = 0 \iff cluster(i) = cluster(j)$.
 \begin{tabbing}
@@
 Each node is assigned a variable $c_i$ that indicates whether the array the associated binding produces is \emph{fully contracted}.
-When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
+When an array is fully contracted it means that all consumers of that array are fused into the same cluster, so we have $c_i = 0 \iff \forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$.
 In the final program, each successive element of a fully contracted array can be stored in a scalar register, rather than requiring an array register or memory storage.
@@
 MMM     \= MMM \= M \= MMM \= M \= MMM \= \kill
-        \> $x_{ij}$    \> $\le$ \> $c_i$  \> \> \\
+        \> $x_{i,j}$    \> $\le$ \> $c_i$  \> \> \\
         \> (for all edges from i)
@@
 Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory.
-As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{ij} = 0$.
+As per Darte's work on array contraction~\cite{darte2002contraction}, we define a variable $c_i$ for each array, and the constraint above ensures that $c_i = 0$ only if $\forall (i',j) \in E.\ i = i' \implies x_{i,j} = 0$.
 By minimizing $c_i$ in the objective function, we favor solutions that reduce the number of intermediate arrays.
@@
 The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
-\TODO{Refer to example code that demonstrates these different opportunities.}
+% \TODO{Refer to example code that demonstrates these different opportunities.}
 
@@
 MMMMM   \= MMMM \= M \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij}$   \\
+Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   \\
            \> \> \> (memory traffic and loop overhead)         \\
@@
 Where      \>                                       \\
-           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
+           \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
            \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
-           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
+           \> $W_{i,j} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
            \> \> \> ($i$ and $j$ share an input array)                                         \\
-           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
+           \> $W_{i,j} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
            \> \> \> (the only benefit is loop overhead)                                        \\
@@
 \subsection{Transitivity of Clustering}
-It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{ij}$.
-This means that for each pair $i,j$ we would only generate an $x_{ij}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
-This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{ij}$.
+It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{i,j}$.
+This means that for each pair $i,j$ we would only generate an $x_{i,j}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
+This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{i,j}$.
 The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.
@@
 \begin{tabbing}
-MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
-Minimise   \> $\Sigma_{(i,j) \in E} W_{ij} \cdot x_{ij} + \Sigma_{i \in V} N \cdot c_i$  \\
+MMMMM   \= MMMM \= M \= MMMM \= M \= MMMM \= \kill
+Minimise   \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j} + \Sigma_{i \in V} N \cdot c_i$  \\
            \> (if $possible(i,j)$)         \\
 Subject to \\
-           \> $-N \cdot x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
+           \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
            \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
 \\
-           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{ij}$ \\
+           \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
            \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     \\
@@
 \\
-           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
+           \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
            \> (if $possible(i,j)$) \\
@@
            \\
-           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
            \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
            \\
-           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$) \\
+           \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
+           \> $x_{j',j}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
+           \> $x_{i',j'}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$ \\
+           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) \\
 
 \\
-           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
+           \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
            \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
Edits to objective function

@@
 
-\eject
 % -----------------------------------------------------------------------------
@@
 Our linear constraints will ensure that if two nodes are fused into the same cluster then their $\pi$ values will be identical --- though nodes in different clusters can also have the same $\pi$ value.
-
 Here is an example of a cyclic clustering:
@@
 \subsection{Linear Constraints}
+\label{s:LinearConstraints}
 The constraints we place on the ILP variables are split into four groups: constraints that ensure the clustering is acyclic; constraints that encode fusion preventing edges; constraints on nodes with different iteration sizes, and constraints involving array contraction.
@@
 \label{s:ObjectiveFunction}
-
-To find the objective function, note that fusing loops can have three main benefits:
+The objective function defines the cost model of the program, and the ILP solver will find the clustering that minimizes this function while satisfying the constraints defined in the previous section.
+The cost model we use in this paper has three components:
 \begin{itemize}
 \item
-reducing memory traffic, such as multiple loops reading from the same array;
+the number of array reads and writes --- an abstraction of the amount of memory bandwidth needed by the program; 
 \item
-removing intermediate arrays, thus reducing the amount of memory required;
+the number of intermediate arrays --- an abstraction of the amount of intermediate memory needed; 
 \item
-and reducing loop overhead, such as when two loops operate on different arrays of the same size.
+the number of distinct clusters --- an abstraction of the cost of loop management instructions, which maintain loop counters and the like.
 \end{itemize}
-However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
+The three components of the cost model are a heuristic abstraction of the true cost of executing the program on current hardware.
+They are ranked in order of importance --- so we prefer to minimize the number of array reads and writes over the number of intermediate arrays, and to minimize the number of intermediate arrays over the number of clusters.
+However, minimizing one component does not necessarily minimize any other.
+For example, as the fused program executes multiple array operations at the same time, in some cases the clustering that requires the least number of array reads and writes uses more intermediate arrays than strictly necessary.
+
+We encode the ordering of the components of the cost model as different weights in the objective function.
+First, note that if the program graph contains $N$ combinators (nodes) then there are at most $N$ opportunities for fusion.
+We then encode the relative cost of loop overhead as weight $1$, the cost of an intermediate array as weight $N$, and the cost of an array read or write as weight $N^2$.
+This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array, and likewise no number of removed intermediate arrays can outweigh a reduction in the number of array reads or writes.
+The final integer linear program is as follows:
+\begin{tabbing}
+MMMMM   \= MMMM \= M \= \kill
+Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   
+\\         \> \> \> (memory traffic and loop overhead)              
+\\         \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$                  
+\\         \> \> \> (removing intermediate arrays)         
+\\ Subject to \> \ldots ~ constraints from \S\ref{s:LinearConstraints} ~ \ldots 
+\\ Where      \>                                       
+\\         \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         
+\\         \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         
+\\         \> $W_{i,j} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     
+\\         \> \> \> ($i$ and $j$ share an input array)
+\\         \> $W_{i,j} = 1$   \> $~|$ \> $@otherwise@$
+\\         \> \> \> (the only benefit is loop overhead)
+\\         \> $N = |V|$
+\end{tabbing}
+
+% However, these benefits cannot be considered in isolation; for example, fusing two loops to reduce loop overhead may remove potential fusion opportunities that reduce memory traffic.
 When operating on large arrays that do not fit in cache, memory traffic dominates execution time.
@@
 The benefits of removing loop overhead are least of all; it should be performed if possible, but must never remove opportunities for other fusion.
-% \TODO{Refer to example code that demonstrates these different opportunities.}
-
-This total ordering can be encoded into an ILP objective function as weights.
-If the program graph contains $N$ combinators, then there are at most $N$ opportunities for fusion.
-The encoding of loop overhead is weight $1$, removing intermediate arrays is weight $N$, and reducing memory traffic is weight $N^2$.
-This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array,
-and likewise no number of removed intermediate arrays can outweigh a reduction in memory traffic.
-Although, it is worth noting that reducing memory traffic does \emph{tend} to remove intermediate arrays, and vice versa.
-
-\begin{tabbing}
-MMMMM   \= MMMM \= M \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   \\
-           \> \> \> (memory traffic and loop overhead)         \\
-           \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$  \\
-           \> \> \> (removing intermediate arrays)         \\
-Subject to \> \ldots                                \\
-Where      \>                                       \\
-           \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
-           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
-           \> $W_{i,j} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
-           \> \> \> ($i$ and $j$ share an input array)                                         \\
-           \> $W_{i,j} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
-           \> \> \> (the only benefit is loop overhead)                                        \\
-           \\
-           \> $N = |V|$
-\end{tabbing}
 
@@
 $possible(a,b)$ \\
-    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$\\
-\\
+    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
+\\[1ex]
 MMMM \= M \=     \= MMMMMMMMMMMM    \=  \kill
@@
            \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
-           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            \\
-\\
+           \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            
+\\[1ex]
            \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
-           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     \\
-\\
+           \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     
+\\[1ex]
            \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
-           \> (if $(i,j,\fusionpreventing) \in E$)    \\
-\\
+           \> (if $(i,j,\fusionpreventing) \in E$)    
+\\[1ex]
            \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
@@
            \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
-           \> (if $\neg possible(i,j)$) \\
-           \\
+           \> (if $\neg possible(i,j)$) 
+\\[1ex]
            \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  \\
-           \\
+           \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  
+\\[1ex]
            \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
@@
            \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge possible'(i,j)$ \\
-           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) \\
-
-\\
+           \> \> $\wedge~parents(i,j) = \{(i',j')\}$) 
+\\[1ex]
            \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
@@
            \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
-           \> \> \> (the only benefit is loop overhead)                                        \\
-           \\
+           \> \> \> (the only benefit is loop overhead)                                        
+\\[1ex]
            \> $N = |V|$

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
kill 4.5 transitivity

@@
 % -----------------------------------------------------------------------------
-\subsection{Transitivity of Clustering}
-It may seem that we could generate far fewer constraints and rely on transitivity of clustering equality $x_{i,j}$.
-This means that for each pair $i,j$ we would only generate an $x_{i,j}$ and constraints, if $i$ or $j$ is a @filter@, or there is an edge between $i$ and $j$.
-This would simplify the integer linear program, but would also modify the objective function not to \emph{minimise} on these removed $x_{i,j}$.
-The objective function of this simplified version would not count the benefits of fusing two non-connected nodes, and would not reduce memory traffic as effectively.
-
-% -----------------------------------------------------------------------------
 \subsection{Fusion-preventing Path Optimisation}

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
Squash onto 10 pages
fix bench normalize2 constraints
rename split to possible - there is a possibility they may be fused
benches talk about closest and quad

@@
 To make the program shorter and easier to work with, we use the optimised version.
-First, we must calculate $split$ -- that is, those nodes which have no fusion-preventing path between them.
+First, we must calculate $possible$ -- that is, those nodes which have no fusion-preventing path between them.
 Any elements in the following sets have no fusion-preventing paths.
@@
 \\
-    \> $x_{sum1,sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
-    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$ \\
-    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$ \\
+    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
+    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
+    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
 \end{tabbing}
@@
 MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$x_{sum1, gts}, x_{sum1, sum2}, x_{gts, sum2}, x_{ys1,  ys2}$
+$x_{sum1, gts}, x_{sum1, sum1}, x_{sum1, sum2}, x_{gts, sum2}, x_{ys1,  ys2}$
     \> $=$ \> $0$ \\
@@
 $c_{gts}, c_{ys1}, c_{ys2}$           
-    \> $=$ \> $0$ \\
+    \> $=$ \> $0$
 \end{tabbing}
@@
 $x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
-    \> $=$ \> $1$ \\
+    \> $=$ \> $1$
 \end{tabbing}
@@
     \> $=$ \> $1$ \\
-\\
 $c_{gts}, c_{ys1}, c_{ys2}$           
-    \> $=$ \> $0$ \\
+    \> $=$ \> $0$
 \end{tabbing}
@@
 
-The objective function for this solution is
+The objective function's value is
 $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
 
+\subsection{Closest Points}
+The closest points benchmark is a divide-and-conquer algorithm that, given an array of points, finds the closest pair of points.
+A midpoint is found, and the points are filtered into those above the midpoint, and those below.
+The closest points in the filtered arrays are then recursively computed, and the two results merged.
+As the filters are operated on recursively, there are no opportunities for fusing the filters, making our clustering the same as Megiddo's.
+Our clustering is able to generate both filtered arrays in a single loop, unlike stream fusion, which requires a separate loop for each.
+
+\subsection{QuadTree}
+The QuadTree benchmark recursively computes a 2-dimensional space partitioning tree of an array of points.
+At each step, the array of points is filtered into four boxes.
+As with closest points, there are no opportunities for fusing the filters, and our clustering is the same as Megiddo's.
+Our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.
+
 \subsection{QuickHull}
-The core of the QuickHull algorithm is very similar to @hull@, below.
+One benchmark we did not implement, but is worth mentioning, is QuickHull.
+The core of the QuickHull algorithm is shown below.
 Given a line and an array of points, the points are filtered to those above the line,
@@
 \begin{code}
+
 hull :: (Point,Point) -> Array Point -> Array Point
@@
        ma   = fold   (maxFrom line) pts'
-       hl   = hull   (l, ma)        pts'
-       hr   = hull   (ma, r)        pts'
-   in  hl  ++ hr
+   in (hull (l, ma) pts') ++ (hull (ma, r) pts')
 \end{code}
 
-Stream fusion is unable to fuse the @pts'@ and @ma@ together, as @pts'@ is used multiple times and cannot be inlined into @ma@.
-Similarly, Megiddo~\cite{megiddo1998optimal} is unable to fuse them together, as their iteration sizes are different.
+Stream fusion cannot fuse @pts'@ and @ma@ together, as @pts'@ is used multiple times and cannot be inlined into @ma@.
+Megiddo cannot fuse them either, as their iteration sizes are different.
 Our formulation is able to fuse these together.
-
 If the @fold@ is changed to operate over @pts@ instead of @pts'@, Megiddo's formulation \emph{is} able to fuse the two.

### 2014-05-16 2014betterfusionforfilters/section/04-ILP.tex
Copy edits to S4.6

@@
 This ensures that no amount of loop overhead reduction can outweigh the benefit of removing an intermediate array, and likewise no number of removed intermediate arrays can outweigh a reduction in the number of array reads or writes.
-The final integer linear program is as follows:
+The integer linear program including the objective function is as follows:
 \begin{tabbing}
@@
 \label{s:OptimisedConstraints}
-We can remove certain constraints and variables, by only checking nodes if there is no \emph{fusion-preventing path} between $i$ and $j$.
-If there is a fusion-preventing path between two nodes, they are known to be in different clusters.
-With our specific knowledge of the problem domain, we can simplify the integer linear program and make the job of the ILP solver easier by omitting these constraints.
-Creating fewer constraints and variables should make it faster for the ILP solver to find a solution.
-
-For the graph, we define a function $possible(a,b)$ to check whether there exists no fusion-preventing path between two nodes, $a$ and $b$.
-If there is no fusion-preventing path, then fusion between the two nodes is possible.
-We also define a function $possible'(a,b)$, to check whether the pair has parents that may also be fused.
-
+The integer linear program defined in the previous section includes more constraints than strictly necessary to define the valid clusterings.
+If two nodes have a path between them which includes a fusion preventing edge, then we know up front that they must be placed in different clusters.
+The following function $possible(a,b)$ determines whether there is any possibility that the two nodes $a$ and $b$ can be fused.
+Similarly the function $possible'(a, b)$ checks whether there is any possibility that the parents of $a$ and $b$ may be fused.
 \begin{tabbing}
@@
 \end{tabbing}
-
-With $possible$ and $possible'$ defined, we can refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
-The entire formulation of the integer linear program follows.
+With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.
+Doing this reduces the total number of constraints, and makes the job of the ILP solver easier.
+The final formulation of the integer linear program follows.
 

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
quickhull

@@
 \subsection{QuickHull}
-One benchmark we did not implement, but is worth mentioning, is QuickHull.
-The core of the QuickHull algorithm is shown below.
-Given a line and an array of points, the points are filtered to those above the line,
-and the farthest point above the line is also found.
+QuickHull is another program that is worth mentioning, as it highlights some fusion opportunities with filters.
+The core of the QuickHull algorithm is shown below: given a line and an array of points, the points are filtered to those above the line, and the farthest point above the line is also found.
 
 \begin{code}
-
 hull :: (Point,Point) -> Array Point -> Array Point

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
Edits to benchmark intro

@@
 
-We have run three benchmarks to constrast our clustering algorithm, Megiddo~\cite{megiddo1998optimal}'s clustering, the clustering produced by stream fusion~\cite{coutts2007streamfusion}, and no fusion at all.
-The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree.
+This section discusses three representative benchmarks, and gives the full ILP program of the first.
+These benchmarks highlight the main differences between our fusion mechanism and related work.
+The runtimes of each benchmark are summarized in Figure~\ref{f:BenchResults}.
+We report times for: the unfused case where each operator is assigned to its own cluster; the clustering implied by stream fusion~\cite{coutts2007streamfusion}; the clustering chosen by Megiddo~\cite{megiddo1998optimal}, and the clustering chosen by our system.
+
+For each benchmark we report the runtimes of hand-fused C code based on the clustering determined by each algorithm.
+Although we also have an implementation of our Data Flow Fusion system in terms of a GHC plugin~\cite{lippmeier2013flow}, we report on hand-fused C code to provide a fair comparison to related work.
+As mentioned in~\cite{lippmeier2013flow}, the current Haskell stream fusion mechanism introduces overhead in terms of a large number of duplicate loop counters, which increases register pressure unnecessarily.
+Hand fusing all code and compiling it with the same compiler (GCC) isolates the true cost of the various clusterings from low level differences in code generation.
+
+The benchmark programs are at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}.
+
+% The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree.
 The benchmark programs are all hand-written, hand-fused C code based on the clustering.
@@
 Interestingly, stream fusion's clustering for @normalize2@ performs better than Megiddo's, despite having more loops, as stream fusion is able to remove the intermediate array.
-
-Benchmark programs are available at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}.
 

### 2014-05-16 2014betterfusionforfilters/section/06-Related.tex
related work: darte formulation

@@
 The idea of using integer linear programming to cluster an operator graph for array fusion was first fully described by Megiddo and Sarkar~\cite{megiddo1998optimal}~(1999).
-This algorithm was then extended to support multiple iteration sizes by Darte and Huard~\cite{darte2002contraction}.
+A simpler formulation, supporting only loops of the same iteration size, but optimizing for array contraction, was then described by Darte and Huard~\cite{darte2002contraction}~(2002).
 Both algorithms were developed in the context of imperative languages (Fortran) and are based around a Loop Dependence Graph (LDG).

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
Copy edits to S5.1: normalize

@@
 Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
-Closest pair    & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
+Closest points  & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
 QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Normalize2}
-To demonstrate the ILP formulation, the clustering of @normalize2@ is derived as follows.
+To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
 \begin{code}
@@
 \end{code}
-
-To make the program shorter and easier to work with, we use the optimised version.
-First, we must calculate $possible$ -- that is, those nodes which have no fusion-preventing path between them.
-Any elements in the following sets have no fusion-preventing paths.
+We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
 \[ \{ \{sum1, gts, sum2\}
@@
 
-Note that, in the objective function, the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.
+The complete ILP program is shown below. Note that in the objective function the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.
 
@@
            \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
-           \> $ 25 \cdot x_{ys1, ys2} + 5 \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
+           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
 Subject to \\
@@
     \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
-    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ \\
-\\
-    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ \\
-\\
+    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
+\\[1ex]
+    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
+\\[1ex]
     \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
-    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$ \\
-\\
-    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$ \\
-\\
+    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
+\\[1ex]
+    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
+\\[1ex]
     \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
@@
 
+\noindent
 One minimal solution to this is:
@@
 MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$x_{sum1, gts}, x_{sum1, sum1}, x_{sum1, sum2}, x_{gts, sum2}, x_{ys1,  ys2}$
+$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
     \> $=$ \> $0$ \\
-$x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
-    \> $=$ \> $1$ \\
-\\
-$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
+$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
+    \> $=$ \> $1$ 
+\\[1ex]
+$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
     \> $=$ \> $0$ \\
-$\pi_{ys1 }, \pi_{ys2 }$
-    \> $=$ \> $1$ \\
-\\
-$c_{gts}, c_{ys1}, c_{ys2}$           
+$\pi_{ys1 },~ \pi_{ys2 }$
+    \> $=$ \> $1$ 
+\\[1ex]
+$c_{gts},~ c_{ys1},~ c_{ys2}$           
     \> $=$ \> $0$
 \end{tabbing}
-This minimal solution is not unique, however in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.
+This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.
+Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
+For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints.
+Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}.
+These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.
 
-Looking at just the non-zero variables in the objective function, the value is
-$25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
-It is worth noting that, for example, the objective could be reduced by fusing $x_{sum1,ys2}$ together.
-However, this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, the constraint requires that $\pi_{sum1} = \pi_{sum2}$, and another constraint requires $\pi_{sum2} < \pi_{ys2}$.
-These constraints may not all hold, so a clustering that fused sum1 and ys2 together would not allow sum1 and sum2 to be fused together.
-
-We can also calculate the objective function for the clustering used by stream fusion, to contrast and show that the constraints are valid.
-To do this, the values of $x_{ij}$ variables are:
+We will now compare the clustering produced by our system, with the one implied by stream fusion.
+As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship.
+The corresponding values of the $x_{ij}$ variables are:
 \begin{tabbing}
@@
 \end{tabbing}
-If we add these extra equality constraints to the integer linear program and solve it, one possible solution is:
+We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
 \begin{tabbing}
@@
 \end{tabbing}
-It is interesting to note that two nodes having equal $\pi_i$ values does not imply that the two nodes are fused together.
-Conversely, however, different $\pi_i$ values do imply that two nodes are not fused together.
+Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused.
+Conversely, if two nodes have different $\pi$ values then they are never fused.
+The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
 
-The objective function's value is
-$25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
-
+% -----------------------------------------------------------------------------
 \subsection{Closest Points}
@@
 
+% -----------------------------------------------------------------------------
 \subsection{QuadTree}
@@
 
+% -----------------------------------------------------------------------------
 \subsection{QuickHull}

### 2014-05-16 2014betterfusionforfilters/Main.tex
typo in abstract...
More desc to benchmarks

@@
 \begin{abstract}
-The key to compiling functional, collection oriented array programs into efficient code is to minimise memory trafffic.
+The key to compiling functional, collection oriented array programs into efficient code is to minimise memory traffic.
 Simply fusing subsequent array operations into a single computation is not sufficient; we also need to cluster \emph{separate} traversals of the same array into a single traversal.
@@
 Robert Clifton-Everest,
+Erik de Castro Lopo,
 Kai Engelhardt,

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
More desc to benchmarks

@@
            \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
-           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
-Subject to \\
+           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
+\\[1ex]
+Subject to 
     \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
@@
 \end{tabbing}
-
 \noindent
@@
 \end{tabbing}
-Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused.
-Conversely, if two nodes have different $\pi$ values then they are never fused.
-The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
+Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.
+
+% The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
 
@@
 \subsection{Closest Points}
-The closest points benchmark is a divide-and-conquer algorithm that, given an array of points, finds the closest pair of points.
-A midpoint is found, and the points are filtered into those above the midpoint, and those below.
-The closest points in the filtered arrays are then recursively computed, and the two results merged.
-As the filters are operated on recursively, there are no opportunities for fusing the filters, making our clustering the same as Megiddo's.
-Our clustering is able to generate both filtered arrays in a single loop, unlike stream fusion, which requires a separate loop for each.
+The closest points benchmark is a divide-and-conquer algorithm that finds the closest pair of 2-dimensional points in an array.
+We first find the midpoint along the Y-axis, and filter the remaining points to those above and below the midpoint.
+We then recursively find the closest pair of points in the two halves, and merge the results.
+As the filtered points are passed directly to the recursive call, there is no further opportunity to fuse them, and our clustering is the same as returned by Megiddo's algorithm.
+However, our clustering generates both filtered arrays in a single loop, unlike stream fusion that requires a separate loop for each.
 
@@
 \subsection{QuadTree}
-The QuadTree benchmark recursively computes a 2-dimensional space partitioning tree of an array of points.
-At each step, the array of points is filtered into four boxes.
-As with closest points, there are no opportunities for fusing the filters, and our clustering is the same as Megiddo's.
-Our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.
+The QuadTree benchmark recursively builds a 2-dimensional space partitioning tree from an array of points.
+At each step the array of points is filtered into four 2-dimensional boxes.
+As with the closest points algorith, there are no further opportunities for fusing the filtered results, and our clustering is the same as Megiddo's.
+However, our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.
 
@@
 \subsection{QuickHull}
-QuickHull is another program that is worth mentioning, as it highlights some fusion opportunities with filters.
-The core of the QuickHull algorithm is shown below: given a line and an array of points, the points are filtered to those above the line, and the farthest point above the line is also found.
+The core of the QuickHull algorithm is shown below: given a line and an array of points, we filter the points to those above the line, and also find the point farthest from that line.
 
@@
 
-Stream fusion cannot fuse @pts'@ and @ma@ together, as @pts'@ is used multiple times and cannot be inlined into @ma@.
-Megiddo cannot fuse them either, as their iteration sizes are different.
-Our formulation is able to fuse these together.
-If the @fold@ is changed to operate over @pts@ instead of @pts'@, Megiddo's formulation \emph{is} able to fuse the two.
-However, a programmer unaware of fusion would expect the original version to be faster, because @pts'@ is smaller than @pts@.
+Stream fusion cannot fuse the @pts'@ and @ma@ bindings because @pts'@ is referred to multiple times and thus cannot be inlined.
+Megiddo's algorithm also cannot fuse the two bindings because their iteration sizes are different.
+If the @ma@ binding was rewritten to operate over the @pts@ array instead of @pts'@, Megiddo's formulation would be able to fuse the two, and the overall program would give the same result.
+However, this performance behavior is counter intuitive because @pts'@ is likely to be smaller than @pts@, so in an unfused program the original version would be faster.
+Our system fuses both versions.

### 2014-06-17 2014betterfusionforfilters/section/01-Introduction.tex
minor changes

@@
 \begin{code}
- normalize2 :: Array Int -> Array Int
+ normalize2 :: Array Int -> (Array Int, Array Int)
  normalize2 xs

### 2014-06-17 2014betterfusionforfilters/section/03-SizeInference.tex
minor changes

@@
 We use the latter to represent the result size of the @cross@ operator discussed in the previous section.
-Constraints maybe either be trivially $\true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$.
+Constraints may either be trivially $\true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$.
 We refer to the trivially true and equality constraints as \emph{atomic constraints}.
@@
 
-Note that $k_xs$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$. Eliminating the duplicates, as well as the trivially $\true$ terms then yields:
+Note that $k_{xs}$ is mentioned twice on the right of an atomic equality constraint, so we can substitute $k_2$ for $k_3$.
+Eliminating the duplicates, as well as the trivially $\true$ terms then yields:
 $$
@@
 % No,
-We now define a function $trans$, to find the parent transducer a combinator application.
+We now define a function $trans$, to find the parent transducer of a combinator application.
 Since each name is bound to at most one combinator, we abuse terminology here slightly and write \emph{combinator $n$} when refering to the combinator occuring in the binding of the name $n$.

### 2014-06-17 2014betterfusionforfilters/section/04-ILP.tex
quickhull
fix c_i in 4.5
More desc to benchmarks
Spacing fix
minor changes

@@
         sum1 = fold (+) 0    us
-        ys   = map  (/ sum) incs
+        ys   = map  (/ sum1) incs
     in  ys
@@
 
-For small programs such as @normalizeInc@ it is possible to naively enumerate all possible clusterings, select just those that are \emph{valid} with respect to fusion preventing edges, and chose the one that maximises a cost metric such as the number of array accesses needed.
+For small programs such as @normalizeInc@ it is possible to naively enumerate all possible clusterings, select just those that are \emph{valid} with respect to fusion preventing edges, and choose the one that maximises a cost metric such as the number of array accesses needed.
 However, as the program size increases the number of possible clusterings becomes too large to naively enumerate.
@@
 The dependency graph contains enough information to determine the possible clusterings of the input program, while abstracting away from the exact operators used to compute each intermediate array.
-The rules for producing a dependency graphs are in Figure~\ref{f:DependencyGraph}.
+The rules for producing dependency graphs are in Figure~\ref{f:DependencyGraph}.
 
@@
 Similarly the function $possible'(a, b)$ checks whether there is any possibility that the parents of $a$ and $b$ may be fused.
+
+\newpage
+
 \begin{tabbing}
@@
 Minimise   \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j} + \Sigma_{i \in V} N \cdot c_i$  \\
-           \> (if $possible(i,j)$)         \\
-Subject to \\
-           \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
+           \> (if $possible(i,j)$)         
+\\[1ex]
+Subject to \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
            \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            
@@
            \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
-           \> (if $possible(i,j)$) \\
+           \> (if $(i,j, \fusible) \in E$) \\
            \> $c_{i }$    \> $ = $ \> $ 1 $           \>       \>            \\
-           \> (if $\neg possible(i,j)$) 
+           \> (if $(i,j, \fusionpreventing) \in E$)
 \\[1ex]
@@
            \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
-           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) \\
+           \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) 
+\\[1ex]
 MMMMM   \= MMMM \= M \= \kill
-Where      \>                                       \\
-           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
+Where      \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
            \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
@@
            \> \> \> (the only benefit is loop overhead)                                        
-\\[1ex]
+\\
            \> $N = |V|$

### 2014-06-19 2014betterfusionforfilters/section/07-Conclusion.tex
conclusion

%!TEX root = ../Main.tex
\section{Conclusion}

We have shown that using integer linear programming to find clusterings is able to produce predictably better clusterings than heuristic and greedy approaches.
Having a predictable clustering algorithm is particularly important when combined with other transformations, such as the vectorisation done by Data Parallel Haskell~\cite{chakravarty2007data}.
In the case of Data Parallel Haskell, the actual combinators to be clustered are not written directly by the programmer, so being able to find a good clustering without the programmer tweaking the combinators is important.

One obvious shortcoming of this work is the limited selection of combinators.
Other combinators can currently be used as external computations, but this is not ideal as they will not be fused.
Combinators such as @append@, @scan@ and @slice@ would be simple to add.

Implementing @length@ is particularly interesting, as it does not require the array to be manifest, but does require some array with the same rate to be manifest.
For example, finding the length of the output of a filter can only be done after the filter is computed.
Once @length@ is implemented, more interesting functions such as @reverse@ can be implemented as a @generate@ followed by a @gather@.

For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold, segmented map, which operate on segmented representations of nested arrays.
Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
For example, segmented fold takes a segmented array, and returns a single array of the outer rate.

### 2014-06-19 2014betterfusionforfilters/section/01-Introduction.tex
mention sequential & parallel

@@
 
-To compile the clusters found by our clustering technique into loops, we use data flow fusion~\cite{lippmeier2013flow}.
-It improved on existing array fusion approaches~\cite{coutts2007streamfusion, keller2010repa} as it guarantees complete fusion into a single loop for programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+To compile the clusters found by our clustering technique into sequential loops, we use data flow fusion~\cite{lippmeier2013flow}.
+It improved on existing array fusion approaches~\cite{coutts2007streamfusion, keller2010repa} as it guarantees fusion into a single loop for programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
+Our clustering technique is also applicable to parallel loops.
 

### 2014-06-19 2014betterfusionforfilters/section/06-Related.tex
minor changes
mention rompf

@@
 Kennedy~\cite{kennedy2001fastgreedy} describes another greedy approach which tries to maximize the reuse of intermediate arrays, and Song~\cite{song2004improving} tries to reduce memory references.
+
+Greedy and heuristic approaches that operate on lists of bindings rather than the graph, such as Rompf~\cite{rompf2013optimizing}, can find optimal clusterings in some cases, but are subject to changes in the order of bindings.
+In these cases, reordering bindings can produce a different clustering, leading to unpredictable runtime performance.
 

### 2014-06-19 2014betterfusionforfilters/section/03-SizeInference.tex
rename function in CNF definition to program: should reduce confusion about function vs worker function

@@
 The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}.
-The top level judgment ~~$\SizeF{\function}{\sigma}$~~ assigns a size scheme to a function.
+The top level judgment ~~$\SizeF{\program}{\sigma}$~~ assigns a size scheme to a function.
 It does this by extracting size constraints and then solving them.
@@
 $$
-Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\function}{\sigma}$.
+Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.
 

### 2014-06-20 2014betterfusionforfilters/section/03-SizeInference.tex
change function to program in size inference. conclusion changes.

@@
 The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}.
-The top level judgment ~~$\SizeF{\program}{\sigma}$~~ assigns a size scheme to a function.
+The top level judgment ~~$\SizeF{\program}{\sigma}$~~ assigns a size scheme to a program.
 It does this by extracting size constraints and then solving them.
@@
 size types when solving constraints, whereas $\exists k$ indicates a  \emph{rigid} size variable that cannot be unified with other sizes.
-We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.
+We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall program.
 
@@
 $$
-Rule~(SFun) also characterises the functions we accept: A function is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.
+Rule~(SProgram) also characterises the programs we accept: a program is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.
 
@@
 \subsection{Constraint Solving and Generalization}
-The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a function by first extracting size constraints, before solving them and generalizing the result.
+The top-level rule in Figure~\ref{f:ConstraintGeneration} assigns a size scheme to a program by first extracting size constraints, before solving them and generalizing the result.
 In the rule, the solving process is indicated by $\textrm{SOLVE}$, and takes an environment and a constraint set, and produces a solved environment and constraint set.
@@
 
-To produce the final size scheme, we look up the sizes of the input and output variables of the original function from the solved constraints and generalize appropriately.
+To produce the final size scheme, we look up the sizes of the input and output variables of the original program from the solved constraints and generalize appropriately.
 This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}.
@@
 When the environment of our size constraints contains rigid variables (indicated by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme.
-Consider the @filterLeft@ function from Section~\ref{s:SizeTypes}
+Consider the @filterLeft@ program from Section~\ref{s:SizeTypes}
 \begin{code}
@@
 \end{code}
-The size constraints for this function, already in solved form, are as follows.
+The size constraints for this program, already in solved form, are as follows.
 $$
@@
 
-Note that, although Rule~(SFun) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule.
+Note that, although Rule~(SProgram) from Figure~\ref{f:ConstraintGeneration} performs a \emph{generalisation} process, there is no corresponding instantiation rule.
 The size inference process works on the entire graph at a time, and there is no mechanism for one operator to invoke another.
@@
 \end{code}
-The above function filters its input array, and then applies @map2@ to the filtered version as well as the original array.
+The above program filters its input array, and then applies @map2@ to the filtered version as well as the original array.
 As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true.
@@
 
-In this case, Rule~(SFun) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).
+In this case, Rule~(SProgram) does not apply, because the parameter variable $xs$ has size $k_1$, but $k_1$ is marked as rigid in the environment (with $\exists k_1$).
 
-As a final example, the following function is ill-sized, because the two filter operators are not guaranteed to produce the same number of elements.
+As a final example, the following program is ill-sized, because the two filter operators are not guaranteed to produce the same number of elements.
 \begin{code}
@@
 
-The initial size constraints for this function are:
+The initial size constraints for this program are:
 $$
@@
 \textbf{Lemma: sole transducers}.
-If a function $f$ is \emph{valid}, then its bindings will have at most one transducer:
+If a program $p$ is \emph{valid}, then its bindings will have at most one transducer:
 \[
-\forall f, \sigma, n.\ f :_s \sigma \implies |trans(binds(f), n)| \le 1
+\forall p, \sigma, n.\ p :_s \sigma \implies |trans(binds(p), n)| \le 1
 \]
@@
 % I replaced 'one set of parents' with 'one pair of parents' here, because that's what the lemma says
-For some bindings $f$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
+For some program $p$ with valid constraints, each pair of names $a$ and $b$ will have at most one pair of parents $parents(a,b)$.
 \[
-\forall f, \sigma, a, b.\ f :_s \sigma \implies |parents(binds(f), a, b)| \le 1
+\forall p, \sigma, a, b.\ p :_s \sigma \implies |parents(binds(p), a, b)| \le 1
 \]
@@
 
-These two lemmas are used in the integer linear programming formulation, when constructing the program.
+These two lemmas are used in the integer linear programming formulation, when generating the constraints.
 When fusing two nodes of different iteration size, at most one pair of parents will need to be checked.

### 2014-06-20 2014betterfusionforfilters/section/07-Conclusion.tex
change function to program in size inference. conclusion changes.

@@
 
-For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold, segmented map, which operate on segmented representations of nested arrays.
+For this work to be useful for Data Parallel Haskell, more combinators are required such as segmented fold and segmented map, which operate on segmented representations of nested arrays.
 Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
 For example, segmented fold takes a segmented array, and returns a single array of the outer rate.
+
+Data flow fusion, which generates the sequential loops, must also be extended with these extra combinators.
+Data flow fusion can also be extended to generate parallel loops for most combinators by simply splitting the workload among processors.
+Merging the output of each loop can be performed by concatenating the results of filters, and merging the folds, assuming the fold operation is associative.
+
+Another interesting possibility is combinators where the output order is not important.
+There are two orderings for cross product: one that requires the second array to be manifest, the other requires the first array to be manifest.
+It may be worthwhile to add a separate cross combinator that does not ensure which ordering is used, but instead uses the ordering that produces the best clustering.
+This could be achieved in the integer linear program by requiring that at least one of the cross combinator's inputs are not fused together.

### 2014-06-20 2014betterfusionforfilters/section/05-Benchmarks.tex
mention GLPK solver etc

@@
 Hand fusing all code and compiling it with the same compiler (GCC) isolates the true cost of the various clusterings from low level differences in code generation.
+
+We used GLPK as an external solver, which is not particularly fast, but has Haskell bindings.
+For these small programs with few combinators, it produces results in under one hundred milliseconds.
+For larger programs, with twenty or more combinators to be clustered, GLPK is inadequate.
+One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in six seconds, and another open source solver CBC took ninety seconds.
 

### 2014-06-20 2014betterfusionforfilters/section/05-Benchmarks.tex
cplex timings

@@
 For larger programs, with twenty or more combinators to be clustered, GLPK is inadequate.
-One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in six seconds, and another open source solver CBC took ninety seconds.
+One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
+For the same example, the unoptimised program with all constraints takes over twenty minutes in CBC, and CPLEX takes one and a half seconds.
 

### 2014-06-22 2014betterfusionforfilters/section/03-SizeInference.tex
typesetting/indentation fixes, etc

@@
 \end{alltt}
+
 However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process.
@@
 $$
+
 Rule~(SProgram) also characterises the programs we accept: a program is \emph{valid} if and only if $\exists \sigma.\ \SizeF{\program}{\sigma}$.
@@
 \end{code}
+
 The size constraints for this program, already in solved form, are as follows.
@@
 \end{code}
+
 The above program filters its input array, and then applies @map2@ to the filtered version as well as the original array.

### 2014-06-22 2014betterfusionforfilters/section/04-ILP.tex
rename function in CNF definition to program: should reduce confusion about function vs worker function
typesetting/indentation fixes, etc

@@
 MMMMM       \= M  \= \kill
-$nodes$     \> @:@ \> $\function \to V$              \\
-$edges$     \> @:@ \> $\function \to E$              \\
+$nodes$     \> @:@ \> $\program \to V$              \\
+$edges$     \> @:@ \> $\program \to E$              \\
 $edge$      \> @:@ \> $\{bind\} \times bind \to E$   \\
@@
 \end{tabbing}
+
 For each pair of nodes with indices $i$ and $j$ we use a boolean variable $x_{i,j}$ which indicates whether those two nodes are fused.
@@
 \end{tabbing}
+
 The second group of variables is used to ensure that the clustering is acyclic.
@@
 \end{code}
+
 There is no fusion-preventing edge directly between the @xs@ and @zs@ bindings, but there is a fusion-preventing edge between @sum@ and @zs@.
@@
 \end{tabbing}
+
 The final group of variables is used to help define the cost model encoded by the objective function.
@@
 \end{tabbing}
+
 As per Megiddo~\cite{megiddo1998optimal} the form of these constraints is determined by whether there is an dependency between nodes $i$ and $j$.
@@
 \end{tabbing}
+
 When combined with the precedence-preserving constraints earlier, setting $x_{i,j} = 1$ also forces $\pi_i < \pi_j$.
@@
 \end{tabbing}
+
 Secondly, if the two nodes have different iteration sizes and no common parent then they also cannot be fused and we set $x_{i,j} = 1$:
@@
 \end{tabbing}
+
 Finally, if the two nodes had different iteration sizes but \emph{do} have parent transducers of the same size, then the two nodes can be fused if they are fused with their respective parents, and the parents themselves are fused:
@@
 \end{tabbing}
+
 This last part is the main difference to existing ILP solutions: we allow nodes with different iteration sizes to be fused when their parent transducers are fused.
@@
 \end{code}
+
 Here $sum1$ and $sum2$ have different iteration sizes and we have that $parents(sum1, sum2) = \{(sum1, gts)\}$.
@@
 \end{tabbing}
+
 Recall that an array is fully contracted when all of the consumers are fused with the nodes that produces it, which means that the array does not need to be fully materialized in memory.
@@
 \end{itemize}
+
 The three components of the cost model are a heuristic abstraction of the true cost of executing the program on current hardware.
@@
 \end{tabbing}
+
 With $possible$ and $possible'$ defined, we refine our formulation to only generate constraints between two nodes if there is a chance they may be fused together.

### 2014-06-22 2014betterfusionforfilters/section/05-Benchmarks.tex
typesetting/indentation fixes, etc

@@
 
-We used GLPK as an external solver, which is not particularly fast, but has Haskell bindings.
+We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings.
 For these small programs with few combinators, it produces results in under one hundred milliseconds.
-For larger programs, with twenty or more combinators to be clustered, GLPK is inadequate.
+For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
 One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
-For the same example, the unoptimised program with all constraints takes over twenty minutes in CBC, and CPLEX takes one and a half seconds.
+For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.
 
@@
 \end{code}
+
 We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
@@
 \noindent
+
 One minimal solution to this is:
@@
 \end{tabbing}
+
 This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.
@@
 \end{tabbing}
+
 We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
@@
 \end{tabbing}
+
 Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.

### 2014-06-22 2014betterfusionforfilters/section/07-Conclusion.tex
typesetting/indentation fixes, etc

@@
 Rate inference will have to be adapted to handle segmented arrays, possibly using an inner and outer rates.
-For example, segmented fold takes a segmented array, and returns a single array of the outer rate.
+For example, segmented fold takes a segmented array of inner and outer rate, and returns a single array of the outer rate.
+Similarly, segmented append takes two segmented arrays with the same outer rate, but different inner rates, and returns a segmented array with a new inner rate and the same outer rate.
 

### 2014-06-22 2014betterfusionforfilters/section/01-Introduction.tex
change function to program in size inference. conclusion changes.
minor contraction of introduction to fit into one page

@@
 \label{s:Introduction}
-A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.
-It is essential, however, that the compiler combines these operations and compiles them into a small number of efficient loops, as a naive translation leads to high memory traffic and poor performance.
+A collection-oriented programming model is ideal for expressing data-parallel computations, as it exposes the communication patterns to the compiler without requiring complicated loop analysis.
+It is essential, however, that the compiler combines these operations into a small number of efficient loops, as a naive translation leads to high memory traffic and poor performance.
 This is a well known problem, and techniques to fuse subsequent array operations together have been presented in~\cite{gill1993short, coutts2007streamfusion, keller2010repa}, to name just a few.
@@
 % : first, we often need to re-arrange the array operations and cluster those that iterate over the same structures.
-As shown in (Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find good clusterings.
+As shown in Megiddo~\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}, Integer Linear Programming can be used to find good clusterings.
 Unfortunately, they cannot handle operations like filter, where the output size differs from the input size.
-The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.
+We present a technique that can handle both multi-loop fragments as well as size-altering operations.
 
@@
 It improved on existing array fusion approaches~\cite{coutts2007streamfusion, keller2010repa} as it guarantees fusion into a single loop for programs that operate on the same size input data and contain no fusion-preventing dependencies between operators.
-Our clustering technique is also applicable to parallel loops.
+% Our clustering technique is also applicable to parallel loops.
 

### 2014-06-24 2014betterfusionforfilters/section/04-ILP.tex
Hide conclusion, edits, pack back to 10 pages

@@
 $edge$      \> @:@ \> $\{bind\} \times bind \to E$              \\
-$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$   \\
-\\
-$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$       \\
-\\
-$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$    \\
-\\
+$inedge$    \> @:@ \> $\{bind\} \times name \times name \to E$
+\\[1ex]
+$nodes(bs)$ \> $= \{(name(b), \iiter_{\Gamma,C}(b)) | b \in bs\}$
+\\[1ex]
+$edges(bs)$ \> $= \bigcup_{b \in bs}edge(bs, b)$
+\\[1ex]
 MM             \= M \= \kill
 $edge(bs, out = @fold@~f~in)$ \\
-    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$
+\\
 $edge(bs, out = @map@~f~in)$  \\
-    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$
+\\
 $edge(bs, out = @filter@~f~in)$ \\
-    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$       \\
+    \> $=$    \> $\{inedge(bs,out,s) | s \in fv(f)\} \cup \{inedge(bs, out, in) \}$
+\\
 $edge(bs, out = @gather@~data~indices)$ \\
-    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$       \\
+    \> $=$    \> $\{(out,data, \fusionpreventing) \} \cup \{inedge(bs, out, indices) \}$      
+\\
 $edge(bs, out = @cross@~a~b)$ \\
-    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$ \\
+    \> $=$    \> $\{inedge(bs, out, a) \}           \cup      \{(out, b, \fusionpreventing) \}$
+\\
 $edge(bs, outs = @external@~ins)$  \\
-    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$ \\
-\\
+    \> $=$    \> $\{(outs,i, \fusionpreventing) | i \in ins \}$
+\\[1ex]
 $inedge(bs,to,\from)$ \\
@@
 % -------------------------------------
-\paragraph{Array contraction} The final group of constraints gives meaning to the $c$ variables that we use to represent whether an array is fully contracted:
+\paragraph{Array contraction} The final group gives meaning to the $c$ variables, which represent whether an array is fully contracted:
 \begin{tabbing}
@@
 MMMMM   \= MMMM \= M \= \kill
-Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   
-\\         \> \> \> (memory traffic and loop overhead)              
-\\         \> $+$ \> $\Sigma_{i \in V} N \cdot c_i$                  
-\\         \> \> \> (removing intermediate arrays)         
-\\ Subject to \> \ldots ~ constraints from \S\ref{s:LinearConstraints} ~ \ldots 
-\\ Where      \>                                       
-\\         \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         
+Minimise   \>     $\Sigma_{(i,j) \in E} W_{i,j} \cdot x_{i,j}$   
+                        ~~ (memory traffic and loop overhead)
+\\ ~~~~~~~~~~~~~ $+$ \> $\Sigma_{i \in V} N \cdot c_i$
+                        ~~~~~~~~~~~~ (removing intermediate arrays)
+\\[1ex]
+   Subject to  \> \ldots ~ constraints from \S\ref{s:LinearConstraints} ~ \ldots 
+\\ Where   \> $W_{i,j} = N^2$ \> $~|$ \> $(i,j) \in E $         
 \\         \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         
@@
 Similarly the function $possible'(a, b)$ checks whether there is any possibility that the parents of $a$ and $b$ may be fused.
-
-\newpage
-
+%
 \begin{tabbing}
@@
 $possible$ \> $:$     \> $name \times name \to \mathbb{B}$      \\
-MM        \= M    \= \kill
-$possible(a,b)$ \\
-    \>$=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
+MMMMMMx        \= M    \= \kill
+$possible(a,b)$   
+        \> $=$  \>$\forall p \in path(a,b) \cup path(b,a).\ \fusionpreventing \not\in p$
 \\[1ex]
@@
 $possible'$     \> $:$ \> $name \times name \to \mathbb{B}$      \\
-MM        \= M    \= \kill
-$possible'(a,b)$ \\
-    \>$=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
-    \>$\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
+MMMMMMx        \= M    \= \kill
+$possible'(a,b)$ 
+        \> $=$   \>$\exists A, B.\  parents(a,b) = \{A,B\} \wedge possible(a,b)$ \\
+        \> $\wedge$ \> $possible(A,a) \wedge possible(B,b) \wedge possible(A,B)$
 \end{tabbing}
@@
 The final formulation of the integer linear program follows.
-
+%
 \begin{tabbing}
@@
            \> (if $possible(i,j)$)         
-\\[1ex]
+\\[0.5ex]
 Subject to \> $-N \cdot x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
            \> (if $possible(i,j) \wedge (i,j) \not\in E \wedge (j,i) \not\in E$)            
-\\[1ex]
+\\[0.5ex]
            \>    $x_{i,j}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N \cdot x_{i,j}$ \\
            \> (if $possible(i,j) \wedge (i,j,\fusible) \in E$)     
-\\[1ex]
+\\[0.5ex]
            \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
            \> (if $(i,j,\fusionpreventing) \in E$)    
-\\[1ex]
+\\[0.5ex]
            \> $x_{i,j}$    \> $\le$ \> $c_i$           \>       \>            \\
@@
            \> (if $(i,j, \fusionpreventing) \in E$)
-\\[1ex]
+\\[0.5ex]
            \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
            \> (if $\bot \in \{\iiter_{\Gamma,C}(i), \iiter_{\Gamma,C}(j)\}$)  
-\\[1ex]
+\\[0.5ex]
            \> $x_{i',i}$   \> $\le$ \> $x_{i,j}$        \>       \>            \\
@@
            \> \> $\wedge~parents(i,j) = \{(i',j')\}$) 
-\\[1ex]
+\\[0.5ex]
            \> $x_{i,j}$    \> $=$   \> $1$             \>       \>            \\
            \> (if $\iiter_{\Gamma,C}(i) \not= \iiter_{\Gamma,C}(j) \wedge \neg possible'(i,j)$) 
-\\[1ex]
+\\[0.5ex]
 MMMMM   \= MMMM \= M \= \kill

### 2014-06-24 2014betterfusionforfilters/section/05-Benchmarks.tex
Hide conclusion, edits, pack back to 10 pages

@@
 
-We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings.
+We have used both GLPK and CPLEX as external ILP solvers.
+For small programs such as @normalizeInc@, both solvers produce solutions in under 100ms.
+For a larger randomly generated example with twenty-five combinators, GLPK took over twenty minutes to produce a solution while the commercial CPLEX solver was able to produce a solution in under one second --- which is still quite usable.
+We will investigate the reason for this wide range in performance in future work.
+
+% We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings.
 For these small programs with few combinators, it produces results in under one hundred milliseconds.
-For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
-One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
-For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.
+% For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
+% One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
+% For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.
 
@@
 % -----------------------------------------------------------------------------
+\pagebreak
 \subsection{Normalize2}
@@
            \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
-\\[1ex]
+\\[0.5ex]
 Subject to 
@@
     \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
-\\[1ex]
+\\[0.5ex]
     \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
-\\[1ex]
+\\[0.5ex]
     \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
     \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
-\\[1ex]
+\\[0.5ex]
     \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
-\\[1ex]
+\\[0.5ex]
     \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\

### 2014-12-19 2015mergingmerges/Main.tex
just some notes

--- 2014-05-15 2014betterfusionforfilters/Main.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{sigplanconf}
 \usepackage{amssymb}
@@
 
-% TODO: Fix up writing to use the words combinator and operator consistently
-% We seem to be using "combinator" for the functions in the original program, 
-% but "operator" for to name the abstract operation that combinator performs.
-
-% \TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
-
-% \TODO{Put this somewhere: Fusion-preventing dependencies are akin to barrier synchronisations in the Bulk Synchronous model of parallel computation. }
-
 % -----------------------------------------------------------------------------
@@
 
-\title{Fusing Filters with Integer Linear Programming}
-% Better title?
-% \title{Better fusion for filters}
+% \exclusivelicense
+% \conferenceinfo{}{}
+% \copyrightyear{2015}
+% \copyrightdata{}
+\doi{}
+% \pagenumbering{gobble}
+
+\title{Merging merges, more or less}
 
@@
   \and Gabriele Keller$^\dagger$ 
+  \and Manuel Chakravarty$^\dagger$ 
 }{
@@
     University of New South Wales, Australia \\[2pt]
-    \textsf{\{amosr,benl,keller\}@cse.unsw.edu.au}
+    \textsf{\{amosr,benl,keller,chak\}@cse.unsw.edu.au}
   }
@@
 \begin{abstract}
-Minimising memory traffic is key to compiling functional, collection-oriented array programs into efficient code.
-It is not sufficient to simply fuse subsequent array operations into a single computation; we also have to cluster \emph{separate} traversals of the same array into a single loop.
-Previous work demonstrated how Integer Linear Programming (ILP) can be used to cluster the operators in a general data-flow graph into subgraphs, which can be individually fused.
-However, these approaches can only handle operations which preserve the size of the array, thereby missing out on some optimisation opportunities.
-This paper addresses this shortcoming by extending the ILP approach with support for size-changing operations, using an external ILP solver to find good clusterings.
 \end{abstract}
@@
 \keywords
-	Arrays, Fusion, Haskell
+	Arrays; Fusion; Haskell
 
 \input{section/01-Introduction.tex}
-\input{section/02-CNF.tex}
-\input{section/03-SizeInference.tex}
-\input{section/04-ILP.tex}
-\input{section/05-Benchmarks.tex}
-\input{section/06-Related.tex}
 
 \section*{Acknowledgements}
-Many thanks are due to
-Manuel Chakravarty,
-Robert Clifton-Everest,
-Kai Engelhardt,
-Bill Kroon,
-Frederik M.\ Madsen,
-Abdallah Saffidine,
-Carter Schonwald,
-and Jingling Xue
-for enlightening discussions relating to this work.
+The audience of the inaugural SAWDAP workshop, especially Thomas Sutton and Patryk Zadarnowski, were very helpful in an early version of this work.
 

### 2014-12-19 2015mergingmerges/section/01-Introduction.tex
just some notes

%!TEX root = ../Main.tex
\section{Related work}
\label{s:Related}

\subsection{StreamIt}
StreamIt\cite{thies2002streamit} uses synchronous dataflow (SDF), which means completely static rates, so the exact number of input elements and output elements for each combinator is always known.
For example, an audio lowpass filter may have an input of one, an output of one, and lookbehind, or peek, of ten.
Static rates make scheduling with bounded buffers easier, but disallow rate-changing operations like predicate-matching filters, or merges.
Translating from StreamIt parlance to functional programming combinators, a ``stateless filter'' corresponds to a map, and a ``stateful filter'' corresponds to a scan.

I think StreamIt extends SFD with asynchronous peeks?

StreamIt's peeking lets you write some stateful computations without state, which can then be parallelised.
It feels a little bit like how requiring an associative operator to fold allows parallelisation that still looks like state.
I don't think the details of these optimisations are particularly important right now.
\cite{gordon2010compiler}

``We plan to support dynamically changing rates in the next version of StreamIt''\cite{thies2002streamit} (I have found no recent evidence of this)

\subsubsection{Dynamic Expressivity with Static Optimization for Streaming Languages (streamit slides)}
Talks about dynamic scheduling: the graph is partitioned into static subgraphs, so that all dynamic edges cross partition boundaries. These static subgraphs are then compiled into kernels as normal, and an overall dynamic scheduler is used.
Nothing about detecting deadlocks or buffer overruns.

\subsubsection{W. Thies, Language and compiler support for stream programs, 2009}
Kahn Process Networks...
``It is undecidable to statically determine the amount of buffering ... or to check whether the computation might deadlock.''\cite{thies2009language}

Synchronous dataflow has static rates, and can therefore find a valid scheduling at compiletime.

CSP has synchronous messages, blocking until a reply message is sent.

StreamIt has a structured way of creating graphs: only splitjoins, vertical (sequential) pipelines, and feedback loops. A bit like a series-parallel graph, with feedbacks added.

\begin{quote}
Boolean dataflow [HL97] is a compromise between these two extremes; it computes a parameterized schedule
of the graph at compile time, and substitutes runtime conditions to decide which paths are taken.
The performance is nearly that of synchronous dataflow while keeping some flexibility of dynamic
dataflow.
\end{quote}

\subsection{Lava}
I don't think Lava is relevant, but Obsidian might be.

\subsection{Advances in Dataflow Programming Languages, W Johnston, Hanna}
Starts mainly about dataflow hardware\cite{johnston2004advances}.
Most disadvantages and criticisms of early dataflow, and hence its decline, can be attributed to the hardware models themselves, rather than the languages.

Static dataflow architecture:
seems like it would be sufficient for us. Check references on what's possible:

Synchronous dataflow (SDF) requires statically known input/output rates, which is not sufficient for us, although I think if you could annotate rates with ``Maybes'' it would be (but suspect that would lose the nice properties of SDF).

I had these citations highlighted but I don't remember why.
\begin{enumerate}
\item
DAVIS, A. L. 1978. The architecture and system method of DDM1: A recursively structured data driven machine. In Proceedings of the 5th Annual Symposion on Computer Architecture (New York). 210–215.

\item
DENNIS, J. B. AND MISUNAS, D. P. 1975. A prelimi- nary architecture for a basic data-flow processor. In Proceedings of the Second Annual Symposium on Computer Architecture. 126–132.

\item
DENNIS, J. B. 1980. Data flow supercomputers. IEEE Comput. 13, 11 (Nov.), 48–56.

\item
DENNIS, J. B. 1974. First version of a data flow pro- cedure language. In Proceedings of the Sympo- sium on Programming (Institut de Programma- tion, University of Paris, Paris, France). 241– 271.

\item
SILC, J., ROBIC, B., AND UNGERER, T. 1998. Asyn- chrony in parallel computing: from dataflow to multithreading. Parallel Distrib. Comput. Pract. 1, 1, 3–30.
\end{enumerate}

\subsection{A survey of stream processing, Robert Stephens, 1997}
This is looking for a general theory of stream processing - sounds useful\cite{stephens1997survey}.
Highlights that dataflow SPSs can be seen as an implementation of abstract STs.
One of dataflow's main aims has always been to avoid the ``von Neumann bottleneck'' by allowing parallelism.

Classifies SPSs along three axes: asynchronous or synchronous; deterministic or non; and uni-directional or bidirectional channels. StreamIt, for example, fits into SDU.
(Although another paper said that one way to implement asynchronous was to use bidirectional channels?)

Functional stream languages:
``several specialized stream orientated functional languages have been developed including ARTIC (see [60]), HOPE (see [52]) and RUTH (see [98])''

FOCUS: ``Within such networks data is exchanged via unbounded FIFO channels that are modelled as streams.''
Unbounded channels, of course, is exactly what we don't want.

Reactive systems: it talks about real time systems such as operating systems, which seem like they would actually be closest to what we need.

\subsubsection{Section 8}
Section 8.2 is a particularly confusing definition of standard list combinators.

Section 8.3 has some prolog definitions of more list combinators, and then some higher-order versions of the combinators.
There is a combinator called ``Merge'', which is actually an interleaving of every second element:

\begin{code}
merge :: [a] -> [a] -> [a]
merge (a:a':as) (b:b':bs)
 = a : b' : merge as bs
\end{code}

However, this and append are the only combinators with multiple stream inputs.

\subsubsection{Lucid}
Section 9.4 is about Lucid. Lucid has some more interesting combinators which take multiple streams.

\begin{code}
attime :: (Time -> a) -> (Time -> Time) -> (Time -> a)
attime as ts t
 = as (ts t)

upon :: [a] -> [Bool] -> [a]
upon (a:as) bs
 = a : rest (a:as) bs

rest (a:as) (b:bs)
 | b
 = head as : rest as bs
 | otherwise
 = a : rest (a:as) bs
\end{code}

So I imagine that these combinators could actually be used for interesting merges.
However, there is no mention of whether compilation without unbounded buffers can be achieved.
It looks like primitive operators such as @nor@ are essentially @zipWith (nor)@.

\subsubsection{LUSTRE}
LUSTRE is related to Lucid, but requires causality and synchronicity.
Their deadlock checking does reject some valid programs.
The primitive combinators are: previous (called Z in signal processing?), followed by (something like @head a ++ tail b@), when (find the next\emph{[sic?]} true value in bool stream and use value then) and current (like a past-looking when, except for initial value find future).
These seem to destroy causality.

This looks promising, but I don't know whether our merges will be allowed by the deadlock checker.

\subsubsection{SIGNAL}
Could be worth looking into. Synchronous. Not sure about deadlock and buffer detection etc.

\subsection{A Co-iterative Characterization of Synchronous Stream Functions}
This is \cite{caspi1998co}, Paul Caspi and Marc Pouzet.

Coiteration for streams: giving a stream an initial state, and a transition function from state to pair of value and new state.

First, looking at length-preserving functions, then non-length preserving such as filter.
Starts off very similar to stream fusion, but the catch is that most streams based on other streams don't actually care about the input stream's guts (transition function), but actually just care about the values!

So instead of

\begin{code}
data Stream a s = (s, s -> (s,a))
map :: Stream a s -> Stream a s
map = ...
\end{code}
we have
\begin{code}
data Stream i o s = (s, i -> s -> (s,o))
map :: (a -> b) -> Stream a b s
map = ...
\end{code}
but these can only handle length-preserving functions.

Definition of a synchronous stream function:
\begin{code}
co_apply :: CStream (a -> Maybe s -> (b, Maybe s))
         -> CStream a
         -> CStream b

 f : Stream a -> Stream b
is synchronous iff there exists
 f' : Stream (a -> s -> (b, s))
such that
 f == co_apply f'
\end{code}

Lambda and recursion are a bit confusing. Read again.

Now, the result type of each CStream becomes @F a s@ instead of just @(a, s)@ where @F a s@ is defined:
\begin{code}
data F a s
 = P a s
 | S   s
\end{code}
which is equivalent to @(Maybe a, s)@. They then write an @extend@ (@map apply@) combinator which consumes two inputs at once, but throws an error if one argument yields when the other doesn't. Put another way, ``if the clocks of the two arguments are the same'', where the clock is @map isJust@.

Their definition of merge is interesting. The list version they give is
\begin{code}
merge (False:cs) xs (y:ys) = y : merge cs xs ys
merge (True :cs) (x:xs) ys = x : merge cs xs ys
\end{code}
which only pulls from one of the true or false streams at a time.
However, the co-iterative version pulls from all streams at each iteration, but requires that only one of the true or false streams actually produce a value: if both produce a value, it is a runtime error.
It feels like there is a step missing between the two.

My initial thought is that this clock test is a bit too restrictive for us, because we could store \emph{one} value in a buffer if the two clocks don't exactly line up.

The actual clock calculus looks very similar to a regular type system, with fairly minor and understandable differences.

\subsection{Lucy-n: a n-Synchronous Extension of Lustre}
This is Pouzet et al again\cite{mandel2010lucy}.
While the synchronous model requires no buffering, n-synchrony relaxes this by allowing communication through buffers whose size is known at compile-time.
They use a similar clock calculus to before, but extend it with subtyping. Buffering is introduced through an explicit @buffer@ primitive, but the required buffer size is inferred.

What is ``Cyclo Static Data Flow''?
Apparently ``synchronous languages'' like Lustre and Lucid Synchrone are not quite as flexible as SDF.
In synchronous languages, two streams can be composed (zipped?) only if their clocks are \emph{exactly} the same. 
N-synchronous relaxes this, and allows composition of streams with unequal clocks, but that can be synchronised with a finite, statically known sized, buffer.

``In this paper, we restrict the clock language ce to define ultimately periodic boolean sequences only''
ie infinite repetitions with some finite prefix.

It looks like the kernel is missing the @on@ operator: @s on e@ which is like a filter. It does have @e when ce@ but here @ce@ is a clock, so can only be a repeating boolean sequence. No - it does have @on@, but it takes a clock instead of an expression.
For some clocks @a@ and @b@, @a on b@ is a sub-clock of @a@.

Section 5 is where it gets interesting: ``In order to overcome this complexity, we propose in this paper not to consider exact periodic clocks but their abstraction''.
They abstract over the clocks with $\langle b^0, b^1 \rangle (r)$, where $b$s are the minimum and maximum shift of @1@s, and $r$ is the proportion of @1@s to @0@s.
So, basically, $r$ is the slope of the line of @1@s. The resulting abstract slope of @a on b@ can be found quite easily.
Two clocks can only be scheduled together if their $r$s are the same, that is they have the same slope.
The size of the buffer needed is the difference between two of the $b$s.

\subsection{Compile-Time Scheduling of Dynamic Constructs in Dataflow Program Graphs}
This is Ha and Lee\cite{ha1997compile}.

### 2014-12-19 2015mergingmerges/section/01-Introduction.tex
ok

@@
 This is Ha and Lee\cite{ha1997compile}.
+``The main purpose of this paper is to show how we can define the profiles of dynamic constructs at compile-time''.
+
+They use simulation or programmer pragmas to approximate the runtime statistics.
+It looks like this is actually just about dynamic length/runtimes of certain nodes, rather than dynamic dependencies.
+I don't think this is relevant.

### 2015-02-01 2015mergingmerges/section/XX-Notes.tex
words

%!TEX root = ../Main.tex
\section{Notes}
\label{s:Notes}

Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.

Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
However, for querying large data sets, these kind of combinators are essential.
The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.

Our goal, then, appears relatively simple: to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

\subsection{Streams and combinators}
First, we define an abstraction of two-input combinators, very similar to stream fusion\CITE or co-iterative stream functions of Caspi and Pouzet\cite{caspi1998co}.
We present a simplified stateless version here, which would not be able to express all useful combinators (such as segmented scan) but suffices for merge, append and zip.
We also only treat combinators that consume both input streams linearly, which excludes combinators like cross product or gather.
In Coq\footnote{The definitions have been reordered for clarity}:

\begin{code}
Definition Stream2 (a b c : Set)
 := option a -> option b -> Step2 c

Inductive Step2 c :=
 | MkStep2 : Move -> Move -> option c -> Step2 c
 | Done    : Step2 c.

Inductive Move :=
 | Forward
 | Stay.
\end{code}

At each evaluation step, a @Stream2@ is called with the value at each input stream, or @None@ if the input stream is finished.
The return of each @Stream2@ evaluation step can be either a @MkStep2@ indicating that further processing is possible, or @Done@, in which case the combinator has finished.
A @MkStep2@ contains two @Move@s, which designate which input streams should be consumed and the next value pulled, and which should stay where they are.
@MkStep2@ also contains an optional output value.

\subsubsection{Examples}

Below, we show example combinators.
The first combinator pairwise zips two streams together, consuming both streams in lock-step.

\begin{code}
Definition Zip (A B : Set)
 :  Stream2 A B (A*B)
 := fun a b =>
    match (a,b) with
    | (Some a, Some b)
    => MkStep2 Forward Forward (Some (a,b))
    | _               
    => Done
    end.
\end{code}

Next, append consumes from the left stream until it is empty, then consumes the right stream.
\begin{code}
Definition Append (A : Set)
 :  Stream2 A A A
 := fun a b =>
    match (a,b) with
    | (Some a, _)
    => MkStep2 Forward Stay    (Some a)
    | (None,   Some b)
    => MkStep2 Stay    Forward (Some b)
    | (None,   None)               
    => Done
    end.
\end{code}

As a final example, the merge join inspects both its input elements, and consumes the one with the smallest value.
There are many variations on this theme, but this particular version that outputs all values in order is actually equivalent to a segmented append.

\begin{code}
Definition MergeJoin (A : Set)
 :  Stream2 (nat*A) (nat*A) (nat*A)
 := fun a b =>
    match (a,b) with
    | (Some (ia,a), Some (ib, b))
    => if   le_dec ia ib
       then MkStep2 Forward Stay (Some (ia, a))
       else MkStep2 Stay Forward (Some (ib, b))
    | (Some (ia,a), None)
    =>      MkStep2 Forward Stay (Some (ia,a))
    | (None, Some (ib,b))
    =>      MkStep2 Stay Forward (Some (ib,b))
    | (None, None)
    =>      Done
    end.
\end{code}

### 2015-02-01 2015mergingmerges/section/XX-RelatedWork.tex
words

--- 2014-12-19 2015mergingmerges/section/01-Introduction.tex
@@
 I don't think this is relevant.
+
+\subsection{Scheduling dynamic dataflow graphs with bounded memory using the token flow model}
+This is Buck\cite{buck1993scheduling}.
+
+Petri nets: Petri nets are directed graphs. Vertices are split into \emph{places} and \emph{transitions}.
+
+Safeness: a Petri net with an initial marking $\mu$ is safe if it is not possible, by any sequence of transition firings, to reach a new marking $\mu'$ where any place has more than one token.
+Adding backwards acknowledgement arcs can force safeness (but I suspect can ruin liveness).
+
+Boundedness: a generalisation of safety. A place is \emph{k-bounded} if the number of tokens never exceeds $k$.
+
+Liveness: the avoidance of deadlock (where nothing can fire).
+
+Conservativeness: strict conservatism is when the number of tokens is never changed by firing.
+Conservative with respect to a weight vector $w$ where the weighted sum of number of tokens never changes.
+
+Karp and Miller computation graphs: nodes are operations and arcs are queues of data.
+Each arc $d_p$ has four natural numbers: $A_p$ size of initial queue, $U_p$ output size of arc's in, $W_p$ input size of arc's out, $T_p$ minimum queue length necessary for out to execute (peek?).
+These computation graphs are determinate (if their functions are). They specify the conditions for termination (rather than deadlock-free!).
+
+Marked graphs: a subset of Petri nets. Every place has exactly one input transition and one output transition. No parallel arcs. I guess a transition can have multiple output places though.
+For a given place, only one output transition exists, so if the place is full there is only one choice of what to execute.
+Marked graphs are much easier to analyse than general Petri nets: deadlocks cannot occur on cycles with at least one token going through them.
+
+Homogeneous dataflow graphs: where all nodes, in each firing, consume exactly one token from each input arc and produce one token on each output arc. This is corresponds to a marked graph.
+
+Synchronous, or regular, dataflow: generalisation of homogeneous where consumption or production can be different to one, but still constant and known.
+
+Dynamic dataflow: where the number of tokens consumed or produced depends on values of certain input tokens (eg SWITCH/SELECT). This is Turing complete, so more powerful than Petri nets.
+Nondeterministic merge is another extension.
+
+Kahn: blocking reads etc, so no nondeterministic merge.
+
+Regular dataflow techniques cannot handle dynamic dependencies, but sometimes conditional branching can be emulated with conditional assignment (executing both sides) but this is not always appropriate.
+There are some extensions to regular dataflow to allow certain dynamic dependencies.
+
+Control flow / dataflow hybrid: Turing complete, so hard to analyse.
+Standard compiler program block DAGs.
+
+Controlled use of dynamic actors: dynamic actors in general are Turing complete, but restricting dynamic actors can make analysis feasible.
+For example, where all conditionals are later merged with the same flag.
+
+Clock calculus: as in LUSTRE etc.
+
+Token flow model: an extension of regular (synchronous) dataflow allowing dynamic actors such as SWITCH and SELECT.
+Boolean-controlled dataflow (BDF): each port is annotated with a symbolic expression (rather than just a constant) denoting how many values are consumed or produced.
+For example, the SWITCH inputs are both still annotated with $1$, but the outputs are changed to $p_i$ and $1 - p_i$, meaning only one of the outputs will produce one value at any time.
+As with regular graphs, a topology matrix is created with these symbolic expressions instead of constants, as a function of the $p$s.
+If the topology matrix function has a nontrivial solution that \emph{does not} depend on the values of $p$s, it is strongly consistent: no matter what, it will be able to be scheduled.
+If there are nontrivial solutions that exist only for particular $p$s, it is weakly consistent: schedules would work for only some sets of data.
+(However, this analysis is undecidable in general)
+The actual value of these $p_i$ values is not important, it's just some abstract symbolic probability, however they can be treated as the fraction of iterations that have values.
+Strong consistency alone does not assure bounded memory.
+
+Chapter 3 may be worth rereading, but chapter 4 is just implementation details.
+
+Chapter 5 is extending BDF: an extension to boolean-controlled dataflow that generalises the boolean controls to integers (IDF). Not too worried.
+
+\subsection{Generic Programming with Adjunctions}
+This is Hinze\cite{hinze2012generic}.
+
+``For instance, append does not have the form of a fold as it takes a second argument that is used later in the base case.''
+An adjoint fold is a generalisation, by somehow allowing the argument of a fold to be wrapped in a functor application.
+The functor must have a right adjoint, or left adjoint for unfolds.
+
+Skipping to S3.
+
+\subsection{Conjugate Hylomorphisms}
+This is Gibbons\cite{gibbons2015conjugate}.
+
+\subsection{Coroutines and networks of parallel processes}
+This is the original Kahn network paper\cite{kahn1976coroutines}.
+It's not actually particularly interesting, and just explains processes as coroutines, then goes through a few example programs.
+Interesting historically as an early (1976?) example of yearning for purity and such.
+
+\subsection{Bounded scheduling of process networks}
+This is \cite{parks1995bounded}.
+About finding a bounded execution of Kahn processes, which is undecidable, but this choice quote is hardly illuminating:
+``Fortunately, because we are interested in programs that never terminate, our scheduler has infinite time and can guarantee that programs execute forever with bounded buffering whenever
+possible.''
+
+``We can let the scheduler work as the program executes. Because the program is designed to run forever without terminating, the scheduler has an infinite amount of time to arrive at an answer''
+I don't quite understand this. Presumably you use a default execution to start with, but if the default execution is fast enough, why bother.
+It seems like it just chooses some bound and executes a slightly modified graph with that bound, and if it fails, tries with a larger bound.
+
+\subsection{Kahn process networks are a flexible alternative to MapReduce}
+MapReduce alone is a bit restrictive, but Kahn processes are more flexible and still composable and whatnot\cite{vrba2009kahn}.
+Does runtime deadlock detection when a processes blocks on a send.
+Not much really conceptual work here.

### 2015-02-02 2015mergingmerges/section/01-Introduction.tex
ok

--- 2014-12-19 2015mergingmerges/section/01-Introduction.tex
@@
 %!TEX root = ../Main.tex
-\section{Related work}
-\label{s:Related}
+\section{Introduction}
+\label{s:Introduction}
 
-\subsection{StreamIt}
-StreamIt\cite{thies2002streamit} uses synchronous dataflow (SDF), which means completely static rates, so the exact number of input elements and output elements for each combinator is always known.
-For example, an audio lowpass filter may have an input of one, an output of one, and lookbehind, or peek, of ten.
-Static rates make scheduling with bounded buffers easier, but disallow rate-changing operations like predicate-matching filters, or merges.
-Translating from StreamIt parlance to functional programming combinators, a ``stateless filter'' corresponds to a map, and a ``stateful filter'' corresponds to a scan.
+Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
+Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
+Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
+At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.
 
-I think StreamIt extends SFD with asynchronous peeks?
+Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
+However, for querying large data sets, these kind of combinators are essential.
+The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
+Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
+Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.
 
-StreamIt's peeking lets you write some stateful computations without state, which can then be parallelised.
-It feels a little bit like how requiring an associative operator to fold allows parallelisation that still looks like state.
-I don't think the details of these optimisations are particularly important right now.
-\cite{gordon2010compiler}
+Our goal, then, appears relatively simple: to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
 
-``We plan to support dynamically changing rates in the next version of StreamIt''\cite{thies2002streamit} (I have found no recent evidence of this)
+\subsection{Motivation}
 
-\subsubsection{Dynamic Expressivity with Static Optimization for Streaming Languages (streamit slides)}
-Talks about dynamic scheduling: the graph is partitioned into static subgraphs, so that all dynamic edges cross partition boundaries.
-These static subgraphs are then compiled into kernels as normal, and an overall dynamic scheduler is used.
-Nothing about detecting deadlocks or buffer overruns.
-
-\subsubsection{W. Thies, Language and compiler support for stream programs, 2009}
-Kahn Process Networks...
-``It is undecidable to statically determine the amount of buffering ... or to check whether the computation might deadlock.''\cite{thies2009language}
-
-Synchronous dataflow has static rates, and can therefore find a valid scheduling at compiletime.
-
-CSP has synchronous messages, blocking until a reply message is sent.
-
-StreamIt has a structured way of creating graphs: only splitjoins, vertical (sequential) pipelines, and feedback loops. A bit like a series-parallel graph, with feedbacks added.
-
-\begin{quote}
-Boolean dataflow [HL97] is a compromise between these two extremes; it computes a parameterized schedule
-of the graph at compile time, and substitutes runtime conditions to decide which paths are taken.
-The performance is nearly that of synchronous dataflow while keeping some flexibility of dynamic
-dataflow.
-\end{quote}
-
-\subsection{Lava}
-I don't think Lava is relevant, but Obsidian might be.
-
-\subsection{Advances in Dataflow Programming Languages, W Johnston, Hanna}
-Starts mainly about dataflow hardware\cite{johnston2004advances}.
-Most disadvantages and criticisms of early dataflow, and hence its decline, can be attributed to the hardware models themselves, rather than the languages.
-
-Static dataflow architecture:
-seems like it would be sufficient for us. Check references on what's possible:
-
-Synchronous dataflow (SDF) requires statically known input/output rates, which is not sufficient for us, although I think if you could annotate rates with ``Maybes'' it would be (but suspect that would lose the nice properties of SDF).
-
-I had these citations highlighted but I don't remember why.
-\begin{enumerate}
-\item
-DAVIS, A. L. 1978. The architecture and system method of DDM1: A recursively structured data driven machine. In Proceedings of the 5th Annual Symposion on Computer Architecture (New York). 210–215.
-
-\item
-DENNIS, J. B. AND MISUNAS, D. P. 1975. A prelimi- nary architecture for a basic data-flow processor. In Proceedings of the Second Annual Symposium on Computer Architecture. 126–132.
-
-\item
-DENNIS, J. B. 1980. Data flow supercomputers. IEEE Comput. 13, 11 (Nov.), 48–56.
-
-\item
-DENNIS, J. B. 1974. First version of a data flow pro- cedure language. In Proceedings of the Sympo- sium on Programming (Institut de Programma- tion, University of Paris, Paris, France). 241– 271.
-
-\item
-SILC, J., ROBIC, B., AND UNGERER, T. 1998. Asyn- chrony in parallel computing: from dataflow to multithreading. Parallel Distrib. Comput. Pract. 1, 1, 3–30.
-\end{enumerate}
-
-\subsection{A survey of stream processing, Robert Stephens, 1997}
-This is looking for a general theory of stream processing - sounds useful\cite{stephens1997survey}.
-Highlights that dataflow SPSs can be seen as an implementation of abstract STs.
-One of dataflow's main aims has always been to avoid the ``von Neumann bottleneck'' by allowing parallelism.
-
-Classifies SPSs along three axes: asynchronous or synchronous; deterministic or non; and uni-directional or bidirectional channels. StreamIt, for example, fits into SDU.
-(Although another paper said that one way to implement asynchronous was to use bidirectional channels?)
-
-Functional stream languages:
-``several specialized stream orientated functional languages have been developed including ARTIC (see [60]), HOPE (see [52]) and RUTH (see [98])''
-
-FOCUS: ``Within such networks data is exchanged via unbounded FIFO channels that are modelled as streams.''
-Unbounded channels, of course, is exactly what we don't want.
-
-Reactive systems: it talks about real time systems such as operating systems, which seem like they would actually be closest to what we need.
-
-\subsubsection{Section 8}
-Section 8.2 is a particularly confusing definition of standard list combinators.
-
-Section 8.3 has some prolog definitions of more list combinators, and then some higher-order versions of the combinators.
-There is a combinator called ``Merge'', which is actually an interleaving of every second element:
+Consider the following program, which partitions its input array by the first element of each pair.
+If we plan on executing this program with data that is too large to execute in memory, what is the best execution strategy?
 
 \begin{code}
-merge :: [a] -> [a] -> [a]
-merge (a:a':as) (b:b':bs)
- = a : b' : merge as bs
+app2 :: [(Bool,a)] -> [(Bool,a)]
+app2 xs
+  = filter     (fst) xs
+ ++ filter (not.fst) xs
+
+main
+  = do xs  <- readMassiveFile
+       sendEmails (app2 xs)
 \end{code}
 
-However, this and append are the only combinators with multiple stream inputs.
+We may wish to iterate through the input array only once, but that would require buffering at least the second half of the append in memory, while the first half is being processed.
+Laziness alone is insufficient here, as the second half of the append would still be referencing the start of the array as the first half is being pulled, leading to the same buffering.
 
-\subsubsection{Lucid}
-Section 9.4 is about Lucid. Lucid has some more interesting combinators which take multiple streams.
+Indeed, we believe that this case is in fact best solved by \emph{duplicating work}, to some extent:
 
 \begin{code}
-attime :: (Time -> a) -> (Time -> Time) -> (Time -> a)
-attime as ts t
- = as (ts t)
+app2 :: [(Bool,a)] -> [(Bool,a)] -> [(Bool,a)]
+app2 xs xs'
+  = filter     (fst) xs
+ ++ filter (not.fst) xs'
 
-upon :: [a] -> [Bool] -> [a]
-upon (a:as) bs
- = a : rest (a:as) bs
-
-rest (a:as) (b:bs)
- | b
- = head as : rest as bs
- | otherwise
- = a : rest (a:as) bs
+main
+  = do xs  <- readMassiveFile
+       xs' <- readMassiveFile
+       sendEmails (app2 xs xs')
 \end{code}
 
-So I imagine that these combinators could actually be used for interesting merges.
-However, there is no mention of whether compilation without unbounded buffers can be achieved.
-It looks like primitive operators such as @nor@ are essentially @zipWith (nor)@.
-
-\subsubsection{LUSTRE}
-LUSTRE is related to Lucid, but requires causality and synchronicity.
-Their deadlock checking does reject some valid programs.
-The primitive combinators are: previous (called Z in signal processing?), followed by (something like @head a ++ tail b@), when (find the next\emph{[sic?]} true value in bool stream and use value then) and current (like a past-looking when, except for initial value find future).
-These seem to destroy causality.
-
-This looks promising, but I don't know whether our merges will be allowed by the deadlock checker.
-
-\subsubsection{SIGNAL}
-Could be worth looking into. Synchronous. Not sure about deadlock and buffer detection etc.
-
-\subsection{A Co-iterative Characterization of Synchronous Stream Functions}
-This is \cite{caspi1998co}, Paul Caspi and Marc Pouzet.
-
-Coiteration for streams: giving a stream an initial state, and a transition function from state to pair of value and new state.
-
-First, looking at length-preserving functions, then non-length preserving such as filter.
-Starts off very similar to stream fusion, but the catch is that most streams based on other streams don't actually care about the input stream's guts (transition function), but actually just care about the values!
-
-So instead of
+This is not always the case, however, and there are many situations where using the same input array multiple times does not require multiple traversals.
 
 \begin{code}
-data Stream a s = (s, s -> (s,a))
-map :: Stream a s -> Stream a s
-map = ...
+zip2 :: [(Bool,a)] -> [(Bool,a)]
+zip2 xs
+  = map f xs `zip` map g xs
+
+filter2 :: [(Bool,a)] -> [(Bool,a)]
+filter2 xs
+  = (filter f xs, filter g xs)
 \end{code}
-we have
-\begin{code}
-data Stream i o s = (s, i -> s -> (s,o))
-map :: (a -> b) -> Stream a b s
-map = ...
-\end{code}
-but these can only handle length-preserving functions.
-
-Definition of a synchronous stream function:
-\begin{code}
-co_apply :: CStream (a -> Maybe s -> (b, Maybe s))
-         -> CStream a
-         -> CStream b
-
- f : Stream a -> Stream b
-is synchronous iff there exists
- f' : Stream (a -> s -> (b, s))
-such that
- f == co_apply f'
-\end{code}
-
-Lambda and recursion are a bit confusing. Read again.
-
-Now, the result type of each CStream becomes @F a s@ instead of just @(a, s)@ where @F a s@ is defined:
-\begin{code}
-data F a s
- = P a s
- | S   s
-\end{code}
-which is equivalent to @(Maybe a, s)@.
-They then write an @extend@ (@map apply@) combinator which consumes two inputs at once, but throws an error if one argument yields when the other doesn't.
-Put another way, ``if the clocks of the two arguments are the same'', where the clock is @map isJust@.
-
-Their definition of merge is interesting. The list version they give is
-\begin{code}
-merge (False:cs) xs (y:ys) = y : merge cs xs ys
-merge (True :cs) (x:xs) ys = x : merge cs xs ys
-\end{code}
-which only pulls from one of the true or false streams at a time.
-However, the co-iterative version pulls from all streams at each iteration, but requires that only one of the true or false streams actually produce a value: if both produce a value, it is a runtime error.
-It feels like there is a step missing between the two.
-
-My initial thought is that this clock test is a bit too restrictive for us, because we could store \emph{one} value in a buffer if the two clocks don't exactly line up.
-
-The actual clock calculus looks very similar to a regular type system, with fairly minor and understandable differences.
-
-\subsection{Lucy-n: a n-Synchronous Extension of Lustre}
-This is Pouzet et al again\cite{mandel2010lucy}.
-While the synchronous model requires no buffering, n-synchrony relaxes this by allowing communication through buffers whose size is known at compile-time.
-They use a similar clock calculus to before, but extend it with subtyping. Buffering is introduced through an explicit @buffer@ primitive, but the required buffer size is inferred.
-
-What is ``Cyclo Static Data Flow''?
-Apparently ``synchronous languages'' like Lustre and Lucid Synchrone are not quite as flexible as SDF.
-In synchronous languages, two streams can be composed (zipped?) only if their clocks are \emph{exactly} the same. 
-N-synchronous relaxes this, and allows composition of streams with unequal clocks, but that can be synchronised with a finite, statically known sized, buffer.
-
-``In this paper, we restrict the clock language ce to define ultimately periodic boolean sequences only''
-ie infinite repetitions with some finite prefix.
-
-It looks like the kernel is missing the @on@ operator: @s on e@ which is like a filter.
-It does have @e when ce@ but here @ce@ is a clock, so can only be a repeating boolean sequence.
-No - it does have @on@, but it takes a clock instead of an expression.
-For some clocks @a@ and @b@, @a on b@ is a sub-clock of @a@.
-
-Section 5 is where it gets interesting: ``In order to overcome this complexity, we propose in this paper not to consider exact periodic clocks but their abstraction''.
-They abstract over the clocks with $\langle b^0, b^1 \rangle (r)$, where $b$s are the minimum and maximum shift of @1@s, and $r$ is the proportion of @1@s to @0@s.
-So, basically, $r$ is the slope of the line of @1@s. The resulting abstract slope of @a on b@ can be found quite easily.
-Two clocks can only be scheduled together if their $r$s are the same, that is they have the same slope.
-The size of the buffer needed is the difference between two of the $b$s.
-
-\subsection{Compile-Time Scheduling of Dynamic Constructs in Dataflow Program Graphs}
-This is Ha and Lee\cite{ha1997compile}.
+The goal of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.

### 2015-02-02 2015mergingmerges/section/02-Streams.tex
ok

%!TEX root = ../Main.tex
\section{Streams and combinators}
\label{s:Streams}
First, we define an abstraction of two-input combinators, very similar to stream fusion\CITE or co-iterative stream functions of Caspi and Pouzet\cite{caspi1998co}.
We present a simplified stateless version here, which would not be able to express all useful combinators (such as segmented scan) but suffices for merge, append and zip.
We also only treat combinators that consume both input streams linearly, which excludes combinators like cross product or gather.
In Coq\footnote{The definitions have been reordered for clarity}:

\begin{code}
Definition Stream2 (a b c : Set)
 := option a -> option b -> Step2 c

Inductive Step2 c :=
 | MkStep2 : Move -> Move -> option c -> Step2 c
 | Done2   : Step2 c.

Inductive Move :=
 | Forward
 | Stay.
\end{code}

At each evaluation step, a @Stream2@ is called with the value at each input stream, or @None@ if the input stream is finished.
The return of each @Stream2@ evaluation step can be either a @MkStep2@ indicating that further processing is possible, or @Done2@, in which case the combinator has finished.
A @MkStep2@ contains two @Move@s, which designate which input streams should be consumed and the next value pulled, and which should stay where they are.
@MkStep2@ also contains an optional output value.

The main difference from stream fusion is that the coordination between streams is lifted out.
In stream fusion, each stream transformer takes a stream as input and evaluates it itself.
This means that, since the stream evaluation is local to a stream, scheduling decisions can only be made locally.
These local schedules cannot guarantee desired properties like bounded memory, as multiple streams may need to be stepped together if they use the same input.
Instead, the co-iterative version is evaluated by an outside scheduler, which can find an evaluation order that satisfies these properties.

For single-input combinators such as map, filter and scan, @Stream1@ is defined below.
\begin{code}
Definition Stream1 (a b : Set)
 := option a -> Step2 b

Inductive Step1 b :=
 | MkStep1 : Move -> option b -> Step1 b
 | Done1   : Step1 b.
\end{code}

\subsection{Examples}

Below, we show example combinators.
The first combinator pairwise zips two streams together, consuming both streams in lock-step.

\begin{code}
Definition Zip (A B : Set)
 :  Stream2 A B (A*B)
 := fun a b =>
    match (a,b) with
    | (Some a, Some b)
    => MkStep2 Forward Forward (Some (a,b))
    | _               
    => Done2
    end.
\end{code}

Next, append consumes from the left stream until it is empty, then consumes the right stream.
\begin{code}
Definition Append (A : Set)
 :  Stream2 A A A
 := fun a b =>
    match (a,b) with
    | (Some a, _)
    => MkStep2 Forward Stay    (Some a)
    | (None,   Some b)
    => MkStep2 Stay    Forward (Some b)
    | (None,   None)               
    => Done2
    end.
\end{code}

As a final example, the merge join inspects both its input elements, and consumes the one with the smallest value.
There are many variations on this theme, but this particular version that outputs all values in order is actually equivalent to a segmented append.

\begin{code}
Definition MergeJoin (A : Set)
 :  Stream2 (nat*A) (nat*A) (nat*A)
 := fun a b =>
    match (a,b) with
    | (Some (ia,a), Some (ib, b))
    => if   le_dec ia ib
       then MkStep2 Forward Stay (Some (ia, a))
       else MkStep2 Stay Forward (Some (ib, b))
    | (Some (ia,a), None)
    =>      MkStep2 Forward Stay (Some (ia,a))
    | (None, Some (ib,b))
    =>      MkStep2 Stay Forward (Some (ib,b))
    | (None, None)
    =>      Done2
    end.
\end{code}

\subsection{Kahn process networks}

We define a simplified Kahn process network that contains no cycles (recursion).

\begin{code}
Inductive Graph
 : list Set -> list Set -> list Set -> Type :=
\end{code}
The graph is parameterised by the list of inputs, the list of interior nodes, and the list of outputs.

\begin{code}
 | Empty  : Graph [] [] []

 | Inp    : forall a i n o,
            Graph i n o
         -> Graph (a::i) (a::n) o

 | Out    : forall a i n o,
            Index n a
         -> Graph i n o
         -> Graph i n (a::o)
\end{code}

A @Graph@ can be @Empty@, in which case there are no inputs, nodes or outputs.
The @Inp@ constructor takes an existing graph, and adds an input node to the start.
Because the graph cannot be recursive, the inner graph of @Inp@ cannot refer to the newly created @Inp@.
Similarly, @Out@ takes an existing graph, and an index into the list of interior nodes, and marks that node as an output.

\begin{code}
 | Trans1 : forall a b i n o,
            Index n a
         -> Stream1 a b
         -> Graph i n o
         -> Graph i (b::n) o.

 | Trans2 : forall a b c i n o,
            Index n a
         -> Index n b
         -> Stream2 a b c
         -> Graph i n o
         -> Graph i (c::n) o.
\end{code}
The two transition constructors, @Trans1@ and @Trans2@ take @Stream1@ and @Stream2@ respectively, which defines the actual computation of the transition. They also extend existing graphs, and their inputs must be in the existing graph, again disallowing recursive edges or cycles.

### 2015-02-02 2015mergingmerges/section/03-Access.tex
ok

%!TEX root = ../Main.tex
\section{Access patterns}
\label{s:Access}

The key insight is to treat combinators' access patterns as regular expressions over the names of streams.
When an input stream is mentioned in a regular expression, it signifies that a blocking read must be made from that stream before the combinator can proceed.
When output, written @E@ for emit, is mentioned, a blocking write is made to the output.
We associate these regular expression access patterns with a combinator, process or network, such that for any given set of inputs, there exists an execution trace that satisfies the regular expression:
\[ \forall g : @Graph@.~ \forall is : @Inputs@.~ \exists t : @Trace@~g~is.~
    \flatten(t) \in \regex(g) \]

Where $\flatten : @Trace@~g~i \to [@Index@~n]$ converts an execution trace of a graph to the list of indices as they fired, and $\regex : @Graph@ \to @Regex@~(@Index@~n)$ retrieves the access pattern of a given graph.

For example, @map a@ reads from @a@ at every step, then emits a value @E@, and repeats until no more data remains in @a@.
We define the operator @:~@ as ``has access pattern''.
\begin{code}
map     a   :~ (aE)*
filter  a   :~ (aE?)*
zip     a b :~ (abE)*
append  a b :~ (aE)*(bE)*
merjoin a b :~ ((aE)*(bE)*)*
\end{code}
The access pattern of the merge join (@merjoin@) also illustrates its relationship with append and segmented append.
These access patterns be composed together to apply to whole graphs:

\begin{code}
append (map a) (filter b) :~ (aE)*(bE?)*
append (zip a b) (zip a b) :~ (abE)*(abE)*
\end{code}
\TODO{but I don't have a good definition for this yet}

\subsection{Traces}
\TODO{Define what a trace is}.

We define filter functions over traces and regular expressions, which, given two graphs and the trace or acess pattern of one, filters it to the intersection of the names.
\[ \filterT(g', t) = \filter (\in flatten(g')) t \]
\[ \filterR(g', r) = \filter (\in flatten(g')) r \]
It also holds that
\[ \forall rx. \forall s. \forall w. w \in rx \implies \filter(\in s) w \in \filter(\in s) rx\]

We say that two traces are \emph{affable} if they agree on common parts:
\[ \affable(t : @Trace@~g~is, t' : @Trace@~g'~is) = \filterT(g', t) == \filterT(g, t') \]

For example, any traces of @map a@ and @zip a b@ would be affable, since after filtering out the @b@s both traces would be equal.

Two graphs are fusible if, for all possible inputs, there exist affable traces:
\begin{align*}
\mathit{fusible}(g, g')
        &=  \forall is : @Inputs@~(g \cap g').  \\
        &   \exists t  : @Trace@~g~is.          \\
        &   \exists t' : @Trace@~g'~is.         \\
        &   \affable(t, t')                     %
\end{align*}

Indeed, if we have two graphs, $\forall g~g' : @Graph@.$
\begin{eqnarray*}
 & \filterR(\regex(g)) = \filterR(\regex(g')) \\
 \implies & \forall w. w \in \filterR(\regex(g)) \iff w \in \filterR(\regex(g')) \\
 \implies & \forall is. \exists t : @Trace@~g~is. \exists t' : @Trace@~g'~is. \\
          & \filterT(\flatten(t)) \in \filterR(\regex(g)) \\
          & \wedge~\filterT(\flatten(t')) \in \filterR(\regex(g')) \\
          & \wedge~\filterT t = \filterT t' \\
 \implies & \affable(t, t') \\
 \implies & \mathit{fusible}(g, g') \\
\end{eqnarray*}

### 2015-02-03 2015mergingmerges/section/03-Access.tex
ok

@@
           & \wedge~\filterT(\flatten(t')) \in \filterR(\regex(g')) \\
+\end{eqnarray*}
+
+But it is not necessarily true that these two traces $t$ and $t'$ have the same filtered flattening.
+I \emph{think} it is true, because they have the same inputs, but I need to think about how to prove it.
+Something about determinism.
+\begin{eqnarray*}
           & \wedge~\filterT t = \filterT t' \\
@@
 \end{eqnarray*}
+
+We can split the combinators into two types: value dependent and independent.
+Filter, merge join and indices are value dependent, whereas map, append, zip, and so on are value independent.
+Value independent combinators have the property that for all inputs, the trace can be determined solely by the length of the inputs, regardless of the values.
+
+\subsubsection{Examples}
+Going back to the original @app2@ example, let us compute its access pattern.
+I use a simplified, canonical form here and discard the worker functions.
+
+\begin{code}
+app2 :: [(Bool,a)] -> [(Bool,a)]
+app2 X
+ = let F = filter X     :~ (XF?)*
+       G = filter X     :~ (XG?)*
+       H = append F G   :~ (XF?)*(XG?)*
+   in  H
+\end{code}
+If we try to fuse @F@ and @G@, we start by finding the common parts, which is just @X@. Filtering both access patterns to only occurrences of @X@ leaves @X*@ and @X*@, which are trivially equal.
+Therefore, @F@ and @G@ can be fused.
+
+However, if we attempt to fuse @F@ with @H@, we must again find the intersection: this being @X@ and @F@.
+We find that @F@'s filtered access pattern remains the same, @(XF?)*@, while @H@'s filtered access pattern is @(XF?)*X*@.
+These two regular expressions are not equal, and thus @F@ and @H@ may not be fused.

### 2015-02-03 2015mergingmerges/section/XX-RelatedWork.tex
ok

@@
 Not much really conceptual work here.
+
+\subsection{Solving out-of-order communication in Kahn Process Networks}
+This is \cite{turjan2002compile}, resulting from a search about regular expressions and KPNs.
+\emph{Compaan} is about converting signal processing Matlab code to KPNs.
+Determine whether, for their given networks, reordering is necessary to limit buffers and deadlocks.
+It looks like this only works for statically known polytope loops.
+
+\subsection{Realtime Signal Processing}
+This is \cite{reekie1995realtime} by Hideki John Reekie at UTS.
+Very interesting work on Visual Haskell, a visual dataflow-esque representation of Haskell programs.
+Talks about static and dynamic dataflow, but really just seems to be about the required combinators, and not compilation.
+Rewrite-rules based optimisations, so only local information like map/map fusion etc.
+
+\subsection{Requirements on the execution of Kahn process networks}
+This is \cite{geilen2003requirements}.
+
+Looks like this is directed at solving some problems in Parks\cite{parks1995bounded} where KPNs are scheduled in bounded memory, but has incorrect behaviour for some KPNs.
+The incorrect behaviour is that ``local artificial deadlocks'' can occur, where a subset of the network cannot proceed as it requires a larger buffer, but other parts can still make progress.
+Parks only deals with ``global artificial deadlocks'', where the entire network can make no progress.
+As a result, the output may not be ``complete''.
+
+Writing is normally non-blocking, but they have writing to a full buffer block.
+Talks about using cycles to detect local deadlocks. 
+Shows that process networks with channel bounds are deterministic - and our subset of KPN has all channels bounded by 1.
+Their scheduling algorithm just enlarges the smallest full buffer whenever any sort of deadlock occurs.

### 2015-03-02 2015mergingmerges/Main.tex
words
ok

@@
 \begin{abstract}
+Fusion is an essential part of optimising high-level array computations.
+Fusion systems for combinators tend to focus on combinators with data-independent access patterns such as @zip@, @filter@ and @map@, while ignoring @merge@ as in merge sort.
+In this paper, we show that describing combinators as deterministic finite automata allows us to express more combinators, while leading to a fusion algorithm based on intersection of automata.
 \end{abstract}
@@
 \input{section/01-Introduction.tex}
+\input{section/02-Streams.tex}
+\input{section/03-Access.tex}
+\input{section/XX-RelatedWork.tex}
 

### 2015-03-02 2015mergingmerges/section/01-Introduction.tex
ok

@@
 
-Consider the following program, which partitions its input array by the first element of each pair.
+Consider the following program, which loads some accounts, sums up the balance of their loans, and sends debt collection emails to some, while raises the credit card limit of others.
+Can we execute this without buffering the entire file in memory? Perhaps not, but we could if the relative ordering of debt collecting and raising limits is not important.
 If we plan on executing this program with data that is too large to execute in memory, what is the best execution strategy?
@@
 \begin{code}
-app2 :: [(Bool,a)] -> [(Bool,a)]
-app2 xs
-  = filter     (fst) xs
- ++ filter (not.fst) xs
+arrearsOffers
+    :: [(AccountId, Account)]
+    -> [(AccountId, Loan)]
+    -> ([Account], [Account])
+arrearsOffers as ls
+ = let als  = merge as ls
+       bals = group_by sum_balance als
+       arrs = filter (<0   . balance) bals
+       offs = filter (>500 . balance) bals
+   in (arrs, offs)
 
 main
-  = do xs  <- readMassiveFile
-       sendEmails (app2 xs)
+ = do accounts <- readAccounts
+      loans    <- readLoans
+      let (arrears, offers)
+                = arrearsOffers accounts loans
+      mapM_ sendDebtCollector arrears
+      mapM_ raiseLimit        offers
 \end{code}

### 2015-03-03 2015mergingmerges/section/01-Introduction.tex
rewrite intro

--- 2015-02-02 2015mergingmerges/section/01-Introduction.tex
@@
 
-Our goal, then, appears relatively simple: to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
+In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing}\TODO{correct?} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
+This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
+Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.
 
-\subsection{Motivation}
+Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
 
-Consider the following program, which partitions its input array by the first element of each pair.
-If we plan on executing this program with data that is too large to execute in memory, what is the best execution strategy?
+\subsection{Buffering}
+
+Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
+However, when dealing with data too large to fit in memory such as tables on disk, intermediate arrays becomes essential rather than just desirable.
+Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
+For these situations, some sort of assurance of total fusion are required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+
+Consider the following program.
+Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:
 
 \begin{code}
-app2 :: [(Bool,a)] -> [(Bool,a)]
-app2 xs
-  = filter     (fst) xs
- ++ filter (not.fst) xs
-
-main
-  = do xs  <- readMassiveFile
-       sendEmails (app2 xs)
+zip1 :: Source Int -> Source Int -> IO ()
+zip1 xs ys
+ = let xs'  = filter (>0) xs
+       ys'  = filter (<0) ys
+       zz   = zip xs' ys'
+   in  mapM_ fire_missile zz
 \end{code}
 
-We may wish to iterate through the input array only once, but that would require buffering at least the second half of the append in memory, while the first half is being processed.
-Laziness alone is insufficient here, as the second half of the append would still be referencing the start of the array as the first half is being pulled, leading to the same buffering.
-
-Indeed, we believe that this case is in fact best solved by \emph{duplicating work}, to some extent:
+This program can be executed with no buffering of either input or any intermediate arrays.
+Now suppose wish to perform another action over the first source:
 
 \begin{code}
-app2 :: [(Bool,a)] -> [(Bool,a)] -> [(Bool,a)]
-app2 xs xs'
-  = filter     (fst) xs
- ++ filter (not.fst) xs'
-
-main
-  = do xs  <- readMassiveFile
-       xs' <- readMassiveFile
-       sendEmails (app2 xs xs')
+zip2 :: Source Int -> Source Int -> IO ()
+zip2 xs ys
+ = let xs'  = filter (>0) xs
+       ys'  = filter (<0) ys
+       xs2  = map    (+1) xs
+       zz   = zip xs' ys'
+   in  do   mapM_ fire_missile zz
+            mapM_ fortify      xs2
 \end{code}
 
-This is not always the case, however, and there are many situations where using the same input array multiple times does not require multiple traversals.
+Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ anf performing the second action afterwards.
+However, we could execute this with no buffering if the relative ordering of the two @mapM_@s was not important, and could be reordered and interspersed as the data is read.
+We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.
 
 \begin{code}
-zip2 :: [(Bool,a)] -> [(Bool,a)]
-zip2 xs
-  = map f xs `zip` map g xs
+zip3 :: Source Int -> Source Int -> IO ()
+zip3 xs ys
+ = let xs'  = filter (>0) xs
+       ys'  = filter (<0) ys
+       xs2  = map    (+1) xs
+       zz   = zip xs' ys'
+   in  When fire_missile zz
+   :&& When fortify      xs2
+\end{code}
 
-filter2 :: [(Bool,a)] -> [(Bool,a)]
-filter2 xs
-  = (filter f xs, filter g xs)
+As a final example, suppose we removed the second data source and filtered the same source according to different predicates, then paired them together.
+Can we still execute this with no buffer, and only iteration over @xs@?
+
+\begin{code}
+zip4 :: Source Int -> IO ()
+zip4 xs
+ = let xs'  = filter (>0) xs
+       xs'' = filter (<0) xs
+       zz   = zip xs' xs''
+   in  mapM_ fire_missile zz
 \end{code}
-The goal of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.
+
+Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
+It may not be immediately obvious just looking at the program that it requires two iterations.
+
+In summary, the purpose of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.

### 2015-03-03 2015mergingmerges/section/03-Machines.tex
ok, some notes on the particular kind of DFA

%!TEX root = ../Main.tex
\section{Machines}
\label{s:Machines}

We use a particular type of deterministic finite automata (DFAs) where each state has a type that restricts the alphabet of output transitions.
We use state types such as @Pull a@ with output transitions @Some a@ and @None@, or type @Out b@ with single output transition @Unit@.

\subsection{Restricted DFA}

We describe a restricted automaton as a eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.

As usual, $Q$ is the set of states, $\Sigma$ the alphabet, but in addition $\Gamma$ is a set of state types.
We define the types of functions as follows:

\[ \gamma : Q \to \Gamma \]
Every state has a type.

\[ \sigma : \Gamma \to \{ \Sigma \} \]
Every type has some subset of the alphabet that transitions are required for.

\[ \delta : \forall q : Q. \sigma(\gamma(q)) \to Q \]
For each state, the transition function need only be defined for those letters of the state's type.

These restricted DFAs can be converted to normal automata by adding a rejecting sink state, and adding transitions for all missing letters of the alphabet to this rejecting state.

\subsection{State types}
We now describe a set of types and alphabet for describing combinator programs as a restricted DFA.
These are parameterised by both the names of sources, sinks and bindings $n$, and the type of worker functions $f$ (such as @Expression@ for code generation).

\begin{tabbing}
MM \= MM \= MMMMMM \= M \= M \kill
$\Gamma$ \> $=$ \> @Pull@     \>     \> $n$         \\
         \>     \> (Attempt to read from a named input) \\

         \> $~|$\> @Release@  \>     \> $n$         \\
         \>     \> (Read inputs must be released) \\

         \> $~|$\> @Out@      \> $f$ \> $n$         \\
         \>     \> (Emit the value computed by $f$ to output channel $n$) \\

         \> $~|$\> @OutDone@  \>     \> $n$         \\
         \>     \> (Signal that output channel $n$ is complete) \\

         \> $~|$\> @If@       \> $f$ \> $n$         \\
         \>     \> (If based on current state of output channel $n$) \\

         \> $~|$\> @Update@   \> $f$ \> $n$         \\
         \>     \> (Update output channel $n$'s current state) \\

         \> $~|$\> @Skip@     \>                    \\
         \>     \> (Simple goto, useful for code generation) \\

         \> $~|$\> @Done@     \>                    \\
         \>     \> (The program is finished) \\
\end{tabbing}

Each of these state types requires a corresponding set of output transitions to be defined.
\begin{tabbing}
MMMMMMMM \= M \= \kill
$\gamma(@Pull @n)$
                            \> $=$ \> \{@Some@ n, @None@\} \\
$\gamma(@Release @n)$
                            \> $=$ \> \{@Unit@\} \\
$\gamma(@Out @   f~n)$
                            \> $=$ \> \{@Unit@\} \\
$\gamma(@OutDone @ n)$
                            \> $=$ \> \{@Unit@\} \\
$\gamma(@If @    f~n)$
                            \> $=$ \> \{@True@, @False@\} \\
$\gamma(@Update @f~n)$
                            \> $=$ \> \{@Unit@\} \\
$\gamma(@Skip@      )$
                            \> $=$ \> \{@Unit@\} \\
$\gamma(@Done@      )$
                            \> $=$ \> \{\} \\
\end{tabbing}

\subsection{Conversion of combinators}
The combinators from section~\ref{s:Combinators} can be converted to these DFAs quite easily.

Well, do it.
I'm still not sure, maybe just showing the pictures would be better.

### 2015-03-04 2015mergingmerges/section/02-Machines.tex
ok

--- 2015-03-03 2015mergingmerges/section/03-Machines.tex
@@
 
-\subsection{Conversion of combinators}
-The combinators from section~\ref{s:Combinators} can be converted to these DFAs quite easily.
+\subsection{Combinators}
+Many interesting combinators can be described using these machines.
+Below, the machines for some important combinators are given.
+The machines are described in an almost BASIC-like syntax with numbered state labels before the colon, followed by the state, and then the transitions.
+If the transitions are not shown, a @Unit@ transition to the next numbered state is implicit.
+The combinator's output channel is assumed to be named @o@.
 
-Well, do it.
-I'm still not sure, maybe just showing the pictures would be better.
+While writing combinators in this form is neither pleasant nor pretty, it only needs to be done once.
+The implementation contains more combinators such as group by, append and indices of segment lengths.
+
+\subsubsection{map f xs}
+\begin{code}
+10: Pull xs         (Some x => 20; None  => 90)
+20: Out o (f x)
+30: Release x       (Unit   => 10)
+
+90: OutDone o
+99: Done
+\end{code}
+
+\subsubsection{filter p xs}
+\begin{code}
+10: Pull xs         (Some x => 20; None  => 90)
+20: If (p x)        (True   => 30; False => 40)
+30: Out o x
+40: Release x       (Unit   => 10)
+
+90: OutDone o
+99: Done
+\end{code}
+
+\subsubsection{zip xs ys}
+\begin{code}
+10: Pull xs         (Some x => 20; None  => 90)
+20: Pull ys         (Some x => 30; None  => 80)
+30: Out o (x, y)
+40: Release x
+50: Release y       (Unit   => 10)
+
+80: Release x
+90: OutDone o
+99: Done
+\end{code}
+
+\subsubsection{merge xs ys}
+\begin{code}
+10: Pull xs         (Some x => 20; None  => 100)
+20: Pull ys         (Some y => 30; None  => 201)
+30: If (x <= y)     (True   => 40; False => 50)
+
+40: Out o x
+41: Release x
+42: Pull x          (Some x => 30; None  => 101)
+
+50: Out o y
+51: Release y
+52: Pull y          (Some y => 30; None  => 201)
+
+-- xs is empty
+100: Pull ys        (Some y => 101; None => 900)
+101: Out o y
+102: Release y      (Unit   => 100)
+
+-- ys is empty
+200: Pull xs        (Some x => 201; None => 900)
+201: Out o x
+202: Release x      (Unit   => 200)
+
+900: OutDone o
+901: Done
+\end{code}
+
+Figures 1-3 show automata for combinators @map@, @filter@ and @zip@.
+To reduce clutter, these figures omit labels for @Unit@ transitions.
+
+\input{figures/Combinators.tex}
+
+\subsection{Code extraction}
+
+Extracting imperative code from these machines is quite straightforward.
+Variables are created for each 
+Each state is converted to a basic block and given a unique identifier as its label.

### 2015-03-04 2015mergingmerges/section/02-Machines.tex
remove progs

--- 2015-03-03 2015mergingmerges/section/03-Machines.tex
@@
 
-\subsection{Conversion of combinators}
-The combinators from section~\ref{s:Combinators} can be converted to these DFAs quite easily.
+\subsection{Combinators}
+Many interesting combinators can be described using these machines.
+In figures 1-4, the machines for some important combinators are given.
+To reduce clutter, these figures omit labels for @Unit@ transitions.
 
-Well, do it.
-I'm still not sure, maybe just showing the pictures would be better.
+While writing combinators in this form is neither pleasant nor pretty, it only needs to be done once.
+The implementation contains more combinators such as group by, append and indices of segment lengths.
+
+\input{figures/Combinators.tex}
+
+\subsection{Code extraction}
+
+Extracting imperative code from these machines is quite straightforward.
+Variables are created for each 
+Each state is converted to a basic block and given a unique identifier as its label.

### 2015-03-04 2015mergingmerges/section/02-Machines.tex
merge section

@@
 
+Note that @Out@, @OutDone@, @If@ and @Update@ are also annotated with the name of the output channel.
+While a single combinator will only have one output, after merging multiple machines together there could be multiple outputs.
+Each combinator has its own local mutable state, accessible by the functions given to @Update@, @Out@ and @If@.
+These machine state types must be annotated with the name of the output channel to designate which local mutable state, if any.
+
 \subsection{Combinators}

### 2015-03-04 2015mergingmerges/section/03-Access.tex
merge section

--- 2015-02-02 2015mergingmerges/section/03-Access.tex
@@
 %!TEX root = ../Main.tex
-\section{Access patterns}
-\label{s:Access}
+\section{Merging machines}
+\label{s:Merging}
 
-The key insight is to treat combinators' access patterns as regular expressions over the names of streams.
-When an input stream is mentioned in a regular expression, it signifies that a blocking read must be made from that stream before the combinator can proceed.
-When output, written @E@ for emit, is mentioned, a blocking write is made to the output.
-We associate these regular expression access patterns with a combinator, process or network, such that for any given set of inputs, there exists an execution trace that satisfies the regular expression:
-\[ \forall g : @Graph@.~ \forall is : @Inputs@.~ \exists t : @Trace@~g~is.~
-    \flatten(t) \in \regex(g) \]
+This section describes how machines can be merged together to form a single machine that computes both.
 
-Where $\flatten : @Trace@~g~i \to [@Index@~n]$ converts an execution trace of a graph to the list of indices as they fired, and $\regex : @Graph@ \to @Regex@~(@Index@~n)$ retrieves the access pattern of a given graph.
+The idea is to create a new machine with the product of the two states; at each state, one or the other of the machines may be able to take a step.
+If so, we update that machine's state and leave the other one as-is.
+For shared inputs or outputs, both machines must take the steps at the same time: if they both pull from the same input, they must pull at the same time.
+Similarly, if one machine produces an input and the other reads it, the first machine can only produce its output if the second is pulling on it.
+If neither machine can make a move, the program is disallowed.
+For example, if the first machine is trying to read from a shared input while the second tries to read from the first machine's input, executing it may require buffering the input until the second machine catches up.
 
-For example, @map a@ reads from @a@ at every step, then emits a value @E@, and repeats until no more data remains in @a@.
-We define the operator @:~@ as ``has access pattern''.
+However, the above explanation is too strict and outlaws programs such as @zip xs ys@ merged with @zip ys xs@, as neither machine can progress until the other does, leading to deadlock.
+To resolve this, we add to the resulting machine's state a set of pending events for each machine.
+The first machine may read from a shared input as long as there is not already a ``read'' event pending for the other machine to deal with.
+Reading from a shared input adds a pending ``read'' event to the other machine, and the other machine may then skip past a @Pull@ for that source, as it is known that the other machine has already pulled on it.
+
+\subsection{General merging}
+Talk about merging with
 \begin{code}
-map     a   :~ (aE)*
-filter  a   :~ (aE?)*
-zip     a b :~ (abE)*
-append  a b :~ (aE)*(bE)*
-merjoin a b :~ ((aE)*(bE)*)*
+data Event
+ = Read n
+ | Finished n
+
+merge :: Machine l1 -> Machine l2
+      -> Machine (l1, l2, Set Event)
 \end{code}
-The access pattern of the merge join (@merjoin@) also illustrates its relationship with append and segmented append.
-These access patterns be composed together to apply to whole graphs:
+
+\subsection{Producer-consumer}
+The general case above is sufficient, but in some cases can generate larger machines than necessary.
+Talk about special case for producer-consumers.
 
 \begin{code}
-append (map a) (filter b) :~ (aE)*(bE?)*
-append (zip a b) (zip a b) :~ (abE)*(abE)*
+data Which = LeftExec | RightExec
+
+mergeV :: Machine l1 -> Machine l2
+       -> Machine (l1, l2, Which)
 \end{code}
-\TODO{but I don't have a good definition for this yet}
-
-\subsection{Traces}
-\TODO{Define what a trace is}.
-
-We define filter functions over traces and regular expressions, which, given two graphs and the trace or acess pattern of one, filters it to the intersection of the names.
-\[ \filterT(g', t) = \filter (\in flatten(g')) t \]
-\[ \filterR(g', r) = \filter (\in flatten(g')) r \]
-It also holds that
-\[ \forall rx. \forall s. \forall w. w \in rx \implies \filter(\in s) w \in \filter(\in s) rx\]
-
-We say that two traces are \emph{affable} if they agree on common parts:
-\[ \affable(t : @Trace@~g~is, t' : @Trace@~g'~is) = \filterT(g', t) == \filterT(g, t') \]
-
-For example, any traces of @map a@ and @zip a b@ would be affable, since after filtering out the @b@s both traces would be equal.
-
-Two graphs are fusible if, for all possible inputs, there exist affable traces:
-\begin{align*}
-\mathit{fusible}(g, g')
-        &=  \forall is : @Inputs@~(g \cap g').  \\
-        &   \exists t  : @Trace@~g~is.          \\
-        &   \exists t' : @Trace@~g'~is.         \\
-        &   \affable(t, t')                     %
-\end{align*}
-
-Indeed, if we have two graphs, $\forall g~g' : @Graph@.$
-\begin{eqnarray*}
- & \filterR(\regex(g)) = \filterR(\regex(g')) \\
- \implies & \forall w. w \in \filterR(\regex(g)) \iff w \in \filterR(\regex(g')) \\
- \implies & \forall is. \exists t : @Trace@~g~is. \exists t' : @Trace@~g'~is. \\
-          & \filterT(\flatten(t)) \in \filterR(\regex(g)) \\
-          & \wedge~\filterT(\flatten(t')) \in \filterR(\regex(g')) \\
-          & \wedge~\filterT t = \filterT t' \\
- \implies & \affable(t, t') \\
- \implies & \mathit{fusible}(g, g') \\
-\end{eqnarray*}

### 2015-03-16 2015mergingmerges/section/01-Introduction.tex
defn of merge

@@
 Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
-However, when dealing with data too large to fit in memory such as tables on disk, intermediate arrays becomes essential rather than just desirable.
+However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
 Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
@@
 
-Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ anf performing the second action afterwards.
+Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
 However, we could execute this with no buffering if the relative ordering of the two @mapM_@s was not important, and could be reordered and interspersed as the data is read.
@@
    in  When fire_missile zz
-   :&& When fortify      xs2
+   :&  When fortify      xs2
 \end{code}
@@
 Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
-It may not be immediately obvious just looking at the program that it requires two iterations.
+It may not be immediately obvious just looking at the program that it requires two iterations, which is why we would like this to be automated.
 

### 2015-03-16 2015mergingmerges/section/02-Machines.tex
defn of merge

@@
          \> $~|$\> @Skip@     \>                    \\
-         \>     \> (Simple goto, useful for code generation) \\
+         \>     \> (Simple goto, simplifies merge algorithm) \\
 
@@
 
+In the proceeding parts, any reference to @Machine@ is a restricted DFA with these state types.
+
 \subsection{Combinators}
@@
 Extracting imperative code from these machines is quite straightforward.
-Variables are created for each 
+Variables are created for each @Pull@ source, to store a single value read from the source, and each output channel has a variable to store its latest output and its current state.
 Each state is converted to a basic block and given a unique identifier as its label.
+
+For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the following:
+\begin{code}
+run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
+run pull_xs when_o = l10
+ where
+  l10 = do  x  <- pull_xs
+            case x of
+             Nothing -> l90
+             Just x' -> l15 x'
+
+  l15 x | p x       = l20 x
+        | otherwise = l30 x
+
+  l20 x =   when_o (Just x)  >> l30 x
+
+  l30 x =   l10
+
+  l90   =   when_o Nothing  >> l91
+
+  l91   =   return ()
+\end{code}

### 2015-03-16 2015mergingmerges/section/03-Access.tex
defn of merge

@@
 Similarly, if one machine produces an input and the other reads it, the first machine can only produce its output if the second is pulling on it.
+
 If neither machine can make a move, the program is disallowed.
@@
 \subsection{General merging}
-Talk about merging with
-\begin{code}
-data Event
- = Read n
- | Finished n
+Let us define the potential pending events on a machine.
+One machine may have read from a shared input, but the other has not.
+Similarly, if one machine has tried to read from a shared input but the input is empty, when the other machine attempts to read it can skip directly to the @None@ case.
 
-merge :: Machine l1 -> Machine l2
-      -> Machine (l1, l2, Set Event)
-\end{code}
+The other cases are when one machine produces a value, and another consumes it.
+Here, the producer may have produced a value, or may have finished producing on that channel.
+
+\begin{tabbing}
+MM \= MM \= MMMMMM \= M\kill
+$\Psi$ \> $=$  \> @Value@     \> $n$         \\
+       \> $~|$ \> @Finished@  \> $n$         \\
+\end{tabbing}
+
+The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times l_2 \times \{\Psi\}$.
+It starts by creating an empty machine, and inserting states one by one, starting from the initial state as the initial state of each machine, and empty pending sets.
+Then, for each output transition of the input machines, states are recursively added.
+
+\begin{tabbing}
+MMMMMMM \= MM \= MMMMMM \= M\kill
+@merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
+        \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
+\\
+@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~(@init @m_2)~\{\}~@empty@$ \\
+\\
+$@merge@'$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\} \times @Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
+           \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
+\\
+MMM \= M \= MMMMMMMMMM \= M \=\kill
+$@merge@'~s_1~s_2~\psi~m$ \\
+ \> $~|$ \> $(s_1, s_2, \psi) \in m$  \\
+ \> $=$  \> $m$ \\
+ \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~s_2~\psi$ \\
+ \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
+\end{tabbing}
+
+The $@merge@'$ function adds the given state and recursively adds the successor states to the given machine.
+If the given state has already been added --- for example the state is reachable by multiple states --- the state need not be added.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM\kill
+$@move@$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\}$ \\
+           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times l_2 \times \{\Psi\})\}$ \\
+\end{tabbing}
+
+The @move@ function computes the states and transitions of the merged machine.
+These are added to the machine by $@merge@'$ above.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
+\\
+ \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
+ \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
+\end{tabbing}
+
+If either of the machines are attempting to update their local state, this can be done easily without interfering with the other machine.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
+\\
+ \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
+ \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
+\end{tabbing}
+
+Skips are dealt with in the same way, affecting only one machine.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@True@$ \\
+ \> $\wedge$ \> $s_1''$     \> $=$ \> $@trans@~m_1~s_1~@False@$ \\
+ \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1',s_2,\psi)), (@False@, (s_1'', s_2, \psi))\})$ \\
+\\
+ \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@True@$ \\
+ \> $\wedge$ \> $s_2''$     \> $=$ \> $@trans@~m_2~s_2~@False@$ \\
+ \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1,s_2',\psi)), (@False@, (s_1, s_2'', \psi))\})$ \\
+\end{tabbing}
+
+Similarly, @If@s only affect one of the machines, leaving the other in its original state.
+If we had an oracle for checking functional equivalence, we could check whether two @If@s had the same function and inputs and potentially allow more programs.
+However, lacking an oracle, even if both machines have @If@s with the same predicate, we generate a machine for all possible result combinations (all four of them).
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Done@$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $=$ \> $(@Done@,~\{\})$ \\
+\end{tabbing}
+
+If both machines are @Done@, the resulting machine will also be @Done@.
+
+The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Pull@, @Release@, @Out@ and @OutDone@ are more complicated.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \setminus \{@Value@~n\} ))\})$ \\
+\end{tabbing}
+
+If the first machine is producing an output on channel $n$ and the second machine uses $n$, the first machine would usually have to wait until the second machine is ready to accept new data.
+However, if the second machine is @Done@, the first machine may output to the channel as frequently as it likes.
+Note that the output channel $n$ may have multiple consumers, so continuing to output after the second machine consumer has finished is not as futile as it may seem at first.
+\TODO{Although, all the combinators so far use the whole input, and anyway checking for @Done@ of machine is too fine-grained if machine already has multiple combinators; perhaps an ``I promise not to read from X any more'' state would be interesting.}
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Value@~n\} ))\})$ \\
+\end{tabbing}
+
+Again, if the first machine is producing and the second is consuming, the first machine may only produce if there is no unhandled value on the channel.
+This is what $\psi$ is for.
+If there is a value in $\psi$, a later case may allow the second machine to pull from it.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
+\end{tabbing}
+
+The final case for @Out@ is when the second machine does not use this output, allowing the first machine to output at any time.
+The cases where the second machine is on an @Out@ are symmetric to these.
+
+The cases for @OutDone@ proceed similarly, and are also symmetric for the first and second machines.
+
+\begin{tabbing}
+MMM \= M \= MMMMMMM \= M \=\kill
+ \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
+ \\
+ \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Finished@~n\}))\})$ \\
+ \\
+ \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
+ \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
+ \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
+ \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
+\end{tabbing}
+
+Pulls now.
+
+Releases.
+
+If none of the above cases apply, it means the machines cannot be fused.
+Either they require more than one element of buffering, or reasoning about function equality would be required to prove that they do not.
+The only unhandled cases above are where the two machines share input or outputs, and there are existing values buffered in $\psi$.
+In which case, we give a compile time error indicating that these two machines cannot be fused.
 

### 2015-03-17 2015mergingmerges/section/02-Machines.tex
add Close state to machines

@@
          \> $~|$\> @Release@  \>     \> $n$         \\
-         \>     \> (Read inputs must be released) \\
+         \>     \> (Values must be released before the input can be pulled again) \\
+
+         \> $~|$\> @Close@    \>     \> $n$         \\
+         \>     \> (A named input may still have data, but we will not read it) \\
 
@@
                             \> $=$ \> \{@Unit@\} \\
+$\gamma(@Close @n)$
+                            \> $=$ \> \{@Unit@\} \\
 $\gamma(@Out @   f~n)$
@@
 In the proceeding parts, any reference to @Machine@ is a restricted DFA with these state types.
+
+\subsection{Invariants}
+Not all machines are valid or meaningful, and we wish to rule them out.
+For example, a machine with an initial state of $@If@~(x>5)~n$ is meaningless because there is no local $x$ that has been read from an input. 
+On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
+An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but the explicit pulls makes synchronising two machines easier.
+
+The invariants are:
+\begin{itemize}
+\item each function mentioned in @Out@, @If@ or @Update@ can only refer to previously read values;
+\item each successful @Pull@ must be @Released@ before another @Pull@ can be made;
+\item all inputs must either be finished by an empty @Pull@ or @Close@d at @Done@;
+\item all outputs must be finished with @OutDone@ at @Done@;
+\item inputs cannot be read after an empty @Pull@ or after they are @Close@d;
+\item outputs cannot be written after they are finished with @OutDone@.
+\end{itemize}
+
+\begin{code}
+check_machine m
+ = check_state (_init m) (M.singleton (_init m) S.empty)
+ where
+  check_state l acc
+   | Just t <- M.lookup l (_trans m)
+   , Just s <- M.lookup l acc
+   = case t of
+      Pull n l1 l2
+       | not $ Value    n `S.member` s
+       , not $ Finished n `S.member` s
+       ->  go l1 (S.insert (Value n) s) acc
+       >>= go l2 (S.insert (Finished n) s)
+
+      Release n l'
+       | Value n `S.member` s
+       -> go l' (S.delete (Value n) s)
+          acc
+
+      Close   n l'
+       | not $ Value n `S.member` s
+       -> go l' (S.insert (Finished n) s)
+          acc
+
+      Update f l'
+       | all (`S.member` s) (fvs f)
+       -> go l' s acc
+
+      If f l1 l2
+       | all (`S.member` s) (fvs f)
+       -> go l1 s acc >>= go l2 s
+
+      Out f l'
+       | all (`S.member` s) (fvs f)
+       -> go l' (S.insert (Value (_state f)) s) acc
+
+      OutFinished n l'
+       | not $ Finished n `S.member` s
+       -> go l' (S.insert (Finished n) s) acc
+       | otherwise
+
+      Skip l'
+       -> go l' s acc
+
+      Done
+       | all (\n -> Finished n `S.member` s) 
+       $ S.toList
+       $ S.union inputs outputs
+       -> Right acc
+
+  go l s m
+   | Just s' <- M.lookup l m
+   = let sI  = S.intersection s s'
+         sD  = S.difference   s s'
+     -- Allow 
+     in  if         s == s'
+         then       Right m
+         else if    all is_output (S.toList sD)
+         then       check_state l (M.insert l sI m)
+         else       Error
+   | otherwise
+   = check_state l (M.insert l s m)
+\end{code}
 

### 2015-03-17 2015mergingmerges/section/03-Access.tex
add Close state to machines

@@
 Note that the output channel $n$ may have multiple consumers, so continuing to output after the second machine consumer has finished is not as futile as it may seem at first.
-\TODO{Although, all the combinators so far use the whole input, and anyway checking for @Done@ of machine is too fine-grained if machine already has multiple combinators; perhaps an ``I promise not to read from X any more'' state would be interesting.}
+\TODO{Although, all the combinators so far use the whole input, and anyway checking for @Done@ of machine is too fine-grained if machine already has multiple combinators; perhaps an ``I promise not to read from X any more'' state would be interesting.
+Actually, @zip@ doesn't consume the whole input.}
 

### 2015-03-20 2015mergingmerges/section/02-Machines.tex
ok

@@
 
+To perform this invariant checking, we annotate each state with a set of available information at that state.
+The available information can be that there exists some value that can be used, or that a channel is finished or closed.
+For invariant checking, we do not distinguish between finished and closed channels.
+
+\begin{tabbing}
+MM \= M \= \kill
+$\Psi$ \> $=$  \> $@Value@~n$ \\
+       \> $~|$ \> $@Finished@~n$ \\
+\end{tabbing}
+
+To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
+Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions.
+If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
+The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
+
+\input{figures/Invariants.tex}
+
+\begin{tabbing}
+MMMMMM \= M \= MMMMMMMMM \= M \= \kill
+
+$@out@(s, t, a)$
+ \> $~|$ \> $s = @Pull@~n $ \> $ \wedge$ \> $t = @Some@~n$ \\
+ \> $=$  \> $ a \cup \{@Value@~n\} $ \\
+\\
+ \> $~|$ \> $s = @Pull@~n $ \> $ \wedge$ \> $t = @None@$ \\
+ \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
+\\
+ \> $~|$ \> $s = @Release@~n $ \\
+ \> $=$  \> $ a \setminus \{@Value@~n\} $ \\
+\\
+ \> $~|$ \> $s = @Close@~n $ \\
+ \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
+\\
+ \> $~|$ \> $s = @Update@~f~n $ \\
+ \> $=$  \> $ a $ \\
+\\
+ \> $~|$ \> $s = @If@~f~n $ \\
+ \> $=$  \> $ a $ \\
+\\
+ \> $~|$ \> $s = @Out@~f~n $ \\
+ \> $=$  \> $ a \cup \{@Value@~n\} $ \\
+\\
+ \> $~|$ \> $s = @OutFinished@~n $ \\
+ \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
+\\
+ \> $~|$ \> $s = @Skip@ $ \\
+ \> $=$  \> $ a $ \\
+\end{tabbing}
+
+After each state is annotated, we check that its annotation allows the state type 
+For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.
+
+\begin{tabbing}
+MMMMMM \= M \= \kill
+
+$@check@(s, a)$
+ \> $~|$ \> $s = @Pull@~n$ \\
+ \> $=$  \> $ @Value@~n \not\in a \wedge @Finished@~n \not\in a $ \\
+\\
+ \> $~|$ \> $s = @Release@~n$ \\
+ \> $=$  \> $ @Value@~n \in a$ \\
+\\
+ \>      \> $\cdots$
+\end{tabbing}
+
 \begin{code}

### 2015-03-20 2015mergingmerges/section/03-Access.tex
ok

--- 2015-03-16 2015mergingmerges/section/03-Access.tex
@@
 Note that the output channel $n$ may have multiple consumers, so continuing to output after the second machine consumer has finished is not as futile as it may seem at first.
-\TODO{Although, all the combinators so far use the whole input, and anyway checking for @Done@ of machine is too fine-grained if machine already has multiple combinators; perhaps an ``I promise not to read from X any more'' state would be interesting.}
+\TODO{Checking for @Done@ of machine is too coarse-grained, need to check if $n$ is closed instead.}
 

### 2015-03-20 2015mergingmerges/section/02-Machines.tex
remove haskell code of invariant checking, leave just inference rules

@@
 
-\begin{tabbing}
-MMMMMM \= M \= MMMMMMMMM \= M \= \kill
-
-$@out@(s, t, a)$
- \> $~|$ \> $s = @Pull@~n $ \> $ \wedge$ \> $t = @Some@~n$ \\
- \> $=$  \> $ a \cup \{@Value@~n\} $ \\
-\\
- \> $~|$ \> $s = @Pull@~n $ \> $ \wedge$ \> $t = @None@$ \\
- \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
-\\
- \> $~|$ \> $s = @Release@~n $ \\
- \> $=$  \> $ a \setminus \{@Value@~n\} $ \\
-\\
- \> $~|$ \> $s = @Close@~n $ \\
- \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
-\\
- \> $~|$ \> $s = @Update@~f~n $ \\
- \> $=$  \> $ a $ \\
-\\
- \> $~|$ \> $s = @If@~f~n $ \\
- \> $=$  \> $ a $ \\
-\\
- \> $~|$ \> $s = @Out@~f~n $ \\
- \> $=$  \> $ a \cup \{@Value@~n\} $ \\
-\\
- \> $~|$ \> $s = @OutFinished@~n $ \\
- \> $=$  \> $ a \cup \{@Finished@~n\} $ \\
-\\
- \> $~|$ \> $s = @Skip@ $ \\
- \> $=$  \> $ a $ \\
-\end{tabbing}
-
 After each state is annotated, we check that its annotation allows the state type 
 For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.
-
-\begin{tabbing}
-MMMMMM \= M \= \kill
-
-$@check@(s, a)$
- \> $~|$ \> $s = @Pull@~n$ \\
- \> $=$  \> $ @Value@~n \not\in a \wedge @Finished@~n \not\in a $ \\
-\\
- \> $~|$ \> $s = @Release@~n$ \\
- \> $=$  \> $ @Value@~n \in a$ \\
-\\
- \>      \> $\cdots$
-\end{tabbing}
-
-\begin{code}
-check_machine m
- = check_state (_init m) (M.singleton (_init m) S.empty)
- where
-  check_state l acc
-   | Just t <- M.lookup l (_trans m)
-   , Just s <- M.lookup l acc
-   = case t of
-      Pull n l1 l2
-       | not $ Value    n `S.member` s
-       , not $ Finished n `S.member` s
-       ->  go l1 (S.insert (Value n) s) acc
-       >>= go l2 (S.insert (Finished n) s)
-
-      Release n l'
-       | Value n `S.member` s
-       -> go l' (S.delete (Value n) s)
-          acc
-
-      Close   n l'
-       | not $ Value n `S.member` s
-       -> go l' (S.insert (Finished n) s)
-          acc
-
-      Update f l'
-       | all (`S.member` s) (fvs f)
-       -> go l' s acc
-
-      If f l1 l2
-       | all (`S.member` s) (fvs f)
-       -> go l1 s acc >>= go l2 s
-
-      Out f l'
-       | all (`S.member` s) (fvs f)
-       -> go l' (S.insert (Value (_state f)) s) acc
-
-      OutFinished n l'
-       | not $ Finished n `S.member` s
-       -> go l' (S.insert (Finished n) s) acc
-       | otherwise
-
-      Skip l'
-       -> go l' s acc
-
-      Done
-       | all (\n -> Finished n `S.member` s) 
-       $ S.toList
-       $ S.union inputs outputs
-       -> Right acc
-
-  go l s m
-   | Just s' <- M.lookup l m
-   = let sI  = S.intersection s s'
-         sD  = S.difference   s s'
-     -- Allow 
-     in  if         s == s'
-         then       Right m
-         else if    all is_output (S.toList sD)
-         then       check_state l (M.insert l sI m)
-         else       Error
-   | otherwise
-   = check_state l (M.insert l s m)
-\end{code}
 

### 2015-03-26 2015mergingmerges/section/02-Machines.tex
start on merging inference rules
working on invariants

@@
 The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
+See figure~\ref{fig:invariants}.
+
+$$ \CheckOut{\Gamma}{s}{a} $$
+Under environment $\Gamma$ and machine $m$, the state $s$ has available information $a$.
+
+$$ \CheckOutTrans{l}{a} $$
+Under environment $\Gamma$ and machine $m$, the transition from state $s$ to $t$ under label $l$ has available information $a$.
+
+$$ \CheckStateType{s}{T} $$
+In machine $m$, state $s$ has state type $T$.
+
+$$ \Trans{s}{l}{t} \quad \TransU{s}{t}$$
+In machine $m$, there is a transition from $s$ to $t$, optionally labelled $l$.
 

### 2015-03-28 2015mergingmerges/section/02-Machines.tex
slightly cleaner invariants

@@
 The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
-See figure~\ref{fig:invariants}.
+\input{figures/Invariants.tex}
+
+After each state is annotated, we check that its annotation allows the state type 
+For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.
+
+\subsubsection{Generation}
+Figure~\ref{fig:inv:generation}.
+
+Each transition modifies its input's annotation, depending on the state type and the transition label.
+For example, the @Some@ output edge of a @Pull@ adds new information that a @Value@ is available.
 
@@
 
-$$ \Trans{s}{l}{t} \quad \TransU{s}{t}$$
-In machine $m$, there is a transition from $s$ to $t$, optionally labelled $l$.
+\subsubsection{Checking}
+Figure~\ref{fig:inv:checking}.
 
-\input{figures/Invariants.tex}
+We check each state type against its final annotation. For example, @Release@ requires that there exists a @Value@ to release.
 
-After each state is annotated, we check that its annotation allows the state type 
-For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.
+$$ \StateOK $$
+Under environment $\Gamma$ and machine $m$, the state $s$ has a valid annotation.
+
+\subsubsection{Environment closure}
+Figure~\ref{fig:inv:closure}.
+
+We must compute the transitive closure of each state's annotation.
+Given two environments, the final environment and unvetted environment, states are pulled from the unvetted environment and checked against the final environment.
+If the state already exists in the final environment and the annotations are equal, that state has been computed and can be removed from the unvetted environment.
+If the state exists in both environments but they are different, either the only differences are output values, and we proceed with the intersection of the two, or if there are other differences, the machine is invalid.
+Finally, if the state does not exist in the final environment, we add it to the final environment, compute all output transitions of the state, and add them to the unvetted environment.
+
+\TODO{Perhaps this would be described better as a function than an inference rule.}
+The problem, of course, is that state machines are not trees, so a derivation tree for typing does not make much sense.
+
+$$ \EnvGrow{\Gamma}{p}{\Gamma'} $$
+Under environment $\Gamma$ and machine $m$, with $p$ as an unvetted environment to be merged into $\Gamma$, the transitive closure of adding transitions is $\Gamma'$.
+
+\subsubsection{Entire machine}
+Figure~\ref{fig:inv:entire}.
+
+First, the environment closure is computed, starting with the initial state of the machine having no available information.
+Next, every state of the machine is checked to be valid against the environment closure.
+
+$$ m~@ok@ $$
+The entire machine is valid.
 
@@
 Many interesting combinators can be described using these machines.
-In figures 1-4, the machines for some important combinators are given.
+In figures~\ref{fig:com:map}-\ref{fig:com:merge}, the machines for some important combinators are given.
 To reduce clutter, these figures omit labels for @Unit@ transitions.

### 2015-03-29 2015mergingmerges/section/01-Introduction.tex
little things

@@
 
-In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing}\TODO{correct?} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
+In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
 This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.

### 2015-03-29 2015mergingmerges/section/01-Introduction.tex
little things

@@
 Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-For these situations, some sort of assurance of total fusion are required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
 
@@
 Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
-However, we could execute this with no buffering if the relative ordering of the two @mapM_@s was not important, and could be reordered and interspersed as the data is read.
+However, we could execute this with no buffering if the relative ordering of the two @mapM_@s were not important, and could be reordered and interspersed as the data is read.
 We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.
@@
 Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
-It may not be immediately obvious just looking at the program that it requires two iterations, which is why we would like this to be automated.
+It may not be immediately obvious just looking at this program alongside the earlier ones that it requires two iterations, which is why we would like this to be automated.
 

### 2015-03-29 2015mergingmerges/section/02-Machines.tex
little things

@@
 
-We describe a restricted automaton as a eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.
+We describe a restricted automaton as an eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.
 
@@
 
-\[ \delta : \forall q : Q. \sigma(\gamma(q)) \to Q \]
+\[ \delta ~:~\forall q : Q.\ \sigma(\gamma(q))~\to~Q \]
 For each state, the transition function need only be defined for those letters of the state's type.
@@
 \subsection{State types}
-We now describe a set of types and alphabet for describing combinator programs as a restricted DFA.
+We now describe the set of types and alphabet for describing combinator programs as a restricted DFA.
 These are parameterised by both the names of sources, sinks and bindings $n$, and the type of worker functions $f$ (such as @Expression@ for code generation).
@@
 On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
-An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but the explicit pulls makes synchronising two machines easier.
+An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but having explicit releases makes synchronising two machines easier.
 
@@
 To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
-Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions.
+Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions respectively.
 If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
 The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
+These @Out@ values are then implicitly released.
+
+For each state, we check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, functions of @Out@ and @If@ require their values to be present, and so on.
+
 \input{figures/Invariants.tex}
-
-After each state is annotated, we check that its annotation allows the state type 
-For each state, we also check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, and so on.
 
@@
 Under environment $\Gamma$ and machine $m$, the state $s$ has a valid annotation.
+
+$$ \fvs{f} \subset a $$
+Any free variables mentioned in a function must exist as @Value@s in the annotation, in order for the function to use these values.
 
@@
 
-For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the following:
+For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the Figure~\ref{fig:extract:filter}.
+
+\begin{figure}
 \begin{code}
@@
 \end{code}
+\caption{Extracted code for filter}
+\label{fig:extract:filter}
+\end{figure}

### 2015-03-29 2015mergingmerges/section/03-Access.tex
start on merging inference rules
working on invariants
rules for general merge
little things

@@
        \> $~|$ \> @Finished@  \> $n$         \\
+       \> $~|$ \> @Closed@    \> $n$         \\
 \end{tabbing}
 
-The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times l_2 \times \{\Psi\}$.
+The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$.
 It starts by creating an empty machine, and inserting states one by one, starting from the initial state as the initial state of each machine, and empty pending sets.
@@
 @merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
-        \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
+        \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
 \\
@@
 \\
-$@merge@'$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\} \times @Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
-           \> $\to$ \> $@Machine @(l_1 \times l_2 \times \{\Psi\})$ \\
+$@merge@'$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$               \\
+           \> $\times$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
+           \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
 \\
 MMM \= M \= MMMMMMMMMM \= M \=\kill
-$@merge@'~s_1~s_2~\psi~m$ \\
- \> $~|$ \> $(s_1, s_2, \psi) \in m$  \\
+$@merge@'~s_1~\psi_1~s_2~\psi_2~m$ \\
+ \> $~|$ \> $(s_1, \psi_1, s_2, \psi_2) \in m$  \\
  \> $=$  \> $m$ \\
- \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~s_2~\psi$ \\
+ \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~\psi_1~s_2~\psi_2$ \\
  \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
@@
 MMM \= M \= MMMMMMM\kill
-$@move@$ \> $:$ \> $l_1 \times l_2 \times \{\Psi\}$ \\
-           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times l_2 \times \{\Psi\})\}$ \\
+$@move@$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$ \\
+           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})\}$ \\
 \end{tabbing}
@@
 
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
-\\
- \> $~|$ \> $@Update@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
- \> $=$ \> $(@Update@~f~n,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
-\end{tabbing}
+See figure~\ref{fig:merge:general}.
+
+\input{figures/MergeGen.tex}
 
@@
 
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
-\\
- \> $~|$ \> $@Skip@$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@Unit@$ \\
- \> $=$ \> $(@Skip@,~\{(@Unit@, (s_1,s_2',\psi))\})$ \\
-\end{tabbing}
-
 Skips are dealt with in the same way, affecting only one machine.
-
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@True@$ \\
- \> $\wedge$ \> $s_1''$     \> $=$ \> $@trans@~m_1~s_1~@False@$ \\
- \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1',s_2,\psi)), (@False@, (s_1'', s_2, \psi))\})$ \\
-\\
- \> $~|$ \> $@If@~f~n$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $\wedge$ \> $s_2'$     \> $=$ \> $@trans@~m_2~s_2~@True@$ \\
- \> $\wedge$ \> $s_2''$     \> $=$ \> $@trans@~m_2~s_2~@False@$ \\
- \> $=$ \> $(@If@~f~n,~\{(@True@, (s_1,s_2',\psi)), (@False@, (s_1, s_2'', \psi))\})$ \\
-\end{tabbing}
 
@@
 
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Done@$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $=$ \> $(@Done@,~\{\})$ \\
-\end{tabbing}
-
 If both machines are @Done@, the resulting machine will also be @Done@.
@@
 
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \setminus \{@Value@~n\} ))\})$ \\
-\end{tabbing}
-
 If the first machine is producing an output on channel $n$ and the second machine uses $n$, the first machine would usually have to wait until the second machine is ready to accept new data.
-However, if the second machine is @Done@, the first machine may output to the channel as frequently as it likes.
-Note that the output channel $n$ may have multiple consumers, so continuing to output after the second machine consumer has finished is not as futile as it may seem at first.
-\TODO{Checking for @Done@ of machine is too coarse-grained, need to check if $n$ is closed instead.}
-
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Value@~n\} ))\})$ \\
-\end{tabbing}
+However, if the second machine has closed $n$, the will never pull on $n$ again, so the first machine may output to the channel as frequently as it likes.
+Note that the output channels may have multiple consumers, so continuing to output after a consumer has finished is not as futile as it may seem at first.
 
 Again, if the first machine is producing and the second is consuming, the first machine may only produce if there is no unhandled value on the channel.
-This is what $\psi$ is for.
-If there is a value in $\psi$, a later case may allow the second machine to pull from it.
-
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@Out@~f~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@Out@~f~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
-\end{tabbing}
+This is what $\psi$ and $\phi$ are for.
+If there is a value in the other machine's event set, $\phi$, a later case may allow the second machine to pull from it.
 
 The final case for @Out@ is when the second machine does not use this output, allowing the first machine to output at any time.
-The cases where the second machine is on an @Out@ are symmetric to these.
 
-The cases for @OutDone@ proceed similarly, and are also symmetric for the first and second machines.
-
-\begin{tabbing}
-MMM \= M \= MMMMMMM \= M \=\kill
- \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $@Done@$ \> $=$ \> $@state@~m_2~s_2$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
- \\
- \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $@Value@~n$ \> $\not\in$ \> $\psi$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi \cup \{@Finished@~n\}))\})$ \\
- \\
- \> $~|$ \> $@OutDone@~n$ \> $=$ \> $@state@~m_1~s_1$ \\
- \> $\wedge$ \> $n$      \> $\not\in$ \> $@inputs@~m_2$ \\
- \> $\wedge$ \> $s_1'$     \> $=$ \> $@trans@~m_1~s_1~@Unit@$ \\
- \> $=$ \> $(@OutDone@~n,~\{(@Unit@, (s_1',s_2,\psi))\})$ \\
-\end{tabbing}
+The cases for @OutDone@ proceed similarly to @Out@.
 
@@
 \end{code}
+
+\subsection{Theorems we want to prove}
+\subsubsection{General preservation}
+$$
+\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @merge@~a~b = c \implies c~@ok@
+$$
+
+\subsubsection{Vertical preservation}
+$$
+\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @mergeV@~a~b = c \implies c~@ok@
+$$
+
+\subsubsection{Vertical progress}
+\begin{tabbing}
+MM \= M \= \kill
+$ \forall a~b.\ $ \> \>
+$a~@ok@~\wedge~b~@ok@$ \\
+\> $\wedge$ \> $@inputs@~a~\cap~@inputs@~b=\emptyset~\wedge~@outputs@~a~\cap~@outputs@~b=\emptyset$ \\
+\> $\wedge~($ \> $\exists n.\ (@inputs@~a~\cap~@outputs@~b)~\cup~(@outputs@~a~\cap~@inputs@~b) = \sgl{n})$ \\
+ $\implies$ \> \> $\exists c.\ @mergeV@~a~b=c$
+\end{tabbing}

### 2015-03-31 2015mergingmerges/section/03-Merging.tex
ok

--- 2015-03-29 2015mergingmerges/section/03-Access.tex
@@
 
-See figure~\ref{fig:merge:general}.
-
 \input{figures/MergeGen.tex}
@@
 \subsection{Theorems we want to prove}
+
+As we do not wish to handle recursive combinators, we also do not need to handle merging cyclic machines.
+This is one case that regular dataflow handles but we do not.
+The predicate below checks if either of the machines does not use the other's input; if both used the other's input, it would create a cycle.
+
+\begin{tabbing}
+MMMMMMM \= MM \= \kill
+$@acyclic@~a~b$
+\> $:=$
+\> $\outputs{a}~\cap~\inputs{b}~=~\emptyset$
+\\
+\> $~\vee$ \> $\outputs{b}~\cap~\inputs{a}~=~\emptyset$
+\end{tabbing}
+
+Two machines cannot be merged if they both provide the same output, or write to the same output channel.
+
+\begin{tabbing}
+MMMMMMM \= MM \= \kill
+$@distinct@~a~b$
+\> $:=$
+\> $\outputs{a}~\cap~\outputs{b}~=\emptyset$
+\end{tabbing}
+
+For the special case of vertical fusion, we require there be one output used as the other machine's sole input, and no other sharing.
+Both machines may have multiple outputs, and the producer may have multiple inputs, but the consumer can only have one input.
+
+\begin{tabbing}
+MMMMMMM \= MM \= MM \= \kill
+$@vertical@~a~b$
+\> $:=$
+\> $\exists n.$
+\> $(n \in \outputs{a} \wedge \sgl{n} = \inputs{b})$
+\\
+\>
+\> $~\vee$
+\> $(n \in \outputs{b} \wedge \sgl{n} = \inputs{a})$
+\end{tabbing}
+
 \subsubsection{General preservation}
-$$
-\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @merge@~a~b = c \implies c~@ok@
-$$
+The result of merging two machines should maintain the pull invariants (\S~\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
+
+\begin{tabbing}
+MMMMMMM \= MM \= MM \= \kill
+$\forall a~b.$
+\>
+\> $a~@ok@~\wedge~b~@ok@$
+\\
+\> $\wedge$
+\> $@acyclic@~a~b$
+\\
+\> $\wedge$
+\> $@distinct@~a~b$
+\\
+\> $\implies$
+\> $\forall c \in (@merge@~a~b).\ c~@ok@$
+\end{tabbing}
+
+\paragraph{Lemma transitions:} for each state $u$ of $@merge@~a~b$ and corresponding states $s$ and $t$ of $a$ and $b$, any output transitions from $u$ either move along $s$'s output transitions or $t$'s output transitions, but not both.
+Similarly, the state type of $u$ will either be the type of $s$, the type of $t$, or a @Skip@.
+
+\paragraph{Lemma appears:} for all states $s$ in $a$, if there is a path from $@initial@~a$ to $s$, then:
+\begin{itemize}
+\item if $s$ is a @Close@, either $s$ or a @Finish@ or a @Close@ from the other machine $b$ will appear in the output ($@merge@~a~b$);
+\item if $s$ is an @Out@, @Update@, @If@, $s$ will appear in the output;
+\item otherwise, $s$ 
+\end{itemize}
+ACTUALLY, \textbf{appears} isn't true.
+
+\paragraph{Proof sketch:} each state in $a$ and $b$ has an associated set of $\Psi$.
+If we deconstruct each state of $c$, it has a state of $a$, a state of $b$, and a $\Psi$ set for each.
+
+Show that in this $(\sptpC)$, the $\psi$ and $\phi$ is something similar to $s$ and $t$'s $\Psi$ set.
+
+Total $\Psi$ is the union of $\psi$ and $\phi$, except if one set has @Closed@ but the other doesn't, that channel is still open.
+
+Then, look at each rule in @move@, and the preconditions should translate to sufficient preconditions on $\Psi$:
+
+@Skip@, @Update@ and @If@ are trivial.
+
+If both machines are @Done@, since both machines are also @ok@, all their inputs and outputs must be closed..
 
 \subsubsection{Vertical preservation}
-$$
-\forall a~b.\ a~@ok@~\wedge~b~@ok@~\wedge \forall c.\ @mergeV@~a~b = c \implies c~@ok@
-$$
+
+\begin{tabbing}
+MMMMMMM \= MM \= MM \= \kill
+$\forall a~b.$
+\>
+\> $a~@ok@~\wedge~b~@ok@$
+\\
+\> $\wedge$
+\> $@acyclic@~a~b$
+\\
+\> $\wedge$
+\> $@vertical@~a~b$
+\\
+\> $\wedge$
+\> $@distinct@~a~b$
+\\
+\> $\implies$
+\> $\forall c \in (@mergeV@~a~b).\ c~@ok@$
+\end{tabbing}
 
 \subsubsection{Vertical progress}
+
 \begin{tabbing}
-MM \= M \= \kill
-$ \forall a~b.\ $ \> \>
-$a~@ok@~\wedge~b~@ok@$ \\
-\> $\wedge$ \> $@inputs@~a~\cap~@inputs@~b=\emptyset~\wedge~@outputs@~a~\cap~@outputs@~b=\emptyset$ \\
-\> $\wedge~($ \> $\exists n.\ (@inputs@~a~\cap~@outputs@~b)~\cup~(@outputs@~a~\cap~@inputs@~b) = \sgl{n})$ \\
- $\implies$ \> \> $\exists c.\ @mergeV@~a~b=c$
+MMMMMMM \= MM \= MM \= \kill
+$\forall a~b.$
+\>
+\> $a~@ok@~\wedge~b~@ok@$
+\\
+\> $\wedge$
+\> $@acyclic@~a~b$
+\\
+\> $\wedge$
+\> $@vertical@~a~b$
+\\
+\> $\wedge$
+\> $@distinct@~a~b$
+\\
+\> $\implies$
+\> $\exists c.\ c \in @mergeV@~a~b$
 \end{tabbing}

### 2015-04-01 2015mergingmerges/section/03-Merging.tex
add example

@@
 
-The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Pull@, @Release@, @Out@ and @OutDone@ are more complicated.
+The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Pull@, @Release@, @Out@ and @OutFinished@ are more complicated.
 
@@
 
-The cases for @OutDone@ proceed similarly to @Out@.
+The cases for @OutFinished@ proceed similarly to @Out@.
 
@@
 In which case, we give a compile time error indicating that these two machines cannot be fused.
+
+\subsubsection{Example}
+Let us merge a @map@ (figure~\ref{fig:com:map}) and a @filter@ (figure~\ref{fig:com:filter}).
+The actual program would be
+\begin{code}
+fun xs
+ = let m  = map    (+1) xs
+       f  = filter (>0) xs
+   in (m, f)
+\end{code}
+
+We need to name each state.
+While the graphical form is easier to understand, a textual form makes looking up a particular state's transitions easier.
+The @map@ state types and transitions are below, with initial state being @m1@.
+
+\begin{code}
+m1 : Pull xs     { Some => m2; None => m4 }
+m2 : Out m (x+1) { Unit => m3 }
+m3 : Release x   { Unit => m1 }
+m4 : OutFinished m{ Unit => m5 }
+m5 : Done
+\end{code}
+
+Likewise, the filter has initial state @f1@.
+
+\begin{code}
+f1 : Pull xs     { Some => f2; None  => f5 }
+f2 : If (x>0)    { True => f3; False => f4 }
+f3 : Out f x     { Unit => f4 }
+f4 : Release x   { Unit => f1 }
+f5 : OutFinished f{ Unit => f6 }
+f6 : Done
+\end{code}
+
+We start by computing $@merge@'$ of an empty machine, empty $\psi$ sets, and initial states of each machine:
+$$
+@merge@'~@m1@~\emptyset~@f1@~\emptyset~@empty@
+$$
+
+In this case, there are two possible moves: either @m@ may take a step, or @f@ may take a step, but both steps are @Pull xs@.
+When two moves are possible, we will take the @m@ one, but this decision has no effect on the outcome \TODO{(prove it)}.
+\begin{code}
+move m f m1 {} f1 {} =
+    Pull xs,
+    { Some => m2, {Value xs},    f1, {Value xs}
+    ; None => m4, {Finished xs}, f1, {Finished xs} }
+(Shared pull, no pending values)
+\end{code}
+
+We insert this state and transitions into the empty machine, noting that the transitions to undefined states will be defined by later moves.
+
+\begin{code}
+m1,{},f1,{}
+    : Pull xs
+    { Some => m2, {Value xs},    f1, {Value xs}
+    ; None => m4, {Finished xs}, f1, {Finished xs} }
+\end{code}
+
+Now, calculate the first undefined move.
+Again, either machine may move: @m@ could compute its output, or @f@ could skip past the @Pull xs@, since this has been performed by @m@, witnessed by the @Value xs@ in @f@'s $\psi$ set.
+
+\begin{code}
+move m f m2 {Value xs} f1 {Value xs} =
+    Out m (x+1)
+    { Unit => m3, {Value xs}, f1, {Value xs} }
+(Local output, free to write)
+\end{code}
+
+This puts @m@ at a @Release@, which is changed to a skip as the later release by @f@ will be the real release.
+
+\begin{code}
+move m f m3 {Value xs} f1 {Value xs} =
+    Skip
+    { Unit => m1, {}, f1, {Value xs} }
+(Shared release: first release is skip)
+\end{code}
+
+Now machine @m@ is back at a @Pull xs@, but it is unable to execute the pull itself as the other machine @f@ still has a pending @Value xs@ to deal with.
+The machine @f@ is also at a @Pull xs@, and is able to skip past since it already has a value.
+
+\begin{code}
+move m f m1 {} f1 {Value xs} =
+    Skip
+    { Unit => m1, {}, f2, {Value xs} }
+(Pull has value)
+\end{code}
+
+Now @f@ is at an @If@, and @m@ still cannot run.
+
+\begin{code}
+move m f m1 {} f2 {Value xs} =
+    If (x>0)
+    { True  => m1, {}, f3, {Value xs}
+    , False => m1, {}, f4, {Value xs} }
+(If)
+\end{code}
+
+The first case of the @If@ is a local output. 
+\begin{code}
+move m f m1 {} f3 {Value xs} =
+    Out f x
+    { Unit => m1, {}, f4, {Value xs} }
+(Local output, free to write)
+\end{code}
+
+Next is a release.
+In this case, the input is shared and the other machine has already released, since there is no @Value xs@ in its set.
+We emit an actual release in this case, since we are sure both machines are finished using it.
+
+\begin{code}
+move m f m1 {} f4 {Value xs} =
+    Release x
+    { Unit => m1, {}, f1, {} }
+(Shared release: other machine already released)
+\end{code}
+
+Now the only missing state is @m4, {Finished xs}, f1, {Finished xs}@ after the @m@'s pull is empty.
+
+\begin{code}
+move m f m4 {Finished xs} f1 {Finished xs} =
+    OutFinished m
+    { Unit => m5, {Finished xs}, f1, {Finished xs} }
+(Finish local output)
+\end{code}
+
+Now @m@ is @Done@, and cannot move, while @f@ is at a @Pull xs@ that it already knows is finished.
+\begin{code}
+move m f m5 {Finished xs} f1 {Finished xs} =
+    Skip
+    { Unit => m5, {Finished xs}, f5, {Finished xs} }
+(Pull of finished)
+\end{code}
+
+Now @m@ is @Done@, and cannot move, while @f@ is at a @Pull xs@ that it already knows is finished.
+\begin{code}
+move m f m5 {Finished xs} f5 {Finished xs} =
+    OutFinished f
+    { Unit => m5, {Finished xs}, f6, {Finished xs} }
+(Finish local output)
+\end{code}
+
+Both machines are now done.
+\begin{code}
+move m f m5 {Finished xs} f6 {Finished xs} =
+    Done
+(Both machines done)
+\end{code}
+
+The entire machine at the end is:
+\begin{code}
+m1,{},f1,{}
+    : Pull xs
+    { Some => m2, {Value xs},    f1, {Value xs}
+    ; None => m4, {Finished xs}, f1, {Finished xs} }
+m2,{Value xs},f1,{Value xs}
+    : Out m (x+1)
+    { Unit => m3, {Value xs},    f1, {Value xs}    }
+m3,{Value xs},f1,{Value xs}
+    : Skip
+    { Unit => m1, {},            f1, {Value xs}    }
+m1,{},f1,{Value xs}
+    : Skip
+    { Unit => m1, {},            f2, {Value xs}    }
+m1,{},f2,{Value xs}
+    : If (x>0)
+    { True => m1, {},            f3, {Value xs}
+    , False=> m1, {},            f4, {Value xs}    }
+m1,{},f3,{Value xs}
+    : Out f x
+    { Unit => m1, {},            f4, {Value xs}    }
+m1,{},f4,{Value xs}
+    : Release x
+    { Unit => m1, {},            f1, {}            }
+m4,{Finished xs},f1,{Finished xs}
+    : OutFinished m
+    { Unit => m5, {Finished xs}, f1, {Finished xs} }
+m5,{Finished xs},f1,{Finished xs}
+    : Skip
+    { Unit => m5, {Finished xs}, f5, {Finished xs} }
+m5,{Finished xs},f5,{Finished xs}
+    : OutFinished f
+    { Unit => m5, {Finished xs}, f6, {Finished xs} }
+m5,{Finished xs},f6,{Finished xs}
+    : Done
+\end{code}
+
+The @Skip@s can then be removed quite easily, leading to the graph below.
+
+\begin{dot2tex}[scale=0.5]
+digraph {
+    rankdir=LR;
+    mindist=0.1
+    nodesep=0.1;
+    ranksep=0.1;
+    node [shape="circle", margin=0];
+    10 [label="Pull xs"];
+    20 [label="Out m (x+1)"];
+    50 [label="If"];
+    60 [label="Out f x"];
+    70 [label="Release x"];
+    900[label="OutFinished m"];
+    920[label="OutFinished f"];
+    930[label="Done"];
+
+    10 -> 20 [label="Some x"];
+    10 -> 900 [label="None"];
+
+    50 -> 60 [label="True"];
+    50 -> 70 [label="False"];
+
+    20 -> 50;
+    60 -> 70;
+    70 -> 10;
+    900 -> 920;
+    920 -> 930;
+}
+\end{dot2tex}
 

### 2015-04-01 2015mergingmerges/section/03-Merging.tex
ok

@@
 
-The other cases are when one machine produces a value, and another consumes it.
-Here, the producer may have produced a value, or may have finished producing on that channel.
+Shared outputs, when one machine produces a value and the other consumes it.
+Here, events are used to indicate that the producer has written a value, and cannot write another until the consumer has dealt with it, or that the producer has finished and the consumer's next read will skip to the @None@ case.
+
+Finally, if one machine closes a shared input but the other is still reading, we do not close the input but instead treat the input as a local input, allowing the reading machine to continue reading at any rate.
 
@@
 
+\subsubsection{Non-interfering states}
 If either of the machines are attempting to update their local state, this can be done easily without interfering with the other machine.
@@
 
-The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Pull@, @Release@, @Out@ and @OutFinished@ are more complicated.
+\subsubsection{Output and finishing}
+The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Out@, @OutFinished@, @Pull@, @Release@ and @Close@ are more complicated.
 
 If the first machine is producing an output on channel $n$ and the second machine uses $n$, the first machine would usually have to wait until the second machine is ready to accept new data.
-However, if the second machine has closed $n$, the will never pull on $n$ again, so the first machine may output to the channel as frequently as it likes.
+However, if the second machine has closed $n$, it will never pull on $n$ again, so the first machine may output to the channel as frequently as it likes.
 Note that the output channels may have multiple consumers, so continuing to output after a consumer has finished is not as futile as it may seem at first.
@@
 
-Pulls now.
+\subsubsection{Pull}
+When the channel is a local input, @Pull@ing proceeds as normal.
+Likewise with a shared input that has been closed by the other machine - the other machine will no longer attempt to pull, so no synchronisation is required.
 
-Releases.
+If there is a @Value@ in the current machine's event set, the channel must either be a shared input, or an output channel on the other machine.
+If it is a shared input, the presence of a @Value@ means the other machine has already executed a @Pull@ and has succeeded.
+Otherwise, for an output channel, the @Value@ means the other machine has produced a value which must be used before another can be produced.
+In either case, we simply proceed to the @Some@ branch of the @Pull@, safe in the knowledge that there is a value to use.
+
+If there is a @Finished@ in either machine's event set, and neither machine has a pending @Value@s to deal with, we know that the other machine has had either an @OutFinished@ or an empty @Pull@.
+We proceed to the @None@ case of the @Pull@.
+
+Finally, if there are no pending @Value@ or @Finished@ events in either machine and it is a shared input, this machine can safely pull on the input, adding pending events for the other machine to deal with.
+
+\subsubsection{Release}
+Pulled values must be released.
+This is used for synchronisation, so that one machine cannot pull on a shared input until after both machines have released the value.
+Otherwise, one machine could pull while the other machine is still using the value, requiring a larger buffer.
+
+For local channels, the release is performed as usual.
+
+If the channel is an output of the other machine, the release is emitted as a skip, but the @Value@ is removed from the event set, which will allow the other machine's @Out@ to proceed.
+
+For shared inputs, there will be two releases of the one value, but only the second should be emitted as an actual release.
+
+For the first release, both machines will have a @Value@ event, and we emit a skip, removing this machine's @Value@.
+
+For the second release, the other machine has already removed its @Value@ event, so we remove this machine's @Value@ and emit a real release.
+At this stage, neither machine will have a @Value@ event, which will allow another @Pull@ to execute.
+
+\subsubsection{Close}
+For @Close@, if the channel is neither an input nor an output of the other machine, then closing the channel does not affect the other machine and acts as a normal @Close@.
+
+If the channel is an output of the other machine, we skip past.
+Because the channel may have other consumers, we do not stop it.
+So that further outputs on this channel do not wait on this machine, we add a @Closed@ event to $\psi$.
 

### 2015-04-03 2015mergingmerges/section/01-Introduction.tex
intro

@@
 \begin{code}
-zip1 :: Source Int -> Source Int -> IO ()
+zip1 :: [Int] -> [Int] -> IO ()
 zip1 xs ys
@@
        zz   = zip xs' ys'
-   in  mapM_ fire_missile zz
+   in  mapM_ print        zz
 \end{code}
@@
 \begin{code}
-zip2 :: Source Int -> Source Int -> IO ()
+zip2 :: [Int] -> [Int] -> IO ()
 zip2 xs ys
@@
        zz   = zip xs' ys'
-   in  do   mapM_ fire_missile zz
-            mapM_ fortify      xs2
+   in  do   mapM_ print   zz
+            mapM_ print   xs2
 \end{code}
 
-Sadly, this program requires buffering the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
+If we were to execute this program either strictly or lazily, we must buffer the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
 However, we could execute this with no buffering if the relative ordering of the two @mapM_@s were not important, and could be reordered and interspersed as the data is read.
-We use the following @When@ syntax to denote performing a reorderable IO action \emph{when}ever data is available.
+We change the input lists to @Source@s and the @mapM_@s to @Perform@s, to denote that the IO actions are reorderable and to execute whenever data is available, and use @:&@ to join performs together.
 
 \begin{code}
-zip3 :: Source Int -> Source Int -> IO ()
+zip3 :: Source Int -> Source Int -> Perform
 zip3 xs ys
@@
        zz   = zip xs' ys'
-   in  When fire_missile zz
-   :&  When fortify      xs2
+   in  Perform print      zz
+   :&  Perform print      xs2
 \end{code}
@@
 \begin{code}
-zip4 :: Source Int -> IO ()
+zip4 :: [Int] -> IO ()
 zip4 xs
@@
        zz   = zip xs' xs''
-   in  mapM_ fire_missile zz
+   in  mapM_ print        zz
 \end{code}
 
-Sadly this last program requires an unbounded buffer or two iterations over @xs@, so we would like to disallow it.
-It may not be immediately obvious just looking at this program alongside the earlier ones that it requires two iterations, which is why we would like this to be automated.
+Sadly this last program requires an unbounded buffer or two iterations over @xs@, as we do not know anything about the contents of @xs@.
+Suppose that the entire list were made of a hundred positive numbers, followed by a hundred negative numbers: the first positive number would need to be kept around until the first negative number was reached, and likewise for the second and subsequent positive numbers -- so all one hundred positive numbers would need to be stored in memory until the negative numbers were reached.
+Now consider if, instead of a hundred positive numbers, there were a thousand: now it is necessary to buffer the thousand numbers in memory instead.
+The important part here is that one cannot compute, just by looking at the program, an adequate buffer size that works for all inputs.
+We wish to disallow this program and others like it, where we cannot ensure execution with a bounded buffer.
 

### 2015-04-03 2015mergingmerges/section/02-Machines.tex
fix release
ok
machines

@@
 
-We use a particular type of deterministic finite automata (DFAs) where each state has a type that restricts the alphabet of output transitions.
-We use state types such as @Pull a@ with output transitions @Some a@ and @None@, or type @Out b@ with single output transition @Unit@.
+A machine is a deterministic finite automata (DFA) that describes a streaming computation of a group of combinators.
+We use a particular type of DFA where each state has a type that restricts the allowed output transitions, meaning that unnecessary edges are omitted.
+A state type such as @Pull a@ with output transitions @Some a@ and @None@ denotes reading from an input source named @a@, and a state type of @Out (a+1) b@ with single output transition @Unit@ for writing @a+1@ to the output channel @b@.
 
-\subsection{Restricted DFA}
-
-We describe a restricted automaton as an eight-tuple, extending the standard description by three elements: $(Q, \Sigma, \delta, q_0, F, \Gamma, \gamma, \sigma)$.
-
-As usual, $Q$ is the set of states, $\Sigma$ the alphabet, but in addition $\Gamma$ is a set of state types.
-We define the types of functions as follows:
-
-\[ \gamma : Q \to \Gamma \]
-Every state has a type.
-
-\[ \sigma : \Gamma \to \{ \Sigma \} \]
-Every type has some subset of the alphabet that transitions are required for.
-
-\[ \delta ~:~\forall q : Q.\ \sigma(\gamma(q))~\to~Q \]
-For each state, the transition function need only be defined for those letters of the state's type.
-
-These restricted DFAs can be converted to normal automata by adding a rejecting sink state, and adding transitions for all missing letters of the alphabet to this rejecting state.
-
-\subsection{State types}
-We now describe the set of types and alphabet for describing combinator programs as a restricted DFA.
+\subsection{State types and transitions}
+Each state has an associated state type.
 These are parameterised by both the names of sources, sinks and bindings $n$, and the type of worker functions $f$ (such as @Expression@ for code generation).
+The state types are as follows:
 
 \begin{tabbing}
-MM \= MM \= MMMMMM \= M \= M \kill
-$\Gamma$ \> $=$ \> @Pull@     \>     \> $n$         \\
+MMMM \= MM \= MMMMMM \= M \= M \kill
+$@Type@$ \> $=$ \> @Pull@     \>     \> $n$         \\
          \>     \> (Attempt to read from a named input) \\
@@
 
-Each of these state types requires a corresponding set of output transitions to be defined.
-\begin{tabbing}
-MMMMMMMM \= M \= \kill
-$\gamma(@Pull @n)$
-                            \> $=$ \> \{@Some@ n, @None@\} \\
-$\gamma(@Release @n)$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@Close @n)$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@Out @   f~n)$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@OutDone @ n)$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@If @    f~n)$
-                            \> $=$ \> \{@True@, @False@\} \\
-$\gamma(@Update @f~n)$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@Skip@      )$
-                            \> $=$ \> \{@Unit@\} \\
-$\gamma(@Done@      )$
-                            \> $=$ \> \{\} \\
-\end{tabbing}
-
 Note that @Out@, @OutDone@, @If@ and @Update@ are also annotated with the name of the output channel.
@@
 These machine state types must be annotated with the name of the output channel to designate which local mutable state, if any.
+During code generation, each local mutable state will be associated with a separate variable.
 
-In the proceeding parts, any reference to @Machine@ is a restricted DFA with these state types.
+Some of these state types could be removed without affecting the expresivity or code generation of machines but are used in the merging algorithm, such as @Release@ which is used for synchronising shared reads.
+
+Each of these state types requires a corresponding set of output transitions, or edges to other states, to be defined.
+Most state types perform no flow control, and so require a single output transition denoting the state to be executed afterwards, like a semicolon in imperative code.
+The states @Pull@ or @If@ need to perform different actions depending on whether the input is empty or not, or whether the predicate is true, so require multiple output transitions.
+No further execution is possible after the @Done@ state, so it has no output transitions.
+\begin{tabbing}
+MMMMMMMMMMMM \= M \= \kill
+$@StateOut@(@Pull @n)$
+                            \> $=$ \> \{@Some@ n, @None@\} \\
+$@StateOut@(@Release @n)$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@Close @n)$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@Out @   f~n)$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@OutDone @ n)$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@If @    f~n)$
+                            \> $=$ \> \{@True@, @False@\} \\
+$@StateOut@(@Update @f~n)$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@Skip@      )$
+                            \> $=$ \> \{@Unit@\} \\
+$@StateOut@(@Done@      )$
+                            \> $=$ \> \{\} \\
+\end{tabbing}
+
+In the proceeding parts, we use $@Machine@~l$ to denote a machine with states labelled by $l$, where each state has a @Type@, and transitions as defined by the @StateOut@ of its type.
+
+\subsection{Combinators}
+As an example of @filter p xs@ combinator expressed as a machine, see figure~\ref{fig:com:filter}.
+Here, the initial state is a @Pull xs@, an attempt to read from the input stream @xs@.
+If the read is successful, we take the @Some xs@ transition leading to $@If@~(p~xs_e)$ - the predicate applied to the currently read element of @xs@.
+If the predicate is true, we execute $@Out@~o~xs_e$, writing the current element of @xs@ to the output channel named @o@; otherwise, we release the element of @xs@ and go back to the @Pull@.
+Again, this release is not strictly necessary for expressivity or code generation, but is essential for synchronisation during merging.
+Finally, if the @Pull xs@ fails and there is no data left to read, we signal that the output stream has finished, and then finish the entire machine.
+
+Many interesting combinators can be described using these machines.
+In figures~\ref{fig:com:map}-\ref{fig:com:merge}, the machines for some important combinators are given.
+To reduce clutter, these figures omit labels for @Unit@ transitions.
+
+While writing combinators in this form is neither pleasant nor pretty, it only needs to be done once.
+The implementation contains more combinators such as group by, append and indices of segment lengths.
+
+\input{figures/Combinators.tex}
 
 \subsection{Invariants}
+\label{s:Machines:Invariants}
 Not all machines are valid or meaningful, and we wish to rule them out.
-For example, a machine with an initial state of $@If@~(x>5)~n$ is meaningless because there is no local $x$ that has been read from an input. 
+For example, a machine with an initial state of $@If@~(xs_e~>~5)~n$ is meaningless because there is no element of $xs$ that has been read from an input. 
 On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
@@
 To perform this invariant checking, we annotate each state with a set of available information at that state.
-The available information can be that there exists some value that can be used, or that a channel is finished or closed.
-For invariant checking, we do not distinguish between finished and closed channels.
+The DFA machines simplify the merging algorithm, but lose binding and scope information which would be implicit in a structured language.
+We need to reconstruct and check the scope information, similar to a type state system\CITE.
+
+The available event information can be that there exists some value that can be used, or that a channel is finished or closed.
 
 \begin{tabbing}
-MM \= M \= \kill
-$\Psi$ \> $=$  \> $@Value@~n$ \\
+MMMM \= M \= \kill
+@Event@ \> $=$  \> $@Value@~n$ \\
        \> $~|$ \> $@Finished@~n$ \\
+       \> $~|$ \> $@Closed@~n$ \\
 \end{tabbing}
@@
 
-\subsection{Combinators}
-Many interesting combinators can be described using these machines.
-In figures~\ref{fig:com:map}-\ref{fig:com:merge}, the machines for some important combinators are given.
-To reduce clutter, these figures omit labels for @Unit@ transitions.
-
-While writing combinators in this form is neither pleasant nor pretty, it only needs to be done once.
-The implementation contains more combinators such as group by, append and indices of segment lengths.
-
-\input{figures/Combinators.tex}
-
 \subsection{Code extraction}

### 2015-04-04 2015mergingmerges/section/01-Introduction.tex
contributions

@@
 
-In summary, the purpose of this work is to determine which programs can be executed with no buffering, and offer an efficient compilation strategy for these.
+In this paper, we present:
+\begin{itemize}
+\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
+\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
+\item invariants required of such machines (\S\ref{s:Machines:Invariants});
+\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
+\end{itemize}

### 2015-04-04 2015mergingmerges/section/03-Merging.tex
remove vertical special case

@@
 
-\subsection{Producer-consumer}
-The general case above is sufficient, but in some cases can generate larger machines than necessary.
-Talk about special case for producer-consumers.
-
-\begin{code}
-data Which = LeftExec | RightExec
-
-mergeV :: Machine l1 -> Machine l2
-       -> Machine (l1, l2, Which)
-\end{code}
-
 \subsection{Theorems we want to prove}
@@
 
-For the special case of vertical fusion, we require there be one output used as the other machine's sole input, and no other sharing.
-Both machines may have multiple outputs, and the producer may have multiple inputs, but the consumer can only have one input.
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$@vertical@~a~b$
-\> $:=$
-\> $\exists n.$
-\> $(n \in \outputs{a} \wedge \sgl{n} = \inputs{b})$
-\\
-\>
-\> $~\vee$
-\> $(n \in \outputs{b} \wedge \sgl{n} = \inputs{a})$
-\end{tabbing}
-
-\subsubsection{General preservation}
+\subsubsection{Preservation}
 The result of merging two machines should maintain the pull invariants (\S~\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
@@
 If both machines are @Done@, since both machines are also @ok@, all their inputs and outputs must be closed..
-
-\subsubsection{Vertical preservation}
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.$
-\>
-\> $a~@ok@~\wedge~b~@ok@$
-\\
-\> $\wedge$
-\> $@acyclic@~a~b$
-\\
-\> $\wedge$
-\> $@vertical@~a~b$
-\\
-\> $\wedge$
-\> $@distinct@~a~b$
-\\
-\> $\implies$
-\> $\forall c \in (@mergeV@~a~b).\ c~@ok@$
-\end{tabbing}
-
-\subsubsection{Vertical progress}
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.$
-\>
-\> $a~@ok@~\wedge~b~@ok@$
-\\
-\> $\wedge$
-\> $@acyclic@~a~b$
-\\
-\> $\wedge$
-\> $@vertical@~a~b$
-\\
-\> $\wedge$
-\> $@distinct@~a~b$
-\\
-\> $\implies$
-\> $\exists c.\ c \in @mergeV@~a~b$
-\end{tabbing}

### 2015-04-04 2015mergingmerges/section/03-Merging.tex
proof attempt

@@
 \subsubsection{Preservation}
-The result of merging two machines should maintain the pull invariants (\S~\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
+The result of merging two machines should maintain the pull invariants (\S\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
 
@@
 
+\paragraph{Lemma values:} 
+for each state $u$ of $@merge@~a~b$, if either $\psi$ or $\phi$ contain a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
+(Note that the converse is not true, as @Value n@ will appear in the invariant set, but not $\psi$ or $\phi$ for local @Out@s)
+
+We proceed by induction over the @move@ rules.
+
+The only way that @Value n@ can be added to $\psi$ or $\phi$ is through the rules (Shared output, ready to read) and (Shared pull, no pending values).
+These rules themselves produce state types of @Out@ and @Pull@ respectively, and by the definition of invariant sets @Value@ is inserted for these transitions.
+
+For rule (Local release), the state type is @Release n@ which removes @Value n@ from the invariant set.
+Due to the locality preconditions $n~\not\in~@inputs@~m_2$ and $n~\not\in~@outputs@~m_2$, we know that neither of the rules to add @n@ could apply, so $\psi$ and $\phi$ must not contain @Value n@.
+
+For the cases that remove a @Value n@ from the $\psi$ set, the first two, (Release of output) and (Shared release: first release is skip) both only emit skips, so do not affect the invariant set.
+This means that while the invariant set may still contain a @Value n@, it is not obliged to.
+We appeal to the induction hypothesis for the input transitions, since earlier input transitions must have added the @Value n@ to both sets.
+
+The last case that removes a @Value n@ from the $\psi$ set is (Shared release: other machine already released).
+Here, we know $\psi$ contains a @Value n@ and that $\phi$ does not.
+This means that after removing the @Value n@ from $\psi$, neither set will have one.
+The produced state type also means that the invariant set will no longer have a @Value n@.
+
+For the remaining cases, the @Value@s inside the $\psi$ and $\phi$ sets are left alone, and their state types do not explicitly remove any @Value n@s from the invariant set either.
+We can conclude that any @Value@s in the sets have been added by other states, and not removed.
+The last thing is that @Value@s can be implicitly removed at join-points of states.
+However, in order for a state to have @Value@s in its sets, all transitions into this state must have either inserted @Value@s itself, or kept @Value@s from its predecessors in turn; otherwise it would not have a transition to a state with @Value@s in its sets.
+
+\paragraph{Lemma closes and finishes:} 
+for each state $u$ of $@merge@~a~b$, if \emph{either} event set contains a @Finish n@, the invariant set at state $u$ will also contain a @Finish n@.
+If \emph{both} event sets contain a @Close n@, the invariant set will contain a @Close n@.
+
+This proof proceeds very similarly to the above lemma.
+For (Finish shared output, no pending events), a @Finished n@ is added to $\phi$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
+For (Shared pull, no pending values), a @Pull@ is emitted, and the @None@ case adds @Finished n@ to both sets, as with the invariant set.
+
+\paragraph{Lemma locals:} for any reads, outs or closes in one machine that are not in @inputs@ or @outputs@ of the other machine, no state in the other machine will affect the invariant set relating to these channels.
+By definition of @inputs@ and @outputs@.
+
 \paragraph{Lemma transitions:} for each state $u$ of $@merge@~a~b$ and corresponding states $s$ and $t$ of $a$ and $b$, any output transitions from $u$ either move along $s$'s output transitions or $t$'s output transitions, but not both.
 Similarly, the state type of $u$ will either be the type of $s$, the type of $t$, or a @Skip@.
+By inspection of each rule of @move@.
 
-\paragraph{Lemma appears:} for all states $s$ in $a$, if there is a path from $@initial@~a$ to $s$, then:
+\paragraph{Lemma paths:} for two given states $u$ and $v$ of $@merge@~a~b$, and their corresponding input machine states $s$ and $t$, for each path between $u$ and $v$, there exists a path between $s$ and $t$ such that all states in the $st$ path are in the $uv$ path, \emph{except}:
 \begin{itemize}
-\item if $s$ is a @Close@, either $s$ or a @Finish@ or a @Close@ from the other machine $b$ will appear in the output ($@merge@~a~b$);
-\item if $s$ is an @Out@, @Update@, @If@, $s$ will appear in the output;
-\item otherwise, $s$ 
+\item @Pull@s can be omitted if every predecessor path of the @Pull@ has a @Pull@, @Out@ or @OutFinished@ on the same channel;
+\item @Release@s can be omitted if the channel is an output of the other machine;
+\item @Release@s can be omitted if every successor path has a @Release@ on the same channel;
+\item @Close@s can be omitted if every successor path has a @Close@ or @Pull@ on the same channel.
 \end{itemize}
-ACTUALLY, \textbf{appears} isn't true.
 
-\paragraph{Proof sketch:} each state in $a$ and $b$ has an associated set of $\Psi$.
-If we deconstruct each state of $c$, it has a state of $a$, a state of $b$, and a $\Psi$ set for each.
+It is easy to see that for non-interfering transitions @Update@, @Skip@, @If@ and @Done@, this holds.
+For the other cases, it is necessary to perform induction over the @move@ rules, and inspect the event sets.
+If one of the input machine's states is a @Pull@, then either a @Pull@ or a @Skip@ may be produced.
+@Skip@s are only produced if the event set already contains a @Value@ or a @Finished@, which can only be produced by an earlier @Pull@, @Out@ or @OutFinished@.
+Likewise for the remaining cases.
 
-Show that in this $(\sptpC)$, the $\psi$ and $\phi$ is something similar to $s$ and $t$'s $\Psi$ set.
+\paragraph{Proof sketch:} 
+We can use the above lemmas to show that if the two input machines are valid, the result of @merge@ will also be valid.
+For non-shared or local channels, it is easy to see that the invariant set for these channels will be exactly as it was in the input machine (lemma locals).
 
-Total $\Psi$ is the union of $\psi$ and $\phi$, except if one set has @Closed@ but the other doesn't, that channel is still open.
+For shared inputs, since each input machine is valid, we know all @Pull@s are paired with @Release@s.
+Given lemma values, and that @Release@s remove @Value@s from the invariant set, we can see that each generated @Pull@ on a shared resource will only be generated when the invariant set has no @Value@s, and will be similarly paired for @Release@s.
 
-Then, look at each rule in @move@, and the preconditions should translate to sufficient preconditions on $\Psi$:
+For shared outputs, we just need to verify that the only way the reading machine can reach a reading state is while there is a @Value n@ in the invariant set, and this also follows easily from lemma values.
 
-@Skip@, @Update@ and @If@ are trivial.
-
-If both machines are @Done@, since both machines are also @ok@, all their inputs and outputs must be closed..
+Finally, we can see as a result of lemma closes and finishes and paths, that since both machines finish or close their inputs and outputs by their @Done@ state, then by the @Done@ of the merged machine, where both machines are also @Done@, all channels must be similarly closed at the end.

### 2015-04-05 2015mergingmerges/section/01-Introduction.tex
Hide material in intro, and shift invariants to the end

@@
 Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
-Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
-At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.
 
-Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
-However, for querying large data sets, these kind of combinators are essential.
-The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
-Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
-Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.
+% -- Use the space to argue why merges are needed, not the other way around.
+% -- This material would be better in the conclusion.
 
-In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
-This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
-Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.
+% Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
+% At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.
 
-Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
+% Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
+
+% For querying large data sets merge-like combinators are essential.
+% The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
+% Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
+% Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.
+
+% In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
+% This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
+% Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.
+
+% Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
+
+\TODO{Insert key example demonstrating some function involving merges that needs to be fused.}
+\TODO{The intro, including the following bulleted list should fit all on the first page.}
+
+We make the following contributions:
+\begin{itemize}
+\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
+\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
+\item invariants required of such machines (\S\ref{s:Machines:Invariants});
+\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
+\end{itemize}
 
@@
 For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
-
 Consider the following program.
@@
 We wish to disallow this program and others like it, where we cannot ensure execution with a bounded buffer.
-
-In this paper, we present:
-\begin{itemize}
-\item a kind of deterministic finite automata, for expressing streaming computations with value-dependent rates (\S\ref{s:Machines});
-\item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
-\item invariants required of such machines (\S\ref{s:Machines:Invariants});
-\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
-\end{itemize}

### 2015-04-05 2015mergingmerges/section/02-Machines.tex
contributions
Hide material in intro, and shift invariants to the end

@@
 \subsection{Combinators}
+\label{s:Machines:Combinators}
 As an example of @filter p xs@ combinator expressed as a machine, see figure~\ref{fig:com:filter}.
@@
 \input{figures/Combinators.tex}
-
-\subsection{Invariants}
-\label{s:Machines:Invariants}
-Not all machines are valid or meaningful, and we wish to rule them out.
-For example, a machine with an initial state of $@If@~(xs_e~>~5)~n$ is meaningless because there is no element of $xs$ that has been read from an input. 
-On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
-An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but having explicit releases makes synchronising two machines easier.
-
-The invariants are:
-\begin{itemize}
-\item each function mentioned in @Out@, @If@ or @Update@ can only refer to previously read values;
-\item each successful @Pull@ must be @Released@ before another @Pull@ can be made;
-\item all inputs must either be finished by an empty @Pull@ or @Close@d at @Done@;
-\item all outputs must be finished with @OutDone@ at @Done@;
-\item inputs cannot be read after an empty @Pull@ or after they are @Close@d;
-\item outputs cannot be written after they are finished with @OutDone@.
-\end{itemize}
-
-To perform this invariant checking, we annotate each state with a set of available information at that state.
-The DFA machines simplify the merging algorithm, but lose binding and scope information which would be implicit in a structured language.
-We need to reconstruct and check the scope information, similar to a type state system\CITE.
-
-The available event information can be that there exists some value that can be used, or that a channel is finished or closed.
-
-\begin{tabbing}
-MMMM \= M \= \kill
-@Event@ \> $=$  \> $@Value@~n$ \\
-       \> $~|$ \> $@Finished@~n$ \\
-       \> $~|$ \> $@Closed@~n$ \\
-\end{tabbing}
-
-To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
-Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions respectively.
-If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
-The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
-These @Out@ values are then implicitly released.
-
-For each state, we check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, functions of @Out@ and @If@ require their values to be present, and so on.
-
-\input{figures/Invariants.tex}
-
-\subsubsection{Generation}
-Figure~\ref{fig:inv:generation}.
-
-Each transition modifies its input's annotation, depending on the state type and the transition label.
-For example, the @Some@ output edge of a @Pull@ adds new information that a @Value@ is available.
-
-$$ \CheckOut{\Gamma}{s}{a} $$
-Under environment $\Gamma$ and machine $m$, the state $s$ has available information $a$.
-
-$$ \CheckOutTrans{l}{a} $$
-Under environment $\Gamma$ and machine $m$, the transition from state $s$ to $t$ under label $l$ has available information $a$.
-
-$$ \CheckStateType{s}{T} $$
-In machine $m$, state $s$ has state type $T$.
-
-\subsubsection{Checking}
-Figure~\ref{fig:inv:checking}.
-
-We check each state type against its final annotation. For example, @Release@ requires that there exists a @Value@ to release.
-
-$$ \StateOK $$
-Under environment $\Gamma$ and machine $m$, the state $s$ has a valid annotation.
-
-$$ \fvs{f} \subset a $$
-Any free variables mentioned in a function must exist as @Value@s in the annotation, in order for the function to use these values.
-
-\subsubsection{Environment closure}
-Figure~\ref{fig:inv:closure}.
-
-We must compute the transitive closure of each state's annotation.
-Given two environments, the final environment and unvetted environment, states are pulled from the unvetted environment and checked against the final environment.
-If the state already exists in the final environment and the annotations are equal, that state has been computed and can be removed from the unvetted environment.
-If the state exists in both environments but they are different, either the only differences are output values, and we proceed with the intersection of the two, or if there are other differences, the machine is invalid.
-Finally, if the state does not exist in the final environment, we add it to the final environment, compute all output transitions of the state, and add them to the unvetted environment.
-
-\TODO{Perhaps this would be described better as a function than an inference rule.}
-The problem, of course, is that state machines are not trees, so a derivation tree for typing does not make much sense.
-
-$$ \EnvGrow{\Gamma}{p}{\Gamma'} $$
-Under environment $\Gamma$ and machine $m$, with $p$ as an unvetted environment to be merged into $\Gamma$, the transitive closure of adding transitions is $\Gamma'$.
-
-\subsubsection{Entire machine}
-Figure~\ref{fig:inv:entire}.
-
-First, the environment closure is computed, starting with the initial state of the machine having no available information.
-Next, every state of the machine is checked to be valid against the environment closure.
-
-$$ m~@ok@ $$
-The entire machine is valid.
-
-\subsection{Code extraction}
-
-Extracting imperative code from these machines is quite straightforward.
-Variables are created for each @Pull@ source, to store a single value read from the source, and each output channel has a variable to store its latest output and its current state.
-Each state is converted to a basic block and given a unique identifier as its label.
-
-For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the Figure~\ref{fig:extract:filter}.
-
-\begin{figure}
-\begin{code}
-run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
-run pull_xs when_o = l10
- where
-  l10 = do  x  <- pull_xs
-            case x of
-             Nothing -> l90
-             Just x' -> l15 x'
-
-  l15 x | p x       = l20 x
-        | otherwise = l30 x
-
-  l20 x =   when_o (Just x)  >> l30 x
-
-  l30 x =   l10
-
-  l90   =   when_o Nothing  >> l91
-
-  l91   =   return ()
-\end{code}
-\caption{Extracted code for filter}
-\label{fig:extract:filter}
-\end{figure}

### 2015-04-05 2015mergingmerges/section/04-Invariants.tex
Hide material in intro, and shift invariants to the end

%!TEX root = ../Main.tex
\section{Invariants}
\label{s:Machines:Invariants}
Not all machines are valid or meaningful, and we wish to rule them out.
For example, a machine with an initial state of $@If@~(xs_e~>~5)~n$ is meaningless because there is no element of $xs$ that has been read from an input. 
On the other hand, some programs could be meaningful, but requiring them to be in a particular form simplifies fusion, later.
An example of such is a program that does not release its input before pulling again; a sympathetic code extraction would be to simply release before every pull, but having explicit releases makes synchronising two machines easier.

The invariants are:
\begin{itemize}
\item each function mentioned in @Out@, @If@ or @Update@ can only refer to previously read values;
\item each successful @Pull@ must be @Released@ before another @Pull@ can be made;
\item all inputs must either be finished by an empty @Pull@ or @Close@d at @Done@;
\item all outputs must be finished with @OutDone@ at @Done@;
\item inputs cannot be read after an empty @Pull@ or after they are @Close@d;
\item outputs cannot be written after they are finished with @OutDone@.
\end{itemize}

To perform this invariant checking, we annotate each state with a set of available information at that state.
The DFA machines simplify the merging algorithm, but lose binding and scope information which would be implicit in a structured language.
We need to reconstruct and check the scope information, similar to a type state system\CITE.

The available event information can be that there exists some value that can be used, or that a channel is finished or closed.

\begin{tabbing}
MMMM \= M \= \kill
@Event@ \> $=$  \> $@Value@~n$ \\
       \> $~|$ \> $@Finished@~n$ \\
       \> $~|$ \> $@Closed@~n$ \\
\end{tabbing}

To begin, we annotate the initial state with an empty set of information: at the start of the machine, we have no read values available, nor is it known (or likely!) that any channels are finished.
Then, for each output transition, we annotate the output state according to the current state type and the transition label: @Pull n@ adds a @Value n@ or @Finished n@ to its @Some n@ and @None@ transitions respectively.
If the output state has already been annotated, we check that the two annotations are the same, modulo values produced by @Out@.
The reason that annotations may be different in the values produced by @Out@ is that the @Out@ values are not explicitly released, and may be referred to by other functions, for example after fusing two combinators together.
These @Out@ values are then implicitly released.

For each state, we check that its annotation allows its execution: a @Pull n@ cannot run if $n$ is already finished; @Release n@ can only run if there is a @Value n@ to release, functions of @Out@ and @If@ require their values to be present, and so on.

\input{figures/Invariants.tex}

\subsubsection{Generation}
Figure~\ref{fig:inv:generation}.

Each transition modifies its input's annotation, depending on the state type and the transition label.
For example, the @Some@ output edge of a @Pull@ adds new information that a @Value@ is available.

$$ \CheckOut{\Gamma}{s}{a} $$
Under environment $\Gamma$ and machine $m$, the state $s$ has available information $a$.

$$ \CheckOutTrans{l}{a} $$
Under environment $\Gamma$ and machine $m$, the transition from state $s$ to $t$ under label $l$ has available information $a$.

$$ \CheckStateType{s}{T} $$
In machine $m$, state $s$ has state type $T$.

\subsubsection{Checking}
Figure~\ref{fig:inv:checking}.

We check each state type against its final annotation. For example, @Release@ requires that there exists a @Value@ to release.

$$ \StateOK $$
Under environment $\Gamma$ and machine $m$, the state $s$ has a valid annotation.

$$ \fvs{f} \subset a $$
Any free variables mentioned in a function must exist as @Value@s in the annotation, in order for the function to use these values.

\subsubsection{Environment closure}
Figure~\ref{fig:inv:closure}.

We must compute the transitive closure of each state's annotation.
Given two environments, the final environment and unvetted environment, states are pulled from the unvetted environment and checked against the final environment.
If the state already exists in the final environment and the annotations are equal, that state has been computed and can be removed from the unvetted environment.
If the state exists in both environments but they are different, either the only differences are output values, and we proceed with the intersection of the two, or if there are other differences, the machine is invalid.
Finally, if the state does not exist in the final environment, we add it to the final environment, compute all output transitions of the state, and add them to the unvetted environment.

\TODO{Perhaps this would be described better as a function than an inference rule.}
The problem, of course, is that state machines are not trees, so a derivation tree for typing does not make much sense.

$$ \EnvGrow{\Gamma}{p}{\Gamma'} $$
Under environment $\Gamma$ and machine $m$, with $p$ as an unvetted environment to be merged into $\Gamma$, the transitive closure of adding transitions is $\Gamma'$.

\subsubsection{Entire machine}
Figure~\ref{fig:inv:entire}.

First, the environment closure is computed, starting with the initial state of the machine having no available information.
Next, every state of the machine is checked to be valid against the environment closure.

$$ m~@ok@ $$
The entire machine is valid.

\subsection{Code extraction}

Extracting imperative code from these machines is quite straightforward.
Variables are created for each @Pull@ source, to store a single value read from the source, and each output channel has a variable to store its latest output and its current state.
Each state is converted to a basic block and given a unique identifier as its label.

For example, the @filter@ in Figure~\ref{fig:com:filter} is compiled to something like the Figure~\ref{fig:extract:filter}.

\begin{figure}
\begin{code}
run :: IO Int -> (Maybe Int -> IO ()) -> IO ()
run pull_xs when_o = l10
 where
  l10 = do  x  <- pull_xs
            case x of
             Nothing -> l90
             Just x' -> l15 x'

  l15 x | p x       = l20 x
        | otherwise = l30 x

  l20 x =   when_o (Just x)  >> l30 x

  l30 x =   l10

  l90   =   when_o Nothing  >> l91

  l91   =   return ()
\end{code}
\caption{Extracted code for filter}
\label{fig:extract:filter}
\end{figure}

### 2015-04-06 2015mergingmerges/section/03-Merging.tex
Add alternate presentation of shared pull rule
remove psis

@@
 \begin{tabbing}
-MM \= MM \= MMMMMM \= M\kill
-$\Psi$ \> $=$  \> @Value@     \> $n$         \\
-       \> $~|$ \> @Finished@  \> $n$         \\
-       \> $~|$ \> @Closed@    \> $n$         \\
+MMMM \= M \= \kill
+@Event@ \> $=$  \> $@Value@~n$ \\
+       \> $~|$ \> $@Finished@~n$ \\
+       \> $~|$ \> $@Closed@~n$ \\
 \end{tabbing}
 
-The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$.
+The @merge@ function takes two machines with states $l_1$ and $l_2$, and produces a new machine with state $l_1 \times \{@Event@\} \times l_2 \times \{@Event@\}$.
 It starts by creating an empty machine, and inserting states one by one, starting from the initial state as the initial state of each machine, and empty pending sets.
@@
 @merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
-        \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
+        \> $\to$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
 \\
@@
 \\
-$@merge@'$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$               \\
-           \> $\times$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
-           \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
+$@merge@'$ \> $:$ \> $l_1 \times \{@Event@\} \times l_2 \times \{@Event@\}$               \\
+           \> $\times$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
+           \> $\to$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
 \\
 MMM \= M \= MMMMMMMMMM \= M \=\kill
-$@merge@'~s_1~\psi_1~s_2~\psi_2~m$ \\
- \> $~|$ \> $(s_1, \psi_1, s_2, \psi_2) \in m$  \\
+$@merge@'~s_1~e_1~s_2~e_2~m$ \\
+ \> $~|$ \> $(s_1, e_1, s_2, e_2) \in m$  \\
  \> $=$  \> $m$ \\
- \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~\psi_1~s_2~\psi_2$ \\
- \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
+ \> $~|$ \> $(\mathit{type},\mathit{transitions}) \in @move@~s_1~e_1~s_2~e_2$ \\
+ \> $=$  \> $\bigcup_{t~\in~\mathit{transitions}}(@merge@'~(@snd@~t)~m')$\\
+~where  \> $m' = m~\cup~((s_1,e_1,s_2,e_2),~\mathit{type})~\cup~\mathit{transitions}$ \\
 \end{tabbing}
@@
 MMM \= M \= MMMMMMM\kill
-$@move@$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$ \\
-           \> $\to$ \> $\Gamma \times \{\Sigma \times (l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})\}$ \\
+$@move@$ \> $:$ \> $l_1 \times \{@Event@\} \times l_2 \times \{@Event@\}$ \\
+           \> $\to$ \> $@Type@ \times \{@StateOut@ \times (l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})\}$ \\
 \end{tabbing}
@@
 \input{figures/MergeGen.tex}
+\input{figures/MergeGenNew.tex}
 
@@
 Again, if the first machine is producing and the second is consuming, the first machine may only produce if there is no unhandled value on the channel.
-This is what $\psi$ and $\phi$ are for.
-If there is a value in the other machine's event set, $\phi$, a later case may allow the second machine to pull from it.
+This is what the two event sets $e_1$ and $e_2$ are for.
+If there is a value in the other machine's event set, $e_2$, a later case may allow the second machine to pull from it.
 
@@
 Because the channel may have other consumers, we do not stop it.
-So that further outputs on this channel do not wait on this machine, we add a @Closed@ event to $\psi$.
+So that further outputs on this channel do not wait on this machine, we add a @Closed@ event to $e_1$.
 
@@
 Either they require more than one element of buffering, or reasoning about function equality would be required to prove that they do not.
-The only unhandled cases above are where the two machines share input or outputs, and there are existing values buffered in $\psi$.
+The only unhandled cases above are where the two machines share input or outputs, and there are existing values buffered in $e_1$.
 In which case, we give a compile time error indicating that these two machines cannot be fused.
@@
 
-We start by computing $@merge@'$ of an empty machine, empty $\psi$ sets, and initial states of each machine:
+We start by computing $@merge@'$ of an empty machine, empty event sets, and initial states of each machine:
 $$
@@
 Now, calculate the first undefined move.
-Again, either machine may move: @m@ could compute its output, or @f@ could skip past the @Pull xs@, since this has been performed by @m@, witnessed by the @Value xs@ in @f@'s $\psi$ set.
+Again, either machine may move: @m@ could compute its output, or @f@ could skip past the @Pull xs@, since this has been performed by @m@, witnessed by the @Value xs@ in @f@'s event set.
 
@@
 \paragraph{Lemma values:} 
-for each state $u$ of $@merge@~a~b$, if either $\psi$ or $\phi$ contain a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
-(Note that the converse is not true, as @Value n@ will appear in the invariant set, but not $\psi$ or $\phi$ for local @Out@s)
+for each state $u$ of $@merge@~a~b$, if either event set contain a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
+(Note that the converse is not true, as @Value n@ will appear in the invariant set, but neither event set for local @Out@s)
 
@@
 
-The only way that @Value n@ can be added to $\psi$ or $\phi$ is through the rules (Shared output, ready to read) and (Shared pull, no pending values).
+The only way that @Value n@ can be added to the event set is through the rules (Shared output, ready to read) and (Shared pull, no pending values).
 These rules themselves produce state types of @Out@ and @Pull@ respectively, and by the definition of invariant sets @Value@ is inserted for these transitions.
@@
 For rule (Local release), the state type is @Release n@ which removes @Value n@ from the invariant set.
-Due to the locality preconditions $n~\not\in~@inputs@~m_2$ and $n~\not\in~@outputs@~m_2$, we know that neither of the rules to add @n@ could apply, so $\psi$ and $\phi$ must not contain @Value n@.
+Due to the locality preconditions $n~\not\in~@inputs@~m_2$ and $n~\not\in~@outputs@~m_2$, we know that neither of the rules to add @n@ could apply, so neither event set can contain @Value n@.
 
-For the cases that remove a @Value n@ from the $\psi$ set, the first two, (Release of output) and (Shared release: first release is skip) both only emit skips, so do not affect the invariant set.
+For the cases that remove a @Value n@ from the event set, the first two, (Release of output) and (Shared release: first release is skip) both only emit skips, so do not affect the invariant set.
 This means that while the invariant set may still contain a @Value n@, it is not obliged to.
@@
 
-The last case that removes a @Value n@ from the $\psi$ set is (Shared release: other machine already released).
-Here, we know $\psi$ contains a @Value n@ and that $\phi$ does not.
-This means that after removing the @Value n@ from $\psi$, neither set will have one.
+The last case that removes a @Value n@ from the event set is (Shared release: other machine already released).
+Here, we know the first machine's event set, $e_1$, contains a @Value n@ and that the second machine's event set, $e_2$, does not.
+This means that after removing the @Value n@ from $e_1$, neither set will have one.
 The produced state type also means that the invariant set will no longer have a @Value n@.
 
-For the remaining cases, the @Value@s inside the $\psi$ and $\phi$ sets are left alone, and their state types do not explicitly remove any @Value n@s from the invariant set either.
+For the remaining cases, the @Value@s inside the event sets are left alone, and their state types do not explicitly remove any @Value n@s from the invariant set either.
 We can conclude that any @Value@s in the sets have been added by other states, and not removed.
@@
 This proof proceeds very similarly to the above lemma.
-For (Finish shared output, no pending events), a @Finished n@ is added to $\phi$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
+For (Finish shared output, no pending events), a @Finished n@ is added to $e_2$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
 For (Shared pull, no pending values), a @Pull@ is emitted, and the @None@ case adds @Finished n@ to both sets, as with the invariant set.

### 2015-04-08 2015mergingmerges/section/01-Introduction.tex
intro

@@
 
-Despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
-Unlike the simple access patterns of append or zip, the merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
+Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
+However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
+Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
+For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+
+In addition, despite the years of work on array combinator fusion, scheduling dataflow networks and streaming languages, it seems one combinator has been cast aside as jetsam: the humble merge join.
+Unlike the simple access patterns of append or zip, merge join consumes from each of its inputs in a value-dependent order, always pulling from the input with the lowest associated key.
+
+Merge joins are the heart of database joins, where two tables are ordered by the same key.
+They are the ``conquer'' part of the divide and conquer merge sort.
+They can even be used to implement segmented appends.
 
@@
 
-\TODO{Insert key example demonstrating some function involving merges that needs to be fused.}
-\TODO{The intro, including the following bulleted list should fit all on the first page.}
+For example, if we have two sets of values over time, one useful operation is a correlation: to find the intersection of the values at given times, and compute some new value based on the two values:
+\begin{code}
+correlate :: [(Date,Value)] -> [(Date,Value)]
+          -> [(Date,Value)]
+correlate c1 c2
+ = let m1 = merge c1 c2
+       g1 = group_by compute m1
+   in  g1
+\end{code}
+
+The core of this algorithm is the @merge@ which intersperses the two inputs so equal dates are next to each other.
+Now suppose we have a third set and wish to compute the correlation of the first and second, and the second and third.
+Since we are already iterating over the second set, it would be ideal to compute this other correlation in the same iteration, fusing both correlations into a single loop.
+
+\begin{code}
+corr2     :: [(Date,Value)] -> [(Date,Value)]
+          -> [(Date,Value)]
+          ->([(Date,Value)],   [(Date,Value)])
+corr2     c1 c2 c3
+ = let m1 = merge c1 c2
+       m2 = merge c2 c3
+       g1 = group_by compute m1
+       g2 = group_by compute m2
+   in (g1, g2)
+\end{code}
+
+If this program were executed strictly, the first merge would be computed in total, reading @c1@ and @c2@, then the second merge would be computed, re-reading @c2@ from the start, and reading @c3@.
+Laziness is no panacea either, as whether @c2@ may still be read if @g1@ were consumed in total, then @g2@ consumed.
+Neither alone is sufficient to implement this as a single iteration.
+Instead, the two merges are fused, and whenever either merge requires input from @c2@, we attempt to execute the other merge until it too attempts to read from @c2@.
 
@@
 \item some common combinators implemented with such machines (\S\ref{s:Machines:Combinators});
+\item an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging});
 \item invariants required of such machines (\S\ref{s:Machines:Invariants});
-\item and an algorithm for merging non-recursive pairs of these machines (\S\ref{s:Merging}).
+\item and proofs of correctness (\S\ref{s:Proofs});
 \end{itemize}
 
-\subsection{Buffering}
-
-Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
-However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
-Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
-Consider the following program.
-Suppose we have two sources of integers from disk, wish to filter each and pair them together, then perform some action for each pair:
-
-\begin{code}
-zip1 :: [Int] -> [Int] -> IO ()
-zip1 xs ys
- = let xs'  = filter (>0) xs
-       ys'  = filter (<0) ys
-       zz   = zip xs' ys'
-   in  mapM_ print        zz
-\end{code}
-
-This program can be executed with no buffering of either input or any intermediate arrays.
-Now suppose wish to perform another action over the first source:
-
-\begin{code}
-zip2 :: [Int] -> [Int] -> IO ()
-zip2 xs ys
- = let xs'  = filter (>0) xs
-       ys'  = filter (<0) ys
-       xs2  = map    (+1) xs
-       zz   = zip xs' ys'
-   in  do   mapM_ print   zz
-            mapM_ print   xs2
-\end{code}
-
-If we were to execute this program either strictly or lazily, we must buffer the entire @xs@ source in memory while performing the first actions, then looping through the buffered @xs@ and performing the second action afterwards.
-However, we could execute this with no buffering if the relative ordering of the two @mapM_@s were not important, and could be reordered and interspersed as the data is read.
-We change the input lists to @Source@s and the @mapM_@s to @Perform@s, to denote that the IO actions are reorderable and to execute whenever data is available, and use @:&@ to join performs together.
-
-\begin{code}
-zip3 :: Source Int -> Source Int -> Perform
-zip3 xs ys
- = let xs'  = filter (>0) xs
-       ys'  = filter (<0) ys
-       xs2  = map    (+1) xs
-       zz   = zip xs' ys'
-   in  Perform print      zz
-   :&  Perform print      xs2
-\end{code}
-
-As a final example, suppose we removed the second data source and filtered the same source according to different predicates, then paired them together.
-Can we still execute this with no buffer, and only iteration over @xs@?
-
-\begin{code}
-zip4 :: [Int] -> IO ()
-zip4 xs
- = let xs'  = filter (>0) xs
-       xs'' = filter (<0) xs
-       zz   = zip xs' xs''
-   in  mapM_ print        zz
-\end{code}
-
-Sadly this last program requires an unbounded buffer or two iterations over @xs@, as we do not know anything about the contents of @xs@.
-Suppose that the entire list were made of a hundred positive numbers, followed by a hundred negative numbers: the first positive number would need to be kept around until the first negative number was reached, and likewise for the second and subsequent positive numbers -- so all one hundred positive numbers would need to be stored in memory until the negative numbers were reached.
-Now consider if, instead of a hundred positive numbers, there were a thousand: now it is necessary to buffer the thousand numbers in memory instead.
-The important part here is that one cannot compute, just by looking at the program, an adequate buffer size that works for all inputs.
-We wish to disallow this program and others like it, where we cannot ensure execution with a bounded buffer.
+\newpage

### 2015-04-08 2015mergingmerges/section/03-Merging.tex
change merge rule names
intro
separate proofs into own section

--- 2015-04-04 2015mergingmerges/section/03-Merging.tex
@@
 @merge@ \> $:$ \> $@Machine @l_1 \times @Machine @l_2$ \\
-        \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
+        \> $\to$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
 \\
-@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~(@init @m_2)~\{\}~@empty@$ \\
+@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~\{\}~(@init @m_2)~\{\}~@empty@$ \\
 \\
-$@merge@'$ \> $:$ \> $l_1 \times \{\Psi\} \times l_2 \times \{\Psi\}$               \\
-           \> $\times$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
-           \> $\to$ \> $@Machine @(l_1 \times \{\Psi\} \times l_2 \times \{\Psi\})$ \\
+$@merge@'$ \> $:$ \> $l_1 \times \{@Event@\} \times l_2 \times \{@Event@\}$               \\
+           \> $\times$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
+           \> $\to$ \> $@Machine @(l_1 \times \{@Event@\} \times l_2 \times \{@Event@\})$ \\
 \\
 MMM \= M \= MMMMMMMMMM \= M \=\kill
-$@merge@'~s_1~\psi_1~s_2~\psi_2~m$ \\
- \> $~|$ \> $(s_1, \psi_1, s_2, \psi_2) \in m$  \\
+$@merge@'~s_1~e_1~s_2~e_2~m$ \\
+ \> $~|$ \> $(s_1, e_1, s_2, e_2) \in m$  \\
  \> $=$  \> $m$ \\
- \> $~|$ \> $(\gamma,\delta) \in @move@~s_1~\psi_1~s_2~\psi_2$ \\
- \> $=$  \> $@fold@~@merge@'~(m \cup \gamma \cup \delta)~\delta$ \\
+ \> $~|$ \> $(\mathit{type},\mathit{transitions}) \in @move@~s_1~e_1~s_2~e_2$ \\
+ \> $=$  \> $\bigcup_{t~\in~\mathit{transitions}}(@merge@'~(@snd@~t)~m')$\\
+~where  \> $m' = m~\cup~((s_1,e_1,s_2,e_2),~\mathit{type})~\cup~\mathit{transitions}$ \\
 \end{tabbing}
@@
 \end{dot2tex}
-
-\subsection{Theorems we want to prove}
-
-As we do not wish to handle recursive combinators, we also do not need to handle merging cyclic machines.
-This is one case that regular dataflow handles but we do not.
-The predicate below checks if either of the machines does not use the other's input; if both used the other's input, it would create a cycle.
-
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@acyclic@~a~b$
-\> $:=$
-\> $\outputs{a}~\cap~\inputs{b}~=~\emptyset$
-\\
-\> $~\vee$ \> $\outputs{b}~\cap~\inputs{a}~=~\emptyset$
-\end{tabbing}
-
-Two machines cannot be merged if they both provide the same output, or write to the same output channel.
-
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@distinct@~a~b$
-\> $:=$
-\> $\outputs{a}~\cap~\outputs{b}~=\emptyset$
-\end{tabbing}
-
-\subsubsection{Preservation}
-The result of merging two machines should maintain the pull invariants (\S~\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.$
-\>
-\> $a~@ok@~\wedge~b~@ok@$
-\\
-\> $\wedge$
-\> $@acyclic@~a~b$
-\\
-\> $\wedge$
-\> $@distinct@~a~b$
-\\
-\> $\implies$
-\> $\forall c \in (@merge@~a~b).\ c~@ok@$
-\end{tabbing}
-
-\paragraph{Lemma transitions:} for each state $u$ of $@merge@~a~b$ and corresponding states $s$ and $t$ of $a$ and $b$, any output transitions from $u$ either move along $s$'s output transitions or $t$'s output transitions, but not both.
-Similarly, the state type of $u$ will either be the type of $s$, the type of $t$, or a @Skip@.
-
-\paragraph{Lemma appears:} for all states $s$ in $a$, if there is a path from $@initial@~a$ to $s$, then:
-\begin{itemize}
-\item if $s$ is a @Close@, either $s$ or a @Finish@ or a @Close@ from the other machine $b$ will appear in the output ($@merge@~a~b$);
-\item if $s$ is an @Out@, @Update@, @If@, $s$ will appear in the output;
-\item otherwise, $s$ 
-\end{itemize}
-ACTUALLY, \textbf{appears} isn't true.
-
-\paragraph{Proof sketch:} each state in $a$ and $b$ has an associated set of $\Psi$.
-If we deconstruct each state of $c$, it has a state of $a$, a state of $b$, and a $\Psi$ set for each.
-
-Show that in this $(\sptpC)$, the $\psi$ and $\phi$ is something similar to $s$ and $t$'s $\Psi$ set.
-
-Total $\Psi$ is the union of $\psi$ and $\phi$, except if one set has @Closed@ but the other doesn't, that channel is still open.
-
-Then, look at each rule in @move@, and the preconditions should translate to sufficient preconditions on $\Psi$:
-
-@Skip@, @Update@ and @If@ are trivial.
-
-If both machines are @Done@, since both machines are also @ok@, all their inputs and outputs must be closed..

### 2015-04-08 2015mergingmerges/section/05-Proofs.tex
separate proofs into own section

\section{Proofs}
\label{s:Proofs}

As we do not wish to handle recursive combinators, we also do not need to handle merging cyclic machines.
This is one case that regular dataflow handles but we do not.
The predicate below checks if either of the machines does not use the other's input; if both used the other's input, it would create a cycle.

\begin{tabbing}
MMMMMMM \= MM \= \kill
$@acyclic@~a~b$
\> $:=$
\> $\outputs{a}~\cap~\inputs{b}~=~\emptyset$
\\
\> $~\vee$ \> $\outputs{b}~\cap~\inputs{a}~=~\emptyset$
\end{tabbing}

Two machines cannot be merged if they both provide the same output, or write to the same output channel.

\begin{tabbing}
MMMMMMM \= MM \= \kill
$@distinct@~a~b$
\> $:=$
\> $\outputs{a}~\cap~\outputs{b}~=\emptyset$
\end{tabbing}

\subsection{Preservation}
The result of merging two machines should maintain the pull invariants (\S\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.

\begin{tabbing}
MMMMMMM \= MM \= MM \= \kill
$\forall a~b.$
\>
\> $a~@ok@~\wedge~b~@ok@$
\\
\> $\wedge$
\> $@acyclic@~a~b$
\\
\> $\wedge$
\> $@distinct@~a~b$
\\
\> $\implies$
\> $\forall c \in (@merge@~a~b).\ c~@ok@$
\end{tabbing}

\paragraph{Lemma values:} 
for each state $u$ of $@merge@~a~b$, if either event set contain a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
(Note that the converse is not true, as @Value n@ will appear in the invariant set, but neither event set for local @Out@s)

We proceed by induction over the @move@ rules.

The only way that @Value n@ can be added to the event set is through the rules (Shared output, ready to read) and (Shared pull, no pending values).
These rules themselves produce state types of @Out@ and @Pull@ respectively, and by the definition of invariant sets @Value@ is inserted for these transitions.

For rule (Local release), the state type is @Release n@ which removes @Value n@ from the invariant set.
Due to the locality preconditions $n~\not\in~@inputs@~m_2$ and $n~\not\in~@outputs@~m_2$, we know that neither of the rules to add @n@ could apply, so neither event set can contain @Value n@.

For the cases that remove a @Value n@ from the event set, the first two, (Release of output) and (Shared release: first release is skip) both only emit skips, so do not affect the invariant set.
This means that while the invariant set may still contain a @Value n@, it is not obliged to.
We appeal to the induction hypothesis for the input transitions, since earlier input transitions must have added the @Value n@ to both sets.

The last case that removes a @Value n@ from the event set is (Shared release: other machine already released).
Here, we know the first machine's event set, $e_1$, contains a @Value n@ and that the second machine's event set, $e_2$, does not.
This means that after removing the @Value n@ from $e_1$, neither set will have one.
The produced state type also means that the invariant set will no longer have a @Value n@.

For the remaining cases, the @Value@s inside the event sets are left alone, and their state types do not explicitly remove any @Value n@s from the invariant set either.
We can conclude that any @Value@s in the sets have been added by other states, and not removed.
The last thing is that @Value@s can be implicitly removed at join-points of states.
However, in order for a state to have @Value@s in its sets, all transitions into this state must have either inserted @Value@s itself, or kept @Value@s from its predecessors in turn; otherwise it would not have a transition to a state with @Value@s in its sets.

\paragraph{Lemma closes and finishes:} 
for each state $u$ of $@merge@~a~b$, if \emph{either} event set contains a @Finish n@, the invariant set at state $u$ will also contain a @Finish n@.
If \emph{both} event sets contain a @Close n@, the invariant set will contain a @Close n@.

This proof proceeds very similarly to the above lemma.
For (Finish shared output, no pending events), a @Finished n@ is added to $e_2$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
For (Shared pull, no pending values), a @Pull@ is emitted, and the @None@ case adds @Finished n@ to both sets, as with the invariant set.

\paragraph{Lemma locals:} for any reads, outs or closes in one machine that are not in @inputs@ or @outputs@ of the other machine, no state in the other machine will affect the invariant set relating to these channels.
By definition of @inputs@ and @outputs@.

\paragraph{Lemma transitions:} for each state $u$ of $@merge@~a~b$ and corresponding states $s$ and $t$ of $a$ and $b$, any output transitions from $u$ either move along $s$'s output transitions or $t$'s output transitions, but not both.
Similarly, the state type of $u$ will either be the type of $s$, the type of $t$, or a @Skip@.
By inspection of each rule of @move@.

\paragraph{Lemma paths:} for two given states $u$ and $v$ of $@merge@~a~b$, and their corresponding input machine states $s$ and $t$, for each path between $u$ and $v$, there exists a path between $s$ and $t$ such that all states in the $st$ path are in the $uv$ path, \emph{except}:
\begin{itemize}
\item @Pull@s can be omitted if every predecessor path of the @Pull@ has a @Pull@, @Out@ or @OutFinished@ on the same channel;
\item @Release@s can be omitted if the channel is an output of the other machine;
\item @Release@s can be omitted if every successor path has a @Release@ on the same channel;
\item @Close@s can be omitted if every successor path has a @Close@ or @Pull@ on the same channel.
\end{itemize}

It is easy to see that for non-interfering transitions @Update@, @Skip@, @If@ and @Done@, this holds.
For the other cases, it is necessary to perform induction over the @move@ rules, and inspect the event sets.
If one of the input machine's states is a @Pull@, then either a @Pull@ or a @Skip@ may be produced.
@Skip@s are only produced if the event set already contains a @Value@ or a @Finished@, which can only be produced by an earlier @Pull@, @Out@ or @OutFinished@.
Likewise for the remaining cases.

\paragraph{Proof sketch:} 
We can use the above lemmas to show that if the two input machines are valid, the result of @merge@ will also be valid.
For non-shared or local channels, it is easy to see that the invariant set for these channels will be exactly as it was in the input machine (lemma locals).

For shared inputs, since each input machine is valid, we know all @Pull@s are paired with @Release@s.
Given lemma values, and that @Release@s remove @Value@s from the invariant set, we can see that each generated @Pull@ on a shared resource will only be generated when the invariant set has no @Value@s, and will be similarly paired for @Release@s.

For shared outputs, we just need to verify that the only way the reading machine can reach a reading state is while there is a @Value n@ in the invariant set, and this also follows easily from lemma values.

Finally, we can see as a result of lemma closes and finishes and paths, that since both machines finish or close their inputs and outputs by their @Done@ state, then by the @Done@ of the merged machine, where both machines are also @Done@, all channels must be similarly closed at the end.

### 2015-04-08 2015mergingmerges/section/06-Conclusion.tex
add conclusion

%!TEX root = ../Main.tex
\section{Conclusion}
\label{s:Conclusion}

In this paper we have described a fusion algorithm that handles combinators with value-dependent access patterns, such as @merge@.
This is achieved by converting combinators to deterministic finite automata, then combining machines in a way that is similar to parallel execution of the machines.

As our algorithm only takes two machines, an entire combinator program must be fused by repeatedly fusing pairs of machines.
However, the order in which machines are fused can affect whether or not fusion is possible.
For example, given two inputs, maps of the inputs, and zipping the results:

\begin{code}
zipf xs ys
 = let xs' = map (+1) xs
       ys' = map (+1) ys
       zs  = zip xs' ys'
   in  zs
\end{code}

In this example, if the machines for @xs'@ and @ys'@ are fused together first, the fusion will prematurely decide on an ordering of both machines: perhaps all @xs'@ will be computed followed by all @ys'@, perhaps the other way around, or maybe they will be interspersed as we wish.
Then, when this arbitrarily-ordered machine is fused with the @zip@, it will fail as the @zip@ would need to read all the @xs'@ before finding a @ys'@.
It is important to note, however, that the ordering of fusion does not affect the semantics of the result, just whether a result is found.

To ensure a fusion result in this case, we must first fuse @zs@ with one of its inputs, then with the other input.
It is possible to attempt all possible permutations of the order and take the first that succeeds.
We are confident that there exists an algorithm to find a valid ordering, but this is left to future work.

Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.

Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.

For querying large data sets merge-like combinators are essential.
The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.

In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.

Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.

### 2015-04-09 2015mergingmerges/section/06-Conclusion.tex
ok

@@
 In this example, if the machines for @xs'@ and @ys'@ are fused together first, the fusion will prematurely decide on an ordering of both machines: perhaps all @xs'@ will be computed followed by all @ys'@, perhaps the other way around, or maybe they will be interspersed as we wish.
-Then, when this arbitrarily-ordered machine is fused with the @zip@, it will fail as the @zip@ would need to read all the @xs'@ before finding a @ys'@.
+If the first is taken, when this arbitrarily-ordered machine is fused with the @zip@, it will fail as the @zip@ would need to read all the @xs'@ before finding a @ys'@.
 It is important to note, however, that the ordering of fusion does not affect the semantics of the result, just whether a result is found.
@@
 We are confident that there exists an algorithm to find a valid ordering, but this is left to future work.
+
+% Kahn process networks\cite{kahn1976coroutines} are a flexible and valuable abstraction\cite{vrba2009kahn}, but they 
+% Our fusion system is able to handle a strict subset of Kahn process networks\cite{kahn1976coroutines}.
+% which are known to be flexible\cite{vrba2009kahn}
 

### 2015-04-09 2015mergingmerges/section/01-Introduction.tex
add conclusion
rough figure of group by

@@
 
-% Most dataflow language optimisations tend to focus on fully static networks where the exact access pattern is known at compile time\cite{thies2002streamit}, but disallowing combinators like merge join, append and filter.
-% At the other side of the spectrum, some dataflow languages such as Lucid\cite{stephens1997survey} focus on expressivity, forgoing any kind of static analyses for optimisations.
-
-% Merge joins, appends, and filters are not typically necessary for the bulk kind of operations such as audio transforms, video compression and so on, that regular dataflow has found its applications in\cite{johnston2004advances}.
-
-% For querying large data sets merge-like combinators are essential.
-% The ``MapReduce'' framework has been touted as a solution to easy distribution of workloads, but has been found to be lacking in flexibility\cite{vrba2009kahn}, in favour of Kahn process networks.
-% Kahn process networks alone are too flexible: many interesting properties we would like to assure, such as the absence of deadlocks, and ability to run in bounded memory, are undecidable.
-% Regular dataflow languages correspond to a subset of Kahn process networks, restricted to the point of keeping these desired properties\cite{thies2009language}.
-
-% In functional languages, fusion systems such as stream fusion\cite{coutts2007stream} and fold/build fusion\cite{jones2001playing} rely on the inliner to move producers into their consumers, after which rewrite rules can remove intermediate allocations.
-% This short-cut fusion works well for vertical fusion when producers have only single consumers, but when a producer is used by multiple consumers it cannot be inlined without duplicating work, so fusion will not occur.
-% Our earlier work on flow fusion\cite{lippmeier2013data} is able to fuse these multiple consumer cases, but only for a small set of combinators; neither @append@ nor @merge@ are handled.
-
-% Our goal is to extend the regular dataflow languages enough to allow this subset of dynamic combinators, while keeping the desired properties, for the purpose of compilation and optimisation.
-
 For example, if we have two sets of values over time, one useful operation is a correlation: to find the intersection of the values at given times, and compute some new value based on the two values:
@@
  = let m1 = merge c1 c2
-       g1 = group_by compute m1
+       g1 = groupby compute m1
    in  g1
@@
        m2 = merge c2 c3
-       g1 = group_by compute m1
-       g2 = group_by compute m2
+       g1 = groupby compute m1
+       g2 = groupby compute m2
    in (g1, g2)

### 2015-04-09 2015mergingmerges/section/02-Machines.tex
remove psis
rough figure of group by

@@
 \begin{tabbing}
-MMMM \= MM \= MMMMMM \= M \= M \kill
+MMM \= MM \= MMMMMM \= M \= M \kill
 $@Type@$ \> $=$ \> @Pull@     \>     \> $n$         \\
@@
          \> $~|$\> @Release@  \>     \> $n$         \\
-         \>     \> (Values must be released before the input can be pulled again) \\
+         \>     \> (Values must be released before the next pull) \\
 
          \> $~|$\> @Close@    \>     \> $n$         \\
-         \>     \> (A named input may still have data, but we will not read it) \\
+         \>     \> (An input may still have data, but we will not read it) \\
 
@@
 Many interesting combinators can be described using these machines.
-In figures~\ref{fig:com:map}-\ref{fig:com:merge}, the machines for some important combinators are given.
+In figures~\ref{fig:com:map}-\ref{fig:com:groupby}, the machines for some important combinators are given.
 To reduce clutter, these figures omit labels for @Unit@ transitions.

### 2015-04-09 2015mergingmerges/section/04-Invariants.tex
remove psis
proof of closure transitions

@@
 
-\subsubsection{Generation}
+\subsection{Generation}
 Figure~\ref{fig:inv:generation}.
@@
 
-\subsubsection{Checking}
+\subsection{Checking}
 Figure~\ref{fig:inv:checking}.
@@
 
-\subsubsection{Environment closure}
+\subsection{Environment closure}
 Figure~\ref{fig:inv:closure}.
@@
 
-\subsubsection{Entire machine}
+$$ \EnvMachine{m}{\Gamma} $$
+The environment closure for an entire machine is computed by starting with the initial state of the machine having no available information.
+
+\subsection{Entire machine}
 Figure~\ref{fig:inv:entire}.
 
-First, the environment closure is computed, starting with the initial state of the machine having no available information.
-Next, every state of the machine is checked to be valid against the environment closure.
+Every state of the machine is checked to be valid against the environment closure.
 

### 2015-04-09 2015mergingmerges/section/05-Proofs.tex
add conclusion
proof of closure transitions

@@
+%!TEX root = ../Main.tex
 \section{Proofs}
@@
 
-\subsection{Preservation}
-The result of merging two machines should maintain the pull invariants (\S\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
+The final requirement for correct fusion is the two machines to satisfy their invariants.
+\begin{tabbing}
+MMMMMMM \= MM \= \kill
+$@oks@~a~b$
+\> $:=$
+\> $@acyclic@~a~b~\wedge~@distinct@~a~b$ \\
+\> $\wedge$ \> $@ok@~a~\wedge~@ok@~b$ \\
+\end{tabbing}
+
+\subsection{Closure transitions}
+The first lemma we require is to show that the environment closure computes the intersection of all transitions into a given state.
+
+We would like to show that this holds for computing the closure for an entire machine:
+
+\begin{tabbing}
+MM \= MM \= MM \= \kill
+$\forall...$ \\
+\> \> $\EnvMachine{m}{\Gamma}$ \\
+\> $\wedge$ \> $u~:_E~e~\in~\Gamma$ \\
+\> $\implies$\> $\bigcap \{e'~|~n~\Rightarrow~u~\in~m$ \\
+\>           \>         \> $\wedge~\CheckOutTransUP{\Gamma'}{n}{u}{e'} \}$ \\
+\> $=$       \> $e$
+\end{tabbing}
+
+However this is not quite true, as the environment closure is only computed for reachable transitions.
+We also need a weaker postcondition in order to perform induction over the closure environment.
+We add the extra constraint that only transitions reachable from the pending state are considered.
+We also take into account the intersection of the initial set and the pending set.
+
+The actual lemma to prove, then, is:
+
+\begin{tabbing}
+MM \= MM \= MM \= \kill
+$\forall...$ \\
+\>          \> $\EnvGrow{\Gamma}{p}{\Gamma'}$ \\
+\> $\wedge$ \> $u~:_E~e~\in~\Gamma'$ \\
+\> $\implies$\> $\bigcap\{e'~|~n~\Rightarrow~u~\in~m$ \\
+\>           \>         \> $\wedge~\CheckOutTransUP{\Gamma'}{n}{u}{e'}$ \\
+\>           \>         \> $\wedge~@exists path from p to u@\}$ \\
+\> $\cap$ \> $\bigcap\{e' ~|~ u~:_E~e'~\in~\Gamma \} $ \\
+\> $\cap$ \> $\bigcap\{e' ~|~ u~:_E~e'~\in~p \} $ \\
+\> $=$       \> $e$
+\end{tabbing}
+
+We can proceed by induction over the closure.
+
+(CloComputed):
+regardless of whether $s~=~u$ or not, we simply use the induction hypothesis; since $a~=~b$, removing $s$ from $p$ does not change the result, as $s~:_E~a$ is left in $\Gamma$.
+
+(CloRemoveOut):
+again, we just use the induction hypothesis;
+$s$ is removed from $\Gamma$ but the intersection of $\Gamma$'s $s$ and $p$'s $s$ is inserted into $p$.
+
+(CloTransitions):
+if $s~=~u$, the intersection is not affected as it is moved from $p$ to $\Gamma$.
+If $u$ is one of the transitions from $s$, it is already in the first part of the intersection, so adding it to $p$ will not change the intersection.
+
+(CloFinished):
+here, $p$ is empty and there is no path from any elements of $p$ to $u$, so neither the first nor the last intersections have any effect, and the result is just $\Gamma$.
+
+\subsection{Values}
+For each state $u$ of $@merge@~a~b$, if either event set contains a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
+(Note that the converse is not true, as @Value n@ will appear in the invariant set, but neither event set for local @Out@s)
 
@@
 MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.$
+$\forall a~b.\ \forall u \in (@merge@~a~b).$
+\\
 \>
-\> $a~@ok@~\wedge~b~@ok@$
+\> $@oks@~a~b$
 \\
 \> $\wedge$
-\> $@acyclic@~a~b$
+\> $\EnvMachine{@merge@~a~b}{\Gamma}$
 \\
 \> $\wedge$
-\> $@distinct@~a~b$
+\> $u~:_E~e~\in~\Gamma$
+\\
+\> $\wedge$
+\> $\Value{n}~\in~(u \cdot e_1~\cup~u \cdot e_2)$
 \\
 \> $\implies$
-\> $\forall c \in (@merge@~a~b).\ c~@ok@$
+\> $\Value{n}~\in~e$
 \end{tabbing}
 
-\paragraph{Lemma values:} 
-for each state $u$ of $@merge@~a~b$, if either event set contain a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
-(Note that the converse is not true, as @Value n@ will appear in the invariant set, but neither event set for local @Out@s)
+Let $m~=~@merge@~a~b$.
+
+We first apply rule (CloMachine) to the assumption.
+
+$$
+\EnvMachine{@merge@~a~b}{\Gamma}
+\implies
+\EnvGrow{\emptyset}{@initial@~m~:_E~\emptyset}{\Gamma}
+$$
+
+We now proceed by induction over the closure.
+
+$$
+\EnvGrow{\Gamma}{p}{\Gamma'}
+$$
+
+(CloComputed): by the induction hypothesis.
 
@@
 However, in order for a state to have @Value@s in its sets, all transitions into this state must have either inserted @Value@s itself, or kept @Value@s from its predecessors in turn; otherwise it would not have a transition to a state with @Value@s in its sets.
+
+\subsection{Preservation}
+The result of merging two machines should maintain the pull invariants (\S\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
+
+\begin{tabbing}
+MMMMMMM \= MM \= MM \= \kill
+$\forall a~b.$
+\>
+\> $a~@ok@~\wedge~b~@ok@$
+\\
+\> $\wedge$
+\> $@acyclic@~a~b$
+\\
+\> $\wedge$
+\> $@distinct@~a~b$
+\\
+\> $\implies$
+\> $\forall c \in (@merge@~a~b).\ c~@ok@$
+\end{tabbing}
 

### 2015-04-12 2015mergingmerges/section/05-Proofs.tex
proof attempts

@@
 
+We also define a predicate for checking whether there exists a path of transitions between two states of a machine.
+\begin{tabbing}
+MMMMMMM \= MM \= \kill
+$@reachable@~m~u~v$ \\
+\> $:=$
+\> $u~=~v$ \\
+\> $\wedge$ \> $\bigvee_{u'|u~\Rightarrow~u'}~@reachable@~m~u'~v$ \\
+\end{tabbing}
+
 \subsection{Closure transitions}
@@
 
+\begin{lemma}
+Closure is intersection.
+
 \begin{tabbing}
@@
 \>           \>         \> $\wedge~\CheckOutTransUP{\Gamma'}{n}{u}{e'}$ \\
-\>           \>         \> $\wedge~@exists path from p to u@\}$ \\
+\>           \>         \> $\wedge~\bigvee_{v~\in~p}@reachable@~m~v~n\}$ \\
 \> $\cap$ \> $\bigcap\{e' ~|~ u~:_E~e'~\in~\Gamma \} $ \\
@@
 \end{tabbing}
+\end{lemma}
 
-We can proceed by induction over the closure.
+\begin{proof}
+By induction over the closure.
 
-(CloComputed):
+\begin{itemize}
+\item \textbf{Case (CloComputed):}
+
 regardless of whether $s~=~u$ or not, we simply use the induction hypothesis; since $a~=~b$, removing $s$ from $p$ does not change the result, as $s~:_E~a$ is left in $\Gamma$.
 
-(CloRemoveOut):
+\item \textbf{Case (CloTransitions):}
+
+if $s~=~u$, the intersection is not affected as it is moved from $p$ to $\Gamma$.
+If $u$ is one of the transitions from $s$, it is already in the first part of the intersection, so adding it to $p$ will not change the intersection.
+
+\item \textbf{Case (CloRemoveOut):}
+
 again, we just use the induction hypothesis;
@@
 
-(CloTransitions):
-if $s~=~u$, the intersection is not affected as it is moved from $p$ to $\Gamma$.
-If $u$ is one of the transitions from $s$, it is already in the first part of the intersection, so adding it to $p$ will not change the intersection.
+\item \textbf{Case (CloFinished):}
 
-(CloFinished):
 here, $p$ is empty and there is no path from any elements of $p$ to $u$, so neither the first nor the last intersections have any effect, and the result is just $\Gamma$.
+\end{itemize}
+
+\end{proof}
+
+Before we can prove that the result of merging satisfies invariants, we must show that the result of merging has an environment closure.
+Computing the environment closure requires all transitions into a state to have the same event set, except for values of outputs.
+
+\begin{lemma}
+Likeness implies closure
+\begin{tabbing}
+MM \= MM \= MM \kill
+ $~($
+\> $\forall u ~|~ \bigcup_{q~\in~p~\cup~\Gamma}(@reachable@~m~q~u).$ \\
+\> $\forall p_0 \in p~\cup~\Gamma.$ \\
+\> $\forall p_1 \in p~\cup~\Gamma.$ \\
+\> $\forall path_0 \in @path@~p_0~u.$ \\
+\> $\forall path_1 \in @path@~p_1~u.$ \\
+\> \> $path_0~:_E~\tau_0$ \\
+\> $\wedge$ \> $path_1~:_E~\tau_1$ \\
+\> $\implies$ \> $@like@~\tau_0~\tau_1)$ \\
+$\implies$ \> $\exists \Gamma'.$ \> $\EnvGrow{\Gamma}{p}{\Gamma'}$ \\
+\end{tabbing}
+\end{lemma}
+
+\begin{proof}
+Induction over $\langle \card{@states@~m~\setminus~\Gamma},~\Sigma_{s~:_E~\tau~\in~\Gamma}\card{\tau},~\card{p} \rangle$, where $\langle \cdots \rangle$ is a lexicographic ordering.
+
+\begin{description}
+\item[Case $\langle 0,~0,~0 \rangle$]
+
+$\card{p}~=~0 \implies p=\emptyset$
+
+apply (CloFinished)
+
+\item[Case $\langle g,~t,~y \rangle$]
+
+destruct $p$
+\begin{description}
+\item[Case $\emptyset$]
+
+apply (CloFinished)
+
+\item[Case $s~:_E~\tau, p$]
+
+destruct $s \in \Gamma$
+\begin{description}
+\item[Case false]
+
+apply (CloTransitions).
+
+apply i.h:
+
+$\card{@states@~m~\setminus~(s~:_E~\tau, \Gamma)} < \card{@states@~m~\setminus~\Gamma}$
+
+\item[Case $s~:_E~a~\in~\Gamma$]
+
+destruct $a~=~\tau$
+\begin{description}
+\item[Case $a~=~\tau$]
+
+apply (CloComputed):
+
+$g$ same. $t$ same. $y$ decreases.
+
+\item[Case $a~\not=~\tau$]
+
+apply (CloUpdateTransitions).
+
+$@like@~a~\tau$ from hypothesis.
+
+$a~\not=~\tau~\implies~\card{a~\cap~\tau}~<~\card{a}$
+\TODO{this is wrong}
+
+$g$ same. $t$ decreases.
+\end{description}
+\end{description}
+\end{description}
+\end{description}
+
+\end{proof}
 

### 2015-04-15 2015mergingmerges/section/05-Proofs.tex
fix proof of "likeness brings closure" by modifying closure rules slightly
define merged event set by input machines events

@@
 
+$g$ decreases, as
 $\card{@states@~m~\setminus~(s~:_E~\tau, \Gamma)} < \card{@states@~m~\setminus~\Gamma}$
@@
 
-destruct $a~=~\tau$
+$@like@~a~\tau$ from hypothesis.
+
+destruct $a~\cap~\tau~=~a$
 \begin{description}
-\item[Case $a~=~\tau$]
+\item[Case $a~\cap~\tau~=~a$]
 
-apply (CloComputed):
+apply (CloComputed).
 
-$g$ same. $t$ same. $y$ decreases.
+apply i.h.
+$g$ same. $t$ same. $y$ decreases, as $\card{p~\setminus~s}~<~\card{p}$
 
-\item[Case $a~\not=~\tau$]
+\item[Case $a~\cap~\tau~\not=~a$]
 
@@
 
-$@like@~a~\tau$ from hypothesis.
+apply i.h.
 
-$a~\not=~\tau~\implies~\card{a~\cap~\tau}~<~\card{a}$
-\TODO{this is wrong}
+$g$ same. $t$ decreases, as
+$a~\cap~\tau~\not=~a~\implies~\card{a~\cap~\tau}~<~\card{a}$
 
-$g$ same. $t$ decreases.
 \end{description}
@@
 \end{proof}
+
+\subsection{Merged event set}
+Assuming that the two inputs to a merged program are valid according to the invariants, we must show that the merged result, if it exists, is also valid.
+We can compute the event set for a merge state from the event sets of the corresponding input machines' states.
+
+For each input or output of either machine, if it is closed and local, it is closed.
+If it is shared and only closed in one machine, the other machine must still be using it.
+If either machine has a value, the result machine has a value.
+We must also check the state sets of the output machine, $e_1$ and $e_2$ here, because the input machine may have made an output and silently discarded it, but if that value is yet to be read by the other machine, it cannot be silently discarded by the result machine.
+If either machine is finished, the result machine is finished.
+
+\begin{tabbing}
+M \= MMMMMM \= M \= MM \kill
+$@mergeT@~(s_1,~e_1,~s_2,~e_2)$ \\
+$\bigcup_{n~\in~ns}$ \\
+ \> $\Closed{n}$ \> $~|~$ \> $\Closed{n} \in e_a \cup e_1 ~\wedge~ @notin@ b n$ \\
+ \>              \> $\vee$\> $\Closed{n} \in e_b \cup e_2 ~\wedge~ @notin@ a n$ \\
+ \>              \> $\vee$\> $\Closed{n} \in e_a \cup e_1 ~\wedge~ \Closed{n} \in e_b \cup e_2$ \\
+\\
+ \> $\Value{n} $ \> $~|~$ \> $\Value{n} \in e_a \cup e_1  ~\vee~   \Value{n} \in e_b \cup e_2$ \\
+\\
+ \> $\Finished{n}$\>$~|~$ \> $\Finished{n} \in e_a \cup e_1  ~\vee~   \Finished{n} \in e_b \cup e_2$ \\
+where \\
+ \> $\EnvMachine{a}{\Gamma_a}$ \\
+ \> $\EnvMachine{b}{\Gamma_b}$ \\
+ \> $s_1~:_E~e_a~\in~\Gamma_a$ \\
+ \> $s_2~:_E~e_b~\in~\Gamma_b$ \\
+ \> $ns~=~@inputs@~a~\cup~@outputs@~a~\cup~@inputs@~b~\cup~@outputs@~b$ \\
+ \> $@notin@~m~n~=~n~\not\in~@inputs@~m~\cup~@outputs@~m$ \\
+\end{tabbing}
 

### 2015-04-16 2015mergingmerges/section/02-Machines.tex
change some @move@ rules over to new style

@@
 \input{figures/Combinators.tex}
+
+\subsection{Definitions}
+\TODO{define some interesting functions:}
+
+$@channel@~t$ takes a state type (eg $@Out@~f~n$) and returns the channel it is operating on ($n$), if applicable.
+States like @Skip@ and @Done@ don't have an associated channel.
+
+$@inputs@~m$ is the set of all channels used in @Pull@s, @Release@s and @Close@s.
+
+$@outputs@~m$ is the set of all channels used in @Out@s, @OutDone@s, @Update@s and @If@s.
+While @Update@s and @If@s do not perform outputs, they must belong to a particular output channel, and including them in @outputs@ simplifies handling of degenerate cases where @Update@s are performed but no @Out@s.
+
+$@initial@~m$ gives the initial state of the machine.
+
+$@states@~m$ gives the set of states of the machine, ignoring transitions.

### 2015-04-16 2015mergingmerges/section/03-Merging.tex
change some @move@ rules over to new style

@@
 \\
-@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@init @m_1)~\{\}~(@init @m_2)~\{\}~@empty@$ \\
+@merge @$m_1~m_2$ \> $=$ \> $@merge@'~(@initial @m_1)~\{\}~(@initial @m_2)~\{\}~@empty@$ \\
 \\
@@
 \subsubsection{Output and finishing}
-The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Out@, @OutFinished@, @Pull@, @Release@ and @Close@ are more complicated.
+The previous cases have been rather trivial as they require no synchronisation between the machines, but the remaining cases of @Out@, @OutDone@, @Pull@, @Release@ and @Close@ are more complicated.
 
@@
 
-The cases for @OutFinished@ proceed similarly to @Out@.
+The cases for @OutDone@ proceed similarly to @Out@.
 
@@
 
-If there is a @Finished@ in either machine's event set, and neither machine has a pending @Value@s to deal with, we know that the other machine has had either an @OutFinished@ or an empty @Pull@.
+If there is a @Finished@ in either machine's event set, and neither machine has a pending @Value@s to deal with, we know that the other machine has had either an @OutDone@ or an empty @Pull@.
 We proceed to the @None@ case of the @Pull@.
@@
 m3 : Release x   { Unit => m1 }
-m4 : OutFinished m{ Unit => m5 }
+m4 : OutDone m   { Unit => m5 }
 m5 : Done
@@
 f4 : Release x   { Unit => f1 }
-f5 : OutFinished f{ Unit => f6 }
+f5 : OutDone m   { Unit => f6 }
 f6 : Done
@@
 move m f m4 {Finished xs} f1 {Finished xs} =
-    OutFinished m
+    OutDone m
     { Unit => m5, {Finished xs}, f1, {Finished xs} }
@@
 move m f m5 {Finished xs} f5 {Finished xs} =
-    OutFinished f
+    OutDone f
     { Unit => m5, {Finished xs}, f6, {Finished xs} }
@@
 m4,{Finished xs},f1,{Finished xs}
-    : OutFinished m
+    : OutDone m
     { Unit => m5, {Finished xs}, f1, {Finished xs} }
@@
 m5,{Finished xs},f5,{Finished xs}
-    : OutFinished f
+    : OutDone f
     { Unit => m5, {Finished xs}, f6, {Finished xs} }
@@
     70 [label="Release x"];
-    900[label="OutFinished m"];
-    920[label="OutFinished f"];
+    900[label="OutDone m"];
+    920[label="OutDone f"];
     930[label="Done"];

### 2015-04-16 2015mergingmerges/section/05-Proofs.tex
change some @move@ rules over to new style

@@
 This proof proceeds very similarly to the above lemma.
-For (Finish shared output, no pending events), a @Finished n@ is added to $e_2$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
+For (Finish shared output, no pending events), a @Finished n@ is added to $e_2$, while @OutDone n@ adds a @Finish n@ to the invariant set.
 For (Shared pull, no pending values), a @Pull@ is emitted, and the @None@ case adds @Finished n@ to both sets, as with the invariant set.
@@
 \begin{itemize}
-\item @Pull@s can be omitted if every predecessor path of the @Pull@ has a @Pull@, @Out@ or @OutFinished@ on the same channel;
+\item @Pull@s can be omitted if every predecessor path of the @Pull@ has a @Pull@, @Out@ or @OutDone@ on the same channel;
 \item @Release@s can be omitted if the channel is an output of the other machine;
@@
 If one of the input machine's states is a @Pull@, then either a @Pull@ or a @Skip@ may be produced.
-@Skip@s are only produced if the event set already contains a @Value@ or a @Finished@, which can only be produced by an earlier @Pull@, @Out@ or @OutFinished@.
+@Skip@s are only produced if the event set already contains a @Value@ or a @Finished@, which can only be produced by an earlier @Pull@, @Out@ or @OutDone@.
 Likewise for the remaining cases.

### 2015-11-14 2016icicle/Main.tex
Icicle paper

--- 2014-12-19 2015mergingmerges/Main.tex
@@
-\documentclass{sigplanconf}
+\documentclass[preprint]{sigplanconf}
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{graphicx}
@@
 \usepackage{mathptmx}
+\usepackage{mathtools}
 \usepackage{stmaryrd}
@@
 
-\title{Merging merges, more or less}
+\title{Icicle: a single-pass query language}
 
 \authorinfo{
-  Amos Robinson$^\dagger$ 
-  \and Ben Lippmeier$^\dagger$
-  \and Gabriele Keller$^\dagger$ 
-  \and Manuel Chakravarty$^\dagger$ 
+  Amos Robinson$^\dagger$$^\ddagger$
+  \and Huw Campbell$^\ddagger$
+  \and Mark Hibberd$^\ddagger$
+  \and Tran Ma$^\ddagger$
+  \and Jacob Stanley$^\ddagger$
 }{
@@
     University of New South Wales, Australia \\[2pt]
-    \textsf{\{amosr,benl,keller,chak\}@cse.unsw.edu.au}
+    \textsf{amosr@cse.unsw.edu.au}
+  }
+  \shortstack{
+    $^\ddagger$Ambiata      \\
+    Big data and shit       \\[2pt]
+    \textsf{\{first.last\}@ambiata.com}
   }
@@
 \begin{abstract}
+When streaming a large amount of data, simply iterating over the data may take hours.
+In this case, the difference between a query that takes one iteration and a query that takes multiple iterations must be very explicit.
+For many purposes, limiting queries to a single pass is sufficient.
+We introduce a type system based on modality types to enforce this single pass restriction.
+By using an appropriate intermediate language we can fuse multiple independent queries together, before extracting efficient single-loop C code.
 \end{abstract}
@@
 \keywords
-	Arrays; Fusion; Haskell
+	Arrays; Fusion
 
 \input{section/01-Introduction.tex}
+\input{section/02-Source.tex}
+\input{section/03-Types.tex}
 
 \section*{Acknowledgements}
-The audience of the inaugural SAWDAP workshop, especially Thomas Sutton and Patryk Zadarnowski, were very helpful in an early version of this work.
+I would like to thank all of Ambiata for allowing me to work on such an interesting and useful problem, and on top of that being the most sensible group of people I have ever had the pleasure of working with.
 

### 2015-11-14 2016icicle/section/01-Introduction.tex
Icicle paper

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}

For queries over a large amount of data, it is important that the queries are as fast as possible.
If a query must read and loop over the data multiple times, that could more than double the query time, as well as potentially requiring large intermediate storage.
It also turns out that with some ingenuity, many interesting algorithms can be expressed or at least approximated in a single pass.
The problem with many existing query languages, however, is that it can be hard to tell how many passes will be used for a given query.
This execution-agnostic property of declarative languages can be both a benefit and a disadvantage.

With a language like SQL, it may not be immediately obvious how many passes a query requires, and it may depend on particular compiler optimisations and the mood of the database.
For example, how many iterations do the following queries require?

\begin{code}
SELECT open - MIN(open)    FROM stocks;
SELECT open - MIN(open, 1) FROM stocks;
SELECT        MIN(open)    FROM stocks;
SELECT open,  MIN(open)    FROM stocks;
\end{code}

For our purposes, we would rather stricter compile-time guarantees: if a query typechecks and compiles, it should be able to be executed efficiently.
The aim is to remove the mystery of database and compiler optimisations with a clear, unambiguous definition of which queries should be accepted and efficient, and which will be rejected.

\TODO{Look into data flow languages like Lucid and how they enforce causality, cite them}

We make the following contributions:
\begin{itemize}
\item 
We present a query language, Icicle, with a type system using modality types\CITE{Rowan} to enforce the single-pass restriction(\S\ref{s:Source}).
\item
We present an intermediate language similar to MapReduce\CITE{?} that simplifies fusion between multiple queries(\S ref).
\item
We describe an algorithm for extracting efficient imperative code from the intermediate language, used to extract C code in our implementation (\S ref).
\item
We show benchmarks of our implementation, which performs favourably even compared to hand-optimised text programs such as @grep@ and @wc@ (\S ref).
\end{itemize}

Our implementation is available at \url{https://github.com/ambiata/icicle}.

### 2015-11-14 2016icicle/section/02-Source.tex
Icicle paper

%!TEX root = ../Main.tex
\section{Icicle Source}
\label{s:Source}

We present a cut down version of our implementation; our implementation has many more primitives, case expressions, as well as primitive contexts for windowing based on time, folding over groups, and so on.

The first example defines @sum@ as a fold over the input data.
The initial value for the fold is $0$, followed by the previous value @s@ added to the current value @v@.
\begin{code}
sum v
 =  fold s = 0; s+v
 ~> s
\end{code}

Next, we can define @count@ and mean in terms of this sum.
\begin{code}
count
 =  sum 1

mean v
 =  sum v / count
\end{code}

We can use @filter@ to count only the days where the close price is higher than the open.
\begin{code}
gap
 =  filter close > open
 ~> count

proportion
 =  gap / count
\end{code}

We can also use @group@ to put into buckets based on the percentage of growth.
Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
\begin{code}
growthPercent
 =  int ((close - open) / open * 100)

growthBy
 =  let total = count
 ~> group growthPercent
 ~> count / total
\end{code}

Finally, we could compare the average growth over the last five entries to the average growth over the last fifty entries, and see which is bigger.
\begin{code}
compareGrowth
 =  (latest 5  ~> mean growthPercent)
 -  (latest 50 ~> mean growthPercent)

\end{code}

\input{figures/Source-Grammar.tex}

### 2015-11-14 2016icicle/section/03-Types.tex
Icicle paper

%!TEX root = ../Main.tex
\section{Types}
\label{s:Types}

\input{figures/Source-Type.tex}

%%% Types
In figure~\ref{fig:source:type:defs} we define the types.
We use $T$ for base types, of which we only have natural numbers and booleans.

Modalities are $M$, with @Aggregate@ meaning the result of some fold, @Element@ being an element of the input stream, and @Pure@ being a pure computation.
The modalities are defined so that @Aggregate@ and @Element@ computations cannot be mixed.
There is no way to turn an aggregate computation into an element computation, as aggregate computations are only available once the entire stream has been read.
Element computations must be explicitly folded over or aggregated in some way, in order to produce an aggregate computation based off it.
\TODO{Staged computation similarities}

The @Pure@ modality is actually the absence of a mode; this is simply made explicit for clarity.

Finally we define function types as $T_\to$, which can be a simple return type with a modality, or a function taking a modality argument and returning a function.
For simplicity of compilation, we have strictly disallowed higher order functions by requiring the argument to be a base type rather than a function type.

At various points in the following rules we will implicitly convert function types $T_\to$ to modal base types $M~T$; at these points function arrows are not allowed.

%%% Function application
In figure~\ref{fig:source:type:wrap} there are two judgments defined, which are both used for the type of function application.
In contrast to traditional modality type systems, we infer the boxing and unboxing of various modalities.

Informally, if a function expects an @Element@ and is passed a @Pure@ argument, we should box the argument into @Element@.
If a function expects a @Pure@ computation and is passed an @Element@, we must unbox the argument, apply the function, then rebox the result.
However, this reboxing can only occur if the result is @Pure@ or @Element@ - an @Aggregate@ result cannot be boxed into an @Element@.

The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to reboxing the function return type $\tau$ into modality $m$, returning $\tau'$ as the boxed return type.
The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
The rule (WrapElement) applies when reboxing an @Element@ result as an @Element@, and similarly for (WrapAggregate).
For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.

The judgment $\WrapApp{\tau}{\phi}{\tau'}$ attempts to apply the function type $\tau$ to argument $\phi$, performing boxing and unboxing if necessary.
If both modalities and types are equal, (WrapApp) applies the function as normal.
When the function expects a @Pure@ argument and is passed an @Element@ or an @Aggregate@, we perform reboxing as in (WrapAppElement) and (WrapAppAggregate).
Finally, no matter what the function expects, if it is passed a @Pure@ it is simply boxed and applied as in (WrapAppPure), with no need to rebox the result.

\TODO{Or should we just use an explicitly box/unbox version?}

%%% Expressions
With function application already defined, typechecking expressions in figure~\ref{fig:source:type:exp} is quite simple.
The judgment $\Typecheck{\Gamma}{x}{\tau}$ means that under a context $\Gamma$, the expression $x$ has type $\tau$.
Rules (TcVar) and (TcPrim) simply look up the variable or primitive in the environment.
For (TcApp) we simply use the boxed function application judgment from figure~\ref{fig:source:type:wrap}.

\TODO{I think we need a box/cast rule for going from @Pure@ to @Element@ etc}

%%% Contexts
Figure~\ref{fig:source:type:ctx} shows the typing rules for contexts.
The judgment $\Typecheck{\Gamma}{c}{\tau}$ means that under a context $\Gamma$, the context $c$ has type $\tau$.
For simplicity, we reuse the same judgment for expressions and contexts.

The rule (TcLet) typechecks its definition and adds it to the environment when typechecking the body.

For (TcFilter), the predicate must be an @Element@, as it is executed for each element of the input.
The rest of the computation must return an @Aggregate@ over the filtered input.

For (TcFold), the form of fold is $@fold@~v~@=@~x@;@~x'$ where $v$ is the name of the fold binding, $x$ is its initial value, and $x'$ computes the successor value, based on the previous value of $v$.
The initial expression $x$ must be @Pure@ of some type $\tau$, and the successor expression $x'$ is typechecked with the fold binding set to $@Element@~\tau$ in the environment.
The result of the successor expression must also have type $@Element@~\tau$.
This is because the current value of the fold is available to the successor expression, as well as the current elements, but the results of other folds are not available.
In the remaining context, the fold value is only available as an @Aggregate@.

For (TcLatestElement), if you have a $@latest@~n$ of @Element@s we fill a circular buffer with the last $n$ values.
Similarly, (TcLatestAggregate) uses a circular buffer, and then applies the aggregate to the last $n$ values.

Finally, (TcGroup) requires its key or group ``by'' to be an @Element@, and the value to be an @Aggregate@.
It builds up a map from key to value, repeatedly applying the aggregate as new values are found.

### 2015-11-14 2016icicle/section/01-Introduction.tex
ok

@@
 
+For feature generation for machine learning, our queries have some quite specific restrictions.
+Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.
+Each query is implicitly partitioned by the entity, the company code in this case, and performs an aggregate over that company's entries, sorted by date.
+In this way, our query language is actually a streaming dataflow language with only one input stream and clock.
+By making use of these restrictions, we are able to generate more efficient code than a general query language.
+
 \TODO{Look into data flow languages like Lucid and how they enforce causality, cite them}

### 2015-11-14 2016icicle/section/02-Source.tex
ok

@@
 
-We can also use @group@ to put into buckets based on the percentage of growth.
+Finally, we can use @group@ to put into buckets based on the percentage of growth.
 Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
@@
 
-Finally, we could compare the average growth over the last five entries to the average growth over the last fifty entries, and see which is bigger.
-\begin{code}
-compareGrowth
- =  (latest 5  ~> mean growthPercent)
- -  (latest 50 ~> mean growthPercent)
-
-\end{code}
+\TODO{Mention resumables and bubblegum?}
 

### 2015-11-24 2016icicle/Main.tex
related work: lustre

@@
   Amos Robinson$^\dagger$$^\ddagger$
-  \and Huw Campbell$^\ddagger$
-  \and Mark Hibberd$^\ddagger$
+  \and Ben Lippmeier$^\dagger$
   \and Tran Ma$^\ddagger$
@@
     University of New South Wales, Australia \\[2pt]
-    \textsf{amosr@cse.unsw.edu.au}
+    \textsf{amosr,benl@cse.unsw.edu.au}
   }
@@
 
+\input{section/XX-Related-DataFlow.tex}
+
 \section*{Acknowledgements}
-I would like to thank all of Ambiata for allowing me to work on such an interesting and useful problem, and on top of that being the most sensible group of people I have ever had the pleasure of working with.
+I would like to thank all of Ambiata, particularly Huw Campbell and Mark Hibberd.
 

### 2015-11-24 2016icicle/section/XX-Related-DataFlow.tex
related work: lustre

%!TEX root = ../Main.tex
\section{Related work}
\label{s:Related}
\subsection{Data flow languages}

The closest related work are synchronous data flow languages such as {\sc Lustre}, Icicle programs are restricted to those that can be computed in bounded memory.
{\sc Lustre}\cite{halbwachs1991synchronous} achieves this in three main ways:

\begin{enumerate}
\item They allow only restricted set of primitives such as @when@ for filtering, @pre@ for a single element buffer, and so on.
While the primitives used are quite different, our primitives must also be carefully chosen to ensure bounded memory.
\item Cycles in the graph must contain at least one @pre@, to break the dependency loop.
In comparison we have only non-recursive definitions and recursion is handled by the @fold@ primitive.
\item Operators such as addition can only be applied to input streams with the same clock or rate; an expression like @(X when C)@ @+@ @(Y when (not C))@ would require an unbounded buffer\CITE{Lustre clock stuff}.
In contrast, because our only clock changing operation (@filter@) must return an aggregate there is no possibility of applying a primitive to different clocks.
\end{enumerate}

Existing data flow languages tend to focus on stream transformers running over potentially infinite data, while our purpose is specifically for performing aggregates that are only available at the end of the stream.
While this means these languages are more expressive, being able to output streams as well as aggregates, it can obscure the desired meaning.
By making the separation between @Element@ and @Aggregate@ computations explicit, a programmer cannot accidentally mistake an intermediate value of an aggregate for  the final value.

For example, checking an element of the stream against the mean of the entire stream is impossible in a single pass, and our typesystem rules this program out:
\begin{code}
   filter value > mean value
-> count
\end{code}

However, a similar program written in {\sc Lustre} would compare the running mean against the current value.
\begin{code}
count when (value > mean(value))
\end{code}

While this may be the desired behaviour in some cases, we believe this decision must be made explicit, by using the @running@ primitive.
\begin{code}
   filter value > running (mean value)
-> count
\end{code}

In summary, while our language is quite similar to existing synchronous data flow languages, the extra restrictions we impose allow us to use a simpler type system.

### 2015-11-27 2016icicle/section/XX-Related-DataFlow.tex
ok

@@
 
-While this may be the desired behaviour in some cases, we believe this decision must be made explicit, by using the @running@ primitive.
+While this may be the desired behaviour in some cases, we believe this decision must be made explicit.
+At the moment this must be hand-written as a fold, but we intend to add a new primitive called @running@ which peeks into a fold and uses its intermediate result.
 \begin{code}

### 2015-12-05 2016icicle/section/02-Source.tex
related work: lustre
ok
changes

@@
 
-The first example defines @sum@ as a fold over the input data.
-The initial value for the fold is $0$, followed by the previous value @s@ added to the current value @v@.
-\begin{code}
-sum v
- =  fold s = 0; s+v
- ~> s
-\end{code}
+The first example defines $\mi{sum}$ as a fold over the input data.
+The initial value for the fold is $0$, followed by the previous value $s$ added to the current value $v$.
 
-Next, we can define @count@ and mean in terms of this sum.
-\begin{code}
-count
- =  sum 1
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$\mi{sum}~v$    \\
+\> $=$  \> $@fold@~s~=~0~@then@~s~+~v$ \\
+\> @in@ \> $s$ \\
+\end{tabbing}
 
-mean v
- =  sum v / count
-\end{code}
+Next, we can define $\mi{count}$ and $\mi{mean}$ in terms of this sum.
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$\mi{count}$                                        \\
+ \> $=$  \> $\mi{sum}~1$                            \\
+                                                    \\
+$\mi{mean}~v$                                       \\
+ \> $=$  \> $\mi{sum}~v~/~\mi{count}$               \\
+\end{tabbing}
 
 We can use @filter@ to count only the days where the close price is higher than the open.
-\begin{code}
-gap
- =  filter close > open
- ~> count
-
-proportion
- =  gap / count
-\end{code}
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$\mi{gap}$                                          \\
+ \> $=$  \> $@filter@~\mi{close}~>~\mi{open}$       \\
+ \> @in@ \> $\mi{count}$                            \\
+                                                    \\
+$\mi{proportion}$                                   \\
+ \> $=$  \> $\mi{gap}~/~\mi{count}$
+\end{tabbing}
 
@@
 Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
-\begin{code}
-growthPercent
- =  int ((close - open) / open * 100)
-
-growthBy
- =  let total = count
- ~> group growthPercent
- ~> count / total
-\end{code}
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$\mi{growthPercent}$                                        \\
+ \> $=$  \> $((\mi{close}-\mi{open}) \times 100 ) / \mi{open}$  \\
+                                                            \\
+$\mi{growthBy}$                                             \\
+ \> $=$  \> $@let@~\mi{total} = \mi{count}$                 \\
+ \> @in@ \> $@group@~\mi{growthPercent}$                    \\
+ \> @in@ \> $\mi{count} / \mi{total}$                    \\
+\end{tabbing}
 
@@
 
+\TODO{Talk about scan}
+
+We formally define the grammar of the language in figure~\ref{fig:source:grammar}.
+Note that general application is not allowed: arguments can only be applied to primitives and named functions.
+Similarly, functions and primitives must be fully applied.
+This, combined with the lack of lambda construction, means that higher order functions are outlawed.
+Outlawing higher order functions is somewhat unsatisfying, but simplifies the typing rules and makes it easier to support our performance guarantees.
+
+While the examples shown allow user defined functions, there is no recursion allowed.
+This means that all function definitions can be inlined into the user query with guaranteed termination.
+When converting to the intermediate language, we assume that this inlining has occurred.
+
+\TODO{Should we go through the evaluation rules here, or after typechecking?}
+If we can write the evaluation rules to be ``agnostic'' of the type, it would be ideal to put them here.
+
 \input{figures/Source-Grammar.tex}
+
+\input{figures/Source-Eval.tex}

### 2015-12-05 2016icicle/section/03-Types.tex
ok
nothing much happening
changes

@@
 
-For (TcFold), the form of fold is $@fold@~v~@=@~x@;@~x'$ where $v$ is the name of the fold binding, $x$ is its initial value, and $x'$ computes the successor value, based on the previous value of $v$.
+For (TcFold), the form of fold is $@fold@~v~@=@~x~@then@~x'$ where $v$ is the name of the fold binding, $x$ is its initial value, and $x'$ computes the successor value, based on the previous value of $v$.
 The initial expression $x$ must be @Pure@ of some type $\tau$, and the successor expression $x'$ is typechecked with the fold binding set to $@Element@~\tau$ in the environment.
@@
 
-For (TcLatestElement), if you have a $@latest@~n$ of @Element@s we fill a circular buffer with the last $n$ values.
-Similarly, (TcLatestAggregate) uses a circular buffer, and then applies the aggregate to the last $n$ values.
-
 Finally, (TcGroup) requires its key or group ``by'' to be an @Element@, and the value to be an @Aggregate@.
 It builds up a map from key to value, repeatedly applying the aggregate as new values are found.
+
+\input{figures/Source-Compile.tex}

### 2015-12-05 2016icicle/section/XX-Related-DataFlow.tex
changes

@@
 While this may be the desired behaviour in some cases, we believe this decision must be made explicit.
-At the moment this must be hand-written as a fold, but we intend to add a new primitive called @running@ which peeks into a fold and uses its intermediate result.
+At the moment this must be hand-written as a fold, but we intend to add a new primitive called @scan@ which peeks into a fold and uses its intermediate result.
 \begin{code}
-   filter value > running (mean value)
+   filter value > scan (mean value)
 -> count

### 2015-12-05 2016icicle/section/02-Source.tex
changes

@@
 This, combined with the lack of lambda construction, means that higher order functions are outlawed.
-Outlawing higher order functions is somewhat unsatisfying, but simplifies the typing rules and makes it easier to support our performance guarantees.
+While a first-order language (one lacking higher order functions) offers somewhat less abstraction, it simplifies the typing rules and makes it easier to support our performance guarantees.
 

### 2015-12-05 2016icicle/section/03-Types.tex
changes

@@
 %%% Expressions
-With function application already defined, typechecking expressions in figure~\ref{fig:source:type:exp} is quite simple.
-The judgment $\Typecheck{\Gamma}{x}{\tau}$ means that under a context $\Gamma$, the expression $x$ has type $\tau$.
+With function application already defined, typechecking expressions in figure~\ref{fig:source:type:ctx} becomes quite standard.
+The judgment $\Typecheck{\Gamma}{x}{\tau}$ means that under a context $\Gamma$, the expression $x$ has type and modality $\tau$.
+Being a first-order language, the typing rule returns a first-order type rather than a function type.
+
 Rules (TcVar) and (TcPrim) simply look up the variable or primitive in the environment.
+
 For (TcApp) we simply use the boxed function application judgment from figure~\ref{fig:source:type:wrap}.
@@
 \TODO{I think we need a box/cast rule for going from @Pure@ to @Element@ etc}
-
-%%% Contexts
-Figure~\ref{fig:source:type:ctx} shows the typing rules for contexts.
-The judgment $\Typecheck{\Gamma}{c}{\tau}$ means that under a context $\Gamma$, the context $c$ has type $\tau$.
 For simplicity, we reuse the same judgment for expressions and contexts.

### 2015-12-05 2016icicle/section/02-Source.tex
ok

@@
 
+This language is not referentially transparent, which may seem odd, but this actually allows us to restrict the language and uphold our performance guarantees.
+The input stream is implicitly passed as an argument to all parts of the program, but cannot be referred to explicitly.
+Additionally, operations like @filter@ create a new context where the input stream is modified.
+This means that a @let@ binding outside of a @filter@ can have different semantics when moved inside the @filter@, despite having the same body.
+However, by passing this argument implicitly we remove a class of programs that would be hard to guarantee performance for.
+By prohibiting the programmer from accessing the underlying stream, they cannot perform certain operations like zipping two streams with different filters together.
+
 We formally define the grammar of the language in figure~\ref{fig:source:grammar}.

### 2015-12-05 2016icicle/section/03-Types.tex
ok

@@
 
+We will now formally describe the typesystem of Icicle, and show how modality types are used to enforce the single-pass restriction.
+
 \input{figures/Source-Type.tex}
 
-%%% Types
+\subsection{Types}
 In figure~\ref{fig:source:type:defs} we define the types.
@@
 
-%%% Function application
+\subsection{Function application with unboxing}
 In figure~\ref{fig:source:type:wrap} there are two judgments defined, which are both used for the type of function application.
 In contrast to traditional modality type systems, we infer the boxing and unboxing of various modalities.
+\TODO{Citation and find something similar.}
+\TODO{Probably talk about comonads or something}
 
@@
 
-The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to reboxing the function return type $\tau$ into modality $m$, returning $\tau'$ as the boxed return type.
+These function application judgments only deal with the unboxing case; boxing is dealt with later as the single rule (TcBox).
+
+The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to unboxing the a function argument of type $m$ and reboxing the function return type $\tau$ into modality $m$, with $\tau'$ as the boxed return type.
 The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
@@
 For (WrapPure) the result type is a @Pure@, so it can be boxed to any mode.
+If these rules do not apply, for example when an @Aggregate@ result is attempted to be reboxed as an @Element@, the function application is an error.
 
-The judgment $\WrapApp{\tau}{\phi}{\tau'}$ attempts to apply the function type $\tau$ to argument $\phi$, performing boxing and unboxing if necessary.
-If both modalities and types are equal, (WrapApp) applies the function as normal.
-When the function expects a @Pure@ argument and is passed an @Element@ or an @Aggregate@, we perform reboxing as in (WrapAppElement) and (WrapAppAggregate).
-Finally, no matter what the function expects, if it is passed a @Pure@ it is simply boxed and applied as in (WrapAppPure), with no need to rebox the result.
+The judgment $\WrapApp{\tau}{\phi\ldots}{\tau'}$ attempts to apply the function type $\tau$ to the argument list $\phi\ldots$, performing unboxing as necessary.
+When the argument list is empty and the function type is a nullary function or simple value type, (WrapAppFinished) simply returns the result type.
 
-\TODO{Or should we just use an explicitly box/unbox version?}
+In the next rules, we use the syntax $a;~b$ for a list with $a$ as its head and $b$ as its tail.
 
-%%% Expressions
+When the actual argument and expected function argument modalities and types are equal, (WrapAppEqual) applies the function as normal, and continues along with any remaining arguments.
+
+When the function expects a @Pure@ argument and is passed an @Element@ or an @Aggregate@, we perform reboxing as in (WrapAppPure).
+This attempts to rebox the function result to the argument modality, and then continues by wrapping the remaining arguments.
+
+\subsection{Expression}
 With function application already defined, typechecking expressions in figure~\ref{fig:source:type:ctx} becomes quite standard.
@@
 Rules (TcVar) and (TcPrim) simply look up the variable or primitive in the environment.
+While the environment contains function types $T_\to$, these rules only apply if the result is a nullary function with no arguments.
 
-For (TcApp) we simply use the boxed function application judgment from figure~\ref{fig:source:type:wrap}.
+The rule (TcBox) allows any @Pure@ computation to be implicitly cast to another mode, such as @Element@ or @Aggregate@.
+This is equivalent to boxing the expression.
 
-\TODO{I think we need a box/cast rule for going from @Pure@ to @Element@ etc}
-For simplicity, we reuse the same judgment for expressions and contexts.
+There are two different application rules (TcAppVar) and (TcAppPrim), which simply look up the function type in different environments before using the boxed function application judgement from figure~\ref{fig:source:type:wrap}.
+In these ruler we use ``$\ldots$'' as shorthand for finding the types of all arguments.
 
@@
 
-\input{figures/Source-Compile.tex}
+\subsection{Single-pass restriction}
+
+We now describe how this typesystem encodes the single-pass restriction: that is, that if a given query type-checks, it only needs to look at each element in the stream once.
+
+First, look at an example that would require two passes over the data: maybe @filter open > mean open in count@.
+In this case, $\mit{open}$ has type $\Elm{\NN}$ while $\mit{mean~open}$ has type $\Agg{\NN}$.
+The primitive $>$ has type $\NN \to \NN \to \NN$ and can be boxed to either @Element@ or @Pure@, but not both.
+This means that $\mit{open} > \mit{mean open}$ is a type error.
+
+In Icicle, folding and boxing pure values are the only ways to produce @Aggregate@ types.
+The other typing rules that refer to @Aggregate@s require an existing @Aggregate@ and do not produce an aggregate on their own.
+This means that all @Aggregate@s must either be the result of folds, which in general require all their input before they can be computed, or pure computations.
+
+There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@, except for the @scan@ primitive which explicitly uses a fold's intermediate value, rather than the value of the entire pass.
+
+The only way to require multiple passes over the data, therefore, is to have a @Pure@ or @Element@ computation that depends on the result of an @Aggregate@ computation.
+And so on.

### 2015-12-05 2016icicle/section/04-Core.tex
ok

%!TEX root = ../Main.tex
\section{Core language}
\label{s:Core}

In figure~\ref{fig:core:grammar} we introduce the intermediate language.
The top-level of this language is $\mit{Program}$ with four stages of computation: pure computations, followed by stream transformers, then folds, and finally extracting the result of the folds.

The program definition may seem more complicated than necessary, as all programs could be expressed as a single fold and an extract, but separating these out turns out to be useful for efficiency.
This will be expanded upon later.

The expression type for Core is much simpler than Source, $\mit{CoreExp}$ having just names, primitive application and let bindings.
More primitives are required for the Core language, as some filters in the Source language are converted to @If@s, and groups are converted to use map primitives such as lookup and insert.
The @scan@ primitive is no longer used in the Core language as it is just converted to a variable reference.

\input{figures/Core-Grammar.tex}

\input{figures/Core-Compile.tex}

### 2015-12-05 2016icicle/section/04-Core.tex
ok

@@
 
+This section presents the Core language, which we use as an intermediate language before converting to imperative code.
+\TODO{Do we need formal semantics and type system?}
+
+\subsection{Grammar}
+
+\input{figures/Core-Grammar.tex}
+
 In figure~\ref{fig:core:grammar} we introduce the intermediate language.
 The top-level of this language is $\mit{Program}$ with four stages of computation: pure computations, followed by stream transformers, then folds, and finally extracting the result of the folds.
-
-The program definition may seem more complicated than necessary, as all programs could be expressed as a single fold and an extract, but separating these out turns out to be useful for efficiency.
-This will be expanded upon later.
+Each stage of computation can only refer to the stages before it.
 
@@
 
-\input{figures/Core-Grammar.tex}
+The first section of the program is the @Befores@ section, which is just binding names to pure expressions.
+These expressions cannot refer to any part of the stream, any computation, nor the results of any folds.
+
+Next are @Streams@: streams and stream transformers.
+The first stream type is @Source@, which is simply the unmodified input stream.
+The two stream transformers take a function and the name of the input stream.
+These are @Map@ which applies a function to each element, and @Filter@ which filters out according to a predicate.
+Neither of the worker functions can refer to any streams, only the bindings in the @Befores@ section.
+All of these stream transformers can be computed \emph{on-line} and by looking at a single element at a time, without any sort of accumulator.
+
+In the third section are the @Folds@, which perform some reduction over an input stream.
+Each fold has a constructor function which is successively applied to the latest accumulator and each stream value; an initialiser or zero value for the accumulator; and the name of the input stream.
+The initialiser is pure, and so can only refer to variables defined in @Befores@.
+The constructor function can refer to @Befores@, previous @Folds@, but not directly to the @Streams@.
+Referring to previous @Folds@ in the constructor body gives the latest value of the accumulator, this is used to implement the @scan@ primitive.
+
+Then the @Extracts@ are bound which can refer to the results of @Folds@ and any @Befores@.
+These are used to perform and post-computation on the folds which cannot be done at each step, for example dividing the sum and the count to get the mean.
+
+Finally, the @Outputs@ are defined, which define which variables are to be returned as the result of the query.
+These can refer to @Befores@, @Folds@ and @Extracts@.
+
+\subsection{Fusion}
+
+When fusing together two queries, we can simply give each unique names, then append each respective section together; the @Befores@ appended to the @Befores@, @Streams@ to @Streams@ and so on.
+After appending them together, we remove duplicate bindings by iterating through each binding set and comparing it to the previous bindings in the set.
+If a duplicate is found, we remove the later binding and change all references in the rest of the program to the earlier binding.
+
+The program definition is more complicated than necessary, as all programs of this form can actually be expressed as a single fold and an extract.
+However, having smaller components makes finding duplicate computations after fusion easier.
+For example, if mean is expressed as two folds (sum and count) then removing the duplicate after fusing with another count is rather simple.
+If mean were expressed as a single fold containing both sum and count, removing the duplicate count would require somehow splitting apart the sum and count or removing duplicates some other way.
+
+\subsection{Conversion from Source}
+
+Yes yes convert from Source fix that
 

### 2015-12-06 2016icicle/section/02-Source.tex
ok

@@
 
-\TODO{Should we go through the evaluation rules here, or after typechecking?}
-If we can write the evaluation rules to be ``agnostic'' of the type, it would be ideal to put them here.
+\subsection{Evaluation}
 
@@
 
+In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.
+We use the shorthand $\ov{\Gamma}$ to mean a list of $\Gamma$, and this same overline syntax is used as a variable containing a list.
+The judgment $\BigstepG{x}{\ov{v}}{v}$ uses a list $\ov{\Gamma}$, with one environment for each element of the input stream.
+The environment $\Gamma$ is used for scalar variables that are not associated with any particular input element.
+Next is the expression $x$ that is to be evaluated.
+The last two are outputs $\ov{v}$ and $v$.
+The output $\ov{v}$ is the intermediate result of the computation for each part of the input stream.
+That is, each element of $\ov{v}$ is the result of evaluating $x$ on the input stream up to that element.
+The final output, $v$, is the result at the end of the stream.
+In most cases it is the same as the last element of $\ov{v}$, unless $\ov{v}$ is empty or for a variable lookup.
+
+The reason for returning the result list $\ov{v}$ is to make implementing the @scan@ primitive easier.
+This allows @scan@ itself to be evalauted effectively as a no-op, but does mean that each other rule has to implement its own @scan@ as well.
+
+\TODO{It seems like we should be able to get rid of the single element return $v$, but then would it be strange with the return value being of length $max~1~|\ov{v}|$?}
+
+When evaluating on an input such as our @stocks@ table, we would create an environment for each row, with open and close as the variables.
+The rule (EvTop) simply uses the empty environment for the scalar environment, and ignores the stream output.
+
+For variables, the rules (EvVar) and (EvVarInput) look up the variable in the contexts.
+In the case of direct inputs as introduced by (EvTop), there is no corresponding entry in the scalar environment, so (EvVarInput) is used which returns an error as the scalar output.
+In this way, if the input stream is used as an aggregate it will be an evaluation error, but if it is used as an element value and folded upon, the error value will be thrown away.
+For normal variables, there should be a binding in all environments.
+
+Let bindings are handled by rule (EvLet), which evaluates its definition, adds part of the stream result to each stream environments, and adds the scalar result to the scalar environment.
+
+For the rule (EvFold), first the zero or initial value of the fold is computed, using only its scalar value.
+The extra rules (ScanNil) and (ScanCons) are then used to repeatedly apply the fold computation to each stream input, starting from the initial value.
+The list of fold values are added to the stream environments, and the last of the fold values, or the initial value if none, is added to the scalar environment.
+
+In (EvFilter) the predicate is evaluated to a list of results. 
+The stream environments are filtered according to this predicate list using $\mit{pack}$, and the rest of the expression is computed.
+However, the result list of the rest of the expression is only the scan of the filtered stream inputs.
+We use $\mit{extend}$ to add back parts of the stream that have been filtered out, but must first find the initial value of the scan to start from.
+
+The primitive operator rules (EvAdd) and (EvGt) apply their operator to each element of the inputs.
+The primitive (EvNat) returns a list of constant numbers, one for each stream element.
+Finally, (EvScan) is a no-op since the hard work of computing the @scan@s is actually implemented in the rest of the rules.
+
 \input{figures/Source-Eval.tex}

### 2015-12-15 2016icicle/Main.tex
ok
Some stuff

@@
 \usepackage{float}
-\usepackage{style/utils}
 \usepackage{style/code}
 \usepackage{style/proof}
+\usepackage{style/utils}
 \usepackage{style/judgements}
@@
 We introduce a type system based on modality types to enforce this single pass restriction.
-By using an appropriate intermediate language we can fuse multiple independent queries together, before extracting efficient single-loop C code.
+By using an appropriate intermediate language we guarantee fusion between queries on the same table, before extracting efficient single-loop C code.
 \end{abstract}
@@
 \input{section/03-Types.tex}
+\input{section/04-Core.tex}
+\input{section/05-Generation.tex}
+\input{section/06-Conclusion.tex}
 
@@
 \section*{Acknowledgements}
-I would like to thank all of Ambiata, particularly Huw Campbell and Mark Hibberd.
+must be anonymous
 

### 2015-12-15 2016icicle/section/01-Introduction.tex
changes
Some stuff

@@
 
-With a language like SQL, it may not be immediately obvious how many passes a query requires, and it may depend on particular compiler optimisations and the mood of the database.
-For example, how many iterations do the following queries require?
-
-\begin{code}
-SELECT open - MIN(open)    FROM stocks;
-SELECT open - MIN(open, 1) FROM stocks;
-SELECT        MIN(open)    FROM stocks;
-SELECT open,  MIN(open)    FROM stocks;
-\end{code}
-
+With a language like SQL, it may not be immediately obvious how many passes a query requires, and it may depend on particular compiler optimisations and the statistics and settings of the database.
 For our purposes, we would rather stricter compile-time guarantees: if a query typechecks and compiles, it should be able to be executed efficiently.
@@
 By making use of these restrictions, we are able to generate more efficient code than a general query language.
-
-\TODO{Look into data flow languages like Lucid and how they enforce causality, cite them}
 

### 2015-12-15 2016icicle/section/02-Source.tex
Some stuff

@@
 
-We present a cut down version of our implementation; our implementation has many more primitives, case expressions, as well as primitive contexts for windowing based on time, folding over groups, and so on.
+In this section we describe the grammar and evaluation semantics of the Icicle language.
 
-The first example defines $\mi{sum}$ as a fold over the input data.
-The initial value for the fold is $0$, followed by the previous value $s$ added to the current value $v$.
+Example data
+\begin{code}
+CODE, OPEN, CLOSE, DATE
+ABC,    10,    12, 2015-01-01
+ABC,    12,    14, 2015-01-02
+IAG,    10,    12, 2015-01-01
+IAG,    12,    14, 2015-01-02
+\end{code}
+
+Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
+Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
+We define the stocks table as follows, leaving out the company name (entity identifier) and date.
 
@@
 MM \= MM \= MMMMMMMMMMM \= \kill
-$\mi{sum}~v$    \\
-\> $=$  \> $@fold@~s~=~0~@then@~s~+~v$ \\
-\> @in@ \> $s$ \\
+$@table@~\mi{stocks}$    \\
+$\{$ \> $\mi{open}$ \> $:~\NN$ \\
+$~,$ \> $\mi{close}$ \> $:~\NN$ \\
+$\}$
 \end{tabbing}
 
+This defines the columns as natural numbers, and allows them to be referred to by name.
+In the definitions and queries below, any reference to these names means the column itself, with the entire data for a given entity.
+This is annotated with the mode @Element@, meaning that it exists for every element of the input.
+
+We now define some useful functions and other definitions that can be used in the queries themselves.
+
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$@definitions@$ \\
+$\{$
+\end{tabbing}
+
+First we define a function to compute the sum of some input data.
+While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
+The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
+This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
+The second parameter is the actual value to sample at the clock.
+In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
+
+By separating the clock from the element values, we guarantee that any references to elements denote the same point in the stream, and can always be used together.
+For example, computing the difference between the two fields above ($\mi{open} - \mi{close}$) refers to the open and close fields at the same point.
+The clock is used when performing a fold, and determine at which points to sample the elements and update the accumulator.
+This is in constrast to traditional streaming dataflow languages, where the element data and the clock are intertwined.
+This requires clock analysis\CITE to tell whether two element streams can be used together with a bounded buffer.
+
+We denote clocks in brackets as they are a different universe to normal expressions.
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+$\mi{sum}[c]~v$    \\
+\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$ \\
+\> @in@ \> $s$. \\
+\end{tabbing}
+
+The result of $\mi{sum}$ above has type $\Agg{\NN}$.
+Aggregate types denote that their value is only available after the entire input stream has been seen.
+By restricting the bodies of folds to be element types, and their output to be aggregate types, it means no reference to the result of folds can be made in the body of a fold.
+
 Next, we can define $\mi{count}$ and $\mi{mean}$ in terms of this sum.
+For 
 \begin{tabbing}
@@
 $\mi{count}$                                        \\
- \> $=$  \> $\mi{sum}~1$                            \\
+ \> $=$  \> $\mi{sum}~1$.                           \\
                                                     \\
 $\mi{mean}~v$                                       \\
- \> $=$  \> $\mi{sum}~v~/~\mi{count}$               \\
+ \> $=$  \> $\mi{sum}~v~/~\mi{count}$.              \\
 \end{tabbing}

### 2015-12-16 2016icicle/section/02-Source.tex
running example and work on explicitly clocked syntax

@@
 
-Example data
-\begin{code}
-CODE, OPEN, CLOSE, DATE
-ABC,    10,    12, 2015-01-01
-ABC,    12,    14, 2015-01-02
-IAG,    10,    12, 2015-01-01
-IAG,    12,    14, 2015-01-02
-\end{code}
-
 Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
 Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
-We define the stocks table as follows, leaving out the company name (entity identifier) and date.
+We define the stocks table as follows, leaving out the company code (entity identifier) and date.
 
@@
 We now define some useful functions and other definitions that can be used in the queries themselves.
-
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-$@definitions@$ \\
-$\{$
-\end{tabbing}
 
@@
 MM \= MM \= MMMMMMMMMMM \= \kill
+@function@ 
 $\mi{sum}[c]~v$    \\
@@
 
-Next, we can define $\mi{count}$ and $\mi{mean}$ in terms of this sum.
-For 
+We can define a query that computes the sum of each company's open prices.
+Query definitions are similar to function definitions, except they only take one argument: the clock describing all available input.
+They also must return aggregates, as an element query could be unbounded in size.
 \begin{tabbing}
 MM \= MM \= MMMMMMMMMMM \= \kill
-$\mi{count}$                                        \\
- \> $=$  \> $\mi{sum}~1$.                           \\
-                                                    \\
-$\mi{mean}~v$                                       \\
- \> $=$  \> $\mi{sum}~v~/~\mi{count}$.              \\
+@query@ 
+$\mi{opensum}[c]$    \\
+\> $=$  \> $\mi{sum}[c]~\mi{open}$. \\
 \end{tabbing}
 
-We can use @filter@ to count only the days where the close price is higher than the open.
+We can run $\mi{opensum}$ over some example data.
+This example data has the company code and date for each entry, as well as the stock's open and close prices.
+Note that the sum only becomes available at the end of each entity, and only applies to a particular entity's elements.
+
+\begin{code}
+CODE, OPEN, CLOSE, DATE,        OPENSUM
+ABC,    10,    12, 2015-01-01,       10
+IAG,    11,    12, 2015-01-01
+IAG,    12,    10, 2015-01-02,       23
+\end{code}
+
+Next, we can define functions $\mi{count}$ and $\mi{mean}$ in terms of $\mi{sum}$.
+Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.
+The input clocks are passed along as-is.
 \begin{tabbing}
 MM \= MM \= MMMMMMMMMMM \= \kill
-$\mi{gap}$                                          \\
- \> $=$  \> $@filter@~\mi{close}~>~\mi{open}$       \\
- \> @in@ \> $\mi{count}$                            \\
+@function@ 
+$\mi{count}[c]$                                     \\
+ \> $=$  \> $\mi{sum}[c]~1$.                           \\
                                                     \\
-$\mi{proportion}$                                   \\
- \> $=$  \> $\mi{gap}~/~\mi{count}$
+@function@ 
+$\mi{mean}[c]~v$                                       \\
+ \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
 \end{tabbing}
 
-Finally, we can use @group@ to put into buckets based on the percentage of growth.
-Here the key for the map is the percentage of growth, and the value is the number of days with that amount of growth, divided by the total number of days.
+Clocks can be reduced to filter out certain elements using @where@.
+In this query, we use @where@ to count the number of days where the close price is higher than the open price.
 \begin{tabbing}
 MM \= MM \= MMMMMMMMMMM \= \kill
-$\mi{growthPercent}$                                        \\
- \> $=$  \> $((\mi{close}-\mi{open}) \times 100 ) / \mi{open}$  \\
-                                                            \\
-$\mi{growthBy}$                                             \\
- \> $=$  \> $@let@~\mi{total} = \mi{count}$                 \\
- \> @in@ \> $@group@~\mi{growthPercent}$                    \\
- \> @in@ \> $\mi{count} / \mi{total}$                    \\
+@query@ 
+$\mi{gap}[c]$                                        \\
+ \> $=$  \> $\mi{count}[c~@where@~\mi{close}~>~\mi{open}]$.       \\
 \end{tabbing}
+
+In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.
+\begin{code}
+CODE, OPEN, CLOSE, CLOSE > OPEN, GAP
+ABC,    10,    12,            T,   1
+IAG,    11,    12,            T
+IAG,    12,    10,            F,   1
+\end{code}
 
@@
 \TODO{Talk about scan}
-
-This language is not referentially transparent, which may seem odd, but this actually allows us to restrict the language and uphold our performance guarantees.
-The input stream is implicitly passed as an argument to all parts of the program, but cannot be referred to explicitly.
-Additionally, operations like @filter@ create a new context where the input stream is modified.
-This means that a @let@ binding outside of a @filter@ can have different semantics when moved inside the @filter@, despite having the same body.
-However, by passing this argument implicitly we remove a class of programs that would be hard to guarantee performance for.
-By prohibiting the programmer from accessing the underlying stream, they cannot perform certain operations like zipping two streams with different filters together.
 

### 2015-12-18 2016icicle/section/02-Source.tex
working on evaluator

@@
 
-\TODO{Talk about scan}
-
 We formally define the grammar of the language in figure~\ref{fig:source:grammar}.
@@
 We use the shorthand $\ov{\Gamma}$ to mean a list of $\Gamma$, and this same overline syntax is used as a variable containing a list.
-The judgment $\BigstepG{x}{\ov{v}}{v}$ uses a list $\ov{\Gamma}$, with one environment for each element of the input stream.
+The judgment $\Bigstep{x}{\ov{v}}{v}$ uses a list $\ov{\Gamma}$, with one environment for each element of the input stream.
 The environment $\Gamma$ is used for scalar variables that are not associated with any particular input element.
@@
 In most cases it is the same as the last element of $\ov{v}$, unless $\ov{v}$ is empty or for a variable lookup.
-
-The reason for returning the result list $\ov{v}$ is to make implementing the @scan@ primitive easier.
-This allows @scan@ itself to be evalauted effectively as a no-op, but does mean that each other rule has to implement its own @scan@ as well.
 

### 2016-01-04 2016icicle/section/02-Source.tex
Add fold rule
eval notes

@@
 $\mi{sum}[c]~v$    \\
-\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$ \\
-\> @in@ \> $s$. \\
+\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
 \end{tabbing}
@@
 In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.
-We use the shorthand $\ov{\Gamma}$ to mean a list of $\Gamma$, and this same overline syntax is used as a variable containing a list.
-The judgment $\Bigstep{x}{\ov{v}}{v}$ uses a list $\ov{\Gamma}$, with one environment for each element of the input stream.
-The environment $\Gamma$ is used for scalar variables that are not associated with any particular input element.
-Next is the expression $x$ that is to be evaluated.
-The last two are outputs $\ov{v}$ and $v$.
-The output $\ov{v}$ is the intermediate result of the computation for each part of the input stream.
-That is, each element of $\ov{v}$ is the result of evaluating $x$ on the input stream up to that element.
-The final output, $v$, is the result at the end of the stream.
-In most cases it is the same as the last element of $\ov{v}$, unless $\ov{v}$ is empty or for a variable lookup.
+The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.
+The evaluation works on a single entity's data at a time, and the column values, such as $\mi{open}$ and $\mi{close}$ in the examples above, must be substituted into the functions and expression beforehand.
+The clock is initially set to true for all values in the input stream.
 
-\TODO{It seems like we should be able to get rid of the single element return $v$, but then would it be strange with the return value being of length $max~1~|\ov{v}|$?}
+Substitution is denoted by $x[n~:=~v]$, where $x$ is the expression to substitute into, $n$ is the variable name to replace, and $v$ is the replacement payload.
 
-When evaluating on an input such as our @stocks@ table, we would create an environment for each row, with open and close as the variables.
-The rule (EvTop) simply uses the empty environment for the scalar environment, and ignores the stream output.
+For example, for the $\mi{opensum}$ query we would have two evaluations:
+\begin{tabbing}
+MMMM \= M \= \kill
+$\mi{clock}_1$ \> $=$ \> $@Elements@~[@T@]$ \\
+$\mi{open}_1$  \> $=$ \> $@Elements@~[10]$ \\
+$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_1,~\mi{open}~:=~\mi{open}_1]}{@Scalar@~10} $ \\
+\\
+$\mi{clock}_2$ \> $=$ \> $@Elements@~[@T@,~@T@]$ \\
+$\mi{open}_2$  \> $=$ \> $@Elements@~[11,~12]$ \\
+$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_2,~\mi{open}~:=~\mi{open}_2]}{@Scalar@~23} $
+\end{tabbing}
 
-For variables, the rules (EvVar) and (EvVarInput) look up the variable in the contexts.
-In the case of direct inputs as introduced by (EvTop), there is no corresponding entry in the scalar environment, so (EvVarInput) is used which returns an error as the scalar output.
-In this way, if the input stream is used as an aggregate it will be an evaluation error, but if it is used as an element value and folded upon, the error value will be thrown away.
-For normal variables, there should be a binding in all environments.
+In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
+For example, the definition of $\mi{sum}$ would actually have $@Scalar@~0$ as its initialiser.
+Scalars can also be thought of as a repeated stream of the same value, of the same length as the input stream.
 
-Let bindings are handled by rule (EvLet), which evaluates its definition, adds part of the stream result to each stream environments, and adds the scalar result to the scalar environment.
+The first rule (EvValue) simply returns its value unchanged.
 
-For the rule (EvFold), first the zero or initial value of the fold is computed, using only its scalar value.
-The extra rules (ScanNil) and (ScanCons) are then used to repeatedly apply the fold computation to each stream input, starting from the initial value.
-The list of fold values are added to the stream environments, and the last of the fold values, or the initial value if none, is added to the scalar environment.
+The next two rules, (EvPrimScalar) and (EvPrimElements) are for primitive binary operators (as the only primitives are binary).
+When both arguments are scalars, we simply apply the operator to both arguments.
+If both arguments are element streams, we pair the streams together and apply the operator to each pair.
+If only one of the arguments is an element stream, the other argument must be a scalar, and the function $\mi{elements}$ is used to extend the scalar to a repeated stream before applying the operator.
 
-In (EvFilter) the predicate is evaluated to a list of results. 
-The stream environments are filtered according to this predicate list using $\mit{pack}$, and the rest of the expression is computed.
-However, the result list of the rest of the expression is only the scan of the filtered stream inputs.
-We use $\mit{extend}$ to add back parts of the stream that have been filtered out, but must first find the initial value of the scan to start from.
+Next, (EvFunApp) performs named function application.
+The function environment $f$ is searched for a function named $n$, and we find the names of the arguments $\mi{args}$ and the function definition $\mi{exp}$.
+The function definitions are non-recursive, and the function definition can only refer to functions defined before it.
+We use $f'$ as the new list of available function bindings.
+All arguments are evaluated to values, then each argument value is substituted into the function body, using the names of the arguments.
 
-The primitive operator rules (EvAdd) and (EvGt) apply their operator to each element of the inputs.
-The primitive (EvNat) returns a list of constant numbers, one for each stream element.
-Finally, (EvScan) is a no-op since the hard work of computing the @scan@s is actually implemented in the rest of the rules.
+For (EvLet) the let-definition is evaluated, then substituted into the let-body.
+
+For (EvWhere), used in clocks, both arguments are streams of booleans.
+We evaluate both arguments and perform a pairwise and of the two, where the result clock is only true for an element if it is true for both inputs on that element.
+
+The final evaluation rule, (EvFold), is the most important and most complex.
+It evaluates a reduction over the stream.
+First the clock and the initial value are evaluated, then a secondary judgment form is used to evaluate the fold.
+
+The fold judgment form takes a list of functions, the index of the current fold value, the stream of boolean clock values, the non-empty list of fold results so far, and the expression to evaluate.
+For (EvFoldNil) the clock is finished, so the last value in the fold result must be the final fold value.
+
+For the other cases, we inspect the first element of the clock stream.
+If the clock is true, we perform the fold with the fold results so far, and compute the rest of the clock stream.
+If the clock is false, we reuse the last value in the fold results, as any elements for this index will not be used.
+
+For (EvFoldClockTrue) we evaluate $k$ with the fold results so far substituted into the accumulator as an element stream.
+The result will be an element stream of at least length $i$, as the result of the fold so far.
+We then get the result at index $i$ and add it as the latest accumulator value, and continue computing the rest of the stream with this new accumulator.
+
+For (EvFoldClockFalse) we can re-use the previous accumulator value, which is the last value of the fold results.
+We add the last fold result to the accumulator, and compute the rest of the stream.
 

### 2016-02-03 2016icicle/section/01-Introduction.tex
Add notes

@@
 
+\ben{What is a concrete example of an SQL query where it's not immediatly obvious how many passes through the data it needs?
+We've claimed that for some queries the number of required passes is not ``immediately obvious'', and now need to back that up}
+
 For feature generation for machine learning, our queries have some quite specific restrictions.
+\ben{The following describes properties that that queries and data have intrinsically, not \emph{restrictions} placed on them}.
 Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.

### 2016-02-03 2016icicle/section/02-Source.tex
type stuff
Add notes

@@
 In this section we describe the grammar and evaluation semantics of the Icicle language.
+\ben{The running example here should be driven by solving the stated problem introduced in the introduction.}
 
 Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
+\ben{Why does the fact that it's a date matter?
+Could we also use any other data type that has an ordering?}
+
 Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
@@
 
+\ben{If this table really has a data field as well then why do we leave it implicit? That's a bit confusing}.
 \begin{tabbing}
@@
 This is annotated with the mode @Element@, meaning that it exists for every element of the input.
+\ben{What is annotated with the mode? The column itself or the type of the column?}
 
@@
 In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
+\ben{Maybe hold off the description of @Element@ and @Aggregate@ until later so we can introduce them both at the same time. At this point in the text it's not clear why we need the modal types.}
 
@@
 This requires clock analysis\CITE to tell whether two element streams can be used together with a bounded buffer.
+\ben{The need for clock analysis needs to be backup up with an example.
+Readers that aren't familiar with the semantics of data flow languages won't understand the phrase ``the element and the clock are intertwined''}
 
 We denote clocks in brackets as they are a different universe to normal expressions.
+\ben{Explain the syntax. What are $v$ $s$ and @then@ about?} \ben{Mention that $s$ is a stream. What is its type? Where are $c$ and $s$ bound?}
+
 \begin{tabbing}
@@
 @function@ 
-$\mi{sum}[c]~v$    \\
+$\mi{sum}[c]~(v~:~\NN)$    \\
 \> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
@@
 The result of $\mi{sum}$ above has type $\Agg{\NN}$.
+
+\ben{Does the concrete syntax support this annotation, eg:}
+\begin{tabbing}
+MM \= MM \= MMMMMMMMMMM \= \kill
+@function@ 
+$\mi{sum}[c]~(v~:~\NN) : \Agg{\NN}$    \\
+\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
+\end{tabbing}
+
 Aggregate types denote that their value is only available after the entire input stream has been seen. 
 By restricting the bodies of folds to be element types, and their output to be aggregate types, it means no reference to the result of folds can be made in the body of a fold.
+\ben{Give a counter-example of what can go wrong it we don't have this restriction}
 
@@
 They also must return aggregates, as an element query could be unbounded in size.
+\ben{The above describes the syntax of the query, but what is the essential reason for defining functions and queries with different language constructs?
+Could we just say that a query is a function with some particular restrictions?}
+\ben{Can we also provide the aggregate result type on the query definition itself?}
+
 \begin{tabbing}
@@
 \end{code}
+\ben{The fact that there is also an implicit group-by operation being performed is not made explicit in the query code.
+The operation being performed above is a segmented fold over the data, rather than a complete fold.
+Suppose there was also a @VOLUME@ field for the number of shares traded per day for each stock, and we wanted to compute the total volume of all shares traded on every day we have on record?
+How would we express that in the query language?}
 
@@
 Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.
+
+\ben{It's interpreted as an element containing all ones, but not a \emph{stream} containing all ones.
+The fact that the input data is implicitly partitioned by the key need to be stated up front, at it seems like a key part of the language design.
+Suppose we go back to the @VOLUME@ example and want to broadcast the value @1@ to \emph{all} the rows in the entire data set.
+How would we do that?}
+
 The input clocks are passed along as-is.
@@
 @function@ 
-$\mi{mean}[c]~v$                                       \\
+$\mi{mean}[c]~(v~:~\NN)$                                       \\
  \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
@@
 In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.
+
+\ben{That's not what \emph{gap} means.
+A stock price has ``gapped up'' when it closes at some price $c$ on day 0, opens at some price $o$ on day 1, and $o > c$.
+The point is that there is a price discontinuity between trading days, and there was no opportunity to buy and sell at any price between $c$ and $o$.}
+
 \begin{code}
@@
 \input{figures/Source-Grammar.tex}
+\ben{Should the set of types also include @Date@? In the first example the table also contained dates}.
 
 In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.
+
 The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.
+
+\ben{The previous section states that ``all function definitions can be inlined into the user query''.
+If this is true then why not always inline them?
+Why does the evaluation semantics need an environment containing a separate set of functions?}
+
 The evaluation works on a single entity's data at a time, and the column values, such as $\mi{open}$ and $\mi{close}$ in the examples above, must be substituted into the functions and expression beforehand.
+\ben{What is a ``column value''?}
 The clock is initially set to true for all values in the input stream.
@@
 
-In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
+\ben{The syntax @Elements@ looks similar to @Element@, which was a type.
+Can we usual list or set syntax to present these?} \ben{I think it would look cleaner to have an overall construct to express the combination of input data expression to execute.
+The examples feel like they're setting up the big-step judgement in a particular way to get around the lack of an overall program/module construct.}
+
+\ben{I don't follow the following paragraph.} In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
 For example, the definition of $\mi{sum}$ would actually have $@Scalar@~0$ as its initialiser.

### 2016-02-03 2016icicle/section/03-Types.tex
type stuff
Add notes

@@
 The @Pure@ modality is actually the absence of a mode; this is simply made explicit for clarity.
+\ben{Standard modal logics use $P$, $\Diamond\,P$ and $\Box\,P$. A proposition $P$ which is universaly true does not have an associated mode. It's a similar story with @IO Int@ and @Int@ in Haskell.}
 
@@
 
-There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@, except for the @scan@ primitive which explicitly uses a fold's intermediate value, rather than the value of the entire pass.
+There is also no way to take an @Aggregate@ and convert it to a @Pure@ or an @Element@.
 

### 2016-02-24 2016icicle/section/01-Introduction.tex
break everything

@@
 
+\amos{Yes, I had some concrete examples earlier but they turned out to be invalid. How about something along these lines?}
+\begin{code}
+SELECT open FROM stocks;
+
+SELECT open / (SELECT AVG(open) FROM stocks)
+FROM stocks;
+
+SELECT open / (SELECT AVG(open) FROM stocks),
+       open - (SELECT MIN(open) FROM stocks)
+FROM stocks;
+
+SELECT open / avgopen, open - minopen
+FROM stocks,
+(SELECT AVG(open) as avgopen,
+        MIN(open) minopen FROM stocks) agg;
+
+SELECT (SELECT AVG(open) FROM stocks)
+       (SELECT MIN(open) FROM stocks)
+FROM ();
+
+SELECT AVG(open) as avgopen,
+       MIN(open) minopen FROM stocks
+\end{code}
+
 For feature generation for machine learning, our queries have some quite specific restrictions.
 \ben{The following describes properties that that queries and data have intrinsically, not \emph{restrictions} placed on them}.
+\amos{I don't understand what this means. Is an SQL query with cross product, join, etc not a query?}
+
 Queries only need to read from one `table', for example the stocks table with company codes, dates and open and close prices for each day.

### 2016-02-24 2016icicle/section/02-Source.tex
break everything

@@
 The second parameter is the actual value to sample at the clock.
-In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
+In Icicle this has type , meaning it is available as a number at any part of the stream.
 \ben{Maybe hold off the description of @Element@ and @Aggregate@ until later so we can introduce them both at the same time. At this point in the text it's not clear why we need the modal types.}
@@
 
-The result of $\mi{sum}$ above has type $\Agg{\NN}$.
+The result of $\mi{sum}$ above has type .
 
@@
 @function@ 
-$\mi{sum}[c]~(v~:~\NN) : \Agg{\NN}$    \\
+$\mi{sum}[c]~(v~:~\NN) : $    \\
 \> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
@@
 
-The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.
-
 \ben{The previous section states that ``all function definitions can be inlined into the user query''.
@@
 $\mi{open}_1$  \> $=$ \> $@Elements@~[10]$ \\
-$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_1,~\mi{open}~:=~\mi{open}_1]}{@Scalar@~10} $ \\
 \\
@@
 $\mi{open}_2$  \> $=$ \> $@Elements@~[11,~12]$ \\
-$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_2,~\mi{open}~:=~\mi{open}_2]}{@Scalar@~23} $
 \end{tabbing}

### 2016-02-24 2016icicle/section/03-Types.tex
break everything

@@
 
-The judgment $\WrapMode{\tau}{m}{\tau'}$, corresponds to unboxing the a function argument of type $m$ and reboxing the function return type $\tau$ into modality $m$, with $\tau'$ as the boxed return type.
 The rule (WrapModeArrow) simply recurses down through the function arguments, applying the judgment to the result.
@@
 
-The judgment $\WrapApp{\tau}{\phi\ldots}{\tau'}$ attempts to apply the function type $\tau$ to the argument list $\phi\ldots$, performing unboxing as necessary.
 When the argument list is empty and the function type is a nullary function or simple value type, (WrapAppFinished) simply returns the result type.
@@
 First, look at an example that would require two passes over the data: maybe @filter open > mean open in count@.
-In this case, $\mit{open}$ has type $\Elm{\NN}$ while $\mit{mean~open}$ has type $\Agg{\NN}$.
 The primitive $>$ has type $\NN \to \NN \to \NN$ and can be boxed to either @Element@ or @Pure@, but not both.

### 2016-03-01 2016icicle/section/02-Source.tex
kill

--- 2016-01-04 2016icicle/section/02-Source.tex
@@
 
-In this section we describe the grammar and evaluation semantics of the Icicle language.
-
-Icicle queries execute over a single table, which must at least contain an entity identifier and a date.
-Each query implicitly groups the data by entity and sorts it by date, before iterating through the data and computing some aggregate.
-We define the stocks table as follows, leaving out the company code (entity identifier) and date.
-
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-$@table@~\mi{stocks}$    \\
-$\{$ \> $\mi{open}$ \> $:~\NN$ \\
-$~,$ \> $\mi{close}$ \> $:~\NN$ \\
-$\}$
-\end{tabbing}
-
-This defines the columns as natural numbers, and allows them to be referred to by name.
-In the definitions and queries below, any reference to these names means the column itself, with the entire data for a given entity.
-This is annotated with the mode @Element@, meaning that it exists for every element of the input.
-
-We now define some useful functions and other definitions that can be used in the queries themselves.
-
-First we define a function to compute the sum of some input data.
-While traditionally sum would take just an input stream of numbers, we separate this into two distinct parameters.
-The first parameter is the clock for the input stream, which defines the parts of the input stream to sample.
-This can be thought of as a stream of boolean flags, where true denotes that the input stream should be sampled at this time.
-The second parameter is the actual value to sample at the clock.
-In Icicle this has type $\Elm{\NN}$, meaning it is available as a number at any part of the stream.
-
-By separating the clock from the element values, we guarantee that any references to elements denote the same point in the stream, and can always be used together.
-For example, computing the difference between the two fields above ($\mi{open} - \mi{close}$) refers to the open and close fields at the same point.
-The clock is used when performing a fold, and determine at which points to sample the elements and update the accumulator.
-This is in constrast to traditional streaming dataflow languages, where the element data and the clock are intertwined.
-This requires clock analysis\CITE to tell whether two element streams can be used together with a bounded buffer.
-
-We denote clocks in brackets as they are a different universe to normal expressions.
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-@function@ 
-$\mi{sum}[c]~v$    \\
-\> $=$  \> $@fold@[c]~s~=~0~@then@~s~+~v$. \\
-\end{tabbing}
-
-The result of $\mi{sum}$ above has type $\Agg{\NN}$.
-Aggregate types denote that their value is only available after the entire input stream has been seen.
-By restricting the bodies of folds to be element types, and their output to be aggregate types, it means no reference to the result of folds can be made in the body of a fold.
-
-We can define a query that computes the sum of each company's open prices.
-Query definitions are similar to function definitions, except they only take one argument: the clock describing all available input.
-They also must return aggregates, as an element query could be unbounded in size.
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-@query@ 
-$\mi{opensum}[c]$    \\
-\> $=$  \> $\mi{sum}[c]~\mi{open}$. \\
-\end{tabbing}
-
-We can run $\mi{opensum}$ over some example data.
-This example data has the company code and date for each entry, as well as the stock's open and close prices.
-Note that the sum only becomes available at the end of each entity, and only applies to a particular entity's elements.
-
-\begin{code}
-CODE, OPEN, CLOSE, DATE,        OPENSUM
-ABC,    10,    12, 2015-01-01,       10
-IAG,    11,    12, 2015-01-01
-IAG,    12,    10, 2015-01-02,       23
-\end{code}
-
-Next, we can define functions $\mi{count}$ and $\mi{mean}$ in terms of $\mi{sum}$.
-Count passes the argument $1$ to sum, and this is interpreted as an element containing all ones.
-The input clocks are passed along as-is.
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-@function@ 
-$\mi{count}[c]$                                     \\
- \> $=$  \> $\mi{sum}[c]~1$.                           \\
-                                                    \\
-@function@ 
-$\mi{mean}[c]~v$                                       \\
- \> $=$  \> $\mi{sum}[c]~v~/~\mi{count}[c]$.              \\
-\end{tabbing}
-
-Clocks can be reduced to filter out certain elements using @where@.
-In this query, we use @where@ to count the number of days where the close price is higher than the open price.
-\begin{tabbing}
-MM \= MM \= MMMMMMMMMMM \= \kill
-@query@ 
-$\mi{gap}[c]$                                        \\
- \> $=$  \> $\mi{count}[c~@where@~\mi{close}~>~\mi{open}]$.       \\
-\end{tabbing}
-
-In the example below we have shown the result of the element expression $\mi{close}~>~\mi{open}$ for clarity only; this is not included in the output of the $\mi{gap}$ query.
-\begin{code}
-CODE, OPEN, CLOSE, CLOSE > OPEN, GAP
-ABC,    10,    12,            T,   1
-IAG,    11,    12,            T
-IAG,    12,    10,            F,   1
-\end{code}
-
-\TODO{Mention resumables and bubblegum?}
-
-We formally define the grammar of the language in figure~\ref{fig:source:grammar}.
-Note that general application is not allowed: arguments can only be applied to primitives and named functions.
-Similarly, functions and primitives must be fully applied.
-This, combined with the lack of lambda construction, means that higher order functions are outlawed.
-While a first-order language (one lacking higher order functions) offers somewhat less abstraction, it simplifies the typing rules and makes it easier to support our performance guarantees.
-
-While the examples shown allow user defined functions, there is no recursion allowed.
-This means that all function definitions can be inlined into the user query with guaranteed termination.
-When converting to the intermediate language, we assume that this inlining has occurred.
-
-\subsection{Evaluation}
-
 \input{figures/Source-Grammar.tex}
 
-In figure~\ref{fig:source:eval} we describe the bigstep evaluation semantics of the language.
-The judgment $\Bigstep{f}{x}{v}$ takes the list of defined functions, $f$, and the expression to evaluate, $x$, and returns the result value $v$.
-The evaluation works on a single entity's data at a time, and the column values, such as $\mi{open}$ and $\mi{close}$ in the examples above, must be substituted into the functions and expression beforehand.
-The clock is initially set to true for all values in the input stream.
-
-Substitution is denoted by $x[n~:=~v]$, where $x$ is the expression to substitute into, $n$ is the variable name to replace, and $v$ is the replacement payload.
-
-For example, for the $\mi{opensum}$ query we would have two evaluations:
-\begin{tabbing}
-MMMM \= M \= \kill
-$\mi{clock}_1$ \> $=$ \> $@Elements@~[@T@]$ \\
-$\mi{open}_1$  \> $=$ \> $@Elements@~[10]$ \\
-$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_1,~\mi{open}~:=~\mi{open}_1]}{@Scalar@~10} $ \\
-\\
-$\mi{clock}_2$ \> $=$ \> $@Elements@~[@T@,~@T@]$ \\
-$\mi{open}_2$  \> $=$ \> $@Elements@~[11,~12]$ \\
-$\Bigstep{\mi{sum}}{\mi{opensum}[c~:=~\mi{clock}_2,~\mi{open}~:=~\mi{open}_2]}{@Scalar@~23} $
-\end{tabbing}
-
-In the evaluation rules we make the distinction between @Scalar@ and @Elements@ explicit, but in the source grammar we omit the @Scalar@ qualifier, and do not allow @Elements@ as values, except when substituted in for evaluation.
-For example, the definition of $\mi{sum}$ would actually have $@Scalar@~0$ as its initialiser.
-Scalars can also be thought of as a repeated stream of the same value, of the same length as the input stream.
-
-The first rule (EvValue) simply returns its value unchanged.
-
-The next two rules, (EvPrimScalar) and (EvPrimElements) are for primitive binary operators (as the only primitives are binary).
-When both arguments are scalars, we simply apply the operator to both arguments.
-If both arguments are element streams, we pair the streams together and apply the operator to each pair.
-If only one of the arguments is an element stream, the other argument must be a scalar, and the function $\mi{elements}$ is used to extend the scalar to a repeated stream before applying the operator.
-
-Next, (EvFunApp) performs named function application.
-The function environment $f$ is searched for a function named $n$, and we find the names of the arguments $\mi{args}$ and the function definition $\mi{exp}$.
-The function definitions are non-recursive, and the function definition can only refer to functions defined before it.
-We use $f'$ as the new list of available function bindings.
-All arguments are evaluated to values, then each argument value is substituted into the function body, using the names of the arguments.
-
-For (EvLet) the let-definition is evaluated, then substituted into the let-body.
-
-For (EvWhere), used in clocks, both arguments are streams of booleans.
-We evaluate both arguments and perform a pairwise and of the two, where the result clock is only true for an element if it is true for both inputs on that element.
-
-The final evaluation rule, (EvFold), is the most important and most complex.
-It evaluates a reduction over the stream.
-First the clock and the initial value are evaluated, then a secondary judgment form is used to evaluate the fold.
-
-The fold judgment form takes a list of functions, the index of the current fold value, the stream of boolean clock values, the non-empty list of fold results so far, and the expression to evaluate.
-For (EvFoldNil) the clock is finished, so the last value in the fold result must be the final fold value.
-
-For the other cases, we inspect the first element of the clock stream.
-If the clock is true, we perform the fold with the fold results so far, and compute the rest of the clock stream.
-If the clock is false, we reuse the last value in the fold results, as any elements for this index will not be used.
-
-For (EvFoldClockTrue) we evaluate $k$ with the fold results so far substituted into the accumulator as an element stream.
-The result will be an element stream of at least length $i$, as the result of the fold so far.
-We then get the result at index $i$ and add it as the latest accumulator value, and continue computing the rest of the stream with this new accumulator.
-
-For (EvFoldClockFalse) we can re-use the previous accumulator value, which is the last value of the fold results.
-We add the last fold result to the accumulator, and compute the rest of the stream.
-
 \input{figures/Source-Eval.tex}

### 2016-03-01 2016icicle/section/04-Core.tex
ok
break everything
kill

--- 2015-12-05 2016icicle/section/04-Core.tex
@@
 
-In figure~\ref{fig:core:grammar} we introduce the intermediate language.
-The top-level of this language is $\mit{Program}$ with four stages of computation: pure computations, followed by stream transformers, then folds, and finally extracting the result of the folds.
+This section presents the Core language, which we use as an intermediate language before converting to imperative code.
 
-The program definition may seem more complicated than necessary, as all programs could be expressed as a single fold and an extract, but separating these out turns out to be useful for efficiency.
-This will be expanded upon later.
-
-The expression type for Core is much simpler than Source, $\mit{CoreExp}$ having just names, primitive application and let bindings.
-More primitives are required for the Core language, as some filters in the Source language are converted to @If@s, and groups are converted to use map primitives such as lookup and insert.
-The @scan@ primitive is no longer used in the Core language as it is just converted to a variable reference.
+\subsection{Grammar}
 
@@
 
-\input{figures/Core-Compile.tex}
+% \input{figures/Core-Compile.tex}

### 2016-03-03 2016icicle/section/02-Source.tex
stuff

@@
 
+The two main types in Icicle are @Stream@ and @Fold@.
+Streams represent data values as they flow through the program.
+Streams do not always have data flowing through them, but have an associated clock which describes when data occurs.
+Two streams with the same clock both have data at the same time.
+Folds, on the other hand, are results of computations over the stream data that has been seen.
+Folds always have a well-defined value.
+Folds can be easily converted to streams, by sampling their current value whenever the stream clock is true.
+It is harder to convert a stream to a fold, as the stream has gaps where it is undefined, and one must specify how to fill the gaps (hold last, fill with zero, etc).
+Finally, folds can be computed recursively based on their previous values.
+
 \input{figures/Source-Grammar.tex}
+The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
+The first four rules are rather standard lambda calculus.
+
+Next are the primitives.
+@when@ takes a stream and a fold, and returns a new fold whose value is the stream \emph{when}ever the stream is defined, and the fold otherwise.
+This is used for defining folds that depend on stream values.
+@sample@ takes a stream of any type and a fold, creates a new stream using the values of the fold.
+@mapS@ and @zipS@ perform map and zip operations on streams of the same rate.
+@mapF@ and @zipF@ perform map and zip operations on folds.
+
+Let expressions such as $@let@~x~=e~@in@~e$ are as usual. 
+Folds are defined using the syntax @let folds@ syntax. 
+Multiple folds can be defined together, and each fold is defined by the initial value, and the ``kons'' part, defining the next value.
+
+For example, a simple fold comprised of nothing but $0$
+\begin{tabbing}
+MM \= \kill
+$@let folds@~\mi{zeros}~=~0,~\mi{zeros}$ \\
+$@in@~\mi{zeros}$
+\\
+\\
+$\mi{zeros}$ \> $=~\langle 0~0~0~0~\cdots~\rangle$ \\
+\end{tabbing}
+
+When computing the sum over another fold, the newly-defined fold and the input fold are zipped and then added together.
+\begin{tabbing}
+$\mi{sum}~=~\lambda{}(\mi{values}~:~@Fold@~\NN).$ \\
+$@let folds@~\mi{s}~=~0,~@mapF@~(+)~(@zipF@~\mi{s}~\mi{values})$ \\
+$@in@~\mi{s}$
+\\
+\\
+$\mi{sum}~\langle 0~1~2~3~\cdots~\rangle$ \\
+$~=~\langle 0~1~3~6~\cdots~\rangle$ \\
+\end{tabbing}
+
+This $\mi{sum}$ function requires a fold argument, but it can be applied to a stream $s$ by creating a fold that is the stream when it is defined, and $0$ otherwise: $@when@~s~zeros$.
+A more general way is to define a new $\mi{sum'}$ function that works directly over streams:
+\begin{tabbing}
+MMMMMM \= MM \= MM \kill
+$\mi{sum'}~=~\lambda{}(\mi{values}~:~@Stream@~c~\NN).$ \\
+$@let folds@~\mi{s}~=~0,$ \\
+\> $@when@$ \> $(@mapS@~(+)~(@zipS@~(@sample@~\mi{values}~\mi{s})~\mi{values}))$ \\
+\> \> $~\mi{s}$ \\
+$@in@~\mi{s}$
+\\
+\\
+$\mi{sum'}~\langle \bot~1~\bot~3~\cdots~\rangle$ \\
+$~=~\langle 0~1~1~4~\cdots~\rangle$ \\
+\end{tabbing}
+
+Here, the fold $s$ is sampled to a stream with the same clock as the input stream.
+Whenever the input stream is defined, the previous value of $s$ and the current value of the $\mi{values}$ will be added together, producing the new sum.
+When the input stream is not defined, the previous value of $s$ is used unchanged.
+
+When referring to the newly defined folds in the right-hand-side of the fold (the kons), all values refer to the immediate previous value.
+The order of the fold bindings has no effect on the program.
+\begin{tabbing}
+MM \= \kill
+$@let folds@$ \\
+\> $\mi{one}~=~1,~\mi{zero}$ \\
+\> $\mi{zero}~=~0,~\mi{one}$ \\
+$@in@~\mi{one}$
+\\
+\\
+$\mi{one}$  \> $=~\langle 1~0~1~0~\cdots~\rangle$ \\
+$\mi{zero}$ \> $=~\langle 0~1~0~1~\cdots~\rangle$ \\
+\end{tabbing}
+
+Yes, I realise the grammar doesn't have @+@ or @/@, and that it doesn't have polymorphism.
+For the examples, assume that a specialisation pass occurs, and that there is a ``sensible set of primitives on natural numbers and lists''.
+
+The chosen primitives are expressive enough, but can be rather cumbersome for programming with.
+The
 
 \input{figures/Source-Eval.tex}
+
+\subsection{Obvious extensions}
+Obvious extensions and things I removed for simplification.
+
+\subsubsection{Fixpoint}
+$@fix@~(x~:~\tau)~@in@~e$.
+By requiring $\tau$ to have kind @Data@ rather than @Flow@, we can allow recursion at runtime, while still ensuring a static dataflow graph.
+I am tempted to kill the kinds for the simple version because with no fixpoint they do not serve much purpose (the dataflow graph is static regardless).
+
+\subsubsection{ Maps and zips }
+@mapS/mapF@, @zipS/zipF@.
+For the simple version these could be merged into a single map/zip operating over just @Fold@s.
+You can convert from a @Stream@ to a @Fold@ and back easily enough - you need a zero element for the @Fold@ but whatever.
+
+\subsubsection{ Subrates }
+$@let-rate@~(c'~:_k~@Clock@)~(w~:~c'~\le~c)~=~(e~:~@Stream@~c~\BB)$ and $@subrate@~(w~:~c'~\le~c)~(e~:~@Stream@~c~\tau)~:~@Stream@~c'~\tau$.
+You can fake this by using Folds anyway.
+
+\subsubsection{ Unpack }
+$@let-unpack@~(x_\tau~:_k~@Data@)~(x_z~:~x_\tau)~(x_k~:~x_\tau~\to~@Fold@~x_\tau)~(x_r~:~x_\tau~\to~\tau)~=~(e~:~@Fold@~\tau)$.
+Pulling apart folds to make new folds.
+Useful for groups, segmented operations, and so on where the fold can be restarted or actually be many folds, as well as for other higher order folds.

### 2016-04-06 2016fhpc/Main.tex
ok

--- 2015-11-24 2016icicle/Main.tex
@@
 
-\title{Icicle: a single-pass query language}
+\title{Icicle: fuse your queries}
 
@@
   \and Ben Lippmeier$^\dagger$
-  \and Tran Ma$^\ddagger$
-  \and Jacob Stanley$^\ddagger$
 }{
@@
     $^\ddagger$Ambiata      \\
-    Big data and shit       \\[2pt]
-    \textsf{\{first.last\}@ambiata.com}
+    Grow with data          \\[2pt]
+    \textsf{amos.robinson@ambiata.com}
   }
@@
 When streaming a large amount of data, simply iterating over the data may take hours.
-In this case, the difference between a query that takes one iteration and a query that takes multiple iterations must be very explicit.
-For many purposes, limiting queries to a single pass is sufficient.
-We introduce a type system based on modality types to enforce this single pass restriction.
-By using an appropriate intermediate language we can fuse multiple independent queries together, before extracting efficient single-loop C code.
+If multiple queries are to be performed, it is important that work is not duplicated.
+Queries that can be performed together must be performed in the same iteration.
+We introduce a simple streaming language for computing queries in a single-pass over the data.
+By using an appropriate intermediate language we guarantee fusion between queries on the same input streams, before extracting efficient C code.
 \end{abstract}
@@
 \input{section/01-Introduction.tex}
-\input{section/02-Source.tex}
-\input{section/03-Types.tex}
-
-\input{section/XX-Related-DataFlow.tex}
+\input{section/02-Core.tex}
 
 \section*{Acknowledgements}
-I would like to thank all of Ambiata, particularly Huw Campbell and Mark Hibberd.
 

### 2016-04-06 2016fhpc/section/01-Introduction.tex
ok

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}

For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
When performing multiple queries, it is important to limit the amount of duplicate work and iterations.

For example, suppose we have a table called @stocks@ with a price for each company, for each day.
Computing the sum and mean over all companies and all time is quite simple in SQL:
\begin{code}
SELECT SUM(price)
FROM stocks;

SELECT SUM(price) / COUNT(price)
FROM stocks;
\end{code}

It is easy to express a query which computes both at once:
\begin{code}
SELECT SUM(price)
     , SUM(price) / COUNT(price)
FROM stocks;
\end{code}

Suppose we also wish to compute the sum of each company:
\begin{code}
SELECT company
     , SUM(price)
FROM stocks
GROUP BY company;
\end{code}

Now, it becomes less obvious how to express both sum over all companies, and sum of each company in a single query.
One can use a nested query, but it is not obvious whether the query optimiser will be able to perform this query as a single iteration over the data. 
\begin{code}
SELECT company
     , SUM(price)
     , (SELECT SUM(price) FROM stocks)
FROM stocks
GROUP BY company;
\end{code}

In our intermediate language:
\begin{code}
loop stocks {
  stage {
    fold sum   = 0
         then    sum + price;
  }

  stage {
    fold keys  = GroupKey.empty
         then    GroupKey.insert keys company;
    fold gsum  = GroupVal.empty
         then    GroupVal.update
                  gsum
                  (GroupKey.lookup keys company)
                  0
                  (\a. a + price);
    let  group = Group.create keys gsum;
  }
}
\end{code}

It could be argued that the above queries need not be fused: with adequate indices, the database engine could store the sum for each company, as well as the total sum, making the actual queries effectively constant time.
This does not necessarily hold true for more interesting queries, such as exponentially smoothed averages.

In this paper, we describe an intermediate language specifically for this kind of horizontal fusion across queries.

### 2016-04-06 2016fhpc/section/02-Core.tex
ok

%!TEX root = ../Main.tex
\section{Icicle Core}
\label{s:IcicleCore}

Horizontal fusion for imperative loops is relatively easily, when the loops have the same number of iterations.
The problem with imperative loops occurs when trying to remove duplicate computations.
Firstly, the `definition' of a computation is split across multiple places: because the initial value of an accumulator and its update expression are in separate parts of the program, some analysis must be done to recover these.
For example, in the following loop, @a@ and @b@ have equivalent `update' expressions, but could denote very different computations depending on the implementation of @function@.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}
\end{code}

Secondly, the order of statements in a loop affects the meaning: when accumulators are mutually dependent on each other, one can reorder the statements to produce a different, but still valid program.
These two loops are not equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}
for (...) {
  b = a - b;
  a = a + b;
}
\end{code}

Neither of these problems are insurmountable, but the analysis required certainly complicates our goal of removing duplicate computation after fusion.

With this in mind, we will introduce our intermediate language based on what we call ``fold nests'', as opposed to loop nests.
Firstly, each accumulator should be defined in one place: this means that if two accumulators are alpha equivalent, they are equivalent.
Secondly, reordering statements (or accumulators) should not affect the meaning: two programs are equivalent if they contain equivalent sets of bindings, regardless of ordering.

We now convert the first example to folds:
\begin{code}
a = 0;
b = 1;
for (...) {
  a = function(a);
  b = function(b);
}

==>
loop ... {
  stage {
    fold a = 0
        then function(a);
  }
  stage {
    fold b = 1
        then function(b);
  }
}
\end{code}

Each fold binding must be grouped inside a @stage@ block, which allows mutual recursion across folds.
In this case, there is no mutual recursion, so the stages only contain one binding each.
The first fold is given the name @a@ with @fold a@.
Its initial value comes straight after the equals sign, so @a@ starts with a value of @0@.
The update or `kons' expression comes after @then@, and in the update expression any reference to @a@ refers to the current value of the fold.

It now becomes easier to see that the two folds are not equivalent, as the entire definitions are in one place.

The next example requires mutual recursion.
\begin{code}
a = 0;
b = 1;
for (...) {
  a = a + b;
  b = a - b;
}

==>
loop ... {
  stage {
    fold a = 0
        then a + b;
    fold b = 1
        then a' - b;
  }
}
\end{code}

Here, the update expression for @a@ refers to a later fold, @b@.
Update expressions can refer to later fold bindings in the same stage, as well as all earlier fold bindings.
The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @a'@.
This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.

Another way to think of the new value restriction above is disallowing cycles in the references to new values:
when @b@ references @a'@, the new value of @a@, it means that @a'@ must be computed before @b'@ can be.
If there were a cycle where @a'@ also depended upon @b'@, there would be no order to compute them in.
Disallowing cycles ensures an execution order.

By making the distinction between `old value' and `new value' explicit, the ordering in the program no longer has any effect on the semantics.
After converting the reordered loop with different semantics, the two programs are no longer alpha-equivalent:
\begin{code}
a = 0;
b = 1;
for (...) {
  b = a - b;
  a = a + b;
}

==>
loop ... {
  stage {
    fold b = 1
        then a - b;
    fold a = 0
        then a + b';
  }
}
\end{code}

\begin{figure}

\begin{code}
N     := a,b,...
FoldN := a,b,...

Exp   := N
       | FoldN  | FoldN'
       | Exp Exp
       | \N. Exp

Bind  := fold FoldN : T = Exp then Exp;
       | let  N     : T = Exp;

Stage := stage { Bind...  }
Query := loop N : T { Stage... }

\end{code}

\caption{Grammar}
\label{fig:grammar}
\end{figure}

\newcommand\TypecheckX[3]
{                   #1
        ~\vdash~    #2~:~#3
}

\newcommand\TypecheckS[3]
{                   #1
        ~\vdash~    #2~\dashv~#3
}

\begin{figure}

$$
\boxed{\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}}
$$

$$
\ruleAx
{
  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{\}}{\emptyset}
}{QueryNil}
$$

$$
\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau}{s}{\Gamma'}
  \quad
  \TypecheckS{\Gamma,~\Gamma'}{@loop@~x~:~\tau~\{s'\}}{\Gamma''}
}{
  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{s;~s'\}}{\Gamma' \cup \Gamma''}
}{QueryCons}
$$

$$
\boxed{\TypecheckS{\Gamma}{\mi{Stage}}{\Gamma'}}
$$

$$
\ruleAx
{
  \TypecheckS{\Gamma}{@stage@~\{\}}{\emptyset}
}{StageNil}
$$

$$
\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau,~x'~:~\tau}{@stage@~\{s'\}}{\Gamma'}
  \quad
  \TypecheckX{\Gamma}{z}{\tau}
  \quad
  \TypecheckX{\Gamma~\cup~\{v~|~v'~\in~\Gamma'\}}{k}{\tau}
}
{
  \TypecheckS{\Gamma}{@stage@~\{@fold@~x~:~\tau~=~z~@then@~k;~s'\}}{\Gamma',~x~:~\tau,~x'~:~\tau}
}{StageFold}
$$

$$
\ruleIN
{
  \TypecheckS{\Gamma,~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}
  \quad
  \TypecheckX{\Gamma}{x}{\tau}
}
{
  \TypecheckS{\Gamma}{@stage@~\{@let@~x~:~\tau~=~e;~s'\}}{\Gamma',~x~:~\tau}
}{StageLet}
$$

\caption{Typing rules}
\label{fig:typing}
\end{figure}

The full grammar is in figure~\ref{fig:grammar}, however primitives and types have been omitted.
It is assumed that there are some primitives such as @1@, @+@ and types such as @Int@, but the elements do not actually matter.

The typing rules are in figure~\ref{fig:typing}.
The rules shown only cover the staging and folds.
As with the grammar, we assume there is a typing rule $\TypecheckX{\Gamma}{\mi{Exp}}{\tau}$ which checks an expression under the given environment.

### 2016-04-07 2016fhpc/section/02-Core.tex
ok

@@
     fold b = 1
-        then a' - b;
+        then new a - b;
   }
@@
 The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
-The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @a'@.
+The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @new a@.
 This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.
@@
 Another way to think of the new value restriction above is disallowing cycles in the references to new values:
-when @b@ references @a'@, the new value of @a@, it means that @a'@ must be computed before @b'@ can be.
-If there were a cycle where @a'@ also depended upon @b'@, there would be no order to compute them in.
+when @b@ references @new a@, the new value of @a@, it means that @new a@ must be computed before @new b@ can be.
+If there were a cycle where @new a@ also depended upon @new b@, there would be no order to compute them in.
 Disallowing cycles ensures an execution order.
@@
     fold a = 0
-        then a + b';
+        then a + new b;
   }
@@
 Exp   := N
-       | FoldN  | FoldN'
+       | FoldN  | new FoldN
        | Exp Exp
@@
 
-\begin{figure}
+\begin{figure*}
 
@@
 }{QueryNil}
-$$
-
-$$
 \ruleIN
@@
 }{StageNil}
-$$
-
-$$
-\ruleIN
-{
-  \TypecheckS{\Gamma,~x~:~\tau,~x'~:~\tau}{@stage@~\{s'\}}{\Gamma'}
-  \quad
-  \TypecheckX{\Gamma}{z}{\tau}
-  \quad
-  \TypecheckX{\Gamma~\cup~\{v~|~v'~\in~\Gamma'\}}{k}{\tau}
-}
-{
-  \TypecheckS{\Gamma}{@stage@~\{@fold@~x~:~\tau~=~z~@then@~k;~s'\}}{\Gamma',~x~:~\tau,~x'~:~\tau}
-}{StageFold}
-$$
-
-$$
 \ruleIN
@@
 
+$$
+\begin{array}{l}
+
+\ruleIN
+{
+  \TypecheckS{\Gamma,~x~:~\tau,~@new@~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}
+
+  \quad
+  \TypecheckX{\Gamma}{z}{\tau}
+  \quad
+  \TypecheckX{\Gamma~\cup~\mi{folds}}{k}{\tau}
+}
+{
+  \TypecheckS{\Gamma}{@stage@~\{@fold@~x~:~\tau~=~z~@then@~k;~s'\}}{\Gamma',~x~:~\tau,~@new@~x~:~\tau}
+}{StageFold}
+
+  \\
+  \mbox{where} \\
+  \quad
+  \mi{folds}~=~\{v~:~\tau'~|~@new@~v~:~\tau'~\in~\Gamma'\}
+  \end{array}
+$$
+
 \caption{Typing rules}
 \label{fig:typing}
-\end{figure}
+\end{figure*}
 
@@
 As with the grammar, we assume there is a typing rule $\TypecheckX{\Gamma}{\mi{Exp}}{\tau}$ which checks an expression under the given environment.
+
+The first judgment, $\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}$

### 2016-04-07 2016fhpc/section/XX-Related.tex
ok

%!TEX root = ../Main.tex
\section{Related}
\label{s:Related}

``most compilers restrict themselves to common subexpression elimination within basic blocks or extended basic blocks''
\cite{debray1992compiler}

Temporal common subexpression elimination in Single Assignment C
allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
This is a more general case of the elimination opportunities that arise from loop unrolling.

For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
\begin{code}
int[] A;
int a = 1;
int b = 0;
while(int i = 0; i != size - 1; i++)  {
  int ax = f(A[i]);
  int bx = f(A[i+1]);
  a += ax;
  b *= bx;
}
\end{code}

I cannot think of a better example than this right now.

### 2016-04-10 2016fhpc/section/03-Elimination.tex
ok

%!TEX root = ../Main.tex
\section{Elimination}
\label{s:Elimination}

\subsection{Preliminary transforms}

A-normalisation is relatively easy, however any expressions in the initialiser of each fold can be lifted out as a separate fold with initialiser and an identity expression (that is, the name of the fold).
Any expressions to be lifted out from the update expression should be created as new @let@ bindings.
They cannot necessarily be created as @fold@ bindings because the expression will not work for the initial expression, if it refers to other lets or loop iterators.

Some simple forwarding can be applied, such as lets of variables, and folds where the initial and update expressions are the same.
Fold initials could be forwarded to other initials, but fold updates cannot always be forwarded to other updates, as the first evaluation of the other update will refer to the forwarder's initial.

So-called ``melting'' can be performed, splitting apart pairs into multiple bindings, arrays of pairs into pairs of arrays, and so on.
Key-value maps can be similarly split into a special map from key to index, and an array for the values.
Splitting these complex structures into their constituent parts exposes more opportunities for elimination removal.
The following program computes the mean of the loop elements. Here, the sum and count are both computed, and stored in a pair.

\begin{code}
loop item {
  stage {
    fold sumcount
     = (0, 0)
     then
       ( fst sumcount + item
       , snd sumcount + 1);

    fold mean
     = fst sumcount / snd sumcount
     then
       fst sumcount / snd sumcount;
  }
}
\end{code}

If another query were to compute the sum as well, it is not obvious looking at the two expressions that @sum@ and @fst sumcount@ are equivalent:
\begin{code}
loop item {
  stage {
    fold sum
     = 0
     then
       sum + item;
  }
}
\end{code}

Now, by splitting the @sumcount@ pair into two bindings named @sumcount1@ and @sumcount2@, it becomes easier to see that the definitions of @sum@ and @sumcount1@ are equivalent.
\begin{code}
loop item {
  stage {
    fold sumcount1
     = 0
     then
       sumcount1 + item;

    fold sumcount
     = 0
     then
       sumcount2 + 1;

    fold mean
     = sumcount1 / sumcount2
     then
       sumcount1 / sumcount2;
  }
}
\end{code}

\subsection{Strongly connected components}
Up to now, the reason for the stages has only been briefly mentioned.

The strongly connected components must be found.

### 2016-04-10 2016fhpc/section/XX-Related.tex
ok

@@
 I cannot think of a better example than this right now.
+
+Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
+By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
+This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
+However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive

### 2016-04-11 2016fhpc/section/XX-Related.tex
ok

@@
 
+Continuous queries are a similar domain, where a single query keeps running and producing results, as data is received.
+In \cite{munagala2007optimization} there are multiple queries based off the same incoming stream.
+Each queries is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
+If these predicates are expensive to compute, it is certainly not ideal to compute these for each query.
+
 Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.

### 2016-04-12 2016fhpc/section/03-Elimination.tex
ok

@@
 
-    fold sumcount
+    fold sumcount2
      = 0
@@
 Up to now, the reason for the stages has only been briefly mentioned.
+The idea is that each stage should contain a single set of mutually recursive folds.
+Then, each set of mutually recursive folds becomes a single unit to be operated on for removing duplicates.
 
-The strongly connected components must be found.
+The type system enforces that mutually recursive folds must be in the same stage, but not that each stage can only contain one set of mutually recursive folds.
+We find the strongly connected components of the program graph, so that each strongly connected component.
+When treating the program as a graph, each binding is a node and references to a fold @f@ as either @f@ for the current value, or @new f@ for the new value, both count as edges to @f@.
+
+As a contrived example, we have bindings @sum@ and @count@ on their own, and another pair of bindings @a@ and @b@ which depend on each other, as well as on @sum@ and @count@.
+\begin{code}
+loop item {
+  stage {
+    fold count = 0 then count + 1;
+
+    fold sum   = 0 then sum   + item;
+
+    fold a     = 1 then count + b;
+    fold b     = 0 then sum + new a;
+  }
+}
+\end{code}
+
+The graph looks something like this.
+\begin{code}
+  count     sum
+    |        |
+    V        V
+    a <----> b
+\end{code}
+
+The strongly connected components are found, and ripped out into separate stages.
+The bindings of each stage are sorted topologically, but this time only @new@ references count as edges in the graph.
+This is because @new@ references require the new, updated value of the fold, but other references use the value that is already available at the start of the iteration.
+Hence, only @new@ references impose an ordering constraint.
+\begin{code}
+loop item {
+  stage {
+    fold count = 0 then count + 1;
+  }
+  stage {
+    fold sum   = 0 then sum   + item;
+  }
+  stage {
+    fold a     = 1 then count + b;
+    fold b     = 0 then sum + new a;
+  }
+}
+\end{code}
+
+\subsection{Stage layers}
+Once we have a list or set of stages, we can perform a topological ordering over the stages themselves.
+In this case, a whole stage becomes a node in the graph, and edges are any references between stages.
+
+The topological ordering is used for denoting execution order between the stages, but also tells us which stages need to be checked against each other, to remove duplicates.
+Each stage could be checked against all other stages, but this would be wasteful.
+
+We can look at the topological ordering of stages as layers of the same depth.
+In the @count/sum/ab@ example, @count@ and @sum@ are together on the top layer, while @ab@ is on the second layer.
+
+Each stage need only be compared with those on the same layer, and those on the layer directly above.
+In fact, of those on the layer above, it only needs to be compared with those it refers to.
+\TODO{Justify this further.}
+Because the stages have already been split apart, we also need only check against those with equal or greater number of bindings.
+
+\subsection{Duplicate removal}
+Because the stages have already been split apart, we can only replace a stage

### 2016-04-20 2016fhpc/section/XX-Related.tex
notes about GVN

@@
 
-``most compilers restrict themselves to common subexpression elimination within basic blocks or extended basic blocks''
-\cite{debray1992compiler}
+The most closely related transforms are induction variable elimination\cite{shivers1988control} and global value numbering\cite{rosen1988global}.
+Induction variable elimination finds and removes accumulators that are linear functions of the iteration number.
+This does not deal with non-linear functions such as sum, or mutually recursive accumulators.
+
+Global value numbering works on an SSA form and is much more general than induction variable elimination.
+Early versions such as Rosen et al\cite{rosen1988global} only worked on extended basic blocks with backedges removed.
+Alpern et al\cite{alpern1988detecting} introduced cases for specific loop backedges, where a fixpoint is performed to find the congruence sets of loops.
+More recently, Gulwani and Necula\cite{gulwani2004polynomial} improved upon this, supporting removal of more expressions, and imposing a polynomial time bound.
+Global value numbering for loops can perform all optimisations here.
+
+Common subexpression elimination is very closely related, but most imperative compilers only perform common subexpression elimination on straight-line computations with no control flow\cite{debray1992compiler}, or on loop invariant expressions\cite{bodik1998complete}.
+Common subexpression elimination for functional languages would be able remove lone expressions, but does not deal with mutually recursive folds.
 
@@
 This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
-However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive
+However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.
+
+The polyhedral model\cite{benabderrahmane2010polyhedral} analyses data dependencies in loops.
+For a given iteration, the polyhedral model computes the iterations which must be executed before the given iteration.
+By knowing the dependent iterations, one can restructure the loop so that the iterations are executed in a different order, while still performing any dependencies in order.
+For example, a loop iterating over @i@, with loop body @A[i] = A[i-10]@ depends on the tenth previous iteration.
+This loop could be executed in an order like @0, 1, 2,...@, but it could also stride by ten: @0, 10, 20..., 1, 11, 21,...@.
+This allows fusion of any two loops if the intersection of the two loops' iteration spaces is not empty.
+In our case the iteration order is fixed, as the program cannot control the order in which stream elements are received.
+The polyhedral model may still be applicable to streaming computations where one could introduce certain-sized buffers, then process the buffer in parallel.
+In this case the polyhedral model may expose more opportunities for duplicate elimination, but does not itself remove duplicates.

### 2016-05-06 2016fhpc/section/01-Introduction.tex
ok

@@
 
-For example, suppose we have a table called @stocks@ with a price for each company, for each day.
-Computing the sum and mean over all companies and all time is quite simple in SQL:
+For example, suppose we have a table called @stocks@ with open and close prices for each company, for each day.
+We may wish to find the number of days where the open price exceeds the close price, and vice versa:
 \begin{code}
-SELECT SUM(price)
-FROM stocks;
+SELECT COUNT(*) AS open_gt_close
+FROM   stocks
+WHERE  open > close;
 
-SELECT SUM(price) / COUNT(price)
+SELECT COUNT(*) AS open_lt_close
+FROM   stocks
+WHERE  open < close;
+\end{code}
+
+In order to fuse these so that both are computed in a single query, some ingenuity is required.
+This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
+\begin{code}
+SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
+     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
 FROM stocks;
@@
 
-It is easy to express a query which computes both at once:
+If we also wish to compute the mean of open price on those days, we can do this like so:
 \begin{code}
-SELECT SUM(price)
-     , SUM(price) / COUNT(price)
+SELECT SUM(open) / COUNT(*) AS mean
+FROM   stocks
+WHERE  open > close;
+\end{code}
+
+Now we can perform a similar trick to fuse all three together.
+\begin{code}
+SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
+     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
+     , SUM  (IF(open > close, open, 0))
+     / COUNT(IF(open > close, 1, 0)) AS mean
 FROM stocks;
@@
 
-Suppose we also wish to compute the sum of each company:
+However, there are some problems with this approach.
+This fusion on SQL source queries is manual and error-prone.
+More importantly, we have no guarantees about which queries can be fused.
+We would like a guarantee that \emph{any} two queries can be fused this way.
+
+Another problem is that we have no guarantee that duplicate computations introduced by fusion will be eliminated.
+Above we are computing the count where open exceeds close twice: once in @open_gt_close@ and again in @mean@.
+
+The purpose of Icicle is two-fold: we wish to ensure that all queries over the same table can be fused together, and guarantee that such duplicate computations will be removed.
+This is achieved by having a very restricted query language with no sorting, nested subqueries, and allowing only a single-pass over the data.
+
+In Icicle, the above queries are expressed as:
 \begin{code}
-SELECT company
-     , SUM(price)
-FROM stocks
-GROUP BY company;
+feature stocks
+in filter open > close
+in count.
+
+feature stocks
+in filter open < close
+in count.
+
+feature stocks
+in filter open > close
+in sum open / count.
 \end{code}
 
-Now, it becomes less obvious how to express both sum over all companies, and sum of each company in a single query.
-One can use a nested query, but it is not obvious whether the query optimiser will be able to perform this query as a single iteration over the data. 
+These are then converted into the following query plans before fusion:
 \begin{code}
-SELECT company
-     , SUM(price)
-     , (SELECT SUM(price) FROM stocks)
-FROM stocks
-GROUP BY company;
-\end{code}
-
-In our intermediate language:
-\begin{code}
-loop stocks {
-  stage {
-    fold sum   = 0
-         then    sum + price;
-  }
-
-  stage {
-    fold keys  = GroupKey.empty
-         then    GroupKey.insert keys company;
-    fold gsum  = GroupVal.empty
-         then    GroupVal.update
-                  gsum
-                  (GroupKey.lookup keys company)
-                  0
-                  (\a. a + price);
-    let  group = Group.create keys gsum;
-  }
+query stocks {
+  fold count  = 0
+           then if open > close then count + 1 else count;
+  output open_gt_close = count;
 }
 \end{code}
-
-It could be argued that the above queries need not be fused: with adequate indices, the database engine could store the sum for each company, as well as the total sum, making the actual queries effectively constant time.
-This does not necessarily hold true for more interesting queries, such as exponentially smoothed averages.
-
-In this paper, we describe an intermediate language specifically for this kind of horizontal fusion across queries.

### 2016-05-12 2016fhpc/Main.tex
ok
introduction and first section (Source), start on Core section

@@
 
-\title{Icicle: fuse your queries}
+\title{Icicle: write once, run once}
 
@@
 \begin{abstract}
-When streaming a large amount of data, simply iterating over the data may take hours.
+When querying a large amount of data, simply iterating over the data may take hours.
 If multiple queries are to be performed, it is important that work is not duplicated.
@@
 We introduce a simple streaming language for computing queries in a single-pass over the data.
-By using an appropriate intermediate language we guarantee fusion between queries on the same input streams, before extracting efficient C code.
+With type and other restrictions we guarantee fusion between queries on the same input table, before extracting efficient C code.
 \end{abstract}
@@
 \input{section/01-Introduction.tex}
-\input{section/02-Core.tex}
+\input{section/02-Source.tex}
+\input{section/03-Core.tex}
+\input{section/04-Benchmarks.tex}
+\input{section/05-Conclusion.tex}
+\input{section/XX-Elimination.tex}
+\input{section/XX-Related.tex}
 

### 2016-05-12 2016fhpc/section/01-Introduction.tex
introduction and first section (Source), start on Core section

@@
 \begin{code}
-SELECT COUNT(*) AS open_gt_close
+SELECT COUNT(*)
 FROM   stocks
@@
 
-SELECT COUNT(*) AS open_lt_close
+SELECT COUNT(*)
 FROM   stocks
@@
 \begin{code}
-SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
-     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
+SELECT COUNT(IF(open > close, 1, 0))
+     , COUNT(IF(open < close, 1, 0))
 FROM stocks;
@@
 \begin{code}
-SELECT SUM(open) / COUNT(*) AS mean
+SELECT SUM(open) / COUNT(*)
 FROM   stocks
@@
 \begin{code}
-SELECT COUNT(IF(open > close, 1, 0)) AS open_gt_close
-     , COUNT(IF(open < close, 1, 0)) AS open_lt_close
+SELECT COUNT(IF(open > close, 1, 0))
+     , COUNT(IF(open < close, 1, 0))
      , SUM  (IF(open > close, open, 0))
-     / COUNT(IF(open > close, 1, 0)) AS mean
+     / COUNT(IF(open > close, 1, 0))
 FROM stocks;
@@
 Another problem is that we have no guarantee that duplicate computations introduced by fusion will be eliminated.
-Above we are computing the count where open exceeds close twice: once in @open_gt_close@ and again in @mean@.
+Above we are computing the count where open exceeds close twice: once in the count, and again for the mean.
 
 The purpose of Icicle is two-fold: we wish to ensure that all queries over the same table can be fused together, and guarantee that such duplicate computations will be removed.
-This is achieved by having a very restricted query language with no sorting, nested subqueries, and allowing only a single-pass over the data.
+This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
 
@@
 \begin{code}
-feature stocks
+query stocks
 in filter open > close
@@
 
-feature stocks
+query stocks
 in filter open < close
@@
 
-feature stocks
+query stocks
 in filter open > close
@@
 
-These are then converted into the following query plans before fusion:
-\begin{code}
-query stocks {
-  fold count  = 0
-           then if open > close then count + 1 else count;
-  output open_gt_close = count;
-}
-\end{code}
+These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.
+
+The contributions of this paper are:
+\begin{itemize}
+\item
+we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
+\item
+we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
+\item
+and we present benchmarks of our generated code, which compares favourably against standard unix utilities and outperforms our existing system by orders of magnitude\sref{s:Benchmarks}.
+\end{itemize}

### 2016-05-12 2016fhpc/section/02-Source.tex
introduction and first section (Source), start on Core section

%!TEX root = ../Main.tex
\section{Icicle}
\label{s:IcicleSource}

\begin{figure}

\begin{code}
Exp   := x | Exp Op Exp | x Exp... | Query

Query := fold x = Exp then Exp
       | let  x = Exp          in Query
       | filter   Exp          in Query
       | group    Exp          in Query
       | Exp

Top   := query x in Query
Fun   := function x x... = Query
\end{code}

\caption{Icicle Grammar}
\label{fig:icicle:grammar}
\end{figure}

The grammar for Icicle is given in figure~\ref{fig:icicle:grammar}.
@Exp@ and @Query@ are mutually recursive, defining the body of a query.
@Top@ defines a top-level query over a particular table, whereas @Fun@ is used for function definitions.

Expressions can be variable names, operators such as addition and division, function application, or nested queries.
Function application @x Exp...@ is the name of a top-level function applied to any number of arguments.

Queries can define folds over the input, let-bindings, filtering according to some predicate, or grouping by some key.
The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
Count can be expressed by the following fold:
\begin{code}
fold count = 0 then count + 1
\end{code}

Let-bindings allow a name to be used in place of the expression, in the rest of some query.
\begin{code}
let diff = open - close
in  mean diff
\end{code}

Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
The following query will count the number of entries where the open price exceeds the close price.
\begin{code}
filter open > close in count
\end{code}

Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
\begin{code}
(filter open > close in count) / count
\end{code}

Group takes a key to group by, and a query to perform on the values of each grouping.
This query groups by the company code, and counts the number of records in each company.
\begin{code}
group company in count
\end{code}

@Fun@ is used for function definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1
\end{code}

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum value
 = fold s = 0 then s + value
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean value
 = sum value / count
\end{code}

\subsection{Single-pass restriction}

All Icicle queries must execute in a single pass over the data, as reading the data multiple times is expensive. 
Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives.

In order to ensure that queries can be executed as a single pass, we use a modal type system inspired by staged computation\cite{davies2001modal}.
We introduce two modalities, called @Element@ and @Aggregate@, which denote when computations must occur.

The @Element@ modality means that a computation is defined for each record in the table, or element in the stream.
Each column in the table is @Element@; for example the @open@ column has type @Element Int@, meaning that for each record in the column there is an @Int@.
This can be thought of as being represented by a stream of values.

The @Aggregate@ modality means that a computation is available only after all records in the table have been seen, or after the end of the stream.
These are used for the results of folds.
When computing count, the final value is not known until all the records have been seen, so @count@ has type @Aggregate Int@.

The modalities are automatically boxed and unboxed, so that if a function expecting a pure value is passed an @Element@ computation, the result of the function becomes an @Element@ computation too.
For example, if @open : Element Int@, then @open == 1@ has type @Element Bool@.
@Element@ and @Aggregate@ are mutually exclusive: a computation cannot be both an element and an aggregate.

Some examples with their types:
\begin{code}
1             :         Int
open          : Element Int
open > 1      : Element Int
sum           : Element Int -> Aggregate Int
sum open      : Aggregate Int
sum open > 1  : Aggregate Bool
\end{code}

The following program attemps to find the number of entries where the open price is above the mean of the close price.
This requires two passes over the data, as we do not know what the mean of all the data is until we have seen all the data.
This program violates the single-pass restriction, and is outlawed by the typesystem.
The left side of the @>@ has type @Element Int@, while the right side has type @Aggregate Int@, and these two are mutually exclusive.
\begin{code}
filter open > mean close
in count
(Error: (>) must be Element and Aggregate)
\end{code}

Here is an example to highlight one benefit of tracking these modalities.
Suppose we have a table with two fields: key and value.
We wish to find the mean of values whose key matches the most recent key.
Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only thosewith the right key.

\begin{code}
   let k    = newest key
in let avg  = filter (key == k) in mean value
in avg
\end{code}

This will not typecheck in Icicle, as @newest key@ has the modality @Aggregate@, while the filter condition @key == k@ attempts to check if an @Element@ is equal to an @Aggregate@.

This query could be translated to another synchronous streaming language such as {\sc Lustre}\cite{halbwachs1991synchronous}, and would typecheck.
However, the semantics of the translated query will not be what was originally desired.
Rather than checking each key against the most recent key as-of the end of the stream,
it will check each key against the current most recent key, which is the key itself.
This means the filter predicate is always be true, and thus returns the mean of the entire stream.
These semantics may be surprising to the novice stream programmer.

Hopefully our programmer is not too discouraged by the @Element /= Aggregate@ type error, and pushes on.
They now know that their query, as formulated, requires multiple passes over the data: the problem now is to reformulate it as a single pass.
With a little ingenuity, we can rewrite it as such: we can group by the key and perform the mean for each group.
After the group, we can perform a lookup by the most recent key.
This requires storing and computing the means for all keys despite only needing one at the end, so we are assuming that the number of keys are bounded in some way.

\begin{code}
   let k    = newest key
in let avgs = group  key in mean value
in avgs ! k
\end{code}

\subsection{Bounded buffer restriction}
As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
Any operation which requires a buffer must be bounded in size, as an unbounded buffer could potentially grow too large to fit in memory.

Here is an example stream program that requires unbounded buffering:
it takes an input stream @xs@, and filters it into those above zero, and those below zero.
These two filtered streams are then joined together pairwise, so the first positive element is paired with the first negative element, and so on.
The program, written in Haskell:

\begin{code}
zip (filter (>0) xs) (filter (<0) xs)
\end{code}

This program requires an unbounded buffer: if the input stream contains ten positive values followed by one negative value, all ten of the positive values must be held onto until the negative value is seen.
Similarly, if the entire stream is positive, all of the elements must be retained until the very end, just in case a negative value shows up.

In Icicle, we outlaw this kind of program by implicitly threading the input stream through operations.
The streams themselves are not materialised in the program: stream operations like @fold@ and @filter@ do not take parameters of the streams, but instead operate on the context stream.
By removing the explicit stream parameter, stream elements can only be joined from the same context, when both elements are available.

For example, in the query @filter p in mean value@, the @mean value@ is only applied to stream values which satisfy the predicate @p@.

This is a different approach than existing synchronous streaming languages\cite{mandel2010lucy} and flow fusion\cite{lippmeier2013data}, which perform clock analysis or `rate inference'.
Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together.
Filters produce a stream with a different output clock to its input, so different filters cannot be zipped together.
Our system requires no clock analysis, but is less expressive as streams are not `first class'.

### 2016-05-12 2016fhpc/section/03-Core.tex
introduction and first section (Source), start on Core section

--- 2016-04-07 2016fhpc/section/02-Core.tex
@@
 %!TEX root = ../Main.tex
-\section{Icicle Core}
+\section{Intermediate language}
 \label{s:IcicleCore}
+
+\begin{figure}
+
+\begin{code}
+Exp   := x | Prim Exp...
+
+Plan  := query  x : T
+       before { x : T = Exp... }
+       folds  { x : T = Exp then Exp... }
+       after  { x : T = Exp... }
+       return { x : T = x... }
+
+T     := bool | int | map T T
+\end{code}
+
+\caption{Grammar}
+\label{fig:core:grammar}
+\end{figure}
+
+The intermediate language can be thought of as similar to a query plan in other databases.
+Queries in the source language are converted to query plans, and then query plans on the same table are fused together.
+Common subexpression elimination, partial evaluation and other optimisations are performed at the query plan level.
+
+The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
+Expressions can be variables or primitive applications: function applications are not allowed in expressions here, as their definitions are inlined before converting to query plans.
+The @Plan@ itself is split into stages of computation, which correspond to the modal stages in the Source query.
+The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
+The next stage, @folds@, defines the @Element@ computations and how they are converted into @Aggregate@ computations.
+Next are @after@ which correspond to @Aggregate@ computations, occurring after the entire stream has been seen.
+Finally, the @return@ specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.
 
@@
 \end{code}
-
-\begin{figure}
-
-\begin{code}
-N     := a,b,...
-FoldN := a,b,...
-
-Exp   := N
-       | FoldN  | new FoldN
-       | Exp Exp
-       | \N. Exp
-
-Bind  := fold FoldN : T = Exp then Exp;
-       | let  N     : T = Exp;
-
-Stage := stage { Bind...  }
-Query := loop N : T { Stage... }
-
-\end{code}
-
-\caption{Grammar}
-\label{fig:grammar}
-\end{figure}
-
-\newcommand\TypecheckX[3]
-{                   #1
-        ~\vdash~    #2~:~#3
-}
-
-\newcommand\TypecheckS[3]
-{                   #1
-        ~\vdash~    #2~\dashv~#3
-}
-
-\begin{figure*}
-
-$$
-\boxed{\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}}
-$$
-
-$$
-\ruleAx
-{
-  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{\}}{\emptyset}
-}{QueryNil}
-\ruleIN
-{
-  \TypecheckS{\Gamma,~x~:~\tau}{s}{\Gamma'}
-  \quad
-  \TypecheckS{\Gamma,~\Gamma'}{@loop@~x~:~\tau~\{s'\}}{\Gamma''}
-}{
-  \TypecheckS{\Gamma}{@loop@~x~:~\tau~\{s;~s'\}}{\Gamma' \cup \Gamma''}
-}{QueryCons}
-$$
-
-$$
-\boxed{\TypecheckS{\Gamma}{\mi{Stage}}{\Gamma'}}
-$$
-
-$$
-\ruleAx
-{
-  \TypecheckS{\Gamma}{@stage@~\{\}}{\emptyset}
-}{StageNil}
-\ruleIN
-{
-  \TypecheckS{\Gamma,~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}
-  \quad
-  \TypecheckX{\Gamma}{x}{\tau}
-}
-{
-  \TypecheckS{\Gamma}{@stage@~\{@let@~x~:~\tau~=~e;~s'\}}{\Gamma',~x~:~\tau}
-}{StageLet}
-$$
-
-$$
-\begin{array}{l}
-
-\ruleIN
-{
-  \TypecheckS{\Gamma,~x~:~\tau,~@new@~x~:~\tau}{@stage@~\{s'\}}{\Gamma'}
-
-  \quad
-  \TypecheckX{\Gamma}{z}{\tau}
-  \quad
-  \TypecheckX{\Gamma~\cup~\mi{folds}}{k}{\tau}
-}
-{
-  \TypecheckS{\Gamma}{@stage@~\{@fold@~x~:~\tau~=~z~@then@~k;~s'\}}{\Gamma',~x~:~\tau,~@new@~x~:~\tau}
-}{StageFold}
-
-  \\
-  \mbox{where} \\
-  \quad
-  \mi{folds}~=~\{v~:~\tau'~|~@new@~v~:~\tau'~\in~\Gamma'\}
-  \end{array}
-$$
-
-\caption{Typing rules}
-\label{fig:typing}
-\end{figure*}
-
-The full grammar is in figure~\ref{fig:grammar}, however primitives and types have been omitted.
-It is assumed that there are some primitives such as @1@, @+@ and types such as @Int@, but the elements do not actually matter.
-
-The typing rules are in figure~\ref{fig:typing}.
-The rules shown only cover the staging and folds.
-As with the grammar, we assume there is a typing rule $\TypecheckX{\Gamma}{\mi{Exp}}{\tau}$ which checks an expression under the given environment.
-
-The first judgment, $\TypecheckS{\Gamma}{\mi{Query}}{\Gamma'}$

### 2016-05-12 2016fhpc/section/04-Benchmarks.tex
introduction and first section (Source), start on Core section

--- 2014-06-22 2014betterfusionforfilters/section/05-Benchmarks.tex
@@
 
-\begin{figure*}
-$$\begin{array}{c}
-
-\begin{tabular}{lrrrrrrrr}
-                & \multicolumn{2}{c}{Unfused}         & \multicolumn{2}{c}{Stream}
-                & \multicolumn{2}{c}{Megiddo} &\multicolumn{2}{c}{\textbf{Ours}} \\
-                & Time & Loops   & Time & Loops      & Time & Loops & Time & Loops   \\
-\hline
-Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
-Closest points  & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
-QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
-\end{tabular}
-
-\end{array}$$
-\caption{Benchmark results}
-\label{f:BenchResults}
-\end{figure*}
-
-This section discusses three representative benchmarks, and gives the full ILP program of the first.
-These benchmarks highlight the main differences between our fusion mechanism and related work.
-The runtimes of each benchmark are summarized in Figure~\ref{f:BenchResults}.
-We report times for: the unfused case where each operator is assigned to its own cluster; the clustering implied by stream fusion~\cite{coutts2007streamfusion}; the clustering chosen by Megiddo~\cite{megiddo1998optimal}, and the clustering chosen by our system.
-
-For each benchmark we report the runtimes of hand-fused C code based on the clustering determined by each algorithm.
-Although we also have an implementation of our Data Flow Fusion system in terms of a GHC plugin~\cite{lippmeier2013flow}, we report on hand-fused C code to provide a fair comparison to related work.
-As mentioned in~\cite{lippmeier2013flow}, the current Haskell stream fusion mechanism introduces overhead in terms of a large number of duplicate loop counters, which increases register pressure unnecessarily.
-Hand fusing all code and compiling it with the same compiler (GCC) isolates the true cost of the various clusterings from low level differences in code generation.
-
-We use GLPK as an external solver, which is not particularly fast, but has readily-available Haskell bindings.
-For these small programs with few combinators, it produces results in under one hundred milliseconds.
-For larger programs with twenty or more combinators to be clustered, GLPK is inadequate.
-One randomly generated example with twenty-five combinators took over twenty minutes in GLPK, while the commercial solver CPLEX was able to solve the same example in a second, and another open source solver CBC took one hundred seconds.
-For the same example, the unoptimised program with all constraints took over twenty minutes in CBC, while CPLEX still only took one and a half seconds.
-
-The benchmark programs are at \url{https://github.com/amosr/papers/tree/master/2014betterfusionforfilters/benches}.
-
-% The three benchmarks are the @normalize2@ example, finding the closest pair of points, and quadtree.
-The benchmark programs are all hand-written, hand-fused C code based on the clustering.
-Each program was run with the same input five times, and the minimum runtime was used.
-Runtimes and the number of loops for each clustering are shown in Figure~\ref{f:BenchResults}.
-In all cases, our clustering performs better than or as good as Megiddo's, and better than stream fusion and unfused.
-Interestingly, stream fusion's clustering for @normalize2@ performs better than Megiddo's, despite having more loops, as stream fusion is able to remove the intermediate array.
-
-% -----------------------------------------------------------------------------
-\subsection{Normalize2}
-To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
-\begin{code}
-  normalize2 :: Array Int -> Array Int
-  normalize2 xs
-   = let sum1 = fold   (+)  0   xs
-         gts  = filter (>   0)  xs
-         sum2 = fold   (+)  0   gts
-         ys1  = map    (/ sum1) xs
-         ys2  = map    (/ sum2) xs
-     in (ys1, ys2)
-\end{code}
-
-We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
-\[ \{ \{sum1, gts, sum2\}
- , \{sum1, ys2\}
- , \{gts, sum2, ys1\}
- , \{ys1, ys2\} \} \]
-
-The complete ILP program is shown below. Note that in the objective function the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.
-
-\begin{tabbing}
-MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
-Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
-           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
-           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $
-\\[1ex]
-Subject to 
-    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
-    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
-    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
-    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
-    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
-    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
-\\[1ex]
-    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
-\\[1ex]
-    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
-    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
-\\[1ex]
-    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
-\\[1ex]
-    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
-    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
-    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
-\end{tabbing}
-\noindent
-
-One minimal solution to this is:
-\begin{tabbing}
-MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
-    \> $=$ \> $0$ \\
-$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
-    \> $=$ \> $1$ 
-\\[1ex]
-$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
-    \> $=$ \> $0$ \\
-$\pi_{ys1 },~ \pi_{ys2 }$
-    \> $=$ \> $1$ 
-\\[1ex]
-$c_{gts},~ c_{ys1},~ c_{ys2}$           
-    \> $=$ \> $0$
-\end{tabbing}
-
-This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.
-Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
-For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints.
-Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}.
-These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.
-
-We will now compare the clustering produced by our system, with the one implied by stream fusion.
-As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship.
-The corresponding values of the $x_{ij}$ variables are:
-\begin{tabbing}
-MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$x_{gts, sum2}$
-    \> $=$ \> $0$ \\
-$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
-    \> $=$ \> $1$
-\end{tabbing}
-
-We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
-\begin{tabbing}
-MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
-$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
-    \> $=$ \> $0$ \\
-$\pi_{ys1 }, \pi_{ys2 }$
-    \> $=$ \> $1$ \\
-$c_{gts}, c_{ys1}, c_{ys2}$           
-    \> $=$ \> $0$
-\end{tabbing}
-
-Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused.
-
-% The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.
-
-% -----------------------------------------------------------------------------
-\subsection{Closest Points}
-The closest points benchmark is a divide-and-conquer algorithm that finds the closest pair of 2-dimensional points in an array.
-We first find the midpoint along the Y-axis, and filter the remaining points to those above and below the midpoint.
-We then recursively find the closest pair of points in the two halves, and merge the results.
-As the filtered points are passed directly to the recursive call, there is no further opportunity to fuse them, and our clustering is the same as returned by Megiddo's algorithm.
-However, our clustering generates both filtered arrays in a single loop, unlike stream fusion that requires a separate loop for each.
-
-% -----------------------------------------------------------------------------
-\subsection{QuadTree}
-The QuadTree benchmark recursively builds a 2-dimensional space partitioning tree from an array of points.
-At each step the array of points is filtered into four 2-dimensional boxes.
-As with the closest points algorith, there are no further opportunities for fusing the filtered results, and our clustering is the same as Megiddo's.
-However, our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.
-
-% -----------------------------------------------------------------------------
-\subsection{QuickHull}
-The core of the QuickHull algorithm is shown below: given a line and an array of points, we filter the points to those above the line, and also find the point farthest from that line.
-
-\begin{code}
-hull :: (Point,Point) -> Array Point -> Array Point
-hull line@(l,r) pts
- = let pts' = filter (above   line) pts
-       ma   = fold   (maxFrom line) pts'
-   in (hull (l, ma) pts') ++ (hull (ma, r) pts')
-\end{code}
-
-Stream fusion cannot fuse the @pts'@ and @ma@ bindings because @pts'@ is referred to multiple times and thus cannot be inlined.
-Megiddo's algorithm also cannot fuse the two bindings because their iteration sizes are different.
-If the @ma@ binding was rewritten to operate over the @pts@ array instead of @pts'@, Megiddo's formulation would be able to fuse the two, and the overall program would give the same result.
-However, this performance behavior is counter intuitive because @pts'@ is likely to be smaller than @pts@, so in an unfused program the original version would be faster.
-Our system fuses both versions.
+Run some benchmarks.

### 2016-05-12 2016fhpc/section/XX-Elimination.tex
introduction and first section (Source), start on Core section

--- 2016-04-12 2016fhpc/section/03-Elimination.tex
@@
 
+\subsection{Stage-local duplicate removal}
+Here is an example of a stage with two counters.
+\begin{code}
+stage {
+  fold a = 0 then c + 1
+  fold b = 0 then c + 1
+  fold c = 0 then a + b
+}
+Execution:
+[ a = 0, b = 0, c = 0 ]
+[ a = 1, b = 1, c = 0 ]
+[ a = 1, b = 1, c = 2 ]
+[ a = 3, b = 3, c = 2 ]
+[ a = 3, b = 3, c = 6 ]
+[ a = 7, b = 7, c = 6 ]
+[ a = 7, b = 7, c = 14 ]
+\end{code}
+
+The bindings for @a@ and @b@ here are identical, and so one can be replaced with the other, by substituting @b := a@ and removing the binding for @b@.
+\begin{code}
+stage {
+  fold a = 0 then c + 1
+  fold c = 0 then a + a
+}
+\end{code}
+
+Another example is a single count, spread across two recursive bindings.
+Here, @a@ and @b@ depend on each other, but both have the same body modulo alpha.
+\begin{code}
+stage {
+  fold a = 0 then b + 1
+  fold b = 0 then a + 1
+}
+Execution:
+[ a = 0, b = 0 ]
+[ a = 1, b = 1 ]
+[ a = 2, b = 2 ]
+\end{code}
+
+One can place them into equivalence groups based on their bodies: @a@ and @b@ go in the same equivalence group, and so any reference to @a@ or @b@ refers to the equivalence group containing both.
+Equivalence groups are then squashed together, with a single binding for each equivalence group.
+\begin{code}
+stage {
+  fold a = 0 then a + 1
+}
+\end{code}
+
 \subsection{Strongly connected components}
@@
 We can look at the topological ordering of stages as layers of the same depth.
-In the @count/sum/ab@ example, @count@ and @sum@ are together on the top layer, while @ab@ is on the second layer.
+In the @count/sum/ab@ example, @count@ and @sum@ are together on the top layer, while the stage containing @a@ and @b@ is on the second layer.
 
-Each stage need only be compared with those on the same layer, and those on the layer directly above.
+When checking for duplicates, each stage need only be compared with those on the same layer, and those on the layer directly above.
 In fact, of those on the layer above, it only needs to be compared with those it refers to.
-\TODO{Justify this further.}
-Because the stages have already been split apart, we also need only check against those with equal or greater number of bindings.
+It is not necessary to check more two layers above, for example, because we know two things:
+this stage refers to a binding in the layer above, while any stage two layers above cannot possibly refer to a lower stage.
+However, in order to be duplicates, they must refer to the same things, or duplicates of the same things.
 
-\subsection{Duplicate removal}
-Because the stages have already been split apart, we can only replace a stage
+\subsection{Inter-stage duplicate removal}
+To check if one stage is a duplicate of another, we go through each binding of the first stage, checking if there is an alpha-equivalent binding in the other stage.
+We also create a substitution between the two stages.
+This substitution is later performed on the rest of the program, so that after the duplicate is removed, references to the duplicate are updated to refer to the first stage.
+
+During the process of checking, there may be multiple possible substitutions that appear to work.
+The following program defines three mutually recursive counters; two of the counters increment by one, and the last increments by two.
+Because of the mutual recursion, this ends up making each counter increment by a repeating pattern: one, one, two.
+\begin{code}
+stage {
+  fold a = 0 then b + 1
+  fold b = 0 then c + 1
+  fold c = 0 then a + 2
+}
+Execution:
+[ a = 0, b = 0, c = 0 ]
+[ a = 1, b = 1, c = 2 ]
+[ a = 2, b = 3, c = 3 ]
+[ a = 4, b = 4, c = 4 ]
+[ a = 5, b = 5, c = 6 ]
+[ a = 6, b = 7, c = 7 ]
+\end{code}
+
+Now, suppose we wish to check if the following stage is a duplicate of the previous counters.
+Note that these bindings are actually in a different order to the previous bindings, whereas the equivalent order would be @x@, @y@ then @z@.
+\begin{code}
+stage {
+  fold y = 0 then z + 1
+  fold x = 0 then y + 1
+  fold z = 0 then x + 2
+}
+\end{code}
+
+We start by inpsecting @y@, the first binding, and check against all bindings in the other set.
+We wish to find a substitution from the names @x@, @y@ and @z@ to the names @a@, @b@ and @c@.
+The substitution cannot mention any names other than those bound by the two stages.
+
+There appear to be two possibilities for @y@: it is alpha-equivalent to both @a@ and @b@, producing two different substitutions:
+\begin{code}
+(Y1)
+y := a
+z := b
+(Y2)
+y := b
+z := c
+\end{code}
+
+We continue checking the remaining bindings, by producing substitutions for @x@.
+Like @y@, @x@ fits both @a@ and @b@, with two possible substitutions:
+\begin{code}
+(X1)
+x := a
+y := b
+(X2)
+x := b
+z := c
+\end{code}
+
+We can cross-product these substitutions together, composing (Y1) with (X1) and so on, producing four possible substitutions.
+The first two contain contradictions and are discarded immediately.
+The next two seem plausible, so far.
+\begin{code}
+(Y1X1)
+y := a
+y := b
+(Y1X2)
+z := b
+z := c
+(Y2X1)
+x := a
+y := b
+z := c
+(Y2X2)
+x := b
+y := b
+z := c
+\end{code}
+
+Now we check the final binding, @z@, which produces only one possible substitution:
+\begin{code}
+(Z1)
+z := c
+x := a
+\end{code}
+
+Composing (Z1) with (Y2X1) and (Y2X2), we find that (Y2X1Z1) is the only remaining possibility, since (Z1) and (Y2X2) are contradictory.
+\begin{code}
+(Y2X1Z1)
+x := a
+y := b
+z := c
+(Y2X2Z1)
+x := a
+x := b
+\end{code}
+
+We can now remove the stage containing @x@, @y@ and @z@, while performing this substitution over the remainder of the program, so that any references to @x@ become references to @a@, and so on.
+
+If, for a different pair of stages, there were multiple possible substitutions after checking all bindings against the others, we could have chosen any of the substitutions.
+They are all equivalent.

### 2016-05-15 2016fhpc/section/03-Core.tex
some example query plans, some preliminary benchmark info

@@
 
-Plan  := query  x : T
+Plan  := query  x with (x : T)...
        before { x : T = Exp... }
@@
 
-Horizontal fusion for imperative loops is relatively easily, when the loops have the same number of iterations.
-The problem with imperative loops occurs when trying to remove duplicate computations.
-Firstly, the `definition' of a computation is split across multiple places: because the initial value of an accumulator and its update expression are in separate parts of the program, some analysis must be done to recover these.
-For example, in the following loop, @a@ and @b@ have equivalent `update' expressions, but could denote very different computations depending on the implementation of @function@.
+One of the simplest queries to convert to a query plan is a count.
+The query plan header specifies the table and the type of each column in the table.
+This defines a single fold called @count@ which starts at zero, and is incremented for each element.
+At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
 \begin{code}
-a = 0;
-b = 1;
-for (...) {
-  a = function(a);
-  b = function(b);
-}
+count_query =
+query stocks
+in    count
+
+==>
+
+query stocks with (open : Int) (close : Int)
+before { }
+folds  { count : Int = 0 then count + 1 }
+after  { }
+return { count_query : Int = count }
 \end{code}
 
-Secondly, the order of statements in a loop affects the meaning: when accumulators are mutually dependent on each other, one can reorder the statements to produce a different, but still valid program.
-These two loops are not equivalent:
+Computing the sum is only slightly more complicated.
+Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
 \begin{code}
-a = 0;
-b = 1;
-for (...) {
-  a = a + b;
-  b = a - b;
-}
-for (...) {
-  b = a - b;
-  a = a + b;
-}
+sum_open =
+query stocks
+in    sum open
+
+==>
+
+query stocks with (open : Int) (close : Int)
+before { }
+folds  { sum : Int = 0 then sum + open }
+after  { }
+return { sum_open : Int = sum }
 \end{code}
 
-Neither of these problems are insurmountable, but the analysis required certainly complicates our goal of removing duplicate computation after fusion.
-
-With this in mind, we will introduce our intermediate language based on what we call ``fold nests'', as opposed to loop nests.
-Firstly, each accumulator should be defined in one place: this means that if two accumulators are alpha equivalent, they are equivalent.
-Secondly, reordering statements (or accumulators) should not affect the meaning: two programs are equivalent if they contain equivalent sets of bindings, regardless of ordering.
-
-We now convert the first example to folds:
+In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
+Here we use two folds, one for the sum, and one for the count.
+While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
 \begin{code}
-a = 0;
-b = 1;
-for (...) {
-  a = function(a);
-  b = function(b);
-}
+mean_open =
+query stocks
+in    sum open / count
 
 ==>
-loop ... {
-  stage {
-    fold a = 0
-        then function(a);
-  }
-  stage {
-    fold b = 1
-        then function(b);
-  }
-}
+
+query stocks with (open : Int) (close : Int)
+before { }
+folds  { sum       : Int = 0 then sum + open 
+         count     : Int = 0 then count + 1 }
+after  { mean      : Int = sum / count }
+return { mean_open : Int = mean }
 \end{code}
 
-Each fold binding must be grouped inside a @stage@ block, which allows mutual recursion across folds.
-In this case, there is no mutual recursion, so the stages only contain one binding each.
-The first fold is given the name @a@ with @fold a@.
-Its initial value comes straight after the equals sign, so @a@ starts with a value of @0@.
-The update or `kons' expression comes after @then@, and in the update expression any reference to @a@ refers to the current value of the fold.
-
-It now becomes easier to see that the two folds are not equivalent, as the entire definitions are in one place.
-
-The next example requires mutual recursion.
+If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
+The names of the returns and the column names stay the same.
+Query plans can only be fused if they act on the same table.
 \begin{code}
-a = 0;
-b = 1;
-for (...) {
-  a = a + b;
-  b = a - b;
-}
-
-==>
-loop ... {
-  stage {
-    fold a = 0
-        then a + b;
-    fold b = 1
-        then new a - b;
-  }
-}
+query stocks with (open : Int) (close : Int)
+before { }
+folds  { c1 : Int = 0 then c1 + 1
+         c2 : Int = 0 then c2 + 1
+         s1 : Int = 0 then s1 + open 
+         s2 : Int = 0 then s2 + open }
+after  { m1 : Int = s2 / c2 }
+return { count_query : Int = c1
+         sum_open    : Int = s1
+         mean_open   : Int = m1 }
 \end{code}
 
-Here, the update expression for @a@ refers to a later fold, @b@.
-Update expressions can refer to later fold bindings in the same stage, as well as all earlier fold bindings.
-The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
-The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @new a@.
-This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.
+The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
+We can perform fairly standard common subexpression elimination to remove these.
+\begin{code}
+query stocks with (open : Int) (close : Int)
+before { }
+folds  { c1 : Int = 0 then c1 + 1
+         s1 : Int = 0 then s1 + open }
+after  { m1 : Int = s1 / c1 }
+return { count_query : Int = c1
+         sum_open    : Int = s1
+         mean_open   : Int = m1 }
+\end{code}
 
-Another way to think of the new value restriction above is disallowing cycles in the references to new values:
-when @b@ references @new a@, the new value of @a@, it means that @new a@ must be computed before @new b@ can be.
-If there were a cycle where @new a@ also depended upon @new b@, there would be no order to compute them in.
-Disallowing cycles ensures an execution order.
+The interesting thing here is just how simple fusion is.
+Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
+The single-pass restriction means that any source query can be expressed as a single fold over the data.
+By keeping the folds separate we have exposed more opportunities for common subexpression elimination.
 
-By making the distinction between `old value' and `new value' explicit, the ordering in the program no longer has any effect on the semantics.
-After converting the reordered loop with different semantics, the two programs are no longer alpha-equivalent:
-\begin{code}
-a = 0;
-b = 1;
-for (...) {
-  b = a - b;
-  a = a + b;
-}
-
-==>
-loop ... {
-  stage {
-    fold b = 1
-        then a - b;
-    fold a = 0
-        then a + new b;
-  }
-}
-\end{code}
+An efficient imperative loop can be generated from this query plan rather easily, in a similar way as previous work on flow fusion\cite{lippmeier2013data}.

### 2016-05-15 2016fhpc/section/04-Benchmarks.tex
some example query plans, some preliminary benchmark info

@@
 
+We are using Icicle in production, over medium-sized datasets that fit on a single disk.
+
+One evaluation we performed was replacing a hand-written R script which had been running in production.
+This R script works over around three hundred gigabytes of PSV data.
+It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.
+
+The R script for this takes around fifteen hours to run and is 3,566 lines, 2,311 of which are code.
+In contrast, our Icicle queries take eleven minutes to run, and the dictionary describing the queries is 191 lines of code.
+
+\begin{code}
+          R       Icicle
+Minutes   900         11
+LOC       2,311      191
+MB/s      6.6        543
+\end{code}
+
 Run some benchmarks.
+
+\begin{code}
+        grep "^$"   wc    wc -l   Icicle
+MB/s      74        251   775       1,060
+\end{code}

### 2016-05-15 2016fhpc/section/XX-Elimination.tex
some example query plans, some preliminary benchmark info

@@
 \label{s:Elimination}
+
+Horizontal fusion for imperative loops is relatively easily, when the loops have the same number of iterations.
+The problem with imperative loops occurs when trying to remove duplicate computations.
+Firstly, the `definition' of a computation is split across multiple places: because the initial value of an accumulator and its update expression are in separate parts of the program, some analysis must be done to recover these.
+For example, in the following loop, @a@ and @b@ have equivalent `update' expressions, but could denote very different computations depending on the implementation of @function@.
+\begin{code}
+a = 0;
+b = 1;
+for (...) {
+  a = function(a);
+  b = function(b);
+}
+\end{code}
+
+Secondly, the order of statements in a loop affects the meaning: when accumulators are mutually dependent on each other, one can reorder the statements to produce a different, but still valid program.
+These two loops are not equivalent:
+\begin{code}
+a = 0;
+b = 1;
+for (...) {
+  a = a + b;
+  b = a - b;
+}
+for (...) {
+  b = a - b;
+  a = a + b;
+}
+\end{code}
+
+Neither of these problems are insurmountable, but the analysis required certainly complicates our goal of removing duplicate computation after fusion.
+
+With this in mind, we will introduce our intermediate language based on what we call ``fold nests'', as opposed to loop nests.
+Firstly, each accumulator should be defined in one place: this means that if two accumulators are alpha equivalent, they are equivalent.
+Secondly, reordering statements (or accumulators) should not affect the meaning: two programs are equivalent if they contain equivalent sets of bindings, regardless of ordering.
+
+We now convert the first example to folds:
+\begin{code}
+a = 0;
+b = 1;
+for (...) {
+  a = function(a);
+  b = function(b);
+}
+
+==>
+loop ... {
+  stage {
+    fold a = 0
+        then function(a);
+  }
+  stage {
+    fold b = 1
+        then function(b);
+  }
+}
+\end{code}
+
+Each fold binding must be grouped inside a @stage@ block, which allows mutual recursion across folds.
+In this case, there is no mutual recursion, so the stages only contain one binding each.
+The first fold is given the name @a@ with @fold a@.
+Its initial value comes straight after the equals sign, so @a@ starts with a value of @0@.
+The update or `kons' expression comes after @then@, and in the update expression any reference to @a@ refers to the current value of the fold.
+
+It now becomes easier to see that the two folds are not equivalent, as the entire definitions are in one place.
+
+The next example requires mutual recursion.
+\begin{code}
+a = 0;
+b = 1;
+for (...) {
+  a = a + b;
+  b = a - b;
+}
+
+==>
+loop ... {
+  stage {
+    fold a = 0
+        then a + b;
+    fold b = 1
+        then new a - b;
+  }
+}
+\end{code}
+
+Here, the update expression for @a@ refers to a later fold, @b@.
+Update expressions can refer to later fold bindings in the same stage, as well as all earlier fold bindings.
+The value referred to is the value of the binding at the start of the iteration, before any updates have occurred.
+The new value of earlier folds, after being updated, can be referred to by suffixing a prime to the name: in the binding for @b@, the updated @a@ is referred to by @new a@.
+This restriction of only accessing the new value of earlier folds is akin to the causality restriction in dataflow languages\cite{mandel2010lucy}.
+
+Another way to think of the new value restriction above is disallowing cycles in the references to new values:
+when @b@ references @new a@, the new value of @a@, it means that @new a@ must be computed before @new b@ can be.
+If there were a cycle where @new a@ also depended upon @new b@, there would be no order to compute them in.
+Disallowing cycles ensures an execution order.
+
+By making the distinction between `old value' and `new value' explicit, the ordering in the program no longer has any effect on the semantics.
+After converting the reordered loop with different semantics, the two programs are no longer alpha-equivalent:
+\begin{code}
+a = 0;
+b = 1;
+for (...) {
+  b = a - b;
+  a = a + b;
+}
+
+==>
+loop ... {
+  stage {
+    fold b = 1
+        then a - b;
+    fold a = 0
+        then a + new b;
+  }
+}
+\end{code}
 

### 2016-05-15 2016fhpc/section/04-Benchmarks.tex
more bench

@@
 
-We are using Icicle in production, over medium-sized datasets that fit on a single disk.
+We are currently using Icicle in production, over medium-sized datasets that fit on a single disk.
+Initial results have been very promising, and we are currently working towards a distributed implementation for larger datasets.
 
@@
 In contrast, our Icicle queries take eleven minutes to run, and the dictionary describing the queries is 191 lines of code.
+
+It is also important to note that, as a very constrained streaming language, memory usage of Icicle queries is more or less constant in the input size, and runtime is linear.
+This is very important, as new data is received regularly and the input size grows.
+This does not appear to be the case with the R script, which continually requires larger computers with more memory, in order to finish in a reasonable time.
+One of the benefits of Icicle is allowing data scientists to focus on queries, without having to worry about performance.
 

### 2016-05-18 2016fhpc/section/01-Introduction.tex
Typechecking etc

@@
 \begin{code}
-query stocks
-in filter open > close
-in count.
-
-query stocks
-in filter open < close
-in count.
-
-query stocks
-in filter open > close
-in sum open / count.
+table stocks { open : Int; close : Int; }
+query more = filter open > close in count;
+query less = filter open < close in count;
+query mean = filter open > close in sum open / count;
 \end{code}

### 2016-05-18 2016fhpc/section/02-Source.tex
Typechecking etc

@@
 
-\begin{figure}
+\input{figures/Source-Grammar.tex}
+\input{figures/Source-Type.tex}
+\input{figures/Source-Eval.tex}
 
-\begin{code}
-Exp   := x | Exp Op Exp | x Exp... | Query
+The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
+Value types are given in $T$ and can be numbers, booleans or maps.
+Modal types, $T_M$ can be a pure value type, or a modality associated with a value type.
+Function types, $T_\to$, can be either a non-function modal type, or a function from modal type to function type.
+As Icicle is a first-order language, function types are not value types.
 
-Query := fold x = Exp then Exp
-       | let  x = Exp          in Query
-       | filter   Exp          in Query
-       | group    Exp          in Query
-       | Exp
+The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
+$\mi{Exp}$ defines the expressions used as query and function bodies.
+$\mi{Prim}$ defines the primitive types.
+Function definitions are given in $\mi{Fun}$.
+Query definitions are given in $\mi{Query}$.
+$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
 
-Top   := query x in Query
-Fun   := function x x... = Query
-\end{code}
-
-\caption{Icicle Grammar}
-\label{fig:icicle:grammar}
-\end{figure}
-
-The grammar for Icicle is given in figure~\ref{fig:icicle:grammar}.
-@Exp@ and @Query@ are mutually recursive, defining the body of a query.
-@Top@ defines a top-level query over a particular table, whereas @Fun@ is used for function definitions.
-
-Expressions can be variable names, operators such as addition and division, function application, or nested queries.
-Function application @x Exp...@ is the name of a top-level function applied to any number of arguments.
+\TODO{update this to fit new grammar}
+Expressions can be variable names, primitive operators such as addition and division, function application, or nested queries.
+Function application @x Exp*@ is the name of a top-level function applied to any number of arguments.
+Primitive application @Prim Exp*@ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
 
@@
 \begin{code}
-function sum value
+function sum (value : Element Int)
  = fold s = 0 then s + value
@@
 \begin{code}
-function mean value
+function mean (value : Element Int)
  = sum value / count
@@
 sum open > 1  : Aggregate Bool
+\end{code}
+
+Folds have a name and two expressions.
+The first expression, the initialiser, is pure.
+The second expression, the update, has the fold's name bound to an @Element@ of the previous value.
+The type of the expression must be @Element@.
+The final return of the fold is an @Aggregate@.
+\begin{code}
+fold x = T then Element T : Aggregate T
+\end{code}
+
+Let-bindings are independent of the modality.
+
+For filters, the predicate is an @Element@, which could be thought of as a stream of booleans.
+The rest of the query must be an @Aggregate@; that is, it must return a scalar rather than a stream, to enforce the bounded buffer restriction explained in \sref{s:IcicleSource:bounded}.
+\begin{code}
+filter (Element Bool) in (Aggregate T)
+\end{code}
+
+Grouping is similar to filters; the key must be an @Element@, while the group value must be an @Aggregate@.
+\begin{code}
+group (Element Key) in (Aggregate Value)
 \end{code}
@@
 in let avgs = group  key in mean value
-in avgs ! k
+in mapLookup avgs k
 \end{code}
@@
 \subsection{Bounded buffer restriction}
+\label{s:IcicleSource:bounded}
 As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.

### 2016-05-18 2016fhpc/section/03-Core.tex
Typechecking etc

@@
 \begin{code}
-Exp   := x | Prim Exp...
+Exp   := x | Prim Exp*
 
-Plan  := query  x with (x : T)...
-       before { x : T = Exp... }
-       folds  { x : T = Exp then Exp... }
-       after  { x : T = Exp... }
-       return { x : T = x... }
-
-T     := bool | int | map T T
+Plan  := plan   x { (x : T;)* }
+       before { x : T = Exp;* }
+       folds  { x : T = Exp then Exp;* }
+       after  { x : T = Exp;* }
+       return { x : T = x;* }
 \end{code}
@@
 \begin{code}
-count_query =
-query stocks
-in    count
+table stocks { open : Int; close : Int; }
+query count_query = count
 
@@
 
-query stocks with (open : Int) (close : Int)
+plan stocks { open : Int; close : Int; }
 before { }
@@
 \begin{code}
-sum_open =
-query stocks
-in    sum open
+query sum_open = sum open
 
@@
 \begin{code}
-mean_open =
-query stocks
-in    sum open / count
+query mean_open = sum open / count
 
@@
 \begin{code}
-query stocks with (open : Int) (close : Int)
+plan stocks { open : Int; close : Int; }
 before { }

### 2016-05-21 2016fhpc/section/01-Introduction.tex
ok

@@
 \item
-we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
+we show how Icicle guarantees any two queries on the same table can be fused together, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+
 \item
 we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
+
 \item

### 2016-05-21 2016fhpc/section/01-Introduction.tex
ok

--- 2016-05-18 2016fhpc/section/01-Introduction.tex
@@
 \item
-we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
+we show a simple typesystem that guarantees any two queries on the same table can be fused together, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+
 \item
 we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
+
 \item

### 2016-05-22 2016fhpc/section/01-Introduction.tex
Evaluation rules

--- 2016-05-18 2016fhpc/section/01-Introduction.tex
@@
 \item
-we describe restrictions on Icicle which guarantee any two queries on the same table can be fused together\sref{s:IcicleSource};
+we show a typesystem that guarantees any two queries on the same table can be fused together, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+
 \item
 we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
+
 \item

### 2016-05-22 2016fhpc/section/02-Source.tex
Evaluation rules

@@
 
-\TODO{update this to fit new grammar}
-Expressions can be variable names, primitive operators such as addition and division, function application, or nested queries.
-Function application @x Exp*@ is the name of a top-level function applied to any number of arguments.
-Primitive application @Prim Exp*@ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
-
-Queries can define folds over the input, let-bindings, filtering according to some predicate, or grouping by some key.
-The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
-Count can be expressed by the following fold:
-\begin{code}
-fold count = 0 then count + 1
-\end{code}
+Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
+The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
+Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
+Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
 
@@
 in  mean diff
+\end{code}
+
+The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
+Count can be expressed by the following fold:
+\begin{code}
+fold count = 0 then count + 1
 \end{code}
@@
 Group takes a key to group by, and a query to perform on the values of each grouping.
-This query groups by the company code, and counts the number of records in each company.
+This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
 \begin{code}
-group company in count
+group ((open - close) / 100) in count
 \end{code}
 
-@Fun@ is used for function definitions.
+$\mi{Fun}$ is used for function definitions.
 The @count@ function takes no arguments and returns a fold counting the number of elements.
@@
 function count
- = fold c = 0 then c + 1
+ = fold c = 0 then c + 1;
 \end{code}
@@
 function sum (value : Element Int)
- = fold s = 0 then s + value
+ = fold s = 0 then s + value;
 \end{code}
@@
 function mean (value : Element Int)
- = sum value / count
+ = sum value / count;
 \end{code}
+
+$\mi{Query}$ defines an aggregate query which must return a single value, and gives it a name:
+\begin{code}
+query count_all_records = count;
+\end{code}
+
+Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
+All queries here operate over the same input table, and will be fused into a single loop over the input data.
 
@@
 
-Some examples with their types:
-\begin{code}
-1             :         Int
-open          : Element Int
-open > 1      : Element Int
-sum           : Element Int -> Aggregate Int
-sum open      : Aggregate Int
-sum open > 1  : Aggregate Bool
-\end{code}
-
-Folds have a name and two expressions.
-The first expression, the initialiser, is pure.
-The second expression, the update, has the fold's name bound to an @Element@ of the previous value.
-The type of the expression must be @Element@.
-The final return of the fold is an @Aggregate@.
-\begin{code}
-fold x = T then Element T : Aggregate T
-\end{code}
-
-Let-bindings are independent of the modality.
-
-For filters, the predicate is an @Element@, which could be thought of as a stream of booleans.
-The rest of the query must be an @Aggregate@; that is, it must return a scalar rather than a stream, to enforce the bounded buffer restriction explained in \sref{s:IcicleSource:bounded}.
-\begin{code}
-filter (Element Bool) in (Aggregate T)
-\end{code}
-
-Grouping is similar to filters; the key must be an @Element@, while the group value must be an @Aggregate@.
-\begin{code}
-group (Element Key) in (Aggregate Value)
-\end{code}
-
-The following program attemps to find the number of entries where the open price is above the mean of the close price.
-This requires two passes over the data, as we do not know what the mean of all the data is until we have seen all the data.
-This program violates the single-pass restriction, and is outlawed by the typesystem.
-The left side of the @>@ has type @Element Int@, while the right side has type @Aggregate Int@, and these two are mutually exclusive.
-\begin{code}
-filter open > mean close
-in count
-(Error: (>) must be Element and Aggregate)
-\end{code}
-
 Here is an example to highlight one benefit of tracking these modalities.
@@
 We wish to find the mean of values whose key matches the most recent key.
-Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only thosewith the right key.
+Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.
 
@@
 Our system requires no clock analysis, but is less expressive as streams are not `first class'.
+
+\subsection{Type system}
+
+The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
+The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{T_M}$ means that under context $\Gamma$, the expression has a given value type and modality.
+$\TypecheckP{\mi{Prim}}{T_\to}$ associates a primitive with its function type.
+$\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
+The next two judgments,
+$\TypecheckS{\Gamma}{\mi{Fun}}{\Gamma}$ and
+$\TypecheckS{\Gamma}{\mi{Exp}}{\Gamma}$
+both takes an input environment and function or query, and returns an environment containing the function or query name and its type.
+Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.
+
+The rule (TcVar) performs variable lookup in the context, while (TcNat) shows that number constants have type $\NN$.
+(TcBox) shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
+The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.
+
+(TcPrimApp) and (TcFunApp) find the result type of primitive and function application.
+They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ to find the return type and modality of the application.
+The set-comprehension syntax $\{e_i\}_i$ means that this rule allows any number of arguments.
+
+(TcLet) is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
+
+(TcFold) finds the type of a @fold@ expression.
+The initial value of the fold must be pure.
+The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
+The final return type of the fold is an @Aggregate@.
+
+(TcFilter) requires its predicate to be a stream of booleans ($@Element@~\BB$).
+The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
+(TcGroup) is similar to filter, except returning a @Map@.
+
+The rules (PrimArith), (PrimRel) and (PrimLookup) give function types for each primitive.
+
+The next two rules show how function application boxes and unboxes modalities.
+
+(AppArgs) applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
+
+(AppRebox) is used when a pure function is applied to a non-pure value.
+If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
+Note that (TcBox) rule can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
+
+(CheckFun) inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
+
+(CheckQuery) checks a query's types, and makes sure it returns an @Aggregate@.
+
+Finally, (CheckTop) checks a whole top-level program.
+It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
+Then, each query is checked, with the table columns and all functions available.
+The final returned context is all query names and their types.
+
+\subsection{Evaluation}
+
+The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
+
+We define simple values $V$ as numbers, booleans, maps of values, or tuples of values.
+Next, stream values $V'$ correspond to modalities: pure computations are represented as a single value;
+@Element@s as a stream transformer from a heap to a value;
+and @Aggregate@s as folds defining an initial state and update and extract functions.
+
+The evaluation rules operate over a slightly modified expression type, $E'$, which is an $\mi{Exp}$ with $V'$ as values instead of just simple values and natural numbers.
+This is to simplify typechecking over $V'$, and as $V'$ only appear in restricted form for evaluation.
+We also assume that all functions have been inlined before evaluation.
+
+The judgment $\SourceStepX{E'}{V'}$ defines a big-step evaluation relation.
+
+(EVal) applies when the expression is alread a stream value.
+This corresponds, roughly, to the (TcNat) typechecking rule.
+
+(EBoxStream) and (EBoxFold) convert a pure value to a stream and fold, respectively.
+The stream transformer ignores the input stream and returns the pure value.
+The fold has a unit accumulator, and the extract function simply returns the pure value.
+
+Next (EPrimValue), (EPrimStream) and (EPrimFold) evaluate primitive applications.
+For pure values, the values are unboxed and the primitive is applied.
+For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
+Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
+At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
+Finally, the extract function extracts the final values of the input folds, and applies the primitive.
+
+The (ELet) rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
+
+(EFold) is one of the most complex rules, as it introduces recursion to the streams.
+It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
+In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
+This action-at-a-distance is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
+Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
+The extract function for this is the identity, simply returning the accumulator.
+
+(EFilter) evaluates its predicate to a stream of booleans, and the rest to a fold.
+The result is a fold which only applies the update function when the predicate stream is true.
+
+(EGroup) is rather similar to filter, except that it returns a map.
+Here, we start with an empty map of bottoms.
+At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
+We then apply the update function to the accumulator, and update that key in the map.
+For the extract function, we simply apply the extract to each element of the map.
+
+In order to execute these evaluation rules on an actual input table, a few steps are necessary.
+First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
+For example, $e[@open@~:=~@Stream@~(\lam{\sigma} \sigma~@open@)]$.
+Next, the query must be converted to a @Fold@.
+Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
+This occurs for every record in the table.
+Finally, the extract function is called, which gives the result of the query.

### 2016-05-23 2016fhpc/section/04-Benchmarks.tex
benchmark plot

@@
 
-\begin{code}
-          R       Icicle
-Minutes   900         11
-LOC       2,311      191
-MB/s      6.6        543
-\end{code}
+\begin{figure}
 
-Run some benchmarks.
+\begin{tikzpicture}
+\begin{axis}[
+	x tick label style={/pgf/number format/1000 sep=},
+	ylabel=MB/s,
+	enlargelimits=0.01,
+	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
+	ybar interval=0.7,
+  symbolic x coords={R, grep, wc, Icicle-372, wc -l, Icicle-1, disk, end}
+]
+\addplot coordinates {(R,6.6) (grep,74) (wc,251) (Icicle-372,543) (wc -l,775) (Icicle-1,1060) (disk,1800) (end,0) };
 
-\begin{code}
-        grep "^$"   wc    wc -l   Icicle
-MB/s      74        251   775       1,060
-\end{code}
+% \legend{MB/s}
+
+\end{axis}
+\end{tikzpicture}
+
+\caption{Benchmark results}
+\label{fig:bench:table}
+\end{figure}
+
+The table in figure~\ref{fig:bench:table} shows the throughput in megabytes per second.
+We compared the throughput of several programs over the same 360GB dataset:
+\begin{itemize}
+\item our original R implementation (R);
+\item Icicle with equivalent queries (Icicle-372);
+\item Icicle with a single query computing the mean (Icicle-1);
+\item finding empty lines with @grep "^$"@;
+\item counting characters, words and lines with @wc@;
+\item counting only lines with @wc -l@; and
+\item the sequential read speed of the disk (@disk@).
+\end{itemize}
+All of the unix utilities were run with the collation set to @LANG=C@ for maximum performance.
+
+It is heartening to see that both Icicle runs outperform @R@, @grep@ and @wc@, while the single query even outperforms @wc -l@.

### 2016-05-24 2016fhpc/section/02-Source.tex
typing rules

@@
 Value types are given in $T$ and can be numbers, booleans or maps.
-Modal types, $T_M$ can be a pure value type, or a modality associated with a value type.
-Function types, $T_\to$, can be either a non-function modal type, or a function from modal type to function type.
+Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
+Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
 As Icicle is a first-order language, function types are not value types.
@@
 The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
-The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{T_M}$ means that under context $\Gamma$, the expression has a given value type and modality.
-$\TypecheckP{\mi{Prim}}{T_\to}$ associates a primitive with its function type.
-$\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
-The next two judgments,
-$\TypecheckS{\Gamma}{\mi{Fun}}{\Gamma}$ and
-$\TypecheckS{\Gamma}{\mi{Exp}}{\Gamma}$
-both takes an input environment and function or query, and returns an environment containing the function or query name and its type.
+Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
+Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
+Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
+Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
 Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.
 
-The rule (TcVar) performs variable lookup in the context, while (TcNat) shows that number constants have type $\NN$.
-(TcBox) shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
+Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$.
+
+Rule TcBox shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
 The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.
 
-(TcPrimApp) and (TcFunApp) find the result type of primitive and function application.
-They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{T_\to}{\ov{T_M}}{T_M}$ to find the return type and modality of the application.
-The set-comprehension syntax $\{e_i\}_i$ means that this rule allows any number of arguments.
+Rules TcPrimApp and TcFunApp find the result type of primitive and function application.
+They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
+The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
 
-(TcLet) is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
+Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
 
-(TcFold) finds the type of a @fold@ expression.
+Rule TcFold finds the type of a @fold@ expression.
 The initial value of the fold must be pure.
@@
 
-(TcFilter) requires its predicate to be a stream of booleans ($@Element@~\BB$).
+Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
 The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
-(TcGroup) is similar to filter, except returning a @Map@.
+Rule TcGroup is similar to filter, except returning a @Map@.
 
-The rules (PrimArith), (PrimRel) and (PrimLookup) give function types for each primitive.
+The rules PrimArith, PrimRel and PrimLookup give function types for each primitive.
 
-The next two rules show how function application boxes and unboxes modalities.
+The next judgment shows how function application boxes and unboxes modalities.
+This judgment takes a function type and all its arguments, and returns the result type of the function.
 
-(AppArgs) applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
+Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
 
-(AppRebox) is used when a pure function is applied to a non-pure value.
+Rule AppRebox is used when a pure function is applied to a non-pure value.
 If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
-Note that (TcBox) rule can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
+Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
+These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.
 
-(CheckFun) inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
+Rule CheckFun inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
 
-(CheckQuery) checks a query's types, and makes sure it returns an @Aggregate@.
+Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.
 
-Finally, (CheckTop) checks a whole top-level program.
+Finally, rule CheckTop checks a whole top-level program.
 It first creates a new context with the table columns bound to @Element@s, and typechecks each function.

### 2016-05-24 2016fhpc/Main.tex
some example query plans, some preliminary benchmark info
Typechecking etc
benchmark plot

@@
 \usepackage{float}
+\usepackage{tikz}
+\usepackage{pgfplots}
 \usepackage{style/code}
@@
 \begin{abstract}
-When querying a large amount of data, simply iterating over the data may take hours.
+We present a pure query language, Icicle, which statically guarantees that multiple queries over the same input data set will be fused.
+We include a modal type system that guarantees these fused results will be evaluated with a single pass over the input data.
+Icicle queries are compiled via a fold-based intermediate language to efficient C code.
+We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.
+
+% When querying a large amount of data, simply iterating over the data may take hours.
 If multiple queries are to be performed, it is important that work is not duplicated.
@@
 \input{section/05-Conclusion.tex}
-\input{section/XX-Elimination.tex}
+% \input{section/XX-Elimination.tex}
 \input{section/XX-Related.tex}

### 2016-05-24 2016fhpc/section/01-Introduction.tex
Work on intro

--- 2016-05-21 2016fhpc/section/01-Introduction.tex
@@
 
-For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
+% \ben{the abstract already says this.}
+% For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
 When performing multiple queries, it is important to limit the amount of duplicate work and iterations.
 
-For example, suppose we have a table called @stocks@ with open and close prices for each company, for each day.
-We may wish to find the number of days where the open price exceeds the close price, and vice versa:
+% \ben{Start with the example.}
+Suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We wish to find the number of days where the open price exceeds the close price, and vice versa:
 \begin{code}
-SELECT COUNT(*)
-FROM   stocks
-WHERE  open > close;
+  SELECT COUNT(*) FROM stocks WHERE open > close;
+  SELECT COUNT(*) FROM stocks WHERE open < close;
+\end{code}
+As both queries are over the same input data set, we would like to evaluate them using a single pass over this data.
+As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:
 
-SELECT COUNT(*)
-FROM   stocks
-WHERE  open < close;
-\end{code}
+% In order to fuse these so that both are computed in a single query, some ingenuity is required.
+This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
 
-In order to fuse these so that both are computed in a single query, some ingenuity is required.
-This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
 \begin{code}
@@
 \end{code}
-
-If we also wish to compute the mean of open price on those days, we can do this like so:
+Suppose we also want to compute the mean of open price on those days:
 \begin{code}
@@
 \end{code}
-
-Now we can perform a similar trick to fuse all three together.
+Now we can perform a similar trick to fuse all three queries together.
+%
 \begin{code}
@@
 
-However, there are some problems with this approach.
-This fusion on SQL source queries is manual and error-prone.
-More importantly, we have no guarantees about which queries can be fused.
-We would like a guarantee that \emph{any} two queries can be fused this way.
+Fusing SQL queries by hand is tedious and error-prone.
+At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple separate queries over hundreds of gigabytes worth of data.
+For such applications we \emph{must} automatically fuse these queries and be sure that the result is correct in terms of both its value and performance charateristics.
 
-Another problem is that we have no guarantee that duplicate computations introduced by fusion will be eliminated.
-Above we are computing the count where open exceeds close twice: once in the count, and again for the mean.
+In the fused query above, note that we are counting the number of times where the open price exceeds the close price twice: once for the first source query and once for the third.
+In practice, the result of query fusion tends to have many common sub expressions, and we must also guarantee that the duplicates will be eliminated.
 
-The purpose of Icicle is two-fold: we wish to ensure that all queries over the same table can be fused together, and guarantee that such duplicate computations will be removed.
-This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
+% This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
+% \ben{Maybe leave mention of sorting until later sections}.
 
+\ben{Mention resumable queries in intro. The contributions mention them.}
+
+\ben{Rewrite the examples using this syntax?}
 In Icicle, the above queries are expressed as:
@@
 \item
-we show how Icicle guarantees any two queries on the same table can be fused together, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
 
 \item
-we show an intermediate fold-based language, which makes fusion a simple matter of appending two programs\sref{s:IcicleCore};
+we present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two programs\sref{s:IcicleCore};
 
 \item
-and we present benchmarks of our generated code, which compares favourably against standard unix utilities and outperforms our existing system by orders of magnitude\sref{s:Benchmarks}.
+and we present production benchmarks of our generated program, which compare favourably against standard unix utilities, and outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
 \end{itemize}

### 2016-05-24 2016fhpc/Main.tex
Work on intro
Reorg to move counter example to the intro

@@
 \authorinfo{
-  Amos Robinson$^\dagger$$^\ddagger$
-  \and Ben Lippmeier$^\dagger$
+  Amos Robinson$^{\alpha \beta}$
+  \and Ben Lippmeier$^{\beta \gamma}$
 }{
@@
   \shortstack{
-    $^\dagger$Computer Science and Engineering \\
-    University of New South Wales, Australia \\[2pt]
-    \textsf{amosr,benl@cse.unsw.edu.au}
+    $^\alpha$Ambiata      \\
+    Grow with data          \\[2pt]
+    \textsf{amos.robinson@ambiata.com}
   }
   \shortstack{
-    $^\ddagger$Ambiata      \\
-    Grow with data          \\[2pt]
-    \textsf{amos.robinson@ambiata.com}
+    $^\beta$UNSW Australia\\
+    \\[2pt]
+    ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
+  }
+  \shortstack{
+    $^\gamma$Vertigo Technology \\
+    \\[2pt]
+    ~~\textsf{benl@vergo.co}
   }
@@
 \begin{abstract}
-We present a pure query language, Icicle, which statically guarantees that multiple queries over the same input data set will be fused.
-We include a modal type system that guarantees these fused results will be evaluated with a single pass over the input data.
+We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input data stream will be fused.
 Icicle queries are compiled via a fold-based intermediate language to efficient C code.
@@
 \end{abstract}
-
-\category
-	{D.3.4}
-	{Programming Languages}
-	{Processors---Compilers; Optimization}
-
-\terms
-	Languages, Performance
-
-\keywords
-	Arrays; Fusion
 

### 2016-05-24 2016fhpc/section/01-Introduction.tex
Reorg to move counter example to the intro

@@
 
-% \ben{the abstract already says this.}
-% For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
-When performing multiple queries, it is important to limit the amount of duplicate work and iterations.
+At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input.
+We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch.
+Our feature generation process is performed in parallel on hundreds of nodes of a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.
 
-% \ben{Start with the example.}
-Suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We wish to find the number of days where the open price exceeds the close price, and vice versa:
-\begin{code}
-  SELECT COUNT(*) FROM stocks WHERE open > close;
-  SELECT COUNT(*) FROM stocks WHERE open < close;
-\end{code}
-As both queries are over the same input data set, we would like to evaluate them using a single pass over this data.
-As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:
-
-% In order to fuse these so that both are computed in a single query, some ingenuity is required.
-This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
+For example queries, suppose we have a table, @stocks@, containing daily open and close prices for a set of companies.
+We want to compute the number of days where the open price exceeded the close price, and vice versa.
+We also want the mean of the open price for days in which the open price exceeded the close price.
+In Icicle we write the three sub queries as follows:
 
 \begin{code}
-  SELECT COUNT(IF(open > close, 1, 0))
-       , COUNT(IF(open < close, 1, 0))
-  FROM stocks;
+table stocks { open : Int, close : Int }
+query 
+  more  = filter open > close in count;
+  less  = filter open < close in count;
+  mean  = filter open > close in sum open / count;
 \end{code}
-Suppose we also want to compute the mean of open price on those days:
-\begin{code}
-  SELECT SUM(open) / COUNT(*)
-  FROM   stocks
-  WHERE  open > close;
-\end{code}
-Now we can perform a similar trick to fuse all three queries together.
+In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered.
+We describe this construct more fully in \REF
+
+A joint query like this can be converted to a back-end language such a SQL, but doing it by hand it tedious and error prone.
+As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
+We instead need to lift each predicate to an expression-level conditional:
 %
@@
 
-Fusing SQL queries by hand is tedious and error-prone.
-At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple separate queries over hundreds of gigabytes worth of data.
-For such applications we \emph{must} automatically fuse these queries and be sure that the result is correct in terms of both its value and performance charateristics.
+As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.
 
-In the fused query above, note that we are counting the number of times where the open price exceeds the close price twice: once for the first source query and once for the third.
-In practice, the result of query fusion tends to have many common sub expressions, and we must also guarantee that the duplicates will be eliminated.
-
-% This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
-% \ben{Maybe leave mention of sorting until later sections}.
-
-\ben{Mention resumable queries in intro. The contributions mention them.}
-
-\ben{Rewrite the examples using this syntax?}
-In Icicle, the above queries are expressed as:
+Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion.
+As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table.
+We might try something like:
+%
 \begin{code}
-table stocks { open : Int; close : Int; }
-query more = filter open > close in count;
-query less = filter open < close in count;
-query mean = filter open > close in sum open / count;
+  table kvs { key : Int; value : Int }
+  query avg = let k = last key
+              in  filter (key == k) in mean value;
 \end{code}
 
-These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.
-
-The contributions of this paper are:
+Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} for doing so implied by the above query cannot be evaluated in a streaming fashion.
+When we are streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start.
+We need a better solution.
+With this in mind, the contributions of this paper are:
 \begin{itemize}
@@
 \end{itemize}
+
+% This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
+% \ben{Maybe leave mention of sorting until later sections}.
+
+% \ben{Rewrite the examples using this syntax?}
+% In Icicle, the above queries are expressed as:
+% \begin{code}
+% table stocks { open : Int; close : Int; }
+% query more = filter open > close in count;
+% query less = filter open < close in count;
+% query mean = filter open > close in sum open / count;
+% \end{code}
+
+% \ben{the abstract already says this.}
+% For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
+When performing multiple queries, it is important to limit the amount of duplicate work and iterations.
+
+% \ben{Start with the example.}
+% \begin{code}
+%   SELECT COUNT(*) FROM stocks WHERE open > close;
+%   SELECT COUNT(*) FROM stocks WHERE open < close;
+% \end{code}
+% As both queries are over the same input data set, we would like to evaluate them using a single pass over this data.
+As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:
+
+% In order to fuse these so that both are computed in a single query, some ingenuity is required.
+This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
+
+% \begin{code}
+%   SELECT COUNT(IF(open > close, 1, 0))
+%        , COUNT(IF(open < close, 1, 0))
+%   FROM stocks;
+% \end{code}
+% Suppose we also want to compute the mean of open price on those days:
+% \begin{code}
+%   SELECT SUM(open) / COUNT(*)
+%   FROM   stocks
+%   WHERE  open > close;
+% \end{code}
+
+% These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.

### 2016-05-24 2016fhpc/section/02-Source.tex
typing rules
Work on intro
Reorg to move counter example to the intro

@@
 %!TEX root = ../Main.tex
-\section{Icicle}
-\label{s:IcicleSource}
 
-\input{figures/Source-Grammar.tex}
-\input{figures/Source-Type.tex}
-\input{figures/Source-Eval.tex}
+\eject
+\section{Elements and Aggregates}
 
-The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
-Value types are given in $T$ and can be numbers, booleans or maps.
-Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
-Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
-As Icicle is a first-order language, function types are not value types.
+All Icicle queries must execute in a single pass over the data.
+Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives.
+However, not all queries can be evaluated in a single pass.
+For example, suppose we have a table with two fields key and value, and we wish to find the mean of values whose key matches the last one in the table.
+We might try something like:
 
-The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
-$\mi{Exp}$ defines the expressions used as query and function bodies.
-$\mi{Prim}$ defines the primitive types.
-Function definitions are given in $\mi{Fun}$.
-Query definitions are given in $\mi{Query}$.
-$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
+\begin{code}
+  table t { key : Int; value : Int }
+  query avg = let k    = last key
+              in  filter (key == k) in mean value;
+\end{code}
+The problem with the above query is that it requires two passes through the data. One to
 
-Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
-The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
-Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
-Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
+Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.
 
-Let-bindings allow a name to be used in place of the expression, in the rest of some query.
-\begin{code}
-let diff = open - close
-in  mean diff
-\end{code}
+This will not typecheck in Icicle, as @newest key@ has the modality @Aggregate@, while the filter condition @key == k@ attempts to check if an @Element@ is equal to an @Aggregate@.
 
-The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
-Count can be expressed by the following fold:
-\begin{code}
-fold count = 0 then count + 1
-\end{code}
-
-Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
-The following query will count the number of entries where the open price exceeds the close price.
-\begin{code}
-filter open > close in count
-\end{code}
-
-Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
-This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
-\begin{code}
-(filter open > close in count) / count
-\end{code}
-
-Group takes a key to group by, and a query to perform on the values of each grouping.
-This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
-\begin{code}
-group ((open - close) / 100) in count
-\end{code}
-
-$\mi{Fun}$ is used for function definitions.
-The @count@ function takes no arguments and returns a fold counting the number of elements.
-\begin{code}
-function count
- = fold c = 0 then c + 1;
-\end{code}
-
-Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
-It returns a fold that starts at zero and for every element, increases the old sum by @value@.
-\begin{code}
-function sum (value : Element Int)
- = fold s = 0 then s + value;
-\end{code}
-
-The definition of mean is just the sum divided by the count.
-\begin{code}
-function mean (value : Element Int)
- = sum value / count;
-\end{code}
-
-$\mi{Query}$ defines an aggregate query which must return a single value, and gives it a name:
-\begin{code}
-query count_all_records = count;
-\end{code}
-
-Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
-All queries here operate over the same input table, and will be fused into a single loop over the input data.
+This query could be translated to another synchronous streaming language such as {\sc Lustre}\cite{halbwachs1991synchronous}, and would typecheck.
+However, the semantics of the translated query will not be what was originally desired.
+Rather than checking each key against the most recent key as-of the end of the stream,
+it will check each key against the current most recent key, which is the key itself.
+This means the filter predicate is always be true, and thus returns the mean of the entire stream.
+These semantics may be surprising to the novice stream programmer.
 
 \subsection{Single-pass restriction}
-
-All Icicle queries must execute in a single pass over the data, as reading the data multiple times is expensive. 
-Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives.
 
@@
 Here is an example to highlight one benefit of tracking these modalities.
-Suppose we have a table with two fields: key and value.
-We wish to find the mean of values whose key matches the most recent key.
-Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.
-
-\begin{code}
-   let k    = newest key
-in let avg  = filter (key == k) in mean value
-in avg
-\end{code}
-
-This will not typecheck in Icicle, as @newest key@ has the modality @Aggregate@, while the filter condition @key == k@ attempts to check if an @Element@ is equal to an @Aggregate@.
-
-This query could be translated to another synchronous streaming language such as {\sc Lustre}\cite{halbwachs1991synchronous}, and would typecheck.
-However, the semantics of the translated query will not be what was originally desired.
-Rather than checking each key against the most recent key as-of the end of the stream,
-it will check each key against the current most recent key, which is the key itself.
-This means the filter predicate is always be true, and thus returns the mean of the entire stream.
-These semantics may be surprising to the novice stream programmer.
 
@@
 Our system requires no clock analysis, but is less expressive as streams are not `first class'.
+
+\section{Source Language}
+\label{s:IcicleSource}
+
+\input{figures/Source-Grammar.tex}
+\input{figures/Source-Type.tex}
+\input{figures/Source-Eval.tex}
+
+The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
+Value types are given in $T$ and can be numbers, booleans or maps.
+Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
+Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
+As Icicle is a first-order language, function types are not value types.
+
+The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
+$\mi{Exp}$ defines the expressions used as query and function bodies.
+$\mi{Prim}$ defines the primitive types.
+Function and query definitions are given in $\mi{Def}$.
+$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
+
+Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
+The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
+Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
+Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
+
+Let-bindings allow a name to be used in place of the expression, in the rest of some query.
+\begin{code}
+let diff = open - close
+in  mean diff
+\end{code}
+
+The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
+Count can be expressed by the following fold:
+\begin{code}
+fold count = 0 then count + 1
+\end{code}
+
+Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
+The following query will count the number of entries where the open price exceeds the close price.
+\begin{code}
+filter open > close in count
+\end{code}
+
+Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
+This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
+\begin{code}
+(filter open > close in count) / count
+\end{code}
+
+Group takes a key to group by, and a query to perform on the values of each grouping.
+This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
+\begin{code}
+group ((open - close) / 100) in count
+\end{code}
+
+$\mi{Def}$ is used for function and query definitions.
+The @count@ function takes no arguments and returns a fold counting the number of elements.
+\begin{code}
+function count
+ = fold c = 0 then c + 1;
+\end{code}
+
+Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
+It returns a fold that starts at zero and for every element, increases the old sum by @value@.
+\begin{code}
+function sum (value : Element Int)
+ = fold s = 0 then s + value;
+\end{code}
+
+The definition of mean is just the sum divided by the count.
+\begin{code}
+function mean (value : Element Int)
+ = sum value / count;
+\end{code}
+
+Queries must return a single value, and are given it a name:
+\begin{code}
+query count_all_records = count;
+\end{code}
+
+Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
+All queries here operate over the same input table, and will be fused into a single loop over the input data.
 

### 2016-05-25 2016fhpc/section/01-Introduction.tex
Wibbles

@@
 We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch.
-Our feature generation process is performed in parallel on hundreds of nodes of a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.
+Our feature generation process is executed in parallel on hundreds of nodes on a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.
 
@@
 We instead need to lift each predicate to an expression-level conditional:
-%
 \begin{code}
@@
 \end{code}
-
 As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.
@@
 We might try something like:
-%
 \begin{code}
@@
 \end{code}
-
-Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} for doing so implied by the above query cannot be evaluated in a streaming fashion.
-When we are streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start.
+Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in a streaming fashion.
+While streaming through the table we always have access to the last key in the stream, but to perform a filter to select rows that match this key would require that we stream through the table again from the start.
 We need a better solution.
@@
 \end{itemize}
+
+\TODO{Touch on related work}
 

### 2016-05-25 2016fhpc/Main.tex
Mention modal type system in abstract
Wibbles to intro

@@
 \begin{abstract}
-We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input data stream will be fused.
-Icicle queries are compiled via a fold-based intermediate language to efficient C code.
+We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input stream will be fused.
+We use a modal type system to ensure that fused queries can be computed in an incremental fashion, and a fold-based intermediate language to compile down to efficient C code.
 We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.

### 2016-05-25 2016fhpc/section/01-Introduction.tex
Wibbles to intro

@@
 
-A joint query like this can be converted to a back-end language such a SQL, but doing it by hand it tedious and error prone.
+A joint query like this can be converted to a back-end language such as SQL, but doing it by hand it tedious and error prone.
 As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
@@
 Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in a streaming fashion.
-While streaming through the table we always have access to the last key in the stream, but to perform a filter to select rows that match this key would require that we stream through the table again from the start.
+While streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start.
 We need a better solution.
@@
 \item
-we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+We present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
 
 \item
-we present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two programs\sref{s:IcicleCore};
+We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two programs\sref{s:IcicleCore};
 
 \item
-and we present production benchmarks of our generated program, which compare favourably against standard unix utilities, and outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
+We present production benchmarks of our generated program, which compare favourably against standard unix utilities, and outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
 \end{itemize}

### 2016-05-25 2016fhpc/section/01-Introduction.tex
Work on S2

@@
 query 
-  more  = filter open > close in count;
-  less  = filter open < close in count;
-  mean  = filter open > close in sum open / count;
+  more  = filter open > close of count;
+  less  = filter open < close of count;
+  mean  = filter open > close of sum open / count;
 \end{code}
@@
 \begin{code}
-  table kvs { key : Int; value : Int }
+  table kvs { key : Text; value : Real }
   query avg = let k = last key
-              in  filter (key == k) in mean value;
+              in  filter (key == k) of mean value;
 \end{code}

### 2016-05-25 2016fhpc/section/02-Source.tex
Work on S2

@@
 \eject
+% ---------------------------------------------------------
 \section{Elements and Aggregates}
 
-All Icicle queries must execute in a single pass over the data.
-Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives.
-However, not all queries can be evaluated in a single pass.
-For example, suppose we have a table with two fields key and value, and we wish to find the mean of values whose key matches the last one in the table.
-We might try something like:
+To allow incremental computation, all Icicle queries must execute in a single pass over the input stream.
+Of course, not all queries can be executed in a single pass: the key examples are anything that require random access indexing, or otherwise need to access data in an order that is different to what the stream will provide.
+However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
+Here is the unstreamable example from the introduction:
+\begin{code}
+  table kvs { key : Text; value : Real }
+  query avg = let k = last key
+              in  filter (key == k) of mean value;
+\end{code}
 
+The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs this value to process the very first element in the same stream.
+We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate} as to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation that only needs access to the current element read from the stream.
+
+The trick to compute our average in an incremental fashion is to recognise the @filter@ operation is serving to select a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information.
+Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
 \begin{code}
-  table t { key : Int; value : Int }
-  query avg = let k    = last key
-              in  filter (key == k) in mean value;
+  table kvs { key : Text; value : Real } 
+  query avg = let k    = last  key in
+              let avgs = group key of mean value
+              in  lookup k avgs
 \end{code}
-The problem with the above query is that it requires two passes through the data. One to
 
-Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.
+Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and to compute running mean of the values of each group.
+The @avgs@ value becomes a map of group keys to their running means.
+Once we have reached the end of the stream we then have access to the last key and can lookup the final result.
 
-This will not typecheck in Icicle, as @newest key@ has the modality @Aggregate@, while the filter condition @key == k@ attempts to check if an @Element@ is equal to an @Aggregate@.
+% ---------------------------------------------------------
+\subsection{The Stage Restriction}
+To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}.
+We use two modalities, @Element@ and @Aggregate@.
+Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed.
+In the expression @filter (key == k) of mean value@, the variable @key@ has type @Element Text@ while the variable @k@ has type @Aggregate Text@.
+Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.
 
-This query could be translated to another synchronous streaming language such as {\sc Lustre}\cite{halbwachs1991synchronous}, and would typecheck.
-However, the semantics of the translated query will not be what was originally desired.
-Rather than checking each key against the most recent key as-of the end of the stream,
-it will check each key against the current most recent key, which is the key itself.
-This means the filter predicate is always be true, and thus returns the mean of the entire stream.
-These semantics may be surprising to the novice stream programmer.
+% ---------------------------------------------------------
+\subsection{Finite Streams and Synchronous Data Flow}
+As opposed to synchronous data flow languages such as {\sc Lustre}\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
+Icicle is fundamentally a query language, which queries finite streams of data held in a non-volatile store.
+In contrast Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like in airplanes).
+In Icicle, the ``last'' element in a stream is the last one that was in the file read from disk.
+In Lustre, the ``last'' element in a stream is the one that was most recently received.
+If the ``unstreamable'' version of the query from \S2 was converted into Lustre then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself.
+This means the filter predicate would always be true, and the query would return the mean of the entire stream.
+Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.
 
-\subsection{Single-pass restriction}
+% ---------------------------------------------------------
+\subsection{Incremental Update}
 
-In order to ensure that queries can be executed as a single pass, we use a modal type system inspired by staged computation\cite{davies2001modal}.
-We introduce two modalities, called @Element@ and @Aggregate@, which denote when computations must occur.
-
-The @Element@ modality means that a computation is defined for each record in the table, or element in the stream.
-Each column in the table is @Element@; for example the @open@ column has type @Element Int@, meaning that for each record in the column there is an @Int@.
-This can be thought of as being represented by a stream of values.
-
-The @Aggregate@ modality means that a computation is available only after all records in the table have been seen, or after the end of the stream.
-These are used for the results of folds.
-When computing count, the final value is not known until all the records have been seen, so @count@ has type @Aggregate Int@.
+\TODO{Point out that query from S2 can be evaluated incrementally}
 
@@
 
-Here is an example to highlight one benefit of tracking these modalities.
+% The @Element@ modality means that a computation is defined for each record in the table, or element in the stream.
+Each column in the table is @Element@; for example the @open@ column has type @Element Int@, meaning that for each record in the column there is an @Int@.
+This can be thought of as being represented by a stream of values.
 
-Hopefully our programmer is not too discouraged by the @Element /= Aggregate@ type error, and pushes on.
+% The @Aggregate@ modality means that a computation is available only after all records in the table have been seen, or after the end of the stream.
+These are used for the results of folds.
+When computing count, the final value is not known until all the records have been seen, so @count@ has type @Aggregate Int@.
+
+% Hopefully our programmer is not too discouraged by the @Element /= Aggregate@ type error, and pushes on.
 They now know that their query, as formulated, requires multiple passes over the data: the problem now is to reformulate it as a single pass.
@@
 
-\begin{code}
-   let k    = newest key
-in let avgs = group  key in mean value
-in mapLookup avgs k
-\end{code}
-
+% ---------------------------------------------------------
 \subsection{Bounded buffer restriction}
 \label{s:IcicleSource:bounded}
+\TODO{Point out that the group uses space proportional to the number of keys in the input, which may be as large as the number of elements in the stream}
+
 As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
@@
 
+\eject
+% ---------------------------------------------------------
 \section{Source Language}

### 2016-05-25 2016fhpc/section/02-Source.tex
Hide syntax examples to save space

@@
 
-Let-bindings allow a name to be used in place of the expression, in the rest of some query.
-\begin{code}
-let diff = open - close
-in  mean diff
-\end{code}
+% Let-bindings allow a name to be used in place of the expression, in the rest of some query.
+% \begin{code}
+% let diff = open - close
+% in  mean diff
+% \end{code}
 
 The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
-Count can be expressed by the following fold:
-\begin{code}
-fold count = 0 then count + 1
-\end{code}
+
+% Count can be expressed by the following fold:
+% \begin{code}
+% fold count = 0 then count + 1
+% \end{code}
 
 Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
-The following query will count the number of entries where the open price exceeds the close price.
-\begin{code}
-filter open > close in count
-\end{code}
 
-Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
-This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
-\begin{code}
-(filter open > close in count) / count
-\end{code}
+% The following query will count the number of entries where the open price exceeds the close price.
+% \begin{code}
+% filter open > close in count
+% \end{code}
+
+% Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
+% This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
+% \begin{code}
+% (filter open > close in count) / count
+% \end{code}
 
 Group takes a key to group by, and a query to perform on the values of each grouping.
-This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
-\begin{code}
-group ((open - close) / 100) in count
-\end{code}
+
+% This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
+% \begin{code}
+% group ((open - close) / 100) in count
+% \end{code}
 
 $\mi{Def}$ is used for function and query definitions.
-The @count@ function takes no arguments and returns a fold counting the number of elements.
-\begin{code}
-function count
- = fold c = 0 then c + 1;
-\end{code}
+% The @count@ function takes no arguments and returns a fold counting the number of elements.
+% \begin{code}
+% function count
+%  = fold c = 0 then c + 1;
+% \end{code}
 
-Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
-It returns a fold that starts at zero and for every element, increases the old sum by @value@.
-\begin{code}
-function sum (value : Element Int)
- = fold s = 0 then s + value;
-\end{code}
+% Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
+% It returns a fold that starts at zero and for every element, increases the old sum by @value@.
+% \begin{code}
+% function sum (value : Element Int)
+%  = fold s = 0 then s + value;
+% \end{code}
 
-The definition of mean is just the sum divided by the count.
-\begin{code}
-function mean (value : Element Int)
- = sum value / count;
-\end{code}
+% The definition of mean is just the sum divided by the count.
+% \begin{code}
+% function mean (value : Element Int)
+%  = sum value / count;
+% \end{code}
 
-Queries must return a single value, and are given it a name:
-\begin{code}
-query count_all_records = count;
-\end{code}
+% Queries must return a single value, and are given it a name:
+% \begin{code}
+% query count_all_records = count;
+% \end{code}
 
@@
 
+% ---------------------------------------------------------
 \subsection{Type system}
-
 The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.

### 2016-05-26 2016fhpc/section/04-Benchmarks.tex
update benchmarks

@@
 	ybar interval=0.7,
-  symbolic x coords={R, grep, wc, Icicle-372, wc -l, Icicle-1, disk, end}
+  symbolic x coords={R, grep, wc, Icicle-372, wc -l, cat, Icicle-1, end}
 ]
-\addplot coordinates {(R,6.6) (grep,74) (wc,251) (Icicle-372,543) (wc -l,775) (Icicle-1,1060) (disk,1800) (end,0) };
+\addplot coordinates {(R,6.0) (grep,74) (wc,182.6) (Icicle-372,846) (wc -l,852) (cat,878) (Icicle-1,1029) (end,0) };
 
@@
 
-\caption{Benchmark results}
-\label{fig:bench:table}
+\caption{Throughput comparisons of Icicle against existing R code and standard unix utilities}
+\label{fig:bench:other}
 \end{figure}
 
-The table in figure~\ref{fig:bench:table} shows the throughput in megabytes per second.
+\begin{figure}
+
+\begin{tikzpicture}
+\begin{axis}[
+	x tick label style={/pgf/number format/1000 sep=},
+	ylabel=MB/s,
+  ymin=0, ymax=1100,
+%  xmax=372,
+  xlabel=Number of queries,
+	enlargelimits=0.01,
+	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
+%	ybar interval=0.7,
+]
+\addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
+
+% \legend{MB/s}
+
+\end{axis}
+\end{tikzpicture}
+
+\caption{Decrease in throughput as queries are added}
+\label{fig:bench:queries}
+\end{figure}
+
+The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
 We compared the throughput of several programs over the same 360GB dataset:

### 2016-05-26 2016fhpc/section/01-Introduction.tex
Work on bounded buffers

@@
 
-A joint query like this can be converted to a back-end language such as SQL, but doing it by hand it tedious and error prone.
+A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone.
 As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
@@
 
-Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion.
+Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion, which allows us to update the result as we receive new data.
 As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table.
@@
 \begin{code}
-  table kvs { key : Text; value : Real }
+  table kvs { key : Date; value : Real }
   query avg = let k = last key
@@
 \end{code}
-Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in a streaming fashion.
+Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an inceremental fashion.
 While streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start.
@@
 \item
-We present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};
+We present a domain specific language and type system that guarantees any two queries on the same input table can be fused, and allows queries to be resumed as new data is received\sref{s:IcicleSource};
 
 \item
-We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two programs\sref{s:IcicleCore};
+We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};
 
 \item
-We present production benchmarks of our generated program, which compare favourably against standard unix utilities, and outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
+We present production benchmarks of Icicle compiled code, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
 \end{itemize}

### 2016-05-26 2016fhpc/section/02-Source.tex
Work on bounded buffers

@@
 To allow incremental computation, all Icicle queries must execute in a single pass over the input stream.
-Of course, not all queries can be executed in a single pass: the key examples are anything that require random access indexing, or otherwise need to access data in an order that is different to what the stream will provide.
+Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order that is different to what a stream can provide.
 However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
-Here is the unstreamable example from the introduction:
+Here is the unstreamable example again from the introduction:
 \begin{code}
-  table kvs { key : Text; value : Real }
+  table kvs { key : Date; value : Real }
   query avg = let k = last key
@@
 
-The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs this value to process the very first element in the same stream.
-We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate} as to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation that only needs access to the current element read from the stream.
+The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs it to process the very first element in the same stream.
+We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate}, because to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation because it only needs access to the current element in the stream.
 
-The trick to compute our average in an incremental fashion is to recognise the @filter@ operation is serving to select a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information.
+The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information.
 Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
 \begin{code}
-  table kvs { key : Text; value : Real } 
+  table kvs { key : Date; value : Real } 
   query avg = let k    = last  key in
@@
 Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed.
-In the expression @filter (key == k) of mean value@, the variable @key@ has type @Element Text@ while the variable @k@ has type @Aggregate Text@.
+In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Text@ while @k@ has type @Aggregate Text@.
 Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.
+
+Note that the types of pure values such as constants are automatically promoted to have the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.
 
@@
 \subsection{Finite Streams and Synchronous Data Flow}
-As opposed to synchronous data flow languages such as {\sc Lustre}\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
-Icicle is fundamentally a query language, which queries finite streams of data held in a non-volatile store.
-In contrast Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like in airplanes).
-In Icicle, the ``last'' element in a stream is the last one that was in the file read from disk.
+As opposed to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
+Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner.
+In contrast, Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes).
+In Icicle, the ``last'' element in a stream is the last one that appears in the file on disk.
 In Lustre, the ``last'' element in a stream is the one that was most recently received.
-If the ``unstreamable'' version of the query from \S2 was converted into Lustre then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself.
+If the ``unstreamable'' version of the query from \S2 was converted into Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself.
 This means the filter predicate would always be true, and the query would return the mean of the entire stream.
@@
 
-\TODO{Point out that query from S2 can be evaluated incrementally}
-
-The modalities are automatically boxed and unboxed, so that if a function expecting a pure value is passed an @Element@ computation, the result of the function becomes an @Element@ computation too.
-For example, if @open : Element Int@, then @open == 1@ has type @Element Bool@.
-@Element@ and @Aggregate@ are mutually exclusive: a computation cannot be both an element and an aggregate.
+Suppose we apply a query large table and record the result.
+Tomorrow morning we receive more data and add it to the table.
+We would like to recompute the result of this query without needing to process all data from the start of the table.
+We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives.
+In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.
 
@@
 \label{s:IcicleSource:bounded}
-\TODO{Point out that the group uses space proportional to the number of keys in the input, which may be as large as the number of elements in the stream}
-
-As Icicle is designed to be a streaming language, the amount of data to be streamed may not necessarily fit in memory.
-Any operation which requires a buffer must be bounded in size, as an unbounded buffer could potentially grow too large to fit in memory.
-
-Here is an example stream program that requires unbounded buffering:
-it takes an input stream @xs@, and filters it into those above zero, and those below zero.
-These two filtered streams are then joined together pairwise, so the first positive element is paired with the first negative element, and so on.
-The program, written in Haskell:
-
+Icicle queries are intended to process tables of arbitrary size that may not fit in memory.
+Due to this, each query must execute without requiring buffer space proportional to the size of the input.
+For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
 \begin{code}
-zip (filter (>0) xs) (filter (<0) xs)
+      zip (filter (> 0) xs) (filter (< 0) xs)
 \end{code}
 
-This program requires an unbounded buffer: if the input stream contains ten positive values followed by one negative value, all ten of the positive values must be held onto until the negative value is seen.
-Similarly, if the entire stream is positive, all of the elements must be retained until the very end, just in case a negative value shows up.
+Suppose @(xs :: Stream Int)@.
+The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero.
+This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.
 
-In Icicle, we outlaw this kind of program by implicitly threading the input stream through operations.
-The streams themselves are not materialised in the program: stream operations like @fold@ and @filter@ do not take parameters of the streams, but instead operate on the context stream.
-By removing the explicit stream parameter, stream elements can only be joined from the same context, when both elements are available.
+The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with a one major cavet that we will discuss in a moment.
+In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly.
+Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context.
+In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only.
+Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query.
+In the expression @filter p in mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.
 
-For example, in the query @filter p in mean value@, the @mean value@ is only applied to stream values which satisfy the predicate @p@.
+Finally, note that our major caveat is that @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
+For our applications the keys are commonly company names, customer names, and days of the year.
+Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory.
+Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.
 
-This is a different approach than existing synchronous streaming languages\cite{mandel2010lucy} and flow fusion\cite{lippmeier2013data}, which perform clock analysis or `rate inference'.
-Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together.
+% Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together.
 Filters produce a stream with a different output clock to its input, so different filters cannot be zipped together.

### 2016-05-26 2016fhpc/section/XX-Related.tex
ok
Evaluation rules
Work on bounded buffers

@@
 \label{s:Related}
+
+\TODO{fold this into related}
+The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
+These systems perform a clock, or \emph{rate} analysis of the streams being processed.
+Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
 
@@
 Global value numbering for loops can perform all optimisations here.
+
+Even better is Nie and Cheng's approach\cite{nie2007efficient}, which does not seem to be more complete or lower complexity than Gulwani and Necula, just lower constant factors.
 

### 2016-05-26 2016fhpc/section/02-Source.tex
some minor changes and remove GVN stuff from related

@@
 Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed.
-In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Text@ while @k@ has type @Aggregate Text@.
+In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Date@ while @k@ has type @Aggregate Date@.
 Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.
@@
 
-Suppose we apply a query large table and record the result.
+Suppose we apply a query to a large table and record the result.
 Tomorrow morning we receive more data and add it to the table.
@@
 \begin{code}
-      zip (filter (> 0) xs) (filter (< 0) xs)
+    unbounded (xs :: Stream Int)
+     = zip (filter (> 0) xs) (filter (< 0) xs)
 \end{code}
 
-Suppose @(xs :: Stream Int)@.
 The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero.
@@
 
-The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with a one major cavet that we will discuss in a moment.
+The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
 In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly.
@@
 Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query.
-In the expression @filter p in mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.
+In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.
 

### 2016-05-26 2016fhpc/section/XX-Related.tex
some minor changes and remove GVN stuff from related

--- 2016-04-11 2016fhpc/section/XX-Related.tex
@@
 
-``most compilers restrict themselves to common subexpression elimination within basic blocks or extended basic blocks''
-\cite{debray1992compiler}
+\TODO{fold this into related}
+The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
+These systems perform a clock, or \emph{rate} analysis of the streams being processed.
+Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
 
-Temporal common subexpression elimination in Single Assignment C
-allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
-This is a more general case of the elimination opportunities that arise from loop unrolling.
+% The most closely related transforms are induction variable elimination\cite{shivers1988control} and global value numbering\cite{rosen1988global}.
+% Induction variable elimination finds and removes accumulators that are linear functions of the iteration number.
+% This does not deal with non-linear functions such as sum, or mutually recursive accumulators.
 
-For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
-On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
-\begin{code}
-int[] A;
-int a = 1;
-int b = 0;
-while(int i = 0; i != size - 1; i++)  {
-  int ax = f(A[i]);
-  int bx = f(A[i+1]);
-  a += ax;
-  b *= bx;
-}
-\end{code}
+% Global value numbering works on an SSA form and is much more general than induction variable elimination.
+% Early versions such as Rosen et al\cite{rosen1988global} only worked on extended basic blocks with backedges removed.
+% Alpern et al\cite{alpern1988detecting} introduced cases for specific loop backedges, where a fixpoint is performed to find the congruence sets of loops.
+% More recently, Gulwani and Necula\cite{gulwani2004polynomial} improved upon this, supporting removal of more expressions, and imposing a polynomial time bound.
+% Global value numbering for loops can perform all optimisations here.
 
-I cannot think of a better example than this right now.
+% Even better is Nie and Cheng's approach\cite{nie2007efficient}, which does not seem to be more complete or lower complexity than Gulwani and Necula, just lower constant factors.
+
+% Common subexpression elimination is very closely related, but most imperative compilers only perform common subexpression elimination on straight-line computations with no control flow\cite{debray1992compiler}, or on loop invariant expressions\cite{bodik1998complete}.
+% Common subexpression elimination for functional languages would be able remove lone expressions, but does not deal with mutually recursive folds.
+
+% Temporal common subexpression elimination in Single Assignment C allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
+% This is a more general case of the elimination opportunities that arise from loop unrolling.
+
+% For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
+% On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
+% \begin{code}
+% int[] A;
+% int a = 1;
+% int b = 0;
+% while(int i = 0; i != size - 1; i++)  {
+%   int ax = f(A[i]);
+%   int bx = f(A[i+1]);
+%   a += ax;
+%   b *= bx;
+% }
+% \end{code}
+% 
+% I cannot think of a better example than this right now.
 
@@
 This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
-However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive
+However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.
+
+% The polyhedral model\cite{benabderrahmane2010polyhedral} analyses data dependencies in loops.
+% For a given iteration, the polyhedral model computes the iterations which must be executed before the given iteration.
+% By knowing the dependent iterations, one can restructure the loop so that the iterations are executed in a different order, while still performing any dependencies in order.
+% For example, a loop iterating over @i@, with loop body @A[i] = A[i-10]@ depends on the tenth previous iteration.
+% This loop could be executed in an order like @0, 1, 2,...@, but it could also stride by ten: @0, 10, 20..., 1, 11, 21,...@.
+% This allows fusion of any two loops if the intersection of the two loops' iteration spaces is not empty.
+% In our case the iteration order is fixed, as the program cannot control the order in which stream elements are received.
+% The polyhedral model may still be applicable to streaming computations where one could introduce certain-sized buffers, then process the buffer in parallel.
+% In this case the polyhedral model may expose more opportunities for duplicate elimination, but does not itself remove duplicates.
+%

### 2016-05-27 2016fhpc/section/02-Source.tex
Change evaluator to take modality; update some benchmarks; hide conclusion

@@
 
-We define simple values $V$ as numbers, booleans, maps of values, or tuples of values.
-Next, stream values $V'$ correspond to modalities: pure computations are represented as a single value;
+Definition $V$ are simple values of numbers, booleans, maps and tuples.
+Definition $V'$ are stream values that correspond to modalities: pure computations are represented as a single value;
 @Element@s as a stream transformer from a heap to a value;
 and @Aggregate@s as folds defining an initial state and update and extract functions.
+Definition $M'$ are the modalities themselves without value types.
 
@@
 
-The judgment $\SourceStepX{E'}{V'}$ defines a big-step evaluation relation.
+The judgment $\SourceStepX{M'}{E'}{V'}$ defines a big-step evaluation relation.
+The rules below use the modality in deciding whether pure values should be converted to streams or folds.
 
-(EVal) applies when the expression is alread a stream value.
-This corresponds, roughly, to the (TcNat) typechecking rule.
+Rule EVal applies when the expression is alread a stream value.
+This corresponds roughly to the TcNat typechecking rule.
 
-(EBoxStream) and (EBoxFold) convert a pure value to a stream and fold, respectively.
+Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the environment modality is not pure.
 The stream transformer ignores the input stream and returns the pure value.
@@
 
-Next (EPrimValue), (EPrimStream) and (EPrimFold) evaluate primitive applications.
+Rules EPrimValue, EPrimStream and EPrimFold evaluate primitive applications.
 For pure values, the values are unboxed and the primitive is applied.
@@
 
-The (ELet) rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
+Rule ELet rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
+The only difference is we inspect the type of the bound expression to find the modality, then evaluate the bound under that modality.
+The rest of the expression is evaluated under the original modality.
 
-(EFold) is one of the most complex rules, as it introduces recursion to the streams.
+Rule EFold is one of the most complex, as it introduces recursion to the streams.
 It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
 In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
-This action-at-a-distance is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
+This sleight of hand is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
 Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
@@
 
-(EFilter) evaluates its predicate to a stream of booleans, and the rest to a fold.
+Rule EFilter evaluates its predicate to a stream of booleans, and the rest to a fold.
 The result is a fold which only applies the update function when the predicate stream is true.
 
-(EGroup) is rather similar to filter, except that it returns a map.
+Rule EGroup is rather similar to filter, except that it returns a map.
 Here, we start with an empty map of bottoms.
@@
 First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
-For example, $e[@open@~:=~@Stream@~(\lam{\sigma} \sigma~@open@)]$.
+For example, $e[@open@~:=~@Stream@~(\lam{\store} \store~@open@)]$.
 Next, the query must be converted to a @Fold@.

### 2016-05-27 2016fhpc/section/04-Benchmarks.tex
Change evaluator to take modality; update some benchmarks; hide conclusion

--- 2016-05-15 2016fhpc/section/04-Benchmarks.tex
@@
 
-We are currently using Icicle in production, over medium-sized datasets that fit on a single disk.
-Initial results have been very promising, and we are currently working towards a distributed implementation for larger datasets.
+We are currently using Icicle in production, over medium-sized datasets (around three hundred gigabytes) that fit on a single disk.
+While future versions will require incremental computation, the runtime processing medium-sized datasets has been small enough to run over the entire dataset each time.
+These initial results have been very promising, and we are currently implementing distribution across multiple machines for datasets that do not fit on a single disk.
+Incremental computation becomes more important for this distributed case, as it can potentially avoid copying terabytes of data over the network.
 
-One evaluation we performed was replacing a hand-written R script which had been running in production.
+We evaluated Icicle by replacing an R script which ran weekly in production.
 This R script works over around three hundred gigabytes of PSV data.
@@
 
-The R script for this takes around fifteen hours to run and is 3,566 lines, 2,311 of which are code.
-In contrast, our Icicle queries take eleven minutes to run, and the dictionary describing the queries is 191 lines of code.
+The R script for this takes fifteen hours to run and is 3,566 lines, 2,311 of which are code.
+In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
 
@@
 
-\begin{code}
-          R       Icicle
-Minutes   900         11
-LOC       2,311      191
-MB/s      6.6        543
-\end{code}
+\begin{figure}
 
-Run some benchmarks.
+\begin{tikzpicture}
+\begin{axis}[
+	x tick label style={/pgf/number format/1000 sep=},
+	ylabel=MB/s,
+	enlargelimits=0.01,
+	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
+	ybar interval=0.7,
+  symbolic x coords={R, wc, grep, wc -l, cat, Run 372, Run 1, end}
+]
+\addplot coordinates {(R,6.0) (wc,182.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
 
-\begin{code}
-        grep "^$"   wc    wc -l   Icicle
-MB/s      74        251   775       1,060
-\end{code}
+% \legend{MB/s}
+
+\end{axis}
+\end{tikzpicture}
+
+\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities}
+\label{fig:bench:other}
+\end{figure}
+
+\begin{figure}
+
+\begin{tikzpicture}
+\begin{axis}[
+	x tick label style={/pgf/number format/1000 sep=},
+	ylabel=MB/s,
+  ymin=800, ymax=1100,
+%  xmax=372,
+  xlabel=Number of queries,
+	enlargelimits=0.01,
+	legend style={legend columns=-1},
+%	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
+%	ybar interval=0.7,
+]
+
+% bench/raw.txt
+% Running with fewer queries
+ \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
+
+% bench/raw.txt
+% Running with fewer queries, better distribution
+\addplot coordinates { (0,1029) (31,1013) (62,1006) (93,998) (124,991) (155,941) (186,956) (217,933) (248,931) (279,915) (279,915) (310,919) (341,916) (372,908) };
+
+% Fusion, no output
+\addplot coordinates { (0,1001) (31,1038) (62,1025) (93,1031) (124,1006) (155,1015) (186,1020) (217,997) (248,1026) (279,1024) (310,1009) (341,1038) (372,1025) };
+
+\legend{CSE, No CSE, dev/null};
+
+\end{axis}
+\end{tikzpicture}
+
+\caption{Decrease in throughput as queries are added, with and without common subexpression elimination (CSE) over queries}
+\label{fig:bench:queries}
+\end{figure}
+
+The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
+We compared the throughput of several programs over the same 317GB dataset:
+\begin{itemize}
+\item our original R implementation (R);
+\item Icicle with equivalent queries (Run 372);
+\item Icicle with a single query computing the mean (Run 1);
+\item finding empty lines with @grep "^$"@;
+\item counting characters, words and lines with @wc@;
+\item counting only lines with @wc -l@; and
+\item reading and throwing away the results with @cat > /dev/null@.
+\end{itemize}
+
+We ran all of the unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
+The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
+The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
+Icicle significantly outperformed R, @grep@ and @wc@, and was roughly on par with @wc -l@ and @cat@.
+Although the computer had 32 CPUs, Icicle was the only benchmark we ran that used multiple CPUs: the other benchmarks were single threaded.
+
+Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.

### 2016-05-28 2016fhpc/section/04-Benchmarks.tex
benchmarks

@@
 
-\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities}
+\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities; higher is faster.}
 \label{fig:bench:other}
@@
 	ylabel=MB/s,
-  ymin=800, ymax=1100,
+  ymin=0, ymax=1200,
 %  xmax=372,
@@
 % Running with fewer queries
- \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
+% \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
 
@@
 
-\legend{CSE, No CSE, dev/null};
+% No fusion, no output
+% \addplot coordinates { (0,987) (31,1037) (62,1024) (93,1002) (124,1037) (155,1024) (186,1004) (217,1040) (248,1023) (279,998) (310,1018) (341,1013) (372,1003) };
+
+\legend{Output to disk, Output to /dev/null};
 
@@
 
-\caption{Decrease in throughput as queries are added, with and without common subexpression elimination (CSE) over queries}
+\caption{Decrease in throughput as queries are added. By comparing writing the output to disk and writing to /dev/null, we see that actual computation time remains constant.}
 \label{fig:bench:queries}
@@
 Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
+We ran with a number of queries twice, with one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
+The graph shows the disk version decreasing roughly linearly in the number of queries, while ignoring the output stays constant.
+This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
+The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled.

### 2016-05-28 2016fhpc/section/04-Benchmarks.tex
just mention size of larger datasets in benchs

@@
 
-We are currently using Icicle in production, over medium-sized datasets (around three hundred gigabytes) that fit on a single disk.
+We are currently using Icicle in production, over medium-sized datasets (around three hundred gigabytes uncompressed) that fit on a single disk.
 While future versions will require incremental computation, the runtime processing medium-sized datasets has been small enough to run over the entire dataset each time.
-These initial results have been very promising, and we are currently implementing distribution across multiple machines for datasets that do not fit on a single disk.
+These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
 Incremental computation becomes more important for this distributed case, as it can potentially avoid copying terabytes of data over the network.

### 2016-05-28 2016fhpc/section/01-Introduction.tex
some changes to introduction, s2, s3

@@
 
-At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input.
-We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch.
+At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over terabytes worth of compressed data.\footnote{In 2016 this was a lot of data.}
+For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input.
+We also ingest tens of gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch.
 Our feature generation process is executed in parallel on hundreds of nodes on a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.
@@
 \item
-We present an fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};
+We present a fold-based intermediate language, which allows the query fusion transformation to be a simple matter of appending two intermediate programs\sref{s:IcicleCore};
 
 \item
-We present production benchmarks of Icicle compiled code, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
+We present production benchmarks of Icicle compiled code on our smaller single-disk datasets, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
 \end{itemize}

### 2016-05-28 2016fhpc/section/02-Source.tex
some changes to introduction, s2, s3

@@
 To allow incremental computation, all Icicle queries must execute in a single pass over the input stream.
-Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order that is different to what a stream can provide.
-However, as we saw in the introduction, although it my not be possible to evaluate a particular \emph{algorithm} in an streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
+Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in a different order to what the stream provides.
+However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
 Here is the unstreamable example again from the introduction:
@@
 
-The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular sub-set of values from the input, but the value computed from this sub-set depends only on the values in that sub-set, and no other information.
+The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information.
 Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
@@
 
+When exposing impure computations whose value changes across successive runs, these must be bound as aggregates.
+For example, we can bind the current time @now@ as an @Aggregate Date@.
+This ensures that reused parts of an incremental computation are the same regardless of which day they are executed.
+
 % The @Element@ modality means that a computation is defined for each record in the table, or element in the stream.
@@
 The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero.
-This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then the positive values must be retained before any output can be produced.
+This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
 
@@
 In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly.
-Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on stream defined in the context.
+Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on the stream defined in the context.
 In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only.
@@
 
-Finally, note that our major caveat is that @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
+Finally, note that our major caveat is that the @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
 For our applications the keys are commonly company names, customer names, and days of the year.
-Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory.
+Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.
+% Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory.
 Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.
@@
 The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
-Value types are given in $T$ and can be numbers, booleans or maps.
+Value types, $T$, can be numbers, booleans or maps.
 Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
-Function types, $\TauFun$, can be either a non-function modal type, or a function from modal type to function type.
+Function types, $\TauFun$, are a list of argument types, and a single return type.
 As Icicle is a first-order language, function types are not value types.
 
-The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
+$\mi{Table}$ defines a table with a name and the names and types of columns.
 $\mi{Exp}$ defines the expressions used as query and function bodies.
+Primitive application is defined as prefix, but we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
 $\mi{Prim}$ defines the primitive types.
@@
 $\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
+All queries in a $\mi{Top}$ are fused into a single loop over the input data.
 
-Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
-The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
-Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
-Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
+We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@.
+We write a function @count@, which takes no arguments.
+Its definition is a fold that starts at zero, and increments by one for every element.
+\begin{code}
+  function count
+   = fold c = 0 then c + 1;
+\end{code}
+
+Next, @sum@ takes one argument, which is the elements to compute the sum of.
+Its definition is a fold that starts at zero, and increments by the sum element for each element.
+We can then define @mean@ as the sum divided by the count.
+\begin{code}
+  function sum  (v : Element Real)
+   = fold s = 0 then s + v;
+  function mean (v : Element Real)
+   = sum v / count;
+\end{code}
+
+Finally, the @last@ function takes the elements of keys, and for every element returns that element.
+Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
+\begin{code}
+  function last (k : Element Date)
+   = fold l = date 1900 1 1 then k;
+\end{code}
+
+% Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
+% The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
+% Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
+% Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
 
@@
 
-The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
+% The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
 
@@
 
-Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
+% Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
 
@@
 
-Group takes a key to group by, and a query to perform on the values of each grouping.
+% Group takes a key to group by, and a query to perform on the values of each grouping.
 
@@
 
-$\mi{Def}$ is used for function and query definitions.
+% $\mi{Def}$ is used for function and query definitions.
 % The @count@ function takes no arguments and returns a fold counting the number of elements.
@@
 
-Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
-All queries here operate over the same input table, and will be fused into a single loop over the input data.
+% Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
+% All queries here operate over the same input table, and will be fused into a single loop over the input data.
 

### 2016-05-30 2016fhpc/section/02-Source.tex
Merge conflicts

@@
 
-The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
-Value types, $T$, can be numbers, booleans or maps.
-Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
-Function types, $\TauFun$, are a list of argument types, and a single return type.
+The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
+Value types $T$ include numbers, booleans or maps.
+Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
+Function types $\TauFun$ include non-function modal types, or a function from modal type to function type.
 As Icicle is a first-order language, function types are not value types.
 
-$\mi{Table}$ defines a table with a name and the names and types of columns.
-$\mi{Exp}$ defines the expressions used as query and function bodies.
-Primitive application is defined as prefix, but we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
-$\mi{Prim}$ defines the primitive types.
+Table definitions $\mi{Table}$ define a table name and the names and types of columns.
+Expressions $\mi{Exp}$ can be variable names, constants, applications of primitives or functions.
+The @fold@ construct defines the name of an accumulator, the expression for the initial value, and the expression used to update the accumulator for each element of the stream.
+The @filter@ construct defines a predicate and an expression to accumulate values for which the predicate is true.
+The @group@ construct defines an expression used to determine the key for each element of the stream, and an expression to accumulate the values that share a common key.
+
+$\mi{Prim}$ defines the primitive operators.
 Function and query definitions are given in $\mi{Def}$.
 $\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
-All queries in a $\mi{Top}$ are fused into a single loop over the input data.
+All queries in a top-level program process the same table.
 
-We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@.
-We write a function @count@, which takes no arguments.
-Its definition is a fold that starts at zero, and increments by one for every element.
-\begin{code}
-  function count
-   = fold c = 0 then c + 1;
-\end{code}
+% We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@. 
+% We write a function @count@, which takes no arguments.
+% Its definition is a fold that starts at zero, and increments by one for every element.
+% \begin{code}
+%   function count
+%    = fold c = 0 then c + 1;
+% \end{code}
 
-Next, @sum@ takes one argument, which is the elements to compute the sum of.
-Its definition is a fold that starts at zero, and increments by the sum element for each element.
-We can then define @mean@ as the sum divided by the count.
-\begin{code}
-  function sum  (v : Element Real)
-   = fold s = 0 then s + v;
-  function mean (v : Element Real)
-   = sum v / count;
-\end{code}
+% Next, @sum@ takes one argument, which is the elements to compute the sum of.
+% Its definition is a fold that starts at zero, and increments by the sum element for each element.
+% We can then define @mean@ as the sum divided by the count.
+% \begin{code}
+%   function sum  (v : Element Real)
+%    = fold s = 0 then s + v;
+%   function mean (v : Element Real)
+%    = sum v / count;
+% \end{code}
 
-Finally, the @last@ function takes the elements of keys, and for every element returns that element.
-Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
-\begin{code}
-  function last (k : Element Date)
-   = fold l = date 1900 1 1 then k;
-\end{code}
+% Finally, the @last@ function takes the elements of keys, and for every element returns that element.
+% Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
+% \begin{code}
+%   function last (k : Element Date)
+%    = fold l = date 1900 1 1 then k;
+% \end{code}
 
 % Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
-% The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
+
+% a predicate, or grouping by a key.
 % Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
@@
 \subsection{Type system}
-The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
-Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
-Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
-Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
-Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
+The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
+The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ reads ``under context $\Gamma$, the expression $\mi{Exp}$ has type $M$.''. 
+The judgment form $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
+The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
+The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
 Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

### 2016-05-30 2016fhpc/section/01-Introduction.tex
Shift cut out text to separate files
Copy edits to intro

@@
 
-Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion, which allows us to update the result as we receive new data.
+Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion, which allows the result to be updated as we receive new data.
 As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table.
@@
 Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an inceremental fashion.
-While streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start.
+While streaming through the table we always have access to the last key in the stream, but to select rows that match this key would require that we stream through the table again from the start.
 We need a better solution.
@@
 \item
-We present a domain specific language and type system that guarantees any two queries on the same input table can be fused, and allows queries to be resumed as new data is received\sref{s:IcicleSource};
+We present a domain specific language that guarantees any set of queries on a shared input table can be fused, and allows the query results to be updated as new data is received\sref{s:IcicleSource};
 
@@
 \item
-We present production benchmarks of Icicle compiled code on our smaller single-disk datasets, which compare favourably against standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
+We present production benchmarks of Icicle compiled code which compare favourably to standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
 \end{itemize}
 
+% on our smaller single-disk datasets
+% BEN: for streaming applications the absolute size of the input size doesn't affect the performance, so we don't need to apologise.
+
 \TODO{Touch on related work}
-
-% This is achieved by having a very restricted query language with no sorting, and allowing only a single-pass over the data.
-% \ben{Maybe leave mention of sorting until later sections}.
-
-% \ben{Rewrite the examples using this syntax?}
-% In Icicle, the above queries are expressed as:
-% \begin{code}
-% table stocks { open : Int; close : Int; }
-% query more = filter open > close in count;
-% query less = filter open < close in count;
-% query mean = filter open > close in sum open / count;
-% \end{code}
-
-% \ben{the abstract already says this.}
-% For queries over a large amount of data, care must be taken to reduce the number of iterations over the input data.
-When performing multiple queries, it is important to limit the amount of duplicate work and iterations.
-
-% \ben{Start with the example.}
-% \begin{code}
-%   SELECT COUNT(*) FROM stocks WHERE open > close;
-%   SELECT COUNT(*) FROM stocks WHERE open < close;
-% \end{code}
-% As both queries are over the same input data set, we would like to evaluate them using a single pass over this data.
-As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:
-
-% In order to fuse these so that both are computed in a single query, some ingenuity is required.
-This is because the @WHERE@ clause filters over the entire query, but when putting both in one query we need two different filters.
-
-% \begin{code}
-%   SELECT COUNT(IF(open > close, 1, 0))
-%        , COUNT(IF(open < close, 1, 0))
-%   FROM stocks;
-% \end{code}
-% Suppose we also want to compute the mean of open price on those days:
-% \begin{code}
-%   SELECT SUM(open) / COUNT(*)
-%   FROM   stocks
-%   WHERE  open > close;
-% \end{code}
-
-% These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.

### 2016-05-30 2016fhpc/section/02-Source.tex
Shift cut out text to separate files
S2 copy edits

@@
 
-To allow incremental computation, all Icicle queries must execute in a single pass over the input stream.
-Of course, not all queries can be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in a different order to what the stream provides.
+To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
+Of course, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
 However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
-Here is the unstreamable example again from the introduction:
+Here is the unstreamable example from the introduction again:
 \begin{code}
@@
 
-The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs it to process the very first element in the same stream.
-We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate}, because to compute it we need to have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation because it only needs access to the current element in the stream.
+The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs this value to process the very first element in the same stream.
+We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate}, because to compute it we must have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation because it only needs access to the current element in the stream.
 
-The trick to compute our average in an incremental fashion is to recognise that the @filter@ operation is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information.
+The trick to compute our average in an incremental fashion is to recognise that @filter@ is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information.
 Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
@@
 
-Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and to compute running mean of the values of each group.
+Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and compute running mean of the values of each group.
 The @avgs@ value becomes a map of group keys to their running means.
-Once we have reached the end of the stream we then have access to the last key and can lookup the final result.
+Once we reach the end of the stream we will have access to the last key and can lookup the final result.
 
@@
 
-% The @Element@ modality means that a computation is defined for each record in the table, or element in the stream.
-Each column in the table is @Element@; for example the @open@ column has type @Element Int@, meaning that for each record in the column there is an @Int@.
-This can be thought of as being represented by a stream of values.
-
-% The @Aggregate@ modality means that a computation is available only after all records in the table have been seen, or after the end of the stream.
-These are used for the results of folds.
-When computing count, the final value is not known until all the records have been seen, so @count@ has type @Aggregate Int@.
-
-% Hopefully our programmer is not too discouraged by the @Element /= Aggregate@ type error, and pushes on.
-They now know that their query, as formulated, requires multiple passes over the data: the problem now is to reformulate it as a single pass.
-With a little ingenuity, we can rewrite it as such: we can group by the key and perform the mean for each group.
-After the group, we can perform a lookup by the most recent key.
-This requires storing and computing the means for all keys despite only needing one at the end, so we are assuming that the number of keys are bounded in some way.
-
 % ---------------------------------------------------------
@@
 Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.
-% Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory.
-Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.
-
-% Here, streams are given clock types denoting when the streams have elements available, and only streams with the same clock can be zipped together.
-Filters produce a stream with a different output clock to its input, so different filters cannot be zipped together.
-Our system requires no clock analysis, but is less expressive as streams are not `first class'.
 
@@
 All queries in a top-level program process the same table.
-
-% We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@. 
-% We write a function @count@, which takes no arguments.
-% Its definition is a fold that starts at zero, and increments by one for every element.
-% \begin{code}
-%   function count
-%    = fold c = 0 then c + 1;
-% \end{code}
-
-% Next, @sum@ takes one argument, which is the elements to compute the sum of.
-% Its definition is a fold that starts at zero, and increments by the sum element for each element.
-% We can then define @mean@ as the sum divided by the count.
-% \begin{code}
-%   function sum  (v : Element Real)
-%    = fold s = 0 then s + v;
-%   function mean (v : Element Real)
-%    = sum v / count;
-% \end{code}
-
-% Finally, the @last@ function takes the elements of keys, and for every element returns that element.
-% Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
-% \begin{code}
-%   function last (k : Element Date)
-%    = fold l = date 1900 1 1 then k;
-% \end{code}
-
-% Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
-
-% a predicate, or grouping by a key.
-% Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
-% Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
-
-% Let-bindings allow a name to be used in place of the expression, in the rest of some query.
-% \begin{code}
-% let diff = open - close
-% in  mean diff
-% \end{code}
-
-% The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
-
-% Count can be expressed by the following fold:
-% \begin{code}
-% fold count = 0 then count + 1
-% \end{code}
-
-% Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
-
-% The following query will count the number of entries where the open price exceeds the close price.
-% \begin{code}
-% filter open > close in count
-% \end{code}
-
-% Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
-% This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
-% \begin{code}
-% (filter open > close in count) / count
-% \end{code}
-
-% Group takes a key to group by, and a query to perform on the values of each grouping.
-
-% This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
-% \begin{code}
-% group ((open - close) / 100) in count
-% \end{code}
-
-% $\mi{Def}$ is used for function and query definitions.
-% The @count@ function takes no arguments and returns a fold counting the number of elements.
-% \begin{code}
-% function count
-%  = fold c = 0 then c + 1;
-% \end{code}
-
-% Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
-% It returns a fold that starts at zero and for every element, increases the old sum by @value@.
-% \begin{code}
-% function sum (value : Element Int)
-%  = fold s = 0 then s + value;
-% \end{code}
-
-% The definition of mean is just the sum divided by the count.
-% \begin{code}
-% function mean (value : Element Int)
-%  = sum value / count;
-% \end{code}
-
-% Queries must return a single value, and are given it a name:
-% \begin{code}
-% query count_all_records = count;
-% \end{code}
-
-% Finally, $\mi{Top}$ puts all these together with a table definition, function definitions, and query definitions in one place.
-% All queries here operate over the same input table, and will be fused into a single loop over the input data.
 

### 2016-05-30 2016fhpc/section/02-Source.tex
Copy edits to bounded buffer restriction

@@
 \section{Elements and Aggregates}
-
 To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
@@
 We use two modalities, @Element@ and @Aggregate@.
-Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed.
+Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are available only once the entire stream has been consumed.
 In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Date@ while @k@ has type @Aggregate Date@.
-Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.
+Attempting to compile the unstreamable query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.
 
-Note that the types of pure values such as constants are automatically promoted to have the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.
+Note that the types of pure values such as constants are automatically promoted to the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.
 
@@
 \subsection{Finite Streams and Synchronous Data Flow}
-As opposed to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
+In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
 Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner.
-In contrast, Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes).
-In Icicle, the ``last'' element in a stream is the last one that appears in the file on disk.
+Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes).
+In Icicle, the ``last'' element in a stream is the last one that appears in the table on disk.
 In Lustre, the ``last'' element in a stream is the one that was most recently received.
-If the ``unstreamable'' version of the query from \S2 was converted into Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself.
-This means the filter predicate would always be true, and the query would return the mean of the entire stream.
+If the unstreamable query from \S2 was converted to Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself.
+The filter predicate would always be true, and the query would return the mean of the entire stream.
 Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.
@@
 \subsection{Incremental Update}
-
-Suppose we apply a query to a large table and record the result.
+Suppose we query a large table and record the result.
 Tomorrow morning we receive more data and add it to the table.
-We would like to recompute the result of this query without needing to process all data from the start of the table.
-We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives.
+We would like to update the result without needing to process all data from the start of the table.
+We can do this by remembering the values of all intermediate aggregates that were computed in the query, and updating them as new data arrives.
 In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.
 
-When exposing impure computations whose value changes across successive runs, these must be bound as aggregates.
-For example, we can bind the current time @now@ as an @Aggregate Date@.
-This ensures that reused parts of an incremental computation are the same regardless of which day they are executed.
+We can also provide impure contextual information to the query such as the current date, by assigning it an aggregate type.
+As element-wise computations cannot depend on aggregate computations, we ensure that reused parts of an incremental computation are the same regardless of which day they are executed.
 
@@
 \label{s:IcicleSource:bounded}
-Icicle queries are intended to process tables of arbitrary size that may not fit in memory.
+Icicle queries process tables of arbitrary size that may not fit in memory.
 Due to this, each query must execute without requiring buffer space proportional to the size of the input.
-For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
+As a counter example, here is a simple function which cannot be applied without reserving a buffer of the same size as the input:
 \begin{code}
-    unbounded (xs :: Stream Int)
+    unbounded (xs : Stream Int)
      = zip (filter (> 0) xs) (filter (< 0) xs)
@@
 
-The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero.
-This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
+This function takes an input stream @xs@, and pairs the elements that are greater than zero with those that are less than zero.This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
 
-The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
+Icicle is designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
 In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly.
-Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on the stream defined in the context.
-In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only.
+Constructs such as @filter@ and @fold@ do not take the name of the stream as an argument, but instead operate on the stream defined in the context.
+In the example Icicle queries presented so far, the name of the table (stream) is not mentioned in the body of the query, and serves as a comment only.
 Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query.
@@
 
-Finally, note that our major caveat is that the @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
+Finally, note that our major caveat is that the @group@ construct we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
 For our applications the keys are commonly company names, customer names, and days of the year.
-Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.
+Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory.
+Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.
+
+% Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.
+
+% BEN: "Can be undesirable" is too imprecise.
+What will happen is that an entry will be added to the map for every key, and if there are more keys than will fit in memory then the query will run out of memory.
+Surely adding the mentioned warning to the compiler would not be difficult?
+If it's easy then saying it's done sounds much better than "we intend to".
+We need to have a solid story about this point.
+One of the key contributions of Icicle is that it does not require ``unbounded buffering'', but if you're grouping by arbitrary values from the input stream then it clearly does.
 

### 2016-05-30 2016fhpc/section/02-Source.tex
Update meta-variables in type rules to match grammar

@@
 The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
-The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ reads ``under context $\Gamma$, the expression $\mi{Exp}$ has type $M$.''. 
+The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ associates an expression $\mi{Exp}$ with its type $M$ under context $\Gamma$.
 The judgment form $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
@@
 
-Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$.
+Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$. Rule TcBox assigns a pure value either @Element@ or @Aggregate@ type.
 
-Rule TcBox shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
-The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.
+% shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
+% The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.
 
-Rules TcPrimApp and TcFunApp find the result type of primitive and function application.
-They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
-The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
+Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments.
+
+% find the result type of primitive and function application.
+% They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
+% The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
 

### 2016-05-30 2016fhpc/section/02-Source.tex
Pack down description of typing rules

@@
 
-Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments.
+Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments. Rule TcLet is standard.
 
@@
 % The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
+% Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
 
-Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
+In rule TcFold the initial value has value type $T$.
+The type of the current element from the stream is added to the context of $e_k$ as an @Element@, and the result of the overall fold is an @Aggregate@.
+Rules TcFilter and TcGroup are similar.
 
-Rule TcFold finds the type of a @fold@ expression.
-The initial value of the fold must be pure.
-The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
-The final return type of the fold is an @Aggregate@.
+% The initial value of the fold must be pure.
+% The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
+% The final return type of the fold is an @Aggregate@.
 
-Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
-The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
-Rule TcGroup is similar to filter, except returning a @Map@.
+% Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
+% The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
+% Rule TcGroup is similar to filter, except returning a @Map@.
 
-The rules PrimArith, PrimRel and PrimLookup give function types for each primitive.
+The rules PrimArith, PrimRel and PrimLookup assign types to primitives.
+Rule AppArgs produces the type of a function or primitive applied to its arguments.
+Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.
 
-The next judgment shows how function application boxes and unboxes modalities.
-This judgment takes a function type and all its arguments, and returns the result type of the function.
+Rule CheckFun builds the type of a user defined function, returning it as an element of the output context.
+Rule CheckQuery is similar, noting that all queries return values of @Aggregate@ type.
+Finally, rule CheckTop checks a whole top-level program.
 
-Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
+% The next judgment shows how function application boxes and unboxes modalities. This judgment takes a function type and all its arguments, and returns the result type of the function.
 
-Rule AppRebox is used when a pure function is applied to a non-pure value.
-If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
-Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
-These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.
+% Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
 
-Rule CheckFun inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
+% Rule AppRebox is used when a pure function is applied to a non-pure value.
+% If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
+% Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
+% These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.
 
-Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.
+% inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
 
-Finally, rule CheckTop checks a whole top-level program.
-It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
-Then, each query is checked, with the table columns and all functions available.
-The final returned context is all query names and their types.
+% Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.
 
+% It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
+% Then, each query is checked, with the table columns and all functions available.
+% The final returned context is all query names and their types.
+
+% ---------------------------------------------------------
 \subsection{Evaluation}
-
 The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.

### 2016-05-30 2016fhpc/section/03-Core.tex
some minor changes and remove GVN stuff from related
Adjust spacing in intermediate language section

@@
 
-\begin{code}
-Exp   := x | Prim Exp*
-
-Plan  := plan   x { (x : T;)* }
-       before { x : T = Exp;* }
-       folds  { x : T = Exp then Exp;* }
-       after  { x : T = Exp;* }
-       return { x : T = x;* }
-\end{code}
+\begin{tabbing}
+MMM \= MM \= MMMM \= MM \= MMMMMM \= \kill
+$\mi{PlanX}$
+\GrammarDef
+  $x~|~@Int@~|~\mi{Prim}~\ov{\mi{PlanX}}$
+\\
+\\
+$\mi{Plan}$
+\GrammarDef
+  $@plan@~x$ \> $\{~\ov{x~:~T;}~\}$
+\\
+  \> \> $@before@$ \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
+  \> \> $@folds@$  \> $\{~\ov{x~:~T~=~\mi{PlanX}~@then@~\mi{PlanX};}~\}$ \\
+  \> \> $@after@$  \> $\{~\ov{x~:~T~=~\mi{PlanX};}~\}$ \\
+  \> \> $@return@$ \> $\{~\ov{x~:~T~=~x;}~\}$ \\
+\end{tabbing}
 
@@
 
-==>
-
-plan stocks { open : Int; close : Int; }
+==> plan stocks { open : Int; close : Int; }
     before { }
@@
 
-An efficient imperative loop can be generated from this query plan rather easily, in a similar way as previous work on flow fusion\cite{lippmeier2013data}.
+An efficient imperative loop can be generated from this query plan rather easily, in a similar way as previous work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex

@@
 	ybar interval=0.7,
-  symbolic x coords={R, wc, grep, wc -l, cat, Run 372, Run 1, end}
+  symbolic x coords={R, wc, CPU1, grep, wc -l, cat, Run 372, Run 1, end}
 ]
-\addplot coordinates {(R,6.0) (wc,182.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
+\addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
 
@@
 
+% \begin{figure}
+% \begin{tikzpicture}
+% \begin{axis}[
+% 	x tick label style={/pgf/number format/1000 sep=},
+% 	ylabel=MB/s,
+%   ymin=0, ymax=1200,
+%   xlabel=Number of threads,
+% 	enlargelimits=0.01,
+% 	legend style={legend columns=-1},
+% ]
+% \addplot coordinates { (1,163) (2,335) (3,442) (4,532) (5,578) (6,615) (7,633) (8,628) (9,631) (10,626) (11,637) (12,643) (13,658) (14,795) (15,781) (16,798) (17,840) (18,864) (19,862) (20,867) (21,847) (22,877) (23,876) (24,884) (25,862) };
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Scaling as threads are added}
+% \label{fig:bench:scaling}
+% \end{figure}
+
+\TODO{update to talk about single CPU number.
+Single CPU version 30 times faster than R, and in the same ballpark as @wc@ and @grep@.
+Because the memory usage is bounded, we have a better idea of how many we can run on a single machine without running out of memory and stepping on each other's toes.
+This means we can scale up to get quite close to the disk speed.
+This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
+Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.}
+
 The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.

### 2016-06-02 2016fhpc/section/02-Source.tex
minor changes
start on core conversion of mean group

@@
 $\mi{Prim}$ defines the primitive operators.
-Function and query definitions are given in $\mi{Def}$.
+$\mi{V}$ defines values.
+$\mi{Def}$ contains both function and query definitions.
 $\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
@@
 
-Rule TcVar performs variable lookup in the context, while TcNat shows that number constants have type $\NN$. Rule TcBox assigns a pure value either @Element@ or @Aggregate@ type.
+Rules TcNat, TcBool, TcMap and TcPair assign types to literal values.
+Rule TcVar performs variable lookup in the context.
+Rule TcBox assigns a pure expression either @Element@ or @Aggregate@ type.
 
@@
 
-The rules PrimArith, PrimRel and PrimLookup assign types to primitives.
+The rules PrimArith, PrimRel, PrimLookup, PrimFst and PrimSnd assign types to primitives.
 Rule AppArgs produces the type of a function or primitive applied to its arguments.
@@
 The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
+Grammar $N$ defines the modes of evaluation, including pure computation.
+Grammar $\Sigma$ defines a heap containing stream values.
+Grammar $V'$ defines a different type of stream value for each evaluation mode:
+\begin{itemize}
+\item
+@Pure@ computations are represented as a single value;
+\item
+@Element@ stream transformers are represented as a function applied to each element; and
+\item
+@Aggregate@ folds are represented as an initial (zero) state, an update (konstrukt) function applied to each stream element and current state, and an eject function applied to the final state.
+\end{itemize}
 
-Definition $V$ are simple values of numbers, booleans, maps and tuples.
-Definition $V'$ are stream values that correspond to modalities: pure computations are represented as a single value;
-@Element@s as a stream transformer from a heap to a value;
-and @Aggregate@s as folds defining an initial state and update and extract functions.
-Definition $M'$ are the modalities themselves without value types.
+The judgment form $\SourceStepX{N}{\Sigma}{e}{V'}$ defines a big-step evaluation relation: under evaluation mode $N$ with heap $\Sigma$, expression $e$ evaluates to stream value $V'$.
+The evaluation mode is used to know when pure values should be converted to streams or folds.
+We assume that all functions have been inlined into the expression before evaluation.
 
-The evaluation rules operate over a slightly modified expression type, $E'$, which is an $\mi{Exp}$ with $V'$ as values instead of just simple values and natural numbers.
-This is to simplify typechecking over $V'$, and as $V'$ only appear in restricted form for evaluation.
-We also assume that all functions have been inlined before evaluation.
+Rule EVal applies when the expression is already a stream value.
+Rule EVar performs variable lookup in the heap.
+Rule ELet uses the bound expression's type to find the evaluation mode, then evaluates the bound expression under that mode.
+The bound expression's value is added to the heap, and the rest of the expression is evaluated in the original evaluation mode.
 
-The judgment $\SourceStepX{M'}{E'}{V'}$ defines a big-step evaluation relation.
-The rules below use the modality in deciding whether pure values should be converted to streams or folds.
-
-Rule EVal applies when the expression is alread a stream value.
-This corresponds roughly to the TcNat typechecking rule.
-
-Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the environment modality is not pure.
-The stream transformer ignores the input stream and returns the pure value.
-The fold has a unit accumulator, and the extract function simply returns the pure value.
+Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the evaluation mode is not pure.
+Conversion to a stream transformer ignores the input stream and returns the pure value.
+Conversion to a fold has no state, and the eject function returns the pure value.
 
@@
 For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
-Folds are the most complicated: the fold accumulators are tupled together, for example $(\times_i~z_i)$ being the tuple of all the initial accumulators.
-At each step, the accumulators for all the input folds are updated with their own update function, and the input heap.
-Finally, the extract function extracts the final values of the input folds, and applies the primitive.
+For folds, the states of all the argument folds are joined together as tuples.
+At each step, the states of the input folds are updated using their own update function.
+The eject function extracts the ejected values of the input folds, and applies the primitive.
 
-Rule ELet rule is fairly standard, evaluating the bound to a value, substituting it in, and evaluating the rest.
-The only difference is we inspect the type of the bound expression to find the modality, then evaluate the bound under that modality.
-The rest of the expression is evaluated under the original modality.
-
-Rule EFold is one of the most complex, as it introduces recursion to the streams.
+Rule EFold introduces recursion and memory to the streams.
 It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
-In the update expression, the fold binding $x$ is substituted with a stream that looks up $x$ in its input heap.
-This sleight of hand is necessary because the value of $x$ depends on the value of the fold, which is not known yet.
-Finally, the return fold is constructed, which binds $x$ to the current fold accumulator.
-The extract function for this is the identity, simply returning the accumulator.
+The update expression has mode @Element@, so it must be evaluated to a stream.
+This stream expects the stream element as an argument, but we need to give it the current fold state as well.
+In order to pass both as a pair, we modify all streams in the heap to use the second half of the pair.
+We also add a stream for the fold state to the heap; this uses the first half of the pair.
+The update function then calls the stream transformer with a pair of the state and the input element.
+The eject function is the identity function.
 
@@
 Rule EGroup is rather similar to filter, except that it returns a map.
-Here, we start with an empty map of bottoms.
-At every update step, we find the key and look up the current accumulator for that key in the map, or the fold initialiser if it is bottom.
+Here, we start with a map of every key to the fold initialiser.
+At every update step, we find the key and look up the current accumulator for that key in the map.
 We then apply the update function to the accumulator, and update that key in the map.
-For the extract function, we simply apply the extract to each element of the map.
+For the eject function, we simply apply the eject to each element of the map.
 
-In order to execute these evaluation rules on an actual input table, a few steps are necessary.
-First, variables referring to column names such as @open@ or @close@ must have stream values substituted in, so that the actual value will be looked up in the input heap.
-For example, $e[@open@~:=~@Stream@~(\lam{\store} \store~@open@)]$.
-Next, the query must be converted to a @Fold@.
-Then, for each record, the update function would be called with an input heap binding the record values: $\{@open@ = 1, @close@ = 2\}$ and so on.
-This occurs for every record in the table.
-Finally, the extract function is called, which gives the result of the query.
+The judgment form $t~|~e~\Downarrow~V$ evaluates an expression over a table input: on input table $t$, aggregate expression $e$ evaluates to value $V$.
+The input table $t$ is a map from column name to a list of all the values for that column.
+Rule ETable creates an initial heap where each column is a stream transformer that pulls out the $n$th element of a nested tuple.
+Tuples are constructed correspondingly, so that each row becomes a nested tuple where the $n$th element contains the value of the $n$th column.

### 2016-06-02 2016fhpc/section/03-Core.tex
start on core conversion of mean group

@@
 %!TEX root = ../Main.tex
+\clearpage
 \section{Intermediate language}
 \label{s:IcicleCore}
+\TODO{PAGE LAYOUT REMOVE CLEARPAGE}
 
@@
 \GrammarDef
-  $x~|~@Int@~|~\mi{Prim}~\ov{\mi{PlanX}}$
+  $x~|~V~|~\mi{PlanP}~\ov{\mi{PlanX}}~|~\lam{x}\mi{PlanX}$
+\\
+$\mi{PlanP}$
+\GrammarDef
+  $\mi{Prim}~|~@mapUpdate@~|~@mapEmpty@~|~@mapMap@~|~@mapZip@$
 \\
@@
 
-\caption{Grammar}
+\caption{Query Plan Grammar}
 \label{fig:core:grammar}
@@
 The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
-Expressions can be variables or primitive applications: function applications are not allowed in expressions here, as their definitions are inlined before converting to query plans.
-The @Plan@ itself is split into stages of computation, which correspond to the modal stages in the Source query.
+Expressions can be variables, values, applications of primitives or anonymous functions.
+Function definitions and uses are not allowed in expressions here, as their definitions are inlined before converting to query plans.
+Anonymous functions are only allowed as arguments to primitives: they cannot be applied or stored in variables.
+The $\mi{Plan}$ itself is split into stages of computation, which correspond to the modal stages in the Source query.
 The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
@@
 
-One of the simplest queries to convert to a query plan is a count.
-The query plan header specifies the table and the type of each column in the table.
-This defines a single fold called @count@ which starts at zero, and is incremented for each element.
-At the end of the query, the final count is assigned to the output @count_query@, which is the query's name.
+We will start with a small example by converting just the @last key@ part of the group-mean query.
+Here, we have the function definition for @last k@, which is defined as a fold with an initial value to describe a missing date @NO_DATE@, and the update value is the current @k@.
+The current fold value is bound to @l@ but is not used.
+
 \begin{code}
-    table stocks { open : Int; close : Int; }
-    query count_query = count
+    table kvs { key : Date; value : Real }
 
-==> plan stocks { open : Int; close : Int; }
-    before { }
-    folds  { count : Int = 0 then count + 1 }
-    after  { }
-    return { count_query : Int = count }
+    function last (d : Element Date)
+     = fold l = NO_DATE then d;
+
+    query last_key = last key
 \end{code}
 
-Computing the sum is only slightly more complicated.
-Here there is again a single fold, @sum@, starting at zero and being incremented by the current column's @open@.
+We now inline function definitions before converting to a query plan, renaming as necessary.
+This is guaranteed to terminate because no recursion is allowed in function definitions.
+
 \begin{code}
-    query sum_open = sum open
-
-==> plan stocks { open : Int; close : Int; }
-    before { }
-    folds  { sum : Int = 0 then sum + open }
-    after  { }
-    return { sum_open : Int = sum }
+==> table kvs { key : Date; value : Real } 
+    query last_key = (fold c = NO_DATE then key)
 \end{code}
 
-In order to compute the mean, we compute the sum and the count, and after all computation has finished, divide the two.
-Here we use two folds, one for the sum, and one for the count.
-While we could have used a single fold containing a pair of the two, keeping the folds separate exposes more opportunities for common subexpression elimination.
+We can now convert this to a query plan.
+
 \begin{code}
-    query mean_open = sum open / count
+==> plan kvs { key : Date; value : Real; }
+    before { }
+    folds  { c : Date = NO_DATE then key }
+    after  { }
+    return { last_key : Date = c }
+\end{code}
 
-==> plan stocks { open : Int; close : Int; }
+We will now convert the mean group.
+Here, we have function definitions for @count@, @sum@ and @mean@.
+Count is defined by starting from zero, and incrementing by one for each element.
+Sum starts at zero and increments its argument.
+Mean is then defined as sum divided by the count.
+
+\begin{code}
+    table kvs { key : Date; value : Real }
+
+    function count
+     = fold c = 0 then c + 1;
+
+    function sum (e : Element Real)
+     = fold s = 0 then s + e;
+
+    function mean (e : Element Real)
+     = sum e / count;
+
+    query mean_all = group key of mean value
+\end{code}
+
+After function inlining, the result is this:
+\begin{code}
+==> table kvs { key : Date; value : Real } 
+    query mean_all = group key
+     of (fold foldS = 0 then foldS + value)
+      / (fold foldC = 0 then foldC + 1)
+\end{code}
+
+Finally, we have the query plan.
+Each fold inside the group becomes its own map which is initialised to empty.
+At each iteration the fold update is applied to the value in the map, or the fold's initial value if the key is not yet present.
+After all iterations are finished, the two maps are joined together and the pairs are divided.
+\begin{code}
+==> plan kvs { key : Date; value : Real; }
+
     before { }
-    folds  { sum       : Int = 0 then sum + open 
-             count     : Int = 0 then count + 1 }
-    after  { mean      : Int = sum / count }
-    return { mean_open : Int = mean }
+
+    folds  { groupS   : Map Date Real
+                = mapEmpty
+             then mapUpdate groupS key 0
+                  (\foldS. foldS + value)
+
+           ; groupC   : Map Date Real
+                = mapEmpty
+             then mapUpdate groupC key 0
+                  (\foldC. foldC + 1) }
+
+    after  { groupM   : Map Date Real
+                = mapMap (/) (mapZip groupS groupC) }
+
+    return { mean_all : Map Date Real = groupM }
 \end{code}
+
+It is possible to convert this as a map of pairs of the sum and count.
+Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
+For example, a query @group key of count@ can reuse the already constructed @groupC@ map.
 
@@
 Query plans can only be fused if they act on the same table.
-\begin{code}
-    plan stocks { open : Int; close : Int; }
-    before { }
-    folds  { c1 : Int = 0 then c1 + 1
-             c2 : Int = 0 then c2 + 1
-             s1 : Int = 0 then s1 + open 
-             s2 : Int = 0 then s2 + open }
-    after  { m1 : Int = s2 / c2 }
-    return { count_query : Int = c1
-             sum_open    : Int = s1
-             mean_open   : Int = m1 }
-\end{code}
 
 The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
-We can perform fairly standard common subexpression elimination to remove these.
-\begin{code}
-    plan stocks { open : Int; close : Int; }
-    before { }
-    folds  { c1 : Int = 0 then c1 + 1
-             s1 : Int = 0 then s1 + open }
-    after  { m1 : Int = s1 / c1 }
-    return { count_query : Int = c1
-             sum_open    : Int = s1
-             mean_open   : Int = m1 }
-\end{code}
+We can perform common subexpression elimination to remove these.
 
@@
 The single-pass restriction means that any source query can be expressed as a single fold over the data.
-By keeping the folds separate we have exposed more opportunities for common subexpression elimination.
 

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
start on core conversion of mean group

@@
 This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
-Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.}
+Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.
+Blah~\cite{mcsherry2015scalability}
+}
 

### 2016-06-02 2016fhpc/section/XX-Related.tex
start on core conversion of mean group

--- 2016-04-20 2016fhpc/section/XX-Related.tex
@@
 
-The most closely related transforms are induction variable elimination\cite{shivers1988control} and global value numbering\cite{rosen1988global}.
-Induction variable elimination finds and removes accumulators that are linear functions of the iteration number.
-This does not deal with non-linear functions such as sum, or mutually recursive accumulators.
+The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
+These systems perform a clock, or \emph{rate} analysis of the streams being processed.
+Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
 
-Global value numbering works on an SSA form and is much more general than induction variable elimination.
-Early versions such as Rosen et al\cite{rosen1988global} only worked on extended basic blocks with backedges removed.
-Alpern et al\cite{alpern1988detecting} introduced cases for specific loop backedges, where a fixpoint is performed to find the congruence sets of loops.
-More recently, Gulwani and Necula\cite{gulwani2004polynomial} improved upon this, supporting removal of more expressions, and imposing a polynomial time bound.
-Global value numbering for loops can perform all optimisations here.
+\TODO{fill}
+Advances in dataflow programming languages~\cite{johnston2004advances}.
+Survey of stream processing~\cite{stephens1997survey}.
+How to architect a query compiler~\cite{shaikhha2016architect}.
+StreamIt~\cite{thies2002streamit}.
 
-Common subexpression elimination is very closely related, but most imperative compilers only perform common subexpression elimination on straight-line computations with no control flow\cite{debray1992compiler}, or on loop invariant expressions\cite{bodik1998complete}.
-Common subexpression elimination for functional languages would be able remove lone expressions, but does not deal with mutually recursive folds.
+Kahn process networks are a flexible alternative to MapReduce~\cite{vrba2009kahn}.
+Scheduling dynamic dataflow~\cite{buck1993scheduling}.
+Co-iterative characterization~\cite{caspi1998co}.
 
-Temporal common subexpression elimination in Single Assignment C
-allows reuse of expressions computed in the previous iteration\cite{imlig2001loop}.
-This is a more general case of the elimination opportunities that arise from loop unrolling.
-
-For example, the program below loops over an array @A@, and stores the product @a@ of some computation on the current element (@f(A[i])@), while performing the same computation over the next element (@f(A[i+1])@) and summing it in @b@.
-On successive iterations, the @bx = f(A[i+1])@ computed from last iteration could be reused as the new @ax = f(A[i])@.
-\begin{code}
-int[] A;
-int a = 1;
-int b = 0;
-while(int i = 0; i != size - 1; i++)  {
-  int ax = f(A[i]);
-  int bx = f(A[i+1]);
-  a += ax;
-  b *= bx;
-}
-\end{code}
-
-I cannot think of a better example than this right now.
+Compare to short-cut fusion~\cite{coutts2007stream}.
 
@@
 By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
-This is a lot closer, and performs CSE on expressions like @x = sum(y)@, where @sum@ is a built-in aggregate function.
 However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.
-
-The polyhedral model\cite{benabderrahmane2010polyhedral} analyses data dependencies in loops.
-For a given iteration, the polyhedral model computes the iterations which must be executed before the given iteration.
-By knowing the dependent iterations, one can restructure the loop so that the iterations are executed in a different order, while still performing any dependencies in order.
-For example, a loop iterating over @i@, with loop body @A[i] = A[i-10]@ depends on the tenth previous iteration.
-This loop could be executed in an order like @0, 1, 2,...@, but it could also stride by ten: @0, 10, 20..., 1, 11, 21,...@.
-This allows fusion of any two loops if the intersection of the two loops' iteration spaces is not empty.
-In our case the iteration order is fixed, as the program cannot control the order in which stream elements are received.
-The polyhedral model may still be applicable to streaming computations where one could introduce certain-sized buffers, then process the buffer in parallel.
-In this case the polyhedral model may expose more opportunities for duplicate elimination, but does not itself remove duplicates.

### 2016-06-02 2016fhpc/section/03-Core.tex
intermediate language

@@
 %!TEX root = ../Main.tex
-\clearpage
 \section{Intermediate language}
 \label{s:IcicleCore}
-\TODO{PAGE LAYOUT REMOVE CLEARPAGE}
 
@@
 
-We will start with a small example by converting just the @last key@ part of the group-mean query.
-Here, we have the function definition for @last k@, which is defined as a fold with an initial value to describe a missing date @NO_DATE@, and the update value is the current @k@.
-The current fold value is bound to @l@ but is not used.
+Before we can convert our example query to a query plan, we must look at the definitions of the functions @count@, @sum@, @mean@ and @last@.
+Here, we have definitions of @count@ and @sum@.
+Count is defined by starting from zero, and incrementing by one for each element.
+Sum starts at zero and increments its argument.
 
 \begin{code}
-    table kvs { key : Date; value : Real }
-
-    function last (d : Element Date)
-     = fold l = NO_DATE then d;
-
-    query last_key = last key
-\end{code}
-
-We now inline function definitions before converting to a query plan, renaming as necessary.
-This is guaranteed to terminate because no recursion is allowed in function definitions.
-
-\begin{code}
-==> table kvs { key : Date; value : Real } 
-    query last_key = (fold c = NO_DATE then key)
-\end{code}
-
-We can now convert this to a query plan.
-
-\begin{code}
-==> plan kvs { key : Date; value : Real; }
-    before { }
-    folds  { c : Date = NO_DATE then key }
-    after  { }
-    return { last_key : Date = c }
-\end{code}
-
-We will now convert the mean group.
-Here, we have function definitions for @count@, @sum@ and @mean@.
-Count is defined by starting from zero, and incrementing by one for each element.
-Sum starts at zero and increments its argument.
-Mean is then defined as sum divided by the count.
-
-\begin{code}
-    table kvs { key : Date; value : Real }
-
     function count
@@
      = fold s = 0 then s + e;
+\end{code}
 
+Mean can then defined as sum divided by the count.
+\begin{code}
     function mean (e : Element Real)
      = sum e / count;
-
-    query mean_all = group key of mean value
 \end{code}
 
-After function inlining, the result is this:
+Here, we have the function definition for @last k@, which is defined as a fold with an initial value to describe a missing date @NO_DATE@\footnote{In our production compiler, @last@ returns a @Maybe@.}, and the update value is the current @k@.
+The current fold value is bound to @l@ but is not used.
+
 \begin{code}
-==> table kvs { key : Date; value : Real } 
-    query mean_all = group key
-     of (fold foldS = 0 then foldS + value)
-      / (fold foldC = 0 then foldC + 1)
+    function last (d : Element Date)
+     = fold l = NO_DATE then d;
 \end{code}
 
-Finally, we have the query plan.
-Each fold inside the group becomes its own map which is initialised to empty.
+% \begin{code}
+%   query avg = let k    = last key in
+%               let avgs = group key of mean value
+%               in  lookup k avgs
+% \end{code}
+
+We now inline these function definitions before converting to a query plan, renaming as necessary.
+This is guaranteed to terminate because no recursion is allowed in function definitions.
+\begin{code}
+  query avg
+   =    let lst = (fold l = NO_DATE then key)
+     in let map = group key of
+                  ( (fold s = 0 then s + value)
+                  / (fold c = 0 then c + 1) )
+     in let ret = lookup lst map
+     in     ret
+\end{code}
+
+We can now convert to the query plan.
+The binding @lst@ becomes a single fold, initialised to @NO_DATE@ and updated with @key@.
+\begin{code}
+  plan kvs { key : Date; value : Real;      }
+  folds    { fL  : Date = NO_DATE then key  }
+  after    { lst : Date = fL                }
+\end{code}
+
+For the @map@ binding, each fold inside the group becomes its own map which is initialised to empty: fold @s@ becomes @gS@, and @c@ becomes @gC@.
 At each iteration the fold update is applied to the value in the map, or the fold's initial value if the key is not yet present.
@@
 \begin{code}
-==> plan kvs { key : Date; value : Real; }
+  folds  { gS  : Map Date Real = mapEmpty
+           then mapUpdate gS key 0 (\s. s + value)
 
-    before { }
+         ; gC  : Map Date Real = mapEmpty
+           then mapUpdate gC key 0 (\c. c + 1) }
 
-    folds  { groupS   : Map Date Real
-                = mapEmpty
-             then mapUpdate groupS key 0
-                  (\foldS. foldS + value)
-
-           ; groupC   : Map Date Real
-                = mapEmpty
-             then mapUpdate groupC key 0
-                  (\foldC. foldC + 1) }
-
-    after  { groupM   : Map Date Real
-                = mapMap (/) (mapZip groupS groupC) }
-
-    return { mean_all : Map Date Real = groupM }
+  after  { map : Map Date Real
+               = mapMap (/) (mapZip gS gC) }
 \end{code}
@@
 Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
-For example, a query @group key of count@ can reuse the already constructed @groupC@ map.
+For example, another query @group key of count@ can reuse the already constructed @gC@ map.
 
-If we now wish to fuse these query plans together, it is as simple as giving the bindings unique names, and appending each part of the plan together.
+Finally, the @ret@ binding contains no folds and is an @Aggregate@ type, so is converted to an @after@.
+The return associates the query's name with its final value.
+\begin{code}
+  after  { ret : Real = lookup lst map }
+  return { avg : Real = ret }
+\end{code}
+
+If we now wish to fuse these query plans together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
 The names of the returns and the column names stay the same.
 Query plans can only be fused if they act on the same table.
-
-The above query plan computes all three queries at once, but has some duplicate work: the counts and the sums are performed twice.
-We can perform common subexpression elimination to remove these.
-
 The interesting thing here is just how simple fusion is.
@@
 
-An efficient imperative loop can be generated from this query plan rather easily, in a similar way as previous work on flow fusion~\cite{lippmeier2013data}.
+An efficient imperative loop can be generated from this query plan in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
remove some stuff from benchmarks

@@
 
-We are currently using Icicle in production, over medium-sized datasets (around three hundred gigabytes uncompressed) that fit on a single disk.
-While future versions will require incremental computation, the runtime processing medium-sized datasets has been small enough to run over the entire dataset each time.
+At Ambiata, we are currently using Icicle in production over medium-sized datasets (around three hundred gigabytes uncompressed) that fit on a single disk.
 These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
-Incremental computation becomes more important for this distributed case, as it can potentially avoid copying terabytes of data over the network.
+Incremental computation is even more important for the distributed case, as it can potentially avoid copying terabytes of data over the network.
 
@@
 In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
-
-It is also important to note that, as a very constrained streaming language, memory usage of Icicle queries is more or less constant in the input size, and runtime is linear.
-This is very important, as new data is received regularly and the input size grows.
-This does not appear to be the case with the R script, which continually requires larger computers with more memory, in order to finish in a reasonable time.
-One of the benefits of Icicle is allowing data scientists to focus on queries, without having to worry about performance.
 

### 2016-06-02 2016fhpc/Main.tex
Change evaluator to take modality; update some benchmarks; hide conclusion
minor changes
benchmarks

@@
     $^\alpha$Ambiata      \\
-    Grow with data          \\[2pt]
     \textsf{amos.robinson@ambiata.com}
@@
     $^\beta$UNSW Australia\\
-    \\[2pt]
     ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
@@
     $^\gamma$Vertigo Technology \\
-    \\[2pt]
     ~~\textsf{benl@vergo.co}
@@
 We use a modal type system to ensure that fused queries can be computed in an incremental fashion, and a fold-based intermediate language to compile down to efficient C code.
-We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.
+We present production benchmarks demonstrating significant speedup over existing queries written in R, and on par with the widely used Unix tools \texttt{grep} and \texttt{wc}.
 
@@
 \input{section/04-Benchmarks.tex}
-\input{section/05-Conclusion.tex}
+% \input{section/05-Conclusion.tex}
 % \input{section/XX-Elimination.tex}
@@
 
+% \input{section/A1-ExampleDerivation.tex}
+
 \end{document}

### 2016-06-02 2016fhpc/section/01-Introduction.tex
benchmarks

@@
 We also want the mean of the open price for days in which the open price exceeded the close price.
-In Icicle we write the three sub queries as follows:
+In Icicle we write the three queries as follows:
 
@@
 In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered.
-We describe this construct more fully in \REF
+% We describe this construct more fully in \REF
 
 A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone.
-As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
+As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
 We instead need to lift each predicate to an expression-level conditional:
@@
 \item
-We present production benchmarks of Icicle compiled code which compare favourably to standard unix utilities, and outperform an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}.
+We present production benchmarks of Icicle compiled code which outperforms an existing R feature generation system by several orders of magnitude, and are on par with standard Unix utilities that conceptually perform less work\sref{s:Benchmarks}.
 \end{itemize}

### 2016-06-02 2016fhpc/section/03-Core.tex
benchmarks

@@
   after  { map : Map Date Real
-               = mapMap (/) (mapZip gS gC) }
+            = mapMap (\s c. s / c) (mapZip gS gC) }
 \end{code}
 
-It is possible to convert this as a map of pairs of the sum and count.
+It would be possible to convert this as a map of pairs of the sum and count.
 Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
@@
 
-If we now wish to fuse these query plans together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
+If we wish to fuse query plans on the same table together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
 The names of the returns and the column names stay the same.
-Query plans can only be fused if they act on the same table.
-The interesting thing here is just how simple fusion is.
-Thanks to the single-pass restriction, we know there are no dependencies between any two query plans: if there were, the source query would have required multiple passes.
-The single-pass restriction means that any source query can be expressed as a single fold over the data.
+It is the single-pass restriction that makes fusion so simple, as it implies there are no fusion-preventing dependencies between any two query plans: if there were, the source query would have required multiple passes.
 
-An efficient imperative loop can be generated from this query plan in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.
+Efficient imperative loops can be generated from these query plans in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
benchmarks

@@
 
-At Ambiata, we are currently using Icicle in production over medium-sized datasets (around three hundred gigabytes uncompressed) that fit on a single disk.
+At Ambiata, we are currently using Icicle in production over medium-sized datasets that fit on a single disk.
 These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
-Incremental computation is even more important for the distributed case, as it can potentially avoid copying terabytes of data over the network.
+Incremental computation is even more important for this distributed case, as we can avoid copying terabytes of data over the network.
 
@@
 
-The R script for this takes fifteen hours to run and is 3,566 lines, 2,311 of which are code.
+The R script for this takes fifteen hours to run and is 3,566 lines of code.
 In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
@@
 	ybar interval=0.7,
-  symbolic x coords={R, wc, CPU1, grep, wc -l, cat, Run 372, Run 1, end}
+  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end}
 ]
-\addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
+\addplot coordinates {(R,6.0) (wc,182.6) (1 CPU,186.6) (grep,261) (cat,878) (32 CPU,908) (end,0) };
+% \addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
 
@@
 
-\caption{Throughput comparisons of Icicle against existing R code and standard Unix utilities; higher is faster.}
+\caption{Throughput comparisons of Icicle (1 CPU and 32 CPU) against existing R code and standard Unix utilities; higher is faster.}
 \label{fig:bench:other}
@@
 
-\TODO{update to talk about single CPU number.
-Single CPU version 30 times faster than R, and in the same ballpark as @wc@ and @grep@.
-Because the memory usage is bounded, we have a better idea of how many we can run on a single machine without running out of memory and stepping on each other's toes.
-This means we can scale up to get quite close to the disk speed.
-This is not true of the R code: assuming perfect scaling, we would need at least 150 threads.
-Furthermore, the single-threaded R script is already using all the memory, and it is not obvious whether memory usage will decrease linearly as the input is divided and more threads are added.
-Blah~\cite{mcsherry2015scalability}
-}
-
 The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
@@
 \item our original R implementation (R);
-\item Icicle with equivalent queries (Run 372);
-\item Icicle with a single query computing the mean (Run 1);
+\item Icicle running single-threaded (1 CPU);
+\item Icicle running on multiple processors (32 CPU);
 \item finding empty lines with @grep "^$"@;
 \item counting characters, words and lines with @wc@;
-\item counting only lines with @wc -l@; and
 \item reading and throwing away the results with @cat > /dev/null@.
@@
 
-We ran all of the unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
+We ran all of the Unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
 The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
 The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
-Icicle significantly outperformed R, @grep@ and @wc@, and was roughly on par with @wc -l@ and @cat@.
-Although the computer had 32 CPUs, Icicle was the only benchmark we ran that used multiple CPUs: the other benchmarks were single threaded.
+Icicle significantly outperformed R, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
+This is despite doing conceptually more work than @wc@ and @grep@.
+By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
+The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
+These results give us confidence that our distributed implementation will be performant as well as scalable~\cite{mcsherry2015scalability}.
 
 Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
-We ran with a number of queries twice, with one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
-The graph shows the disk version decreasing roughly linearly in the number of queries, while ignoring the output stays constant.
+We ran two versions with each number of queries; one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
+The graph shows the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output stays constant.
 This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.

### 2016-06-02 2016fhpc/figures/Bench-All.tex
bench graphs

--- 2016-06-02 2016fhpc/section/04-Benchmarks.tex
@@
 %!TEX root = ../Main.tex
-\section{Benchmarks}
-\label{s:Benchmarks}
-
-At Ambiata, we are currently using Icicle in production over medium-sized datasets that fit on a single disk.
-These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
-Incremental computation is even more important for this distributed case, as we can avoid copying terabytes of data over the network.
-
-We evaluated Icicle by replacing an R script which ran weekly in production.
-This R script works over around three hundred gigabytes of PSV data.
-It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.
-
-The R script for this takes fifteen hours to run and is 3,566 lines of code.
-In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
 
@@
 \begin{axis}[
-	x tick label style={/pgf/number format/1000 sep=},
 	ylabel=MB/s,
+  ymin=0, ymax=1000,
 	enlargelimits=0.01,
-	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
 	ybar interval=0.7,
-  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end}
+  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end},
+  width=8.5cm, height=5cm,
 ]
@@
 \end{figure}
-
-\begin{figure}
-
-\begin{tikzpicture}
-\begin{axis}[
-	x tick label style={/pgf/number format/1000 sep=},
-	ylabel=MB/s,
-  ymin=0, ymax=1200,
-%  xmax=372,
-  xlabel=Number of queries,
-	enlargelimits=0.01,
-	legend style={legend columns=-1},
-%	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
-%	ybar interval=0.7,
-]
-
-% bench/raw.txt
-% Running with fewer queries
-% \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
-
-% bench/raw.txt
-% Running with fewer queries, better distribution
-\addplot coordinates { (0,1029) (31,1013) (62,1006) (93,998) (124,991) (155,941) (186,956) (217,933) (248,931) (279,915) (279,915) (310,919) (341,916) (372,908) };
-
-% Fusion, no output
-\addplot coordinates { (0,1001) (31,1038) (62,1025) (93,1031) (124,1006) (155,1015) (186,1020) (217,997) (248,1026) (279,1024) (310,1009) (341,1038) (372,1025) };
-
-% No fusion, no output
-% \addplot coordinates { (0,987) (31,1037) (62,1024) (93,1002) (124,1037) (155,1024) (186,1004) (217,1040) (248,1023) (279,998) (310,1018) (341,1013) (372,1003) };
-
-\legend{Output to disk, Output to /dev/null};
-
-\end{axis}
-\end{tikzpicture}
-
-\caption{Decrease in throughput as queries are added. By comparing writing the output to disk and writing to /dev/null, we see that actual computation time remains constant.}
-\label{fig:bench:queries}
-\end{figure}
-
-% \begin{figure}
-% \begin{tikzpicture}
-% \begin{axis}[
-% 	x tick label style={/pgf/number format/1000 sep=},
-% 	ylabel=MB/s,
-%   ymin=0, ymax=1200,
-%   xlabel=Number of threads,
-% 	enlargelimits=0.01,
-% 	legend style={legend columns=-1},
-% ]
-% \addplot coordinates { (1,163) (2,335) (3,442) (4,532) (5,578) (6,615) (7,633) (8,628) (9,631) (10,626) (11,637) (12,643) (13,658) (14,795) (15,781) (16,798) (17,840) (18,864) (19,862) (20,867) (21,847) (22,877) (23,876) (24,884) (25,862) };
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Scaling as threads are added}
-% \label{fig:bench:scaling}
-% \end{figure}
-
-The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
-We compared the throughput of several programs over the same 317GB dataset:
-\begin{itemize}
-\item our original R implementation (R);
-\item Icicle running single-threaded (1 CPU);
-\item Icicle running on multiple processors (32 CPU);
-\item finding empty lines with @grep "^$"@;
-\item counting characters, words and lines with @wc@;
-\item reading and throwing away the results with @cat > /dev/null@.
-\end{itemize}
-
-We ran all of the Unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
-The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
-The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
-Icicle significantly outperformed R, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
-This is despite doing conceptually more work than @wc@ and @grep@.
-By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
-The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
-These results give us confidence that our distributed implementation will be performant as well as scalable~\cite{mcsherry2015scalability}.
-
-Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
-We ran two versions with each number of queries; one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
-The graph shows the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output stays constant.
-This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
-The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled.

### 2016-06-02 2016fhpc/section/04-Benchmarks.tex
benchmarks
bench graphs

@@
 \label{s:Benchmarks}
+\input{figures/Bench-All.tex}
 
@@
 In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
-
-\begin{figure}
-
-\begin{tikzpicture}
-\begin{axis}[
-	x tick label style={/pgf/number format/1000 sep=},
-	ylabel=MB/s,
-	enlargelimits=0.01,
-	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
-	ybar interval=0.7,
-  symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end}
-]
-\addplot coordinates {(R,6.0) (wc,182.6) (1 CPU,186.6) (grep,261) (cat,878) (32 CPU,908) (end,0) };
-% \addplot coordinates {(R,6.0) (wc,182.6) (CPU1,186.6) (grep,261) (wc -l,852) (cat,878) (Run 372,908) (Run 1,1029) (end,0) };
-
-% \legend{MB/s}
-
-\end{axis}
-\end{tikzpicture}
-
-\caption{Throughput comparisons of Icicle (1 CPU and 32 CPU) against existing R code and standard Unix utilities; higher is faster.}
-\label{fig:bench:other}
-\end{figure}
-
-\begin{figure}
-
-\begin{tikzpicture}
-\begin{axis}[
-	x tick label style={/pgf/number format/1000 sep=},
-	ylabel=MB/s,
-  ymin=0, ymax=1200,
-%  xmax=372,
-  xlabel=Number of queries,
-	enlargelimits=0.01,
-	legend style={legend columns=-1},
-%	legend style={at={(0.5,-0.1)},anchor=north,legend columns=-1},
-%	ybar interval=0.7,
-]
-
-% bench/raw.txt
-% Running with fewer queries
-% \addplot coordinates {(0,1025.5) (1,1029.6) (48,978) (96,941) (144,943) (192,936) (240,904) (276,895.5) (324,881) (372,846.2) };
-
-% bench/raw.txt
-% Running with fewer queries, better distribution
-\addplot coordinates { (0,1029) (31,1013) (62,1006) (93,998) (124,991) (155,941) (186,956) (217,933) (248,931) (279,915) (279,915) (310,919) (341,916) (372,908) };
-
-% Fusion, no output
-\addplot coordinates { (0,1001) (31,1038) (62,1025) (93,1031) (124,1006) (155,1015) (186,1020) (217,997) (248,1026) (279,1024) (310,1009) (341,1038) (372,1025) };
-
-% No fusion, no output
-% \addplot coordinates { (0,987) (31,1037) (62,1024) (93,1002) (124,1037) (155,1024) (186,1004) (217,1040) (248,1023) (279,998) (310,1018) (341,1013) (372,1003) };
-
-\legend{Output to disk, Output to /dev/null};
-
-\end{axis}
-\end{tikzpicture}
-
-\caption{Decrease in throughput as queries are added. By comparing writing the output to disk and writing to /dev/null, we see that actual computation time remains constant.}
-\label{fig:bench:queries}
-\end{figure}
-
-% \begin{figure}
-% \begin{tikzpicture}
-% \begin{axis}[
-% 	x tick label style={/pgf/number format/1000 sep=},
-% 	ylabel=MB/s,
-%   ymin=0, ymax=1200,
-%   xlabel=Number of threads,
-% 	enlargelimits=0.01,
-% 	legend style={legend columns=-1},
-% ]
-% \addplot coordinates { (1,163) (2,335) (3,442) (4,532) (5,578) (6,615) (7,633) (8,628) (9,631) (10,626) (11,637) (12,643) (13,658) (14,795) (15,781) (16,798) (17,840) (18,864) (19,862) (20,867) (21,847) (22,877) (23,876) (24,884) (25,862) };
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Scaling as threads are added}
-% \label{fig:bench:scaling}
-% \end{figure}
 
@@
 
+\input{figures/Bench-Queries.tex}
+
 Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.

### 2016-06-03 2016fhpc/section/02-Source.tex
related

@@
 
-This function takes an input stream @xs@, and pairs the elements that are greater than zero with those that are less than zero.This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
+This function takes an input stream @xs@, and pairs the elements that are greater than zero with those that are less than zero.
+This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
 

### 2016-06-03 2016fhpc/section/04-Benchmarks.tex
related

@@
 The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
-These results give us confidence that our distributed implementation will be performant as well as scalable~\cite{mcsherry2015scalability}.
+These results give us confidence that our distributed implementation will be fast as well as scalable~\cite{mcsherry2015scalability}.
 
@@
 This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
-The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled.
+The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled if a better output format is used.

### 2016-06-03 2016fhpc/section/XX-Related.tex
related

@@
 %!TEX root = ../Main.tex
-\section{Related}
-\label{s:Related}
+\eject
+\section{Conclusion}
+\label{s:Conclusion}
+This paper shows Icicle, a streaming query language.
+The streaming, single-pass nature of Icicle allows all queries over the same table to be fused into a single loop over the data.
+Icicle's modal type system allows incremental computation, ensuring that queries give the same value regardless of how the input is sliced up, and when the increments are performed.
+The modal types encode when computations are available, and disallows using the results of folds before they are finished.
 
-The Icicle approach to bounded buffering is simpler but less expressive than both existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
-These systems perform a clock, or \emph{rate} analysis of the streams being processed.
-Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same rate --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
+In Icicle there is only one stream, the input table, which cannot be directly referred to or modified.
+This approach is simpler but less expressive than existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
+These systems use clock types and causal analysis to ensure programs can be executed with bounded buffers~\cite{stephens1997survey}.
+Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same clock --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
 
-\TODO{fill}
-Advances in dataflow programming languages~\cite{johnston2004advances}.
-Survey of stream processing~\cite{stephens1997survey}.
-How to architect a query compiler~\cite{shaikhha2016architect}.
-StreamIt~\cite{thies2002streamit}.
+StreamIt~\cite{thies2002streamit} is an imperative streaming language with explicit push and pop operations on streams, which has been extended with dynamic scheduling~\cite{soule2013dynamic}. 
+Dynamic scheduling is a trade-off: it is required for stream transformers such as groups and filters, but is slower than static scheduling.
+In Icicle we require that all groups and filters feed into a fold aggregate.
+This allows us to retain fully static scheduling while supporting these transformers, albeit in a limited way.
 
-Kahn process networks are a flexible alternative to MapReduce~\cite{vrba2009kahn}.
-Scheduling dynamic dataflow~\cite{buck1993scheduling}.
-Co-iterative characterization~\cite{caspi1998co}.
+It is well known that streaming dataflow languages correspond to Kahn process networks, and synchronous languages to restricted networks with bounded buffers~\cite{johnston2004advances}.
+Kahn process networks have also been touted as a more flexible and intuitive abstraction than MapReduce~\cite{vrba2009kahn}.
+Synchronous streaming dataflow languages such as Icicle and Lucy, then, should prove a useful abstraction for distributed computing.
 
-Compare to short-cut fusion~\cite{coutts2007stream}.
+% How to architect a query compiler~\cite{shaikhha2016architect}.
+% Scheduling dynamic dataflow~\cite{buck1993scheduling}.
+% Co-iterative characterization~\cite{caspi1998co}.
 
-Continuous queries are a similar domain, where a single query keeps running and producing results, as data is received.
-In \cite{munagala2007optimization} there are multiple queries based off the same incoming stream.
-Each queries is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
+Short-cut fusion techniques such as foldr/build~\cite{gill1993short} and stream fusion~\cite{coutts2007stream} rely on inlining to expose fusion opportunities.
+Because inlining relies on heuristics, there are no guarantees about when fusion will occur.
+This sort of fusion is useful for general programs where it is considered a ``bonus'' optimisation, but in our case we require absolute guarantees that fusion will occur.
+
+Continuous queries~\cite{chandrasekaran2003telegraphcq, arasu2003cql, babu2001continuous, madden2002continuously,shirsath2016query,munagala2007optimization} are a similar domain, where a single query keeps running and producing results, as data is received.
+In Munagala et al there are multiple queries based off the same incoming stream.
+Each query is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
 If these predicates are expensive to compute, it is certainly not ideal to compute these for each query.

### 2016-06-03 2016fhpc/section/XX-Related.tex
more related work

@@
 \label{s:Conclusion}
-This paper shows Icicle, a streaming query language.
+This paper has introduced Icicle, a streaming query language.
 The streaming, single-pass nature of Icicle allows all queries over the same table to be fused into a single loop over the data.
@@
 Because inlining relies on heuristics, there are no guarantees about when fusion will occur.
-This sort of fusion is useful for general programs where it is considered a ``bonus'' optimisation, but in our case we require absolute guarantees that fusion will occur.
+This sort of fusion is useful for general programs where it is considered a ``bonus'' optimisation, but insufficient when we require absolute guarantees that fusion will occur.
 
-Continuous queries~\cite{chandrasekaran2003telegraphcq, arasu2003cql, babu2001continuous, madden2002continuously,shirsath2016query,munagala2007optimization} are a similar domain, where a single query keeps running and producing results, as data is received.
-In Munagala et al there are multiple queries based off the same incoming stream.
+Continuous queries are very closely related to our incremental computation, where a single query keeps running and producing results as rows are added to the table.
+Continuous queries can be added and removed at any time, so any optimisations must be done dynamically.
+Continuous queries must also handle the case where rows are deleted and updated.
+In Icicle, the queries are statically known, and the data is immutable once it is received.
+
+In Munagala et al~\cite{munagala2007optimization}, the queries are filters with no grouping or aggregates.
 Each query is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
-If these predicates are expensive to compute, it is certainly not ideal to compute these for each query.
+A predicate index can be used to avoid recomputing the predicates, in case they are expensive to compute~\cite{madden2002continuously}.
+
+Continuous Query Language (CQL)~\cite{arasu2002abstract,stream2003stream} allows aggregates in its queries, but they must be builtin aggregate functions.
+It is not obvious how arbitrary aggregate functions could be supported while allowing deletion, except by recomputing the entire aggregate again.
 
@@
 By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
-However, this is only for a small set of built-in aggregates, and does not deal with mutually recursive definitions.
+However, this is again only for a small set of built-in aggregates.

### 2016-06-04 2016fhpc/section/01-Introduction.tex
ok

@@
 
-\TODO{Touch on related work}
+This work is related to stream processing languages such as Lucy~\cite{mandel2010lucy} and Streamit~\cite{thies2002streamit}, but without the need for clock or deadlock analyses.
+It is also related to continuous queries~\cite{arasu2003cql} where queries produce results as rows are inserted or deleted, except we can compute arbitrary folds and do not support deletion.
+
+Our implementation is available at \url{https://github.com/ambiata/icicle}.

### 2016-06-04 2016fhpc/section/04-Benchmarks.tex
ok

@@
 We evaluated Icicle by replacing an R script which ran weekly in production.
-This R script works over around three hundred gigabytes of PSV data.
+This R script works over around three hundred gigabytes of pipe-separated values (PSV) data.
 It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.
@@
 
+We have been able to achieve such good results by generating parsing code specialised to the schema, and using data-only flattening~\cite{bergstrom2013data} as an efficient representation of structures such as arrays of sum types and tuples.
+By using a query plan that is close to a functional language, we are able to apply many standard optimisations such as common subexpression elimination and dead code removal.
+
 \input{figures/Bench-Queries.tex}

### 2016-06-07 2016fhpc/section/01-Introduction.tex
Copy edits to intro

--- 2016-06-02 2016fhpc/section/01-Introduction.tex
@@
 
-At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over terabytes worth of compressed data.\footnote{In 2016 this was a lot of data.}
+At Ambiata we perform feature generation for machine learning applications by executing many thousands of simple queries over terabytes worth of compressed data.\footnote{In 2016 this was a lot of data.}
 For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input.
@@
 In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered.
-% We describe this construct more fully in \REF
 
-A joint query like this can be converted to a back-end language such as SQL, but doing so by hand is tedious and error prone.
+Such a joint query can be converted to a back-end language like SQL, but doing so by hand is tedious and error prone.
 As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
-We instead need to lift each predicate to an expression-level conditional:
+We must instead lift each predicate to an expression-level conditional:
 \begin{code}
@@
 \end{code}
-As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.
+As we see, the result of query fusion tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.
 
@@
 \end{code}
-Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an inceremental fashion.
+Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an incremental fashion.
 While streaming through the table we always have access to the last key in the stream, but to select rows that match this key would require that we stream through the table again from the start.
@@
 
-% on our smaller single-disk datasets
-% BEN: for streaming applications the absolute size of the input size doesn't affect the performance, so we don't need to apologise.
-
-\TODO{Touch on related work}
+Icicle is related to stream processing languages such as Lucy~\cite{mandel2010lucy} and Streamit~\cite{thies2002streamit}, except we forgo the need for clock and deadlock analysis.
+Icicle is also related to work on continuous queries~\cite{arasu2003cql}, where query results are updated as rows are inserted into the source table, except we can also compute arbitrary reductions and do not need to handle deleted source rows.
+Our implementation is available at \url{https://github.com/ambiata/icicle}.

### 2016-06-07 2016fhpc/Main.tex
related
more related work
Wibble to email formatting

@@
     $^\beta$UNSW Australia\\
-    ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
+    ~~~~\textsf{\{amosr,benl\}@cse.unsw.edu.au}
   }
@@
 \input{section/04-Benchmarks.tex}
-% \input{section/05-Conclusion.tex}
-% \input{section/XX-Elimination.tex}
 \input{section/XX-Related.tex}
@@
 \section*{Acknowledgements}
+We are grateful to the entire team at Ambiata; in particular Tran Ma and Jacob Stanley have been indispensable in compiler engineering and performance optimisations.
 
@@
 
-% \input{section/A1-ExampleDerivation.tex}
-
 \end{document}

### 2016-06-07 2016fhpc/section/01-Introduction.tex
More edits to intro

@@
 \item
-We present production benchmarks of Icicle compiled code which outperforms an existing R feature generation system by several orders of magnitude, and are on par with standard Unix utilities that conceptually perform less work\sref{s:Benchmarks}.
+We present production benchmarks of Icicle compiled code which outperforms an existing R feature generation system by several orders of magnitude\sref{s:Benchmarks}. 
 \end{itemize}
+
+% , and are on par with standard Unix utilities.
+% that conceptually perform less work.
+%\ben{I think this clause confuses the contribution.
+If if Icicle is "on par" with standard unix utilities, then why not just use those standard utilities?
+The point is that with Icicle we can also fuse multiple queries, and the fused code should be at least as good as the unix utils.
+However, fusing two instances of 'wc' won't be exactly 2x fast then running them sequentially, but still better than actually running them sequentially.
+There's no space to discuss the details in the intro.
 

### 2016-06-07 2016fhpc/section/02-Source.tex
Copy edits to S2

@@
 To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
-Of course, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
+Sadly, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
 However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
@@
 
-The problem with this is that the value of @last key@ is only availble once we have reached the end of the stream, but the @filter@ operation needs this value to process the very first element in the same stream.
+The problem is that the value of @last key@ is only available once we have reached the end of the stream, but @filter@ needs this value to process the very first element in the same stream.
 We distinguish between these two access patterns by giving them different names: we say that @last key@ is an \emph{aggregate}, because to compute it we must have consumed the \emph{entire stream}, whereas the filter predicate is an \emph{element}-wise computation because it only needs access to the current element in the stream.
 
-The trick to compute our average in an incremental fashion is to recognise that @filter@ is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information.
+The trick to compute our average in a streaming fashion is to recognize that @filter@ is selecting a particular subset of values from the input, but the value computed from this subset depends only on the values in that subset, and no other information.
 Instead of computing the mean of a single subset whose identity is only known at the end of the stream, we can instead compute the mean of \emph{all possible subsets}, and return the required one once we know what that is:
@@
 
-Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and compute running mean of the values of each group.
+Here we use the @group@ construct to assign key-value pairs to groups as we obtain them, and compute the running mean of the values of each group.
 The @avgs@ value becomes a map of group keys to their running means.
@@
 
-Note that the types of pure values such as constants are automatically promoted to the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.
+Note that the types of pure values such as constants are automatically promoted to the required modality.
+For example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.
 
@@
 
-We can also provide impure contextual information to the query such as the current date, by assigning it an aggregate type.
-As element-wise computations cannot depend on aggregate computations, we ensure that reused parts of an incremental computation are the same regardless of which day they are executed.
+We also provide impure contextual information to the query such as the current date, by assigning it an aggregate type.
+As element-wise computations cannot depend on aggregate computations we ensure that reused parts of an incremental computation are the same regardless of which day they are executed.
 
 % ---------------------------------------------------------
-\subsection{Bounded buffer restriction}
+\subsection{Bounded Buffer Restriction}
 \label{s:IcicleSource:bounded}
@@
 This function takes an input stream @xs@, and pairs the elements that are greater than zero with those that are less than zero.
-This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.
+This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all positive values must be buffered until we reach the negative ones, which allow output to be produced.
 
-Icicle is designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
-In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly.
+In Icicle, queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
+In Icicle, the stream being processed (such as @xs@ above) is implicit in each query.
 Constructs such as @filter@ and @fold@ do not take the name of the stream as an argument, but instead operate on the stream defined in the context.
-In the example Icicle queries presented so far, the name of the table (stream) is not mentioned in the body of the query, and serves as a comment only.
 Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query.
-In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.
+In the expression @(filter p of mean value)@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.
 
-Finally, note that our major caveat is that the @group@ construct we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
+Finally, our major caveat is that the @group@ construct we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
 For our applications the keys are commonly company names, customer names, and days of the year.

### 2016-06-07 2016fhpc/section/02-Source.tex
Wibbles on S3.1

@@
 The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
-Value types $T$ include numbers, booleans or maps.
+Value types $T$ include numbers, booleans and maps.
 Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
-Function types $\TauFun$ include non-function modal types, or a function from modal type to function type.
+Function types $\TauFun$ include non-function modal types, and functions from modal type to function type.
 As Icicle is a first-order language, function types are not value types.
@@
 Table definitions $\mi{Table}$ define a table name and the names and types of columns.
-Expressions $\mi{Exp}$ can be variable names, constants, applications of primitives or functions.
+Expressions $\mi{Exp}$ include variable names, constants, applications of primitives and functions.
 The @fold@ construct defines the name of an accumulator, the expression for the initial value, and the expression used to update the accumulator for each element of the stream.
@@
 % ---------------------------------------------------------
-\subsection{Type system}
+\subsection{Type System}
 The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
-The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ associates an expression $\mi{Exp}$ with its type $M$ under context $\Gamma$.
-The judgment form $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
-The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
-The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
-Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.
+The judgment form $\Typecheck{\Gamma}{\mi{e}}{\TauMode}$ associates an expression $\mi{e}$ with its type $M$ under context $\Gamma$.
+The judgment form $\TypecheckP{\mi{p}}{\TauFun}$ associates a primitive with its function type.
+The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used to lift function application to modal types: a function type applied to a list of modal argument types produces a result type and matching mode.
+The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and produces an environment containing the function or query name and its type.
+Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing only the types of the queries.
 
@@
 Rule TcVar performs variable lookup in the context.
-Rule TcBox assigns a pure expression either @Element@ or @Aggregate@ type.
+Rule TcBox assigns an expression either @Element@ or @Aggregate@ type.
 
-% shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
-% The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.
-
-Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments. Rule TcLet is standard.
-
-% find the result type of primitive and function application.
-% They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
-% The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
-% Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.
+Rules TcPrimApp and TcFunApp produce the type a primitive or function applied to its arguments. Rule TcLet is standard.
 
@@
 
-% The initial value of the fold must be pure.
-% The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
-% The final return type of the fold is an @Aggregate@.
-
-% Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
-% The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
-% Rule TcGroup is similar to filter, except returning a @Map@.
-
-The rules PrimArith, PrimRel, PrimLookup, PrimFst and PrimSnd assign types to primitives.
+Rules PrimArith, PrimRel, PrimLookup, PrimFst and \mbox{PrimSnd} assign types to primitives.
 Rule AppArgs produces the type of a function or primitive applied to its arguments.
-Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.
+Rule AppRebox is used when the arguments have modal type $m$ --- applying a function to arguments of mode $m$ produces a result of the same mode.
 
@@
 
-% The next judgment shows how function application boxes and unboxes modalities. This judgment takes a function type and all its arguments, and returns the result type of the function.
-
-% Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.
-
-% Rule AppRebox is used when a pure function is applied to a non-pure value.
-% If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
-% Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
-% These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.
-
-% inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.
-
-% Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.
-
-% It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
-% Then, each query is checked, with the table columns and all functions available.
-% The final returned context is all query names and their types.
-
 % ---------------------------------------------------------
 \subsection{Evaluation}
-The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
+Evaluation rules for Icicle are given in Figure~\ref{fig:source:eval}.
 Grammar $N$ defines the modes of evaluation, including pure computation.

### 2016-06-08 2016fhpc/section/02-Source.tex
Work on description of evaluation rules

@@
 Grammar $\Sigma$ defines a heap containing stream values.
-Grammar $V'$ defines a different type of stream value for each evaluation mode:
+Grammar $V'$ defines the results that can be produced by evaluation, depending on the mode:
 \begin{itemize}
 \item
-@Pure@ computations are represented as a single value;
+@Pure@ computation results are a single value;
 \item
-@Element@ stream transformers are represented as a function applied to each element; and
+@Element@ computation results are stream transformers, which are represented by meta functions that take a value and produces a new value; and
 \item
-@Aggregate@ folds are represented as an initial (zero) state, an update (konstrukt) function applied to each stream element and current state, and an eject function applied to the final state.
+@Aggregate@ computation results consist of an initial (zero) state, an update (konstrukt) meta function to be applied to each stream element and current state, and an eject meta function to be applied to the final state.
 \end{itemize}
 
-The judgment form $\SourceStepX{N}{\Sigma}{e}{V'}$ defines a big-step evaluation relation: under evaluation mode $N$ with heap $\Sigma$, expression $e$ evaluates to stream value $V'$.
-The evaluation mode is used to know when pure values should be converted to streams or folds.
+In the grammar $V'$ we write $\stackrel{\bullet}{\to}$ to highlight that the objects in those positions are meta-functions, rather than abstract syntax.
+To actually process data from the input table we will need to apply the produced meta-functions to this data.
+
+The judgment form $\SourceStepX{N}{\Sigma}{e}{V'}$ defines a big-step evaluation relation: under evaluation mode $N$ with heap $\Sigma$, expression $e$ evaluates to result $V'$.
+The evaluation mode $N$ controls whether pure values should be promoted to element (stream) or aggregate (fold) results. 
 We assume that all functions have been inlined into the expression before evaluation.
 
-Rule EVal applies when the expression is already a stream value.
+Rule EVal applies when the expression is already a completed result.
 Rule EVar performs variable lookup in the heap.
-Rule ELet uses the bound expression's type to find the evaluation mode, then evaluates the bound expression under that mode.
-The bound expression's value is added to the heap, and the rest of the expression is evaluated in the original evaluation mode.
+Rule ELet evaluates the bound expression under the given mode.
 
-Rules EBoxStream and EBoxFold convert a pure value to a stream and fold respectively, when the evaluation mode is not pure.
-Conversion to a stream transformer ignores the input stream and returns the pure value.
-Conversion to a fold has no state, and the eject function returns the pure value.
+% uses the bound expression's type to find the evaluation mode, then evaluates the bound expression under that mode.
 
-Rules EPrimValue, EPrimStream and EPrimFold evaluate primitive applications.
-For pure values, the values are unboxed and the primitive is applied.
-For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
-For folds, the states of all the argument folds are joined together as tuples.
-At each step, the states of the input folds are updated using their own update function.
-The eject function extracts the ejected values of the input folds, and applies the primitive.
+% The bound expression's value is added to the heap, and the rest of the expression is evaluated in the original evaluation mode.
 
-Rule EFold introduces recursion and memory to the streams.
-It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
-The update expression has mode @Element@, so it must be evaluated to a stream.
-This stream expects the stream element as an argument, but we need to give it the current fold state as well.
-In order to pass both as a pair, we modify all streams in the heap to use the second half of the pair.
-We also add a stream for the fold state to the heap; this uses the first half of the pair.
-The update function then calls the stream transformer with a pair of the state and the input element.
-The eject function is the identity function.
+Rules EBoxStream and EBoxFold lift constant values to stream results and aggregate results respectively.
+To lift a constant to a stream result we produce a meta-function that always returns the value.
+To lift a constant to an aggregate result we set the update meta-function to return a dummy, and have the eject meta-function return the value of interest.
 
-Rule EFilter evaluates its predicate to a stream of booleans, and the rest to a fold.
-The result is a fold which only applies the update function when the predicate stream is true.
+%   a pure value to a stream and fold respectively, when the evaluation mode is not pure.
+% Conversion to a stream transformer ignores the input stream and returns the pure value.
+% Conversion to a fold has no state, and the eject function returns the pure value.
 
-Rule EGroup is rather similar to filter, except that it returns a map.
-Here, we start with a map of every key to the fold initialiser.
-At every update step, we find the key and look up the current accumulator for that key in the map.
-We then apply the update function to the accumulator, and update that key in the map.
-For the eject function, we simply apply the eject to each element of the map.
+Rules EPrimValue, EPrimStream and EPrimFold apply primitive operators to constant values, streams and aggregations respectively.
+In EPrimStream the result is a new stream transformer that applies the primitive to each of the elements gained from the input streams.
+In EPrimFold the result consists of new update and eject functions that get their input values by applying the update and eject functions gained by evaluating the arguments.
+
+% For pure values, the values are unboxed and the primitive is applied.
+% For streams, a new stream is created and the input heap applied to all streams before applying the primitive.
+% For folds, the states of all the argument folds are joined together as tuples.
+% At each step, the states of the input folds are updated using their own update function.
+% The eject function extracts the ejected values of the input folds, and applies the primitive.
+
+Rule EFilter first evaluates the predicate $e$ to a stream transformer $f$, and the body $e'$ to an aggregation.
+The result is a new aggregation where the update function applies the predicate stream transformer $f$ to the input element $s$ to yield a boolean flag which specifies whether the current aggregation state should be updated.
+
+Rule EGroup is similar to EFilter, except that the stream transformer $f$ produces group keys rather than boolean flags, and we maintain a finite map of aggregation states for each key.
+In the result aggregation the update function updates the appropriate element in the map, and the eject function is applied to every accumulator.
+
+% Rule EFold introduces recursion and memory to the streams.
+% It evaluates its initial to a pure value, $z'$, which is used as the initial state for the fold.
+% The update expression has mode @Element@, so it must be evaluated to a stream.
+% This stream expects the stream element as an argument, but we need to give it the current fold state as well.
+% In order to pass both as a pair, we modify all streams in the heap to use the second half of the pair.
+% We also add a stream for the fold state to the heap; this uses the first half of the pair.
+% The update function then calls the stream transformer with a pair of the state and the input element.
+% The eject function is the identity function.
+
+Rule EFold introduces a new accumulator which is visible in the context of the body $k$.
+Evaluating the body $k$ produces a body stream transformer $k'$ whose job is to update this new accumulator each time it is applied.
+In the conclusion of EFold we pass this stream transformer a tuple $(v, s)$ where $v$ is the new accumulator and $s$ is the current element of the stream we get from the context of the overall @fold@ expression.
+The heap used when evaluating $k$ is updated so that references to either the stream elements or new accumulator access the appropriate side of the tuple.
 
@@
 The input table $t$ is a map from column name to a list of all the values for that column.
-Rule ETable creates an initial heap where each column is a stream transformer that pulls out the $n$th element of a nested tuple.
-Tuples are constructed correspondingly, so that each row becomes a nested tuple where the $n$th element contains the value of the $n$th column.
+Rule ETable creates an initial heap where each column name $x_i$ is bound to an expression which projects out the appropriate element from a single row in the input table.
+Evaluating the expression $e$ produces an aggregation result where the update function $k$ accepts each row from the table and updates all the accumulators defined by $e$.
+The actual computation is driven by the $\mi{fold}$ meta-function.
+
+% column is a stream transformer that pulls out the $n$th element of a nested tuple.
+% Tuples are constructed correspondingly, so that each row becomes a nested tuple where the $n$th element contains the value of the $n$th column.

### 2016-06-08 2016fhpc/section/03-Core.tex
Copy edits to intermediate language

@@
 
-The intermediate language can be thought of as similar to a query plan in other databases.
-Queries in the source language are converted to query plans, and then query plans on the same table are fused together.
-Common subexpression elimination, partial evaluation and other optimisations are performed at the query plan level.
+The Icicle intermediate language is similar to a physical query plan for a database system.
+We convert each source level query to a query plan, then fuse together the plans for queries on the same table.
+Once we have the fused query plan we then perform standard optimsiations such as common subexpression elimination and partial evaluation.
 
-The grammar for the intermediate language is given in figure~\ref{fig:core:grammar}.
-Expressions can be variables, values, applications of primitives or anonymous functions.
-Function definitions and uses are not allowed in expressions here, as their definitions are inlined before converting to query plans.
-Anonymous functions are only allowed as arguments to primitives: they cannot be applied or stored in variables.
-The $\mi{Plan}$ itself is split into stages of computation, which correspond to the modal stages in the Source query.
-The first stage, @before@, is used for pure computations which do not depend on any stream values at all.
-The next stage, @folds@, defines the @Element@ computations and how they are converted into @Aggregate@ computations.
-Next are @after@ which correspond to @Aggregate@ computations, occurring after the entire stream has been seen.
-Finally, the @return@ specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.
+The grammar for the Icicle intermediate language is given in Figure~\ref{fig:core:grammar}.
+Expressions $PlanX$ include variables, values, applications of primitives and anonymous functions.
+% Function definitions and uses are not allowed in expressions here, as their definitions are inlined before converting to query plans.
+% Anonymous functions are only allowed as arguments to primitives: they cannot be applied or stored in variables.
+The $\mi{Plan}$ itself is split into a five stage \emph{loop anatomy}.
+First we have the name of the table and the names and element types of each column.
+The @before@ stage then defines pure values which do not depend on any table data.
+The @folds@ stage defines element computations and how they are converted to aggregate results.
+The @after@ stage defines aggregate computations that combine multiple aggregations after the entire table has been processed.
+Finally, the @return@ stage specifies the output values of the query; a single query will have only one output value, but the result of fusion can have many outputs.
 
-Before we can convert our example query to a query plan, we must look at the definitions of the functions @count@, @sum@, @mean@ and @last@.
-Here, we have definitions of @count@ and @sum@.
-Count is defined by starting from zero, and incrementing by one for each element.
-Sum starts at zero and increments its argument.
-
+Before we discuss an example query plan we first define the @count@, @sum@, @mean@ and @last@ functions used in earlier sections. Both @count@ and @sum@ are simple folds:
 \begin{code}
@@
 \end{code}
-
-Mean can then defined as sum divided by the count.
+The @mean@ function then divides the @sum@ by the @count@.
 \begin{code}
@@
 
-Here, we have the function definition for @last k@, which is defined as a fold with an initial value to describe a missing date @NO_DATE@\footnote{In our production compiler, @last@ returns a @Maybe@.}, and the update value is the current @k@.
-The current fold value is bound to @l@ but is not used.
-
+The @last@ function uses a @fold@ that initializes the accumulator to the empty date value @NO_DATE@\footnote{In our production compiler, @last@ returns a @Maybe@.}, then updates it with the date gained from the current element in the stream.
 \begin{code}
@@
 
-We now inline these function definitions before converting to a query plan, renaming as necessary.
-This is guaranteed to terminate because no recursion is allowed in function definitions.
+Inlining the above functions into the query from \label{s:ElementsAndAggregates} yields the following:
+% This is guaranteed to terminate because no recursion is allowed in function definitions.
 \begin{code}
@@
 
-We can now convert to the query plan.
-The binding @lst@ becomes a single fold, initialised to @NO_DATE@ and updated with @key@.
+To convert this source query to a plan in the intermediate language we convert each of the let-bindings separately then simply concatenate the corresponding parts of the loop anatomy.
+The @lst@ binding becomes a single fold, initialized to @NO_DATE@ and updated with the current @key@.
 \begin{code}
@@
 
-For the @map@ binding, each fold inside the group becomes its own map which is initialised to empty: fold @s@ becomes @gS@, and @c@ becomes @gC@.
-At each iteration the fold update is applied to the value in the map, or the fold's initial value if the key is not yet present.
-After all iterations are finished, the two maps are joined together and the pairs are divided.
-\begin{code}
+For the @map@ binding, each fold accumulator inside the body of the @group@ construct is associate with its own finite map.
+The @s@ accumulator is associated with map @gS@, and the @c@ accumulator with @gC@.
+Each time we receive a row from the table the accumulator associated with the @key@ is updated, using the default value @0@ if an entry for that key is not yet present.
+After we have processed the entire table each sum with its corresponding count to yield a map of means for each key.
+\begin{alltt}
   folds  { gS  : Map Date Real = mapEmpty
-           then mapUpdate gS key 0 (\s. s + value)
+           then mapUpdate gS key 0 (\(\lambda\)s. s + value)
 
          ; gC  : Map Date Real = mapEmpty
-           then mapUpdate gC key 0 (\c. c + 1) }
+           then mapUpdate gC key 0 (\(\lambda\)c. c + 1) }
 
   after  { map : Map Date Real
-            = mapMap (\s c. s / c) (mapZip gS gC) }
-\end{code}
+            = mapMap (\(\lambda\)s c. s / c) (mapZip gS gC) }
+\end{alltt}
 
-It would be possible to convert this as a map of pairs of the sum and count.
-Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
-For example, another query @group key of count@ can reuse the already constructed @gC@ map.
+% It would be possible to convert this as a map of pairs of the sum and count.
+% Keeping this as two separate maps rather than a map of pairs exposes more opportunities for common subexpression elimination when fusion occurs.
+% For example, another query @group key of count@ can reuse the already constructed @gC@ map.
 
-Finally, the @ret@ binding contains no folds and is an @Aggregate@ type, so is converted to an @after@.
-The return associates the query's name with its final value.
+Finally, the @ret@ binding from the original query is evaluated in the @after@ stage. In the @return@ stage we specify that the result of the overall query @avg@ is the result of the @ret@ binding.
 \begin{code}
@@
 
-If we wish to fuse query plans on the same table together, it is as simple as giving the @before@, @folds@ and @after@ bindings unique names and appending each part of the plan together.
-The names of the returns and the column names stay the same.
-It is the single-pass restriction that makes fusion so simple, as it implies there are no fusion-preventing dependencies between any two query plans: if there were, the source query would have required multiple passes.
+To combine the plans from each binding we simply concatenate the corresponding parts of the anatomy.
+To fuse multiple plans we freshen the names of each binding and also concatenate the corresponding parts of the anatomy.
+The single-pass restriction on queries makes the fusion process so simple.
+The single-pass restriction implies that there are no fusion-preventing dependencies between any two query plans.
 
-Efficient imperative loops can be generated from these query plans in a similar way to previous work on flow fusion~\cite{lippmeier2013data}.
+Given a fused query plan we the convert it to an imperative loop nest in a similar way to our prior work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-08 2016fhpc/section/04-Benchmarks.tex
Copy edits to benchmarks section

--- 2016-06-03 2016fhpc/section/04-Benchmarks.tex
@@
 
-At Ambiata, we are currently using Icicle in production over medium-sized datasets that fit on a single disk.
-These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
-Incremental computation is even more important for this distributed case, as we can avoid copying terabytes of data over the network.
+At Ambiata we are currently using Icicle in production over medium-sized data sets that fit on a single disk.
+We are also currenly implementing a scheduler to distribute larger data sets across multiple nodes.
+The data we are working with is tens of terabytes compressed, which in 2016 does not fit on a single disk.
+However, each row has a natural primary key and the features we need to compute depend only on the data within single key groups, which makes the workload very easy to distribute.
 
-We evaluated Icicle by replacing an R script which ran weekly in production.
-This R script works over around three hundred gigabytes of PSV data.
-It computes twelve queries over each of the thirty-one input tables, computing 372 queries in total.
+% These initial results have been very promising, and we are currently implementing distribution across multiple machines to handle datasets that are tens of terabytes compressed, and do not fit on a single disk.
+% Incremental computation is even more important for this distributed case, as we can avoid copying terabytes of data over the network.
 
-The R script for this takes fifteen hours to run and is 3,566 lines of code.
-In contrast, our Icicle queries take seven minutes to run, and the dictionary describing the queries is 191 lines of code.
+In our proof of concept tesing we replaced an existing R script that performed feature generation with new Icicle code.
+The R script computed features from a 317GB data set.
+It computed 12 queries over each of 31 input tables, for 372 query evaluations in total.
+The R script took 15 hours to run and consisted of 3,566 lines of code.
+The replacement Icicle version is only 191 lines of code and takes seven minutes to run.
 
-The table in figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
-We compared the throughput of several programs over the same 317GB dataset:
+The table in Figure~\ref{fig:bench:other} shows the throughput in megabytes per second.
+We compared the throughput of several programs over the same dataset:
 \begin{itemize}
@@
 
-We ran all of the Unix utilities with unicode disabled using @LANG=C LC_COLLATE@ for maximum performance.
-The computer we used for most of these was an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
-The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
+We ran all the Unix utilities with unicode decoding disabled using @LANG=C LC_COLLATE@ for maximum performance. The input data does not contain unicode characters. 
+We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
 Icicle significantly outperformed R, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
-This is despite doing conceptually more work than @wc@ and @grep@.
+This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@.
 By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
@@
 
+% The R code requires an @i2.8xlarge@ with 244GB of memory, but we were unable to perform our other benchmarks on such a large machine.
+% \BEN{I wouldn't worry about stating this. You can't give out your R code so no one will test it themselves, and if even if you could run it on the smaller machine it wouldn't be faster}
+
+% We were able to achieve such good results by generating parsing code specialised to the schema, and using data-only flattening~\cite{bergstrom2013data} as an efficient representation of structures such as arrays of sum types and tuples.
+% \BEN{This isn't part of the contribution of this paper, so there's no real reason to discuss it}.
+
+% By using a query plan that is close to a functional language, we are able to apply many standard optimisations such as common subexpression elimination and dead code removal.
+% \BEN{We already said this before}.
+
 \input{figures/Bench-Queries.tex}
 
-Figure~\ref{fig:bench:queries} shows how Icicle scales in throughput as more queries are added.
-We ran two versions with each number of queries; one version writing output to disk, and the other throwing away the result to @/dev/null@ rather than spending time writing.
-The graph shows the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output stays constant.
-This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
-The upshot of this is that the time spent computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled if a better output format is used.
+Figure~\ref{fig:bench:queries} shows how the total througput scales as the number of fused queries is increased.
+For each number of queries, we ran two versions of the fused result: one version that wrote the output to disk, and the other that piped the result to @/dev/null@.
+The graph shows the throughput of the disk version decreasing roughly linearly in the number of queries, while the version ignoring the output remains constant.
+This suggests that we are IO bound on the write side.
+The time spent evaluating the queries themselves is small relative to our current IO load, and we are curently scaling up our system to use a larger query set.
+
+% This suggests that the output code is the bottleneck, which is unsurprising given the output format is text-based PSV.
+% computing the queries appears constant as hundreds of queries are added, which suggests that many more queries can be handled if a better output format is used.
+% \BEN{I don't understand this, surely the program would produce the PSV even if it was just sent to /dev/null?}

### 2016-06-08 2016fhpc/section/04-Benchmarks.tex
Work on related work section

@@
 At Ambiata we are currently using Icicle in production over medium-sized data sets that fit on a single disk.
-We are also currenly implementing a scheduler to distribute larger data sets across multiple nodes.
+We are also currently implementing a scheduler to distribute larger data sets across multiple nodes.
 The data we are working with is tens of terabytes compressed, which in 2016 does not fit on a single disk.
@@
 We used an Amazon EC2 @c3.8xlarge@ with 32 CPUs, 60GB of RAM, and striped, RAIDed SSD storage.
-Icicle significantly outperformed R, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
+The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
 This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@.

### 2016-06-08 2016fhpc/section/XX-Related.tex
Work on related work section

@@
 \eject
-\section{Conclusion}
+\section{Related Work}
 \label{s:Conclusion}
-This paper has introduced Icicle, a streaming query language.
-The streaming, single-pass nature of Icicle allows all queries over the same table to be fused into a single loop over the data.
-Icicle's modal type system allows incremental computation, ensuring that queries give the same value regardless of how the input is sliced up, and when the increments are performed.
-The modal types encode when computations are available, and disallows using the results of folds before they are finished.
 
-In Icicle there is only one stream, the input table, which cannot be directly referred to or modified.
-This approach is simpler but less expressive than existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
-These systems use clock types and causal analysis to ensure programs can be executed with bounded buffers~\cite{stephens1997survey}.
-Stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same clock --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
+% This paper has introduced Icicle, a streaming query language. The streaming, single-pass nature of Icicle allows all queries over the same table to be fused into a single loop over the data.
+% Icicle's modal type system allows incremental computation, ensuring that queries give the same value regardless of how the input is sliced up, and when the increments are performed.
+% The modal types encode when computations are available, and disallows using the results of folds before they are finished.
+% \ben{the above paragraph doesn't add any new information}.
+
+In Icicle there is only one stream, the input table, which is implicit in the body of the queries.
+This approach is intentionally simpler than existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
+These latter systems use clock types and causal analysis to ensure that programs can be executed with bounded buffers~\cite{stephens1997survey}.
+Synchronous data flow languages implement Kahn networks that are restricted to use bounded buffering~\cite{johnston2004advances}.
+In such languages, stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same clock --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
+In Icicle the fact that the input stream is implicit and distributed to all combinators means that we can forgo clock analysis.
+All queries in a program execute in lock-step on the same element at the same moment, which ensures that fusion is a simple matter of concatenating the components of the loop anatomy of each query.
 
@@
 
-It is well known that streaming dataflow languages correspond to Kahn process networks, and synchronous languages to restricted networks with bounded buffers~\cite{johnston2004advances}.
-Kahn process networks have also been touted as a more flexible and intuitive abstraction than MapReduce~\cite{vrba2009kahn}.
-Synchronous streaming dataflow languages such as Icicle and Lucy, then, should prove a useful abstraction for distributed computing.
+% It is well known that streaming dataflow languages correspond to Kahn process networks.
+
+% Kahn process networks have also been touted as a more flexible and intuitive abstraction than MapReduce~\cite{vrba2009kahn}.
+% This doesn't say anything about Icicle.
+
+% Synchronous streaming dataflow languages such as Icicle and Lucy, then, should prove a useful abstraction for distributed computing.
 

### 2016-06-09 2016fhpc/section/XX-Related.tex
More on dynamic scheduling

@@
 
-In Icicle there is only one stream, the input table, which is implicit in the body of the queries.
+In Icicle there is only one stream, sourced from the input table, which is implicit in the bodies of queries.
 This approach is intentionally simpler than existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
-These latter systems use clock types and causal analysis to ensure that programs can be executed with bounded buffers~\cite{stephens1997survey}.
-Synchronous data flow languages implement Kahn networks that are restricted to use bounded buffering~\cite{johnston2004advances}.
+Synchronous data flow languages implement Kahn networks that are restricted to use bounded buffering~\cite{johnston2004advances} by clock typing and causal analysis~\cite{stephens1997survey}.
 In such languages, stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same clock --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
@@
 
-StreamIt~\cite{thies2002streamit} is an imperative streaming language with explicit push and pop operations on streams, which has been extended with dynamic scheduling~\cite{soule2013dynamic}. 
-Dynamic scheduling is a trade-off: it is required for stream transformers such as groups and filters, but is slower than static scheduling.
-In Icicle we require that all groups and filters feed into a fold aggregate.
-This allows us to retain fully static scheduling while supporting these transformers, albeit in a limited way.
+Short-cut fusion techniques such as foldr/build~\cite{gill1993short} and stream fusion~\cite{coutts2007stream} rely on inlining to expose fusion opportunities.
+In Haskell compilers such as GHC, the decision of when to inline is made by internal compiler heuristics, which makes it difficult for the programmer to predict when fusion will occur.
+In this environment, array fusion is considered a ``bonus'' optimization rather than integral part of the compilation method.
+In contrast, for our feature generation application we really must ensure that multiple queries over the same table are fused, so we cannot rely on heuristics.
+
+StreamIt~\cite{thies2002streamit} is an imperative streaming language which has been extended with dynamic scheduling~\cite{soule2013dynamic}.
+Dynamic scheduling handles data flow graphs where the transfer rate between different stream operators is not known at compile time.
+Dynamic scheduling is trade-off: it is required for stream operators such as grouping and filtering where the output data rate is not known statically, but using dynamic techniques for graphs with static transfer rates tends to have a performance cost.
+Icicle includes grouping and filtering operators where the output rates are statically unknown, however the associated language constructs require grouped and filtered data to be aggregated rather than passed as as the input to another stream operator.
+This allows Icicle to retain fully static scheduling, so the compiled queries consist of straight line code with no buffering.
+
+Continuous queries are very closely related to our incremental computation, where a single query keeps running and producing results as rows are added to the table.
+Continuous queries can be added and removed at any time, so any optimisations must be done dynamically.
+Continuous queries must also handle the case where rows are deleted and updated.
+In Icicle, the queries are statically known, and the data is immutable once it is received.
+
+Continuous Query Language (CQL)~\cite{arasu2002abstract,stream2003stream} allows aggregates in its queries, but they must be builtin aggregate functions.
+It is not obvious how arbitrary aggregate functions could be supported while allowing deletion, except by recomputing the entire aggregate again.
+
+In Munagala et al~\cite{munagala2007optimization}, the queries are filters with no grouping or aggregates.
+Each query is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
+A predicate index can be used to avoid recomputing the predicates, in case they are expensive to compute~\cite{madden2002continuously}.
+
+Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
+By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
+However, this is again only for a small set of built-in aggregates.
 
@@
 % Kahn process networks have also been touted as a more flexible and intuitive abstraction than MapReduce~\cite{vrba2009kahn}.
-% This doesn't say anything about Icicle.
+% \ben{This doesn't say anything about Icicle.}
 
@@
 % Co-iterative characterization~\cite{caspi1998co}.
-
-Short-cut fusion techniques such as foldr/build~\cite{gill1993short} and stream fusion~\cite{coutts2007stream} rely on inlining to expose fusion opportunities.
-Because inlining relies on heuristics, there are no guarantees about when fusion will occur.
-This sort of fusion is useful for general programs where it is considered a ``bonus'' optimisation, but insufficient when we require absolute guarantees that fusion will occur.
-
-Continuous queries are very closely related to our incremental computation, where a single query keeps running and producing results as rows are added to the table.
-Continuous queries can be added and removed at any time, so any optimisations must be done dynamically.
-Continuous queries must also handle the case where rows are deleted and updated.
-In Icicle, the queries are statically known, and the data is immutable once it is received.
-
-In Munagala et al~\cite{munagala2007optimization}, the queries are filters with no grouping or aggregates.
-Each query is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
-A predicate index can be used to avoid recomputing the predicates, in case they are expensive to compute~\cite{madden2002continuously}.
-
-Continuous Query Language (CQL)~\cite{arasu2002abstract,stream2003stream} allows aggregates in its queries, but they must be builtin aggregate functions.
-It is not obvious how arbitrary aggregate functions could be supported while allowing deletion, except by recomputing the entire aggregate again.
-
-Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
-By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
-However, this is again only for a small set of built-in aggregates.

### 2016-06-09 2016fhpc/section/XX-Related.tex
Copy edits to related

@@
 This approach is intentionally simpler than existing synchronous data flow languages such as Lucy~\cite{mandel2010lucy}, as well as our prior work on flow fusion~\cite{lippmeier2013data}.
-Synchronous data flow languages implement Kahn networks that are restricted to use bounded buffering~\cite{johnston2004advances} by clock typing and causal analysis~\cite{stephens1997survey}.
+Synchronous data flow languages implement Kahn networks~\cite{vrba2009kahn} that are restricted to use bounded buffering~\cite{johnston2004advances} by clock typing and causal analysis~\cite{stephens1997survey}.
 In such languages, stream combinators with multiple inputs, such as @zip@, are assigned types that require their stream arguments to have the same clock --- meaning that elements always arrive in lockstep and the combinators themselves do not need to perform their own buffering.
@@
 
-Continuous queries are very closely related to our incremental computation, where a single query keeps running and producing results as rows are added to the table.
-Continuous queries can be added and removed at any time, so any optimisations must be done dynamically.
-Continuous queries must also handle the case where rows are deleted and updated.
-In Icicle, the queries are statically known, and the data is immutable once it is received.
-
-Continuous Query Language (CQL)~\cite{arasu2002abstract,stream2003stream} allows aggregates in its queries, but they must be builtin aggregate functions.
-It is not obvious how arbitrary aggregate functions could be supported while allowing deletion, except by recomputing the entire aggregate again.
-
-In Munagala et al~\cite{munagala2007optimization}, the queries are filters with no grouping or aggregates.
-Each query is filtered by a conjunction of predicates, where each single predicate may be used by multiple queries.
-A predicate index can be used to avoid recomputing the predicates, in case they are expensive to compute~\cite{madden2002continuously}.
-
-Multi-query data analysis\cite{andrade2003efficient} describes a similar problem, where they have multiple queries over the same data that run periodically.
-By grouping the multiple queries together, more optimisations can be performed than when treating the queries separately.
-However, this is again only for a small set of built-in aggregates.
-
-% It is well known that streaming dataflow languages correspond to Kahn process networks.
-
-% Kahn process networks have also been touted as a more flexible and intuitive abstraction than MapReduce~\cite{vrba2009kahn}.
-% \ben{This doesn't say anything about Icicle.}
-
-% Synchronous streaming dataflow languages such as Icicle and Lucy, then, should prove a useful abstraction for distributed computing.
+Icicle is closely related to work in continuous and shared queries.
+A continuous query is one that processes input data which may have new records added or removed from it at any time.
+The result of the continuous query must be updated as soon as the input data changes.
+Shared queries are ones in which the same sub expressions occur in several individual queries over the same data, and we wish to share the results of these sub expressions among all individuals that use them.
+For example, in Munagala \emph{et al}~\cite{munagala2007optimization}, input records are filtered by a conjunction of predicates, and the predicates occur in multiple queries.
+Madden \emph{et al}~\cite{madden2002continuously} uses a predicate index to avoid recomputing them.
+Andrade \emph{et al} describes a compiler for queries over geospacial imagery~\cite{andrade2003efficient} that shares the results of several pre-defined aggregation functions between queries.
+Continuous Query Language (CQL)~\cite{arasu2002abstract,stream2003stream} again allows aggregates in its queries, but they must be builtin aggregate functions.
+Icicle addresses a computationally similar problem, except that our input data sets can only have new records added rather than deleted, which allows us to support general aggregations rather than just filter predicates.
+It is not obvious how arbitrary aggregate functions could be supported while also allowing deletion of records from the input data --- other than by recomputing the entire aggregation after each deletion.
 

### 2016-06-09 2016fhpc/section/03-Core.tex
fix typo in S4
Cite anatomy of a loop

@@
 % Anonymous functions are only allowed as arguments to primitives: they cannot be applied or stored in variables.
-The $\mi{Plan}$ itself is split into a five stage \emph{loop anatomy}.
+The $\mi{Plan}$ itself is split into a five stage \emph{loop anatomy}~\cite{Shivers:Anatomy}.
 First we have the name of the table and the names and element types of each column.
@@
 Each time we receive a row from the table the accumulator associated with the @key@ is updated, using the default value @0@ if an entry for that key is not yet present.
-After we have processed the entire table each sum with its corresponding count to yield a map of means for each key.
+After we have processed the entire table we divide each sum with its corresponding count to yield a map of means for each key.
 \begin{alltt}

### 2016-06-09 2016fhpc/figures/Bench-All.tex
minor changes

@@
   symbolic x coords={R, wc, 1 CPU, grep, cat, 32 CPU, end},
-  width=8.5cm, height=5cm,
+  width=8.5cm, height=6.5cm,
 ]

### 2016-06-09 2016fhpc/section/01-Introduction.tex
minor changes

@@
 As the three queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter.
-We must instead lift each predicate to an expression-level conditional:
+We must instead lift each predicate to an expression-level conditional and compute the count by summing the conditional:
 \begin{code}
-  SELECT COUNT(IF(open > close, 1, 0))
-       , COUNT(IF(open < close, 1, 0))
-       , SUM  (IF(open > close, open, 0))
-       / COUNT(IF(open > close, 1, 0))
+  SELECT SUM(IF(open > close, 1,    0))
+       , SUM(IF(open < close, 1,    0))
+       , SUM(IF(open > close, open, 0))
+       / SUM(IF(open > close, 1,    0))
   FROM stocks;

### 2016-06-09 2016fhpc/section/02-Source.tex
Copy edits to intermediate language
minor changes

@@
 \section{Elements and Aggregates}
+\label{s:ElementsAndAggregates}
 To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
@@
 \subsection{Finite Streams and Synchronous Data Flow}
-In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length.
+In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, the streams processed by Icicle are conceptually finite in length.
 Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner.
@@
 The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and produces an environment containing the function or query name and its type.
-Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing only the types of the queries.
+Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing the types of all the definitions.
 
@@
 
-Rules TcPrimApp and TcFunApp produce the type a primitive or function applied to its arguments. Rule TcLet is standard.
+Rules TcPrimApp and TcFunApp produce the type of a primitive or function applied to its arguments. Rule TcLet is standard.
 
@@
 To lift a constant to a stream result we produce a meta-function that always returns the value.
-To lift a constant to an aggregate result we set the update meta-function to return a dummy, and have the eject meta-function return the value of interest.
+To lift a constant to an aggregate result we set the update meta-function to return a dummy value, and have the eject meta-function return the value of interest.
 

### 2016-06-09 2016fhpc/section/03-Core.tex
minor changes

@@
 
-Inlining the above functions into the query from \label{s:ElementsAndAggregates} yields the following:
+Inlining the above functions into the query from \S\ref{s:ElementsAndAggregates} yields the following:
 % This is guaranteed to terminate because no recursion is allowed in function definitions.
@@
 
-For the @map@ binding, each fold accumulator inside the body of the @group@ construct is associate with its own finite map.
+For the @map@ binding, each fold accumulator inside the body of the @group@ construct is associated with its own finite map.
 The @s@ accumulator is associated with map @gS@, and the @c@ accumulator with @gC@.
@@
 \begin{alltt}
-  folds  { gS  : Map Date Real = mapEmpty
+  folds  \{ gS  : Map Date Real = mapEmpty
            then mapUpdate gS key 0 (\(\lambda\)s. s + value)
@@
          ; gC  : Map Date Real = mapEmpty
-           then mapUpdate gC key 0 (\(\lambda\)c. c + 1) }
+           then mapUpdate gC key 0 (\(\lambda\)c. c + 1) \}
 
-  after  { map : Map Date Real
-            = mapMap (\(\lambda\)s c. s / c) (mapZip gS gC) }
+  after  \{ map : Map Date Real
+            = mapMap (\(\lambda\)s c. s / c) (mapZip gS gC) \}
 \end{alltt}
@@
 To fuse multiple plans we freshen the names of each binding and also concatenate the corresponding parts of the anatomy.
-The single-pass restriction on queries makes the fusion process so simple.
-The single-pass restriction implies that there are no fusion-preventing dependencies between any two query plans.
+The single-pass restriction on queries makes the fusion process so simple, because it ensures that there are no fusion-preventing dependencies between any two query plans.
 
-Given a fused query plan we the convert it to an imperative loop nest in a similar way to our prior work on flow fusion~\cite{lippmeier2013data}.
+Given a fused query plan we then convert it to an imperative loop nest in a similar way to our prior work on flow fusion~\cite{lippmeier2013data}.

### 2016-06-09 2016fhpc/section/04-Benchmarks.tex
minor changes

@@
 By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.
+The memory usage of Icicle starts at around 200MB of RAM for a single thread, but as more threads are added approaches 15MB per thread.
+The memory usage is constant in the input size and depends on the number of queries.
 The R code is single threaded and would require at least 150 processors to reach similar speeds, assuming perfect scaling.
@@
 
-Figure~\ref{fig:bench:queries} shows how the total througput scales as the number of fused queries is increased.
+Figure~\ref{fig:bench:queries} shows how the total read througput scales as the number of fused queries is increased.
 For each number of queries, we ran two versions of the fused result: one version that wrote the output to disk, and the other that piped the result to @/dev/null@.

### 2016-07-26 2016fhpc/section/02-Source.tex
Some notes from reviewers

@@
 Once we reach the end of the stream we will have access to the last key and can lookup the final result.
+%% Review #1:
+%% In Section 3 I kept looking for a definition of the “last” function, which takes an important role in the Introduction.
+%% In Section 4, I found the definition.
+It would help the reader to give a small note early in Section 2 or 3 that “last”, “mean”, “sum” are user-level (i.e.
+non-primitive) functions whose definition will be given in Section 4.
+Evaluation and typing rules are defined in~\S\ref{s:IcicleSource}, while the user functions @last@ and @mean@ are defined in~\S\ref{s:IcicleCore}.
 
@@
 
-In Icicle, queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment.
+%% Review #1
+%% It’s not clear what’s meant by “In Icicle, queries that would require unbounded buffering cannot be written”.
+%% My understanding is that they can be written, but we would get a runtime error because the buffer would eventually become too big to fit the memory. Is this right?
+% change to "statically outlawed by the typesystem": I think that's unambiguous.
+
+In Icicle, queries that would require unbounded buffering are statically outlawed by the typesystem, with one major caveat that we will discuss in a moment.
 In Icicle, the stream being processed (such as @xs@ above) is implicit in each query.
@@
 Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
-Function types $\TauFun$ include non-function modal types, and functions from modal type to function type.
+%% Review 3:
+%% In Section 3, the description of function types is confusing.
+%% The authors state, "Function types F include non-function modal types, and functions from modal type to function type."
+%% It appears from Figure 1 that function types do not include non-function modal types (syntactic category M), and include only functions from modal type(s) to *modal* type.
+Function types $\TauFun$ include functions with any number of modal type arguments to a modal return type.
 As Icicle is a first-order language, function types are not value types.

### 2016-07-26 2016fhpc/section/04-Benchmarks.tex
Some notes from reviewers

@@
 
-In our proof of concept tesing we replaced an existing R script that performed feature generation with new Icicle code.
+In our proof of concept testing we replaced an existing R script that performed feature generation with new Icicle code.
 The R script computed features from a 317GB data set.
@@
 The fused Icicle version significantly outperformed the R version of the queries, and the single-threaded version was on par with @wc@, while only a little slower than @grep@.
-This is despite the fact that the Icicle queries perform more computational work than than @wc@ and @grep@.
+This is despite the fact that the Icicle queries perform more computational work than @wc@ and @grep@.
 By using multiple processors, we were able to scale up to perform as well as @cat@, approaching the disk speed.

### 2016-07-26 2016fhpc/section/XX-Related.tex
Some notes from reviewers

@@
 Dynamic scheduling is trade-off: it is required for stream operators such as grouping and filtering where the output data rate is not known statically, but using dynamic techniques for graphs with static transfer rates tends to have a performance cost.
-Icicle includes grouping and filtering operators where the output rates are statically unknown, however the associated language constructs require grouped and filtered data to be aggregated rather than passed as as the input to another stream operator.
+Icicle includes grouping and filtering operators where the output rates are statically unknown, however the associated language constructs require grouped and filtered data to be aggregated rather than passed as the input to another stream operator.
 This allows Icicle to retain fully static scheduling, so the compiled queries consist of straight line code with no buffering.

### 2016-07-26 2016fhpc/Main.tex
Some notes from reviewers
Add the ACM category rubbish

@@
 \begin{document}
+\toappear{}
+\pagenumbering{gobble}
 
-% \exclusivelicense
-% \conferenceinfo{}{}
-% \copyrightyear{2015}
-% \copyrightdata{}
-\doi{}
-% \pagenumbering{gobble}
-
-\title{Icicle: write once, run once}
+\title{Icicle: Write Once, Run Once}
 
@@
 \begin{abstract}
-We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input stream will be fused.
-We use a modal type system to ensure that fused queries can be computed in an incremental fashion, and a fold-based intermediate language to compile down to efficient C code.
-We present production benchmarks demonstrating significant speedup over existing queries written in R, and on par with the widely used Unix tools \texttt{grep} and \texttt{wc}.
+We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input stream are fused.
+We use a modal type system to ensure that fused queries can be computed incrementalally, and a fold-based intermediate language to compile down to efficient C code.
+We present production benchmarks demonstrating significant speedup over existing queries written in R, and on par with the Unix tools \texttt{grep} and \texttt{wc}.
 
@@
 \end{abstract}
+
+\category
+	{D.3.4}
+	{Programming Languages}
+	{Processors---Compilers; Optimization}
+
+\keywords{Query, fusion, stream}
 

### 2016-07-26 2016fhpc/section/01-Introduction.tex
Add the ACM category rubbish

@@
 \end{code}
-In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered.
+
+In the above code, @open > close@ and @close < open@ are filter predicates, and @count@ counts how many times the predicate is true.
 
@@
 \end{code}
+
 As we see, the result of query fusion tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.
@@
 \end{code}
-Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated in an incremental fashion.
-While streaming through the table we always have access to the last key in the stream, but to select rows that match this key would require that we stream through the table again from the start.
+
+Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} implied by the above query cannot be evaluated incrementally.
+While streaming through the table we always have access to the last key in the stream, but finding the rows that match this key requires streaming the table again from the start.
 We need a better solution.
-With this in mind, the contributions of this paper are:
+The contributions of this paper are:
 \begin{itemize}

### 2016-12-05 2017mergingmerges/section/scrub.tex
really rough notes

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}

Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.

Furthermore, existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
Pull can thus express @merge@, but not @partition@.
Push fusion supports splits (@partition@), but not joins (@merge@).

We make the following contributions:
\begin{itemize}
\item a process calculus for streaming programs (\S\ref{s:Process});
\item some common combinators implemented as processes (\S\ref{s:Combinators});
\item and even more exotic combinators such as segmented operations (\S\ref{s:Combinators});
\item an algorithm for fusing pairs of processes (\S\ref{s:Fusion});
\item and a Coq proof of soundness (\S\ref{s:Proofs});
\end{itemize}

\section{Processes}
\label{s:Process}
Actually this is the process calculus version.

We will start by looking at a simple combinator that increments each value.
Given an input stream @ins@, @increment ins@ creates a new process with one output stream called @outs@.
The process blocks reading the value from @ins@.
After reading the value @i@, it pushes @i + 1@ to the @outs@ stream.
The pulled value must be dropped or released before another value can be performed.
Operationally this is a no-op, but is used for coordinating between producer and consumer.
Finally, the process loops back to the start.

\begin{code}
increment ins = process outs.
  let loop =
     i <- pull ins
     push outs (i + 1)
     drop ins
     jump loop
  in jump loop
\end{code}

What is interesting here?
Streams are named and infinite.
Output streams are owned by a particular process and can only be written to by it.
Pushes and pulls are blocking.
Pulls can be performed anywhere.
Drops are required for coordination.
This can be thought of as yielding control back to the producer after the value has been used.

\begin{code}
V := Scalar
C := Channel

Process :=
  | V <- pull C
  | push C Exp
  | drop C
  | skip
  | case Exp then Process else Process
  | jump F V*

Program :=
  process C*.
  let F V* = Process
  in  Process
\end{code}

Put in some evaluation rules.

\section{Combinators}
\label{s:Combinators}

Some simple combinator definitions.
Map

\begin{code}
map f xs = process ys.
  let loop =
     x <- pull xs
     push ys (f x)
     drop xs
     jump loop
  in jump loop
\end{code}

\begin{code}
filter f xs = process ts.
  let loop =
     x <- pull xs
     case f x
       then push ys x
       else skip
     drop xs
     jump loop
  in jump loop
\end{code}

\begin{code}
partition f xs = process ts fs.
  let loop =
     x <- pull xs
     case f x
       then push ts x
       else push fs x
     drop xs
     jump loop
  in jump loop
\end{code}

\begin{code}
zip xs ys = process xys.
  let loop =
     x <- pull xs
     y <- pull ys
     push xys (x,y)
     drop xs
     drop ys
     jump loop
  in jump loop
\end{code}

\begin{code}
merge xs ys = process xys.
  let loop x y =
     case x < y
       then push xys x
            drop xs
            x' <- pull xs
            jump loop x' y
       else push xys y
            drop ys
            y' <- pull ys
            jump loop x y'
  in x <- pull xs
     y <- pull ys
     jump loop x y
\end{code}

\section{Fusion}
\label{s:Fusion}

\section{Proofs}
\label{s:Proofs}
Proof of soundness is formalised in Coq.

\section{The problem with synchronised product}
Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.

This sort of coordination is a bit too coarse and causes deadlocks.
This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.

\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let abs = zip as bs
      bas = zip bs as
  in  abs, bas
\end{code}

After inlining the definition of zip, we have:
\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let abs = process abs.
       let loop1 =
          a <- pull as
          b <- pull bs
          push abs (a,b)
          drop as
          drop bs
          jump loop1
       in jump loop1
      bas = process bas.
       let loop2 =
          b <- pull bs
          a <- pull as
          push abs (b,a)
          drop bs
          drop as
          jump loop2
       in jump loop2
  in  abs, bas
\end{code}

Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
Thus, the two machines are deadlocked, waiting for the other one to agree.

For this case, a simple solution suffices to fix this: adding extra processes for splits. So we would add a new combinator @dup@ which takes one input stream and creates two output streams. It reads a single element from the input and pushes it to both outputs.

\begin{code}
dup ins = process l r.
  let loop =
     i <- pull ins
     push l i
     push r i
     drop ins
     jump loop
  in jump loop
\end{code}
And the modified zip uses this like so:

\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let as1, as2 = dup as
      bs1, bs2 = dup bs
      abs = zip as1 bs1
      bas = zip bs2 as2
  in  abs, bas
\end{code}

This has removed the deadlock, allowing a trace such as:
\begin{code}
as, as1, bs, bs1, bs2, as2
\end{code}

However it is not hard to modify the program so that the deadlock still exists.
By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let as1, as2 = dup as
      bs1, bs2 = dup bs
      abs = zip as2 bs1
      bas = zip bs2 as1
  in  abs, bas
\end{code}

Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
\begin{code}
zipself :: [a] -> [a*a]
zipself as =
  let as1, as2 = dup as
      aas = zip as2 as1
  in  aas
\end{code}
Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.

\section{Fully abstract case interpretation}

This cannot be fused because it requires an unbounded buffer.
\begin{code}
zipltgts :: [a] -> [a*a]
zipltgts as =
  let as1 = filter (<0) as
      as2 = filter (>0) as
      aas = zip as1 as2
  in  aas
\end{code}

You might think this can be fused.
It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
So while it cannot be true that the first filter reaches @>0 = true@ case and the second filter reaches @>0 = false@, we try both combinations and treat it as unfusable.
\begin{code}
zipgts :: [a] -> [a*a]
zipgts as =
  let as1 = filter (>0) as
      as2 = filter (>0) as
      aas = zip as1 as2
  in  aas
\end{code}

\section{Finite streams}
So far we have only treated infinite streams.
Finite streams could easily be encoded as infinite streams padded with a sentinel "end of file" element.
However, this plays poorly with the fully abstract case statements, as we would require a case for each element read to check whether it is the end of stream.
Adding these cases, we would end up with similar problems as @zipgts@ above.

One solution is to embed this special case into @pull@ itself, by giving it two actions to perform; one after successful read, one after end of stream.
It may be simpler and more useful to instead perform more sophisticated tracing of the @case@ conditionals though, as this would also solve the @zipgts@ example above.

### 2016-12-17 2017mergingmerges/section/scrub.tex
ugh

@@
 Push fusion supports splits (@partition@), but not joins (@merge@).
+Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections.
+However this requires manual plumbing.
 
@@
 
+\section{Streams vs Push/Pull}
+
+Types
+\begin{code}
+Stream  : Data -> Static
+Source  : Data -> Static
+Sink    : Data -> Static
+Process :         Static
+\end{code}
+
+Streams have the advantage of being able to be passed between producer and consumer.
+Stream combinators are easily composed as normal functions.
+
+The polar representation of Source/Sink has the advantage of having obvious bufferless execution semantics.
+Composition is harder.
+
 \section{Processes}
 \label{s:Process}
-Actually this is the process calculus version.
+We will start by looking at a combinator that increments each value.
+Given an input stream @ins@, @increment ins@ returns a new stream with the incremented values.
+The function @stream_1_1@ takes a process program with one pull input and one push output, and converts it to a stream program.
 
-We will start by looking at a simple combinator that increments each value.
-Given an input stream @ins@, @increment ins@ creates a new process with one output stream called @outs@.
 The process blocks reading the value from @ins@.
-After reading the value @i@, it pushes @i + 1@ to the @outs@ stream.
+After reading the value @i@, it pushes @i + 1@ to the @out@ stream.
 The pulled value must be dropped or released before another value can be performed.
@@
 \begin{code}
-increment ins = process outs.
-  let loop =
-     i <- pull ins
-     push outs (i + 1)
-     drop ins
-     jump loop
-  in jump loop
+increment : Stream Int -> Stream Int
+increment = stream_1_1 \ins out.
+  letrec
+    p1   = pull ins       p2
+    p2 i = push out (i+1) p3
+    p3   = drop ins       p1
+  in p1
 \end{code}
@@
 \begin{code}
-V := Scalar
-C := Channel
+stream_1_1 : (Source a -> Sink b
+              -> Process)
+          -> Stream a -> Stream b
+stream_2_1 : (Source a -> Source b -> Sink c
+              -> Process)
+          -> Stream a -> Stream b -> Stream c
+stream_1_2 : (Source a -> Sink b -> Sink c
+              -> Process)
+          -> Stream a -> (Stream b, Stream c)
 
-Process :=
-  | V <- pull C
-  | push C Exp
-  | drop C
-  | skip
-  | case Exp then Process else Process
-  | jump F V*
-
-Program :=
-  process C*.
-  let F V* = Process
-  in  Process
+pull : Source a -> (a -> Process) -> Process
+push : Sink   a ->  a -> Process  -> Process
+drop : Source a ->       Process  -> Process
+case : Bool -> Process -> Process -> Process
 \end{code}
@@
 \begin{code}
-map f xs = process ys.
-  let loop =
-     x <- pull xs
-     push ys (f x)
-     drop xs
-     jump loop
-  in jump loop
+map f = stream_1_1 \is os.
+  letrec
+    p1   = pull is p2
+    p2 i = push os (f i) p3
+    p3   = drop is p1
+  in p1
 \end{code}
@@
 \begin{code}
-filter f xs = process ts.
-  let loop =
-     x <- pull xs
-     case f x
-       then push ys x
-       else skip
-     drop xs
-     jump loop
-  in jump loop
+filter f = stream_1_1 \is os.
+  letrec
+    p1   = pull is p2
+    p2 i = case (f x) (p3 i) p4
+    p3 i = push os i p4
+    p4   = drop is p1
+  in p1
 \end{code}
@@
 \begin{code}
-partition f xs = process ts fs.
-  let loop =
-     x <- pull xs
-     case f x
-       then push ts x
-       else push fs x
-     drop xs
-     jump loop
-  in jump loop
+partition f = stream_1_2 \is ots ofs.
+  letrec
+    p1   = pull is p2
+    p2 i = case (f x)
+            (push ts i p3)
+            (push fs i p3)
+    p3   = drop is p1
+  in p1
 \end{code}
@@
 \begin{code}
-zip xs ys = process xys.
-  let loop =
-     x <- pull xs
-     y <- pull ys
-     push xys (x,y)
-     drop xs
-     drop ys
-     jump loop
-  in jump loop
+zip = stream_2_1 \xs ys xys.
+  letrec
+    p1     = pull xs        p2
+    p2 x   = pull ys        p3
+    p3 x y = push xys (x,y) p4
+    p4     = drop xs        p5
+    p5     = drop ys        p1
+  in p1
 \end{code}
@@
 \begin{code}
-merge xs ys = process xys.
-  let loop x y =
-     case x < y
-       then push xys x
-            drop xs
-            x' <- pull xs
-            jump loop x' y
-       else push xys y
-            drop ys
-            y' <- pull ys
-            jump loop x y'
-  in x <- pull xs
-     y <- pull ys
-     jump loop x y
+merge = stream_2_1 \xs ys xys.
+  letrec
+    go x y = case (x < y)
+             (pX x y)
+             (pY x y)
+    pX x y = push xys x
+            (drop xs
+            (pull xs (\x'. go x' y)))
+    pY x y = push xys y
+            (drop ys
+            (pull ys (\y'. go x y')))
+  in pull xs (\x. pull ys (\y. go x y))
 \end{code}
@@
 \label{s:Fusion}
+
+\begin{code}
+channels : Process -> ({Source}, {Sink})
+channels (pull i   p) = ({i}, {})  U channels p
+channels (drop i   p) = ({i}, {})  U channels p
+channels (push o _ p) = ({}, {o})  U channels p
+channels (case _ p q) = channels p U channels q
+\end{code}
 
@@
 \section{The problem with synchronised product}
+Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
 Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
@@
 \begin{code}
-dup ins = process l r.
-  let loop =
-     i <- pull ins
-     push l i
-     push r i
-     drop ins
-     jump loop
-  in jump loop
+dup = stream_1_2 \is ols ors.
+  letrec
+    p1   = pull is p2
+    p2 i = push ols i (p3 i)
+    p3 i = push ors i p4
+    p4   = drop is p1
+  in p1
 \end{code}

### 2016-12-26 2017mergingmerges/section/scrub.tex
just some thing

@@
 
+Another way to characterise fusion systems is whether they are \emph{local} (shortcut) or \emph{global} fusion.
+Local fusion systems such as stream fusion~\cite{coutts2007stream} use local rewrite rules to remove intermediate arrays, and tend to rely on general purpose compiler optimisations to expose opportunities for these rewrites to occur.
+Generally, the local rewrite rule can only be applied once the producer has been inlined into the consumer.
+Because of the local nature of these rewrites and the reliance on inlining, local fusion can only perform producer-consumer fusion with a single consumer, as inlining the producer into multiple consumers would duplicate work.
+Local fusion is fragile because heuristics are used to determine whether inlining occurs~\cite{lippmeier2013data}.
+This fragility poses serious problems to programmers who require fusion for adequate performance, as one must have deep knowledge of the internal compiler phases and library design in order to predict whether fusion will occur -- or worse yet, they must scour the intermediate code to count the number of loops produced.
+In contrast, global fusion systems such as those traditionally used in imperative compilers make use of specific optimisations which must be implemented separately, but are able to perform horizontal fusion and are less prone to the fragility of local fusion.
+
+This local/global distinction is not a hard classification, as some systems fit somewhere in between.
+For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but only performs producer-consumer fusion with a single local rule.
+
 We make the following contributions:
@@
 \item a process calculus for streaming programs (\S\ref{s:Process});
-\item some common combinators implemented as processes (\S\ref{s:Combinators});
-\item and even more exotic combinators such as segmented operations (\S\ref{s:Combinators});
+\item some common combinators implemented as processes (\S\ref{s:Combinators:Common});
+\item and as an appendix more exotic combinators such as segmented operations (\S\ref{s:Combinators:Exotic});
 \item an algorithm for fusing pairs of processes (\S\ref{s:Fusion});
-\item and a Coq proof of soundness (\S\ref{s:Proofs});
+\item and a formalisation and proof of soundness in Coq (\S\ref{s:Proofs});
 \end{itemize}
@@
 \section{Combinators}
-\label{s:Combinators}
+
+\subsection{Common}
+\label{s:Combinators:Common}
 
@@
   in p1
-\end{code}
-
-\begin{code}
-merge = stream_2_1 \xs ys xys.
-  letrec
-    go x y = case (x < y)
-             (pX x y)
-             (pY x y)
-    pX x y = push xys x
-            (drop xs
-            (pull xs (\x'. go x' y)))
-    pY x y = push xys y
-            (drop ys
-            (pull ys (\y'. go x y')))
-  in pull xs (\x. pull ys (\y. go x y))
 \end{code}
@@
 So far we have only treated infinite streams.
-Finite streams could easily be encoded as infinite streams padded with a sentinel "end of file" element.
+Finite streams could easily be encoded as infinite streams padded with a sentinel ``end of file'' element.
 However, this plays poorly with the fully abstract case statements, as we would require a case for each element read to check whether it is the end of stream.
@@
 It may be simpler and more useful to instead perform more sophisticated tracing of the @case@ conditionals though, as this would also solve the @zipgts@ example above.
+
+\appendix
+\section{Exotic Combinators}
+\label{s:Combinators:Exotic}
+
+\begin{code}
+merge = stream_2_1 \xs ys xys.
+  letrec
+    go x y = case (x < y)
+             (pX x y)
+             (pY x y)
+    pX x y = push xys x
+            (drop xs
+            (pull xs (\x'. go x' y)))
+    pY x y = push xys y
+            (drop ys
+            (pull ys (\y'. go x y')))
+  in pull xs (\x. pull ys (\y. go x y))
+\end{code}

### 2016-12-28 2017mergingmerges/section/XX-Processes.tex
there there

%!TEX root = ../Main.tex
\section{Processes}
\label{s:Processes}

\begin{figure}
\begin{minipage}[t]{0.4\textwidth}
\begin{tabbing}
\Instr \TABDEF @MMMM@  \TABSKIP $\Exp$ \TABSKIP $\Exp$ \TABSKIP $\Exp$ \kill

\Chan \> $\to$ \> (channel/stream name) \\
\Label \> $\to$ \> (label name) \\
\Value \> $\to$ \> (value) \\
\\

\Proc \>:=\> \\
M \= M \= \kill
\> @process@ \\
\> \> @  name:@  (process name) \\
\> \> @   ins:@  $\MapType{\Chan}{\InputState}$ \\
\> \> @  outs:@  $\sgl{\Chan}$ \\
\> \> @  heap:@  $\Sigma$ \\
\> \> @ label:@  \Label \\
\> \> @blocks:@  $\MapType{Label}{\Instr}$ \\
\\
\Instr \TABDEF \kill
\InputState \> := \> @pending@~\Value~$|$~@have@~$|$~@none@

\end{tabbing}
\end{minipage}
\begin{minipage}[t]{0.05\textwidth}
\quad
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}
\begin{tabbing}
\Instr \TABDEF @MMMM@  \TABSKIP $\Exp$ \TABSKIP $\Exp$ \TABSKIP $\Exp$ \kill

\Exp \> $\to$ \> (value expression) \\
\Var \> $\to$ \> (value variable) \\
$\Sigma$ \> $\to$ \> (value heap) \\
\\

\Instr
    \> :=\> @pull@  \> \Chan  \> \Var  \> \Goto \\
    \TABALT @drop@  \> \Chan  \>       \> \Goto \\
    \TABALT @push@  \> \Chan  \> \Exp  \> \Goto \\
    \TABALT @case@  \> \Exp   \> \Goto \> \Goto \\
    \TABALT @jump@  \>        \>       \> \Goto \\
\\
\\
\Goto \> := \> \Label $[\MapType{\Var}{\Exp}]$ \\
\\
\Nest \> := \> $\Proc\ldots$ \\
\end{tabbing}
\end{minipage}

\caption{Process definitions}
\label{fig:Process:Def}
\end{figure}

The grammar for processes is shown in figure~\ref{fig:Process:Def}.
Channels, labels and variables are specified by some external, globally unique set of names.
We assume that values and expressions are specified externally to the core calculus.
The actual details of the external computation are not important, except that there should be an evaluation form taking a heap and an expression and producing a value, as well as some way to destruct booleans.

A process defines a stream computation, taking any number of input streams and producing at least one output stream.
The process name is just for fun.
The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
The input states are explained in detail later in \S\ref{s:Process:Eval}.

The output streams are in some sense ``owned'' by the process that produces them.
While a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
A process may, however, produce multiple output streams.

The heap is used for evaluation - it is used when evaluating the external value calculus.
Each process has its own private heap, meaning that the only communication between processes occurs by streams.

The label is the current block, and each block is the instruction to be executed in the current state.
Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

Pulling from a stream is blocking, and as usual with Kahn processes~\cite{kahn1976coroutines}, there is no way to tell whether a stream currently has a value.
Pull takes as arguments the channel to read from, the variable to put the read value in, and the next label state to jump to after a successful pull.

After values have been pulled, they must be disposed of with @drop@.
This has no real evaluation semantics other than for coordination between processes.

All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
Combining variable update with stream instructions simplifies the fusion process~\REF, as some parts of fusion need to perform both at once.

A process nest is a set of multiple processes that can be evaluated concurrently.
The intersection of all process outputs should be empty - there should be no overlap.
Any inputs that are not mentioned as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

\subsection{Map/map}
\label{s:Process:MapMap}

One of the simplest combinators is @map@.
This might need to go elsewhere.
As well as needing a better example than map/map.
Let's start with the process definition for @map@.
The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
The goto labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
Mention that the initial heap has the names but no values, but they could be initialised to whatever.
It doesn't matter since they'll be written before anything is read.

\begin{code}
map f = process
    name: (map f)
     ins: as
    outs: bs
    heap: {a = 0}
   label: map0
  blocks: map0 = pull as    a  map1
          map1 = push bs (f a) map2
          map2 = drop as       map0
\end{code}

As well as a ``combinator nest'', a function comprised exclusively of process combinators.
The input streams are supplied as arguments, and output streams as return values.
\begin{code}
mapMap f g xs
 = let ys = map f xs
       zs = map g ys
   in  zs
\end{code}

It is not hard to assume that, given the process definitions and a combinator nest, we can produce a process nest.
This is simple enough for a paragraph prose description.
It's just a bit of inlining and renaming everything to be unique.

\begin{code}
process
    name: (map f)
     ins: xs
    outs: ys
    heap: {x}
   label: p0
  blocks: p0 = pull xs    x  p1
          p1 = push ys (f x) p2
          p2 = drop xs       p0
process
    name: (map g)
     ins: ys
    outs: zs
    heap: {y}
   label: q0
  blocks: q0 = pull ys    y  q1
          q1 = push zs (g y) q2
          q2 = drop ys       q0
\end{code}

Now we can perform some kind of fusion on this nest, resulting in one process that computes, as output, both @ys@ and @zs@.
Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator nest did not return them.

\begin{code}
process
    name: (map f / map g)
     ins: xs
    outs: ys zs
    heap: {x, y, _ys}
   label: p0q0
  blocks: p0q0            = pull xs    x  p1q0
          p1q0            = push ys (f x) p2q0-pending-ys { _ys = f x }
          p2q0-pending-ys = drop xs       p0q0-pending-ys
          p0q0-pending-ys = jump          p0q1-have-ys    { y = _ys }
          p0q1-have-ys    = push zs (g y) p0q2-have-ys
          p0q2-have-ys    = jump          p0q0
\end{code}

\subsection{Evaluation}
\label{s:Process:Eval}

### 2016-12-28 2017mergingmerges/section/scrub.tex
there there

@@
 This local/global distinction is not a hard classification, as some systems fit somewhere in between.
-For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but only performs producer-consumer fusion with a single local rule.
+For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but seems to only perform producer-consumer fusion with a single local rule.
 
@@
 \begin{itemize}
-\item a process calculus for streaming programs (\S\ref{s:Process});
+\item a process calculus for streaming programs (\S\ref{s:Processes});
 \item some common combinators implemented as processes (\S\ref{s:Combinators:Common});
@@
 \end{itemize}
+
+\input{section/XX-Processes.tex}
 
@@
 \section{Processes}
-\label{s:Process}
+\label{s:ProcessZZZ}
+
 We will start by looking at a combinator that increments each value.
@@
 \section{Combinators}
+\label{s:Combinators}
 

### 2016-12-31 2017mergingmerges/section/XX-Processes.tex
rough shape of evaluation rules

@@
 \label{s:Process:Eval}
+
+\begin{figure}
+
+$$
+\arrLR{
+  \boxed{\ProcInject{\Proc}{\Chan}{\Value}{\Proc}}
+}{
+  \boxed{\ProcsInject{\sgl{\Proc}}{\Chan}{\Value}{\sgl{\Proc}}}
+}
+$$
+
+$$
+\ruleIN{
+  c=@none@ \in @ins@~p
+}{
+  \ProcInject{p}{c}{v}{p ~@ins@~ \{ c = @pending@~v\}}
+}{InjectValue}
+\ruleIN{
+  c \not\in @ins@~p
+}{
+  \ProcInject{p}{c}{v}{p}
+}{InjectIgnore}
+$$
+
+$$
+\ruleIN{
+  \forall i.~ \ProcInject{p_i}{c}{v}{p'_i}
+}{
+  \ProcsInject{\sgl{p_i}}{c}{v}{\sgl{p'_i}}
+}{ProcessesInject}
+$$
+
+\caption{Process evaluation: inject}
+\label{fig:Process:Eval:Inject}
+\end{figure}
+
+\begin{figure}
+
+$$
+\alpha~@:=@~ \Push~\Chan~\Value ~|~ \tau
+$$
+
+$$
+@let@~@block@~p~=~@blocks@~p~(@label@~p)
+$$
+
+$$
+\arrLR{
+  \boxed{\ProcShake{\Proc}{\alpha}{\Proc}}
+}{
+  \boxed{\ProcsShake{\sgl{\Proc}}{\alpha}{\sgl{\Proc}}}
+}
+$$
+
+\TODO{These could be split into smaller judgment forms with something like @block;ins;sigma => label;ins;sigma;alpha@}
+
+\TODO{Missing the update heap on output. Should be @pull c x l[u] => p heap[u]@}
+
+$$
+\ruleIN{
+  c=@pending@~v \in @ins@~p
+  \quad
+  @pull@~c~x~l'~=~@block@~p
+}{
+  \ProcShake{p}{\tau}{p ~@ins@~ \{ c = @have@\},~@label@~=~l', @heap@ \{ x = v\}}
+}{Pull}
+\ruleIN{
+  c=@have@ \in @ins@~p
+  \quad
+  @drop@~c~l'~=~@block@~p
+}{
+  \ProcShake{p}{\tau}{p ~@ins@~ \{ c = @none@\},~@label@~=~l'}
+}{Drop}
+$$
+
+$$
+\ruleIN{
+  @push@~c~e~l'~=~@block@~p
+}{
+  \ProcShake{p}{\Push~c~(e~(@heap@~p))}{p~@label@~=~l'}
+}{Push}
+\ruleIN{
+  @jump@~l'~=~@block@~p
+}{
+  \ProcShake{p}{\tau}{p~@label@~=~l'}
+}{Jump}
+$$
+
+$$
+\ruleIN{
+  @case@~e~l_t~l_f~=~@block@~p
+  \quad
+  e~(@heap@~p)~=~@t@
+}{
+  \ProcShake{p}{\tau}{p~@label@~=~l_t}
+}{CaseT}
+\ruleIN{
+  @case@~e~l_t~l_f~=~@block@~p
+  \quad
+  e~(@heap@~p)~=~@f@
+}{
+  \ProcShake{p}{\tau}{p~@label@~=~l_f}
+}{CaseF}
+$$
+
+$$
+\ruleIN{
+  \ProcShake{p_i}{\tau}{p'_i}
+}{
+  \ProcsShake{
+    \sgl{p_0 \ldots p_i \ldots p_n}
+  }{\tau}{
+    \sgl{p_0 \ldots p'_i \ldots p_n}
+  }
+}{ProcessesInternal}
+$$
+
+$$
+\ruleIN{
+  \ProcShake{p_i}{\Push~c~v}{p'_i}
+  \quad
+  \forall j~|~j \neq i.~
+  \ProcInject{p_j}{c}{v}{p'_j}
+}{
+  \ProcsShake{
+    \sgl{p_0 \ldots p_i \ldots p_n}
+  }{\Push~c~v}{
+    \sgl{p'_0 \ldots p'_i \ldots p'_n}
+  }
+}{ProcessesPush}
+$$
+
+\caption{Process evaluation: shake}
+\label{fig:Process:Eval:Shake}
+\end{figure}
+
+We use the ``inject and shake'' method of evaluation. (This isn't a real thing.)
+
+Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
+The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall nest's input streams.
+For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
+Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.
+
+The rules in figure~\ref{fig:Process:Eval:Shake} are the `shake' part of evaluation, where actual computation occurs. 
+As usual, $\alpha$ denotes the message type, with $\tau$ being an internal message. The \Push~ message is a single value being output on a channel.

### 2017-01-01 2017mergingmerges/section/XX-Processes.tex
evaluation rules

--- 2016-12-28 2017mergingmerges/section/XX-Processes.tex
@@
 
-\begin{figure}
-\begin{minipage}[t]{0.4\textwidth}
-\begin{tabbing}
-\Instr \TABDEF @MMMM@  \TABSKIP $\Exp$ \TABSKIP $\Exp$ \TABSKIP $\Exp$ \kill
-
-\Chan \> $\to$ \> (channel/stream name) \\
-\Label \> $\to$ \> (label name) \\
-\Value \> $\to$ \> (value) \\
-\\
-
-\Proc \>:=\> \\
-M \= M \= \kill
-\> @process@ \\
-\> \> @  name:@  (process name) \\
-\> \> @   ins:@  $\MapType{\Chan}{\InputState}$ \\
-\> \> @  outs:@  $\sgl{\Chan}$ \\
-\> \> @  heap:@  $\Sigma$ \\
-\> \> @ label:@  \Label \\
-\> \> @blocks:@  $\MapType{Label}{\Instr}$ \\
-\\
-\Instr \TABDEF \kill
-\InputState \> := \> @pending@~\Value~$|$~@have@~$|$~@none@
-
-\end{tabbing}
-\end{minipage}
-\begin{minipage}[t]{0.05\textwidth}
-\quad
-\end{minipage}
-\begin{minipage}[t]{0.4\textwidth}
-\begin{tabbing}
-\Instr \TABDEF @MMMM@  \TABSKIP $\Exp$ \TABSKIP $\Exp$ \TABSKIP $\Exp$ \kill
-
-\Exp \> $\to$ \> (value expression) \\
-\Var \> $\to$ \> (value variable) \\
-$\Sigma$ \> $\to$ \> (value heap) \\
-\\
-
-\Instr
-    \> :=\> @pull@  \> \Chan  \> \Var  \> \Goto \\
-    \TABALT @drop@  \> \Chan  \>       \> \Goto \\
-    \TABALT @push@  \> \Chan  \> \Exp  \> \Goto \\
-    \TABALT @case@  \> \Exp   \> \Goto \> \Goto \\
-    \TABALT @jump@  \>        \>       \> \Goto \\
-\\
-\\
-\Goto \> := \> \Label $[\MapType{\Var}{\Exp}]$ \\
-\\
-\Nest \> := \> $\Proc\ldots$ \\
-\end{tabbing}
-\end{minipage}
-
-\caption{Process definitions}
-\label{fig:Process:Def}
-\end{figure}
+\input{figures/ProcessDef.tex}
 
@@
 \label{s:Process:Eval}
+
+\input{figures/ProcessEval.tex}
+
+We use the ``inject and shake'' method of evaluation. (This isn't a real thing.)
+
+Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
+The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall nest's input streams.
+For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
+Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.
+
+Injection can only happen when a process is ready to receive more input.
+A process has a single element buffer for each input, stored in its input state.
+This can be either @none@ meaning an empty buffer, @pending@ meaning a single value has been added to the buffer but has not been read yet, or @have@ meaning the value was added and in the process of being used.
+
+(InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
+An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded (or at least multiple element) buffer.
+Injecting the value puts the value as @pending@ in the buffer.
+
+(InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.
+
+(ProcessesInject) perform injection over a process nest.
+Every process in the nest must have the value injected into it.
+This means if multiple processes read from that stream, all input buffers for that stream must be empty.
+
+The rules in figure~\ref{fig:Process:Eval:Shake} are the `shake' part of evaluation, where actual computation occurs. 
+As usual, $\alpha$ denotes the message type, with $\tau$ being an internal message. The \Push~ message is a single value being output on a channel.
+
+The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{\phi'}$
+executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
+The output message $\alpha$ can be an internal state change or an emitted value.
+The result also has the new label, the new input state buffers, and the substitution to apply to the heap.
+
+The two judgment forms for shaking processes are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p}}$.
+The process shaking just shakes a single instruction and updates the process.
+Shaking a process nest chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the nest.
+
+(Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
+The result substitution sets the variable to the pulled value, as well as any substitutions in the \Goto~ of the instruction.
+
+(Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull.
+
+(Push) evaluates the push expression $e$ under the heap, and sends the value as the message.
+
+(Jump) simply returns the new label and substitution.
+
+(CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.
+
+(Shake) unwraps a single process and evaluates the instruction.
+It updates the process with the new label, input state and heap.
+The shorthand $\Sigma[\phi]$ is used to denote the heap with the substitution simultaneously applied.
+
+(ProcessesInternal) chooses one process from the nest and evaluates it.
+When the process evaluates with an internal message ($\tau$), the entire nest evaluates by replacing that process.
+
+(ProcessesPush) chooses one process from the nest and evaluates it, where the process evaluates with a push message.
+The emitted push message is then injected into all other processes in the nest, which means they must either ignore the channel or be ready to add it to their buffer.
+The entire nest then emits the same message.

### 2017-01-02 2017mergingmerges/section/01-Introduction.tex
start on fusion definition

--- 2016-12-28 2017mergingmerges/section/scrub.tex
@@
 \end{itemize}
-
-\input{section/XX-Processes.tex}
-
-\section{Streams vs Push/Pull}
-
-Types
-\begin{code}
-Stream  : Data -> Static
-Source  : Data -> Static
-Sink    : Data -> Static
-Process :         Static
-\end{code}
-
-Streams have the advantage of being able to be passed between producer and consumer.
-Stream combinators are easily composed as normal functions.
-
-The polar representation of Source/Sink has the advantage of having obvious bufferless execution semantics.
-Composition is harder.
-
-\section{Processes}
-\label{s:ProcessZZZ}
-
-We will start by looking at a combinator that increments each value.
-Given an input stream @ins@, @increment ins@ returns a new stream with the incremented values.
-The function @stream_1_1@ takes a process program with one pull input and one push output, and converts it to a stream program.
-
-The process blocks reading the value from @ins@.
-After reading the value @i@, it pushes @i + 1@ to the @out@ stream.
-The pulled value must be dropped or released before another value can be performed.
-Operationally this is a no-op, but is used for coordinating between producer and consumer.
-Finally, the process loops back to the start.
-
-\begin{code}
-increment : Stream Int -> Stream Int
-increment = stream_1_1 \ins out.
-  letrec
-    p1   = pull ins       p2
-    p2 i = push out (i+1) p3
-    p3   = drop ins       p1
-  in p1
-\end{code}
-
-What is interesting here?
-Streams are named and infinite.
-Output streams are owned by a particular process and can only be written to by it.
-Pushes and pulls are blocking.
-Pulls can be performed anywhere.
-Drops are required for coordination.
-This can be thought of as yielding control back to the producer after the value has been used.
-
-\begin{code}
-stream_1_1 : (Source a -> Sink b
-              -> Process)
-          -> Stream a -> Stream b
-stream_2_1 : (Source a -> Source b -> Sink c
-              -> Process)
-          -> Stream a -> Stream b -> Stream c
-stream_1_2 : (Source a -> Sink b -> Sink c
-              -> Process)
-          -> Stream a -> (Stream b, Stream c)
-
-pull : Source a -> (a -> Process) -> Process
-push : Sink   a ->  a -> Process  -> Process
-drop : Source a ->       Process  -> Process
-case : Bool -> Process -> Process -> Process
-\end{code}
-
-Put in some evaluation rules.
-
-\section{Combinators}
-\label{s:Combinators}
-
-\subsection{Common}
-\label{s:Combinators:Common}
-
-Some simple combinator definitions.
-Map
-
-\begin{code}
-map f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = push os (f i) p3
-    p3   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-filter f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x) (p3 i) p4
-    p3 i = push os i p4
-    p4   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-partition f = stream_1_2 \is ots ofs.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x)
-            (push ts i p3)
-            (push fs i p3)
-    p3   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-zip = stream_2_1 \xs ys xys.
-  letrec
-    p1     = pull xs        p2
-    p2 x   = pull ys        p3
-    p3 x y = push xys (x,y) p4
-    p4     = drop xs        p5
-    p5     = drop ys        p1
-  in p1
-\end{code}
-
-\section{Fusion}
-\label{s:Fusion}
-
-\begin{code}
-channels : Process -> ({Source}, {Sink})
-channels (pull i   p) = ({i}, {})  U channels p
-channels (drop i   p) = ({i}, {})  U channels p
-channels (push o _ p) = ({}, {o})  U channels p
-channels (case _ p q) = channels p U channels q
-\end{code}
-
-\section{Proofs}
-\label{s:Proofs}
-Proof of soundness is formalised in Coq.
-
-\section{The problem with synchronised product}
-Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
-Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
-So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.
-
-This sort of coordination is a bit too coarse and causes deadlocks.
-This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
-Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = zip as bs
-      bas = zip bs as
-  in  abs, bas
-\end{code}
-
-After inlining the definition of zip, we have:
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = process abs.
-       let loop1 =
-          a <- pull as
-          b <- pull bs
-          push abs (a,b)
-          drop as
-          drop bs
-          jump loop1
-       in jump loop1
-      bas = process bas.
-       let loop2 =
-          b <- pull bs
-          a <- pull as
-          push abs (b,a)
-          drop bs
-          drop as
-          jump loop2
-       in jump loop2
-  in  abs, bas
-\end{code}
-
-Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
-Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
-So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
-Thus, the two machines are deadlocked, waiting for the other one to agree.
-
-For this case, a simple solution suffices to fix this: adding extra processes for splits.
-So we would add a new combinator @dup@ which takes one input stream and creates two output streams.
-It reads a single element from the input and pushes it to both outputs.
-
-\begin{code}
-dup = stream_1_2 \is ols ors.
-  letrec
-    p1   = pull is p2
-    p2 i = push ols i (p3 i)
-    p3 i = push ors i p4
-    p4   = drop is p1
-  in p1
-\end{code}
-And the modified zip uses this like so:
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as1 bs1
-      bas = zip bs2 as2
-  in  abs, bas
-\end{code}
-
-This has removed the deadlock, allowing a trace such as:
-\begin{code}
-as, as1, bs, bs1, bs2, as2
-\end{code}
-
-However it is not hard to modify the program so that the deadlock still exists.
-By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as2 bs1
-      bas = zip bs2 as1
-  in  abs, bas
-\end{code}
-
-Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
-\begin{code}
-zipself :: [a] -> [a*a]
-zipself as =
-  let as1, as2 = dup as
-      aas = zip as2 as1
-  in  aas
-\end{code}
-Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.
-
-\section{Fully abstract case interpretation}
-
-This cannot be fused because it requires an unbounded buffer.
-\begin{code}
-zipltgts :: [a] -> [a*a]
-zipltgts as =
-  let as1 = filter (<0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-You might think this can be fused.
-It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
-So while it cannot be true that the first filter reaches @>0 = true@ case and the second filter reaches @>0 = false@, we try both combinations and treat it as unfusable.
-\begin{code}
-zipgts :: [a] -> [a*a]
-zipgts as =
-  let as1 = filter (>0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-\section{Finite streams}
-So far we have only treated infinite streams.
-Finite streams could easily be encoded as infinite streams padded with a sentinel ``end of file'' element.
-However, this plays poorly with the fully abstract case statements, as we would require a case for each element read to check whether it is the end of stream.
-Adding these cases, we would end up with similar problems as @zipgts@ above.
-
-One solution is to embed this special case into @pull@ itself, by giving it two actions to perform; one after successful read, one after end of stream.
-It may be simpler and more useful to instead perform more sophisticated tracing of the @case@ conditionals though, as this would also solve the @zipgts@ example above.
-
-\appendix
-\section{Exotic Combinators}
-\label{s:Combinators:Exotic}
-
-\begin{code}
-merge = stream_2_1 \xs ys xys.
-  letrec
-    go x y = case (x < y)
-             (pX x y)
-             (pY x y)
-    pX x y = push xys x
-            (drop xs
-            (pull xs (\x'. go x' y)))
-    pY x y = push xys y
-            (drop ys
-            (pull ys (\y'. go x y')))
-  in pull xs (\x. pull ys (\y. go x y))
-\end{code}

### 2017-01-02 2017mergingmerges/section/XX-Fusion.tex
start on fusion definition

%!TEX root = ../Main.tex
\section{Fusion}
\label{s:Fusion}

\input{figures/FusionDef.tex}

Figure~\ref{fig:Fusion:Types} shows the type definitions and types of functions for fusion.
Fusion proceeds by taking two processes in a nest, and creating a single process that computes both.
The new process uses labels made of the product of the two input processes, as well as the static part of the input state - that is, whether the input buffer is pending after injection, has been pulled, or is empty, but without the actual injected value.

$\InputState_1$ defines the statis input state. $\Label_1$ is a single process label with the set of input states. We add the constructor $@label@'$ to the $\Label$ type which is the pair of process labels with input states.

$\ChanType_2$ classifies the kind of channels and the communications between two processes.
Two processes can read from the same channel (@in2@), in which case pulling must be coordinated together.
When one process reads from a channel and the other ignores it (@in1@) no coordination is required.
When one process writes to a channel and the other reads (@in1out1@) the reading process must wait for the other to write.
Finally, when one process writes and the other ignores (@out1@) no coordination is necessary.
As explained in~\S\ref{s:Process:Eval}, it is not allowed for two processes to write to the same channel.

Before looking at the definitions, it is worth noting the types and purposes of important functions:
\begin{itemize}
\item
\ti{fuseNest} takes a process nest and repeatedly fuses pairs together. 
It chooses arbitrary processes with shared channels and fuses them together.

\item
\ti{fusePair} fuses a pair of processes together.
The heap variables must be distinct (this can be ensured by renaming).
The two processes must have some shared channels; otherwise no coordination is required at all, and the fusion process will choose an arbitrary interleaving of the processes.
This restriction is not an issue in practice since there is no benefit to fusing unrelated processes.

\item
\ti{tryStep} is used by \ti{fusePair} to emit an instruction for one of the processes.
It is called with an instruction for one process, as well as both processes' current states, and returns a value if that process can take a step (emit the instruction).
If either of the processes can take a step, the fused program will take a step.
If neither can take a step, fusion fails.
\end{itemize}

Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions.

Figure~\ref{fig:Fusion:Def:Step} contains the definitions of step fusion functions.

\input{figures/FusionUtils.tex}

Figure~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
\ti{channels} computes the $\ChanType_2$ map for a pair of processes.
\ti{connected} checks whether two processes have any shared inputs or outputs - basically whether it is worth fusing them.
\ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
\ti{swaplabels} flips the order of the compound labels in an instruction.

### 2017-01-02 2017mergingmerges/section/scrub.tex
start on fusion definition

--- 2016-12-17 2017mergingmerges/section/scrub.tex
@@
 %!TEX root = ../Main.tex
-\section{Introduction}
-\label{s:Introduction}
-
-Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
-However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
-Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
-
-Furthermore, existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
-Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
-Pull can thus express @merge@, but not @partition@.
-Push fusion supports splits (@partition@), but not joins (@merge@).
-Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections.
-However this requires manual plumbing.
-
-We make the following contributions:
-\begin{itemize}
-\item a process calculus for streaming programs (\S\ref{s:Process});
-\item some common combinators implemented as processes (\S\ref{s:Combinators});
-\item and even more exotic combinators such as segmented operations (\S\ref{s:Combinators});
-\item an algorithm for fusing pairs of processes (\S\ref{s:Fusion});
-\item and a Coq proof of soundness (\S\ref{s:Proofs});
-\end{itemize}
-
 \section{Streams vs Push/Pull}
@@
 
+\subsection{Common}
+\label{s:Combinators:Common}
+
 Some simple combinator definitions.

### 2017-01-04 2017mergingmerges/section/XX-Fusion.tex
fusion definition etc

@@
 
-Figure~\ref{fig:Fusion:Types} shows the type definitions and types of functions for fusion.
+Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
 Fusion proceeds by taking two processes in a nest, and creating a single process that computes both.
@@
 This restriction is not an issue in practice since there is no benefit to fusing unrelated processes.
-
-\item
-\ti{tryStep} is used by \ti{fusePair} to emit an instruction for one of the processes.
-It is called with an instruction for one process, as well as both processes' current states, and returns a value if that process can take a step (emit the instruction).
-If either of the processes can take a step, the fused program will take a step.
-If neither can take a step, fusion fails.
 \end{itemize}
@@
 
-Figure~\ref{fig:Fusion:Def:Step} contains the definitions of step fusion functions.
+Figures~\ref{fig:Fusion:Def:Step} and~\ref{fig:Fusion:Def:StepPair} define the judgment rules for fusing a single instruction at a time.
 

### 2017-01-07 2017mergingmerges/section/A1-Combinators.tex
some reorg

--- 2017-01-02 2017mergingmerges/section/scrub.tex
@@
 %!TEX root = ../Main.tex
-\section{Streams vs Push/Pull}
-
-Types
-\begin{code}
-Stream  : Data -> Static
-Source  : Data -> Static
-Sink    : Data -> Static
-Process :         Static
-\end{code}
-
-Streams have the advantage of being able to be passed between producer and consumer.
-Stream combinators are easily composed as normal functions.
-
-The polar representation of Source/Sink has the advantage of having obvious bufferless execution semantics.
-Composition is harder.
-
-\section{Processes}
-\label{s:ProcessZZZ}
-
-We will start by looking at a combinator that increments each value.
-Given an input stream @ins@, @increment ins@ returns a new stream with the incremented values.
-The function @stream_1_1@ takes a process program with one pull input and one push output, and converts it to a stream program.
-
-The process blocks reading the value from @ins@.
-After reading the value @i@, it pushes @i + 1@ to the @out@ stream.
-The pulled value must be dropped or released before another value can be performed.
-Operationally this is a no-op, but is used for coordinating between producer and consumer.
-Finally, the process loops back to the start.
-
-\begin{code}
-increment : Stream Int -> Stream Int
-increment = stream_1_1 \ins out.
-  letrec
-    p1   = pull ins       p2
-    p2 i = push out (i+1) p3
-    p3   = drop ins       p1
-  in p1
-\end{code}
-
-What is interesting here?
-Streams are named and infinite.
-Output streams are owned by a particular process and can only be written to by it.
-Pushes and pulls are blocking.
-Pulls can be performed anywhere.
-Drops are required for coordination.
-This can be thought of as yielding control back to the producer after the value has been used.
-
-\begin{code}
-stream_1_1 : (Source a -> Sink b
-              -> Process)
-          -> Stream a -> Stream b
-stream_2_1 : (Source a -> Source b -> Sink c
-              -> Process)
-          -> Stream a -> Stream b -> Stream c
-stream_1_2 : (Source a -> Sink b -> Sink c
-              -> Process)
-          -> Stream a -> (Stream b, Stream c)
-
-pull : Source a -> (a -> Process) -> Process
-push : Sink   a ->  a -> Process  -> Process
-drop : Source a ->       Process  -> Process
-case : Bool -> Process -> Process -> Process
-\end{code}
-
-Put in some evaluation rules.
-
+\appendix
 \section{Combinators}
 \label{s:Combinators}
-
-\subsection{Common}
-\label{s:Combinators:Common}
 
@@
 
-\section{Fusion}
-\label{s:Fusion}
-
-\begin{code}
-channels : Process -> ({Source}, {Sink})
-channels (pull i   p) = ({i}, {})  U channels p
-channels (drop i   p) = ({i}, {})  U channels p
-channels (push o _ p) = ({}, {o})  U channels p
-channels (case _ p q) = channels p U channels q
-\end{code}
-
-\section{Proofs}
-\label{s:Proofs}
-Proof of soundness is formalised in Coq.
-
-\section{The problem with synchronised product}
-Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
-Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
-So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.
-
-This sort of coordination is a bit too coarse and causes deadlocks.
-This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
-Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = zip as bs
-      bas = zip bs as
-  in  abs, bas
-\end{code}
-
-After inlining the definition of zip, we have:
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = process abs.
-       let loop1 =
-          a <- pull as
-          b <- pull bs
-          push abs (a,b)
-          drop as
-          drop bs
-          jump loop1
-       in jump loop1
-      bas = process bas.
-       let loop2 =
-          b <- pull bs
-          a <- pull as
-          push abs (b,a)
-          drop bs
-          drop as
-          jump loop2
-       in jump loop2
-  in  abs, bas
-\end{code}
-
-Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
-Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
-So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
-Thus, the two machines are deadlocked, waiting for the other one to agree.
-
-For this case, a simple solution suffices to fix this: adding extra processes for splits.
-So we would add a new combinator @dup@ which takes one input stream and creates two output streams.
-It reads a single element from the input and pushes it to both outputs.
-
-\begin{code}
-dup = stream_1_2 \is ols ors.
-  letrec
-    p1   = pull is p2
-    p2 i = push ols i (p3 i)
-    p3 i = push ors i p4
-    p4   = drop is p1
-  in p1
-\end{code}
-And the modified zip uses this like so:
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as1 bs1
-      bas = zip bs2 as2
-  in  abs, bas
-\end{code}
-
-This has removed the deadlock, allowing a trace such as:
-\begin{code}
-as, as1, bs, bs1, bs2, as2
-\end{code}
-
-However it is not hard to modify the program so that the deadlock still exists.
-By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as2 bs1
-      bas = zip bs2 as1
-  in  abs, bas
-\end{code}
-
-Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
-\begin{code}
-zipself :: [a] -> [a*a]
-zipself as =
-  let as1, as2 = dup as
-      aas = zip as2 as1
-  in  aas
-\end{code}
-Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.
-
-\section{Fully abstract case interpretation}
-
-This cannot be fused because it requires an unbounded buffer.
-\begin{code}
-zipltgts :: [a] -> [a*a]
-zipltgts as =
-  let as1 = filter (<0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-You might think this can be fused.
-It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
-So while it cannot be true that the first filter reaches @>0 = true@ case and the second filter reaches @>0 = false@, we try both combinations and treat it as unfusable.
-\begin{code}
-zipgts :: [a] -> [a*a]
-zipgts as =
-  let as1 = filter (>0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-\section{Finite streams}
-So far we have only treated infinite streams.
-Finite streams could easily be encoded as infinite streams padded with a sentinel ``end of file'' element.
-However, this plays poorly with the fully abstract case statements, as we would require a case for each element read to check whether it is the end of stream.
-Adding these cases, we would end up with similar problems as @zipgts@ above.
-
-One solution is to embed this special case into @pull@ itself, by giving it two actions to perform; one after successful read, one after end of stream.
-It may be simpler and more useful to instead perform more sophisticated tracing of the @case@ conditionals though, as this would also solve the @zipgts@ example above.
-
-\appendix
-\section{Exotic Combinators}
-\label{s:Combinators:Exotic}
-
 \begin{code}

### 2017-01-07 2017mergingmerges/section/XX-Related.tex
some reorg

%!TEX root = ../Main.tex
\section{Related work}

While it is not explicitly mentioned in the paper, Kiselyov et al~\cite{kiselyov2016stream}'s {\bf strymonas} system seems to only allow streams to be used once.
Furthermore, the entire stream computation must be terminated with a single fold which takes a single stream.
This means multiple outputs with different rates cannot be treated, for example @partition@ which returns multiple arrays filtered with different predicates.

This innocent-looking yet useless program, which should pair each line with itself, actually pairs each line with the successive line.
\begin{code}
let lines = linesFromFile filename
in  zip lines lines
\end{code}

One would expect the above program to have the same behaviour as the following - or at least produce an error otherwise:
\begin{code}
zip (linesFromFile filename) (linesFromFile filename)
\end{code}

The one feature they support that we do not is concatMap/flatMap, which concatenates all produced substreams.
This is a limitation in the \emph{conversion} to process calculus, not in the process calculus itself.
Our process calculus can handle concrete instantiations of concatMap when specialised to a particular subprocess, but not the general case where the subprocess is statically unknown.
By using a staging restriction as they do, we could modify the conversion to create a specialised version of concatMap for the given argument.

\subsection{The problem with synchronised product}
Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.

This sort of coordination is a bit too coarse and causes deadlocks.
This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.

\begin{code}
zips :: [a] -> [b] -> [a*b] * [b*a]
zips as bs =
  let abs = zip as bs
      bas = zip bs as
  in  abs, bas
\end{code}

% After inlining the definition of zip, we have:
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let abs = process abs.
%        let loop1 =
%           a <- pull as
%           b <- pull bs
%           push abs (a,b)
%           drop as
%           drop bs
%           jump loop1
%        in jump loop1
%       bas = process bas.
%        let loop2 =
%           b <- pull bs
%           a <- pull as
%           push abs (b,a)
%           drop bs
%           drop as
%           jump loop2
%        in jump loop2
%   in  abs, bas
% \end{code}
% 
% Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
% Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
% So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
% Thus, the two machines are deadlocked, waiting for the other one to agree.
% 
% For this case, a simple solution suffices to fix this: adding extra processes for splits. So we would add a new combinator @dup@ which takes one input stream and creates two output streams. It reads a single element from the input and pushes it to both outputs.
% 
% \begin{code}
% dup = stream_1_2 \is ols ors.
%   letrec
%     p1   = pull is p2
%     p2 i = push ols i (p3 i)
%     p3 i = push ors i p4
%     p4   = drop is p1
%   in p1
% \end{code}
% And the modified zip uses this like so:
% 
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let as1, as2 = dup as
%       bs1, bs2 = dup bs
%       abs = zip as1 bs1
%       bas = zip bs2 as2
%   in  abs, bas
% \end{code}
% 
% This has removed the deadlock, allowing a trace such as:
% \begin{code}
% as, as1, bs, bs1, bs2, as2
% \end{code}
% 
% However it is not hard to modify the program so that the deadlock still exists.
% By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
% \begin{code}
% zips :: [a] -> [b] -> [a*b] * [b*a]
% zips as bs =
%   let as1, as2 = dup as
%       bs1, bs2 = dup bs
%       abs = zip as2 bs1
%       bas = zip bs2 as1
%   in  abs, bas
% \end{code}
% 
% Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
% \begin{code}
% zipself :: [a] -> [a*a]
% zipself as =
%   let as1, as2 = dup as
%       aas = zip as2 as1
%   in  aas
% \end{code}
% Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.

### 2017-01-07 2017mergingmerges/section/scrub.tex
some reorg

@@
 %!TEX root = ../Main.tex
-\section{Streams vs Push/Pull}
-
-Types
-\begin{code}
-Stream  : Data -> Static
-Source  : Data -> Static
-Sink    : Data -> Static
-Process :         Static
-\end{code}
-
-Streams have the advantage of being able to be passed between producer and consumer.
-Stream combinators are easily composed as normal functions.
-
-The polar representation of Source/Sink has the advantage of having obvious bufferless execution semantics.
-Composition is harder.
 
@@
 \label{s:ProcessZZZ}
-
-We will start by looking at a combinator that increments each value.
-Given an input stream @ins@, @increment ins@ returns a new stream with the incremented values.
-The function @stream_1_1@ takes a process program with one pull input and one push output, and converts it to a stream program.
-
-The process blocks reading the value from @ins@.
-After reading the value @i@, it pushes @i + 1@ to the @out@ stream.
-The pulled value must be dropped or released before another value can be performed.
-Operationally this is a no-op, but is used for coordinating between producer and consumer.
-Finally, the process loops back to the start.
-
-\begin{code}
-increment : Stream Int -> Stream Int
-increment = stream_1_1 \ins out.
-  letrec
-    p1   = pull ins       p2
-    p2 i = push out (i+1) p3
-    p3   = drop ins       p1
-  in p1
-\end{code}
 
@@
 
-\begin{code}
-stream_1_1 : (Source a -> Sink b
-              -> Process)
-          -> Stream a -> Stream b
-stream_2_1 : (Source a -> Source b -> Sink c
-              -> Process)
-          -> Stream a -> Stream b -> Stream c
-stream_1_2 : (Source a -> Sink b -> Sink c
-              -> Process)
-          -> Stream a -> (Stream b, Stream c)
-
-pull : Source a -> (a -> Process) -> Process
-push : Sink   a ->  a -> Process  -> Process
-drop : Source a ->       Process  -> Process
-case : Bool -> Process -> Process -> Process
-\end{code}
-
-Put in some evaluation rules.
-
-\section{Combinators}
-\label{s:Combinators}
-
-\subsection{Common}
-\label{s:Combinators:Common}
-
-Some simple combinator definitions.
-Map
-
-\begin{code}
-map f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = push os (f i) p3
-    p3   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-filter f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x) (p3 i) p4
-    p3 i = push os i p4
-    p4   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-partition f = stream_1_2 \is ots ofs.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x)
-            (push ts i p3)
-            (push fs i p3)
-    p3   = drop is p1
-  in p1
-\end{code}
-
-\begin{code}
-zip = stream_2_1 \xs ys xys.
-  letrec
-    p1     = pull xs        p2
-    p2 x   = pull ys        p3
-    p3 x y = push xys (x,y) p4
-    p4     = drop xs        p5
-    p5     = drop ys        p1
-  in p1
-\end{code}
-
-\section{Fusion}
-\label{s:Fusion}
-
-\begin{code}
-channels : Process -> ({Source}, {Sink})
-channels (pull i   p) = ({i}, {})  U channels p
-channels (drop i   p) = ({i}, {})  U channels p
-channels (push o _ p) = ({}, {o})  U channels p
-channels (case _ p q) = channels p U channels q
-\end{code}
-
 \section{Proofs}
@@
 Proof of soundness is formalised in Coq.
-
-\section{The problem with synchronised product}
-Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
-Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
-So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.
-
-This sort of coordination is a bit too coarse and causes deadlocks.
-This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
-Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = zip as bs
-      bas = zip bs as
-  in  abs, bas
-\end{code}
-
-After inlining the definition of zip, we have:
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = process abs.
-       let loop1 =
-          a <- pull as
-          b <- pull bs
-          push abs (a,b)
-          drop as
-          drop bs
-          jump loop1
-       in jump loop1
-      bas = process bas.
-       let loop2 =
-          b <- pull bs
-          a <- pull as
-          push abs (b,a)
-          drop bs
-          drop as
-          jump loop2
-       in jump loop2
-  in  abs, bas
-\end{code}
-
-Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
-Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
-So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
-Thus, the two machines are deadlocked, waiting for the other one to agree.
-
-For this case, a simple solution suffices to fix this: adding extra processes for splits.
-So we would add a new combinator @dup@ which takes one input stream and creates two output streams.
-It reads a single element from the input and pushes it to both outputs.
-
-\begin{code}
-dup = stream_1_2 \is ols ors.
-  letrec
-    p1   = pull is p2
-    p2 i = push ols i (p3 i)
-    p3 i = push ors i p4
-    p4   = drop is p1
-  in p1
-\end{code}
-And the modified zip uses this like so:
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as1 bs1
-      bas = zip bs2 as2
-  in  abs, bas
-\end{code}
-
-This has removed the deadlock, allowing a trace such as:
-\begin{code}
-as, as1, bs, bs1, bs2, as2
-\end{code}
-
-However it is not hard to modify the program so that the deadlock still exists.
-By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let as1, as2 = dup as
-      bs1, bs2 = dup bs
-      abs = zip as2 bs1
-      bas = zip bs2 as1
-  in  abs, bas
-\end{code}
-
-Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
-\begin{code}
-zipself :: [a] -> [a*a]
-zipself as =
-  let as1, as2 = dup as
-      aas = zip as2 as1
-  in  aas
-\end{code}
-Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.
 
@@
 
-You might think this can be fused.
+You might think the following can be fused.
 It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
@@
 It may be simpler and more useful to instead perform more sophisticated tracing of the @case@ conditionals though, as this would also solve the @zipgts@ example above.
-
-\appendix
-\section{Exotic Combinators}
-\label{s:Combinators:Exotic}
-
-\begin{code}
-merge = stream_2_1 \xs ys xys.
-  letrec
-    go x y = case (x < y)
-             (pX x y)
-             (pY x y)
-    pX x y = push xys x
-            (drop xs
-            (pull xs (\x'. go x' y)))
-    pY x y = push xys y
-            (drop ys
-            (pull ys (\y'. go x y')))
-  in pull xs (\x. pull ys (\y. go x y))
-\end{code}

### 2017-01-07 2017mergingmerges/section/A1-Combinators.tex
some combinators

@@
 
-Some simple combinator definitions.
-Map
+Here we show the definitions of some combinators.
+We start with simple combinators supported by most streaming systems, and progress to more interesting combinators.
+Some combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but can be implemented in our finite stream extension described in~\S\REF.
+The fact that segmented versions of these combinators can be implemented is compelling evidence of this.
+
+\subsection{Standard combinators}
+
+\paragraph{Map} is one of the simplest combinators, and is supported almost everywhere.
 
 \begin{code}
-map f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = push os (f i) p3
-    p3   = drop is p1
-  in p1
+map f = process
+     ins: as
+    outs: bs
+    heap: {a}
+   label: p0
+  blocks: p0 = pull as    a  p1
+          p1 = push bs (f a) p2
+          p2 = drop as       p0
 \end{code}
 
+\paragraph{Filter} is another ubiquitous combinator.
+
 \begin{code}
-filter f = stream_1_1 \is os.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x) (p3 i) p4
-    p3 i = push os i p4
-    p4   = drop is p1
-  in p1
+filter f = process
+     ins: as
+    outs: bs
+    heap: {a}
+   label: p0
+  blocks: p0 = pull as    a  p1
+          p1 = case    (f a) p2 p3
+          p2 = push bs    a  p2
+          p3 = drop as       p0
 \end{code}
 
+\paragraph{Partition} is an interesting combinator since it produces two output streams at once.
+The two output streams have different rates.
+This sort of splitting is unsupported in pull fusion systems.
+
 \begin{code}
-partition f = stream_1_2 \is ots ofs.
-  letrec
-    p1   = pull is p2
-    p2 i = case (f x)
-            (push ts i p3)
-            (push fs i p3)
-    p3   = drop is p1
-  in p1
+partition f = process
+     ins: as
+    outs: ts fs
+    heap: {a}
+   label: p0
+  blocks: p0 = pull as    a  p1
+          p1 = case    (f a) pT pF
+          pT = push ts    a  p3
+          pF = push fs    a  p3
+          p3 = drop as       p0
 \end{code}
 
+\paragraph{Zip} takes two streams and combines them into one.
+It is fairly ubiquitous in pull fusion systems, but cannot be supported by push systems.
+
 \begin{code}
-zip = stream_2_1 \xs ys xys.
-  letrec
-    p1     = pull xs        p2
-    p2 x   = pull ys        p3
-    p3 x y = push xys (x,y) p4
-    p4     = drop xs        p5
-    p5     = drop ys        p1
-  in p1
+zip = process
+     ins: as bs
+    outs: abs
+    heap: {a, b}
+   label: p0
+  blocks: p0 = pull as    a   p1
+          p1 = pull bs    b   p2
+          p2 = push abs (a,b) p3
+          p3 = drop as        p4
+          p4 = drop bs        p0
 \end{code}
 
+\paragraph{Scan} performs a prefix sum over an input stream, much like a fold that returns every intermediate value.
+For simplicity we are performing just prefix sums, rather than arbitrary scans, but this is easy to generalise.
+Here we use two values in the heap; @a@ stores the last seen input and @s@ stores the current sum.
+We initialise @s@ to 0 at the start.
+Note that this prefix scan includes the seed value 0 in the output; updating the current sum to include the last pulled value happens after the push.
+
 \begin{code}
-merge = stream_2_1 \xs ys xys.
-  letrec
-    go x y = case (x < y)
-             (pX x y)
-             (pY x y)
-    pX x y = push xys x
-            (drop xs
-            (pull xs (\x'. go x' y)))
-    pY x y = push xys y
-            (drop ys
-            (pull ys (\y'. go x y')))
-  in pull xs (\x. pull ys (\y. go x y))
+scan = process
+     ins: as
+    outs: bs
+    heap: {a, s = 0}
+   label: p0
+  blocks: p0 = pull as a  p1
+          p1 = push bs s  p2[s = s + a]
+          p2 = drop as    p0
 \end{code}
+
+\paragraph{Unfold} creates a new stream from a seed and an iteration function.
+The seed has type @s@, the iteration has type @s -> (a,s)@ and the output stream contains @a@s.
+This process has only one state which pushes the output, calls the iteration function, and updates the state.
+The state/heap is initialised to the first result of the iteration function.
+
+\begin{code}
+unfold z f = process
+     ins: 
+    outs: as
+    heap: {s = f z}
+   label: p0
+  blocks: p0 = push as (fst s) p0[s = f (snd s)]
+\end{code}
+
+\paragraph{Merge} takes two sorted inputs and produces a sorted input containing both inputs.
+It is the main worker in merge sort.
+Merge is a particularly interesting combinator since the access patterns are data dependent; that is, which order it pulls from inputs depends on the values of the inputs.
+This is the infinite version of merge; the finite version would have a separate `fixup' stage where after one input has finished, it copies the rest of the other input.
+
+\begin{code}
+merge = process
+     ins: as bs
+    outs: abs
+    heap: {a, b}
+   label: pI0
+  blocks: pI0 = pull as a  pI1
+          pI1 = pull bs b  pC0
+          pC0 = case (a < b) pA0 pB0
+
+          pA0 = push abs a pA1
+          pA1 = drop as    pA2
+          pA2 = pull as  a pC0
+
+          pB0 = push abs b pB1
+          pB1 = drop bs    pB2
+          pB2 = pull bs  b pC0
+\end{code}
+
+\subsection{Segmented combinators}
+
+\paragraph{Segmented fold}
+takes two input streams; one for the segment lengths, one for the values.
+The output stream has the same rate as the segment lengths.
+
+\begin{code}
+folds k z = process
+     ins: lens vals
+    outs: outs
+    heap: {count, state, val}
+   label: init0
+  blocks: init0 = pull lens count go0[state = z]
+          go0   = case (count > 0) go1 done0
+          go1   = pull vals val    go2
+          go2   = jump go0[count = count - 1, state = k state val]
+          done0 = push outs state init0
+\end{code}
+
+\paragraph{Segmented take}
+takes two input streams; the segment descriptor lengths, as well as the underlying values.
+It outputs another two streams for the output segment descriptor and values.
+The number of elements to take is constant across all segments.
+
+\paragraph{Segmented append} is very similar to merge.

### 2017-01-08 2017mergingmerges/section/A1-Combinators.tex
a few combinators. just for fun

@@
 
+\begin{code}
+takes n = process
+     ins: lens vals
+    outs: olens ovals
+    heap: {count, takecount, ix, val}
+   label: init0
+  blocks: init0 = pull lens count lens0
+          lens0 = jump lens1
+                [ ix = 0
+                , takecount = min count n ]
+          lens1 = push olens takecount take0
+          take0 = case (ix < takecount) take1 drop0
+          take1 = pull vals val take2
+          take2 = push ovals val take0[ix = ix + 1]
+          drop0 = case (ix < count) drop1 init0
+          drop1 = pull vals val drop0[ix = ix + 1]
+\end{code}
+
 \paragraph{Segmented append} is very similar to merge.
+
+\begin{code}
+apps = process
+     ins: Alens Avals Blens Bvals
+    outs: olens ovals
+    heap: {Alen, Blen, val}
+   label: init0
+  blocks: init0 = pull Alens Alen init1
+          init1 = pull Blens Blen init2
+          init2 = push olens  (Alen + Blen) copyA0
+          copyA0 = case (Alen > 0) copyA1 copyB0
+          copyA1 = pull Avals val copyA2
+          copyA2 = push ovals val copyA0[Alen = Alen - 1]
+          copyB0 = case (Blen > 0) copyB1 init0
+          copyB1 = pull Bvals val copyB2
+          copyB2 = push ovals val copyB0[Blen = Blen - 1]
+\end{code}

### 2017-01-08 2017mergingmerges/code/example/MapMap.hs
code: add basic types and start on evaluator

@@ -1,0 +1,55 @@
+
+module MapMap where
+import Machine.Base
+import Machine.Eval
+import Control.Monad.State.Strict
+import qualified Data.Map       as Map
+
+
+
+map1 :: Process
+map1    
+ = Process
+ { processName   = "map1"
+ , processIns    = Map.fromList [(Channel "as", None)]
+ , processOuts   = [Channel "bs"]
+ , processHeap   = Heap (Map.fromList [(Var "a1", VInt 0)])
+ , processLabel  = Label "map10"
+ , processBlocks  =
+        [ ( Label "map10"
+          , Pull (Channel "as") (Var "a1")
+                 (Next (Label "map11") Map.empty))
+
+        , ( Label "map11"
+          , Push (Channel "bs") (XApp (XVar (Var "f")) (XVar (Var "a1")))     
+                 (Next (Label "map12") Map.empty))
+
+        , ( Label "map12"
+          , Drop (Channel "as")
+                 (Next (Label "map10") Map.empty))
+        ]
+ }
+
+
+map2 :: Process
+map2    
+ = Process
+ { processName   = "map2"
+ , processIns    = Map.fromList [(Channel "bs", None)]
+ , processOuts   = [ Channel "cs"]
+ , processHeap   = Heap (Map.fromList [(Var "b1", VInt 0)])
+ , processLabel  = Label "map20"
+ , processBlocks  =
+        [ ( Label "map20"
+          , Pull (Channel "bs") (Var "b1")
+                 (Next (Label "map21") Map.empty))
+
+        , ( Label "map21"
+          , Push (Channel "cs") (XApp (XVar (Var "g")) (XVar (Var "b1")))     
+                 (Next (Label "map22") Map.empty))
+
+        , ( Label "map22"
+          , Drop (Channel "as")
+                 (Next (Label "map20") Map.empty))
+        ]
+ }

### 2017-01-08 2017mergingmerges/code/src/Machine/Base.hs

@@ -1,0 +1,108 @@
+
+module Machine.Base where
+import qualified Data.Map       as Map
+import Data.Map                 (Map)
+
+
+-- | Generic name type.
+type Name
+        = String
+
+
+-- | Channel name.
+data Channel
+        = Channel Name
+        deriving (Show, Eq, Ord)
+
+
+-- | Variable name.
+data Var        
+        = Var Name
+        deriving (Show, Eq, Ord)
+
+-- | Code label.
+data Label      
+        = Label Name
+        deriving (Show, Eq)
+        
+
+-- | Value heap?
+data Heap
+        = Heap (Map Var Value)
+        deriving Show
+
+
+
+-- | Describes the state of the input buffer for each channel.
+data InputState
+        -- | No element is buffered.
+        = None
+
+        -- | A single value has been added to the buffer,
+        --   but has not been read from the buffer yet.
+        | Pending Value
+
+        -- | A single value has been added to the buffer,
+        --   and is currently being used.
+        | Have
+        deriving (Show, Eq)
+
+
+-- | A nest of processes.
+data Nest
+        = Nest [Process]
+        deriving (Show)
+
+
+-- | A stream process.
+data Process
+        = Process
+        { -- | Name of process.
+          processName   :: String
+
+          -- | Map of input channels to process to what state they're in.
+        , processIns    :: Map Channel InputState
+
+          -- | Map of output channel names.
+        , processOuts   :: [Channel]
+
+          -- | Value heap?
+        , processHeap   :: Heap
+
+          -- | Process label?
+        , processLabel  :: Label
+
+          -- | Labeled instructions.
+        , processBlocks :: [(Label, Instruction)]
+        }
+        deriving Show
+
+
+-- | A single instruction in the process.
+data Instruction
+        = Pull  Channel Var     Next
+        | Drop  Channel         Next
+        | Push  Channel Expr    Next
+        | Case  Expr            Next Next
+        | Jump                  Next
+        deriving (Show, Eq)
+
+
+-- | Move to next label and set variables to the given expressions.
+data Next
+        = Next Label (Map Var Expr)
+        deriving (Show, Eq)
+
+
+data Expr
+        = XVal Value
+        | XVar Var
+        | XAdd Expr Expr
+        | XApp Expr Expr
+        deriving (Show, Eq)
+
+data Value
+        = VInt  Int
+        | VAbs  Var Expr
+        deriving (Show, Eq)
+

### 2017-01-08 2017mergingmerges/code/src/Machine/Eval.hs

@@ -1,0 +1,119 @@
+
+module Machine.Eval where
+import Machine.Base
+import Data.Map                         (Map)
+import qualified Data.Map               as Map
+
+
+-- | Evaluate an expression in the current heap.
+eval   :: Heap -> Expr -> Value
+eval heap@(Heap hsHeap) xx
+ = case xx of
+        XVal value
+         -> value
+
+        XVar var
+         -> case Map.lookup var hsHeap of
+                Just value      -> value
+                Nothing         -> error $ "eval: unbound variable " ++ show var
+
+        XAdd x1 x2
+         |  VInt i1             <- eval heap x1
+         ,  VInt i2             <- eval heap x2
+         -> VInt (i1 + i2)
+
+        XApp x1 x2
+         |  VAbs var1 xBody     <- eval heap x1
+         ,  v2                  <- eval heap x2
+         -> error "eval: finish me"
+
+
+---------------------------------------------------------------------------------------------------
+-- | Inject a value into an input channel of a process.
+inject :: Channel -> Value -> Process -> Maybe Process
+inject c v p
+ = case Map.lookup c (processIns p) of
+        -- Process is ready to receive input on this channel.
+        Just None  -> Just p { processIns = Map.insert c (Pending v) (processIns p) }
+
+        -- Process was not ready to receive input on this channel.
+        Just _     -> Nothing
+
+        -- Process does not have an input of this name,
+        -- so just return the original process.
+        Nothing    -> Just p
+
+
+---------------------------------------------------------------------------------------------------
+-- | An output action from a Shake evaluation.
+data Action
+        = ActionPush Channel Value
+
+
+-- | Shake an instruction of a process.
+shake   :: Instruction                          -- ^ Instruction to shake.
+        -> Map Channel InputState               -- ^ States of input channels.
+        -> Heap                                 -- ^ Current heap.
+        -> Maybe ( Label                        --   New label to jump to.
+                 , Map Channel InputState       --   New channel input states.
+                 , Heap                         --   New heap.
+                 , Maybe Action)                --   Output action.
+
+shake instr sInput heap@(Heap hvHeap)
+ = case instr of
+        -- Pull a value from a channel.
+        Pull channel var (Next label' hxUpdate)
+         -- If there is a pending value then write it to the heap,
+         -- and set the channel state to 'Have' to indicate we're currently using it.
+         |  Just (Pending value) <- Map.lookup channel sInput
+         -> Just ( label'
[…]

### 2017-01-08 2017mergingmerges/code/example/MapMap.hs
Add feeding code

@@ -6,6 +6,9 @@ import Control.Monad.State.Strict
 import qualified Data.Map       as Map
 
 
+inputs
+ = Map.fromList
+ [ (Channel "as", [VInt 1, VInt 2, VInt 3]) ]
 
 map1 :: Process
 map1    

### 2017-01-08 2017mergingmerges/code/src/Machine/Base.hs

@@ -103,6 +103,7 @@ data Expr
 
 data Value
         = VInt  Int
+        | VBool Bool
         | VAbs  Var Expr
         deriving (Show, Eq)
 

### 2017-01-08 2017mergingmerges/code/src/Machine/Eval.hs

@@ -3,8 +3,9 @@ module Machine.Eval where
 import Machine.Base
 import Data.Map                         (Map)
 import qualified Data.Map               as Map
+import Data.List
 
-
+---------------------------------------------------------------------------------------------------
 -- | Evaluate an expression in the current heap.
 eval   :: Heap -> Expr -> Value
 eval heap@(Heap hsHeap) xx
@@ -30,18 +31,18 @@ eval heap@(Heap hsHeap) xx
 
 ---------------------------------------------------------------------------------------------------
 -- | Inject a value into an input channel of a process.
-inject :: Channel -> Value -> Process -> Maybe Process
+inject :: Channel -> Value -> Process -> Process
 inject c v p
  = case Map.lookup c (processIns p) of
         -- Process is ready to receive input on this channel.
-        Just None  -> Just p { processIns = Map.insert c (Pending v) (processIns p) }
+        Just None  -> p { processIns = Map.insert c (Pending v) (processIns p) }
 
         -- Process was not ready to receive input on this channel.
-        Just _     -> Nothing
+        Just _     -> error "inject: process not ready"
 
         -- Process does not have an input of this name,
         -- so just return the original process.
-        Nothing    -> Just p
+        Nothing    -> p
 
 
 ---------------------------------------------------------------------------------------------------
@@ -65,8 +66,8 @@ shake instr sInput heap@(Heap hvHeap)
         Pull channel var (Next label' hxUpdate)
          -- If there is a pending value then write it to the heap,
          -- and set the channel state to 'Have' to indicate we're currently using it.
-         |  Just (Pending value) <- Map.lookup channel sInput
-         -> Just ( label'
+         |   Just (Pending value) <- Map.lookup channel sInput
+         ->  Just ( label'
                  , Map.insert channel Have sInput
                  , Heap (Map.unions 
                           [ hvHeap
@@ -74,7 +75,7 @@ shake instr sInput heap@(Heap hvHeap)
                           , Map.singleton var value ])
                  , Nothing)
 
-         | otherwise
+         |  otherwise
          -> Nothing
 
 
@@ -110,8 +111,115 @@ shake instr sInput heap@(Heap hvHeap)
                           , Map.map (eval heap) hxUpdate ])
                  , Nothing)
 
+
+        -- Case branching.
+        Case xScrut (Next labelThen hxThen) (Next labelElse hxElse)
+         -> case eval heap xScrut of
[…]

### 2017-01-08 2017mergingmerges/code/example/MapMap.hs
code: make eval for map

@@ -4,6 +4,7 @@ import Machine.Base
 import Machine.Eval
 import Control.Monad.State.Strict
 import qualified Data.Map       as Map
+import qualified Text.Show.Pretty       as P
 
 
 inputs
@@ -24,7 +25,7 @@ map1
                  (Next (Label "map11") Map.empty))
 
         , ( Label "map11"
-          , Push (Channel "bs") (XApp (XVar (Var "f")) (XVar (Var "a1")))     
+          , Push (Channel "bs") (XApp (XVal VSucc) (XVar (Var "a1")))
                  (Next (Label "map12") Map.empty))
 
         , ( Label "map12"

### 2017-01-08 2017mergingmerges/code/src/Machine/Base.hs

@@ -77,7 +77,6 @@ data Process
         }
         deriving Show
 
-
 -- | A single instruction in the process.
 data Instruction
         = Pull  Channel Var     Next
@@ -104,6 +103,12 @@ data Expr
 data Value
         = VInt  Int
         | VBool Bool
+        | VSucc 
         | VAbs  Var Expr
         deriving (Show, Eq)
+
+data Component
+        = ComponentProcess Process
+        | ComponentInput   Channel [Value]
+        | ComponentOuptut  Channel [Value]     
 

### 2017-01-08 2017mergingmerges/code/src/Machine/Eval.hs

@@ -2,7 +2,7 @@
 module Machine.Eval where
 import Machine.Base
 import Data.Map                         (Map)
-import qualified Data.Map               as Map
+import qualified Data.Map.Strict        as Map
 import Data.List
 
 ---------------------------------------------------------------------------------------------------
@@ -28,6 +28,10 @@ eval heap@(Heap hsHeap) xx
          ,  v2                  <- eval heap x2
          -> error "eval: finish me"
 
+         |  VSucc               <- eval heap x1
+         ,  VInt i2             <- eval heap x2
+         -> VInt (i2 + 1)
+
 
 ---------------------------------------------------------------------------------------------------
 -- | Inject a value into an input channel of a process.
@@ -49,6 +53,7 @@ inject c v p
 -- | An output action from a Shake evaluation.
 data Action
         = ActionPush Channel Value
+        deriving Show
 
 
 -- | Shake an instruction of a process.
@@ -70,9 +75,9 @@ shake instr sInput heap@(Heap hvHeap)
          ->  Just ( label'
                  , Map.insert channel Have sInput
                  , Heap (Map.unions 
-                          [ hvHeap
+                          [ Map.singleton var value
                           , Map.map (eval heap) hxUpdate
-                          , Map.singleton var value ])
+                          , hvHeap ])
                  , Nothing)
 
          |  otherwise
@@ -87,8 +92,8 @@ shake instr sInput heap@(Heap hvHeap)
          -> Just ( label'
                  , Map.insert channel None sInput
                  , Heap (Map.unions
-                          [ hvHeap
-                          , Map.map (eval heap) hxUpdate ])
+                          [ Map.map (eval heap) hxUpdate
+                          , hvHeap ])
                  , Nothing)
 
 
@@ -97,8 +102,8 @@ shake instr sInput heap@(Heap hvHeap)
          -> Just ( label'
                  , sInput
                  , Heap (Map.unions
-                          [ hvHeap
-                          , Map.map (eval heap) hxUpdate ])
+                          [ Map.map (eval heap) hxUpdate
+                          , hvHeap ])
                  , Just (ActionPush channel (eval heap expr)))
 
 
@@ -107,8 +112,8 @@ shake instr sInput heap@(Heap hvHeap)
          -> Just ( label'
                  , sInput
                  , Heap (Map.unions
-                          [ hvHeap
-                          , Map.map (eval heap) hxUpdate ])
[…]

### 2017-01-08 2017mergingmerges/code/example/MapMap.hs
code: cleanups

@@ -1,7 +1,7 @@
 
 module MapMap where
 import Machine.Base
-import Machine.Eval
+import Machine.Execute
 import Control.Monad.State.Strict
 import qualified Data.Map       as Map
 import qualified Text.Show.Pretty       as P

### 2017-01-08 2017mergingmerges/code/src/Machine/Eval.hs

@@ -5,7 +5,7 @@ import Data.Map                         (Map)
 import qualified Data.Map.Strict        as Map
 import Data.List
 
----------------------------------------------------------------------------------------------------
+
 -- | Evaluate an expression in the current heap.
 eval   :: Heap -> Expr -> Value
 eval heap@(Heap hsHeap) xx
@@ -31,202 +31,5 @@ eval heap@(Heap hsHeap) xx
          |  VSucc               <- eval heap x1
          ,  VInt i2             <- eval heap x2
          -> VInt (i2 + 1)
-
-
----------------------------------------------------------------------------------------------------
--- | Inject a value into an input channel of a process.
-inject :: Channel -> Value -> Process -> Process
-inject c v p
- = case Map.lookup c (processIns p) of
-        -- Process is ready to receive input on this channel.
-        Just None  -> p { processIns = Map.insert c (Pending v) (processIns p) }
-
-        -- Process was not ready to receive input on this channel.
-        Just _     -> error "inject: process not ready"
-
-        -- Process does not have an input of this name,
-        -- so just return the original process.
-        Nothing    -> p
-
-
----------------------------------------------------------------------------------------------------
--- | An output action from a Shake evaluation.
-data Action
-        = ActionPush Channel Value
-        deriving Show
-
-
--- | Shake an instruction of a process.
-shake   :: Instruction                          -- ^ Instruction to shake.
-        -> Map Channel InputState               -- ^ States of input channels.
-        -> Heap                                 -- ^ Current heap.
-        -> Maybe ( Label                        --   New label to jump to.
-                 , Map Channel InputState       --   New channel input states.
-                 , Heap                         --   New heap.
-                 , Maybe Action)                --   Output action.
-
-shake instr sInput heap@(Heap hvHeap)
- = case instr of
-        -- Pull a value from a channel.
-        Pull channel var (Next label' hxUpdate)
-         -- If there is a pending value then write it to the heap,
-         -- and set the channel state to 'Have' to indicate we're currently using it.
-         |   Just (Pending value) <- Map.lookup channel sInput
-         ->  Just ( label'
-                 , Map.insert channel Have sInput
-                 , Heap (Map.unions 
-                          [ Map.singleton var value
-                          , Map.map (eval heap) hxUpdate
[…]

### 2017-01-08 2017mergingmerges/code/src/Machine/Execute.hs

@@ -1,0 +1,28 @@
+
+module Machine.Execute where
+import Machine.Base
+import Machine.Shake
+import Machine.Feed
+import Data.Map                         (Map)
+import Data.List
+import qualified Data.Map.Strict        as Map
+
+
+execute 
+        :: Map Channel [Value]          -- Input  channels values.
+        -> Map Channel [Value]          -- Output channel values.
+        -> [Process]                    -- Processes.
+        -> [Action]                     -- Actions
+        -> (Map Channel [Value], [Process], [Action])     
+                                        -- Processes after execution.
+
+execute cvsIn cvsOut ps acc
+ | all null $ map snd $ Map.toList cvsIn
+ = (cvsIn, ps, acc)
+
+ | otherwise
+ = let  (cvsIn', ps')   = feedProcesses cvsIn ps
+        (ps'',   as')   = shakeSteps [] ps' []
+   in   execute cvsIn' cvsOut ps'' (acc ++ as')
+
+

### 2017-01-08 2017mergingmerges/code/src/Machine/Feed.hs

@@ -1,0 +1,39 @@
+
+module Machine.Feed where
+import Machine.Base
+import Data.Map                         (Map)
+import Data.List
+import qualified Data.Map.Strict        as Map
+
+
+type ChannelValueS a    
+        =   Map Channel [Value] -> a
+        -> (Map Channel [Value], a)
+
+
+-- | Feed input into a channel, if it needs it.
+feedInputState ::  ChannelValueS (Channel, InputState)
+feedInputState cvs (c, state)
+ = case state of
+        None    
+         -> case Map.lookup c cvs of
+                Nothing         -> (cvs, (c, state))
+                Just (v:vs)     -> (Map.insert c vs cvs, (c, Pending v))
+
+        _ -> (cvs, (c, state))
+
+
+-- | Feed input into the given process.
+feedProcess :: ChannelValueS Process
+feedProcess cvs p
+ = let  (cvs', cis')    = mapAccumL feedInputState cvs 
+                        $ Map.toList $ processIns p
+
+        p'              = p { processIns = Map.fromList cis' }
+   in   (cvs', p')
+
+
+-- | Feed input into the given proceses.
+feedProcesses :: ChannelValueS [Process]
+feedProcesses cvs ps
+        = mapAccumL feedProcess cvs ps

### 2017-01-08 2017mergingmerges/code/src/Machine/Shake.hs

@@ -1,0 +1,159 @@
+
+module Machine.Shake where
+import Machine.Base
+import Machine.Eval
+import Data.Map                         (Map)
+import Data.List
+import qualified Data.Map.Strict        as Map
+
+
+-- | An output action from a Shake evaluation.
+data Action
+        = ActionPush Channel Value
+        deriving Show
+
+
+---------------------------------------------------------------------------------------------------
+-- | Inject a value into an input channel of a process.
+inject :: Channel -> Value -> Process -> Process
+inject c v p
+ = case Map.lookup c (processIns p) of
+        -- Process is ready to receive input on this channel.
+        Just None  -> p { processIns = Map.insert c (Pending v) (processIns p) }
+
+        -- Process was not ready to receive input on this channel.
+        Just _     -> error "inject: process not ready"
+
+        -- Process does not have an input of this name,
+        -- so just return the original process.
+        Nothing    -> p
+
+
+---------------------------------------------------------------------------------------------------
+shakeSteps 
+        :: [Process]            -- ^ Processes which are currently stalled.
+        -> [Process]            -- ^ Processes we can still try to execute.
+        -> [Action]             -- ^ Actions produced so far.
+        -> ( [Process]          --   Result processes.
+           , [Action])          --   Result actions.
+
+shakeSteps stalled [] acc
+ = (stalled, acc)
+
+shakeSteps stalled (p : psRest) acc
+ = case shakeStep p of
+        Just (p', mAction)
+         -> case mAction of
+                Nothing
+                 -> shakeSteps [] (stalled ++ (p' : psRest)) acc
+
+                Just action@(ActionPush channel value)
+                 -> let stalled' = map (inject channel value) stalled
+                        psRest'  = map (inject channel value) psRest
+                    in  shakeSteps [] (stalled' ++ (p' : psRest')) (acc ++ [action])
+
+        Nothing
+         -> shakeSteps (stalled ++ [p]) psRest acc
+
+
+---------------------------------------------------------------------------------------------------
+shakeStep :: Process -> Maybe (Process, Maybe Action)
+shakeStep p
+ |  Just instr  <- lookup (processLabel p) (processBlocks p)
+ =  case shake instr (processIns p) (processHeap p) of
+        Nothing 
+         -> Nothing
+
+        Just (label', sInput', heap', mAction)
+         -> Just ( p    { processLabel  = label'
+                        , processIns    = sInput'
+                        , processHeap   = heap' }
+                 , mAction)
+
+
[…]

### 2017-01-08 2017mergingmerges/code/src/Machine/Shake.hs
code: wait until consumers are ready to accept outputs from producers

@@ -15,18 +15,18 @@ data Action
 
 ---------------------------------------------------------------------------------------------------
 -- | Inject a value into an input channel of a process.
-inject :: Channel -> Value -> Process -> Process
+inject :: Channel -> Value -> Process -> Maybe Process
 inject c v p
  = case Map.lookup c (processIns p) of
         -- Process is ready to receive input on this channel.
-        Just None  -> p { processIns = Map.insert c (Pending v) (processIns p) }
+        Just None  -> Just $ p { processIns = Map.insert c (Pending v) (processIns p) }
 
         -- Process was not ready to receive input on this channel.
-        Just _     -> error "inject: process not ready"
+        Just _     -> Nothing
 
         -- Process does not have an input of this name,
         -- so just return the original process.
-        Nothing    -> p
+        Nothing    -> Just p
 
 
 ---------------------------------------------------------------------------------------------------
@@ -48,9 +48,15 @@ shakeSteps stalled (p : psRest) acc
                  -> shakeSteps [] (stalled ++ (p' : psRest)) acc
 
                 Just action@(ActionPush channel value)
-                 -> let stalled' = map (inject channel value) stalled
-                        psRest'  = map (inject channel value) psRest
-                    in  shakeSteps [] (stalled' ++ (p' : psRest')) (acc ++ [action])
+                 -- The process wants to push to one of its output channels,
+                 -- so the consumers need to be ready to accept it for the step to complete.
+                 |  Just stalled' <- sequence $ map (inject channel value) stalled
+                 ,  Just psRest'  <- sequence $ map (inject channel value) psRest
+                 -> shakeSteps [] (stalled' ++ (p' : psRest')) (acc ++ [action])
+
+                 -- One of the consumers wasn't ready to accept the output.
+                 |  otherwise
+                 -> shakeSteps (stalled ++ [p]) psRest acc
 
         Nothing
          -> shakeSteps (stalled ++ [p]) psRest acc

### 2017-01-08 2017mergingmerges/code/example/MapMap.hs
core: fix wrong var name in mapmap demo

@@ -49,11 +49,11 @@ map2
                  (Next (Label "map21") Map.empty))
 
         , ( Label "map21"
-          , Push (Channel "cs") (XApp (XVar (Var "g")) (XVar (Var "b1")))     
+          , Push (Channel "cs") (XApp (XVal VSucc) (XVar (Var "b1")))     
                  (Next (Label "map22") Map.empty))
 
         , ( Label "map22"
-          , Drop (Channel "as")
+          , Drop (Channel "bs")
                  (Next (Label "map20") Map.empty))
         ]
  }

### 2017-01-08 2017mergingmerges/section/01-Introduction.tex
prose

@@
 
-Fusion for array programs is important for removing intermediate arrays, reducing memory traffic as well as allocations.
-However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
-Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
 
-Furthermore, existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
+% However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
+% Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
+% For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+
+Suppose we have two files containing identifiers.
+Both files are sorted, but may contain duplicates.
+We wish to retrieve the unique set from each file, as well as the union of both files.
+Can we do this in a single pass over each file?
+\begin{code}
+uniques file1 file2
+ = let one = group file1
+       two = group file2
+       all = group (merge file1 file2)
+   in (one, two, all)
+\end{code}
+
+The @group@ combinator removes consecutive duplicates, while @merge@ concatenates sorted inputs such that the output remains sorted.
+
+Existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
 Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
@@
 \begin{itemize}
-\item a process calculus for streaming programs (\S\ref{s:Processes});
-\item some common combinators implemented as processes (\S\ref{s:Combinators:Common});
-\item and as an appendix more exotic combinators such as segmented operations (\S\ref{s:Combinators:Exotic});
-\item an algorithm for fusing pairs of processes (\S\ref{s:Fusion});
-\item and a formalisation and proof of soundness in Coq (\S\ref{s:Proofs});
+\item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
+\item an algorithm for fusing processes, the first which supports splits, joins and arbitrary combinators (\S\ref{s:Fusion});
+\item a formalisation and proof of soundness in Coq (\S\ref{s:Proofs});
+\item an informal description of the required extensions to support finite streams (\S\ref{s:Finite});
+\item and show that our process calculus is general enough to encode common combinators as well as segmented operations (\S\ref{s:Combinators}).
 \end{itemize}

### 2017-01-09 2017mergingmerges/section/XX-Processes.tex
prose
feed evaluation with external input values

@@
 \input{figures/ProcessEval.tex}
+\input{figures/ProcessFeed.tex}
 
@@
 It updates the process with the new label, input state and heap.
-The shorthand $\Sigma[\phi]$ is used to denote the heap with the substitution simultaneously applied.
 
@@
 The emitted push message is then injected into all other processes in the nest, which means they must either ignore the channel or be ready to add it to their buffer.
+If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
 The entire nest then emits the same message.
+
+The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process nest and output values are accumulated.
+The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{nest}}{\ti{streams}}{\ti{nest}'}$.
+The input map $\ti{inputs}$ contains values for the nest inputs: nest outputs are not allowed, but ignored channels can have values.
+The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
+Feeding evaluates the process nest until all input values have been injected.
+
+Note that the result stream and nest are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
+The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
+
+(FeedStart) is the axiom form where all input values have been injected and there are no input values left.
+This is the start of evaluation.
+
+(FeedInternal) first recursively feeds its input accumulator and process nest, then allows the resulting nest to take an internal step.
+The internal step does not affect the accumulators.
+The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.
+
+(FeedPush) works similarly to (FeedInternal) except that the process nest emits a push message.
+The pushed value is added to the end of the accumulator for that channel.
+
+(FeedExternal) allows inputs to be injected into the process nest.
+For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
+The recursive feed is evaluated with the last value removed from the accumulators.
+The last value is then injected into the nest, and added back to the result accumulators.

### 2017-01-09 2017mergingmerges/section/XX-Proof.tex
feed evaluation with external input values

--- 2015-04-12 2015mergingmerges/section/05-Proofs.tex
@@
 %!TEX root = ../Main.tex
+
 \section{Proofs}
@@
 
-As we do not wish to handle recursive combinators, we also do not need to handle merging cyclic machines.
-This is one case that regular dataflow handles but we do not.
-The predicate below checks if either of the machines does not use the other's input; if both used the other's input, it would create a cycle.
-
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@acyclic@~a~b$
-\> $:=$
-\> $\outputs{a}~\cap~\inputs{b}~=~\emptyset$
-\\
-\> $~\vee$ \> $\outputs{b}~\cap~\inputs{a}~=~\emptyset$
-\end{tabbing}
-
-Two machines cannot be merged if they both provide the same output, or write to the same output channel.
-
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@distinct@~a~b$
-\> $:=$
-\> $\outputs{a}~\cap~\outputs{b}~=\emptyset$
-\end{tabbing}
-
-The final requirement for correct fusion is the two machines to satisfy their invariants.
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@oks@~a~b$
-\> $:=$
-\> $@acyclic@~a~b~\wedge~@distinct@~a~b$ \\
-\> $\wedge$ \> $@ok@~a~\wedge~@ok@~b$ \\
-\end{tabbing}
-
-We also define a predicate for checking whether there exists a path of transitions between two states of a machine.
-\begin{tabbing}
-MMMMMMM \= MM \= \kill
-$@reachable@~m~u~v$ \\
-\> $:=$
-\> $u~=~v$ \\
-\> $\wedge$ \> $\bigvee_{u'|u~\Rightarrow~u'}~@reachable@~m~u'~v$ \\
-\end{tabbing}
-
-\subsection{Closure transitions}
-The first lemma we require is to show that the environment closure computes the intersection of all transitions into a given state.
-
-We would like to show that this holds for computing the closure for an entire machine:
-
-\begin{tabbing}
-MM \= MM \= MM \= \kill
-$\forall...$ \\
-\> \> $\EnvMachine{m}{\Gamma}$ \\
-\> $\wedge$ \> $u~:_E~e~\in~\Gamma$ \\
-\> $\implies$\> $\bigcap \{e'~|~n~\Rightarrow~u~\in~m$ \\
-\>           \>         \> $\wedge~\CheckOutTransUP{\Gamma'}{n}{u}{e'} \}$ \\
-\> $=$       \> $e$
-\end{tabbing}
-
-However this is not quite true, as the environment closure is only computed for reachable transitions.
-We also need a weaker postcondition in order to perform induction over the closure environment.
-We add the extra constraint that only transitions reachable from the pending state are considered.
-We also take into account the intersection of the initial set and the pending set.
-
-The actual lemma to prove, then, is:
-
-\begin{lemma}
-Closure is intersection.
-
-\begin{tabbing}
-MM \= MM \= MM \= \kill
-$\forall...$ \\
-\>          \> $\EnvGrow{\Gamma}{p}{\Gamma'}$ \\
-\> $\wedge$ \> $u~:_E~e~\in~\Gamma'$ \\
-\> $\implies$\> $\bigcap\{e'~|~n~\Rightarrow~u~\in~m$ \\
-\>           \>         \> $\wedge~\CheckOutTransUP{\Gamma'}{n}{u}{e'}$ \\
-\>           \>         \> $\wedge~\bigvee_{v~\in~p}@reachable@~m~v~n\}$ \\
-\> $\cap$ \> $\bigcap\{e' ~|~ u~:_E~e'~\in~\Gamma \} $ \\
-\> $\cap$ \> $\bigcap\{e' ~|~ u~:_E~e'~\in~p \} $ \\
-\> $=$       \> $e$
-\end{tabbing}
-\end{lemma}
-
-\begin{proof}
-By induction over the closure.
-
-\begin{itemize}
-\item \textbf{Case (CloComputed):}
-
-regardless of whether $s~=~u$ or not, we simply use the induction hypothesis; since $a~=~b$, removing $s$ from $p$ does not change the result, as $s~:_E~a$ is left in $\Gamma$.
-
-\item \textbf{Case (CloTransitions):}
-
-if $s~=~u$, the intersection is not affected as it is moved from $p$ to $\Gamma$.
-If $u$ is one of the transitions from $s$, it is already in the first part of the intersection, so adding it to $p$ will not change the intersection.
-
-\item \textbf{Case (CloRemoveOut):}
-
-again, we just use the induction hypothesis;
-$s$ is removed from $\Gamma$ but the intersection of $\Gamma$'s $s$ and $p$'s $s$ is inserted into $p$.
-
-\item \textbf{Case (CloFinished):}
-
-here, $p$ is empty and there is no path from any elements of $p$ to $u$, so neither the first nor the last intersections have any effect, and the result is just $\Gamma$.
-\end{itemize}
-
-\end{proof}
-
-Before we can prove that the result of merging satisfies invariants, we must show that the result of merging has an environment closure.
-Computing the environment closure requires all transitions into a state to have the same event set, except for values of outputs.
-
-\begin{lemma}
-Likeness implies closure
-\begin{tabbing}
-MM \= MM \= MM \kill
- $~($
-\> $\forall u ~|~ \bigcup_{q~\in~p~\cup~\Gamma}(@reachable@~m~q~u).$ \\
-\> $\forall p_0 \in p~\cup~\Gamma.$ \\
-\> $\forall p_1 \in p~\cup~\Gamma.$ \\
-\> $\forall path_0 \in @path@~p_0~u.$ \\
-\> $\forall path_1 \in @path@~p_1~u.$ \\
-\> \> $path_0~:_E~\tau_0$ \\
-\> $\wedge$ \> $path_1~:_E~\tau_1$ \\
-\> $\implies$ \> $@like@~\tau_0~\tau_1)$ \\
-$\implies$ \> $\exists \Gamma'.$ \> $\EnvGrow{\Gamma}{p}{\Gamma'}$ \\
-\end{tabbing}
-\end{lemma}
-
-\begin{proof}
-Induction over $\langle \card{@states@~m~\setminus~\Gamma},~\Sigma_{s~:_E~\tau~\in~\Gamma}\card{\tau},~\card{p} \rangle$, where $\langle \cdots \rangle$ is a lexicographic ordering.
-
-\begin{description}
-\item[Case $\langle 0,~0,~0 \rangle$]
-
-$\card{p}~=~0 \implies p=\emptyset$
-
-apply (CloFinished)
-
-\item[Case $\langle g,~t,~y \rangle$]
-
-destruct $p$
-\begin{description}
-\item[Case $\emptyset$]
-
-apply (CloFinished)
-
-\item[Case $s~:_E~\tau, p$]
-
-destruct $s \in \Gamma$
-\begin{description}
-\item[Case false]
-
-apply (CloTransitions).
-
-apply i.h:
-
-$\card{@states@~m~\setminus~(s~:_E~\tau, \Gamma)} < \card{@states@~m~\setminus~\Gamma}$
-
-\item[Case $s~:_E~a~\in~\Gamma$]
-
-destruct $a~=~\tau$
-\begin{description}
-\item[Case $a~=~\tau$]
-
-apply (CloComputed):
-
-$g$ same. $t$ same. $y$ decreases.
-
-\item[Case $a~\not=~\tau$]
-
-apply (CloUpdateTransitions).
-
-$@like@~a~\tau$ from hypothesis.
-
-$a~\not=~\tau~\implies~\card{a~\cap~\tau}~<~\card{a}$
-\TODO{this is wrong}
-
-$g$ same. $t$ decreases.
-\end{description}
-\end{description}
-\end{description}
-\end{description}
-
-\end{proof}
-
-\subsection{Values}
-For each state $u$ of $@merge@~a~b$, if either event set contains a @Value n@, then the machine's invariant set will contain @Value n@ at state $u$.
-(Note that the converse is not true, as @Value n@ will appear in the invariant set, but neither event set for local @Out@s)
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.\ \forall u \in (@merge@~a~b).$
-\\
-\>
-\> $@oks@~a~b$
-\\
-\> $\wedge$
-\> $\EnvMachine{@merge@~a~b}{\Gamma}$
-\\
-\> $\wedge$
-\> $u~:_E~e~\in~\Gamma$
-\\
-\> $\wedge$
-\> $\Value{n}~\in~(u \cdot e_1~\cup~u \cdot e_2)$
-\\
-\> $\implies$
-\> $\Value{n}~\in~e$
-\end{tabbing}
-
-Let $m~=~@merge@~a~b$.
-
-We first apply rule (CloMachine) to the assumption.
-
-$$
-\EnvMachine{@merge@~a~b}{\Gamma}
-\implies
-\EnvGrow{\emptyset}{@initial@~m~:_E~\emptyset}{\Gamma}
-$$
-
-We now proceed by induction over the closure.
-
-$$
-\EnvGrow{\Gamma}{p}{\Gamma'}
-$$
-
-(CloComputed): by the induction hypothesis.
-
-We proceed by induction over the @move@ rules.
-
-The only way that @Value n@ can be added to the event set is through the rules (Shared output, ready to read) and (Shared pull, no pending values).
-These rules themselves produce state types of @Out@ and @Pull@ respectively, and by the definition of invariant sets @Value@ is inserted for these transitions.
-
-For rule (Local release), the state type is @Release n@ which removes @Value n@ from the invariant set.
-Due to the locality preconditions $n~\not\in~@inputs@~m_2$ and $n~\not\in~@outputs@~m_2$, we know that neither of the rules to add @n@ could apply, so neither event set can contain @Value n@.
-
-For the cases that remove a @Value n@ from the event set, the first two, (Release of output) and (Shared release: first release is skip) both only emit skips, so do not affect the invariant set.
-This means that while the invariant set may still contain a @Value n@, it is not obliged to.
-We appeal to the induction hypothesis for the input transitions, since earlier input transitions must have added the @Value n@ to both sets.
-
-The last case that removes a @Value n@ from the event set is (Shared release: other machine already released).
-Here, we know the first machine's event set, $e_1$, contains a @Value n@ and that the second machine's event set, $e_2$, does not.
-This means that after removing the @Value n@ from $e_1$, neither set will have one.
-The produced state type also means that the invariant set will no longer have a @Value n@.
-
-For the remaining cases, the @Value@s inside the event sets are left alone, and their state types do not explicitly remove any @Value n@s from the invariant set either.
-We can conclude that any @Value@s in the sets have been added by other states, and not removed.
-The last thing is that @Value@s can be implicitly removed at join-points of states.
-However, in order for a state to have @Value@s in its sets, all transitions into this state must have either inserted @Value@s itself, or kept @Value@s from its predecessors in turn; otherwise it would not have a transition to a state with @Value@s in its sets.
-
-\subsection{Preservation}
-The result of merging two machines should maintain the pull invariants (\S\ref{s:Machines:Invariants}), provided that the input machines also maintain their invariants.
-
-\begin{tabbing}
-MMMMMMM \= MM \= MM \= \kill
-$\forall a~b.$
-\>
-\> $a~@ok@~\wedge~b~@ok@$
-\\
-\> $\wedge$
-\> $@acyclic@~a~b$
-\\
-\> $\wedge$
-\> $@distinct@~a~b$
-\\
-\> $\implies$
-\> $\forall c \in (@merge@~a~b).\ c~@ok@$
-\end{tabbing}
-
-\paragraph{Lemma closes and finishes:} 
-for each state $u$ of $@merge@~a~b$, if \emph{either} event set contains a @Finish n@, the invariant set at state $u$ will also contain a @Finish n@.
-If \emph{both} event sets contain a @Close n@, the invariant set will contain a @Close n@.
-
-This proof proceeds very similarly to the above lemma.
-For (Finish shared output, no pending events), a @Finished n@ is added to $e_2$, while @OutFinished n@ adds a @Finish n@ to the invariant set.
-For (Shared pull, no pending values), a @Pull@ is emitted, and the @None@ case adds @Finished n@ to both sets, as with the invariant set.
-
-\paragraph{Lemma locals:} for any reads, outs or closes in one machine that are not in @inputs@ or @outputs@ of the other machine, no state in the other machine will affect the invariant set relating to these channels.
-By definition of @inputs@ and @outputs@.
-
-\paragraph{Lemma transitions:} for each state $u$ of $@merge@~a~b$ and corresponding states $s$ and $t$ of $a$ and $b$, any output transitions from $u$ either move along $s$'s output transitions or $t$'s output transitions, but not both.
-Similarly, the state type of $u$ will either be the type of $s$, the type of $t$, or a @Skip@.
-By inspection of each rule of @move@.
-
-\paragraph{Lemma paths:} for two given states $u$ and $v$ of $@merge@~a~b$, and their corresponding input machine states $s$ and $t$, for each path between $u$ and $v$, there exists a path between $s$ and $t$ such that all states in the $st$ path are in the $uv$ path, \emph{except}:
-\begin{itemize}
-\item @Pull@s can be omitted if every predecessor path of the @Pull@ has a @Pull@, @Out@ or @OutFinished@ on the same channel;
-\item @Release@s can be omitted if the channel is an output of the other machine;
-\item @Release@s can be omitted if every successor path has a @Release@ on the same channel;
-\item @Close@s can be omitted if every successor path has a @Close@ or @Pull@ on the same channel.
-\end{itemize}
-
-It is easy to see that for non-interfering transitions @Update@, @Skip@, @If@ and @Done@, this holds.
-For the other cases, it is necessary to perform induction over the @move@ rules, and inspect the event sets.
-If one of the input machine's states is a @Pull@, then either a @Pull@ or a @Skip@ may be produced.
-@Skip@s are only produced if the event set already contains a @Value@ or a @Finished@, which can only be produced by an earlier @Pull@, @Out@ or @OutFinished@.
-Likewise for the remaining cases.
-
-\paragraph{Proof sketch:} 
-We can use the above lemmas to show that if the two input machines are valid, the result of @merge@ will also be valid.
-For non-shared or local channels, it is easy to see that the invariant set for these channels will be exactly as it was in the input machine (lemma locals).
-
-For shared inputs, since each input machine is valid, we know all @Pull@s are paired with @Release@s.
-Given lemma values, and that @Release@s remove @Value@s from the invariant set, we can see that each generated @Pull@ on a shared resource will only be generated when the invariant set has no @Value@s, and will be similarly paired for @Release@s.
-
-For shared outputs, we just need to verify that the only way the reading machine can reach a reading state is while there is a @Value n@ in the invariant set, and this also follows easily from lemma values.
-
-Finally, we can see as a result of lemma closes and finishes and paths, that since both machines finish or close their inputs and outputs by their @Done@ state, then by the @Done@ of the merged machine, where both machines are also @Done@, all channels must be similarly closed at the end.
+Proof of soundness is formalised in Coq.

### 2017-01-09 2017mergingmerges/section/scrub.tex
prose
feed evaluation with external input values

@@
 %!TEX root = ../Main.tex
-
-\section{Processes}
-\label{s:ProcessZZZ}
-
-What is interesting here?
-Streams are named and infinite.
-Output streams are owned by a particular process and can only be written to by it.
-Pushes and pulls are blocking.
-Pulls can be performed anywhere.
-Drops are required for coordination.
-This can be thought of as yielding control back to the producer after the value has been used.
-
-\section{Proofs}
-\label{s:Proofs}
-Proof of soundness is formalised in Coq.
 
@@
 \section{Finite streams}
+\label{s:Finite}
 So far we have only treated infinite streams.

### 2017-01-09 2017mergingmerges/section/XX-Proof.tex
prose: talk about proofs, talk about unidirectional soundness

@@
 
-Proof of soundness is formalised in Coq.
+Our fusion system is formalised in Coq, where we have proved soundness of fusion: if the fused program evaluates to a particular output, then the original programs also evaluate to that output.
+It is important to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
+This is because evaluation of a process nest is non-deterministic, and fusion commits to a particular evaluation order.
+
+The problem with commiting to an evaluation order is most easily explained with a process nest containing two infinitely pushing processes.
+Process @A@ is repeatedly pushing to a stream called @X@, while process @B@ repeatedly pushes to @Y@.
+When evaluating this pair as a process nest, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.
+
+When fusion is performed on this process pair, an arbitrary but \emph{particular} order will be chosen.
+Thus, the chosen order will be one of the valid ones, but not all valid orders will be the chosen one.
+
+We know of no `realistic' examples where combinators have multiple evaluation orders, so believe this is not an issue in practice.
+
+The system described here has some differences to our Coq formalisation.

### 2017-01-13 2017mergingmerges/section/01-Introduction.tex
prose

@@
 
-Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
-
+% Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
 % However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
@@
 
-Suppose we have two files containing identifiers.
+Suppose we have two large files containing identifiers, and wish to perform some analysis on them.
 Both files are sorted, but may contain duplicates.
-We wish to retrieve the unique set from each file, as well as the union of both files.
-Can we do this in a single pass over each file?
+We wish to retrieve the unique set from the first file, as well as the union of both files.
+Can we perform both of these tasks at once, without reading the input files multiple times?
+
 \begin{code}
-uniques file1 file2
- = let one = group file1
-       two = group file2
-       all = group (merge file1 file2)
-   in (one, two, all)
+uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
+ = let uniques = group file1
+       union   = group (merge file1 file2)
+   in (uniques, union)
 \end{code}
 
-The @group@ combinator removes consecutive duplicates, while @merge@ concatenates sorted inputs such that the output remains sorted.
-
-Existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
-Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
-Pull can thus express @merge@, but not @partition@.
-Push fusion supports splits (@partition@), but not joins (@merge@).
-Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections.
-However this requires manual plumbing.
-
-Another way to characterise fusion systems is whether they are \emph{local} (shortcut) or \emph{global} fusion.
-Local fusion systems such as stream fusion~\cite{coutts2007stream} use local rewrite rules to remove intermediate arrays, and tend to rely on general purpose compiler optimisations to expose opportunities for these rewrites to occur.
-Generally, the local rewrite rule can only be applied once the producer has been inlined into the consumer.
-Because of the local nature of these rewrites and the reliance on inlining, local fusion can only perform producer-consumer fusion with a single consumer, as inlining the producer into multiple consumers would duplicate work.
-Local fusion is fragile because heuristics are used to determine whether inlining occurs~\cite{lippmeier2013data}.
-This fragility poses serious problems to programmers who require fusion for adequate performance, as one must have deep knowledge of the internal compiler phases and library design in order to predict whether fusion will occur -- or worse yet, they must scour the intermediate code to count the number of loops produced.
-In contrast, global fusion systems such as those traditionally used in imperative compilers make use of specific optimisations which must be implemented separately, but are able to perform horizontal fusion and are less prone to the fragility of local fusion.
-
-This local/global distinction is not a hard classification, as some systems fit somewhere in between.
-For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but seems to only perform producer-consumer fusion with a single local rule.
+In this example the @group@ combinator removes consecutive duplicates leaving only the first value, while @merge@ concatenates two sorted inputs such that the output remains sorted.
+There are a few interesting things about this example.
+First, the access pattern of @merge@ is \emph{value-dependent}: the order in which @merge@ reads values from @file1@ or @file2@ depends on the values in the files.
+If all the values in @file1@ are smaller than the values in @file2@, then @merge@ will pull values from @file1@ before pulling from @file2@, and vice versa.
+Second, note that @file1@ is used twice.
+If we wish to ensure that each value is only read from the file once, we must coordinate between the two use sites: when @uniques@ requires a new value it must ensure that @union@ is ready to receive a new value, and vice versa.
+It is important to note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
 
@@
 \item an informal description of the required extensions to support finite streams (\S\ref{s:Finite});
-\item and show that our process calculus is general enough to encode common combinators as well as segmented operations (\S\ref{s:Combinators}).
+\item and show that our calculus is general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
 \end{itemize}

### 2017-01-13 2017mergingmerges/section/XX-Proof.tex
prose

@@
 
-Our fusion system is formalised in Coq, where we have proved soundness of fusion: if the fused program evaluates to a particular output, then the original programs also evaluate to that output.
+Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair} if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
 It is important to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
@@
 The system described here has some differences to our Coq formalisation.
+First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Goto~label of any instruction.
+This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
+This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
+Ideally, a future version of the formalisation would have this change.
+Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
+Instead we sequentially evaluate both processes separately with the same input values and outputs.
+Finally, we have not implemented \ti{fuseNest} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.
+
+Despite these differences, we believe the current formalisation gives sufficient confidence in correctness.

### 2017-01-13 2017mergingmerges/section/XX-Related.tex
prose

@@
 
-While it is not explicitly mentioned in the paper, Kiselyov et al~\cite{kiselyov2016stream}'s {\bf strymonas} system seems to only allow streams to be used once.
+Existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
+Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
+Pull can thus express @merge@, but not @partition@.
+Push fusion supports splits (@partition@), but not joins (@merge@).
+Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections.
+However this requires manual plumbing.
+
+Another way to characterise fusion systems is whether they are \emph{local} (shortcut) or \emph{global} fusion.
+Local fusion systems such as stream fusion~\cite{coutts2007stream} use local rewrite rules to remove intermediate arrays, and tend to rely on general purpose compiler optimisations to expose opportunities for these rewrites to occur.
+Generally, the local rewrite rule can only be applied once the producer has been inlined into the consumer.
+Because of the local nature of these rewrites and the reliance on inlining, local fusion can only perform producer-consumer fusion with a single consumer, as inlining the producer into multiple consumers would duplicate work.
+Local fusion is fragile because heuristics are used to determine whether inlining occurs~\cite{lippmeier2013data}.
+This fragility poses serious problems to programmers who require fusion for adequate performance, as one must have deep knowledge of the internal compiler phases and library design in order to predict whether fusion will occur -- or worse yet, they must scour the intermediate code to count the number of loops produced.
+In contrast, global fusion systems such as those traditionally used in imperative compilers make use of specific optimisations which must be implemented separately, but are able to perform horizontal fusion and are less prone to the fragility of local fusion.
+
+This local/global distinction is not a hard classification, as some systems fit somewhere in between.
+For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but only performs producer-consumer fusion with a single local rule.
+While not explicitly mentioned in the paper, Kiselyov et al~\cite{kiselyov2016stream}'s {\bf strymonas} system only allows streams to be used once.
 Furthermore, the entire stream computation must be terminated with a single fold which takes a single stream.
@@
 \end{code}
+
+When stream programs are encoded as labelled transition systems, some fusion can be done by computing the synchronised product of two transition systems, but this suffers deadlock when the two programs share multiple inputs and read them in different orders.
+By annotating transition states with the status of each input channel, our fusion algorithm works for these cases where synchronised product does not.
 

### 2017-01-13 2017mergingmerges/code/src/Machine/Base.hs
code: start on fusion rules

@@ -22,7 +22,8 @@ data Var
 
 -- | Code label.
 data Label      
-        = Label Name
+        = Label         Name
+        | LabelJoint    (Label, Map Channel InputMode) (Label, Map Channel InputMode)
         deriving (Show, Eq)
         
 
@@ -31,6 +32,12 @@ data Heap
         = Heap (Map Var Value)
         deriving Show
 
+
+data InputMode
+        = ModeNone
+        | ModePending
+        | ModeHave
+        deriving (Show, Eq)
 
 
 -- | Describes the state of the input buffer for each channel.

### 2017-01-13 2017mergingmerges/code/src/Machine/Fuse.hs

@@ -1,0 +1,65 @@
+
+module Machine.Fuse where
+import Machine.Base
+import Data.Map                         (Map)
+import qualified Data.Map               as Map
+
+
+data ChannelMode        
+        = ModeInputExclusive
+        | ModeInputShared
+        | ModeOutput
+        | ModeConnected
+        deriving (Show, Eq)
+
+(?) m x = Map.lookup x m
+
+tryStep 
+        :: Map Channel ChannelMode         -- ^ Structural mode of each channel.
+        -> Map Channel Var                 -- ^ Buffer variable for each channel.
+        -> (Label, Map Channel InputMode)  -- ^ Label and input channel states for first process.
+        -> (Label, Map Channel InputMode)  -- ^ Label and input channel states for second process.
+        -> Instruction                     -- ^ Instruction from first process.
+        -> Maybe Instruction               -- ^ Result instruction for first process.
+
+tryStep csMode xsBuffer
+        (label1, csState1)
+        (label2, csState2)
+        instr
+
+ = case instr of
+
+        Pull c x (Next label1' xvsUpdate)
+         -- First process has exclusive use of the input channel.
+         |  Just ModeInputExclusive <- csMode ? c
+         -> Just $ Pull c x 
+                 $ Next (LabelJoint (label1', csState1) 
+                                    (label2,  csState2))
+                        xvsUpdate
+
+         -- Input channel is used by both processes,
+         -- and there is a pending value on the channel.
+         |   (Just ModeInputShared == csMode   ? c)
+          || (Just ModeConnected   == csMode   ? c)
+         ,  Just ModePending       <- csState1 ? c
+         ,  Just xBuffer           <- Map.lookup c xsBuffer
+         -> Just $ Jump
+                 $ Next (LabelJoint (label1', Map.insert c ModeHave csState1)
+                                    (label2,  csState2))
+                        (Map.insert x (XVar xBuffer) xvsUpdate)
+
+         -- Input channel is used by both processes,
+         -- and neither has pulled a value yet.
+         |  Just ModeInputShared   <- csMode   ? c
+         ,  Just ModeNone          <- csState1 ? c
+         ,  Just ModeNone          <- csState2 ? c
+         ,  Just xBuffer           <- Map.lookup c xsBuffer
+         -> Just $ Pull c xBuffer
+                 $ Next (LabelJoint (label1, Map.insert c ModePending csState1)
+                                    (label2, Map.insert c ModePending csState2))
+                         xvsUpdate
+
+
+        _ -> Nothing
+
+

### 2017-01-14 2017mergingmerges/section/XX-FutureWork.tex
prose explanation of finite streams extension

%!TEX root = ../Main.tex
\section{Extensions and future work}
\label{s:FutureWork}

We now discuss some of the shortcomings of the system, and extensions and future work to ameliorate this.

\subsection{Finite streams}
\label{s:Finite}

The processes we have seen so far deal with infinite streams, but in practice most streams are finite.
Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output.
We have focussed on the infinite stream version because it is somewhat simpler to explain and prove, but the extensions required to support finite streams do not require substantial conceptual changes.

We now describe the extensions required to support finite streams.
We add a new @closed@ constructor to the \InputState~ to encode the end of the stream.
Once an input stream is in the closed state, it can never change to another state: it remains closed thereafter.

We modify the @pull@ instruction so that it has two output labels (like @case@).
The first label, the read branch, is executed as before when the pull succeeds and a value is read from the stream.
The second label, the close branch, is executed when the stream is closed, and no more values will ever be available.
After a pull takes the close branch, any subsequent pulls from that stream will also take the close branch.

We add two new instructions for closing output streams and disconnecting from input streams.
Closing an output stream $(@close@~\Chan~\Goto)$ is similar to pushing an end-of-file marker to all readers.
As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
After a stream has been closed, no further values can be pushed.

Disconnecting from input streams $(@disconnect@~\Chan~\Goto)$ signals that a process is no longer interested in the values of a stream.
This can be used when a process requires the first values of a stream, but does not require the whole stream.
If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
After a process disconnects from an input channel, it can no longer pull from that channel.

We also add an instruction for terminating the process (@done@).
After all input streams have been read to completion or disconnected and output streams closed, the process may execute @done@ to signal that processing is complete.

The fusion definition must be extended to deal with these new instructions.
The static input state has a @closed@ constructor added and disconnection is encoded by removal from the input state, and the \ti{tryStep} changes more or less follow the evaluation changes.
Shared and connected pulls now deal with two more possibilities in the input state: the input may be closed in which case the close branch of the pull is taken; or the other process may have disconnected in which case the pull is executed as in the non-shared non-connected case.
Connected pushes must also deal with when the other process has disconnected in which case the push is executed as if it were non-connected.
For @in1@ and @out1@ channels, the new @close@ and @disconnect@ instructions are used as normal with no coordination required.
For connected @close@, as with @push@, the receiving process must have @none@ and the next step performs the @close@ and sets the input state to @closed@.
For shared @disconnect@, the @disconnect@ is only performed after both processes have disconnected; otherwise the entry is just removed from the input state.
For connected @disconnect@, the @disconnect@ is not performed and the entry is removed from the input state.

Finally, \ti{tryStepPair} is modified so that @done@ is performed when both machines are @done@.

These modifications allow our system to fuse finite streams as well as infinite.
We have implemented an initial prototype that supports finite streams, but future work is required to prove them correct.

\subsection{Fully abstract case interpretation}

This cannot be fused because it requires an unbounded buffer.
\begin{code}
zipltgts :: [a] -> [a*a]
zipltgts as =
  let as1 = filter (<0) as
      as2 = filter (>0) as
      aas = zip as1 as2
  in  aas
\end{code}

You might think the following can be fused.
It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
So while it cannot be true that the first filter reaches @>0 = true@ case and the second filter reaches @>0 = false@, we try both combinations and treat it as unfusable.
\begin{code}
zipgts :: [a] -> [a*a]
zipgts as =
  let as1 = filter (>0) as
      as2 = filter (>0) as
      aas = zip as1 as2
  in  aas
\end{code}

### 2017-01-16 2017mergingmerges/section/XX-FutureWork.tex
non-determinism and fusion order

@@
 %!TEX root = ../Main.tex
-\section{Extensions and future work}
+\section{Conclusions and future work}
 \label{s:FutureWork}
@@
 \subsection{Fully abstract case interpretation}
+\label{s:FullyAbstractCase}
 
-This cannot be fused because it requires an unbounded buffer.
+The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
+This can cause an issue with coordination and buffering, as the processes may dynamically require only a bounded buffer, the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
+
+For example, suppose two processes have the same case condition @x > 0@.
+The fusion algorithm will try all possibilities including contradictory ones, such as where the first process has @x > 0 = true@ and the second has @x > 0 = false@.
+If any of these possibilities require an unbounded buffer, the fusion algorithm fails.
+This means that the following program which pairs all positive elements with themselves, while unlikely to be written by a human, is not able to be fused.
 \begin{code}
-zipltgts :: [a] -> [a*a]
-zipltgts as =
-  let as1 = filter (<0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-You might think the following can be fused.
-It cannot because we treat @case@ conditions as fully abstract and make no attempt to filter out impossible combinations.
-So while it cannot be true that the first filter reaches @>0 = true@ case and the second filter reaches @>0 = false@, we try both combinations and treat it as unfusable.
-\begin{code}
-zipgts :: [a] -> [a*a]
 zipgts as =
@@
 \end{code}
+
+In general, if the example above used two different predicates for each filter, the fusion system would be right to outlaw it as requiring unbounded buffers.
+If the same program is rewritten to use only a single filter and use its result twice, the program is able to be fused.
+
+A possible extension is to somehow cull these contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
+It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit some kind of failure instruction rather than failing to produce a process.
+A separate postprocessing step could then perform analyses and remove statically unreachable process states.
+After postprocessing, if any failure instructions are reachable, the fusion process would fail as before.
+
+\subsection{Non-determinism and fusion order}
+\label{s:FusionOrder}
+
+The main fusion algorithm here works on pairs of processes.
+When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
+The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
+Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process.
+In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes.
+The simplest example of this has two input streams, a function applied to both, then zipped together.
+
+\begin{code}
+zipMap as bs =
+  let as' = map (+1) as
+      bs' = map (+1) bs
+      abs = zip as' bs'
+  in  abs
+\end{code}
+
+There are three combinators here, so after converting each combinator to its process there are three orders we can fuse.
+The two main options are to fuse the two maps together and then add the zip, or to fuse the zip with one of the maps, then add the other map.
+If we start by fusing the zip with one of its maps, the zip ensures that its inputs are produced in lock-step pairs, and then adding the other map will succeed.
+However if we try to fuse the two maps together, there are many possible interleavings: the fused program could read all of @as'@ first; it could read all of @bs'@ first; it read the two in lock-step pairs; or any combination of these.
+When the zip is added, fusion will fail if the wrong interleaving was chosen.
+
+% The example above can be solved by fusing connected processes first, but it is possible to construct a connected process that still relies on the order of fusion.
+% 
+% \begin{code}
+% zipApps as bs cs =
+%   let as' = as ++ bs
+%       bs' = as ++ cs
+%       abs = zip as' bs'
+%   in  abs
+% \end{code}
+
+Our current solution to this is to try all permutations of fusing processes and use the first one that succeeds.
+A more principled solution may be to allow non-determinism in a single process by adding a non-deterministic choice instruction.
+Then when fusing two processes together, if both processes are pulling from unrelated streams, the result would be a non-deterministic choice between pulling from the first process and executing the first, or pulling from the second process and executing the second.
+In this way we could defer committing to a particular evaluation order until the required order is known.
+This may produce larger intermediate programs, but the same deterministic program could be extracted at the end.

### 2017-01-16 2017mergingmerges/section/XX-Fusion.tex
rejigging syntax etc

@@
 Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
-Fusion proceeds by taking two processes in a nest, and creating a single process that computes both.
+Fusion proceeds by taking two processes in a network, and creating a single process that computes both.
 The new process uses labels made of the product of the two input processes, as well as the static part of the input state - that is, whether the input buffer is pending after injection, has been pulled, or is empty, but without the actual injected value.
@@
 \item
-\ti{fuseNest} takes a process nest and repeatedly fuses pairs together. 
+\ti{fuseNetwork} takes a process network and repeatedly fuses pairs together. 
 It chooses arbitrary processes with shared channels and fuses them together.

### 2017-01-16 2017mergingmerges/section/XX-FutureWork.tex
rejigging syntax etc

@@
 We add two new instructions for closing output streams and disconnecting from input streams.
-Closing an output stream $(@close@~\Chan~\Goto)$ is similar to pushing an end-of-file marker to all readers.
+Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
 As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
@@
 
-Disconnecting from input streams $(@disconnect@~\Chan~\Goto)$ signals that a process is no longer interested in the values of a stream.
+Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
 This can be used when a process requires the first values of a stream, but does not require the whole stream.

### 2017-01-16 2017mergingmerges/section/XX-Processes.tex
prose
rejigging syntax etc

@@
 All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
-Combining variable update with stream instructions simplifies the fusion process~\REF, as some parts of fusion need to perform both at once.
+Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.
 
-A process nest is a set of multiple processes that can be evaluated concurrently.
+A process network is a set of multiple processes that can be evaluated concurrently.
 The intersection of all process outputs should be empty - there should be no overlap.
@@
 The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
-The goto labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
+The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
 Mention that the initial heap has the names but no values, but they could be initialised to whatever.
@@
 
-As well as a ``combinator nest'', a function comprised exclusively of process combinators.
+As well as a ``combinator network'', a function comprised exclusively of process combinators.
 The input streams are supplied as arguments, and output streams as return values.
@@
 
-It is not hard to assume that, given the process definitions and a combinator nest, we can produce a process nest.
+It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
 This is simple enough for a paragraph prose description.
@@
 
-Now we can perform some kind of fusion on this nest, resulting in one process that computes, as output, both @ys@ and @zs@.
-Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator nest did not return them.
+Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
+Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.
 
@@
 Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
-The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall nest's input streams.
+The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall network's input streams.
 For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
@@
 
-(ProcessesInject) perform injection over a process nest.
-Every process in the nest must have the value injected into it.
+(ProcessesInject) performs injection over a process network.
+Every process in the network must have the value injected into it.
 This means if multiple processes read from that stream, all input buffers for that stream must be empty.
@@
 
-The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{\phi'}$
+The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
 executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
@@
 The process shaking just shakes a single instruction and updates the process.
-Shaking a process nest chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the nest.
+Shaking a process network chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the network.
 
 (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
-The result substitution sets the variable to the pulled value, as well as any substitutions in the \Goto~ of the instruction.
+The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.
 
@@
 (Shake) unwraps a single process and evaluates the instruction.
+The instruction updates are evaluated and updated in the process heap.
 It updates the process with the new label, input state and heap.
 
-(ProcessesInternal) chooses one process from the nest and evaluates it.
-When the process evaluates with an internal message ($\tau$), the entire nest evaluates by replacing that process.
+(ProcessesInternal) chooses one process from the network and evaluates it.
+When the process evaluates with an internal message ($\tau$), the entire network evaluates by replacing that process.
 
-(ProcessesPush) chooses one process from the nest and evaluates it, where the process evaluates with a push message.
-The emitted push message is then injected into all other processes in the nest, which means they must either ignore the channel or be ready to add it to their buffer.
+(ProcessesPush) chooses one process from the network and evaluates it, where the process evaluates with a push message.
+The emitted push message is then injected into all other processes in the network, which means they must either ignore the channel or be ready to add it to their buffer.
 If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
-The entire nest then emits the same message.
+The entire network then emits the same message.
 
-The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process nest and output values are accumulated.
-The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{nest}}{\ti{streams}}{\ti{nest}'}$.
-The input map $\ti{inputs}$ contains values for the nest inputs: nest outputs are not allowed, but ignored channels can have values.
+The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process network and output values are accumulated.
+The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
+The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
 The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
-Feeding evaluates the process nest until all input values have been injected.
+Feeding evaluates the process network until all input values have been injected.
 
-Note that the result stream and nest are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
+Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
 The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
@@
 
-(FeedInternal) first recursively feeds its input accumulator and process nest, then allows the resulting nest to take an internal step.
+(FeedInternal) first recursively feeds its input accumulator and process network, then allows the resulting network to take an internal step.
 The internal step does not affect the accumulators.
@@
 
-(FeedPush) works similarly to (FeedInternal) except that the process nest emits a push message.
+(FeedPush) works similarly to (FeedInternal) except that the process network emits a push message.
 The pushed value is added to the end of the accumulator for that channel.
 
-(FeedExternal) allows inputs to be injected into the process nest.
+(FeedExternal) allows inputs to be injected into the process network.
 For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
 The recursive feed is evaluated with the last value removed from the accumulators.
-The last value is then injected into the nest, and added back to the result accumulators.
+The last value is then injected into the network, and added back to the result accumulators.

### 2017-01-16 2017mergingmerges/section/XX-Proof.tex
rejigging syntax etc

@@
 
-Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair} if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
-It is important to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
-This is because evaluation of a process nest is non-deterministic, and fusion commits to a particular evaluation order.
+Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair}: if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
+It is interesting to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
+This is because evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.
 
-The problem with commiting to an evaluation order is most easily explained with a process nest containing two infinitely pushing processes.
+The problem with commiting to an evaluation order is most easily explained with a process network containing two infinitely pushing processes.
 Process @A@ is repeatedly pushing to a stream called @X@, while process @B@ repeatedly pushes to @Y@.
-When evaluating this pair as a process nest, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.
+When evaluating this pair as a process network, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.
 
@@
 The system described here has some differences to our Coq formalisation.
-First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Goto~label of any instruction.
+First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Next~label of any instruction.
 This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
@@
 Instead we sequentially evaluate both processes separately with the same input values and outputs.
-Finally, we have not implemented \ti{fuseNest} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.
+Finally, we have not implemented \ti{fuseNetwork} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.
 

### 2017-01-16 2017mergingmerges/section/01-Introduction.tex
intro prose

@@
  = let uniques = group file1
-       union   = group (merge file1 file2)
+       merged  = merge file1 file
+       union   = group merged
    in (uniques, union)
@@
 
+Sadly, stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot fuse this example because there are multiple consumers.
+Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired fusion~\cite{lippmeier2013data} cannot fuse this because they do not support value-dependent input patterns.
+Polarised data flow fusion~\cite{lippmeier2016polarized} is able to fuse this case, but the program must be manually rewritten by categorising streams into pull or push and adding extra duplicate combinators.
+
+Our approach is to treat each combinator as a separate process with communication channels between them.
+Each process is a simple sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing an output.
+Multiple processes form a network where processes are executed concurrently.
+Fusion, then, involves taking a network of concurrent processes and choosing a particular evaluation order, condensing them to a single sequential process.
+
+The process for @uniques@ is defined by the @group@ combinator, and operates thus: first it reads from its input stream @file1@ and stores that in a local variable.
+It also keeps track of the last pulled value, and compares the just read value.
+If they are different, it pushes the new value to its output stream @uniques@ and updates the last pulled value.
+In either case, it loops back indefinitely to pull from @file1@ again.
+
+Fusion is then rather like executing two processes concurrently by taking the (synchronised) product of two state machines: either the both machines can take the same step (both pulling from @file1@, or one pushing to @merged@ while the other pulls), or one of the machines can take an independent step which isn't in the alphabet of the other while the other `rests'.
+Deadlock can occur with synchronous product when each machine is waiting on the other, and neither can progress.
+We solve this by also storing in the state label information about which input channels are known to have values.
+This way, rather than requiring both machines to take the same step at the same time, executing a @push@ from one machine fills the buffer of the other, allowing the other machine to read the value from the buffer at its leisure.
+In order to push, the other machine must have an empty buffer.
+In this way we have extended synchronised product to allow more cases without deadlock.
+
 We make the following contributions:

### 2017-01-16 2017mergingmerges/section/XX-Related.tex
intro prose

@@
 \subsection{The problem with synchronised product}
+\label{s:Synchro}
 Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.

### 2017-01-17 2017mergingmerges/section/01-Introduction.tex
mostly rejigging

@@
 Can we perform both of these tasks at once, without reading the input files multiple times?
+\TODO{More compelling justification. Why would we want to do this?}
 
+\TODO{Should this example also have readFiles?}
 \begin{code}
-uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
- = let uniques = group file1
-       merged  = merge file1 file
-       union   = group merged
-   in (uniques, union)
+uniquesUnion file1 file2
+ = do   let uniques = group file1
+        let merged  = merge file1 file2
+        let union   = group merged
+        writeFile "uniques.txt" uniques
+        writeFile "union.txt"   unions
 \end{code}
+
+% This example is somewhat incorrect because the streams here are pull streams,
+% but the fused result cannot return pull streams (since we don't know what order they will be pulled)
+% \begin{code}
+% uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
+%  = let uniques = group file1
+%        merged  = merge file1 file2
+%        union   = group merged
+%    in (uniques, union)
+% \end{code}
 
@@
 It is important to note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
+
+In order to fuse this example, we require both pull \emph{and} push streams.
+The input streams must be pull streams since the order values are required is determined by the @merge@ combinator.
+For the same reason, the outputs sent to each @writeFile@ must be push streams.
 
@@
 
+\TODO{Should this be a concrete example of how they are fused?}
+\TODO{Is the comparison to synchronised product helpful or just going to cause more confusion?}
 Fusion is then rather like executing two processes concurrently by taking the (synchronised) product of two state machines: either the both machines can take the same step (both pulling from @file1@, or one pushing to @merged@ while the other pulls), or one of the machines can take an independent step which isn't in the alphabet of the other while the other `rests'.
@@
 \item and show that our calculus is general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
+  \TODO{Is this list of combinators helpful or just boring?}
 \end{itemize}

### 2017-01-17 2017mergingmerges/section/A1-Combinators.tex
mostly rejigging

@@
 We start with simple combinators supported by most streaming systems, and progress to more interesting combinators.
-Some combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but can be implemented in our finite stream extension described in~\S\REF.
+Some combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but can be implemented in our finite stream extension described in~\S\ref{s:Finite}.
 The fact that segmented versions of these combinators can be implemented is compelling evidence of this.
@@
 
+\paragraph{Group} removes filters out consecutive duplicates from a stream.
+If the stream is sorted, the output will contain only the unique entries in the stream.
+
+\begin{code}
+group = process
+     ins: as
+    outs: bs
+    heap: {last, val}
+   label: I0
+  blocks: I0 = pull as val I1
+          I1 = push bs val I2[last = val]
+          I2 = drop as     R0
+
+          R0 = pull as val R1
+          R1 = case (last /= val) R2 R3
+          R2 = push bs val R3[last = val]
+          R3 = drop as     R0
+\end{code}
+
 \paragraph{Merge} takes two sorted inputs and produces a sorted input containing both inputs.
@@
     heap: {a, b}
-   label: pI0
-  blocks: pI0 = pull as a  pI1
-          pI1 = pull bs b  pC0
-          pC0 = case (a < b) pA0 pB0
+   label: I0
+  blocks: I0 = pull as a  I1
+          I1 = pull bs b  C0
+          C0 = case (a < b) A0 B0
 
-          pA0 = push abs a pA1
-          pA1 = drop as    pA2
-          pA2 = pull as  a pC0
+          A0 = push abs a A1
+          A1 = drop as    A2
+          A2 = pull as  a C0
 
-          pB0 = push abs b pB1
-          pB1 = drop bs    pB2
-          pB2 = pull bs  b pC0
+          B0 = push abs b B1
+          B1 = drop bs    B2
+          B2 = pull bs  b C0
 \end{code}

### 2017-01-17 2017mergingmerges/section/XX-Processes.tex
mostly rejigging

@@
 A process defines a stream computation, taking any number of input streams and producing at least one output stream.
-The process name is just for fun.
+Processes have an optional name in parentheses used only for decsriptive purposes, and does not need to be unique.
 The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
@@
 \begin{code}
-map f = process
-    name: (map f)
+map f = process (map f)
      ins: as
@@
 \begin{code}
-process
-    name: (map f)
+process (map f)
      ins: xs
@@
           p2 = drop xs       p0
-process
-    name: (map g)
+process (map g)
      ins: ys
@@
 \begin{code}
-process
-    name: (map f / map g)
+process (map f / map g)
      ins: xs

### 2017-01-18 2017mergingmerges/section/01-Introduction.tex
introduction

--- 2017-01-13 2017mergingmerges/section/01-Introduction.tex
@@
 Can we perform both of these tasks at once, without reading the input files multiple times?
-
+\TODO{More compelling justification. Why would someone want to do this?}
+\TODO{Should this example also have readFiles?}
 \begin{code}
-uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
- = let uniques = group file1
-       union   = group (merge file1 file2)
-   in (uniques, union)
+uniquesUnion file1 file2
+ = do   let uniques = group file1
+        let merged  = merge file1 file2
+        let union   = group merged
+        writeFile "uniques.txt" uniques
+        writeFile "union.txt"   unions
 \end{code}
 
-In this example the @group@ combinator removes consecutive duplicates leaving only the first value, while @merge@ concatenates two sorted inputs such that the output remains sorted.
+% This example is somewhat incorrect because the streams here are pull streams,
+% but the fused result cannot return pull streams (since we don't know what order they will be pulled)
+% \begin{code}
+% uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
+%  = let uniques = group file1
+%        merged  = merge file1 file2
+%        union   = group merged
+%    in (uniques, union)
+% \end{code}
+
+In this example the @group@ combinator filters out consecutive duplicates, while @merge@ concatenates two sorted inputs such that the output remains sorted.
 There are a few interesting things about this example.
@@
 If we wish to ensure that each value is only read from the file once, we must coordinate between the two use sites: when @uniques@ requires a new value it must ensure that @union@ is ready to receive a new value, and vice versa.
-It is important to note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
+Note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
+
+In order to fuse this example, we require both pull \emph{and} push streams.
+The input streams must be pull streams since the order values are required is determined by the @merge@ combinator.
+For the same reason, the outputs sent to each @writeFile@ must be push streams.
+
+Sadly, stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot fuse this example because there are multiple consumers.
+Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired fusion~\cite{lippmeier2013data} cannot fuse this because they do not support value-dependent input patterns.
+Polarised data flow fusion~\cite{lippmeier2016polarized} is able to fuse this case, but the program must be rewritten with explicit push/pull annotations and duplication nodes.
+
+Our approach is to treat each combinator as a separate process with communication channels between them.
+Each process is a very restricted sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing an output.
+Multiple processes form a network where processes are executed concurrently.
+Fusion takes a concurrent network and \emph{sequentialises} it into a single process by choosing a particular evaluation order that requires no buffering.
+
+The example above has three combinators, so the process network has three processes.
+The two @writeFile@s outputs are treated as sinks that values can be pushed to at any time, and are not converted to processes.
+During code generation, any output values from the @uniques@ and @union@ streams are sent to the corresponding @writeFile@ sink, but we do not address code generation in this paper.
+
+The process for @uniques@ is defined by the @group@ combinator, and can be thought of as an imperative loop: first it reads from its input stream @file1@ and stores that in a local variable.
+It also keeps track of the last pulled value, and compares that against the newly read value.
+If they are different, it pushes the new value to its output stream @uniques@.
+In either case, it updates the last pulled value and loops back to the start to pull from @file1@ again.
+
+The process for @merged@ is defined by the @merge@ combinator, which starts by reading from both @file1@ and @file2@ and storing these in local variables.
+It then compares its two values to see which is the smaller.
+If the value from @file1@ is smaller, it pushes that value and pulls a new value from @file1@, otherwise it pushes the value from @file2@ and pulls from @file2@.
+This is performed in a loop.
+
+We fuse these two processes by interleaving the two such that the shared input @file1@ is only pulled from when both processes agree.
+The new process pulls from @file1@, which is copied to the variables for both processes.
+The @uniques@ process now has all it needs to execute, so it checks the value against the last pulled value, pushes if necessary, and goes back to try to pull from @file1@ again.
+At this stage the @merged@ process still has a value from @file1@ that it has pulled but not used, so @uniques@ cannot pull from @file1@ again.
+We now let @merged@ run, pulling from @file2@ and checking which is smaller.
+If the value from @file1@ is smaller, the value is emitted and @merged@ wishes to pull a new value from @file1@.
+Both processes now agree on pulling from @file1@ again, so the new value is pulled and @uniques@ can run again.
+Otherwise if the value from @file1@ is not smaller, the value from @file2@ is emitted and @merged@ pulls from @file2@ with no coordination required.
 

### 2017-01-25 2017mergingmerges/section/01-Introduction.tex
processes eval etc

@@
 Sadly, stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot fuse this example because there are multiple consumers.
-Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired fusion~\cite{lippmeier2013data} cannot fuse this because they do not support value-dependent input patterns.
-Polarised data flow fusion~\cite{lippmeier2016polarized} is able to fuse this case, but the program must be rewritten with explicit push/pull annotations and duplication nodes.
+Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired fusion~\cite{lippmeier2013data} cannot fuse this because they do not support value-dependent input patterns, and furthermore data flow fusion only supports a restricted set of baked-in combinators.
+Polarised data flow fusion~\cite{lippmeier2016polarized} is able to fuse this case, but requires the program to be rewritten with explicitly push/pull annotations on each stream.
 
@@
 \item an informal description of the required extensions to support finite streams (\S\ref{s:Finite});
-\item and show that our calculus is general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
-  \TODO{Is this list of combinators helpful or just boring?}
+\item and show our processes are general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
 \end{itemize}

### 2017-01-25 2017mergingmerges/section/A1-Combinators.tex
processes eval etc

@@
     outs: bs
-    heap: {last, val}
-   label: I0
-  blocks: I0 = pull as val I1
-          I1 = push bs val I2[last = val]
-          I2 = drop as     R0
-
-          R0 = pull as val R1
-          R1 = case (last /= val) R2 R3
-          R2 = push bs val R3[last = val]
+    heap: {first = True, last, val}
+   label: R0
+  blocks: R0 = pull as val R1
+          R1 = case (first || last /= val) R2 R3
+          R2 = push bs val R3[last = val, first = False]
           R3 = drop as     R0

### 2017-01-25 2017mergingmerges/section/XX-Processes.tex
introduction
processes eval etc

--- 2017-01-16 2017mergingmerges/section/XX-Processes.tex
@@
 
+This section describes processes in detail.
+We start with an informal explanation of processes before formally defining the syntax and dynamic semantics.
+
+A process defines a stream computation, taking any number of input streams and producing at least one output stream.
+Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
+It takes no inputs, and produces one output stream @xs@.
+Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
+This variable will be incremented and pushed.
+Each process also has a current label, which denotes the instruction to perform next.
+The initial label for @iota@ is @L0@.
+Each process has a mapping from labels to instructions.
+In this case we have two instructions, @L0@ and @L1@.
+The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move to label @L1@.
+Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
+
+\begin{code}
+process (iota)
+     ins: 
+    outs: xs
+    heap: {i = 0}
+   label: L0
+  blocks: L0 = push xs i  L1
+          L1 = jump       L0{i = i + 1}
+\end{code}
+
+When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+
+The @uniques = group file1@ is a more interesting example.
+Recall that @group@ removes consecutive duplicates from its input stream.
+It has one input, @file1@, and one output, @uniques@.
+There are three variables in the heap: @first@ is initialised to @True@ as the first pulled value has no last value to compare against; @value@ stores the most recently pulled value, and @last@ stores the last pulled value.
+The @last@ and @value@ variables are initialised to $0$, but could be initialised to any value: these initial values will not be used.
+The initial label is @L0@, which pulls from the input @file1@. This blocks waiting for an input value, and when one is received, it is stored in @value@ and the process moves to @L1@.
+Instruction @L1@ performs a case analysis on a boolean: if it is the first read value, or the last value is not equal to the most recent value, it jumps to @L2@; otherwise it jumps to @L3@.
+Instruction @L2@ pushes the most recent value to the output and jumps to @L3@, updating the last value to the most recent value, and setting first to @False@.
+Finally, the instruction for @L3@ `drops' the recently pulled value from @file1@ and jumps back to @L0@.
+This dropping is required to coordinate between multiple processes that read from the same input: now that the read value from @file1@ has been dropped, another process is free to pull the next value from @file1@ if it so wishes.
+
+\begin{code}
+process (group)
+     ins: file1
+    outs: uniques
+    heap: {first = True, last = 0, value = 0}
+   label: L0
+  blocks: L0 = pull file1   value              L1
+          L1 = case (first || last /= value)   L2               L3
+          L2 = push uniques value              L3{last = value, first = False}
+          L3 = drop file1                      L0
+\end{code}
+
+When executed on an input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
+
+Finally, we will look at the process for @merge@.
+Merge takes two input streams, @file1@ and @file2@, and produces the output @merged@.
+The two heap variables @a@ and @b@ are for storing the currently read values from both inputs.
+Again, they are initialised to $0$, but the initial values will not be used.
+Merge requires more complicated control flow than before.
+Instructions @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
+Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
+Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
+Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.
+
+\begin{code}
+process (merge)
+     ins: file1 file2
+    outs: merged
+    heap: {a = 0, b = 0}
+   label: L0
+  blocks: L0 = pull file1 a  L1
+          L1 = pull file2 b  C0
+
+          C0 = case (a < b)  A0     B0
+
+          A0 = push merged a A1
+          A1 = drop file1    A2
+          A2 = pull file1  a C0
+
+          B0 = push merged b B1
+          B1 = drop file2    B2
+          B2 = pull file2  b C0
+\end{code}
+
+%% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
 \input{figures/ProcessDef.tex}
+
+Because this process is written for infinite streams, executing on finite inputs leads to strange results.
+Executing this process with the inputs $[1, 4]$ and $[2, 3, 100]$ will produce the first four elements $[1, 2, 3, 4]$ as expected, but note that the $100$ is missing.
+This is because after the $4$ is pushed, the process tries to pull from @file1@ again, blocking until further input is available.
+We address finite streams in~\S\ref{s:Finite}.
+
+We are now ready to take our @group@ and @merge@ processes and fuse them together. 
+We create a new process with inputs @file1@ and @file2@ and outputs @uniques@ and @merged@.
+In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
+
+\begin{code}
+process (group/merge)
+     ins: file1 file2
+    outs: uniques merged
+    heap: {first = True, last, value_a, b}
+   label: L0
+  blocks: L0 = pull file1   value_a             L1
+          L1 = case (first || last /= value_a)  L2    L3
+          L2 = push uniques value_a             L3{last = value_a, first = False}
+          L3 = pull file2 b                     C0
+
+          C0 = case (value_a < b)               A0    B0
+
+          A0 = push merged  value_a             A1
+          A1 = drop file1                       A2
+          A2 = pull file1   value_a             A3
+          A3 = case (first || last /= value_a)  A4    C0
+          A4 = push uniques value_a             C0{last = value_a, first=False}
+
+          B0 = push merged b                    B1
+          B1 = drop file2                       B2
+          B2 = pull file2  b                    C0
+\end{code}
+
+\subsection{Process definitions}
 
@@
 
+\TODO{Settle on a terminology for channels / streams. Are both necessary?}
+Streams are the abstract data flowing through while channels are particular endpoints, so a stream can have multiple channels.
+A stream can have at most one output channel, and any number of input channels.
+
 A process defines a stream computation, taking any number of input streams and producing at least one output stream.
-The process name is just for fun.
+Processes have an optional name in parentheses used only for descriptive purposes, and does not need to be unique.
 The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
@@
 
-\subsection{Map/map}
-\label{s:Process:MapMap}
-
-One of the simplest combinators is @map@.
-This might need to go elsewhere.
-As well as needing a better example than map/map.
-Let's start with the process definition for @map@.
-The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
-The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
-Mention that the initial heap has the names but no values, but they could be initialised to whatever.
-It doesn't matter since they'll be written before anything is read.
-
-\begin{code}
-map f = process
-    name: (map f)
-     ins: as
-    outs: bs
-    heap: {a = 0}
-   label: map0
-  blocks: map0 = pull as    a  map1
-          map1 = push bs (f a) map2
-          map2 = drop as       map0
-\end{code}
-
-As well as a ``combinator network'', a function comprised exclusively of process combinators.
-The input streams are supplied as arguments, and output streams as return values.
-\begin{code}
-mapMap f g xs
- = let ys = map f xs
-       zs = map g ys
-   in  zs
-\end{code}
-
-It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
-This is simple enough for a paragraph prose description.
-It's just a bit of inlining and renaming everything to be unique.
-
-\begin{code}
-process
-    name: (map f)
-     ins: xs
-    outs: ys
-    heap: {x}
-   label: p0
-  blocks: p0 = pull xs    x  p1
-          p1 = push ys (f x) p2
-          p2 = drop xs       p0
-process
-    name: (map g)
-     ins: ys
-    outs: zs
-    heap: {y}
-   label: q0
-  blocks: q0 = pull ys    y  q1
-          q1 = push zs (g y) q2
-          q2 = drop ys       q0
-\end{code}
-
-Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
-Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.
-
-\begin{code}
-process
-    name: (map f / map g)
-     ins: xs
-    outs: ys zs
-    heap: {x, y, _ys}
-   label: p0q0
-  blocks: p0q0            = pull xs    x  p1q0
-          p1q0            = push ys (f x) p2q0-pending-ys { _ys = f x }
-          p2q0-pending-ys = drop xs       p0q0-pending-ys
-          p0q0-pending-ys = jump          p0q1-have-ys    { y = _ys }
-          p0q1-have-ys    = push zs (g y) p0q2-have-ys
-          p0q2-have-ys    = jump          p0q0
-\end{code}
+% \subsection{Map/map}
+% \label{s:Process:MapMap}
+% 
+% One of the simplest combinators is @map@.
+% This might need to go elsewhere.
+% As well as needing a better example than map/map.
+% Let's start with the process definition for @map@.
+% The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
+% The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
+% Mention that the initial heap has the names but no values, but they could be initialised to whatever.
+% It doesn't matter since they'll be written before anything is read.
+% 
+% \begin{code}
+% map f = process (map f)
+%      ins: as
+%     outs: bs
+%     heap: {a = 0}
+%    label: map0
+%   blocks: map0 = pull as    a  map1
+%           map1 = push bs (f a) map2
+%           map2 = drop as       map0
+% \end{code}
+% 
+% As well as a ``combinator network'', a function comprised exclusively of process combinators.
+% The input streams are supplied as arguments, and output streams as return values.
+% \begin{code}
+% mapMap f g xs
+%  = let ys = map f xs
+%        zs = map g ys
+%    in  zs
+% \end{code}
+% 
+% It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
+% This is simple enough for a paragraph prose description.
+% It's just a bit of inlining and renaming everything to be unique.
+% 
+% \begin{code}
+% process (map f)
+%      ins: xs
+%     outs: ys
+%     heap: {x}
+%    label: p0
+%   blocks: p0 = pull xs    x  p1
+%           p1 = push ys (f x) p2
+%           p2 = drop xs       p0
+% process (map g)
+%      ins: ys
+%     outs: zs
+%     heap: {y}
+%    label: q0
+%   blocks: q0 = pull ys    y  q1
+%           q1 = push zs (g y) q2
+%           q2 = drop ys       q0
+% \end{code}
+% 
+% Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
+% Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.
+% 
+% \begin{code}
+% process (map f / map g)
+%      ins: xs
+%     outs: ys zs
+%     heap: {x, y, _ys}
+%    label: p0q0
+%   blocks: p0q0            = pull xs    x  p1q0
+%           p1q0            = push ys (f x) p2q0-pending-ys { _ys = f x }
+%           p2q0-pending-ys = drop xs       p0q0-pending-ys
+%           p0q0-pending-ys = jump          p0q1-have-ys    { y = _ys }
+%           p0q1-have-ys    = push zs (g y) p0q2-have-ys
+%           p0q2-have-ys    = jump          p0q0
+% \end{code}
 
@@
 
-We use the ``inject and shake'' method of evaluation. (This isn't a real thing.)
+We now describe evaluation of processes and process networks.
+We split evaluation into three main parts:
+\begin{itemize}
+\item Injection, where values are inserted into a process' input buffer.
+Injection is only possible when the process input buffer is empty.
+\item Shaking, where a process takes a step from one label to another.
+Shaking a process results in a new process as well as an output message.
+If a process pushes to a stream, the push value must be able to be injected to other processes.
+\item Feeding, where an environment of input values are fed to the processes, and output values are collected.
+This is the `top level' of evaluation that uses both injection and shaking.
+\end{itemize}
+
+Evaluation of a process network is non-deterministic, in that at any point there are many possible processes that can take a step.
+However, because each process itself is deterministic and has blocking reads, overall evaluation should be deterministic as per Kahn process networks.
+That is: the order in which values are pushed to different output streams is not deterministic, but the order and values for a particular output stream \emph{are} deterministic.
+
+Note that while process nests evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
 

### 2017-01-27 2017mergingmerges/section/XX-Processes.tex
some things

@@
 
-Because this process is written for infinite streams, executing on finite inputs leads to strange results.
-Executing this process with the inputs $[1, 4]$ and $[2, 3, 100]$ will produce the first four elements $[1, 2, 3, 4]$ as expected, but note that the $100$ is missing.
-This is because after the $4$ is pushed, the process tries to pull from @file1@ again, blocking until further input is available.
+As this process merges infinite streams, evaluating with finite streams will block waiting for more input once the end of either stream is reached.
+This is not an issue for infinite streams, and we use examples involving finite streams are much easier to describe.
 We address finite streams in~\S\ref{s:Finite}.
+Executing this process with the inputs $[1, 4]$ and $[2, 3, 100]$ will produce the first four elements $[1, 2, 3, 4]$ as expected, but note that the $100$ is missing because the process is blocked waiting for the next value from @file1@.
+\amos{I don't think this is necessary}
 
@@
 In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
+Rather than explaining the process in detail, we simply note that the fused process is an interleaved scheduling of the two, starting with executing @group@, then executing @merge@, and going back to @group@ whenever pulling from @file1@ again.
+The labels @L0@-@L2@ and @A1@-@A4@ are much the same as those in @group@, while @C0@, @A0@-@A2@ and @B0@-@B2@ are similar to those in @merge@.
 
@@
 
-\TODO{Settle on a terminology for channels / streams. Are both necessary?}
+\amos{Settle on a terminology for channels / streams. Are both necessary?}
 Streams are the abstract data flowing through while channels are particular endpoints, so a stream can have multiple channels.
@@
 Evaluation of a process network is non-deterministic, in that at any point there are many possible processes that can take a step.
-However, because each process itself is deterministic and has blocking reads, overall evaluation should be deterministic as per Kahn process networks.
+However, because each process itself is deterministic and has blocking reads, overall evaluation is deterministic as per Kahn process networks.
 That is: the order in which values are pushed to different output streams is not deterministic, but the order and values for a particular output stream \emph{are} deterministic.
 
-Note that while process nests evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
+Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
 
@@
 The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall network's input streams.
-For the map/map example, the values for @xs@ would be injected externally, but the values for @ys@ will be injected into the second process from the first process.
 Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.
@@
 
-Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
-The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
+% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
+% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
 
@@
 The internal step does not affect the accumulators.
-The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.
+% The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.
 

### 2017-01-28 2017mergingmerges/section/XX-Fusion.tex
some things
fusion prose

--- 2017-01-04 2017mergingmerges/section/XX-Fusion.tex
@@
 
+The core fusion algorithm works by constructing a static schedule for a pair of processes, which can be repeated over the whole network until only one process remains.
+The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
 Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
-Fusion proceeds by taking two processes in a nest, and creating a single process that computes both.
-The new process uses labels made of the product of the two input processes, as well as the static part of the input state - that is, whether the input buffer is pending after injection, has been pulled, or is empty, but without the actual injected value.
-
-$\InputState_1$ defines the statis input state.
-$\Label_1$ is a single process label with the set of input states.
-We add the constructor $@label@'$ to the $\Label$ type which is the pair of process labels with input states.
+Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
+The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
+If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.
 
@@
 Finally, when one process writes and the other ignores (@out1@) no coordination is necessary.
-As explained in~\S\ref{s:Process:Eval}, it is not allowed for two processes to write to the same channel.
+Recall that each output channel is uniquely owned and cannot be written by another process, so it is not possible for both processes to write to the same output.
 
-Before looking at the definitions, it is worth noting the types and purposes of important functions:
-\begin{itemize}
-\item
-\ti{fuseNest} takes a process nest and repeatedly fuses pairs together. 
-It chooses arbitrary processes with shared channels and fuses them together.
+Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
+\ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
+The order processes are fused can affect whether fusion succeeds or fails, so all permutations are tried.
+We address this in more detail in~\S\ref{s:FusionOrder}.
 
-\item
-\ti{fusePair} fuses a pair of processes together.
-The heap variables must be distinct (this can be ensured by renaming).
-The two processes must have some shared channels; otherwise no coordination is required at all, and the fusion process will choose an arbitrary interleaving of the processes.
-This restriction is not an issue in practice since there is no benefit to fusing unrelated processes.
-\end{itemize}
+\ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
+The blocks are constructed by computing a fixpoint, starting at the initial labels of each process with empty input states.
+Each block is computed with \ti{tryStepPair}, which statically chooses one of the two processes to execute, and any reachable blocks are added recursively until fixpoint is reached.
+This implementation assumes the heap variables are distinct, which can easily be ensured by renaming.
 
-Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions.
+Figure~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
+It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.
+The \ti{tryStepPair} function defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
+If both processes can execute, \ti{tryStepPair} uses a simple priority heuristic of deferring pulls: this is because pulls may block, while other actions are more likely to produce immediate results.
 
-Figures~\ref{fig:Fusion:Def:Step} and~\ref{fig:Fusion:Def:StepPair} define the judgment rules for fusing a single instruction at a time.
+\ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
+If the process is pulling or pushing a non-shared channel categorised as @in1@ or @out1@, no coordination is required and injection is ignored.
+For shared and connected channels, the process can only pull or push if the other process would be willing to accept the injection at the same time.
+Simple instructions such as @case@ and @jump@ require no coordination and can be executed at any point.
 

### 2017-01-29 2017mergingmerges/code/example/MapMap.hs
code: adding fn to get channel modes

@@ -2,6 +2,7 @@
 module MapMap where
 import Machine.Base
 import Machine.Execute
+import Machine.Fuse
 import Control.Monad.State.Strict
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
@@ -12,8 +13,8 @@ inputs
  = Map.fromList
  [ (Channel "as", [VInt 1, VInt 2, VInt 3]) ]
 
-map1 :: Process
-map1    
+pMap1 :: Process
+pMap1
  = Process
  { processName          = "map1"
  , processIns           = Set.fromList [Channel "as"]
@@ -37,8 +38,8 @@ map1
  }
 
 
-map2 :: Process
-map2    
+pMap2 :: Process
+pMap2   
  = Process
  { processName          = "map2"
  , processIns           = Set.fromList [Channel "bs"]
@@ -60,3 +61,4 @@ map2
                  (Next (Label "map20") Map.empty))
         ]
  }
+

### 2017-01-29 2017mergingmerges/code/src/Machine/Fuse.hs

@@ -1,24 +1,60 @@
-
+{-# LANGUAGE TupleSections #-}
 module Machine.Fuse where
 import Machine.Base
-import Data.Map                         (Map)
-import qualified Data.Map               as Map
+import Data.Map                 (Map)
+import Data.Set                 (Set)
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
 
 
 data ChannelMode        
         = ModeInputExclusive
         | ModeInputShared
-        | ModeOutput
         | ModeConnected
-        deriving (Show, Eq)
+        | ModeOutput
+        deriving (Show, Eq, Ord)
 
 (?) m x = Map.lookup x m
 
+
 ---------------------------------------------------------------------------------------------------
-channelModes
-        :: Process
-        -> Process
-        -> 
+-- | Yield the set of channel usage modes relative to the given pair of processes.
+processChannelModes :: Process -> Process -> Set (Channel, ChannelMode)
+processChannelModes process1 process2
+ = let  ins1    = processIns  process1
+        ins2    = processIns  process2
+        outs1   = processOuts process1
+        outs2   = processOuts process2
+   in Set.unions
+        [ Set.map (,ModeInputShared)    
+                $ Set.intersection ins1 ins2
+
+        , Set.map (,ModeInputExclusive)
+                $ Set.filter (\c -> not $ Set.member c (Set.union outs1 outs2))
+                $ Set.union ins1 ins2
+
+        , Set.map (,ModeConnected)
+                $ Set.intersection
+                        (Set.union ins1  ins2)
+                        (Set.union outs1 outs2)
+
+        , Set.map (,ModeOutput)
+                $ Set.filter (\c -> not $ Set.member c (Set.union ins1 ins2))
+                $ Set.union outs1 outs2
+        ]
+
+
+-- | Check whether two processes are directly connected via any channel.
+processesAreConnected :: Process -> Process -> Bool
+processesAreConnected process1 process2
+ = let  ins1    = processIns  process1
+        ins2    = processIns  process2
+        outs1   = processOuts process1
+        outs2   = processOuts process2
+   in not $ Set.null
+          $ Set.intersection
+               (Set.union ins1 ins2)
+               (Set.union ins1 outs2)
 
 
 

### 2017-01-29 2017mergingmerges/section/01-Introduction.tex
some things
Work through introduction

--- 2017-01-17 2017mergingmerges/section/01-Introduction.tex
@@
 
-% Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
-% However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
-% Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-% For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+Suppose we have two large arrays containing textual identifiers, and wish to perform some analysis on them. Both arrays are sorted, but may contain duplicate identifiers. 
+We wish to retrieve the unique set of identifiers from the first array, as well as produce the union of both ararys.
+Can we perform both of these tasks at once, without reading the input arrays multiple times?
+\begin{code}
+  uniquesUnion : Array Text -> Array Text -> (Array Text, Array Text)
+  uniquesUnion arr1 arr2
+   = let  uniques1 = group arr1
+          merged   = merge arr1 arr2
+          union    = group merged
+     in   (uniques1, union)
+\end{code}
 
-Suppose we have two large files containing identifiers, and wish to perform some analysis on them.
-Both files are sorted, but may contain duplicates.
-We wish to retrieve the unique set from the first file, as well as the union of both files.
-Can we perform both of these tasks at once, without reading the input files multiple times?
-\TODO{More compelling justification. Why would we want to do this?}
-
-\TODO{Should this example also have readFiles?}
-\begin{code}
-uniquesUnion file1 file2
- = do   let uniques = group file1
-        let merged  = merge file1 file2
-        let union   = group merged
-        writeFile "uniques.txt" uniques
-        writeFile "union.txt"   unions
-\end{code}
+% \amos{More compelling justification. Why would someone want to do this?}
+% \amos{Should this example also have readFiles?}
 
@@
 
-In this example the @group@ combinator removes consecutive duplicates leaving only the first value, while @merge@ concatenates two sorted inputs such that the output remains sorted.
+In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays such that the output remains sorted.
 There are a few interesting things about this example.
-First, the access pattern of @merge@ is \emph{value-dependent}: the order in which @merge@ reads values from @file1@ or @file2@ depends on the values in the files.
-If all the values in @file1@ are smaller than the values in @file2@, then @merge@ will pull values from @file1@ before pulling from @file2@, and vice versa.
-Second, note that @file1@ is used twice.
-If we wish to ensure that each value is only read from the file once, we must coordinate between the two use sites: when @uniques@ requires a new value it must ensure that @union@ is ready to receive a new value, and vice versa.
-It is important to note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
+Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ reads values from @arr1@ or @arr2@ depends on the values in the arrays themselves.
+If all the values in @arr1@ are smaller than the values in @arr2@, then @merge@ will pull values from @arr1@ before pulling from @arr2@, and vice versa.
+Secondly, although @arr1@ occurs twice in the program, at runtime we only want to read through the elements of this array once.
+To achieve this the compiled program must coordinate between the two uses of @arr1@, so when the implementation of @uniques@ requires a new value the implementation of @union@ is also ready to receive a new value, and vice versa.
+Although the program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.
 
-In order to fuse this example, we require both pull \emph{and} push streams.
-The input streams must be pull streams since the order values are required is determined by the @merge@ combinator.
-For the same reason, the outputs sent to each @writeFile@ must be push streams.
+The above example is seeming innocuous, but existing fusion transformations cannot handle it.
+Stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot avoid performing duplicate memory access when a particular array or intermediate result is used by multiple consumers.
+We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
+Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired array fusion~\cite{lippmeier2013data} also cannot fuse this example because they do not support combinators with multiple inputs that have value-dependent access patterns.
+We refer to such as situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
+Polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
 
-Sadly, stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot fuse this example because there are multiple consumers.
-Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired fusion~\cite{lippmeier2013data} cannot fuse this because they do not support value-dependent input patterns.
-Polarised data flow fusion~\cite{lippmeier2016polarized} is able to fuse this case, but the program must be manually rewritten by categorising streams into pull or push and adding extra duplicate combinators.
-
-Our approach is to treat each combinator as a separate process with communication channels between them.
-Each process is a simple sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing an output.
-Multiple processes form a network where processes are executed concurrently.
-Fusion, then, involves taking a network of concurrent processes and choosing a particular evaluation order, condensing them to a single sequential process.
-
-The process for @uniques@ is defined by the @group@ combinator, and operates thus: first it reads from its input stream @file1@ and stores that in a local variable.
-It also keeps track of the last pulled value, and compares the just read value.
-If they are different, it pushes the new value to its output stream @uniques@ and updates the last pulled value.
-In either case, it loops back indefinitely to pull from @file1@ again.
-
-\TODO{Should this be a concrete example of how they are fused?}
-\TODO{Is the comparison to synchronised product helpful or just going to cause more confusion?}
-Fusion is then rather like executing two processes concurrently by taking the (synchronised) product of two state machines: either the both machines can take the same step (both pulling from @file1@, or one pushing to @merged@ while the other pulls), or one of the machines can take an independent step which isn't in the alphabet of the other while the other `rests'.
-Deadlock can occur with synchronous product when each machine is waiting on the other, and neither can progress.
-We solve this by also storing in the state label information about which input channels are known to have values.
-This way, rather than requiring both machines to take the same step at the same time, executing a @push@ from one machine fills the buffer of the other, allowing the other machine to read the value from the buffer at its leisure.
-In order to push, the other machine must have an empty buffer.
-In this way we have extended synchronised product to allow more cases without deadlock.
+In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communications channels between the processes.
+Each combinator is expressed as a restricted, sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing to an output channel.
+The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
 
@@
 \item an algorithm for fusing processes, the first which supports splits, joins and arbitrary combinators (\S\ref{s:Fusion});
-\item a formalisation and proof of soundness in Coq (\S\ref{s:Proofs});
-\item an informal description of the required extensions to support finite streams (\S\ref{s:Finite});
-\item and show that our calculus is general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
-  \TODO{Is this list of combinators helpful or just boring?}
+\item a description of the required extensions to support finite streams (\S\ref{s:Finite});
+\item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs});
+\item and show our processes are general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
 \end{itemize}
+
+\ben{Add a few more sentences on related work.
+Explain how this work extends the old flow fusion paper.
+It is not short-cut fusion like Oleg's recent work.
+We are not in the same space as Fortran style array fusion transformations like polyhedral}
+
+% BL: describe this later.
+% Furthermore, the data-flow fusion system of~\cite{lippmeier2013data} only deals with a fixed set of baked-in combinators.
+
+% BL: Shift the detailed description into a later section.
+% The example above has three combinators, so the process network has three processes.
+% The two @writeFile@s outputs are treated as sinks that values can be pushed to at any time, and are not converted to processes.
+% During code generation, any output values from the @uniques@ and @union@ streams are sent to the corresponding @writeFile@ sink, but we do not address code generation in this paper.
+
+% The process for @uniques@ is defined by the @group@ combinator, and can be thought of as an imperative loop: first it reads from its input stream @file1@ and stores that in a local variable.
+% It also keeps track of the last pulled value, and compares that against the newly read value.
+% If they are different, it pushes the new value to its output stream @uniques@.
+% In either case, it updates the last pulled value and loops back to the start to pull from @file1@ again.
+
+% The process for @merged@ is defined by the @merge@ combinator, which starts by reading from both @file1@ and @file2@ and storing these in local variables.
+% It then compares its two values to see which is the smaller.
+% If the value from @file1@ is smaller, it pushes that value and pulls a new value from @file1@, otherwise it pushes the value from @file2@ and pulls from @file2@.
+% This is performed in a loop.
+
+% We fuse these two processes by interleaving the two such that the shared input @file1@ is only pulled from when both processes agree.
+% The new process pulls from @file1@, which is copied to the variables for both processes.
+% The @uniques@ process now has all it needs to execute, so it checks the value against the last pulled value, pushes if necessary, and goes back to try to pull from @file1@ again.
+% At this stage the @merged@ process still has a value from @file1@ that it has pulled but not used, so @uniques@ cannot pull from @file1@ again.
+% We now let @merged@ run, pulling from @file2@ and checking which is smaller.
+% If the value from @file1@ is smaller, the value is emitted and @merged@ wishes to pull a new value from @file1@.
+% Both processes now agree on pulling from @file1@ again, so the new value is pulled and @uniques@ can run again.
+% Otherwise if the value from @file1@ is not smaller, the value from @file2@ is emitted and @merged@ pulls from @file2@ with no coordination required.
+
+% If we wish to ensure that each value is only read from the file once, we must coordinate between the two use sites: when @uniques@ requires a new value it must ensure that @union@ is ready to receive a new value, and vice versa.
+Note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
+% In order to fuse this example, we require both pull \emph{and} push streams.
+% The input streams must be pull streams since the order values are required is determined by the @merge@ combinator.
+% For the same reason, the outputs sent to each @writeFile@ must be push streams.
+
+% Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
+% However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
+% Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
+% For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
+
+% Fusion eliminates intermediate array buffers and converts pipelines of array combinators into low-level iteration based loop code. Different fusion systems can handle
+
+% When comparing fusion systems, three important criteria to consider are: whether the system supports splits, where a stream is used multiple times; whether it supports joins, where a combinator has multiple inputs; and whether arbitrary combinators such as @merge@ and segmented appends can be encoded.
+Existing fusion systems support one or two of these, but not three.
+We present a fusion system based on process calculus that supports all three: splits, joins and arbitrary combinators.
+
+% Our system has been formalised in Coq where we have proved soundness of the fusion algorithm. It is expressive enough to encode a wide range of combinators including operations on segmented arrays.
+
+% \amos{``Arbitrary combinators'' is not quite true. How can we distinguish combinators we support from 2013 data flow fusion paper? Perhaps by mentioning value-dependent input / access patterns.}
+
+% Leave this to the description of the algorithm, not the abstract.
+% We encode each combinator as a separate process with any number of input and output channels.
+Each process is sequential but multiple processes can be executed concurrently.
+We give a concurrent execution semantics for multiple processes, but these are used only as a specification for how the fused program must behave.
+The fused program itself is sequential and can easily be converted to simple imperative code.
+
+% Our fusion algorithm takes two concurrently executable processes and creates a sequential interleaving of the two such that they execute with no unbounded buffers.
+% If fusion would require unbounded buffers (or the fusion algorithm wrongly infers that it would) then fusion fails.
+% If fusion fails, the user can be presented with an error message telling them which combinators could not be fused.
+% For scenarios where fusion is required, this is a great advantage over fragile shortcut fusion systems.
+
+% BL: leave the apologies to the conclusion.
+% The version presented here deals with infinite streams, and we informally describe the extensions required to support finite streams.
+
+% BL: leave this to the main intro.
+% optimising high-level array and streaming computations, as it reduces memory traffic and intermediate arrays.
+The benefits of removing intermediate arrays are even more important as data sizes approach the size of memory.

### 2017-01-31 2017mergingmerges/section/XX-Processes.tex
Work through introduction
start on concrete expression syntax

@@
 
-This section describes processes in detail.
-We start with an informal explanation of processes before formally defining the syntax and dynamic semantics.
+% BL: Just start into the description.
+%     Readers will work out what's coming by flipping through the paper themselves.
+% This section describes processes in detail. We start with an informal explanation of processes before formally defining the syntax and dynamic semantics.
 
@@
   blocks: L0 = pull file1   value               L1
-          L1 = case (first || last /= value)   L2               L3
+          L1 = case (or first (last /= value))  L2               L3
           L2 = push uniques value               L3{last = value, first = False}
@@
   blocks: L0 = pull file1   value_a               L1
-          L1 = case (first || last /= value_a)  L2    L3
+          L1 = case (or first (last /= value_a))  L2    L3
           L2 = push uniques value_a               L3{last = value_a, first = False}
@@
           A2 = pull file1   value_a               A3
-          A3 = case (first || last /= value_a)  A4    C0
+          A3 = case (or first (last /= value_a))  A4    C0
           A4 = push uniques value_a               C0{last = value_a, first=False}

### 2017-02-02 2017mergingmerges/section/01-Introduction.tex
nothing useful

@@
 Suppose we have two large arrays containing textual identifiers, and wish to perform some analysis on them. Both arrays are sorted, but may contain duplicate identifiers. 
-We wish to retrieve the unique set of identifiers from the first array, as well as produce the union of both ararys.
+We wish to retrieve the unique set of identifiers from the first array, as well as produce the union of both arrays.
 Can we perform both of these tasks at once, without reading the input arrays multiple times?
@@
 
-The above example is seeming innocuous, but existing fusion transformations cannot handle it.
+The above example seems innocuous, but existing fusion transformations cannot handle it.
 Stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot avoid performing duplicate memory access when a particular array or intermediate result is used by multiple consumers.
 We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
+Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in the presence of splits this causes excessive inlining which duplicates work.
 Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired array fusion~\cite{lippmeier2013data} also cannot fuse this example because they do not support combinators with multiple inputs that have value-dependent access patterns.
@@
 Polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
+
+\amos{The split/join distinction is actually more about pushes vs pulls, as \emph{some} joins can be expressed with synchronous dataflow eg non-deterministic merge}
 

### 2017-02-04 2017mergingmerges/section/01-Introduction.tex
introduction

@@
 The above example seems innocuous, but existing fusion transformations cannot handle it.
-Stream fusion~\cite{coutts2007stream} and other forms of shortcut fusion~\cite{jones2001playing} cannot avoid performing duplicate memory access when a particular array or intermediate result is used by multiple consumers.
+As observed by \citet{kay2009you}, both pull- and push-based fusion have fundamental limitations.
+Pull-based systems such as stream fusion~\cite{coutts2007stream} cannot perform fusion when a particular array or intermediate result is used by multiple consumers.
 We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
-Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in the presence of splits this causes excessive inlining which duplicates work.
-Synchronous dataflow languages~\cite{mandel2010lucy} and data flow inspired array fusion~\cite{lippmeier2013data} also cannot fuse this example because they do not support combinators with multiple inputs that have value-dependent access patterns.
-We refer to such as situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
-Polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
-
-\amos{The split/join distinction is actually more about pushes vs pulls, as \emph{some} joins can be expressed with synchronous dataflow eg non-deterministic merge}
+Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in this case splits can cause excessive inlining which duplicates work.
+Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse this example because they do not support combinators with multiple inputs.
+We refer to such a situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
+Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of combinators.
+More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
+Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent combinators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
 
@@
 We are not in the same space as Fortran style array fusion transformations like polyhedral}
+\amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}
 

### 2017-02-04 2017mergingmerges/section/XX-Fusion.tex
ok

@@
 
-The core fusion algorithm works by constructing a static schedule for a pair of processes, which can be repeated over the whole network until only one process remains.
+The core fusion algorithm works by constructing a static schedule for a pair of processes, which is repeated over the whole network until only one process remains.
 The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.

### 2017-02-04 2017mergingmerges/section/XX-Processes.tex
ok

@@
 Channels, labels and variables are specified by some external, globally unique set of names.
-We assume that values and expressions are specified externally to the core calculus.
-The actual details of the external computation are not important, except that there should be an evaluation form taking a heap and an expression and producing a value, as well as some way to destruct booleans.
+For values and expressions we use a simple untyped lambda calculus with a few primitives chosen to facilitate the examples.
 
@@
 
-The heap is used for evaluation - it is used when evaluating the external value calculus.
+The heap is used for evaluation of expressions.
 Each process has its own private heap, meaning that the only communication between processes occurs by streams.
@@
 
-Pulling from a stream is blocking, and as usual with Kahn processes~\cite{kahn1976coroutines}, there is no way to tell whether a stream currently has a value.
-Pull takes as arguments the channel to read from, the variable to put the read value in, and the next label state to jump to after a successful pull.
-
-After values have been pulled, they must be disposed of with @drop@.
-This has no real evaluation semantics other than for coordination between processes.
+As usual with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
+Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
+After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer process to push to that stream.
 

### 2017-02-06 2017mergingmerges/section/XX-Fusion.tex
Fusion section prose

@@
 
-\input{figures/FusionDef.tex}
+\input{figures/Fusion-types.tex}
 
@@
 
+\input{figures/Fusion-fusePair.tex}
+
 Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
@@
 
+\input{figures/Fusion-tryStepPair.tex}
+
+The \ti{tryStepPair} function defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
+It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.
+
+Clause (DeferPull1) applies when both processes are able to take a step.
+If both processes can execute, \ti{tryStepPair} uses a simple priority heuristic of deferring pulls: this is because pulls may block, while other actions are more likely to produce immediate results.
+In this case the instruction for $p$ is \emph{not} pulling, so it can be executed.
+
+Clause (DeferPull2) is much the same as (DeferPull1) except this time the instruction for $q$ is not a pull.
+Because we are calling $\ti{tryStep}$ with the instruction from process $q$, the output labels it generates are pairs of $(q,p)$ rather than $(p,q)$.
+We use $\ti{swaplabels}$ to convert them back to $(p,q)$ and keep them consistent with the other labels.
+
+Clauses (Run1) and (Run2) apply when only one process can run, or both processes are pulls.
+If both are pulls, we make the arbitrary choice to execute $p$ with (Run1).
+For clause (Run2) we again need to swap the labels as with (DeferPull2).
+
+\input{figures/Fusion-tryStep.tex}
+
 Figure~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
 It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.
-The \ti{tryStepPair} function defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
-If both processes can execute, \ti{tryStepPair} uses a simple priority heuristic of deferring pulls: this is because pulls may block, while other actions are more likely to produce immediate results.
 
-\ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
+Function \ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
 If the process is pulling or pushing a non-shared channel categorised as @in1@ or @out1@, no coordination is required and injection is ignored.
@@
 Simple instructions such as @case@ and @jump@ require no coordination and can be executed at any point.
+
+Clause (LocalJump) applies when the process is trying to jump.
+In this case, the fused instruction simply performs the jump, leaving the other process as-is.
+When the fused instruction is evaluated, it corresponds to the same (Jump) shake rule on the input process.
+
+Clause (LocalCase) is equally simple, and the fused instruction performs the case.
+
+Clause (LocalPush) applies when the process is trying to push to a non-shared, local channel.
+In this case the push can be performed as usual, with no coordination required.
+
+Clause (SharedPush) applies when the process is trying to push to a shared channel.
+Pushing to a shared channel requires the other process to be ready to accept the injected value.
+In terms of evaluating the input processes, this corresponds to the (ProcessesPush) and (InjectValue) rules where push can only succeed if the other process' dynamic input state is ``@none@''.
+Here we encode the injection rules inside the fused process by requiring the static input state to be ``$@none@_S$''.
+When this is satisfied, the push also stores the pushed value in a local variable ``$@chan@~c$'' and sets the static input state to ``$@pending@_S$'' denoting that the value is available.
+
+Clause (LocalPull) applies when the process is trying to pull from a local channel with no coordination required.
+
+Clause (SharedPull) applies when the process is trying to pull from a shared channel that the other process either pulls from or pushes to.
+The static state of the process has a ``$@pending@_S$'' value for the channel, which corresponds to a dynamic state of ``$@pending@~(@chan@~c)$''.
+This means that one of the processes has pulled a value, or the other process has pushed a value, and in either case there is a value ready to use.
+The key observation here is that when this jump is evaluated it will use the pending value from ``$(@chan@~c)$'' and act equivalently to the shake rule (Pull) on a single process.
+
+Clause (SharedPullInject) applies when the process is trying to pull from a shared channel that both processes pull from, and neither process has a value.
+The static state of both processes are ``$@none@_S$'', so the value can be injected into both.
+We encode the injection rules for the input processes inside the fused process, so that the fused process has explicit control over injection.
+To do this, we pull from the input channel, setting both static states to ``$@pending@_S$''.
+Now the dynamic state of the fused process is ``@have@'' as it has successfully pulled, but we set the static state of the input processes to ``$@pending@_S$'' as from their perspective the value has been injected but not yet pulled.
+This step leaves both processes at their current label, allowing the next step to be (SharedPull).
+
+Clause (LocalDrop) applies when the process is trying to drop a local channel with no coordination required.
+
+Clause (ConnectedDrop) applies when dropping a value pushed by the other process.
+Because the value was not pulled through a channel but instead sent by a local variable, the channel does not need to be dropped.
+We set the static input state to ``$@none@_S$'', essentially performing the (Drop) shake rule statically.
+
+Clauses (SharedDropOne) and (SharedDropBoth) apply when dropping a value that both have pulled.
+When both processes pull from the same input, the first drop is a fake drop as with (ConnectedDrop), but the second one is a real drop.
+If the other process' static state is not ``$@none@_S$'', the other process has not dropped yet so we perform a fake drop.
+Otherwise, we perform the real drop.
+After the real drop is performed, both static states will be ``$@none@_S$'' and either process will be able to pull again.
 

### 2017-02-06 2017mergingmerges/section/XX-Fusion.tex
just citations

@@
 Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
-\ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
+Function \ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
 The order processes are fused can affect whether fusion succeeds or fails, so all permutations are tried.
@@
 
-\ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
+Function \ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
 The blocks are constructed by computing a fixpoint, starting at the initial labels of each process with empty input states.
@@
 
-The \ti{tryStepPair} function defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
+Function \ti{tryStepPair} defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
 It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.

### 2017-02-06 2017mergingmerges/section/XX-Related.tex
just citations

@@
 However this requires manual plumbing.
+
+The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}, by explicitly separating the types of pull and push arrays.
+Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations like append on pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
+% Here push arrays are `defunctionalized' and represented as a datatype containing the combinator and its arguments, which allows specialised and efficient code generation.
+% However this work only supports a limited number of baked-in push combinators such as append, mapping, interleaving and reversing.
+
+Streaming IO libraries have blossomed in the Haskell ecosystem, generally based on Iteratees \cite{kiselyov2012iteratees}.
+Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers.
+However, these libraries are not designed for absolute performance and programs tend to be written to operate over streams of chunks rather than streams of values.
+For the most part they support only simple straight-line computations.
 

### 2017-02-07 2017mergingmerges/section/XX-Processes.tex
hide iota combinator and abstract group

@@
 
-% BL: Just start into the description.
-%     Readers will work out what's coming by flipping through the paper themselves.
-% This section describes processes in detail. We start with an informal explanation of processes before formally defining the syntax and dynamic semantics.
+Each combinator is defined as a stream process and expressed in terms of a simple imperative program with a local heap.
+The process pulls data from an aribrary number if input streams and pushes data to at least one output stream.
+The definition of the @group@ combinator which removes consecutive elements from its input stream is as follows:
 
-A process defines a stream computation, taking any number of input streams and producing at least one output stream.
-Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
-It takes no inputs, and produces one output stream @xs@.
-Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
-This variable will be incremented and pushed.
-Each process also has a current label, which denotes the instruction to perform next.
-The initial label for @iota@ is @L0@.
-Each process has a mapping from labels to instructions.
-In this case we have two instructions, @L0@ and @L1@.
-The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move to label @L1@.
-Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
-
+\begin{alltt}
+ group 
+   = \(\lambda\) (s1: Stream a) (s2: Stream a). 
+     \(\exists\) (first: Bool)  (last: a) (value: a).
+\end{alltt}\vspace{-1em}
 \begin{code}
-process (iota)
-     ins: 
-    outs: xs
-    heap: {i = 0}
-   label: L0
-  blocks: L0 = push xs i  L1
-          L1 = jump       L0{i = i + 1}
+     process
+     { ins:    { s1 }
+     , outs:   { s2 }
+     , heap:   { first = True, last = 0, value = 0 }
+     , label:  L0
+     , instrs: { L0 = pull s1 value                    L1 {}
+               , L1 = case (first || (last /= value))  L2 {}  L3 {}
+               , L2 = push s2 value                    L3 { last = value, first = False }
+               , L3 = drop s1                          L0 {} } }
 \end{code}
 
-When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+The @group@ combinator has two parameters, @s1@ and @s2@ which are the bind the input and output streams respectively.
+The combinator also defines three local variables @first@, @last@ and @value@.
+The body of the combinator is a record that defines the process.
+The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams.
+The @heap@ field gives the starting state of each of the local variables.
+The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the first instruction.
 
-The @uniques = group file1@ is a more interesting example.
 Recall that @group@ removes consecutive duplicates from its input stream.
@@
 
-\begin{code}
-process (group)
-     ins: file1
-    outs: uniques
-    heap: {first = True, last = 0, value = 0}
-   label: L0
-  blocks: L0 = pull file1   value               L1
-          L1 = case (or first (last /= value))  L2               L3
-          L2 = push uniques value               L3{last = value, first = False}
-          L3 = drop file1                       L0
-\end{code}
+When executed on an input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
 
-When executed on an input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
+% BL: I don't think describing iota works at this point. This combinator is not used in the motivating example, so skipping to it seems disjointed.
+
+% Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
+% It takes no inputs, and produces one output stream @xs@.
+% Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
+% This variable will be incremented and pushed.
+% Each process also has a current label, which denotes the instruction to perform next.
+% The initial label for @iota@ is @L0@.
+% Each process has a mapping from labels to instructions.
+% In this case we have two instructions, @L0@ and @L1@.
+% The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move %to label @L1@.
+Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
+
+% \begin{code}
+% process (iota)
+%      ins: 
+%     outs: xs
+%     heap: {i = 0}
+%    label: L0
+%   blocks: L0 = push xs i  L1
+%           L1 = jump       L0{i = i + 1}
+% \end{code}
+
+% When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+
+% The @uniques = group file1@ is a more interesting example.
 

### 2017-02-07 2017mergingmerges/section/XX-Processes.tex
Reformat merge combinator

@@
 The process pulls data from an aribrary number if input streams and pushes data to at least one output stream.
-The definition of the @group@ combinator which removes consecutive elements from its input stream is as follows:
+\ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.
 
+\subsection{Grouping}
+The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}
+
+\begin{figure}
 \begin{alltt}
@@
    = \(\lambda\) (s1: Stream a) (s2: Stream a). 
-     \(\exists\) (first: Bool)  (last: a) (value: a).
-\end{alltt}\vspace{-1em}
+     \(\nu\) (first: Bool)  (last: a) (value: a) (L0..L3: Label).
+\end{alltt}
 \begin{code}
@@
 
+\medskip
+
+\begin{alltt}
+ merge
+   = \(\lambda\) (s1: Stream a) (s2: Stream a) (s3: Stream a). 
+     \(\nu\) (a: a)  (b: a) (value: a) (L0..L3: Label) (FIXME).
+\end{alltt}
+\begin{code}
+     process
+     { ins:    { s1, s2 }
+     , outs:   { s3 }
+     , heap:   { a = 0, b = 0 }
+     , label:  L0
+     , instrs: { L0 = pull s1 a  L1 {}
+               , L1 = pull s2 b  C0 {}
+               , C0 = case (a < b)  A0 {}  B0 {}
+               , A0 = push s3 a A1 {}
+               , A1 = drop s1    A2 {}
+               , A2 = pull s1  a C0 {}
+               , B0 = push merged b B1 {}
+               , B1 = drop s2       B2 {}
+               , B2 = pull s2  b    C0 {} } }
+\end{code}
+\caption{Grouping and Merging}
+\label{fig:Process:Group}
+\end{figure}
+
 The @group@ combinator has two parameters, @s1@ and @s2@ which are the bind the input and output streams respectively.
-The combinator also defines three local variables @first@, @last@ and @value@.
+The \emph{nu-binders} \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.
+
 The body of the combinator is a record that defines the process.
@@
 
-Recall that @group@ removes consecutive duplicates from its input stream.
-It has one input, @file1@, and one output, @uniques@.
-There are three variables in the heap: @first@ is initialised to @True@ as the first pulled value has no last value to compare against; @value@ stores the most recently pulled value, and @last@ stores the last pulled value.
-The @last@ and @value@ variables are initialised to $0$, but could be initialised to any value: these initial values will not be used.
-The initial label is @L0@, which pulls from the input @file1@. This blocks waiting for an input value, and when one is received, it is stored in @value@ and the process moves to @L1@.
-Instruction @L1@ performs a case analysis on a boolean: if it is the first read value, or the last value is not equal to the most recent value, it jumps to @L2@; otherwise it jumps to @L3@.
-Instruction @L2@ pushes the most recent value to the output and jumps to @L3@, updating the last value to the most recent value, and setting first to @False@.
-Finally, the instruction for @L3@ `drops' the recently pulled value from @file1@ and jumps back to @L0@.
-This dropping is required to coordinate between multiple processes that read from the same input: now that the read value from @file1@ has been dropped, another process is free to pull the next value from @file1@ if it so wishes.
+The instruction @(pull s1 value L1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @value@, then continues with the instruction at label @L1@.
+The set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.
 
-When executed on an input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
+The instruction @(case (first || (last /= value)) L2 {} L3 {})@ checks whether the predicate @(first || last /= value)@ is true and if so continues with the instruction at label @L2@, otherwise continues with the instruction at label @L3@.
 
-% BL: I don't think describing iota works at this point. This combinator is not used in the motivating example, so skipping to it seems disjointed.
+The instruction @(push s2 value L3 { last = value, first = False })@ pushes the value @value@ to the stream @s2@ and continues with the instruction at label @L3@, once the heap has been updated to set variable @last@ to @value@ and @first@ to @False@.
 
-% Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
-% It takes no inputs, and produces one output stream @xs@.
-% Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
-% This variable will be incremented and pushed.
-% Each process also has a current label, which denotes the instruction to perform next.
-% The initial label for @iota@ is @L0@.
-% Each process has a mapping from labels to instructions.
-% In this case we have two instructions, @L0@ and @L1@.
-% The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move %to label @L1@.
-Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
+Finally, the instruction @(drop s1 L0 {})@ signals that the current element that we pulled from stream @s1@ is no longer required, and contines with the instruction at @L0@.
+This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm but has no direct operational meaning.
+We discuss @drop@ further in \TODO{ref}.
 
-% \begin{code}
-% process (iota)
-%      ins: 
-%     outs: xs
-%     heap: {i = 0}
-%    label: L0
-%   blocks: L0 = push xs i  L1
-%           L1 = jump       L0{i = i + 1}
-% \end{code}
+Overall, the @first@ variable tracks whether we are dealing with the first value from the stream, @last@ holds the last value pulled from the stream (or 0 if none have been read yet), and @value@ holds the current value pulled from the stream.
+The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
+For example, when executed on the input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
 
-% When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+Note that we refer to @group@ itself as a \emph{combinator} as it is parameterised by the names of the streams it deals with (@s1@ and @s2@).
+Once the combinator has been instantiated to work on particular streams the result is a concrete \emph{process}.
 
-% The @uniques = group file1@ is a more interesting example.
+% Recall that @group@ removes consecutive duplicates from its input stream.
+% It has one input, @file1@, and one output, @uniques@.
+
+% There are three variables in the heap: @first@ is initialised to @True@ as the first pulled value has no last value to compare against; @value@ stores the most recently pulled value, and @last@ stores the last pulled value.
+
+% The @last@ and @value@ variables are initialised to $0$, but could be initialised to any value: these initial values will not be used.
+% The initial label is @L0@, which pulls from the input @file1@. This blocks waiting for an input value, and when one is received, it is stored in @value@ and the process moves to @L1@.
+% Instruction @L1@ performs a case analysis on a boolean: if it is the first read value, or the last value is not equal to the most recent value, it jumps to @L2@; otherwise it jumps to @L3@.
+% Instruction @L2@ pushes the most recent value to the output and jumps to @L3@, updating the last value to the most recent value, and setting first to @False@.
+% Finally, the instruction for @L3@ `drops' the recently pulled value from @file1@ and jumps back to @L0@.
+% This dropping is required to coordinate between multiple processes that read from the same input: now that the read value from @file1@ has been dropped, another process is free to pull the next value from @file1@ if it so wishes.
+
+\subsection{Merging}
 
@@
 
-\begin{code}
-process (merge)
-     ins: file1 file2
-    outs: merged
-    heap: {a = 0, b = 0}
-   label: L0
-  blocks: L0 = pull file1 a  L1
-          L1 = pull file2 b  C0
-
-          C0 = case (a < b)  A0     B0
-
-          A0 = push merged a A1
-          A1 = drop file1    A2
-          A2 = pull file1  a C0
-
-          B0 = push merged b B1
-          B1 = drop file2    B2
-          B2 = pull file2  b C0
-\end{code}
-
 %% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
@@
 
+\subsection{Fusion}
 We are now ready to take our @group@ and @merge@ processes and fuse them together. 
@@
 The last value is then injected into the network, and added back to the result accumulators.
+
+% -- cuts ---------------------------------------------------------------------
+% BL: I don't think describing iota works at this point. This combinator is not used in the motivating example, so skipping to it seems disjointed.
+
+% Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
+% It takes no inputs, and produces one output stream @xs@.
+% Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
+% This variable will be incremented and pushed.
+% Each process also has a current label, which denotes the instruction to perform next.
+% The initial label for @iota@ is @L0@.
+% Each process has a mapping from labels to instructions.
+% In this case we have two instructions, @L0@ and @L1@.
+% The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move %to label @L1@.
+Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
+
+% \begin{code}
+% process (iota)
+%      ins: 
+%     outs: xs
+%     heap: {i = 0}
+%    label: L0
+%   blocks: L0 = push xs i  L1
+%           L1 = jump       L0{i = i + 1}
+% \end{code}
+
+% When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+
+% The @uniques = group file1@ is a more interesting example.

### 2017-02-08 2017mergingmerges/section/01-Introduction.tex
Edits to S2

@@
 
-In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays such that the output remains sorted.
+In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays so that the output remains sorted.
 There are a few interesting things about this example.
@@
 
-In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communications channels between the processes.
+In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communication channels between the processes.
 Each combinator is expressed as a restricted, sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing to an output channel.
@@
 \amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}
+\ben{I agree. I meant that we should cite related work to say that's also called ``fusion'', but in a completely different domain.}
 

### 2017-02-08 2017mergingmerges/section/XX-Processes.tex
Edits to S2

@@
 
-Each combinator is defined as a stream process and expressed in terms of a simple imperative program with a local heap.
+Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap.
 The process pulls data from an aribrary number if input streams and pushes data to at least one output stream.
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Grouping}
-The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}
-
 \begin{figure}
@@
 \end{code}
-
-\medskip
-
-\begin{alltt}
- merge
-   = \(\lambda\) (s1: Stream a) (s2: Stream a) (s3: Stream a). 
-     \(\nu\) (a: a)  (b: a) (value: a) (L0..L3: Label) (FIXME).
-\end{alltt}
-\begin{code}
-     process
-     { ins:    { s1, s2 }
-     , outs:   { s3 }
-     , heap:   { a = 0, b = 0 }
-     , label:  L0
-     , instrs: { L0 = pull s1 a  L1 {}
-               , L1 = pull s2 b  C0 {}
-               , C0 = case (a < b)  A0 {}  B0 {}
-               , A0 = push s3 a A1 {}
-               , A1 = drop s1    A2 {}
-               , A2 = pull s1  a C0 {}
-               , B0 = push merged b B1 {}
-               , B1 = drop s2       B2 {}
-               , B2 = pull s2  b    C0 {} } }
-\end{code}
-\caption{Grouping and Merging}
+\caption{The group combinator}
 \label{fig:Process:Group}
@@
 
-The @group@ combinator has two parameters, @s1@ and @s2@ which are the bind the input and output streams respectively.
-The \emph{nu-binders} \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.
+The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}.
+The @group@ combinator has two parameters, @s1@ and @s2@ which bind the input and output streams respectively.
+The \emph{nu-binders} like \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Merging}
+\begin{figure}
+\begin{alltt}
+ merge
+   = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat) (s3: Stream Nat). 
+     \(\nu\) (x1: Nat)  (x2: Nat) (L0..L3: Label) (FIXME).
+\end{alltt}
+\begin{code}
+     process
+     { ins:    { s1, s2 }
+     , outs:   { s3 }
+     , heap:   { x1 = 0, x2 = 0 }
+     , label:  L0
+     , instrs: { L0 = pull s1 x1     L1 {}
+               , L1 = pull s2 x2     C0 {}
+               , C0 = case (x1 < x2) A0 {}  B0 {}
+               , A0 = push s3 x1     A1 {}
+               , A1 = drop s1        A2 {}
+               , A2 = pull s1 x1     C0 {}
+               , B0 = push s3 x2     B1 {}
+               , B1 = drop s2        B2 {}
+               , B2 = pull s2  x2    C0 {} } }
+\end{code}
+\caption{The merge combinator}
+\label{fig:Process:Merge}
+\end{figure}
 
-Finally, we will look at the process for @merge@.
-Merge takes two input streams, @file1@ and @file2@, and produces the output @merged@.
-The two heap variables @a@ and @b@ are for storing the currently read values from both inputs.
-Again, they are initialised to $0$, but the initial values will not be used.
-Merge requires more complicated control flow than before.
-Instructions @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
-Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
-Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
-Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.
+The definition of the @merge@ combinator which merges two input streams is given in Figure~\ref{fig:Process:Merge}.
+The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
+The two heap variables @x1@ and @x2@ are used to store the currently read values from each input.
+The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section.
+We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.
 
-%% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
-\input{figures/ProcessDef.tex}
+As this process merges infinite streams, if we execute it using a finite prefix then the final state will be an intermediate one that may not yet have pushed all available output.
+For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
+The machine will arrive at instruction @L?@ which blocks waiting for the next value to be pulled from @arr1@.
+We discuss how to handle finite streams later in ~\S\ref{s:Finite}
 
-As this process merges infinite streams, evaluating with finite streams will block waiting for more input once the end of either stream is reached.
-This is not an issue for infinite streams, and we use examples involving finite streams are much easier to describe.
-We address finite streams in~\S\ref{s:Finite}.
-Executing this process with the inputs $[1, 4]$ and $[2, 3, 100]$ will produce the first four elements $[1, 2, 3, 4]$ as expected, but note that the $100$ is missing because the process is blocked waiting for the next value from @file1@.
-\amos{I don't think this is necessary}
+% The instructions at @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
+% Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
+% Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
+% Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.
 
+% -----------------------------------------------------------------------------
 \subsection{Fusion}
@@
 
+\clearpage{}
 \subsection{Process definitions}
+
+%% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
+\input{figures/ProcessDef.tex}
 

### 2017-02-11 2017mergingmerges/section/XX-Processes.tex
Add group process state diagram
Redo group diagram
Add merge state diagram
Add process diagrams to the figures with the code description

@@
 \begin{figure}
+
+\begin{center}
 \begin{alltt}
@@
        = \(\lambda\) (s1: Stream a) (s2: Stream a). 
-     \(\nu\) (first: Bool)  (last: a) (value: a) (L0..L3: Label).
+         \(\nu\) (first: Bool)  (last: a) (value: a) (A0..A3: Label).
 \end{alltt}
@@
          , outs:   { s2 }
-     , heap:   { first = True, last = 0, value = 0 }
-     , label:  L0
-     , instrs: { L0 = pull s1 value                    L1 {}
-               , L1 = case (first || (last /= value))  L2 {}  L3 {}
-               , L2 = push s2 value                    L3 { last = value, first = False }
-               , L3 = drop s1                          L0 {} } }
+         , heap:   { f = True, l = 0, v = 0 }
+         , label:  A0
+         , instrs: { A0 = pull s1 v            A1 {}
+                   , A1 = case (f || (l /= v)) A2 {}  A3 {}
+                   , A2 = push s2 v            A3 { l = v, f = F }
+                   , A3 = drop s1              A0 {} } }
 \end{code}
+\end{center}
+\vspace{1em}
+\includegraphics[scale=1.1]{figures/state-group.pdf}
 \caption{The group combinator}
@@
 
-The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Figure~\ref{fig:Process:Group}.
+The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}.
+We include the concrete code representation as well as a diagram of the state machine defined by the associated stream process.
+
 The @group@ combinator has two parameters, @s1@ and @s2@ which bind the input and output streams respectively.
@@
 
-The instruction @(pull s1 value L1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @value@, then continues with the instruction at label @L1@.
-The set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.
+The instruction @(pull s1 value A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@.
+In the syntax the set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.
 
-The instruction @(case (first || (last /= value)) L2 {} L3 {})@ checks whether the predicate @(first || last /= value)@ is true and if so continues with the instruction at label @L2@, otherwise continues with the instruction at label @L3@.
+The instruction @(case (f || (l /= v)) L2 {} L3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@.
+We use variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
 
-The instruction @(push s2 value L3 { last = value, first = False })@ pushes the value @value@ to the stream @s2@ and continues with the instruction at label @L3@, once the heap has been updated to set variable @last@ to @value@ and @first@ to @False@.
+The instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the stream @s2@ and continues with the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
 
-Finally, the instruction @(drop s1 L0 {})@ signals that the current element that we pulled from stream @s1@ is no longer required, and contines with the instruction at @L0@.
+Finally, the instruction @(drop s1 A0 {})@ signals that the current element that we pulled from stream @s1@ is no longer required, and contines with the instruction at @A0@.
 This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm but has no direct operational meaning.
@@
 
-Overall, the @first@ variable tracks whether we are dealing with the first value from the stream, @last@ holds the last value pulled from the stream (or 0 if none have been read yet), and @value@ holds the current value pulled from the stream.
+Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
 The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
@@
 Note that we refer to @group@ itself as a \emph{combinator} as it is parameterised by the names of the streams it deals with (@s1@ and @s2@).
-Once the combinator has been instantiated to work on particular streams the result is a concrete \emph{process}.
+Once the combinator has been instantiated to work on particular streams the result is a concrete \emph{process}, which is the code representation of a state machine.
 
@@
                    , heap:   { x1 = 0, x2 = 0 }
-     , label:  L0
-     , instrs: { L0 = pull s1 x1     L1 {}
-               , L1 = pull s2 x2     C0 {}
-               , C0 = case (x1 < x2) A0 {}  B0 {}
-               , A0 = push s3 x1     A1 {}
-               , A1 = drop s1        A2 {}
-               , A2 = pull s1 x1     C0 {}
-               , B0 = push s3 x2     B1 {}
-               , B1 = drop s2        B2 {}
-               , B2 = pull s2  x2    C0 {} } }
+                   , label:  B0
+                   , instrs: { B0 = pull s1 x1     B1 {}
+                             , B1 = pull s2 x2     C0 {}
+                             , C0 = case (x1 < x2) D0 {}  E0 {}
+                             , D0 = push s3 x1     D1 {}
+                             , D1 = drop s1        D2 {}
+                             , D2 = pull s1 x1     C0 {}
+                             , E0 = push s3 x2     E1 {}
+                             , E1 = drop s2        E2 {}
+                             , E2 = pull s2  x2    C0 {} } }
 \end{code}
+
+\medskip
+\includegraphics[scale=1.1]{figures/state-merge.pdf}
 \caption{The merge combinator}
@@
 
-The definition of the @merge@ combinator which merges two input streams is given in Figure~\ref{fig:Process:Merge}.
+The definition of the @merge@ combinator which merges two input streams is given in Fig.~\ref{fig:Process:Merge}.
 The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
@@
 \subsection{Fusion}
+
 We are now ready to take our @group@ and @merge@ processes and fuse them together. 

### 2017-02-11 2017mergingmerges/section/XX-Processes.tex
Update state diagrams
Start describing fusion of two pull instructions

@@
                  = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat) (s3: Stream Nat). 
-                   \(\nu\) (x1: Nat)  (x2: Nat) (L0..L3: Label) (FIXME).
+                   \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
 \end{alltt}
@@
 % -----------------------------------------------------------------------------
+\clearpage{}
 \subsection{Fusion}
+Our fusion algorithm takes two process state machines and produces a new one that performs the job of both.
+The algorithm proceeds by considering pairs of states, one in each of the machines to be fused.
+For example, suppose we need to produce a single machine that performs the job of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
+We want a result machine that performs the job of a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
+Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr2@) and a stream named @s2@ with the second.
+
+Now, both the @group@ machine and the @merge@ machine have a @pull@ instruction as their first one, so we have the situation shown in Fig.\ref{fig:Fusion:Pulls}.
+The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
+In the result machine we produce three new instructions, that transition between four result states, @F0@ to @F3@.
+Each of the result states represents a combination of two original states, one from each of the input machines.
+For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being its own initial state @B0@.
+
+We also associate each of the result states with information describing whether or not we have already pulled a value from each of the input streams, and if so, whether both machines are finished with that value and it is time to move to the next one.
+For the @F0@ case shown in Fig.
+\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
+This means that the result state @F0@ represents a combination of the two input states @A0@ and @B0@.
+As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.
+
+\begin{figure}
+\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
+\caption{Fusing pulls}
+\label{fig:Fusion:Pulls}
+\end{figure}
+
+\subsection{Fused Result}
 

### 2017-02-11 2017mergingmerges/section/XX-Processes.tex
Start describing fusion with case instructions and instr priority

@@
 Our fusion algorithm takes two process state machines and produces a new one that performs the job of both.
-The algorithm proceeds by considering pairs of states, one in each of the machines to be fused.
 For example, suppose we need to produce a single machine that performs the job of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
@@
 
-Now, both the @group@ machine and the @merge@ machine have a @pull@ instruction as their first one, so we have the situation shown in Fig.\ref{fig:Fusion:Pulls}.
+% -----------------------------------------------------------------------------
+\subsubsection{Fusing Pulls}
+\begin{figure}
+\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
+\caption{Fusing pull instructions}
+\label{fig:Fusion:Pulls}
+\end{figure}
+
+The algorithm proceeds by considering pairs of states, one in each of the machines to be fused.
+Both the @group@ machine and the @merge@ machine have a @pull@ instruction as their first one, so we have the situation shown in Fig.\ref{fig:Fusion:Pulls}.
 The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
-In the result machine we produce three new instructions, that transition between four result states, @F0@ to @F3@.
+In the result machine we produce three new instructions that transition between four result states, @F0@ to @F3@.
 Each of the result states represents a combination of two original states, one from each of the input machines.
@@
 
+From the joint state @F0@, both of the input machines need to pull from stream @s1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
+In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
+For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.
+After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
+
+Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
+The `none` state means that we have not yet pulled a value from the associated stream.
+The `pending` state means we have pulled a value into the stream buffer variable (@b1@ in this case).
+The `have` state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
+In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1` is set to `have` in @F3@ after we have set `x1 = b1'.
+
+\eject{}
+% -----------------------------------------------------------------------------
+\subsubsection{Fusing Cases}
 \begin{figure}
-\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
-\caption{Fusing pulls}
-\label{fig:Fusion:Pulls}
+\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
+\caption{Fusing case instructions}
+\label{fig:Fusion:Case}
 \end{figure}
 
+Once the result machine has arrived in the joint state @F3@ this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
+Fig.
+\ref{fig:Fusion:Case} shows the next few transitions of these machines.
+From state @A1@ the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @s1@ to its output stream, which we name @s3@ --- or to just move onto the next value from its input.
+From state @B1@ the merge machine needs to pull a value from its second input stream @s2@.
+
+% -----------------------------------------------------------------------------
 \subsection{Fused Result}
-
 We are now ready to take our @group@ and @merge@ processes and fuse them together. 

### 2017-02-14 2017mergingmerges/section/01-Introduction.tex
mainly prose

@@
 Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent combinators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
+The polyhedral model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out combinators such as @group@ and @merge@.
 
@@
 We are not in the same space as Fortran style array fusion transformations like polyhedral}
-\amos{I'm not convinced anybody reading this cares about Fortran or polyhedral}
-\ben{I agree. I meant that we should cite related work to say that's also called ``fusion'', but in a completely different domain.}
 

### 2017-02-14 2017mergingmerges/section/XX-Processes.tex
mainly prose

@@
 Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap.
-The process pulls data from an aribrary number if input streams and pushes data to at least one output stream.
-\ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.
+The process pulls data from an arbitrary number of input streams and pushes data to at least one output stream.
+Each array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
+% \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.
+
+% make it explicit that the processes are only an intermediate repr
+The process language is used for the implementation of the combinators and as an intermediate representation for fusion, while the orignal program using these combinators can be written without any knowledge of processes.
 
@@
            group 
-             = \(\lambda\) (s1: Stream a) (s2: Stream a). 
-               \(\nu\) (first: Bool)  (last: a) (value: a) (A0..A3: Label).
+             = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat). 
+               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
 \end{alltt}
@@
                , outs:   { s2 }
-               , heap:   { f = True, l = 0, v = 0 }
+               , heap:   { f = T, l = 0, v = 0 }
                , label:  A0
@@
 
-The @group@ combinator has two parameters, @s1@ and @s2@ which bind the input and output streams respectively.
-The \emph{nu-binders} like \mbox{$\nu$ @(first: Bool)@...} indicate that each time we instantiate the @group@ combinator we should create fresh names for @first@, @last@ and so on that do not conflict with other instantiations.
+The @group@ combinator has two parameters, @s1@ and @s2@, which bind the input and output streams respectively.
+The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
 
@@
 The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams.
-The @heap@ field gives the starting state of each of the local variables.
+The @heap@ field gives the initial values of each of the local variables.
 The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the first instruction.
 
-The instruction @(pull s1 value A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@.
+The instruction @(pull s1 v A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@.
 In the syntax the set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.
 
-The instruction @(case (f || (l /= v)) L2 {} L3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@.
+The instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@.
 We use variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
@@
 
-Finally, the instruction @(drop s1 A0 {})@ signals that the current element that we pulled from stream @s1@ is no longer required, and contines with the instruction at @A0@.
-This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm but has no direct operational meaning.
-We discuss @drop@ further in \TODO{ref}.
+Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and continues with the instruction at @A0@.
+This @drop@ instruction is used to coordinate processes when performing fusion and concurrent evaluation; the next element of a stream may only be pulled after all consuming processes have pulled and dropped the current element.
+% This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm and concurrent evaluation, but has no direct operational meaning.
+% We discuss @drop@ further in \TODO{ref}.
 
@@
 
-As this process merges infinite streams, if we execute it using a finite prefix then the final state will be an intermediate one that may not yet have pushed all available output.
+As this process merges infinite streams, if we execute it using a finite prefix then the result state will be an intermediate one that may not yet have pushed all available output.
 For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-The machine will arrive at instruction @L?@ which blocks waiting for the next value to be pulled from @arr1@.
+The process will arrive at instruction @E2@ which blocks waiting for the next value to be pulled from @s2@.
 We discuss how to handle finite streams later in ~\S\ref{s:Finite}
@@
 We want a result machine that performs the job of a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr2@) and a stream named @s2@ with the second.
+Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second.
 
@@
 Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
-The `none` state means that we have not yet pulled a value from the associated stream.
-The `pending` state means we have pulled a value into the stream buffer variable (@b1@ in this case).
-The `have` state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
-In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1` is set to `have` in @F3@ after we have set `x1 = b1'.
+The `none' state means that we have not yet pulled a value from the associated stream.
+The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case).
+The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
+In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.
 
@@
 \subsection{Fused Result}
-We are now ready to take our @group@ and @merge@ processes and fuse them together. 
-We create a new process with inputs @file1@ and @file2@ and outputs @uniques@ and @merged@.
-In order to make the fused process easier to understand, we have performed some minor optimisations on the output of the fusion algorithm described in~\S\ref{s:Fusion}, such as removing @jump@s and merging the two variables @value@ and @a@ into a single variable @value_a@.
-Rather than explaining the process in detail, we simply note that the fused process is an interleaved scheduling of the two, starting with executing @group@, then executing @merge@, and going back to @group@ whenever pulling from @file1@ again.
-The labels @L0@-@L2@ and @A1@-@A4@ are much the same as those in @group@, while @C0@, @A0@-@A2@ and @B0@-@B2@ are similar to those in @merge@.
 
+Fig.~\ref{fig:Process:Fused} shows the result of fusing @group@ and @merge@ together, after some manual simplification.
+The redundant @jump@s at @F1@ and @F2@ have been removed, and the variables @v@ and @x1@ have been merged into a single variable, @x1@.
+
+\amos{Or is it better to have the actual unsimplified result of fusion? It will be bigger, but don't have to handwave.}
+
+\begin{figure}
 \begin{code}
-process (group/merge)
-     ins: file1 file2
-    outs: uniques merged
-    heap: {first = True, last, value_a, b}
-   label: L0
-  blocks: L0 = pull file1   value_a               L1
-          L1 = case (or first (last /= value_a))  L2    L3
-          L2 = push uniques value_a               L3{last = value_a, first = False}
-          L3 = pull file2 b                       C0
+            process
+            { ins:    { s1, s2 }
+            , outs:   { s3, s4 }
+            , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
+            , label:  F0
+            , instrs: { F0 = pull s1 x1              F3 {}
+                      , F3 = case (f || (l /= x1))   F4 {} F5 {}
+                      , F4 = push s3 x1              F5 { l = x1, f = F }
+                      , F5 = pull s2 x2              F6 {}
 
-          C0 = case (value_a < b)                 A0    B0
+                      , F6 = case (x1 < x2)          A0 {} B0 {}
 
-          A0 = push merged  value_a               A1
-          A1 = drop file1                         A2
-          A2 = pull file1   value_a               A3
-          A3 = case (or first (last /= value_a))  A4    C0
-          A4 = push uniques value_a               C0{last = value_a, first=False}
+                      , A0 = push s4 x1              A1 {}
+                      , A1 = drop s1                 A2 {}
+                      , A2 = pull s1 x1              A3 {}
+                      , A3 = case (f || (l /= x1))   A4 {} F6 {}
+                      , A4 = push s3 x1              F6 { l = x1, f = F }
 
-          B0 = push merged b                      B1
-          B1 = drop file2                         B2
-          B2 = pull file2  b                      C0
+                      , B0 = push s4 x2              B1 {}
+                      , B1 = drop s2                 B2 {}
+                      , B2 = pull s2 x2              F6 {} } }
 \end{code}
+\caption{Fusion of group and merge}
+\label{fig:Process:Fused}
+\end{figure}
 
@@
 A process defines a stream computation, taking any number of input streams and producing at least one output stream.
-Processes have an optional name in parentheses used only for descriptive purposes, and does not need to be unique.
 The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
@@
 
-The output streams are in some sense ``owned'' by the process that produces them.
-While a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
+The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
 This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
@@
 The heap is used for evaluation of expressions.
-Each process has its own private heap, meaning that the only communication between processes occurs by streams.
+Each process has its own private heap, therefore the only communication between processes occurs by streams.
 
-The label is the current block, and each block is the instruction to be executed in the current state.
+The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
 Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
@@
 
-All instructions take as an argument the next label state to jump to, which includes any variable updates that should be performed on the private heap at the same time.
-Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.
+All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
+% Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.
 
 A process network is a set of multiple processes that can be evaluated concurrently.
-The intersection of all process outputs should be empty - there should be no overlap.
-Any inputs that are not mentioned as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
+Any inputs that are not produced as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
 Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
@@
 \item Shaking, where a process takes a step from one label to another.
-Shaking a process results in a new process as well as an output message.
+Shaking a process results in an updated process as well as an output message.
 If a process pushes to a stream, the push value must be able to be injected to other processes.

### 2017-02-15 2017mergingmerges/section/XX-Processes.tex
unsimplified fusion result

@@
             , outs:   { s3, s4 }
-            , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
+            , heap:   { f = T, l = 0, v = 0, x1 = 0, x2 = 0, b1 = 0}
             , label:  F0
-            , instrs: { F0 = pull s1 x1              F3 {}
-                      , F3 = case (f || (l /= x1))   F4 {} F5 {}
-                      , F4 = push s3 x1              F5 { l = x1, f = F }
-                      , F5 = pull s2 x2              F6 {}
+            , instrs:
 
-                      , F6 = case (x1 < x2)          A0 {} B0 {}
+  (shared pull)
+              { F0  = pull s1 b1            F1  {}
 
-                      , A0 = push s4 x1              A1 {}
-                      , A1 = drop s1                 A2 {}
-                      , A2 = pull s1 x1              A3 {}
-                      , A3 = case (f || (l /= x1))   A4 {} F6 {}
-                      , A4 = push s3 x1              F6 { l = x1, f = F }
+  (merge label B0)
+              , F1  = jump                  F2  {v  = b1}
 
-                      , B0 = push s4 x2              B1 {}
-                      , B1 = drop s2                 B2 {}
-                      , B2 = pull s2 x2              F6 {} } }
+  (group labels A0-A3)
+              , F2  = jump                  F3  {x1 = b1}
+              , F3  = case (f || (l /= v))  F4  {}                F5  {}
+              , F4  = push s3 v             F5  {l  = v, f  = F}
+              , F5  = jump                  F6  {}
+
+  (merge label B1)
+              , F6  = pull s2 x2            F7  {}
+
+  (merge label C0)
+              , F7  = case (x1 < x2)        F8  {}                F16 {}
+
+  (merge labels D0-D1)
+              , F8  = push s4 x1            F9  {}
+              , F9  = drop s1               F10 {}
+
+  (shared pull)
+              , F10 = pull s1 b1            F11 {}
+
+  (group labels A0-A3)
+              , F11 = jump                  F12 {v  = b1}
+              , F12 = case (f || (l /= v))  F13 {}                F14 {}
+              , F13 = push s3 v             F14 {l  = v, f  = F}
+              , F14 = jump                  F15 {}
+
+  (merge label D2)
+              , F15 = jump                  F7  {x1 = b1}
+
+  (merge labels E0-E2)
+              , F16 = push s4 x2            F17 {}
+              , F17 = drop s2               F18 {}
+              , F18 = pull s2 x2            F7  {} } }
 \end{code}
+
+% With original label names and 'n', 'h' and 'p' for input state
+%\begin{code}
+%A0nB0n = pull s1 b1  A0pB0p{}
+%A0pB0p = jump        A1hB0p{v  = b1}
+%A1hB0p = jump        A1hB1h{x1 = b1}
+%
+%A1hB1h = case (f || (l /= v))
+%                     A2hB1h{}
+%                     A3hB1h{}
+%
+%A2hB1h = push s3 v   A3hB1h{l  = v, f  = F}
+%A3hB1h = jump        A0nB1h{}
+%A0nB1h = pull s2 x2  A0nC0h{}
+%
+%A0nC0h = case (x1 < x2)
+%                     A0nD0h{}
+%                     A0nE0h{}
+%
+%A0nD0h = push s4 x1  A0nD1h{}
+%A0nD1h = drop s1     A0nD2n{}
+%A0nD2n = pull s1 b1  A0pD2p{}
+%A0pD2p = jump        A1hD2p{v  = b1}
+%
+%A1hD2p = case (f || (l /= v))
+%                     A2hD2p{}
+%                     A3hD2p{}
+%
+%A2hD2p = push s3 v   A3hD2p{l  = v, f  = F}
+%A3hD2p = jump        A0nD2p{}
+%A0nD2p = jump        A0nC0h{x1 = b1}
+%
+%A0nE0h = push s4 x2  A0nE1h{}
+%A0nE1h = drop s2     A0nE2h{}
+%A0nE2h = pull s2 x2  A0nC0h{}
+%\end{code}
+
+% Cleaned up, simplified, variables collated
+% \begin{code}
+%             process
+%             { ins:    { s1, s2 }
+%             , outs:   { s3, s4 }
+%             , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
+%             , label:  F0
+%             , instrs: { F0 = pull s1 x1              F3 {}
+%                       , F3 = case (f || (l /= x1))   F4 {} F5 {}
+%                       , F4 = push s3 x1              F5 { l = x1, f = F }
+%                       , F5 = pull s2 x2              F6 {}
+% 
+%                       , F6 = case (x1 < x2)          A0 {} B0 {}
+% 
+%                       , A0 = push s4 x1              A1 {}
+%                       , A1 = drop s1                 A2 {}
+%                       , A2 = pull s1 x1              A3 {}
+%                       , A3 = case (f || (l /= x1))   A4 {} F6 {}
+%                       , A4 = push s3 x1              F6 { l = x1, f = F }
+% 
+%                       , B0 = push s4 x2              B1 {}
+%                       , B1 = drop s2                 B2 {}
+%                       , B2 = pull s2 x2              F6 {} } }
+% \end{code}
 \caption{Fusion of group and merge}

### 2017-02-18 2017mergingmerges/section/XX-Processes.tex
Make map process creation combinato

@@
 From state @B1@ the merge machine needs to pull a value from its second input stream @s2@.
+
+In terms of fusion, we could construct the result machine in multiple ways.
+One option is to perform the case branch first and then pull from @s2@.
+We could alternately pull from @s2@ first and then perform the case branch.
+By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
+As the set of variables to not overlap then either ordering is correct.
+For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \TODO{ref}.
 

### 2017-02-18 2017mergingmerges/code/example/MapMap.hs

@@ -1,64 +1,0 @@
-
-module MapMap where
-import Machine.Base
-import Machine.Execute
-import Machine.Fuse
-import Control.Monad.State.Strict
-import qualified Data.Map       as Map
-import qualified Data.Set       as Set
-import qualified Text.Show.Pretty       as P
-
-
-inputs
- = Map.fromList
- [ (Channel "as", [VInt 1, VInt 2, VInt 3]) ]
-
-pMap1 :: Process
-pMap1
- = Process
- { processName          = "map1"
- , processIns           = Set.fromList [Channel "as"]
- , processInStates      = Map.fromList [(Channel "as", None)]
- , processOuts          = Set.fromList [Channel "bs"]
- , processHeap          = Heap (Map.fromList [(Var "a1", VInt 0)])
- , processLabel         = Label "map10"
- , processBlocks        =
-        [ ( Label "map10"
-          , Pull (Channel "as") (Var "a1")
-                 (Next (Label "map11") Map.empty))
-
-        , ( Label "map11"
-          , Push (Channel "bs") (XApp (XVal VSucc) (XVar (Var "a1")))
-                 (Next (Label "map12") Map.empty))
-
-        , ( Label "map12"
-          , Drop (Channel "as")
-                 (Next (Label "map10") Map.empty))
-        ]
- }
-
-
-pMap2 :: Process
-pMap2   
- = Process
- { processName          = "map2"
- , processIns           = Set.fromList [Channel "bs"]
- , processInStates      = Map.fromList [(Channel "bs", None)]
- , processOuts          = Set.fromList [Channel "cs"]
- , processHeap          = Heap (Map.fromList [(Var "b1", VInt 0)])
- , processLabel         = Label "map20"
- , processBlocks        =
-        [ ( Label "map20"
-          , Pull (Channel "bs") (Var "b1")
-                 (Next (Label "map21") Map.empty))
-
-        , ( Label "map21"
-          , Push (Channel "cs") (XApp (XVal VSucc) (XVar (Var "b1")))     
-                 (Next (Label "map22") Map.empty))
-
-        , ( Label "map22"
-          , Drop (Channel "bs")
-                 (Next (Label "map20") Map.empty))
-        ]
- }
-

### 2017-02-18 2017mergingmerges/code/src/Machine/Combinator.hs

@@ -1,0 +1,9 @@
+
+module Machine.Combinator
+        ( module Machine.Base
+        , module Machine.New
+        , mkMap)
+where
+import Machine.Combinator.Map
+import Machine.Base
+import Machine.New

### 2017-02-18 2017mergingmerges/code/src/Machine/Combinator/Map.hs

@@ -1,0 +1,40 @@
+
+module Machine.Combinator.Map
+where
+import Machine.New
+import Machine.Base
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-- | Construct a new `map` process.
+mkMap   :: Channel              -- ^ Input  Channel.
+        -> Channel              -- ^ Output Channel.
+        -> (Expr -> Expr)       -- ^ Worker function.
+        -> New Process
+
+mkMap cIn cOut f
+ = do
+        [l0, l1, l2]    <- replicateM 3 (newLabel "M")
+        v0              <- newVar "V"
+
+        return  
+         $ Process
+         { processName          = "map"
+         , processIns           = Set.fromList [cIn]
+         , processInStates      = Map.fromList [(cIn, None)]
+         , processOuts          = Set.fromList [cOut]
+         , processHeap          = Heap (Map.fromList [(v0, VInt 0)])
+         , processLabel         = l0
+         , processBlocks        =
+                [ ( l0
+                  , Pull cIn v0 (Next l1 Map.empty))
+
+                , ( l1
+                  , Push cOut   (f (XVar v0))
+                                (Next l2 Map.empty))
+
+                , ( l2
+                  , Drop cIn    (Next l0 Map.empty))
+                ]
+         }

### 2017-02-18 2017mergingmerges/code/src/Machine/New.hs

@@ -1,0 +1,44 @@
+
+module Machine.New
+        ( module Control.Monad.State.Strict
+        , New (..)
+        , evalNew
+        , newIx
+        , newLabel
+        , newVar)
+where
+import Machine.Base
+import Control.Monad.State.Strict
+
+
+-- | Type of computations that can allocate new names.
+type New a 
+ = State Int a
+
+-- | Produce a new index.
+newIx :: New Int
+newIx 
+ = do   i   <- get
+        put (i + 1)
+        return i
+
+
+-- | Produce a new label name.
+newLabel  :: String -> New Label
+newLabel prefix
+ = do   i   <- newIx
+        return $ Label $ "l" ++ prefix ++ show i
+
+
+-- | Produce a new variable.
+newVar  :: String -> New Var
+newVar prefix
+ = do   i   <- newIx
+        return $ Var   $ "x" ++ prefix ++ show i
+
+
+-- | Evaluate a `New` computation.
+evalNew  :: New a -> a
+evalNew n
+ = evalState n 0
+

### 2017-02-18 2017mergingmerges/code/src/Machine/Execute.hs
Fix feeding for shared input channels

@@ -6,6 +6,7 @@ import Machine.Feed
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map
+import Data.Maybe
 
 
 execute 
@@ -21,8 +22,13 @@ execute cvsIn cvsOut ps acc
  = (cvsIn, ps, acc)
 
  | otherwise
- = let  (cvsIn', ps')   = feedProcesses cvsIn ps
-        (ps'',   as')   = shakeSteps [] ps' []
+ = let  (cvsIn', ps')   
+         = fromMaybe (error "execute: feed failed") 
+                     (feedProcesses cvsIn ps)
+
+        (ps'',   as')
+         = shakeSteps [] ps' []
+
    in   execute cvsIn' cvsOut ps'' (acc ++ as')
 
 

### 2017-02-18 2017mergingmerges/code/src/Machine/Feed.hs

@@ -1,39 +1,52 @@
 
 module Machine.Feed where
 import Machine.Base
+import Machine.Inject
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map
 
 
-type ChannelValueS a    
-        =   Map Channel [Value] -> a
-        -> (Map Channel [Value], a)
-
-
--- | Feed input into a channel, if it needs it.
-feedInputState ::  ChannelValueS (Channel, InputState)
-feedInputState cvs (c, state)
- = case state of
-        None    
-         -> case Map.lookup c cvs of
-                Nothing         -> (cvs, (c, state))
-                Just (v:vs)     -> (Map.insert c vs cvs, (c, Pending v))
-
-        _ -> (cvs, (c, state))
-
-
--- | Feed input into the given process.
-feedProcess :: ChannelValueS Process
-feedProcess cvs p
- = let  (cvs', cis')    = mapAccumL feedInputState cvs 
-                        $ Map.toList $ processInStates p
-
-        p'              = p { processInStates = Map.fromList cis' }
-   in   (cvs', p')
-
-
 -- | Feed input into the given proceses.
-feedProcesses :: ChannelValueS [Process]
+feedProcesses 
+ ::  Map Channel [Value]         -- ^ Values on input channels.
+ -> [Process]                    -- ^ Current processes.
+ ->  Maybe ( Map Channel [Value] -- New processes and remaining values.
+           , [Process])
+
 feedProcesses cvs ps
-        = mapAccumL feedProcess cvs ps
+  = do  (cvs', ps') <- feedProcessList (Map.toList cvs) ps
+        return (Map.fromList cvs', ps')
+
+
+feedProcessList 
+ :: [(Channel, [Value])]
+ -> [Process]
+ -> Maybe ( [(Channel, [Value])]
+          , [Process])
+
+feedProcessList []  ps
+ =      return ([], ps)
+
+feedProcessList (cvs : cvss) ps 
+ = do   (cvs',  ps')    <- feedProcess1    cvs  ps  
+        (cvss', ps'')   <- feedProcessList cvss ps' 
+        return (cvs' : cvss', ps'')
+
+
+feedProcess1 
+ :: (Channel, [Value]) 
+ -> [Process] 
+ -> Maybe ( (Channel, [Value])
+          , [Process])
+
+feedProcess1 (c, []) ps 
[…]

### 2017-02-18 2017mergingmerges/code/src/Machine/Inject.hs

@@ -1,0 +1,34 @@
+
+module Machine.Inject where
+import Machine.Base
+import Data.Map                 (Map)
+import qualified Data.Map       as Map
+
+
+-- | Inject a value into a possibly shared input channel of the given
+--   processeses. All processes which have the named channel as one 
+--   of their input channels must accept the value, otherwise `Nothing`.
+injects :: [Process] -> Channel -> Value -> Maybe [Process]
+injects ps c v
+ = mapM (\p -> inject p c v) ps
+
+
+-- | Inject a value into an input channel of a process.
+inject  :: Process   -> Channel -> Value -> Maybe Process
+inject p c v 
+
+ -- InjectValue
+ | Just None    <- Map.lookup c (processInStates p)
+ = Just $ p { processInStates
+               = Map.insert c (Pending v) (processInStates p) }
+
+ -- InjectIgnore
+ | Nothing      <- Map.lookup c (processInStates p)
+ = Just $ p
+
+ -- No rule matches.
+ -- The process was not ready to receive input on this channel.
+ | otherwise
+ = Nothing 
+
+

### 2017-02-18 2017mergingmerges/code/src/Machine/Shake.hs

@@ -1,6 +1,7 @@
 
 module Machine.Shake where
 import Machine.Base
+import Machine.Inject
 import Machine.Eval
 import Data.Map                         (Map)
 import Data.List
@@ -11,24 +12,6 @@ import qualified Data.Map.Strict        as Map
 data Action
         = ActionPush Channel Value
         deriving Show
-
-
----------------------------------------------------------------------------------------------------
--- | Inject a value into an input channel of a process.
-inject :: Channel -> Value -> Process -> Maybe Process
-inject c v p
- = case Map.lookup c (processInStates p) of
-        -- Process is ready to receive input on this channel.
-        Just None  
-         -> Just $ p { processInStates 
-                        = Map.insert c (Pending v) (processInStates p) }
-
-        -- Process was not ready to receive input on this channel.
-        Just _     -> Nothing
-
-        -- Process does not have an input of this name,
-        -- so just return the original process.
-        Nothing    -> Just p
 
 
 ---------------------------------------------------------------------------------------------------
@@ -52,8 +35,8 @@ shakeSteps stalled (p : psRest) acc
                 Just action@(ActionPush channel value)
                  -- The process wants to push to one of its output channels,
                  -- so the consumers need to be ready to accept it for the step to complete.
-                 |  Just stalled' <- sequence $ map (inject channel value) stalled
-                 ,  Just psRest'  <- sequence $ map (inject channel value) psRest
+                 |  Just stalled' <- injects stalled channel value
+                 ,  Just psRest'  <- injects psRest  channel value
                  -> shakeSteps [] (stalled' ++ (p' : psRest')) (acc ++ [action])
 
                  -- One of the consumers wasn't ready to accept the output.

### 2017-02-20 2017mergingmerges/section/01-Introduction.tex
prose

@@
 
-% \amos{More compelling justification. Why would someone want to do this?}
-% \amos{Should this example also have readFiles?}
-
-% This example is somewhat incorrect because the streams here are pull streams,
-% but the fused result cannot return pull streams (since we don't know what order they will be pulled)
-% \begin{code}
-% uniquesUnion (file1 file2 : Stream Id) : (Stream Id, Stream Id)
-%  = let uniques = group file1
-%        merged  = merge file1 file2
-%        union   = group merged
-%    in (uniques, union)
-% \end{code}
-
 In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays so that the output remains sorted.
-There are a few interesting things about this example.
+This example has a few interesting attributes.
 Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ reads values from @arr1@ or @arr2@ depends on the values in the arrays themselves.
@@
 Secondly, although @arr1@ occurs twice in the program, at runtime we only want to read through the elements of this array once.
-To achieve this the compiled program must coordinate between the two uses of @arr1@, so when the implementation of @uniques@ requires a new value the implementation of @union@ is also ready to receive a new value, and vice versa.
-Although the program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.
+To achieve this, the compiled program must coordinate between the two uses of @arr1@, such that values are only read when both @uniques1@ and @merged@ are ready to receive a new value.
+Finally, although this program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.
 
@@
 We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
-Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but in this case splits can cause excessive inlining which duplicates work.
+Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work - in this case reading the array multiple times.
 Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse this example because they do not support combinators with multiple inputs.
@@
 
-\ben{Add a few more sentences on related work.
+% \ben{Add a few more sentences on related work.
 Explain how this work extends the old flow fusion paper.

### 2017-02-20 2017mergingmerges/section/XX-Processes.tex
prose

@@
 The process pulls data from an arbitrary number of input streams and pushes data to at least one output stream.
-Each array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
+Each input array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
 % \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.
@@
 % make it explicit that the processes are only an intermediate repr
-The process language is used for the implementation of the combinators and as an intermediate representation for fusion, while the orignal program using these combinators can be written without any knowledge of processes.
+The process language is used for the implementation of combinators and as an intermediate representation for fusion, while the original program using these combinators can be written without any knowledge of processes.
 
@@
 The @heap@ field gives the initial values of each of the local variables.
-The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the first instruction.
+The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
 
-The instruction @(pull s1 v A1 {})@ pulls the next element from stream @s1@, writes it into the heap variable @v@ (value), then continues with the instruction at label @A1@.
-In the syntax the set @{}@ at the end of the instruction can be used to update values in the heap, but as we do not need to perform heap updates in this instance we leave it empty.
+The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
+The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates here we leave it empty.
 
-The instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || l /= v)@ is true and if so continues with the instruction at label @A2@, otherwise continues with the instruction at label @A3@.
-We use variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
+Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to the instruction at label @A3@.
+We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
 
-The instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the stream @s2@ and continues with the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
+When the above predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
 
-Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and continues with the instruction at @A0@.
+Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@.
 This @drop@ instruction is used to coordinate processes when performing fusion and concurrent evaluation; the next element of a stream may only be pulled after all consuming processes have pulled and dropped the current element.
@@
 
-The definition of the @merge@ combinator which merges two input streams is given in Fig.~\ref{fig:Process:Merge}.
+The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}.
 The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
 The two heap variables @x1@ and @x2@ are used to store the currently read values from each input.
-The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section.
-We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.
+The process starts by pulling from both input streams, then compares the two values, and pushes the smaller of the values to the output stream.
+It then drops the input stream corresponding to the smaller value, then pulls from the same stream, and performs the same comparison again.
+% The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section.
+% We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.
 
-As this process merges infinite streams, if we execute it using a finite prefix then the result state will be an intermediate one that may not yet have pushed all available output.
+As this process merges infinite streams, if we execute it with a finite input prefix, the result will be an intermediate state that may not yet have pushed all available output.
 For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-The process will arrive at instruction @E2@ which blocks waiting for the next value to be pulled from @s2@.
+After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@.
 We discuss how to handle finite streams later in ~\S\ref{s:Finite}
@@
 \subsection{Fusion}
-Our fusion algorithm takes two process state machines and produces a new one that performs the job of both.
-For example, suppose we need to produce a single machine that performs the job of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
-We want a result machine that performs the job of a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-Relative to the @uniquesUnion@ example we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second.
+Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
+For example, suppose we need to produce a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
+We want a result machine that computes both a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
+Compared to the @uniquesUnion@ example, we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second array (@arr2@).
 
@@
 
-The algorithm proceeds by considering pairs of states, one in each of the machines to be fused.
-Both the @group@ machine and the @merge@ machine have a @pull@ instruction as their first one, so we have the situation shown in Fig.\ref{fig:Fusion:Pulls}.
+The algorithm proceeds by considering pairs of states: one in each of the machines to be fused.
+Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Fig.~\ref{fig:Fusion:Pulls}.
 The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
@@
 Each of the result states represents a combination of two original states, one from each of the input machines.
-For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being its own initial state @B0@.
+For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.
 
-We also associate each of the result states with information describing whether or not we have already pulled a value from each of the input streams, and if so, whether both machines are finished with that value and it is time to move to the next one.
-For the @F0@ case shown in Fig.
-\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
+We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream, which is used to determine whether another value can be pulled.
+For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
 This means that the result state @F0@ represents a combination of the two input states @A0@ and @B0@.
@@
 The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
-In Fig.\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.
+In Fig.~\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.
 
@@
 
-Once the result machine has arrived in the joint state @F3@ this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
-Fig.
-\ref{fig:Fusion:Case} shows the next few transitions of these machines.
-From state @A1@ the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @s1@ to its output stream, which we name @s3@ --- or to just move onto the next value from its input.
-From state @B1@ the merge machine needs to pull a value from its second input stream @s2@.
+Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
+Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
+From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @s1@ to its output stream, which we name @s3@ --- or to just move on to the next value from its input.
+From state @B1@, the @merge@ machine needs to pull a value from its second input stream @s2@.
+
+In the fused result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
+At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
+
+Finally, at @F5@ the @merge@ machine pulls from @s2@, moving from @B1@ to @C0@.
+Because the stream @s2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
+In this case, we simply perform a @pull@ instruction.
 

### 2017-02-20 2017mergingmerges/section/XX-Processes.tex
ok

@@
 After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@.
-We discuss how to handle finite streams later in ~\S\ref{s:Finite}
+We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
 
@@
 By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
-As the set of variables to not overlap then either ordering is correct.
-For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \TODO{ref}.
+As the set of variables does not overlap, either ordering is correct.
+For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.
 
@@
 
-Fig.~\ref{fig:Process:Fused} shows the result of fusing @group@ and @merge@ together, after some manual simplification.
-The redundant @jump@s at @F1@ and @F2@ have been removed, and the variables @v@ and @x1@ have been merged into a single variable, @x1@.
-
-\amos{Or is it better to have the actual unsimplified result of fusion? It will be bigger, but don't have to handwave.}
+Fig.~\ref{fig:Process:Fused} shows the result of fusing @group@ and @merge@ together.
+The process has two inputs, @s1@ and @s2@, which correspond to the original input streams.
+Of the two outputs, @s3@ is the @group@'s output stream, while @s4@ is the @merge@'s output stream.
+The heap contains the variables from both processes, as well as the new buffer variable @b1@, which has been initialised to $0$.
 

### 2017-02-21 2017mergingmerges/section/XX-Processes.tex
ok
fig 5. fusion of group and merge

@@
 
-Fig.~\ref{fig:Process:Fused} shows the result of fusing @group@ and @merge@ together.
+Fig.~\ref{fig:Process:Fused} shows the entire result of fusing @group@ and @merge@ together.
 The process has two inputs, @s1@ and @s2@, which correspond to the original input streams.
 Of the two outputs, @s3@ is the @group@'s output stream, while @s4@ is the @merge@'s output stream.
-The heap contains the variables from both processes, as well as the new buffer variable @b1@, which has been initialised to $0$.
+The heap contains the variables from both processes, as well as the new buffer variable @b1@, which is used to store the current element pulled from @s1@.
+
+For the most part, the fused process simply alternates between performing the instructions of @group@ and those of @merge@.
+For the shared input stream @s1@, the two processes must be coordinated so that values are only pulled or dropped when both agree.
+
+The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
+
+The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
+These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
+Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.
+
+Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
+As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
+In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped.
+The actual @drop@ is only performed once both processes are `none', at label @F9@.
 
@@
 , instrs:
+\end{code}
 
-  (shared pull)
-              { F0  = pull s1 b1            F1  {}
+\newcommand\annot[5]{
+  \tiny ((#1, \> \tiny \{s1 =      #2\}), \> \tiny (#3, \> \tiny \{s1 =      #4, \> \tiny s2 =      #5\}))
+}
+\newcommand\icase[7]{
+ \tt{#1} \> \tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \tt{#5} \{\tt{#6}\} \> \tiny #1 \> \tiny = #7 \\
+}
+\newcommand\instr[5]{
+ \tt{#1}\>\tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \> \tiny #1 \> \tiny = #5 \\
+}
+\begin{tabbing}
+@  @ \=
+@, F17@ \= = @case (f || (l /= v)) @
+      \= @F17{x1 = b1, f = F}@ \= @F17{ }@
+@      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
+\> @{@ \instr{F0}{pull s1 b1}{F1}{}{\annot{A0}{none}{B0}{none}{none}}
 
-  (merge label B0)
-              , F1  = jump                  F2  {v  = b1}
+\> @,@
+\instr{F1}{jump}{F2}{v  = b1}
+      {\annot{A0}{pending}{B0}{pending}{none}}
+\> @,@
+\instr{F2}{jump}{F3}{x1 = b1}
+      {\annot{A1}{have}{B0}{pending}{none}}
+\> @,@
+\icase{F3}{case (f || (l /= v))}{F4}{}{F5}{}
+      {\annot{A1}{have}{B1}{have}{none}}
 
-  (group labels A0-A3)
-              , F2  = jump                  F3  {x1 = b1}
-              , F3  = case (f || (l /= v))  F4  {}                F5  {}
-              , F4  = push s3 v             F5  {l  = v, f  = F}
-              , F5  = jump                  F6  {}
+\> @,@
+\instr{F4}{push s3 v}{F5}{l = v, f = F}
+      {\annot{A2}{have}{B1}{have}{none}}
 
-  (merge label B1)
-              , F6  = pull s2 x2            F7  {}
+\> @,@
+\instr{F5}{jump}{F6}{}
+      {\annot{A3}{have}{B1}{have}{none}}
 
-  (merge label C0)
-              , F7  = case (x1 < x2)        F8  {}                F16 {}
+\> @,@
+\instr{F6}{pull s2 x2}{F7}{}
+      {\annot{A0}{none}{B1}{have}{none}}
+\\
 
-  (merge labels D0-D1)
-              , F8  = push s4 x1            F9  {}
-              , F9  = drop s1               F10 {}
+\> @,@
+\icase{F7}{case (x1 < x2)}{F8}{}{F16}{}
+      {\annot{A0}{none}{C0}{have}{have}}
 
-  (shared pull)
-              , F10 = pull s1 b1            F11 {}
+\\
+\> @,@
+\instr{F8}{push s4 x1}{F9}{}
+      {\annot{A0}{none}{D0}{have}{have}}
+\> @,@
+\instr{F9}{drop s1}{F10}{}
+      {\annot{A0}{none}{D1}{none}{have}}
+\> @,@
+\instr{F10}{pull s1 b1}{F11}{}
+      {\annot{A0}{none}{D2}{none}{have}}
+\> @,@
+\instr{F11}{jump}{F11}{v = b1}
+      {\annot{A0}{pending}{D2}{pending}{have}}
+\> @,@
+\icase{F12}{case (f || (l /= v))}{F13}{}{F14}{}
+      {\annot{A1}{have}{D2}{pending}{have}}
+\> @,@
+\instr{F13}{push s3 v}{F14}{l = v, f = F}
+      {\annot{A2}{have}{D2}{pending}{have}}
+\> @,@
+\instr{F14}{jump}{F15}{}
+      {\annot{A3}{have}{D2}{pending}{have}}
+\> @,@
+\instr{F15}{jump}{F7}{x1 = b1}
+      {\annot{A0}{none}{D2}{pending}{have}}
 
-  (group labels A0-A3)
-              , F11 = jump                  F12 {v  = b1}
-              , F12 = case (f || (l /= v))  F13 {}                F14 {}
-              , F13 = push s3 v             F14 {l  = v, f  = F}
-              , F14 = jump                  F15 {}
+\\
+\> @,@
+\instr{F16}{push s4 x2}{F17}{}
+      {\annot{A0}{none}{E0}{have}{have}}
+\> @,@
+\instr{F17}{drop s2}{F18}{}
+      {\annot{A0}{none}{E1}{have}{have}}
+\> @,@
+\instr{F18}{pull s2}{F7}{}
+      {\annot{A0}{none}{E2}{have}{none}}
 
-  (merge label D2)
-              , F15 = jump                  F7  {x1 = b1}
-
-  (merge labels E0-E2)
-              , F16 = push s4 x2            F17 {}
-              , F17 = drop s2               F18 {}
-              , F18 = pull s2 x2            F7  {} } }
-\end{code}
+@} }@
+\end{tabbing}
 

### 2017-02-21 2017mergingmerges/code/src/Machine/Base.hs
code: cleanup expr language

@@ -1,4 +1,4 @@
-
+{-# LANGUAGE PatternSynonyms #-}
 module Machine.Base where
 import Data.Map                 (Map)
 import Data.Set                 (Set)
@@ -7,7 +7,7 @@ import qualified Data.Set       as Set
 
 
 -------------------------------------------------------------------------------
--- | Generic name type.
+-- | Generic name.
 type Name
         = String
 
@@ -20,7 +20,10 @@ data Channel
 
 -- | Variable name.
 data Var        
+        -- | Variable with the given name.
         = Var    Name
+
+        -- | Variable used as the buffer for some channel.
         | VarBuf Channel
         deriving (Show, Eq, Ord)
 
@@ -28,23 +31,54 @@ data Var
 -------------------------------------------------------------------------------
 -- | Expressions.
 data Expr
-        = XVal Value
-        | XVar Var
-        | XAdd Expr Expr
-        | XApp Expr Expr
+        = XVal Value            -- ^ Values.
+        | XVar Var              -- ^ Variables.
+        | XApp Expr Expr        -- ^ Generic application.
         deriving (Show, Eq)
+
+pattern XBool b = XVal (VLit (PBool b))
+pattern XInt  i = XVal (VLit (PInt i))
+pattern XAdd    = XVal (VLit PAdd)
 
 
 -- | Values.
 data Value
-        = VInt  Int
-        | VBool Bool
-        | VSucc 
-        | VAbs  Var Expr
+        = VAbs  Var  Expr       -- ^ Function abstraction.
+        | VLit  Prim            -- ^ Literal primitive value.
+        | VPAP  Prim [Value]    -- ^ Partially applied primitive.
+        deriving (Show, Eq)
+
+pattern VBool b = VLit (PBool b)
+pattern VInt  i = VLit (PInt  i)
+pattern VAdd    = VLit PAdd
+
+
+-- | Primitives.
+data Prim
+        = PBool Bool            -- ^ Boolean literal.
+        | PInt  Int             -- ^ Integer literal.
+        | PAdd                  -- ^ Addition primitive.
         deriving (Show, Eq)
 
 
+
 -------------------------------------------------------------------------------
+-- | Describes the state of the current value available from a channel.
+data InputState
+        -- | The value has not yet been read from the channel.
+        = None
+
+        -- | The value has been read from the channel 
+        --   and stored in the associated single-element buffer,
+        --   but has not yet been stored in a process-specific variable.
+        | Pending Value
+
+        -- | The value has been stored in a process specific variable,
[…]

### 2017-02-21 2017mergingmerges/code/src/Machine/Eval.hs

@@ -18,18 +18,15 @@ eval heap@(Heap hsHeap) xx
                 Just value      -> value
                 Nothing         -> error $ "eval: unbound variable " ++ show var
 
-        XAdd x1 x2
-         |  VInt i1             <- eval heap x1
-         ,  VInt i2             <- eval heap x2
-         -> VInt (i1 + i2)
-
         XApp x1 x2
-         |  VAbs var1 xBody     <- eval heap x1
-         ,  v2                  <- eval heap x2
-         -> error "eval: finish me"
+         -> case eval heap x1 of
+                VAbs var2 xBody -> error "eval: finish me"
 
-         |  VSucc               <- eval heap x1
-         ,  VInt i2             <- eval heap x2
-         -> VInt (i2 + 1)
+                VAdd
+                 |  v2          <- eval heap x2
+                 -> VPAP PAdd [v2]
 
+                VPAP PAdd [VInt i1]
+                 |  VInt i2     <- eval heap x2
+                 -> VInt (i1 + i2)
 

### 2017-02-21 2017mergingmerges/code/src/Machine/Base.hs
code: add def of group combinator

@@ -12,9 +12,10 @@ type Name
         = String
 
 
--- | Channel name.
+-- | Channel name,
+--   along with the type of the values that flow through the channel.
 data Channel
-        = Channel Name
+        = Channel Name Type
         deriving (Show, Eq, Ord)
 
 
@@ -28,7 +29,20 @@ data Var
         deriving (Show, Eq, Ord)
 
 
+-- | Get the type associated with a channel.
+typeOfChannel :: Channel -> Type
+typeOfChannel (Channel _ t) = t
+
+
+
 -------------------------------------------------------------------------------
+-- | Value types.
+data Type
+        = TBool                 -- ^ Type of booleans.
+        | TInt                  -- ^ Type of integers.
+        deriving (Show, Eq, Ord)
+
+
 -- | Expressions.
 data Expr
         = XVal Value            -- ^ Values.
@@ -36,10 +50,8 @@ data Expr
         | XApp Expr Expr        -- ^ Generic application.
         deriving (Show, Eq)
 
-pattern XBool b = XVal (VLit (PBool b))
-pattern XInt  i = XVal (VLit (PInt i))
-pattern XAdd    = XVal (VLit PAdd)
-
+infixl 5 @@
+(@@) = XApp
 
 -- | Values.
 data Value
@@ -48,18 +60,52 @@ data Value
         | VPAP  Prim [Value]    -- ^ Partially applied primitive.
         deriving (Show, Eq)
 
-pattern VBool b = VLit (PBool b)
-pattern VInt  i = VLit (PInt  i)
-pattern VAdd    = VLit PAdd
-
 
 -- | Primitives.
 data Prim
         = PBool Bool            -- ^ Boolean literal.
+        | POr                   -- ^ Boolean or.
+        | PAnd                  -- ^ Boolean and.
+
         | PInt  Int             -- ^ Integer literal.
-        | PAdd                  -- ^ Addition primitive.
+        | PAdd                  -- ^ Integer addition.
+
+        | PEq                   -- ^ Primitive equality   check.
+        | PNeq                  -- ^ Primitive inequality check.
         deriving (Show, Eq)
 
+
+pattern XBool b = XVal (VLit (PBool b))
+pattern XOr     = XVal (VLit  POr)
+pattern XAnd    = XVal (VLit  PAnd)
+pattern XInt  i = XVal (VLit (PInt i))
+pattern XAdd    = XVal (VLit  PAdd)
[…]

### 2017-02-21 2017mergingmerges/code/src/Machine/Combinator.hs

@@ -5,5 +5,6 @@ module Machine.Combinator
         , mkMap)
 where
 import Machine.Combinator.Map
+import Machine.Combinator.Group
 import Machine.Base
 import Machine.New

### 2017-02-21 2017mergingmerges/code/src/Machine/Combinator/Group.hs

@@ -1,0 +1,56 @@
+
+module Machine.Combinator.Group where
+import Machine.New
+import Machine.Base
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-- | Construct a new `group` process.
+--
+--   Group filters out consecutive duplicate values from a stream.
+--   If the stream is sorted then the output will contain only
+--   the unique entries in the input stream.
+--
+mkGroup :: Channel              -- ^ Input  Channel.
+        -> Channel              -- ^ Output Channel.
+        -> New Process
+
+mkGroup cIn cOut
+ = do
+        [l0, l1, l2, l3]
+                <- replicateM 4 (newLabel "G")
+        vFirst  <- newVar "First"
+        vLast   <- newVar "Last"
+        vVal    <- newVar "Val"
+
+        return  
+         $ Process
+         { processName          = "group"
+         , processIns           = Map.fromList [(cIn, None)]
+         , processOuts          = Set.fromList [cOut]
+
+         , processHeap          
+            = Heap $ Map.fromList 
+                [ (vFirst,      VBool True)
+                , (vLast,       defaultValueOfChannel cIn)
+                , (vVal,        defaultValueOfChannel cIn) ]
+
+         , processLabel         = l0
+
+         , processBlocks        =
+                [ (l0,  Pull    cIn vVal (next l1))
+
+                , (l1,  Case    (XOr @@ (XVar vFirst) 
+                                     @@ (XNeq @@ (XVar vLast) @@ (XVar vVal)))
+                                (next l2) (next l3))
+
+                , (l2,  Push    cOut (XVar vVal)
+                                (next' l3 [ (vLast,  XVar  vVal)
+                                          , (vFirst, XBool False)]))
+
+                , (l3,  Drop    cIn (next l0))
+                ]
+         }
+
+

### 2017-02-21 2017mergingmerges/code/src/Machine/Base.hs
code: add def of merge combinator

@@ -70,8 +70,12 @@ data Prim
         | PInt  Int             -- ^ Integer literal.
         | PAdd                  -- ^ Integer addition.
 
-        | PEq                   -- ^ Primitive equality   check.
-        | PNeq                  -- ^ Primitive inequality check.
+        | PEq                   -- ^ Polymorphic equality   check.
+        | PNeq                  -- ^ Polymorphic inequality check.
+        | PLt                   -- ^ Polymorphic less-than.
+        | PLe                   -- ^ Polymorphic less-than-equal.
+        | PGt                   -- ^ Polymorphic greater-than.
+        | PGe                   -- ^ Polymorphic greather-than-equal.
         deriving (Show, Eq)
 
 
@@ -82,6 +86,10 @@ pattern XInt  i = XVal (VLit (PInt i))
 pattern XAdd    = XVal (VLit  PAdd)
 pattern XEq     = XVal (VLit  PEq)
 pattern XNeq    = XVal (VLit  PNeq)
+pattern XLt     = XVal (VLit  PLt)
+pattern XLe     = XVal (VLit  PLe)
+pattern XGt     = XVal (VLit  PGt)
+pattern XGe     = XVal (VLit  PGe)
 
 pattern VBool b = VLit (PBool b)
 pattern VOr     = VLit  POr
@@ -90,6 +98,10 @@ pattern VInt  i = VLit (PInt  i)
 pattern VAdd    = VLit  PAdd
 pattern VEq     = VLit  PEq
 pattern VNeq    = VLit  PNeq
+pattern VLt     = VLit  PLt
+pattern VLe     = VLit  PLe
+pattern VGt     = VLit  PGt
+pattern VGe     = VLit  PGe
 
 
 -- | Get a default value of the given type.

### 2017-02-21 2017mergingmerges/code/src/Machine/Combinator.hs

@@ -6,5 +6,6 @@ module Machine.Combinator
 where
 import Machine.Combinator.Map
 import Machine.Combinator.Group
+import Machine.Combinator.Merge
 import Machine.Base
 import Machine.New

### 2017-02-21 2017mergingmerges/code/src/Machine/Combinator/Group.hs

@@ -12,8 +12,8 @@ import qualified Data.Set       as Set
 --   If the stream is sorted then the output will contain only
 --   the unique entries in the input stream.
 --
-mkGroup :: Channel              -- ^ Input  Channel.
-        -> Channel              -- ^ Output Channel.
+mkGroup :: Channel      -- ^ Input  Channel.
+        -> Channel      -- ^ Output Channel.
         -> New Process
 
 mkGroup cIn cOut
@@ -26,9 +26,9 @@ mkGroup cIn cOut
 
         return  
          $ Process
-         { processName          = "group"
-         , processIns           = Map.fromList [(cIn, None)]
-         , processOuts          = Set.fromList [cOut]
+         { processName  = "group"
+         , processIns   = Map.fromList [(cIn, None)]
+         , processOuts  = Set.fromList [cOut]
 
          , processHeap          
             = Heap $ Map.fromList 
@@ -36,9 +36,9 @@ mkGroup cIn cOut
                 , (vLast,       defaultValueOfChannel cIn)
                 , (vVal,        defaultValueOfChannel cIn) ]
 
-         , processLabel         = l0
+         , processLabel = l0
 
-         , processBlocks        =
+         , processBlocks =
                 [ (l0,  Pull    cIn vVal (next l1))
 
                 , (l1,  Case    (XOr @@ (XVar vFirst) 

### 2017-02-21 2017mergingmerges/section/XX-Fusion.tex
process evaluation

@@
 The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
-Figure~\ref{fig:Fusion:Types} shows the type definitions for fusion.
+Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.
 Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
@@
 
-Figure~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
+Fig.~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
 Function \ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
@@
 
-Function \ti{tryStepPair} defined in figure~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
+Function \ti{tryStepPair} defined in Fig.~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
 It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.
@@
 
-Figure~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
+Fig.~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
 It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.
@@
 
-Figure~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
+Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
 \ti{channels} computes the $\ChanType_2$ map for a pair of processes.

### 2017-02-21 2017mergingmerges/section/XX-Processes.tex
process evaluation

@@
 
-The grammar for processes is shown in figure~\ref{fig:Process:Def}.
+The grammar for processes is shown in Fig.~\ref{fig:Process:Def}.
 Channels, labels and variables are specified by some external, globally unique set of names.
-For values and expressions we use a simple untyped lambda calculus with a few primitives chosen to facilitate the examples.
+For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples.
 
-\amos{Settle on a terminology for channels / streams. Are both necessary?}
-Streams are the abstract data flowing through while channels are particular endpoints, so a stream can have multiple channels.
-A stream can have at most one output channel, and any number of input channels.
+We refer to the \emph{endpoint} of a stream as a channel.
+A stream may flow through many input channels on different processes, but can only be produced from at most one output channel.
 
-A process defines a stream computation, taking any number of input streams and producing at least one output stream.
-The input streams are paired with an input state which is used for coordinating multiple processes during evaluation; in process definitions before any evaluation has occurred this should be @none@.
-The input states are explained in detail later in \S\ref{s:Process:Eval}.
+Each input channel is paired with an input state, which can be either @none@ (no value), @have@ (a value has been pulled), or ($@pending@~\Value$) (a value is available to be pulled).
+These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
+These should be initialised to @none@ for a new process.
+Input states are explained in detail later in \S\ref{s:Process:Eval}.
 
@@
 The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
-Instructions can pull from a stream, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
+Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
 
@@
 Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
-After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer process to push to that stream.
+After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.
 
 All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
-% Combining variable update with stream instructions simplifies the fusion process in~\S\ref{s:Fusion}, as some parts of fusion need to perform both at once.
 
 A process network is a set of multiple processes that can be evaluated concurrently.
-Any inputs that are not produced as outputs of processes are assumed to be external inputs - their values will be provided by the environment.
+Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
 Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
@@
 
-We now describe evaluation of processes and process networks.
-We split evaluation into three main parts:
+Evaluation of processes and process networks is split into three parts:
 \begin{itemize}
-\item Injection, where values are inserted into a process' input buffer.
-Injection is only possible when the process input buffer is empty.
+\item Injection, where a single value for a stream is inserted into a process' input buffer.
+Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
+Injection for a process network proceeds by injecting the same value into all processes in the network.
+
 \item Shaking, where a process takes a step from one label to another.
-Shaking a process results in an updated process as well as an output message.
-If a process pushes to a stream, the push value must be able to be injected to other processes.
-\item Feeding, where an environment of input values are fed to the processes, and output values are collected.
-This is the `top level' of evaluation that uses both injection and shaking.
+Shaking a process results in an updated process and an optional pushed value.
+When a process pushes to a channel, the value is injected to other processes.
+Shaking a process network chooses any process that can be shaked, and updates it in the process network.
+
+\item Feeding, where an environment of input values are fed to the process network, and output values are collected.
+Feeding alternates between injecting and shaking: injecting values from the environment when processes are ready to receive values, and repeatedly shaking the processes to collect their outputs.
 \end{itemize}
@@
 
-Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
+% Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
 
-Rules in figure~\ref{fig:Process:Eval:Inject} are about injecting values into a process; these are the values used when the process performs a @pull@.
-The injected values may be pushed values from other processes for internal streams, or may come from an external source for the overall network's input streams.
-Injection is just about orchestrating values between processes, and no actual computation happens here; it just makes values available to be pulled.
+\subsubsection{Injection}
+Fig.~\ref{fig:Process:Eval:Inject} defines the judgment forms and rules for injection.
+Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.
 
-Injection can only happen when a process is ready to receive more input.
 A process has a single element buffer for each input, stored in its input state.
-This can be either @none@ meaning an empty buffer, @pending@ meaning a single value has been added to the buffer but has not been read yet, or @have@ meaning the value was added and in the process of being used.
+This can be either @none@, in which case the buffer is empty and a value may be injected; @pending@, where a single value has been injected to the buffer but has not been pulled yet; or @have@, where a value has already been injected and is currently being used.
 
-(InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
-An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded (or at least multiple element) buffer.
-Injecting the value puts the value as @pending@ in the buffer.
+The judgment form ($\ProcInject{p}{c}{v}{p'}$) means that the process $p$ has the value $v$ inserted to the input buffer for channel $c$, resulting in the updated process $p'$.
 
-(InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.
+Rule (InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
+An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded, or at least multiple element, buffer.
+We use the syntax ($\HeapUpdateOne{@ins@~p}{c}{@pending@~v}$) to perform a map update, in this case removing the ($c=@none@$) from ($@ins@~p$) and then inserting ($c=@pending@~v$).
+This sets the input buffer as pending, meaning that any future attempts to inject to this channel will fail, until the process pulls and then drops the value.
 
-(ProcessesInject) performs injection over a process network.
+Rule (InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.
+
+Rule (ProcessesInject) performs injection over a process network.
 Every process in the network must have the value injected into it.
@@
 
-The rules in figure~\ref{fig:Process:Eval:Shake} are the `shake' part of evaluation, where actual computation occurs. 
-As usual, $\alpha$ denotes the message type, with $\tau$ being an internal message. The \Push~ message is a single value being output on a channel.
+\subsubsection{Shaking}
+Fig.~\ref{fig:Process:Eval:Shake} defines the judgment forms and rules for shaking a process, allowing it to take a step, and produce an output message.
+% I don't like them, but alpha and tau are standard names in process calculi. 
+The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.
 
-The judgment form for shaking a single instruction $\ProcBlockShake{b}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
-executes an instruction $b$ with the input states $i$ and the heap $\Sigma$.
-The output message $\alpha$ can be an internal state change or an emitted value.
-The result also has the new label, the new input state buffers, and the substitution to apply to the heap.
+The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
+executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
+The result contains an output message $\alpha$, as well as the new label $l'$, the new input state $i'$, and the update substitution to apply to the heap $u'$.
 
-The two judgment forms for shaking processes are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p}}$.
-The process shaking just shakes a single instruction and updates the process.
-Shaking a process network chooses a single process to shake, then if the result is an emitted value, that value is injected into all the other processes in the network.
-
-(Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
+Rule (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
 The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.
 
-(Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull.
+Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.
 
-(Push) evaluates the push expression $e$ under the heap, and sends the value as the message.
+Rule (Push) evaluates the push expression $e$ under the heap, and sends that value as output the message.
 
-(Jump) simply returns the new label and substitution.
+Rule (Jump) simply returns the new label and substitution.
 
-(CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.
+Rules (CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.
 
-(Shake) unwraps a single process and evaluates the instruction.
-The instruction updates are evaluated and updated in the process heap.
-It updates the process with the new label, input state and heap.
+The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
+The process shaking shakes a single instruction and updates the process.
+Shaking a process network chooses a single process to shake and updates it in the network.
+If shaking the process resulted in an output value, that value is injected into all the other processes in the network.
 
-(ProcessesInternal) chooses one process from the network and evaluates it.
-When the process evaluates with an internal message ($\tau$), the entire network evaluates by replacing that process.
+Rule (Shake) takes a single process and shakes its instruction.
+The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
+The process steps to the new label with the new input state, and the values are updated in the heap.
 
-(ProcessesPush) chooses one process from the network and evaluates it, where the process evaluates with a push message.
-The emitted push message is then injected into all other processes in the network, which means they must either ignore the channel or be ready to add it to their buffer.
+Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces an internal message ($\tau$).
+
+Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces a push message.
+The emitted push message is then injected into all other processes in the network, meaning they must either ignore the channel or be ready to receive it in their buffer.
 If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
@@
 
-The rules in figure~\ref{fig:Process:Eval:Feed} are the `feed' part of evaluation, where external input values are fed into a process network and output values are accumulated.
+\subsubsection{Feeding}
+Fig.~\ref{fig:Process:Eval:Feed} defines the judgment forms and rules for feeding, where external input values are fed into a process network and output values are accumulated.
 The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
@@
 
-(FeedStart) is the axiom form where all input values have been injected and there are no input values left.
-This is the start of evaluation.
+Rule (FeedStart) applies when all input values have been injected and there are no input values left.
+In this case, the output values are the same as the input values.
 
-(FeedInternal) first recursively feeds its input accumulator and process network, then allows the resulting network to take an internal step.
-The internal step does not affect the accumulators.
-% The recursive closure is performed \emph{before} the internal step rather than after for proof engineering reasons: it allows an extra step to be added to the end of a feed evaluation relatively easily.
+Rule (FeedInternal) allows the process network to take an internal step.
+It first feeds its input accumulator and process network, then allows the resulting network to take an internal step.
 
-(FeedPush) works similarly to (FeedInternal) except that the process network emits a push message.
-The pushed value is added to the end of the accumulator for that channel.
+Rule (FeedPush) allows the process network to emit a push message.
+As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message.
+The pushed value is collected in the accumulator list for that stream.
 
-(FeedExternal) allows inputs to be injected into the process network.
+Rule (FeedExternal) allows inputs to be injected into the process network.
 For any channel $c$ which is not an output of one of the processes, we take the last value off its list.

### 2017-02-21 2017mergingmerges/code/src/Machine/Combinator.hs
code: add fusion rules for jump and case

@@ -2,7 +2,9 @@
 module Machine.Combinator
         ( module Machine.Base
         , module Machine.New
-        , mkMap)
+        , mkMap
+        , mkGroup
+        , mkMerge)
 where
 import Machine.Combinator.Map
 import Machine.Combinator.Group

### 2017-02-21 2017mergingmerges/code/src/Machine/Fuse.hs

@@ -78,7 +78,7 @@ processesAreConnected process1 process2
 
 
 ---------------------------------------------------------------------------------------------------
--- | Fuse to processes.
+-- | Fuse two processes.
 fusePair :: Process -> Process -> Either String Process
 fusePair process1 process2
  = let  
@@ -191,17 +191,56 @@ tryStep csMode
 
  = case instr1 of
 
+        -- Jump ---------------------------------------------------------------
+        Jump (Next label1' xvsUpdate)
+         -> Just $ Jump 
+                 $ Next (LabelJoint (label1', csState1)
+                                    (label2,  csState2))
+                        xvsUpdate
+
+
+        -- Case ---------------------------------------------------------------
+        Case xScrut (Next labelA xvsUpdateA) 
+                    (Next labelB xvsUpdateB)
+         -> Just
+         $  Case xScrut
+                  (Next (LabelJoint (labelA, csState1) (label2, csState2)) xvsUpdateA)
+                  (Next (LabelJoint (labelB, csState1) (label2, csState2)) xvsUpdateB)
+
+
+        -- Push ---------------------------------------------------------------
+        Push c xx (Next label1' xvsUpdate)
+         -- (LocalPush)
+         --   Exclusive output, so no other coordination is required.
+         |  Just ModeOutput       <- csMode   ? c
+         -> Just $ Push c xx
+                 $ Next (LabelJoint (label1', csState1)
+                                    (label2,  csState2))
+                        xvsUpdate
+
+         -- (SharedPush)
+         --   Connected output, and the downstream process is ready for the value.
+         |  Just ModeConnected    <- csMode   ? c
+         ,  Just ModeNone         <- csState2 ? c
+         -> Just $ Jump 
+                 $ Next (LabelJoint (label1', csState1)
+                                    (label2,  Map.insert c ModePending csState2))
+                        (Map.insert (VarBuf c) xx xvsUpdate)
+
+
         -- Pull ---------------------------------------------------------------
         Pull c x (Next label1' xvsUpdate)
-         -- First process has exclusive use of the input channel.
+         -- (LocalPull)
+         --   First process has exclusive use of the input channel.
          |  Just ModeInputExclusive <- csMode ? c
          -> Just $ Pull c x 
                  $ Next (LabelJoint (label1', csState1) 
                                     (label2,  csState2))
                         xvsUpdate
 
-         -- Input channel is used by both processes,
[…]

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute.hs
code: cleanup module hierarchy, and update tests

@@ -1,8 +1,8 @@
 
 module Machine.Execute where
+import Machine.Execute.Shake
+import Machine.Execute.Feed
 import Machine.Base
-import Machine.Shake
-import Machine.Feed
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Eval.hs

@@ -1,5 +1,5 @@
 
-module Machine.Eval where
+module Machine.Execute.Eval where
 import Machine.Base
 import Data.Map                         (Map)
 import qualified Data.Map.Strict        as Map

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Feed.hs

@@ -1,7 +1,7 @@
 
-module Machine.Feed where
+module Machine.Execute.Feed where
+import Machine.Execute.Inject
 import Machine.Base
-import Machine.Inject
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Inject.hs

@@ -1,5 +1,5 @@
 
-module Machine.Inject where
+module Machine.Execute.Inject where
 import Machine.Base
 import Data.Map                 (Map)
 import qualified Data.Map       as Map

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Shake.hs

@@ -1,8 +1,8 @@
 
-module Machine.Shake where
+module Machine.Execute.Shake where
+import Machine.Execute.Inject
+import Machine.Execute.Eval
 import Machine.Base
-import Machine.Inject
-import Machine.Eval
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map

### 2017-02-21 2017mergingmerges/code/src/Machine/Transform/Fuse.hs

@@ -1,5 +1,5 @@
 {-# LANGUAGE TupleSections #-}
-module Machine.Fuse where
+module Machine.Transform.Fuse where
 import Machine.Base
 import Data.Map                 (Map)
 import Data.Set                 (Set)

### 2017-02-21 2017mergingmerges/code/src/Test.hs

@@ -1,0 +1,12 @@
+
+import Machine.Base
+
+import Test.Process.Map
+import Test.Eval.Map
+import Test.Fuse.GroupMerge
+import Test.Fuse.Map
+
+
+main 
+ = putStrLn "test"
+ 

### 2017-02-21 2017mergingmerges/code/src/Test/Eval/Map.hs

@@ -1,0 +1,95 @@
+
+module Test.Eval.Map where
+import Machine.Transform.Fuse
+import Machine.Combinator
+import Machine.Execute
+import Text.Show.Pretty
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-------------------------------------------------------------------------------
+-- | Test evaluation of single map process.
+testEvalMap
+ = putStr $ ppShow 
+ $ let  
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+        pMap1   = evalNew $ mkMap cAs cBs xSucc
+
+        cvsInput
+         = Map.fromList
+                [ (cAs, map VInt [1, 2, 3])]
+
+        cvsOutput
+         = Map.fromList
+                [ (cBs, []) ]
+
+        (inputs', processes', actions')
+         = execute cvsInput cvsOutput [pMap1] []
+
+   in   (inputs', processes', actions')
+
+
+-------------------------------------------------------------------------------
+-- | Test evaluation of pipelined map map.
+testEvalPipeMapMap
+ = putStr $ ppShow 
+ $ let  
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        cCs     = Channel "cs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+        (pMap1, pMap2)
+         = evalNew
+         $ do   pMap1   <- mkMap cAs cBs xSucc
+                pMap2   <- mkMap cBs cCs xSucc
+                return (pMap1, pMap2)
+
+        cvsInput
+         =  Map.fromList
+                [ (cAs, map VInt [1, 2, 3])]
+
+        cvsOutput
+         = Map.fromList
+                [ (cCs, [])]
+
+        (inputs', processes', actions')
+         = execute cvsInput cvsOutput [pMap1, pMap2] []
+
+   in   (inputs', processes', actions')
+
+
+-------------------------------------------------------------------------------
+-- | Test evaluation of split map-map
+testEvalSplitMapMap
+ = putStr $ ppShow 
+ $ let  
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        cCs     = Channel "cs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+        (pMap1, pMap2)
[…]

### 2017-02-21 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs

@@ -1,0 +1,46 @@
+
+module Test.Fuse.GroupMerge where
+import Machine.Transform.Fuse
+import Machine.Combinator
+import Machine.Execute
+import Text.Show.Pretty
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-------------------------------------------------------------------------------
+testFuseSplitGroupMerge
+ = putStr $ ppShow 
+ $ let  
+        cIn1    = Channel "In2"    TInt
+        cIn2    = Channel "In1"    TInt
+        cUniq   = Channel "Uniq1"  TInt
+        cMerged = Channel "Merged" TInt
+
+        (pGroup, pMerge)
+         = evalNew 
+         $ do   pGroup  <- mkGroup cIn1 cUniq
+                pMerge  <- mkMerge cIn1 cIn2 cMerged
+                return (pGroup, pMerge)
+
+        result -- Right pFused
+         = fusePair pGroup pMerge
+
+   in   (pGroup, pMerge, result)
+
+{-   
+        cvsInput
+         =  Map.fromList
+                [ (cIn1, map VInt [1, 3, 3, 6, 7, 7, 8])
+                , (cIn2, map VInt [1, 2, 3, 4, 5, 6, 7, 8]) ]
+
+        cvsOutput
+         = Map.fromList
+                [ (cUniq,  [])
+                , (cMerged, []) ]
+
+        (_inputs, _processes, actions)
+         = execute cvsInput cvsOutput [pFused] []
+
+   in   (pGroup, pMerge, pFused, actions)
+-}

### 2017-02-21 2017mergingmerges/code/src/Test/Fuse/Map.hs

@@ -1,0 +1,80 @@
+
+module Test.Fuse.Map where
+import Machine.Transform.Fuse
+import Machine.Combinator
+import Machine.Execute
+import Text.Show.Pretty
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-------------------------------------------------------------------------------
+-- | Test evaluation of pipelined map map.
+--   Note that with a finite input stream the last value is still
+--   in the channel buffer when the process ends.
+testFusePipeMapMap
+ = putStr $ ppShow 
+ $ let  
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        cCs     = Channel "cs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+        (pMap1, pMap2)
+         = evalNew 
+         $ do   pMap1   <- mkMap cAs cBs xSucc 
+                pMap2   <- mkMap cBs cCs xSucc
+                return (pMap1, pMap2)
+
+        modes           = processChannelModes   pMap1 pMap2 
+        bConn           = processesAreConnected pMap1 pMap2
+        Right pOut      = fusePair pMap1 pMap2
+
+        cvsInput
+         =  Map.fromList
+                [ (cAs, [VInt 1, VInt 2, VInt 3, VInt 4, VInt 5]) ]
+
+        cvsOutput
+         = Map.fromList
+                [ (cCs, []) ]
+
+        (inputs', processes', actions')
+         = execute cvsInput cvsOutput [pOut] []
+
+   in   (pMap1, pMap2, modes, bConn, pOut, actions')
+
+
+-------------------------------------------------------------------------------
+-- | Test evaluation of split map map.
+testFuseSplitMapMap
+ = putStr $ ppShow 
+ $ let  
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        cCs     = Channel "cs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+        (pMap1, pMap2)
+         = evalNew 
+         $ do   pMap1   <- mkMap cAs cBs xSucc
+                pMap2   <- mkMap cAs cCs xSucc
+                return (pMap1, pMap2)
+
+        modes           = processChannelModes   pMap1 pMap2 
+        bConn           = processesAreConnected pMap1 pMap2
+        Right pOut      = fusePair pMap1 pMap2
+
[…]

### 2017-02-21 2017mergingmerges/code/src/Test/Main.hs

@@ -1,0 +1,14 @@
+
+import Machine.Base
+
+import Test.Process.Map
+
+import Test.Eval.Map
+
+import Test.Fuse.GroupMerge
+import Test.Fuse.Map
+
+
+main 
+ = putStrLn "test"
+ 

### 2017-02-21 2017mergingmerges/code/src/Test/Process/Map.hs

@@ -1,0 +1,36 @@
+
+module Test.Process.Map where
+import Machine.Execute
+import Machine.Combinator
+import Text.Show.Pretty
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-- | Test construction of map process.
+testProcessMap :: IO ()
+testProcessMap
+ = putStr $ ppShow 
+ $ let  cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+   in   evalNew $ mkMap cAs cBs xSucc
+
+
+
+-- | Test contruction of pipelined map map process.
+testProcessMapMap :: IO ()
+testProcessMapMap
+ = putStr $ ppShow 
+ $ let
+        cAs     = Channel "as" TInt
+        cBs     = Channel "bs" TInt
+        cCs     = Channel "cs" TInt
+        xSucc x = XApp (XApp XAdd (XInt 1)) x
+
+   in   evalNew
+         $ do   pMap1   <- mkMap cAs cBs xSucc
+                pMap2   <- mkMap cBs cCs xSucc
+                return (pMap1, pMap2)
+

### 2017-02-21 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs
code: add joint label stripper

@@ -1,6 +1,7 @@
 
 module Test.Fuse.GroupMerge where
 import Machine.Transform.Fuse
+import Machine.Transform.StripLabels
 import Machine.Combinator
 import Machine.Execute
 import Text.Show.Pretty
@@ -11,22 +12,24 @@ import qualified Data.Set       as Set
 -------------------------------------------------------------------------------
 testFuseSplitGroupMerge
  = putStr $ ppShow 
- $ let  
-        cIn1    = Channel "In2"    TInt
-        cIn2    = Channel "In1"    TInt
-        cUniq   = Channel "Uniq1"  TInt
-        cMerged = Channel "Merged" TInt
+ $ evalNew
+ $ do
+        let cIn1    = Channel "In2"    TInt
+        let cIn2    = Channel "In1"    TInt
+        let cUniq   = Channel "Uniq1"  TInt
+        let cMerged = Channel "Merged" TInt
 
-        (pGroup, pMerge)
-         = evalNew 
-         $ do   pGroup  <- mkGroup cIn1 cUniq
-                pMerge  <- mkMerge cIn1 cIn2 cMerged
-                return (pGroup, pMerge)
+        pGroup  <- mkGroup cIn1 cUniq
+        pMerge  <- mkMerge cIn1 cIn2 cMerged
 
-        result -- Right pFused
-         = fusePair pGroup pMerge
+        let Right pResult
+                = fusePair pGroup pMerge
 
-   in   (pGroup, pMerge, result)
+        let pResult' 
+                = evalNew 
+                $ stripLabels "F" pResult
+
+        return $ (pGroup, pMerge, pResult')
 
 {-   
         cvsInput

### 2017-02-21 2017mergingmerges/code/src/Machine/Pretty.hs
code: add pretty printer for processes

@@ -1,0 +1,203 @@
+{-# LANGUAGE TypeSynonymInstances #-}
+{-# LANGUAGE FlexibleInstances    #-}
+
+module Machine.Pretty where
+import Machine.Base
+import Text.PrettyPrint.Leijen
+import Data.Map                 (Map)
+import Data.Set                 (Set)
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+instance (Pretty a, Pretty b) 
+       => Pretty (Map a b) where
+ pretty mp
+        =   parens
+        $   string "map" <+> indent 0
+        (   vsep [parens $ string "def" <+> pretty x <+> pretty y
+                 | (x, y) <- Map.toList mp])
+
+
+instance Pretty a 
+      => Pretty (Set a) where
+ pretty st
+        =   parens
+        $   string "set" <+> indent 0
+        (   vsep $ map pretty $ Set.toList st)
+
+
+
+instance Pretty Name where
+ pretty str     
+  = string $ show str
+
+
+instance Pretty Channel where
+ pretty (Channel n t)
+  = parens 
+        $   string "channel" 
+        <+> pretty n
+        <+> pretty t
+
+
+instance Pretty Var where
+ pretty vv
+  = case vv of
+        Var n           -> pretty n
+        VarBuf c        -> parens $ string "buf" <+> pretty c
+
+
+instance Pretty Type where
+ pretty tt
+  = case tt of
+        TBool           -> string "bool"
+        TInt            -> string "int"
+
+
+instance Pretty Expr where
+ pretty xx
+  = case xx of
+        XVal v          -> pretty v
+        XVar n          -> parens $ string "var" <+> pretty n
+        XApp x1 x2      -> parens $ string "app" <+> pretty x1 <+> pretty x2
+
+
+instance Pretty Value where
+ pretty vv 
+  = case vv of
+        VAbs v x        -> parens $ string "abs" <+> pretty v  <+> pretty x
+        VLit l          -> pretty l
+        VPAP p vs       -> parens $ string "pap" <+> pretty p  <+> pretty vs
+
+
+instance Pretty Prim where
+ pretty pp
+  = case pp of
+        PBool True      -> parens $ string "bool true" 
+        PBool False     -> parens $ string "bool false"
+        POr             -> string "or"
+        PAnd            -> string "and"
+        PInt i          -> parens $ string "int" <+> int i
+        PAdd            -> string "add"
+        PEq             -> string "eq"
[…]

### 2017-02-21 2017mergingmerges/code/src/Test.hs

@@ -1,10 +1,12 @@
 
 import Machine.Base
+import Machine.Pretty
 
 import Test.Process.Map
 import Test.Eval.Map
 import Test.Fuse.GroupMerge
 import Test.Fuse.Map
+
 
 
 main 

### 2017-02-21 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs

@@ -4,16 +4,26 @@ import Machine.Transform.Fuse
 import Machine.Transform.StripLabels
 import Machine.Combinator
 import Machine.Execute
+import Machine.Pretty
 import Text.Show.Pretty
+import Text.PrettyPrint.Leijen
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
-
+import System.IO
 
 -------------------------------------------------------------------------------
 testFuseSplitGroupMerge
- = putStr $ ppShow 
- $ evalNew
- $ do
+ = do   let (_pGroup, _pMerge, lsDef, pResult')
+                = evalNew $ testFuseSplitGroupMerge'
+
+        displayIO stdout
+         $ renderPretty 0 100
+         $ vcat [ pretty lsDef,         empty
+                , pretty pResult',      empty]
+
+
+testFuseSplitGroupMerge'
+ = do 
         let cIn1    = Channel "In2"    TInt
         let cIn2    = Channel "In1"    TInt
         let cUniq   = Channel "Uniq1"  TInt
@@ -25,11 +35,11 @@ testFuseSplitGroupMerge
         let Right pResult
                 = fusePair pGroup pMerge
 
-        let pResult' 
+        let (pResult', lsDef)
                 = evalNew 
                 $ stripLabels "F" pResult
 
-        return $ (pGroup, pMerge, pResult')
+        return $ (pGroup, pMerge, lsDef, pResult')
 
 {-   
         cvsInput

### 2017-02-21 2017mergingmerges/code/src/Machine/Base.hs
code: finishing evaluator

@@ -55,8 +55,7 @@ infixl 5 @@
 
 -- | Values.
 data Value
-        = VAbs  Var  Expr       -- ^ Function abstraction.
-        | VLit  Prim            -- ^ Literal primitive value.
+        = VLit  Prim            -- ^ Literal primitive value.
         | VPAP  Prim [Value]    -- ^ Partially applied primitive.
         deriving (Show, Eq)
 
@@ -102,6 +101,31 @@ pattern VLt     = VLit  PLt
 pattern VLe     = VLit  PLe
 pattern VGt     = VLit  PGt
 pattern VGe     = VLit  PGe
+
+
+-- | Get the literal primitive from a value, 
+--   if this is one.
+takePrimOfValue :: Value -> Maybe Prim
+takePrimOfValue vv
+ = case vv of
+        VLit p          -> Just p
+        _               -> Nothing
+
+-- | Get the number of parameters of a primitive.
+arityOfPrim :: Prim -> Int
+arityOfPrim p
+ = case p of
+        PBool{}         -> 0
+        POr             -> 2
+        PAnd            -> 2
+        PInt{}          -> 0
+        PAdd            -> 2
+        PEq             -> 2
+        PNeq            -> 2
+        PLt             -> 2
+        PLe             -> 2
+        PGt             -> 2
+        PGe             -> 2
 
 
 -- | Get a default value of the given type.

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Eval.hs

@@ -11,7 +11,7 @@ eval   :: Heap -> Expr -> Value
 eval heap@(Heap hsHeap) xx
  = case xx of
         XVal value
-         -> value
+         ->  value
 
         XVar var
          -> case Map.lookup var hsHeap of
@@ -20,13 +20,43 @@ eval heap@(Heap hsHeap) xx
 
         XApp x1 x2
          -> case eval heap x1 of
-                VAbs var2 xBody -> error "eval: finish me"
+                VLit p
+                 -> evalPrimValue p [eval heap x2]
 
-                VAdd
-                 |  v2          <- eval heap x2
-                 -> VPAP PAdd [v2]
+                VPAP p vs
+                 -> evalPrimValue p (vs ++ [eval heap x2])
 
-                VPAP PAdd [VInt i1]
-                 |  VInt i2     <- eval heap x2
-                 -> VInt (i1 + i2)
+
+-- | Evaluate a primitive applied to some values.
+evalPrimValue :: Prim -> [Value] -> Value
+evalPrimValue p vs
+ | length vs == arityOfPrim p
+ = case (do ps <- sequence $ map takePrimOfValue vs
+            evalPrimPrim p ps) of
+        Nothing -> error $ "evalValue: prim evaluation failure " ++ show (p, vs)
+        Just p' -> VLit p'
+
+
+ | length vs < arityOfPrim p
+ = VPAP p vs
+
+ | otherwise
+ = error $ "evalValue: over application of primitive " ++ show (p, vs)
+
+
+-- | Evaluate a primitive operator applied to some primitive arguments.
+evalPrimPrim  :: Prim -> [Prim]  -> Maybe Prim
+evalPrimPrim pp ps
+ = case (pp, ps) of
+        (POr,  [PBool b1, PBool b2])    -> Just $ PBool (b1 || b2)
+        (PAnd, [PBool b1, PBool b2])    -> Just $ PBool (b1 && b2)
+        (PAdd, [PInt  i1, PInt  i2])    -> Just $ PInt  (i1 +  i2)
+        (PEq,  [PInt  i1, PInt  i2])    -> Just $ PBool (i1 == i2)
+        (PNeq, [PInt  i1, PInt  i2])    -> Just $ PBool (i1 /= i2)
+        (PLt,  [PInt  i1, PInt  i2])    -> Just $ PBool (i1 <  i2)
[…]

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Shake.hs

@@ -2,17 +2,21 @@
 module Machine.Execute.Shake where
 import Machine.Execute.Inject
 import Machine.Execute.Eval
+import Machine.Pretty
 import Machine.Base
 import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map
-
+import Text.PrettyPrint.Leijen
 
 -- | An output action from a Shake evaluation.
 data Action
         = ActionPush Channel Value
         deriving Show
 
+instance Pretty Action where
+ pretty (ActionPush c v)
+  = parens $ text "push" <+> pretty c <+> pretty v
 
 ---------------------------------------------------------------------------------------------------
 shakeSteps 

### 2017-02-21 2017mergingmerges/code/src/Machine/Pretty.hs

@@ -66,7 +66,6 @@ instance Pretty Expr where
 instance Pretty Value where
  pretty vv 
   = case vv of
-        VAbs v x        -> parens $ string "abs" <+> pretty v  <+> pretty x
         VLit l          -> pretty l
         VPAP p vs       -> parens $ string "pap" <+> pretty p  <+> pretty vs
 
@@ -136,7 +135,7 @@ instance Pretty Next where
 
 prettySetVar (v, x)
  = parens
-        $   pretty "set"
+        $   string "set"
         <+> pretty v
         <+> pretty x
 

### 2017-02-21 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs

@@ -13,21 +13,25 @@ import System.IO
 
 -------------------------------------------------------------------------------
 testFuseSplitGroupMerge
- = do   let (_pGroup, _pMerge, lsDef, pResult')
+ = do   let (lsDef, pFused, lsActions)
                 = evalNew $ testFuseSplitGroupMerge'
 
         displayIO stdout
          $ renderPretty 0 100
-         $ vcat [ pretty lsDef,         empty
-                , pretty pResult',      empty]
+         $ vcat [ pretty lsDef
+                , empty
+                , pretty pFused
+                , empty
+                , pretty (vcat $ map pretty lsActions)
+                , empty ]
 
 
 testFuseSplitGroupMerge'
  = do 
-        let cIn1    = Channel "In2"    TInt
-        let cIn2    = Channel "In1"    TInt
-        let cUniq   = Channel "Uniq1"  TInt
-        let cMerged = Channel "Merged" TInt
+        let cIn1    = Channel "cIn2"    TInt
+        let cIn2    = Channel "cIn1"    TInt
+        let cUniq   = Channel "cUniq1"  TInt
+        let cMerged = Channel "cMerged" TInt
 
         pGroup  <- mkGroup cIn1 cUniq
         pMerge  <- mkMerge cIn1 cIn2 cMerged
@@ -35,25 +39,21 @@ testFuseSplitGroupMerge'
         let Right pResult
                 = fusePair pGroup pMerge
 
-        let (pResult', lsDef)
+        let (pFused, lsDef)
                 = evalNew 
                 $ stripLabels "F" pResult
 
-        return $ (pGroup, pMerge, lsDef, pResult')
-
-{-   
-        cvsInput
-         =  Map.fromList
+        let cvsInput
+                =  Map.fromList
                 [ (cIn1, map VInt [1, 3, 3, 6, 7, 7, 8])
                 , (cIn2, map VInt [1, 2, 3, 4, 5, 6, 7, 8]) ]
 
-        cvsOutput
-         = Map.fromList
-                [ (cUniq,  [])
+        let cvsOutput
+                = Map.fromList
+                [ (cUniq,   [])
                 , (cMerged, []) ]
 
-        (_inputs, _processes, actions)
-         = execute cvsInput cvsOutput [pFused] []
+        let (_inputs, _processes, actions)
+                = execute cvsInput cvsOutput [pFused] []
 
-   in   (pGroup, pMerge, pFused, actions)
--}
+        return (lsDef, pFused, actions)

### 2017-02-21 2017mergingmerges/code/src/Machine/Execute/Feed.hs
code: fix execution to make fused group/merge example work

@@ -26,12 +26,19 @@ feedProcessList
           , [Process])
 
 feedProcessList []  ps
- =      return ([], ps)
+ =      Nothing
 
 feedProcessList (cvs : cvss) ps 
- = do   (cvs',  ps')    <- feedProcess1    cvs  ps  
-        (cvss', ps'')   <- feedProcessList cvss ps' 
-        return (cvs' : cvss', ps'')
+ = case feedProcess1 cvs ps of
+        Just (cvs', ps') 
+         -> Just (cvs' : cvss, ps')
+
+        Nothing 
+         -> case feedProcessList cvss ps of
+                Just (cvss', ps')
+                 -> Just (cvs : cvss', ps')
+
+                Nothing -> Nothing
 
 
 feedProcess1 

### 2017-02-23 2017mergingmerges/section/XX-Related.tex
temporary notes about synchronous languages and regular dataflow

@@
 % Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.
+
+\subsection{Scratch: dataflow}
+Synchronous languages (LUSTRE, Lucy-n, SIGNAL) view streams as -- abstractly -- functions from time to values.
+(Or time to maybe values, if they allow holes).
+This forces them into an inherently \emph{push} based system, but what is pushing here is the procession of time.
+If you have two streams, you cannot pull from one and the other: when time advances, it advances for all streams.
+Thus, synchronous languages cannot encode value-dependent pulling - nor can they encode append.
+Append could be seen as a special kind of value-dependent access pattern, if you treat the end of the stream as a special kind of value.
+
+Non-synchronous streaming languages, such as Lucid, may allow some kind of value-dependent pulling, because they do not enforce causality.
+However, they rely on dynamic scheduling.
+
+Do not to confuse regular dataflow, also known as \emph{synchronous data flow}, with synchronous languages.
+Regular dataflow graphs are dataflow graphs where each edge is annotated with the rate.
+
+Dataflow languages:
+\begin{itemize}
+\item Lucid Synchrone
+\item LUSTRE \cite{halbwachs1991synchronous}.
+Dataflow language: causal analysis, clock calculus, synchronous.
+Uses more or less syntactic equality to check equivalence of clocks.
+\item Lucy-n: extension of LUSTRE with periodic clocks to ensure static scheduling \cite{mandel2010lucy}, \cite{benveniste2003synchronous}
+\item SIGNAL \cite{le2003polychrony}
+causal, clocks, synchronous
+\item StreaMIT \cite{thies2002streamit}:
+Regular/synchronous dataflow, ie statically known rates.
+Only allows limited splits and joins: round robin and duplication for splits, round robin and combination for joins. 
+Does not support fully general graphs - instead using combinators to introduce a (split/join) and a combinator for a feedback loop.
+\item Prelude: extension of LUSTRE, seems very similar to Lucy-n \cite{pagetti2011multi}
+\item Ptolemy (II)
+\end{itemize}
+
+Models:
+\begin{itemize}
+\item Synchronous dataflow (SDF):
+all rates are statically known,
+\item Cyclo-static dataflow (CSDF) where the number of produced/consumed tokens varies periodically, sounds very similar to Lucy-n's periodic clocks \cite{mandel2010lucy}
+\item Boolean dataflow (BDF), generally undecidable to statically schedule, so must fall back to dynamic scheduling \cite{buck1993scheduling}
+\item Integer dataflow (IDF), (also falls back to dynamic) \cite{buck1994static}
+Clustering to find control structure: tries to recover structures like ifs, cases and loops from the dataflow graph.
+Extends previous BDF \cite{buck1993scheduling} with loops that repeat some number of times, and multi-way ifs.
+Very rigid and only supports limited control flow structures.
+Unclear how a merge or append would be clustered by this system.
+\item Parameterized dataflow (PDF),  \cite{bhattacharya2001parameterized}
+\item Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}
+\end{itemize}
+
+Dataflow languages tend to fall back on less performant dynamic scheduling, when programs cannot be statically scheduled.
+For example \citet{bouakaz2013real} uses static scheduling for special cases like cyclo-static dataflow graphs, but otherwise uses dynamic earliest deadline first (EDF) scheduling.
+
+When dataflow languages talk about a `merge' operator, they are talking about \emph{non-deterministic merge}. These are not Kahn processes.

### 2017-02-24 2017mergingmerges/code/src/Machine/Combinator.hs
code: add Filter, Scan and state counting code

@@ -2,11 +2,19 @@
 module Machine.Combinator
         ( module Machine.Base
         , module Machine.New
+
+        -- Single input, single output.
         , mkMap
+        , mkFilter
+        , mkScan
         , mkGroup
+
+        -- Multi-input, single output.
         , mkMerge)
 where
 import Machine.Combinator.Map
+import Machine.Combinator.Filter
+import Machine.Combinator.Scan
 import Machine.Combinator.Group
 import Machine.Combinator.Merge
 import Machine.Base

### 2017-02-24 2017mergingmerges/code/src/Test.hs

@@ -8,6 +8,7 @@ import Test.Eval.Map
 import Test.Fuse.GroupMerge
 import Test.Fuse.Map
 import Test.Comb
+import Test.Stat
 
 import Machine.Combinator.Group
 import Machine.Combinator.Map
@@ -22,7 +23,30 @@ main
  = putStrLn "test"
  
 
+combs1 
+ =      [ combMapSucc
+        , combFilterPos
+        , combScanAdd
+        , combGroup 
+        , combMerge ]
+
+counts cs
+        = putStrLn
+        $ unlines
+        $ map (show . statOfProcess) 
+        $ map (evalNew . mkComb) 
+        $ cs
 
 
+countPipes n
+        = counts 
+        $ manyPipeAB n combs1
 
+countSplits n 
+        = counts 
+        $ manySplitAB n combs1
 
+countSplitPipe n
+        = counts
+        $ manySplitAB n
+        $ concat [manyPipeAB i combs1 | i <- [0..n]]

### 2017-02-24 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs

@@ -28,8 +28,8 @@ testFuseSplitGroupMerge
 
 testFuseSplitGroupMerge'
  = do 
-        let cIn1    = Channel "cIn2"    TInt
-        let cIn2    = Channel "cIn1"    TInt
+        let cIn1    = Channel "cIn1"    TInt
+        let cIn2    = Channel "cIn2"    TInt
         let cUniq   = Channel "cUniq1"  TInt
         let cMerged = Channel "cMerged" TInt
 

### 2017-02-24 2017mergingmerges/section/01-Introduction.tex
Work on intro

@@
 
-Suppose we have two large arrays containing textual identifiers, and wish to perform some analysis on them. Both arrays are sorted, but may contain duplicate identifiers. 
-We wish to retrieve the unique set of identifiers from the first array, as well as produce the union of both arrays.
-Can we perform both of these tasks at once, without reading the input arrays multiple times?
+Suppose we have two input streams of textual identifiers, and wish to perform some analysis on these identifiers.
+The identifiers from both streams arrive sorted, but may include duplicates.
+We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of identifiers from both streams.
+Can we perform both of these tasks at once, without needing to read through the stream data multiple times, and without needing unbounded buffering?
+Here is how we might write the source code, where @S@ is for @S@-tream.
 \begin{code}
-  uniquesUnion : Array Text -> Array Text -> (Array Text, Array Text)
-  uniquesUnion arr1 arr2
-   = let  uniques1 = group arr1
-          merged   = merge arr1 arr2
-          union    = group merged
-     in   (uniques1, union)
+  uniquesUnion : S Text -> S Text -> (S Text, S Text)
+  uniquesUnion sIn1 sIn2
+   = let  sUnique = group sIn1
+          sMerged = merge sIn1 sIn2
+          sUnion  = group sMerged
+     in   (sUnique, sUnion)
 \end{code}
 
-In this implementation the @group@ combinator filters out consecutive duplicates, while @merge@ combines two sorted arrays so that the output remains sorted.
-This example has a few interesting attributes.
-Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ reads values from @arr1@ or @arr2@ depends on the values in the arrays themselves.
-If all the values in @arr1@ are smaller than the values in @arr2@, then @merge@ will pull values from @arr1@ before pulling from @arr2@, and vice versa.
-Secondly, although @arr1@ occurs twice in the program, at runtime we only want to read through the elements of this array once.
-To achieve this, the compiled program must coordinate between the two uses of @arr1@, such that values are only read when both @uniques1@ and @merged@ are ready to receive a new value.
-Finally, although this program is expressed in terms of bulk \emph{arrays}, at runtime we want it to execute using some version of \emph{streaming data flow}, in order to save memory bandwidth, avoid repeated computation, and eliminate the need for intermediate buffers such as @merged@.
+In this implementation the @group@ operator filters out consecutive duplicates, while @merge@ combines two sorted streams so that the output remains sorted.
+This example has a few interesting properties.
+Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which this operator pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
+If all the values from @sIn1@ are smaller than the values in @sIn2@, then @merge@ will pull all values from @sIn1@ before pulling the rest from @sIn2@, and vice versa.
+Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once.
+To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that values are only read when both the @group@ and @merge@ operators are ready to receive a new value.
+Finally, as the stream length is assumed to be unbounded, we cannot buffer an aribrary number of elements read from either stream, or risk running out of local storage space.
 
-The above example seems innocuous, but existing fusion transformations cannot handle it.
-As observed by \citet{kay2009you}, both pull- and push-based fusion have fundamental limitations.
-Pull-based systems such as stream fusion~\cite{coutts2007stream} cannot perform fusion when a particular array or intermediate result is used by multiple consumers.
-We refer to this situation as a \emph{split} --- when the example above is viewed as a data-flow network the flow from the argument array @arr1@ is split into both the @group@ and @merge@ consumers.
-Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work - in this case reading the array multiple times.
-Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse this example because they do not support combinators with multiple inputs.
-We refer to such a situation as a \emph{join} --- in the example above the @merge@ combinator expresses a join in the data-flow graph.
-Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of combinators.
-More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse this case, but requires the program to be rewritten to use explicitly polarized stream types, rather than the more familiar arrays.
-Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent combinators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
-The polyhedral model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out combinators such as @group@ and @merge@.
+For an implementation which does \emph{not} use stream fusion, we might implement each of the operators as a separate concurent process, and send each identifier value using an intra-process communication mechanism.
+Developing such an implementation could be easy or hard, depending on what language features are available for concurrency.
+However, worring about the \emph{performance tuning} of such a system, such as whether we need backpressure, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
 
-In our new system we view the program as a concurrent process network, where each combinator is a separate process, and the streams are communication channels between the processes.
-Each combinator is expressed as a restricted, sequential imperative program with commands such as @pull@ for reading from an input channel and @push@ for writing to an output channel.
+We might instead define some sort of uniform interface for data sources, with a single `pull' function that provides the next value in each stream.
+Each operator could be given this interface, so that the next value from each result stream is computed on demand.
+This is approach is commonly taken with implementations of physical operators in data base systems.
+However, this `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
+If a consumer pulls many elements from the result @sUnique@ stream, then the implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@.
+It needs to buffer an aribrary number of elements from @sIn2@ because there is no guarantee of when a consumer will also pull from the @sUnion@ result stream.
+Once that happens the elements from @sIn2@ no longer need to be retained, but not before.
+
+Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requring unbounded buffering.
+Sadly, existing stream fusion transformations cannot handle our example.
+As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations.
+Pull-based systems such as short cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers.
+We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.
+Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.
+
+Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs.
+We refer to such a situation as a \emph{join} --- in our example the @merge@ combinator expresses a join in the data-flow graph.
+Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
+More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.
+
+% The mechanism that combines the implementations of both operators, to yield efficient imperative code also depends on the general purpose compiler optimisations implemented by GHC, and it can be difficult to tell if these have ``worked'' without inspecting the intermediate representations of the compiler.
+
+Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
+The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out operators such as @group@ and @merge@.
+
+In our new system we still view the program as a concurrent process network, where each combinator is a separate process, and the stream data flows through communication channels between the processes.
+Each combinator is then expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
 The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
+When the fusion transformation succeeds we know worked, so there is no need to inspect intermediate representations of the compiler to debug performance problems, which is a common problem with fusion systems based on rewrite rules and other general program transformations \cite{lippmeier2012:guiding}
 
-We make the following contributions:
+In summary, we make the following contributions:
 \begin{itemize}
 \item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
-\item an algorithm for fusing processes, the first which supports splits, joins and arbitrary combinators (\S\ref{s:Fusion});
-\item a description of the required extensions to support finite streams (\S\ref{s:Finite});
+\item an algorithm for fusing processes, which is the first to support splits, joins and arbitrary combinators (\S\ref{s:Fusion});
 \item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs});
-\item and show our processes are general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
 \end{itemize}
+
+Our fusion transformation for infinite stream programms could also serve as the basis for an \emph{array} fusion system, by extending it to work with finite streams. We discuss 
+this extension in \S\ref{s:Finite}.
+
+% TODO: We can't make the appendix a contribution because the reviewers are not required to read appendices.
+% \item and show our processes are general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
 

### 2017-02-24 2017mergingmerges/section/01-Introduction.tex
Make running example take Nats to match the next section

@@
 
-Suppose we have two input streams of textual identifiers, and wish to perform some analysis on these identifiers.
+Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers.
 The identifiers from both streams arrive sorted, but may include duplicates.
@@
 \begin{code}
-  uniquesUnion : S Text -> S Text -> (S Text, S Text)
+  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
   uniquesUnion sIn1 sIn2

### 2017-02-24 2017mergingmerges/section/XX-Related.tex
Notes on synchronous/passive dataflow and scenario-aware dataflow

@@
 (Or time to maybe values, if they allow holes).
+Synchrony is described as ``time advances in lockstep with one or more clocks'' \cite{benveniste2003synchronous}.
 This forces them into an inherently \emph{push} based system, but what is pushing here is the procession of time.
@@
 Append could be seen as a special kind of value-dependent access pattern, if you treat the end of the stream as a special kind of value.
+
+Contrast between our case, where we have \emph{active} processes which perform the pulling and pushing, with synchronous cases which are more passive, in that they don't request input - the values trickle in from outside.
 
@@
 Regular dataflow graphs are dataflow graphs where each edge is annotated with the rate.
+The closest thing to fusion in dataflow computation is static scheduling, where the order in which different actors are fired at runtime is chosen statically.
+This is somewhat similar to fusion, in fusion we are also interested in merging the separate actors into a single one that computes them all.
+Choosing a static schedule is required to fuse actors together, but it alone is not fusion.
+In our system, we are performing both static scheduling and fusion at the same time.
 
@@
 \item Lucid Synchrone
+``takes advantage of the (first-order) functional aspect of Lustre so as to generalize it to higher order constructs in an ML-like style.
+In particular, the Lustre clock calculus is extended and inferred as an ML-type system.'' \cite{benveniste2003synchronous}
 \item LUSTRE \cite{halbwachs1991synchronous}.
@@
 Uses more or less syntactic equality to check equivalence of clocks.
-\item Lucy-n: extension of LUSTRE with periodic clocks to ensure static scheduling \cite{mandel2010lucy}, \cite{benveniste2003synchronous}
+\item Lucy-n: extension of LUSTRE with periodic clocks to ensure static scheduling \cite{mandel2010lucy}
 \item SIGNAL \cite{le2003polychrony}
@@
 Does not support fully general graphs - instead using combinators to introduce a (split/join) and a combinator for a feedback loop.
-\item Prelude: extension of LUSTRE, seems very similar to Lucy-n \cite{pagetti2011multi}
 \item Ptolemy (II)
@@
 \item Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}
+\item Scenario aware dataflow (FSM-SADF) \cite{stuijk2011scenario}
+Somehow use a separate finite state machine, which can control the rates of dataflows.
+Apparently more expressive/succinct than boolean dataflow (BDF) but also easier to analyze.
+`General SADF' must fall back to runtime scheduling, while FSM-SADF can be done statically.
+In FSM-SADF, however, the currently executing scenario can only be switched in between iterations of the whole graph.
+So after the whole graph executes one scenario, the FSM is used to find the next scenario to execute.
+This means that the currently executing scenario cannot depend on values read during the current scenario - which rules out paritioning, for example, because the choice of whether to emit on the true or false branch cannot be made until after the value has been emitted.
+It is possible to indirectly encode this by storing the value to emit in a buffer, and deferring the emit until next scenario - however this loses locality benefits of fusion.
+General SDF could describe a partition, but uses runtime scheduling.
+The focus is more about analysing worst-case execution time in order to ensure real time deadlines can be met \cite{van2015scenario}.
+
+Basically each node of the FSM is a dataflow graph.
+After the current dataflow graph is executed, the FSM can take a step based on the output of that graph.
+Then the graph at the new node is executed, and so on.
+Because the FSM is separate to the dataflow graph, it is not as composable as a combinator-based approach: adding the equivalent to a new combinator could require modifying many of the dataflow graphs, and the FSM, in non-trivial ways.
 \end{itemize}

### 2017-02-24 2017mergingmerges/section/01-Introduction.tex

@@
 To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that values are only read when both the @group@ and @merge@ operators are ready to receive a new value.
-Finally, as the stream length is assumed to be unbounded, we cannot buffer an aribrary number of elements read from either stream, or risk running out of local storage space.
+Finally, as the stream length is assumed to be unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.
 
-For an implementation which does \emph{not} use stream fusion, we might implement each of the operators as a separate concurent process, and send each identifier value using an intra-process communication mechanism.
+For an implementation which does \emph{not} use stream fusion, we might implement each of the operators as a separate concurrent process, and send each identifier value using an intra-process communication mechanism.
 Developing such an implementation could be easy or hard, depending on what language features are available for concurrency.
-However, worring about the \emph{performance tuning} of such a system, such as whether we need backpressure, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
+However, worrying about the \emph{performance tuning} of such a system, such as whether we need back-pressure, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
 
@@
 
-Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requring unbounded buffering.
+Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requiring unbounded buffering.
 Sadly, existing stream fusion transformations cannot handle our example.
@@
 We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.
-Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.
+
+% Leave this to related work. We've already mentioned a canonical pull-based system.
+% Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.
 
 Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs.
-We refer to such a situation as a \emph{join} --- in our example the @merge@ combinator expresses a join in the data-flow graph.
+We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow graph.
 Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
@@
 
-In our new system we still view the program as a concurrent process network, where each combinator is a separate process, and the stream data flows through communication channels between the processes.
-Each combinator is then expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
+In our new system we still view the program as a concurrent process network, where each operator is a separate process, and the stream data flows through communication channels between the processes.
+Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
 The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
-When the fusion transformation succeeds we know worked, so there is no need to inspect intermediate representations of the compiler to debug performance problems, which is a common problem with fusion systems based on rewrite rules and other general program transformations \cite{lippmeier2012:guiding}
+When the fusion transformation succeeds we know it has worked.
+There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.
 
@@
 \item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
-\item an algorithm for fusing processes, which is the first to support splits, joins and arbitrary combinators (\S\ref{s:Fusion});
+\item an algorithm for fusing these processes, the first to support arbitrary splits and joins (\S\ref{s:Fusion});
+\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. \TODO{Ref}
 \item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs});
@@
 
-Our fusion transformation for infinite stream programms could also serve as the basis for an \emph{array} fusion system, by extending it to work with finite streams. We discuss 
-this extension in \S\ref{s:Finite}.
+Our fusion transformation for infinite stream programs could also serve as the basis for an \emph{array} fusion system, using a natural extension to finite streams.
+We discuss this extension in \S\ref{s:Finite}.
 

### 2017-02-24 2017mergingmerges/section/01-Introduction.tex
Fixing start of S2, also pagination

@@
 However, this `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
-If a consumer pulls many elements from the result @sUnique@ stream, then the implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@.
+Suppose a consumer pulls many elements from the result @sUnique@ stream.
+The implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@.
 It needs to buffer an aribrary number of elements from @sIn2@ because there is no guarantee of when a consumer will also pull from the @sUnion@ result stream.
@@
 We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow graph.
-Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
+Some systems support both pull and push: data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
 More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.
@@
 Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
-The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out operators such as @group@ and @merge@.
+The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level.
+The polyhedral model is based around affine loops, which makes it difficult to support for filter-like operators such as @group@ and @merge@.
 
-In our new system we still view the program as a concurrent process network, where each operator is a separate process, and the stream data flows through communication channels between the processes.
+In our new system we still view the program as a concurrent process network.
+Each operator is a separate process, and the stream data flows through communication channels between the processes.
 Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
-The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
+The fusion transform takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
 When the fusion transformation succeeds we know it has worked.

### 2017-02-24 2017mergingmerges/section/XX-Processes.tex
Fixing start of S2, also pagination

@@
 %!TEX root = ../Main.tex
-\section{Processes}
+\section{Processes, Combinators and Operators}
 \label{s:Processes}
 
-Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap.
-The process pulls data from an arbitrary number of input streams and pushes data to at least one output stream.
-Each input array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
-% \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.
-
-% make it explicit that the processes are only an intermediate repr
-The process language is used for the implementation of combinators and as an intermediate representation for fusion, while the original program using these combinators can be written without any knowledge of processes.
+A \emph{process} in our system is a simple imperative program with a local heap.
+A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream.
+The process language is an intermediate representation we use when fusing the overall dataflow network.
+A \emph{combinator} is a template for a particular \emph{process} which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
+Each process implements a logical \emph{operator}, so we use ``operator'' when describing the values being computed, but ``process'' when referring to the implementation.
 
@@
 \subsection{Grouping}
-\begin{figure}
-
-\begin{center}
-\begin{alltt}
-           group 
-             = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat). 
-               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
-\end{alltt}
-\begin{code}
-               process
-               { ins:    { s1 }
-               , outs:   { s2 }
-               , heap:   { f = T, l = 0, v = 0 }
-               , label:  A0
-               , instrs: { A0 = pull s1 v            A1 {}
-                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
-                         , A2 = push s2 v            A3 { l = v, f = F }
-                         , A3 = drop s1              A0 {} } }
-\end{code}
-\end{center}
-\vspace{1em}
-\includegraphics[scale=1.1]{figures/state-group.pdf}
-\caption{The group combinator}
-\label{fig:Process:Group}
-\end{figure}
 
 The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}.
-We include the concrete code representation as well as a diagram of the state machine defined by the associated stream process.
+We include the concrete code representation and a diagram of the process viewed as a state machine.
 
@@
 The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
-The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates here we leave it empty.
+The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates yet we leave it empty.
 
@@
 
-When the above predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
+When the predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
 
 Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@.
-This @drop@ instruction is used to coordinate processes when performing fusion and concurrent evaluation; the next element of a stream may only be pulled after all consuming processes have pulled and dropped the current element.
-% This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm and concurrent evaluation, but has no direct operational meaning.
-% We discuss @drop@ further in \TODO{ref}.
+This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
+The next element of a stream may only be pulled after all consuming processes have pulled and then and dropped the current element.
 
@@
 The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
-For example, when executed on the input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.
+For example, when executed on the input stream $[1, 2, 2, 3]$, the process will produce the output $[1, 2, 3]$.
 
-Note that we refer to @group@ itself as a \emph{combinator} as it is parameterised by the names of the streams it deals with (@s1@ and @s2@).
-Once the combinator has been instantiated to work on particular streams the result is a concrete \emph{process}, which is the code representation of a state machine.
+\begin{figure}
 
-% Recall that @group@ removes consecutive duplicates from its input stream.
-% It has one input, @file1@, and one output, @uniques@.
-
-% There are three variables in the heap: @first@ is initialised to @True@ as the first pulled value has no last value to compare against; @value@ stores the most recently pulled value, and @last@ stores the last pulled value.
-
-% The @last@ and @value@ variables are initialised to $0$, but could be initialised to any value: these initial values will not be used.
-% The initial label is @L0@, which pulls from the input @file1@. This blocks waiting for an input value, and when one is received, it is stored in @value@ and the process moves to @L1@.
-% Instruction @L1@ performs a case analysis on a boolean: if it is the first read value, or the last value is not equal to the most recent value, it jumps to @L2@; otherwise it jumps to @L3@.
-% Instruction @L2@ pushes the most recent value to the output and jumps to @L3@, updating the last value to the most recent value, and setting first to @False@.
-% Finally, the instruction for @L3@ `drops' the recently pulled value from @file1@ and jumps back to @L0@.
-% This dropping is required to coordinate between multiple processes that read from the same input: now that the read value from @file1@ has been dropped, another process is free to pull the next value from @file1@ if it so wishes.
+\begin{center}
+\begin{alltt}
+           group 
+             = \(\lambda\) (sG1: Stream Nat) (sG2: Stream Nat). 
+               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
+\end{alltt}
+\begin{code}
+               process
+               { ins:    { sG1 }
+               , outs:   { sG2 }
+               , heap:   { f = T, l = 0, v = 0 }
+               , label:  A0
+               , instrs: { A0 = pull sG1 v           A1 {}
+                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
+                         , A2 = push sG2 v           A3 { l = v, f = F }
+                         , A3 = drop sG1             A0 {} } }
+\end{code}
+\end{center}
+\vspace{1em}
+\includegraphics[scale=1.1]{figures/state-group.pdf}
+\caption{The group combinator}
+\label{fig:Process:Group}
+\end{figure}
 
@@
                merge
-                 = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat) (s3: Stream Nat). 
+                 = \(\lambda\) (sM1: Stream Nat) (sM2: Stream Nat) (sM3: Stream Nat). 
                    \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
@@
 The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
-The two heap variables @x1@ and @x2@ are used to store the currently read values from each input.
-The process starts by pulling from both input streams, then compares the two values, and pushes the smaller of the values to the output stream.
-It then drops the input stream corresponding to the smaller value, then pulls from the same stream, and performs the same comparison again.
-% The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section.
-% We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.
+The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream.
+The process starts by pulling from each input streams.
+It then compares the two pulled values, and pushes the smaller of the values to the output stream.
+The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.
 
@@
 For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@.
-We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
-
-% The instructions at @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
-% Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
-% Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
-% Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.
+After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
 
 % -----------------------------------------------------------------------------
-\clearpage{}
 \subsection{Fusion}
 Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
-For example, suppose we need to produce a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
-We want a result machine that computes both a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-Compared to the @uniquesUnion@ example, we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second array (@arr2@).
+For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
+The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
+For ease of comparison we will assume the parameters in each combinator are instantiated with arguments of the same names.
 
@@
 \subsubsection{Fusing Pulls}
-\begin{figure}
-\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
-\caption{Fusing pull instructions}
-\label{fig:Fusion:Pulls}
-\end{figure}
 
@@
 For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
-This means that the result state @F0@ represents a combination of the two input states @A0@ and @B0@.
+The result state @F0@ represents a combination of the two input states @A0@ and @B0@.
 As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.
@@
 For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.
-After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
 
@@
 
-\eject{}
+\begin{figure}
+\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
+\caption{Fusing pull instructions}
+\label{fig:Fusion:Pulls}
+\end{figure}
+
 % -----------------------------------------------------------------------------
@@
 
+\subsection{Optimisation}
+After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
+
+\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}
+
+% ----------------------------------------------------------------------------------
 \clearpage{}

### 2017-02-24 2017mergingmerges/section/XX-Definition.tex
Split into some more sections, shift appendix to the end

--- 2017-02-24 2017mergingmerges/section/XX-Processes.tex
@@
 %!TEX root = ../Main.tex
-\section{Processes, Combinators and Operators}
-\label{s:Processes}
 
-A \emph{process} in our system is a simple imperative program with a local heap.
-A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream.
-The process language is an intermediate representation we use when fusing the overall dataflow network.
-A \emph{combinator} is a template for a particular \emph{process} which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
-Each process implements a logical \emph{operator}, so we use ``operator'' when describing the values being computed, but ``process'' when referring to the implementation.
-
-% -----------------------------------------------------------------------------
-\subsection{Grouping}
-
-The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}.
-We include the concrete code representation and a diagram of the process viewed as a state machine.
-
-The @group@ combinator has two parameters, @s1@ and @s2@, which bind the input and output streams respectively.
-The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
-
-The body of the combinator is a record that defines the process.
-The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams.
-The @heap@ field gives the initial values of each of the local variables.
-The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
-
-The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
-The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates yet we leave it empty.
-
-Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to the instruction at label @A3@.
-We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
-
-When the predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
-
-Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@.
-This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
-The next element of a stream may only be pulled after all consuming processes have pulled and then and dropped the current element.
-
-Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
-The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
-For example, when executed on the input stream $[1, 2, 2, 3]$, the process will produce the output $[1, 2, 3]$.
-
-\begin{figure}
-
-\begin{center}
-\begin{alltt}
-           group 
-             = \(\lambda\) (sG1: Stream Nat) (sG2: Stream Nat). 
-               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
-\end{alltt}
-\begin{code}
-               process
-               { ins:    { sG1 }
-               , outs:   { sG2 }
-               , heap:   { f = T, l = 0, v = 0 }
-               , label:  A0
-               , instrs: { A0 = pull sG1 v           A1 {}
-                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
-                         , A2 = push sG2 v           A3 { l = v, f = F }
-                         , A3 = drop sG1             A0 {} } }
-\end{code}
-\end{center}
-\vspace{1em}
-\includegraphics[scale=1.1]{figures/state-group.pdf}
-\caption{The group combinator}
-\label{fig:Process:Group}
-\end{figure}
-
-% -----------------------------------------------------------------------------
-\subsection{Merging}
-\begin{figure}
-\begin{alltt}
-               merge
-                 = \(\lambda\) (sM1: Stream Nat) (sM2: Stream Nat) (sM3: Stream Nat). 
-                   \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
-\end{alltt}
-\begin{code}
-                   process
-                   { ins:    { s1, s2 }
-                   , outs:   { s3 }
-                   , heap:   { x1 = 0, x2 = 0 }
-                   , label:  B0
-                   , instrs: { B0 = pull s1 x1     B1 {}
-                             , B1 = pull s2 x2     C0 {}
-                             , C0 = case (x1 < x2) D0 {}  E0 {}
-                             , D0 = push s3 x1     D1 {}
-                             , D1 = drop s1        D2 {}
-                             , D2 = pull s1 x1     C0 {}
-                             , E0 = push s3 x2     E1 {}
-                             , E1 = drop s2        E2 {}
-                             , E2 = pull s2 x2     C0 {} } }
-\end{code}
-
-\medskip
-\includegraphics[scale=1.1]{figures/state-merge.pdf}
-\caption{The merge combinator}
-\label{fig:Process:Merge}
-\end{figure}
-
-The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}.
-The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
-The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream.
-The process starts by pulling from each input streams.
-It then compares the two pulled values, and pushes the smaller of the values to the output stream.
-The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.
-
-As this process merges infinite streams, if we execute it with a finite input prefix, the result will be an intermediate state that may not yet have pushed all available output.
-For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
-
-% -----------------------------------------------------------------------------
-\subsection{Fusion}
-Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
-For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
-The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-For ease of comparison we will assume the parameters in each combinator are instantiated with arguments of the same names.
-
-% -----------------------------------------------------------------------------
-\subsubsection{Fusing Pulls}
-
-The algorithm proceeds by considering pairs of states: one in each of the machines to be fused.
-Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Fig.~\ref{fig:Fusion:Pulls}.
-The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
-In the result machine we produce three new instructions that transition between four result states, @F0@ to @F3@.
-Each of the result states represents a combination of two original states, one from each of the input machines.
-For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.
-
-We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream, which is used to determine whether another value can be pulled.
-For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
-The result state @F0@ represents a combination of the two input states @A0@ and @B0@.
-As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.
-
-From the joint state @F0@, both of the input machines need to pull from stream @s1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
-In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
-For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.
-
-Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
-The `none' state means that we have not yet pulled a value from the associated stream.
-The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case).
-The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
-In Fig.~\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.
-
-\begin{figure}
-\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
-\caption{Fusing pull instructions}
-\label{fig:Fusion:Pulls}
-\end{figure}
-
-% -----------------------------------------------------------------------------
-\subsubsection{Fusing Cases}
-\begin{figure}
-\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
-\caption{Fusing case instructions}
-\label{fig:Fusion:Case}
-\end{figure}
-
-Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
-Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
-From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @s1@ to its output stream, which we name @s3@ --- or to just move on to the next value from its input.
-From state @B1@, the @merge@ machine needs to pull a value from its second input stream @s2@.
-
-In the fused result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
-At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
-
-Finally, at @F5@ the @merge@ machine pulls from @s2@, moving from @B1@ to @C0@.
-Because the stream @s2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
-In this case, we simply perform a @pull@ instruction.
-
-In terms of fusion, we could construct the result machine in multiple ways.
-One option is to perform the case branch first and then pull from @s2@.
-We could alternately pull from @s2@ first and then perform the case branch.
-By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
-As the set of variables does not overlap, either ordering is correct.
-For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.
-
-% -----------------------------------------------------------------------------
-\subsection{Fused Result}
-
-Fig.~\ref{fig:Process:Fused} shows the entire result of fusing @group@ and @merge@ together.
-The process has two inputs, @s1@ and @s2@, which correspond to the original input streams.
-Of the two outputs, @s3@ is the @group@'s output stream, while @s4@ is the @merge@'s output stream.
-The heap contains the variables from both processes, as well as the new buffer variable @b1@, which is used to store the current element pulled from @s1@.
-
-For the most part, the fused process simply alternates between performing the instructions of @group@ and those of @merge@.
-For the shared input stream @s1@, the two processes must be coordinated so that values are only pulled or dropped when both agree.
-
-The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
-
-The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
-These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
-Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.
-
-Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
-As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
-In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped.
-The actual @drop@ is only performed once both processes are `none', at label @F9@.
-
-\begin{figure}
-\begin{code}
-process
-{ ins:    { s1, s2 }
-, outs:   { s3, s4 }
-, heap:   { f = T, l = 0, v = 0, x1 = 0, x2 = 0, b1 = 0}
-, label:  F0
-, instrs:
-\end{code}
-
-\newcommand\annot[5]{
-  \tiny ((#1, \> \tiny \{s1 =      #2\}), \> \tiny (#3, \> \tiny \{s1 =      #4, \> \tiny s2 =      #5\}))
-}
-\newcommand\icase[7]{
- \tt{#1} \> \tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \tt{#5} \{\tt{#6}\} \> \tiny #1 \> \tiny = #7 \\
-}
-\newcommand\instr[5]{
- \tt{#1}\>\tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \> \tiny #1 \> \tiny = #5 \\
-}
-\begin{tabbing}
-@  @ \=
-@, F17@ \= = @case (f || (l /= v)) @
-      \= @F17{x1 = b1, f = F}@ \= @F17{ }@
-@      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
-\> @{@ \instr{F0}{pull s1 b1}{F1}{}{\annot{A0}{none}{B0}{none}{none}}
-
-\> @,@
-\instr{F1}{jump}{F2}{v  = b1}
-      {\annot{A0}{pending}{B0}{pending}{none}}
-\> @,@
-\instr{F2}{jump}{F3}{x1 = b1}
-      {\annot{A1}{have}{B0}{pending}{none}}
-\> @,@
-\icase{F3}{case (f || (l /= v))}{F4}{}{F5}{}
-      {\annot{A1}{have}{B1}{have}{none}}
-
-\> @,@
-\instr{F4}{push s3 v}{F5}{l = v, f = F}
-      {\annot{A2}{have}{B1}{have}{none}}
-
-\> @,@
-\instr{F5}{jump}{F6}{}
-      {\annot{A3}{have}{B1}{have}{none}}
-
-\> @,@
-\instr{F6}{pull s2 x2}{F7}{}
-      {\annot{A0}{none}{B1}{have}{none}}
-\\
-
-\> @,@
-\icase{F7}{case (x1 < x2)}{F8}{}{F16}{}
-      {\annot{A0}{none}{C0}{have}{have}}
-
-\\
-\> @,@
-\instr{F8}{push s4 x1}{F9}{}
-      {\annot{A0}{none}{D0}{have}{have}}
-\> @,@
-\instr{F9}{drop s1}{F10}{}
-      {\annot{A0}{none}{D1}{none}{have}}
-\> @,@
-\instr{F10}{pull s1 b1}{F11}{}
-      {\annot{A0}{none}{D2}{none}{have}}
-\> @,@
-\instr{F11}{jump}{F11}{v = b1}
-      {\annot{A0}{pending}{D2}{pending}{have}}
-\> @,@
-\icase{F12}{case (f || (l /= v))}{F13}{}{F14}{}
-      {\annot{A1}{have}{D2}{pending}{have}}
-\> @,@
-\instr{F13}{push s3 v}{F14}{l = v, f = F}
-      {\annot{A2}{have}{D2}{pending}{have}}
-\> @,@
-\instr{F14}{jump}{F15}{}
-      {\annot{A3}{have}{D2}{pending}{have}}
-\> @,@
-\instr{F15}{jump}{F7}{x1 = b1}
-      {\annot{A0}{none}{D2}{pending}{have}}
-
-\\
-\> @,@
-\instr{F16}{push s4 x2}{F17}{}
-      {\annot{A0}{none}{E0}{have}{have}}
-\> @,@
-\instr{F17}{drop s2}{F18}{}
-      {\annot{A0}{none}{E1}{have}{have}}
-\> @,@
-\instr{F18}{pull s2}{F7}{}
-      {\annot{A0}{none}{E2}{have}{none}}
-
-@} }@
-\end{tabbing}
-
-% With original label names and 'n', 'h' and 'p' for input state
-%\begin{code}
-%A0nB0n = pull s1 b1  A0pB0p{}
-%A0pB0p = jump        A1hB0p{v  = b1}
-%A1hB0p = jump        A1hB1h{x1 = b1}
-%
-%A1hB1h = case (f || (l /= v))
-%                     A2hB1h{}
-%                     A3hB1h{}
-%
-%A2hB1h = push s3 v   A3hB1h{l  = v, f  = F}
-%A3hB1h = jump        A0nB1h{}
-%A0nB1h = pull s2 x2  A0nC0h{}
-%
-%A0nC0h = case (x1 < x2)
-%                     A0nD0h{}
-%                     A0nE0h{}
-%
-%A0nD0h = push s4 x1  A0nD1h{}
-%A0nD1h = drop s1     A0nD2n{}
-%A0nD2n = pull s1 b1  A0pD2p{}
-%A0pD2p = jump        A1hD2p{v  = b1}
-%
-%A1hD2p = case (f || (l /= v))
-%                     A2hD2p{}
-%                     A3hD2p{}
-%
-%A2hD2p = push s3 v   A3hD2p{l  = v, f  = F}
-%A3hD2p = jump        A0nD2p{}
-%A0nD2p = jump        A0nC0h{x1 = b1}
-%
-%A0nE0h = push s4 x2  A0nE1h{}
-%A0nE1h = drop s2     A0nE2h{}
-%A0nE2h = pull s2 x2  A0nC0h{}
-%\end{code}
-
-% Cleaned up, simplified, variables collated
-% \begin{code}
-%             process
-%             { ins:    { s1, s2 }
-%             , outs:   { s3, s4 }
-%             , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
-%             , label:  F0
-%             , instrs: { F0 = pull s1 x1              F3 {}
-%                       , F3 = case (f || (l /= x1))   F4 {} F5 {}
-%                       , F4 = push s3 x1              F5 { l = x1, f = F }
-%                       , F5 = pull s2 x2              F6 {}
-% 
-%                       , F6 = case (x1 < x2)          A0 {} B0 {}
-% 
-%                       , A0 = push s4 x1              A1 {}
-%                       , A1 = drop s1                 A2 {}
-%                       , A2 = pull s1 x1              A3 {}
-%                       , A3 = case (f || (l /= x1))   A4 {} F6 {}
-%                       , A4 = push s3 x1              F6 { l = x1, f = F }
-% 
-%                       , B0 = push s4 x2              B1 {}
-%                       , B1 = drop s2                 B2 {}
-%                       , B2 = pull s2 x2              F6 {} } }
-% \end{code}
-\caption{Fusion of group and merge}
-\label{fig:Process:Fused}
-\end{figure}
-
-\subsection{Optimisation}
-After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
-
-\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}
-
-% ----------------------------------------------------------------------------------
 \clearpage{}
-\subsection{Process definitions}
+\section{Process definitions}
 
@@
 
-\subsection{Evaluation}
+% -----------------------------------------------------------------------------
+\subsection{Execution}
 \label{s:Process:Eval}

### 2017-02-24 2017mergingmerges/section/XX-Evaluation.tex
Split into some more sections, shift appendix to the end

%!TEX root = ../Main.tex

\section{Evaluation}

\TODO{Discuss the performance of the system. Why this is a good idea. Give graphs that show the number of states is sensible when more processes are fused in.}

### 2017-02-24 2017mergingmerges/section/XX-Processes.tex
Split into some more sections, shift appendix to the end

@@
 \TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}
-
-% ----------------------------------------------------------------------------------
-\clearpage{}
-\subsection{Process definitions}
-
-%% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
-\input{figures/ProcessDef.tex}
-
-The grammar for processes is shown in Fig.~\ref{fig:Process:Def}.
-Channels, labels and variables are specified by some external, globally unique set of names.
-For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples.
-
-We refer to the \emph{endpoint} of a stream as a channel.
-A stream may flow through many input channels on different processes, but can only be produced from at most one output channel.
-
-Each input channel is paired with an input state, which can be either @none@ (no value), @have@ (a value has been pulled), or ($@pending@~\Value$) (a value is available to be pulled).
-These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
-These should be initialised to @none@ for a new process.
-Input states are explained in detail later in \S\ref{s:Process:Eval}.
-
-The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
-This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
-A process may, however, produce multiple output streams.
-
-The heap is used for evaluation of expressions.
-Each process has its own private heap, therefore the only communication between processes occurs by streams.
-
-The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
-Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
-
-As usual with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
-Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
-After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.
-
-All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
-
-A process network is a set of multiple processes that can be evaluated concurrently.
-Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
-Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
-By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.
-
-% \subsection{Map/map}
-% \label{s:Process:MapMap}
-% 
-% One of the simplest combinators is @map@.
-% This might need to go elsewhere.
-% As well as needing a better example than map/map.
-% Let's start with the process definition for @map@.
-% The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
-% The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
-% Mention that the initial heap has the names but no values, but they could be initialised to whatever.
-% It doesn't matter since they'll be written before anything is read.
-% 
-% \begin{code}
-% map f = process (map f)
-%      ins: as
-%     outs: bs
-%     heap: {a = 0}
-%    label: map0
-%   blocks: map0 = pull as    a  map1
-%           map1 = push bs (f a) map2
-%           map2 = drop as       map0
-% \end{code}
-% 
-% As well as a ``combinator network'', a function comprised exclusively of process combinators.
-% The input streams are supplied as arguments, and output streams as return values.
-% \begin{code}
-% mapMap f g xs
-%  = let ys = map f xs
-%        zs = map g ys
-%    in  zs
-% \end{code}
-% 
-% It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
-% This is simple enough for a paragraph prose description.
-% It's just a bit of inlining and renaming everything to be unique.
-% 
-% \begin{code}
-% process (map f)
-%      ins: xs
-%     outs: ys
-%     heap: {x}
-%    label: p0
-%   blocks: p0 = pull xs    x  p1
-%           p1 = push ys (f x) p2
-%           p2 = drop xs       p0
-% process (map g)
-%      ins: ys
-%     outs: zs
-%     heap: {y}
-%    label: q0
-%   blocks: q0 = pull ys    y  q1
-%           q1 = push zs (g y) q2
-%           q2 = drop ys       q0
-% \end{code}
-% 
-% Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
-% Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.
-% 
-% \begin{code}
-% process (map f / map g)
-%      ins: xs
-%     outs: ys zs
-%     heap: {x, y, _ys}
-%    label: p0q0
-%   blocks: p0q0            = pull xs    x  p1q0
-%           p1q0            = push ys (f x) p2q0-pending-ys { _ys = f x }
-%           p2q0-pending-ys = drop xs       p0q0-pending-ys
-%           p0q0-pending-ys = jump          p0q1-have-ys    { y = _ys }
-%           p0q1-have-ys    = push zs (g y) p0q2-have-ys
-%           p0q2-have-ys    = jump          p0q0
-% \end{code}
-
-\subsection{Evaluation}
-\label{s:Process:Eval}
-
-\input{figures/ProcessEval.tex}
-\input{figures/ProcessFeed.tex}
-
-Evaluation of processes and process networks is split into three parts:
-\begin{itemize}
-\item Injection, where a single value for a stream is inserted into a process' input buffer.
-Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
-Injection for a process network proceeds by injecting the same value into all processes in the network.
-
-\item Shaking, where a process takes a step from one label to another.
-Shaking a process results in an updated process and an optional pushed value.
-When a process pushes to a channel, the value is injected to other processes.
-Shaking a process network chooses any process that can be shaked, and updates it in the process network.
-
-\item Feeding, where an environment of input values are fed to the process network, and output values are collected.
-Feeding alternates between injecting and shaking: injecting values from the environment when processes are ready to receive values, and repeatedly shaking the processes to collect their outputs.
-\end{itemize}
-
-Evaluation of a process network is non-deterministic, in that at any point there are many possible processes that can take a step.
-However, because each process itself is deterministic and has blocking reads, overall evaluation is deterministic as per Kahn process networks.
-That is: the order in which values are pushed to different output streams is not deterministic, but the order and values for a particular output stream \emph{are} deterministic.
-
-% Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
-
-\subsubsection{Injection}
-Fig.~\ref{fig:Process:Eval:Inject} defines the judgment forms and rules for injection.
-Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.
-
-A process has a single element buffer for each input, stored in its input state.
-This can be either @none@, in which case the buffer is empty and a value may be injected; @pending@, where a single value has been injected to the buffer but has not been pulled yet; or @have@, where a value has already been injected and is currently being used.
-
-The judgment form ($\ProcInject{p}{c}{v}{p'}$) means that the process $p$ has the value $v$ inserted to the input buffer for channel $c$, resulting in the updated process $p'$.
-
-Rule (InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
-An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded, or at least multiple element, buffer.
-We use the syntax ($\HeapUpdateOne{@ins@~p}{c}{@pending@~v}$) to perform a map update, in this case removing the ($c=@none@$) from ($@ins@~p$) and then inserting ($c=@pending@~v$).
-This sets the input buffer as pending, meaning that any future attempts to inject to this channel will fail, until the process pulls and then drops the value.
-
-Rule (InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.
-
-Rule (ProcessesInject) performs injection over a process network.
-Every process in the network must have the value injected into it.
-This means if multiple processes read from that stream, all input buffers for that stream must be empty.
-
-\subsubsection{Shaking}
-Fig.~\ref{fig:Process:Eval:Shake} defines the judgment forms and rules for shaking a process, allowing it to take a step, and produce an output message.
-% I don't like them, but alpha and tau are standard names in process calculi. 
-The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.
-
-The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
-executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
-The result contains an output message $\alpha$, as well as the new label $l'$, the new input state $i'$, and the update substitution to apply to the heap $u'$.
-
-Rule (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
-The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.
-
-Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.
-
-Rule (Push) evaluates the push expression $e$ under the heap, and sends that value as output the message.
-
-Rule (Jump) simply returns the new label and substitution.
-
-Rules (CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.
-
-The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
-The process shaking shakes a single instruction and updates the process.
-Shaking a process network chooses a single process to shake and updates it in the network.
-If shaking the process resulted in an output value, that value is injected into all the other processes in the network.
-
-Rule (Shake) takes a single process and shakes its instruction.
-The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
-The process steps to the new label with the new input state, and the values are updated in the heap.
-
-Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces an internal message ($\tau$).
-
-Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces a push message.
-The emitted push message is then injected into all other processes in the network, meaning they must either ignore the channel or be ready to receive it in their buffer.
-If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
-The entire network then emits the same message.
-
-\subsubsection{Feeding}
-Fig.~\ref{fig:Process:Eval:Feed} defines the judgment forms and rules for feeding, where external input values are fed into a process network and output values are accumulated.
-The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
-The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
-The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
-Feeding evaluates the process network until all input values have been injected.
-
-% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
-% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
-
-Rule (FeedStart) applies when all input values have been injected and there are no input values left.
-In this case, the output values are the same as the input values.
-
-Rule (FeedInternal) allows the process network to take an internal step.
-It first feeds its input accumulator and process network, then allows the resulting network to take an internal step.
-
-Rule (FeedPush) allows the process network to emit a push message.
-As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message.
-The pushed value is collected in the accumulator list for that stream.
-
-Rule (FeedExternal) allows inputs to be injected into the process network.
-For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
-The recursive feed is evaluated with the last value removed from the accumulators.
-The last value is then injected into the network, and added back to the result accumulators.
-
-% -- cuts ---------------------------------------------------------------------
-% BL: I don't think describing iota works at this point. This combinator is not used in the motivating example, so skipping to it seems disjointed.
-
-% Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
-% It takes no inputs, and produces one output stream @xs@.
-% Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
-% This variable will be incremented and pushed.
-% Each process also has a current label, which denotes the instruction to perform next.
-% The initial label for @iota@ is @L0@.
-% Each process has a mapping from labels to instructions.
-% In this case we have two instructions, @L0@ and @L1@.
-% The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move %to label @L1@.
-Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
-
-% \begin{code}
-% process (iota)
-%      ins: 
-%     outs: xs
-%     heap: {i = 0}
-%    label: L0
-%   blocks: L0 = push xs i  L1
-%           L1 = jump       L0{i = i + 1}
-% \end{code}
-
-% When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
-
-% The @uniques = group file1@ is a more interesting example.

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator.hs
Add diamond graph fusion tests

@@ -8,6 +8,7 @@ module Machine.Combinator
         , mkFilter
         , mkScan
         , mkGroup
+        , mkZip
 
         -- Multi-input, single output.
         , mkMerge)
@@ -17,5 +18,6 @@ import Machine.Combinator.Filter
 import Machine.Combinator.Scan
 import Machine.Combinator.Group
 import Machine.Combinator.Merge
+import Machine.Combinator.Zip
 import Machine.Base
 import Machine.New

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator/Zip.hs

@@ -1,0 +1,40 @@
+
+module Machine.Combinator.Zip where
+import Machine.New
+import Machine.Base
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-- | Construct a new `zip` process.
+mkZip   :: Channel              -- ^ Input  Channel A.
+        -> Channel              -- ^ Input  Channel B.
+        -> Channel              -- ^ Output Channel.
+        -> New Process
+
+mkZip cInA cInB cOut
+ = do
+        [l0, l1, l2, l3, l4]    
+                <- replicateM 5 (newLabel "Z")
+        vA      <- newVar "A"
+        vB      <- newVar "B"
+
+        return  
+         $ Process
+         { processName          = "map"
+         , processIns           = Map.fromList [(cInA, None), (cInB, None)]
+         , processOuts          = Set.fromList [cOut]
+         , processHeap          
+                = Heap $ Map.fromList 
+                        [ (vA, defaultValueOfChannel cInA)
+                        , (vB, defaultValueOfChannel cInB)] 
+
+         , processLabel         = l0
+
+         , processBlocks        =
+                [ (l0, Pull cInA vA (next l1))
+                , (l1, Pull cInB vB (next l2))
+                , (l2, Push cOut (XTuple 2 @@ XVar vA @@ XVar vB) (next l3))
+                , (l3, Drop cInA    (next l3))
+                , (l4, Drop cInB    (next l0)) ]
+         }

### 2017-02-25 2017mergingmerges/code/src/Machine/Pretty.hs

@@ -81,6 +81,7 @@ instance Pretty Prim where
         PAnd            -> string "and"
         PInt i          -> parens $ string "int" <+> int i
         PAdd            -> string "add"
+        PTuple n        -> string "tuple" <+> int n
         PEq             -> string "eq"
         PNeq            -> string "neq"
         PLt             -> string "lt"

### 2017-02-25 2017mergingmerges/code/src/Test/Fuse/Merge.hs

@@ -36,10 +36,8 @@ testFuseSplitMergeMerge
                 let Right pResult   = fusePair pMerge1 pMerge2
                 pFused_raw        <- fmap fst $ stripLabels "F" pResult
 
-
                 let modes'         = processChannelModes   pMerge1 pMerge2
                 let bConn'         = processesAreConnected pMerge1 pMerge2
-
 
                 return (modes', bConn', pFused_raw)
 

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator.hs
Add DiamondAlt example show different fusion orders work, or not

@@ -9,6 +9,7 @@ module Machine.Combinator
         , mkScan
         , mkGroup
         , mkZip
+        , mkAlt2
 
         -- Multi-input, single output.
         , mkMerge)
@@ -19,5 +20,6 @@ import Machine.Combinator.Scan
 import Machine.Combinator.Group
 import Machine.Combinator.Merge
 import Machine.Combinator.Zip
+import Machine.Combinator.Alt
 import Machine.Base
 import Machine.New

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator/Alt.hs

@@ -1,0 +1,56 @@
+
+module Machine.Combinator.Alt where
+import Machine.New
+import Machine.Base
+import qualified Data.Map       as Map
+import qualified Data.Set       as Set
+
+
+-- | Construct a new `alt2` process.
+mkAlt2  :: Channel              -- ^ Input  Channel A.
+        -> Channel              -- ^ Input  Channel B.
+        -> Channel              -- ^ Output Channel.
+        -> New Process
+
+mkAlt2 cInA cInB cOut
+ = do
+        [l0, l1, l2, l3, l4, l5, l6, l7, l8, l9, l10, l11]   
+                <- replicateM 12 (newLabel "L")
+        vA1     <- newVar "A1"
+        vA2     <- newVar "A2"
+        vB1     <- newVar "B1"
+        vB2     <- newVar "B2"
+
+        return  
+         $ Process
+         { processName          = "alt"
+         , processIns           = Map.fromList [(cInA, None), (cInB, None)]
+         , processOuts          = Set.fromList [cOut]
+         , processHeap          
+                = Heap $ Map.fromList 
+                        [ (vA1, defaultValueOfChannel cInA)
+                        , (vA2, defaultValueOfChannel cInA)
+                        , (vB1, defaultValueOfChannel cInB)
+                        , (vB2, defaultValueOfChannel cInB)] 
+
+         , processLabel         = l0
+
+         , processBlocks        =
+                [ (l0,  Pull cInA vA1 (next l1))
+                , (l1,  Pull cInB vB1 (next l2))
+
+                , (l2,  Drop cInA     (next l3))
+                , (l3,  Pull cInA vA2 (next l4))
+
+                , (l4,  Drop cInB     (next l5))
+                , (l5,  Pull cInB vB2 (next l6))
+
+                , (l6,  Push cOut     (XVar vA1) (next l7))
+                , (l7,  Push cOut     (XVar vA2) (next l8))
+                , (l8,  Push cOut     (XVar vB1) (next l9))
+                , (l9,  Push cOut     (XVar vB2) (next l10))
+
+                , (l10, Drop cInA (next l11))
+                , (l11, Drop cInB (next l0)) ]
+
+         }

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator/Zip.hs

@@ -21,7 +21,7 @@ mkZip cInA cInB cOut
 
         return  
          $ Process
-         { processName          = "map"
+         { processName          = "zip"
          , processIns           = Map.fromList [(cInA, None), (cInB, None)]
          , processOuts          = Set.fromList [cOut]
          , processHeap          
@@ -35,6 +35,6 @@ mkZip cInA cInB cOut
                 [ (l0, Pull cInA vA (next l1))
                 , (l1, Pull cInB vB (next l2))
                 , (l2, Push cOut (XTuple 2 @@ XVar vA @@ XVar vB) (next l3))
-                , (l3, Drop cInA    (next l3))
+                , (l3, Drop cInA    (next l4))
                 , (l4, Drop cInB    (next l0)) ]
          }

### 2017-02-25 2017mergingmerges/code/src/Machine/Transform/Fuse.hs

@@ -79,7 +79,11 @@ processesAreConnected process1 process2
 
 ---------------------------------------------------------------------------------------------------
 -- | Fuse two processes.
-fusePair :: Process -> Process -> Either String Process
+fusePair 
+        :: Process 
+        -> Process 
+        -> Either Label Process
+
 fusePair process1 process2
  = let  
         -- Compute how shared channels are used.
@@ -111,9 +115,7 @@ fusePair process1 process2
                     in  foldM go (instrs ++ [(ll, instr')]) lsOut
 
                 Nothing     
-                 -> Left ("cannot step " ++ show ll)
-
-                -- instrs
+                 -> Left ll 
 
          | otherwise
          = error "missing instruction"

### 2017-02-25 2017mergingmerges/section/XX-Related.tex
related work

--- 2017-02-06 2017mergingmerges/section/XX-Related.tex
@@
 
-Existing systems for combinator fusion only support a subset of combinators, and only in limited cases.
-Pull fusion supports joins where one combinator pulls from multiple inputs, but not splits where one output is used my multiple combinators.
-Pull can thus express @merge@, but not @partition@.
-Push fusion supports splits (@partition@), but not joins (@merge@).
-Polarised fusion~\cite{lippmeier2016polarized} addresses these shortcomings by separating the computation into pull producers and push consumers, with explicit drain loops to convert between the two sections.
-However this requires manual plumbing.
+This work aims to address the problems with existing combinator-based array fusion systems.
+As stated in the introduction, neither pull-based or push-based fusion alone is sufficient.
+Some combinators are inherently push-based, particularly those with multiple outputs such as @unzip@; while others are inherently pull-based, such as @zip@.
 
-The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}, by explicitly separating the types of pull and push arrays.
-Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations like append on pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
-% Here push arrays are `defunctionalized' and represented as a datatype containing the combinator and its arguments, which allows specialised and efficient code generation.
-% However this work only supports a limited number of baked-in push combinators such as append, mapping, interleaving and reversing.
+Short cut fusion is an attractive idea, as it allows fusion systems to be specified with a single rewrite rule.
+However, short cut fusion relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
+Thus, short cut fusion is inherently biased towards pull fusion.
+Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support only the intersection of pull and push: that is, neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
+
+Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
+For effectful input streams, for example reading from the network, this duplicated work can change the overall meaning of the program.
+% I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general
+
+Data flow fusion~\cite{lippmeier2013data} is neither pull-based nor push-based, and supports arbitrary splits and joins.
+It supports standard combinators such as @map@, @filter@ and @fold@, and converts each stream to a series with explicit rate types, similar to the clock types of Lucid Synchrone \cite{benveniste2003synchronous}.
+These rate types ensure that well-typed programs can be fused without introducing unbounded buffers.
+This allows unfusable programs to be caught at compile time.
+The main drawback of data flow fusion is that it only supports a limited set of combinators, and adding more is non-trivial.
+
+One way to address the difference between pull and push streams is to explicitly support both separately, as seen in \citet{bernardy2015duality} and \citet{lippmeier2016polarized}.
+Here, pull streams have the type @Source@ and are represented as a source that is always available to be pulled from, while push streams have the type @Sink@ and are represented as a sink that can always accept input.
+Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers and ordering of effects.
+Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
+Sinks, however, are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
+It has the type @Sink a -> Sink b -> Sink (a,b)@.
+This system requires the streaming computation to be manually split into sources and sinks, and be joined together by a loop that `drains' values from the source and pushes them into the sink.
+
+The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}.
+Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
 
@@
 Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers.
-However, these libraries are not designed for absolute performance and programs tend to be written to operate over streams of chunks rather than streams of values.
-For the most part they support only simple straight-line computations.
+However, these libraries provide no fusion guarantees, and as such programs tend to be written over chunks of data to make up for the communication overhead.
+For the most part they support only straight-line computations, with only limited forms of branching.
 
-Another way to characterise fusion systems is whether they are \emph{local} (shortcut) or \emph{global} fusion.
-Local fusion systems such as stream fusion~\cite{coutts2007stream} use local rewrite rules to remove intermediate arrays, and tend to rely on general purpose compiler optimisations to expose opportunities for these rewrites to occur.
-Generally, the local rewrite rule can only be applied once the producer has been inlined into the consumer.
-Because of the local nature of these rewrites and the reliance on inlining, local fusion can only perform producer-consumer fusion with a single consumer, as inlining the producer into multiple consumers would duplicate work.
-Local fusion is fragile because heuristics are used to determine whether inlining occurs~\cite{lippmeier2013data}.
-This fragility poses serious problems to programmers who require fusion for adequate performance, as one must have deep knowledge of the internal compiler phases and library design in order to predict whether fusion will occur -- or worse yet, they must scour the intermediate code to count the number of loops produced.
-In contrast, global fusion systems such as those traditionally used in imperative compilers make use of specific optimisations which must be implemented separately, but are able to perform horizontal fusion and are less prone to the fragility of local fusion.
+In relation to process calculi, synchronised product has been suggested as a method for fusing Kahn process networks together~\cite{fradet2004network}, but does not appear to have been implemented or evaluated.
+The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
+This is a much simpler fusion method than ours, but is also much stricter.
+When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
+Our system can be seen as an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.
 
-This local/global distinction is not a hard classification, as some systems fit somewhere in between.
-For example, {\bf strymonas}~\cite{kiselyov2016stream} uses staged compilation for introspection of the code and to ensure appropriate inlining, but only performs producer-consumer fusion with a single local rule.
-While not explicitly mentioned in the paper, Kiselyov et al~\cite{kiselyov2016stream}'s {\bf strymonas} system only allows streams to be used once.
-Furthermore, the entire stream computation must be terminated with a single fold which takes a single stream.
-This means multiple outputs with different rates cannot be treated, for example @partition@ which returns multiple arrays filtered with different predicates.
+Synchronous languages such as LUSTRE~\cite{halbwachs1991synchronous}, Lucy-n~\cite{mandel2010lucy} and SIGNAL~\cite{le2003polychrony} all use some form of clock calculus and causality analysis to ensure that programs can be statically scheduled with bounded buffers.
+These languages describe \emph{passive} processes where values are fed in to streams from outside environments, such as data coming from sensors.
+In this case, the passive process has no control over the rate of input coming in, and if they support multiple input streams, they must accept values from them in any order.
+In contrast, the processes we describe are \emph{active} processes that have control over the input that is coming in.
+This is necessary for combinators such as mergesort-style @merge@, as well as @append@.
+Note that in the synchronous language literature, it is common to refer to a different merge operation, also known as @default@, which computes a stream that is defined whenever either input is defined.
 
-This innocent-looking yet useless program, which should pair each line with itself, actually pairs each line with the successive line.
-\begin{code}
-let lines = linesFromFile filename
-in  zip lines lines
-\end{code}
+Synchronous dataflow (not to be confused with synchronous languages above) is a dataflow graph model of computation where each dataflow actor has constant, statically known input and output rates.
+The main advantage of synchronous dataflow is that it is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
+StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
+Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend synchronous dataflow with boolean and integer valued control ports, and attempt to recover the structure of ifs and loops from select and switch actors.
+These systems allow some dynamic structures to be scheduled statically, but are very rigid and only support limited control flow structures: it is unclear how merge or append could be scheduled by this system.
+Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is still quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
+A finite state machine is constructed, where each node of the FSM denotes its own synchronous dataflow graph.
+The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph.
+For example, a filter is represented with two nodes in the FSM.
+The dataflow graph for the initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
+The dataflow graph for the next state emits the value, and moves back to the first state.
+This does appear to be able to express value-dependent operations such as merge, but lacks the composability - and familiarity - of combinators.
 
-One would expect the above program to have the same behaviour as the following - or at least produce an error otherwise:
-\begin{code}
-zip (linesFromFile filename) (linesFromFile filename)
-\end{code}
-
-The one feature they support that we do not is concatMap/flatMap, which concatenates all produced substreams.
-This is a limitation in the \emph{conversion} to process calculus, not in the process calculus itself.
-Our process calculus can handle concrete instantiations of concatMap when specialised to a particular subprocess, but not the general case where the subprocess is statically unknown.
-By using a staging restriction as they do, we could modify the conversion to create a specialised version of concatMap for the given argument.
-
-\subsection{The problem with synchronised product}
-\label{s:Synchro}
-Related work on Network Fusion~\cite{fradet2004network} allows fusion of Kahn Process Networks using synchronised product to fuse pairs of processes together.
-Synchronised product is: both machines can take independent steps so long as it is not in the alphabet of the other, but have to agree on any shared actions.
-So if both machines pull from @xs@, the @pull xs@ message can only be executed when both machines agree.
-
-This sort of coordination is a bit too coarse and causes deadlocks.
-This example cannot be fused by synchronised product: one zip is trying to pull from @as@ and the other from @bs@, but neither can proceed without the other.
-Note that while this contrived example is unlikely to be written by hand, it is plausible that it would be part of a larger program after inlining has occurred.
-
-\begin{code}
-zips :: [a] -> [b] -> [a*b] * [b*a]
-zips as bs =
-  let abs = zip as bs
-      bas = zip bs as
-  in  abs, bas
-\end{code}
-
-When stream programs are encoded as labelled transition systems, some fusion can be done by computing the synchronised product of two transition systems, but this suffers deadlock when the two programs share multiple inputs and read them in different orders.
-By annotating transition states with the status of each input channel, our fusion algorithm works for these cases where synchronised product does not.
-
-% After inlining the definition of zip, we have:
-% \begin{code}
-% zips :: [a] -> [b] -> [a*b] * [b*a]
-% zips as bs =
-%   let abs = process abs.
-%        let loop1 =
-%           a <- pull as
-%           b <- pull bs
-%           push abs (a,b)
-%           drop as
-%           drop bs
-%           jump loop1
-%        in jump loop1
-%       bas = process bas.
-%        let loop2 =
-%           b <- pull bs
-%           a <- pull as
-%           push abs (b,a)
-%           drop bs
-%           drop as
-%           jump loop2
-%        in jump loop2
-%   in  abs, bas
-% \end{code}
+% StreamIt:
+% Only allows limited splits and joins: round robin and duplication for splits, round robin and combination for joins. 
+% Does not support fully general graphs - instead using combinators to introduce a (split/join) and a combinator for a feedback loop.
 % 
-% Here, in order to compute the synchronised product of these two, we need to know which are the common or shared actions.
-% Shared are: @pull as@, @pull bs@, @drop as@ and @drop bs@.
-% So both processes have to execute these together: if one wants to @pull as@, it can only act when the other one wants to @pull as@ as well.
-% Thus, the two machines are deadlocked, waiting for the other one to agree.
-% 
-% For this case, a simple solution suffices to fix this: adding extra processes for splits.
-So we would add a new combinator @dup@ which takes one input stream and creates two output streams.
-It reads a single element from the input and pushes it to both outputs.
-% 
-% \begin{code}
-% dup = stream_1_2 \is ols ors.
-%   letrec
-%     p1   = pull is p2
-%     p2 i = push ols i (p3 i)
-%     p3 i = push ors i p4
-%     p4   = drop is p1
-%   in p1
-% \end{code}
-% And the modified zip uses this like so:
-% 
-% \begin{code}
-% zips :: [a] -> [b] -> [a*b] * [b*a]
-% zips as bs =
-%   let as1, as2 = dup as
-%       bs1, bs2 = dup bs
-%       abs = zip as1 bs1
-%       bas = zip bs2 as2
-%   in  abs, bas
-% \end{code}
-% 
-% This has removed the deadlock, allowing a trace such as:
-% \begin{code}
-% as, as1, bs, bs1, bs2, as2
-% \end{code}
-% 
-% However it is not hard to modify the program so that the deadlock still exists.
-% By changing the each zip to pull from @as2@ and @bs2@ first, the streams are still used linearly, but deadlock occurs as @as1@ cannot be produced until after @bs2@ has been consumed, and vice versa.
-% \begin{code}
-% zips :: [a] -> [b] -> [a*b] * [b*a]
-% zips as bs =
-%   let as1, as2 = dup as
-%       bs1, bs2 = dup bs
-%       abs = zip as2 bs1
-%       bas = zip bs2 as1
-%   in  abs, bas
-% \end{code}
-% 
-% Using @dup@ we can concoct a simpler example with two processes and one input stream where synchronised product deadlocks.
-% \begin{code}
-% zipself :: [a] -> [a*a]
-% zipself as =
-%   let as1, as2 = dup as
-%       aas = zip as2 as1
-%   in  aas
-% \end{code}
-% Here @dup@ is attempting to push to @as1@, but cannot until @zip@ pulls, while @zip@ is attempting to pull from @as2@, which cannot proceed until the @dup@ pushes.
+% Parameterized dataflow (PDF),  \cite{bhattacharya2001parameterized}
+% Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator.hs
Add another version of alt2

@@ -9,7 +9,8 @@ module Machine.Combinator
         , mkScan
         , mkGroup
         , mkZip
-        , mkAlt2
+        , mkAlt2_int
+        , mkAlt2_blk
 
         -- Multi-input, single output.
         , mkMerge)

### 2017-02-25 2017mergingmerges/code/src/Machine/Combinator/Alt.hs

@@ -7,12 +7,12 @@ import qualified Data.Set       as Set
 
 
 -- | Construct a new `alt2` process.
-mkAlt2  :: Channel              -- ^ Input  Channel A.
+mkAlt2_int  :: Channel              -- ^ Input  Channel A.
         -> Channel              -- ^ Input  Channel B.
         -> Channel              -- ^ Output Channel.
         -> New Process
 
-mkAlt2 cInA cInB cOut
+mkAlt2_int cInA cInB cOut
  = do
         [l0, l1, l2, l3, l4, l5, l6, l7, l8, l9, l10, l11]   
                 <- replicateM 12 (newLabel "L")
@@ -42,6 +42,56 @@ mkAlt2 cInA cInB cOut
                 , (l2,  Drop cInA     (next l3))
                 , (l3,  Pull cInA vA2 (next l4))
 
+                , (l4,  Drop cInB     (next l5))
+                , (l5,  Pull cInB vB2 (next l6))
+
+                , (l6,  Push cOut     (XVar vA1) (next l7))
+                , (l7,  Push cOut     (XVar vA2) (next l8))
+                , (l8,  Push cOut     (XVar vB1) (next l9))
+                , (l9,  Push cOut     (XVar vB2) (next l10))
+
+                , (l10, Drop cInA (next l11))
+                , (l11, Drop cInB (next l0)) ]
+
+         }
+
+
+-- | Construct a new `alt2` process.
+mkAlt2_blk  
+        :: Channel              -- ^ Input  Channel A.
+        -> Channel              -- ^ Input  Channel B.
+        -> Channel              -- ^ Output Channel.
+        -> New Process
+
+mkAlt2_blk cInA cInB cOut
+ = do
+        [l0, l1, l2, l3, l4, l5, l6, l7, l8, l9, l10, l11]   
+                <- replicateM 12 (newLabel "L")
+        vA1     <- newVar "A1"
+        vA2     <- newVar "A2"
+        vB1     <- newVar "B1"
+        vB2     <- newVar "B2"
+
+        return  
+         $ Process
+         { processName          = "alt2_blk"
+         , processIns           = Map.fromList [(cInA, None), (cInB, None)]
+         , processOuts          = Set.fromList [cOut]
+         , processHeap          
+                = Heap $ Map.fromList 
+                        [ (vA1, defaultValueOfChannel cInA)
+                        , (vA2, defaultValueOfChannel cInA)
[…]

### 2017-02-25 2017mergingmerges/code/src/Test/Fuse/DiamondAlt.hs

@@ -24,15 +24,16 @@ testFuseDiamondAltOk
          = evalNew 
          $ do   
                 pZip3   <- mkZip  cA1  cA2  cOut
-                pAlt1   <- mkAlt2 cIn1 cIn2 cA1
+                pAlt1   <- mkAlt2_blk cIn1 cIn2 cA1
                 let pAlt13  = fusePair' [pZip3, pAlt1] pZip3 pAlt1
 
-                pAlt2   <- mkAlt2  cIn2 cIn3 cA2
+                pAlt2   <- mkAlt2_blk  cIn2 cIn3 cA2
                 let pZipped = fusePair' [pZip3, pAlt1, pAlt2] pAlt13 pAlt2
 
                 pFused'         <- fmap fst $ stripLabels "F" pZipped
                 return pFused'
    in   pFused
+
 
 testFuseDiamondAltFail
  = putStrLn $ show $ pretty
@@ -48,14 +49,12 @@ testFuseDiamondAltFail
         pFused
          = evalNew 
          $ do   
-                pAlt1             <- mkAlt2 cIn1 cIn2 cA1
-                pAlt2             <- mkAlt2 cIn2 cIn3 cA2
+                pAlt1             <- mkAlt2_blk cIn1 cIn2 cA1
+                pAlt2             <- mkAlt2_blk cIn2 cIn3 cA2
                 let pAlt12      = fusePair' [pAlt1, pAlt2] pAlt1 pAlt2
 
                 pZip              <- mkZip  cA1 cA2  cOut
                 let pZipped     = fusePair' [pAlt1, pAlt2, pZip] pAlt12 pZip
-
-
 
                 pFused'         <- fmap fst $ stripLabels "F" pZipped
                 return pFused'

### 2017-02-25 2017mergingmerges/section/01-Introduction.tex
Add another version of alt2
Don't claim we support fully arbitrary splits and joins

@@
 The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level.
-The polyhedral model is based around affine loops, which makes it difficult to support for filter-like operators such as @group@ and @merge@.
+The polyhedral model is based around affine loops, which makes it difficult to support filter-like operators such as @group@ and @merge@.
 
@@
 \item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
-\item an algorithm for fusing these processes, the first to support arbitrary splits and joins (\S\ref{s:Fusion});
+\item an algorithm for fusing these processes, the first to support splits and joins (\S\ref{s:Fusion});
 \item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. \TODO{Ref}

### 2017-02-25 2017mergingmerges/section/XX-Processes.tex
End S2 where it is, that is enough intuition before going onto the defs

@@
 
-The @group@ combinator has two parameters, @s1@ and @s2@, which bind the input and output streams respectively.
+The @group@ combinator has two parameters, @sIn1@ and @sOut1@, which bind the input and output streams respectively.
 The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
@@
 
-The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
-The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates yet we leave it empty.
+The initial instruction @(pull sIn1 v A1 {})@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
+The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to update anything yet we leave it empty.
 
@@
 
-When the predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
+When the predicate is true, the instruction @(push sOut1 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
 
-Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@.
+Finally, the instruction @(drop sIn1 A0 {})@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
 This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
-The next element of a stream may only be pulled after all consuming processes have pulled and then and dropped the current element.
+The next element of a stream may only be pulled after all consumers have pulled and then and dropped the current element.
 
@@
            group 
-             = \(\lambda\) (sG1: Stream Nat) (sG2: Stream Nat). 
+             = \(\lambda\) (sIn1: Stream Nat) (sOut1: Stream Nat). 
                \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
@@
                process
-               { ins:    { sG1 }
-               , outs:   { sG2 }
+               { ins:    { sIn1  }
+               , outs:   { sOut1 }
                , heap:   { f = T, l = 0, v = 0 }
                , label:  A0
-               , instrs: { A0 = pull sG1 v           A1 {}
+               , instrs: { A0 = pull sIn1 v          A1 {}
                          , A1 = case (f || (l /= v)) A2 {}  A3 {}
-                         , A2 = push sG2 v           A3 { l = v, f = F }
-                         , A3 = drop sG1             A0 {} } }
+                         , A2 = push sOut1 v         A3 { l = v, f = F }
+                         , A3 = drop sIn1            A0 {} } }
 \end{code}
@@
                merge
-                 = \(\lambda\) (sM1: Stream Nat) (sM2: Stream Nat) (sM3: Stream Nat). 
+                 = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
                    \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
@@
                    process
-                   { ins:    { s1, s2 }
-                   , outs:   { s3 }
+                   { ins:    { sM1, sM2 }
+                   , outs:   { sM3 }
                    , heap:   { x1 = 0, x2 = 0 }
                    , label:  B0
-                   , instrs: { B0 = pull s1 x1     B1 {}
-                             , B1 = pull s2 x2     C0 {}
+                   , instrs: { B0 = pull sIn1  x1   B1 {}
+                             , B1 = pull sIn2  x2   C0 {}
                              , C0 = case (x1 < x2)  D0 {}  E0 {}
-                             , D0 = push s3 x1     D1 {}
-                             , D1 = drop s1        D2 {}
-                             , D2 = pull s1 x1     C0 {}
-                             , E0 = push s3 x2     E1 {}
-                             , E1 = drop s2        E2 {}
-                             , E2 = pull s2 x2     C0 {} } }
+                             , D0 = push sOut2 x1   D1 {}
+                             , D1 = drop sIn1       D2 {}
+                             , D2 = pull sIn1  x1   C0 {}
+                             , E0 = push sOut2 x2   E1 {}
+                             , E1 = drop sIn2       E2 {}
+                             , E2 = pull sIn2 x2    C0 {} } }
 \end{code}
@@
 The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}.
-The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
+The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@.
 The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream.
@@
 
-As this process merges infinite streams, if we execute it with a finite input prefix, the result will be an intermediate state that may not yet have pushed all available output.
+As this process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
 For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @s2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
+After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
 
@@
 The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-For ease of comparison we will assume the parameters in each combinator are instantiated with arguments of the same names.
+For ease of comparison we will assume the parameters in each combinator are instantiated by arguments with the same names.
 
@@
 
-We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream, which is used to determine whether another value can be pulled.
-For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
+We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream.
+For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
 The result state @F0@ represents a combination of the two input states @A0@ and @B0@.
-As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.
+As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @sIn1@ and @sIn2@ map to `none'.
 
-From the joint state @F0@, both of the input machines need to pull from stream @s1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
+From the joint state @F0@, both of the input machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
 In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
@@
 The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
-In Fig.~\ref{fig:Fusion:Pulls},  `s1' is set to `have' in @F2@ after we have set `v = b1', while `s1' is set to `have' in @F3@ after we have set `x1 = b1'.
+In Fig.~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.
 
@@
 Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
-From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @s1@ to its output stream, which we name @s3@ --- or to just move on to the next value from its input.
-From state @B1@, the @merge@ machine needs to pull a value from its second input stream @s2@.
-
-In the fused result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
+From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to its output stream @sOut1@, or to just move on to the next value from its input.
+From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@.
+In the result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
 At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
 
-Finally, at @F5@ the @merge@ machine pulls from @s2@, moving from @B1@ to @C0@.
-Because the stream @s2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
-In this case, we simply perform a @pull@ instruction.
+Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
+Because the stream @sIn22@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
 
-In terms of fusion, we could construct the result machine in multiple ways.
-One option is to perform the case branch first and then pull from @s2@.
-We could alternately pull from @s2@ first and then perform the case branch.
+Note that we could construct the fused machine in multiple ways.
+One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch.
 By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
@@
 Fig.~\ref{fig:Process:Fused} shows the entire result of fusing @group@ and @merge@ together.
-The process has two inputs, @s1@ and @s2@, which correspond to the original input streams.
-Of the two outputs, @s3@ is the @group@'s output stream, while @s4@ is the @merge@'s output stream.
-The heap contains the variables from both processes, as well as the new buffer variable @b1@, which is used to store the current element pulled from @s1@.
+There are other rules for handling different combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
+The result process has two inputs, @sIn1@ and @sIn2@, which correspond to the original input streams.
+Of the two outputs, @sOut1@ is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.
 
-For the most part, the fused process simply alternates between performing the instructions of @group@ and those of @merge@.
-For the shared input stream @s1@, the two processes must be coordinated so that values are only pulled or dropped when both agree.
+Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequental result process that implements both.
+We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes.
+As with the Flow Fusion system of \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
+However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are.
+This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.
 
-The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
+To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
+The order in which processes are fused together does matter, as does the order in which the instructions are emitted --- we discuss both points further in \S\ref{s:Evaluation}.
 
-The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
+Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere.
+When we instantiate combinators to create the original processes we use fresh names at that point.
+The stream buffer variables we aditionally introduce for coordination are freshly created during fusion.
+
+% BL: This is too much low level detail at this point, we need more intuition.
+
+% The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
+
+% The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
 These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
@@
 
-Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
+% Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
 As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
@@
 process
-{ ins:    { s1, s2 }
-, outs:   { s3, s4 }
+{ ins:    { sIn1,  sIn2 }
+, outs:   { sOut1, sOut2 }
 , heap:   { f = T, l = 0, v = 0, x1 = 0, x2 = 0, b1 = 0}
@@
 \newcommand\annot[5]{
-  \tiny ((#1, \> \tiny \{s1 =      #2\}), \> \tiny (#3, \> \tiny \{s1 =      #4, \> \tiny s2 =      #5\}))
+  \tiny ((#1,   \> \tiny \{sIn1 =      #2\}), 
+                \> \tiny (#3, \> \tiny \{sIn1 =      #4, \> \tiny sIn2 =      #5\}))
 }
@@
 @      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
-\> @{@ \instr{F0}{pull s1 b1}{F1}{}{\annot{A0}{none}{B0}{none}{none}}
+\> @{@ \instr{F0}{pull sIn1 b1}{F1}{}
+      {\annot{A0}{none}{B0}{none}{none}}
 
@@
 \> @,@
-\instr{F4}{push s3 v}{F5}{l = v, f = F}
+\instr{F4}{push sOut1 v}{F5}{l = v, f = F}
       {\annot{A2}{have}{B1}{have}{none}}
@@
 \> @,@
-\instr{F6}{pull s2 x2}{F7}{}
+\instr{F6}{pull sIn2 x2}{F7}{}
       {\annot{A0}{none}{B1}{have}{none}}
@@
 \> @,@
-\instr{F8}{push s4 x1}{F9}{}
+\instr{F8}{push sOut2 x1}{F9}{}
       {\annot{A0}{none}{D0}{have}{have}}
 \> @,@
-\instr{F9}{drop s1}{F10}{}
+\instr{F9}{drop sIn1}{F10}{}
       {\annot{A0}{none}{D1}{none}{have}}
 \> @,@
-\instr{F10}{pull s1 b1}{F11}{}
+\instr{F10}{pull sIn1 b1}{F11}{}
       {\annot{A0}{none}{D2}{none}{have}}
@@
 \> @,@
-\instr{F13}{push s3 v}{F14}{l = v, f = F}
+\instr{F13}{push sOut1 v}{F14}{l = v, f = F}
       {\annot{A2}{have}{D2}{pending}{have}}
@@
 \> @,@
-\instr{F16}{push s4 x2}{F17}{}
+\instr{F16}{push sOut2 x2}{F17}{}
       {\annot{A0}{none}{E0}{have}{have}}
 \> @,@
-\instr{F17}{drop s2}{F18}{}
+\instr{F17}{drop sIn2}{F18}{}
       {\annot{A0}{none}{E1}{have}{have}}
 \> @,@
-\instr{F18}{pull s2}{F7}{}
+\instr{F18}{pull sIn2}{F7}{}
       {\annot{A0}{none}{E2}{have}{none}}
@@
 
-% With original label names and 'n', 'h' and 'p' for input state
-%\begin{code}
-%A0nB0n = pull s1 b1  A0pB0p{}
-%A0pB0p = jump        A1hB0p{v  = b1}
-%A1hB0p = jump        A1hB1h{x1 = b1}
-%
-%A1hB1h = case (f || (l /= v))
-%                     A2hB1h{}
-%                     A3hB1h{}
-%
-%A2hB1h = push s3 v   A3hB1h{l  = v, f  = F}
-%A3hB1h = jump        A0nB1h{}
-%A0nB1h = pull s2 x2  A0nC0h{}
-%
-%A0nC0h = case (x1 < x2)
-%                     A0nD0h{}
-%                     A0nE0h{}
-%
-%A0nD0h = push s4 x1  A0nD1h{}
-%A0nD1h = drop s1     A0nD2n{}
-%A0nD2n = pull s1 b1  A0pD2p{}
-%A0pD2p = jump        A1hD2p{v  = b1}
-%
-%A1hD2p = case (f || (l /= v))
-%                     A2hD2p{}
-%                     A3hD2p{}
-%
-%A2hD2p = push s3 v   A3hD2p{l  = v, f  = F}
-%A3hD2p = jump        A0nD2p{}
-%A0nD2p = jump        A0nC0h{x1 = b1}
-%
-%A0nE0h = push s4 x2  A0nE1h{}
-%A0nE1h = drop s2     A0nE2h{}
-%A0nE2h = pull s2 x2  A0nC0h{}
-%\end{code}
-
-% Cleaned up, simplified, variables collated
-% \begin{code}
-%             process
-%             { ins:    { s1, s2 }
-%             , outs:   { s3, s4 }
-%             , heap:   { f = T, l = 0, x1 = 0, x2 = 0}
-%             , label:  F0
-%             , instrs: { F0 = pull s1 x1              F3 {}
-%                       , F3 = case (f || (l /= x1))   F4 {} F5 {}
-%                       , F4 = push s3 x1              F5 { l = x1, f = F }
-%                       , F5 = pull s2 x2              F6 {}
-% 
-%                       , F6 = case (x1 < x2)          A0 {} B0 {}
-% 
-%                       , A0 = push s4 x1              A1 {}
-%                       , A1 = drop s1                 A2 {}
-%                       , A2 = pull s1 x1              A3 {}
-%                       , A3 = case (f || (l /= x1))   A4 {} F6 {}
-%                       , A4 = push s3 x1              F6 { l = x1, f = F }
-% 
-%                       , B0 = push s4 x2              B1 {}
-%                       , B1 = drop s2                 B2 {}
-%                       , B2 = pull s2 x2              F6 {} } }
-% \end{code}
 \caption{Fusion of group and merge}
@@
 \end{figure}
-
-\subsection{Optimisation}
-After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
-
-\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}

### 2017-02-25 2017mergingmerges/section/01-Introduction.tex
Wibbles
wibble to intro

@@
 
-We might instead define some sort of uniform interface for data sources, with a single `pull' function that provides the next value in each stream.
+We might instead define some sort of uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
 Each operator could be given this interface, so that the next value from each result stream is computed on demand.
@@
 Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs.
-We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow graph.
+We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network.
 Some systems support both pull and push: data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
@@
 \item an algorithm for fusing these processes, the first to support splits and joins (\S\ref{s:Fusion});
-\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. \TODO{Ref}
+\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. (\S\ref{s:Evaluation}).
 \item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs});

### 2017-02-25 2017mergingmerges/section/01-Introduction.tex
copy edits to intro

@@
 We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of identifiers from both streams.
-Can we perform both of these tasks at once, without needing to read through the stream data multiple times, and without needing unbounded buffering?
 Here is how we might write the source code, where @S@ is for @S@-tream.
@@
 
+% Can we perform both of these tasks at once, without needing to read through the stream data multiple times, and without needing unbounded buffering?
+
 In this implementation the @group@ operator filters out consecutive duplicates, while @merge@ combines two sorted streams so that the output remains sorted.
@@
 
-For an implementation which does \emph{not} use stream fusion, we might implement each of the operators as a separate concurrent process, and send each identifier value using an intra-process communication mechanism.
+We might try to implement each of the operators as a separate concurrent process, and send each identifier value using an intra-process communication mechanism.
 Developing such an implementation could be easy or hard, depending on what language features are available for concurrency.
-However, worrying about the \emph{performance tuning} of such a system, such as whether we need back-pressure, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
+However, worrying about the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
 
@@
 
-Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requiring unbounded buffering.
+Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requiring unbounded buffering.
 Sadly, existing stream fusion transformations cannot handle our example.
@@
 
-In our new system we still view the program as a concurrent process network.
-Each operator is a separate process, and the stream data flows through communication channels between the processes.
+Our new system takes a different approach.
 Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
-The fusion transform takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers.
-When the fusion transformation succeeds we know it has worked.
+We view each operator as a process in a concurrent process network.
+Our fusion transform then \emph{sequentializes} the network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
+When the fusion transform succeeds we know it has worked.
 There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.
@@
 
-Our fusion transformation for infinite stream programs could also serve as the basis for an \emph{array} fusion system, using a natural extension to finite streams.
+Our fusion transform for infinite stream programs could also serve as the basis for an \emph{array} fusion system, using a natural extension to finite streams.
 We discuss this extension in \S\ref{s:Finite}.

### 2017-02-25 2017mergingmerges/section/XX-Processes.tex
squeeze fused figure into columns

@@
 %!TEX root = ../Main.tex
-\section{Processes, Combinators and Operators}
+\section{Processes, Machines, Combinators and Operators}
 \label{s:Processes}
@@
 The process language is an intermediate representation we use when fusing the overall dataflow network.
-A \emph{combinator} is a template for a particular \emph{process} which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
-Each process implements a logical \emph{operator}, so we use ``operator'' when describing the values being computed, but ``process'' when referring to the implementation.
+When describing the fusion transform we describe the control flow of the process as a state machine, hence Machine Fusion.
+
+A \emph{combinator} is a template for a particular process which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
+Each process implements a logical \emph{operator} --- so we use ``operator'' when describing the values being computed, but ``process'' and ``machine'' when referring to the implementation.
 
@@
 The initial instruction @(pull sIn1 v A1 {})@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
-The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to update anything yet we leave it empty.
+The empty set @{}@ after the target label @A1@ can be used to update heap variables, but as we do not need to update anything yet we leave it empty.
 
-Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to the instruction at label @A3@.
-We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
+Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@.
+We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) for whether this is the first element.
 
@@
 The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream.
-The process starts by pulling from each input streams.
+The process starts by pulling from each of the input streams.
 It then compares the two pulled values, and pushes the smaller of the values to the output stream.
@@
 
-As this process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
+As this the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
 For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-After pushing the last $4$, the process will block at instruction @E2@, waiting for the next value to be pulled from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
+After pushing the last value $4$, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@.
+We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
 
@@
 For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
-The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
+The result will be a machine that computes the result of both the @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
 For ease of comparison we will assume the parameters in each combinator are instantiated by arguments with the same names.
@@
 The result state @F0@ represents a combination of the two input states @A0@ and @B0@.
-As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @sIn1@ and @sIn2@ map to `none'.
+As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.
 
 From the joint state @F0@, both of the input machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
-In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
-For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.
+In the result machine this is managed by first storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
+For this we attach updates to a @jump@ instruction, which otherwise transitions between states without affecting any of the input or output streams.
 
-Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
+Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'.
 The `none' state means that we have not yet pulled a value from the associated stream.
@@
 Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
-Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
+The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
 From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to its output stream @sOut1@, or to just move on to the next value from its input.
 From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@.
-In the result, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
+In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
 At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
@@
 Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
-Because the stream @sIn22@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
+Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
 
-Note that we could construct the fused machine in multiple ways.
+Note that we could construct the fused result machine in several ways.
 One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch.
-By construction, the predicate used in the branch only refers to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
+By construction, the predicate used in the branch refers only to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine.
 As the set of variables does not overlap, either ordering is correct.
@@
 
-Fig.~\ref{fig:Process:Fused} shows the entire result of fusing @group@ and @merge@ together.
+Fig.~\ref{fig:Process:Fused} shows the end result of fusing @group@ and @merge@ together.
 There are other rules for handling different combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
 The result process has two inputs, @sIn1@ and @sIn2@, which correspond to the original input streams.
-Of the two outputs, @sOut1@ is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.
+It also has two outputs, @sOut1@ which is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.
 
@@
 We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes.
-As with the Flow Fusion system of \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
+As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
 However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are.
@@
 Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere.
-When we instantiate combinators to create the original processes we use fresh names at that point.
-The stream buffer variables we aditionally introduce for coordination are freshly created during fusion.
+When we instantiated combinators to create the original processes we introduced fresh names at that point.
+The stream buffer variables we aditionally introduced for coordination were freshly created during fusion.
 
@@
 @, F17@  \= = @case (f || (l /= v)) @
-         \= @F17{x1 = b1, f = F}@ \= @F17{ }@
-@      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
+         \= @F17{}         @ \= @F17{ }@
+@   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
 \> @{@ \instr{F0}{pull sIn1 b1}{F1}{}

### 2017-02-26 2017mergingmerges/section/01-Introduction.tex
Fix description of motivating example

@@
 Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once.
-To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that values are only read when both the @group@ and @merge@ operators are ready to receive a new value.
+To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it.
 Finally, as the stream length is assumed to be unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.
@@
 Each operator could be given this interface, so that the next value from each result stream is computed on demand.
-This is approach is commonly taken with implementations of physical operators in data base systems.
-However, this `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
-Suppose a consumer pulls many elements from the result @sUnique@ stream.
-The implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@.
-It needs to buffer an aribrary number of elements from @sIn2@ because there is no guarantee of when a consumer will also pull from the @sUnion@ result stream.
-Once that happens the elements from @sIn2@ no longer need to be retained, but not before.
+This is approach is commonly taken with implementations of physical operators in database systems.
+However, the `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
+Suppose a consumer pulls many elements from the @sUnique@ output stream.
+In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering a arbitrary number of them.
+It needs to buffer these elements because they are also needed to perform the @merge@ operation.
+When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.
 
@@
 As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations.
-Pull-based systems such as short cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers.
-We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.
+Pull-based systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers.
+We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.
 
@@
 
-Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs.
+Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example either, because they do not support operators with multiple inputs.
 We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network.
-Some systems support both pull and push: data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
+Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
 More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.
@@
 
-Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
+Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
 The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level.
@@
 
-Our new system takes a different approach.
+In this paper we present Machine Fusion, a new approach.
 Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
-We view each operator as a process in a concurrent process network.
-Our fusion transform then \emph{sequentializes} the network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
+We view each operator as a process in a concurrent process network, and represent the control flow of the process as a simple state machine.
+Our fusion transform then \emph{sequentializes} the concurrent data flow network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
 When the fusion transform succeeds we know it has worked.

### 2017-02-26 2017mergingmerges/section/XX-Processes.tex
Sort out terminology, process, machine, combinator, operator

--- 2017-02-25 2017mergingmerges/section/XX-Processes.tex
@@
 
-Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether the predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to the instruction at label @A3@.
-We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
+Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@.
+We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) for whether this is the first element.
 

### 2017-02-26 2017mergingmerges/section/XX-Processes.tex
Fold in written notes

--- 2017-02-25 2017mergingmerges/section/XX-Processes.tex
@@
 We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes.
-As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
+As with the Flow Fusion system of \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
 However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are.
@@
 @, F17@  \= = @case (f || (l /= v)) @
-         \= @F17{}         @ \= @F17{ }@
-@   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
+         \= @F17{x1 = b1, f = F}@ \= @F17{ }@
+@      @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{s1 = pending\}), \= \tiny (B0, \= \tiny \{s1 = pending, \= \tiny s2 = pending\})) \kill
 \> @{@ \instr{F0}{pull sIn1 b1}{F1}{}

### 2017-02-26 2017mergingmerges/section/XX-FutureWork.tex
future work

@@
 %!TEX root = ../Main.tex
-\section{Conclusions and future work}
+\section{Future work}
 \label{s:FutureWork}
@@
 Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output.
-We have focussed on the infinite stream version because it is somewhat simpler to explain and prove, but the extensions required to support finite streams do not require substantial conceptual changes.
+We have focussed on the infinite stream version because it is simpler to explain and prove, but the extensions required to support finite streams do not require substantial conceptual changes.
 
-We now describe the extensions required to support finite streams.
-We add a new @closed@ constructor to the \InputState~ to encode the end of the stream.
-Once an input stream is in the closed state, it can never change to another state: it remains closed thereafter.
+Unlike infinite streams, pulling from a finite stream can fail, meaning the stream is finished.
+We therefore modify the @pull@ instruction to have two output labels: one to execute when a value is pulled, and the other to execute when the stream is finished.
+On the pushing end, we also need some way of finishing streams, so we add a new instruction to close an output stream.
 
-We modify the @pull@ instruction so that it has two output labels (like @case@).
-The first label, the read branch, is executed as before when the pull succeeds and a value is read from the stream.
-The second label, the close branch, is executed when the stream is closed, and no more values will ever be available.
-After a pull takes the close branch, any subsequent pulls from that stream will also take the close branch.
+During evaluation we need some way of knowing whether a stream is closed, which can be added as an extra constructor in the \InputState~ type.
+The same constructor is added to the static input state used by fusion.
+In this way, for any changes made to evaluation, the analogous static change must be made in the fusion transform.
 
-We add two new instructions for closing output streams and disconnecting from input streams.
-Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
-As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
-After a stream has been closed, no further values can be pushed.
+For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel.
+After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow.
 
-Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
-This can be used when a process requires the first values of a stream, but does not require the whole stream.
-If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
-Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
-The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
-After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
-After a process disconnects from an input channel, it can no longer pull from that channel.
+We have implemented an early prototype that supports finite streams, but further work is required to prove it correct.
 
-We also add an instruction for terminating the process (@done@).
-After all input streams have been read to completion or disconnected and output streams closed, the process may execute @done@ to signal that processing is complete.
+It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements.
+However, this requires the fusion transform to analyse and reason about case statements and their predicates.
+It seems obvious that if two consumers read the same value and check if it is EOF, they are both going to give the same result, but this is of course undecidable in general.
+By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for semi-decidable equality of predicates.
+Case analysis is still a useful extension and is described in the next section.
 
-The fusion definition must be extended to deal with these new instructions.
-The static input state has a @closed@ constructor added and disconnection is encoded by removal from the input state, and the \ti{tryStep} changes more or less follow the evaluation changes.
-Shared and connected pulls now deal with two more possibilities in the input state: the input may be closed in which case the close branch of the pull is taken; or the other process may have disconnected in which case the pull is executed as in the non-shared non-connected case.
-Connected pushes must also deal with when the other process has disconnected in which case the push is executed as if it were non-connected.
-For @in1@ and @out1@ channels, the new @close@ and @disconnect@ instructions are used as normal with no coordination required.
-For connected @close@, as with @push@, the receiving process must have @none@ and the next step performs the @close@ and sets the input state to @closed@.
-For shared @disconnect@, the @disconnect@ is only performed after both processes have disconnected; otherwise the entry is just removed from the input state.
-For connected @disconnect@, the @disconnect@ is not performed and the entry is removed from the input state.
+% We now describe the extensions required to support finite streams.
+% We add a new @closed@ constructor to the \InputState~ to encode the end of the stream.
+% Once an input stream is in the closed state, it can never change to another state: it remains closed thereafter.
+% 
+% We modify the @pull@ instruction so that it has two output labels (like @case@).
+% The first label, the read branch, is executed as before when the pull succeeds and a value is read from the stream.
+% The second label, the close branch, is executed when the stream is closed, and no more values will ever be available.
+% After a pull takes the close branch, any subsequent pulls from that stream will also take the close branch.
+% 
+% We add two new instructions for closing output streams and disconnecting from input streams.
+% Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
+% As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
+% After a stream has been closed, no further values can be pushed.
+% 
+% Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
+% This can be used when a process requires the first values of a stream, but does not require the whole stream.
+% If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
+% Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
+% The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
+% After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
+% After a process disconnects from an input channel, it can no longer pull from that channel.
+% 
+% We also add an instruction for terminating the process (@done@).
+% After all input streams have been read to completion or disconnected and output streams closed, the process may execute @done@ to signal that processing is complete.
+% 
+% The fusion definition must be extended to deal with these new instructions.
+% The static input state has a @closed@ constructor added and disconnection is encoded by removal from the input state, and the \ti{tryStep} changes more or less follow the evaluation changes.
+% Shared and connected pulls now deal with two more possibilities in the input state: the input may be closed in which case the close branch of the pull is taken; or the other process may have disconnected in which case the pull is executed as in the non-shared non-connected case.
+% Connected pushes must also deal with when the other process has disconnected in which case the push is executed as if it were non-connected.
+% For @in1@ and @out1@ channels, the new @close@ and @disconnect@ instructions are used as normal with no coordination required.
+% For connected @close@, as with @push@, the receiving process must have @none@ and the next step performs the @close@ and sets the input state to @closed@.
+% For shared @disconnect@, the @disconnect@ is only performed after both processes have disconnected; otherwise the entry is just removed from the input state.
+% For connected @disconnect@, the @disconnect@ is not performed and the entry is removed from the input state.
+% 
+% Finally, \ti{tryStepPair} is modified so that @done@ is performed when both machines are @done@.
+% 
+% These modifications allow our system to fuse finite streams as well as infinite.
 
-Finally, \ti{tryStepPair} is modified so that @done@ is performed when both machines are @done@.
-
-These modifications allow our system to fuse finite streams as well as infinite.
-We have implemented an initial prototype that supports finite streams, but future work is required to prove them correct.
-
-\subsection{Fully abstract case interpretation}
+\subsection{Case analysis}
 \label{s:FullyAbstractCase}
@@
 The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
-This can cause an issue with coordination and buffering, as the processes may dynamically require only a bounded buffer, the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
+This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
 
-For example, suppose two processes have the same case condition @x > 0@.
-The fusion algorithm will try all possibilities including contradictory ones, such as where the first process has @x > 0 = true@ and the second has @x > 0 = false@.
-If any of these possibilities require an unbounded buffer, the fusion algorithm fails.
-This means that the following program which pairs all positive elements with themselves, while unlikely to be written by a human, is not able to be fused.
-\begin{code}
-zipgts as =
-  let as1 = filter (>0) as
-      as2 = filter (>0) as
-      aas = zip as1 as2
-  in  aas
-\end{code}
-
-In general, if the example above used two different predicates for each filter, the fusion system would be right to outlaw it as requiring unbounded buffers.
-If the same program is rewritten to use only a single filter and use its result twice, the program is able to be fused.
+For example, suppose two processes have the same case condition (@x > 0@).
+The fusion algorithm will generate all four possibilities, including the contradictory ones such as where the first process has (@x > 0 = true@) and the second has (@x > 0 = false@).
+If any of these possibilities require an unbounded buffer, the fusion algorithm will fail.
 

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Work on process definitions

--- 2017-02-24 2017mergingmerges/section/XX-Definition.tex
@@
 
-%% This figure is referenced below, in 'Process definitions', but putting it up here before the big fused process stops the fused one from splitting across multiple pages.
 \input{figures/ProcessDef.tex}
 
-The grammar for processes is shown in Fig.~\ref{fig:Process:Def}.
-Channels, labels and variables are specified by some external, globally unique set of names.
+The grammar for process definitions is given in Fig.~\ref{fig:Process:Def}.
+Variables, Channels and Labels are specified by unique names.
+We refer to the \emph{endpoint} of a stream as a channel.
+A particular stream may flow into the input channels of several different processes, but can only be produced by a single output channel.
 For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples.
+The `$||$' operator is boolean-or, `+' addition, `/=' not-equal, and `$<$' less-then.
 
-We refer to the \emph{endpoint} of a stream as a channel.
-A stream may flow through many input channels on different processes, but can only be produced from at most one output channel.
+A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the currently being executed instruction, annd the @instrs@ the set of labeled instructions.
+We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics.
+When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed.
+Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$.
+If an $\InputState$ is not specified we assume it is `none'.
 
-Each input channel is paired with an input state, which can be either @none@ (no value), @have@ (a value has been pulled), or ($@pending@~\Value$) (a value is available to be pulled).
-These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
-These should be initialised to @none@ for a new process.
-Input states are explained in detail later in \S\ref{s:Process:Eval}.
+In the grammar of Fig.~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable.
+The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes.
+During fusion, the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state.
+We will use a different version of $\InputState$ in \S\ref{s:Fusion} when we define the fusion algorithm.
 
-The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
-This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
-A process may, however, produce multiple output streams.
+The @instrs@ field of the $\Proc$ maps labels to instructions.
+The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel; @case@ which branches depending on the result of a boolean expression; @jump@ causes control to move to a new instruction, and @drop@ which indicates that the current value pulled from a channel is no longer needed.
 
-The heap is used for evaluation of expressions.
-Each process has its own private heap, therefore the only communication between processes occurs by streams.
+All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings which are used to update the heap.
+The update bindings are attached directly to instructions to make the fusion rules easier to specify, in contrast to a presentation with a separate @update@ instruction.
 
-The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
-Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
+As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
+Unlike Kahn processes, pushing to a channel can also block.
+Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.
 
-As usual with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
-Unlike normal Kahn processes however, pushing to a channel can also block: each consumer has a single-element buffer and pushing only succeeds when all buffers are empty.
-After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.
-
-All instructions take as an argument the next label to jump to, as well as any variable updates that should be performed on the private heap at the same time.
-
-A process network is a set of multiple processes that can be evaluated concurrently.
-Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
-Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
-By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.
-
-% \subsection{Map/map}
-% \label{s:Process:MapMap}
-% 
-% One of the simplest combinators is @map@.
-% This might need to go elsewhere.
-% As well as needing a better example than map/map.
-% Let's start with the process definition for @map@.
-% The inputs here is actually a map with @as=none@, but we leave the value off when it is @none@.
-% The next labels for the instructions are also shortened as @map0@ instead of writing the empty heap update afterwards.
-% Mention that the initial heap has the names but no values, but they could be initialised to whatever.
-% It doesn't matter since they'll be written before anything is read.
-% 
-% \begin{code}
-% map f = process (map f)
-%      ins: as
-%     outs: bs
-%     heap: {a = 0}
-%    label: map0
-%   blocks: map0 = pull as    a  map1
-%           map1 = push bs (f a) map2
-%           map2 = drop as       map0
-% \end{code}
-% 
-% As well as a ``combinator network'', a function comprised exclusively of process combinators.
-% The input streams are supplied as arguments, and output streams as return values.
-% \begin{code}
-% mapMap f g xs
-%  = let ys = map f xs
-%        zs = map g ys
-%    in  zs
-% \end{code}
-% 
-% It is not hard to assume that, given the process definitions and a combinator network, we can produce a process network.
-% This is simple enough for a paragraph prose description.
-% It's just a bit of inlining and renaming everything to be unique.
-% 
-% \begin{code}
-% process (map f)
-%      ins: xs
-%     outs: ys
-%     heap: {x}
-%    label: p0
-%   blocks: p0 = pull xs    x  p1
-%           p1 = push ys (f x) p2
-%           p2 = drop xs       p0
-% process (map g)
-%      ins: ys
-%     outs: zs
-%     heap: {y}
-%    label: q0
-%   blocks: q0 = pull ys    y  q1
-%           q1 = push zs (g y) q2
-%           q2 = drop ys       q0
-% \end{code}
-% 
-% Now we can perform some kind of fusion on this network, resulting in one process that computes, as output, both @ys@ and @zs@.
-% Later, when producing imperative code for this, the output pushes to @ys@ can be ignored and changed to jumps, as the original combinator network did not return them.
-% 
-% \begin{code}
-% process (map f / map g)
-%      ins: xs
-%     outs: ys zs
-%     heap: {x, y, _ys}
-%    label: p0q0
-%   blocks: p0q0            = pull xs    x  p1q0
-%           p1q0            = push ys (f x) p2q0-pending-ys { _ys = f x }
-%           p2q0-pending-ys = drop xs       p0q0-pending-ys
-%           p0q0-pending-ys = jump          p0q1-have-ys    { y = _ys }
-%           p0q1-have-ys    = push zs (g y) p0q2-have-ys
-%           p0q2-have-ys    = jump          p0q0
-% \end{code}
+When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely ignore the @drop@ instructions.
+The @drop@ instructions are used to help control how processes being fused should be synchronized, but do not affect the execution of a single process.
+We will discuss @drop@s further in \S\ref{s:Optimisation}.
 
@@
 % -- cuts ---------------------------------------------------------------------
-% BL: I don't think describing iota works at this point. This combinator is not used in the motivating example, so skipping to it seems disjointed.
+% BL: Discuss this during def of evaluation.
 
-% Before describing the @group@ process, we start by looking at one of the simplest combinators, @iota@, which produces a stream of increasing numbers.
-% It takes no inputs, and produces one output stream @xs@.
-% Each process has its own local heap where the values are stored, and in this case we initialise the local variable @i@ to @0@.
-% This variable will be incremented and pushed.
-% Each process also has a current label, which denotes the instruction to perform next.
-% The initial label for @iota@ is @L0@.
-% Each process has a mapping from labels to instructions.
-% In this case we have two instructions, @L0@ and @L1@.
-% The instruction for @L0@ pushes the current value of variable @i@ onto the output stream, then proceeds to move %to label @L1@.
-Instruction @L1@ moves back to @L0@, while also incrementing the variable @i@.
+% These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
 
-% \begin{code}
-% process (iota)
-%      ins: 
-%     outs: xs
-%     heap: {i = 0}
-%    label: L0
-%   blocks: L0 = push xs i  L1
-%           L1 = jump       L0{i = i + 1}
-% \end{code}
+% Claim "functional dataflow" or some such.
+Each process only updates values in its local heap, the only intra-process communication is via streams.
+This is unlike StreamIt which can send out-of-band messages between its processes.
 
-% When executed, this program produces an infinite stream of increasing numbers: $0, 1, 2\ldots$ while the label alternates between @L0@ and @L1@.
+% The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
+This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
+A process may, however, produce multiple output streams.
 
-% The @uniques = group file1@ is a more interesting example.
+% Each process has its own private heap, therefore the only communication between processes occurs by streams.
+
+% The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
+Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
+
+% After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.
+
+% A process network is a set of multiple processes that can be evaluated concurrently.
+Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
+Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
+By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
ins instrs and the next updates need to be maps not sets, as we cant have elems with duplicate keys

@@
 \clearpage{}
+% -----------------------------------------------------------------------------
 \section{Process definitions}
@@
 
-A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the currently being executed instruction, annd the @instrs@ the set of labeled instructions.
+A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the currently being executed instruction, and the @instrs@ a map of labels to instructions.
 We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics.
@@
 The @instrs@ field of the $\Proc$ maps labels to instructions.
-The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel; @case@ which branches depending on the result of a boolean expression; @jump@ causes control to move to a new instruction, and @drop@ which indicates that the current value pulled from a channel is no longer needed.
+The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel; @case@ which branches depending on the result of a boolean expression; @jump@ which causes control to move to a new instruction, and @drop@ which indicates that the current value pulled from a channel is no longer needed.
 
-All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings which are used to update the heap.
-The update bindings are attached directly to instructions to make the fusion rules easier to specify, in contrast to a presentation with a separate @update@ instruction.
+All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a map of heap $\Var$ to $\Exp$ which is used to update the heap.
+The map of update expressions is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.
 
@@
 Evaluation of processes and process networks is split into three parts:
+
 \begin{itemize}
@@
 
+% -----------------------------------------------------------------------------
 \subsubsection{Injection}
@@
 
+% Injecting requires the input buffer to be empty or the process to ignore that channel.
+For networks, all processes must successfully inject the value.
+
+% -----------------------------------------------------------------------------
 \subsubsection{Shaking}
@@
 
+% -----------------------------------------------------------------------------
 \subsubsection{Feeding}

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Flab on what an update is

@@
 
-All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a map of heap $\Var$ to $\Exp$ which is used to update the heap.
-The map of update expressions is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.
+All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap.
+The list of update bindings is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.
 

### 2017-02-26 2017mergingmerges/section/XX-FutureWork.tex
semidecidable wibble

@@
 It seems obvious that if two consumers read the same value and check if it is EOF, they are both going to give the same result, but this is of course undecidable in general.
-By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for semi-decidable equality of predicates.
-Case analysis is still a useful extension and is described in the next section.
+By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
+Case analysis is still a useful extension and is described next.
 

### 2017-02-26 2017mergingmerges/section/02-Processes.tex
Kill commented out stuff in S1 S2 that we've already handled
change to list syntax for variable updates

--- 2017-02-25 2017mergingmerges/section/XX-Processes.tex
@@
 %!TEX root = ../Main.tex
+
+% -----------------------------------------------------------------------------
 \section{Processes, Machines, Combinators and Operators}
@@
 
-The initial instruction @(pull sIn1 v A1 {})@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
-The empty set @{}@ after the target label @A1@ can be used to update heap variables, but as we do not need to update anything yet we leave it empty.
+The initial instruction @(pull sIn1 v A1 [])@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
+The empty list @[]@ after the target label @A1@ can be used to update heap variables, but as we do not need to update anything yet we leave it empty.
 
-Next, the instruction @(case (f || (l /= v)) A2 {} A3 {})@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@.
+Next, the instruction @(case (f || (l /= v)) A2 [] A3 [])@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@.
 We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) for whether this is the first element.
 
-When the predicate is true, the instruction @(push sOut1 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
+When the predicate is true, the instruction @(push sOut1 v A3 [ l = v, f = F ])@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
 
-Finally, the instruction @(drop sIn1 A0 {})@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
+Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
 This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
@@
                , label:  A0
-               , instrs: { A0 = pull sIn1 v          A1 {}
-                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
-                         , A2 = push sOut1 v         A3 { l = v, f = F }
-                         , A3 = drop sIn1            A0 {} } }
+               , instrs: { A0 = pull sIn1 v          A1 []
+                         , A1 = case (f || (l /= v)) A2 []  A3 []
+                         , A2 = push sOut1 v         A3 [ l = v, f = F ]
+                         , A3 = drop sIn1            A0 [] } }
 \end{code}
@@
                    , label:  B0
-                   , instrs: { B0 = pull sIn1  x1   B1 {}
-                             , B1 = pull sIn2  x2   C0 {}
-                             , C0 = case (x1 < x2)  D0 {}  E0 {}
-                             , D0 = push sOut2 x1   D1 {}
-                             , D1 = drop sIn1       D2 {}
-                             , D2 = pull sIn1  x1   C0 {}
-                             , E0 = push sOut2 x2   E1 {}
-                             , E1 = drop sIn2       E2 {}
-                             , E2 = pull sIn2 x2    C0 {} } }
+                   , instrs: { B0 = pull sIn1  x1   B1 []
+                             , B1 = pull sIn2  x2   C0 []
+                             , C0 = case (x1 < x2)  D0 []  E0 []
+                             , D0 = push sOut2 x1   D1 []
+                             , D1 = drop sIn1       D2 []
+                             , D2 = pull sIn1  x1   C0 []
+                             , E0 = push sOut2 x2   E1 []
+                             , E1 = drop sIn2       E2 []
+                             , E2 = pull sIn2 x2    C0 [] } }
 \end{code}
@@
 \subsection{Fusion}
+
 Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
@@
 
-% BL: This is too much low level detail at this point, we need more intuition.
-
-% The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
-
-% The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
-These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
-Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.
-
-% Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
-As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
-In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped.
-The actual @drop@ is only performed once both processes are `none', at label @F9@.
-
 \begin{figure}
@@
 \newcommand\icase[7]{
- \tt{#1} \> \tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \tt{#5} \{\tt{#6}\} \> \tiny #1 \> \tiny = #7 \\
+ \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7 \\
 }
 \newcommand\instr[5]{
- \tt{#1}\>\tt{= #2} \> \tt{#3} \{\tt{#4}\} \> \> \tiny #1 \> \tiny = #5 \\
+ \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5 \\
 }
@@
 @, F17@  \= = @case (f || (l /= v)) @
-         \= @F17{}         @ \= @F17{ }@
+         \= @F17@ \= @[ ]     @ \= @F17@ \= @[ ]@
 @   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
@@
 \end{figure}
+
+% -----------------------------------------------------------------------------
+% BL: This is too much low level detail at this point, we need more intuition.
+
+% The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
+
+% The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
+These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
+Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.
+
+% Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
+As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
+In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped.
+The actual @drop@ is only performed once both processes are `none', at label @F9@.

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Edits to execution

@@
 
-Evaluation of processes and process networks is split into three parts:
+Execution of process networks consists of three aspects:
 
-\begin{itemize}
-\item Injection, where a single value for a stream is inserted into a process' input buffer.
-Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
-Injection for a process network proceeds by injecting the same value into all processes in the network.
+\begin{enumerate}
+\item \emph{Injection}, which attempts to provide a single value gained from a stream to a process, or a set of processes.
+Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.
 
-\item Shaking, where a process takes a step from one label to another.
-Shaking a process results in an updated process and an optional pushed value.
-When a process pushes to a channel, the value is injected to other processes.
-Shaking a process network chooses any process that can be shaked, and updates it in the process network.
+\item \emph{Shaking}, which advances a single process from one instruction to another. Shaking a set of processes succeeds when \emph{any} of the processes in the set can advance.
 
-\item Feeding, where an environment of input values are fed to the process network, and output values are collected.
-Feeding alternates between injecting and shaking: injecting values from the environment when processes are ready to receive values, and repeatedly shaking the processes to collect their outputs.
-\end{itemize}
+\item \emph{Feeding}, which manages communication between separate processes in the network.
+Feeding alternates between Injecting and Shaking.
+When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input.
+If they all accept it then we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
 
-Evaluation of a process network is non-deterministic, in that at any point there are many possible processes that can take a step.
-However, because each process itself is deterministic and has blocking reads, overall evaluation is deterministic as per Kahn process networks.
-That is: the order in which values are pushed to different output streams is not deterministic, but the order and values for a particular output stream \emph{are} deterministic.
+\end{enumerate}
 
-% Note that while process network evaluation is non-deterministic and concurrent, evaluating a single process is sequential and deterministic: code generation for fused processes only needs to deal with the sequential case.
+Evaluation of a process network is non-deterministic.
+At any moment several processes may be able to take a step, while others are blocked trying to pull or push to their channels.
+However, because each process itself is deterministic, and uses blocking reads, the sequence of values pushed to each stream is deterministic, as per Kahn process networks.
+
+Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not.
+When we fuse two processes together we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.
+
+Each output channel may be pushed to by a single process only, so in a sense each output channel is ``owned'' by a single process.
+The only intra-process communication is via channels and streams.
+Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes.
+This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to the formal input and output streams.
 
@@
 This can be either @none@, in which case the buffer is empty and a value may be injected; @pending@, where a single value has been injected to the buffer but has not been pulled yet; or @have@, where a value has already been injected and is currently being used.
+
+% Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
+Injection for a process network proceeds by injecting the same value into all processes in the network.
 

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Treat records like maps from labels to fields

@@
 % -----------------------------------------------------------------------------
+\eject{}
 \subsubsection{Injection}
-Fig.~\ref{fig:Process:Eval:Inject} defines the judgment forms and rules for injection.
-Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.
+Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
+The judgment form $\ProcInject{p}{v}{c}{p'}$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
+The @injects@ form is similar, but for sets of processes rather than a single one.
+
+Rule (InjectValue) handles the injection of a single value into a single process.
+
+% Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.
 
@@
 Injection for a process network proceeds by injecting the same value into all processes in the network.
-
-The judgment form ($\ProcInject{p}{c}{v}{p'}$) means that the process $p$ has the value $v$ inserted to the input buffer for channel $c$, resulting in the updated process $p'$.
 

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
S3: Copy edits to injection

@@
 The judgment form $\ProcInject{p}{v}{c}{p'}$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
-The @injects@ form is similar, but for sets of processes rather than a single one.
+The @injects@ form is similar, but operates on sets of processes rather than a single one.
 
-Rule (InjectValue) handles the injection of a single value into a single process.
+Rule (InjectValue) injects of a single value into a single process.
+The value is stored as a (@pending@~ v) binding in the $\InputState$ for the associated channel of the process.
+The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.
 
-% Injection is used by the other parts of evaluation to \emph{inject} stream values into a process' input buffer.
+Rule (InjectIgnore) allows processes that do not use a particular named channel to ignore values injected into that channel.
 
-A process has a single element buffer for each input, stored in its input state.
-This can be either @none@, in which case the buffer is empty and a value may be injected; @pending@, where a single value has been injected to the buffer but has not been pulled yet; or @have@, where a value has already been injected and is currently being used.
-
-% Injection is only possible when the process' input buffer is empty, or the process ignores that stream.
-Injection for a process network proceeds by injecting the same value into all processes in the network.
-
-Rule (InjectValue) allows a value to be injected only when the input state is @none@, meaning the buffer is empty.
-An attempt to inject a value while the buffer is @pending@ or @have@ would require an unbounded, or at least multiple element, buffer.
-We use the syntax ($\HeapUpdateOne{@ins@~p}{c}{@pending@~v}$) to perform a map update, in this case removing the ($c=@none@$) from ($@ins@~p$) and then inserting ($c=@pending@~v$).
-This sets the input buffer as pending, meaning that any future attempts to inject to this channel will fail, until the process pulls and then drops the value.
-
-Rule (InjectIgnore) allows processes that do not use a particular input stream to ignore an injected input.
-
-Rule (ProcessesInject) performs injection over a process network.
-Every process in the network must have the value injected into it.
-This means if multiple processes read from that stream, all input buffers for that stream must be empty.
-
-% Injecting requires the input buffer to be empty or the process to ignore that channel.
-For networks, all processes must successfully inject the value.
+Rule (InjectMany) attempts to inject a single value into a set of processes.
+We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of the.
+Once a value has been injected into all consuming processes that require it, it is then safe for a producing process to discard it.
 

### 2017-02-26 2017mergingmerges/section/A1-Combinators.tex
clean up appendix

--- 2017-01-08 2017mergingmerges/section/A1-Combinators.tex
@@
 We start with simple combinators supported by most streaming systems, and progress to more interesting combinators.
-Some combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but can be implemented in our finite stream extension described in~\S\REF.
+Some standard combinators such as @fold@, @take@ and @append@ are missing due to the infinite nature of our streams, but could be implemented with the finite stream extension described in~\S\ref{s:Finite}.
 The fact that segmented versions of these combinators can be implemented is compelling evidence of this.
 
-\subsection{Standard combinators}
+Many of these combinators take a ``default'' argument, which is used to initialise the heap, but the stored value is never actually read.
+Ideally these could be left unspecified, or the heap left uninitialised in cases where it is never read.
 
-\paragraph{Map} is one of the simplest combinators, and is supported almost everywhere.
+The map combinator applies a function to every element of the stream.
 
+\begin{alltt}
+    map 
+        = \(\lambda\) (f : \(\alpha \to \beta\)) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\beta\)). 
+          \(\nu\) (a: \(\alpha\)) (L0..L2: Label).
+\end{alltt}
 \begin{code}
-map f = process
-     ins: as
-    outs: bs
-    heap: {a}
-   label: p0
-  blocks: p0 = pull as    a  p1
-          p1 = push bs (f a) p2
-          p2 = drop as       p0
+         process
+         { ins:    { sIn }
+         , outs:   { sOut }
+         , heap:   { a = default }
+         , label:  L0
+         , instrs: { L0 = pull sIn     a  L1 []
+                   , L1 = push sOut (f a) L2 []
+                   , L2 = drop sIn        L0 [] } }
 \end{code}
 
-\paragraph{Filter} is another ubiquitous combinator.
+Filter returns a new stream containing only the elements that satisfy some predicate.
 
+\begin{alltt}
+    filter 
+        = \(\lambda\) (f : \(\alpha \to\) Bool) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\alpha\)). 
+          \(\nu\) (a: \(\alpha\)) (L0..L3: Label).
+\end{alltt}
 \begin{code}
-filter f = process
-     ins: as
-    outs: bs
-    heap: {a}
-   label: p0
-  blocks: p0 = pull as    a  p1
-          p1 = case    (f a) p2 p3
-          p2 = push bs    a  p2
-          p3 = drop as       p0
+         process
+         { ins:    { sIn }
+         , outs:   { sOut }
+         , heap:   { a = default }
+         , label:  L0
+         , instrs: { L0 = pull sIn  a     L1 []
+                   , L1 = case   (f a)    L2 []  L3 []
+                   , L2 = push sOut a     L3 []
+                   , L3 = drop sIn        L0 [] } }
 \end{code}
 
-\paragraph{Partition} is an interesting combinator since it produces two output streams at once.
-The two output streams have different rates.
-This sort of splitting is unsupported in pull fusion systems.
+\newpage
 
+Partition is similar to filter, but has two output streams: those that satisfy the predicate, and those that do not.
+Partition is an inherently push-based operation, and cannot be supported by pull streams without buffering.
+
+\begin{alltt}
+    partition 
+        = \(\lambda\) (f : \(\alpha \to\) Bool) (default : \(\alpha\))
+            (sIn: Stream \(\alpha\)) (sOut1: Stream \(\alpha\)) (sOut2: Stream \(\alpha\)). 
+          \(\nu\) (a: \(\alpha\)) (L0..L4: Label).
+\end{alltt}
 \begin{code}
-partition f = process
-     ins: as
-    outs: ts fs
-    heap: {a}
-   label: p0
-  blocks: p0 = pull as    a  p1
-          p1 = case    (f a) pT pF
-          pT = push ts    a  p3
-          pF = push fs    a  p3
-          p3 = drop as       p0
+         process
+         { ins:    { sIn }
+         , outs:   { sOut1, sOut2 }
+         , heap:   { a = default }
+         , label:  L0
+         , instrs: { L0 = pull sIn   a    L1 []
+                   , L1 = case    (f a)   L2 []  L3 []
+                   , L2 = push sOut1 a    L4 []
+                   , L3 = push sOut2 a    L4 []
+                   , L4 = drop sIn        L0 [] } }
 \end{code}
 
-\paragraph{Zip} takes two streams and combines them into one.
-It is fairly ubiquitous in pull fusion systems, but cannot be supported by push systems.
+Zip, or zip-with, pairwise combines two input streams.
+Zipping is an inherently pull-based operation.
 
+\begin{alltt}
+    zipWith 
+        = \(\lambda\) (f : \(\alpha \to \beta \to \gamma\)) (default1 : \(\alpha\)) (default2 : \(\beta\))
+            (sIn1: Stream \(\alpha\)) (sIn2: Stream \(\beta\)) (sOut: Stream \(\gamma\)). 
+          \(\nu\) (a: \(\alpha\)) (b : \(\beta\)) (L0..L4: Label).
+\end{alltt}
 \begin{code}
-zip = process
-     ins: as bs
-    outs: abs
-    heap: {a, b}
-   label: p0
-  blocks: p0 = pull as    a   p1
-          p1 = pull bs    b   p2
-          p2 = push abs (a,b) p3
-          p3 = drop as        p4
-          p4 = drop bs        p0
+         process
+         { ins:    { sIn1, sIn2 }
+         , outs:   { sOut }
+         , heap:   { a = default1, b = default2 }
+         , label:  L0
+         , instrs: { L0 = pull sIn1 a       L1 []
+                   , L1 = pull sIn2 b       L2 []
+                   , L2 = push sOut (f a b) L3 []
+                   , L3 = drop sIn1         L4 []
+                   , L4 = drop sIn2         L0 [] } }
 \end{code}
 
-\paragraph{Scan} performs a prefix sum over an input stream, much like a fold that returns every intermediate value.
-For simplicity we are performing just prefix sums, rather than arbitrary scans, but this is easy to generalise.
-Here we use two values in the heap; @a@ stores the last seen input and @s@ stores the current sum.
-We initialise @s@ to 0 at the start.
-Note that this prefix scan includes the seed value 0 in the output; updating the current sum to include the last pulled value happens after the push.
+Scan is similar to a fold, but instead of returning a single value at the end, it returns an intermediate value for each element of the stream.
 
+\begin{alltt}
+    scan 
+        = \(\lambda\) (k : \(\alpha \to \beta \to \beta\)) (z : \(\beta\)) (default : \(\alpha\)) (sIn: Stream \(\alpha\)) (sOut: Stream \(\beta\)).
+          \(\nu\) (a: \(\alpha\)) (s : \(\beta\)) (L0..L2: Label).
+\end{alltt}
 \begin{code}
-scan = process
-     ins: as
-    outs: bs
-    heap: {a, s = 0}
-   label: p0
-  blocks: p0 = pull as a  p1
-          p1 = push bs s  p2[s = s + a]
-          p2 = drop as    p0
+         process
+         { ins:    { sIn  }
+         , outs:   { sOut }
+         , heap:   { a = default, s = z }
+         , label:  L0
+         , instrs: { L0 = pull sIn  a       L1 []
+                   , L1 = push sOut s       L2 [ s = f a s ]
+                   , L2 = drop sIn          L0 [] } }
 \end{code}
 
-\paragraph{Unfold} creates a new stream from a seed and an iteration function.
-The seed has type @s@, the iteration has type @s -> (a,s)@ and the output stream contains @a@s.
-This process has only one state which pushes the output, calls the iteration function, and updates the state.
-The state/heap is initialised to the first result of the iteration function.
+\newpage
 
+Segmented fold performs a fold over each nested stream, using a segmented representation.
+Here we are representing nested streams using one stream for the lengths of each substream, and another stream containing the values.
+The output stream has the same rate as the lengths stream.
+It reads a count (@c@) from the lengths stream, setting the fold state to zero (@z@).
+Then it reads count times from the values stream, updating the fold state.
+Afterwards, it pushes the final fold state, and continues to read a new count.
+
+\begin{alltt}
+    folds 
+        = \(\lambda\) (k : \(\alpha \to \beta \to \beta\)) (z : \(\beta\)) (default : \(\alpha\))
+            (sLens: Stream Nat) (sVals: Stream \(\alpha\)) (sOut: Stream \(\beta\)).
+          \(\nu\) (c : Nat) (a: \(\alpha\)) (s : \(\beta\)) (L0..L5: Label).
+\end{alltt}
 \begin{code}
-unfold z f = process
-     ins: 
-    outs: as
-    heap: {s = f z}
-   label: p0
-  blocks: p0 = push as (fst s) p0[s = f (snd s)]
+         process
+         { ins:    { sLens, sVals }
+         , outs:   { sOut }
+         , heap:   { c = 0, a = default, s = z }
+         , label:  L0
+         , instrs: { L0 = pull sLens c      L1 [ s = z ]
+                   , L1 = case (c > 0)      L2 []  L4 []
+                   , L2 = pull sVals a      L3 []
+                   , L3 = drop sVals        L1 [ c = c - 1, s = k s a ]
+                   , L4 = push sOut  s      L5 []
+                   , L5 = drop sLens        L0 [] } }
 \end{code}
 
-\paragraph{Merge} takes two sorted inputs and produces a sorted input containing both inputs.
-It is the main worker in merge sort.
-Merge is a particularly interesting combinator since the access patterns are data dependent; that is, which order it pulls from inputs depends on the values of the inputs.
-This is the infinite version of merge; the finite version would have a separate `fixup' stage where after one input has finished, it copies the rest of the other input.
+Segmented take computes an @n@-length prefix of each nested stream.
+It starts by reading a count from the lengths stream, then copies at most @n@ elements.
+If there are leftovers, it pulls and discards them, then pulls the next length.
 
+\begin{alltt}
+    takes 
+        = \(\lambda\) (n : Nat) (default : \(\alpha\))
+            (sLens: Stream Nat) (sVals: Stream \(\alpha\))
+            (oLens: Stream Nat) (oVals: Stream \(\alpha\)).
+          \(\nu\) (c : Nat) (take : Nat) (ix : Nat) (a: \(\alpha\)) (L0..L9: Label).
+\end{alltt}
 \begin{code}
-merge = process
-     ins: as bs
-    outs: abs
-    heap: {a, b}
-   label: pI0
-  blocks: pI0 = pull as a  pI1
-          pI1 = pull bs b  pC0
-          pC0 = case (a < b) pA0 pB0
+         process
+         { ins:    { sLens, sVals }
+         , outs:   { oLens, oVals }
+         , heap:   { c = 0, take = 0, ix = 0, a = default }
+         , label:  L0
+         , instrs: { L0 = pull sLens c      L1 [ ix = 0, take = min count n ]
+                   , L1 = push oLens take   L2 []
 
-          pA0 = push abs a pA1
-          pA1 = drop as    pA2
-          pA2 = pull as  a pC0
+                   , L2 = case (ix < take)  L3 []  L6 []
+                   , L3 = pull sVals a      L4 []
+                   , L4 = push oVals a      L5 []
+                   , L5 = drop sVals        L2 [ ix = ix + 1 ]
 
-          pB0 = push abs b pB1
-          pB1 = drop bs    pB2
-          pB2 = pull bs  b pC0
+                   , L6 = case (ix < c)     L7 []  L9 []
+                   , L7 = pull sVals a      L8 []
+                   , L8 = drop sVals        L6 [ ix = ix + 1 ]
+
+                   , L9 = drop sLens        L0 [] } }
 \end{code}
 
-\subsection{Segmented combinators}
+\newpage
 
-\paragraph{Segmented fold}
-takes two input streams; one for the segment lengths, one for the values.
-The output stream has the same rate as the segment lengths.
+Segmented append takes two segmented streams as input, and appends each nested stream.
+It starts by reading a length from both lengths streams into @a@ and @b@, and pushes the sum of both lengths.
+It then copies over @a@ elements from the first values stream, then copies over @b@ elements from the second values stream.
 
+\begin{alltt}
+    appends 
+        = \(\lambda\) (default : \(\alpha\))
+            (aLens: Stream Nat) (aVals: Stream \(\alpha\)) (bLens: Stream Nat) (bVals: Stream \(\alpha\))
+            (oLens: Stream Nat) (oVals: Stream \(\alpha\)).
+          \(\nu\) (a : Nat) (b : Nat) (v: \(\alpha\)) (L0..L12: Label).
+\end{alltt}
 \begin{code}
-folds k z = process
-     ins: lens vals
-    outs: outs
-    heap: {count, state, val}
-   label: init0
-  blocks: init0 = pull lens count go0[state = z]
-          go0   = case (count > 0) go1 done0
-          go1   = pull vals val    go2
-          go2   = jump go0[count = count - 1, state = k state val]
-          done0 = push outs state init0
+         process
+         { ins:    { aLens, aVals, bLens, bVals }
+         , outs:   { oLens, oVals }
+         , heap:   { a = 0, b = 0, v = default }
+         , label:  L0
+         , instrs: { L0  = pull aLens  a     L1 []
+                   , L1  = pull bLens  b     L2 []
+                   , L2  = push oLens (a+b)  L3 []
+
+                   , L3  = case (a > 0)      L4 []  L7 []
+                   , L4  = pull aVals v      L5 []
+                   , L5  = push oVals v      L6 []
+                   , L6  = drop aVals        L3 [ a = a - 1 ]
+
+                   , L7  = case (b > 0)      L8 []  L11[]
+                   , L8  = pull bVals v      L9 []
+                   , L9  = push oVals v      L10[]
+                   , L10 = drop bVals        L7 [ b = b - 1 ]
+
+                   , L11 = drop aLens        L12[]
+                   , L12 = drop bLens        L0 [] } }
 \end{code}
-
-\paragraph{Segmented take}
-takes two input streams; the segment descriptor lengths, as well as the underlying values.
-It outputs another two streams for the output segment descriptor and values.
-The number of elements to take is constant across all segments.
-
-\begin{code}
-takes n = process
-     ins: lens vals
-    outs: olens ovals
-    heap: {count, takecount, ix, val}
-   label: init0
-  blocks: init0 = pull lens count lens0
-          lens0 = jump lens1
-                [ ix = 0
-                , takecount = min count n ]
-          lens1 = push olens takecount take0
-          take0 = case (ix < takecount) take1 drop0
-          take1 = pull vals val take2
-          take2 = push ovals val take0[ix = ix + 1]
-          drop0 = case (ix < count) drop1 init0
-          drop1 = pull vals val drop0[ix = ix + 1]
-\end{code}
-
-\paragraph{Segmented append} is very similar to merge.
-
-\begin{code}
-apps = process
-     ins: Alens Avals Blens Bvals
-    outs: olens ovals
-    heap: {Alen, Blen, val}
-   label: init0
-  blocks: init0 = pull Alens Alen init1
-          init1 = pull Blens Blen init2
-          init2 = push olens  (Alen + Blen) copyA0
-          copyA0 = case (Alen > 0) copyA1 copyB0
-          copyA1 = pull Avals val copyA2
-          copyA2 = push ovals val copyA0[Alen = Alen - 1]
-          copyB0 = case (Blen > 0) copyB1 init0
-          copyB1 = pull Bvals val copyB2
-          copyB2 = push ovals val copyB0[Blen = Blen - 1]
-\end{code}

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Shift Action grammar to process def fig
Only talk about single process evaluation in 'Shaking'

@@
 Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
-The judgment form $\ProcInject{p}{v}{c}{p'}$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
+A judgment of form $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
 The @injects@ form is similar, but operates on sets of processes rather than a single one.
@@
 \subsubsection{Shaking}
-Fig.~\ref{fig:Process:Eval:Shake} defines the judgment forms and rules for shaking a process, allowing it to take a step, and produce an output message.
-% I don't like them, but alpha and tau are standard names in process calculi. 
-The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.
+Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process.
+The first set of rules handle specific instructions.
+A judgment of form $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs action $\alpha$.'' An action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
+We write ~$\cdot$~ for an empty action.
 
-The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$
-executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
+% The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.
+
+% The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$ executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
 The result contains an output message $\alpha$, as well as the new label $l'$, the new input state $i'$, and the update substitution to apply to the heap $u'$.
 
+\eject{}
 Rule (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Fixing pagination in feeding

@@
 The first set of rules handle specific instructions.
-A judgment of form $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs action $\alpha$.'' An action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
+A judgment of form $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs an output action $\alpha$.'' An output action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
 We write ~$\cdot$~ for an empty action.
+
+\eject{}
+Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update that will copy this value into the variable $x$ named in the @pull@ instruction.
+We use the syntax $us \rhd x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$.
+In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied.
+
+Rule (Push) evaluates the expression $e$ under heap $\Sigma$ to a value $v$, and produces a corresponding action which carries this value.
+The judgment $(\Sigma \vdash e \Downarrow v)$ is standard simply typed lambda calculus reduction using the heap $\Sigma$ for the values of free variables.
+As this evaluation is completely standard we omit it to save space.
+
+Rule (Drop) changes the input buffer state from @have@ to @none@.
+A drop can only be executed after pull has set the input buffer to @have@.
+Rule (Jump) simply produces a new label and update expressions.
+Rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and jump to the appropriate label.
+
+A judgment of form $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ transitions to new process $p'$, yielding action $\alpha$''.
+Rule (Shake) advances a whole process.
+We lookup the current instruction pointed to by the processes @label@ and pass it, along with the current channel states and heap to the previous single-instruction judgment.
+The update expressions @us@ that the single-instruction judgment yields are first reduced to values before updating the heap.
+The update expressions themselves are all pure, so the evaluation can safely be done in parallel (or in arbitrary order).
+
+% The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
+The process shaking shakes a single instruction and updates the process.
+Shaking a process network chooses a single process to shake and updates it in the network.
+If shaking the process resulted in an output value, that value is injected into all the other processes in the network.
+
+% Rule (Shake) takes a single process and shakes its instruction.
+The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
+The process steps to the new label with the new input state, and the values are updated in the heap.
 
@@
 
-\eject{}
-Rule (Pull) takes an already injected value from the input buffer, which changes its state from @pending@ to @have@.
-The result substitution sets the variable to the pulled value, as well as any substitutions in the \Next~ of the instruction.
+% -----------------------------------------------------------------------------
+\subsubsection{Feeding}
+Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting the output actions and feeding the carried values to other processes.
+The first set of rules concerns feeding output to other processes within the same group, while the second manages input and output values exchange with the environment the process network is running in.
 
-Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.
-
-Rule (Push) evaluates the push expression $e$ under the heap, and sends that value as output the message.
-
-Rule (Jump) simply returns the new label and substitution.
-
-Rules (CaseT) and (CaseF) evaluate the case expression $e$ and jump to the true or false label depending on the value.
-
-The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
-The process shaking shakes a single instruction and updates the process.
-Shaking a process network chooses a single process to shake and updates it in the network.
-If shaking the process resulted in an output value, that value is injected into all the other processes in the network.
-
-Rule (Shake) takes a single process and shakes its instruction.
-The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
-The process steps to the new label with the new input state, and the values are updated in the heap.
+The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
+The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
+The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
+Feeding evaluates the process network until all input values have been injected.
 
@@
 
-% -----------------------------------------------------------------------------
-\subsubsection{Feeding}
-Fig.~\ref{fig:Process:Eval:Feed} defines the judgment forms and rules for feeding, where external input values are fed into a process network and output values are accumulated.
-The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
-The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
-The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
-Feeding evaluates the process network until all input values have been injected.
+% Evaluation: feeding evaluates a process network on a list of input values and collects the outputs.
+% Feeding alternates between injecting input values and shaking the processes to collect the output messages.
 

### 2017-02-26 2017mergingmerges/code/src/Machine/Transform/Fuse.hs
Code: tryStepPair: prefer jumps, defer pulls.

@@ -161,6 +161,29 @@ tryStepPair
         (label1, csState1) instr1
         (label2, csState2) instr2
 
+ -- If both processes can advance, use some heuristics to decide which to perform first.
+ -- If one instruction is a jump, prefer that one
+ -- If one instruction is a pull, defer that one
+ | Just instr1  <- tryStep csMode
+                        (label1, csState1) instr1
+                        (label2, csState2)
+ , Just instr2  <- tryStep csMode
+                        (label2, csState2) instr2
+                        (label1, csState1)
+ = case (instr1, instr2) of
+    -- PreferJump
+    (Jump{}, _)     -> Just instr1
+    (_, Jump{})     -> Just $ swapLabelsOfInstruction instr2
+
+    -- DeferPull: if both are trying to pull, perform the first one.
+    -- If the second is trying to pull, perform the first.
+    (Pull{},Pull{}) -> Just instr1
+    (_,Pull{})      -> Just instr1
+    (Pull{},_)      -> Just $ swapLabelsOfInstruction instr2
+
+    -- Finally, if both can advance, perform the first
+    (_,_)           -> Just instr1
+
  -- Try to advance the first instruction.
  | Just instr'  <- tryStep csMode 
                         (label1, csState1) instr1

### 2017-02-26 2017mergingmerges/code/src/Test/Comb.hs

@@ -45,7 +45,7 @@ fusePipeAB    comb1 comb2
                 proc1     <- combMk comb1 csIns1     [cInternal]
                 proc2     <- combMk comb2 [cInternal] csOuts2
                 case fusePair proc1 proc2 of
-                 Left err       -> error err
+                 Left err       -> error (show err)
                  Right proc'    
                   -> return $ tagHow "pipe" proc'
    ]
@@ -86,7 +86,7 @@ fuseSplitAB    comb1 comb2
                 proc1       <- combMk comb1 [cIn] csOuts1
                 proc2       <- combMk comb2 [cIn] csOuts2
                 case fusePair proc1 proc2 of
-                 Left  err      -> error err
+                 Left  err      -> error (show err)
                  Right proc'    
                   -> return $ tagHow "split" proc'
    ]

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Edits to feeding

@@
 
-A judgment of form $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ transitions to new process $p'$, yielding action $\alpha$''.
+A judgment of form $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $\alpha$''.
 Rule (Shake) advances a whole process.
@@
 
-% The judgment forms for shaking processes and process networks are $\ProcShake{p}{\alpha}{p'}$ and $\ProcsShake{\sgl{p}}{\alpha}{\sgl{p'}}$.
-The process shaking shakes a single instruction and updates the process.
-Shaking a process network chooses a single process to shake and updates it in the network.
-If shaking the process resulted in an output value, that value is injected into all the other processes in the network.
-
-% Rule (Shake) takes a single process and shakes its instruction.
-The judgment ($\ExpEval{@heap@~p}{u}{\Sigma}$) evaluates all the expressions in the update substitution $u$, under the process heap ($@heap@~p$).
-The process steps to the new label with the new input state, and the values are updated in the heap.
-
-% The type of messages is denoted by $\alpha$, which can either be ($\Push~\Chan~\Value$) for pushing a value to a particular channel, or $\tau$ for an `internal' message which only performs local state updates.
-
-% The judgment form for shaking a single instruction $\ProcBlockShake{x}{i}{\Sigma}{\alpha}{l'}{i'}{u'}$ executes an instruction $x$ with the input states $i$ and the heap $\Sigma$.
-The result contains an output message $\alpha$, as well as the new label $l'$, the new input state $i'$, and the update substitution to apply to the heap $u'$.
-
 % -----------------------------------------------------------------------------
 \subsubsection{Feeding}
-Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting the output actions and feeding the carried values to other processes.
-The first set of rules concerns feeding output to other processes within the same group, while the second manages input and output values exchange with the environment the process network is running in.
+Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting the output actions and feeding output values to other processes.
+The first set of rules concerns feeding values to other processes within the same process network, while the second exchanges input and output values with the environment the process network is running in.
+
+\eject{}
+A judgment of form $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$.
+A process ``group'' is just a set of processes.
+
+Rule (ProcessInternal) allows an arbitrary process in the group to advance to a new state at any time, provided it does not produce an output action.
+This allows processes to perform internal computation, without needing to synchronize with the rest of the group.
+
+Rule (ProcessPush) allows an arbitrary process in the group to advance to a new state, while producing an output action (@push@ c v).
+For this to happen it must be possible to inject the output value @v@ into any processes that has channel @c@ as one of its inputs.
+As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process.
+When any process in the group produces an output action then we take that as the action of the whole group, as it might also need to be sent to the environment.
+
+\smallskip
 
@@
 
-Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces an internal message ($\tau$).
-
-Rule (ProcessesInternal) allows the process network to shake, when one of the processes inside it produces a push message.
-The emitted push message is then injected into all other processes in the network, meaning they must either ignore the channel or be ready to receive it in their buffer.
-If the process tries to emit a push message but it cannot be injected into all other processes, the push fails and another process will be tried.
-The entire network then emits the same message.
-
 % Evaluation: feeding evaluates a process network on a list of input values and collects the outputs.
-% Feeding alternates between injecting input values and shaking the processes to collect the output messages.
+Feeding alternates between injecting input values and shaking the processes to collect the output messages.
 

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Shifting figs around

@@
 \section{Process definitions}
-
 \input{figures/ProcessDef.tex}
@@
 
-\input{figures/ProcessEval.tex}
-\input{figures/ProcessFeed.tex}
-
 Execution of process networks consists of three aspects:
@@
 This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to the formal input and output streams.
+
+\input{figures/ProcessEval.tex}
+\input{figures/ProcessFeed.tex}
 
@@
 
-The judgment form for feeding is $\ProcsFeed{\ti{inputs}}{\ti{network}}{\ti{streams}}{\ti{network}'}$.
-The input map $\ti{inputs}$ contains values for the network inputs: network outputs are not allowed, but ignored channels can have values.
-The result $\ti{streams}$ contains the original inputs as well as accumulated output values.
-Feeding evaluates the process network until all input values have been injected.
-
-% Evaluation: feeding evaluates a process network on a list of input values and collects the outputs.
-Feeding alternates between injecting input values and shaking the processes to collect the output messages.
-
-% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
-% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
+A judgment of the form $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, process group $ps$ takes a step and produces new channel values $cvs'$ and group $ps'$''.
+The channel values $cvs$ map channel names to a list of values.
+For input channels of the overall group, we initialize the map to contain a list of input values for each channel.
+For output channels of the overall group, values pushed to those channels are also collected in the same channel map.
+In a concrete implementation the input and output values would be transported over some IO device, but the simplified semantics here is sufficient to describe the abstract behavior of our system.
 
@@
 The last value is then injected into the network, and added back to the result accumulators.
+
+% Evaluation: feeding evaluates a process network on a list of input values and collects the outputs.
+Feeding alternates between injecting input values and shaking the processes to collect the output messages.
+
+% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
+% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
 

### 2017-02-26 2017mergingmerges/section/03-Definition.tex
Split Inject fig into separate .tex and shift figs in doc
Work on alt2 example to describe diff eval orders

@@
 
-\input{figures/ProcessEval.tex}
-\input{figures/ProcessFeed.tex}
-
 % -----------------------------------------------------------------------------
-\eject{}
 \subsubsection{Injection}
 Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
-A judgment of form $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
+The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
 The @injects@ form is similar, but operates on sets of processes rather than a single one.
@@
 
+\input{figures/ProcessInject}
+\input{figures/ProcessEval.tex}
+\input{figures/ProcessFeed.tex}
+
 % -----------------------------------------------------------------------------
@@
 The first set of rules handle specific instructions.
-A judgment of form $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs an output action $\alpha$.'' An output action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
+The statement $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs an output action $\alpha$.'' An output action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
 We write ~$\cdot$~ for an empty action.
@@
 Rule (Push) evaluates the expression $e$ under heap $\Sigma$ to a value $v$, and produces a corresponding action which carries this value.
-The judgment $(\Sigma \vdash e \Downarrow v)$ is standard simply typed lambda calculus reduction using the heap $\Sigma$ for the values of free variables.
+The judgment $(\Sigma \vdash e \Downarrow v)$ expresses standard simply typed lambda calculus reduction using the heap $\Sigma$ for the values of free variables.
 As this evaluation is completely standard we omit it to save space.
 
-Rule (Drop) changes the input buffer state from @have@ to @none@.
-A drop can only be executed after pull has set the input buffer to @have@.
-Rule (Jump) simply produces a new label and update expressions.
-Rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and jump to the appropriate label.
+Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.
 
-A judgment of form $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $\alpha$''.
+Rule (Jump) simply produces a new label and update expressions and rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and jump to the appropriate label.
+
+The statement $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $\alpha$''.
 Rule (Shake) advances a whole process.
@@
 \eject{}
-A judgment of form $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$.
-A process ``group'' is just a set of processes.
+The statement $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$. A process ``group'' is just a set of processes.
 
@@
 \smallskip
-
-A judgment of the form $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, process group $ps$ takes a step and produces new channel values $cvs'$ and group $ps'$''.
+The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, process group $ps$ takes a step and produces new channel values $cvs'$ and group $ps'$''.
 The channel values $cvs$ map channel names to a list of values.
@@
 
-Rule (FeedStart) applies when all input values have been injected and there are no input values left.
-In this case, the output values are the same as the input values.
+Rule (FeedInternal) allows the process group to perform local computation in the context of the channel values.
 
-Rule (FeedInternal) allows the process network to take an internal step.
-It first feeds its input accumulator and process network, then allows the resulting network to take an internal step.
+Rule (FeedPush) collects an output action produced by a process group and adds it to the list corresponding to the output channel.
 
-Rule (FeedPush) allows the process network to emit a push message.
+Rule (FeedExternal) injects values from the external environment.
+This rule also has the side condition that values cannot be injected from the environment into output channels that are already owned by some process.
+This constraint is used during formal proofs of correctness, but would not need to be checked dynamically in a concrete implementation.
+The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.
+
+\subsection{Evaluation Order}
+
+Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
+\begin{code}
+  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
+  alternates sIn1 sIn2 sIn3
+   = let  sA1     = alt2 sIn1 sIn2
+          sA2     = alt2 sIn2 sIn3
+     in   zip sA1 sA2
+\end{code}
+
+Given the inputs @sIn1@ = $[a1, a2]$, @sIn2@ = $[b1, b2]$ and @sIn3@ = $[c1, c2]$ the \\ output will be $[ (a1, b1), (a2, b2), (b1, c1), (b2, c2) ]$
+
+\TODO{Explain why this needs a specific evaluation order, and some do not work. Needs to fit on this page. Lead into the fusion order talk later.}
+
+Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately.
+At evaluation time we need the results pushed to sA1 and sA2 in the sA1 sA2 sA1 sA2 order, not sA1 sA1 sA2 sA2.
+Writing the rules nondeterministically allows the elaborator to discover a usable order, if there is one.
+This also affects fusion, we don't want to commit to the wrong order too early.
+We shall see that if we fuse the two alt processes first fusion will not work.
+We need to start with zip so that the order in which input elems arrive is constrained.
+
+% Rule (FeedPush) allows the process network to emit a push message.
 As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message.
@@
 
-Rule (FeedExternal) allows inputs to be injected into the process network.
-For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
+% Rule (FeedExternal) allows inputs to be injected into the process network.
+% For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
 The recursive feed is evaluated with the last value removed from the accumulators.
 The last value is then injected into the network, and added back to the result accumulators.
+
+% Rule (FeedStart) applies when all input values have been injected and there are no input values left. In this case, the output values are the same as the input values.
 

### 2017-02-26 2017mergingmerges/code/src/Machine/Base.hs

@@ -125,6 +125,7 @@ arityOfPrim p
         PAnd            -> 2
         PInt{}          -> 0
         PAdd            -> 2
+        PTuple n        -> n
         PEq             -> 2
         PNeq            -> 2
         PLt             -> 2

### 2017-02-26 2017mergingmerges/code/src/Machine/Execute.hs

@@ -22,8 +22,15 @@ execute cvsIn cvsOut ps acc
  | all null $ map snd $ Map.toList cvsIn
  = (cvsIn, ps, acc)
 
+ | Just (cvsIn', ps')   <- feedProcesses cvsIn ps
+ , (ps'', as')          <- shakeSteps [] ps' []
+ = execute cvsIn' cvsOut ps'' (acc ++ as')
+
  | otherwise
- = let  (cvsIn', ps')   
+ = (cvsIn, ps, acc)
+
+ {-
+
          = fromMaybe (error $ unlines
                             [ "execute: feed failed"
                             , P.ppShow cvsIn
@@ -34,7 +41,7 @@ execute cvsIn cvsOut ps acc
          = shakeSteps [] ps' []
 
    in   execute cvsIn' cvsOut ps'' (acc ++ as')
-
+-}
 
 
 

### 2017-02-26 2017mergingmerges/code/src/Machine/Execute/Eval.hs

@@ -30,6 +30,10 @@ eval heap@(Heap hsHeap) xx
 -- | Evaluate a primitive applied to some values.
 evalPrimValue :: Prim -> [Value] -> Value
 evalPrimValue p vs
+ | PTuple n <- p
+ , length vs == n
+ = VTuple vs
+
  | length vs == arityOfPrim p
  = case (do ps <- sequence $ map takePrimOfValue vs
             evalPrimPrim p ps) of
@@ -51,6 +55,7 @@ evalPrimPrim pp ps
         (POr,  [PBool b1, PBool b2])    -> Just $ PBool (b1 || b2)
         (PAnd, [PBool b1, PBool b2])    -> Just $ PBool (b1 && b2)
         (PAdd, [PInt  i1, PInt  i2])    -> Just $ PInt  (i1 +  i2)
+
         (PEq,  [PInt  i1, PInt  i2])    -> Just $ PBool (i1 == i2)
         (PNeq, [PInt  i1, PInt  i2])    -> Just $ PBool (i1 /= i2)
         (PLt,  [PInt  i1, PInt  i2])    -> Just $ PBool (i1 <  i2)

### 2017-02-26 2017mergingmerges/code/src/Machine/Pretty.hs

@@ -70,7 +70,7 @@ instance Pretty Value where
   = case vv of
         VLit l          -> pretty l
         VPAP p vs       -> parens $ string "pap" <+> pretty p  <+> pretty vs
-
+        VTuple vs       -> parens $ string "tuple" <+> pretty vs
 
 instance Pretty Prim where
  pretty pp

### 2017-02-26 2017mergingmerges/code/src/Test/Fuse/DiamondAlt.hs

@@ -32,7 +32,58 @@ testFuseDiamondAltOk
 
                 pFused'         <- fmap fst $ stripLabels "F" pZipped
                 return pFused'
-   in   pFused
+
+        cvsInput
+                =  Map.fromList
+                [ (cIn1, map VInt [10, 11, 12, 13])
+                , (cIn2, map VInt [20, 21, 22, 23])
+                , (cIn3, map VInt [30, 31, 32, 33]) ]
+
+        cvsOutput
+                = Map.fromList
+                [ (cOut,   []) ]
+
+        (_inputs, _processes, actions)
+                = execute cvsInput cvsOutput [pFused] []
+
+   in   (pFused, actions)
+
+
+testFuseDiamondAltFront
+ = putStrLn $ show $ pretty
+ $ let  
+        cIn1     = Channel "cIn1"  TInt
+        cIn2     = Channel "cIn2"  TInt
+        cIn3     = Channel "cIn3"  TInt
+
+        cA1      = Channel "cA1"   TInt
+        cA2      = Channel "cA2"   TInt
+
+        pFused
+         = evalNew 
+         $ do   
+                pAlt1             <- mkAlt2_blk cIn1 cIn2 cA1
+                pAlt2             <- mkAlt2_blk cIn2 cIn3 cA2
+                let pAlt12      = fusePair' [pAlt1, pAlt2] pAlt1 pAlt2
+
+                pFused' <- fmap fst $ stripLabels "F" pAlt12
+                return pFused'
+
+        cvsInput
+                =  Map.fromList
+                [ (cIn1, map VInt [10, 11, 12, 13, 14, 15, 16])
+                , (cIn2, map VInt [20, 21, 22, 23, 24, 25, 26])
+                , (cIn3, map VInt [30, 31, 32, 33, 34, 35, 36]) ]
+
+        cvsOutput
+                = Map.fromList
+                [ (cA1,   [])
+                , (cA2,   []) ]
+
+        (_inputs, _processes, actions)
+                = execute cvsInput cvsOutput [pFused] []
+
+   in   (pFused, actions)
 
 
 testFuseDiamondAltFail

### 2017-02-26 2017mergingmerges/code/src/Main.hs
count combinators up to 10 - takes a long time

@@ -19,9 +19,21 @@ import Machine.Transform.StripLabels
 import Data.Maybe
 import Text.PrettyPrint.Leijen  hiding ((<$>))
 
-main 
- = print $ countsMax $ manySplitAB 6 combs1
+main = countBothUpTo 10
 
+countBothUpTo n = do
+  putStrLn "Splits"
+  countUpTo n manySplitAB
+  putStrLn ""
+  putStrLn "Pipes"
+  countUpTo n manyPipeAB
+
+countUpTo n f
+ = mapM_ countUpTo' [1..n]
+ where
+  countUpTo' num = do
+    putStr ("Combinators: " ++ show num ++ "\t\tStates: ")
+    print $ countsMax $ f num combs1
 
 combs1 
  =      [ combMapSucc

### 2017-02-26 2017mergingmerges/section/XX-Evaluation.tex
End S2 where it is, that is enough intuition before going onto the defs
Work on process definitions
add graph for combinator states. have computed up to 7 so far

@@
 \section{Evaluation}
+\label{s:Evaluation}
 
 \TODO{Discuss the performance of the system. Why this is a good idea. Give graphs that show the number of states is sensible when more processes are fused in.}
+
+\subsection{Optimisation}
+\label{s:Optimisation}
+After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
+
+\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}
+
+\begin{figure}
+
+\begin{minipage}[t]{0.4\textwidth}
+\begin{tikzpicture}
+\begin{axis}[
+	ylabel=Process states,
+	xlabel=Split combinators,
+  ymin=0, ymax=100,
+	enlargelimits=0.01,
+	ybar interval=0.7,
+  width=7.5cm, height=6.5cm,
+]
+\addplot coordinates {(1,9) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
+  % Last one doesn't show for some reason, so add one above the max 
+    (8,100) };
+\end{axis}
+\end{tikzpicture}
+\end{minipage}
+\begin{minipage}[t]{0.10\textwidth}
+\quad
+\end{minipage}
+\begin{minipage}[t]{0.4\textwidth}
+\begin{tikzpicture}
+\begin{axis}[
+%	ylabel=Process states,
+	xlabel=Pipelined combinators,
+  ymin=0, ymax=100,
+	enlargelimits=0.01,
+	ybar interval=0.7,
+  width=7.5cm, height=6.5cm,
+]
+\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)
+  % Last one doesn't show for some reason, so add one above the max 
+  (8,100) };
+\end{axis}
+\end{tikzpicture}
+\end{minipage}
+
+\caption{Maximum number of process states for all combinations of up to $n$ combinators.}
+\label{fig:bench:other}
+\end{figure}

### 2017-02-26 2017mergingmerges/code/src/Main.hs
code: change graph computation to only count up to 8

@@ -19,7 +19,7 @@ import Machine.Transform.StripLabels
 import Data.Maybe
 import Text.PrettyPrint.Leijen  hiding ((<$>))
 
-main = countBothUpTo 10
+main = countBothUpTo 8
 
 countBothUpTo n = do
   putStrLn "Splits"

### 2017-02-27 2017mergingmerges/section/XX-Evaluation.tex
update graph with +merge and -merge counts

@@
   width=7.5cm, height=6.5cm,
+	legend pos=north west,
+	%legend style={at={(0.01,0.99)},anchor=north west},
 ]
-\addplot coordinates {(1,9) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
+\addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
   % Last one doesn't show for some reason, so add one above the max 
     (8,100) };
+
+\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
+  % Last one doesn't show for some reason, so add one above the max 
+    (8,100) };
+
+\legend{All combinators, Without merge};
+
 \end{axis}
@@
   width=7.5cm, height=6.5cm,
+	legend pos=north west,
 ]
@@
   (8,100) };
+
+\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
+  % Last one doesn't show for some reason, so add one above the max 
+    (8,100) };
+
+\legend{All combinators, Without merge};
+
 \end{axis}
@@
 
-\caption{Maximum number of process states for all combinations of up to $n$ combinators.}
+\caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
 \label{fig:bench:other}

### 2017-02-27 2017mergingmerges/code/src/Test/Comb.hs
code: implement more flexible split/pipe combiners that allow multiple inputs

@@ -53,19 +53,52 @@ fusePipeAB    comb1 comb2
  | otherwise
  = []
 
+-------------------------------------------------------------------------------
+-- | Fuse first and second combinators into a semi-pipeline, if possible.
+--    Requires the first combinator to have a single output,
+--    and allows the second combinator to have many inputs.
+--    The first input of the second combinator is connected, but other inputs are left.
+fusePipeMultipleInputs :: Comb -> Comb -> [Comb]
+fusePipeMultipleInputs    comb1 comb2
+ | [tOut1]      <- combOutputs comb1
+ , tIn2 : tIns2 <- combInputs  comb2
+ , tOut1 == tIn2
+ = [ Comb (combInputs  comb1 ++ tIns2) 
+          (combOutputs comb2)
+        $  \csIns csOuts2 
+        -> do   let csIns1 = take (length $ combInputs comb1) csIns
+                let csIns2 = drop (length $ combInputs comb1) csIns
+                cInternal <- newChannel "Internal" tOut1
+                proc1     <- combMk comb1 csIns1     [cInternal]
+                proc2     <- combMk comb2 (cInternal : csIns2) csOuts2
+                case fusePair proc1 proc2 of
+                 Left err       -> error (show err)
+                 Right proc'    
+                  -> return $ tagHow "pipe" proc'
+   ]
+
+ | otherwise
+ = []
+
+
+-- | Given a set of base combinators and a combining function,
+--   create combinators using all the possible ways of combining
+--   the given number of instances.
+manyCombineWith :: (Comb -> Comb -> [Comb]) -> Int -> [Comb] -> [Comb]
+manyCombineWith f 0 cs = []
+manyCombineWith f 1 cs = cs
+manyCombineWith f 2 cs 
+ =  concatMap (\c -> concatMap (f c) cs) cs
+
+manyCombineWith f n cs
+ =  concatMap (\c -> concatMap (f c) (manyCombineWith f (n - 1) cs)) cs
+ ++ concatMap (\c -> concatMap (f c) cs) (manyCombineWith f (n - 1) cs)
 
 -- | Given a set of base combinators,
 --   create combinators using all the possible ways of pipelining
 --   the given number of instances.
 manyPipeAB :: Int -> [Comb] -> [Comb]
-manyPipeAB 0 cs = []
-manyPipeAB 1 cs = cs
[…]

### 2017-02-27 2017mergingmerges/code/src/Main.hs
code: compute counts for graph

@@ -19,52 +19,65 @@ import Machine.Transform.StripLabels
 import Data.Maybe
 import Text.PrettyPrint.Leijen  hiding ((<$>))
 
-main = countBothUpTo 8
+main = countUpTo 7
 
-countBothUpTo n = do
-  putStrLn "Splits"
-  countUpTo n manySplitAB
-  putStrLn ""
-  putStrLn "Pipes"
-  countUpTo n manyPipeAB
-
-countUpTo n f
+countUpTo n
  = mapM_ countUpTo' [1..n]
  where
   countUpTo' num = do
-    putStr ("Combinators: " ++ show num ++ "\t\tStates: ")
-    print $ countsMax $ f num combs1
+    let countWith str f combs = do
+          putStr ("\t" ++ str)
+          print $ countsMax $ f num combs
+    putStrLn ("Combinators: " ++ show num)
+    countWith "Splits all:       " (manyCombineWith fuseSplitFirstManyInputs) combsAll
+    countWith "Splits no merge:  " (manyCombineWith fuseSplitFirstManyInputs) combsNoMerge
+    countWith "SplitMany merge:  " (manyCombineWith fuseSplitManyInputs) combsOnlyMerge
+    countWith "Pipes all:        " manyPipeAB combsAll
+    countWith "Pipes no merge:   " manyPipeAB combsNoMerge
+    countWith "PipeMany merge:   " (manyCombineWith fusePipeMultipleInputs) combsOnlyMerge
 
-combs1 
+
+combsAll
  =      [ combMapSucc
         , combFilterPos
         , combScanAdd
-        , combGroup 
+        , combGroup
         , combMerge ]
+
+combsNoMerge
+ =      [ combMapSucc
+        , combFilterPos
+        , combScanAdd
+        , combGroup ]
+
+combsOnlyMerge
+ =      [ combMerge ]
+
+
 
 counts cs
         = putStrLn
         $ unlines
-        $ map (show    . statOfProcess) 
-        $ map (evalNew . mkComb) 
+        $ map (show    . statOfProcess)
+        $ map (evalNew . mkComb)
         $ cs
 
 
 countsMax cs
         = maximum
         $ map (statInstrs . statOfProcess)
-        $ map (evalNew . mkComb) 
+        $ map (evalNew . mkComb)
         $ cs
 
 countPipes n
-        = counts 
-        $ manyPipeAB n combs1
+        = counts
+        $ manyPipeAB n combsAll
 
[…]

### 2017-02-27 2017mergingmerges/section/XX-Evaluation.tex
graph wibble

@@
 
+% I thought these two looked nicer separately, but let me know if you want them in a single graph
 \begin{minipage}[t]{0.4\textwidth}
@@
 	legend pos=north west,
-	%legend style={at={(0.01,0.99)},anchor=north west},
 ]
 \addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
-  % Last one doesn't show for some reason, so add one above the max 
-    (8,100) };
+  % Last bar doesn't show for some reason, so need to add a dummy value for the next one
+    (8,1) };
 
-\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
-  % Last one doesn't show for some reason, so add one above the max 
-    (8,100) };
+\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
 
-\legend{All combinators, Without merge};
+\legend{With merge, Without merge};
 
@@
 \begin{axis}[
+% Hide the label on the second graph
 %	ylabel=Process states,
@@
 ]
-\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)
-  % Last one doesn't show for some reason, so add one above the max 
-  (8,100) };
+\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
 
-\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)
-  % Last one doesn't show for some reason, so add one above the max 
-    (8,100) };
+\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
 
-\legend{All combinators, Without merge};
+\legend{With merge, Without merge};
 
@@
 \caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
-\label{fig:bench:other}
+\label{fig:bench:outputsize}
 \end{figure}
+
+\begin{figure}
+
+\begin{tikzpicture}
+\begin{axis}[
+	ylabel=Process states,
+	xlabel=Number of merges,
+%  ymode=log,
+  ymin=0, ymax=1500,
+	enlargelimits=0.01,
+	ybar interval=0.7,
+  width=7.5cm, height=6.5cm,
+	legend pos=north west,
+]
+% These are the values for splitting.
+% They are smaller than the 'chaining', but look much nicer on the linear graph.
+\addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
+  % (8,2880)
+  (8,1)
+  };
+
+% These are the values for chaining
+% \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };
+
+\end{axis}
+\end{tikzpicture}
+
+\caption{Exponential blowup occurs when splitting or chaining merges together.}
+\label{fig:bench:exponential}
+\end{figure}

### 2017-02-27 2017mergingmerges/code/src/Main.hs
code: rename fusePipeMultipleInputs to fuseChainManyInputs

@@ -34,7 +34,7 @@ countUpTo n
     countWith "SplitMany merge:  " (manyCombineWith fuseSplitManyInputs) combsOnlyMerge
     countWith "Pipes all:        " manyPipeAB combsAll
     countWith "Pipes no merge:   " manyPipeAB combsNoMerge
-    countWith "PipeMany merge:   " (manyCombineWith fusePipeMultipleInputs) combsOnlyMerge
+    countWith "ChainMany merge:  " (manyCombineWith fuseChainManyInputs) combsOnlyMerge
 
 
 combsAll

### 2017-02-27 2017mergingmerges/code/src/Test/Comb.hs

@@ -54,12 +54,12 @@ fusePipeAB    comb1 comb2
  = []
 
 -------------------------------------------------------------------------------
--- | Fuse first and second combinators into a semi-pipeline, if possible.
---    Requires the first combinator to have a single output,
---    and allows the second combinator to have many inputs.
---    The first input of the second combinator is connected, but other inputs are left.
-fusePipeMultipleInputs :: Comb -> Comb -> [Comb]
-fusePipeMultipleInputs    comb1 comb2
+-- | Fuse first and second combinators into a "chain": similar to a pipeline,
+--    but allow combinators with multiple inputs.
+--    The first combinator's output is connected to the first input of the second,
+--    and any other inputs are lifted up as inputs to the resulting combinator.
+fuseChainManyInputs :: Comb -> Comb -> [Comb]
+fuseChainManyInputs    comb1 comb2
  | [tOut1]      <- combOutputs comb1
  , tIn2 : tIns2 <- combInputs  comb2
  , tOut1 == tIn2

### 2017-02-27 2017mergingmerges/section/XX-Fusion.tex
Shifting figs around
Work on alt2 example to describe diff eval orders
remove fuseNetwork function

@@
 %!TEX root = ../Main.tex
+
+\clearpage{}
 \section{Fusion}
 \label{s:Fusion}
-
-\input{figures/Fusion-types.tex}
 
@@
 
+\input{figures/Fusion-types.tex}
+
 \input{figures/Fusion-fusePair.tex}
 
-Fig.~\ref{fig:Fusion:Def:Top} contains the definitions of top-level fusion functions to fuse a network and a pair of processes.
-Function \ti{fuseNetwork} takes a process network and repeatedly fuses pairs together until all processes in the network have been fused into a single process.
-The order processes are fused can affect whether fusion succeeds or fails, so all permutations are tried.
-We address this in more detail in~\S\ref{s:FusionOrder}.
-
+Fig.~\ref{fig:Fusion:Def:Top} defines the top-level function that fuses a pair of processes.
 Function \ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
@@
 Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
-\ti{channels} computes the $\ChanType_2$ map for a pair of processes.
-\ti{connected} checks whether two processes have any shared inputs or outputs - basically whether it is worth fusing them.
-\ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
-\ti{swaplabels} flips the order of the compound labels in an instruction.
+Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
+Function \ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
+Function \ti{swaplabels} flips the order of the compound labels in an instruction.

### 2017-02-27 2017mergingmerges/section/XX-Proof.tex
mainly prose
remove fuseNetwork function

@@
 
-We know of no `realistic' examples where combinators have multiple evaluation orders, so believe this is not an issue in practice.
+We know of no realistic examples where combinators have multiple evaluation orders, so we believe this is not an issue in practice.
 
@@
 Instead we sequentially evaluate both processes separately with the same input values and outputs.
-Finally, we have not implemented \ti{fuseNetwork} for fusing multiple processes at a time, only \ti{fusePair} for fusing pairs of processes.
 

### 2017-02-27 2017mergingmerges/section/XX-Fusion.tex
PreferJump heuristic

@@
 
-Clause (DeferPull1) applies when both processes are able to take a step.
-If both processes can execute, \ti{tryStepPair} uses a simple priority heuristic of deferring pulls: this is because pulls may block, while other actions are more likely to produce immediate results.
-In this case the instruction for $p$ is \emph{not} pulling, so it can be executed.
+Clauses (PreferJump1) and (PreferJump2) apply heuristics to prioritise which process should take a step.
+If one of the processes can perform a jump, it is given priority.
+This is an attempt to perform multiple jumps together, as many consecutive jumps could later be contracted into a single jump.
+Because clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in the swapped order, it returns an instruction with the output labels swapped. We use $\ti{swaplabels}$ to swap them back.
 
-Clause (DeferPull2) is much the same as (DeferPull1) except this time the instruction for $q$ is not a pull.
-Because we are calling $\ti{tryStep}$ with the instruction from process $q$, the output labels it generates are pairs of $(q,p)$ rather than $(p,q)$.
-We use $\ti{swaplabels}$ to convert them back to $(p,q)$ and keep them consistent with the other labels.
+Clauses (DeferPull1) and (DeferPull2) perform similar heuristics, deferring pulls as much as possible.
+This is because pulls may block, while other actions are more likely to produce immediate results.
 
@@
 If both are pulls, we make the arbitrary choice to execute $p$ with (Run1).
-For clause (Run2) we again need to swap the labels as with (DeferPull2).
 

### 2017-02-27 2017mergingmerges/section/03-Definition.tex
S3.2: talk about non-deterministic evaluation order, and lead into fusion

@@
 
-\subsection{Evaluation Order}
+% -----------------------------------------------------------------------------
+\subsection{Non-deterministic Evaluation Order}
 
-Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
+The evaluation rules of Fig.
+\ref{fig:Process:Eval:Feed} are non-deterministic in several ways.
+Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the group.
+More importantly, (ProcessPush) allows any process to perform a push action at any time, provided the other processes in the group are ready to accept the pushed value.
+Rule (FeedExternal) also allows new values pushed into an input channel from the environment, provided all processes that use that channel are ready to accept the value.
+
+In our system, allowing the evaluation order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
+Consider the following example, where the @alt2@ operator pulls two elements form its first input stream, then two from the second, before pushing all four to its output.
 \begin{code}
   alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
-  alternates sIn1 sIn2 sIn3
-   = let  sA1     = alt2 sIn1 sIn2
-          sA2     = alt2 sIn2 sIn3
-     in   zip sA1 sA2
+  alternates sInA sInB sInC
+   = let  s1   = alt2 sInA sInB
+          s2   = alt2 sInB sInC
+          sOut = zip s1 s2
+     in   sOut
 \end{code}
 
-Given the inputs @sIn1@ = $[a1, a2]$, @sIn2@ = $[b1, b2]$ and @sIn3@ = $[c1, c2]$ the \\ output will be $[ (a1, b1), (a2, b2), (b1, c1), (b2, c2) ]$
+Note that the middle stream @sIn2@ is shared, and the result streams from both @alt2@ operators are zipped into tuples.
+Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.
 
-\TODO{Explain why this needs a specific evaluation order, and some do not work. Needs to fit on this page. Lead into the fusion order talk later.}
+Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time.
+However, the downstream @zip@ process needs to input one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams.
+This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively.
+They can still push two values at a time to their own outputs, but the downstream @zip@ process needs receive one from each process alternately.
+Here is a table of intermediate values to help make the explanation clearer:
 
-Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately.
+\begin{code}
+    sInA = [a1, a2, a3, a4, a5 ...]
+    sInB = [b1, b2, b3, b4, b5 ...]
+    sInC = [c1, c2, c3, c4, c5 ...]
+
+    s1   = alt2 sInA sInB 
+         = [a1, a2, b1, b2, a3, a4, b3, b4 ...]
+
+    s2   = alt2 sInB sInC
+         = [b1, b2, c1, c2, b3, b4, c3, c4 ...]
+
+    sOut = zip s1 s2
+         = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
+\end{code}
+
+Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on.
+The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it.
+While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@.
+The evaluation order of the overall process group is constrained by communication patterns of processes in that group.
+
+As we cannot encode all possible evaluation orderings into the definition of the processes themselves, we have defined the evaluation rules to admit many possible orderings.
+In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, evaluation order would be discovered dynamically at runtime.
+In contrast, the role of our fusion transform is to determine one of these working evaluation orders statically.
+In the fused result process, the instructions will be scheduled so that they run in one of the orders that we would have gotten if the process group was executed dynamically.
+In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.
+
+% Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately.
 At evaluation time we need the results pushed to sA1 and sA2 in the sA1 sA2 sA1 sA2 order, not sA1 sA1 sA2 sA2.

### 2017-02-27 2017mergingmerges/section/03-Definition.tex
Rename XX-Fusion.tex to 04-Fusion.tex

@@
 In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, evaluation order would be discovered dynamically at runtime.
-In contrast, the role of our fusion transform is to determine one of these working evaluation orders statically.
-In the fused result process, the instructions will be scheduled so that they run in one of the orders that we would have gotten if the process group was executed dynamically.
+In contrast, the role of our fusion transform is to construct one of these working evaluation orders statically.
+In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the process group was executed dynamically.
 In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

### 2017-02-27 2017mergingmerges/section/04-Fusion.tex
Rename XX-Fusion.tex to 04-Fusion.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate

--- 2017-02-27 2017mergingmerges/section/XX-Fusion.tex
@@
 
-\clearpage{}
+% -----------------------------------------------------------------------------
 \section{Fusion}
@@
 
-The core fusion algorithm works by constructing a static schedule for a pair of processes, which is repeated over the whole network until only one process remains.
-The static schedule mirrors the dynamic evaluation by statically performing injection and shaking on the two processes.
+The core fusion algorithm works by constructing a static schedule for a single pair of processes. To fuse a whole process group we fuse pairs until only one process remains.
+
 Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.

### 2017-02-27 2017mergingmerges/section/XX-Evaluation.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate

@@
 
-\section{Evaluation}
+% -----------------------------------------------------------------------------
+\section{Fusion in Practice}
 \label{s:Evaluation}
+
+Stream fusion is ultimately performed for practical reasons.
+We want the fused result program to run faster than unfused source program.
+Fused result programs like the one in Fig.
+\ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell (maybe using Template Haskell), C or LLVM abstract assembly.
+The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.
+
+% -----------------------------------------------------------------------------
+\subsection{Fusability}
+\label{s:FusionOrder}
+When we fuse a pair of processes we commit to a particular interleaving of instructions from each process.
+When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process.
+Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
+\begin{code}
+  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
+  alternates sInA sInB sInC
+   = let  s1   = alt2 sInA sInB
+          s2   = alt2 sInB sInC
+          sOut = zip s1 s2
+     in   sOut
+\end{code}
+
+In our current system, if fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails.
+This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
+As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.
+
+Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock.
+Statically, when we try to fuse the result process with the downstrem @zip@ process then the deadlock is discovered and fusion fails.
+Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}.
+The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.
+
+On the upside, fusion failure is easy to detect.
+It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
+The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
+The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
+This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and may other systems based on general purpose program transformations.
+In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
+spelunking: Exploration of caves, especially as a hobby.
+Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.
+
+In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network.
+For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds.
+The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock.
+Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
+When the dataflow network has a single output stream then we use the method of starting from the process closes to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
+This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels where need to be synchronized.
+In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.
+
+% We discuss other possible approaches in \S\ref{s:Future:FusionOrder}.
+
+% When the two @alt2@ processes are fused we end up with a sequential result process.
+Trying to then fuse that result into the downstream @zip@ process fails.
+Alternately, if we were to try to concurrently evaluate the fused result process with the downstream @zip@ process would deadlock because we would reach a state where neither process could advance.
+To state this again: fusion fails when concurrent evaluation of the two source processes would deadlock.
+
+% The main fusion algorithm here works on pairs of processes.
+% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
+The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
+Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process.
+In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes.
+The simplest example of this has two input streams, a function applied to both, then zipped together.
+
+% -----------------------------------------------------------------------------
+\subsection{Result Size}
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Optimisation}

### 2017-02-27 2017mergingmerges/section/XX-FutureWork.tex
Wibble pagination to see how much space we have left
Shift discussion of fusion order to Fusion In Practice section, and elaborate

@@
 %!TEX root = ../Main.tex
+
+% -----------------------------------------------------------------------------
 \section{Future work}
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Finite streams}
@@
 
-\subsection{Non-determinism and fusion order}
-\label{s:FusionOrder}
-
-The main fusion algorithm here works on pairs of processes.
-When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
+% -----------------------------------------------------------------------------
+% \subsection{Non-determinism and fusion order}
+% \label{s:FusionOrder}
+% The main fusion algorithm here works on pairs of processes.
+% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
 The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
@@
 The simplest example of this has two input streams, a function applied to both, then zipped together.
-
-\begin{code}
-zipMap as bs =
-  let as' = map (+1) as
-      bs' = map (+1) bs
-      abs = zip as' bs'
-  in  abs
-\end{code}
-
-There are three combinators here, so after converting each combinator to its process there are three orders we can fuse.
+% \begin{code}
+% zipMap as bs =
+%   let as' = map (+1) as
+%       bs' = map (+1) bs
+%       abs = zip as' bs'
+%   in  abs
+% \end{code}
+% There are three combinators here, so after converting each combinator to its process there are three orders we can fuse.
 The two main options are to fuse the two maps together and then add the zip, or to fuse the zip with one of the maps, then add the other map.
@@
 
-Our current solution to this is to try all permutations of fusing processes and use the first one that succeeds.
+% \subsection{Non-determinism and fusion order}
+% \label{s:Future:FusionOrder}
+% Our current solution to this is to try all permutations of fusing processes and use the first one that succeeds.
 A more principled solution may be to allow non-determinism in a single process by adding a non-deterministic choice instruction.

### 2017-02-27 2017mergingmerges/section/05-Evaluation.tex
Rename file to 05-Evaluation.tex

--- 2017-02-27 2017mergingmerges/section/XX-Evaluation.tex
@@
 	ybar interval=0.7,
-  width=7.5cm, height=6.5cm,
+  width=7.5cm, height=4.5cm,
 	legend pos=north west,

### 2017-02-27 2017mergingmerges/section/04-Fusion.tex
blocks -> instructions

@@
 Function \ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
-The blocks are constructed by computing a fixpoint, starting at the initial labels of each process with empty input states.
-Each block is computed with \ti{tryStepPair}, which statically chooses one of the two processes to execute, and any reachable blocks are added recursively until fixpoint is reached.
+The instructions are constructed by computing a fixpoint, starting at the initial labels of each process with empty input states.
+Each instruction is computed with \ti{tryStepPair}, which statically chooses one of the two processes to execute, and any reachable instructions are added recursively until fixpoint is reached.
 This implementation assumes the heap variables are distinct, which can easily be ensured by renaming.
@@
 Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
-Function \ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the blocks map.
+Function \ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the instructions map.
 Function \ti{swaplabels} flips the order of the compound labels in an instruction.

### 2017-02-27 2017mergingmerges/section/05-Evaluation.tex
Put pipelined combinators graph first
S5.2 leave talk of result size for now

@@
 When the dataflow network has a single output stream then we use the method of starting from the process closes to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
-This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels where need to be synchronized.
+This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
 In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.
@@
 
-\TODO{Discuss the performance of the system. Why this is a good idea. Give graphs that show the number of states is sensible when more processes are fused in.}
+As with fusion transformation system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
+The left of Fig.\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
+To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes.
+The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start.
+The right of Fig.\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
+In both cases the number of states in the result process grows linearly with the number of processes.
+In all combinations, with up to 7 processes there are less than 100 states in the result process.
+
+The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes.
+This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
+
+On the other hand, Fig.\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number if input processes.
+The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream.
+The output of each merge process is the input of the next, forming a chain of merges.
+In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
+
+When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product.
+During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition.
+The use of the finite map ensures that identical states are always combined, but genuinly different states always make it into the result.
+
+In the worst case fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
+If we assume the two processes have about the same number of states then this is O($n^2$).
+Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.
+
+In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction.
+During fusion, the states for the consuming process are inlined at each occurrence of @push@.
+These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.
 
@@
 
+% -----------------------------------------------------------------------------
 \begin{figure}
-
-% I thought these two looked nicer separately, but let me know if you want them in a single graph
+\begin{minipage}[t]{0.10\textwidth}
+\end{minipage}
 \begin{minipage}[t]{0.4\textwidth}
@@
 \begin{axis}[
+% Hide the label on the second graph
 	ylabel=Process states,
-	xlabel=Split combinators,
+	xlabel=Pipelined combinators,
+  ymin=0, ymax=100,
+	enlargelimits=0.01,
+	ybar interval=0.7,
+  width=7.5cm, height=4.5cm,
+	legend pos=north west,
+]
+\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
+\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
+\legend{With merge, Without merge};
+\end{axis}
+\end{tikzpicture}
+\end{minipage}
+\quad
+\quad
+\quad
+\begin{minipage}[t]{0.4\textwidth}
+\begin{tikzpicture}
+\begin{axis}[
+%       ylabel=Process states,
+        xlabel=Parallel combinators,
   ymin=0, ymax=100,
@@
 \legend{With merge, Without merge};
-
-\end{axis}
-\end{tikzpicture}
-\end{minipage}
-\begin{minipage}[t]{0.10\textwidth}
-\quad
-\end{minipage}
-\begin{minipage}[t]{0.4\textwidth}
-\begin{tikzpicture}
-\begin{axis}[
-% Hide the label on the second graph
-%	ylabel=Process states,
-	xlabel=Pipelined combinators,
-  ymin=0, ymax=100,
-	enlargelimits=0.01,
-	ybar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
-
-\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
-
-\legend{With merge, Without merge};
-
 \end{axis}
@@
 
+% -----------------------------------------------------------------------------
 \begin{figure}

### 2017-02-27 2017mergingmerges/section/05-Evaluation.tex
Start opt. section

@@
 % -----------------------------------------------------------------------------
-\subsection{Optimisation}
+\subsection{Optimisation with Drop Instructions}
 \label{s:Optimisation}
-After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.
+After we have fused two processes together, it may be possible to simplify the result before fusing in a third.
+Consider the result of fusing @group@ and @merge@ which we saw back in Fig.\ref{fig:Process:Fused}.
+At labels @F1@ and @F2@ are @jump@ instructions, where the first passes control to the second.
+The update expressions attached to these instructions are also non-interferring, which means we can safely combine these consecutive @jump@ instructions into a single @jump@ to @F4@.
+There are also @jump@ instructions like the one at @F5@ which have no associated update expressions, and thus can be eliminated completely.
+We can also perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.
 
-\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}
+Minimising the number of states in intermediate processes has a follow-on effect in that the final fused result has a lower number of states.
+Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusability of the overall process network will not be affected.
+
+Another useful optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
+In Fig.\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
+Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable.
+Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@.
+When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.
+
+When the two processes to be fused are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
+In this case we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
 

### 2017-02-27 2017mergingmerges/section/05-Evaluation.tex
Talk about drop instructions

@@
 % -----------------------------------------------------------------------------
-\subsection{Optimisation with Drop Instructions}
+\subsection{Optimisation and Drop Instructions}
 \label{s:Optimisation}
@@
 The update expressions attached to these instructions are also non-interferring, which means we can safely combine these consecutive @jump@ instructions into a single @jump@ to @F4@.
+In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code.
+The (PreferJump) clauses of Fig.\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
+
 There are also @jump@ instructions like the one at @F5@ which have no associated update expressions, and thus can be eliminated completely.
-We can also perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.
+Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.
 
-Minimising the number of states in intermediate processes has a follow-on effect in that the final fused result has a lower number of states.
+Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states.
 Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusability of the overall process network will not be affected.
 
-Another useful optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
+Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
 In Fig.\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
@@
 
-When the two processes to be fused are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
-In this case we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
+When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
+This is the case in Fig.\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
+In the result, the interleaved instructions from both source processes then share the same heap variable.
+
+To increase the chance that we can perform this above copy-propagation, we need both processess to want to pull from the same stream at the same time.
+In the definition of a particular process, moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.
+In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream.
 

### 2017-02-27 2017mergingmerges/section/XX-FutureWork.tex
Talk about Finite Streams last as it ends on a happier note

@@
 
-We now discuss some of the shortcomings of the system, and extensions and future work to ameliorate this.
+% We now discuss some of the shortcomings of the system, and extensions and future work to ameliorate this.
+
+\subsection{Case analysis}
+\label{s:FullyAbstractCase}
+
+The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
+This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
+
+For example, suppose two processes have the same case condition (@x > 0@).
+The fusion algorithm will generate all four possibilities, including the contradictory ones such as where the first process has (@x > 0 = true@) and the second has (@x > 0 = false@).
+If any require an unbounded buffer, the fusion algorithm will fail.
+
+A possible extension is to somehow cull these contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
+It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit some kind of failure instruction rather than failing to produce a process.
+A separate postprocessing step could then perform analyses and remove statically unreachable process states.
+After postprocessing, if any failure instructions are reachable, fusion fails as before.
 
@@
 After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow.
-
-We have implemented an early prototype that supports finite streams, but further work is required to prove it correct.
+We have implemented an early prototype that supports finite streams, our mechanized formalization does not cover it.
 
@@
 By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
-Case analysis is still a useful extension and is described next.
 
@@
 % These modifications allow our system to fuse finite streams as well as infinite.
-
-\subsection{Case analysis}
-\label{s:FullyAbstractCase}
-
-The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
-This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
-
-For example, suppose two processes have the same case condition (@x > 0@).
-The fusion algorithm will generate all four possibilities, including the contradictory ones such as where the first process has (@x > 0 = true@) and the second has (@x > 0 = false@).
-If any of these possibilities require an unbounded buffer, the fusion algorithm will fail.
-
-A possible extension is to somehow cull these contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
-It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit some kind of failure instruction rather than failing to produce a process.
-A separate postprocessing step could then perform analyses and remove statically unreachable process states.
-After postprocessing, if any failure instructions are reachable, the fusion process would fail as before.
 

### 2017-02-27 2017mergingmerges/section/01-Introduction.tex
Kill commented out stuff in S1 S2 that we've already handled
typos S1

@@
 
-% Can we perform both of these tasks at once, without needing to read through the stream data multiple times, and without needing unbounded buffering?
-
 In this implementation the @group@ operator filters out consecutive duplicates, while @merge@ combines two sorted streams so that the output remains sorted.
@@
 Each operator could be given this interface, so that the next value from each result stream is computed on demand.
-This is approach is commonly taken with implementations of physical operators in database systems.
+This approach is commonly taken with implementations of physical operators in database systems.
 However, the `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
 Suppose a consumer pulls many elements from the @sUnique@ output stream.
-In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering a arbitrary number of them.
+In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them.
 It needs to buffer these elements because they are also needed to perform the @merge@ operation.
@@
 
-% Leave this to related work. We've already mentioned a canonical pull-based system.
-% Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.
-
 Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example either, because they do not support operators with multiple inputs.
@@
 More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.
-
-% The mechanism that combines the implementations of both operators, to yield efficient imperative code also depends on the general purpose compiler optimisations implemented by GHC, and it can be difficult to tell if these have ``worked'' without inspecting the intermediate representations of the compiler.
 
@@
 We discuss this extension in \S\ref{s:Finite}.
-
-% TODO: We can't make the appendix a contribution because the reviewers are not required to read appendices.
-% \item and show our processes are general enough for many combinators, including segmented operations (\S\ref{s:Combinators}).
-
-% \ben{Add a few more sentences on related work.
-Explain how this work extends the old flow fusion paper.
-It is not short-cut fusion like Oleg's recent work.
-We are not in the same space as Fortran style array fusion transformations like polyhedral}
-
-% BL: describe this later.
-% Furthermore, the data-flow fusion system of~\cite{lippmeier2013data} only deals with a fixed set of baked-in combinators.
-
-% BL: Shift the detailed description into a later section.
-% The example above has three combinators, so the process network has three processes.
-% The two @writeFile@s outputs are treated as sinks that values can be pushed to at any time, and are not converted to processes.
-% During code generation, any output values from the @uniques@ and @union@ streams are sent to the corresponding @writeFile@ sink, but we do not address code generation in this paper.
-
-% The process for @uniques@ is defined by the @group@ combinator, and can be thought of as an imperative loop: first it reads from its input stream @file1@ and stores that in a local variable.
-% It also keeps track of the last pulled value, and compares that against the newly read value.
-% If they are different, it pushes the new value to its output stream @uniques@.
-% In either case, it updates the last pulled value and loops back to the start to pull from @file1@ again.
-
-% The process for @merged@ is defined by the @merge@ combinator, which starts by reading from both @file1@ and @file2@ and storing these in local variables.
-% It then compares its two values to see which is the smaller.
-% If the value from @file1@ is smaller, it pushes that value and pulls a new value from @file1@, otherwise it pushes the value from @file2@ and pulls from @file2@.
-% This is performed in a loop.
-
-% We fuse these two processes by interleaving the two such that the shared input @file1@ is only pulled from when both processes agree.
-% The new process pulls from @file1@, which is copied to the variables for both processes.
-% The @uniques@ process now has all it needs to execute, so it checks the value against the last pulled value, pushes if necessary, and goes back to try to pull from @file1@ again.
-% At this stage the @merged@ process still has a value from @file1@ that it has pulled but not used, so @uniques@ cannot pull from @file1@ again.
-% We now let @merged@ run, pulling from @file2@ and checking which is smaller.
-% If the value from @file1@ is smaller, the value is emitted and @merged@ wishes to pull a new value from @file1@.
-% Both processes now agree on pulling from @file1@ again, so the new value is pulled and @uniques@ can run again.
-% Otherwise if the value from @file1@ is not smaller, the value from @file2@ is emitted and @merged@ pulls from @file2@ with no coordination required.
-
-% If we wish to ensure that each value is only read from the file once, we must coordinate between the two use sites: when @uniques@ requires a new value it must ensure that @union@ is ready to receive a new value, and vice versa.
-Note that we cannot just execute @uniques@ while storing the read values in a buffer, as this may require more memory than is available.
-% In order to fuse this example, we require both pull \emph{and} push streams.
-% The input streams must be pull streams since the order values are required is determined by the @merge@ combinator.
-% For the same reason, the outputs sent to each @writeFile@ must be push streams.
-
-% Fusion for array programs is important for removing intermediate arrays, reducing memory traffic and reducing allocations.
-% However, when dealing with data too large to fit in memory such as tables on disk, removing intermediate arrays becomes essential rather than just desirable.
-% Attempting to create an intermediate array of such amounts of data would lead to thrashing and swapping to disk, or perhaps even running out of swap.
-% For these situations, some sort of assurance of total fusion is required: either the program can be fused with no intermediate arrays or unbounded buffers, or it will not compile at all.
-
-% Fusion eliminates intermediate array buffers and converts pipelines of array combinators into low-level iteration based loop code. Different fusion systems can handle
-
-% When comparing fusion systems, three important criteria to consider are: whether the system supports splits, where a stream is used multiple times; whether it supports joins, where a combinator has multiple inputs; and whether arbitrary combinators such as @merge@ and segmented appends can be encoded.
-Existing fusion systems support one or two of these, but not three.
-We present a fusion system based on process calculus that supports all three: splits, joins and arbitrary combinators.
-
-% Our system has been formalised in Coq where we have proved soundness of the fusion algorithm. It is expressive enough to encode a wide range of combinators including operations on segmented arrays.
-
-% \amos{``Arbitrary combinators'' is not quite true. How can we distinguish combinators we support from 2013 data flow fusion paper? Perhaps by mentioning value-dependent input / access patterns.}
-
-% Leave this to the description of the algorithm, not the abstract.
-% We encode each combinator as a separate process with any number of input and output channels.
-Each process is sequential but multiple processes can be executed concurrently.
-We give a concurrent execution semantics for multiple processes, but these are used only as a specification for how the fused program must behave.
-The fused program itself is sequential and can easily be converted to simple imperative code.
-
-% Our fusion algorithm takes two concurrently executable processes and creates a sequential interleaving of the two such that they execute with no unbounded buffers.
-% If fusion would require unbounded buffers (or the fusion algorithm wrongly infers that it would) then fusion fails.
-% If fusion fails, the user can be presented with an error message telling them which combinators could not be fused.
-% For scenarios where fusion is required, this is a great advantage over fragile shortcut fusion systems.
-
-% BL: leave the apologies to the conclusion.
-% The version presented here deals with infinite streams, and we informally describe the extensions required to support finite streams.
-
-% BL: leave this to the main intro.
-% optimising high-level array and streaming computations, as it reduces memory traffic and intermediate arrays.
-The benefits of removing intermediate arrays are even more important as data sizes approach the size of memory.

### 2017-02-28 2017mergingmerges/section/02-Processes.tex
update fig 5 (fused group and merge) to fit implementation
Start opt. section
notes
fig 5: a touch of colour on the running process

@@
 The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
+
+%%% AR: Should we explain intuition behind how process works and @f@, @l@, @v@ before detailed instructions?
 
@@
 \subsubsection{Fusing Pulls}
+\label{s:Fusion:FusingPulls}
 
@@
 
+%%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
 \begin{figure}
@@
 
+\definecolor{runningc}{HTML}{307020}
+\newcommand\runningcolour[1]{\bf{\textcolor{runningc}{#1}}}
+
 \newcommand\annot[5]{
@@
 }
+\newcommand\annotL[5]{\annot{\runningcolour{#1}}{#2}{#3}{#4}{#5}}
+\newcommand\annotR[5]{\annot{#1}{#2}{\runningcolour{#3}}{#4}{#5}}
+\newcommand\annotLR[5]{\annot{\runningcolour{#1}}{#2}{\runningcolour{#3}}{#4}{#5}}
 \newcommand\icase[7]{
@@
          \= @F17@ \= @[ ]     @ \= @F17@ \= @[ ]@
-@   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
+@   @ \= \tiny @   @F17 \= \tiny = ((\runningcolour{A0,} \= \tiny \{sIn1 = pending\}), \= \tiny (\runningcolour{B0,} \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
 \> @{@ \instr{F0}{pull sIn1 b1}{F1}{}
@@
 \instr{F1}{jump}{F2}{v  = b1}
-      {\annot{A0}{pending}{B0}{pending}{none}}
+      {\annotL{A0}{pending}{B0}{pending}{none}}
 \> @,@
 \instr{F2}{jump}{F3}{x1 = b1}
-      {\annot{A1}{have}{B0}{pending}{none}}
+      {\annotR{A1}{have}{B0}{pending}{none}}
 \> @,@
 \icase{F3}{case (f || (l /= v))}{F4}{}{F5}{}
-      {\annot{A1}{have}{B1}{have}{none}}
+      {\annotL{A1}{have}{B1}{have}{none}}
 
@@
 \instr{F4}{push sOut1 v}{F5}{l = v, f = F}
-      {\annot{A2}{have}{B1}{have}{none}}
+      {\annotL{A2}{have}{B1}{have}{none}}
 
@@
 \instr{F5}{jump}{F6}{}
-      {\annot{A3}{have}{B1}{have}{none}}
+      {\annotL{A3}{have}{B1}{have}{none}}
 
@@
 \instr{F6}{pull sIn2 x2}{F7}{}
-      {\annot{A0}{none}{B1}{have}{none}}
+      {\annotR{A0}{none}{B1}{have}{none}}
 \\
@@
 \icase{F7}{case (x1 < x2)}{F8}{}{F16}{}
-      {\annot{A0}{none}{C0}{have}{have}}
+      {\annotR{A0}{none}{C0}{have}{have}}
 
@@
 \instr{F8}{push sOut2 x1}{F9}{}
-      {\annot{A0}{none}{D0}{have}{have}}
+      {\annotR{A0}{none}{D0}{have}{have}}
 \> @,@
 \instr{F9}{drop sIn1}{F10}{}
-      {\annot{A0}{none}{D1}{none}{have}}
+      {\annotR{A0}{none}{D1}{none}{have}}
 \> @,@
@@
 \> @,@
-\instr{F11}{jump}{F11}{v = b1}
-      {\annot{A0}{pending}{D2}{pending}{have}}
+\instr{F11}{jump}{F12}{v = b1}
+      {\annotL{A0}{pending}{D2}{pending}{have}}
 \> @,@
-\icase{F12}{case (f || (l /= v))}{F13}{}{F14}{}
-      {\annot{A1}{have}{D2}{pending}{have}}
+\instr{F12}{jump}{F13}{x1 = b1}
+      {\annotR{A1}{have}{D2}{pending}{have}}
 \> @,@
-\instr{F13}{push sOut1 v}{F14}{l = v, f = F}
-      {\annot{A2}{have}{D2}{pending}{have}}
+\icase{F13}{case (f || (l /= v))}{F14}{}{F15}{}
+      {\annotL{A1}{have}{C0}{have}{have}}
 \> @,@
-\instr{F14}{jump}{F15}{}
-      {\annot{A3}{have}{D2}{pending}{have}}
+\instr{F14}{push sOut1 v}{F15}{l = v, f = F}
+      {\annotL{A2}{have}{C0}{have}{have}}
 \> @,@
-\instr{F15}{jump}{F7}{x1 = b1}
-      {\annot{A0}{none}{D2}{pending}{have}}
+\instr{F15}{jump}{F7}{}
+      {\annotL{A3}{have}{C0}{have}{have}}
 
@@
 \instr{F16}{push sOut2 x2}{F17}{}
-      {\annot{A0}{none}{E0}{have}{have}}
+      {\annotR{A0}{none}{E0}{have}{have}}
 \> @,@
 \instr{F17}{drop sIn2}{F18}{}
-      {\annot{A0}{none}{E1}{have}{have}}
+      {\annotR{A0}{none}{E1}{have}{have}}
 \> @,@
 \instr{F18}{pull sIn2}{F7}{}
-      {\annot{A0}{none}{E2}{have}{none}}
+      {\annotR{A0}{none}{E2}{have}{none}}
 

### 2017-02-28 2017mergingmerges/section/02-Processes.tex
fig 5. fused graph colours

--- 2017-02-26 2017mergingmerges/section/02-Processes.tex
@@
 , label:  F0
-, instrs:
 \end{code}
+
+\definecolor{groupc}{HTML}{308030}
+\definecolor{mergec}{HTML}{800000}
+\definecolor{sharec}{HTML}{000080}
 
@@
 }
+
 \newcommand\icase[7]{
- \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7 \\
+ \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7
 }
 \newcommand\instr[5]{
- \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5 \\
+ \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5
 }
+
 \begin{tabbing}
@@
          \= @F17@ \= @[ ]     @ \= @F17@ \= @[ ]@
-@   @ \= \tiny F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
-\> @{@ \instr{F0}{pull sIn1 b1}{F1}{}
+@   @ \= \tiny @   @F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill
+
+@, instrs:@
+
+% I have no idea how this works, but you need this particular incantation to colour the whole line in a tabbing environment.
+% It's not ideal that the { and , are coloured too, but that can't be helped.
+\\[0pt \color{sharec}]
+\> @{@
+\instr{F0}{pull sIn1 b1}{F1}{}
       {\annot{A0}{none}{B0}{none}{none}}
 
+\\[0pt \color{groupc}]
 \> @,@
@@
       {\annot{A0}{pending}{B0}{pending}{none}}
+
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A1}{have}{B0}{pending}{none}}
+
+\\[0pt \color{groupc}]
 \> @,@
@@
 
+\\[0pt \color{groupc}]
 \> @,@
@@
 
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A0}{none}{B1}{have}{none}}
+
 \\
-
+\\[0pt \color{mergec}]
 \> @,@
@@
 \\
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A0}{none}{D0}{have}{have}}
+
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A0}{none}{D1}{none}{have}}
+
+\\[0pt \color{sharec}]
 \> @,@
@@
       {\annot{A0}{none}{D2}{none}{have}}
+
+\\[0pt \color{groupc}]
 \> @,@
-\instr{F11}{jump}{F11}{v = b1}
+\instr{F11}{jump}{F12}{v = b1}
       {\annot{A0}{pending}{D2}{pending}{have}}
+
+\\[0pt \color{mergec}]
 \> @,@
-\icase{F12}{case (f || (l /= v))}{F13}{}{F14}{}
+\instr{F12}{jump}{F13}{x1 = b1}
       {\annot{A1}{have}{D2}{pending}{have}}
+
+\\[0pt \color{groupc}]
 \> @,@
-\instr{F13}{push sOut1 v}{F14}{l = v, f = F}
-      {\annot{A2}{have}{D2}{pending}{have}}
+\icase{F13}{case (f || (l /= v))}{F14}{}{F15}{}
+      {\annot{A1}{have}{C0}{have}{have}}
+
+\\[0pt \color{groupc}]
 \> @,@
-\instr{F14}{jump}{F15}{}
-      {\annot{A3}{have}{D2}{pending}{have}}
+\instr{F14}{push sOut1 v}{F15}{l = v, f = F}
+      {\annot{A2}{have}{C0}{have}{have}}
+
+\\[0pt \color{groupc}]
 \> @,@
-\instr{F15}{jump}{F7}{x1 = b1}
-      {\annot{A0}{none}{D2}{pending}{have}}
+\instr{F15}{jump}{F7}{}
+      {\annot{A3}{have}{C0}{have}{have}}
 
 \\
+
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A0}{none}{E0}{have}{have}}
+
+\\[0pt \color{mergec}]
 \> @,@
@@
       {\annot{A0}{none}{E1}{have}{have}}
+
+\\[0pt \color{mergec}]
 \> @,@
@@
 
+\\[0pt \color{black}]
 @} }@

### 2017-02-28 2017mergingmerges/section/02-Processes.tex
I am the king of colours

@@
 \begin{figure}
-\begin{code}
-process
-{ ins:    { sIn1,  sIn2 }
-, outs:   { sOut1, sOut2 }
-, heap:   { f = T, l = 0, v = 0, x1 = 0, x2 = 0, b1 = 0}
-, label:  F0
-\end{code}
-
 \definecolor{groupc}{HTML}{308030}
@@
 \definecolor{sharec}{HTML}{000080}
+
+\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}
+\begin{alltt}
+process
+\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
+, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
+, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
+, label:  \tctt{sharec}{F0}
+\end{alltt}
 
@@
 
-\caption{Fusion of group and merge}
+\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
 \label{fig:Process:Fused}

### 2017-02-28 2017mergingmerges/code/src/Machine/Transform/Fuse.hs
code: modify fusion rules to emit (SharedPush) as a push instruction

@@ -135,7 +135,8 @@ fusePair process1 process2
 
         , processOuts   = Set.fromList
                         $ [ c        | (c, m) <- Map.toList csModes
-                                     ,    m == ModeOutput ]
+                                     ,    m == ModeOutput
+                                       || m == ModeConnected ]
 
         , processHeap   = let Heap h1   = processHeap process1
                               Heap h2   = processHeap process2
@@ -247,7 +248,7 @@ tryStep csMode
          --   Connected output, and the downstream process is ready for the value.
          |  Just ModeConnected    <- csMode   ? c
          ,  Just ModeNone         <- csState2 ? c
-         -> Just $ Jump 
+         -> Just $ Push c xx
                  $ Next (LabelJoint (label1', csState1)
                                     (label2,  Map.insert c ModePending csState2))
                         (Map.insert (VarBuf c) xx xvsUpdate)

### 2017-02-28 2017mergingmerges/section/XX-Proof.tex
notes
S6: proofs: remove cruft

@@
 It is interesting to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
-This is because evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.
-
-The problem with commiting to an evaluation order is most easily explained with a process network containing two infinitely pushing processes.
-Process @A@ is repeatedly pushing to a stream called @X@, while process @B@ repeatedly pushes to @Y@.
-When evaluating this pair as a process network, there are an infinite number of possible interleavings: all @X@s, all @Y@s, pairs of @X@s followed by @Y@s and so on.
-
-When fusion is performed on this process pair, an arbitrary but \emph{particular} order will be chosen.
-Thus, the chosen order will be one of the valid ones, but not all valid orders will be the chosen one.
-
-We know of no realistic examples where combinators have multiple evaluation orders, so we believe this is not an issue in practice.
+This is because, as explained in~\S\ref{s:EvaluationOrder}, evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.
 
@@
 This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
-Ideally, a future version of the formalisation would have this change.
+Ideally, a future version of the formalisation would include this change.
 Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
@@
 
-Despite these differences, we believe the current formalisation gives sufficient confidence in correctness.
+Despite these differences, we believe the Coq formalisation gives sufficient confidence in the correctness of the version presented here.

### 2017-02-28 2017mergingmerges/section/01-Introduction.tex
Edits to S1

@@
 
-In this implementation the @group@ operator filters out consecutive duplicates, while @merge@ combines two sorted streams so that the output remains sorted.
+The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted.
 This example has a few interesting properties.
-Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which this operator pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
-If all the values from @sIn1@ are smaller than the values in @sIn2@, then @merge@ will pull all values from @sIn1@ before pulling the rest from @sIn2@, and vice versa.
+Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
+If all values from @sIn1@ are smaller than the ones from @sIn2@, then @merge@ will pull all values from @sIn1@ before pulling any from @sIn2@, and vice versa.
 Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once.
 To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it.
-Finally, as the stream length is assumed to be unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.
+Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.
 
-We might try to implement each of the operators as a separate concurrent process, and send each identifier value using an intra-process communication mechanism.
+For an implementation, we might try coding each of the operators as a separate concurrent process, and send each stream element using an intra-process communication mechanism.
 Developing such an implementation could be easy or hard, depending on what language features are available for concurrency.
@@
 Each operator could be given this interface, so that the next value from each result stream is computed on demand.
-This approach is commonly taken with implementations of physical operators in database systems.
+This approach is commonly taken with implementations of physical operators in database systems \cite{Graefe:Volcano}.
 However, the `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
@@
 
-Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example either, because they do not support operators with multiple inputs.
+Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs.
 We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network.
@@
 Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
-We view each operator as a process in a concurrent process network, and represent the control flow of the process as a simple state machine.
-Our fusion transform then \emph{sequentializes} the concurrent data flow network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
+We view each operator as a process in a concurrent process network, and the control flow of each process as a simple state machine.
+Our fusion transform then \emph{sequentializes} the concurrent process network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
 When the fusion transform succeeds we know it has worked.
@@
 \item an algorithm for fusing these processes, the first to support splits and joins (\S\ref{s:Fusion});
-\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. (\S\ref{s:Evaluation}).
-\item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs});
+\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. (\S\ref{s:Evaluation});
+\item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs}).
 \end{itemize}
 
-Our fusion transform for infinite stream programs could also serve as the basis for an \emph{array} fusion system, using a natural extension to finite streams.
+Our fusion transform for infinite stream programs also serves as the basis for an \emph{array} fusion system, using a natural extension to finite streams.
 We discuss this extension in \S\ref{s:Finite}.

### 2017-02-28 2017mergingmerges/section/02-Processes.tex
Edits to S2

@@
 A \emph{process} in our system is a simple imperative program with a local heap.
-A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream.
+A process pulls source from an arbitrary number of input streams and pushes result values to at least one output stream.
 The process language is an intermediate representation we use when fusing the overall dataflow network.
@@
 
-A \emph{combinator} is a template for a particular process which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
+A \emph{combinator} is a template for a particular process which parameterizes it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
 Each process implements a logical \emph{operator} --- so we use ``operator'' when describing the values being computed, but ``process'' and ``machine'' when referring to the implementation.
@@
 
-The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}.
-We include the concrete code representation and a diagram of the process viewed as a state machine.
+The definition of the @group@ combinator which detects groups of successive identical elements in input stream is given in Fig.~\ref{fig:Process:Group}.
+The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
+For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@.
+We include the concrete representation and a diagram of the process viewed as a state machine.
 
@@
 The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
+Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
 
@@
 The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
-
-%%% AR: Should we explain intuition behind how process works and @f@, @l@, @v@ before detailed instructions?
+Note that in this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the source representation of @uniquesUnion@ from \S\ref{s:Introduction}.
 
@@
 Next, the instruction @(case (f || (l /= v)) A2 [] A3 [])@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@.
-We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) for whether this is the first element.
+We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.
 
-When the predicate is true, the instruction @(push sOut1 v A3 [ l = v, f = F ])@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).
+When the predicate is true, the instruction @(push sOut1 v A3 [ l = v, f = F ])@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).
 
@@
 This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
-The next element of a stream may only be pulled after all consumers have pulled and then and dropped the current element.
-
-Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
-The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
-For example, when executed on the input stream $[1, 2, 2, 3]$, the process will produce the output $[1, 2, 3]$.
+The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.
 
@@
 
-As this the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
-For example, if we execute the process with the input streams $[1, 4]$ and $[2, 3, 100]$ then the values $[1, 2, 3, 4]$ will be pushed to the output.
-After pushing the last value $4$, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@.
+As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
+For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output.
+After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@.
 We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
@@
 
-Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
-For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
-The result will be a machine that computes the result of both the @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@.
-For ease of comparison we will assume the parameters in each combinator are instantiated by arguments with the same names.
+Our fusion algorithm takes two processes and produces a new one that computes the output of both.
+For example, suppose we need a single process that produces the output of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
+The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process.
+For ease of reference, when describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.
 
@@
 
-The algorithm proceeds by considering pairs of states: one in each of the machines to be fused.
+The algorithm proceeds by considering pairs of states: one in each of the source process state machines to be fused.
 Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Fig.~\ref{fig:Fusion:Pulls}.
 The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
-In the result machine we produce three new instructions that transition between four result states, @F0@ to @F3@.
-Each of the result states represents a combination of two original states, one from each of the input machines.
+In the result machine we produce three new instructions that transition between four joint result states, @F0@ to @F3@.
+Each of the joint result states represents a combination of two source states, one from each of the source machines.
 For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.
 
-We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream.
+We also associate each of the joint result states with information describing whether or not each source state machine has already pulled a value from each of its input streams.
 For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
-The result state @F0@ represents a combination of the two input states @A0@ and @B0@.
+The result state @F0@ represents a combination of the two source states @A0@ and @B0@.
 As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.
 
-From the joint state @F0@, both of the input machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
-In the result machine this is managed by first storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
-For this we attach updates to a @jump@ instruction, which otherwise transitions between states without affecting any of the input or output streams.
+From the result state @F0@, both of the source machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
+In the result machine this is managed by first storing the pulled value in a fresh, shared buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
+To perform the copies we attach updates to a @jump@ instruction, which otherwise transitions between states without affecting any of the input or output streams.
 
@@
 
-Once the result machine has arrived in the joint state @F3@, this is equivalent to the two input machines arriving in states @A1@ and @B1@ respectively.
-The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of these machines.
-From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to its output stream @sOut1@, or to just move on to the next value from its input.
+Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively.
+The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of the source machines.
+From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input.
 From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@.
 In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
-At @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
+From state @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
 
 Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
-Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between @merge@ and @group@ for this pull.
+Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.
 
@@
 
-Fig.~\ref{fig:Process:Fused} shows the end result of fusing @group@ and @merge@ together.
-There are other rules for handling different combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
-The result process has two inputs, @sIn1@ and @sIn2@, which correspond to the original input streams.
-It also has two outputs, @sOut1@ which is the @group@ processes output stream, while @sOut2@ is the @merge@ processes output stream.
+Fig.~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together.
+There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
+The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is the @group@ processes output stream, and @sOut2@ which is the @merge@ processes output stream.
 
-Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequental result process that implements both.
-We have chosen a \emph{specific sequential order} in which to interleave instructions that implement both original processes.
+Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequential result process that implements both.
+We have chosen a \emph{specific sequential order} in which to interleave the instructions that implement both source processes.
 As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
@@
 To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
-The order in which processes are fused together does matter, as does the order in which the instructions are emitted --- we discuss both points further in \S\ref{s:Evaluation}.
+The order in which pairs of processes are fused together does matter, as does the order in which the instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
 
 Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere.
-When we instantiated combinators to create the original processes we introduced fresh names at that point.
-The stream buffer variables we aditionally introduced for coordination were freshly created during fusion.
+When we instantiated combinators to create the original source processes we introduced fresh names at that point.
+The stream buffer variables we additionally introduced for coordination were freshly created during fusion.
 

### 2017-02-28 2017mergingmerges/section/XX-Proof.tex
proof very rough

@@
 
+\begin{code}
+Theorem Soundness (P1 : Program L1 C V1) (P2 : Program L2 C V2) (ss : Streams) (h : Heap)
+    (l1 : L1) (is1 : InputStates) (l2 : L2) (is2 : InputStates) :
+  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
+  -> EvalOriginal Var1 P1 P2 is1 ss h l1 /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
+\end{code}
+
+The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program: if it evaluates to stream @ss@ and heap @h@, then @EvalOriginal@ ensures that the original program evaluates with that program's subset of the heap.
+
 The system described here has some differences to our Coq formalisation.
@@
 This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
-Ideally, a future version of the formalisation would include this change.
+% Ideally, a future version of the formalisation would include this change.
 Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
-Instead we sequentially evaluate both processes separately with the same input values and outputs.
-
-Despite these differences, we believe the Coq formalisation gives sufficient confidence in the correctness of the version presented here.
+Instead we sequentially evaluate both processes with the same input values and outputs.
+Despite these differences, the Coq formalisation gives us sufficient confidence in the correctness of the version presented here.

### 2017-02-28 2017mergingmerges/section/XX-Related.tex
Kill commented out stuff in S1 S2 that we've already handled
proof very rough

@@
 
-This work aims to address the problems with existing combinator-based array fusion systems.
-As stated in the introduction, neither pull-based or push-based fusion alone is sufficient.
+This work aims to address the problems with current combinator-based array fusion systems.
+As stated in the introduction, neither pull-based or push-based fusion is sufficient.
 Some combinators are inherently push-based, particularly those with multiple outputs such as @unzip@; while others are inherently pull-based, such as @zip@.
@@
 % Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}
+
+% Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.

### 2017-02-28 2017mergingmerges/section/XX-Proof.tex
proof: talk about subset of fused heap

@@
 
-The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program: if it evaluates to stream @ss@ and heap @h@, then @EvalOriginal@ ensures that the original program evaluates with that program's subset of the heap.
+The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the fused heap.
+%%% AR: Really would like to say something about this but no room to explain properly
+% For shared inputs when one program has pulled, the other program must be evaluated with the other value removed from the end of the stream.
 

### 2017-02-28 2017mergingmerges/section/03-Definition.tex
Shift discussion of fusion order to Fusion In Practice section, and elaborate
notes
Editing S3.1

@@
 
-A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the currently being executed instruction, and the @instrs@ a map of labels to instructions.
+A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions.
 We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics.
@@
 The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes.
-During fusion, the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state.
-We will use a different version of $\InputState$ in \S\ref{s:Fusion} when we define the fusion algorithm.
+When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state.
+When defining the fusion transform in \S\ref{s:Fusion} we will used a version of $\InputState$ with only this staticaly known information.
 
 The @instrs@ field of the $\Proc$ maps labels to instructions.
-The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel; @case@ which branches depending on the result of a boolean expression; @jump@ which causes control to move to a new instruction, and @drop@ which indicates that the current value pulled from a channel is no longer needed.
+The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel;  @drop@ which indicates that the current value pulled from a channel is no longer needed; @case@ which branches based on the result of a boolean expression, and @jump@ which causes control to move to a new instruction.
 
@@
 
-As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
-Unlike Kahn processes, pushing to a channel can also block.
-Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.
+\eject{}
+When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions.
+The @drop@s instructions are used to control how processes should be synchronized, but do not affect the execution of a single process.
+We will discuss @drop@s further in \S\ref{s:Optimisation}.
 
-When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely ignore the @drop@ instructions.
-The @drop@ instructions are used to help control how processes being fused should be synchronized, but do not affect the execution of a single process.
-We will discuss @drop@s further in \S\ref{s:Optimisation}.
+% This allows us to \emph{deliberately} introduce artificial deadlocks when a process network would require more than one element of buffering.
+%%% AR: added to highlight that this rules out networks that require unbounded buffers
+%%% BL: We don't have any examples of explicitly introducing deadlocks. The process networks just happen to have them when viewed abstractly.
+
+%%% AR: feels a bit disjointed because drops were only mentioned once a few paragraphs ago.
+Maybe reword to talk about lowering in general is obvious for most instructions, and drops are just treated as jumps.
+Or move up.
 
@@
 
-Execution of process networks consists of three aspects:
+The dynamic execution of process networks consists of three aspects:
 
 \begin{enumerate}
-\item \emph{Injection}, which attempts to provide a single value gained from a stream to a process, or a set of processes.
+\item \emph{Injection} of a single value from a stream into a process, or a set of processes.
 Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.
 
-\item \emph{Shaking}, which advances a single process from one instruction to another. Shaking a set of processes succeeds when \emph{any} of the processes in the set can advance.
+\item \emph{Advancing} a single process from one state to another. Advancing a set of processes suceeds when \emph{any} of the processes in the set can advance.
 
-\item \emph{Feeding}, which manages communication between separate processes in the network.
-Feeding alternates between Injecting and Shaking.
+\item \emph{Feeding} outputs of some processes to the inputs of others.
+Feeding alternates between Injecting and Advancing.
 When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input.
 If they all accept it then we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
-
 \end{enumerate}
 
-Evaluation of a process network is non-deterministic.
-At any moment several processes may be able to take a step, while others are blocked trying to pull or push to their channels.
-However, because each process itself is deterministic, and uses blocking reads, the sequence of values pushed to each stream is deterministic, as per Kahn process networks.
+Execution of a process network is non-deterministic.
+At any moment several processes may be able to take a step, while others are blocked.
+As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
+This enables the overall sequence of values on each output channel to be deterministic.
+Unlike Kahn processes, pushing to a channel can also block.
+Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.
+
+%%% AR: what is the distinction between 'execution' and 'evaluation'?
+I only have a vague feeling that execution is something a computer does, while evaluation is the mathematical rules.
+Either way, these should probably be consistent.
+%%% BL: "Evaluation" is pure.  E-"value"-ation. Execution has visible actions, like pushing to streams.
 
 Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not.
-When we fuse two processes together we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.
+When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.
 
-Each output channel may be pushed to by a single process only, so in a sense each output channel is ``owned'' by a single process.
+Each output channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process.
 The only intra-process communication is via channels and streams.
 Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes.
-This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to the formal input and output streams.
+This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to via the formal input and output streams.
 
@@
 Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
-The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields a new process $p'$''.
+The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
 The @injects@ form is similar, but operates on sets of processes rather than a single one.
@@
 Rule (InjectValue) injects of a single value into a single process.
-The value is stored as a (@pending@~ v) binding in the $\InputState$ for the associated channel of the process.
+The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process.
 The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.
@@
 Rule (InjectMany) attempts to inject a single value into a set of processes.
-We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of the.
-Once a value has been injected into all consuming processes that require it, it is then safe for a producing process to discard it.
+We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of them.
+Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.
 
@@
 
+\eject{}
 % -----------------------------------------------------------------------------
-\subsubsection{Shaking}
+\subsubsection{Advancing}
 Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process.
 The first set of rules handle specific instructions.
-The statement $(\ProcBlockShake{i}{is}{\Sigma}{\alpha}{l}{is'}{us})$ reads ``instruction $i$, given channel states $is$ and the heap $\Sigma$, passes control to instruction at label $l$ and yields new channel states $is$, heap update expressions $us'$, and performs an output action $\alpha$.'' An output action $\alpha$ is a list of messages of the form $(\Push~\Chan~\Value)$, which encode the values a process pushes to its output channels.
+The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels.
 We write ~$\cdot$~ for an empty action.
@@
 The statement $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$. A process ``group'' is just a set of processes. 
+%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology
 
@@
 \subsection{Non-deterministic Evaluation Order}
+\label{s:EvaluationOrder}
+%%% AR: Can we simplify this?
+% Maybe we can start by talking about the different evaluation orders for just two alt2s:
+%
+% let s1 = alt2 sInA sInB
+% let s2 = alt2 sInA sInC
+%
+% Here we have many evaluation orders. If we just look at the order in which inputs sInA etc are pulled, some options are:
+% [ A, A, B, C, B, C ]
+% [ A, A, B, B, C, C ]
+% [ A, A, C, C, B, B ]
+%
+% Now if we add a new operator which zips the two together, it actually restricts the pulling order
+%
+% let s1 = alt2 sInA sInB
+% let s2 = alt2 sInA sInC
+% let s3 = zip  s1 s2
+%
+% This restricts the pulling order to:
+% [ A, A, B, C, B, C ]
+%
+%%%
+% Or, maybe we don't even need the zip here, and can leave the zip until S5.1.
+%%%
 

### 2017-02-28 2017mergingmerges/code/src/Machine/Combinator/Alt.hs
code: add cabal file to specify dependencies, remove cruft and reorganise

@@ -7,7 +7,8 @@ import qualified Data.Set       as Set
 
 
 -- | Construct a new `alt2` process.
-mkAlt2_int  :: Channel              -- ^ Input  Channel A.
+mkAlt2_int  
+        :: Channel              -- ^ Input  Channel A.
         -> Channel              -- ^ Input  Channel B.
         -> Channel              -- ^ Output Channel.
         -> New Process

### 2017-02-28 2017mergingmerges/code/src/Machine/Execute.hs

@@ -7,7 +7,6 @@ import Data.Map                         (Map)
 import Data.List
 import qualified Data.Map.Strict        as Map
 import Data.Maybe
-import qualified Text.Show.Pretty       as P
 
 
 execute 
@@ -28,20 +27,6 @@ execute cvsIn cvsOut ps acc
 
  | otherwise
  = (cvsIn, ps, acc)
-
- {-
-
-         = fromMaybe (error $ unlines
-                            [ "execute: feed failed"
-                            , P.ppShow cvsIn
-                            , P.ppShow ps ])
-                     (feedProcesses cvsIn ps)
-
-        (ps'',   as')
-         = shakeSteps [] ps' []
-
-   in   execute cvsIn' cvsOut ps'' (acc ++ as')
--}
 
 
 

### 2017-02-28 2017mergingmerges/code/src/Test/Eval/Map.hs

@@ -3,7 +3,6 @@ module Test.Eval.Map where
 import Machine.Transform.Fuse
 import Machine.Combinator
 import Machine.Execute
-import Text.Show.Pretty
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
 
@@ -11,7 +10,7 @@ import qualified Data.Set       as Set
 -------------------------------------------------------------------------------
 -- | Test evaluation of single map process.
 testEvalMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
@@ -36,7 +35,7 @@ testEvalMap
 -------------------------------------------------------------------------------
 -- | Test evaluation of pipelined map map.
 testEvalPipeMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
@@ -66,7 +65,7 @@ testEvalPipeMapMap
 -------------------------------------------------------------------------------
 -- | Test evaluation of split map-map
 testEvalSplitMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt

### 2017-02-28 2017mergingmerges/code/src/Test/Fuse/GroupMerge.hs

@@ -5,7 +5,6 @@ import Machine.Transform.StripLabels
 import Machine.Combinator
 import Machine.Execute
 import Machine.Pretty
-import Text.Show.Pretty
 import Text.PrettyPrint.Leijen
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set

### 2017-02-28 2017mergingmerges/code/src/Test/Fuse/Map.hs

@@ -4,7 +4,6 @@ import Machine.Transform.Fuse
 import Machine.Transform.StripLabels
 import Machine.Combinator
 import Machine.Execute
-import Text.Show.Pretty
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
 
@@ -14,7 +13,7 @@ import qualified Data.Set       as Set
 --   Note that with a finite input stream the last value is still
 --   in the channel buffer when the process ends.
 testFusePipeMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
@@ -48,7 +47,7 @@ testFusePipeMapMap
 -------------------------------------------------------------------------------
 -- | Test evaluation of split map map.
 testFuseSplitMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
@@ -84,7 +83,7 @@ testFuseSplitMapMap
 -------------------------------------------------------------------------------
 -- | Test evaluation of fused unrelated map map.
 testFuseUnrelatedMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
@@ -115,9 +114,6 @@ testFuseUnrelatedMapMap
                 [ (cCs, [])
                 , (cDs, []) ]
 
---        (inputs', processes', actions')
---         = execute cvsInput cvsOutput [pOut] []
-
-   in   (pMap1, pMap2, modes, bConn, pFused_stripped) -- , actions')
+   in   (pMap1, pMap2, modes, bConn, pFused_stripped)
 
 

### 2017-02-28 2017mergingmerges/code/src/Test/Fuse/Merge.hs

@@ -6,7 +6,6 @@ import Machine.Transform.StripLabels
 import Machine.Combinator
 import Machine.Execute
 import Machine.Pretty
-import Text.Show.Pretty
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
 

### 2017-02-28 2017mergingmerges/code/src/Test/Main.hs

@@ -1,14 +1,0 @@
-
-import Machine.Base
-
-import Test.Process.Map
-
-import Test.Eval.Map
-
-import Test.Fuse.GroupMerge
-import Test.Fuse.Map
-
-
-main 
- = putStrLn "test"
- 

### 2017-02-28 2017mergingmerges/code/src/Test/Process/Map.hs

@@ -2,7 +2,6 @@
 module Test.Process.Map where
 import Machine.Execute
 import Machine.Combinator
-import Text.Show.Pretty
 import qualified Data.Map       as Map
 import qualified Data.Set       as Set
 
@@ -10,7 +9,7 @@ import qualified Data.Set       as Set
 -- | Test construction of map process.
 testProcessMap :: IO ()
 testProcessMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let  cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt
         xSucc x = XApp (XApp XAdd (XInt 1)) x
@@ -22,7 +21,7 @@ testProcessMap
 -- | Test contruction of pipelined map map process.
 testProcessMapMap :: IO ()
 testProcessMapMap
- = putStr $ ppShow 
+ = putStr $ show 
  $ let
         cAs     = Channel "as" TInt
         cBs     = Channel "bs" TInt

### 2017-02-28 2017mergingmerges/section/03-Definition.tex
Finish edits to S3

@@
 Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update that will copy this value into the variable $x$ named in the @pull@ instruction.
-We use the syntax $us \rhd x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$.
-In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied.
+We use the syntax $us,x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$.
+In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied into the local variable.
 
-Rule (Push) evaluates the expression $e$ under heap $\Sigma$ to a value $v$, and produces a corresponding action which carries this value.
-The judgment $(\Sigma \vdash e \Downarrow v)$ expresses standard simply typed lambda calculus reduction using the heap $\Sigma$ for the values of free variables.
+Rule (Push) evaluates the expression $e$ under heap bindings $bs$ to a value $v$, and produces a corresponding action which carries this value.
+The judgment $(bs \vdash e \Downarrow v)$ expresses standard untyped lambda calculus reduction using the heap $bs$ for the values of free variables.
 As this evaluation is completely standard we omit it to save space.
 
-Rule (Drop) changes the input buffer state from @have@ to @none@. A drop can only be executed after pull has set the input buffer to @have@.
+Rule (Drop) changes the input channel state from @have@ to @none@. A @drop@ instruction can only be executed after @pull@ has set the input channel state to @have@.
 
-Rule (Jump) simply produces a new label and update expressions and rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and jump to the appropriate label.
+Rule (Jump) produces a new label and associated update expressions and rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and emit the appropriate label.
 
-The statement $\ProcShake{p}{\alpha}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $\alpha$''.
-Rule (Shake) advances a whole process.
-We lookup the current instruction pointed to by the processes @label@ and pass it, along with the current channel states and heap to the previous single-instruction judgment.
-The update expressions @us@ that the single-instruction judgment yields are first reduced to values before updating the heap.
+The statement $\ProcShake{p}{a}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $a$''.
+Rule (Advance) advances a single process.
+We lookup the current instruction pointed to by the processes @label@ and pass it, along with the current channel states and heap to the previous single instruction judgment.
+The update expressions @us@ that the single instruction judgment yields are first reduced to values before updating the heap.
+We use $(us \lhd bs)$ to replace bindings in $us$ with new ones from $bs$.
 The update expressions themselves are all pure, so the evaluation can safely be done in parallel (or in arbitrary order).
@@
 \subsubsection{Feeding}
-Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting the output actions and feeding output values to other processes.
+Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes.
 The first set of rules concerns feeding values to other processes within the same process network, while the second exchanges input and output values with the environment the process network is running in.
@@
 \eject{}
-The statement $\ProcShake{ps}{\alpha}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $\alpha$. A process ``group'' is just a set of processes. 
-%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology
+The statement $\ProcShake{ps}{a}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $a$. A process ``group'' is just a set of processes.
 
-Rule (ProcessInternal) allows an arbitrary process in the group to advance to a new state at any time, provided it does not produce an output action.
+%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology.
+%%% BL: Structurally it's just a set of processes, but I at some point I used "set" for setting bindings in the heap. If we change group -> network make sure to get all occurrences in the paper.
+
+Rule (ProcessInternal) allows an arbitrary process in the group to advance to a new state at any time, provided it does not yield an output action.
 This allows processes to perform internal computation, without needing to synchronize with the rest of the group.
 
-Rule (ProcessPush) allows an arbitrary process in the group to advance to a new state, while producing an output action (@push@ c v).
-For this to happen it must be possible to inject the output value @v@ into any processes that has channel @c@ as one of its inputs.
+Rule (ProcessPush) allows an arbitrary process in the group to advance to a new state while yielding an output action (@push@ c v).
+For this to happen it must be possible to inject the output value @v@ into all processes that have channel @c@ as one of their inputs.
 As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process.
-When any process in the group produces an output action then we take that as the action of the whole group, as it might also need to be sent to the environment.
+When any process in the group produces an output action we take that as the action of the whole group.
 
@@
 For output channels of the overall group, values pushed to those channels are also collected in the same channel map.
-In a concrete implementation the input and output values would be transported over some IO device, but the simplified semantics here is sufficient to describe the abstract behavior of our system.
+In a concrete implementation the input and output values would be transported over some IO device, but for the semantics here is sufficient to describe the abstract behavior of our system.
 
@@
 
-Rule (FeedPush) collects an output action produced by a process group and adds it to the list corresponding to the output channel.
+Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a process group and appends the value $v$ to the list corresponding to the output channel $c$.
 
@@
 % -----------------------------------------------------------------------------
-\subsection{Non-deterministic Evaluation Order}
+\subsection{Non-deterministic Execution Order}
 \label{s:EvaluationOrder}
@@
 
-The evaluation rules of Fig.
+The execution rules of Fig.
 \ref{fig:Process:Eval:Feed} are non-deterministic in several ways.
 Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the group.
-More importantly, (ProcessPush) allows any process to perform a push action at any time, provided the other processes in the group are ready to accept the pushed value.
-Rule (FeedExternal) also allows new values pushed into an input channel from the environment, provided all processes that use that channel are ready to accept the value.
+More importantly, (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the group are ready to accept the pushed value.
+Rule (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the associated channel are ready to accept the value.
 
-In our system, allowing the evaluation order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
-Consider the following example, where the @alt2@ operator pulls two elements form its first input stream, then two from the second, before pushing all four to its output.
+In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
+Consider the following example, where the @alt2@ operator pulls two elements from its first input stream, then two from the second, before pushing all four to its output stream.
 \begin{code}
@@
 
-Note that the middle stream @sIn2@ is shared, and the result streams from both @alt2@ operators are zipped into tuples.
+Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples.
 Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.
@@
 Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time.
-However, the downstream @zip@ process needs to input one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams.
+However, the downstream @zip@ process needs to pull one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams.
 This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively.
@@
 While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@.
-The evaluation order of the overall process group is constrained by communication patterns of processes in that group.
+The execution order of the overall process group is constrained by communication patterns of processes in that group.
 
-As we cannot encode all possible evaluation orderings into the definition of the processes themselves, we have defined the evaluation rules to admit many possible orderings.
-In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, evaluation order would be discovered dynamically at runtime.
-In contrast, the role of our fusion transform is to construct one of these working evaluation orders statically.
+As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings.
+In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.
+In contrast, the role of our fusion transform is to construct one of these working orders statically.
 In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the process group was executed dynamically.

### 2017-02-28 2017mergingmerges/code/src/Test/Fuse/DiamondMerge.hs
Proof reorg

@@ -1,66 +1,0 @@
-
-
-module Test.Fuse.DiamondMerge where
-import Machine.Transform.Fuse
-import Machine.Transform.StripLabels
-import Machine.Combinator
-import Machine.Execute
-import Machine.Pretty
-import qualified Data.Map       as Map
-import qualified Data.Set       as Set
-
-
--------------------------------------------------------------------------------
--- | Test evaluation of complex network.
---
---
---    cIn1      cIn2       cIn3
---     \      /      \     /
---      (merge)     (merge)
---          |         |
---         cM1       cM2
---          |         |
---           \       /  
---            (merge)
---               |
---              cOut
---
-testFuseDiamondMerge
- = let
-        cIn1    = Channel "cIn1"  TInt
-        cIn2    = Channel "cIn2"  TInt
-        cIn3    = Channel "cIn3"  TInt
-        cM1     = Channel "cM1"   TInt
-        cM2     = Channel "cM2"   TInt
-        cOut    = Channel "cOut" TInt
-
-        pFused
-         = evalNew 
-         $ do   
-                p1_merge1       <- mkMerge cM1  cM2  cOut
-                p2_merge2       <- mkMerge cIn1 cIn2 cM1
-                let Right p12_merge12    = fusePair p1_merge1   p2_merge2
-
-                p3_merge3       <- mkMerge cIn2 cIn3 cM2
-                let Right p123_merge123  = fusePair  p12_merge12 p3_merge3
-
-                pFused'         <- fmap fst $ stripLabels "F" p123_merge123
-
-                return pFused'
-
-        cvsInput
-         =  Map.fromList
-                [ (cIn1, [VInt 1, VInt 2, VInt 3, VInt 4, VInt 5])
-                , (cIn2, [VInt 1, VInt 2, VInt 3, VInt 4, VInt 5]) 
-                , (cIn3, [VInt 1, VInt 2, VInt 3, VInt 4, VInt 5]) ]
-
-        cvsOutput
-         = Map.fromList
-                [ (cM1, [])
-                , (cM2, []) ]
-
-   in do
-        putStrLn $ show $ pretty pFused
-
-        executeTraceIO 100 cvsInput cvsOutput [pFused] []
-

### 2017-02-28 2017mergingmerges/section/04-Fusion.tex
notes
Half way though cleaning S4

@@
 
-The core fusion algorithm works by constructing a static schedule for a single pair of processes. To fuse a whole process group we fuse pairs until only one process remains.
+The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process group we fuse pairs until only one process remains.
 
-Fig.~\ref{fig:Fusion:Types} shows the type definitions for fusion.
-Because the fused process is \emph{statically} evaluating both processes, we use its labels to encode static information about the dynamic state of the original processes.
-The fused label contains the pair of both original labels, as well as the static part of the $\InputState$ for each input channel.
-If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, but rather than knowing its actual value as in the dynamic evaluation, we know its value is stored in the $@chan@$ variable for that channel.
+Fig.~\ref{fig:Fusion:Types} shows the auxilliary grammar used during fusion.
+We use the labels in the result process to encode statically known information about the state of the source processes.
+A label in the result process consists of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
+If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, do not know the actual value.
+We use the (@chan@ c) variable form to represent the stream buffer variable associated with channel @c@ (we used a fresh variable @b1@ back in \S\ref{fig:Process:Fused})
 
-$\ChanType_2$ classifies the kind of channels and the communications between two processes.
-Two processes can read from the same channel (@in2@), in which case pulling must be coordinated together.
-When one process reads from a channel and the other ignores it (@in1@) no coordination is required.
-When one process writes to a channel and the other reads (@in1out1@) the reading process must wait for the other to write.
-Finally, when one process writes and the other ignores (@out1@) no coordination is necessary.
-Recall that each output channel is uniquely owned and cannot be written by another process, so it is not possible for both processes to write to the same output.
+$\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
+Type @in2@ indicates that two processes @pull@ from the same channel, which must then be coordinated.
+Type @in1@ indicates that only a single process pulls from a channel.
+Type @in1out1@ indicates that one processes pushes to a channel that the other pulls from.
+Type @out1@ indicates that the channel is pushed to by a single process.
+Recall that each output channel is uniquely owned and cannot be pushed to by more than one process.
 
 \input{figures/Fusion-types.tex}
-
 \input{figures/Fusion-fusePair.tex}
 
-Fig.~\ref{fig:Fusion:Def:Top} defines the top-level function that fuses a pair of processes.
-Function \ti{fusePair} fuses a pair of processes together, constructing a new process that computes the outputs of both.
-The instructions are constructed by computing a fixpoint, starting at the initial labels of each process with empty input states.
-Each instruction is computed with \ti{tryStepPair}, which statically chooses one of the two processes to execute, and any reachable instructions are added recursively until fixpoint is reached.
-This implementation assumes the heap variables are distinct, which can easily be ensured by renaming.
+Fig.~\ref{fig:Fusion:Def:Top} defines the top-level function \ti{fusePair} that fuses a pair of processes together, constructing a new process that computes the outputs of both.
+We start with a joint label $l_0$ formed from the initial labels of the two source processes.
+The joint label encodes the source states which both processes will be in at that point.
+We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next.
+The possible destination labels of that instruction (computed with $outlabels$) define new reachable states.
+As we discover new reachable states we add them to a map $bs$ of joint label to corresponding instruction, and repeat the process to a fixpoint where no more result states can be discovered.
 
@@
 
-Function \ti{tryStepPair} defined in Fig.~\ref{fig:Fusion:Def:StepPair} calls this for both processes, and if either machine can execute it will execute that process.
-It takes the channel state map, and for both input processes the label with static input state, as well as the instruction at that label.
+Function \ti{tryStepPair} defined in Fig.~\ref{fig:Fusion:Def:StepPair} decides which of the two processes to advance.
 
-Clauses (PreferJump1) and (PreferJump2) apply heuristics to prioritise which process should take a step.
-If one of the processes can perform a jump, it is given priority.
-This is an attempt to perform multiple jumps together, as many consecutive jumps could later be contracted into a single jump.
-Because clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in the swapped order, it returns an instruction with the output labels swapped. We use $\ti{swaplabels}$ to swap them back.
+Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@.
+This heuristic helps to collect jump instructions together in the result process, which are then easier to optimize in post-fusion optimization (\S\ref{s:Optimisation}).
+As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ to re-swap the labels in the result.
 
-Clauses (DeferPull1) and (DeferPull2) perform similar heuristics, deferring pulls as much as possible.
-This is because pulls may block, while other actions are more likely to produce immediate results.
+\eject{}
+Clauses (DeferPull1) and (DeferPull2) are similar heuristics, deferring @pull@ instructions if possible.
+We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
 
-Clauses (Run1) and (Run2) apply when only one process can run, or both processes are pulls.
-If both are pulls, we make the arbitrary choice to execute $p$ with (Run1).
+Clauses (Run1) and (Run2) apply when one process can advance using some other instruction.
+We try the first process first, and if it can advance then so be it.
+This fact means that fusion is left-biased, preferring advancement of the left (first) process over the second.
 
@@
 
-Fig.~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which checks if one of the processes can take a step.
-It takes the channel types, the label with static input states and instruction at that label, as well as the other process' label with static input states.
+\smallskip
+Fig.~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which schedules a single instruction.
+The \ti{tryStep} function takes the map channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.
 
-Function \ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
-If the process is pulling or pushing a non-shared channel categorised as @in1@ or @out1@, no coordination is required and injection is ignored.
+% the label with static input states and instruction at that label, as well as the other process' label with static input states.
+
+% Function \ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
+
+% If the process is pulling or pushing a non-shared channel categorised as @in1@ or @out1@, no coordination is required and injection is ignored.
 For shared and connected channels, the process can only pull or push if the other process would be willing to accept the injection at the same time.
@@
 
-Clause (LocalJump) applies when the process is trying to jump.
-In this case, the fused instruction simply performs the jump, leaving the other process as-is.
-When the fused instruction is evaluated, it corresponds to the same (Jump) shake rule on the input process.
+%%% AR: I'm sure all these clauses can be cleaned up. I think the real point to make is that each of these correspond more or less to an injection and shake rule.
+Clause (LocalJump) applies when the left process wants to jump.
+In this case, the result instruction simply performs the corresponding jump, leaving the other process where it is.
 
-Clause (LocalCase) is equally simple, and the fused instruction performs the case.
+% When the fused instruction is evaluated, it corresponds to the same (Jump) shake rule on the input process.
 
-Clause (LocalPush) applies when the process is trying to push to a non-shared, local channel.
-In this case the push can be performed as usual, with no coordination required.
+Clause (LocalCase) is similar, except that there are two destination labels.
 
-Clause (SharedPush) applies when the process is trying to push to a shared channel.
-Pushing to a shared channel requires the other process to be ready to accept the injected value.
-In terms of evaluating the input processes, this corresponds to the (ProcessesPush) and (InjectValue) rules where push can only succeed if the other process' dynamic input state is ``@none@''.
-Here we encode the injection rules inside the fused process by requiring the static input state to be ``$@none@_S$''.
-When this is satisfied, the push also stores the pushed value in a local variable ``$@chan@~c$'' and sets the static input state to ``$@pending@_S$'' denoting that the value is available.
+Clause (LocalPush) applies when the left process wants to push to a non-shared, local channel.
+In this case the push can be performed directly, with no additional coordination required.
 
-Clause (LocalPull) applies when the process is trying to pull from a local channel with no coordination required.
+Clause (SharedPush) applies when the left process wants to push to a shared channel.
+Pushing to a shared channel requires the other process to be ready to accept a value on that channel at the same time.
+We encode the state constraint of the injection rules by requiring the static input state to be ``$@none@_S$''.
+When this is satisfied, the push stores the pushed value in a local variable ``$@chan@~c$'' and sets the static input state to ``$@pending@_S$'' which indicates that the value is available.
 
-Clause (SharedPull) applies when the process is trying to pull from a shared channel that the other process either pulls from or pushes to.
+% In terms of evaluating the input processes, this corresponds to the (ProcessesPush) and (InjectValue) rules where push can only succeed if the other process' dynamic input state is ``@none@''.
+
+\eject{}
+Clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.
+
+Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to.
 The static state of the process has a ``$@pending@_S$'' value for the channel, which corresponds to a dynamic state of ``$@pending@~(@chan@~c)$''.
-This means that one of the processes has pulled a value, or the other process has pushed a value, and in either case there is a value ready to use.
-The key observation here is that when this jump is evaluated it will use the pending value from ``$(@chan@~c)$'' and act equivalently to the shake rule (Pull) on a single process.
+When the shared channel has this state it means that one of the processes has already pulled the current value, or the other process has pushed a value.
+In either case there is a value ready to use.
+Once the result instruction is executed, successive instructions can use the pending value from ``$(@chan@~c)$'', similarly to the single process (Pull) execution rule.
 
-Clause (SharedPullInject) applies when the process is trying to pull from a shared channel that both processes pull from, and neither process has a value.
-The static state of both processes are ``$@none@_S$'', so the value can be injected into both.
-We encode the injection rules for the input processes inside the fused process, so that the fused process has explicit control over injection.
-To do this, we pull from the input channel, setting both static states to ``$@pending@_S$''.
-Now the dynamic state of the fused process is ``@have@'' as it has successfully pulled, but we set the static state of the input processes to ``$@pending@_S$'' as from their perspective the value has been injected but not yet pulled.
-This step leaves both processes at their current label, allowing the next step to be (SharedPull).
+Clause (SharedPullInject) applies when the left process wants to pull from a shared channel that both processes pull from, and neither process already has a value.
 
-Clause (LocalDrop) applies when the process is trying to drop a local channel with no coordination required.
+% The static state of both processes is ``$@none@_S$'', so the value can be injected into both.
+% We encode the injection rules for the input processes inside the fused process, so that the fused process has explicit control over injection.
+% To do this, we pull from the input channel, setting both static states to ``$@pending@_S$''.
+% Now the dynamic state of the fused process is ``@have@'' as it has successfully pulled, but we set the static state of the input processes to ``$@pending@_S$'' as from their perspective the value has been injected but not yet pulled.
+% This step leaves both processes at their current label, allowing the next step to be (SharedPull).
+
+Clause (LocalDrop) applies when the left process wants to drop the current value from a local channel, which requires no coordination.
 
@@
 Because the value was not pulled through a channel but instead sent by a local variable, the channel does not need to be dropped.
-We set the static input state to ``$@none@_S$'', essentially performing the (Drop) shake rule statically.
+
+% We set the static input state to ``$@none@_S$'', essentially performing the (Drop) shake rule statically.
 
@@
 
+\smallskip
 Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.

### 2017-02-28 2017mergingmerges/section/02-Processes.tex
S2 wibble prose

@@
 A \emph{process} in our system is a simple imperative program with a local heap.
-A process pulls source from an arbitrary number of input streams and pushes result values to at least one output stream.
+A process pulls source values from an arbitrary number of input streams and pushes result values to at least one output stream.
 The process language is an intermediate representation we use when fusing the overall dataflow network.
@@
 
-The definition of the @group@ combinator which detects groups of successive identical elements in input stream is given in Fig.~\ref{fig:Process:Group}.
+The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Fig.~\ref{fig:Process:Group}.
 The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.

### 2017-02-28 2017mergingmerges/section/04-Fusion.tex
More copy edits to S4

@@
 
-The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process group we fuse pairs until only one process remains.
+The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process group we fuse successive pairs of processes until only one remains.
 
-Fig.~\ref{fig:Fusion:Types} shows the auxilliary grammar used during fusion.
-We use the labels in the result process to encode statically known information about the state of the source processes.
-A label in the result process consists of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
-If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, do not know the actual value.
-We use the (@chan@ c) variable form to represent the stream buffer variable associated with channel @c@ (we used a fresh variable @b1@ back in \S\ref{fig:Process:Fused})
+Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion.
+We extend the exiting $Label$ grammar with an additional alternative, $Label_S \times Label_S$ which we use for the labels in a fused result process.
+Each $Label_S$ consists of a $Label$ from one of the  source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$.
+The definition of $Label$ is now recursive.
+While fusing a whole process group, as we fuse pairs of individual processes the labels in the result collect more and more information.
+Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.
 
-$\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
-Type @in2@ indicates that two processes @pull@ from the same channel, which must then be coordinated.
-Type @in1@ indicates that only a single process pulls from a channel.
-Type @in1out1@ indicates that one processes pushes to a channel that the other pulls from.
+% These new labels $Label_S$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
+
+% If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, do not know the actual value.
+
+We also extend the existing $Variable$ grammar with a (@chan@ c) form which represents the stream buffer variable associated with channel @c@.
+We only need one stream buffer variable for each channel, and naming them like this saves us from inventing fresh variable names in the definition of the fusion rules (we used a fresh name @b1@ back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion)
+
+Still in Fig.~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
+Type @in2@ indicates that the two processes @pull@ from the same channel, and these actions must then be coordinated.
+Type @in1@ indicates that only a single process pulls from the channel.
+Type @in1out1@ indicates that one process pushes to the channel and the other pulls from the same.
 Type @out1@ indicates that the channel is pushed to by a single process.
-Recall that each output channel is uniquely owned and cannot be pushed to by more than one process.
+Each output channel is uniquely owned and cannot be pushed to by more than one process.
 
@@
 
-Fig.~\ref{fig:Fusion:Def:Top} defines the top-level function \ti{fusePair} that fuses a pair of processes together, constructing a new process that computes the outputs of both.
+\smallskip
+% -------------------------------------
+Fig.~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes together, constructing a result process that does the job of both.
 We start with a joint label $l_0$ formed from the initial labels of the two source processes.
-The joint label encodes the source states which both processes will be in at that point.
 We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next.
-The possible destination labels of that instruction (computed with $outlabels$) define new reachable states.
-As we discover new reachable states we add them to a map $bs$ of joint label to corresponding instruction, and repeat the process to a fixpoint where no more result states can be discovered.
+The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states.
+As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no more states can be discovered.
 
@@
 
-Function \ti{tryStepPair} defined in Fig.~\ref{fig:Fusion:Def:StepPair} decides which of the two processes to advance.
+\eject{}
+% -------------------------------------
+Fig Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
 
 Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@.
-This heuristic helps to collect jump instructions together in the result process, which are then easier to optimize in post-fusion optimization (\S\ref{s:Optimisation}).
-As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ to re-swap the labels in the result.
+This heuristic helps to collect @jump@ instructions together in the result process, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle.
+As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.
 
-\eject{}
-Clauses (DeferPull1) and (DeferPull2) are similar heuristics, deferring @pull@ instructions if possible.
+Clauses (DeferPull1) and (DeferPull2) are similar heuristics, which defer @pull@ instructions if possible.
 We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
@@
 We try the first process first, and if it can advance then so be it.
-This fact means that fusion is left-biased, preferring advancement of the left (first) process over the second.
+This priority means that fusion is left-biased, preferring advancement of the left process over the second.
 
@@
 
+% -------------------------------------
 \smallskip
-Fig.~\ref{fig:Fusion:Def:Step} defines the \ti{tryStep} function which schedules a single instruction.
+Fig.~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction.
 The \ti{tryStep} function takes the map channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.
@@
 
+% -------------------------------------
 \input{figures/FusionUtils.tex}

### 2017-02-28 2017mergingmerges/section/XX-Proof.tex
mention Var1
S6 proof, S7 related some cuts

@@
 
-The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the fused heap.
+The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the fused heap, using @Var1@ and @Var2@ to extract the variables.
+Care must be taken to remove stream values that the other process has pulled but this one has not yet.
 %%% AR: Really would like to say something about this but no room to explain properly

### 2017-02-28 2017mergingmerges/section/XX-Related.tex
S6 proof, S7 related some cuts

@@
 
-This work aims to address the problems with current combinator-based array fusion systems.
+This work aims to address the limitations of current combinator-based array fusion systems.
 As stated in the introduction, neither pull-based or push-based fusion is sufficient.
@@
 
-Short cut fusion is an attractive idea, as it allows fusion systems to be specified with a single rewrite rule.
+Short cut fusion is an attractive idea, as it allows fusion systems to be specified by a single rewrite rule.
 However, short cut fusion relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
 Thus, short cut fusion is inherently biased towards pull fusion.
-Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support only the intersection of pull and push: that is, neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
+Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
 
 Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
-For effectful input streams, for example reading from the network, this duplicated work can change the overall meaning of the program.
+For effectful inputs such as reading from the network, duplicating work changes the semantics.
 % I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general
@@
 This allows unfusable programs to be caught at compile time.
-The main drawback of data flow fusion is that it only supports a limited set of combinators, and adding more is non-trivial.
+However, it only supports a limited set of combinators, and adding more is non-trivial.
 
 One way to address the difference between pull and push streams is to explicitly support both separately, as seen in \citet{bernardy2015duality} and \citet{lippmeier2016polarized}.
-Here, pull streams have the type @Source@ and are represented as a source that is always available to be pulled from, while push streams have the type @Sink@ and are represented as a sink that can always accept input.
-Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers and ordering of effects.
+Here, pull streams have the type @Source@ and represent a source that is always available to be pulled from, while push streams have the type @Sink@ and represent a sink that can always accept input.
+Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers.
 Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.

### 2017-02-28 2017mergingmerges/section/04-Fusion.tex
Finish edits of fusion

@@
 Clauses (Run1) and (Run2) apply when one process can advance using some other instruction.
-We try the first process first, and if it can advance then so be it.
+We try the first process first, and if that can advance then so be it.
 This priority means that fusion is left-biased, preferring advancement of the left process over the second.
@@
 Fig.~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction.
-The \ti{tryStep} function takes the map channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.
+This function takes the map of channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.
 
-% the label with static input states and instruction at that label, as well as the other process' label with static input states.
-
-% Function \ti{tryStep} statically performs shaking for the current process, as well as injection for both processes.
-
-% If the process is pulling or pushing a non-shared channel categorised as @in1@ or @out1@, no coordination is required and injection is ignored.
-For shared and connected channels, the process can only pull or push if the other process would be willing to accept the injection at the same time.
-Simple instructions such as @case@ and @jump@ require no coordination and can be executed at any point.
-
-%%% AR: I'm sure all these clauses can be cleaned up. I think the real point to make is that each of these correspond more or less to an injection and shake rule.
 Clause (LocalJump) applies when the left process wants to jump.
-In this case, the result instruction simply performs the corresponding jump, leaving the other process where it is.
-
-% When the fused instruction is evaluated, it corresponds to the same (Jump) shake rule on the input process.
+In this case, the result instruction simply performs the corresponding jump, leaving the right process where it is.
 
@@
 
-Clause (LocalPush) applies when the left process wants to push to a non-shared, local channel.
+Clause (LocalPush) applies when the left process wants to push to a non-shared output channel.
 In this case the push can be performed directly, with no additional coordination required.
@@
 Clause (SharedPush) applies when the left process wants to push to a shared channel.
-Pushing to a shared channel requires the other process to be ready to accept a value on that channel at the same time.
-We encode the state constraint of the injection rules by requiring the static input state to be ``$@none@_S$''.
-When this is satisfied, the push stores the pushed value in a local variable ``$@chan@~c$'' and sets the static input state to ``$@pending@_S$'' which indicates that the value is available.
-
-% In terms of evaluating the input processes, this corresponds to the (ProcessesPush) and (InjectValue) rules where push can only succeed if the other process' dynamic input state is ``@none@''.
+Pushing to a shared channel requires the downstream process to be ready to accept the value at the same time.
+We encode this constraint by requiring the static input state of the downstream channel to be $@none@_S$.
+When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $@chan@~c$'' and sets the static input state to ``$@pending@_S$'', which indicates that the new value is now available.
 
 \eject{}
-Clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.
+Still in Fig.~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.
 
 Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to.
-The static state of the process has a ``$@pending@_S$'' value for the channel, which corresponds to a dynamic state of ``$@pending@~(@chan@~c)$''.
-When the shared channel has this state it means that one of the processes has already pulled the current value, or the other process has pushed a value.
-In either case there is a value ready to use.
-Once the result instruction is executed, successive instructions can use the pending value from ``$(@chan@~c)$'', similarly to the single process (Pull) execution rule.
+We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$.
+The result instruction copies the value from the stream buffer variable into variable specific to the left source process, and the corresponding $@have_S@$ channel state in the result label records that it has done so.
 
-Clause (SharedPullInject) applies when the left process wants to pull from a shared channel that both processes pull from, and neither process already has a value.
+Clause (SharedPullInject) applies when the left process wants to pull from a shared channel that both processes pull from, and neither already has a value.
+The result instruction is a @pull@ that loads the stream buffer variable.
 
-% The static state of both processes is ``$@none@_S$'', so the value can be injected into both.
-% We encode the injection rules for the input processes inside the fused process, so that the fused process has explicit control over injection.
-% To do this, we pull from the input channel, setting both static states to ``$@pending@_S$''.
-% Now the dynamic state of the fused process is ``@have@'' as it has successfully pulled, but we set the static state of the input processes to ``$@pending@_S$'' as from their perspective the value has been injected but not yet pulled.
-% This step leaves both processes at their current label, allowing the next step to be (SharedPull).
+Clause (LocalDrop) applies when the left process wants to drop the current value from that it read from an unshared input channel, which requires no coordination.
 
-Clause (LocalDrop) applies when the left process wants to drop the current value from a local channel, which requires no coordination.
+Clause (ConnectedDrop) applies when the left process wants to drop the current value that it received from an upstream process.
+As the value will have been sent via a heap variable instead of a still extant channel, the result instruction just performs a @jump@ while updating the static channel state.
 
-Clause (ConnectedDrop) applies when dropping a value pushed by the other process.
-Because the value was not pulled through a channel but instead sent by a local variable, the channel does not need to be dropped.
-
-% We set the static input state to ``$@none@_S$'', essentially performing the (Drop) shake rule statically.
-
-Clauses (SharedDropOne) and (SharedDropBoth) apply when dropping a value that both have pulled.
-When both processes pull from the same input, the first drop is a fake drop as with (ConnectedDrop), but the second one is a real drop.
-If the other process' static state is not ``$@none@_S$'', the other process has not dropped yet so we perform a fake drop.
-Otherwise, we perform the real drop.
-After the real drop is performed, both static states will be ``$@none@_S$'' and either process will be able to pull again.
+Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop a value it gained from a channel which is shared by the other process.
+In (SharedDropOne) the channel states reveal that the other process is still using the value.
+In this case the result instruction is a @jump@ that only updates the static information that the left source process no longer requires this value.
+In (SharedDropBoth) then channel states reveal that the other process no longer needs the value either.
+In this case the result instruction is a real @drop@, because we are sure that neither process requires the value any longer.
 
@@
 \smallskip
-Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which are not specific to fusion.
+Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which we have already mentioned.
 Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
-Function \ti{outlabels} gets the set of output labels for an instruction - this is used when computing the fixpoint of the instructions map.
+Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states.
 Function \ti{swaplabels} flips the order of the compound labels in an instruction.

### 2017-02-28 2017mergingmerges/section/XX-Proof.tex
Copy edits to proof

--- 2017-02-28 2017mergingmerges/section/XX-Proof.tex
@@
 
-Our fusion system is formalised in Coq, where we have proved soundness of \ti{fusePair}: if the fused program evaluates to a particular output, then the two original programs also evaluate to that output.
-It is interesting to note that the converse is not necessarily true: just because two programs can evaluate to a particular output does not mean the fused program will evaluate to that.
+Our fusion system is formalized in Coq, and we have proved soundness of \ti{fusePair}: if the fused result process produces a particular sequence of values on its output channels then the two source processes may also produce that same sequence.
+Note that due to non-determinism of process execution the converse is not true in practice: just because the two concurrent processes can produce a particular output sequence does not mean the fused result process will as well --- the fused result process uses only one of the many possible orders.
+
+% Note that the converse is not necessarily true: just because two processes can evaluate to a particular output does not mean the fused program will evaluate to that.
 This is because, as explained in~\S\ref{s:EvaluationOrder}, evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.
@@
 Theorem Soundness (P1 : Program L1 C V1) (P2 : Program L2 C V2) (ss : Streams) (h : Heap)
-    (l1 : L1) (is1 : InputStates) (l2 : L2) (is2 : InputStates) :
-  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
+                  (l1 : L1) (is1 : InputStates) (l2 : L2) (is2 : InputStates) 
+  :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
   -> EvalOriginal Var1 P1 P2 is1 ss h l1 /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
@@
 
-The @Soundness@ theorem above uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the fused heap.
+The @Soundness@ theorem uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the result heap, using @Var1@ and @Var2@ to extract the variables.
+
+% Care must be taken to remove stream values that the other process has pulled but this one has not yet.
 %%% AR: Really would like to say something about this but no room to explain properly
@@
 
-The system described here has some differences to our Coq formalisation.
-First, the Coq formalisation has a separate @update@ instruction which modifies a variable in the local heap, rather than allowing heap updates in the output \Next~label of any instruction.
-This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
+To aid mechanization, the Coq formalization has some small differences from the system presented in this paper.
+Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction.
+Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition.
+Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.\ref{fig:Process:Eval:Feed}.
+Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process.
+This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
+
+% This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update.
 This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
 % Ideally, a future version of the formalisation would include this change.
-Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
+% Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process.
 Instead we sequentially evaluate both processes with the same input values and outputs.

### 2017-02-28 2017mergingmerges/section/04-Fusion.tex
prose S4 S5

@@
 % -------------------------------------
-Fig Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
+Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
 
@@
 We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$.
-The result instruction copies the value from the stream buffer variable into variable specific to the left source process, and the corresponding $@have_S@$ channel state in the result label records that it has done so.
+The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_S$ channel state in the result label records that it has done so.
 

### 2017-02-28 2017mergingmerges/section/05-Evaluation.tex
notes
Half way though cleaning S4
prose S4 S5

@@
 
+\eject{}
 % -----------------------------------------------------------------------------
@@
 % -----------------------------------------------------------------------------
-\subsection{Fusability}
+\subsection{Fusibility}
 \label{s:FusionOrder}
@@
 
-In our current system, if fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails.
+In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails.
 This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
@@
 Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock.
-Statically, when we try to fuse the result process with the downstrem @zip@ process then the deadlock is discovered and fusion fails.
+Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails.
 Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}.
@@
 The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
-This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and may other systems based on general purpose program transformations.
+This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
 In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
@@
 Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
-When the dataflow network has a single output stream then we use the method of starting from the process closes to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
+When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
 This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
@@
 
-As with fusion transformation system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
-The left of Fig.\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
+As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
+The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
 To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes.
 The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start.
-The right of Fig.\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
+The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
 In both cases the number of states in the result process grows linearly with the number of processes.
 In all combinations, with up to 7 processes there are less than 100 states in the result process.
+%%% AR: Note that the split version also has only one merge?
 
@@
 
-On the other hand, Fig.\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number if input processes.
+On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
 The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream.
@@
 During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition.
-The use of the finite map ensures that identical states are always combined, but genuinly different states always make it into the result.
+The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result.
 
-In the worst case fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
+In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
 If we assume the two processes have about the same number of states then this is O($n^2$).
@@
 Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states.
-Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusability of the overall process network will not be affected.
+Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusibility of the overall process network will not be affected.
 

### 2017-04-22 2017mergingmerges/section/03-Definition.tex
attempt to fit in sigconf 12p. format

@@
 
-\clearpage{}
 % -----------------------------------------------------------------------------
@@
 
-\eject{}
 When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions.
@@
 
-\eject{}
 % -----------------------------------------------------------------------------
@@
 
-\eject{}
 Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update that will copy this value into the variable $x$ named in the @pull@ instruction.
@@
 
-\eject{}
 The statement $\ProcShake{ps}{a}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $a$. A process ``group'' is just a set of processes.

### 2017-04-22 2017mergingmerges/section/04-Fusion.tex
attempt to fit in sigconf 12p. format

@@
 
-\eject{}
 % -------------------------------------
@@
 
-\eject{}
 Still in Fig.~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.

### 2017-04-29 2017mergingmerges/section/05A-Benchmarks.tex
initial benchmarks

%!TEX root = ../Main.tex

% -----------------------------------------------------------------------------
\section{Results}
\label{s:Benchmarks}

We have implemented this system using Template Haskell for code generation in a library called @folderol@\footnote{\url{https://github.com/amosr/folderol}}.
To better compare against existing fusion systems, we have chosen to use the finite stream extension mentioned in~\S\ref{s:Finite} for our benchmarks.
We present three benchmarks: one array benchmark, and two file-based benchmarks.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Store, Recompute, Ours, Hand, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=200,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for Quickhull; lower is faster.}
\label{fig:bench:quickhull}
\end{figure}

\subsection{Quickhull}
Quickhull is a divide-and-conquer spatial algorithm to find the smallest convex hull containing all points.
At its core is an operation called `filterMax', which takes a line and an array of points, and computes the maximum distance from the line, as well as all points that are above the line.
In our system this can be fused into a single loop, while shortcut-fusion systems do not support horizontal fusion.

We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
For our benchmarks we implemented both, and recomputing was significantly faster.
However, it is worth noting that we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh the cost of array allocation.

\autoref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the recursive case.
It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=360,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(360,Conduit) (310,Pipes) (235,Streaming) (160,Ours) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for append2; lower is faster.}
\label{fig:bench:append}
\end{figure}

\subsection{Append files}
The first file-based benchmark consists of appending two files, while counting the number of lines.
We have benchmarked against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
\autoref{fig:bench:append} shows the runtimes for appending 1.8MB of data.
The absolute performance here is poor here because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would be the same.

\begin{figure}
\begin{tikzpicture}
\begin{axis}[
  symbolic y coords={Streaming, Conduit, Pipes, Ours, end},
	xlabel=Runtime (ms),
  xmin=0, xmax=1570,
	enlargelimits=0.01,
	xbar interval=0.7,
  width=7.5cm, height=4.5cm,
	legend pos=north west,
]
\addplot coordinates {(1570,Streaming) (980,Conduit) (940,Pipes) (590,Ours) (0,end) };
\end{axis}
\end{tikzpicture}
\caption{Runtime for partition; lower is faster.}
\label{fig:bench:part}
\end{figure}

\subsection{Partition file}
The second file-based benchmark takes a single input file and partitions it into two files: one with even-length lines, and one with odd-length lines.
The number of lines in each output file is also counted.
The three streaming libraries are pull-based, and so do not support multiple outputs; the program must be written in a convoluted way or at least partially hand-fused.
Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
\autoref{fig:bench:part} shows the runtimes for partitioning a file into two.

### 2017-05-01 2017mergingmerges/section/05A-Benchmarks.tex
Update benchmark graphs

@@
 	xlabel=Runtime (ms),
-  xmin=0, xmax=360,
+  xmin=0, xmax=290,
 	enlargelimits=0.01,
@@
 ]
-\addplot coordinates {(360,Conduit) (310,Pipes) (235,Streaming) (160,Ours) (0,end) };
+\addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
 \end{axis}
@@
 \autoref{fig:bench:append} shows the runtimes for appending 1.8MB of data.
-The absolute performance here is poor here because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would be the same.
+The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
 
@@
 \begin{axis}[
-  symbolic y coords={Streaming, Conduit, Pipes, Ours, end},
+  symbolic y coords={Streaming, Pipes, Conduit, Ours, end},
 	xlabel=Runtime (ms),
-  xmin=0, xmax=1570,
+  xmin=0, xmax=1200,
 	enlargelimits=0.01,
@@
 ]
-\addplot coordinates {(1570,Streaming) (980,Conduit) (940,Pipes) (590,Ours) (0,end) };
+\addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit) (288,Ours) (0,end) };
 \end{axis}

### 2017-05-01 2017mergingmerges/section/01-Introduction.tex
typesetting to cut down

@@
 Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers.
-The identifiers from both streams arrive sorted, but may include duplicates.
-We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of identifiers from both streams.
+The identifiers from both streams are sorted, but may include duplicates.
+We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of both streams.
 Here is how we might write the source code, where @S@ is for @S@-tream.
@@
 This example has a few interesting properties.
-Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
-If all values from @sIn1@ are smaller than the ones from @sIn2@, then @merge@ will pull all values from @sIn1@ before pulling any from @sIn2@, and vice versa.
+Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
 Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once.
@@
 
-For an implementation, we might try coding each of the operators as a separate concurrent process, and send each stream element using an intra-process communication mechanism.
-Developing such an implementation could be easy or hard, depending on what language features are available for concurrency.
-However, worrying about the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.
+To implement this, we might write each operator as its own concurrent process, sending stream elements over intra-process channels.
+Developing this could be easy or hard, depending on the available language features for concurrency.
+However, the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk stream data to reduce communication overhead, is invariably a headache.
 
-We might instead define some sort of uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
-Each operator could be given this interface, so that the next value from each result stream is computed on demand.
-This approach is commonly taken with implementations of physical operators in database systems \cite{Graefe:Volcano}.
-However, the `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering.
+We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
+Each operator could be given this interface, so that values from each stream are computed on demand.
+This approach is commonly taken in database systems \cite{Graefe:Volcano}.
+However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering.
 Suppose a consumer pulls many elements from the @sUnique@ output stream.
@@
 
-Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requiring unbounded buffering.
+Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering.
 Sadly, existing stream fusion transformations cannot handle our example.
@@
 Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs.
-We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network.
+We refer to this as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network.
 Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators.
@@
 
-Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
-The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level.
-The polyhedral model is based around affine loops, which makes it difficult to support filter-like operators such as @group@ and @merge@.
+Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
+The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.
 
 In this paper we present Machine Fusion, a new approach.
-Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream.
-We view each operator as a process in a concurrent process network, and the control flow of each process as a simple state machine.
+Each operator is expressed as a restricted, sequential imperative program which \emph{pulls} from input streams, and \emph{pushes} to output streams.
+We view each operator as a process in a concurrent process network.
 Our fusion transform then \emph{sequentializes} the concurrent process network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers.
@@
 \begin{itemize}
-\item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
-\item an algorithm for fusing these processes, the first to support splits and joins (\S\ref{s:Fusion});
-\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. (\S\ref{s:Evaluation});
-\item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs}).
+\item a process calculus for infinite streaming programs (\S\ref{s:Processes});
+\item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
+\item benchmarks showing significant performance gains (\S\ref{s:Benchmarks});
+\item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).
 \end{itemize}

### 2017-05-01 2017mergingmerges/section/02-Processes.tex
attempt to fit in sigconf 12p. format
typesetting to cut down

@@
 
-When the predicate is true, the instruction @(push sOut1 v A3 [ l = v, f = F ])@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).
+% AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
+When the predicate is true, the instruction at label @A2@ executes:
+@(push sOut1 v A3 [ l = v, f = F ])@
+pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).
 
@@
 
-\begin{figure}
+\begin{figure*}
 
-\begin{center}
+\begin{minipage}{0.6\textwidth}
 \begin{alltt}
@@
 \end{code}
-\end{center}
-\vspace{1em}
-\includegraphics[scale=1.1]{figures/state-group.pdf}
+\end{minipage}
+\begin{minipage}{0.29\textwidth}
+\includegraphics[scale=1.0]{figures/state-group.pdf}
+\end{minipage}
+
 \caption{The group combinator}
 \label{fig:Process:Group}
-\end{figure}
+\end{figure*}
 
@@
 \subsection{Merging}
-\begin{figure}
+\begin{figure*}
 \begin{alltt}
@@
       , label:  B0
-                    , instrs: { B0 = pull sIn1  x1   B1 []
-                              , B1 = pull sIn2  x2   C0 []
-                              , C0 = case (x1 < x2)  D0 []  E0 []
-                              , D0 = push sOut2 x1   D1 []
-                              , D1 = drop sIn1       D2 []
-                              , D2 = pull sIn1  x1   C0 []
-                              , E0 = push sOut2 x2   E1 []
-                              , E1 = drop sIn2       E2 []
+      , instrs: { B0 = pull sIn1  x1   B1 []             , B1 = pull sIn2  x2   C0 []
+                , C0 = case (x1 < x2)  D0 []  E0 []      , D0 = push sOut2 x1   D1 []
+                , D1 = drop sIn1       D2 []             , D2 = pull sIn1  x1   C0 []
+                , E0 = push sOut2 x2   E1 []             , E1 = drop sIn2       E2 []
                 , E2 = pull sIn2 x2    C0 [] } }
 \end{code}
+% \begin{minipage}{0.29\textwidth}
+\includegraphics[scale=1.0]{figures/state-merge.pdf}
+% \includegraphics[scale=1.0,angle=-90,origin=c]{figures/state-merge.pdf}
+% \end{minipage}
 
-\medskip
-\includegraphics[scale=1.1]{figures/state-merge.pdf}
 \caption{The merge combinator}
 \label{fig:Process:Merge}
-\end{figure}
+\end{figure*}
 
@@
 
-We also associate each of the joint result states with information describing whether or not each source state machine has already pulled a value from each of its input streams.
+We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams.
 For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
@@
 
-\begin{figure}
+\begin{figure*}
 \includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
@@
 \label{fig:Fusion:Pulls}
-\end{figure}
+\end{figure*}
 
@@
 \subsubsection{Fusing Cases}
-\begin{figure}
+\begin{figure*}
 \includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
@@
 \label{fig:Fusion:Case}
-\end{figure}
+\end{figure*}
 
@@
 There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
-The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is the @group@ processes output stream, and @sOut2@ which is the @merge@ processes output stream.
+The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.
 
@@
 We have chosen a \emph{specific sequential order} in which to interleave the instructions that implement both source processes.
-As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
+As with Flow Fusion \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
 However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are.
@@
 To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
-The order in which pairs of processes are fused together does matter, as does the order in which the instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
+The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
 
@@
 %%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
-\begin{figure}
+\begin{figure*}
 \definecolor{groupc}{HTML}{308030}
@@
 \label{fig:Process:Fused}
-\end{figure}
+\end{figure*}
 

### 2017-05-01 2017mergingmerges/section/03-Definition.tex
typesetting to cut down

@@
 \item \emph{Injection} of a single value from a stream into a process, or a set of processes.
-Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.
+Each individual process only needs to accept the value when it is ready for it, and injection into a set of processes succeeds only when they \emph{all} accept it.
 
@@
 Unlike Kahn processes, pushing to a channel can also block.
-Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.
+Each consumer has a single element buffer, and pushing only succeeds when that buffer is empty.
 
@@
 
-Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not.
+Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not.
 When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.
 
-Each output channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process.
+Each channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process.
 The only intra-process communication is via channels and streams.
-Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes.
-This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to via the formal input and output streams.
+Our model is ``pure data flow'' as there are no side-channels between processes.
+This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where processes can communicate via asynchronous messages as well as channels.
 
@@
 The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
-The @injects@ form is similar, but operates on sets of processes rather than a single one.
+The @injects@ form is similar, operating on sets of processes.
 
-Rule (InjectValue) injects of a single value into a single process.
+Rule (InjectValue) injects a single value into a single process.
 The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process.
-The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.
+The $\InputState$ acts as a single element buffer, and must be empty (@none@) for injection to succeed.
 

### 2017-05-04 2017mergingmerges/section/03-Definition.tex
serious cutting

@@
 If an $\InputState$ is not specified we assume it is `none'.
+A network is a set of processes, able to communicate to each other.
 
@@
 
-All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap.
+Instructions include a $\Next$ field containing the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap.
 The list of update bindings is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.
@@
 \begin{enumerate}
-\item \emph{Injection} of a single value from a stream into a process, or a set of processes.
-Each individual process only needs to accept the value when it is ready for it, and injection into a set of processes succeeds only when they \emph{all} accept it.
+\item \emph{Injection} of a single value from a stream into a process, or a network.
+Each individual process only needs to accept the value when it is ready for it, and injection into a network succeeds only when they \emph{all} accept it.
 
-\item \emph{Advancing} a single process from one state to another. Advancing a set of processes suceeds when \emph{any} of the processes in the set can advance.
+\item \emph{Advancing} a single process from one state to another. Advancing a network suceeds when \emph{any} of the processes in the network can advance.
 
@@
 
-Execution of a process network is non-deterministic.
+Execution of a network is non-deterministic.
 At any moment several processes may be able to take a step, while others are blocked.
@@
 Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not.
-When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.
+When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the network to advance without requiring unbounded buffering.
 
@@
 The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
-The @injects@ form is similar, operating on sets of processes.
+The @injects@ form is similar, operating on a network.
 
@@
 
-Rule (InjectMany) attempts to inject a single value into a set of processes.
-We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of them.
-Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.
+Rule (InjectMany) attempts to inject a single value into a network. We use the single process judgment form to inject the value into all processes, which must succeed for all of them.
+% Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.
 
@@
 
-Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update that will copy this value into the variable $x$ named in the @pull@ instruction.
+Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update to copy this value into the variable $x$ in the @pull@ instruction.
 We use the syntax $us,x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$.
@@
 Rule (Advance) advances a single process.
-We lookup the current instruction pointed to by the processes @label@ and pass it, along with the current channel states and heap to the previous single instruction judgment.
-The update expressions @us@ that the single instruction judgment yields are first reduced to values before updating the heap.
+We look up the current instruction for the process @label@ and pass it, along with the channel states and heap, to the above single instruction judgment.
+The update expressions $us$ from the single instruction judgment are reduced to values before updating the heap.
 We use $(us \lhd bs)$ to replace bindings in $us$ with new ones from $bs$.
-The update expressions themselves are all pure, so the evaluation can safely be done in parallel (or in arbitrary order).
+As the update expressions are pure, the evaluation can be done in any order.
 
@@
 Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes.
-The first set of rules concerns feeding values to other processes within the same process network, while the second exchanges input and output values with the environment the process network is running in.
+The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.
 
-The statement $\ProcShake{ps}{a}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $a$. A process ``group'' is just a set of processes.
+The statement $\ProcShake{ps}{a}{ps'}$ reads ``the network $ps$ advances to the network $ps'$ yielding output action $a$''.
 
-%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology.
-%%% BL: Structurally it's just a set of processes, but I at some point I used "set" for setting bindings in the heap. If we change group -> network make sure to get all occurrences in the paper.
+Rule (ProcessInternal) allows an arbitrary process in the network to advance to a new state at any time, provided it does not yield an output action.
+This allows processes to perform internal computation.
+% , without needing to synchronize with the rest of the network.
 
-Rule (ProcessInternal) allows an arbitrary process in the group to advance to a new state at any time, provided it does not yield an output action.
-This allows processes to perform internal computation, without needing to synchronize with the rest of the group.
+Rule (ProcessPush) allows an arbitrary process in the network to advance to a new state while yielding an output action $(@push@~c~v)$.
+For this to succeed it must be possible to inject the output value $v$ into all processes that have channel $c$ as one of their inputs.
+% As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process.
+When any process in the network produces an output action we take that as the action of the whole network.
 
-Rule (ProcessPush) allows an arbitrary process in the group to advance to a new state while yielding an output action (@push@ c v).
-For this to happen it must be possible to inject the output value @v@ into all processes that have channel @c@ as one of their inputs.
-As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process.
-When any process in the group produces an output action we take that as the action of the whole group.
+The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, network $ps$ takes a step and produces new channel values $cvs'$ and network $ps'$''.
+The channel values $cvs$ map channel names to a list of values.
+For input channels of the overall network, we initialize the map to contain a list of input values for each channel.
+For output channels of the overall network, values pushed to those channels are also collected in the same channel map.
+In a concrete implementation the input and output values would be transported over some IO device, but for the semantics we describe the abstract behavior only.
 
-\smallskip
-The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, process group $ps$ takes a step and produces new channel values $cvs'$ and group $ps'$''.
-The channel values $cvs$ map channel names to a list of values.
-For input channels of the overall group, we initialize the map to contain a list of input values for each channel.
-For output channels of the overall group, values pushed to those channels are also collected in the same channel map.
-In a concrete implementation the input and output values would be transported over some IO device, but for the semantics here is sufficient to describe the abstract behavior of our system.
+Rule (FeedInternal) allows the network to perform local computation in the context of the channel values.
 
-Rule (FeedInternal) allows the process group to perform local computation in the context of the channel values.
-
-Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a process group and appends the value $v$ to the list corresponding to the output channel $c$.
+Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a network and appends the value $v$ to the list corresponding to the output channel $c$.
 
@@
 This rule also has the side condition that values cannot be injected from the environment into output channels that are already owned by some process.
-This constraint is used during formal proofs of correctness, but would not need to be checked dynamically in a concrete implementation.
-The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.
+This constraint is required for correctness proofs, but can be ensured by construction in a concrete implementation.
+% The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.
 
@@
 \label{s:EvaluationOrder}
-%%% AR: Can we simplify this?
-% Maybe we can start by talking about the different evaluation orders for just two alt2s:
-%
-% let s1 = alt2 sInA sInB
-% let s2 = alt2 sInA sInC
-%
-% Here we have many evaluation orders. If we just look at the order in which inputs sInA etc are pulled, some options are:
-% [ A, A, B, C, B, C ]
-% [ A, A, B, B, C, C ]
-% [ A, A, C, C, B, B ]
-%
-% Now if we add a new operator which zips the two together, it actually restricts the pulling order
-%
-% let s1 = alt2 sInA sInB
-% let s2 = alt2 sInA sInC
-% let s3 = zip  s1 s2
-%
-% This restricts the pulling order to:
-% [ A, A, B, C, B, C ]
-%
-%%%
-% Or, maybe we don't even need the zip here, and can leave the zip until S5.1.
-%%%
 
@@
 \ref{fig:Process:Eval:Feed} are non-deterministic in several ways.
-Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the group.
-More importantly, (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the group are ready to accept the pushed value.
-Rule (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the associated channel are ready to accept the value.
+Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.
 
 In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
+In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.
+In contrast, the role of our fusion transform is to construct one of these working orders statically.
+In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network were executed dynamically.
+In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.
+
+% In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
 Consider the following example, where the @alt2@ operator pulls two elements from its first input stream, then two from the second, before pushing all four to its output stream.
-\begin{code}
-  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
-  alternates sInA sInB sInC
-   = let  s1   = alt2 sInA sInB
-          s2   = alt2 sInB sInC
-          sOut = zip s1 s2
-     in   sOut
-\end{code}
-
-Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples.
+% \begin{code}
+%   alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
+%   alternates sInA sInB sInC
+%    = let  s1   = alt2 sInA sInB
+%           s2   = alt2 sInB sInC
+%           sOut = zip s1 s2
+%      in   sOut
+% \end{code}
+% 
+% Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples.
 Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.
-
-Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time.
+% 
+% Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time.
 However, the downstream @zip@ process needs to pull one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams.
@@
 Here is a table of intermediate values to help make the explanation clearer:
-
-\begin{code}
-    sInA = [a1, a2, a3, a4, a5 ...]
-    sInB = [b1, b2, b3, b4, b5 ...]
-    sInC = [c1, c2, c3, c4, c5 ...]
-
-    s1   = alt2 sInA sInB 
-         = [a1, a2, b1, b2, a3, a4, b3, b4 ...]
-
-    s2   = alt2 sInB sInC
-         = [b1, b2, c1, c2, b3, b4, c3, c4 ...]
-
-    sOut = zip s1 s2
-         = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
-\end{code}
-
-Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on.
+% 
+% \begin{code}
+%     sInA = [a1, a2, a3, a4, a5 ...]
+%     sInB = [b1, b2, b3, b4, b5 ...]
+%     sInC = [c1, c2, c3, c4, c5 ...]
+% 
+%     s1   = alt2 sInA sInB 
+%          = [a1, a2, b1, b2, a3, a4, b3, b4 ...]
+% 
+%     s2   = alt2 sInB sInC
+%          = [b1, b2, c1, c2, b3, b4, c3, c4 ...]
+% 
+%     sOut = zip s1 s2
+%          = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
+% \end{code}
+% 
+% Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on.
 The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it.
 While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@.
-The execution order of the overall process group is constrained by communication patterns of processes in that group.
+The execution order of the overall network is constrained by communication patterns of processes in that network.
 
-As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings.
+% As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings.
 In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.
 In contrast, the role of our fusion transform is to construct one of these working orders statically.
-In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the process group was executed dynamically.
+In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network was executed dynamically.
 In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.
@@
 We need to start with zip so that the order in which input elems arrive is constrained.
-
-% Rule (FeedPush) allows the process network to emit a push message.
-As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message.
-The pushed value is collected in the accumulator list for that stream.
-
-% Rule (FeedExternal) allows inputs to be injected into the process network.
-% For any channel $c$ which is not an output of one of the processes, we take the last value off its list.
-The recursive feed is evaluated with the last value removed from the accumulators.
-The last value is then injected into the network, and added back to the result accumulators.
-
-% Rule (FeedStart) applies when all input values have been injected and there are no input values left. In this case, the output values are the same as the input values.
-
-% Evaluation: feeding evaluates a process network on a list of input values and collects the outputs.
-Feeding alternates between injecting input values and shaking the processes to collect the output messages.
-
-% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
-% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.
-
-% -- cuts ---------------------------------------------------------------------
-% BL: Discuss this during def of evaluation.
-
-% These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.
-
-% Claim "functional dataflow" or some such.
-Each process only updates values in its local heap, the only intra-process communication is via streams.
-This is unlike StreamIt which can send out-of-band messages between its processes.
-
-% The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process.
-This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order.
-A process may, however, produce multiple output streams.
-
-% Each process has its own private heap, therefore the only communication between processes occurs by streams.
-
-% The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction.
-Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.
-
-% After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.
-
-% A process network is a set of multiple processes that can be evaluated concurrently.
-Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment.
-Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language.
-By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

### 2017-05-04 2017mergingmerges/section/04-Fusion.tex
serious cutting

@@
 
-The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process group we fuse successive pairs of processes until only one remains.
+The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.
 
 Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion.
-We extend the exiting $Label$ grammar with an additional alternative, $Label_S \times Label_S$ which we use for the labels in a fused result process.
-Each $Label_S$ consists of a $Label$ from one of the  source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$.
+We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process.
+Each $Label_F$ consists of a $Label$ from one of the source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$.
 The definition of $Label$ is now recursive.
-While fusing a whole process group, as we fuse pairs of individual processes the labels in the result collect more and more information.
+While fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information.
 Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.
 
-% These new labels $Label_S$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
+% These new labels $Label_F$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
 
-% If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, do not know the actual value.
+% If the static $\InputState_F$ is $@pending@_F$, there is a value waiting to be pulled, do not know the actual value.
 
-We also extend the existing $Variable$ grammar with a (@chan@ c) form which represents the stream buffer variable associated with channel @c@.
-We only need one stream buffer variable for each channel, and naming them like this saves us from inventing fresh variable names in the definition of the fusion rules (we used a fresh name @b1@ back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion)
+We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with channel @c@.
+We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
+(We used a fresh name back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion.)
 
@@
 Type @in1@ indicates that only a single process pulls from the channel.
-Type @in1out1@ indicates that one process pushes to the channel and the other pulls from the same.
+Type @in1out1@ indicates that one process pushes to the channel and the other pulls.
 Type @out1@ indicates that the channel is pushed to by a single process.
@@
 The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states.
-As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no more states can be discovered.
+As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no new states can be discovered.
 
@@
 Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@.
-This heuristic helps to collect @jump@ instructions together in the result process, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle.
+This helps collect @jump@ instructions together, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle.
 As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.
 
-Clauses (DeferPull1) and (DeferPull2) are similar heuristics, which defer @pull@ instructions if possible.
-We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
+Similarly, clauses (DeferPull1) and (DeferPull2) defer @pull@ instructions. We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
 
@@
 
-Clause (LocalCase) is similar, except that there are two destination labels.
+Clause (LocalCase) is similar, except there are two $\Next$ labels.
 
@@
 Pushing to a shared channel requires the downstream process to be ready to accept the value at the same time.
-We encode this constraint by requiring the static input state of the downstream channel to be $@none@_S$.
-When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $@chan@~c$'' and sets the static input state to ``$@pending@_S$'', which indicates that the new value is now available.
+We encode this constraint by requiring the static input state of the downstream channel to be $@none@_F$.
+When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $(@chan@~c)$ and sets the static input state to $@pending@_F$, which indicates that the new value is now available.
 
@@
 Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to.
-We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$.
-The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_S$ channel state in the result label records that it has done so.
+We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_F$.
+The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_F$ channel state in the result label records that it has done so.
 
@@
 
-Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop a value it gained from a channel which is shared by the other process.
+Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop from a channel shared by both processes.
 In (SharedDropOne) the channel states reveal that the other process is still using the value.
-In this case the result instruction is a @jump@ that only updates the static information that the left source process no longer requires this value.
-In (SharedDropBoth) then channel states reveal that the other process no longer needs the value either.
-In this case the result instruction is a real @drop@, because we are sure that neither process requires the value any longer.
+In this case the result is a @jump@ updating the channel state to note that the left process has dropped.
+In (SharedDropBoth) the channel states reveal that the other process no longer needs the value.
+In this case the result is a real @drop@, because we are sure that neither process requires the value any longer.
 
@@
 \smallskip
-Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which we have already mentioned.
+Fig.~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
 Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.

### 2017-05-04 2017mergingmerges/section/05-Evaluation.tex
attempt to fit in sigconf 12p. format
serious cutting

@@
 
-\eject{}
 % -----------------------------------------------------------------------------
-\section{Fusion in Practice}
+\section{Results}
 \label{s:Evaluation}
 
-Stream fusion is ultimately performed for practical reasons.
-We want the fused result program to run faster than unfused source program.
-Fused result programs like the one in Fig.
-\ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell (maybe using Template Haskell), C or LLVM abstract assembly.
+Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program.
+
+\subsection{Benchmarks}
+We have implemented this system using Template Haskell in a library called @folderol@\footnote{\url{https://github.com/amosr/folderol}}.
+To show practical examples, we use the finite stream extension mentioned in~\S\ref{s:Finite}.
+We present three benchmarks: one array algorithm, and two simple file operations.
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[
+  symbolic y coords={Store, Recompute, Ours, Hand, end},
+	xlabel=Runtime (ms),
+  xmin=0, xmax=200,
+	enlargelimits=0.01,
+	xbar interval=0.7,
+  width=7.5cm, height=4.5cm,
+	legend pos=north west,
+]
+\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
+\end{axis}
+\end{tikzpicture}
+\caption{Runtime for Quickhull; lower is faster.}
+\label{fig:bench:quickhull}
+\end{figure}
+
+Quickhull is a divide-and-conquer spatial algorithm to find the smallest convex hull containing all points.
+At its core is an operation called `filterMax' which takes a line and an array of points, and finds the farthest point above the line, as well as all points above the line.
+
+We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
+The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
+This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
+We compared both, and recomputing was significantly faster.
+However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.
+
+Fig.~\ref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
+The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
+The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
+It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[
+  symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
+	xlabel=Runtime (ms),
+  xmin=0, xmax=290,
+	enlargelimits=0.01,
+	xbar interval=0.7,
+  width=7.5cm, height=4.5cm,
+	legend pos=north west,
+]
+\addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
+\end{axis}
+\end{tikzpicture}
+\caption{Runtime for append2; lower is faster.}
+\label{fig:bench:append}
+\end{figure}
+
+The first file benchmark simply appends two files, while counting the number of lines.
+We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
+Fig.~\ref{fig:bench:append} shows the runtimes for appending 2MB of data.
+The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[
+  symbolic y coords={Streaming, Pipes, Conduit hand, Ours, end},
+	xlabel=Runtime (ms),
+  xmin=0, xmax=1200,
+	enlargelimits=0.01,
+	xbar interval=0.7,
+  width=7.5cm, height=4.5cm,
+	legend pos=north west,
+]
+\addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
+\end{axis}
+\end{tikzpicture}
+\caption{Runtime for partition; lower is faster.}
+\label{fig:bench:part}
+\end{figure}
+
+The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
+The output lines are also counted.
+The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or partially hand-fused.
+Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
+Fig.~\ref{fig:bench:part} shows the runtimes for partitioning a 6MB file.
+
+% Fused result programs like the one in Fig.
+\ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly.
 The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.
@@
 When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process.
-Consider our @alternatives@ example from \S\ref{s:EvaluationOrder}, here it is again:
+Consider our @alternatives@ example from \S\ref{s:EvaluationOrder} again:
+
 \begin{code}
@@
 
-In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, then this final fusion transform fails.
+In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, this final fusion transform fails.
 This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
-As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.
+% As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.
 
@@
 Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails.
-Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.\ref{fig:Fusion:Def:StepPair}.
-The $tryStepPair$ function determines which of the instructions from either process can be executed next, and when execution is deadlocked there are none.
+Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.~\ref{fig:Fusion:Def:StepPair}.
+The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.
 
@@
 % -----------------------------------------------------------------------------
-\subsection{Result Size}
-
-As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
+%\subsection{Result Size}
+%
+%As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
 The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
@@
 In all combinations, with up to 7 processes there are less than 100 states in the result process.
-%%% AR: Note that the split version also has only one merge?
-
-The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes.
+%%%% AR: Note that the split version also has only one merge?
+%
+%The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes.
 This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
-
-On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
+%
+%On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
 The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream.
@@
 In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
-
-When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product.
+%
+%When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product.
 During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition.
 The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result.
-
-In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
+%
+%In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
 If we assume the two processes have about the same number of states then this is O($n^2$).
 Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.
-
-In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction.
+%
+%In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction.
 During fusion, the states for the consuming process are inlined at each occurrence of @push@.
@@
 After we have fused two processes together, it may be possible to simplify the result before fusing in a third.
-Consider the result of fusing @group@ and @merge@ which we saw back in Fig.\ref{fig:Process:Fused}.
-At labels @F1@ and @F2@ are @jump@ instructions, where the first passes control to the second.
-The update expressions attached to these instructions are also non-interferring, which means we can safely combine these consecutive @jump@ instructions into a single @jump@ to @F4@.
+Consider the result of fusing @group@ and @merge@ which we saw back in Fig.~\ref{fig:Process:Fused}.
+At labels @F1@ and @F2@ are two consecutive @jump@ instructions.
+The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@.
 In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code.
-The (PreferJump) clauses of Fig.\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
+The (PreferJump) clauses of Fig.~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
 
-There are also @jump@ instructions like the one at @F5@ which have no associated update expressions, and thus can be eliminated completely.
+Other @jump@ instructions like the one at @F5@ have no associated update expressions, and thus can be eliminated completely.
 Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.
 
-Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has lower number of states.
-Provided we do not change the order of instructions that do not require synchronization with other processes, meaning @pull@ @push@ or @drop@, the fusibility of the overall process network will not be affected.
+Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has fewer states.
+Provided we do not change the order of instructions that require synchronization with other processes (@pull@, @push@ or @drop@), the fusibility of the overall process network will not be affected.
 
 Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
-In Fig.\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
+In Fig.~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
 Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable.
@@
 When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
-This is the case in Fig.\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
+This is the case in Fig.~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
 In the result, the interleaved instructions from both source processes then share the same heap variable.
@@
 % -----------------------------------------------------------------------------
-\begin{figure}
-\begin{minipage}[t]{0.10\textwidth}
-\end{minipage}
-\begin{minipage}[t]{0.4\textwidth}
-\begin{tikzpicture}
-\begin{axis}[
-% Hide the label on the second graph
-	ylabel=Process states,
-	xlabel=Pipelined combinators,
-  ymin=0, ymax=100,
-	enlargelimits=0.01,
-	ybar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-\addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
-\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
-\legend{With merge, Without merge};
-\end{axis}
-\end{tikzpicture}
-\end{minipage}
-\quad
-\quad
-\quad
-\begin{minipage}[t]{0.4\textwidth}
-\begin{tikzpicture}
-\begin{axis}[
+% \begin{figure}
+% \begin{minipage}[t]{0.10\textwidth}
+% \end{minipage}
+% \begin{minipage}[t]{0.4\textwidth}
+% \begin{tikzpicture}
+% \begin{axis}[
+% % Hide the label on the second graph
 % 	ylabel=Process states,
-        xlabel=Parallel combinators,
-  ymin=0, ymax=100,
-        enlargelimits=0.01,
-        ybar interval=0.7,
-  width=7.5cm, height=4.5cm,
-        legend pos=north west,
-]
-\addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
-  % Last bar doesn't show for some reason, so need to add a dummy value for the next one
-    (8,1) };
-
-\addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
-
-\legend{With merge, Without merge};
-\end{axis}
-\end{tikzpicture}
-\end{minipage}
-
-\caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
-\label{fig:bench:outputsize}
-\end{figure}
-
-% -----------------------------------------------------------------------------
-\begin{figure}
-
-\begin{tikzpicture}
-\begin{axis}[
-	ylabel=Process states,
-	xlabel=Number of merges,
-%  ymode=log,
-  ymin=0, ymax=1500,
-	enlargelimits=0.01,
-	ybar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-% These are the values for splitting.
-% They are smaller than the 'chaining', but look much nicer on the linear graph.
-\addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
-  % (8,2880)
-  (8,1)
-  };
-
-% These are the values for chaining
-% \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };
-
-\end{axis}
-\end{tikzpicture}
-
-\caption{Exponential blowup occurs when splitting or chaining merges together.}
-\label{fig:bench:exponential}
-\end{figure}
+% 	xlabel=Pipelined combinators,
+%   ymin=0, ymax=100,
+% 	enlargelimits=0.01,
+% 	ybar interval=0.7,
+%   width=7.5cm, height=4.5cm,
+% 	legend pos=north west,
+% ]
+% \addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
+% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
+% \legend{With merge, Without merge};
+% \end{axis}
+% \end{tikzpicture}
+% \end{minipage}
+% \quad
+% \quad
+% \quad
+% \begin{minipage}[t]{0.4\textwidth}
+% \begin{tikzpicture}
+% \begin{axis}[
+% %       ylabel=Process states,
+%         xlabel=Parallel combinators,
+%   ymin=0, ymax=100,
+%         enlargelimits=0.01,
+%         ybar interval=0.7,
+%   width=7.5cm, height=4.5cm,
+%         legend pos=north west,
+% ]
+% \addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
+%   % Last bar doesn't show for some reason, so need to add a dummy value for the next one
+%     (8,1) };
+% 
+% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
+% 
+% \legend{With merge, Without merge};
+% \end{axis}
+% \end{tikzpicture}
+% \end{minipage}
+% 
+% \caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
+% \label{fig:bench:outputsize}
+% \end{figure}
+% 
+% 
+% % -----------------------------------------------------------------------------
+% \begin{figure}
+% 
+% \begin{tikzpicture}
+% \begin{axis}[
+% 	ylabel=Process states,
+% 	xlabel=Number of merges,
+% %  ymode=log,
+%   ymin=0, ymax=1500,
+% 	enlargelimits=0.01,
+% 	ybar interval=0.7,
+%   width=7.5cm, height=4.5cm,
+% 	legend pos=north west,
+% ]
+% % These are the values for splitting.
+% % They are smaller than the 'chaining', but look much nicer on the linear graph.
+% \addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
+%   % (8,2880)
+%   (8,1)
+%   };
+% 
+% % These are the values for chaining
+% % \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };
+% 
+% \end{axis}
+% \end{tikzpicture}
+% 
+% \caption{Exponential blowup occurs when splitting or chaining merges together.}
+% \label{fig:bench:exponential}
+% \end{figure}

### 2017-05-04 2017mergingmerges/section/XX-FutureWork.tex
notes
typesetting to cut down
serious cutting

@@
 
-% We now discuss some of the shortcomings of the system, and extensions and future work to ameliorate this.
+%%% AR: Can we fit a few sentences here just to conclude, say we've described a fusion algorithm, with promising future work too
 
@@
 
-The fusion algorithm treats all @case@ conditions as fully abstract, by exploring all possible combinations for both processes.
+The fusion algorithm treats all @case@ conditions as fully abstract by exploring all possible combinations for both processes.
 This can cause issues for processes that dynamically require only a bounded buffer, but the fusion algorithm statically tries every combination and wrongly asserts that unbounded buffering is required.
 
-For example, suppose two processes have the same case condition (@x > 0@).
-The fusion algorithm will generate all four possibilities, including the contradictory ones such as where the first process has (@x > 0 = true@) and the second has (@x > 0 = false@).
-If any require an unbounded buffer, the fusion algorithm will fail.
+For example, if two processes have the same case condition (@x > 0@), the fusion algorithm generates all four possibilities, including contradictory ones where one process is true and the other is false.
+If any require an unbounded buffer, the fusion algorithm fails.
 
-A possible extension is to somehow cull these contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
-It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit some kind of failure instruction rather than failing to produce a process.
-A separate postprocessing step could then perform analyses and remove statically unreachable process states.
+One solution may be to cull contradictory states so that if it is statically known that a state is unreachable, it does not matter if it requires unbounded buffers.
+It may be possible to achieve this with relatively little change to the fusion algorithm itself, by having it emit a failure instruction rather than failing to produce a process.
+A separate postprocessing step could then remove statically unreachable process states.
 After postprocessing, if any failure instructions are reachable, fusion fails as before.
@@
 Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output.
-We have focussed on the infinite stream version because it is simpler to explain and prove, but the extensions required to support finite streams do not require substantial conceptual changes.
+We have focussed on the infinite stream version as it is simpler to explain and prove, but supporting finite streams does not require substantial conceptual changes.
 
@@
 
-For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel.
+% For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel.
 After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow.
-We have implemented an early prototype that supports finite streams, our mechanized formalization does not cover it.
+Our implementation supports finite streams, but our mechanized formalization does not cover it.
 
@@
 However, this requires the fusion transform to analyse and reason about case statements and their predicates.
-It seems obvious that if two consumers read the same value and check if it is EOF, they are both going to give the same result, but this is of course undecidable in general.
-By making the structure of finite streams explicit and constraining processes to use finite streams in particular ways, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
+It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
+By making the structure of finite streams explicit and constraining how processes use finite streams, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
 

### 2017-05-04 2017mergingmerges/section/XX-Proof.tex
attempt to fit in sigconf 12p. format
serious cutting

@@
 Our fusion system is formalized in Coq, and we have proved soundness of \ti{fusePair}: if the fused result process produces a particular sequence of values on its output channels then the two source processes may also produce that same sequence.
-Note that due to non-determinism of process execution the converse is not true in practice: just because the two concurrent processes can produce a particular output sequence does not mean the fused result process will as well --- the fused result process uses only one of the many possible orders.
+The converse is not true, however: concurrent evaluation represents all possible interleavings, of which the fused process is only one.
 
@@
 \begin{code}
-Theorem Soundness (P1 : Program L1 C V1) (P2 : Program L2 C V2) (ss : Streams) (h : Heap)
-                  (l1 : L1) (is1 : InputStates) (l2 : L2) (is2 : InputStates) 
+Theorem Soundness (P1 : Program L1 C V1)
+                  (P2 : Program L2 C V2)
+                  (ss : Streams) (h : Heap)
+                  (l1 : L1) (is1 : InputStates)
+                  (l2 : L2) (is2 : InputStates) 
   :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
-  -> EvalOriginal Var1 P1 P2 is1 ss h l1 /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
+  -> EvalOriginal Var1 P1 P2 is1 ss h l1
+  /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
 \end{code}
@@
 Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition.
-Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.\ref{fig:Process:Eval:Feed}.
+Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Fig.~\ref{fig:Process:Eval:Feed}.
 Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process.

### 2017-05-04 2017mergingmerges/section/XX-Related.tex
initial benchmarks
serious cutting

@@
 \section{Related work}
+
+\TODO{Cite meta-repa \cite{ankner2013edsl}}
 
@@
 
-Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
+Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
 For effectful inputs such as reading from the network, duplicating work changes the semantics.
@@
 
-One way to address the difference between pull and push streams is to explicitly support both separately, as seen in \citet{bernardy2015duality} and \citet{lippmeier2016polarized}.
-Here, pull streams have the type @Source@ and represent a source that is always available to be pulled from, while push streams have the type @Sink@ and represent a sink that can always accept input.
-Both systems rely on stream bindings being used linearly to ensure correctness, including boundedness of buffers.
+One way to address the difference between pull and push streams is to explicitly support both \cite{bernardy2015duality, lippmeier2016polarized}.
+Here, pull streams have the type @Source@ and represent a source, always ready to be pulled, while push streams have the type @Sink@ and represent a sink, always ready to accept data.
+% Both systems rely on stream bindings being used linearly to ensure correctness.
 Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
-Sinks, however, are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
-It has the type @Sink a -> Sink b -> Sink (a,b)@.
-This system requires the streaming computation to be manually split into sources and sinks, and be joined together by a loop that `drains' values from the source and pushes them into the sink.
+Sinks are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
+% It has the type @Sink a -> Sink b -> Sink (a,b)@.
+This system requires the streaming computation to be manually split into sources and sinks.
+% and be joined together by a loop that `drains' values from the source and pushes them into the sink.
 
-The duality between pull and push arrays has also been explored in Obsidian by \citep{claessen2012expressive} and later in \citep{svensson2014defunctionalizing}.
+The duality between pull and push arrays has also been explored in Obsidian \cite{claessen2012expressive, svensson2014defunctionalizing}.
 Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
@@
 When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
-Our system can be seen as an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.
+Our system is an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.
 
@@
 
-Synchronous dataflow (not to be confused with synchronous languages above) is a dataflow graph model of computation where each dataflow actor has constant, statically known input and output rates.
-The main advantage of synchronous dataflow is that it is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
+Synchronous dataflow (SDF; not to be confused with synchronous languages above) is a dataflow graph model of computation where each node has constant, statically known input and output rates.
+SDF is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
 StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
-Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend synchronous dataflow with boolean and integer valued control ports, and attempt to recover the structure of ifs and loops from select and switch actors.
-These systems allow some dynamic structures to be scheduled statically, but are very rigid and only support limited control flow structures: it is unclear how merge or append could be scheduled by this system.
-Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is still quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
-A finite state machine is constructed, where each node of the FSM denotes its own synchronous dataflow graph.
+Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow structures.
+Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
+A finite state machine is constructed, where each node of the FSM denotes its own SDF graph.
 The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph.
 For example, a filter is represented with two nodes in the FSM.
-The dataflow graph for the initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
-The dataflow graph for the next state emits the value, and moves back to the first state.
-This does appear to be able to express value-dependent operations such as merge, but lacks the composability - and familiarity - of combinators.
+The initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state.
+The next state emits the value, and returns to the start.
+This appears to be able to express value-dependent operations such as merge, but lacks the composability and familiarity of combinators.
 

### 2017-05-06 2017mergingmerges/section/01-Introduction.tex
serious cutting
Make intro fit on the first page

@@
 
-We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
+% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
 Each operator could be given this interface, so that values from each stream are computed on demand.
@@
 Sadly, existing stream fusion transformations cannot handle our example.
+
 As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations.
@@
 We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.
-
 Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs.
@@
 
-Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
-The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.
+% Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
+
+% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.
 
@@
 \item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
-\item benchmarks showing significant performance gains (\S\ref{s:Benchmarks});
+\item benchmarks showing significant performance gains (\S\ref{s:Evaluation});
 \item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).

### 2017-05-06 2017mergingmerges/section/01-Introduction.tex
Edits to S2, get state diag and fused result on same page

@@
 Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
+
 Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once.
@@
 
-Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering.
+Instead, we would prefer to use \emph{stream fusion}, which is a program transformation that takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering.
 Sadly, existing stream fusion transformations cannot handle our example.

### 2017-05-06 2017mergingmerges/section/02-Processes.tex
Edits to S2, get state diag and fused result on same page

@@
 % -----------------------------------------------------------------------------
-\section{Processes, Machines, Combinators and Operators}
+\section{Processes and Machines}
 \label{s:Processes}
@@
 
-The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Fig.~\ref{fig:Process:Group}.
+The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Figure~\ref{fig:Process:Group}.
 The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
@@
 The body of the combinator is a record that defines the process.
-The @ins@ field of the record defines the set of input streams and the @outs@ field the set of output streams.
+The @ins@ field defines the set of input streams and the @outs@ field the set of output streams.
 The @heap@ field gives the initial values of each of the local variables.
 The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
-Note that in this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the source representation of @uniquesUnion@ from \S\ref{s:Introduction}.
+In this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.
 
@@
 @(push sOut1 v A3 [ l = v, f = F ])@
-pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).
+which pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).
 
 Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
-This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
+
+% TODO: Talk about what drop is really for later.
+% This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
 The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.
@@
 
-The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}.
+The definition of the @merge@ combinator, which merges two input streams, is given in Figure~\ref{fig:Process:Merge}.
 The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@.
@@
 
-As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
+% TODO: Shift this later.
+% As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
 For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output.
@@
 The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process.
-For ease of reference, when describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.
+When describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.
 
@@
 The algorithm proceeds by considering pairs of states: one in each of the source process state machines to be fused.
-Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Fig.~\ref{fig:Fusion:Pulls}.
+Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Figure~\ref{fig:Fusion:Pulls}.
 The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
@@
 We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams.
-For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
+For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
 The result state @F0@ represents a combination of the two source states @A0@ and @B0@.
@@
 The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.
-In Fig.~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.
 
-\begin{figure*}
-\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
-\caption{Fusing pull instructions}
-\label{fig:Fusion:Pulls}
-\end{figure*}
+% BL: Dropped this to fix pagination.
+% In Figure~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.
 
@@
 \subsubsection{Fusing Cases}
-\begin{figure*}
-\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
-\caption{Fusing case instructions}
-\label{fig:Fusion:Case}
-\end{figure*}
-
 Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively.
-The left of Fig.~\ref{fig:Fusion:Case} shows the next few transitions of the source machines.
+The left of Figure~\ref{fig:Fusion:Case} shows the next few transitions of the source machines.
 From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input.
@@
 From state @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
-
 Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
@@
 
+% -----------------------------------------------------------------------------
+\subsection{Fused Result}
+
+% TODO: Blend this in.
 Note that we could construct the fused result machine in several ways.
@@
 
-% -----------------------------------------------------------------------------
-\subsection{Fused Result}
-
-Fig.~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together.
+Figure~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together.
 There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
@@
 
+% -----------------------------------------------------------------------------
+\begin{figure*}
+\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
+
+\vspace{2em}
+
+\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
+\caption{Fusing pull (top) and case (bottom) instructions}
+\label{fig:Fusion:Pulls}
+\label{fig:Fusion:Case}
+\end{figure*}
+
+% BL: Merged these figures togather to save space.
+% \begin{figure*}
+% \caption{Fusing case instructions}
+% \label{fig:Fusion:Case}
+% \end{figure*}
+
 %%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
@@
 \definecolor{sharec}{HTML}{000080}
-
-\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}
-\begin{alltt}
-process
-\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
-, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
-, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
-, label:  \tctt{sharec}{F0}
-\end{alltt}
 
@@
 }
+
 \newcommand\instr[5]{
@@
 }
+
+\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}
+
+\hspace{5em}
+
+\begin{adjustbox}{minipage=0.8\textwidth,margin=0pt \smallskipamount,center}
+\begin{alltt}
+process
+\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
+, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
+, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
+, label:  \tctt{sharec}{F0}
+\end{alltt}
 
@@
 \end{tabbing}
-
+\end{adjustbox}
 \caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
@@
 \end{figure*}
+% -----------------------------------------------------------------------------
 

### 2017-05-06 2017mergingmerges/section/02-Processes.tex
Add a new section Breaking it Down to give a high level overview of how the system works

@@
 
+Figure~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together.
+There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
+The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.
+
+To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
+The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
+Although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere.
+When we instantiated combinators to create the original source processes we introduced fresh names at that point.
+
+% The stream buffer variables we additionally introduced for coordination were freshly created during fusion.
+
+\subsection{Breaking It Down}
+Overall, we started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
+Importantly, we needed to ``break down'' the definition of each of the operators into imperative statements so that we could interleave their execution appropriately.
+We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way.
+Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@.
+Instead, in the fused result program we have effectively performed the job of a concurrent scheduler at compile time.
+The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.
+
+The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion \cite{lippmeier2013data}.
+However, in contrast to Flow Fusion and similar systems, with Machine Fusion we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are.
+This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.
+
+Note that trying to use lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem.
+Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists.
+As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirity before demanding any of the elements from the second list (@sUnion@).
+If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before remanding any of @sIn2@, causing a space leak.
+Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer.
+The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers are not at liberty to defer handling of these values.
+
 % TODO: Blend this in.
-Note that we could construct the fused result machine in several ways.
+% Note that we could construct the fused result machine in several ways.
 One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch.
@@
 For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.
-
-Figure~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together.
-There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
-The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.
-
-Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequential result process that implements both.
-We have chosen a \emph{specific sequential order} in which to interleave the instructions that implement both source processes.
-As with Flow Fusion \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time.
-However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are.
-This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.
-
-To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
-The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
-
-Finally, although the result process has a single shared heap, the bindings are guaranteed not to interfere.
-When we instantiated combinators to create the original source processes we introduced fresh names at that point.
-The stream buffer variables we additionally introduced for coordination were freshly created during fusion.
 

### 2017-05-06 2017mergingmerges/section/02-Processes.tex
Fix sigma -> bs in advance rules

@@
 \subsection{Breaking It Down}
-Overall, we started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
-Importantly, we needed to ``break down'' the definition of each of the operators into imperative statements so that we could interleave their execution appropriately.
+We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
+Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately.
 We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way.
 Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@.
-Instead, in the fused result program we have effectively performed the job of a concurrent scheduler at compile time.
+Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time.
 The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.
 
-The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion \cite{lippmeier2013data}.
-However, in contrast to Flow Fusion and similar systems, with Machine Fusion we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are.
+The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion~\cite{lippmeier2013data}.
+However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are.
 This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.

### 2017-05-06 2017mergingmerges/section/03-Definition.tex
Fix sigma -> bs in advance rules

@@
 
-The grammar for process definitions is given in Fig.~\ref{fig:Process:Def}.
+The formal grammar for process definitions is given in Figure~\ref{fig:Process:Def}.
 Variables, Channels and Labels are specified by unique names.
@@
 If an $\InputState$ is not specified we assume it is `none'.
-A network is a set of processes, able to communicate to each other.
+A network is a set of processes that are able to communicate with each other.
 
-In the grammar of Fig.~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable.
+In the grammar of Figure~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable.
 The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes.
 When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state.
-When defining the fusion transform in \S\ref{s:Fusion} we will used a version of $\InputState$ with only this staticaly known information.
+When defining the fusion transform in \S\ref{s:Fusion} we will use a version of $\InputState$ with only this staticaly known information.
 
@@
 When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions.
-The @drop@s instructions are used to control how processes should be synchronized, but do not affect the execution of a single process.
-We will discuss @drop@s further in \S\ref{s:Optimisation}.
+The @drop@ instructions are used to control how processes should be synchronized, but do not affect the execution of a single process.
+We discuss @drop@s further in \S\ref{s:Optimisation}.
 
@@
 
+% TODO BL: Mention what happens if we choose a different ordering, 
+% and how the particular ordering chosen is decided upon.
 Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not.
@@
 \subsubsection{Injection}
-Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
+Figure~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
 The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
@@
 \subsubsection{Advancing}
-Fig.~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process.
+Figure~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process.
 The first set of rules handle specific instructions.
@@
 \subsubsection{Feeding}
-Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes.
+Figure~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes.
 The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.
@@
 
-The execution rules of Fig.
-\ref{fig:Process:Eval:Feed} are non-deterministic in several ways.
+The execution rules of Figure~\ref{fig:Process:Eval:Feed} are non-deterministic in several ways.
 Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.

### 2017-05-06 2017mergingmerges/section/05-Evaluation.tex
Suppress two of the diags to work out pagination, data should be moved to a single diagram

@@
 Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program.
+
+% -----------------------------------------------------------------------------
+\subsection{Finite streams}
+\label{s:Finite}
+
+The processes we have seen so far deal with infinite streams, but in practice most streams are finite.
+Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output.
+We have focussed on the infinite stream version as it is simpler to explain and prove, but supporting finite streams does not require substantial conceptual changes.
+
+Unlike infinite streams, pulling from a finite stream can fail, meaning the stream is finished.
+We therefore modify the @pull@ instruction to have two output labels: one to execute when a value is pulled, and the other to execute when the stream is finished.
+On the pushing end, we also need some way of finishing streams, so we add a new instruction to close an output stream.
+
+During evaluation we need some way of knowing whether a stream is closed, which can be added as an extra constructor in the \InputState~ type.
+The same constructor is added to the static input state used by fusion.
+In this way, for any changes made to evaluation, the analogous static change must be made in the fusion transform.
+
+% For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel.
+After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow.
+Our implementation supports finite streams, but our mechanized formalization does not cover it.
+
+It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements.
+However, this requires the fusion transform to analyse and reason about case statements and their predicates.
+It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
+By making the structure of finite streams explicit and constraining how processes use finite streams, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
+
+% We now describe the extensions required to support finite streams.
+% We add a new @closed@ constructor to the \InputState~ to encode the end of the stream.
+% Once an input stream is in the closed state, it can never change to another state: it remains closed thereafter.
+% 
+% We modify the @pull@ instruction so that it has two output labels (like @case@).
+% The first label, the read branch, is executed as before when the pull succeeds and a value is read from the stream.
+% The second label, the close branch, is executed when the stream is closed, and no more values will ever be available.
+% After a pull takes the close branch, any subsequent pulls from that stream will also take the close branch.
+% 
+% We add two new instructions for closing output streams and disconnecting from input streams.
+% Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
+% As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
+% After a stream has been closed, no further values can be pushed.
+% 
+% Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
+% This can be used when a process requires the first values of a stream, but does not require the whole stream.
+% If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
+% Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
+% The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
+% After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
+% After a process disconnects from an input channel, it can no longer pull from that channel.
+% 
+% We also add an instruction for terminating the process (@done@).
+% After all input streams have been read to completion or disconnected and output streams closed, the process may execute @done@ to signal that processing is complete.
+% 
+% The fusion definition must be extended to deal with these new instructions.
+% The static input state has a @closed@ constructor added and disconnection is encoded by removal from the input state, and the \ti{tryStep} changes more or less follow the evaluation changes.
+% Shared and connected pulls now deal with two more possibilities in the input state: the input may be closed in which case the close branch of the pull is taken; or the other process may have disconnected in which case the pull is executed as in the non-shared non-connected case.
+% Connected pushes must also deal with when the other process has disconnected in which case the push is executed as if it were non-connected.
+% For @in1@ and @out1@ channels, the new @close@ and @disconnect@ instructions are used as normal with no coordination required.
+% For connected @close@, as with @push@, the receiving process must have @none@ and the next step performs the @close@ and sets the input state to @closed@.
+% For shared @disconnect@, the @disconnect@ is only performed after both processes have disconnected; otherwise the entry is just removed from the input state.
+% For connected @disconnect@, the @disconnect@ is not performed and the entry is removed from the input state.
+% 
+% Finally, \ti{tryStepPair} is modified so that @done@ is performed when both machines are @done@.
+% 
+% These modifications allow our system to fuse finite streams as well as infinite.
 
@@
 
-\begin{figure}
-\begin{tikzpicture}
-\begin{axis}[
-  symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
-	xlabel=Runtime (ms),
-  xmin=0, xmax=290,
-	enlargelimits=0.01,
-	xbar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-\addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
-\end{axis}
-\end{tikzpicture}
-\caption{Runtime for append2; lower is faster.}
-\label{fig:bench:append}
-\end{figure}
+% TODO Figure suppressed to work out pagination. 
+%      Move all data to a single figure.
+% \begin{figure}
+% \begin{tikzpicture}
+% \begin{axis}[
+%   symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
+% 	xlabel=Runtime (ms),
+%   xmin=0, xmax=290,
+% 	enlargelimits=0.01,
+% 	xbar interval=0.7,
+%   width=7.5cm, height=4.5cm,
+% 	legend pos=north west,
+% ]
+% \addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Runtime for append2; lower is faster.}
+% \label{fig:bench:append}
+% \end{figure}
 
@@
 We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
-Fig.~\ref{fig:bench:append} shows the runtimes for appending 2MB of data.
+Fig.TODO 
+% ~\ref{fig:bench:append} 
+shows the runtimes for appending 2MB of data.
 The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
 
-\begin{figure}
-\begin{tikzpicture}
-\begin{axis}[
-  symbolic y coords={Streaming, Pipes, Conduit hand, Ours, end},
-	xlabel=Runtime (ms),
-  xmin=0, xmax=1200,
-	enlargelimits=0.01,
-	xbar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-\addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
-\end{axis}
-\end{tikzpicture}
-\caption{Runtime for partition; lower is faster.}
-\label{fig:bench:part}
-\end{figure}
+% TODO: Figure suppressed to work out pagination.
+%       Move all data to a single figure.
+% \begin{figure}
+% \begin{tikzpicture}
+% \begin{axis}[
+%   symbolic y coords={Streaming, Pipes, Conduit hand, Ours, end},
+% 	xlabel=Runtime (ms),
+%   xmin=0, xmax=1200,
+% 	enlargelimits=0.01,
+% 	xbar interval=0.7,
+%   width=7.5cm, height=4.5cm,
+% 	legend pos=north west,
+% ]
+% \addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
+% \end{axis}
+% \end{tikzpicture}
+% \caption{Runtime for partition; lower is faster.}
+% \label{fig:bench:part}
+% \end{figure}
 
@@
 Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
-Fig.~\ref{fig:bench:part} shows the runtimes for partitioning a 6MB file.
+Fig.TODO 
+% ~\ref{fig:bench:part}
+shows the runtimes for partitioning a 6MB file.
 

### 2017-05-06 2017mergingmerges/section/05-Evaluation.tex
Cut back discussion of fusability, pack proof, suppress future work on cases

@@
 The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
-This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
-We compared both, and recomputing was significantly faster.
-However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.
+% This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
+% We compared both, and recomputing was significantly faster.
+% However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.
 
@@
 The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
-It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.
+% It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.
 
@@
 
-On the upside, fusion failure is easy to detect.
-It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
-The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
+On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
+
+% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
 The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
-This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
+
+% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
 In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
@@
 Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
+
 When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
 This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
-In the worst case the fallback approach is to try all possible orderings of processes to fuse, assuming the client programmer is willing to wait for the search to complete.
+In the worst case the fallback approach is to try all possible orderings of processes to fuse.
+
+% assuming the client programmer is willing to wait for the search to complete.
 

### 2017-05-06 2017mergingmerges/section/XX-Proof.tex
Cut back discussion of fusability, pack proof, suppress future work on cases

@@
 \begin{code}
-Theorem Soundness (P1 : Program L1 C V1)
-                  (P2 : Program L2 C V2)
-                  (ss : Streams) (h : Heap)
-                  (l1 : L1) (is1 : InputStates)
-                  (l2 : L2) (is2 : InputStates) 
+Theorem Soundness 
+  (P1 : Program L1 C V1) (P2 : Program L2 C V2)
+  (ss : Streams) (h : Heap) (l1 : L1) (l2 : L2)
+  (is1 : InputStates) (is2 : InputStates) 
   :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
@@
 To aid mechanization, the Coq formalization has some small differences from the system presented in this paper.
-Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction.
+Instead of implementing non-deterministic evaluation we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process.
+This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
+
+% Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction.
 Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition.

### 2017-05-06 2017mergingmerges/section/05-Evaluation.tex
Cut back optimisation and proof sections to save space

@@
 
-Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
+% Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
 In Fig.~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
@@
 This is the case in Fig.~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
-In the result, the interleaved instructions from both source processes then share the same heap variable.
+To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time.
+Moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.
 
-To increase the chance that we can perform this above copy-propagation, we need both processess to want to pull from the same stream at the same time.
-In the definition of a particular process, moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.
-In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream.
+% In the result, the interleaved instructions from both source processes then share the same heap variable.
+
+% In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream.
 

### 2017-05-06 2017mergingmerges/section/XX-Proof.tex
Cut back optimisation and proof sections to save space

@@
 
-The @Soundness@ theorem uses @EvalBs@ to evaluate the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the result heap, using @Var1@ and @Var2@ to extract the variables.
+@EvalBs@ evaluates the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the result heap.
+The Coq formalization has some small differences from the system in this paper.
+Instead of implementing non-deterministic evaluation we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process.
+This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
 
@@
 % For shared inputs when one program has pulled, the other program must be evaluated with the other value removed from the end of the stream.
-
-To aid mechanization, the Coq formalization has some small differences from the system presented in this paper.
-Instead of implementing non-deterministic evaluation we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process.
-This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.
 

### 2017-05-06 2017mergingmerges/section/XX-Related.tex
Cut back related work to get 12 pages

@@
 
-Short cut fusion is an attractive idea, as it allows fusion systems to be specified by a single rewrite rule.
-However, short cut fusion relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
-Thus, short cut fusion is inherently biased towards pull fusion.
+% Short cut fusion is an attractive idea, as it allows fusion systems to be specified by a single rewrite rule. Thus, short cut fusion is inherently biased towards pull fusion.
+
+However, short cut fusion CITE relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
 Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
@@
 Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
-For effectful inputs such as reading from the network, duplicating work changes the semantics.
+
+% For effectful inputs such as reading from the network, duplicating work changes the semantics.
 % I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general
@@
 Data flow fusion~\cite{lippmeier2013data} is neither pull-based nor push-based, and supports arbitrary splits and joins.
-It supports standard combinators such as @map@, @filter@ and @fold@, and converts each stream to a series with explicit rate types, similar to the clock types of Lucid Synchrone \cite{benveniste2003synchronous}.
-These rate types ensure that well-typed programs can be fused without introducing unbounded buffers.
-This allows unfusable programs to be caught at compile time.
-However, it only supports a limited set of combinators, and adding more is non-trivial.
+It supports only a fixed set of standard combinators such as @map@, @filter@ and @fold@, and converts each stream to a series with explicit rate types, similar to the clock types of Lucid Synchrone \cite{benveniste2003synchronous}.
+% These rate types ensure that well-typed programs can be fused without introducing unbounded buffers.
+% This allows unfusable programs to be caught at compile time.
+% However, it only supports a limited set of combinators, and adding more is non-trivial.
 
@@
 Here, pull streams have the type @Source@ and represent a source, always ready to be pulled, while push streams have the type @Sink@ and represent a sink, always ready to accept data.
+The addition of push streams allows more programs to be fused than pull-only systems, but the computation must be manually split into sources and sinks.
+
+% This system requires the streaming computation to be manually split into sources and sinks.
 % Both systems rely on stream bindings being used linearly to ensure correctness.
-Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
-Sinks are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
+% Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
+% Sinks are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
 % It has the type @Sink a -> Sink b -> Sink (a,b)@.
-This system requires the streaming computation to be manually split into sources and sinks.
 % and be joined together by a loop that `drains' values from the source and pushes them into the sink.
@@
 Streaming IO libraries have blossomed in the Haskell ecosystem, generally based on Iteratees \cite{kiselyov2012iteratees}.
-Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers.
-However, these libraries provide no fusion guarantees, and as such programs tend to be written over chunks of data to make up for the communication overhead.
-For the most part they support only straight-line computations, with only limited forms of branching.
+Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers, but do not guarantee fusion.
+
+% However, these libraries do not guarantee fusion guarantees, and as such programs tend to be written over chunks of data to make up for the communication overhead.
+
+% For the most part they support only straight-line computations, with only limited forms of branching.
 
@@
 The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
-This is a much simpler fusion method than ours, but is also much stricter.
 When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
-Our system is an extension of synchronised product that allows some leeway in when processes must take shared steps: they do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.
+In our system the use of stream buffer variables allows some leeway in when processes must take shared steps.
+
+% This is a much simpler fusion method than ours, but is also much stricter.
+
+% They do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.
 
@@
 These languages describe \emph{passive} processes where values are fed in to streams from outside environments, such as data coming from sensors.
-In this case, the passive process has no control over the rate of input coming in, and if they support multiple input streams, they must accept values from them in any order.
-In contrast, the processes we describe are \emph{active} processes that have control over the input that is coming in.
-This is necessary for combinators such as mergesort-style @merge@, as well as @append@.
-Note that in the synchronous language literature, it is common to refer to a different merge operation, also known as @default@, which computes a stream that is defined whenever either input is defined.
+In contrast, our processes are \emph{active} and have control over when data is pulled from the source streams, which is nessesary for multiple input combinators such as @merge@ and  @append@.
+
+% In this case, the passive process has no control over the rate of input coming in, and if they support multiple input streams, they must accept values from them in any order.
+
+% Note that in the synchronous language literature, it is common to refer to a different merge operation, also known as @default@, which computes a stream that is defined whenever either input is defined.
 
 Synchronous dataflow (SDF; not to be confused with synchronous languages above) is a dataflow graph model of computation where each node has constant, statically known input and output rates.
-SDF is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
 StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
 Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow structures.
-Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
+
+% SDF is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.
+
+% Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling.
 A finite state machine is constructed, where each node of the FSM denotes its own SDF graph.

### 2017-05-07 2017mergingmerges/section/XX-Related.tex
Cite sync kahn networks

@@
 
-Synchronous languages such as LUSTRE~\cite{halbwachs1991synchronous}, Lucy-n~\cite{mandel2010lucy} and SIGNAL~\cite{le2003polychrony} all use some form of clock calculus and causality analysis to ensure that programs can be statically scheduled with bounded buffers.
+Synchronous languages such as LUSTRE~\cite{halbwachs1991synchronous}, Lucy-n~\cite{mandel2010lucy} and SIGNAL~\cite{le2003polychrony} all use some form of clock calculus and causality analysis to ensure that programs can be statically scheduled with bounded buffers~\cite{caspi1996:kahn}.
 These languages describe \emph{passive} processes where values are fed in to streams from outside environments, such as data coming from sensors.
 In contrast, our processes are \emph{active} and have control over when data is pulled from the source streams, which is nessesary for multiple input combinators such as @merge@ and  @append@.
+Synchronous dataflow languages reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling \cite{bouakaz2013real}.
 

### 2017-05-07 2017mergingmerges/section/01-Introduction.tex
Fix sigma -> bs in advance rules
Cite sync kahn networks
edits and adjust page order

@@
 
-Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers.
+Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on them.
 The identifiers from both streams are sorted, but may include duplicates.
@@
 
-To implement this, we might write each operator as its own concurrent process, sending stream elements over intra-process channels.
+To implement this program we might write each operator as its own concurrent process, sending stream elements over intra-process channels.
 Developing this could be easy or hard, depending on the available language features for concurrency.
-However, the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk stream data to reduce communication overhead, is invariably a headache.
-
-% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
-Each operator could be given this interface, so that values from each stream are computed on demand.
-This approach is commonly taken in database systems \cite{Graefe:Volcano}.
-However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering.
-Suppose a consumer pulls many elements from the @sUnique@ output stream.
-In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them.
-It needs to buffer these elements because they are also needed to perform the @merge@ operation.
-When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.
+However, the \emph{performance tuning} of such a system, such as using back-pressure to prevent buffers from being overrun, or how to chunk stream data to amortize communication overhead, is invariably a headache.
 
@@
 More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.
-
-% Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
-
-% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.
 
@@
 We discuss this extension in \S\ref{s:Finite}.
+
+% BL: This is now discussed in S2.5 Breaking It Down.
+% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream.
+Each operator could be given this interface, so that values from each stream are computed on demand.
+This approach is commonly taken in database systems \cite{Graefe:Volcano}.
+However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering.
+Suppose a consumer pulls many elements from the @sUnique@ output stream.
+In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them.
+It needs to buffer these elements because they are also needed to perform the @merge@ operation.
+When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.
+
+% TODO: move to related work.
+% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

### 2017-05-07 2017mergingmerges/section/02-Processes.tex
edits and adjust page order

@@
 For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@.
-We include the concrete representation and a diagram of the process viewed as a state machine.
+We include the concrete representation and a diagram of the process when viewed as a state machine.
 
@@
 The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction.
-In this form the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.
+In this form, the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.
 
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Breaking It Down}

### 2017-05-07 2017mergingmerges/section/02-Processes.tex
Edits to S2

@@
 Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
-
-% TODO: Talk about what drop is really for later.
-% This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
-The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.
 
@@
 \end{code}
-% \begin{minipage}{0.29\textwidth}
 \includegraphics[scale=1.0]{figures/state-merge.pdf}
-% \includegraphics[scale=1.0,angle=-90,origin=c]{figures/state-merge.pdf}
-% \end{minipage}
-
 \caption{The merge combinator}
@@
 The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.
-
-% TODO: Shift this later.
-% As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
-For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output.
-After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@.
-We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
 
@@
 
-We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams.
+We also associate each of the joint result states with an description of whether each source machine has already pulled a value from each of its input streams.
 For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
@@
 
-Finally, note that in the result states @F0@ through @F3@ the state of the input streams transition from `none', to `pending' then to `have'.
+Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
 The `none' state means that we have not yet pulled a value from the associated stream.
@@
 
-% The stream buffer variables we additionally introduced for coordination were freshly created during fusion.
-
 % -----------------------------------------------------------------------------
@@
 We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
-Importantly, we needed to \emph{break down} the definition of each of the operators into imperative statements so that we could interleave their execution appropriately.
+Importantly, we needed to \emph{break down} the definition of each operator into imperative statements so that we could interleave their execution appropriately.
 We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way.
 Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@.
-Instead, in the fused result program we have performed the job of a concurrent scheduler at compile time.
-The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching.
+Instead, during fusion we have performed the job of a concurrent scheduler at compile time.
+The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching --- as there would be in a standard concurrent implementation using multiple threads.
 
-The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative code that implements the operators was also used in prior work on Flow Fusion~\cite{lippmeier2013data}.
+The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative statements that implement each operator was also used in prior work on Flow Fusion~\cite{lippmeier2013data}.
 However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are.
-This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss in \S\ref{s:FutureWork}.
+This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams.
 
-Note that trying to use lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem.
+Note that relying on lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem.
 Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists.
 As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirity before demanding any of the elements from the second list (@sUnion@).
-If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before remanding any of @sIn2@, causing a space leak.
+If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before demanding any of @sIn2@, causing a space leak.
 Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer.
-The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers are not at liberty to defer handling of these values.
+The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers cannot defer the handling of these values.
 
@@
 For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.
+
+% TODO: Shift this later.
+% As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output.
+For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output.
+After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@.
+We discuss how to handle finite streams later in ~\S\ref{s:Finite}.
+
+% TODO: Talk about what drop is really for later.
+% This @drop@ instruction is used to coordinate concurrent processes when performing fusion.
+The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.
 
@@
 % -----------------------------------------------------------------------------
-
-% -----------------------------------------------------------------------------
-% BL: This is too much low level detail at this point, we need more intuition.
-
-% The instructions of @merge@ can be split into four categories, which can be identified in the fused process: labels @B0@-@B2@ perform initialisation, and are mapped to @F0@-@F6@; label @C0@ performs case analysis to find the smaller value and is mapped to @F7@; labels @D0@-@D2@ push the value from the first stream, @s1@, and are mapped to @F8@-@F15@; and labels @E0@-@E2@ push the value from the second stream, @s2@, and are mapped to @F16@-@F18@.
-
-% The @group@ process can also be identified in the fused process: the fused labels @F0@-@F5@ perform the pulling, case analysis and pushing from instructions @A0@-@A2@.
-These instructions are seen again in @F10@-@F14@, when the @merge@ process is pulling from the @s1@ stream.
-Thus the @group@ instructions have been duplicated, and are performed once when @merge@ initialises, and again every time @merge@ pulls from the first stream.
-
-% Labels @F5@ and @F14@ correspond to the (@drop s1@) instruction at @A3@ in @group@, but note that they are fused as @jump@ instructions.
-As @s1@ is shared between the two processes, the element from @s1@ can only be dropped once both processes agree to drop.
-In the fused labels on the right hand column, the other process still has @s1@ as `pending' or `have', so the element cannot yet be dropped.
-The actual @drop@ is only performed once both processes are `none', at label @F9@.

### 2017-05-07 2017mergingmerges/section/03-Definition.tex
Edits to S4

@@
 
-The dynamic execution of process networks consists of three aspects:
+The dynamic execution of a process network consists of:
 
@@
 \item \emph{Injection} of a single value from a stream into a process, or a network.
-Each individual process only needs to accept the value when it is ready for it, and injection into a network succeeds only when they \emph{all} accept it.
+Each individual process only needs to accept an injected  value when it is ready for it, and injection into a network succeeds only when they \emph{all} processes accept it.
 
@@
 Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not.
-When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the network to advance without requiring unbounded buffering.
+When we fuse two processes we choose one particular instruction ordering that enables the network to advance without requiring unbounded buffering.
+The single ordering is chosen by heuristically deciding which pair of states to merge during fusion, and is discussed in \S\ref{s:EvaluationOrder}.
 
@@
 The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
-The @injects@ form is similar, operating on a network.
+The @injects@ form is similar, operating on a process network.
 
@@
 
-Rule (Jump) produces a new label and associated update expressions and rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and emit the appropriate label.
+Rule (Jump) produces a new label and associated update expressions. Rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and emit the appropriate label.
 
@@
 This allows processes to perform internal computation.
+
 % , without needing to synchronize with the rest of the network.
@@
 For this to succeed it must be possible to inject the output value $v$ into all processes that have channel $c$ as one of their inputs.
+
 % As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process.
@@
 
-In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
+In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require unbounded buffering.
 In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.

### 2017-05-07 2017mergingmerges/section/04-Fusion.tex
Get benchmark diagram on a better page

@@
 Function \ti{swaplabels} flips the order of the compound labels in an instruction.
+
+\subsection{Fusibility}
+\label{s:FusionOrder}
+When we fuse a pair of processes we commit to a particular interleaving of instructions from each process.
+When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process.
+Consider our @alternatives@ example from \S\ref{s:EvaluationOrder} again:
+
+\begin{code}
+  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
+  alternates sInA sInB sInC
+   = let  s1   = alt2 sInA sInB
+          s2   = alt2 sInB sInC
+          sOut = zip s1 s2
+     in   sOut
+\end{code}
+
+In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, this final fusion transform fails.
+This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
+% As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.
+
+Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock.
+Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails.
+Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.~\ref{fig:Fusion:Def:StepPair}.
+The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.
+
+On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
+
+% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
+The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
+
+% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
+In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
+spelunking: Exploration of caves, especially as a hobby.
+Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.
+
+In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network.
+For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds.
+The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock.
+Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
+
+% TODO BL: Get this back in the discussion.
+% When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
+This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
+In the worst case the fallback approach is to try all possible orderings of processes to fuse.
+
+% assuming the client programmer is willing to wait for the search to complete. 
+% We discuss other possible approaches in \S\ref{s:Future:FusionOrder}.
+
+% When the two @alt2@ processes are fused we end up with a sequential result process.
+Trying to then fuse that result into the downstream @zip@ process fails.
+Alternately, if we were to try to concurrently evaluate the fused result process with the downstream @zip@ process would deadlock because we would reach a state where neither process could advance.
+To state this again: fusion fails when concurrent evaluation of the two source processes would deadlock.
+
+% The main fusion algorithm here works on pairs of processes.
+% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
+The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
+Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process.
+In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes.
+The simplest example of this has two input streams, a function applied to both, then zipped together.

### 2017-05-07 2017mergingmerges/section/05-Evaluation.tex
Get benchmark diagram on a better page

@@
 % -----------------------------------------------------------------------------
+\eject
 \section{Results}
@@
 
+% -----------------------------------------------------------------------------
 \subsection{Benchmarks}
@@
 We present three benchmarks: one array algorithm, and two simple file operations.
-
-\begin{figure}
-\begin{tikzpicture}
-\begin{axis}[
-  symbolic y coords={Store, Recompute, Ours, Hand, end},
-	xlabel=Runtime (ms),
-  xmin=0, xmax=200,
-	enlargelimits=0.01,
-	xbar interval=0.7,
-  width=7.5cm, height=4.5cm,
-	legend pos=north west,
-]
-\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
-\end{axis}
-\end{tikzpicture}
-\caption{Runtime for Quickhull; lower is faster.}
-\label{fig:bench:quickhull}
-\end{figure}
 
@@
 % -----------------------------------------------------------------------------
-\subsection{Fusibility}
-\label{s:FusionOrder}
-When we fuse a pair of processes we commit to a particular interleaving of instructions from each process.
-When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process.
-Consider our @alternatives@ example from \S\ref{s:EvaluationOrder} again:
-
-\begin{code}
-  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
-  alternates sInA sInB sInC
-   = let  s1   = alt2 sInA sInB
-          s2   = alt2 sInB sInC
-          sOut = zip s1 s2
-     in   sOut
-\end{code}
-
-In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, this final fusion transform fails.
-This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before then pushing values to @s2@.
-% As we discussed in \S\ref{s:EvaluationOrder}, the downstream @zip@ process needs to pull single values from @s1@ and @s2@ alternatively.
-
-Dynamically, if we were to try to execute the result process and the downstream @zip@ process concurrently, then the execution would deadlock.
-Statically, when we try to fuse the result process with the downstream @zip@ process then the deadlock is discovered and fusion fails.
-Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this will manifest as the failure of the $tryStepPair$ function from Fig.~\ref{fig:Fusion:Def:StepPair}.
-The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.
-
-On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
-
-% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
-The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
-
-% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
-In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
-spelunking: Exploration of caves, especially as a hobby.
-Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.
-
-In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network.
-For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds.
-The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock.
-Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
-
-When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
-This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
-In the worst case the fallback approach is to try all possible orderings of processes to fuse.
-
-% assuming the client programmer is willing to wait for the search to complete.
-
-% We discuss other possible approaches in \S\ref{s:Future:FusionOrder}.
-
-% When the two @alt2@ processes are fused we end up with a sequential result process.
-Trying to then fuse that result into the downstream @zip@ process fails.
-Alternately, if we were to try to concurrently evaluate the fused result process with the downstream @zip@ process would deadlock because we would reach a state where neither process could advance.
-To state this again: fusion fails when concurrent evaluation of the two source processes would deadlock.
-
-% The main fusion algorithm here works on pairs of processes.
-% When there are more than two processes, there are multiple orders in which the pairs of processes can be fused.
-The order in which pairs of processes are fused does not affect the output values, but it does affect the access pattern: the order in which outputs are produced and inputs read.
-Importantly, the access pattern also affects whether fusion succeeds or fails to produce a process.
-In other words, while evaluating multiple processes is non-deterministic, the act of fusing two processes \emph{commits} to a particular deterministic interleaving of the two processes.
-The simplest example of this has two input streams, a function applied to both, then zipped together.
 
@@
 % In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream.
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[
+  symbolic y coords={Store, Recompute, Ours, Hand, end},
+        xlabel=Runtime (ms),
+  xmin=0, xmax=200,
+        enlargelimits=0.01,
+        xbar interval=0.7,
+  width=7.5cm, height=4.5cm,
+        legend pos=north west,
+]
+\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
+\end{axis}
+\end{tikzpicture}
+\caption{Runtime for Quickhull; lower is faster.}
+\label{fig:bench:quickhull}
+\end{figure}
 

### 2017-05-07 2017mergingmerges/section/04-Fusion.tex
Fix section on Fusability after the Great Cutting

--- 2017-05-04 2017mergingmerges/section/04-Fusion.tex
@@
 
-The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.
+Our core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.
 
-Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion.
+Figure~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion.
 We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process.
@@
 
-We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with channel @c@.
+We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with \mbox{channel @c@}.
 We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
@@
 
-Still in Fig.~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
-Type @in2@ indicates that the two processes @pull@ from the same channel, and these actions must then be coordinated.
+Still in Figure~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
+Type @in2@ indicates that the two processes @pull@ from the same channel, so these actions must be coordinated.
 Type @in1@ indicates that only a single process pulls from the channel.
@@
 % -------------------------------------
-Fig.~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes together, constructing a result process that does the job of both.
+Figure~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes, constructing a result process that does the job of both.
 We start with a joint label $l_0$ formed from the initial labels of the two source processes.
 We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next.
-The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states.
+The possible destination labels of that instruction (computed with $outlabels$ from Figure~\ref{fig:Fusion:Utils}) define new joint labels and reachable states.
 As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no new states can be discovered.
@@
 % -------------------------------------
-Fig.~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
+Figure~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
 
 Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@.
-This helps collect @jump@ instructions together, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle.
-As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.
+This helps collect @jump@ instructions together, which are then easier for post-fusion optimization to handle (\S\ref{s:Optimisation}).
+As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Figure~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.
 
@@
 \smallskip
-Fig.~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction.
+Figure~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction.
 This function takes the map of channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.
@@
 
-Still in Fig.~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.
+Still in Figure~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.
 
@@
 
-Clause (LocalDrop) applies when the left process wants to drop the current value from that it read from an unshared input channel, which requires no coordination.
+Clause (LocalDrop) applies when the left process wants to drop the current value that it read from an unshared input channel, which requires no coordination.
 
@@
 
-\smallskip
-Fig.~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
+% NOTE BL: commented out to save space. The figure is mentioned earlier.
+% \smallskip
+% Figure~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
 Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
@@
 Function \ti{swaplabels} flips the order of the compound labels in an instruction.
+
+% -----------------------------------------------------------------------------
+\subsection{Fusibility}
+\label{s:FusionOrder}
+When we fuse a pair of processes we commit to a particular interleaving of instructions from each process.
+When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process.
+Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
+\begin{code}
+  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
+  alternates sInA sInB sInC
+   = let  s1   = alt2 sInA sInB
+          s2   = alt2 sInB sInC
+          sOut = zip s1 s2
+     in   sOut
+\end{code}
+
+In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, the final fusion transform fails.
+This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before pushing values to @s2@.
+On the other hand, @zip@ needs to pull a \emph{single} value from each of its inputs alternately.
+
+Dynamically, if we were to execute the first fused result process, and the downstream @zip@ process concurrently, then the execution would deadlock.
+Statically, when we try to fuse the result process with the downstream @zip@ process the deadlock is discovered and fusion fails.
+Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this manifests as the failure of the $tryStepPair$ function from Figure~\ref{fig:Fusion:Def:StepPair}.
+The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.
+
+% On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.
+
+% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code.
+The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.
+
+% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations.
+In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def.
+spelunking: Exploration of caves, especially as a hobby.
+Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.
+
+In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network.
+For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds.
+The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possiblity of deadlock.
+Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.
+
+When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking to towards the input streams, and fusing in successive processes as they occur.
+This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized.
+In the worst case the fallback approach is to try all possible orderings of processes to fuse.

### 2017-05-08 2017mergingmerges/section/05-Evaluation.tex
Rename Evaluation -> Implementation
Fix section on Fusability after the Great Cutting
benchmark figure

@@
 % -----------------------------------------------------------------------------
-\eject
-\section{Results}
+\section{Implementation}
 \label{s:Evaluation}
@@
 
-We compare hand-fused against ours, as well as two @Data.Vector@ versions, which uses shortcut fusion.
-The shortcut fusion system cannot fuse both operations into a single loop, and both operations require the distances between the line and each point.
+We compare against a hand-fused program and a @Data.Vector@ program, which uses shortcut fusion.
+The shortcut fusion system cannot fuse both operations into a single loop, and both operations recompute the distances between the line and each point.
 % This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
@@
 
-Fig.~\ref{fig:bench:quickhull} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
-The hand-fused version is the fastest, while our version is slower, but still faster than the two vector versions.
+The first group in Fig.~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
+The hand-fused version is the fastest, while our version is slower, but still faster than the vector version.
 The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
@@
 
-% TODO Figure suppressed to work out pagination. 
-%      Move all data to a single figure.
-% \begin{figure}
-% \begin{tikzpicture}
-% \begin{axis}[
-%   symbolic y coords={Conduit, Pipes, Streaming, Ours, end},
-% 	xlabel=Runtime (ms),
-%   xmin=0, xmax=290,
-% 	enlargelimits=0.01,
-% 	xbar interval=0.7,
-%   width=7.5cm, height=4.5cm,
-% 	legend pos=north west,
-% ]
-% \addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Runtime for append2; lower is faster.}
-% \label{fig:bench:append}
-% \end{figure}
-
-The first file benchmark simply appends two files, while counting the number of lines.
+The first file benchmark simply appends two files, while counting the lines.
 We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
-Fig.TODO 
-% ~\ref{fig:bench:append} 
-shows the runtimes for appending 2MB of data.
+The second group in Fig.~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
 The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
-
-% TODO: Figure suppressed to work out pagination.
-%       Move all data to a single figure.
-% \begin{figure}
-% \begin{tikzpicture}
-% \begin{axis}[
-%   symbolic y coords={Streaming, Pipes, Conduit hand, Ours, end},
-% 	xlabel=Runtime (ms),
-%   xmin=0, xmax=1200,
-% 	enlargelimits=0.01,
-% 	xbar interval=0.7,
-%   width=7.5cm, height=4.5cm,
-% 	legend pos=north west,
-% ]
-% \addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
-% \end{axis}
-% \end{tikzpicture}
-% \caption{Runtime for partition; lower is faster.}
-% \label{fig:bench:part}
-% \end{figure}
 
@@
 The output lines are also counted.
-The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or partially hand-fused.
+The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or at least partially hand-fused.
 Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
-Fig.TODO 
-% ~\ref{fig:bench:part}
-shows the runtimes for partitioning a 6MB file.
+The final group in Fig.~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
 
@@
 \begin{axis}[
-  symbolic y coords={Store, Recompute, Ours, Hand, end},
-        xlabel=Runtime (ms),
-  xmin=0, xmax=200,
+  symbolic x coords={Quickhull, Append, Partition, end},
+        ylabel=Runtime (ms),
+  ymin=0, ymax=300,
         enlargelimits=0.01,
-        xbar interval=0.7,
+        ybar interval=0.7,
   width=7.5cm, height=4.5cm,
-        legend pos=north west,
+  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3}
 ]
-\addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
+\addplot coordinates { (Quickhull, 75) (Append, 88) (Partition, 60) (end, 0) };
+\addplot coordinates {(Quickhull, 110) (Append, 86) (Partition, 61) (end,0)};
+\addplot coordinates { (Quickhull, 130) (Append, 0)  };
+% \addplot coordinates { (Quickhull, 200) (Append, 0) };
+\addplot coordinates {  (Append, 282) (Partition, 136) (end,0) };
+\addplot coordinates {  (Append, 270) (Partition, 166) (end,0) };
+\addplot coordinates {  (Append, 168) (Partition, 268) (end,0) };
+\legend{Handfused, Ours, Vector, Conduit, Pipes, Streaming}
 \end{axis}
 \end{tikzpicture}
-\caption{Runtime for Quickhull; lower is faster.}
-\label{fig:bench:quickhull}
+\caption{Runtime for benchmarks; lower is faster.}
+\label{fig:bench:all}
 \end{figure}
 
-% -----------------------------------------------------------------------------
-% \begin{figure}
-% \begin{minipage}[t]{0.10\textwidth}
-% \end{minipage}
-% \begin{minipage}[t]{0.4\textwidth}
-% \begin{tikzpicture}
-% \begin{axis}[
-% % Hide the label on the second graph
-% 	ylabel=Process states,
-% 	xlabel=Pipelined combinators,
-%   ymin=0, ymax=100,
-% 	enlargelimits=0.01,
-% 	ybar interval=0.7,
-%   width=7.5cm, height=4.5cm,
-% 	legend pos=north west,
-% ]
-% \addplot coordinates {(1,9) (2,17) (3,27) (4,37) (5,47) (6,57) (7,67)   (8,1) };
-% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
-% \legend{With merge, Without merge};
-% \end{axis}
-% \end{tikzpicture}
-% \end{minipage}
-% \quad
-% \quad
-% \quad
-% \begin{minipage}[t]{0.4\textwidth}
-% \begin{tikzpicture}
-% \begin{axis}[
-% %       ylabel=Process states,
-%         xlabel=Parallel combinators,
-%   ymin=0, ymax=100,
-%         enlargelimits=0.01,
-%         ybar interval=0.7,
-%   width=7.5cm, height=4.5cm,
-%         legend pos=north west,
-% ]
-% \addplot coordinates {(1,9) (2,27) (3,40) (4,53) (5,66) (6,79) (7,92)
-%   % Last bar doesn't show for some reason, so need to add a dummy value for the next one
-%     (8,1) };
-% 
-% \addplot coordinates {(1,4) (2,9) (3,14) (4,19) (5,24) (6,29) (7,34)    (8,1) };
-% 
-% \legend{With merge, Without merge};
-% \end{axis}
-% \end{tikzpicture}
-% \end{minipage}
-% 
-% \caption{Maximum output process size for fusing all combinations of up to $n$ combinators.}
-% \label{fig:bench:outputsize}
-% \end{figure}
-% 
-% 
-% % -----------------------------------------------------------------------------
-% \begin{figure}
-% 
-% \begin{tikzpicture}
-% \begin{axis}[
-% 	ylabel=Process states,
-% 	xlabel=Number of merges,
-% %  ymode=log,
-%   ymin=0, ymax=1500,
-% 	enlargelimits=0.01,
-% 	ybar interval=0.7,
-%   width=7.5cm, height=4.5cm,
-% 	legend pos=north west,
-% ]
-% % These are the values for splitting.
-% % They are smaller than the 'chaining', but look much nicer on the linear graph.
-% \addplot coordinates {(1,9) (2,42) (3,97) (4,196) (5,383) (6,746) (7,1461)
-%   % (8,2880)
-%   (8,1)
-%   };
-% 
-% % These are the values for chaining
-% % \addplot coordinates {(1,4) (2,48) (3,194) (4,760) (5,2814) (6,10064) (7,1) };
-% 
-% \end{axis}
-% \end{tikzpicture}
-% 
-% \caption{Exponential blowup occurs when splitting or chaining merges together.}
-% \label{fig:bench:exponential}
-% \end{figure}
+% Quickhull
+% \addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
+% Append
+% \addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
+% Partition
+% \addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };

### 2017-05-08 2017mergingmerges/section/05-Evaluation.tex
consistently use Figure N instead of Fig. N

@@
 
-The first group in Fig.~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
+The first group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
 The hand-fused version is the fastest, while our version is slower, but still faster than the vector version.
@@
 We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
-The second group in Fig.~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
+The second group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
 The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
@@
 Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
-The final group in Fig.~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
+The final group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
 
-% Fused result programs like the one in Fig.
-\ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly.
+% Fused result programs like the one in Figure \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly.
 The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.
@@
 %As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
-The left of Fig.~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
+The left of Figure~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
 To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes.
 The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start.
-The right of Fig.~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
+The right of Figure~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
 In both cases the number of states in the result process grows linearly with the number of processes.
@@
 %
-%On the other hand, Fig.~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
+%On the other hand, Figure~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
 The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream.
@@
 After we have fused two processes together, it may be possible to simplify the result before fusing in a third.
-Consider the result of fusing @group@ and @merge@ which we saw back in Fig.~\ref{fig:Process:Fused}.
+Consider the result of fusing @group@ and @merge@ which we saw back in Figure~\ref{fig:Process:Fused}.
 At labels @F1@ and @F2@ are two consecutive @jump@ instructions.
@@
 In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code.
-The (PreferJump) clauses of Fig.~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
+The (PreferJump) clauses of Figure~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
 
@@
 % Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
-In Fig.~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
+In Figure~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
 Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable.
@@
 When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
-This is the case in Fig.~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
+This is the case in Figure~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
 To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time.

### 2017-05-08 2017mergingmerges/section/02-Processes.tex
Appendix: finite stream rules

@@
 The @group@ combinator has two parameters, @sIn1@ and @sOut1@, which bind the input and output streams respectively.
-The \emph{nu-binders} \mbox{$\nu$ @(f: Bool) (l: Nat)@...} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
+The \emph{nu-binders} \mbox{($\nu$ @(f: Bool) (l: Nat)@...)} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations.
 Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
@@
 The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process.
-When describing the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.
+In our informal description of the fusion algorithm we instantiate the parameters of each combinator with arguments of the same names.
 
@@
 
-The algorithm proceeds by considering pairs of states: one in each of the source process state machines to be fused.
-Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in Figure~\ref{fig:Fusion:Pulls}.
+The algorithm proceeds by considering pairs of states: one from each of the source process state machines to be fused.
+Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in the top of Figure~\ref{fig:Fusion:Pulls}.
 The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@.
@@
 
-We also associate each of the joint result states with an description of whether each source machine has already pulled a value from each of its input streams.
-For the @F0@ case shown in Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
+We also associate each of the joint result states with a description of whether each source machine has already pulled a value from each of its input streams.
+For the @F0@ case in the top of Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
 The result state @F0@ represents a combination of the two source states @A0@ and @B0@.
@@
 Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively.
-The left of Figure~\ref{fig:Fusion:Case} shows the next few transitions of the source machines.
+The lower half of Figure~\ref{fig:Fusion:Case} shows the next few transitions of the source machines.
 From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input.

### 2017-05-08 2017mergingmerges/section/A1-Finite.tex
Appendix: finite stream rules

%!TEX root = ../Main.tex
\eject
\appendix

\section{Finite streams}
\label{s:FiniteDetails}

This appendix briefly describes the finite streams extension, showing the changes to instructions and the fusion algorithm.
The finite evaluation rules are not shown, as the changes follow the same structure as the changes to the fusion algorithm.

\begin{figure}
\begin{tabbing}
MMMMMM \TABDEF @MMMMM@  \TABSKIP $\Chan$ \TABSKIP $\Chan$ \TABSKIP $\Next$ \TABSKIP \kill
\Instr
    \> ::=\> @push@  \> \Chan  \> \Exp  \> \Next \\
    \TABALT  @drop@  \> \Chan  \>       \> \Next \\
    \TABALT  @case@  \> \Exp   \> \Next \> \Next \\
    \TABALT  @jump@  \>        \>       \> \Next \\
    \\
    \TABALT  @pull@  \> \Chan  \> \Var  \> \Next \> \Next \\
    \TABALT  @close@ \> \Chan  \>       \> \Next \\
    \TABALT  @exit@ 
\\
\\
$\InputState_F$ \> ::=  \> ~~~ $@none@_F ~|~ @pending@_F ~|~ @have@_F ~|~ @closed@_F$
\end{tabbing}
\caption{Finite instructions}
\label{fig:Finite:Instr}
\end{figure}

Figure~\ref{fig:Finite:Instr} shows the grammar for instructions and the static input state. The first group of instructions containing @push@, @drop@, @case@ and @jump@ are unchanged.

The @pull@ instruction is modified to have two output labels, similar to @case@. The first, the success branch, is used when the input stream is still open and pulling succeeds, in which case the variable is set to the pulled value as before. The second output label, the closed branch, is used when the input stream has been closed, and the variable is not written to. This new @pull@ is analogous to a @pull@ followed by a @case@ in the infinite stream version.

The @close@ instruction is used by a pushing process to close or end an output stream. Any subsequent pulls from this channel in other processes will take the closed branch. After an output channel is closed, it cannot be pushed to and remains closed forever.

Finally, the @exit@ instruction is used once a process is finished with all its streams, and has nothing left to do. All output streams must be closed before the process finishes. This instruction has no output labels, as there is nothing further to execute.

Also in Figure~\ref{fig:Finite:Instr}, the static input state used for fusion ($\InputState_F$) must now track closed streams. The new constructor $@closed@_F$ denotes that the stream is closed, while the rest is unchanged.

For the fusion algorithm, the top-level function $\ti{fusePair}$ remains unchanged. The functions $\ti{outlabels}$ and $\ti{swaplabels}$ are not shown as they are easily modified by adding cases for the new instructions.

Figure~\ref{fig:Finite:tryStepPair} shows the modified $\ti{tryStepPair}$ function. This function uses the same heuristics to decide which process to execute when both can progress, but now that the processes can finish with @exit@, we must take care to only finish the fused process once \emph{both} source processes are finished. The (DeferExit1) and (DeferExit2) clauses achieve this by allowing the other process if one is an @exit@. Once both processes are finished, both new clauses will fail while (DeferPull1) succeeds, using the @exit@ from the first process. Another way to think of this is that if either process has work to do, the fused process still has work to do.

\begin{figure}
\begin{tabbing}
M \= M \= M \= MMMMMMMMMMMMMMMMMMM \= \kill
$\ti{tryStepPair} ~:~ \ChanTypeMap$ \\
\> \> $\to \Label_1 \to \Instr$ ~~~
            $\to \Label_1 \to \Instr$ \\
\> \> $\to \Maybe~\Instr$ \\
$\ti{tryStepPair} ~\cs~l_p~i_p~l_q~i_q$ \\

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~ i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
\> ~~ $\wedge~@exit@~\not\in~i_p'$ 
   ~~~~~ $\to~i_p'$ 
\> \> \> \note{DeferExit1} 
\\[0.5ex]

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
\> ~~ $\wedge~@exit@~\not\in~i_q'$ 
   ~~~~~ $\to~\ti{swaplabels}~i_q'$ 
\> \> \> \note{DeferExit2} 
\\[0.5ex]

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~@jump@~(l,u)~\in~i_p'$ \\
\> $\to~i_p'$
\> \> \> \note{PreferJump1} 
\\[0.5ex]

\> $~|~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p ~\wedge~@jump@~(l,u)~\in~i_q'$ \\
\> $\to~\ti{swaplabels}~i_q'$ 
\> \> \> \note{PreferJump2}
\\[0.5ex]

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~ i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
\> ~~ $\wedge~@pull@~c~x~(l,u)~\not\in~i_p'$ 
   ~~~~~ $\to~i_p'$ 
\> \> \> \note{DeferPull1} 
\\[0.5ex]

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
\> ~~ $\wedge~@pull@~c~x~(l,u)~\not\in~i_q'$ 
   ~~~~~ $\to~\ti{swaplabels}~i_q'$ 
\> \> \> \note{DeferPull2} 
\\[0.5ex]

\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q$ ~~~~~ $\to~i_p'$ 
\> \> \> \note{Run1} 
\\[0.5ex]

\> $~|~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ ~~~~~ $\to~\ti{swaplabels}~i_q'$
\> \> \> \note{Run2} \\

\end{tabbing}
\caption{Fusion step coordination for a pair of processes.}
\label{fig:Finite:tryStepPair}
\end{figure}

Figure~\ref{fig:Finite:tryStep} shows the modified $\ti{tryStep}$ function.
The clauses for the unchanged instructions @push@, @drop@, @case@ and @jump@ remain unchanged; these are reordered to the top of the function.

The @pull@ clauses use $l'_o$ for the open output label, and $l'_c$ for the closed label.
Clause (LocalPull) now uses two output labels, and leaves the other process as-is.

Clause (SharedPull) applies when the channel state is @pending@, meaning there is already a value available. This means that the channel is not yet closed, and the success branch can be taken.

Clause (SharedPullInject) applies when both processes need to pull from a shared input. As before we execute a real @pull@, this time with two branches. In the success branch, the input states are set to @pending@ as before. In the closed branch, the input states are set to @closed@ so the next and subsequent pulls take the closed branch.

Clause (SharedPullClosed) applies when the channel state is @closed@, which means either the other process has pulled and discovered that the channel is closed, or in case of connected input, the other process has closed the channel. Either way we simply jump, taking the closed branch of the @pull@.

Clause (LocalClose) applies when closing a local output.

Clause (SharedClose) applies when closing a connected output. As with (SharedPush), the other input state for the other process must be empty and ready to pull from the channel. The input state for the other process is then set to @closed@, forcing its next pull to take the closed branch.

Finally, clause (LocalExit) allows the process to finish. However, recall that the $\ti{tryStepPair}$ function has been modified to only @exit@ when both processes are ready to finish.

\begin{figure*}
\begin{tabbing}
M \= M \= MMMMMMMMMMMMMMMMMMMMMM \= MMMMMMMMMMMMMMMMMMMMMMMMMMMMMM \= \kill
$\ti{tryStep} ~:~ \ChanTypeMap \to \Label_1 \to \Instr \to \Label_1 \to \Maybe~\Instr$ \\
$\ti{tryStep} ~\cs~(l_p,s_p)~i_p~(l_q,s_q)~=~@match@~i_p~@with@$ \\

\> $@jump@~(l',u')$ 
\> \> $\to~@jump@~
      \nextStep
        {l'}{s_p}
        {l_q}{s_q}
        {u'}
      $ 
\> \note{LocalJump}
\\[1ex]

\> $@case@~e~(l'_t,u'_t)~(l'_f,u'_f)$
\> \> $\to~@case@~e~
      \nextStep
        {l'_t}{s_p}
        {l_q}{s_q}
        {u'_t}
      ~
      \nextStep
        {l'_f}{s_p}
        {l_q}{s_q}
        {u'_f}
      $ 
\> \note{LocalCase}
\\[1ex]

\> $@push@~c~e~(l',u')$ \\
\> \> $~|~\cs[c]=@out1@$ 
\> $\to~@push@~c~e~
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {s_q}
        {u'}
      $ 
\> \note{LocalPush}\\

\> \> $~|~\cs[c]=@in1out1@ ~\wedge~ s_q[c]=@none@_F$ 
\> $\to~@push@~c~e~
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {\HeapUpdateOne{c}{@pending@_F}{s_q}}
        {\HeapUpdateOne{@chan@~c}{e}{u'}}
      $
\> \note{SharedPush}
\\[1ex]

\> $@drop@~c~(l',u')$ \\
\> \> $~|~\cs[c]=@in1@$
\> \hspace{2em} $\to~@drop@~c~
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {s_q}
        {u'}
      $
\> \note{LocalDrop} \\

\> \> $~|~\cs[c]=@in1out1@$
\> \hspace{2em} $\to~@jump@~
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'}
      $
\> \note{ConnectedDrop}\\

\> \> $~|~\cs[c]=@in2@ ~\wedge~ (s_q[c]=@have@_F \vee s_q[c]=@pending@_F)$ 
\> \hspace{2em} $\to~@jump@~
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'}
      $
\> \note{SharedDropOne}\\

\> \> $~|~\cs[c]=@in2@ ~\wedge~ s_q[c]=@none@_F$
\> \hspace{2em} $\to~@drop@~c~
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'}
      $
\> \note{SharedDropBoth}
\\[1ex]

\\

\> $@pull@~c~x~(l'_o,u'_o)~(l'_c,u'_c)$ \\
\> \> $~|~\cs[c]=@in1@$ 
\> $\to~@pull@~c~x~
      \nextStep
        {l'_o}{s_p}
        {l_q}{s_q}
        {u'_o}
      ~
      \nextStep
        {l'_c}{s_p}
        {l_q}{s_q}
        {u'_c}
    $ 
\> \note{LocalPull}
\\[1ex]

\> \> $~|~(\cs[c]=@in2@ \vee \cs[c]=@in1out1@) ~\wedge~ s_p[c]=@pending@_F$ \\
\> \> $\to~@jump@~
      \nextStep
        {l'_o}
          {\HeapUpdateOne{c}{@have@_F}{s_p}}
        {l_q}
          {s_q}
        {\HeapUpdateOne{x}{@chan@~c}{u'_o}}
        $ 
\> \> \note{SharedPull} 
\\[1ex]

\> \> $~|~\cs[c]=@in2@ ~\wedge~ s_p[c]=@none@_F ~\wedge~ s_q[c]=@none@_F$ \\
\> \> $\to~@pull@~c~(@chan@~c)~
      \nextStep
        {l_p}
          {\HeapUpdateOne{c}{@pending@_F}{s_p}}
        {l_q}
          {\HeapUpdateOne{c}{@pending@_F}{s_q}}
        {[]}$
      \\
\> \> @                @
      $\nextStep
        {l_p}
          {\HeapUpdateOne{c}{@closed@_F}{s_p}}
        {l_q}
          {\HeapUpdateOne{c}{@closed@_F}{s_q}}
        {[]}
  $
\> \> \note{SharedPullInject}
\\[1ex]

\> \> $~|~(\cs[c]=@in2@ \vee \cs[c]=@in1out1@) ~\wedge~ s_p[c]=@closed@_F$ \\
\> \> $\to~@jump@~
      \nextStep
        {l'_c}{s_p}
        {l_q}{s_q}
        {u'_c}
  $
\> \> \note{SharedPullClosed}
\\[1ex]

\> $@close@~c~(l',u')$ \\
\> \> $~|~\cs[c]=@out1@$ 
\> $\to~@close@~c~
      \nextStep
        {l'}{s_p}
        {l_q}{s_q}
        {u'}
    $ 
\> \note{LocalClose}
\\

\> \> $~|~\cs[c]=@in1out1@ ~\wedge~ s_q[c]=@none@_F$ 
\> $\to~@close@~c~
      \nextStep
        {l'}{s_p}
        {l_q}{\HeapUpdateOne{c}{@closed@_F}{s_q}}
        {u'}
    $ 
\> \note{SharedClose}
\\[1ex]

\> $@exit@$
\> 
\> $\to~@exit@$
\> \note{LocalExit}
\\

\end{tabbing}

\caption{Fusion step for a single process of the pair.}

\label{fig:Finite:tryStep}
\end{figure*}

### 2017-05-09 2017mergingmerges/section/05-Evaluation.tex
ref meta-repa. put "extended version" for finite streams def.

@@
 However, this requires the fusion transform to analyse and reason about case statements and their predicates.
-It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
-By making the structure of finite streams explicit and constraining how processes use finite streams, we may be able to give more guarantees than by relying on heuristics for deciding equality of predicates.
+% It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
+By making the structure of finite streams explicit and constraining how processes use finite streams, it is not necessary to rely on heuristics for deciding equality of predicates.
+
+% Waffle waffle to get url on its own line
+This finite stream extension is described in more detail in the appendix of the extended version of this paper, which is available at \url{http://cse.unsw.edu.au/~amosr/papers/merges.pdf}.
 

### 2017-05-09 2017mergingmerges/section/XX-Related.tex
Get benchmark diagram on a better page
ref meta-repa. put "extended version" for finite streams def.

@@
 %!TEX root = ../Main.tex
+\eject
 \section{Related work}
-
-\TODO{Cite meta-repa \cite{ankner2013edsl}}
 
@@
 
-However, short cut fusion CITE relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
-Push-based short cut fusion systems \emph{do} exist \cite{gill1993short}, but support neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
+However, short cut fusion \cite{gill1993short} relies on inlining which, like pull-based streams, only occurs when there is a single consumer.
+Push-based short cut fusion systems \emph{do} exist, notably the original foldr/build formulation, but support neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.
 
-Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined, but when streams are used multiple times this causes excessive inlining, which duplicates work.
+Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined.
+When streams are used multiple times this causes excessive inlining, which duplicates work.
+This can change the semantics for effectful streams.
 
@@
 Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.
+
+Meta-repa \cite{ankner2013edsl} supports both array types in a similar way, using Template Haskell for code generation.
+It supports fusion on both array types.
+When arrays are used multiple times, they must be explicitly reified with a `force' operation to avoid duplication.
 

### 2017-05-12 2017mergingmerges/section/01-Introduction.tex
very minor type

@@
 To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it.
-Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.
+Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or we risk running out of local storage space.
 

### 2017-05-12 2017mergingmerges/section/02-Processes.tex
very minor type

@@
 There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
-The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.
+The result has two input streams, @sIn1@ and @sIn2@, and two output streams: @sOut1@ from @group@, and @sOut2@ from @merge@. 
+The shared input @sIn1@ is pulled by @merge@ at two places, and since both of these need to agree with when @group@ pulls, the @group@ instructions are duplicated at labels @F3@-@F5@ and @F13@-@F15@.
+The first set of instructions could be simplified to a single @push@ with constant propagation, as @f@ is initially true.
 
 To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
-The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.
-Although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere.
-When we instantiated combinators to create the original source processes we introduced fresh names at that point.
+Although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere: when we instantiated combinators to create the original source processes we introduced fresh names.
+
+% AR: Cut to make room for explanation why two copies of group pull
+% The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:FusionOrder} and \S\ref{s:Evaluation}.
 
@@
 Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists.
-As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirity before demanding any of the elements from the second list (@sUnion@).
+As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirety before demanding any of the elements from the second list (@sUnion@).
 If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before demanding any of @sIn2@, causing a space leak.

### 2017-05-12 2017mergingmerges/section/03-Definition.tex
Appendix: finite stream rules
very minor type

@@
 When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state.
-When defining the fusion transform in \S\ref{s:Fusion} we will use a version of $\InputState$ with only this staticaly known information.
+When defining the fusion transform in \S\ref{s:Fusion} we will use a version of $\InputState$ with only this statically known information.
 
@@
 When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input.
-If they all accept it then we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
+If they all accept it, we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
 \end{enumerate}

### 2017-05-12 2017mergingmerges/section/05-Evaluation.tex
very minor type

@@
 
-Stream fusion is ultimately performed for practical reasons. We want the fused result program to run faster than unfused source program.
+Stream fusion is ultimately performed for practical reasons: we want the fused result program to run faster than the original unfused program.
 
@@
 The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or at least partially hand-fused.
-Even with hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
+Even with partial hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
 The final group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.

### 2017-05-12 2017mergingmerges/section/A1-Finite.tex
very minor type

@@
 This function uses the same heuristics to decide which process to execute when both can progress, but now that the processes can finish with @exit@, we must take care to only finish the fused process once \emph{both} source processes are finished.
-The (DeferExit1) and (DeferExit2) clauses achieve this by allowing the other process if one is an @exit@.
+The (DeferExit1) and (DeferExit2) clauses achieve this by forcing the other process to run if one is an @exit@.
 Once both processes are finished, both new clauses will fail while (DeferPull1) succeeds, using the @exit@ from the first process.
@@
 Clause (SharedPullInject) applies when both processes need to pull from a shared input.
-As before we execute a real @pull@, this time with two branches.
+As before, we execute a real @pull@, this time with two branches.
 In the success branch, the input states are set to @pending@ as before.

### 2017-05-20 2017mergingmerges/section/04-Fusion.tex
Appendix: finite stream rules
maybe

@@
 Figure~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion.
-We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process.
-Each $Label_F$ consists of a $Label$ from one of the source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$.
-The definition of $Label$ is now recursive.
-While fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information.
+We extend the $\Label$ grammar with a new alternative, $\LabelF \times \LabelF$ for the labels in a fused result process.
+Each $\LabelF$ consists of a $\Label$ from a source process, paired with a map from $\Chan$ to the statically known part of that channel's current $\InputState$.
+When fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information.
 Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.
 
-% These new labels $Label_F$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
+% The definition of $\Label$ is now recursive.
+% These new labels $LabelF$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.
 
-% If the static $\InputState_F$ is $@pending@_F$, there is a value waiting to be pulled, do not know the actual value.
+% If the static $\InputStateF$ is $@pending@_F$, there is a value waiting to be pulled, do not know the actual value.
 
@@
 
-Still in Figure~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes.
+Still in Figure~\ref{fig:Fusion:Types}, $\ChanTypeTwo$ classifies how channels are used, and possibly shared, between two processes.
 Type @in2@ indicates that the two processes @pull@ from the same channel, so these actions must be coordinated.
@@
 % -------------------------------------
-Figure~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which of two processes to advance.
+Figure~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which process to advance.
+It starts by calling \ti{tryStep} for both processes.
+If both can advance, we use heuristics to decide which one to run first.
 
-Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@.
-This helps collect @jump@ instructions together, which are then easier for post-fusion optimization to handle (\S\ref{s:Optimisation}).
-As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Figure~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.
+Clauses (PreferJump1) and (PreferJump2) prioritize processes that can perform a @jump@.
+This helps collect @jump@ instructions together so they are easier for post-fusion optimization to handle (\S\ref{s:Optimisation}).
+The instruction for the second process was computed by calling \ti{tryStep} with the label arguments swapped, so in (PreferJump2) we need to swap the labels back with $\ti{swaplabels}$ (from Figure~\ref{fig:Fusion:Utils}).
 
-Similarly, clauses (DeferPull1) and (DeferPull2) defer @pull@ instructions. We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
+Similarly, clauses (DeferPull1) and (DeferPull2) defer @pull@ instructions: if one of the instructions is a @pull@, we advance the other one.
+We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.
 
-Clauses (Run1) and (Run2) apply when one process can advance using some other instruction.
-We try the first process first, and if that can advance then so be it.
-This priority means that fusion is left-biased, preferring advancement of the left process over the second.
+Clauses (Run1) and (Run2) apply when the above heuristics do not apply, or only one of the processes can advance.
+% We try the first process first, and if that can advance then so be it. This priority means that fusion is left-biased, preferring advancement of the left process over the second.
+
+Clause (Deadlock) applies when neither process can advance, in which case the processes cannot be fused together and fusion fails.
 
@@
 
+Clause (Blocked) returns @Nothing@ when no other clauses apply, meaning that this process is waiting for the other process to advance.
+
 % -------------------------------------
@@
 % Figure~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
-Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
+Function \ti{channels} computes the $\ChanTypeTwo$ map for a pair of processes.
 Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states.

### 2017-05-21 2017mergingmerges/section/02-Processes.tex

@@
 
-A \emph{combinator} is a template for a particular process which parameterizes it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
+A \emph{combinator} is a template for a process which parameterizes it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process.
 Each process implements a logical \emph{operator} --- so we use ``operator'' when describing the values being computed, but ``process'' and ``machine'' when referring to the implementation.
@@
 % AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
-When the predicate is true, the instruction at label @A2@ executes:
+When the predicate is true, the instruction at label @A2@ executes
 @(push sOut1 v A3 [ l = v, f = F ])@
@@
 The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@.
-The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream.
+The two heap variables @x1@ and @x2@ store the last values read from each input stream.
 The process starts by pulling from each of the input streams.
@@
 The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process.
-In our informal description of the fusion algorithm we instantiate the parameters of each combinator with arguments of the same names.
+In our informal description of the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.
 
@@
 We also associate each of the joint result states with a description of whether each source machine has already pulled a value from each of its input streams.
-For the @F0@ case in the top of Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
+For the @F0@ case at the top of Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})).
 The result state @F0@ represents a combination of the two source states @A0@ and @B0@.
@@
 
-Finally, note that in the result states @F0@ through @F3@ the state of the input streams transitions from `none', to `pending' then to `have'.
+Finally, note that in the result states @F0@ through @F3@, the state of the input streams transitions from `none', to `pending' then to `have'.
 The `none' state means that we have not yet pulled a value from the associated stream.
@@
 There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}.
-The result has two input streams, @sIn1@ and @sIn2@, and two output streams: @sOut1@ from @group@, and @sOut2@ from @merge@. 
-The shared input @sIn1@ is pulled by @merge@ at two places, and since both of these need to agree with when @group@ pulls, the @group@ instructions are duplicated at labels @F3@-@F5@ and @F13@-@F15@.
-The first set of instructions could be simplified to a single @push@ with constant propagation, as @f@ is initially true.
+The result process has two input streams, @sIn1@ and @sIn2@, and two output streams: @sOut1@ from @group@, and @sOut2@ from @merge@. 
+The shared input @sIn1@ is pulled by @merge@ instructions at two places, and since both of these need to agree with when @group@ pulls, the @group@ instructions are duplicated at @F3@-@F5@ and @F13@-@F15@.
+The first set of instructions could be simplified by constant propagation to a single @push@, as @f@ is initially true.
 
-To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process.
-Although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere: when we instantiated combinators to create the original source processes we introduced fresh names.
+To complete the implementation of our example from \S\ref{s:Introduction} we would now fuse this result process with a process from the final line of the example (also a @group@).
+Note that although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere, as when we instantiate combinators to create source processes we introduce fresh names.
 
@@
 \subsection{Breaking It Down}
-We started with a pure functional program as in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
-Importantly, we needed to \emph{break down} the definition of each operator into imperative statements so that we could interleave their execution appropriately.
+We started with a pure functional program in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph.
+We needed to \emph{break down} the definition of each operator into imperative statements so that we could interleave their execution appropriately.
 We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way.
 Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@.
-Instead, during fusion we have performed the job of a concurrent scheduler at compile time.
-The flow of control alternates between the implementation of each of the operators, but as the instructions are interleaved directly there is no overhead due to context switching --- as there would be in a standard concurrent implementation using multiple threads.
+Instead, during fusion we perform the job of a concurrent scheduler at compile time.
+In the result process the flow of control alternates between the instructions for both the @group@ and @merge@ operators, but as the instructions are interleaved directly there is no overhead due to context switching --- as there would be in a standard concurrent implementation using multiple threads.
 
@@
 However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are.
-This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams.
+This allows us to implement a wider range of operators, including ones with nested loops that work on segmented streams.
 
-Note that relying on lazy evaluation for an example such as @uniquesUnion@ does not solve the buffering problem.
+Note that relying on lazy evaluation for @uniquesUnion@ does not eliminate the need for unbounded buffering.
 Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists.
@@
 Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer.
-The space efficiency of our fused program relies critically on the fact that operators can also \emph{push} their result values directly to their consumers, and that the consumers cannot defer the handling of these values.
+The space efficiency of our fused program relies critically on the fact that processes can also \emph{push} their result values directly to their consumers, and that the consumers cannot defer the handling of these values.
 

### 2017-05-21 2017mergingmerges/section/A1-Finite.tex

@@
 \\
-$\InputState_F$ \> ::=  \> ~~~ $@none@_F ~|~ @pending@_F ~|~ @have@_F ~|~ @closed@_F$
+$\InputStateF$ \> ::=  \> ~~~ $@none@_F ~|~ @pending@_F ~|~ @have@_F ~|~ @closed@_F$
 \end{tabbing}
@@
 
-Also in Figure~\ref{fig:Finite:Instr}, the static input state used for fusion ($\InputState_F$) must now track closed streams.
+Also in Figure~\ref{fig:Finite:Instr}, the static input state used for fusion ($\InputStateF$) must now track closed streams.
 The new constructor $@closed@_F$ denotes that the stream is closed, while the rest is unchanged.
@@
 The (DeferExit1) and (DeferExit2) clauses achieve this by forcing the other process to run if one is an @exit@.
-Once both processes are finished, both new clauses will fail while (DeferPull1) succeeds, using the @exit@ from the first process.
+Once both processes are finished, both new clauses will fail while (Run1) succeeds, using the @exit@ from the first process.
 Another way to think of this is that if either process has work to do, the fused process still has work to do.
@@
 \begin{tabbing}
-M \= M \= M \= MMMMMMMMMMMMMMMMMMM \= \kill
+\ti{tryStepPair}\=$~\to~$\=M\kill
+
 $\ti{tryStepPair} ~:~ \ChanTypeMap$ \\
-\> \> $\to \Label_1 \to \Instr$ ~~~
-            $\to \Label_1 \to \Instr$ \\
-\> \> $\to \Maybe~\Instr$ \\
-$\ti{tryStepPair} ~\cs~l_p~i_p~l_q~i_q$ \\
+\> $\to$ \> $\LabelF \to \Instr \to \LabelF \to \Instr$ \\
+\> $\to$ \> $\Maybe~\Instr$ \\
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~ i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
-\> ~~ $\wedge~@exit@~\not\in~i_p'$ 
-   ~~~~~ $\to~i_p'$ 
-\> \> \> \note{DeferExit1} 
-\\[0.5ex]
+M \= $~|~(@pull@~\_~\_~\_ \gets i_p')~$ \= $\to$ \= $\Just (\ti{swaplabels}~i_q')~$ \= M \=\kill
+$\ti{tryStepPair} ~\cs~l_p~i_p~l_q~i_q~=$ \\
+\> $@match@~ (\ti{tryStep}~\cs~l_p~i_p~l_q,~\ti{tryStep}~\cs~l_q~i_q~l_p) ~@with@$ \\
+\> $(\Just i_p',~\Just i_q')$ \\
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
-\> ~~ $\wedge~@exit@~\not\in~i_q'$ 
-   ~~~~~ $\to~\ti{swaplabels}~i_q'$ 
-\> \> \> \note{DeferExit2} 
-\\[0.5ex]
+\> @ @$|~@exit@ \gets i_q'$ \> $\to$ \> $\Just i_p'$
+\> \note{DeferExit1} \\[0.5ex]
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~@jump@~(l,u)~\in~i_p'$ \\
-\> $\to~i_p'$
-\> \> \> \note{PreferJump1} 
-\\[0.5ex]
+\> @ @$|~@exit@ \gets i_p'$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
+\> \note{DeferExit2} \\[0.5ex]
 
-\> $~|~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p ~\wedge~@jump@~(l,u)~\in~i_q'$ \\
-\> $\to~\ti{swaplabels}~i_q'$ 
-\> \> \> \note{PreferJump2}
-\\[0.5ex]
+\> @ @$|~@jump@~\_ \gets i_p'$ \> $\to$ \> $\Just i_p'$
+\> \note{PreferJump1} \\[0.5ex]
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~ i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
-\> ~~ $\wedge~@pull@~c~x~(l,u)~\not\in~i_p'$ 
-   ~~~~~ $\to~i_p'$ 
-\> \> \> \note{DeferPull1} 
-\\[0.5ex]
+\> @ @$|~@jump@~\_ \gets i_q'$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
+\> \note{PreferJump2} \\[0.5ex]
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q ~\wedge~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ \\
-\> ~~ $\wedge~@pull@~c~x~(l,u)~\not\in~i_q'$ 
-   ~~~~~ $\to~\ti{swaplabels}~i_q'$ 
-\> \> \> \note{DeferPull2} 
-\\[0.5ex]
+\> @ @$|~@pull@~\_~\_~\_ \gets i_q'$ \> $\to$ \> $\Just i_p'$
+\> \note{DeferPull1} \\[0.5ex]
 
-\> $~|~i_p'~\in~\ti{tryStep}~\cs~l_p~i_p~l_q$ ~~~~~ $\to~i_p'$ 
-\> \> \> \note{Run1} 
-\\[0.5ex]
+\> @ @$|~@pull@~\_~\_~\_ \gets i_p'$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
+\> \note{DeferPull2} \\[0.5ex]
 
-\> $~|~i_q'~\in~\ti{tryStep}~\cs~l_q~i_q~l_p$ ~~~~~ $\to~\ti{swaplabels}~i_q'$
-\> \> \> \note{Run2} \\
+\> $(\Just i_p',~\_)$ \> $\to$ \> $\Just i_p'$
+\> \note{Run1} \\[0.5ex]
 
+\> $(\_, ~\Just i_q')$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
+\> \note{Run2} \\[0.5ex]
+
+\> $(\Nothing, ~\Nothing)$ \> $\to$ \> $\Nothing$
+\> \note{Deadlock}
 \end{tabbing}
@@
 M \= M \= MMMMMMMMMMMMMMMMMMMMMM \= MMMMMMMMMMMMMMMMMMMMMMMMMMMMMM \= \kill
-$\ti{tryStep} ~:~ \ChanTypeMap \to \Label_1 \to \Instr \to \Label_1 \to \Maybe~\Instr$ \\
+$\ti{tryStep} ~:~ \ChanTypeMap \to \LabelF \to \Instr \to \LabelF \to \Maybe~\Instr$ \\
 $\ti{tryStep} ~\cs~(l_p,s_p)~i_p~(l_q,s_q)~=~@match@~i_p~@with@$ \\
@@
 \> $@jump@~(l',u')$ 
-\> \> $\to~@jump@~
+\> \> $\to~\Just (@jump@~
       \nextStep
@@
         {l_q}{s_q}
-        {u'}
+        {u'})
       $ 
@@
 \> $@case@~e~(l'_t,u'_t)~(l'_f,u'_f)$
-\> \> $\to~@case@~e~
+\> \> $\to~\Just (@case@~e~
       \nextStep
@@
         {l_q}{s_q}
-        {u'_f}
+        {u'_f})
       $ 
@@
 \> \> $~|~\cs[c]=@out1@$ 
-\> $\to~@push@~c~e~
+\\
+\> \> $\to~\Just (@push@~c~e~
       \nextStep
@@
           {s_q}
-        {u'}
+        {u'})
       $ 
-\> \note{LocalPush}\\
+\> \> \note{LocalPush}\\
 
 \> \> $~|~\cs[c]=@in1out1@ ~\wedge~ s_q[c]=@none@_F$ 
-\> $\to~@push@~c~e~
+\\
+\> \> $\to~\Just (@push@~c~e~
       \nextStep
@@
           {\HeapUpdateOne{c}{@pending@_F}{s_q}}
-        {\HeapUpdateOne{@chan@~c}{e}{u'}}
+        {\HeapUpdateOne{@chan@~c}{e}{u'}})
       $
-\> \note{SharedPush}
+\> \> \note{SharedPush}
 \\[1ex]
@@
 \> \> $~|~\cs[c]=@in1@$
-\> \hspace{2em} $\to~@drop@~c~
+\> \hspace{2em} $\to~\Just (@drop@~c~
       \nextStep
@@
           {s_q}
-        {u'}
+        {u'})
       $
@@
 \> \> $~|~\cs[c]=@in1out1@$
-\> \hspace{2em} $\to~@jump@~
+\> \hspace{2em} $\to~\Just (@jump@~
       \nextStep
@@
 \> \> $~|~\cs[c]=@in2@ ~\wedge~ (s_q[c]=@have@_F \vee s_q[c]=@pending@_F)$ 
-\> \hspace{2em} $\to~@jump@~
+\> \hspace{2em} $\to~\Just (@jump@~
       \nextStep
@@
 \> \> $~|~\cs[c]=@in2@ ~\wedge~ s_q[c]=@none@_F$
-\> \hspace{2em} $\to~@drop@~c~
+\> \hspace{2em} $\to~\Just (@drop@~c~
       \nextStep
@@
 \> \> $~|~\cs[c]=@in1@$ 
-\> $\to~@pull@~c~x~
+\\
+\> \> $\to~\Just (@pull@~c~x~
       \nextStep
@@
         {l_q}{s_q}
-        {u'_c}
+        {u'_c})
     $ 
-\> \note{LocalPull}
+\> \> \note{LocalPull}
 \\[1ex]
@@
 \> \> $~|~(\cs[c]=@in2@ \vee \cs[c]=@in1out1@) ~\wedge~ s_p[c]=@pending@_F$ \\
-\> \> $\to~@jump@~
+\> \> $\to~\Just (@jump@~
       \nextStep
@@
           {s_q}
-        {\HeapUpdateOne{x}{@chan@~c}{u'_o}}
+        {\HeapUpdateOne{x}{@chan@~c}{u'_o}})
         $ 
@@
 \> \> $~|~\cs[c]=@in2@ ~\wedge~ s_p[c]=@none@_F ~\wedge~ s_q[c]=@none@_F$ \\
-\> \> $\to~@pull@~c~(@chan@~c)~
+\> \> $\to~\Just (@pull@~c~(@chan@~c)~
       \nextStep
@@
           {\HeapUpdateOne{c}{@closed@_F}{s_q}}
-        {[]}
+        {[]})
   $
@@
 \> \> $~|~(\cs[c]=@in2@ \vee \cs[c]=@in1out1@) ~\wedge~ s_p[c]=@closed@_F$ \\
-\> \> $\to~@jump@~
+\> \> $\to~\Just (@jump@~
       \nextStep
@@
         {l_q}{s_q}
-        {u'_c}
+        {u'_c})
   $
@@
 \> \> $~|~\cs[c]=@out1@$ 
-\> $\to~@close@~c~
+\> $\to~\Just (@close@~c~
       \nextStep
@@
         {l_q}{s_q}
-        {u'}
+        {u'})
     $ 
@@
 \> \> $~|~\cs[c]=@in1out1@ ~\wedge~ s_q[c]=@none@_F$ 
-\> $\to~@close@~c~
+\> $\to~\Just (@close@~c~
       \nextStep
@@
         {l_q}{\HeapUpdateOne{c}{@closed@_F}{s_q}}
-        {u'}
+        {u'})
     $ 
@@
 \> 
-\> $\to~@exit@$
+\> $\to~\Just @exit@$
 \> \note{LocalExit}
-\\
+\\[1ex]
+
+% \> @otherwise@ \>
+\> $\_$ \> $~|~ @otherwise@ $
+\> $\to ~ \Nothing$
+\> \note{Blocked}
 

### 2017-05-22 2017mergingmerges/section/04-Fusion.tex
Address reviewer comments

@@
 \end{code}
-
-In our current system, if we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, the final fusion transform fails.
+If we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, the final fusion transform fails.
 This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before pushing values to @s2@.
@@
 The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none.
+Fusability is an under-approximation for \emph{deadlock freedom} of the network.
 
@@
 
-In practice, the likelihood of fusion suceeding depends on the particular dataflow network being used, as well as the form of the processes in that network.
+In practice, the likelihood of fusion suceeding depends on the particular dataflow network.
 For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds.

### 2017-05-22 2017mergingmerges/section/03-Definition.tex
Refine discussion of non-det fusion order

@@
 
-In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require unbounded buffering.
-In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.
-In contrast, the role of our fusion transform is to construct one of these working orders statically.
+In the semantics, allowing the execution order of processes to be non-deterministic is critical, as it defines a search space where we might find an order that does not require unbounded buffering.
+For a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime.
+In contrast, the role of our fusion system is to construct one of these working orders statically.
 In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network were executed dynamically.
-In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.
+Fusion also eliminates the need to pass messages between processes --- once they are fused we can just copy values between heap locations.
 

### 2017-05-22 2017mergingmerges/section/03-Definition.tex
Say that injection is a meta-level operation

@@
 At any moment several processes may be able to take a step, while others are blocked.
-As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking.
-This enables the overall sequence of values on each output channel to be deterministic.
+As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking, which enables the overall sequence of values on each output channel to be deterministic.
 Unlike Kahn processes, pushing to a channel can also block.
@@
 The only intra-process communication is via channels and streams.
-Our model is ``pure data flow'' as there are no side-channels between processes.
-This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where processes can communicate via asynchronous messages as well as channels.
+Our model is ``pure data flow'' as there are no side-channels between processes --- in contrast to ``impure data flow'' systems such as StreamIt~\cite{thies2002streamit}.
 
@@
 Figure~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes.
+Injection is a meta-level operation, in contrast to @pull@ and @push@ which are instructions in the object language.
 The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''.
@@
 Figure~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process.
-The first set of rules handle specific instructions.
 The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels.

### 2017-05-22 2017mergingmerges/section/01-Introduction.tex
Fix wibble in Intro

@@
 When the fusion transform succeeds we know it has worked.
-There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.
+There is no need to inspect intermediate representations of the compiled code to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.
 

### 2017-05-22 2017mergingmerges/section/02-Processes.tex
Fix phrasing in discussion of fusing cases

@@
 From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@.
-In the result machine, @F3@ performs the case analysis from @A1@, moving to either @A2@ or @A3@, corresponding to @F4@ and @F5@ respectively.
-From state @F4@, the push at @A2@ is executed and moves to @A3@, corresponding to @F5@.
-Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @B1@ to @C0@.
-Because the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.
+In the result machine, @F3@ performs the case analysis from @A1@, moving to either @F4@ or @F5@, corresponding to @A2@ and @A3@ respectively.
+From state @F4@, the push at @A2@ is executed and moves to @F5@, corresponding to @A3@.
+Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @F5@ to @F6@, corresponding to @B1@ and @C0@ respectively.
+As the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.
 

### 2017-05-22 2017mergingmerges/section/02-Processes.tex
Fixing typos and wibbles

@@
 
-Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first the instruction at @A0@.
+Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first instruction at @A0@.
 

### 2017-05-22 2017mergingmerges/section/05-Evaluation.tex
Cut old commented out material on fusion for finite streams
benchmarks

@@
 
-% For a combinator such as @take@ which only needs a finite prefix of the input stream, it may be useful to add an instruction that allows a pulling process to explicitly disconnect from an input channel.
-After disconnecting, the process no longer needs to be coordinated with the producer, leaving the producer free to push as often as the other consumers allow.
-Our implementation supports finite streams, but our mechanized formalization does not cover it.
-
 It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements.
-However, this requires the fusion transform to analyse and reason about case statements and their predicates.
-% It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
+However, this requires the fusion transform to reason about case statements' predicates.
 By making the structure of finite streams explicit and constraining how processes use finite streams, it is not necessary to rely on heuristics for deciding equality of predicates.
@@
 
-% We now describe the extensions required to support finite streams.
-% We add a new @closed@ constructor to the \InputState~ to encode the end of the stream.
-% Once an input stream is in the closed state, it can never change to another state: it remains closed thereafter.
-% 
-% We modify the @pull@ instruction so that it has two output labels (like @case@).
-% The first label, the read branch, is executed as before when the pull succeeds and a value is read from the stream.
-% The second label, the close branch, is executed when the stream is closed, and no more values will ever be available.
-% After a pull takes the close branch, any subsequent pulls from that stream will also take the close branch.
-% 
-% We add two new instructions for closing output streams and disconnecting from input streams.
-% Closing an output stream $(@close@~\Chan~\Next)$ is similar to pushing an end-of-file marker to all readers.
-% As with @push@, the evaluation semantics of @close@ can only proceed if all readers are in a position to accept the end-of-file, but instead of setting the new \InputState~ to @pending@ with a value, the \InputState~ is set to @closed@.
-% After a stream has been closed, no further values can be pushed.
-% 
-% Disconnecting from input streams $(@disconnect@~\Chan~\Next)$ signals that a process is no longer interested in the values of a stream.
-% This can be used when a process requires the first values of a stream, but does not require the whole stream.
-% If a process read the first values of a stream and then stopped pulling, its \InputState~ buffer would fill up and never be cleared, so no other process would be able to continue pulling from that stream.
-% Disconnecting the stream allows other processes to use the stream without the disconnected process getting in the way of computation.
-% The evaluation semantics for @disconnect@ remove the channel from the inputs of the process.
-% After removing the channel from the inputs, when a writing process tries to inject values, this process will just be ignored rather than inserting into the \InputState~ buffer and potentially causing writing to block.
-% After a process disconnects from an input channel, it can no longer pull from that channel.
-% 
-% We also add an instruction for terminating the process (@done@).
-% After all input streams have been read to completion or disconnected and output streams closed, the process may execute @done@ to signal that processing is complete.
-% 
-% The fusion definition must be extended to deal with these new instructions.
-% The static input state has a @closed@ constructor added and disconnection is encoded by removal from the input state, and the \ti{tryStep} changes more or less follow the evaluation changes.
-% Shared and connected pulls now deal with two more possibilities in the input state: the input may be closed in which case the close branch of the pull is taken; or the other process may have disconnected in which case the pull is executed as in the non-shared non-connected case.
-% Connected pushes must also deal with when the other process has disconnected in which case the push is executed as if it were non-connected.
-% For @in1@ and @out1@ channels, the new @close@ and @disconnect@ instructions are used as normal with no coordination required.
-% For connected @close@, as with @push@, the receiving process must have @none@ and the next step performs the @close@ and sets the input state to @closed@.
-% For shared @disconnect@, the @disconnect@ is only performed after both processes have disconnected; otherwise the entry is just removed from the input state.
-% For connected @disconnect@, the @disconnect@ is not performed and the entry is removed from the input state.
-% 
-% Finally, \ti{tryStepPair} is modified so that @done@ is performed when both machines are @done@.
-% 
-% These modifications allow our system to fuse finite streams as well as infinite.
+% It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
 
@@
 To show practical examples, we use the finite stream extension mentioned in~\S\ref{s:Finite}.
-We present three benchmarks: one array algorithm, and two simple file operations.
+We present three benchmarks: two file-based benchmarks, and one array algorithm.
+
+For the file benchmarks, we compare against three Haskell streaming libraries: `Conduit', `Pipes', and `Streaming'.
+These streaming libraries are pull-based, and do not naturally support multiple outputs: the split in the dataflow graph must be hand-fused, or somehow rewritten as a straight-line computation.
+These libraries also have a monadic interface, which allows the structure of the dataflow graph to depend on the values.
+This expressiveness comes at a cost: if the dataflow graph can change dynamically, then we cannot statically fuse it.
+% This price is paid even when the dataflow graph is static, as the same monadic structure is used: because repetition is expressed as an unfolding dynamic graph, even computations that would be static in other systems must be expressed dynamically, reducing the possibility for fusion.
+Note that Conduit uses the term `fusion' for connecting pipelines, but this is not `fusion' in the sense of optimisation.
+% Unfortunate terminology.. not sure.
+
+The first file benchmark simply appends two files while counting the lines.
+In Pipes and Conduit, counting the lines is implemented as a pipe which counts each line before passing it along.
+The first group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
+
+The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
+The output lines are also counted.
+Even with partial hand-fusion because of the multiple outputs, the Pipes and Conduit programs are slower than ours, as well as losing the abstraction benefits from using a high-level library.
+The `Streaming' library allows streams to be shared in a fairly straightforward way and does not require hand-fusion, but is also the slowest in this benchmark.
+The second group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
 
@@
 We compare against a hand-fused program and a @Data.Vector@ program, which uses shortcut fusion.
-The shortcut fusion system cannot fuse both operations into a single loop, and both operations recompute the distances between the line and each point.
+The shortcut fusion system cannot fuse both operations into a single loop, and both operations must recompute the distances between the line and each point.
 % This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
@@
 
-The first group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 80MB of data, or five million points.
-The hand-fused version is the fastest, while our version is slower, but still faster than the vector version.
-The fact that our version is slower than the hand-fused version is surprising, as the generated GHC core is almost identical: the only difference is an extra continuation bound inside the loop, which acts as a sort of indirect jump in the loop body.
-% It is possible that recent work on optimising continuations in GHC~\cite{downen2016sequent} will solve this.
-
-The first file benchmark simply appends two files, while counting the lines.
-We compared against three Haskell streaming libraries: Conduit, Pipes, and Streaming.
-The second group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
-The absolute performance here is poor because all are using line-buffered IO; in practice one would use chunked IO, but the overhead per chunk would remain.
-
-The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
-The output lines are also counted.
-The three streaming libraries are pull-based, and do not support multiple outputs: the program must be written in a convoluted way or at least partially hand-fused.
-Even with partial hand-fusion, the Pipes and Conduit programs are slower than ours, as well as losing any abstraction benefits from using a streaming library.
-The final group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
-
-% Fused result programs like the one in Figure \ref{fig:Process:Fused} can be lowered directly to a target language, such as Haskell, C or LLVM abstract assembly.
-The details of such lowering surely affect the performance of the final program, but as they are largely unrelated to the fusion algorithm itself, we focus on the form of the fused program expressed directly in the process language.
-
-% -----------------------------------------------------------------------------
+The final group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 120MB of data, or eight million points.
 
@@
 \begin{axis}[
-  symbolic x coords={Quickhull, Append, Partition, end},
+  symbolic x coords={Append, Partition, Quickhull, end},
         ylabel=Runtime (ms),
@@
 ]
-\addplot coordinates { (Quickhull, 75) (Append, 88) (Partition, 60) (end, 0) };
-\addplot coordinates {(Quickhull, 110) (Append, 86) (Partition, 61) (end,0)};
-\addplot coordinates { (Quickhull, 130) (Append, 0)  };
-% \addplot coordinates { (Quickhull, 200) (Append, 0) };
-\addplot coordinates {  (Append, 282) (Partition, 136) (end,0) };
-\addplot coordinates {  (Append, 270) (Partition, 166) (end,0) };
-\addplot coordinates {  (Append, 168) (Partition, 268) (end,0) };
+\addplot coordinates {(Append,  88) (Partition,  60) (Quickhull, 126) (end, 0) };
+\addplot coordinates {(Append,  86) (Partition,  61) (Quickhull, 202) (end, 0) };
+\addplot coordinates {                               (Quickhull, 274) (end, 0) };
+\addplot coordinates {(Append, 282) (Partition, 136) (Quickhull,   0) };
+\addplot coordinates {(Append, 270) (Partition, 166) (Quickhull,   0) };
+\addplot coordinates {(Append, 168) (Partition, 268) (Quickhull,   0) };
 \legend{Handfused, Ours, Vector, Conduit, Pipes, Streaming}

### 2017-05-22 2017mergingmerges/section/05-Evaluation.tex
Split appendices into separate document and add discussion of result size
attempt to clean up graph, not quite right

@@
 % -----------------------------------------------------------------------------
-%\subsection{Result Size}
-%
-%As with any fusion system, we must be careful that the size of the result code does not become too large when more and more processes are fused together.
-The left of Figure~\ref{fig:bench:outputsize} shows the maximum number of output states in the result when a particular number of processes are fused together in a pipelined-manner.
-To produce this graph we programmatically generated dataflow networks for \emph{all possible} pipelined combinations of the @map@, @filter@, @scan@, @group@ and @merge@ combinators, and tried all possible fusion orders consiting of adjacent pairs of processes.
-The @merge@ combinator itself has two inputs, so only works at the very start of the pipeline --- we present result for pipelines with and without a @merge@ at the start.
-The right of Figure~\ref{fig:bench:outputsize} shows the number of states in the result when the various combinations of combinators are fused in parallel, for example, we might have a @map@ and a @filter@ processing the same input stream.
-In both cases the number of states in the result process grows linearly with the number of processes.
-In all combinations, with up to 7 processes there are less than 100 states in the result process.
-%%%% AR: Note that the split version also has only one merge?
-%
-%The size of the result process is roughly what one would get when inlining the definitions of each of the original source processes.
-This is common with other systems based on inlining and/or template meta-programming, and is not prohibitive.
-%
-%On the other hand, Figure~\ref{fig:bench:exponential} shows the results for a pathological case where the size of the output program is exponential in the number of input processes.
-The source dataflow networks consists of N merge processes, N+1 input streams, and a single output stream.
-The output of each merge process is the input of the next, forming a chain of merges.
-In source notation the network for N = 3 is @sOut = merge sIn1 (merge sIn2 (merge sIn3 sIn4))@.
-%
-%When fusing two processes the fusion algorithm essentially compares every state in the first process with every state in the second, computing a cross product.
-During the fusion transform, as states in the result process are generated they are added to a finite map --- the @instrs@ field of the process definition.
-The use of the finite map ensures that identical states are always combined, but genuinely different states always make it into the result.
-%
-%In the worst case, fusion of two processes produces O($n*m$) different states, where $n$ and $m$ are the number of states in each.
-If we assume the two processes have about the same number of states then this is O($n^2$).
-Fusing the next process into this result yields O($n^3$), so overall the worst case number of states in the result will be O($n^k$), where $k$ is the number of processes fused.
-%
-%In the particular case of @merge@, the implementation has two occurrences of the @push@ instruction.
-During fusion, the states for the consuming process are inlined at each occurrence of @push@.
-These states are legitimately different because at each occurence of @push@ the input channels of the merge process are in different channel states, and these channel states are included in the overall process state.
-
-% -----------------------------------------------------------------------------
 \subsection{Optimisation and Drop Instructions}
@@
         enlargelimits=0.01,
-        ybar interval=0.7,
-  width=7.5cm, height=4.5cm,
+        ybar interval=0.8,
+  width=8.0cm, height=4.5cm,
   legend style={at={(0.5,-0.2)},anchor=north, legend columns=3}
@@
 \addplot coordinates {(Append, 168) (Partition, 268) (Quickhull,   0) };
+\addplot coordinates {(Append, 0)   };
 \legend{Handfused, Ours, Vector, Conduit, Pipes, Streaming}

### 2017-05-23 2017mergingmerges/section/05-Evaluation.tex
benchmark graph

@@
 \begin{tikzpicture}
+\usetikzlibrary{patterns}
+
+% Bar chart for benchmarks:
+% This is a complicated bar chart because there are six series (conduit, pipes, vector..), but they are not all defined at all points (append, partition, quickhull).
+% We need to split it into three different pictures, overlaid on top of each other.
+% The first picture just defines the legend, the axis labels, etc.
 \begin{axis}[
   symbolic x coords={Append, Partition, Quickhull, end},
-        ylabel=Runtime (ms),
+  xtick=data, ylabel=Runtime (ms),
+  % All pictures need to specify same upper and lower bounds so they line up together
   ymin=0, ymax=300,
+  xmin=Append, xmax=Quickhull,
         enlargelimits=0.01,
-        ybar interval=0.8,
-  width=8.0cm, height=4.5cm,
-  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3}
+  enlarge x limits=0.30,
+        ybar=1pt,
+        bar width=8pt,
+  width=7.5cm, height=4.5cm,
+  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3},
 ]
-\addplot coordinates {(Append,  88) (Partition,  60) (Quickhull, 126) (end, 0) };
-\addplot coordinates {(Append,  86) (Partition,  61) (Quickhull, 202) (end, 0) };
-\addplot coordinates {                               (Quickhull, 274) (end, 0) };
-\addplot coordinates {(Append, 282) (Partition, 136) (Quickhull,   0) };
-\addplot coordinates {(Append, 270) (Partition, 166) (Quickhull,   0) };
-\addplot coordinates {(Append, 168) (Partition, 268) (Quickhull,   0) };
-\addplot coordinates {(Append, 0)   };
-\legend{Handfused, Ours, Vector, Conduit, Pipes, Streaming}
+% The colours need to be specified manually, to force the legend entries to show up..
+% The colours also need to match the colours specified for each series data.
+\addlegendimage{red,fill=red!40!white}
+\addlegendimage{brown,pattern=north east lines,pattern color=brown}
+\addlegendimage{blue,fill=blue!40!white}
+\addlegendimage{teal,fill=teal!40!white}
+\addlegendimage{orange,fill=orange!40!white}
+\addlegendimage{gray,fill=gray!40!white}
+\legend{Handfused, Ours, Conduit, Pipes, Streaming, Vector}
 \end{axis}
+
+% This picture shows the values for Append and Partition, since they have the same series defined
+\begin{axis}[
+  symbolic x coords={Append, Partition, Quickhull, end},
+  ticks=none,
+  ymin=0, ymax=300,
+  xmin=Append, xmax=Quickhull,
+  enlargelimits=0.01,
+  enlarge x limits=0.30,
+  ybar=1pt,
+  bar width=8pt,
+  width=7.5cm, height=4.5cm,
+]
+% Handfused
+\addplot[red,fill=red!40!white] coordinates {(Append,  88) (Partition,  60)   };
+% Ours
+\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {(Append,  86) (Partition,  61)   };
+% Conduit
+\addplot[blue,fill=blue!40!white] coordinates {(Append, 282) (Partition, 136)   };
+% Pipes
+\addplot[teal,fill=teal!40!white] coordinates {(Append, 270) (Partition, 166)  };
+% Streaming
+\addplot[orange,fill=orange!40!white] coordinates {(Append, 168) (Partition, 268)  };
+\end{axis}
+
+% This picture shows the Quickhull results
+\begin{axis}[
+  symbolic x coords={Append, Partition, Quickhull, end},
+  ticks=none,
+  ymin=0, ymax=300,
+  xmin=Append, xmax=Quickhull,
+  enlargelimits=0.01,
+  enlarge x limits=0.30,
+  ybar=1pt,
+  bar width=8pt,
+  width=7.5cm, height=4.5cm,
+]
+% Handfused
+\addplot[red,fill=red!40!white] coordinates {  (Quickhull, 126)  };
+% Ours
+\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {  (Quickhull, 202)  };
+% Vector
+\addplot[gray,fill=gray!40!white] coordinates {(Quickhull, 274)  };
+\end{axis}
+
 \end{tikzpicture}
@@
 \end{figure}
-
-% Quickhull
-% \addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
-% Append
-% \addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
-% Partition
-% \addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };

### 2017-05-24 2017mergingmerges/figures/bench.tex
split out bench figure to separate file

--- 2017-05-23 2017mergingmerges/section/05-Evaluation.tex
@@
-%!TEX root = ../Main.tex
-
-% -----------------------------------------------------------------------------
-\section{Implementation}
-\label{s:Evaluation}
-
-Stream fusion is ultimately performed for practical reasons: we want the fused result program to run faster than the original unfused program.
-
-% -----------------------------------------------------------------------------
-\subsection{Finite streams}
-\label{s:Finite}
-
-The processes we have seen so far deal with infinite streams, but in practice most streams are finite.
-Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output.
-We have focussed on the infinite stream version as it is simpler to explain and prove, but supporting finite streams does not require substantial conceptual changes.
-
-Unlike infinite streams, pulling from a finite stream can fail, meaning the stream is finished.
-We therefore modify the @pull@ instruction to have two output labels: one to execute when a value is pulled, and the other to execute when the stream is finished.
-On the pushing end, we also need some way of finishing streams, so we add a new instruction to close an output stream.
-
-During evaluation we need some way of knowing whether a stream is closed, which can be added as an extra constructor in the \InputState~ type.
-The same constructor is added to the static input state used by fusion.
-In this way, for any changes made to evaluation, the analogous static change must be made in the fusion transform.
-
-It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements.
-However, this requires the fusion transform to reason about case statements' predicates.
-By making the structure of finite streams explicit and constraining how processes use finite streams, it is not necessary to rely on heuristics for deciding equality of predicates.
-
-% Waffle waffle to get url on its own line
-This finite stream extension is described in more detail in the appendix of the extended version of this paper, which is available at \url{http://cse.unsw.edu.au/~amosr/papers/merges.pdf}.
-
-% It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.
-
-% -----------------------------------------------------------------------------
-\subsection{Benchmarks}
-We have implemented this system using Template Haskell in a library called @folderol@\footnote{\url{https://github.com/amosr/folderol}}.
-To show practical examples, we use the finite stream extension mentioned in~\S\ref{s:Finite}.
-We present three benchmarks: two file-based benchmarks, and one array algorithm.
-
-For the file benchmarks, we compare against three Haskell streaming libraries: `Conduit', `Pipes', and `Streaming'.
-These streaming libraries are pull-based, and do not naturally support multiple outputs: the split in the dataflow graph must be hand-fused, or somehow rewritten as a straight-line computation.
-These libraries also have a monadic interface, which allows the structure of the dataflow graph to depend on the values.
-This expressiveness comes at a cost: if the dataflow graph can change dynamically, then we cannot statically fuse it.
-% This price is paid even when the dataflow graph is static, as the same monadic structure is used: because repetition is expressed as an unfolding dynamic graph, even computations that would be static in other systems must be expressed dynamically, reducing the possibility for fusion.
-Note that Conduit uses the term `fusion' for connecting pipelines, but this is not `fusion' in the sense of optimisation.
-% Unfortunate terminology.. not sure.
-
-The first file benchmark simply appends two files while counting the lines.
-In Pipes and Conduit, counting the lines is implemented as a pipe which counts each line before passing it along.
-The first group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.
-
-The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
-The output lines are also counted.
-Even with partial hand-fusion because of the multiple outputs, the Pipes and Conduit programs are slower than ours, as well as losing the abstraction benefits from using a high-level library.
-The `Streaming' library allows streams to be shared in a fairly straightforward way and does not require hand-fusion, but is also the slowest in this benchmark.
-The second group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.
-
-Quickhull is a divide-and-conquer spatial algorithm to find the smallest convex hull containing all points.
-At its core is an operation called `filterMax' which takes a line and an array of points, and finds the farthest point above the line, as well as all points above the line.
-
-We compare against a hand-fused program and a @Data.Vector@ program, which uses shortcut fusion.
-The shortcut fusion system cannot fuse both operations into a single loop, and both operations must recompute the distances between the line and each point.
-% This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
-% We compared both, and recomputing was significantly faster.
-% However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.
-
-The final group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over roughly 120MB of data, or eight million points.
-
-% -----------------------------------------------------------------------------
-\subsection{Optimisation and Drop Instructions}
-\label{s:Optimisation}
-After we have fused two processes together, it may be possible to simplify the result before fusing in a third.
-Consider the result of fusing @group@ and @merge@ which we saw back in Figure~\ref{fig:Process:Fused}.
-At labels @F1@ and @F2@ are two consecutive @jump@ instructions.
-The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@.
-In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code.
-The (PreferJump) clauses of Figure~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.
-
-Other @jump@ instructions like the one at @F5@ have no associated update expressions, and thus can be eliminated completely.
-Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.
-
-Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has fewer states.
-Provided we do not change the order of instructions that require synchronization with other processes (@pull@, @push@ or @drop@), the fusibility of the overall process network will not be affected.
-
-% Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable.
-In Figure~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction.
-Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable.
-Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@.
-When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.
-
-When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable.
-This is the case in Figure~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@.
-To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time.
-Moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.
-
-% In the result, the interleaved instructions from both source processes then share the same heap variable.
-
-% In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream.
-
 \begin{figure}
@@
 % The colours also need to match the colours specified for each series data.
+\addlegendimage{brown,pattern=north east lines,pattern color=brown}
 \addlegendimage{red,fill=red!40!white}
-\addlegendimage{brown,pattern=north east lines,pattern color=brown}
 \addlegendimage{blue,fill=blue!40!white}
@@
 \addlegendimage{gray,fill=gray!40!white}
-\legend{Handfused, Ours, Conduit, Pipes, Streaming, Vector}
+\legend{Ours, Handfused, Conduit, Pipes, Streaming, Vector}
 \end{axis}
@@
 ]
+% Ours
+\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {(Append,  86) (Partition,  61)   };
 % Handfused
 \addplot[red,fill=red!40!white] coordinates {(Append,  88) (Partition,  60)   };
-% Ours
-\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {(Append,  86) (Partition,  61)   };
 % Conduit
@@
 ]
+% Ours
+\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {  (Quickhull, 202)  };
 % Handfused
 \addplot[red,fill=red!40!white] coordinates {  (Quickhull, 126)  };
-% Ours
-\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {  (Quickhull, 202)  };
 % Vector
@@
 \end{axis}
+
+% Runtimes for Quickhull 40MB:
+% Conduit 2-pass:     300ms
+% Conduit 1-pass:     264ms
+% Hand:                 3ms
+% Folderol:             6ms
+% Vector recompute:     8ms
+% Vector store:        11ms
 

### 2017-05-24 2017mergingmerges/section/05-Evaluation.tex
benchmark: put us first
Add Conduit Quickhull runtime as comment
Not sure what to do: showing Conduit runtimes makes difference between
Folderol and Vector insignificant.
split out bench figure to separate file

--- 2017-05-22 2017mergingmerges/section/05-Evaluation.tex
@@
 % In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream. 
-
-\begin{figure}
-\begin{tikzpicture}
-\begin{axis}[
-  symbolic x coords={Append, Partition, Quickhull, end},
-        ylabel=Runtime (ms),
-  ymin=0, ymax=300,
-        enlargelimits=0.01,
-        ybar interval=0.8,
-  width=8.0cm, height=4.5cm,
-  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3}
-]
-\addplot coordinates {(Append,  88) (Partition,  60) (Quickhull, 126) (end, 0) };
-\addplot coordinates {(Append,  86) (Partition,  61) (Quickhull, 202) (end, 0) };
-\addplot coordinates {                               (Quickhull, 274) (end, 0) };
-\addplot coordinates {(Append, 282) (Partition, 136) (Quickhull,   0) };
-\addplot coordinates {(Append, 270) (Partition, 166) (Quickhull,   0) };
-\addplot coordinates {(Append, 168) (Partition, 268) (Quickhull,   0) };
-\addplot coordinates {(Append, 0)   };
-\legend{Handfused, Ours, Vector, Conduit, Pipes, Streaming}
-\end{axis}
-\end{tikzpicture}
-\caption{Runtime for benchmarks; lower is faster.}
-\label{fig:bench:all}
-\end{figure}
-
-% Quickhull
-% \addplot coordinates {(200,Store) (130,Recompute) (110,Ours) (75,Hand) (0,end) };
-% Append
-% \addplot coordinates {(282,Conduit) (270,Pipes) (168,Streaming) (86,Ours) (0,end) };
-% Partition
-% \addplot coordinates {(1200,Streaming) (764,Pipes) (658,Conduit hand) (288,Ours) (0,end) };
+\input{figures/bench.tex}

### 2017-08-02 2017mergingmerges/section/01-Introduction.tex
start on final edits

@@
 This example has a few interesting properties.
-Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
+Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
+%: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
 
@@
 
-To implement this program we might write each operator as its own concurrent process, sending stream elements over intra-process channels.
-Developing this could be easy or hard, depending on the available language features for concurrency.
-However, the \emph{performance tuning} of such a system, such as using back-pressure to prevent buffers from being overrun, or how to chunk stream data to amortize communication overhead, is invariably a headache.
+% To implement this program we might write each operator as its own concurrent process, sending stream elements over inter-process channels.
+To implement this we might translate each operator into its own process in a concurrent network.
+This could be easy or hard, depending on the available language features for concurrency.
+However, the \emph{performance tuning} of such a system, such as using back-pressure to prevent buffers from being overrun, or chunking stream data to amortize communication overhead, is invariably a headache.
 

### 2017-08-02 2017mergingmerges/section/XX-Related.tex
start on final edits

@@
 
-In relation to process calculi, synchronised product has been suggested as a method for fusing Kahn process networks together~\cite{fradet2004network}, but does not appear to have been implemented or evaluated.
-The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
-When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
-In our system the use of stream buffer variables allows some leeway in when processes must take shared steps.
+Filter Fusion~\cite{proebsting1996filter} uses a similar algorithm, but only supports processes with at most one input and one output channel.
+Each output channel may only be used once.
+This means combinators such as @zip@, @unzip@, @append@, and @partition@ cannot be expressed.
+Here a consumer and a producer are fused by alternating between the two; when the consumer needs input, control is passed to the producer, and when the producer has output, control is passed to the consumer.
+This is sufficient when only one channel is shared, but when multiple channels are shared finer-grained coordination is required.
+
+% This is sufficient for producer-consumer fusion, but two consumers pulling from the same input channel need not follow the same alternation.
+% This is sufficient when only one , but when multiple channels are shared between two processes, as may be the case in an intermediate fused result of a larger program, it is not obvious whether simply alternating the two processes is correct.
+
+% The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
+% When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
+% In our system the use of stream buffer variables allows some leeway in when processes must take shared steps.
 
@@
 StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}.
-Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow structures.
+Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow.
 

### 2017-08-15 2017mergingmerges/section/XX-Related.tex
filter fusion supports straight line computations, no splits or joins
Copy edits to related work

@@
 
-Filter Fusion~\cite{proebsting1996filter} uses a similar algorithm, but only supports processes with at most one input and one output channel.
-Each output channel may only be used once.
-This means combinators such as @zip@, @unzip@, @append@, and @partition@ cannot be expressed.
-Here a consumer and a producer are fused by alternating between the two; when the consumer needs input, control is passed to the producer, and when the producer has output, control is passed to the consumer.
-This is sufficient when only one channel is shared, but when multiple channels are shared finer-grained coordination is required.
+Like our own Machine Fusion,  Filter Fusion~\cite{proebsting1996filter} also statically interleaves the code of producer and consumer processes.
+Each process must have a single input and output channel, so common operators like @zip@, @unzip@, @append@, @partition@ and so on are not supported.
+Given an adjacent producer and consumer pair, Filter Fusion alternately assigns control to the code of each.
+When the consumer needs input control is passed to the producer, and when the producer produces its value control is passed back to the consumer.
+This simple scheduling algorithm works only for straight line pipelines of processes.
+Machine fusion provides a finer grained interleaving of code, which is nessesary to support branching dataflows that contain both splits and joins.
+
+%Filter Fusion~\cite{proebsting1996filter} uses a similar algorithm, but only supports processes with at most one input and one output channel.
+% This means combinators such as @zip@, @unzip@, @append@, and @partition@ cannot be expressed.
+%Also, the process network must be a straight line, with no splits or joins.
+%Here a consumer and a producer are fused by alternating between the two; when the consumer needs input, control is passed to the producer, and when the producer has output, control is passed to the consumer.
+%This is sufficient when only one channel is shared, but when multiple channels are shared finer-grained coordination is required.
 

### 2017-08-15 2017mergingmerges/distro/merges-PPDP2017-final.tex
add final submission and remove dot2tex import

\documentclass[sigconf]{acmart}
% \documentclass[sigconf,review]{acmart}\settopmatter{printfolios=true}

\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{mathptmx}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{adjustbox}
\usepackage{hyperref}
\usepackage{alltt}
\usepackage{url}
\usepackage{float}
\usepackage{minipage-marginpar}
\usepackage{style/utils}
\usepackage{style/code}
\usepackage{style/proof}
\usepackage{style/keywords}
\usepackage{style/layout}
\usepackage{style/judgements}

% for combinator pictures
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{shapes,arrows}

%\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}
% \citestyle{acmauthoryear}

% -----------------------------------------------------------------------------
\begin{document}

\copyrightyear{2017}
\acmYear{2017}
\setcopyright{acmlicensed}
\acmConference{PPDP'17}{October 9--11, 2017}{Namur, Belgium}\acmPrice{15.00}\acmDOI{10.1145/3131851.3131865}
\acmISBN{978-1-4503-5291-8/17/10}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003753.10003760</concept_id>
<concept_desc>Theory of computation~Streaming models</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\renewcommand{\textrightarrow}{$\rightarrow$}

\ccsdesc[500]{Theory of computation~Streaming models}

\title{Machine Fusion}
\subtitle{Merging merges, more or less}

\author{Amos Robinson}
\affiliation{Ambiata and UNSW (Australia)}
\email{amosr@cse.unsw.edu.au}

\author{Ben Lippmeier}
\affiliation{Digital Asset and UNSW (Australia)}
\email{benl@ouroborus.net}

\makeatactive
\begin{abstract}
Compilers for stream programs often rely on a fusion transformation to convert the implied dataflow network into low-level iteration based code. Different fusion transformations handle different sorts of networks, with the distinguishing criteria being whether the network may contain splits and joins, and whether the set of fusible operators can be extended. We present the first fusion system that simultaneously satisfies all three of these criteria: networks can contain splits, joins, and new operators can be added to the system without needing to modify the overall fusion transformation.
\end{abstract}

\maketitle

\section{Introduction}
\label{s:Introduction}

Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on them. The identifiers from both streams are sorted, but may include duplicates. We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of both streams. Here is how we might write the source code, where @S@ is for @S@-tream.
\begin{code}
  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
  uniquesUnion sIn1 sIn2
   = let  sUnique = group sIn1
          sMerged = merge sIn1 sIn2
          sUnion  = group sMerged
     in   (sUnique, sUnion)
\end{code}

The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted. This example has a few interesting properties. Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves.
%: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.

Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle its elements once. To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both @group@ and @merge@ are ready to receive it. Finally, as the stream length is unbounded, we cannot buffer arbitrarily many elements, or we risk running out of memory.

% To implement this program we might write each operator as its own concurrent process, sending stream elements over inter-process channels.
To implement this we might translate each operator into its own process in a concurrent network. This could be easy or hard, depending on the available language features for concurrency. However, the \emph{performance tuning} of such a system, such as using back-pressure to prevent buffers from being overrun, or how to chunk stream data to amortize communication overhead, is invariably a headache.

Instead, we would prefer to use \emph{stream fusion}, which is a program transformation that takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example.

As observed by \citet{kay2009you}, both pull and push fusion have fundamental limitations. Pull systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a stream is used by multiple consumers. We refer to this situation as a \mbox{\emph{split} --- in } our example the input stream @sIn1@ is split into both the @group@ and @merge@ consumers. Push systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs. We refer to this as a \emph{join} --- in our example the @merge@ operator expresses a join. Some systems support both: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

In this paper we present Machine Fusion, a new approach. Each operator is expressed as a sequential imperative program which \emph{pulls} from input streams, and \emph{pushes} to output streams. We view each operator as a process in a concurrent process network. Our fusion transform then \emph{sequentializes} the concurrent process network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers. When fusion succeeds we know it has worked. There is no need to inspect the compiled code to debug poor performance, which is a common problem in systems based on general purpose transformations \cite{lippmeier2012:guiding}.

In summary, we make the following contributions:
\begin{itemize}
\item a process calculus for infinite streaming programs (\S\ref{s:Processes});
\item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
\item benchmarks showing significant performance gains (\S\ref{s:Evaluation});
\item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).
\end{itemize}

Our fusion transform for infinite stream programs also serves as the basis for an \emph{array} fusion system, using a natural extension to finite streams. We discuss this extension in \S\ref{s:Finite}.

% BL: This is now discussed in S2.5 Breaking It Down.
% We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

% TODO: move to related work.
% The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

% -----------------------------------------------------------------------------
\section{Processes and Machines}
\label{s:Processes}

A \emph{process} in our system is a simple imperative program with a local heap. A process pulls source values from an arbitrary number of input streams and pushes result values to at least one output stream. The process language is an intermediate representation we use when fusing the overall dataflow network. When describing the fusion transform we describe the control flow of the process as a state machine, hence Machine Fusion.

A \emph{combinator} is a template for a process which parameterizes it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process. Each process implements a logical \emph{operator} --- so we use ``operator'' when describing the values being computed, but ``process'' and ``machine'' when referring to the implementation.

% -----------------------------------------------------------------------------
\subsection{Grouping}

The definition of the @group@ combinator which detects groups of successive identical elements in the input stream is given in Figure~\ref{fig:Process:Group}. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream @[1,2,2,3]@, the process will produce the output @[1,2,3]@. We include the concrete representation and a diagram of the process when viewed as a state machine.

The @group@ combinator has two parameters, @sIn1@ and @sOut1@, which bind the input and output streams respectively. The \emph{nu-binders} \mbox{($\nu$ @(f: Bool) (l: Nat)@...)} indicate that each time the @group@ combinator is instantiated, fresh names must be given to @f@, @l@ and so on, that do not conflict with other instantiations. Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.

The body of the combinator is a record that defines the process. The @ins@ field defines the set of input streams and the @outs@ field the set of output streams. The @heap@ field gives the initial values of each of the local variables. The @instrs@ field contains a set of labeled instructions that define the program, while the @label@ field gives the label of the initial instruction. In this form, the output stream is defined via a parameter, rather than being the result of the combinator, as in the representation of @uniquesUnion@ from \S\ref{s:Introduction}.

The initial instruction @(pull sIn1 v A1 [])@ pulls the next element from the stream @sIn1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@. The empty list @[]@ after the target label @A1@ can be used to update heap variables, but as we do not need to update anything yet we leave it empty.

Next, the instruction @(case (f || (l /= v)) A2 [] A3 [])@ checks whether predicate @(f || (l /= v))@ is true; if so it proceeds to the instruction at label @A2@, otherwise it proceeds to @A3@. We use the variable @l@ (last) to track the last value read from the stream, and the boolean @f@ (first) to track whether this is the first element.

% AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
When the predicate is true, the instruction at label @A2@ executes
@(push sOut1 v A3 [ l = v, f = F ])@
which pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).

Finally, the instruction @(drop sIn1 A0 [])@ signals that the current element that was pulled from stream @sIn1@ is no longer required, and goes back to the first instruction at @A0@.

\begin{figure*}

\begin{minipage}{0.6\textwidth}
\begin{alltt}
 group 
   = \(\lambda\) (sIn1: Stream Nat) (sOut1: Stream Nat). 
     \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
     process
      { ins:    { sIn1  }
      , outs:   { sOut1 }
      , heap:   { f = T, l = 0, v = 0 }
      , label:  A0
      , instrs: { A0 = pull sIn1 v          A1 []
                , A1 = case (f || (l /= v)) A2 []  A3 []
                , A2 = push sOut1 v         A3 [ l = v, f = F ]
                , A3 = drop sIn1            A0 [] } }
\end{code}
\end{minipage}
\begin{minipage}{0.29\textwidth}
\includegraphics[scale=1.0]{figures/state-group.pdf}
\end{minipage}

\caption{The group combinator}
\label{fig:Process:Group}
\end{figure*}

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure*}
\begin{alltt}
       merge
         = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
           \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
           process
            { ins:    { sIn1, sIn2 }
            , outs:   { sOut2 }
            , heap:   { x1 = 0, x2 = 0 }
            , label:  B0
            , instrs: { B0 = pull sIn1  x1   B1 []             , B1 = pull sIn2  x2   C0 []
                      , C0 = case (x1 < x2)  D0 []  E0 []      , D0 = push sOut2 x1   D1 []
                      , D1 = drop sIn1       D2 []             , D2 = pull sIn1  x1   C0 []
                      , E0 = push sOut2 x2   E1 []             , E1 = drop sIn2       E2 []
                      , E2 = pull sIn2 x2    C0 [] } }
\end{code}
\includegraphics[scale=1.0]{figures/state-merge.pdf}
\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure*}

The definition of the @merge@ combinator, which merges two input streams, is given in Figure~\ref{fig:Process:Merge}. The combinator binds the two input streams to @sIn1@ and @sIn2@, while the output stream is @sOut2@. The two heap variables @x1@ and @x2@ store the last values read from each input stream. The process starts by pulling from each of the input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

% -----------------------------------------------------------------------------
\subsection{Fusion}

Our fusion algorithm takes two processes and produces a new one that computes the output of both. For example, suppose we need a single process that produces the output of the first two lines of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a process that computes the result of both @group@ and @merge@ as if they were executed concurrently, where the first input stream of the @merge@ process is the same as the input stream of the @group@ process. In our informal description of the fusion algorithm we will instantiate the parameters of each combinator with arguments of the same names.

% -----------------------------------------------------------------------------
\subsubsection{Fusing Pulls}
\label{s:Fusion:FusingPulls}

The algorithm proceeds by considering pairs of states: one from each of the source process state machines to be fused. Both the @group@ machine and the @merge@ machine pull from the same stream as their initial instruction, so we have the situation shown in the top of Figure~\ref{fig:Fusion:Pulls}. The @group@ machine needs to transition from label @A0@ to label @A1@, and the @merge@ machine from @B0@ to @B1@. In the result machine we produce three new instructions that transition between four joint result states, @F0@ to @F3@.
Each of the joint result states represents a combination of two source states, one from each of the source machines. For example, the first result state @F0@ represents a combination of the @group@ machine being in its initial state @A0@ and the @merge@ machine being in its own initial state @B0@.

We also associate each of the joint result states with a description of whether each source machine has already pulled a value from each of its input streams. For the @F0@ case at the top of Figure~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

From the result state @F0@, both of the source machines then need to pull from stream @sIn1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@. In the result machine this is managed by first storing the pulled value in a fresh, shared buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@. To perform the copies we attach updates to a @jump@ instruction, which otherwise transitions between states without affecting any of the input or output streams.

Finally, note that in the result states @F0@ through @F3@, the state of the input streams transitions from `none', to `pending' then to `have'. The `none' state means that we have not yet pulled a value from the associated stream. The `pending' state means we have pulled a value into the stream buffer variable (@b1@ in this case). The `have' state means that we have copied the pulled value from the stream buffer variable into the local variable used by each machine.

% BL: Dropped this to fix pagination.
% In Figure~\ref{fig:Fusion:Pulls},  `sIn1' is set to `have' for the first machine in @F2@ after we have set `v = b1', while `sIn1' is set to `have' for the second machine in @F3@ after we have set `x1 = b1'.

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
Once the result machine has arrived at the joint state @F3@, this is equivalent to the two source machines arriving in states @A1@ and @B1@ respectively. The lower half of Figure~\ref{fig:Fusion:Case} shows the next few transitions of the source machines. From state @A1@, the @group@ machine needs to perform a @case@ branch to determine whether to push the current value it has from its input stream @sIn1@ to output stream @sOut1@, or to just pull the next value from its input. From state @B1@, the @merge@ machine needs to pull a value from its second input stream @sIn2@. In the result machine, @F3@ performs the case analysis from @A1@, moving to either @F4@ or @F5@, corresponding to @A2@ and @A3@ respectively. From state @F4@, the push at @A2@ is executed and moves to @F5@, corresponding to @A3@. Finally, at @F5@ the @merge@ machine pulls from @sIn2@, moving from @F5@ to @F6@, corresponding to @B1@ and @C0@ respectively. As the stream @sIn2@ is only pulled from by the @merge@ machine, no coordination is required between the @merge@ and @group@ machines for this pull.

% -----------------------------------------------------------------------------
\subsection{Fused Result}

Figure~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together. There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two input streams, @sIn1@ and @sIn2@, and two output streams: @sOut1@ from @group@, and @sOut2@ from @merge@. The shared input @sIn1@ is pulled by @merge@ instructions at two places, and since both of these need to agree with when @group@ pulls, the @group@ instructions are duplicated at @F3@-@F5@ and @F13@-@F15@. The first set of instructions could be simplified by constant propagation to a single @push@, as @f@ is initially true.

To complete the implementation of our example from \S\ref{s:Introduction} we would now fuse this result process with a process from the final line of the example (also a @group@). Note that although the result process has a single shared heap, the heap bindings from each fused process are guaranteed not to interfere, as when we instantiate combinators to create source processes we introduce fresh names.

% AR: Cut to make room for explanation why two copies of group pull
% The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:FusionOrder} and \S\ref{s:Evaluation}.

% -----------------------------------------------------------------------------
\subsection{Breaking It Down}
We started with a pure functional program in \S\ref{s:Introduction}, reimagined it as a dataflow graph, then interleaved imperative code that implemented two of the operators in that dataflow graph. We needed to \emph{break down} the definition of each operator into imperative statements so that we could interleave their execution appropriately. We do this because the standard, single-threaded evaluation semantics of functional programs does not allow us to evaluate stream programs that contain both splits and joins in a space efficient way. Returning to the definition of @uniquesUnion@ from \S\ref{s:Introduction}, we cannot simply execute the @group@ operator on its entire input @sIn1@ before executing the @merge@ operator, as that would require us to buffer all data read from @sIn1@. Instead, during fusion we perform the job of a concurrent scheduler at compile time. In the result process the flow of control alternates between the instructions for both the @group@ and @merge@ operators, but as the instructions are interleaved directly there is no overhead due to context switching --- as there would be in a standard concurrent implementation using multiple threads.

The general approach of converting a pure functional program to a dataflow graph, then interleaving imperative statements that implement each operator was also used in prior work on Flow Fusion~\cite{lippmeier2013data}. However, in contrast to Flow Fusion and similar systems, with \mbox{Machine Fusion} we do not need to organize statements into a fixed \emph{loop anatomy} --- we simply merge them as they are. This allows us to implement a wider range of operators, including ones with nested loops that work on segmented streams.

Note that relying on lazy evaluation for @uniquesUnion@ does not eliminate the need for unbounded buffering. Suppose we converted each of the streams to lazy lists, and used definitions of @group@ and @merge@ that worked over these lists. As @uniquesUnion@ returns a pair of results, there is nothing preventing a consumer from demanding the first list (@sUnique@) in its entirety before demanding any of the elements from the second list (@sUnion@). If this were to happen then the runtime implementation would need to retain all elements of @sIn1@ before demanding any of @sIn2@, causing a space leak. Lazy evaluation is \emph{pull only} meaning that evaluation is driven by the consumer. The space efficiency of our fused program relies critically on the fact that processes can also \emph{push} their result values directly to their consumers, and that the consumers cannot defer the handling of these values.

% TODO: Blend this in.
% Note that we could construct the fused result machine in several ways. One option is to perform the case branch first and then pull from @sIn2@, another is to pull from @sIn2@ first and then perform the branch. By construction, the predicate used in the branch refers only to variables local to the @group@ machine, and the pull instruction from @B1@ stores its result in a variable local to the @merge@ machine. As the set of variables does not overlap, either ordering is correct. For this example we choose to perform the branch first, though will discuss the ramifications of this choice further in \S\ref{s:FusionOrder}.

% TODO: Shift this later.
% As the @merge@ process merges infinite streams, if we execute it with a finite input prefix, it will arrive at an intermediate state that may not yet have pushed all available output. For example, if we execute the process with the input streams @[1,4]@ and @[2,3,100]@ then the values @[1,2,3,4]@ will be pushed to the output. After pushing the last value @4@, the process will block at instruction @E2@, waiting for the next value to become available from @sIn2@. We discuss how to handle finite streams later in ~\S\ref{s:Finite}.

% TODO: Talk about what drop is really for later.
% This @drop@ instruction is used to coordinate concurrent processes when performing fusion. The next element of a stream may only be pulled after all consumers of that stream have pulled and then and dropped the current element.

% -----------------------------------------------------------------------------
\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}

\vspace{2em}

\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing pull (top) and case (bottom) instructions}
\label{fig:Fusion:Pulls}
\label{fig:Fusion:Case}
\end{figure*}

% BL: Merged these figures togather to save space.
% \begin{figure*}
% \caption{Fusing case instructions}
% \label{fig:Fusion:Case}
% \end{figure*}

%%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
\begin{figure*}
\definecolor{groupc}{HTML}{308030}
\definecolor{mergec}{HTML}{800000}
\definecolor{sharec}{HTML}{000080}

\newcommand\annot[5]{
  \tiny ((#1,   \> \tiny \{sIn1 =      #2\}), 
                \> \tiny (#3, \> \tiny \{sIn1 =      #4, \> \tiny sIn2 =      #5\}))
}

\newcommand\icase[7]{
 \tt{#1} \> \tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \tt{#5} \> \tt{[ #6 ]} \> \tiny #1 \> \tiny = #7
}

\newcommand\instr[5]{
 \tt{#1}\>\tt{= #2} \> \tt{#3} \> \tt{[ #4 ]} \> \> \> \tiny #1 \> \tiny = #5
}

\newcommand\tctt[2]{\textcolor{#1}{\tt{#2}}}

\hspace{5em}

\begin{adjustbox}{minipage=0.8\textwidth,margin=0pt \smallskipamount,center}
\begin{alltt}
process
\string{ ins:    \string{ \tctt{sharec}{sIn1},  \tctt{mergec}{sIn2} \string}
, outs:   \string{ \tctt{groupc}{sOut1}, \tctt{mergec}{sOut2} \string}
, heap:   \string{ \tctt{groupc}{f = T, l = 0, v = 0}, \tctt{mergec}{x1 = 0, x2 = 0}, \tctt{sharec}{b1 = 0} \string}
, label:  \tctt{sharec}{F0}
\end{alltt}

\begin{tabbing}
@  @ \=
@, F17@  \= = @case (f || (l /= v)) @
         \= @F17@ \= @[ ]     @ \= @F17@ \= @[ ]@
@   @ \= \tiny @   @F17 \= \tiny = ((A0, \= \tiny \{sIn1 = pending\}), \= \tiny (B0, \= \tiny \{sIn1 = pending, \= \tiny sIn2 = pending\})) \kill

@, instrs:@

% I have no idea how this works, but you need this particular incantation to colour the whole line in a tabbing environment.
% It's not ideal that the { and , are coloured too, but that can't be helped.
\\[0pt \color{sharec}]
\> @{@
\instr{F0}{pull sIn1 b1}{F1}{}
      {\annot{A0}{none}{B0}{none}{none}}

\\[0pt \color{groupc}]
\> @,@
\instr{F1}{jump}{F2}{v  = b1}
      {\annot{A0}{pending}{B0}{pending}{none}}

\\[0pt \color{mergec}]
\> @,@
\instr{F2}{jump}{F3}{x1 = b1}
      {\annot{A1}{have}{B0}{pending}{none}}

\\[0pt \color{groupc}]
\> @,@
\icase{F3}{case (f || (l /= v))}{F4}{}{F5}{}
      {\annot{A1}{have}{B1}{have}{none}}

\\[0pt \color{groupc}]
\> @,@
\instr{F4}{push sOut1 v}{F5}{l = v, f = F}
      {\annot{A2}{have}{B1}{have}{none}}

\\[0pt \color{groupc}]
\> @,@
\instr{F5}{jump}{F6}{}
      {\annot{A3}{have}{B1}{have}{none}}

\\[0pt \color{mergec}]
\> @,@
\instr{F6}{pull sIn2 x2}{F7}{}
      {\annot{A0}{none}{B1}{have}{none}}

\\
\\[0pt \color{mergec}]
\> @,@
\icase{F7}{case (x1 < x2)}{F8}{}{F16}{}
      {\annot{A0}{none}{C0}{have}{have}}

\\
\\[0pt \color{mergec}]
\> @,@
\instr{F8}{push sOut2 x1}{F9}{}
      {\annot{A0}{none}{D0}{have}{have}}

\\[0pt \color{mergec}]
\> @,@
\instr{F9}{drop sIn1}{F10}{}
      {\annot{A0}{none}{D1}{none}{have}}

\\[0pt \color{sharec}]
\> @,@
\instr{F10}{pull sIn1 b1}{F11}{}
      {\annot{A0}{none}{D2}{none}{have}}

\\[0pt \color{groupc}]
\> @,@
\instr{F11}{jump}{F12}{v = b1}
      {\annot{A0}{pending}{D2}{pending}{have}}

\\[0pt \color{mergec}]
\> @,@
\instr{F12}{jump}{F13}{x1 = b1}
      {\annot{A1}{have}{D2}{pending}{have}}

\\[0pt \color{groupc}]
\> @,@
\icase{F13}{case (f || (l /= v))}{F14}{}{F15}{}
      {\annot{A1}{have}{C0}{have}{have}}

\\[0pt \color{groupc}]
\> @,@
\instr{F14}{push sOut1 v}{F15}{l = v, f = F}
      {\annot{A2}{have}{C0}{have}{have}}

\\[0pt \color{groupc}]
\> @,@
\instr{F15}{jump}{F7}{}
      {\annot{A3}{have}{C0}{have}{have}}

\\

\\[0pt \color{mergec}]
\> @,@
\instr{F16}{push sOut2 x2}{F17}{}
      {\annot{A0}{none}{E0}{have}{have}}

\\[0pt \color{mergec}]
\> @,@
\instr{F17}{drop sIn2}{F18}{}
      {\annot{A0}{none}{E1}{have}{have}}

\\[0pt \color{mergec}]
\> @,@
\instr{F18}{pull sIn2}{F7}{}
      {\annot{A0}{none}{E2}{have}{none}}

\\[0pt \color{black}]
@} }@
\end{tabbing}
\end{adjustbox}
\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
\label{fig:Process:Fused}
\end{figure*}
% -----------------------------------------------------------------------------

% -----------------------------------------------------------------------------
\section{Process definitions}
\begin{figure*}
\begin{minipage}[t]{0.4\textwidth}
\begin{tabbing}
MMMMMx \TABDEF @MMMMM@  \TABSKIP $\Exp$ \TABSKIP $\Exp$ \TABSKIP $\Exp$ \kill

\Exp,~$e$       \> ::= \> ~~~ $x~|~v~|~e~e $ \\
                \> $\enskip|~$ \> ~~ $ (e~||~e) ~|~ e+e ~|~ e~@/=@~e ~|~ e < e$ \\
\Value,~$v$     \> ::= \> ~~~ $\mathbb{N}~|~\mathbb{B}~|~(\lambda{}x.~e)$ \\
\Heap,~$bs$     \> ::= \> ~~~ $\cdot~|~bs,~x~=~v$ \\
\Updates,~$us$  \> ::= \> ~~~ $\cdot~|~us,~x~=~e$
\\[0.5em]

\Proc,~$p$      \> ::=\> @process@ \\
MMMMMM \= M \= \kill
\> \> @ins:   @  $(\Chan ~\mapsto~ \InputState)$ \\
\> \> @outs:  @  $\sgl{\Chan}$ \\
\> \> @heap:  @  \Heap \\
\> \> @label: @  \Label \\
\> \> @instrs:@  $(\Label ~\mapsto~ \Instr)$ 
\\[0.5em]
\InputState \> ::= \> ~~ @none@~$|$~@pending@~\Value~$|$~@have@

\end{tabbing}
\end{minipage}
\begin{minipage}[t]{0.05\textwidth}
\quad
\end{minipage}
\begin{minipage}[t]{0.4\textwidth}
\begin{tabbing}
MMMMMM \TABDEF @MMMM@  \TABSKIP $\Chan$ \TABSKIP $\Chan$ \TABSKIP $\Exp$ \kill

\Var,~$x$       \> $\to$ \> ~~~ (value variable) \\
\Chan,~$c$      \> $\to$ \> ~~~ (channel name) \\
\Label,~$l$     \> $\to$ \> ~~~ (label name) \\ 
\ChannelStates  \> ~ =   \> ~~~ $(\Chan \mapsto \InputState)$ \\
\Action,~$a$   \> ::=    \> ~~~ $\cdot ~~|~~ \Push~\Chan~\Value$ \\[0.5em]

\Instr
    \> ::=\> @pull@  \> \Chan  \> \Var  \> \Next \\
    \TABALT  @push@  \> \Chan  \> \Exp  \> \Next \\
    \TABALT  @drop@  \> \Chan  \>       \> \Next \\
    \TABALT  @case@  \> \Exp   \> \Next \> \Next \\
    \TABALT  @jump@  \>        \>       \> \Next \\
\\[0.5em]

\Next \> = \> $\Label~\times~\Updates$ 
\end{tabbing}
\end{minipage}
\caption{Process definitions}
\label{fig:Process:Def}
\end{figure*}

The formal grammar for process definitions is given in Figure~\ref{fig:Process:Def}. Variables, Channels and Labels are specified by unique names. We refer to the \emph{endpoint} of a stream as a channel. A particular stream may flow into the input channels of several different processes, but can only be produced by a single output channel. For values and expressions we use an untyped lambda calculus with a few primitives chosen to facilitate the examples. The `$||$' operator is boolean-or, `+' addition, `/=' not-equal, and `$<$' less-than.

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'. A network is a set of processes that are able to communicate with each other.

In the grammar of Figure~\ref{fig:Process:Def} the $\InputState$ has three options: @none@, which means no value is currently stored in the associated stream buffer variable, $(@pending@~\Value)$ which gives the current value in the stream buffer variable and indicates that it has not yet been copied into a process-local variable, and @have@ which means the pending value has been copied into a process-local variable. The $\Value$ attached to the @pending@ state is used when specifying the evaluation semantics of processes. When performing the fusion transform the $\Value$ itself will not be known, but we can still reason statically that a process must be in the @pending@ state. When defining the fusion transform in \S\ref{s:Fusion} we will use a version of $\InputState$ with only this statically known information.

The @instrs@ field of the $\Proc$ maps labels to instructions. The possible instructions are: @pull@, which pulls the next value from a channel into a given heap variable; @push@, which pushes the value of an expression to an output channel;  @drop@ which indicates that the current value pulled from a channel is no longer needed; @case@ which branches based on the result of a boolean expression, and @jump@ which causes control to move to a new instruction.

Instructions include a $\Next$ field containing the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap.
The list of update bindings is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.

When lowering process code to a target language, such as C, LLVM, or some sort of assembly code, we can safely convert @drop@ to plain @jump@ instructions. The @drop@ instructions are used to control how processes should be synchronized, but do not affect the execution of a single process. We discuss @drop@s further in \S\ref{s:Optimisation}.

% This allows us to \emph{deliberately} introduce artificial deadlocks when a process network would require more than one element of buffering.
%%% AR: added to highlight that this rules out networks that require unbounded buffers
%%% BL: We don't have any examples of explicitly introducing deadlocks. The process networks just happen to have them when viewed abstractly.

%%% AR: feels a bit disjointed because drops were only mentioned once a few paragraphs ago. Maybe reword to talk about lowering in general is obvious for most instructions, and drops are just treated as jumps. Or move up.

% -----------------------------------------------------------------------------
\subsection{Execution}
\label{s:Process:Eval}

The dynamic execution of a process network consists of:

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a network. Each individual process only needs to accept an injected  value when it is ready for it, and injection into a network succeeds only when they \emph{all} processes accept it.

\item \emph{Advancing} a single process from one state to another. Advancing a network succeeds when \emph{any} of the processes in the network can advance.

\item \emph{Feeding} outputs of some processes to the inputs of others. Feeding alternates between Injecting and Advancing. When a process pushes a value to an output channel we attempt to inject this value into all processes that have that same channel as an input. If they all accept it, we then advance their programs as far as they will go, which may cause more values to be pushed to output channels, and so on.
\end{enumerate}

Execution of a network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking, which enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, and pushing only succeeds when that buffer is empty.

%%% AR: what is the distinction between 'execution' and 'evaluation'?  I only have a vague feeling that execution is something a computer does, while evaluation is the mathematical rules. Either way, these should probably be consistent.
%%% BL: "Evaluation" is pure.  E-"value"-ation. Execution has visible actions, like pushing to streams.

% TODO BL: Mention what happens if we choose a different ordering, 
% and how the particular ordering chosen is decided upon.
Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we choose one particular instruction ordering that enables the network to advance without requiring unbounded buffering. The single ordering is chosen by heuristically deciding which pair of states to merge during fusion, and is discussed in \S\ref{s:EvaluationOrder}.

Each channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process. The only intra-process communication is via channels and streams. Our model is ``pure data flow'' as there are no side-channels between processes --- in contrast to ``impure data flow'' systems such as StreamIt~\cite{thies2002streamit}.

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Figure~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. Injection is a meta-level operation, in contrast to @pull@ and @push@ which are instructions in the object language. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on a process network.

Rule (InjectValue) injects a single value into a single process. The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process. The $\InputState$ acts as a single element buffer, and must be empty (@none@) for injection to succeed.

Rule (InjectIgnore) allows processes that do not use a particular named channel to ignore values injected into that channel.

Rule (InjectMany) attempts to inject a single value into a network. We use the single process judgment form to inject the value into all processes, which must succeed for all of them.
% Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.

\begin{figure*}
\begin{minipage}[t]{1\textwidth}

$$
\arrLR{
  \boxed{\ProcInject{\Proc}{\Value}{\Chan}{\Proc}}
}{
  \boxed{\ProcsInject{\sgl{\Proc}}{\Value}{\Chan}{\sgl{\Proc}}}
}
$$

$$
\ruleIN{
  p[@ins@][c] = @none@
}{
  \ProcInject{p}{v}{c}{p~[@ins@ \mapsto (p[@ins@][c \mapsto @pending@~v]) ] }
}{InjectValue}
$$

$$
\ruleIN{
  c \not\in p[@ins@]
}{
  \ProcInject{p}{v}{c}{p}
}{InjectIgnore}
%
\quad
%
\ruleIN{
  \{~ \ProcInject{p_i}{v}{c}{p'_i} ~\}^i
}{
  \ProcsInject{\sgl{p_i}^i}{v}{c}{\sgl{p'_i}^i}
}{InjectMany}
$$

\end{minipage}
\caption{Injection of values into input channels}
\label{fig:Process:Eval:Inject}
\end{figure*}

% -----------------------------------------------------------------------------
\begin{figure*}
\begin{minipage}[t]{1\textwidth}

$$
  \boxed{
    \ProcBlockShake
      {\Instr}
      {\ChannelStates}
      {\Heap}
      {\Action}
      {\Label}
      {\ChannelStates}
      {\Updates}
  }
$$

$$
\ruleIN{
  is[c] = @pending@~v
}{
  \ProcBlockShake
        {@pull@~c~x~(l,us)}
        {is}
        {bs}
        {\cdot}
        {l}
        {is[c \mapsto @have@]}
        {(us, x = v)}
}{Pull}
%
\quad
%
\ruleIN{
  \ExpEval{bs}{e}{v}
}{
  \ProcBlockShake
        {@push@~c~e~(l,us)}
        {is}
        {bs}
        {\Push~c~v}
        {l}
        {is}
        {us}
}{Push}
$$

$$
\ruleIN{
  is[c] = @have@
}{
  \ProcBlockShake
        {@drop@~c~(l,us)}
        {is}
        {bs}
        {\cdot}
        {l}
        {is[c \mapsto @none@]}
        {us}
}{Drop}
\ruleIN{
}{
  \ProcBlockShake
        {@jump@~(l,us)}
        {is}
        {bs}
        {\cdot}
        {l}
        {is}
        {us}
}{Jump}
$$

$$
\ruleIN{
  \ExpEval{bs}{e}{@True@}
}{
  \ProcBlockShake
        {@case@~e~(l_t,us_t)~(l_f,us_f)}
        {is}
        {bs}
        {\cdot}
        {l_t}
        {is}
        {us_t}
}{CaseT}
\ruleIN{
  \ExpEval{bs}{e}{@False@}
}{
  \ProcBlockShake
        {@case@~e~(l_t,us_t)~(l_f,us_f)}
        {is}
        {bs}
        {\cdot}
        {l_f}
        {is}
        {us_f}
}{CaseF}
$$

\vspace{2em}

% ----------------
$$
  \boxed{\ProcShake{\Proc}{\Action}{\Proc}}
$$
$$
\ruleIN{
  \ProcBlockShake
    {p[@instrs@][p[@label@]]} 
    {p[@ins@]}
    {p[@heap@]}
    {a}
    {l}
    {is}
    {us}
  \quad
    \ExpEval{p[@heap@]}{us}{bs}
}{
  \ProcShake
        {p}
        {a}
        {p~[    @label@~ \mapsto ~l
           , ~~ @heap@~  \mapsto (p[@heap@] \lhd bs)
           , ~~ @ins@~   \mapsto ~is]}
}{Advance}
$$

\end{minipage}
\caption{Advancing processes}

% Evaluation: shaking allows proceses to take a step from one label to another as well as produce an output message. If the message is a push, the value is injected to all other processes in the network; otherwise it is an internal step.}
\label{fig:Process:Eval:Shake}
\end{figure*}

\begin{figure*}
\begin{minipage}[t]{1\textwidth}

% ---------------------------------------------------------
$$
  \boxed{\ProcsShake{\sgl{\Proc}}{\Action}{\sgl{\Proc}}}
$$

$$
\ruleIN{
  \ProcShake{p_i}{\cdot}{p'_i}
}{
  \ProcsShake{
    \sgl{p_0 \ldots p_i \ldots p_n}
  }{\cdot}{
    \sgl{p_0 \ldots p'_i \ldots p_n}
  }
}{ProcessesInternal}
%
\quad
%
\ruleIN{
  \ProcShake{p_i}{\Push~c~v}{p'_i}
  \quad
  \forall j~|~j \neq i.~
  \ProcInject{p_j}{c}{v}{p'_j}
}{
  \ProcsShake{
    \sgl{p_0 \ldots p_i \ldots p_n}
  }{\Push~c~v}{
    \sgl{p'_0 \ldots p'_i \ldots p'_n}
  }
}{ProcessesPush}
$$

% ---------------------------------------------------------
\vspace{1em}

\newcommand\vs {\ti{vs}}
\newcommand\accs {\ti{accs}}
\newcommand\network {\ti{ps}}

$$
  \boxed{
    \ProcsFeed
      {(\Chan \mapsto \overline{Value})~}
      {\sgl{\Proc}}
      {(\Chan \mapsto \overline{Value})~}
      {\sgl{\Proc}}
  }
$$
$$
\ruleIN{
  \ProcsShake
    {ps}
    {\cdot}
    {ps'}
}{
  \ProcsFeed
    {cvs}
    {ps}
    {cvs}
    {ps'}
}{FeedInternal}
%
\quad
%
\ruleIN{
  \ProcsShake
    {ps}
    {\Push~c~v}
    {ps'}
}{
  \ProcsFeed
    {cvs}
    {ps}
    {cvs[c \mapsto (cvs[c] \listappend v)]}
    {ps'}
}{FeedPush}
$$

% $$
% \ruleIN{
%   (\forall p \in \network.~c \not\in @outs@~p)
% \quad
%   \ProcsFeed
%     {c=\vs, \accs}
%     {\network}
%     {\accs'}
%     {\network'}
% \quad
%   \ProcsInject
%     {\network'}
%     {c}{v}
%     {\network''}
% }{
%   \ProcsFeed
%     {c=\vs \listappend [v], \accs}
%     {\network}
%     {c=\vs \listappend [v], \accs'}
%     {\network''}
% }{FeedExternal}
% $$

$$
\ruleIN{
  (\forall p \in \network.~c \not\in p[@outs@])
\quad
  \ProcsInject
    {ps}
    {c}
    {v}
    {ps'}
}{
  \ProcsFeed
    {cvs[c \mapsto ([v] \listappend vs)]}
    {ps}
    {cvs[c \mapsto vs~]}
    {ps'}
}{FeedExternal}
$$

\end{minipage}
\caption{Feeding Process Networks}
\label{fig:Process:Eval:Feed}
\end{figure*}

% -----------------------------------------------------------------------------
\subsubsection{Advancing}
Figure~\ref{fig:Process:Eval:Shake} gives the rules for advancing a single process. The statement $(\ProcBlockShake{i}{is}{bs}{a}{l}{is'}{us'})$ reads ``instruction $i$, given channel states $is$ and the heap bindings $bs$, passes control to instruction at label $l$ and yields new channel states $is'$, heap update expressions $us'$, and performs an output action $a$.'' An output action $a$ is an optional message of the form $(\Push~\Chan~\Value)$, which encodes the value a process pushes to one of its output channels. We write ~$\cdot$~ for an empty action.

Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update to copy this value into the variable $x$ in the @pull@ instruction. We use the syntax $us,x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$. In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied into the local variable.

Rule (Push) evaluates the expression $e$ under heap bindings $bs$ to a value $v$, and produces a corresponding action which carries this value. The judgment $(bs \vdash e \Downarrow v)$ expresses standard untyped lambda calculus reduction using the heap $bs$ for the values of free variables. As this evaluation is completely standard we omit it to save space.

Rule (Drop) changes the input channel state from @have@ to @none@. A @drop@ instruction can only be executed after @pull@ has set the input channel state to @have@.

Rule (Jump) produces a new label and associated update expressions. Rules (CaseT) and (CaseF) evaluate the scrutinee $e$ and emit the appropriate label.

The statement $\ProcShake{p}{a}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $a$''. Rule (Advance) advances a single process. We look up the current instruction for the process' @label@ and pass it, along with the channel states and heap, to the above single instruction judgment. The update expressions $us$ from the single instruction judgment are reduced to values before updating the heap. We use $(us \lhd bs)$ to replace bindings in $us$ with new ones from $bs$. As the update expressions are pure, the evaluation can be done in any order.

% -----------------------------------------------------------------------------
\subsubsection{Feeding}
Figure~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes. The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.

The statement $\ProcShake{ps}{a}{ps'}$ reads ``the network $ps$ advances to the network $ps'$ yielding output action $a$''.

Rule (ProcessInternal) allows an arbitrary process in the network to advance to a new state at any time, provided it does not yield an output action. This allows processes to perform internal computation.

% , without needing to synchronize with the rest of the network.

Rule (ProcessPush) allows an arbitrary process in the network to advance to a new state while yielding an output action $(@push@~c~v)$. For this to succeed it must be possible to inject the output value $v$ into all processes that have channel $c$ as one of their inputs.

% As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process. When any process in the network produces an output action we take that as the action of the whole network.

The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, network $ps$ takes a step and produces new channel values $cvs'$ and network $ps'$''. The channel values $cvs$ map channel names to a list of values. For input channels of the overall network, we initialize the map to contain a list of input values for each channel. For output channels of the overall network, values pushed to those channels are also collected in the same channel map. In a concrete implementation the input and output values would be transported over some IO device, but for the semantics we describe the abstract behavior only.

Rule (FeedInternal) allows the network to perform local computation in the context of the channel values.

Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a network and appends the value $v$ to the list corresponding to the output channel $c$.

Rule (FeedExternal) injects values from the external environment. This rule also has the side condition that values cannot be injected from the environment into output channels that are already owned by some process. This constraint is required for correctness proofs, but can be ensured by construction in a concrete implementation.
% The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Execution Order}
\label{s:EvaluationOrder}

The execution rules of Figure~\ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.

In the semantics, allowing the execution order of processes to be non-deterministic is critical, as it defines a search space where we might find an order that does not require unbounded buffering. For a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime. In contrast, the role of our fusion system is to construct one of these working orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network were executed dynamically. Fusion also eliminates the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

% In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering. Consider the following example, where the @alt2@ operator pulls two elements from its first input stream, then two from the second, before pushing all four to its output stream.
% \begin{code}
%   alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
%   alternates sInA sInB sInC
%    = let  s1   = alt2 sInA sInB
%           s2   = alt2 sInB sInC
%           sOut = zip s1 s2
%      in   sOut
% \end{code}
% 
% Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples. Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.
% 
% Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time. However, the downstream @zip@ process needs to pull one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams. This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively. They can still push two values at a time to their own outputs, but the downstream @zip@ process needs receive one from each process alternately. Here is a table of intermediate values to help make the explanation clearer:
% 
% \begin{code}
%     sInA = [a1, a2, a3, a4, a5 ...]
%     sInB = [b1, b2, b3, b4, b5 ...]
%     sInC = [c1, c2, c3, c4, c5 ...]
% 
%     s1   = alt2 sInA sInB 
%          = [a1, a2, b1, b2, a3, a4, b3, b4 ...]
% 
%     s2   = alt2 sInB sInC
%          = [b1, b2, c1, c2, b3, b4, c3, c4 ...]
% 
%     sOut = zip s1 s2
%          = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
% \end{code}
% 
% Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on. The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it. While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@. The execution order of the overall network is constrained by communication patterns of processes in that network.

% As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings. In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime. In contrast, the role of our fusion transform is to construct one of these working orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network was executed dynamically. In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

% Although alt2 produces output elems two at a time, the consumer zip need its input elements to arrive alternately. At evaluation time we need the results pushed to sA1 and sA2 in the sA1 sA2 sA1 sA2 order, not sA1 sA1 sA2 sA2. Writing the rules nondeterministically allows the elaborator to discover a usable order, if there is one. This also affects fusion, we don't want to commit to the wrong order too early. We shall see that if we fuse the two alt processes first fusion will not work. We need to start with zip so that the order in which input elems arrive is constrained.

% -----------------------------------------------------------------------------
\section{Fusion}
\label{s:Fusion}

Our core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.

Figure~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion. We extend the $\Label$ grammar with a new alternative, $\LabelF \times \LabelF$ for the labels in a fused result process. Each $\LabelF$ consists of a $\Label$ from a source process, paired with a map from $\Chan$ to the statically known part of that channel's current $\InputState$. When fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information. Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.

% The definition of $\Label$ is now recursive.
% These new labels $LabelF$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.

% If the static $\InputStateF$ is $@pending@_F$, there is a value waiting to be pulled, do not know the actual value.

We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with \mbox{channel @c@}. We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
We used a fresh name back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion.

Still in Figure~\ref{fig:Fusion:Types}, $\ChanTypeTwo$ classifies how channels are used, and possibly shared, between two processes. Type @in2@ indicates that the two processes @pull@ from the same channel, so these actions must be coordinated. Type @in1@ indicates that only a single process pulls from the channel. Type @in1out1@ indicates that one process pushes to the channel and the other pulls. Type @out1@ indicates that the channel is pushed to by a single process. Each output channel is uniquely owned and cannot be pushed to by more than one process.

\begin{figure}

\begin{tabbing}
@MMMMMMMMMMMM@   \TABDEF \kill

$\Label$        \> ::=  \> ~~~ $\ldots ~|~\LabelF~\times~\LabelF ~|~ \ldots$ \\
$\LabelF$       \> =    \> ~~~ $\Label~\times~\MapType{\Chan}{\InputStateF}$  \\
$\InputStateF$  \> ::=  \> ~~~ $@none@_F ~|~ @pending@_F ~|~ @have@_F$    \\
$\Var$          \> ::=  \> ~~~ $\ldots ~|~@chan@~\Chan ~|~ \ldots$ \\
$\ChanTypeTwo$  \> ::=  \> ~~~ $@in2@~|~@in1@~|~@in1out1@~|~@out1@$
\end{tabbing}
\caption{Fusion type definitions.}
\label{fig:Fusion:Types}
\end{figure}
% The labels for a fused program consists of both of the original program labels, as well as the statically known part of the input state for each channel. The channels of both processes are classified into inputs and outputs, this describes what coordination is required between the two.

\begin{figure}

\begin{tabbing}
M\=M\=M\=M\=M\kill

$\ti{fusePair} ~:~ \Proc \to \Proc \to  \Maybe~\Proc$ \\
$\ti{fusePair}~p~q$ \\
\> $~|$ \> $\Just \ti{is} \gets \ti{go}~\sgl{}~l_0$ \\
\> $=$ \> $\Just ($@process@ \\
@             ins: @ $\sgl{c~|~c=t \in \cs,~t \in \sgl{@in1@,@in2@}} $ \\
@            outs: @ $\sgl{c~|~c=t \in \cs,~t \in \sgl{@in1out1@,@out1@}} $ \\
@            heap: @ $p[@heap@]~\cup~q[@heap@]$ \\
@           label: @ $l_0$ \\
@          instrs: @ $\ti{is})$ \\
\> $~|$ \> $@otherwise@~=~@Nothing@$ \\
@ where@ \\
M\=MM\=M\=~=~\=\kill
 \> \cs \> $=$ \> $\ti{channels}~p~q$ \\[0.5ex]

 \> $l_0$  \> $=$ \> $
      \big($ \= $(p[@label@],$ \= $\sgl{c=@none@_F~|~c~\in~p[@ins@]})$
\\ \> \> \>$,$ \> $(q[@label@],$ \> $\sgl{c=@none@_F~|~c~\in~q[@ins@]})
      \big)$ \\[0.5ex]

 \> $\ti{go}~\ti{bs}~(l_p,l_q)$ \\
 \> \> $~|$ \> $(l_p,l_q)~\in~\ti{bs}$ \\
 \> \> $=$ \> $\Just\ti{bs}$ \\[0.5ex]
 \> \> $~|$ \> $\Just b \gets \ti{tryStepPair}~\cs~l_p~p[@instrs@][l_p]~l_q~q[@instrs@][l_q]$ \\ 
 \> \> $=$ \> $\ti{foldM}~\ti{go}~(\ti{bs}~\cup~\sgl{(l_p,l_q)=b})~(\ti{outlabels}~b)$ \\[0.5ex]
 \> \> $~|$ \> $@otherwise@~=~@Nothing@$
\end{tabbing}

\caption{Fusion of pairs of processes}

% Two processes are fused together by starting at the initial label for each process and computing the instruction based on one of the original process' instructions at that label. Instructions are added recursively until all reachable instructions are included.

\label{fig:Fusion:Def:Top}
\end{figure}

\smallskip
% -------------------------------------
Figure~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes, constructing a result process that does the job of both. We start with a joint label $l_0$ formed from the initial labels of the two source processes. We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next. The possible destination labels of that instruction (computed with $outlabels$ from Figure~\ref{fig:Fusion:Utils}) define new joint labels and reachable states. As we discover reachable states we add them to a map $bs$ of joint label to the corresponding instruction, and repeat the process to a fixpoint where no new states can be discovered.

\begin{figure}
\begin{tabbing}
\ti{tryStepPair}\=$~\to~$\=M\kill

$\ti{tryStepPair} ~:~ \ChanTypeMap$ \\
\> $\to$ \> $\LabelF \to \Instr \to \LabelF \to \Instr$ \\
\> $\to$ \> $\Maybe~\Instr$ \\

M \= $~|~(@pull@~\_~\_~\_ \gets i_p')~$ \= $\to$ \= $\Just (\ti{swaplabels}~i_q')~$ \= M \=\kill
$\ti{tryStepPair} ~\cs~l_p~i_p~l_q~i_q~=$ \\
\> $@match@~ (\ti{tryStep}~\cs~l_p~i_p~l_q,~\ti{tryStep}~\cs~l_q~i_q~l_p) ~@with@$ \\
\> $(\Just i_p',~\Just i_q')$ \\

\> @ @$|~@jump@~\_ \gets i_p'$ \> $\to$ \> $\Just i_p'$
\> \note{PreferJump1} \\[0.5ex]

\> @ @$|~@jump@~\_ \gets i_q'$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
\> \note{PreferJump2} \\[0.5ex]

\> @ @$|~@pull@~\_~\_~\_ \gets i_q'$ \> $\to$ \> $\Just i_p'$
\> \note{DeferPull1} \\[0.5ex]

\> @ @$|~@pull@~\_~\_~\_ \gets i_p'$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
\> \note{DeferPull2} \\[0.5ex]

\> $(\Just i_p',~\_)$ \> $\to$ \> $\Just i_p'$
\> \note{Run1} \\[0.5ex]

\> $(\_, ~\Just i_q')$ \> $\to$ \> $\Just (\ti{swaplabels}~i_q')$
\> \note{Run2} \\[0.5ex]

\> $(\Nothing, ~\Nothing)$ \> $\to$ \> $\Nothing$
\> \note{Deadlock}
\end{tabbing}
\caption{Fusion step coordination for a pair of processes.}
% Statically compute the instruction to perform at a particular fused label. Try to execute either process, preferring jumps and other instructions over pulling, as pulling can block while other instructions may perform ``useful work'' without blocking. If neither machine can execute, fusion fails.
\label{fig:Fusion:Def:StepPair}
\end{figure}

% -------------------------------------
Figure~\ref{fig:Fusion:Def:StepPair} defines function \ti{tryStepPair} which decides which process to advance. It starts by calling \ti{tryStep} for both processes. If both can advance, we use heuristics to decide which one to run first.

Clauses (PreferJump1) and (PreferJump2) prioritize processes that can perform a @jump@. This helps collect @jump@ instructions together so they are easier for post-fusion optimization to handle (\S\ref{s:Optimisation}).
The instruction for the second process was computed by calling \ti{tryStep} with the label arguments swapped, so in (PreferJump2) we need to swap the labels back with $\ti{swaplabels}$ (from Figure~\ref{fig:Fusion:Utils}).

Similarly, clauses (DeferPull1) and (DeferPull2) defer @pull@ instructions: if one of the instructions is a @pull@, we advance the other one. We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.

Clauses (Run1) and (Run2) apply when the above heuristics do not apply, or only one of the processes can advance.
% We try the first process first, and if that can advance then so be it. This priority means that fusion is left-biased, preferring advancement of the left process over the second.

Clause (Deadlock) applies when neither process can advance, in which case the processes cannot be fused together and fusion fails.

\begin{figure*}
\begin{tabbing}
M \= M \= MMMMMMMMMMMMMMMMMM \= @Just (push c e @ \= MMMMMMMMMMMMMMMMMMMMMMMMM \= \kill
$\ti{tryStep} ~:~ \ChanTypeMap \to \LabelF \to \Instr \to \LabelF \to \Maybe~\Instr$ \\
$\ti{tryStep} ~\cs~(l_p,s_p)~i_p~(l_q,s_q)~=~@match@~i_p~@with@$ \\

\> $@jump@~(l',u')$ 
\> \> $\to~\Just (@jump@$ \> $
      \nextStep
        {l'}{s_p}
        {l_q}{s_q}
        {u'})
      $ 
\> \note{LocalJump}
\\[1ex]

\> $@case@~e~(l'_t,u'_t)~(l'_f,u'_f)$
\> \> $\to~\Just (@case@~e$ \> $
      \nextStep
        {l'_t}{s_p}
        {l_q}{s_q}
        {u'_t}
      ~
      \nextStep
        {l'_f}{s_p}
        {l_q}{s_q}
        {u'_f})
      $ 
\> \note{LocalCase}
\\[1ex]

\> $@push@~c~e~(l',u')$ \\
\> \> $~|~\cs[c]=@out1@$ 
\> $\to~\Just (@push@~c~e$ \> $
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {s_q}
        {u'})
      $ 
\> \note{LocalPush}\\

\> \> $~|~\cs[c]=@in1out1@ ~\wedge~ s_q[c]=@none@_F$ 
\> $\to~\Just (@push@~c~e$ \> $
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {\HeapUpdateOne{c}{@pending@_F}{s_q}}
        {\HeapUpdateOne{@chan@~c}{e}{u'}})
      $
\> \note{SharedPush}
\\[1ex]

\> $@pull@~c~x~(l',u')$ \\
\> \> $~|~\cs[c]=@in1@$ 
\> $\to~\Just (@pull@~c~x$ \> $
      \nextStep
        {l'}{s_p}
        {l_q}{s_q}
        {u'})
    $ 
\> \note{LocalPull}
\\[1ex]

\> \> $~|~(\cs[c]=@in2@ \vee \cs[c]=@in1out1@) ~\wedge~ s_p[c]=@pending@_F$ \\
\> \> $\to~\Just (@jump@~
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@have@_F}{s_p}}
        {l_q}
          {s_q}
        {\HeapUpdateOne{x}{@chan@~c}{u'}})
        $ 
\> \> \> \note{SharedPull} 
\\[1ex]

\> \> $~|~\cs[c]=@in2@ ~\wedge~ s_p[c]=@none@_F ~\wedge~ s_q[c]=@none@_F$ \\
\> \> $\to~\Just (@pull@~c~(@chan@~c)~
      \nextStep
        {l_p}
          {\HeapUpdateOne{c}{@pending@_F}{s_p}}
        {l_q}
          {\HeapUpdateOne{c}{@pending@_F}{s_q}}
        {[]})
  $
\> \> \> \note{SharedPullInject}
\\[1ex]

\> $@drop@~c~(l',u')$ \\
\> \> $~|~\cs[c]=@in1@$
\> \hspace{5em} $\to~\Just (@drop@~c$
\> \hspace{5em} $
      \nextStep
        {l'}
          {s_p}
        {l_q}
          {s_q}
        {u'})
      $
\> \note{LocalDrop} \\

\> \> $~|~\cs[c]=@in1out1@$
\> \hspace{5em} $\to~\Just (@jump@$
\> \hspace{5em} $
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'})
      $
\> \note{ConnectedDrop}\\

\> \> $~|~\cs[c]=@in2@ ~\wedge~ (s_q[c]=@have@_F \vee s_q[c]=@pending@_F)$ 
\> \hspace{5em} $\to~\Just (@jump@$
\> \hspace{5em} $
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'})
      $
\> \note{SharedDropOne}\\

\> \> $~|~\cs[c]=@in2@ ~\wedge~ s_q[c]=@none@_F$
\> \hspace{5em} $\to~\Just (@drop@~c$
\> \hspace{5em} $
      \nextStep
        {l'}
          {\HeapUpdateOne{c}{@none@_F}{s_p}}
        {l_q}
          {s_q}
        {u'})
      $
\> \note{SharedDropBoth}\\[1ex]

% \> @otherwise@ \>
\> $\_$ \> $~|~ @otherwise@ $
\> \hspace{5em} $\to ~ \Nothing$
\> \> \note{Blocked}

\end{tabbing}

\caption{Fusion step for a single process of the pair.}

% Given the state of both processes, compute the instruction this process can perform. This is analogous to statically evaluating the pair of processes. If this process cannot execute, the other process may still be able to.
\label{fig:Fusion:Def:Step}
\end{figure*}

% -------------------------------------
\smallskip
Figure~\ref{fig:Fusion:Def:Step} defines function \ti{tryStep} which schedules a single instruction. This function takes the map of channel types, along with the current label and associated instruction of the first (left) process, and the current label of the other (right) process.

Clause (LocalJump) applies when the left process wants to jump.
In this case, the result instruction simply performs the corresponding jump, leaving the right process where it is.

Clause (LocalCase) is similar, except there are two $\Next$ labels.

Clause (LocalPush) applies when the left process wants to push to a non-shared output channel.
In this case the push can be performed directly, with no additional coordination required.

Clause (SharedPush) applies when the left process wants to push to a shared channel. Pushing to a shared channel requires the downstream process to be ready to accept the value at the same time. We encode this constraint by requiring the static input state of the downstream channel to be $@none@_F$. When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $(@chan@~c)$ and sets the static input state to $@pending@_F$, which indicates that the new value is now available.

Still in Figure~\ref{fig:Fusion:Def:Step}, clause (LocalPull) applies when the left process wants to pull from a local channel, which requires no coordination.

Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to. We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_F$. The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_F$ channel state in the result label records that it has done so.

Clause (SharedPullInject) applies when the left process wants to pull from a shared channel that both processes pull from, and neither already has a value. The result instruction is a @pull@ that loads the stream buffer variable.

Clause (LocalDrop) applies when the left process wants to drop the current value that it read from an unshared input channel, which requires no coordination.

Clause (ConnectedDrop) applies when the left process wants to drop the current value that it received from an upstream process. As the value will have been sent via a heap variable instead of a still extant channel, the result instruction just performs a @jump@ while updating the static channel state.

Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop from a channel shared by both processes. In (SharedDropOne) the channel states reveal that the other process is still using the value. In this case the result is a @jump@ updating the channel state to note that the left process has dropped. In (SharedDropBoth) the channel states reveal that the other process no longer needs the value. In this case the result is a real @drop@, because we are sure that neither process requires the value any longer.

Clause (Blocked) returns @Nothing@ when no other clauses apply, meaning that this process is waiting for the other process to advance.

% -------------------------------------
\begin{figure*}

\begin{minipage}{0.45\textwidth}
\begin{tabbing}
$\ChanTypeTwo$   \TABDEF \kill

\ti{channels} \> $:$ \> $\Proc \to \Proc \to \MapType{\Chan}{\ChanTypeTwo}$ \\

  \> $=$    \> $\{ c=@MMMMMMM@~$\= \kill
$\ti{channels}~p~q$
  \> $=$    \> $\{ c=@in2@$
            \> $|~c$ \= $\in$ \= $(@ins@~p~\cap~@ins@~q) \}$ 
            \\

  \> $\cup$ \> $\{ c=@in1@$
            \> $ |~c$ \> $\in$ \> $(@ins@~p~\cup~@ins@~q)~\wedge~c~\not\in$ \= $(@outs@~p~\cup~@outs@~q) \}$ \\

  \> $\cup$ \> $\{ c=@in1out1@$
            \> $|~c$ \> $\in$ \> $(@ins@~p~\cup~@ins@~q)~\wedge~c~\in$ \> $(@outs@~p~\cup~@outs@~q) \}$ \\

  \> $\cup$ \> $\{ c=@out1@$
            \> $ |~c$ \> $\not\in$ \> $(@ins@~p~\cup~@ins@~q)~\wedge~c~\in$ \> $(@outs@~p~\cup~@outs@~q) \}$ 
\end{tabbing}
\end{minipage}

\newcommand\funClauseDef[3]
{ $\ti{#1}~(#2)$ \> $=$ \> $#3$
}
\newcommand\outlabelsDef[2]
{ \funClauseDef{outlabels}{#1}{\sgl{#2}} 
}

\vspace{1ex}
\begin{minipage}{0.45\textwidth}
\begin{tabbing}
$\ti{outlabels}~(@case@~e~(l_t,u_t)~(l_t,u_f))$ \TABSKIP $=$ \TABSKIP \kill
$\ti{outlabels} ~~:~~ \Instr \to \sgl{\Label}$ \\
\outlabelsDef{@pull@~c~x~(l,u)}{l}              \\
\outlabelsDef{@drop@~c~(l,u)}{l}                \\
\outlabelsDef{@push@~c~e~(l,u)}{l}              \\
\outlabelsDef{@case@~e~(l,u)~(l',u')}{l, l'}    \\
\outlabelsDef{@jump@~(l,u)}{l}
\end{tabbing}
\end{minipage}
\begin{minipage}{0.1\textwidth}
~
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{tabbing}
$\ti{swaplabels}~(@case@~e~((l_1,l_2),u)~((l'_1,l'_2),u'))$ \TABSKIP $=$ \TABSKIP \kill
$\ti{swaplabels} ~~:~~ \Instr \to \Instr$ \\
\funClauseDef{swaplabels}
  {@pull@~c~x~((l_1,l_2),u)}
  {@pull@~c~x~((l_2,l_1),u)}    \\
\funClauseDef{swaplabels}
  {@drop@~c~((l_1,l_2),u)}
  {@drop@~c~((l_2,l_1),u)}      \\
\funClauseDef{swaplabels}
  {@push@~c~e~((l_1,l_2),u)}
  {@push@~c~e~((l_2,l_1),u)}    \\
\funClauseDef{swaplabels}
  {@case@~e~((l_1,l_2),u)~((l'_1,l'_2),u')}
  {@case@~e~((l_2,l_1),u)~((l'_2,l'_1),u')}     \\
\funClauseDef{swaplabels}
  {@jump@~((l_1,l_2),u)}
  {@jump@~((l_2,l_1),u)}
\end{tabbing}
\end{minipage}
\caption{Utility functions}
\label{fig:Fusion:Utils}
\end{figure*}

% NOTE BL: commented out to save space. The figure is mentioned earlier.
% \smallskip
% Figure~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned. Function \ti{channels} computes the $\ChanTypeTwo$ map for a pair of processes. Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states. Function \ti{swaplabels} flips the order of the compound labels in an instruction.

% -----------------------------------------------------------------------------
\subsection{Fusibility}
\label{s:FusionOrder}
When we fuse a pair of processes we commit to a particular interleaving of instructions from each process. When we have at least three processes to fuse, the choice of which two to handle first can determine whether this fused result can then be fused with the third process. Consider the following example, where @alt2@ pulls two elements from its first input stream, then two from its second, before pushing all four to its output.
\begin{code}
 alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
 alternates sInA sInB sInC
  = let  s1   = alt2 sInA sInB
         s2   = alt2 sInB sInC
         sOut = zip s1 s2
    in   sOut
\end{code}
If we fuse the two @alt2@ processes together first, then try to fuse this result process with the downstream @zip@ process, the final fusion transform fails. This happens because the first fusion transform commits to a sequential instruction interleaving where two output values \emph{must} be pushed to stream @s1@ first, before pushing values to @s2@. On the other hand, @zip@ needs to pull a \emph{single} value from each of its inputs alternately.

Dynamically, if we were to execute the first fused result process, and the downstream @zip@ process concurrently, then the execution would deadlock. Statically, when we try to fuse the result process with the downstream @zip@ process the deadlock is discovered and fusion fails. Deadlock happens when neither process can advance to the next instruction, and in the fusion algorithm this manifests as the failure of the $tryStepPair$ function from Figure~\ref{fig:Fusion:Def:StepPair}. The $tryStepPair$ function determines which instruction from either process to execute next, and when execution is deadlocked there are none. Fusibility is an under-approximation for \emph{deadlock freedom} of the network.

% On the upside, fusion failure is easy to detect. It is also easy to provide a report to the client programmer that describes why two particular processes could not be fused.

% The report is phrased in terms of the process definitions visible to the client programmer, instead of partially fused intermediate code. The joint labels used in the fusion algorithm represent which states each of the original processes would be in during a concurrent execution, and we provide the corresponding instructions as well as the abstract states of all the input channels.

% This reporting ability is \emph{significantly better} than that of prior fusion systems such as Repa~\cite{lippmeier2012:guiding}, as well as the co-recursive stream fusion of \cite{coutts2007stream}, and many other systems based on general purpose program transformations. In such systems it is usually not clear whether the fusion transformation even succeeded, and debugging why it might not have succeeded involves spelunking\footnote{def. spelunking: Exploration of caves, especially as a hobby. Usually not a science.} through many pages (sometimes hundreds of pages) of compiler intermediate representations.

In practice, the likelihood of fusion succeeding depends on the particular dataflow network. For fusion of pipelines of standard combinators such as @map@, @fold@, @filter@, @scan@ and so on, fusion always succeeds. The process implementations of each of these combinators only pull one element at a time from their source streams, before pushing the result to the output stream, so there is no possibility of deadlock. Deadlock can only happen when multiple streams fan-in to a process with multiple inputs, such as with @merge@.

When the dataflow network has a single output stream then we use the method of starting from the process closest to the output stream, walking towards the input streams, and fusing in successive processes as they occur. This allows the interleaving of the intermediate fused process to be dominated by the consumers, rather than producers, as consumers are more likely to have multiple input channels which need to be synchronized. In the worst case the fall back approach is to try all possible orderings of processes to fuse.

% -----------------------------------------------------------------------------
\section{Implementation}
\label{s:Evaluation}

Stream fusion is ultimately performed for practical reasons: we want the fused result program to run faster than the original unfused program.

% -----------------------------------------------------------------------------
\subsection{Finite streams}
\label{s:Finite}

The processes we have seen so far deal with infinite streams, but in practice most streams are finite. Certain combinators such as @fold@ and @append@ only make sense on finite streams, and others like @take@ produce inherently finite output. We have focussed on the infinite stream version as it is simpler to explain and prove, but supporting finite streams does not require substantial conceptual changes.

Unlike infinite streams, pulling from a finite stream can fail, meaning the stream is finished. We therefore modify the @pull@ instruction to have two output labels: one to execute when a value is pulled, and the other to execute when the stream is finished. On the pushing end, we also need some way of finishing streams, so we add a new instruction to close an output stream.

During evaluation we need some way of knowing whether a stream is closed, which can be added as an extra constructor in the \InputState~ type. The same constructor is added to the static input state used by fusion. In this way, for any changes made to evaluation, the analogous static change must be made in the fusion transform.

It is also possible to encode finite streams as infinite streams with an explicit end-of-stream marker (EOF) and case statements. However, this requires the fusion transform to reason about case statements' predicates.
By making the structure of finite streams explicit and constraining how processes use finite streams, it is not necessary to rely on heuristics for deciding equality of predicates.

% Waffle waffle to get url on its own line
This finite stream extension is described in more detail in the appendix of the extended version of this paper, which is available at \url{http://cse.unsw.edu.au/~amosr/papers/merges.pdf}.

% It seems obvious that if two consumers read the same value and check if it is EOF, they will both be the same, but this is undecidable in general.

% -----------------------------------------------------------------------------
\subsection{Benchmarks}
We have implemented this system using Template Haskell in a library called @folderol@\footnote{\url{https://github.com/amosr/folderol}}.
To show practical examples, we use the finite stream extension mentioned in~\S\ref{s:Finite}.
We present three benchmarks: two file-based benchmarks, and one array algorithm.

For the file benchmarks, we compare against three Haskell streaming libraries: `Conduit', `Pipes', and `Streaming'.
These streaming libraries are pull-based, and do not naturally support multiple outputs: the split in the dataflow graph must be hand-fused, or somehow rewritten as a straight-line computation.
These libraries also have a monadic interface, which allows the structure of the dataflow graph to depend on the values. This expressiveness has a price: if the dataflow graph can change dynamically, we cannot statically fuse it.
% This price is paid even when the dataflow graph is static, as the same monadic structure is used: because repetition is expressed as an unfolding dynamic graph, even computations that would be static in other systems must be expressed dynamically, reducing the possibility for fusion.

The first file benchmark simply appends two files while counting the lines.
In Pipes and Conduit, counting the lines is implemented as a pipe which counts each line before passing it along.
The first group in Figure~\ref{fig:bench:all} shows the runtimes for appending 2MB of data.

The second file benchmark takes a file and partitions it into two files: one with even-length lines, and one with odd-length lines.
The output lines are also counted.
Even with partial hand-fusion because of the multiple outputs, the Pipes and Conduit programs are slower than ours, as well as losing the abstraction benefits from using a high-level library.
The `Streaming' library allows streams to be shared in a fairly straightforward way and does not require hand-fusion, but is also the slowest in this benchmark.
The second group in Figure~\ref{fig:bench:all} shows the runtimes for partitioning a 1MB file.

Quickhull is a divide-and-conquer spatial algorithm to find the smallest convex hull containing all points.
At its core is an operation called `filterMax' which takes a line and an array of points, and finds the farthest point above the line, as well as all points above the line.

Here we also compare against a @Data.Vector@ program, which uses shortcut fusion.
The shortcut fusion system cannot fuse both operations into a single loop, and both operations must recompute the distances between the line and each point.
% This means a choice must be made: either compute the distances upfront and share them, or recompute the distances in each operation.
% We compared both, and recomputing was significantly faster.
% However, we only benchmarked with two-dimensional points: at higher dimensions, the cost of recomputing distances may outweigh array allocation.
As before, the Conduit and Pipes programs must be partially hand-fused.
The third group in Figure~\ref{fig:bench:all} shows the runtimes for Quickhull over 40MB of data, or half a million points, while the final group uses more data (120MB) to compare directly against @Data.Vector@.

% -----------------------------------------------------------------------------
\subsection{Optimisation and Drop Instructions}
\label{s:Optimisation}
After we have fused two processes together, it may be possible to simplify the result before fusing in a third. Consider the result of fusing @group@ and @merge@ which we saw back in Figure~\ref{fig:Process:Fused}. At labels @F1@ and @F2@ are two consecutive @jump@ instructions. The update expressions attached to these instructions are also non-interfering, which means we can safely combine these instructions into a single @jump@. In general, we prefer to have @jump@ instructions from separate processes scheduled into consecutive groups, rather than spread out through the result code. The (PreferJump) clauses of Figure~\ref{fig:Fusion:Def:StepPair} implement a heuristic that causes jump instructions to be scheduled before all others, so they tend to end up in these groups.

Other @jump@ instructions like the one at @F5@ have no associated update expressions, and thus can be eliminated completely. Another simple optimization is to perform constant propagation, which in this case would allow us to eliminate the first @case@ instruction.

Minimising the number of states in an intermediate process has the follow-on effect that the final fused result also has fewer states. Provided we do not change the order of instructions that require synchronization with other processes (@pull@, @push@ or @drop@), the fusibility of the overall process network will not be affected.

% Another optimization is to notice that in some cases, when a heap variable is updated it is always assigned the value of another variable. In Figure~\ref{fig:Process:Fused}, the @v@ and @x1@ variables are only ever assigned the value of @b1@, and @b1@ itself is only ever loaded via a @pull@ instruction. Remember from \S\ref{s:Fusion:FusingPulls} that the variable @b1@ is the stream buffer variable. Values pulled from stream @sIn1@ are first stored in @b1@ before being copied to @v@ and @x1@. When the two processes to be fused share a common input stream, use of stream buffer variable allows one process to continue using the value that was last pulled from the stream, while the other moves onto the next one.

When the two processes are able to accept the next variable from the stream at the same time, there is no need for the separate stream buffer variable. This is the case in Figure~\ref{fig:Process:Fused}, and we can perform a copy-propagation optimisation, replacing all occurrences of @v@ and @x1@ with the single variable @b1@. To increase the chance that we can perform copy-propagation, we need both processess to want to pull from the same stream at the same time. Moving the @drop@ instruction for a particular stream as late as possible prevents a @pull@ instruction from a second process being scheduled in too early.

% In the result, the interleaved instructions from both source processes then share the same heap variable.

% In general, the @drop@ for a particlar stream should be placed just before a @pull@ from the same stream. 
\begin{figure}
\usetikzlibrary{patterns}
\begin{tikzpicture}

% Bar chart for benchmarks:
% This is a complicated bar chart because there are six series (conduit, pipes, vector..), but they are not all defined at all points (append, partition, quickhull).
% We need to split it into three different pictures, overlaid on top of each other.
% The first picture just defines the legend, the axis labels, etc.
\begin{axis}[
  symbolic x coords={Append, Partition, Hull 40MB, Hull 120MB, end},
  xtick=data, ylabel=Runtime (ms),
y label style={at={(axis description cs:0.1,.5)},anchor=south},
  % All pictures need to specify same upper and lower bounds so they line up together
  ymin=0, ymax=300,
  xmin=Append, xmax=Hull 120MB,
        enlargelimits=0.01,
  enlarge x limits=0.20,
        ybar=1pt,
        bar width=7pt,
        small,
  width=8.5cm, height=4.5cm,
  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3},
]
% The colours need to be specified manually, to force the legend entries to show up..
% The colours also need to match the colours specified for each series data.
\addlegendimage{brown,pattern=north east lines,pattern color=brown}
\addlegendimage{red,fill=red!40!white}
\addlegendimage{gray,fill=gray!40!white}
\addlegendimage{blue,fill=blue!40!white}
\addlegendimage{teal,fill=teal!40!white}
\addlegendimage{orange,fill=orange!40!white}
\legend{Ours, Handfused, Vector, Conduit, Pipes, Streaming}
\addplot[white] coordinates {(Append,  0) (Partition,  0)  (Hull 40MB, 0) (Hull 120MB, 0)  };
\end{axis}

\begin{axis}[
  symbolic x coords={Append, Partition, Hull 40MB, Hull 120MB, end},
  ticks=none,
  ymin=0, ymax=300,
  xmin=Append, xmax=Hull 120MB,
        enlargelimits=0.01,
  enlarge x limits=0.20,
        ybar=1pt,
        bar width=7pt,
        small,
  width=8.5cm, height=4.5cm,
  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3},
]
% Ours
\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {(Append,  86) (Partition,  61)   };
% Handfused
\addplot[red,fill=red!40!white] coordinates {(Append,  88) (Partition,  60)   };
% Conduit
\addplot[blue,fill=blue!40!white] coordinates {(Append, 282) (Partition, 136)   };
% Pipes
\addplot[teal,fill=teal!40!white] coordinates {(Append, 270) (Partition, 166)  };
% Streaming
\addplot[orange,fill=orange!40!white] coordinates {(Append, 168) (Partition, 268)  };
\end{axis}

\begin{axis}[
  symbolic x coords={Append, Partition, Hull 40MB, Hull 120MB, end},
  ticks=none,
  ymin=0, ymax=300,
  xmin=Append, xmax=Hull 120MB,
        enlargelimits=0.01,
  enlarge x limits=0.20,
        ybar=1pt,
        bar width=7pt,
        small,
  width=8.5cm, height=4.5cm,
  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3},
]
% Ours
\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {(Hull 40MB, 6) };
% Handfused
\addplot[red,fill=red!40!white] coordinates {(Hull 40MB,  3) };
% Vector
\addplot[gray,fill=gray!40!white] coordinates {(Hull 40MB, 8)  };
% Conduit
\addplot[blue,fill=blue!40!white] coordinates {(Hull 40MB, 275)  };
% Pipes
\addplot[teal,fill=teal!40!white] coordinates {(Hull 40MB, 132) };
% Streaming
\addplot[orange,fill=orange!40!white] coordinates {(Hull 40MB, 150) };
\end{axis}

\begin{axis}[
  symbolic x coords={Append, Partition, Hull 40MB, Hull 120MB, end},
  ticks=none,
  ymin=0, ymax=300,
  xmin=Append, xmax=Hull 120MB,
        enlargelimits=0.01,
  enlarge x limits=0.20,
        ybar=1pt,
        bar width=7pt,
        small,
  width=8.5cm, height=4.5cm,
  legend style={at={(0.5,-0.2)},anchor=north, legend columns=3},
]
% Ours
\addplot[brown,pattern=north east lines,pattern color=brown] coordinates {  (Hull 120MB, 202)  };
% Handfused
\addplot[red,fill=red!40!white] coordinates {  (Hull 120MB, 126)  };
% Vector
\addplot[gray,fill=gray!40!white] coordinates {(Hull 120MB, 274)  };
\end{axis}

\end{tikzpicture}
\caption{Runtime for benchmarks; lower is faster.}
\label{fig:bench:all}
\end{figure}

\section{Proofs}
\label{s:Proofs}

Our fusion system is formalized in Coq, and we have proved soundness of \ti{fusePair}: if the fused result process produces a particular sequence of values on its output channels then the two source processes may also produce that same sequence. The converse is not true, however: concurrent evaluation represents all possible interleavings, of which the fused process is only one.

% Note that the converse is not necessarily true: just because two processes can evaluate to a particular output does not mean the fused program will evaluate to that. This is because, as explained in~\S\ref{s:EvaluationOrder}, evaluation of a process network is non-deterministic, and fusion commits to a particular evaluation order.

\begin{code}
Theorem Soundness 
  (P1 : Program L1 C V1) (P2 : Program L2 C V2)
  (ss : Streams) (h : Heap) (l1 : L1) (l2 : L2)
  (is1 : InputStates) (is2 : InputStates) 
  :  EvalBs (fuse P1 P2) ss h (LX l1 l2 is1 is2)
  -> EvalOriginal Var1 P1 P2 is1 ss h l1
  /\ EvalOriginal Var2 P2 P1 is2 ss h l2.
\end{code}

@EvalBs@ evaluates the fused program, and @EvalOriginal@ ensures that the original program evaluates with that program's subset of the result heap. The Coq formalization has some small differences from the system in this paper. Instead of implementing non-deterministic evaluation we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.

% Care must be taken to remove stream values that the other process has pulled but this one has not yet.
%%% AR: Really would like to say something about this but no room to explain properly
% For shared inputs when one program has pulled, the other program must be evaluated with the other value removed from the end of the stream.

% Firstly, the Coq formalization uses a separate @update@ instruction to modify variables in the local heap, rather than attaching heap updates to the \Next~ label of every instruction. Performing this desugaring makes the low level lemmas easier to prove, but we find attaching the updates to each instruction makes for an easier exposition. Secondly, our formalization only implements sequential evaluation for a single process, rather than non-deterministic evaluation for whole process groups as per Figure~\ref{fig:Process:Eval:Feed}. Instead, we sequentially evaluate each source processes independently, and compare the output values to the ones produced by sequential evaluation of the fused result process. This is sufficient for our purposes because we are mainly interested in the value correctness of the fused program, rather than making a statement about the possible execution orders of the source processes when run concurrently.

% This causes the fusion definition to be slightly more complicated, as two output instructions must be emitted when performing a push or pull followed by an update. This is a fairly minor difference, and we have made this change in the paper version for ease of exposition.
% Ideally, a future version of the formalisation would include this change.
% Secondly, our formalisation does not implement the concurrent evaluation semantics for processes, only sequential evaluation for a single process. Instead we sequentially evaluate both processes with the same input values and outputs. Despite these differences, the Coq formalisation gives us sufficient confidence in the correctness of the version presented here.

\eject
\section{Related work}

This work aims to address the limitations of current combinator-based array fusion systems. As stated in the introduction, neither pull-based or push-based fusion is sufficient. Some combinators are inherently push-based, particularly those with multiple outputs such as @unzip@; while others are inherently pull-based, such as @zip@.

% Short cut fusion is an attractive idea, as it allows fusion systems to be specified by a single rewrite rule. Thus, short cut fusion is inherently biased towards pull fusion.

However, short cut fusion \cite{gill1993short} relies on inlining which, like pull-based streams, only occurs when there is a single consumer. Push-based short cut fusion systems \emph{do} exist, notably the original foldr/build formulation, but support neither @zip@ nor @unzip@ \cite{svenningsson2002shortcut,lippmeier2013data}.

Recent work on stream fusion \cite{kiselyov2016stream} uses staged computation in a push-based system to ensure all combinators are inlined. When streams are used multiple times this causes excessive inlining, which duplicates work. This can change the semantics for effectful streams.

% For effectful inputs such as reading from the network, duplicating work changes the semantics.
% I could write more about this eg only supporting a single output, but the other points probably apply to push streams in general

Data flow fusion~\cite{lippmeier2013data} is neither pull-based nor push-based, and supports arbitrary splits and joins. It supports only a fixed set of standard combinators such as @map@, @filter@ and @fold@, and converts each stream to a series with explicit rate types, similar to the clock types of Lucid Synchrone \cite{benveniste2003synchronous}.
% These rate types ensure that well-typed programs can be fused without introducing unbounded buffers.
% This allows unfusable programs to be caught at compile time.
% However, it only supports a limited set of combinators, and adding more is non-trivial.

One way to address the difference between pull and push streams is to explicitly support both \cite{bernardy2015duality, lippmeier2016polarized}. Here, pull streams have the type @Source@ and represent a source, always ready to be pulled, while push streams have the type @Sink@ and represent a sink, always ready to accept data. The addition of push streams allows more programs to be fused than pull-only systems, but the computation must be manually split into sources and sinks.

% This system requires the streaming computation to be manually split into sources and sinks.
% Both systems rely on stream bindings being used linearly to ensure correctness.
% Operations over sources are expressed fairly naturally compared to streams, for example the @zip@ combinator has the type @Source a -> Source b -> Source (a,b)@.
% Sinks are co-variant, and operations must be performed somewhat backwards, so that the @unzip@ combinator takes the two output sinks to push into and returns a new sink that pushes into these.
% It has the type @Sink a -> Sink b -> Sink (a,b)@.
% and be joined together by a loop that `drains' values from the source and pushes them into the sink.

The duality between pull and push arrays has also been explored in Obsidian \cite{claessen2012expressive, svensson2014defunctionalizing}. Here the distinction is made for the purpose of code generation for GPUs rather than fusion, as operations such as appending pull arrays require conditionals inside the loop, whereas using push arrays moves these conditionals outside the loop.

Meta-repa \cite{ankner2013edsl} supports both array types in a similar way, using Template Haskell for code generation. It supports fusion on both array types. When arrays are used multiple times, they must be explicitly reified with a `force' operation to avoid duplication.

Streaming IO libraries have blossomed in the Haskell ecosystem, generally based on Iteratees \cite{kiselyov2012iteratees}. Libraries such as @conduit@ \cite{hackage:conduit}, @enumerator@ \cite{hackage:enumerator}, @machines@ \cite{hackage:machines} and @pipes@ \cite{hackage:pipes} are all designed to write stream computations with bounded buffers, but do not guarantee fusion.

% However, these libraries do not guarantee fusion guarantees, and as such programs tend to be written over chunks of data to make up for the communication overhead.

% For the most part they support only straight-line computations, with only limited forms of branching.

Like our own Machine Fusion,  Filter Fusion~\cite{proebsting1996filter} also statically interleaves the code of producer and consumer processes. Each process must have a single input and output channel, so common operators like @zip@, @unzip@, @append@, @partition@ and so on are not supported. Given an adjacent producer and consumer pair, Filter Fusion alternately assigns control to the code of each. When the consumer needs input, control is passed to the producer, and when the producer produces its value control is passed back to the consumer. This simple scheduling algorithm works only for straight line pipelines of processes. Machine fusion provides a finer grained interleaving of code, which is nessesary to support branching dataflows that contain both splits and joins.

%Filter Fusion~\cite{proebsting1996filter} uses a similar algorithm, but only supports processes with at most one input and one output channel.
% This means combinators such as @zip@, @unzip@, @append@, and @partition@ cannot be expressed.
%Also, the process network must be a straight line, with no splits or joins.
%Here a consumer and a producer are fused by alternating between the two; when the consumer needs input, control is passed to the producer, and when the producer has output, control is passed to the consumer.
%This is sufficient when only one channel is shared, but when multiple channels are shared finer-grained coordination is required.

% This is sufficient for producer-consumer fusion, but two consumers pulling from the same input channel need not follow the same alternation.
% This is sufficient when only one , but when multiple channels are shared between two processes, as may be the case in an intermediate fused result of a larger program, it is not obvious whether simply alternating the two processes is correct.

% The synchronised product of two processes allows either process to take independent or local steps at any time, but shared actions, such as when both processes communicate on the same channel, must be taken in both processes at the same time.
% When two processes share multiple channels, synchronised product will fail unless both processes read the channels in exactly the same order.
% In our system the use of stream buffer variables allows some leeway in when processes must take shared steps.

% This is a much simpler fusion method than ours, but is also much stricter.

% They do not have to take shared steps at the same time, but if one process lags behind the other, it must catch up before the other one gets too far ahead.

Synchronous languages such as LUSTRE~\cite{halbwachs1991synchronous}, Lucy-n~\cite{mandel2010lucy} and SIGNAL~\cite{le2003polychrony} all use some form of clock calculus and causality analysis to ensure that programs can be statically scheduled with bounded buffers~\cite{caspi1996:kahn}. These languages describe \emph{passive} processes where values are fed in to streams from outside environments, such as data coming from sensors. In contrast, our processes are \emph{active} and have control over when data is pulled from the source streams, which is nessesary for multiple input combinators such as @merge@ and  @append@.
Synchronous dataflow languages reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling \cite{bouakaz2013real}.

% In this case, the passive process has no control over the rate of input coming in, and if they support multiple input streams, they must accept values from them in any order.

% Note that in the synchronous language literature, it is common to refer to a different merge operation, also known as @default@, which computes a stream that is defined whenever either input is defined.

Synchronous dataflow (SDF; not to be confused with synchronous languages above) is a dataflow graph model of computation where each node has constant, statically known input and output rates. StreamIt~\cite{thies2002streamit} uses synchronous dataflow for scheduling when possible, otherwise falling back to dynamic scheduling~\cite{soule2013dynamic}. Boolean dataflow and integer dataflow~\cite{buck1993scheduling,buck1994static} extend SDF with boolean and integer valued control ports, but only support limited control flow.

% SDF is simple enough for static scheduling to be decidable, but this comes at a cost of expressivity.

% Finite state machine-based scenario aware dataflow (FSM-SADF)~\cite{stuijk2011scenario,van2015scenario} is quite expressive compared to boolean and integer dataflow, while still ensuring static scheduling. A finite state machine is constructed, where each node of the FSM denotes its own SDF graph. The FSM transitions from one dataflow graph to another based on control outputs of the currently executing dataflow graph. For example, a filter is represented with two nodes in the FSM. The initial state executes the predicate, and the value of the predicate is used to determine which transition the FSM takes: either the predicate is false and the FSM stays where it is, or the predicate is true and moves to the next state. The next state emits the value, and returns to the start. This appears to be able to express value-dependent operations such as merge, but lacks the composability and familiarity of combinators.

% StreamIt:
% Only allows limited splits and joins: round robin and duplication for splits, round robin and combination for joins.
% Does not support fully general graphs - instead using combinators to introduce a (split/join) and a combinator for a feedback loop.
%
% Parameterized dataflow (PDF),  \cite{bhattacharya2001parameterized}
% Schedulable parametric dataflow (SPDF),  \cite{fradet2012spdf}

% Recent work on stream fusion by \citet{kiselyov2016stream} uses staged computation to ensure all combinators are inlined, but for splits this causes excessive inlining which duplicates work, due to values of the source arrays being read multiple times.

\bibliography{robinson-2}

\end{document}

### 2023-05-03 2023pipit/Main.tex
start writing pipit

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
\documentclass[sigplan,screen]{acmart}

% any of these necessary?
% \usepackage{amssymb}
\usepackage{amsthm}
% \usepackage{graphicx}
\usepackage{amsmath}
% \usepackage{mathptmx}
% \usepackage{mathtools}
% \usepackage{stmaryrd}
% \usepackage{adjustbox}
% \usepackage{hyperref}
% \usepackage{alltt}
% \usepackage{url}
% \usepackage{float}
% \usepackage{minipage-marginpar}

\usepackage{style/utils}
\usepackage{style/code}
\usepackage{style/proof}
\usepackage{style/keywords}
\usepackage{style/judgements}

\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

\acmConference[TyDe '23]{In submission}{2023}{In submission}
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
% \acmPrice{15.00}
% \acmISBN{978-1-4503-XXXX-X/18/06}
%%\acmSubmissionID{123-A56-BU3}

\begin{document}

\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}

\newcommand{\affiliationANU}{
  \affiliation{%
  \institution{Australian National University}
  \city{Canberra}
  \country{Australia}
}
}

\author{Amos Robinson}
% \authornote{}
\email{amos.robinson@anu.edu.au}
\orcid{0009-0004-4837-4981}
\affiliationANU{}
\author{Alex Potanin}
% \authornotemark[1]
\email{alex.potanin@anu.edu.au}
\orcid{0000-0002-4242-2725}
\affiliationANU{}

% \renewcommand{\shortauthors}{Robinson et al.}

\begin{abstract}
  Safety-critical control systems, such as the anti-lock braking systems in most cars today, need to be correct and execute in real-time.
  We could write these systems directly in an imperative language such as C or Rust and verify them, but it can be difficult to state high-level specifications about such low-level implementations.
  Another approach, favoured by large parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre or SCADE, and verify that the implementations satisfy the high-level specification using a model-checker.
  These model-checkers can prove many interesting properties automatically, but do not provide many options when the automated proof techniques fail.
  We introduce Pipit, an embedded domain-specific language for implementing and verifying such controllers in \fstar.
  Pipit aims to provide a high-level language similar to Lustre, while reusing \fstar's proof automation and manual proofs for verifying controllers, and using LowStar's C-code generation for real-time execution.
  Pipit translates its expression language to a transition system for k-inductive proofs, which is mostly verified; verifying the translation to imperative code is future work.
\end{abstract}

%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
\begin{CCSXML}
<ccs2012>
    <concept>
        <concept_id>10010520.10010553.10010562.10010564</concept_id>
        <concept_desc>Computer systems organization~Embedded software</concept_desc>
        <concept_significance>300</concept_significance>
        </concept>
    <concept>
        <concept_id>10010520.10010570.10010572</concept_id>
        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
    <concept>
        <concept_id>10003752.10010124.10010138.10010142</concept_id>
        <concept_desc>Theory of computation~Program verification</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
    <concept>
        <concept_id>10003752.10003790.10003793</concept_id>
        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
        <concept_significance>100</concept_significance>
        </concept>
    <concept>
        <concept_id>10011007.10011006.10011050.10011023</concept_id>
        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
        <concept_significance>500</concept_significance>
        </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computer systems organization~Embedded software}
\ccsdesc[500]{Computer systems organization~Real-time languages}
\ccsdesc[500]{Theory of computation~Program verification}
\ccsdesc[100]{Theory of computation~Modal and temporal logics}
\ccsdesc[500]{Software and its engineering~Specialized application languages}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Lustre, streaming, reactive, control}

\received{1 June 2023}
% \received[revised]{12 March 2009}
% \received[accepted]{5 June 2009}

\maketitle

% enable @verbatim@ syntax
\makeatactive

\section{Introduction}

Cite: Kind2, Lustre, Vélus.
Antilock brakes paper.
The original Kind~\cite{hagen2008verifying}.

\begin{verbatim}
  let veh_speed_estimate ()
\end{verbatim}

\section{Core language}
\label{s:core}

\input{figures/core-grammar.tex}
\input{figures/core-typing.tex}
\input{figures/core-bigstep.tex}

The grammar of Pipit is defined in \autoref{f:core-grammar}.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \CITE.
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
To ensure that streams are productive, recursive equations can only refer to their previous values and must be \emph{guarded} by a delay; this restriction is formalised by the causality check in \REF{}.
This form of recursion differs somewhat from standard Lustre, which uses a set of mutually recursive bindings.
In \autoref{ss:operational-semantics} we will see that this form of recursion allows a syntax-directed semantics, but at the cost of potentially duplicating work.

Let-expressions ($\xlet{x}{e}{e'}$) are standard.
Finally, check-expressions ($\xcheck{e}{e'}$) are used to describe properties that the program should satisfy.
This operation corresponds roughly to the @assert@ primitive in \fstar{}; however we avoid this name, as @assert@ is historically treated as an unchecked assumption in many Lustre languages.

The grammar for values $v$ includes natural numbers, integers and functions embedded from the \fstar{} meta-language.
We denote functions in the meta-language by the explicit meta-lambda notation ($\mlamX{}$) to avoid any potential ambiguity between Pipit expressions ($e$) and meta-expressions ($\metafy{e}$).

\subsection{Typing rules}
\label{ss:typing-rules}

The typing rules of Pipit are defined in \autoref{f:core-typing}, with the judgment form $\typing{\Gamma}{e}{\tau}$ that checks expression $e$ under streaming context $\Gamma$.
The grammar of types (\autoref{f:core-grammar}) distinguishes between bounded-size primitive types ($\primty$) and potentially higher-order types ($\tau$), which allows us to restrict streaming operations to only require bounded buffers.
\TODO{Not sure if this restriction is useful, as arbitrary functions can still use however much memory and time as they want.}

For values (rule TValue), we reuse the \fstar{} meta-language's typing judgment ($\mtyping{}{v}{\tau}$) for typing closed values $v$.
Otherwise, most of the rules are standard.

\subsection{Operational semantics}
\label{ss:operational-semantics}

The operational semantics, expressed as a bigstep relation, are defined in \autoref{f:core-bigstep}.
The judgment form $\bigstep{\Sigma}{e}{v}$ takes a \emph{stream history} ($\Sigma$), which is a non-empty sequence of stores, as well as an expression to evaluate ($e$) and the final value ($v$).

Rule Value denotes that a value evaluates to itself under any streaming history.
Rule Var denotes that evaluating a variable requires looking it up in the right-most (most recent) store; the syntax $\Sigma_\bot; \sigma$ means that this rule applies to any streaming history with one or more steps.
Rule App-Meta allows applying functions on the meta-level; the syntax $(\mlamX{}) v'$ refers to meta-level application of some meta-term.

The rules $\mbox{Fby}_1$ and $\mbox{Fby}_S$ describe the semantics of delayed streams.
Rule $\mbox{Fby}_1$ denotes that when a delay $\xfby{v}{e}$ is evaluated in a context with exactly one step --- that is, at the very start of program execution --- it simply returns the initial value $v$.
Rule $\mbox{Fby}_S$ applies when the same delay is evaluated in a context with strictly more than one step ($\Sigma; \sigma$), in which case, the most recent step is discarded, and the original expression $e$ is evaluated under the previous history.

Next, rules ($\xthenarrow_1$) and ($\xthenarrow_S$) describe the semantics of streams composed by \emph{then}-expressions; as with the rules for delayed streams, the first rule applies when the streaming history contains exactly one step, while the second applies for histories of strictly more than one step.

Rule ($\mu{}$) describes the semantics of recursive streams.
This rule performs a single unfolding of a recursive expression $\xrec{x}{e[x]}$ so that all occurrences of $x$ inside $e$ are replaced with the recursive expression itself.
This rule is not trivially terminating, but so long as any occurrences of the recursive binder are guarded by a delay, then all occurrences of the recursive binder will have a shorter stream history.
This intuition is formalised in \REF{}.

Rule Let uses a standard substitution-based semantics for let-expressions.
Finally, rule Check evaluates a property and requires it to be true ($\top$).

The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
In the reactive semantics of \CITEME{88}, as well as the coinductive semantics of Vélus \CITEME{Vélus}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.

In contrast, the semantics presented here is syntax-directed and completely deterministic.
We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \REF.
However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
For this reason, the semantics here is not suitable for actual execution.

\pagebreak
% \begin{acks}
%   acknowledgement
% \end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{Main}

% \appendix

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

### 2023-05-03 2023pipit/figures/core-grammar.tex
start writing pipit

\begin{figure}
  \[
  \begin{array}{lrl}
    e, e' & := & v ~|~ x ~|~ e~e' \\
          & | & \xfby{v}{e} \\
          & | & \xthen{e}{e'} \\
          & | & \xrec{x}{e[x]} \\
          & | & \xlet{x}{e}{e'[x]} \\
          & | & \xcheck{e}{e'} \\
    \\
    v & := & n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ \mlamX
    \\
    \sigma & := & \sgl{\ov{x \mapsto v}} \\
    \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
      \\
    \primty & := & \mathbb{N} ~|~ \mathbb{B} \\
    \tau, \tau' & := & \primty ~|~ \hdots ~|~ \tau \to \tau' \\
    \Gamma & := & \cdot ~|~ x : \tau, \Gamma \\
    \end{array}
  \]
  \TODO{compact grammar}
  \caption{Pipit language grammar, including expressions ($e$); values and host functions embedded from the \fstar{} meta-language ($v$); stores ($\sigma$); non-empty stream histories ($\Sigma$); bounded-size types ($\primty$); unbounded types ($\tau$); and typing contexts ($\Gamma$).}
  \label{f:core-grammar}
\end{figure}

### 2023-05-03 2023pipit/Main.tex
initial writeup

@@
 
-Cite: Kind2, Lustre, Vélus.
-Antilock brakes paper.
-The original Kind~\cite{hagen2008verifying}.
+We are interested in verifying an antilock braking system for a motorcycle\cite{huang2010design}.
 
-\begin{verbatim}
-  let veh_speed_estimate ()
-\end{verbatim}
+\begin{tabbing}
+  @MM@\= @let @\= \kill
+  @let@ veh\_speed\_estimate $\omega_F$ $\omega_R$ $a_z$ = \\
+    \> @let@ $v_F$ = $\omega_F \cdot $ circumference @in@ \\
+    \> @let@ $v_R$ = $\omega_R \cdot $ circumference @in@ \\
+    \\
+    \> @let rec@ $\floor{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
+    \> \> @then@ $\min v_F v_R$ \\
+    \> \> @else@ $(\xthen{\min v_F v_R}{\xfby{0}{\floor{\hat{v}}}}) + a_z - \epsilon$ @in@ \\
+    \> @let rec@ $\ceil{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
+    \> \> @then@ $\max v_F v_R$ \\
+    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@
+    \\
+    \> $(\floor{\hat{v}}, \ceil{\hat{v}})$
+\end{tabbing}
+
+Cite: Kind2\cite{champion2016kind2}, Lustre\cite{caspi1995functional}, Vélus\cite{bourke2017formally}.
+The original Kind\cite{hagen2008scaling}.
+
+Past-time modal temporal logic \cite{lichtenstein1985glory}
+Lustre support for temporal logic \cite{halbwachs1993executable}
 
@@
 The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
-Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \CITE.
+Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \cite{caspi1995functional}.
 The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
@@
 The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
-In the reactive semantics of \CITEME{88}, as well as the coinductive semantics of Vélus \CITEME{Vélus}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
+In the reactive semantics of \cite{caspi1995functional}, as well as the coinductive semantics of Vélus \CITE{bourke2017formally find best paper}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
 This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.
@@
 In contrast, the semantics presented here is syntax-directed and completely deterministic.
-We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \REF.
+We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \autoref{ss:meta-theory}.
 However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
 For this reason, the semantics here is not suitable for actual execution.
+
+\subsection{Causality}
+\label{ss:causality}
+
+Talk about the causality restriction.
+Does expression $e$ have a dependency on the most-recent value of expression $e'$?
+
+\begin{code}
+let rec direct_dependency (e e': exp) : bool =
+  if e = e' then true
+  else
+  match e with
+  | XVal _ -> false
+  | XVar x' -> false
+  | XPrim2 _ e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
+  | XIte ep et ef -> direct_dependency ep e' || direct_dependency et e' || direct_dependency ef e'
+  | XThen e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
+  | XPre _ -> false
+  | XMu e1 -> direct_dependency e1 (lift e' 0)
+  | XLet e1 e2 -> direct_dependency e1 e' || direct_dependency e2 (lift e' 0)
+
+let rec causal (e: exp): bool =
+  match e with
+  | XVal _ -> true
+  | XVar x' -> true
+  | XPrim2 _ e1 e2 -> causal e1 && causal e2
+  | XIte ep e1 e2 -> causal ep && causal e1 && causal e2
+  | XThen e1 e2 -> causal e1 && causal e2
+  | XPre e1 -> causal e1
+  | XMu e1 -> causal e1 && not (direct_dependency e1 (XVar 0))
+  | XLet e1 e2 -> causal e1 && causal e2
+\end{code}
+
+\begin{lemma}[no dependency, no difference]
+If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.
+
+For all expressions $e$ and well-scoped variables $x$,
+if $\neg \mbox{directly} ~e~ x$ and
+$\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
+then for any $v_x'$, evaluation is unchanged
+$\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
+\end{lemma}
+
+We also need:
+\begin{lemma}[Bigstep recursive XMu]
+  If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:
+
+  if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
+  then for any $v'$,
+  $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
+\end{lemma}
+
+% let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
+%   (streams: C.table (outer + 1) inner)
+%   (vs: C.vector value outer) (v v': value)
+%   (hBSmu: bigstep streams (XMu e) (v :: vs)):
+%     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =
+
+\begin{lemma}[progress]
+  if expression is causal, then for any valid stream history it should evaluate to a value:
+
+  if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
+  $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
+\end{lemma}
+
+\begin{lemma}[monotonic progress]
+  if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
+  there exists some $v'$ such that
+  $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
+\end{lemma}
+
+\subsection{Meta-theory}
+\label{ss:meta-theory}
+
+We use a typed representation, which ensures an intrinsic variant of preservation.
+For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.
+
+\begin{code}
+let bigstep_deterministic
+  (#streams: list (C.row 'c))
+  (#e: exp 'c 'a)
+  (#v1 #v2: 'a)
+  (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
+    Lemma (ensures (v1 == v2)) (decreases hBS1) =
+  bigstep_proof_equivalence hBS1 hBS2
+\end{code}
+
+Monotonicity
+
+Substitution-closure
+\begin{code}
+let rec bigstep_substitute_as_var (e p: exp)
+  (streams1: C.table outer inner1)
+  (streams2: C.table outer inner2)
+  (vsp vsep: C.vector value outer)
+  (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
+  (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
+    Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =
+
+\end{code}
+
+\section{Proving it with transition systems}
+
+Describe the translation from Pipit expressions to labelled transition systems.
+
+K-induction
+
+Statement of LTS soundness:
+\begin{code}
+let system_eval_complete'
+    (#outer: nat) (#vars: nat)
+    (e: exp { causal e /\ wf e vars })
+    (streams: C.table outer vars)
+    (vs: C.vector value outer)
+    (hBS: bigstep streams e vs):
+      Lemma (exists (s': option (state_of_exp e)).
+        system_stepn'
+          (system_of_exp e vars)
+          (C.Table?._0 streams) vs s') =
+    match outer with
+    | 0 -> assert
+      (system_stepn'
+        (system_of_exp e vars)
+        (C.Table?._0 streams) vs None)
+    | _ -> system_eval_complete e streams vs hBS
+\end{code}
+
+Tactics: \cite{martinez2019meta}.
+
+Invariant generation: ICE\cite{garg2014ice}; template-based\cite{kahsai2011instantiation}; PDR explanation\cite{een2011efficient}; IC3 first introduced\cite{bradley2011sat}
+
+\section{Execution}
+
+We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support\cite{protzenko2017verified}.
+% This transform generates \fstar{} functions in the
+
+This allows us to generate C code which executes in bounded space on embedded devices.
+
+Other compilers for Lustre-style languages, such as Vélus\cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
+With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
+For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
+(Is this the same as Lucid Synchrone and Zélus? \CITE)
+
+\section{Related work}
+
+hardware language in Agda \cite{harrison2021mechanized}
+
+Refinement types for Zélus \cite{chen2022synchronous}
+
+\section{Conclusion and future work}
+
+Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
 

### 2023-05-03 2023pipit/Main.tex
remove prim ty

@@
     \> \> @then@ $\max v_F v_R$ \\
-    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@
+    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@ \\
     \\
@@
 The typing rules of Pipit are defined in \autoref{f:core-typing}, with the judgment form $\typing{\Gamma}{e}{\tau}$ that checks expression $e$ under streaming context $\Gamma$.
-The grammar of types (\autoref{f:core-grammar}) distinguishes between bounded-size primitive types ($\primty$) and potentially higher-order types ($\tau$), which allows us to restrict streaming operations to only require bounded buffers.
-\TODO{Not sure if this restriction is useful, as arbitrary functions can still use however much memory and time as they want.}
+% The grammar of types (\autoref{f:core-grammar}) distinguishes between bounded-size primitive types ($\primty$) and potentially higher-order types ($\tau$), which allows us to restrict streaming operations to only require bounded buffers.
+% \TODO{Not sure if this restriction is useful, as arbitrary functions can still use however much memory and time as they want.}
 

### 2023-05-03 2023pipit/Main.tex
remove prim ty
intro notes

@@
 We are interested in verifying an antilock braking system for a motorcycle\cite{huang2010design}.
+Antilock brakes are designed to stop wheels from locking up in emergency braking events.
+Antilock braking controllers generally work by computing the \emph{wheel slip} for the wheels, which is a ratio of the wheel's rotational speed compared to the vehicle's speed.
+To compute the wheel slip, then, the controller must first estimate the vehicle's speed.
+
+\begin{itemize}
+  \item example: abs / vehicle speed estimate
+  \item $\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$
+\end{itemize}
 

### 2023-05-09 2023pipit/Main.tex
ok

@@
 % \usepackage{minipage-marginpar}
+
+% \usepackage{cite}
 
@@
 
-We are interested in verifying an antilock braking system for a motorcycle\cite{huang2010design}.
+\TODO{Required for submission: writing}
+\begin{itemize}
+  \item introduction
+  \item running example: vehicle speed estimate with ``error tolerance limit'' property ($\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$)?
+    + uses modal operators
+    + property is k-inductive for small $t$, requires explicit invariant for larger $t$, motivates k-induction
+    - not super interesting
+    - proof requires CSE, which isn't implemented yet
+  \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
+\end{itemize}
+\TODO{Required for submission: implementation}
+\begin{itemize}
+  \item finish proof of system correctness
+  \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
+  \item implement unverified CSE pass
+  \item verify running example, with CSE and k-induction
+\end{itemize}
+
+We are interested in verifying an antilock braking system for a motorcycle \cite{huang2010design}.
 Antilock brakes are designed to stop wheels from locking up in emergency braking events.
@@
 To compute the wheel slip, then, the controller must first estimate the vehicle's speed.
+When the wheels are travelling freely with no braking pressure and no wheel slip, one can accurately estimate the vehicle's speed by multiplying the wheel rotational speed by the wheel radius.
+Unfortunately, once the wheels begin to slip, it becomes harder to accurately estimate the wheel speed.
+For this reason, brake controllers often use a combination of sensors.
+Some brake controllers \cite{kobayashi1995estimation} incorporate both wheel speed sensors and accelerometer sensors; the readings are combined with a Kalman filter.
 
-\begin{itemize}
-  \item example: abs / vehicle speed estimate
-  \item $\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$
-\end{itemize}
-
+\begin{figure}
 \begin{tabbing}
@@
   @let@ veh\_speed\_estimate $\omega_F$ $\omega_R$ $a_z$ = \\
-    \> @let@ $v_F$ = $\omega_F \cdot $ circumference @in@ \\
-    \> @let@ $v_R$ = $\omega_R \cdot $ circumference @in@ \\
+    \> @let@ $v_F$ = $\omega_F \cdot $ radius @in@ \\
+    \> @let@ $v_R$ = $\omega_R \cdot $ radius @in@ \\
     \\
@@
 
-Cite: Kind2\cite{champion2016kind2}, Lustre\cite{caspi1995functional}, Vélus\cite{bourke2017formally}.
-The original Kind\cite{hagen2008scaling}.
+\caption{Implementation of \tt{veh\_speed\_estimate}.}\label{f:veh-speed-estimate}
+\end{figure}
+
+\begin{itemize}
+  \item example: abs / vehicle speed estimate
+  \item $\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$
+\end{itemize}
+
+Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
+The original Kind \cite{hagen2008scaling}.
+
+\TODO{\fstar{} $\to \mbox{F}^{\star}$}
+\CITEME{2022 thesis on Agda semantics of Lustre?}
 
@@
 The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
-In the reactive semantics of \cite{caspi1995functional}, as well as the coinductive semantics of Vélus \CITE{bourke2017formally find best paper}, the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
+In the reactive semantics of \citet{caspi1995functional}, as well as the coinductive semantics of Vélus \cite{bourke2017formally} (\CITEME{find best paper}), the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
 This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.
@@
 
-Invariant generation: ICE\cite{garg2014ice}; template-based\cite{kahsai2011instantiation}; PDR explanation\cite{een2011efficient}; IC3 first introduced\cite{bradley2011sat}
+Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
 
@@
 
-We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support\cite{protzenko2017verified}.
+We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support \cite{protzenko2017verified}.
 % This transform generates \fstar{} functions in the
@@
 
-Other compilers for Lustre-style languages, such as Vélus\cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
+Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
 With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.

### 2023-05-10 2023pipit/Main.tex
intro page 1

@@
 \begin{abstract}
-  Safety-critical control systems, such as the anti-lock braking systems in most cars today, need to be correct and execute in real-time.
-  We could write these systems directly in an imperative language such as C or Rust and verify them, but it can be difficult to state high-level specifications about such low-level implementations.
-  Another approach, favoured by large parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre or SCADE, and verify that the implementations satisfy the high-level specification using a model-checker.
-  These model-checkers can prove many interesting properties automatically, but do not provide many options when the automated proof techniques fail.
-  We introduce Pipit, an embedded domain-specific language for implementing and verifying such controllers in \fstar.
-  Pipit aims to provide a high-level language similar to Lustre, while reusing \fstar's proof automation and manual proofs for verifying controllers, and using LowStar's C-code generation for real-time execution.
-  Pipit translates its expression language to a transition system for k-inductive proofs, which is mostly verified; verifying the translation to imperative code is future work.
+  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
+  We introduce Pipit, a simple reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
+  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved using k-induction on the transition system.
+  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
+  This translation to imperative code has been proved to preserve types; the proof that the generated imperative code preserves semantics is future work.
 \end{abstract}
@@
 \section{Introduction}
+
+Safety-critical control systems, such as the anti-lock braking systems that occur in most cars today, need to be correct and execute in real-time.
+One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
+These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
+Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
+This mismatch may occur even when the compiler has been verified correct, as in the case of Vélus~\cite{bourke2017formally}.
+For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.
+
+To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
+In the following, we introduce Pipit, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
+Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
+% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
+Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is future work.
+
+Our contributions are as follows:
+\begin{itemize}
+  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
+  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
+    % TODO: CUT WAFFLE
+    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
+    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
+  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
+\end{itemize}
+
+% \begin{itemize}
+%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
+%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
+%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
+%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
+% \end{itemize}
+
+\section{Pipit in flight}
+
+\begin{tabbing}
+  @MM@\= @MMMMMM@ \= \kill
+  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
+    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
+    \> @in @ $p'$
+\end{tabbing}
+
+  \begin{tabbing}
+    @MM@\= @MMMMMM@ \= \kill
+    @let@ fir (coefficients: list $\RRgez$) (s: stream $\RR$): stream $\RR$ = \\
+      \> @match@ coefficients @with@ \\
+      \> @|@ $[]$ \> $\to 0$ \\
+      \> @|@ $c :: \cs $ \> $\to (s \cdot c) + \mbox{fir~} \cs ~(\xfby{0}{s})$ \\
+  \end{tabbing}
+
+$\blacksquare(s = 0) \implies \mbox{fir~} \cs~s = 0 \blacksquare \Box \blacklozenge \Diamond$
 
@@
 
+\section{Coffee machine}
+\label{s:coffee-machine}
+
 \section{Related work}
@@
 Refinement types for Zélus \cite{chen2022synchronous}
+
+\TODO{discuss TCB: big, much bigger than Vélus}
 

### 2023-05-10 2023pipit/Main.tex
shelve

@@
 
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
-    \> @in @ $p'$
-\end{tabbing}
+One common application in controllers is to implement filters, such as a \emph{finite impulse response} (FIR) filter.
+An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
+For discrete signals, it is equivalent to a weighted moving average.
 
@@
   @MM@\= @MMMMMM@ \= \kill
-    @let@ fir (coefficients: list $\RRgez$) (s: stream $\RR$): stream $\RR$ = \\
+  @let@ fir (coefficients: list $\RR$) (s: stream $\RR$): stream $\RR$ = \\
     \> @match@ coefficients @with@ \\
@@
   \item introduction
-  \item running example: vehicle speed estimate with ``error tolerance limit'' property ($\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$)?
-    + uses modal operators
-    + property is k-inductive for small $t$, requires explicit invariant for larger $t$, motivates k-induction
-    - not super interesting
-    - proof requires CSE, which isn't implemented yet
+  \item running example: FIR?
   \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
@@
 
-We are interested in verifying an antilock braking system for a motorcycle \cite{huang2010design}.
-Antilock brakes are designed to stop wheels from locking up in emergency braking events.
-Antilock braking controllers generally work by computing the \emph{wheel slip} for the wheels, which is a ratio of the wheel's rotational speed compared to the vehicle's speed.
-To compute the wheel slip, then, the controller must first estimate the vehicle's speed.
-When the wheels are travelling freely with no braking pressure and no wheel slip, one can accurately estimate the vehicle's speed by multiplying the wheel rotational speed by the wheel radius.
-Unfortunately, once the wheels begin to slip, it becomes harder to accurately estimate the wheel speed.
-For this reason, brake controllers often use a combination of sensors.
-Some brake controllers \cite{kobayashi1995estimation} incorporate both wheel speed sensors and accelerometer sensors; the readings are combined with a Kalman filter.
-
-\begin{figure}
-\begin{tabbing}
-  @MM@\= @let @\= \kill
-  @let@ veh\_speed\_estimate $\omega_F$ $\omega_R$ $a_z$ = \\
-    \> @let@ $v_F$ = $\omega_F \cdot $ radius @in@ \\
-    \> @let@ $v_R$ = $\omega_R \cdot $ radius @in@ \\
-    \\
-    \> @let rec@ $\floor{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
-    \> \> @then@ $\min v_F v_R$ \\
-    \> \> @else@ $(\xthen{\min v_F v_R}{\xfby{0}{\floor{\hat{v}}}}) + a_z - \epsilon$ @in@ \\
-    \> @let rec@ $\ceil{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
-    \> \> @then@ $\max v_F v_R$ \\
-    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@ \\
-    \\
-    \> $(\floor{\hat{v}}, \ceil{\hat{v}})$
-\end{tabbing}
-
-\caption{Implementation of \tt{veh\_speed\_estimate}.}\label{f:veh-speed-estimate}
-\end{figure}
-
-\begin{itemize}
-  \item example: abs / vehicle speed estimate
-  \item $\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$
-\end{itemize}
-
 Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
 The original Kind \cite{hagen2008scaling}.
-
-\TODO{\fstar{} $\to \mbox{F}^{\star}$}
 \CITEME{2022 thesis on Agda semantics of Lustre?}
@@
 
-\pagebreak
+% \pagebreak
 % \begin{acks}

### 2023-05-10 2023pipit/shelve.tex
shelve

%%%%%% BRAKE STUFF

We are interested in verifying an antilock braking system for a motorcycle \cite{huang2010design}.
Antilock brakes are designed to stop wheels from locking up in emergency braking events.
Antilock braking controllers generally work by computing the \emph{wheel slip} for the wheels, which is a ratio of the wheel's rotational speed compared to the vehicle's speed.
To compute the wheel slip, then, the controller must first estimate the vehicle's speed.
When the wheels are travelling freely with no braking pressure and no wheel slip, one can accurately estimate the vehicle's speed by multiplying the wheel rotational speed by the wheel radius.
Unfortunately, once the wheels begin to slip, it becomes harder to accurately estimate the wheel speed.
For this reason, brake controllers often use a combination of sensors.
Some brake controllers \cite{kobayashi1995estimation} incorporate both wheel speed sensors and accelerometer sensors; the readings are combined with a Kalman filter.

\begin{figure}
\begin{tabbing}
  @MM@\= @let @\= \kill
  @let@ veh\_speed\_estimate $\omega_F$ $\omega_R$ $a_z$ = \\
    \> @let@ $v_F$ = $\omega_F \cdot $ radius @in@ \\
    \> @let@ $v_R$ = $\omega_R \cdot $ radius @in@ \\
    \\
    \> @let rec@ $\floor{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
    \> \> @then@ $\min v_F v_R$ \\
    \> \> @else@ $(\xthen{\min v_F v_R}{\xfby{0}{\floor{\hat{v}}}}) + a_z - \epsilon$ @in@ \\
    \> @let rec@ $\ceil{\hat{v}}$ = @if@ $v_F \approx_\epsilon v_R$ \\
    \> \> @then@ $\max v_F v_R$ \\
    \> \> @else@ $(\xthen{\max v_F v_R}{\xfby{0}{\ceil{\hat{v}}}}) + a_z + \epsilon$ @in@ \\
    \\
    \> $(\floor{\hat{v}}, \ceil{\hat{v}})$
\end{tabbing}

\caption{Implementation of \tt{veh\_speed\_estimate}.}\label{f:veh-speed-estimate}
\end{figure}

property ($\blacklozenge_{t} (v_F \approx_{\epsilon} v_R) \implies \lfloor \hat{v}' \rfloor \approx_{t\epsilon} \lceil \hat{v}' \rceil$)

%%%%% MODAL

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
    \> @in @ $p'$
\end{tabbing}

### 2023-05-10 2023pipit/Main.tex
pipit

@@
 
-Our contributions are as follows:
+In this extended abstract, we will briefly describe the following preliminary results, which we intend to publish pending further evaluation:
 \begin{itemize}
@@
 
-One common application in controllers is to implement filters, such as a \emph{finite impulse response} (FIR) filter.
+A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
 An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
-For discrete signals, it is equivalent to a weighted moving average.
+We can implement an FIR filter in Pipit as follows:
 
@@
   @MM@\= @MMMMMM@ \= \kill
-  @let@ fir (coefficients: list $\RR$) (s: stream $\RR$): stream $\RR$ = \\
-    \> @match@ coefficients @with@ \\
+  @let@ fir ($\textit{coefficients}$: list $\RR$) ($s$: stream $\RR$): stream $\RR$ = \\
+    \> @match@ $\textit{coefficients}$ @with@ \\
     \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + \mbox{fir~} \cs ~(\xfby{0}{s})$ \\
+    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~s)})$
 \end{tabbing}
+
+Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
+The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream.
+The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
+If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($s \cdot c$) and delay the result ($\xfby{0}{\ldots}$) of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ s$), summing the two parts.
+When there is no previous value, as is the case at the start of execution, the delay is given the initial value of zero.
+
+Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as different tools support different features.
+To write this program in Lustre, we would likely need to reformulate the program to use arrays: Vélus does not support arrays, so verified compilation is out; Kind2 uses a custom syntax for arrays that no compiler supports, so model-checking with Kind2 is out; and Lustre v6 does support arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous}, only supports reasoning with boolean abstractions.
+
+\newcommand\ii{\textit{input}}
+
+Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
+By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program, which can be easily executed in statically-bounded runtime and space.
+For example, the weighted moving average $\mbox{fir}~[0.5; 0.3; 0.2] \ii$ becomes $(0.5 \cdot \ii) + \xfby{0}{((0.3 \cdot \ii) + \xfby{0}{((0.2 \cdot \ii) + \xfby{0}{0})})}$.
 
@@
 
-\TODO{Required for submission: writing}
-\begin{itemize}
-  \item introduction
-  \item running example: FIR?
-  \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
-\end{itemize}
-\TODO{Required for submission: implementation}
-\begin{itemize}
-  \item finish proof of system correctness
-  \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
-  \item implement unverified CSE pass
-  \item verify running example, with CSE and k-induction
-\end{itemize}
+% \TODO{Required for submission: writing}
+% \begin{itemize}
+%   \item introduction
+%   \item running example: FIR?
+%   \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
+% \end{itemize}
+% \TODO{Required for submission: implementation}
+% \begin{itemize}
+%   \item finish proof of system correctness
+%   \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
+%   \item implement unverified CSE pass
+%   \item verify running example, with CSE and k-induction
+% \end{itemize}
 
-Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
-The original Kind \cite{hagen2008scaling}.
+% Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
+Cite: the original Kind \cite{hagen2008scaling}.
 \CITEME{2022 thesis on Agda semantics of Lustre?}
-
 Past-time modal temporal logic \cite{lichtenstein1985glory}

### 2023-05-10 2023pipit/Main.tex
pipit

@@
 % \usepackage{mathptmx}
-% \usepackage{mathtools}
+\usepackage{mathtools}
 % \usepackage{stmaryrd}
@@
 Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream.
+The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream\footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}.
 The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
@@
 Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as different tools support different features.
-To write this program in Lustre, we would likely need to reformulate the program to use arrays: Vélus does not support arrays, so verified compilation is out; Kind2 uses a custom syntax for arrays that no compiler supports, so model-checking with Kind2 is out; and Lustre v6 does support arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous}, only supports reasoning with boolean abstractions.
+To write this program in Lustre, we would likely need to reformulate the program to use arrays.
+Vélus does not support arrays, so verified compilation is out; Kind2 uses a custom syntax for arrays that no compiler supports, so model-checking with Kind2 is out; and Lustre v6 does support compiling arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous}, only supports reasoning with boolean abstractions.
+
+Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
+By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
+For example, we can compute the weighted moving average with two coefficients:
 
 \newcommand\ii{\textit{input}}
+\newcommand\oo{\textit{output}}
+\newcommand\ok{\textit{ok}}
+\begin{tabbing}
+  @MM@\= @MMMMMM@ \= \kill
+  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
+    \> $\mbox{fir}~[0.7; 0.3] \ii$
+\end{tabbing}
 
-Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
-By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program, which can be easily executed in statically-bounded runtime and space.
-For example, the weighted moving average $\mbox{fir}~[0.5; 0.3; 0.2] \ii$ becomes $(0.5 \cdot \ii) + \xfby{0}{((0.3 \cdot \ii) + \xfby{0}{((0.2 \cdot \ii) + \xfby{0}{0})})}$.
+Which then normalises to the following simple definition:
 
-$\blacksquare(s = 0) \implies \mbox{fir~} \cs~s = 0 \blacksquare \Box \blacklozenge \Diamond$
+\begin{tabbing}
+    @MM@\= @MMMMMM@ \= \kill
+      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
+    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
+\end{tabbing}
+
+We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have been within some particular range for long enough, then the outputs are also within the range.
+This property can occur as an internal invariant in a larger proof.
+
+\begin{tabbing}
+  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
+  @let@ $\mbox{bibo}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
+  \> @let@ $\oo$ \> $= \mbox{fir}_2~\ii~$ \> @in@ \\
+  \> @let@ $\ok$ \> $= |\ii| \le 1 $ \> @in@ \\
+  \> @check@ $((\ok \wedge (\xfby{\top}{\ok})) \implies |\oo| \le 1)$ \\
+\end{tabbing}
+
+This property states that if the current input and the previous input are both in the inclusive range $[-1, 1]$, then the output is also within the range $[-1, 1]$.
+To prove that this property holds, we translate to a transition system.
+In this case, simple induction over the transition relation is sufficient to prove the property.
+There is some boilerplate required to perform the induction, but the step case and base case are automatically proved by \fstar{}.
+
+\begin{verbatim}
+// CLEAN
+let fir_prop_prove (fv: sem_freevars): Lemma (ensures induct1 (bibo_2 fv)) =
+  assert (base_case (bibo_2 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
+  assert (step_case (bibo_2 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
+\end{verbatim}
+
+However, if we wish to prove a similar bounded-input-bounded-output property for three or more coefficients, standard induction over the transition relation is not sufficient.
+The issue is that, in the transition system, each delay, such as $\xfby{\top}{|\ii| \le 1}$ or $\xfby{0}{\ii}$ is given a fresh state variable.
+TODO motivate need for sofar to show rec construct:
+
+\begin{tabbing}
+  @MM@\= @MMMMMM@ \= \kill
+  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
+    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
+    \> @in @ $p'$
+\end{tabbing}
+
+XXX show k-inductive proof of $bibo_3$?
+\begin{verbatim}
+// CLEAN
+let fir_prop_prove (fv: sem_freevars): Lemma (ensures induct2 (bibo3 fv)) =
+  assert (base_case\_2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
+  assert (step_case\_2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
+\end{verbatim}
 

### 2023-05-10 2023pipit/Main.tex
ok

@@
 \begin{verbatim}
-// CLEAN
 let fir_prop_prove (fv: sem_freevars): Lemma (ensures induct2 (bibo3 fv)) =
-  assert (base_case\_2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
-  assert (step_case\_2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
+  assert (base_case2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
+  assert (step_case2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
 \end{verbatim}

### 2023-05-11 2023pipit/Main.tex
example sec

@@
 
-Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as different tools support different features.
+Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
 To write this program in Lustre, we would likely need to reformulate the program to use arrays.
-Vélus does not support arrays, so verified compilation is out; Kind2 uses a custom syntax for arrays that no compiler supports, so model-checking with Kind2 is out; and Lustre v6 does support compiling arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous}, only supports reasoning with boolean abstractions.
+Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support compiling arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous} does not.
 
@@
 
+\newcommand\bibo{\mbox{bibo}}
+\newcommand\fir{\mbox{fir}}
 \newcommand\ii{\textit{input}}
@@
 
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have been within some particular range for long enough, then the outputs are also within the range.
-This property can occur as an internal invariant in a larger proof.
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-  \> @let@ $\oo$ \> $= \mbox{fir}_2~\ii~$ \> @in@ \\
-  \> @let@ $\ok$ \> $= |\ii| \le 1 $ \> @in@ \\
-  \> @check@ $((\ok \wedge (\xfby{\top}{\ok})) \implies |\oo| \le 1)$ \\
-\end{tabbing}
-
-This property states that if the current input and the previous input are both in the inclusive range $[-1, 1]$, then the output is also within the range $[-1, 1]$.
-To prove that this property holds, we translate to a transition system.
-In this case, simple induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but the step case and base case are automatically proved by \fstar{}.
-
-\begin{verbatim}
-// CLEAN
-let fir_prop_prove (fv: sem_freevars): Lemma (ensures induct1 (bibo_2 fv)) =
-  assert (base_case (bibo_2 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
-  assert (step_case (bibo_2 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
-\end{verbatim}
-
-However, if we wish to prove a similar bounded-input-bounded-output property for three or more coefficients, standard induction over the transition relation is not sufficient.
-The issue is that, in the transition system, each delay, such as $\xfby{\top}{|\ii| \le 1}$ or $\xfby{0}{\ii}$ is given a fresh state variable.
-TODO motivate need for sofar to show rec construct:
+To state properties about our filter, it is useful to be able to state temporal properties.
+We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
+We name the past-globally operator \emph{sofar}, as in \emph{the property has been true so far}:
 
@@
   @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> @let rec@ $p'$ $= (\xfby{\top}{p'}) \wedge p$ \\
-    \> @in @ $p'$
+    \> $\xrec{p'}{p \wedge (\xfby{\top}{p'})}$
 \end{tabbing}
 
-XXX show k-inductive proof of $bibo_3$?
-\begin{verbatim}
-let fir_prop_prove (fv: sem_freevars): Lemma (ensures induct2 (bibo3 fv)) =
-  assert (base_case2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ());
-  assert (step_case2 (bibo3 fv)) by (tac_nbe (); tac_intros_smash (); tac_nbe ())
-\end{verbatim}
+This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
+At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{\top}{p'}$).
+If there is no previous value, it defaults to true.
+
+We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
+Such properties can occur as an internal invariant in a larger proof.
+
+\begin{tabbing}
+  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
+  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\RR$ = \\
+  \> @check@ $(\mbox{sofar}(|\ii| \le n)) \implies |\mbox{fir}_2~\ii| \le n)$
+\end{tabbing}
+
+This property states that, for this particular set of coefficients, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
+Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
+To prove that this property holds, we translate to a transition system.
+In this case, simple induction over the transition relation is sufficient to prove the property.
+There is some boilerplate required to perform the induction, but the step case and base case are automatically proved by \fstar{}:
+
+\begin{tabbing}
+  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
+  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
+  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
+  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
+\end{tabbing}
+
+This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
+The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
+
+If we wish to prove a similar BIBO property for three or more coefficients, this standard induction over the transition system is no longer sufficient, and we need a stronger invariant.
+One simple automated way to strengthen the invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which looks at multiple steps of the transition relation at a time.
+For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:
+
+\begin{tabbing}
+  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
+  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
+  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
+  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
+\end{tabbing}
+
+% The issue is that each delayed values inside the FIR filter becomes a separate state variable in the transition system.
+% When looking at the result of a single step of the transition relation in isolation, we can prove that $(\xfby{0}{\ii}) \le n$ is equivalent to $\xfby{\top}{(\ii \le n)}$.
+% When nesting delays, however, the same sort of distributive property is not clear from just a single step of the transition relation; each delay gets its own state variable, and the relationship between the different state variables is lost.
+% We really need a stronger invariant to do this.
 
@@
 % Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
-Cite: the original Kind \cite{hagen2008scaling}.
-\CITEME{2022 thesis on Agda semantics of Lustre?}
-Past-time modal temporal logic \cite{lichtenstein1985glory}
-Lustre support for temporal logic \cite{halbwachs1993executable}
+% Cite: the original Kind \cite{hagen2008scaling}.
+% \CITEME{2022 thesis on Agda semantics of Lustre?}
+% Past-time modal temporal logic \cite{lichtenstein1985glory}
+% Lustre support for temporal logic \cite{halbwachs1993executable}
 

### 2023-05-11 2023pipit/Main.tex
shelving

@@
 \section{Pipit in flight}
+\label{s:tut}
 
@@
 \input{figures/core-grammar.tex}
-\input{figures/core-typing.tex}
+% \input{figures/core-typing.tex}
 \input{figures/core-bigstep.tex}
@@
 The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
-Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \cite{caspi1995functional}.
+Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
 The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
 Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-To ensure that streams are productive, recursive equations can only refer to their previous values and must be \emph{guarded} by a delay; this restriction is formalised by the causality check in \REF{}.
-This form of recursion differs somewhat from standard Lustre, which uses a set of mutually recursive bindings.
-In \autoref{ss:operational-semantics} we will see that this form of recursion allows a syntax-directed semantics, but at the cost of potentially duplicating work.
-
-Let-expressions ($\xlet{x}{e}{e'}$) are standard.
-Finally, check-expressions ($\xcheck{e}{e'}$) are used to describe properties that the program should satisfy.
-This operation corresponds roughly to the @assert@ primitive in \fstar{}; however we avoid this name, as @assert@ is historically treated as an unchecked assumption in many Lustre languages.
-
-The grammar for values $v$ includes natural numbers, integers and functions embedded from the \fstar{} meta-language.
-We denote functions in the meta-language by the explicit meta-lambda notation ($\mlamX{}$) to avoid any potential ambiguity between Pipit expressions ($e$) and meta-expressions ($\metafy{e}$).
-
-\subsection{Typing rules}
-\label{ss:typing-rules}
-
-The typing rules of Pipit are defined in \autoref{f:core-typing}, with the judgment form $\typing{\Gamma}{e}{\tau}$ that checks expression $e$ under streaming context $\Gamma$.
-% The grammar of types (\autoref{f:core-grammar}) distinguishes between bounded-size primitive types ($\primty$) and potentially higher-order types ($\tau$), which allows us to restrict streaming operations to only require bounded buffers.
-% \TODO{Not sure if this restriction is useful, as arbitrary functions can still use however much memory and time as they want.}
-
-For values (rule TValue), we reuse the \fstar{} meta-language's typing judgment ($\mtyping{}{v}{\tau}$) for typing closed values $v$.
-Otherwise, most of the rules are standard.
 

### 2023-05-11 2023pipit/figures/core-grammar.tex
remove prim ty
shelving

@@
   \[
-  \begin{array}{lrl}
+  \begin{array}{lrlll}
     e, e' & := & v ~|~ x ~|~ e~e' \\
-          & | & \xfby{v}{e} \\
-          & | & \xthen{e}{e'} \\
-          & | & \xrec{x}{e[x]} \\
-          & | & \xlet{x}{e}{e'[x]} \\
-          & | & \xcheck{e}{e'} \\
+          & | & \xfby{v}{e} & | & \xthen{e}{e'} \\
+          & | & \xrec{x}{e[x]} & | & \xlet{x}{e}{e'[x]} \\
+          & | & \xcheck{e} \\
     \\
-    v & := & n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ \mlamX
+    v & := & \multicolumn{3}{l}{n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots ~|~ \mlamX}
     \\
@@
     \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
-      \\
-    \primty & := & \mathbb{N} ~|~ \mathbb{B} \\
-    \tau, \tau' & := & \primty ~|~ \hdots ~|~ \tau \to \tau' \\
-    \Gamma & := & \cdot ~|~ x : \tau, \Gamma \\
+    %   \\
+    % \tau, \tau' & := & \multicolumn{3}{l}{\mathbb{N} ~|~ \mathbb{B} ~|~ \hdots ~|~ \tau \to \tau'} \\
+    % \Gamma & := & \cdot ~|~ x : \tau, \Gamma \\
     \end{array}
   \]
-  \TODO{compact grammar}
-  \caption{Pipit language grammar, including expressions ($e$); values and host functions embedded from the \fstar{} meta-language ($v$); stores ($\sigma$); non-empty stream histories ($\Sigma$); bounded-size types ($\primty$); unbounded types ($\tau$); and typing contexts ($\Gamma$).}
+  \caption{Pipit language grammar, including expressions ($e$); values and host functions embedded from the \fstar{} meta-language ($v$); stores ($\sigma$); and non-empty stream histories ($\Sigma$).}
   \label{f:core-grammar}

### 2023-05-11 2023pipit/shelve.tex
shelving

@@
 \end{tabbing}
+
+%%% GRAMMAR
+
+The grammar of Pipit is defined in \autoref{f:core-grammar}.
+The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
+Most of the expression forms are equivalent to a simplified, clock-free primitives from standard Lustre-style languages \cite{caspi1995functional}.
+The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
+Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.
+
+Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
+To ensure that streams are productive, recursive equations can only refer to their previous values and must be \emph{guarded} by a delay; this restriction is formalised by the causality check in \REF{}.
+This form of recursion differs somewhat from standard Lustre, which uses a set of mutually recursive bindings.
+In \autoref{ss:operational-semantics} we will see that this form of recursion allows a syntax-directed semantics, but at the cost of potentially duplicating work.
+
+Let-expressions ($\xlet{x}{e}{e'}$) are standard.
+Finally, check-expressions ($\xcheck{e}$) are used to describe properties that the program should satisfy.
+This operation corresponds roughly to the @assert@ primitive in \fstar{}; however we avoid this name, as @assert@ is historically treated as an unchecked assumption in many Lustre languages.
+
+The grammar for values $v$ includes natural numbers, integers and functions embedded from the \fstar{} meta-language.
+We denote functions in the meta-language by the explicit meta-lambda notation ($\mlamX{}$) to avoid any potential ambiguity between Pipit expressions ($e$) and meta-expressions ($\metafy{e}$).
+
+%%% OPERATIONAL SEMANTICS
+The operational semantics, expressed as a bigstep relation, are defined in \autoref{f:core-bigstep}.
+The judgment form $\bigstep{\Sigma}{e}{v}$ takes a \emph{stream history} ($\Sigma$), which is a non-empty sequence of stores, as well as an expression to evaluate ($e$) and the final value ($v$).
+
+Rule Value denotes that a value evaluates to itself under any streaming history.
+Rule Var denotes that evaluating a variable requires looking it up in the right-most (most recent) store; the syntax $\Sigma_\bot; \sigma$ means that this rule applies to any streaming history with one or more steps.
+Rule App-Meta allows applying functions on the meta-level; the syntax $(\mlamX{}) v'$ refers to meta-level application of some meta-term.
+
+The rules $\mbox{Fby}_1$ and $\mbox{Fby}_S$ describe the semantics of delayed streams.
+Rule $\mbox{Fby}_1$ denotes that when a delay $\xfby{v}{e}$ is evaluated in a context with exactly one step --- that is, at the very start of program execution --- it simply returns the initial value $v$.
+Rule $\mbox{Fby}_S$ applies when the same delay is evaluated in a context with strictly more than one step ($\Sigma; \sigma$), in which case, the most recent step is discarded, and the original expression $e$ is evaluated under the previous history.
+
+Next, rules ($\xthenarrow_1$) and ($\xthenarrow_S$) describe the semantics of streams composed by \emph{then}-expressions; as with the rules for delayed streams, the first rule applies when the streaming history contains exactly one step, while the second applies for histories of strictly more than one step.
+
+Rule ($\mu{}$) describes the semantics of recursive streams.
+This rule performs a single unfolding of a recursive expression $\xrec{x}{e[x]}$ so that all occurrences of $x$ inside $e$ are replaced with the recursive expression itself.
+This rule is not trivially terminating, but so long as any occurrences of the recursive binder are guarded by a delay, then all occurrences of the recursive binder will have a shorter stream history.
+This intuition is formalised in \REF{}.
+
+Rule Let uses a standard substitution-based semantics for let-expressions.
+Finally, rule Check evaluates a property and requires it to be true ($\top$).
+
+The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
+In the reactive semantics of \citet{caspi1995functional}, as well as the coinductive semantics of Vélus \cite{bourke2017formally} (\CITEME{find best paper}), the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
+This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.
+
+In contrast, the semantics presented here is syntax-directed and completely deterministic.
+We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \autoref{ss:meta-theory}.
+However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
+For this reason, the semantics here is not suitable for actual execution.
+
+%%% TYPING RULES
+
+\subsection{Typing rules}
+\label{ss:typing-rules}
+
+The typing rules of Pipit are defined in \autoref{f:core-typing}, with the judgment form $\typing{\Gamma}{e}{\tau}$ that checks expression $e$ under streaming context $\Gamma$.
+% The grammar of types (\autoref{f:core-grammar}) distinguishes between bounded-size primitive types ($\primty$) and potentially higher-order types ($\tau$), which allows us to restrict streaming operations to only require bounded buffers.
+% \TODO{Not sure if this restriction is useful, as arbitrary functions can still use however much memory and time as they want.}
+
+For values (rule TValue), we reuse the \fstar{} meta-language's typing judgment ($\mtyping{}{v}{\tau}$) for typing closed values $v$.
+Otherwise, most of the rules are standard.

### 2023-05-11 2023pipit/Main.tex
almost

@@
 
-In this extended abstract, we will briefly describe the following preliminary results, which we intend to publish pending further evaluation:
+In this extended abstract, we will briefly describe the following preliminary results:
+
 \begin{itemize}
-  \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-  \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-    % TODO: CUT WAFFLE
-    This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-    Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-  \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
+  \item embedded language reduces the need for fancy features (\autoref{s:tut});
+  \item normalisation and proof automation allow us to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
+  \item normalisation and extraction allows us to generate unverified executable code (\autoref{s:extraction});
 \end{itemize}
+
+We will also touch upon the following work-in-progress, which we intend to publish upon completion:
+
+\begin{itemize}
+  \item we describe the syntax (\autoref{s:core});
+  \item a unified semantics for theorem proving and computation;
+  \item verified imperative code generation;
+  \item the ability to mix manual proofs with automated proofs
+  \item further evaluation with the verification of a realistic controller
+\end{itemize}
+
+%CUT
+% \begin{itemize}
+%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
+%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
+%     % TODO: CUT WAFFLE
+%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
+%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
+%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
+% \end{itemize}
 
@@
 
-\section{Pipit in flight}
+% \section{Pipit in flight}
+\pagebreak
+\section{Programming and verifying in Pipit}
 \label{s:tut}
@@
 To write this program in Lustre, we would likely need to reformulate the program to use arrays.
-Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support compiling arrays and recursive nodes, but its model-checker, Lesar~\cite{raymond2008synchronous} does not.
+Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support arrays, but its model-checker Lesar~\cite{raymond2008synchronous} does not.
 
-Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}).
+Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}), inspired by Bedrock~\cite{chlipala2013bedrock}.
 By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
@@
 We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the property has been true so far}:
+We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
 
@@
 
+\subsection{Bounded input, bounded output}
+\label{ss:tut:bibo}
 We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
@@
 In this case, simple induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but the step case and base case are automatically proved by \fstar{}:
+There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
 
@@
 
-If we wish to prove a similar BIBO property for three or more coefficients, this standard induction over the transition system is no longer sufficient, and we need a stronger invariant.
-One simple automated way to strengthen the invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which looks at multiple steps of the transition relation at a time.
+If we wish to prove a similar BIBO property for more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
+One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
 For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:
@@
 
-% The issue is that each delayed values inside the FIR filter becomes a separate state variable in the transition system.
-% When looking at the result of a single step of the transition relation in isolation, we can prove that $(\xfby{0}{\ii}) \le n$ is equivalent to $\xfby{\top}{(\ii \le n)}$.
-% When nesting delays, however, the same sort of distributive property is not clear from just a single step of the transition relation; each delay gets its own state variable, and the relationship between the different state variables is lost.
-% We really need a stronger invariant to do this.
+Although the properties we have shown here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.
 
-% \TODO{Required for submission: writing}
-% \begin{itemize}
-%   \item introduction
-%   \item running example: FIR?
-%   \item describe translation to transition system? detail probably won't fit, translation invariant is the most interesting part.
-% \end{itemize}
-% \TODO{Required for submission: implementation}
-% \begin{itemize}
-%   \item finish proof of system correctness
-%   \item state end-to-end / capstone theorem that k-inductively verified systems apply to original semantics
-%   \item implement unverified CSE pass
-%   \item verify running example, with CSE and k-induction
-% \end{itemize}
-
-% Cite: Kind2 \cite{champion2016kind2}, Lustre \cite{caspi1995functional}, Vélus \cite{bourke2017formally}.
-% Cite: the original Kind \cite{hagen2008scaling}.
-% \CITEME{2022 thesis on Agda semantics of Lustre?}
-% Past-time modal temporal logic \cite{lichtenstein1985glory}
-% Lustre support for temporal logic \cite{halbwachs1993executable}
-
+\pagebreak
 \section{Core language}
@@
 % \input{figures/core-typing.tex}
-\input{figures/core-bigstep.tex}
+% \input{figures/core-bigstep.tex}
 
@@
 Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the very first step, followed by the latest values from stream $e'$ for subsequent steps.
+% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
+Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the latest values from stream $e'$ for subsequent steps.
 
-\subsection{Operational semantics}
-\label{ss:operational-semantics}
+Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
+% To ensure that streams are productive,
+As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
+This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
+We use this form to specify an operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
+The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
+Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
 
-The operational semantics, expressed as a bigstep relation, are defined in \autoref{f:core-bigstep}.
-The judgment form $\bigstep{\Sigma}{e}{v}$ takes a \emph{stream history} ($\Sigma$), which is a non-empty sequence of stores, as well as an expression to evaluate ($e$) and the final value ($v$).
+\section{Extraction}
+\label{s:extraction}
 
-Rule Value denotes that a value evaluates to itself under any streaming history.
-Rule Var denotes that evaluating a variable requires looking it up in the right-most (most recent) store; the syntax $\Sigma_\bot; \sigma$ means that this rule applies to any streaming history with one or more steps.
-Rule App-Meta allows applying functions on the meta-level; the syntax $(\mlamX{}) v'$ refers to meta-level application of some meta-term.
+Pipit can generate executable code which is suitable for real-time execution on embedded devices.
+The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates an imperative representation of the program.
+\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}.
+The result of our translation to imperative code fits in this subset.
+During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, similar to staged compilation.
 
-The rules $\mbox{Fby}_1$ and $\mbox{Fby}_S$ describe the semantics of delayed streams.
-Rule $\mbox{Fby}_1$ denotes that when a delay $\xfby{v}{e}$ is evaluated in a context with exactly one step --- that is, at the very start of program execution --- it simply returns the initial value $v$.
-Rule $\mbox{Fby}_S$ applies when the same delay is evaluated in a context with strictly more than one step ($\Sigma; \sigma$), in which case, the most recent step is discarded, and the original expression $e$ is evaluated under the previous history.
+% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
+% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
 
-Next, rules ($\xthenarrow_1$) and ($\xthenarrow_S$) describe the semantics of streams composed by \emph{then}-expressions; as with the rules for delayed streams, the first rule applies when the streaming history contains exactly one step, while the second applies for histories of strictly more than one step.
+% noextract
+% let expr = Sugar.run2 Pump.controller'
 
-Rule ($\mu{}$) describes the semantics of recursive streams.
-This rule performs a single unfolding of a recursive expression $\xrec{x}{e[x]}$ so that all occurrences of $x$ inside $e$ are replaced with the recursive expression itself.
-This rule is not trivially terminating, but so long as any occurrences of the recursive binder are guarded by a delay, then all occurrences of the recursive binder will have a shorter stream history.
-This intuition is formalised in \REF{}.
+% noextract
+% let system: Pipit.Exec.Exp.xexec expr =
+%   XX.exec_of_exp expr
 
-Rule Let uses a standard substitution-based semantics for let-expressions.
-Finally, rule Check evaluates a property and requires it to be true ($\top$).
+% [@@(Tac.postprocess_with XL.tac_extract)]
+% let reset = XL.mk_reset system
 
-The semantics here, particularly the recursive rule, differ from previous semantics of Lustre.
-In the reactive semantics of \citet{caspi1995functional}, as well as the coinductive semantics of Vélus \cite{bourke2017formally} (\CITEME{find best paper}), the semantics for mutually recursive definitions is locally non-deterministic, in that the semantics does not define which particular order mutual definitions are evaluated in.
-This local non-determinism complicates certain proofs, even though the overall evaluation \emph{is} deterministic \TODO{vague}.
+% [@@(Tac.postprocess_with XL.tac_extract)]
+% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
+% This allows us to generate C code which executes in bounded space on embedded devices.
 
-In contrast, the semantics presented here is syntax-directed and completely deterministic.
-We have a mechanised proof of variant of progress theorem for well-formed (\emph{causal}) programs as well as closure under substitution and determinism, which is described in \autoref{ss:meta-theory}.
-However, the substitution-based semantics here can duplicate expressions, and encoding mutually recursive definitions requires multiple substitutions.
-For this reason, the semantics here is not suitable for actual execution.
+\section{Evaluation}
+\label{s:coffee-machine}
 
-\subsection{Causality}
-\label{ss:causality}
+To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
+This system controls a water-flow solenoid to fill the reservoir of a coffee machine with multiple safeguards to reduce the risk of flooding.
+The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
+The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
+The second safeguard is that the solenoid should not be engaged for more than a minute or so as this may indicate a leak; if so, the controller enters a terminal \emph{stuck} state.
+Finally, to avoid banging the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.
 
-Talk about the causality restriction.
-Does expression $e$ have a dependency on the most-recent value of expression $e'$?
+\newcommand\estop{\textit{stop}}
+\newcommand\low{\textit{low}}
+\newcommand\soltry{\textit{try}}
+\newcommand\stuck{\textit{stuck}}
+\newcommand\solen{\textit{engage}}
+\begin{tabbing}
+  @MM@\= @let@ engage \= @MMMM@ \= \kill
+  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
+  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
+  \> @let@ \stuck{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
+  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \stuck{} @in@ \\
+  \> @check@ (\solen{} $\implies$ not \estop{}); \\
+  \> @check@ (\solen{} $\implies$ \low{}); \\
+  \> \solen{}
+\end{tabbing}
 
-\begin{code}
-let rec direct_dependency (e e': exp) : bool =
-  if e = e' then true
-  else
-  match e with
-  | XVal _ -> false
-  | XVar x' -> false
-  | XPrim2 _ e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
-  | XIte ep et ef -> direct_dependency ep e' || direct_dependency et e' || direct_dependency ef e'
-  | XThen e1 e2 -> direct_dependency e1 e' || direct_dependency e2 e'
-  | XPre _ -> false
-  | XMu e1 -> direct_dependency e1 (lift e' 0)
-  | XLet e1 e2 -> direct_dependency e1 e' || direct_dependency e2 (lift e' 0)
-
-let rec causal (e: exp): bool =
-  match e with
-  | XVal _ -> true
-  | XVar x' -> true
-  | XPrim2 _ e1 e2 -> causal e1 && causal e2
-  | XIte ep e1 e2 -> causal ep && causal e1 && causal e2
-  | XThen e1 e2 -> causal e1 && causal e2
-  | XPre e1 -> causal e1
-  | XMu e1 -> causal e1 && not (direct_dependency e1 (XVar 0))
-  | XLet e1 e2 -> causal e1 && causal e2
-\end{code}
-
-\begin{lemma}[no dependency, no difference]
-If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.
-
-For all expressions $e$ and well-scoped variables $x$,
-if $\neg \mbox{directly} ~e~ x$ and
-$\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
-then for any $v_x'$, evaluation is unchanged
-$\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
-\end{lemma}
-
-We also need:
-\begin{lemma}[Bigstep recursive XMu]
-  If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:
-
-  if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
-  then for any $v'$,
-  $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
-\end{lemma}
-
-% let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
-%   (streams: C.table (outer + 1) inner)
-%   (vs: C.vector value outer) (v v': value)
-%   (hBSmu: bigstep streams (XMu e) (v :: vs)):
-%     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =
-
-\begin{lemma}[progress]
-  if expression is causal, then for any valid stream history it should evaluate to a value:
-
-  if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
-  $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
-\end{lemma}
-
-\begin{lemma}[monotonic progress]
-  if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
-  there exists some $v'$ such that
-  $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
-\end{lemma}
-
-\subsection{Meta-theory}
-\label{ss:meta-theory}
-
-We use a typed representation, which ensures an intrinsic variant of preservation.
-For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.
-
-\begin{code}
-let bigstep_deterministic
-  (#streams: list (C.row 'c))
-  (#e: exp 'c 'a)
-  (#v1 #v2: 'a)
-  (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
-    Lemma (ensures (v1 == v2)) (decreases hBS1) =
-  bigstep_proof_equivalence hBS1 hBS2
-\end{code}
-
-Monotonicity
-
-Substitution-closure
-\begin{code}
-let rec bigstep_substitute_as_var (e p: exp)
-  (streams1: C.table outer inner1)
-  (streams2: C.table outer inner2)
-  (vsp vsep: C.vector value outer)
-  (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
-  (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
-    Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =
-
-\end{code}
-
-\section{Proving it with transition systems}
-
-Describe the translation from Pipit expressions to labelled transition systems.
-
-K-induction
-
-Statement of LTS soundness:
-\begin{code}
-let system_eval_complete'
-    (#outer: nat) (#vars: nat)
-    (e: exp { causal e /\ wf e vars })
-    (streams: C.table outer vars)
-    (vs: C.vector value outer)
-    (hBS: bigstep streams e vs):
-      Lemma (exists (s': option (state_of_exp e)).
-        system_stepn'
-          (system_of_exp e vars)
-          (C.Table?._0 streams) vs s') =
-    match outer with
-    | 0 -> assert
-      (system_stepn'
-        (system_of_exp e vars)
-        (C.Table?._0 streams) vs None)
-    | _ -> system_eval_complete e streams vs hBS
-\end{code}
-
-Tactics: \cite{martinez2019meta}.
-
-Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-\section{Execution}
-
-We have implemented an unverified imperative code generation transform, by making use of \fstar{}'s existing C code extraction support \cite{protzenko2017verified}.
-% This transform generates \fstar{} functions in the
-
-This allows us to generate C code which executes in bounded space on embedded devices.
-
-Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-(Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Coffee machine}
-\label{s:coffee-machine}
+As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
+We can also generate executable C code, which runs in real-time on a microcontroller in the first author's kitchen.
 
@@
 
-hardware language in Agda \cite{harrison2021mechanized}
+Earlier work verifying reactive systems used model-checking via explicit state-space exploration~\cite{raymond2008synchronous}; more recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or check refinement types~\cite{chen2022synchronous}.
+These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples that disprove properties; they can also implement advanced invariant generation techniques such as ICE~\cite{garg2014ice} and IC3/PDR~\cite{bradley2011sat,een2011efficient}.
+However, these special-purpose model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
+We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
 
-Refinement types for Zélus \cite{chen2022synchronous}
+Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
+There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
+We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
+however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, and our currently-unverified imperative code generator.
 
-\TODO{discuss TCB: big, much bigger than Vélus}
+% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
 
-\section{Conclusion and future work}
+% cite also:
+% hardware language in Agda \cite{harrison2021mechanized}
+% Refinement types for Zélus \cite{chen2022synchronous}
+% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
 
+% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
+% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
+% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
+% (Is this the same as Lucid Synchrone and Zélus? \CITE)
+
+\section{Conclusion}
+
+We believe that these preliminary results show promise.
+In the intermediate future, we intend to verify the imperative code generation.
+
+TODO
 Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
@@
 % \pagebreak
-% \begin{acks}
-%   acknowledgement
-% \end{acks}
 

### 2023-05-11 2023pipit/figures/core-grammar.tex
almost

@@
           & | & \xfby{v}{e} & | & \xthen{e}{e'} \\
-          & | & \xrec{x}{e[x]} & | & \xlet{x}{e}{e'[x]} \\
-          & | & \xcheck{e} \\
+          & | & \xrec{x}{e[x]} & | & \xcheck{e} \\
+          & | & \multicolumn{3}{l}{\xlet{x}{e}{e'[x]}} \\
     \\
-    v & := & \multicolumn{3}{l}{n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots ~|~ \mlamX}
-    \\
-    \sigma & := & \sgl{\ov{x \mapsto v}} \\
-    \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
+    v & := & \multicolumn{3}{l}{n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots ~|~ \mlamX} \\
+    % \sigma & := & \sgl{\ov{x \mapsto v}} \\
+    % \Sigma & := & \sigma ~|~ \Sigma;\sigma \\
     %   \\
@@
   \]
-  \caption{Pipit language grammar, including expressions ($e$); values and host functions embedded from the \fstar{} meta-language ($v$); stores ($\sigma$); and non-empty stream histories ($\Sigma$).}
+  \caption{Pipit language grammar including expressions~($e$); and values and host functions embedded from the \fstar{} meta-language~($v$).}
   \label{f:core-grammar}

### 2023-05-11 2023pipit/shelve.tex
almost

@@
 Otherwise, most of the rules are standard.
+
+%%% META-THEORY
+
+\begin{lemma}[no dependency, no difference]
+  If $e$ does not have a direct depend on $x$, then any evaluation of $e$ cannot depend on the current value of $x$.
+
+  For all expressions $e$ and well-scoped variables $x$,
+  if $\neg \mbox{directly} ~e~ x$ and
+  $\bigstep{\Sigma; (\sigma, x \mapsto v_x)}{e}{v}$
+  then for any $v_x'$, evaluation is unchanged
+  $\bigstep{\Sigma; (\sigma, x \mapsto v_x')}{e}{v}$
+  \end{lemma}
+
+  We also need:
+  \begin{lemma}[Bigstep recursive XMu]
+    If a recursive expression $\xrec{x}{e}$ evaluates to some sequence of values $v$, then evaluating the recursive expression doesn't depend on the last value in the sequence:
+
+    if $\bigsteps{\Sigma}{\xrec{x}{e}}{v_0 v_1 \cdots v_{n-1} v_n }$
+    then for any $v'$,
+    $\bigsteps{\Sigma \triangleright^{\uparrow} (v_0 v_1 \cdots v_{n-1} v')}{e}{v_0 v_1 \cdots v_{n-1} v_n}$
+  \end{lemma}
+
+  % let bigstep_recursive_XMu (#outer #inner: nat) (e: exp)
+  %   (streams: C.table (outer + 1) inner)
+  %   (vs: C.vector value outer) (v v': value)
+  %   (hBSmu: bigstep streams (XMu e) (v :: vs)):
+  %     bigstep (C.table_map_append (C.table_of_values (v' :: vs)) streams) e (v :: vs) =
+
+  \begin{lemma}[progress]
+    if expression is causal, then for any valid stream history it should evaluate to a value:
+
+    if $\typing{\Gamma}{e}{\tau}$ and $e \mbox{causal}$ and $\Gamma \vdash* \Sigma$, then exists $v$ such that
+    $\bigstep{\Sigma}{e}{v}$ and $\mtyping{}{v}{\tau}$
+  \end{lemma}
+
+  \begin{lemma}[monotonic progress]
+    if an expression $e$ evaluates to a sequence $vs$ under a stream history $\Sigma$ and a new stream store $\sigma$ is valid, then
+    there exists some $v'$ such that
+    $\bigsteps{\Sigma; \sigma}{e}{vs; v}$
+  \end{lemma}
+
+  \subsection{Meta-theory}
+  \label{ss:meta-theory}
+
+  We use a typed representation, which ensures an intrinsic variant of preservation.
+  For programs that satisfy the causality restriction, we have proved a variant of the progress theorem.
+
+  \begin{code}
+  let bigstep_deterministic
+    (#streams: list (C.row 'c))
+    (#e: exp 'c 'a)
+    (#v1 #v2: 'a)
+    (hBS1: bigstep streams e v1) (hBS2: bigstep streams e v2):
+      Lemma (ensures (v1 == v2)) (decreases hBS1) =
+    bigstep_proof_equivalence hBS1 hBS2
+  \end{code}
+
+  Monotonicity
+
+  Substitution-closure
+  \begin{code}
+  let rec bigstep_substitute_as_var (e p: exp)
+    (streams1: C.table outer inner1)
+    (streams2: C.table outer inner2)
+    (vsp vsep: C.vector value outer)
+    (hBSp: bigstep (C.table_map_append streams1 streams2) p vsp)
+    (hBSep: bigstep (C.table_map_append streams1 streams2) (subst e inner1 p) vsep):
+      Tot (bigstep (C.table_map_append streams1 (C.table_map_append (C.table_of_values vsp) streams2)) e vsep) (decreases hBSep) =
+
+  \end{code}
+
+  \section{Proving it with transition systems}
+
+  Describe the translation from Pipit expressions to labelled transition systems.
+
+  K-induction
+
+  Statement of LTS soundness:
+  \begin{code}
+  let system_eval_complete'
+      (#outer: nat) (#vars: nat)
+      (e: exp { causal e /\ wf e vars })
+      (streams: C.table outer vars)
+      (vs: C.vector value outer)
+      (hBS: bigstep streams e vs):
+        Lemma (exists (s': option (state_of_exp e)).
+          system_stepn'
+            (system_of_exp e vars)
+            (C.Table?._0 streams) vs s') =
+      match outer with
+      | 0 -> assert
+        (system_stepn'
+          (system_of_exp e vars)
+          (C.Table?._0 streams) vs None)
+      | _ -> system_eval_complete e streams vs hBS
+  \end{code}

### 2023-05-12 2023pipit/Main.tex
address page limit

@@
 %% for your publication.
-\documentclass[sigplan,screen]{acmart}
+\documentclass[sigplan,screen, review]{acmart}
 
@@
   Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a simple reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved using k-induction on the transition system.
+  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
+  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
   Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code has been proved to preserve types; the proof that the generated imperative code preserves semantics is future work.
+  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
 \end{abstract}
@@
 To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In the following, we introduce Pipit, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
+In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
 Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is future work.
+Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is ongoing.
 
-In this extended abstract, we will briefly describe the following preliminary results:
+In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
 
 \begin{itemize}
-  \item embedded language reduces the need for fancy features (\autoref{s:tut});
-  \item normalisation and proof automation allow us to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item normalisation and extraction allows us to generate unverified executable code (\autoref{s:extraction});
-\end{itemize}
-
-We will also touch upon the following work-in-progress, which we intend to publish upon completion:
-
-\begin{itemize}
-  \item we describe the syntax (\autoref{s:core});
-  \item a unified semantics for theorem proving and computation;
-  \item verified imperative code generation;
-  \item the ability to mix manual proofs with automated proofs
-  \item further evaluation with the verification of a realistic controller
+  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
+  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
+  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
+  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
+  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
 \end{itemize}
@@
 % \section{Pipit in flight}
-\pagebreak
+% \pagebreak
 \section{Programming and verifying in Pipit}
@@
 A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
+An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
 We can implement an FIR filter in Pipit as follows:
 
+\newcommand\signal{\textit{signal}}
 \begin{tabbing}
   @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($s$: stream $\RR$): stream $\RR$ = \\
+  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
     \> @match@ $\textit{coefficients}$ @with@ \\
     \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~s)})$
+    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
 \end{tabbing}
 
-Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream\footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}.
+% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
+The coefficient vector is represented by a list of reals here, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
+% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
 The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($s \cdot c$) and delay the result ($\xfby{0}{\ldots}$) of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ s$), summing the two parts.
-When there is no previous value, as is the case at the start of execution, the delay is given the initial value of zero.
+If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
+At the start of execution the delay is initialised to zero.
 
-Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
+Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
 To write this program in Lustre, we would likely need to reformulate the program to use arrays.
-Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support arrays, but its model-checker Lesar~\cite{raymond2008synchronous} does not.
+Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays that no compiler supports~\cite{champion2016kind2}; and Lustre v6 does support arrays, but its model-checker Lesar does not~\cite{raymond2008synchronous}.
 
-Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}), inspired by Bedrock~\cite{chlipala2013bedrock}.
+Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit with a small core language embedded inside a theorem prover (\fstar{} in our case).
 By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
-For example, we can compute the weighted moving average with two coefficients:
+For example, we can easily compute the weighted moving average with concrete coefficients:
 
@@
 
-Which then normalises to the following simple definition:
+Normalising the above definition of $\mbox{fir}_2$ removes all lists.
+The result fits in the core language defined in \autoref{s:core}, for which we can generate statically-allocated imperative code:
 
@@
 
-To state properties about our filter, it is useful to be able to state temporal properties.
-We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
+The properties that we want to state about reactive programs usually involve some temporal aspect.
+Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
 We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
@@
   @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{\top}{p'})}$
+    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
 \end{tabbing}
@@
 This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{\top}{p'}$).
+At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
 If there is no previous value, it defaults to true.
@@
   @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\RR$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n)) \implies |\mbox{fir}_2~\ii| \le n)$
+  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
 \end{tabbing}
 
-This property states that, for this particular set of coefficients, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
+This property states that for the coefficients used by $\fir_2$, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
 Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
@@
 
-If we wish to prove a similar BIBO property for more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
+If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
 One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:
+We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
 
@@
 
-Although the properties we have shown here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.
+Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.
 
-\pagebreak
+% \pagebreak
 \section{Core language}
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to specify an operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
+We use this form to specify a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
 The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
@@
 To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine with multiple safeguards to reduce the risk of flooding.
+This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
 The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
 The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The second safeguard is that the solenoid should not be engaged for more than a minute or so as this may indicate a leak; if so, the controller enters a terminal \emph{stuck} state.
-Finally, to avoid banging the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.
+The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
+Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.
 
@@
 \newcommand\soltry{\textit{try}}
-\newcommand\stuck{\textit{stuck}}
+\newcommand\error{\textit{error}}
 \newcommand\solen{\textit{engage}}
@@
   \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \stuck{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \stuck{} @in@ \\
+  \> @let@ \error{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
+  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
   \> @check@ (\solen{} $\implies$ not \estop{}); \\
@@
 As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
-We can also generate executable C code, which runs in real-time on a microcontroller in the first author's kitchen.
+Pipit generates real-time C code for this example.
 
@@
 
-Earlier work verifying reactive systems used model-checking via explicit state-space exploration~\cite{raymond2008synchronous}; more recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples that disprove properties; they can also implement advanced invariant generation techniques such as ICE~\cite{garg2014ice} and IC3/PDR~\cite{bradley2011sat,een2011efficient}.
-However, these special-purpose model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
+Earlier work model-checking reactive systems used explicit state-space exploration~\cite{raymond2008synchronous}; more-recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
+These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples; they can also implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
+However, these model-checkers do not provide high assurance that the semantics they use for proofs matches the compiled code.
 We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
+
+The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
 
@@
 
-We believe that these preliminary results show promise.
-In the intermediate future, we intend to verify the imperative code generation.
+Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
+Next, we intend to verify the imperative code generation.
+To prove large programs, we also need some way to compose smaller proofs together, such as contracts.
+Finally, we need to evaluate Pipit on larger control systems before extending the language with clocks.
 
-TODO
-Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
+% TODO
+% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
 

### 2023-05-12 2023pipit/Main.tex
good

@@
 % \usepackage{minipage-marginpar}
-
-% \usepackage{cite}
 
@@
 To write this program in Lustre, we would likely need to reformulate the program to use arrays.
-Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays that no compiler supports~\cite{champion2016kind2}; and Lustre v6 does support arrays, but its model-checker Lesar does not~\cite{raymond2008synchronous}.
+Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
 
 Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit with a small core language embedded inside a theorem prover (\fstar{} in our case).
-By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
-For example, we can easily compute the weighted moving average with concrete coefficients:
+By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
+% which can be easily executed in statically-bounded runtime and space.
+For example, we can compute the weighted moving average with the following concrete coefficients:
 
@@
 We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
-Such properties can occur as an internal invariant in a larger proof.
+Properties like this can occur as an internal invariant in a larger system:
 
@@
   @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\RR$ = \\
+  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
   \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
@@
 
-This property states that for the coefficients used by $\fir_2$, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
+This property states that
+% for the coefficients used by $\fir_2$,
+  if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
 Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system.
-In this case, simple induction over the transition relation is sufficient to prove the property.
+To prove that this property holds, we translate to a transition system and show that the stream is always true.
+In this case, induction over the transition relation is sufficient to prove the property.
 There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to specify a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
+We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
 The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
@@
 Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates an imperative representation of the program.
+The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
 \fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}.
@@
   \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
+  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
   \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
@@
 
-As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
+% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
+As with the examples in \autoref{ss:tut:bibo}, the above properties can be automatically verified.
+% by \fstar{}'s proof automation.
 Pipit generates real-time C code for this example.
@@
 Next, we intend to verify the imperative code generation.
-To prove large programs, we also need some way to compose smaller proofs together, such as contracts.
+To verify large programs, we also need some way to compose smaller proofs together, such as contracts.
 Finally, we need to evaluate Pipit on larger control systems before extending the language with clocks.

### 2023-05-16 2023pipit/Main.tex
copyediting

@@
   Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
+  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
   This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
@@
 
-Safety-critical control systems, such as the anti-lock braking systems that occur in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
+Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
+One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
 These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
@@
 For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.
+% TODO: link to Kind2 issue?
 
@@
 In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
+Pipit aims to provide a high-level language based on a subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is ongoing.
+Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
 
@@
 % Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals here, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
+The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
 % \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
@@
 Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
+%TODO defer explanation to related work
 
-Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit with a small core language embedded inside a theorem prover (\fstar{} in our case).
+Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit as a small core language embedded inside a theorem prover (\fstar{} in our case).
 By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
-% which can be easily executed in statically-bounded runtime and space.
 For example, we can compute the weighted moving average with the following concrete coefficients:
@@
 Normalising the above definition of $\mbox{fir}_2$ removes all lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate statically-allocated imperative code:
+The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
 
@@
 
-This property states that
-% for the coefficients used by $\fir_2$,
-  if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
+This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
 Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
@@
 We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
+% TODO: intuition, (0 fby x) < n = true fby (x < n)
 
@@
 The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions other than reusing closed functions from the \fstar{} meta-language.
+The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
 Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
@@
 % Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the examples in \autoref{ss:tut:bibo}, the above properties can be automatically verified.
+As with the examples in \autoref{ss:tut:bibo}, the two properties can be automatically verified.
 % by \fstar{}'s proof automation.

### 2023-05-16 2023pipit/Main.tex
shuffling

@@
 In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on a subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
+Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
@@
 % \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
-At the start of execution the delay is initialised to zero.
+The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
+If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
+At the start of execution the delay is initially zero.
 
-Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
-To write this program in Lustre, we would likely need to reformulate the program to use arrays.
-Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-%TODO defer explanation to related work
-
-Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit as a small core language embedded inside a theorem prover (\fstar{} in our case).
-By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
-For example, we can compute the weighted moving average with the following concrete coefficients:
+Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match is part of the \fstar{} meta-language.
+To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
 
@@
 
-Normalising the above definition of $\mbox{fir}_2$ removes all lists.
+Normalising this definition evaluates away all of the lists.
 The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
@@
 \label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
-Properties like this can occur as an internal invariant in a larger system:
+We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
+% Properties like this can occur as an internal invariant in a larger system:
 
@@
 
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
+If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient; the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
 One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
 We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-% TODO: intuition, (0 fby x) < n = true fby (x < n)
 
@@
 To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
+This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
 The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
@@
 \end{tabbing}
+\pagebreak
 
@@
 % by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example.
+Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
 
 \section{Related work}
+\label{s:related-work}
+% TODO arrays
+
+Although the FIR filter from \autoref{s:tut} is a simple definition, verifying it \emph{and} executing it with existing Lustre tools is impractical.
+To write this program in Lustre we would likely need to reformulate the program to use arrays, as dynamically-allocated data structures are not well-suited to real-time execution.
+Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
 

### 2023-05-16 2023pipit/Main.tex
looks alright

@@
 % Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the examples in \autoref{ss:tut:bibo}, the two properties can be automatically verified.
+As with the previous examples, the two properties can be automatically verified.
 % by \fstar{}'s proof automation.
@@
 
-Although the FIR filter from \autoref{s:tut} is a simple definition, verifying it \emph{and} executing it with existing Lustre tools is impractical.
-To write this program in Lustre we would likely need to reformulate the program to use arrays, as dynamically-allocated data structures are not well-suited to real-time execution.
-Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
+Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
+Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
+To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
+However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
 
-Earlier work model-checking reactive systems used explicit state-space exploration~\cite{raymond2008synchronous}; more-recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples; they can also implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
+In terms of model-checking reactive systems,
+% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
+recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
+These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
 However, these model-checkers do not provide high assurance that the semantics they use for proofs matches the compiled code.
@@
 We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, and our currently-unverified imperative code generator.
+however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
 
@@
 Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to compose smaller proofs together, such as contracts.
-Finally, we need to evaluate Pipit on larger control systems before extending the language with clocks.
+To verify large programs, we also need some way to separately prove smaller pieces which can be composed together, such as contracts~\cite{champion2016kind2}.
+Finally, we need to evaluate Pipit on larger control systems before extending the language with more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
 

### 2023-05-16 2023pipit/Main.tex
polish

@@
 
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient; the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
+If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
 One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
@@
 
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.
+Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
 
@@
 % The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the latest values from stream $e'$ for subsequent steps.
+Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
 
@@
 The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}.
-The result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, similar to staged compilation.
+\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
+During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
 
@@
 The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.
+Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
 
@@
 
-% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
+Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
+% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
 As with the previous examples, the two properties can be automatically verified.
@@
 These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide high assurance that the semantics they use for proofs matches the compiled code.
+However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
 We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
@@
 Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language with more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
+To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
+Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
 

### 2023-05-17 2023pipit/Main.tex
cite github issue

@@
 Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in the Vélus compiler, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB}.
-% TODO: link to Kind2 issue?
+This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
+For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
 

### 2023-05-19 2023pipit/Main.tex
tried to add more related work but it will not fit!

@@
 
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match is part of the \fstar{} meta-language.
+Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
 To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
-The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
+We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
+Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
 Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

### 2023-12-08 2023pipit/Main.tex
add missing files

@@
 \section{Introduction}
-
 Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.

### 2023-12-08 2023pipit/code/therm.c

@@ -1,0 +1,61 @@
+/** Example of using driver in therm.lus from C code */
+
+/** The driver generates C code that defines a state type and a return type: */
+
+typedef struct driver_state {
+  // ...
+} driver_state;
+
+typedef struct driver_return {
+  TCMD cmd;
+  bool temp_read_success;
+  bool temp_good;
+  int  temp;
+} driver_return;
+
+/** And reset and step functions. */
+void driver_reset(driver_state *state);
+/** The step function performs a single step of the transition system, given input `rsp` and storing the result in `output` */
+void driver_step(driver_state *state, driver_return *output, TRSP rsp);
+
+/** Our driver will call the step function and interface with the hardware.
+  It's conceptually just a big while loop, but in practice we'd probably split
+  it out to a timer function that's run every 100ms or so because we'd have
+  other work to do in the meantime.
+ */
+
+/** write one byte on i2c bus */
+bool i2c_write1(char address, char register, char value);
+/** read any number of bytes on i2c bus, returning number of bytes actually read */
+int i2c_read(char address, char register, int nbytes, char *output);
+
+void driver_c_loop() {
+  driver_state state;
+  driver_reset(&state);
+  TRSP last_rsp = { false, 0, 0 };
+  while (true) {
+    driver_return r;
+    driver_step(&state, &r, last_rsp);
+
+    switch (r.cmd) {
+      case CMD_RESET:
+        last_rsp.cmd_ok = i2c_write1(I2C_TEMP, I2C_REG_RESET, 1);
+        break;
+      case CMD_SET_INT_ENABLE:
+        last_rsp.cmd_ok = i2c_write1(I2C_TEMP, I2C_REG_INT, 1);
+        break;
+      case CMD_SET_INT_DISABLE:
+        last_rsp.cmd_ok = i2c_write1(I2C_TEMP, I2C_REG_INT, 0);
+        break;
+      case CMD_READ:
+        char bytes[2];
+        int nbytes = i2c_read(I2C_TEMP, I2C_REG_TEMP, 2, &bytes);
+        last_rsp.cmd_ok = (nbytes == 2);
+        last_rsp.temp  = bytes[0];
+        last_rsp.fresh = bytes[1];
+        break;
+
+      case CMD_NONE:
+    }
+  }
+}

### 2023-12-08 2024pipit/Main.tex
start pipit paper

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+% To anonymise:
+% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}
 
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
+% \pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
+\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+
+\bibliographystyle{plainurl}
+
+% is this package necessary?
 \usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
 
@@
 
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
+\title{Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems}
 
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
+\author{Amos Robinson}{Australian National University, Canberra, Australia}{amos.robinson@anu.edu.au}{https://orcid.org/0009-0004-4837-4981}{}
+
+\author{Alex Potanin}{Australian National University, Canberra, Australia}{alex.potanin@anu.edu.au}{https://orcid.org/0000-0002-4242-2725}{}
+
+\authorrunning{A. Robinson and A. Potanin}
+\Copyright{Amos Robinson and Alex Potanin}
+% \ccsdesc[300]{Computer systems organization~Embedded software}
+\ccsdesc[500]{Computer systems organization~Real-time languages}
+\ccsdesc[500]{Theory of computation~Program verification}
+% \ccsdesc[100]{Theory of computation~Modal and temporal logics}
+\ccsdesc[500]{Software and its engineering~Specialized application languages}
+\keywords{Lustre, streaming, reactive, verification}
+
+% \category{}
 
@@
 
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
+\maketitle
 
@@
 
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
 % enable @verbatim@ syntax
@@
 
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
+\input{section/01-Introduction.tex}
 
@@
 \input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
+\input{figures/core-typing.tex}
+\input{figures/core-bigstep.tex}
 
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
+We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
 Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
@@
 
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{Main}
@@
 \endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-08 2024pipit/section/01-Introduction.tex
start pipit paper

--- 2023-12-08 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
+%!TEX root = ../Main.tex
 \section{Introduction}
@@
 To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
+In this paper we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
 Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
@@
 
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
+In this paper, we make the following contributions:
 
 \begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
+  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
+  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; by checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\TODO{ref});
+  \item we describe the translation to transition system (\TODO{ref});
+  \item we identify the invariants and lemmas required to prove preservation of both dynamic (bigstep) semantics and checked semantics, required to transform properties proven on transition systems to the original program (\TODO{ref});
+  \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\TODO{ref}).
 \end{itemize}
+
+% \begin{itemize}
+%   \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
+%   \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
+%   \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
+%   % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
+%   \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
+% \end{itemize}
 
@@
 % \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-12 2024pipit/Main.tex
start checked semantics

@@
 \usepackage{mathtools}
+\usepackage{mathpartir}
+
+% checked box
+\usepackage{wasysym}
 
@@
 \input{figures/core-bigstep.tex}
+\input{figures/core-check.tex}
 
 The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
+The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(x)$.
 Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.

### 2023-12-16 2024pipit/Main.tex
ok

@@
 \usepackage{wasysym}
+\usepackage{underscore}
 
@@
 
+Big TODOs:
+\begin{itemize}
+  \item example: trigger-fetch from ttcan;
+  \item transition systems and translation;
+  \item C codegen: note about duplicating work, sketch potential solution;
+  \item C codegen: future work: sketch potential solution for separate compilation, compare to Accelerate FFI \cite{clifton2014embedding};
+  \item cite Liam \cite{oconnor2019deferring} similar idea, but we defer the \emph{context} rather than just syntactically deferring the proof;
+  % \item 
+\end{itemize}
+
 \input{section/01-Introduction.tex}
+
+\section{Time-triggered}
+
+% \newcommand\signal{\textit{signal}}
+\begin{tabbing}
+  @MM@\= @MM@ \= \kill
+  @let@ count_when ($\textit{limit}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
+    \> @let rec@ $\textit{count}$ @=@ \\
+    \> \> @if@ $\textit{count} >= \textit{limit}$ @then@ $\textit{limit}$ \\
+    \> \> @else@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ \\
+    \> @in@ $\textit{count}$ \\
+\end{tabbing}
+
+\begin{tabbing}
+  @MM@\= @MM@ \= \kill
+  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
+    \> @contract@ \\
+    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
+    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
+    \> \> \{ trigger.~$\true$ \} \\
+\end{tabbing}
 

### 2023-12-18 2024pipit/Main.tex
pipit2024

--- 2023-12-12 2024pipit/Main.tex
@@
 
+\usepackage{stmaryrd}
+
 % checked box
 \usepackage{wasysym}
+\usepackage{underscore}
 
@@
 
+Big TODOs:
+\begin{itemize}
+  \item example: trigger-fetch from ttcan;
+  \item transition systems and translation;
+  \item C codegen: note about duplicating work, sketch potential solution;
+  \item C codegen: future work: sketch potential solution for separate compilation, compare to Accelerate FFI \cite{clifton2014embedding};
+  \item cite Liam \cite{oconnor2019deferring} similar idea, but we defer the \emph{context} rather than just syntactically deferring the proof;
+  \item oracle transition system in the semantics: does this have a standard name? references? not really related to prophecy variables.
+  % \item 
+\end{itemize}
+
 \input{section/01-Introduction.tex}
 
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-\input{figures/core-typing.tex}
-\input{figures/core-bigstep.tex}
-\input{figures/core-check.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(x)$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
+\input{section/20-Motivation.tex}
+\input{section/30-Core.tex}
+\input{section/40-Transition.tex}
+\input{section/50-Extraction.tex}
+\input{section/60-Evaluation.tex}
+\input{section/70-Related.tex}
+\input{section/80-Conclusion.tex}
 

### 2023-12-18 2024pipit/section/01-Introduction.tex
pipit2024

@@
 \begin{itemize}
+  \item we describe a key issue with prior program logics for Lustre-style languages (\autoref{s:motivation});
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
-  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; by checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\TODO{ref});
-  \item we describe the translation to transition system (\TODO{ref});
+  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\TODO{ref});
+  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified contracts (\autoref{s:transition});
   \item we identify the invariants and lemmas required to prove preservation of both dynamic (bigstep) semantics and checked semantics, required to transform properties proven on transition systems to the original program (\TODO{ref});
-  \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\TODO{ref}).
+  \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
 \end{itemize}

### 2023-12-18 2024pipit/section/20-Motivation.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
+%!TEX root = ../Main.tex
 
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
+\section{Time-triggered}
+\label{s:motivation}
 
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
+Talk about the problem.
 
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
+% \newcommand\signal{\textit{signal}}
+\begin{tabbing}
+  @MM@\= @MM@ \= \kill
+  @let@ count_when ($\textit{limit}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
+    \> @let rec@ $\textit{count}$ @=@ \\
+    \> \> @if@ $\textit{count} >= \textit{limit}$ @then@ $\textit{limit}$ \\
+    \> \> @else@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ \\
+    \> @in@ $\textit{count}$ \\
+\end{tabbing}
 
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
+\begin{tabbing}
+  @MM@\= @MM@ \= \kill
+  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
+    \> @contract@ \\
+    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
+    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
+    \> \> \{ trigger.~$\true$ \} \\
+\end{tabbing}
 
@@
 \label{s:tut}
+
+\TODO{KILLMERGE}
 
@@
 Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-18 2024pipit/section/30-Core.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
+%!TEX root = ../Main.tex
 
@@
 \input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
+\input{figures/core-typing.tex}
+\input{figures/core-bigstep.tex}
+\input{figures/core-check.tex}
 
 The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
+The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(x)$.
 Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
+We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
 Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
 Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-18 2024pipit/section/40-Transition.tex
pipit2024

%!TEX root = ../Main.tex

\section{Transition systems}
\label{s:transition}
\input{figures/system-types.tex}

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input environment and the result type.
It also contains two internal environments: firstly, the state environment describes the private state required to execute the machine; secondly, the free environment contains any extra input values that the transition system would like to quantify over.
The free environment is used to allow the system to concoct arbitrary values, when it would not otherwise be able to return a concrete value.
For contracts which abstract over the implementation, the natural definition of the step function would involve an existential quantifier: there exists some value that satisfies the specification.
Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation with existential quantifiers complicates the resulting transition, as the quantifiers get between the value and its definition.
Instead, the free environment provides the step function with a fresh unconstrained value of the desired type, and the step function itself can constrain it however it wishes.

The free environment is used to generate similar systems to the transition systems generated by existing systems such as Kind2 \cite{champion2016kind2}, which can generate fresh variables at will.
The main distinction is that embedding such a translation inside a theorem prover and proving it correct requires some ingenuity.

\input{figures/system-translation.tex}

### 2023-12-18 2024pipit/section/50-Extraction.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
+%!TEX root = ../Main.tex
 
@@
 % This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-18 2024pipit/section/60-Evaluation.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
+%!TEX root = ../Main.tex
 
 \section{Evaluation}
-\label{s:coffee-machine}
+\label{s:evaluation}
 
@@
 Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-18 2024pipit/section/70-Related.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
+%!TEX root = ../Main.tex
 
@@
 % (Is this the same as Lucid Synchrone and Zélus? \CITE)
-
-\section{Conclusion}
-
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
-
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-18 2024pipit/section/80-Conclusion.tex
pipit2024

--- 2023-05-19 2023pipit/Main.tex
@@
-%% Commands for TeXCount
-%TC:macro \cite [option:text,text]
-%TC:macro \citep [option:text,text]
-%TC:macro \citet [option:text,text]
-%TC:envir table 0 1
-%TC:envir table* 0 1
-%TC:envir tabular [ignore] word
-%TC:envir displaymath 0 word
-%TC:envir math 0 word
-%TC:envir comment 0 0
-%%
-%% For submission and review of your manuscript please change the
-%% command to \documentclass[manuscript, screen, review]{acmart}.
-%%
-%% When submitting camera ready or to TAPS, please change the command
-%% to \documentclass[sigconf]{acmart} or whichever template is required
-%% for your publication.
-\documentclass[sigplan,screen, review]{acmart}
-
-% any of these necessary?
-% \usepackage{amssymb}
-\usepackage{amsthm}
-% \usepackage{graphicx}
-\usepackage{amsmath}
-% \usepackage{mathptmx}
-\usepackage{mathtools}
-% \usepackage{stmaryrd}
-% \usepackage{adjustbox}
-% \usepackage{hyperref}
-% \usepackage{alltt}
-% \usepackage{url}
-% \usepackage{float}
-% \usepackage{minipage-marginpar}
-
-\usepackage{style/utils}
-\usepackage{style/code}
-\usepackage{style/proof}
-\usepackage{style/keywords}
-\usepackage{style/judgements}
-
-\setcopyright{acmcopyright}
-\copyrightyear{2023}
-\acmYear{2023}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-\acmConference[TyDe '23]{In submission}{2023}{In submission}
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-% \acmPrice{15.00}
-% \acmISBN{978-1-4503-XXXX-X/18/06}
-%%\acmSubmissionID{123-A56-BU3}
-
-\begin{document}
-
-\title{Pipit: Reactive Systems in \fstar ~(Extended Abstract)}
-
-\newcommand{\affiliationANU}{
-  \affiliation{%
-  \institution{Australian National University}
-  \city{Canberra}
-  \country{Australia}
-}
-}
-
-\author{Amos Robinson}
-% \authornote{}
-\email{amos.robinson@anu.edu.au}
-\orcid{0009-0004-4837-4981}
-\affiliationANU{}
-\author{Alex Potanin}
-% \authornotemark[1]
-\email{alex.potanin@anu.edu.au}
-\orcid{0000-0002-4242-2725}
-\affiliationANU{}
-
-% \renewcommand{\shortauthors}{Robinson et al.}
-
-\begin{abstract}
-  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
-  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
-  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
-\end{abstract}
-
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-\begin{CCSXML}
-<ccs2012>
-    <concept>
-        <concept_id>10010520.10010553.10010562.10010564</concept_id>
-        <concept_desc>Computer systems organization~Embedded software</concept_desc>
-        <concept_significance>300</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10010520.10010570.10010572</concept_id>
-        <concept_desc>Computer systems organization~Real-time languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10010124.10010138.10010142</concept_id>
-        <concept_desc>Theory of computation~Program verification</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10003752.10003790.10003793</concept_id>
-        <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
-        <concept_significance>100</concept_significance>
-        </concept>
-    <concept>
-        <concept_id>10011007.10011006.10011050.10011023</concept_id>
-        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
-        <concept_significance>500</concept_significance>
-        </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[300]{Computer systems organization~Embedded software}
-\ccsdesc[500]{Computer systems organization~Real-time languages}
-\ccsdesc[500]{Theory of computation~Program verification}
-\ccsdesc[100]{Theory of computation~Modal and temporal logics}
-\ccsdesc[500]{Software and its engineering~Specialized application languages}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{Lustre, streaming, reactive, control}
-
-\received{1 June 2023}
-% \received[revised]{12 March 2009}
-% \received[accepted]{5 June 2009}
-
-\maketitle
-
-% enable @verbatim@ syntax
-\makeatactive
-
-\section{Introduction}
-
-Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
-One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
-Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
-This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
-For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.
-
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
-Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
-% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
-
-In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:
-
-\begin{itemize}
-  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-\end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}
-
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
-
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
-
-\newcommand\signal{\textit{signal}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
-\end{tabbing}
-
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
-
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
-\end{tabbing}
-
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
-
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
-
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
-
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
-
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
-
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
-
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
-
-% \pagebreak
-\section{Core language}
-\label{s:core}
-
-\input{figures/core-grammar.tex}
-% \input{figures/core-typing.tex}
-% \input{figures/core-bigstep.tex}
-
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(\mlamX{})$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
-
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-% To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
-This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
-
-\section{Extraction}
-\label{s:extraction}
-
-Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
-
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
-
-% noextract
-% let expr = Sugar.run2 Pump.controller'
-
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
-
-\section{Evaluation}
-\label{s:coffee-machine}
-
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
-
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
-\begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
-\end{tabbing}
-\pagebreak
-
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
-
-\section{Related work}
-\label{s:related-work}
-% TODO arrays
-
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
-
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
-
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
-
-Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
-There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
-
-% Refinement types have also been used to verify reactive systems~\cite{chen2022synchronous}, but this work does not address the issue of correct compilation.
-
-% cite also:
-% hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
-
-% Other compilers for Lustre-style languages, such as Vélus \cite{bourke2017formally} and Heptagon \cite{gerard2012modular} ensure that the generated code can be executed in bounded memory and that each step takes a bounded time.
-% With a carefully chosen set of primitives, the core streaming operations shown here would ensure bounded memory and time.
-% For practicality, however, we currently allow embedding arbitrary total functions from the \fstar{} meta-language, which are not necessarily bounded in time.
-% (Is this the same as Lucid Synchrone and Zélus? \CITE)
+%!TEX root = ../Main.tex
 
@@
 % Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
-
-% \pagebreak
-
-%%
-%% The next two lines define the bibliography style to be used, and
-%% the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{Main}
-
-% \appendix
-
-\end{document}
-\endinput
-%%
-%% End of file `sample-sigplan.tex'.

### 2023-12-19 2024pipit/Main.tex
ok

@@
 \bibliographystyle{plainurl}
+
+\ifx\authoranonymous\relax
+\newcommand{\GITHUBURL}{(anonymised)}
+\else
+\newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
+\fi
 
@@
   \item transition systems and translation;
-  \item C codegen: note about duplicating work, sketch potential solution;
-  \item C codegen: future work: sketch potential solution for separate compilation, compare to Accelerate FFI \cite{clifton2014embedding};
-  \item cite Liam \cite{oconnor2019deferring} similar idea, but we defer the \emph{context} rather than just syntactically deferring the proof;
-  \item oracle transition system in the semantics: does this have a standard name? references? not really related to prophecy variables.
-  % \item 
 \end{itemize}

### 2023-12-19 2024pipit/section/01-Introduction.tex
ok

@@
 One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
-These model-checkers can prove many interesting properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
+These model-checkers can prove many interesting safety properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
 Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
@@
 To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
-In this paper we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
+In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
 Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.

### 2023-12-19 2024pipit/section/70-Related.tex
ok

@@
 % (Is this the same as Lucid Synchrone and Zélus? \CITE)
+
+The deferred aspect of our proofs is similar to the deferred proofs of verification conditions for imperative programs, such as \cite{oconnor2019deferring}.
+However, such verification conditions are \emph{syntactically} deferred so that the verification condition can be proved later; in our case, the verification conditions are \emph{semantically} deferred, so that more knowledge of the enclosing program can be exploited in the proof.
+In imperative programs, this sort of extra knowledge is generally provided explicitly as loop invariants, and non-looping statements have their weakest precondition computed automatically.
+In Lustre-style reactive languages such as ours, programs tend to be composed of many nested recursive streams, which perform a similar function to loops.
+Explicitly specifying an invariant for each recursive stream would be cumbersome; deferring the proof allows such invariants to be implicit.

### 2023-12-19 2024pipit/section/80-Conclusion.tex
ok

@@
 Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
+%  To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
 Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
@@
 %  Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
+
+We are interested in further pursuing the intersection of model-checking with interactive theorem proving.
+A smart contract called Djed \cite{zahnentferner2023djed} currently uses a mixture of Kind2 \cite{champion2016kind2} and manual Isabelle/HOL proofs to show that the contract is well-behaved.
+In future work, we would like to further investigate whether Pipit's integration of streaming proofs with \fstar{}'s automated proof system would be able to provide similar proofs, without introducing any semantic gap between the two systems.
+
+Our current array support is limited: constant arrays provide a pure index function, while we only support fixed-size mutable arrays by wrapping bit-vectors.
+Array support is an obvious direction for future work; integrating with a verified array-fusion system such as \cite{robinson2017machine} would be an interesting and useful extension.
+
+Future work: verified imperative code generation.

### 2023-12-19 2024pipit/section/20-Motivation.tex
ok

@@
 
-Talk about the problem.
+Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
+Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
+Like imperative programs, they describe how state evolves over time.
+Reasoning about them is somewhat tricky, then, as one must reason about finite prefixes of streams.
+
+As a concrete example to introduce Pipit, we consider a program with a static schedule of \emph{triggers}, or actions to be performed at a particular time.
+These triggers are stored in a constant array.
+The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
+This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time}; we will discuss further in \autoref{s:evaluation}.
+
+We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
+At each time step, $\mbox{count_when}~\textit{max}~\textit{inc}$ checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
 
@@
   @MM@\= @MM@ \= \kill
-  @let@ count_when ($\textit{limit}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
-    \> @let rec@ $\textit{count}$ @=@ \\
-    \> \> @if@ $\textit{count} >= \textit{limit}$ @then@ $\textit{limit}$ \\
-    \> \> @else@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ \\
-    \> @in@ $\textit{count}$ \\
+  @let@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
+    \> $@rec@~\rawbind{\textit{count}}{}$ \\
+    \> \> @let@ $\textit{count'}$ @=@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ @in@ \\
+    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
+    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
+    % \> $\textit{count}$ \\
 \end{tabbing}
+
+The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
+The syntax $0~@fby@~\textit{count}$ can be read as zero \emph{followed by} the previous count.
+
+The syntax $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$ asserts that the count is within the range $[0, \textit{max}]$.
+The subscript $\PSUnknown$ on the check is the \emph{property status}, which in this case denotes that the assertion has been stated, but it is not yet known whether it holds.
+A property status of $\PSValid$, on the other hand, denotes that a property has been proved to hold.
+These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.
+
+Pipit is an embedded domain-specific language. % we usually present Pipit programs using \fstar{} syntactic sugar rather than showing the desugared core program.
+The program above is really syntactic sugar for a \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable; we will discuss the details of the core language in \autoref{s:core}.
+
+To actually prove the property above, we translate the program into a transition system and prove the property inductively on the system.
+Finally, we \emph{bless} the expression, which marks the properties as valid.
+Blessing is an intensional operation: it modifies the internal structure of the expression, but does not affect the operational semantics.
+
+\begin{tabbing}
+  @MM@\= @MM@ \= \kill
+  @let@ $\mbox{count_when}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\BB$ $\to$ stream $\NN$ = \\
+    \> @let@ $\textit{system} = \mbox{System.translate}_1 (\mbox{count_when}~\textit{max})$ @in@ \\
+    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
+    \> $@bless@_1$ $(\mbox{count_when}~\textit{max})$
+\end{tabbing}
+
+The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
+This proof does require some boilerplate; in the future we plan to minimise this with metaprogramming.
+
+Callers of count_when can now use the valid variant without needing to re-prove the count-range property.
 
@@
   @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
-    \> @contract@ \\
+    \> $@contract@_{\PSUnknown}$ \\
     \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\

### 2023-12-20 2024pipit/section/20-Motivation.tex
ok

@@
 Like imperative programs, they describe how state evolves over time.
-Reasoning about them is somewhat tricky, then, as one must reason about finite prefixes of streams.
+Additionally, Lustre programs tend to be composed of many nested recursive loops.
+When verifying both pure and imperative programs, recursion or loops generally require explicit loop invariants.
+Requiring an explicit invariant for each recursive stream quickly becomes unwieldy for Lustre programs.
 
@@
 The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
-This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time}; we will discuss further in \autoref{s:evaluation}.
+This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
 
 We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
-At each time step, $\mbox{count_when}~\textit{max}~\textit{inc}$ checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
+At each time step, count_when checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
 
-% \newcommand\signal{\textit{signal}}
 \begin{tabbing}
@@
     \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
-    % \> $\textit{count}$ \\
 \end{tabbing}
@@
 The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
-The syntax $0~@fby@~\textit{count}$ can be read as zero \emph{followed by} the previous count.
+The syntax $0~@fby@~\textit{count}$ can be read as the initial value of zero \emph{followed by} the previous count.
 
@@
 These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.
+In practice, the user does not explicitly specify property statuses in the source language.
+The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true.
+Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are impliciitly lifted to constant streams.
+
+We need to defer the proof of the property here, because at the point of stating the property we don't yet have a concrete definition for the count variable.
+In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the count and putting the @check@ outside of the @rec@ combinator.
+However, it is not always possible to defer property statements: for example, when calling other streaming functions that have their own preconditions, it may not be possible to move the function call outside of its enclosing @rec@.
 
@@
 
-To actually prove the property above, we translate the program into a transition system and prove the property inductively on the system.
+To actually prove the property above, we use \fstar{}'s tactics and proof automation to translate the program into a transition system and prove the property inductively on the system.
 Finally, we \emph{bless} the expression, which marks the properties as valid.
-Blessing is an intensional operation: it modifies the internal structure of the expression, but does not affect the operational semantics.
+Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the operational semantics.
 
@@
 The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
-This proof does require some boilerplate; in the future we plan to minimise this with metaprogramming.
+The pipit_simplify tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof automation can then solve the inductive check directly.
+This proof does require some boilerplate; in the future we hope to minimise this.
 
 Callers of count_when can now use the valid variant without needing to re-prove the count-range property.
+In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
+By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we also reduce the semantic gap between compilation and proofs, as well as allowing manual proof steps when the automated proofs get stuck.
+As an embedded language, we also benefit from writing our programs in a rich metalanguage.
+
+For the trigger fetch logic, we will use count_when as the index into the triggers array.
+To retrieve the element from the triggers array, we assume a pure, non-streaming \fstar{} function for array indexing.
+We use an unsafe indexing function, which we lift to a streaming primitive and wrap in a contract to make it safe:
 
@@
   @MM@\= @MM@ \= \kill
-  @let@ trigger_fetch ($\textit{index}$: stream $\NN$): stream trigger = \\
+  @let@ array_index ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: stream $\NN$): stream $\alpha$ = \\
   \> $@contract@_{\PSUnknown}$ \\
-    \> \> \{ $0 \le \textit{index} \le \textit{maximum}$ \} \\
-    \> \> Array.index_const \textit{triggers_array} \textit{index} \\
-    \> \> \{ trigger.~$\true$ \} \\
+  \> \> \{ $0 \le \textit{index} < \mbox{Array.length}~\textit{array}$ \} \\
+  \> \> Array.index_unsafe \textit{array} \textit{index} \\
+  \> \> \{ $v.~v~=~\mbox{Array.index_unsafe}~\textit{array}~\textit{index}$ \} \\
 \end{tabbing}
 
-% \section{Pipit in flight}
-% \pagebreak
-\section{Programming and verifying in Pipit}
-\label{s:tut}
+Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
+The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
+In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
+The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
+This is a somewhat pathological contract, as the guarantee does not abstract away the details of the implementation at all.
 
-\TODO{KILLMERGE}
+We use the unsafe indexing function, rather than a safe variant with a precondition, as we cannot currently lift primitives with preconditions; we discuss future work to remove this limitation in \autoref{s:core:primitive-tables}.
 
-A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
-An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
-We can implement an FIR filter in Pipit as follows:
+We can now combine the count_when and index stream functions to implement a (much simplified) time-triggered trigger fetch controller.
 
-\newcommand\signal{\textit{signal}}
 \begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
-    \> @match@ $\textit{coefficients}$ @with@ \\
-    \> @|@ $[]$ \> $\to 0$ \\
-    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
+  @MM@\= @MM@ \= @let index@ \= \kill
+  @let@ trigger_fetch ($\textit{time}$: stream $\NN$): stream trigger = \\
+    \> $@rec @ \textit{trigger}.$ \\
+    \> \> $@let @ \textit{inc} = \text{false} @ fby @ (\textit{trigger.time_mark} < \textit{time}) @ in@$\\
+    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{Array.size}~\textit{trigger_array} - 1)~\textit{inc} @ in@$ \\
+    \> \> $\text{array_index}~\textit{trigger_array}~\textit{index}$\\
 \end{tabbing}
 
-% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
-The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
-% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
-The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
-If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
-At the start of execution the delay is initially zero.
+The trigger_fetch function takes a stream denoting the current time-since-reset in microseconds and returns a stream of the current active trigger.
+Each trigger also has a time mark, which denotes when it should start.
+To compute the index, we use the count_when function, so that it increments when the \emph{previous} trigger's time is in the past.
+We then use the array_index function to look up the new trigger in the constant array of triggers.
 
-Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match are part of the \fstar{} meta-language.
-To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:
-
-\newcommand\bibo{\mbox{bibo}}
-\newcommand\fir{\mbox{fir}}
-\newcommand\ii{\textit{input}}
-\newcommand\oo{\textit{output}}
-\newcommand\ok{\textit{ok}}
+Finally, we need to prove that trigger_fetch's contract instantiation is valid.
+This proof script is much the same as the previous one:
 \begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $\mbox{fir}~[0.7; 0.3] \ii$
+  @MM@\= @MM@ \= \kill
+  @let@ $\mbox{trigger_fetch}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\NN$ $\to$ stream trigger = \\
+    \> @let@ $\textit{system} = \mbox{System.translate}_1~\text{trigger_fetch}$ @in@ \\
+    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
+    \> $@bless@_1$ trigger_fetch
 \end{tabbing}
 
-Normalising this definition evaluates away all of the lists.
-The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:
+Since the implementation of trigger_fetch used $\text{count_valid}_{\PSValid}$ instead of count_valid, the proof did not need to re-prove the property about the range of count.
 
-\begin{tabbing}
-    @MM@\= @MMMMMM@ \= \kill
-      @let@ $\mbox{fir}_2$ ($\ii$: stream $\RR$): stream $\RR$ = \\
-    \> $(0.7 \cdot \ii) + (\xfby{0}{((0.3 \cdot \ii) + (\xfby{0}{0}))})$
-\end{tabbing}
+We have only proved a simple indexing property here but have seen the important pieces of checks, contracts and deferred proofs.
+In \autoref{s:evaluation} we will extend this small example to the real system, which uses checked machine-integer arithmetic and proves that scheduling is never late.
 
-The properties that we want to state about reactive programs usually involve some temporal aspect.
-Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
-We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:
+% \subsection{Program logics and dependent types}
 
-\begin{tabbing}
-  @MM@\= @MMMMMM@ \= \kill
-  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
-    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
-\end{tabbing}
+% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
+% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.
 
-This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
-At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
-If there is no previous value, it defaults to true.
+% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
+% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
+% We could imagine specifying count_when with such a type:
 
-\subsection{Bounded input, bounded output}
-\label{ss:tut:bibo}
-We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
-% Properties like this can occur as an internal invariant in a larger system:
+% \begin{tabbing}
+%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
+% \end{tabbing}
 
-\begin{tabbing}
-  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
-  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\BB$ = \\
-  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
-\end{tabbing}
+% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.
 
-This property states that if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
-Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
-To prove that this property holds, we translate to a transition system and show that the stream is always true.
-In this case, induction over the transition relation is sufficient to prove the property.
-There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_2$ (n: $\RRgez$): Lemma (induct ($\bibo_2$ n)) = \\
-  \> @assert@ (base\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case ($\bibo_2 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-This definition uses \fstar{}'s \emph{lemma} syntax to state that the BIBO property holds inductively for any $n$.
-The two assertions prove the inductive cases separately, using our \emph{simplify} tactic to ensure that the translation to transition system is normalised away, and any translation artefacts are removed.
-
-If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient: the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
-One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
-We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
-
-\begin{tabbing}
-  @MM@\= @MMMMMMMMMMMMMMMMMMMMMMMMMMM@ \= \kill
-  @let@ $\mbox{proof}_3$ (n: $\RRgez$): Lemma (induct\_k 2 ($\bibo_3$ n)) = \\
-  \> @assert@ (base\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ()); \\
-  \> @assert@ (step\_case\_k 2 ($\bibo_3 ~n$)) \> @by@ (pipit\_simplify ())
-\end{tabbing}
-
-Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation to verify reactive systems.
+% We believe that such type-based approaches are useful for composing programs together, but 
+%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
+% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
+% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$

### 2023-12-20 2024pipit/section/30-Core.tex
ok

@@
 \input{figures/core-grammar.tex}
-\input{figures/core-typing.tex}
-\input{figures/core-bigstep.tex}
-\input{figures/core-check.tex}
+
+We now introduce the core Pipit language.
+Note that this form differs slightly from the surface syntax presented earlier in \autoref{s:motivation}, which used the syntax of the metalanguage \fstar{}, as well as including proofs in \fstar{} itself.
 
 The grammar of Pipit is defined in \autoref{f:core-grammar}.
-The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and applications ($e~e'$); however, it does not include any form for defining functions except reusing closed functions from the \fstar{} meta-language $(x)$.
-Most of the expression forms were introduced informally in \autoref{s:tut} and correspond to the clock-free primitives in Lustre \cite{caspi1995functional}.
-% The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
-Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
+The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and primitive applications ($p(\ov{e})$).
+Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre \cite{caspi1995functional}.
+
+Values, variables and primitives are standard.
+The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
+% Streams can also be composed together using the \emph{then} notation ($\xthen{e}{e'}$) which denotes that the value of stream $e$ is used for the first step, followed by the values from stream $e'$ for subsequent steps.
 
@@
 Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.
+Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a single tuple.
+
+Checked properties and contracts are annotated with their property status.
+
+We support a fixed set of values and primitives.
+
+Streams $V$ are represented as a non-empty sequence of values; streaming history environments $\Sigma$ are streams of heaps.
+Types $\tau$ and type environments $\Gamma$ are standard.
+
+\input{figures/core-typing.tex}
+
+We define the typing judgments for Pipit in \autoref{f:core-typing}.
+Most of the typing rules are fairly standard for an unclocked Lustre.
+The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
+This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type.
+For the core language, we assume that everything is a stream.
+
+For values, we use an auxiliary judgment form $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
+Likewise, for primitives we use the auxiliary judgment form $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
+Primitives are pure, non-streaming functions.
+
+Rules \textsc{TValue}, \textsc{TVar}, \textsc{TPrim} and \textsc{TLet} are standard.
+
+Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.
+
+Rule \textsc{TRec} states that a recursive stream $\xrec{x}{e}$ has the recursive stream bound inside $e$.
+The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \REF{}.
+
+Rule \textsc{TCheck} states that checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.
+
+Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\eimpl}{\eguar}$ with a body expression of some type $\tau$.
+The overall expression has result type $\tau$.
+Both rely and guarantee clauses must be boolean expressions.
+Additionally, the guarantee clause can refer to the result value by $x$.
+
+\subsection{Dynamic semantics}
+\input{figures/core-bigstep.tex}
+
+The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
+We present our semantics in a big-step form.
+This differs somewhat from traditional \emph{reactive} semantics of Lustre presented in \cite{caspi1995functional}.
+We believe that the big-step semantics is better for performing equational reasoning, as it is substitution-based, while the reactive semantics better emphasises the finite-state execution of the system.
+For reasoning about the finite-state execution, we instead use transition systems (\autoref{s:transition}).
+
+The judgment form $\bigstep{\Sigma}{e}{v}$ denotes that expression $e$ evaluates to value $v$ under streaming history $\Sigma$.
+The streaming history is a stream of heaps.
+
+\subsection{Checked semantics}
+\input{figures/core-check.tex}
+
+We define the checked semantics of Pipit in \autoref{f:core-check}.
+\TODO{}
+
+\TODO{Define bless function}

### 2023-12-20 2024pipit/section/35-Primitive-Tables.tex
ok

\subsection{Primitive tables}
\label{s:core:primitive-tables}

The core language presented in \autoref{f:core-grammar} has a restricted set of primitives.
Our actual implementation is parameterised by the set of primitive types and operations.

In addition to a standard deeply-embedded primitive table which supports integers, tuples, booleans and reals, we also provide a shallowly-embedded primitive table, where operations can be arbitrary pure \fstar{} functions.
This shallow embedding also allows straightforward integration of \fstar{} types and functions, as well as proofs inside streaming functions.

This parameterisation serves mainly as a convenience.
For the formalisation presented here, we assume a simple static set of primitives.

In our current implementation, we cannot encode primitives that have preconditions, as the preconditions are checked eagerly by \fstar{}, before anything is known about the surrounding streaming values.

To work around this limitation, we encode certain primitive preconditions as implications on the postcondition.
For example, to encode the array indexing operation in \autoref{s:motivation}, we use an unsafe variant of the indexing function:
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @val@ Array.index_unsafe ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: $\NN$): \\
  \> $v: \alpha \{ \textit{index} < \mbox{Array.length}~\textit{array} \implies v = \mbox{Array.index}~\textit{array}~\textit{index} \}$
\end{tabbing}

By wrapping this unsafe index function inside a contract that requires the index to be valid, we can prove the postcondition on the streaming level.

In the future, we plan to investigate allowing preconditions on non-streaming functions, so that safe variants can be used directly.

### 2024-01-04 2024pipit/section/20-Motivation.tex
ok

@@
 
-As a concrete example to introduce Pipit, we consider a program with a static schedule of \emph{triggers}, or actions to be performed at a particular time.
-These triggers are stored in a constant array.
-The program maintains an index that refers to the current active trigger; at each instant in time, the program checks if the active trigger has expired, and if so, it increments the index.
-This trigger logic is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
+To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
+In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
+The triggers are stored in a constant array, sorted by their associated time-mark.
+Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.
+
+The time-marks are strict deadlines on the order of microseconds, which prohibits the driver from looping through the whole array at each time step to find the next active trigger.
+Instead, the driver maintains an index that refers to the current trigger; at each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
+Multiple triggers can have the same time-mark, as long as they are active in different cycles; triggers that are active on the same cycle must have a sufficiently wide gap in their time-marks to allow the driver time to iterate through the inactive triggers between them and reach the next active trigger before it starts.
+The implementation itself is simple, but formalising exactly what constitutes a ``sufficiently wide gap'' and proving that the schedule is never late requires care.
 
@@
 The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
-The syntax $0~@fby@~\textit{count}$ can be read as the initial value of zero \emph{followed by} the previous count.
+The syntax $0~@fby@~\textit{count}$ can be read as ``the initial value of zero \emph{followed by} the previous count''.
 
@@
 In practice, the user does not explicitly specify property statuses in the source language.
-The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true.
-Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are impliciitly lifted to constant streams.
+The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true\footnote{We limit our scope to safety properties: in real-time systems, liveness properties tend to be easily restated as bounded liveness properties, which are a form of safety property \cite{manna2012temporal}}.
+Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are implicitly lifted to constant streams.
 
-We need to defer the proof of the property here, because at the point of stating the property we don't yet have a concrete definition for the count variable.
+We defer the proof of the property here because, at the point of stating the property inside the @rec@ combinator, we don't yet have a concrete definition for the count variable.
 In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the count and putting the @check@ outside of the @rec@ combinator.
 However, it is not always possible to defer property statements: for example, when calling other streaming functions that have their own preconditions, it may not be possible to move the function call outside of its enclosing @rec@.
+In such cases, it is necessary to defer the proof of the precondition.
 
-Pipit is an embedded domain-specific language. % we usually present Pipit programs using \fstar{} syntactic sugar rather than showing the desugared core program.
-The program above is really syntactic sugar for a \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable; we will discuss the details of the core language in \autoref{s:core}.
+Pipit is an embedded domain-specific language.
+The program above is really syntactic sugar for an \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable.
+We will discuss the details of the core language in \autoref{s:core}, but for now we will show the source program with some minor embedding details omitted.
 
-To actually prove the property above, we use \fstar{}'s tactics and proof automation to translate the program into a transition system and prove the property inductively on the system.
-Finally, we \emph{bless} the expression, which marks the properties as valid.
-Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the operational semantics.
+To actually prove the property above, we
+% use \fstar{}'s tactics and proof automation to
+translate the program into a transition system and prove the property inductively on the system.
+Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
+Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the runtime semantics.
 
@@
 
-The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one parameter.
-The pipit_simplify tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof automation can then solve the inductive check directly.
-This proof does require some boilerplate; in the future we hope to minimise this.
+The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one stream parameter.
+The pipit_simplify tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof-by-SMT can then solve the inductive check directly.
+% This proof does require some boilerplate; in the future we hope to minimise this.
 
-Callers of count_when can now use the valid variant without needing to re-prove the count-range property.
+Callers of count_when can now use the validated variant without needing to re-prove the count-range property.
 In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
@@
 As an embedded language, we also benefit from writing our programs in a rich metalanguage.
+
+The trigger-fetch logic requires such a combination of automatic and manual proofs.
+The trigger-fetch logic uses the count_when streaming function to define the index; the increment flag given to count_when will increment the index if the previous index has expired or is inactive in the current cycle.
+We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity, such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
+
+\begin{tabbing}
+  @MM@\= @MM@ \= @let index@ \= \kill
+  @let@ trigger_fetch ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
+    \> $@rec @ \textit{index}.$ \\
+    \> \> $@let @ \textit{inc} = \text{false} @ fby @ ((\text{time_mark}~\textit{cfg}~\textit{index} \le \textit{time}) \vee \neg (\text{enabled}~\textit{cfg}~\textit{index}~\textit{cycle})) @ in@$\\
+    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{trigger_count}~\textit{cfg})~\textit{inc} @ in@$ \\
+    \> \> $\xcheckP{\PSUnknown}{(\text{time_no_skips}~\textit{time} \implies \text{can_reach_next_active}~\textit{cfg}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
+    \> \> $\text{pose_forall}_{\PSValid}~(\forall \rawbind{i} \text{can_reach_next}~\textit{cfg}~\textit{cycle}~i)~\textit{index}$; \\
+    \> \> $\textit{index}$\\
+\end{tabbing}
+
+The implementation of trigger_fetch defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
+The static trigger configuration contains the constant triggers array, as well as proofs that the configuration is schedulable: for example, that the triggers are sorted by time-mark and have adequate gaps between their times.
+
+The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
+
+We state, with a checked property of unknown status, that an invariant holds for the current time and index.
+Informally, this invariant states that, if the time stream never misses any steps, then the next active trigger is in the future, and we can reach it in time if we increment the index at every time-step.
+
+Finally, we use the pose_forall combinator to instantiate a non-streaming proof with a streaming argument: given a proof of $\forall \rawbind{(x: \tau)} p~x$ and a stream $s: \stream{\tau}$, we can infer that $p$ holds for all elements of the stream $s$.
+We annotate the call with $\PSValid$ here to emphasise that the proof is not deferred, but we do not include this in practice.
+The pose_forall function is in fact just a helper function implemented using the check, transition-system and bless functions already seen.
+
+\subsection{Scratch}
 
@@
 The trigger_fetch function takes a stream denoting the current time-since-reset in microseconds and returns a stream of the current active trigger.
-Each trigger also has a time mark, which denotes when it should start.
+Each trigger also has a time-mark, which denotes when it should start.
 To compute the index, we use the count_when function, so that it increments when the \emph{previous} trigger's time is in the past.

### 2024-01-04 2024pipit/section/20-Motivation.tex
kill scratch

@@
 
-\subsection{Scratch}
-
-For the trigger fetch logic, we will use count_when as the index into the triggers array.
-To retrieve the element from the triggers array, we assume a pure, non-streaming \fstar{} function for array indexing.
-We use an unsafe indexing function, which we lift to a streaming primitive and wrap in a contract to make it safe:
-
 \begin{tabbing}
@@
 The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
-This is a somewhat pathological contract, as the guarantee does not abstract away the details of the implementation at all.
-
-We use the unsafe indexing function, rather than a safe variant with a precondition, as we cannot currently lift primitives with preconditions; we discuss future work to remove this limitation in \autoref{s:core:primitive-tables}.
-
-We can now combine the count_when and index stream functions to implement a (much simplified) time-triggered trigger fetch controller.
-
-\begin{tabbing}
-  @MM@\= @MM@ \= @let index@ \= \kill
-  @let@ trigger_fetch ($\textit{time}$: stream $\NN$): stream trigger = \\
-    \> $@rec @ \textit{trigger}.$ \\
-    \> \> $@let @ \textit{inc} = \text{false} @ fby @ (\textit{trigger.time_mark} < \textit{time}) @ in@$\\
-    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{Array.size}~\textit{trigger_array} - 1)~\textit{inc} @ in@$ \\
-    \> \> $\text{array_index}~\textit{trigger_array}~\textit{index}$\\
-\end{tabbing}
-
-The trigger_fetch function takes a stream denoting the current time-since-reset in microseconds and returns a stream of the current active trigger.
-Each trigger also has a time-mark, which denotes when it should start.
-To compute the index, we use the count_when function, so that it increments when the \emph{previous} trigger's time is in the past.
-We then use the array_index function to look up the new trigger in the constant array of triggers.
-
-Finally, we need to prove that trigger_fetch's contract instantiation is valid.
-This proof script is much the same as the previous one:
-\begin{tabbing}
-  @MM@\= @MM@ \= \kill
-  @let@ $\mbox{trigger_fetch}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\NN$ $\to$ stream trigger = \\
-    \> @let@ $\textit{system} = \mbox{System.translate}_1~\text{trigger_fetch}$ @in@ \\
-    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
-    \> $@bless@_1$ trigger_fetch
-\end{tabbing}
-
-Since the implementation of trigger_fetch used $\text{count_valid}_{\PSValid}$ instead of count_valid, the proof did not need to re-prove the property about the range of count.
-
-We have only proved a simple indexing property here but have seen the important pieces of checks, contracts and deferred proofs.
-In \autoref{s:evaluation} we will extend this small example to the real system, which uses checked machine-integer arithmetic and proves that scheduling is never late.
 

### 2024-01-05 2024pipit/section/20-Motivation.tex
trigger index

@@
   @MM@\= @MM@ \= @let index@ \= \kill
-  @let@ trigger_fetch ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
+  @let@ trigger_index ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
     \> $@rec @ \textit{index}.$ \\
@@
 
-The implementation of trigger_fetch defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
-The static trigger configuration contains the constant triggers array, as well as proofs that the configuration is schedulable: for example, that the triggers are sorted by time-mark and have adequate gaps between their times.
+The implementation of trigger_index defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
+The static trigger configuration contains the constant triggers array, as well as proofs that the configuration describes a valid schedule: for example, that the triggers are sorted by time-mark and have adequate gaps between their times.
 
@@
 
-Finally, we use the pose_forall combinator to instantiate a non-streaming proof with a streaming argument: given a proof of $\forall \rawbind{(x: \tau)} p~x$ and a stream $s: \stream{\tau}$, we can infer that $p$ holds for all elements of the stream $s$.
+We then use the pose_forall combinator to instantiate a non-streaming proof with a streaming argument: given a proof of $(\forall \rawbind{(x: \tau)} p~x)$ and a stream $(s: \stream{\tau})$, we can infer that $p$ holds for all elements of the stream $s$.
 We annotate the call with $\PSValid$ here to emphasise that the proof is not deferred, but we do not include this in practice.
 The pose_forall function is in fact just a helper function implemented using the check, transition-system and bless functions already seen.
+The property we state here is that if the trigger at the current index is active, then there is a sufficient time gap to be able to reach the next active trigger after the current index.
+
+After explicitly instantiating this lemma, the streaming invariant can be proved by induction on the transition system; the proof script is almost identical to that of count_when.
 

### 2024-01-05 2024pipit/section/20-Motivation.tex
example, core language mostly

@@
     \> @let@ $\textit{system} = \mbox{System.translate}_1 (\mbox{count_when}~\textit{max})$ @in@ \\
-    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
+    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit_simplify ()); \\
     \> $@bless@_1$ $(\mbox{count_when}~\textit{max})$
@@
   @MM@\= @MM@ \= @let index@ \= \kill
-  @let@ trigger_index ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
+  @let@ trigger_fetch ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
     \> $@rec @ \textit{index}.$ \\
@@
     \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{trigger_count}~\textit{cfg})~\textit{inc} @ in@$ \\
-    \> \> $\xcheckP{\PSUnknown}{(\text{time_no_skips}~\textit{time} \implies \text{can_reach_next_active}~\textit{cfg}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
+    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cfg}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
+    % \> \> $\xcheckP{\PSUnknown}{(\text{time_no_skips}~\textit{time} \implies \text{can_reach_next_active}~\textit{cfg}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
     \> \> $\text{pose_forall}_{\PSValid}~(\forall \rawbind{i} \text{can_reach_next}~\textit{cfg}~\textit{cycle}~i)~\textit{index}$; \\
@@
 
-The implementation of trigger_index defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
+The implementation of trigger_fetch defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
 The static trigger configuration contains the constant triggers array, as well as proofs that the configuration describes a valid schedule: for example, that the triggers are sorted by time-mark and have adequate gaps between their times.
@@
 We state, with a checked property of unknown status, that an invariant holds for the current time and index.
-Informally, this invariant states that, if the time stream never misses any steps, then the next active trigger is in the future, and we can reach it in time if we increment the index at every time-step.
+Informally, this invariant states that the next active trigger at or after the current index is now or in the future, and we can reach it in time if we increment the index at every time-step.
 
@@
 
-After explicitly instantiating this lemma, the streaming invariant can be proved by induction on the transition system; the proof script is almost identical to that of count_when.
+After explicitly instantiating this lemma, we prove the streaming invariant by induction on the transition system.
+To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for trigger_fetch.
+The contract we state is that if the environment always ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that whenever the index points to a valid, active trigger, then the trigger has not yet expired.
+Note that this contract is not a full functional specification.
 
 \begin{tabbing}
-  @MM@\= @MM@ \= \kill
-  @let@ array_index ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: stream $\NN$): stream $\alpha$ = \\
-  \> $@contract@_{\PSUnknown}$ \\
-  \> \> \{ $0 \le \textit{index} < \mbox{Array.length}~\textit{array}$ \} \\
-  \> \> Array.index_unsafe \textit{array} \textit{index} \\
-  \> \> \{ $v.~v~=~\mbox{Array.index_unsafe}~\textit{array}~\textit{index}$ \} \\
+  @MM@\= @MM@ \= @MMMMMM@ \= \kill
+  @let@ $\text{trigger_fetch}_{\PSValid}$ ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
+  \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
+  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} )$ \\
+  \> \> @guar@ = $(\lambda \textit{time index}.~ \text{index_valid}~\textit{cfg}~\textit{index} \wedge \text{enabled}~\textit{cfg}~\textit{index}~\textit{cycle})$ \\
+  % \> \> @guar@ = $(\lambda \textit{time index}.~ (\textit{index} < (\text{trigger_count}~\textit{cfg}) \wedge \text{enabled}~\textit{cfg}~\textit{index}~\textit{cycle})$ \\
+  \> \> \> $\implies \text{time_mark}~\textit{cfg}~\textit{index} \ge \textit{now})$ \\
+  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cfg}~\textit{cycle}~\textit{time} )$ \\
+  \> \} @in@ \\
+  \> @assert@ $(\text{Contract.inductive_check}~\textit{contract})$ @by@ (pipit_simplify ()); \\
+  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
 \end{tabbing}
 
-Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
-The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
-In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
-The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
+In the implementation of the validated variant of trigger_fetch, we first construct the contract from streaming functions.
+The $\text{Contract.contract_of_stream}_1$ combinator describes a contract with one input (the time stream), and takes stream transformers for each of the rely, guarantee and body.
+The combinator transforms the surface syntax into core expressions.
+The assertion $(\text{Contract.inductive_check}~\textit{contract})$ then translates the expressions into a transition system, and checks that if the rely always holds then the guarantee always holds, and that the as-yet-unchecked subproperties hold.
+Finally, $\text{Contract.stream_of_contract}_1$ blesses the core expression and converts it back to a stream transformer, so it can be easily used by other parts of the program.
 
-% \subsection{Program logics and dependent types}
+When this function is used in other parts of the program, the caller must ensure that the environment satisfies the rely.
+In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.
+Deferring the proof of the rely clause is crucial for function calls where satisfying the precondition at a particular point in time somehow depends on the value of the output at a previous step.
+These kinds of self-dependencies and mutual-dependencies are fairly common in Lustre-style programs: for example, the trigger_fetch's invocation of count_when, whose input $\textit{inc}$ depends on the previous output $\textit{index}$.
+
+\subsection{Program logics and dependent types}
+\TODO{compare to refinement types, verification conditions, etc here, or wait to \autoref{s:related-work}?}
 

### 2024-01-05 2024pipit/section/30-Core.tex
ok
example, core language mostly

@@
 
-Checked properties and contracts are annotated with their property status.
+Checked properties and contracts are annotated with their property status $\PStatus$, which can either be valid ($\PSValid$) or unknown ($\PSUnknown$).
+For checked properies $\xcheckP{\PStatus}{e}$, the property status denotes whether the property has been proved to be valid.
 
-We support a fixed set of values and primitives.
+Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
+Firstly, when a contract is \emph{defined}, the definer must prove that the body $\ebody$ satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
+Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
+Conceptually, then, a contract could have two property statuses: one for the definer, and one for the instantiation.
+However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
+For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.
+
+For the presentation of the formal grammar, we consider only a fixed set of values and primitives.
 
@@
 
-Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\eimpl}{\eguar}$ with a body expression of some type $\tau$.
+Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
 The overall expression has result type $\tau$.
@@
 
+Rule {\sc Value} states that evaluating a value results in the value itself.
+
+Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.
+
+Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then use the prim-sem metafunction to actually apply the primitive.
+
+Rule $\textsc{Fby}_1$ evaluates a followed-by expression when the streaming history contains only a single element.
+Here, $\xfby{v}{e}$ evaluates to $v$, as there is no previous value of $e$ to use.
+
+Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
+In this case, $\xfby{v}{e}$ proceeds to evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.
+
+Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
+For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding will eventually terminate, as the follow-by shortens the streaming history.
+
+Rule {\sc Let} is standard.
+
+Rule {\sc Check} states that check expressions always evaluate to unit.
+We do not perform a dynamic check that the property is true here; checking the truth of properties is dealt with in the checked semantics (\autoref{s:core:checked}).
+
+Rule {\sc Contract} states that contracts evaluate by just evaluating their body.
+Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
+This ensures that the dynamic semantics correspond directly to the program we intend to execute.
+
+We also use two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.
+
+Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under streaming history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
+This judgment is an iterated application of the single-value big-step form.
+
+Judgment form $\bigstepalways{\Sigma}{e}$ denotes that expression $e$, which must be a boolean, evaluates to the stream of trues under history $\Sigma$.
+Informally, it can be read as ``in streaming history $\Sigma$, $e$ is always true''.
+
 \subsection{Checked semantics}
+\label{s:core:checked}
 \input{figures/core-check.tex}
 
+In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
+We call these the \emph{checked} semantics.
+Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.
+
+The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties of $e$ with status $\PStatus$ hold.
+The property status dictates which properties should be checked and which should be ignored.
+
+To show that an expression $e$'s unknown properties hold, we prove that for all streaming histories $\Sigma$, assuming the valid properties hold ($\semcheck{\Sigma}{\PSValid}{e}$), then the unknown properties ($\semcheck{\Sigma}{\PSUnknown}{e}$) hold.
+
+To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
+Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
+The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
+
+% and the unknown properties in the body and guarantee both hold ($\semcheck{\Sigma}{\PSUnknown}{\ebody} \wedge \semcheck{\Sigma}{\PSUnknown}{\eguar[x := \ebody]}$).
+% and assuming that the valid properties in the rely, body and guarantee hold ($\semcheck{\Sigma}{\PSValid}{\erely} \wedge \semcheck{\Sigma}{\PSValid}{\ebody} \wedge \semcheck{\Sigma}{\PSValid}{\eguar[x := \ebody]}$)
+
+To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
+Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
+This sort of feedback is necessary for proving properties of mutually-dependent calls.
+Although this feedback appears circular, we enforce causality by requiring that occurrences of recursive streams are guarded by delays (\autoref{s:core:causality}).
+
 We define the checked semantics of Pipit in \autoref{f:core-check}.
-\TODO{}
+The checked semantics mostly follows the structure of the dynamic semantics, checking any properties and contracts as they are encountered.
 
-\TODO{Define bless function}
+Rules {\sc ChkValue} and {\sc ChkVar} state that values and variables are always valid.
+
+Rule {\sc ChkPrim} checks a primitive application by descending into the subexpressions.
+
+Rules $\textsc{ChkFby}_1$ and $\textsc{ChkFby}_S$ follow the structure of the big-step rules  $\textsc{Fby}_1$ and $\textsc{Fby}_S$.
+At the start of the stream, $\textsc{ChkFby}_1$ checks the initial value.
+At subsequent parts of the stream, $\textsc{ChkFby}_S$ discards the most recent stream history and checks the subexpression.
+
+Rules {\sc ChkRec} and {\sc ChkLet} are similar, performing the same unfolding as the corresponding big-step rules, then checking the resulting expression.
+
+Finally, rules {\sc ChkCheck} and {\sc ChkContract} perform the real work.
+
+Rule {\sc ChkCheck} applies when checking property status $\PStatus$ of an expression $\xcheckP{\PStatus'}{e}$.
+If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we perform the actual check by evaluating the expression $e$ and requiring it to evaluate to a stream of trues.
+Otherwise, we ignore the check-expression.
+In both cases, we descend into the expression and check its subexpressions, as they may have nested properties.
+
+Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
+Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
+To check the caller property when $\PStatus = PStatus'$, we evaluate the rely $\erely$ and require it to be true.
+To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
+We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
+
+\subsubsection{Blessing}
+\TODO{blessing}
+
+\subsection{Causality}
+\label{s:core:causality}
+
+\TODO{causality}

### 2024-01-05 2024pipit/section/70-Related.tex
ok
example, core language mostly

@@
 
-Although the FIR filter from \autoref{s:tut} is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
+Although the FIR filter from XXX is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
 Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
@@
 However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
+
+\TODO{Verified compilation via Frama-C} \cite{brun2023equation}
 
@@
 Explicitly specifying an invariant for each recursive stream would be cumbersome; deferring the proof allows such invariants to be implicit.
+
+\subsection{Counterexamples and invariant generation}
+\TODO{integration with model-checker}

### 2024-01-08 2024pipit/section/40-Transition.tex
ok

@@
 \input{figures/system-translation.tex}
+
+\TODO{describe translation}
+The substitution-based semantics is better for equational reasoning, and (subjectively perhaps) simpler.
+The transition system is better for certain automated proofs: a surprising number of interesting properties can be proven with k-induction alone.
+
+\subsection{Translation correctness proofs}
+
+\TODO{describe the proofs of translation correctness}
+The translation to a transition system above is an abstraction.
+We prove it.
+Interesting lemmas etc.

### 2024-01-08 2024pipit/section/50-Extraction.tex
ok

@@
 % This allows us to generate C code which executes in bounded space on embedded devices.
+
+\TODO{}
+\begin{itemize}
+  \item C codegen: note about duplicating work, sketch potential solution;
+  \item C codegen: future work: sketch potential solution for separate compilation, compare to Accelerate FFI \cite{clifton2014embedding};
+  \item staging: similarity to \cite{gallego2021w};
+  \item this extraction is very similar to the transition system translation, except translating contracts uses the body and ignores the rely and guarantee;
+  \item talk about proof of correctness
+\end{itemize}

### 2024-01-08 2024pipit/section/70-Related.tex
ok

@@
 \label{s:related-work}
+
 % TODO arrays
+\TODO{requires rewrite}
 
@@
 
-\subsection{Counterexamples and invariant generation}
+\subsection{Model-checking, counterexamples and invariant generation}
 \TODO{integration with model-checker}
+Model-checkers do have distinct advantages over interactive theorem-provers: they are more usable, can find strengthening invariants automatically, and provide counterexamples.
+However, they are limited to comparatively simple properties.
+The time-triggered example from \autoref{s:motivation} would be difficult to encode in a model-checker with a limited specification language, as the trigger invariant uses an auxiliary recursive function to find the next active trigger.
+
+In future work, we would like to explore integrating Pipit with an existing model-checker.
+For simple first-order nonrecursive programs and specifications that fit within the bounds of another model-checker, we could generate an encoding to use for counterexample generation.
+We would also like to investigate automatically lifting the strengthened invariants and suggesting the corresponding Pipit invariant.
+Such a translation and lifting would not form part of the trusted computing base, as generated invariants would still be proved separately in \fstar{}.
+In this way, we hope that it may be possible to obtain some of the usability benefits of model-checkers, with the added power of manual proofs when necessary, while retaining trust that the compiled code satisfies the original specification.

### 2024-01-08 2024pipit/section/01-Introduction.tex
ok

@@
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.
+Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; we also prove that the imperative code is a refinement of the verified transition system.
 
@@
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
-  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\TODO{ref});
-  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified contracts (\autoref{s:transition});
-  \item we identify the invariants and lemmas required to prove preservation of both dynamic (bigstep) semantics and checked semantics, required to transform properties proven on transition systems to the original program (\TODO{ref});
+  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\autoref{f:core-check});
+  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
+  \item we identify the invariants and lemmas required to prove preservation of the dynamic (bigstep) semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
+  \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
   \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).

### 2024-01-10 2024pipit/section/01-Introduction.tex
time-triggered example

@@
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; we also prove that the imperative code is a refinement of the verified transition system.
+Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics; we also prove that the imperative code preserves the semantics.
 
@@
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
-  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their property lifted to valid status (\autoref{f:core-check});
+  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
   \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});

### 2024-01-10 2024pipit/section/20-Motivation.tex
time-triggered example

--- 2024-01-04 2024pipit/section/20-Motivation.tex
@@
 
-\section{Time-triggered}
+\section{Pipit for time-triggered networks}
 \label{s:motivation}
 
-Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
-Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
-Like imperative programs, they describe how state evolves over time.
-Additionally, Lustre programs tend to be composed of many nested recursive loops.
-When verifying both pure and imperative programs, recursion or loops generally require explicit loop invariants.
-Requiring an explicit invariant for each recursive stream quickly becomes unwieldy for Lustre programs.
+% Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
+% Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
+% Like imperative programs, they describe how state evolves over time.
+% Additionally, Lustre programs tend to be composed of many nested recursive loops.
+% When verifying both pure and imperative programs, recursion or loops generally require explicit loop invariants.
+% Requiring an explicit invariant for each recursive stream quickly becomes unwieldy for Lustre programs.
 
 To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
-In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
+% In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
 The triggers are stored in a constant array, sorted by their associated time-mark.
-Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.
+% Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.
 
-The time-marks are strict deadlines on the order of microseconds, which prohibits the driver from looping through the whole array at each time step to find the next active trigger.
-Instead, the driver maintains an index that refers to the current trigger; at each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
-Multiple triggers can have the same time-mark, as long as they are active in different cycles; triggers that are active on the same cycle must have a sufficiently wide gap in their time-marks to allow the driver time to iterate through the inactive triggers between them and reach the next active trigger before it starts.
-The implementation itself is simple, but formalising exactly what constitutes a ``sufficiently wide gap'' and proving that the schedule is never late requires care.
+\subsection{Deferring and proving properties}
 
+The driver maintains an index that refers to the current trigger.
+At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
 We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
@@
     \> $@rec@~\rawbind{\textit{count}}{}$ \\
+    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
     \> \> @let@ $\textit{count'}$ @=@ $(0~@fby@~\textit{count}) + (@if@~\textit{inc}~@then@~1~@else@~0)$ @in@ \\
-    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
-    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$ \\
+    \> \> @if @ $\textit{count'} \ge \textit{max}$ @then@ $\textit{max}$  @else@ $\textit{count'}$
 \end{tabbing}
 
-The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which in turn defines a local incremented stream \emph{count'}.
-The syntax $0~@fby@~\textit{count}$ can be read as ``the initial value of zero \emph{followed by} the previous count''.
+The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which states an invariant about the count before defining the incremented stream \emph{count'}.
+Inside \emph{count'}, the syntax $0~@fby@~\textit{count}$ is read as ``the initial value of zero \emph{followed by} the previous count''.
 
@@
 In practice, the user does not explicitly specify property statuses in the source language.
-The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true\footnote{We limit our scope to safety properties: in real-time systems, liveness properties tend to be easily restated as bounded liveness properties, which are a form of safety property \cite{manna2012temporal}}.
+The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true.
+% \footnote{We limit our scope to safety properties: in real-time systems, liveness properties tend to be easily restated as bounded liveness properties, which are a form of safety property \cite{manna2012temporal}}
 Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are implicitly lifted to constant streams.
@@
 We defer the proof of the property here because, at the point of stating the property inside the @rec@ combinator, we don't yet have a concrete definition for the count variable.
-In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the count and putting the @check@ outside of the @rec@ combinator.
+In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the recursive count and putting the @check@ outside of the @rec@ combinator.
 However, it is not always possible to defer property statements: for example, when calling other streaming functions that have their own preconditions, it may not be possible to move the function call outside of its enclosing @rec@.
-In such cases, it is necessary to defer the proof of the precondition.
+% In such cases, it is necessary to defer the proof of the precondition.
 
@@
 
-To actually prove the property above, we
-% use \fstar{}'s tactics and proof automation to
-translate the program into a transition system and prove the property inductively on the system.
+To actually prove the property above, we use the meta-language \fstar{}'s tactics to translate the program into a transition system and prove the property inductively on the system.
 Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
@@
 In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
-By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we also reduce the semantic gap between compilation and proofs, as well as allowing manual proof steps when the automated proofs get stuck.
-As an embedded language, we also benefit from writing our programs in a rich metalanguage.
+By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we have increased confidence that the compiled code matches the verified code, as well as allowing manual proof steps when the automated proofs get stuck.
+% As an embedded language, we also benefit from writing our programs in a rich metalanguage.
 
-The trigger-fetch logic requires such a combination of automatic and manual proofs.
+\subsection{The time-triggered system matrix}
+
+The schedule of the time-triggered network is conceptually described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
+Each basic cycle is identified by its cycle index and each transmission column has an associated time-mark.
+
+\autoref{f:tt-systemmatrix-ok} (left) shows an example system matrix with cycles @C0@ and @C1@ and transmission columns at time-marks 0, 1 and 2.
+For this example, we assume that one message can be sent per clock cycle.
+To execute this system matrix, we synchronise the local time to zero at the start of basic cycle @C0@.
+After a basic cycle completes, the nodes on the network synchronise before execution continues to the next basic cycle.
+
+\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding configuration for the triggers array.
+The enabled set denotes the basic cycles for which a trigger is active.
+
+The system has strict timing requirements, which restrict how triggers can be defined.
+In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
+Additionally, the schedule must allow sufficient time for the driver to skip over any disabled triggers.
+Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2 instead.
+However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per time-mark, and it takes at least two steps to reach trigger 2 from the start of the array.
+
+\begin{figure}
+  \begin{minipage}{0.38\textwidth}
+\begin{tabular}{r|ccc}
+   & TM0 & TM1 & TM2 \\
+  \hline
+  C0 & MSG A & MSG B & - \\
+  C1 & MSG A & - & MSG C
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.6\textwidth}
+\small
+\begin{verbatim}
+  0: { time_mark = 0; enabled = {C0,C1}; msg = A; }
+  1: { time_mark = 1; enabled = {C0};    msg = B; }
+  2: { time_mark = 2; enabled = {C1};    msg = C; }
+\end{verbatim}
+\end{minipage}
+
+\caption{Left: system matrix; right: corresponding triggers array configuration}
+\label{f:tt-systemmatrix-ok}
+\end{figure}
+
+% The time-marks are strict deadlines on the order of microseconds, which prohibits the driver from looping through the whole array at each time step to find the next active trigger.
+% Instead, the driver maintains an index that refers to the current trigger; at each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
+% Multiple triggers can have the same time-mark, as long as they are active in different cycles; triggers that are active on the same cycle must have a sufficiently wide gap in their time-marks to allow the driver time to iterate through the inactive triggers between them and reach the next active trigger before it starts.
+% The implementation itself is simple, but formalising exactly what constitutes a ``sufficiently wide gap'' and proving that the schedule is never late requires care.
+
+We impose three main restrictions on the triggers array: it must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
+
+With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
+The proof of this lemma is straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
+
+% \begin{figure}
+%   \begin{minipage}{0.38\textwidth}
+% \begin{tabular}{r|c}
+%    & TM0 \\
+%   \hline
+%   C0 & MSG A \\
+%   C1 & MSG B
+% \end{tabular}
+% \end{minipage}
+% \begin{minipage}{0.6\textwidth}
+%   \small
+%   \begin{verbatim}
+%     { time_mark =  0; enabled = {C0}; msg = A; }
+%     { time_mark =  0; enabled = {C1}; msg = B; }
+%   \end{verbatim}
+%   \end{minipage}
+
+% \caption{Invalid trigger configuration}
+% \label{f:tt-systemmatrix-bad}
+% \end{figure}
+
+\subsection{Instantiating lemmas and defining contracts}
+
+% The trigger-fetch logic requires such a combination of automatic and manual proofs.
+We can now implement the trigger-fetch logic, which maintains an index that refers to the next-scheduled trigger.
 The trigger-fetch logic uses the count_when streaming function to define the index; the increment flag given to count_when will increment the index if the previous index has expired or is inactive in the current cycle.
-We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity, such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
+We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
 
@@
   @MM@\= @MM@ \= @let index@ \= \kill
-  @let@ trigger_fetch ($\textit{cfg}$: triggers) ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
+  @let@ trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
     \> $@rec @ \textit{index}.$ \\
-    \> \> $@let @ \textit{inc} = \text{false} @ fby @ ((\text{time_mark}~\textit{cfg}~\textit{index} \le \textit{time}) \vee \neg (\text{enabled}~\textit{cfg}~\textit{index}~\textit{cycle})) @ in@$\\
-    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{trigger_count}~\textit{cfg})~\textit{inc} @ in@$ \\
-    \> \> $\xcheckP{\PSUnknown}{(\text{time_no_skips}~\textit{time} \implies \text{can_reach_next_active}~\textit{cfg}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
-    \> \> $\text{pose_forall}_{\PSValid}~(\forall \rawbind{i} \text{can_reach_next}~\textit{cfg}~\textit{cycle}~i)~\textit{index}$; \\
+    \> \> $@let @ \textit{inc} = \text{false} @ fby @ ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) @ in@$\\
+    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} @ in@$ \\
+    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index}$); \\
+    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
     \> \> $\textit{index}$\\
@@
 
-The implementation of trigger_fetch defines a streaming function which takes a static trigger configuration, the cycle index (static for now), and the stream denoting the current time.
-The static trigger configuration contains the constant triggers array, as well as proofs that the configuration is schedulable: for example, that the triggers are sorted by time-mark and have adequate gaps between their times.
+The implementation of \emph{trigger_fetch} defines a streaming function which takes a static cycle index and the stream denoting the current time.
+The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
+We allow the index to go one past the end of the array to denote that there are no more triggers, although this situation is unlikely to occur in practice.
 
-The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
+We use the $\textit{pose}$ combinator to instantiate the \emph{lemma_can_reach_next} lemma in the streaming context.
+% The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
+The $\textit{pose}$ function is a helper function implemented using the @check@ and @bless@ functions from earlier.
+% The property we state here is that if the trigger at the current index is active, then there is a sufficient time gap to be able to reach the next active trigger after the current index.
 
-We state, with a checked property of unknown status, that an invariant holds for the current time and index.
-Informally, this invariant states that, if the time stream never misses any steps, then the next active trigger is in the future, and we can reach it in time if we increment the index at every time-step.
+We then state an invariant as a deferred property.
+Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.
 
-Finally, we use the pose_forall combinator to instantiate a non-streaming proof with a streaming argument: given a proof of $\forall \rawbind{(x: \tau)} p~x$ and a stream $s: \stream{\tau}$, we can infer that $p$ holds for all elements of the stream $s$.
-We annotate the call with $\PSValid$ here to emphasise that the proof is not deferred, but we do not include this in practice.
-The pose_forall function is in fact just a helper function implemented using the check, transition-system and bless functions already seen.
+With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
+To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
+The contract we state is that if the environment always ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
 
 \begin{tabbing}
-  @MM@\= @MM@ \= \kill
-  @let@ array_index ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: stream $\NN$): stream $\alpha$ = \\
-  \> $@contract@_{\PSUnknown}$ \\
-  \> \> \{ $0 \le \textit{index} < \mbox{Array.length}~\textit{array}$ \} \\
-  \> \> Array.index_unsafe \textit{array} \textit{index} \\
-  \> \> \{ $v.~v~=~\mbox{Array.index_unsafe}~\textit{array}~\textit{index}$ \} \\
+  @MM@\= @MM@ \= @MMMMMMMMMMMMM@ \= \kill
+  @let@ $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
+  \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
+  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} )$ \\
+  \> \> @guar@ = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
+  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{now})$ \\
+  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} )$ \\
+  \> \} @in@ \\
+  \> @assert@ $(\text{Contract.inductive_check}~\textit{contract})$ @by@ (pipit_simplify ()); \\
+  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
 \end{tabbing}
 
-Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
-The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
-In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
-The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
+In the implementation of the validated variant of trigger_fetch, we first construct the contract from streaming functions.
+The $\text{Contract.contract_of_stream}_1$ combinator describes a contract with one input (the time stream), and takes stream transformers for each of the rely, guarantee and body.
+The combinator transforms the surface syntax into core expressions.
+The assertion $(\text{Contract.inductive_check}~\textit{contract})$ then translates the expressions into a transition system, and checks that if the rely always holds then the guarantee always holds, and that the as-yet-unchecked subproperties hold.
+Finally, $\text{Contract.stream_of_contract}_1$ blesses the core expression and converts it back to a stream transformer, so it can be easily used by other parts of the program.
+
+When this function is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
+In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.
+% Deferring the proof of the rely clause is crucial for function calls where satisfying the precondition at a particular point in time somehow depends on the value of the output at a previous step.
+% These kinds of self-dependencies and mutual-dependencies are fairly common in Lustre-style programs: for example, \emph{trigger_fetch}'s invocation of \emph{count_when}, whose input $\textit{inc}$ depends on the previous output $\textit{index}$.
 
 % \subsection{Program logics and dependent types}
+% \TODO{compare to refinement types, verification conditions, etc here, or wait to \autoref{s:related-work}?}
 

### 2024-01-10 2024pipit/section/20-Motivation.tex
work on time-triggered

@@
 % In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
-The triggers are stored in a constant array, sorted by their associated time-mark.
 % Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.
@@
 
-The driver maintains an index that refers to the current trigger.
+The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark; the driver maintains an index that refers to the current trigger.
 At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
 We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
-At each time step, count_when checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
+At each time step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
 
@@
 The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one stream parameter.
-The pipit_simplify tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof-by-SMT can then solve the inductive check directly.
+The \emph{pipit_simplify} tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof-by-SMT can then solve the inductive check directly.
 % This proof does require some boilerplate; in the future we hope to minimise this.
 
-Callers of count_when can now use the validated variant without needing to re-prove the count-range property.
+Callers of \emph{count_when} can now use the validated variant without needing to re-prove the count-range property.
 In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
-By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we have increased confidence that the compiled code matches the verified code, as well as allowing manual proof steps when the automated proofs get stuck.
+By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we have increased confidence that the compiled code matches the verified code and, as we shall see, we also have access to a rich specification language.
 % As an embedded language, we also benefit from writing our programs in a rich metalanguage.
@@
 
-The schedule of the time-triggered network is conceptually described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
+The schedule of the time-triggered network is abstractly described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
 Each basic cycle is identified by its cycle index and each transmission column has an associated time-mark.
@@
 \autoref{f:tt-systemmatrix-ok} (left) shows an example system matrix with cycles @C0@ and @C1@ and transmission columns at time-marks 0, 1 and 2.
-For this example, we assume that one message can be sent per clock cycle.
+For this example, we assume that one message can be sent per clock cycle. % , though in practice a message could take on the order of ten microseconds
 To execute this system matrix, we synchronise the local time to zero at the start of basic cycle @C0@.
@@
 Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2 instead.
-However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per time-mark, and it takes at least two steps to reach trigger 2 from the start of the array.
+However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per time-mark, and it takes two steps to reach trigger 2 from the start of the array.
 
@@
 With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
-The proof of this lemma is straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
+These properties are fairly straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
 
@@
 % The trigger-fetch logic requires such a combination of automatic and manual proofs.
-We can now implement the trigger-fetch logic, which maintains an index that refers to the next-scheduled trigger.
-The trigger-fetch logic uses the count_when streaming function to define the index; the increment flag given to count_when will increment the index if the previous index has expired or is inactive in the current cycle.
+We can now implement the trigger-fetch logic, which maintains the index that refers to the next-scheduled trigger.
+The trigger-fetch logic uses the \emph{count_when} streaming function to define the index; the increment flag given to count_when will increment the index if the previous index has expired or is inactive in the current cycle.
 We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
@@
     \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} @ in@$ \\
-    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index}$); \\
+    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
     \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
@@
 
-The implementation of \emph{trigger_fetch} defines a streaming function which takes a static cycle index and the stream denoting the current time.
+The \emph{trigger_fetch} function takes a static cycle index and the stream denoting the current time.
 The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
-We allow the index to go one past the end of the array to denote that there are no more triggers, although this situation is unlikely to occur in practice.
+We allow the index to go one past the end of the array to denote that there are no more triggers.
 
-We use the $\textit{pose}$ combinator to instantiate the \emph{lemma_can_reach_next} lemma in the streaming context.
+We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 refers to the lemma taking one streaming argument (the index).
 % The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
-The $\textit{pose}$ function is a helper function implemented using the @check@ and @bless@ functions from earlier.
+% The $\textit{pose}$ function is a helper function implemented using the @check@ and @bless@ functions from earlier.
 % The property we state here is that if the trigger at the current index is active, then there is a sufficient time gap to be able to reach the next active trigger after the current index.
-
 We then state an invariant as a deferred property.
@@
 
-In the implementation of the validated variant of trigger_fetch, we first construct the contract from streaming functions.
+In the implementation of the validated variant of \emph{trigger_fetch}, we first construct the contract from streaming functions.
 The $\text{Contract.contract_of_stream}_1$ combinator describes a contract with one input (the time stream), and takes stream transformers for each of the rely, guarantee and body.

### 2024-01-10 2024pipit/section/30-Core.tex
ok
causality etc

@@
 
-The grammar of Pipit is defined in \autoref{f:core-grammar}.
+\autoref{f:core-grammar} defines the grammar of Pipit.
 The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and primitive applications ($p(\ov{e})$).
-Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre \cite{caspi1995functional}.
+Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre~\cite{caspi1995functional}.
 
-Values, variables and primitives are standard.
 The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.
@@
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
-We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
-Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
+% We use this form to define a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \cite{caspi1995functional} which is not syntax-directed.
+% Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
 Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a single tuple.
@@
 
-For the presentation of the formal grammar, we consider only a fixed set of values and primitives.
+% For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table \autoref{s:core:primitive-tables}.
 
@@
 We define the typing judgments for Pipit in \autoref{f:core-typing}.
-Most of the typing rules are fairly standard for an unclocked Lustre.
+Most of the typing rules are standard for an unclocked Lustre.
 The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
-This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type.
-For the core language, we assume that everything is a stream.
+This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type; for the core language, we instead assume that everything is a stream.
 
@@
 Rule \textsc{TRec} states that a recursive stream $\xrec{x}{e}$ has the recursive stream bound inside $e$.
-The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \REF{}.
+The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \autoref{s:core:causality}.
 
-Rule \textsc{TCheck} states that checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.
+Rule \textsc{TCheck} states that statically checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.
 
@@
 We present our semantics in a big-step form.
-This differs somewhat from traditional \emph{reactive} semantics of Lustre presented in \cite{caspi1995functional}.
-We believe that the big-step semantics is better for performing equational reasoning, as it is substitution-based, while the reactive semantics better emphasises the finite-state execution of the system.
-For reasoning about the finite-state execution, we instead use transition systems (\autoref{s:transition}).
+This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
+Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based, while the reactive semantics emphasises the finite-state streaming execution of the system.
+We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
 
 The judgment form $\bigstep{\Sigma}{e}{v}$ denotes that expression $e$ evaluates to value $v$ under streaming history $\Sigma$.
-The streaming history is a stream of heaps.
+The streaming history is a stream of heaps; in practice, we only evaluate expressions with a non-empty streaming history.
 
@@
 
-Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then use the prim-sem metafunction to actually apply the primitive.
+Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then use the prim-sem metafunction to apply the primitive.
 
@@
 Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
-This ensures that the dynamic semantics correspond directly to the program we intend to execute.
+% This ensures that the dynamic semantics correspond directly to the program we intend to execute.
 
@@
 To show that an expression $e$'s unknown properties hold, we prove that for all streaming histories $\Sigma$, assuming the valid properties hold ($\semcheck{\Sigma}{\PSValid}{e}$), then the unknown properties ($\semcheck{\Sigma}{\PSUnknown}{e}$) hold.
+The assumption here means that we do not have to re-check properties after proving them once.
 
+Contracts involve two proofs: one for the definition and one for the instantiation.
 To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
@@
 The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
+\TODO{WAFFLE? This is expanded on in \autoref{s:core:blessing}; kill here?}
 
@@
 
-Rules $\textsc{ChkFby}_1$ and $\textsc{ChkFby}_S$ follow the structure of the big-step rules  $\textsc{Fby}_1$ and $\textsc{Fby}_S$.
-At the start of the stream, $\textsc{ChkFby}_1$ checks the initial value.
-At subsequent parts of the stream, $\textsc{ChkFby}_S$ discards the most recent stream history and checks the subexpression.
+Rules $\textsc{ChkFby}_1$ and $\textsc{ChkFby}_S$ are derived from the structure of the big-step rules  $\textsc{Fby}_1$ and $\textsc{Fby}_S$.
+At an input stream of length one, $\textsc{ChkFby}_1$ asserts that all subproperties hold for the (non-existent) previous values in the stream.
+At subsequent parts of the stream, $\textsc{ChkFby}_S$ discards the most recent element of the stream history and checks the subexpression with the previous inputs.
+% One may imagine replacing both rules for $\xfby{v}{e}$ with a single rule that simply checks $e$ under the same streaming context.
+% However, such a rule would require infinite derivations for recursive streams.
 
-Rules {\sc ChkRec} and {\sc ChkLet} are similar, performing the same unfolding as the corresponding big-step rules, then checking the resulting expression.
+Rules {\sc ChkRec} and {\sc ChkLet} both perform the same unfolding as the corresponding big-step rules and check the resulting expression.
 
-Finally, rules {\sc ChkCheck} and {\sc ChkContract} perform the real work.
+Finally, the heavy lifting is performed by rules {\sc ChkCheck} and {\sc ChkContract}.
 
@@
 If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we perform the actual check by evaluating the expression $e$ and requiring it to evaluate to a stream of trues.
-Otherwise, we ignore the check-expression.
+Otherwise, we do not need to evaluate the check-expression.
 In both cases, we descend into the expression and check its subexpressions, as they may have nested properties.
+Such nested properties are unlikely to be written directly by the user, but might occur after program transformations such as inlining.
 
@@
 Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
-To check the caller property when $\PStatus = PStatus'$, we evaluate the rely $\erely$ and require it to be true.
+To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to be true.
 To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
 We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
+Unfortunately, this rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.
 
-\subsubsection{Blessing}
-\TODO{blessing}
+\subsubsection{Blessing expressions and contracts}
+\label{s:core:blessing}
 
-\subsection{Causality}
+Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
+Blessing an expression requires a proof that the checked semantics hold for all input streams:
+
+$$
+\ruleIN{
+  \forall \Sigma.~
+  \semcheck{\Sigma}{\PSValid}{e}
+  \implies
+  \semcheck{\Sigma}{\PSUnknown}{e}
+}{\text{bless}~e}{BlessExpression}
+$$
+
+Blessing is slightly different for contract definitions, as we need to separate the definition of the contract from the instantiation.
+To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
+We also assume that the valid properties in the rely, body and guarantee hold, and show the corresponding unknown properties:
+
+\begin{tabbing}
+  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
+  @let@ $\text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}: \text{prop}$ = \\
+  \> $\forall \Sigma.$
+  \> $ (
+    \semcheck{\Sigma}{\PSValid}{(\erely, \ebody, \eguar[x := \ebody])}
+    ~\wedge~
+    \bigstepalways{\Sigma}{\erely}
+  ) $ \\
+  \> $\implies$
+  \> $(
+    \semcheck{\Sigma}{\PSUnknown}{(\erely, \ebody, \eguar[x := \ebody])}
+    ~\wedge~
+    \bigstepalways{\Sigma}{\eguar[x := \ebody]}
+    )$
+\end{tabbing}
+
+After proving that the contract is valid for all inputs, we can bless the contract definition.
+Blessing the contract definition blesses the subexpressions for the rely, body and guarantee, but leaves the contract's \emph{instantiation} property status as unknown:
+$$
+\inferrule{
+  \text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}
+}{\text{bless_contract}~\{\erely\}~\ebody~\{ \eguar\}}(\textsc{BlessContract})
+$$
+
+% The contract we saw in \autoref{s:motivation:contract} did not explicitly bless its contract, but the
+
+\subsection{Causality and metatheory}
 \label{s:core:causality}
 
-\TODO{causality}
+To ensure that recursive streams have a computational interpretation, we require that all recursive streams are guarded by a followed-by delay.
+We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
+This check is stricter than necessary: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ does mention the recursive stream $x$ outside of the delay, but after inlining the let, it would be causal.
+We hope to relax this restriction somewhat in future work.
+
+The causality restriction gives us some important properties about the metatheory.
+The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.
+
+% \begin{theorem}[bigstep-deterministic]
+%   For any streaming history $\Sigma$, if expression $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$ and also evaluates to $v'$ $(\bigstep{\Sigma}{e}{v'})$, then $v = v'$.
+% \end{theorem}
+
+\begin{theorem}[bigstep-is-total]
+  For any non-empty streaming history $\Sigma$ and causal expression $e$, there exists some value $v$ such that $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$.
+\end{theorem}
+
+The relationship between substitution and the streaming history is also important.
+In general, we have a substitution property that states that evaluating a substituted expression $e[x := e']$ under some context $\Sigma$ is equivalent to evaluating $e'$ and adding it to the context $\Sigma$:
+
+\begin{theorem}[bigstep-substitute]
+  For a streaming history $\Sigma$ and a causal expression $e$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
+  The converse is also true.
+\end{theorem}
+
+% \begin{theorem}[bigstep-rec-substitute-elim]
+%   For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if the recursive expression evaluates to a stream $(\bigsteps{\Sigma; \sigma}{\xrec{x}{e}}{V; v})$, then we can evaluate $e$ to the same value by binding $x$ to the same stream in the streaming history $(\bigstep{(\Sigma; \sigma)[x \mapsto (V; v)]}{e}{v})$.
+% \end{theorem}
+
+% Recursive expressions are the essence of computation, so their properties are particularly important.
+Our semantics for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
+An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
+\TODO{XYZ}
+Finally, we also have that evaluation of a causal recursive expression is not affected by the most recent recursive value:
+\begin{theorem}[bigstep-rec-causal]
+  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, the evaluation of $e$ does not depend on the most recent value of $x$ in the streaming history: if $(\bigstep{\Sigma; \sigma}{e}{v})$, then substituting $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x := v']}{e}{v})$.
+\end{theorem}

### 2024-01-10 2024pipit/section/30-Core.tex
core/causality lemmas

@@
 
-% Recursive expressions are the essence of computation, so their properties are particularly important.
-Our semantics for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
+The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
 An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
-\TODO{XYZ}
-Finally, we also have that evaluation of a causal recursive expression is not affected by the most recent recursive value:
+The above substitution theorem can be used to show that these two semantics are equivalent.
+Thanks to causality, we can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
+This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
 \begin{theorem}[bigstep-rec-causal]
-  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, the evaluation of $e$ does not depend on the most recent value of $x$ in the streaming history: if $(\bigstep{\Sigma; \sigma}{e}{v})$, then substituting $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x := v']}{e}{v})$.
+  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x := v']}{e}{v})$.
 \end{theorem}

### 2024-01-10 2024pipit/section/01-Introduction.tex
tr

@@
 \begin{itemize}
-  \item we describe a key issue with prior program logics for Lustre-style languages (\autoref{s:motivation});
+  \item we motivate our work, which allows us to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
@@
   \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
-  \item we identify the invariants and lemmas required to prove preservation of the dynamic (bigstep) semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
+  \item we identify the invariants and lemmas required to prove preservation of the dynamic semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
   \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
-  \item we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
 \end{itemize}
-
-% \begin{itemize}
-%   \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
-%   \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
-%   \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
-%   % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
-%   \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-%CUT
-% \begin{itemize}
-%   \item We identify the invariants and lemmas required to prove correctness of the translation from a clock-free Lustre-style language to a transition system (\autoref{ss:meta-theory}).
-%   \item We present a syntax-directed operational semantics for a clock-free Lustre-style language (\autoref{ss:operational-semantics}).
-%     % TODO: CUT WAFFLE
-%     This syntax-directed semantics is in contrast to previous \emph{relational} operational semantics of Lustre, which are not syntax-directed and require detailed proofs of determinism.
-%     Proving determinism on our semantics requires only straightforward induction; we believe that this semantics may be better for proving other properties too.
-%   \item We show preliminary results that demonstrate the effectiveness of Pipit by implementing a verified controller for a simple embedded system (\autoref{s:coffee-machine}).
-% \end{itemize}
-
-% \begin{itemize}
-%   \item an embedded language makes it easier to express interesting programs with a small real-time synchronous core language;
-%   \item by embedding in a language with good proof automation support, we can reason about embedded programs without much manual proof effort;
-%   \item by embedding in a language with manual proof support, we can prove harder problems than existing model checkers; and
-%   \item by embedding in a language with a well-defined extraction to (heap) allocation-free C code, we can get a partially verified compiler; so far we have only proved type preservation
-% \end{itemize}

### 2024-01-10 2024pipit/section/30-Core.tex
tr

@@
 The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.
+These properties are mechanised in \fstar{}.
 
@@
 \begin{theorem}[bigstep-rec-causal]
-  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x := v']}{e}{v})$.
+  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
 \end{theorem}

### 2024-01-10 2024pipit/section/35-Primitive-Tables.tex
ok
tr

@@
-\subsection{Primitive tables}
-\label{s:core:primitive-tables}
+% KILL I don't think this is really useful
 
-The core language presented in \autoref{f:core-grammar} has a restricted set of primitives.
-Our actual implementation is parameterised by the set of primitive types and operations.
+% \subsection{Primitive tables}
+% \label{s:core:primitive-tables}
 
-In addition to a standard deeply-embedded primitive table which supports integers, tuples, booleans and reals, we also provide a shallowly-embedded primitive table, where operations can be arbitrary pure \fstar{} functions.
-This shallow embedding also allows straightforward integration of \fstar{} types and functions, as well as proofs inside streaming functions.
+% The core language presented in \autoref{f:core-grammar} has a restricted set of primitives.
+% Our actual implementation is parameterised by the set of primitive types and operations.
 
-This parameterisation serves mainly as a convenience.
-For the formalisation presented here, we assume a simple static set of primitives.
+% In addition to a standard deeply-embedded primitive table which supports integers, tuples, booleans and reals, we also provide a shallowly-embedded primitive table, where operations can be arbitrary pure \fstar{} functions.
+% This shallow embedding also allows straightforward integration of \fstar{} types and functions, as well as proofs inside streaming functions.
 
-In our current implementation, we cannot encode primitives that have preconditions, as the preconditions are checked eagerly by \fstar{}, before anything is known about the surrounding streaming values.
-
-To work around this limitation, we encode certain primitive preconditions as implications on the postcondition.
-For example, to encode the array indexing operation in \autoref{s:motivation}, we use an unsafe variant of the indexing function:
-\begin{tabbing}
-  @MM@\= @MM@ \= \kill
-  @val@ Array.index_unsafe ($\textit{array}: \mbox{const_array}~\alpha$) ($\textit{index}$: $\NN$): \\
-  \> $v: \alpha \{ \textit{index} < \mbox{Array.length}~\textit{array} \implies v = \mbox{Array.index}~\textit{array}~\textit{index} \}$
-\end{tabbing}
-
-By wrapping this unsafe index function inside a contract that requires the index to be valid, we can prove the postcondition on the streaming level.
-
-In the future, we plan to investigate allowing preconditions on non-streaming functions, so that safe variants can be used directly.
+% This parameterisation serves mainly as a convenience.
+% For the formalisation presented here, we assume a simple static set of primitives.

### 2024-01-10 2024pipit/section/40-Transition.tex
ok
tr

@@
 
-\section{Transition systems}
+\section{Abstract transition systems}
 \label{s:transition}
@@
 
+To prove properties about Pipit programs, we translate to an \emph{abstract} transition system, so-called because it abstracts away the implementation details of contract instantiations.
+For extraction we also translate to \emph{executable} transition systems, which we discuss in \autoref{s:extraction}.
+
 \autoref{f:system-types} shows the types of transition systems.
-A transition system is parameterised by its input environment and the result type.
-It also contains two internal environments: firstly, the state environment describes the private state required to execute the machine; secondly, the free environment contains any extra input values that the transition system would like to quantify over.
-The free environment is used to allow the system to concoct arbitrary values, when it would not otherwise be able to return a concrete value.
-For contracts which abstract over the implementation, the natural definition of the step function would involve an existential quantifier: there exists some value that satisfies the specification.
+A transition system is parameterised by its input context and the result type.
+It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to quantify over.
+The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.
+
+For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: there exists some value that satisfies the specification.
 Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
-Using a step relation with existential quantifiers complicates the resulting transition, as the quantifiers get between the value and its definition.
-Instead, the free environment provides the step function with a fresh unconstrained value of the desired type, and the step function itself can constrain it however it wishes.
+Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block normalisation and result in a more complex transition system.
+Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.
 
-The free environment is used to generate similar systems to the transition systems generated by existing systems such as Kind2 \cite{champion2016kind2}, which can generate fresh variables at will.
-The main distinction is that embedding such a translation inside a theorem prover and proving it correct requires some ingenuity.
+As usual, the step-result contains the updated state for the transition system, as well as the result value.
+The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
+For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.
+
+% There is a slight difference with how followed-by delays are handled --- the checked semantics checks the delayed expression in the delayed context, while the transition system checks one step ahead ---
+
+% The free context is used to generate similar systems to the transition systems generated by existing systems such as Kind2 \cite{champion2016kind2}, which can generate fresh variables at will.
+% The main distinction is that embedding such a translation inside a theorem prover and proving it correct requires some ingenuity.
 
 \input{figures/system-translation.tex}
+
+\autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
+For most expression forms, the state and free contexts are defined by taking the union of the contexts of sub-expressions.
+Followed-by delays introduce a local state variable $x$, in which to store the most recent stream value.
+We generate a fresh variable here, though the implementation uses de Bruijn indices.
+Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are fresh.
 
@@
 \subsection{Translation correctness proofs}
+\label{s:transition:proof}
 

### 2024-01-10 2024pipit/section/60-Evaluation.tex
ok
tr

@@
 \label{s:evaluation}
+
+\TODO{EVALUATION NEEDS DOING:}
+\begin{itemize}
+  \item describe time-triggered CAN implementation
+  \item whole implementation includes the extended version of trigger logic
+  \item verified machine arithmetic
+  \item for which we can generate real-time C code
+  \item describe coffee machine controller
+  \item simple, but fully verified according to an abstract spec
+  \item and for which we generate real-time C code
+\end{itemize}
 
@@
 \end{tabbing}
-\pagebreak
 
@@
 % by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
+Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \COFFEEURL}.

### 2024-01-10 2024pipit/section/40-Transition.tex
transition systems

@@
 
+Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics.
+The proof that the rely and guarantee propositions correspond to the checked semantics is future work.
+
 % There is a slight difference with how followed-by delays are handled --- the checked semantics checks the delayed expression in the delayed context, while the transition system checks one step ahead ---
@@
 \autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
-For most expression forms, the state and free contexts are defined by taking the union of the contexts of sub-expressions.
-Followed-by delays introduce a local state variable $x$, in which to store the most recent stream value.
+For most expression forms, the state and free contexts are defined by taking the union of the contexts of subexpressions.
+Followed-by delays introduce a local state variable $x_{@fby@}$ in which to store the most recent stream value.
 We generate a fresh variable here, though the implementation uses de Bruijn indices.
-Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are fresh.
+Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are unique.
 
-\TODO{describe translation}
-The substitution-based semantics is better for equational reasoning, and (subjectively perhaps) simpler.
-The transition system is better for certain automated proofs: a surprising number of interesting properties can be proven with k-induction alone.
+\autoref{f:system-translation} defines the translation for expressions.
+Values and expressions have no internal state.
+For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
+We omit the rely and guarantee definitions here; both are trivially true.
+
+To translate primitives, we union together the initial states of the subexpressions; updating the state is similar.
+For the rely and guarantee definitions, we take the conjunction: we can assume that all subexpressions rely clauses hold, and must show that all guarantees hold.
+
+To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@}$ to $v$.
+At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
+The rely and guarantee differ from the checked semantics here: in the checked semantics, we check the subexpression on the previous inputs, but here we check the current subexpression.
+This means that a single step of the rely and guarantee do not exactly correspond to the checked semantics; however, we posit that they are equivalent for a rely and guarantee that has been proven to hold for any sequence of inputs.
+
+To translate a recursive expression $\xrec{x}{e}$ of type $\tau$, we require an arbitrary value $x: \tau$ in the free heap.
+The rely proposition constrains the free variable $x$ to be the result of evaluating $e$ with the binding for $x$ passed along, thus closing the recursive loop.
+
+To translate let-expressions $\xlet{x}{e}{e'}$, we extend the input heap with the value of $e$ before evaluating $e'$.
+The presentation here duplicates the computation of the value of $e$, but this is not an issue in practice.
+
+To translate a check property, we inspect the property status.
+If the property is known to be valid, then we can assume the property is true in the rely clause.
+Otherwise, we include the property as an obligation in the guarantee clause.
+In either case, we also include the subexpression's rely and guarantee clauses.
+
+Finally, to translate contract instantiations, we use the contract's rely and guarantee and ignore the body.
+As with recursive expressions, we require an arbitrary value $x: \tau$ in the free heap.
+The translation's rely allows us to assume that the contract definition holds: that is, the contract's rely implies the contract's guarantee.
+If the contract instantiation is known to be valid, we can also assume that the contract's rely holds.
+Otherwise, we include the contract's rely as an obligation by putting it in the translation's guarantee.
+
+In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
+The true semantics of the contract, however, only holds if the contract rely is true \emph{at every step so far}.
+This simplification is justified as our definition of validity for a transition system also requires the translation rely to be true at every step.
+This simplification is essentially an application of the $\Box \Box p \implies \Box p$ axiom of modal logic.
+The mechanised proof of this simplification is future work.
 
@@
 
-\TODO{describe the proofs of translation correctness}
-The translation to a transition system above is an abstraction.
-We prove it.
-Interesting lemmas etc.
+We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
+The transition system itself is deterministic, but the free context provides the non-determinism; our theorem  statement existentially quantifies the free heap.
+
+\input{figures/system-invariant.tex}
+
+The results presented here rely heavily on the totality and substitution metaproperties described in \autoref{s:core:causality}.
+\autoref{f:system-invariant} defines the invariant for the abstraction proof; the judgment form $\sysinv{\Sigma}{e}{s}$ checks that $s$ is a valid state heap.
+We use the invariant to state that, if executing the transition system for $e$ on the entire streaming history $\Sigma$ results in state heap $s$, then $s$ is a valid state.
+
+As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
+Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
+The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
+
+\begin{theorem}[translation-abstraction]
+  For a well-typed causal expression $e$ and streaming history $\Sigma$, if $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$, then there exists a sequence of free heaps $\Sigma_{F}$ such that repeated application of the transition system's step results in $v$.
+\end{theorem}

### 2024-01-11 2024pipit/Main.tex
ok
time-triggered example
tr
extraction

@@
 \newcommand{\GITHUBURL}{(anonymised)}
+\newcommand{\COFFEEURL}{(anonymised)}
 \else
 \newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
+\newcommand{\COFFEEURL}{\url{https://youtu.be/6IybbQFPOl8}}
 \fi
@@
   Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
-  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
-  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
+  Pipit can also generate executable code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
+  The executable code is deterministic and total and preserves the semantics of the original program.
 \end{abstract}
@@
 
-Big TODOs:
-\begin{itemize}
-  \item example: trigger-fetch from ttcan;
-  \item transition systems and translation;
-\end{itemize}
-
 \input{section/01-Introduction.tex}
-
 \input{section/20-Motivation.tex}
 \input{section/30-Core.tex}
+\input{section/35-Primitive-Tables.tex}
 \input{section/40-Transition.tex}

### 2024-01-11 2024pipit/section/01-Introduction.tex
extraction

@@
 % By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
-Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics; we also prove that the imperative code preserves the semantics.
+To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
+To execute programs, Pipit can generate executable code, which is total and semantics-preserving.
 
@@
   \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
-  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
-  \item we identify the invariants and lemmas required to prove preservation of the dynamic semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
-  \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
-  \item finally, we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
+  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions, rather than relations, which results in simpler transition systems (\autoref{s:transition});
+  \item we identify and mechanise the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:transition:proof});
+  \item similarly, we verify that the executable transition system preserves the original semantics (\autoref{s:extraction});
+  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
 \end{itemize}

### 2024-01-11 2024pipit/section/20-Motivation.tex
causality etc
extraction

@@
 The program above is really syntactic sugar for an \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable.
-We will discuss the details of the core language in \autoref{s:core}, but for now we will show the source program with some minor embedding details omitted.
+We will discuss the details of the core language in \autoref{s:core}, but for now we focus on the source program with some minor embedding details omitted.
 
@@
 
-The system has strict timing requirements, which restrict how triggers can be defined.
+The system has strict timing requirements which restrict how triggers can be defined.
 In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
-Additionally, the schedule must allow sufficient time for the driver to skip over any disabled triggers.
-Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2 instead.
-However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per time-mark, and it takes two steps to reach trigger 2 from the start of the array.
+Additionally, the schedule must allow sufficient time for the driver to skip over the disabled triggers.
+Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2, as triggers 1 and 2 have distinct cycles.
+However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per tick, and it takes two steps to reach trigger 2 from the start of the array.
 
@@
 \subsection{Instantiating lemmas and defining contracts}
+\label{s:motivation:contract}
 
@@
 
-We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 refers to the lemma taking one streaming argument (the index).
+We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 indicates that the lemma is being applied to one streaming argument (the index).
 % The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
@@
 To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
-The contract we state is that if the environment always ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
+The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
 
@@
 % These kinds of self-dependencies and mutual-dependencies are fairly common in Lustre-style programs: for example, \emph{trigger_fetch}'s invocation of \emph{count_when}, whose input $\textit{inc}$ depends on the previous output $\textit{index}$.
-
-% \subsection{Program logics and dependent types}
-% \TODO{compare to refinement types, verification conditions, etc here, or wait to \autoref{s:related-work}?}
-
-% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
-% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.
-
-% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
-% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
-% We could imagine specifying count_when with such a type:
-
-% \begin{tabbing}
-%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
-% \end{tabbing}
-
-% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.
-
-% We believe that such type-based approaches are useful for composing programs together, but 
-%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
-% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
-% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$

### 2024-01-11 2024pipit/section/50-Extraction.tex
ok
extraction

--- 2023-12-18 2024pipit/section/50-Extraction.tex
@@
 Pipit can generate executable code which is suitable for real-time execution on embedded devices.
-The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
-\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
-During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.
+The code extraction uses a variation of the abstract transition system described in \autoref{s:transition}, with two main differences to ensure that the result is executable without relying on the environment to provide values for the free context.
+Contracts are straightforward to execute by using the body of the contract rather than abstracting over the implementation.
 
-% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
-% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.
+To execute recursive expressions $\xrec{x}{e} : \tau$, we require an arbitrary value of type $\tau$ to seed the fixpoint, as described in \autoref{s:core:causality}.
+We first call the step function to evaluate $e$ with $x$ bound to $\bot_\tau$.
+This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
+To get the correct state, we call the step function again, this time with $e$ bound to $v$.
 
-% noextract
-% let expr = Sugar.run2 Pump.controller'
+This translation to transition systems is verified to preserve the original semantics; however, it can duplicate work by requiring two calls to the step function.
+To extract the above transition systems, we use \fstar{}'s normalisation-by-evaluation to generate step functions that update their mutable state in-place and which \lowstar{} can extract to C.
+This normalisation strategy inlines the two step functions and is often able to remove the duplicate work.
 
-% noextract
-% let system: Pipit.Exec.Exp.xexec expr =
-%   XX.exec_of_exp expr
+Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
+In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
+We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
 
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let reset = XL.mk_reset system
-
-% [@@(Tac.postprocess_with XL.tac_extract)]
-% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
-% This allows us to generate C code which executes in bounded space on embedded devices.
+% \item staging: similarity to \cite{gallego2021w};

### 2024-01-11 2024pipit/section/01-Introduction.tex
some intro

@@
 \begin{itemize}
-  \item we motivate our work, which allows us to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
+  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
   \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
-  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions, rather than relations, which results in simpler transition systems (\autoref{s:transition});
-  \item we identify and mechanise the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:transition:proof});
-  \item similarly, we verify that the executable transition system preserves the original semantics (\autoref{s:extraction});
+  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
+  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
+  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
   \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).

### 2024-01-11 2024pipit/section/20-Motivation.tex
some intro

@@
 
-The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark; the driver maintains an index that refers to the current trigger.
+The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark.
+The driver maintains an index that refers to the current trigger.
 At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
@@
 
-We impose three main restrictions on the triggers array: it must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
+We impose three main restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
 
 With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
-These properties are fairly straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
+These properties are straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
 
@@
 We can now implement the trigger-fetch logic, which maintains the index that refers to the next-scheduled trigger.
-The trigger-fetch logic uses the \emph{count_when} streaming function to define the index; the increment flag given to count_when will increment the index if the previous index has expired or is inactive in the current cycle.
+The trigger-fetch logic uses the \emph{count_when} streaming function to define the index; the increment flag given to \emph{count_when} will increment the index if the previous index has expired or is inactive in the current cycle.
 We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
@@
 With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
-To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
+To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
 The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.

### 2024-01-13 2024pipit/section/01-Introduction.tex
evaluation

@@
   \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
-  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver, which we have partially verified (\autoref{s:evaluation}).
 \end{itemize}

### 2024-01-13 2024pipit/section/20-Motivation.tex
evaluation

@@
 
-To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered CAN bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
+To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered Controller Area Network (CAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
 % In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
@@
 % The trigger-fetch logic requires such a combination of automatic and manual proofs.
-We can now implement the trigger-fetch logic, which maintains the index that refers to the next-scheduled trigger.
-The trigger-fetch logic uses the \emph{count_when} streaming function to define the index; the increment flag given to \emph{count_when} will increment the index if the previous index has expired or is inactive in the current cycle.
+We can now implement the trigger-fetch logic, which keeps track of the current trigger.
+The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
 We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
@@
     \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
-    \> \> $\textit{index}$\\
+    \> \> $\textit{index}$
 \end{tabbing}
 
-The \emph{trigger_fetch} function takes a static cycle index and the stream denoting the current time.
+The \emph{trigger_fetch} function takes a static cycle index and a stream denoting the current time.
 The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.

### 2024-01-13 2024pipit/section/50-Extraction.tex
some intro
evaluation

@@
 
-Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
+Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
 In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
@@
 % \item staging: similarity to \cite{gallego2021w};
+\CITE{Noise* code extraction} \cite{ho2022noise} hybrid embedding

### 2024-01-13 2024pipit/section/60-Evaluation.tex
bosch
evaluation

--- 2023-12-18 2024pipit/section/60-Evaluation.tex
@@
 
-To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
-This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
-The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
-The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
-The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
-Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage:
+As a preliminary evaluation of Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
+The CAN bus is commonly found in safety-critical automotive and industrial settings.
+The time-triggered network architecture defines a static schedule of network traffic.
+All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.
 
-\newcommand\estop{\textit{stop}}
-\newcommand\low{\textit{low}}
-\newcommand\soltry{\textit{try}}
-\newcommand\error{\textit{error}}
-\newcommand\solen{\textit{engage}}
+At a high level, the schedule is described by a system matrix which consists of rows of basic cycles.
+Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
+Actions in the schedule may not be relevant to all nodes, so each node has its own local array containing the relevant triggers; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
+The trigger action for receiving an application-specific message checks that a particular message has been received since the trigger was last executed; depending on this, the driver increments or decrements a message-status-counter, which will in turn signal an error once the upper limit is reached.
+Reference messages start a new basic cycle and are used to synchronise the nodes.
+Watch alerts are generally placed after the expected end of the cycle and are used to signal an error if no reference message is received.
+
+The TTCAN protocol can be implemented in two levels of increasing complexity.
+In the first level, reference messages contain the index of the newly-started cycle.
+In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
+We implement the first level as it is more amenable to software implementation~\cite{hartwich2002integration}.
+
+The implementation defines a streaming function that takes a stream describing the current time, the state of the hardware, and any received messages.
+It returns a stream of commands to be performed, such as sending a particular reference message.
+The implementation defines a pure streaming function.
+To actually interact with the hardware we assume a small hardware-interop layer that reads from the hardware registers and translates the commands to hardware-register writes, but we have not yet implemented this.
+We package the driver's inputs into a record for convenience:
+
 \begin{tabbing}
-  @MM@\= @let@ engage \= @MMMM@ \= \kill
-  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
-  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
-  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
-  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
-  \> @check@ (\solen{} $\implies$ not \estop{}); \\
-  \> @check@ (\solen{} $\implies$ \low{}); \\
-  \> \solen{}
+  MM \= bus_status: \= \kill
+  @type@ driver_input = \{ \\
+    \> local_time: \> network_time_unit; \\
+    \> mode_cmd: \> option mode; \\
+    \> tx_status: \> tx_status; \\
+    \> bus_status: \> bus_status; \\
+    \> rx_ref: \> option ref_message; \\
+    \> rx_app: \> option app_message_index; \\
+    \}
 \end{tabbing}
-\pagebreak
 
-Predicate \emph{true\_for t} is true if a signal has been true for time $t$; \emph{any} is true if a signal has ever been true.
-% The \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
-As with the previous examples, the two properties can be automatically verified.
-% by \fstar{}'s proof automation.
-Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
+Here, the local-time field denotes the time-since-boot in \emph{network time-units}, which matches the bitrate of the underlying network bus.
+The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
+The transmission-status describes the status of the last transmission request and may be none, success, or various error conditions.
+The bus-status describes whether the bus is currently idle, busy, or in an error state.
+The two receive fields denote messages received from the bus; for application-specific messages the time-triggered logic only needs the message identifier.
+
+The driver-logic returns a stream of commands for the hardware-interop layer to perform:
+
+\begin{tabbing}
+  MM \= enable_acks: \= \kill
+  @type@ commands = \{ \\
+  \> enable_acks: \> bool; \\
+  \> tx_ref: \>       option ref_message; \\
+  \> tx_app: \> option app_message_index; \\
+  \> tx_delay: \>     network_time_unit; \\
+ \}
+\end{tabbing}
+
+The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
+The transmit fields denote whether to send a reference message or an application-specific message.
+For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
+To meet the schedule as closely as possible, the driver predicts the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.
+
+The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
+We generate real-time C code as described in \autoref{s:extraction}.
+We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
+The runtime of the driver logic is fairly stable: over 5,000 executions, the worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
+Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
+For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent, which our implementation satisfies.
+For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$; however, as our runtime does not include the time required for the hardware-interop layer, we are not confident that our implementation will satisfy full-speed CAN bus.
+
+Our code generation can be improved in a few ways.
+A common optimisation in Lustre is to fuse consecutive if-statements with the same condition~\cite{bourke2017formally}; such an optimisation seems useful here, as our treatment of optional values introduces repeated unpacking and repacking.
+Some form of array fusion~\cite{robinson2017machine} may also be useful for removing redundant array operations.
+Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
+Composing these step functions together requires repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
+In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.
+
+% The Bosch M-TTCAN IP module limits to 64 trigger elements \cite{bosch2019mttcan}.
+
+We have verified a simplification of the trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
+We plan to verify the remainder of the TTCAN implementation and publish it separately.
+Prior work formalising TTCAN has variously modeled the protocol itself~\cite{saha2007finite, pan2014modeling,li2018formal},
+instances of the protocol~\cite{guo2020model},
+and abstract models of TTCAN implementations~\cite{leen2006modeling}, but we are unaware of any prior work that has verified an \emph{executable} implementation of TTCAN.
+
+Separately, Pipit has also been used to implement and verify a simple real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
+The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
+The specification includes a simple model of the water reservoir and shows that the reservoir will not exceed the maximum level.

### 2024-01-15 2024pipit/section/30-Core.tex
bosch
eval: kind2

@@
 
-Streams $V$ are represented as a non-empty sequence of values; streaming history environments $\Sigma$ are streams of heaps.
+Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
 Types $\tau$ and type environments $\Gamma$ are standard.
@@
 We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
+Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital signal processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.
 

### 2024-01-15 2024pipit/section/50-Extraction.tex
eval: kind2

--- 2024-01-11 2024pipit/section/50-Extraction.tex
@@
 
-This translation to transition systems is verified to preserve the original semantics; however, it can duplicate work by requiring two calls to the step function.
-To extract the above transition systems, we use \fstar{}'s normalisation-by-evaluation to generate step functions that update their mutable state in-place and which \lowstar{} can extract to C.
-This normalisation strategy inlines the two step functions and is often able to remove the duplicate work.
+This translation to transition systems is verified to preserve the original semantics.
+To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
+The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
+We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
+This specialisation results in a concrete transition system that fits in the \lowstar{}~\cite{protzenko2017verified} subset of \fstar{}, which can then be extracted to statically-allocated C code.
 
-Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
-In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
-We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
-
-% \item staging: similarity to \cite{gallego2021w};
+The translation for recursive streams described above calls the step function of the sub-stream twice, which can duplicate work.
+The normalisation strategy used to partially-evaluate the translation inlines the two occurrences of the step function, and is often able to remove the duplicate work, but this removal is not guaranteed.
+Our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
+In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
+Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
+% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

### 2024-01-15 2024pipit/section/60-Evaluation.tex
eval: kind2

@@
 
-Here, the local-time field denotes the time-since-boot in \emph{network time-units}, which matches the bitrate of the underlying network bus.
+Here, the local-time field denotes the time-since-boot in \emph{network time units}, which are based on the bitrate of the underlying network bus.
 The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
@@
 For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
-To meet the schedule as closely as possible, the driver predicts the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.
+To meet the schedule as closely as possible, the driver anticipates the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.
+
+\subsection{Runtime}
 
@@
 We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
-The runtime of the driver logic is fairly stable: over 5,000 executions, the worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
+The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
 Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
-For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent, which our implementation satisfies.
-For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$; however, as our runtime does not include the time required for the hardware-interop layer, we are not confident that our implementation will satisfy full-speed CAN bus.
+For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent.
+For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$.
+Further evaluation is required to ensure that the complete runtime including the hardware-interop layer is sufficient for full-speed CAN.
 
@@
 Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
-Composing these step functions together requires repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
+Composing these step functions together results in repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
 In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.
 
-% The Bosch M-TTCAN IP module limits to 64 trigger elements \cite{bosch2019mttcan}.
+\subsection{Verification}
 
-We have verified a simplification of the trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
+We have verified a simplified trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
+For comparison, we implemented the same logic in the Kind2 model-checker~\cite{champion2016kind2}.
+The restrictions placed on the triggers array --- that triggers are sorted by time-mark, that there must be an adequate time-gap between a trigger and its next-enabled, and that a trigger's time-mark must be greater-than-or-equal-to its index --- are naturally expressed with quantifiers.
+The Kind2 model-checker includes experimental array and quantifier support~\cite{kind2userdoc}.
+% but uses a custom syntax for arrays with no compiler support.
+Due to the experimental nature of these features, we had to work around some limitations: for example, the use of arrays and quantifiers disables IC3-based invariant generation; quantified variables cannot be used in function calls; and the use of top-level constant arrays caused runtime errors that rendered most properties invalid~\cite{kind2024toparray}.
+
+We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays up to size sixteen.
+Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after \TODO{hours}.
+In comparison, the Pipit version is verified \emph{for all sizes of arrays} and the checking took 5s of CPU time.
+The M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
+Both developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+We ran these experiments on a 2020 MacBook Air M1 with 16 gigabytes of RAM.
+
+A critical simplification we made to the Kind2 implementation was to modify the trigger-enabled set to be a single cycle index.
+In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
+Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
+In our Pipit development, we can treat the definition of the cycle set abstractly.
+However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
+
+% \begin{tabular}{rrr}
+%   size & wall-clock & CPU time \\
+%   4 & 8s & 11s \\
+%   8 & 8s & 13s \\
+%   16 & 125s & 276s \\
+%   32 & - & - \\
+% \end{tabular}
+
 We plan to verify the remainder of the TTCAN implementation and publish it separately.
@@
 
-Separately, Pipit has also been used to implement and verify a simple real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
+Separately, Pipit has also been used to implement and verify a real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
 The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
-The specification includes a simple model of the water reservoir and shows that the reservoir will not exceed the maximum level.
+The specification includes a simple model of the water reservoir and shows that the reservoir does not exceed the maximum level under different failure-mode assumptions.

### 2024-01-15 2024pipit/section/70-Related.tex
causality etc
eval: kind2

@@
 
-% TODO arrays
-\TODO{requires rewrite}
+Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
+Compiling the triggers array with an unverified compiler such as Lustre V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
 
-Although the FIR filter from XXX is quite simple, verifying it \emph{and} executing it with existing Lustre tools is nontrivial.
-Lustre itself does not support lists, as dynamically-allocated data structures are not well-suited to real-time execution.
-To write this filter in Lustre we would either need to unroll the lists ourselves or reformulate the program to use arrays.
-However, Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and the Lustre V6 compiler does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
+Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
+The Kind2 model-checker does include experimental array and quantifier support~\cite{kind2userdoc}, but uses a custom syntax for arrays with no compiler support.
+This experimental support has some limitations: the use of arrays and quantifiers disables IC3-based invariant generation, quantified variables cannot be used in function calls, and top-level arrays are not supported~\cite{kind2024toparray}.
+
+Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous} and JKind~\cite{gacek2018jkind} do not support quantifiers.
+It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
 
@@
 
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.
+The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}.
+Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
+However, the model-checking suffers from the same semantic gap.
 
@@
 % hardware language in Agda \cite{harrison2021mechanized}
-% Refinement types for Zélus \cite{chen2022synchronous}
-% Invariant generation: ICE \cite{garg2014ice}; template-based \cite{kahsai2011instantiation}; PDR explanation \cite{een2011efficient}; IC3 first introduced \cite{bradley2011sat}
+% template-based invariant generation \cite{kahsai2011instantiation}
 
@@
 In this way, we hope that it may be possible to obtain some of the usability benefits of model-checkers, with the added power of manual proofs when necessary, while retaining trust that the compiled code satisfies the original specification.
+
+% \subsection{Model-checking, counterexamples and invariant generation}
+\subsection{Comparison to a combinator-based verification approach}
+\TODO{compare to refinement types, verification conditions, etc here}
+
+% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
+% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.
+
+% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
+% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
+% We could imagine specifying count_when with such a type:
+
+% \begin{tabbing}
+%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
+% \end{tabbing}
+
+% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.
+
+% We believe that such type-based approaches are useful for composing programs together, but 
+%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
+% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
+% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$

### 2024-01-15 2024pipit/section/60-Evaluation.tex
eval runtimes

@@
 
-We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays up to size sixteen.
-Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after \TODO{hours}.
+We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to sixteen elements.
+Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
 In comparison, the Pipit version is verified \emph{for all sizes of arrays} and the checking took 5s of CPU time.
@@
 Both developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
-We ran these experiments on a 2020 MacBook Air M1 with 16 gigabytes of RAM.
+We ran these experiments on a 2020 M1 MacBook Air with 16 gigabytes of RAM.
 
-A critical simplification we made to the Kind2 implementation was to modify the trigger-enabled set to be a single cycle index.
+The Kind2 implementation required a critical simplification, which was to modify the trigger-enabled set to be a single cycle index.
 In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
@@
 However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
+With the concrete definition of cycle-set, Kind2 verified an 8-size array in 90s of CPU time and timed out with runtime errors on a 16-size array.
 
-% \begin{tabular}{rrr}
-%   size & wall-clock & CPU time \\
-%   4 & 8s & 11s \\
-%   8 & 8s & 13s \\
-%   16 & 125s & 276s \\
-%   32 & - & - \\
-% \end{tabular}
+\begin{figure}
+\begin{tabular}{r|rr|rr|rr}
+  & \multicolumn{4}{c}{Kind2} & Pipit \\
+  & \multicolumn{2}{c}{simple enable-set} & \multicolumn{2}{c}{full enable-set} & \\
+  size & wall-clock & CPU & wall-clock & CPU & wall-clock & CPU \\
+  \hline
+  1 & 2s &  3s  & 6s & 12s & 6s & 6s \\
+  2 & 3s &  3s  & 8s & 12s & 6s & 6s \\
+  4 & 5s & 11s  & 12s & 24s & 6s & 6s \\
+  8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
+  16 & 125s & 276s & & (error) & 6s & 6s \\
+  32 & & (error) & & (error) & 6s & 6s \\
+\end{tabular}
+\caption{Proof runtime for simplified trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bit arithmetic and matches the specification.}
+\label{f:evaluation:kind2-runtime}
+\end{figure}
 

### 2024-01-15 2024pipit/section/60-Evaluation.tex
clean up eval

--- 2024-01-15 2024pipit/section/60-Evaluation.tex
@@
 
-We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays up to size sixteen.
-Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after \TODO{hours}.
-In comparison, the Pipit version is verified \emph{for all sizes of arrays} and the checking took 5s of CPU time.
-The M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
-Both developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
-We ran these experiments on a 2020 MacBook Air M1 with 16 gigabytes of RAM.
+We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, a 32-size array reported multiple runtime errors.
+% Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
+For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
 
-A critical simplification we made to the Kind2 implementation was to modify the trigger-enabled set to be a single cycle index.
+We made a critical simplification in the Kind2 implementation, which was to modify the trigger-enabled set to be a single cycle index.
 In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
@@
 
-% \begin{tabular}{rrr}
-%   size & wall-clock & CPU time \\
-%   4 & 8s & 11s \\
-%   8 & 8s & 13s \\
-%   16 & 125s & 276s \\
-%   32 & - & - \\
-% \end{tabular}
+\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
+We ran these experiments on a 2020 M1 MacBook Air with 16 gigabytes of RAM.
+Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+
+\begin{figure}
+  \center
+\begin{tabular}{r|rr|rr|rr}
+  & \multicolumn{4}{c|}{Kind2} & Pipit \\
+  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
+  size & wall-clock & CPU & wall-clock & CPU & wall-clock & CPU \\
+  \hline
+  1 & 2s &  3s  & 6s & 12s & 6s & 6s \\
+  2 & 3s &  3s  & 8s & 12s & 6s & 6s \\
+  4 & 5s & 11s  & 12s & 24s & 6s & 6s \\
+  8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
+  16 & 125s & 276s & \multicolumn{2}{c|}{(error)} & 6s & 6s \\
+  32 & \multicolumn{2}{c|}{(error)} & \multicolumn{2}{c|}{(error)} & 6s & 6s \\
+\end{tabular}
+\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.}
+\label{f:evaluation:kind2-runtime}
+\end{figure}
 

### 2024-01-15 2024pipit/section/60-Evaluation.tex
eval

@@
 
-We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, a 32-size array reported multiple runtime errors.
+We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, a 32-size array reported multiple runtime errors and did not terminate after several hours.
 % Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
@@
 However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
+This limitation also makes the specification quite unwieldy, as functions must be manually inlined.
 
@@
   8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
-  16 & 125s & 276s & \multicolumn{2}{c|}{(error)} & 6s & 6s \\
-  32 & \multicolumn{2}{c|}{(error)} & \multicolumn{2}{c|}{(error)} & 6s & 6s \\
+  16 & 125s & 276s & \multicolumn{2}{c|}{error} & 6s & 6s \\
+  32 & \multicolumn{2}{c|}{error} & \multicolumn{2}{c|}{error} & 6s & 6s \\
 \end{tabular}

### 2024-01-15 2024pipit/section/70-Related.tex
related

@@
 
+% SMT-based model-checkers~\cite{hagen2008scaling}
+
 Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
-Compiling the triggers array with an unverified compiler such as Lustre V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
+Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
+Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.
 
 Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
-The Kind2 model-checker does include experimental array and quantifier support~\cite{kind2userdoc}, but uses a custom syntax for arrays with no compiler support.
-This experimental support has some limitations: the use of arrays and quantifiers disables IC3-based invariant generation, quantified variables cannot be used in function calls, and top-level arrays are not supported~\cite{kind2024toparray}.
+As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but is limited to verifying small arrays up to 8 or 16 triggers.
+Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in multiple copies of the entire triggers array on the stack.
 
-Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous} and JKind~\cite{gacek2018jkind} do not support quantifiers.
+Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
 It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
 
-\TODO{Verified compilation via Frama-C} \cite{brun2023equation}
+These model-checkers have definite usability advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
+However, even when the problem can be expressed, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
+In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
+If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not expand the trusted computing base.
 
-In terms of model-checking reactive systems,
-% earlier model-checkers used explicit state-space exploration~\cite{raymond2008synchronous};
-recent work uses SMT solvers to check inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
-These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
-However, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
-We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.
+Recent work has also introduced a form of refinement types for Lustre~\cite{chen2022synchronous}.
+Rather than using transition systems, this work generates self-contained verification conditions based on the types of streams.
+Such a type-based approach promises to allow abstraction of the implementation details.
+However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide be insufficient for verifying the rest of the system.
+For this function, the best specification is likely to include a re-statement of the implementation itself.
 
-The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}.
+The embedded language Copilot generates real-time C code for runtime monitoring~\cite{laurent2015assuring}.
 Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
-However, the model-checking suffers from the same semantic gap.
+Copilot supports model-checking via Kind2; however, the model-checking has a limited specification language and does not support contracts.
 
@@
 There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
-We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs;
-however, the trusted computing base of Pipit is much larger than Vélus, as we depend on all of \fstar{}, \lowstar{}'s C code extraction, the SMT solver, as well as our currently-unverified imperative code generator.
+We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
+Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s C code extraction and the Z3 SMT solver.
 
@@
 
-\subsection{Model-checking, counterexamples and invariant generation}
-\TODO{integration with model-checker}
-Model-checkers do have distinct advantages over interactive theorem-provers: they are more usable, can find strengthening invariants automatically, and provide counterexamples.
-However, they are limited to comparatively simple properties.
-The time-triggered example from \autoref{s:motivation} would be difficult to encode in a model-checker with a limited specification language, as the trigger invariant uses an auxiliary recursive function to find the next active trigger.
-
-In future work, we would like to explore integrating Pipit with an existing model-checker.
-For simple first-order nonrecursive programs and specifications that fit within the bounds of another model-checker, we could generate an encoding to use for counterexample generation.
-We would also like to investigate automatically lifting the strengthened invariants and suggesting the corresponding Pipit invariant.
-Such a translation and lifting would not form part of the trusted computing base, as generated invariants would still be proved separately in \fstar{}.
-In this way, we hope that it may be possible to obtain some of the usability benefits of model-checkers, with the added power of manual proofs when necessary, while retaining trust that the compiled code satisfies the original specification.
-
-% \subsection{Model-checking, counterexamples and invariant generation}
-\subsection{Comparison to a combinator-based verification approach}
-\TODO{compare to refinement types, verification conditions, etc here}
+% \subsection{Comparison to a combinator-based verification approach}
+% \TODO{compare to refinement types, verification conditions, etc here}
 

### 2024-01-15 2024pipit/section/70-Related.tex
conclusion

@@
 As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but is limited to verifying small arrays up to 8 or 16 triggers.
-Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in multiple copies of the entire triggers array on the stack.
+Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
 
@@
 
+% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.
 % Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
-% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.
 

### 2024-01-15 2024pipit/section/80-Conclusion.tex
ok
extraction
conclusion

--- 2023-12-18 2024pipit/section/80-Conclusion.tex
@@
 
-Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
-Next, we intend to verify the imperative code generation.
-To verify large programs, we also need some way to separately prove smaller pieces which can then be composed together, such as contracts~\cite{champion2016kind2}.
-Finally, we need to evaluate Pipit on larger control systems before extending the language to support more features, such as Lustre's clocks for describing partially-defined streams~\cite{caspi1995functional}.
+We have presented Pipit, a verified compiler and proof system for reactive systems.
+Our implementation of the TTCAN driver logic shows that Pipit can express programs which are currently unsupported by other verified Lustre compilers.
+Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
+Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
+We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that any translated property the user proves on the transition system holds for the original property on the checked semantics.
 
-% TODO
-% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
+In the future, we intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
+Clocks are important for composing complex systems together and avoiding unnecessary computation.
+
+We are interested in further pursuing the intersection of model-checking with interactive theorem proving.
+A smart-contract called Djed \cite{zahnentferner2023djed} currently uses a mixture of Kind2 \cite{champion2016kind2} and manual Isabelle/HOL proofs to show that the contract is well-behaved.
+In future work, we would like to further investigate whether Pipit's integration of streaming proofs with \fstar{}'s automated proof system would be able to provide similar proofs, without introducing any semantic gap between the two systems.

### 2024-01-16 2024pipit/figures/core-structure.tex
diagram

%!TEX root = ../Main.tex
\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels next to the arrows denote properties of the translation. The refinement label ($\subseteq$) indicates a verified abstraction relation; the negated entailment ($\not\vdash$) indicates that the entailment relation --- contracts and checked properties proven on abstract system apply to the checked semantics --- is not yet verified; the equivalence ($=$) denotes a verified equivalence relation.}
\label{f:core:structure}
\end{figure}

### 2024-01-16 2024pipit/section/30-Core.tex
diagram

@@
 
-\input{figures/core-grammar.tex}
+\input{figures/core-structure.tex}
 
@@
 
+\autoref{f:core:structure} shows the architecture of Pipit.
+On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
+The translation from the surface syntax to the core language is trusted.
+There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
+The translation from the abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
+One property that still remains to be proven is that if the abstract transition system says a contract or checked property holds, then the same contract or property holds on the original semantics; this is denoted as negated entailment in the figure ($\not\vdash$).
+The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
+The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.
+
 \autoref{f:core-grammar} defines the grammar of Pipit.
@@
 Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre~\cite{caspi1995functional}.
+
+\input{figures/core-grammar.tex}
 
@@
 
-% For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table \autoref{s:core:primitive-tables}.
+For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.
 
@@
 \subsection{Dynamic semantics}
+\label{s:core:dynamic}
 \input{figures/core-bigstep.tex}
@@
 The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
-\TODO{WAFFLE? This is expanded on in \autoref{s:core:blessing}; kill here?}
 

### 2024-01-16 2024pipit/section/60-Evaluation.tex
diagram

@@
 
-As a preliminary evaluation of Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
+To evaluate Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
 The CAN bus is commonly found in safety-critical automotive and industrial settings.
@@
 \end{tabular}
-\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.}
+\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
+The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.}
 \label{f:evaluation:kind2-runtime}

### 2024-01-16 2024pipit/section/80-Conclusion.tex
diagram

@@
 We have presented Pipit, a verified compiler and proof system for reactive systems.
-Our implementation of the TTCAN driver logic shows that Pipit can express programs which are currently unsupported by other verified Lustre compilers.
+Our implementation of the TTCAN driver logic shows that, by embedding pure \fstar{} functions for array operations, Pipit can express programs which are currently unsupported by other verified Lustre compilers.
 Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
 Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
-We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that any translated property the user proves on the transition system holds for the original property on the checked semantics.
+We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that the proof obligations generated by the abstract transition system match the checked semantics.
 
-In the future, we intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
-Clocks are important for composing complex systems together and avoiding unnecessary computation.
+In the future, we intend to verify the remainder of the TTCAN driver logic.
+We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
+Clocks are important for composing complex systems together and avoiding unnecessary computation; clocks may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.
 

### 2024-01-16 2024pipit/figures/core-structure.tex
rework missing proof obligations prose

@@
 The labels next to the arrows denote properties of the translation.
-The refinement label ($\subseteq$) indicates a verified abstraction relation; the negated entailment ($\not\vdash$) indicates that the entailment relation --- contracts and checked properties proven on abstract system apply to the checked semantics --- is not yet verified; the equivalence ($=$) denotes a verified equivalence relation.}
+% The refinement label ($\subseteq$) indicates a verified abstraction relation; the negated entailment ($\not\vdash$) indicates that the entailment relation --- that the proof obligations on the abstract transition system entail the original proof obligations --- is not yet verified; the equivalence ($=$) denotes a verified equivalence relation.
+}
 \label{f:core:structure}

### 2024-01-16 2024pipit/section/30-Core.tex
rework missing proof obligations prose

@@
 
-\autoref{f:core:structure} shows the architecture of Pipit.
+\autoref{f:core:structure} shows the high-level architecture of Pipit.
 On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
@@
 The translation from the abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
-One property that still remains to be proven is that if the abstract transition system says a contract or checked property holds, then the same contract or property holds on the original semantics; this is denoted as negated entailment in the figure ($\not\vdash$).
+One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
 The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.

### 2024-01-16 2024pipit/section/20-Motivation.tex
eval: kind2
copy

@@
 Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
-Blessing is an intensional operation: it traverses the expression and updates the internal metadata, but it does not affect the runtime semantics.
+Blessing is an intensional operation that traverses the expression and updates the internal metadata, but does not affect the runtime semantics.
 
@@
 
-We impose three main restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
+We impose three restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
 
 With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
-These properties are straightforward in a theorem prover, but would be difficult to state in a model-checker with a limited specification language.
+These properties are straightforward in a theorem prover, but are difficult to state in a model-checker with a limited specification language.
 
@@
   \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
-  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} )$ \\
+  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} );$ \\
   \> \> @guar@ = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
-  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{now})$ \\
-  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} )$ \\
+  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{time});$ \\
+  \> \> @body@ = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} );$ \\
   \> \} @in@ \\

### 2024-01-16 2024pipit/section/30-Core.tex
copy

@@
 There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
-The translation from the abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
+The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
 One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
@@
 % Our semantics has a simpler proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
-Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a single tuple.
+Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.
 
@@
 Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
-Firstly, when a contract is \emph{defined}, the definer must prove that the body $\ebody$ satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
+Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
 Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
-Conceptually, then, a contract could have two property statuses: one for the definer, and one for the instantiation.
+Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
 However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
@@
 
-For values, we use an auxiliary judgment form $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
-Likewise, for primitives we use the auxiliary judgment form $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
+For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
+Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
 Primitives are pure, non-streaming functions.
@@
 Both rely and guarantee clauses must be boolean expressions.
-Additionally, the guarantee clause can refer to the result value by $x$.
+Additionally, the guarantee clause can refer to the result value as $x$.
 
@@
 This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
-Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based, while the reactive semantics emphasises the finite-state streaming execution of the system.
+Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based and syntax-directed, while the reactive semantics emphasises the finite-state streaming execution of the system.
 We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
-Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital signal processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.
+Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital-signal-processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.
 
@@
 
-Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then use the prim-sem metafunction to apply the primitive.
+Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
 
@@
 Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
-In this case, $\xfby{v}{e}$ proceeds to evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.
+In this case, $\xfby{v}{e}$ evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
 
 Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
-For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding will eventually terminate, as the follow-by shortens the streaming history.
+For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding eventually terminates as each followed-by shortens the history.
 
@@
 
-We also use two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.
+We also define two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.
 
-Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under streaming history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
-This judgment is an iterated application of the single-value big-step form.
+Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
+This judgment performs iterated application of single-value evaluation.
 
-Judgment form $\bigstepalways{\Sigma}{e}$ denotes that expression $e$, which must be a boolean, evaluates to the stream of trues under history $\Sigma$.
-Informally, it can be read as ``in streaming history $\Sigma$, $e$ is always true''.
+Judgment form $\bigstepalways{\Sigma}{e}$ denotes that a boolean expression $e$ evaluates to the stream of trues under history $\Sigma$.
+Informally, it can be read as ``$e$ is always true in history $\Sigma$''.
 
@@
 This sort of feedback is necessary for proving properties of mutually-dependent calls.
-Although this feedback appears circular, we enforce causality by requiring that occurrences of recursive streams are guarded by delays (\autoref{s:core:causality}).
+This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
 
 We define the checked semantics of Pipit in \autoref{f:core-check}.
-The checked semantics mostly follows the structure of the dynamic semantics, checking any properties and contracts as they are encountered.
+The checked semantics mostly follows the structure of the dynamic semantics, additionally checking any properties and contracts as they are encountered.
 
@@
 We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.
-Unfortunately, this rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.
+% This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.
 
@@
 
-Blessing is slightly different for contract definitions, as we need to separate the definition of the contract from the instantiation.
+Blessing is different for contract definitions, as we need to separate the definition of the contract from the instantiation.
 To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
@@
 \begin{theorem}[bigstep-substitute]
-  For a streaming history $\Sigma$ and a causal expression $e$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
+  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
   The converse is also true.
@@
 The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
-An alternative semantics would be to have the environment outside the semantics invent a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
-The above substitution theorem can be used to show that these two semantics are equivalent.
-Thanks to causality, we can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
+An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
+The above substitution theorem can be used to show that, for causal expressions, these two semantics are equivalent.
+We can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
 This fact can be used to ``seed'' evaluation by starting with an arbitrary value:

### 2024-01-16 2024pipit/section/40-Transition.tex
copy

@@
 A transition system is parameterised by its input context and the result type.
-It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to quantify over.
+It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to existentially quantify over.
 The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.
 
-For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: there exists some value that satisfies the specification.
+For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: ``there exists some value that satisfies the specification''.
 Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
-Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block normalisation and result in a more complex transition system.
+Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
 Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.
 
-As usual, the step-result contains the updated state for the transition system, as well as the result value.
+Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
 The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
@@
 For most expression forms, the state and free contexts are defined by taking the union of the contexts of subexpressions.
-Followed-by delays introduce a local state variable $x_{@fby@}$ in which to store the most recent stream value.
+Followed-by delays introduce a local state variable $x_{@fby@(e)}$ in which to store the most recent stream value.
 We generate a fresh variable here, though the implementation uses de Bruijn indices.
-Recursive streams and contracts both introduce new bindings into the free context, assuming that their binders $x$ are unique.
+Recursive streams and contracts both introduce new bindings into the free context; we assume that their binders $x$ are unique.
 
@@
 
-To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@}$ to $v$.
+To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@(e)}$ to $v$.
 At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
@@
 To translate let-expressions $\xlet{x}{e}{e'}$, we extend the input heap with the value of $e$ before evaluating $e'$.
-The presentation here duplicates the computation of the value of $e$, but this is not an issue in practice.
+The presentation here duplicates the computation of the value of $e$, but the actual implementation introduces a single binding.
 
@@
 In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
-The true semantics of the contract, however, only holds if the contract rely is true \emph{at every step so far}.
-This simplification is justified as our definition of validity for a transition system also requires the translation rely to be true at every step.
-This simplification is essentially an application of the $\Box \Box p \implies \Box p$ axiom of modal logic.
-The mechanised proof of this simplification is future work.
+The real semantics of the contract, however, requires the contract rely to be true \emph{at every step so far}.
+This difference is benign, as the inductive case of the proof of transition system validity assumes that both the rely and guarantee held at previous steps; if the rely also holds now, then it has held at every step so far.
+% to prove that a transition system satisfies its guarantee, we show that if the rely is true \emph{at every step so far}, then the guarantee is also true.
+% This simplification is essentially an application of the $\Box \Box p \iff \Box p$ equivalence of modal logic.
+% The mechanised proof of this simplification is future work.
 

### 2024-01-16 2024pipit/section/50-Extraction.tex
copy

@@
 This translation to transition systems is verified to preserve the original semantics.
+The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
+For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
+
 To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
@@
 We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
-This specialisation results in a concrete transition system that fits in the \lowstar{}~\cite{protzenko2017verified} subset of \fstar{}, which can then be extracted to statically-allocated C code.
+This specialisation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.
 

### 2024-01-16 2024pipit/section/60-Evaluation.tex
copy

@@
 
-The enable-acks field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
+The enable-acknowledgements field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
 The transmit fields denote whether to send a reference message or an application-specific message.

### 2024-01-16 2024pipit/section/70-Related.tex
copy

@@
 Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
-As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but is limited to verifying small arrays up to 8 or 16 triggers.
+As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but in our experiments was limited to verifying small arrays up to 8 or 16 triggers.
 Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
@@
 Such a type-based approach promises to allow abstraction of the implementation details.
-However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide be insufficient for verifying the rest of the system.
-For this function, the best specification is likely to include a re-statement of the implementation itself.
+However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide too much and be insufficient for verifying the rest of the system.
+For such functions, the best specification is likely to include a re-statement of the implementation itself.
 
@@
 We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
-Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s C code extraction and the Z3 SMT solver.
+Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s unverified C code extraction and the Z3 SMT solver; in comparison, Vélus' C code generation is verified and does not depend on any SMT solver.
 

### 2024-01-16 2024pipit/section/80-Conclusion.tex
copy

@@
 We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
-Clocks are important for composing complex systems together and avoiding unnecessary computation; clocks may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.
+Clocks are important for composing complex systems together and avoiding unnecessary computation; they may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.
 

### 2024-01-17 2024pipit/section/40-Transition.tex
copy

@@
 \begin{theorem}[translation-abstraction]
-  For a well-typed causal expression $e$ and streaming history $\Sigma$, if $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$, then there exists a sequence of free heaps $\Sigma_{F}$ such that repeated application of the transition system's step results in $v$.
+  For a well-typed causal expression $e$ and streaming history $\Sigma$, if $e$ evaluates to stream $V$ $(\bigsteps{\Sigma}{e}{V})$, then there exists a sequence of free heaps $\Sigma_{F}$ such that repeated application of the transition system's step results in $V$.
 \end{theorem}

### 2024-01-18 2024pipit/section/60-Evaluation.tex
update evaluation

@@
 
-We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, a 32-size array reported multiple runtime errors and did not terminate after several hours.
+We were able to reliably verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, we ran into intermittent runtime errors.
 % Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
@@
 \autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
-We ran these experiments on a 2020 M1 MacBook Air with 16 gigabytes of RAM.
+We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
 Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
@@
   & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
-  size & wall-clock & CPU & wall-clock & CPU & wall-clock & CPU \\
+  size & wall-clock & user-time & wall-clock & user-time & wall-clock & user-time \\
   \hline
-  1 & 2s &  3s  & 6s & 12s & 6s & 6s \\
-  2 & 3s &  3s  & 8s & 12s & 6s & 6s \\
-  4 & 5s & 11s  & 12s & 24s & 6s & 6s \\
-  8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
-  16 & 125s & 276s & \multicolumn{2}{c|}{error} & 6s & 6s \\
-  32 & \multicolumn{2}{c|}{error} & \multicolumn{2}{c|}{error} & 6s & 6s \\
+
+  1 & 1.51s&1.33s
+  & 1.61s&2.26s
+  & 5.28s&5.03s \\
+  2 & 1.50s&1.29s
+  & 1.68s&2.85s
+  & 5.28s&5.03s \\
+  4 & 1.56s&1.59s
+  & 2.10s&5.05s
+  & 5.28s&5.03s \\
+  8 & 1.80s&3.12s
+  & 4.29s&18.12s
+  & 5.28s&5.03s \\
+  16 & 3.41s&12.29s
+  & 16.41s&71.23s
+  & 5.28s&5.03s \\
+  32 & 11.24s&67.27s
+  & 941.64s&3853.16s
+  & 5.28s&5.03s \\
 \end{tabular}
 \caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
-The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.}
+The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
+We were unable to verify arrays of size 64 with Kind2 as our verification timed out after three hours.}
 \label{f:evaluation:kind2-runtime}

### 2024-01-18 2024pipit/section/01-Introduction.tex
fix evaluation: ref to kind2 evaluation

--- 2024-01-11 2024pipit/section/01-Introduction.tex
@@
   \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
-  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver and verifying an abstract model of a key operation (\autoref{s:evaluation}).
 \end{itemize}

### 2024-01-18 2024pipit/section/70-Related.tex
fix evaluation: ref to kind2 evaluation

@@
 Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
-As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but in our experiments was limited to verifying small arrays up to 8 or 16 triggers.
+As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify arrays up to the 64 triggers supported by hardware implementations.
 Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.

### 2024-03-20 2024pipit/section/01-Introduction.tex
contributions fix

--- 2024-01-11 2024pipit/section/01-Introduction.tex
@@
   \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
-  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
 \end{itemize}

### 2024-03-20 2024pipit/reviews/reviews-ecoop24-response.md
reviews

# Outline

We thank the reviewers for their insightful reviews.
A common criticism is the lack of intuition and examples given in the technical presentation, and the lack of context given to motivate the technical development.
We are currently three pages under the page limit; we believe that we have sufficient space and time to address this issue and propose to add further explanations and examples.

Reviewer A observed that we are missing the key definition of k-induction.
We propose to include a description of this in section 4.

Reviewers B and C observed that the evaluation which compares to Kind2 is lacking some critical details.
We propose to expand this section to answer the reviewers' questions.

Additionally, we propose to incorporate the now-finished proof of entailment (Figure 2, edge labelled $\not\vdash$) which shows that proof obligations on the abstract system agree with the checked semantics.
This proof was in-progress at the time of the submission, and is now completed.
The proof exposed some small issues which we also propose to amend in the paper formalism.
We believe that this change strengthens the paper and brings us closer to the long-term goal of an end-to-end verified workflow.

# Proposed changes to address reviewer comments
Concretely, we propose to make the following changes to address reviewer comments.
We describe the proposed changes to include the newly-finished proofs separately at the end of this document.

## Section 2: time-triggered networks

* provide a better description of the TTCAN protocol and driver, as the current one is too terse. It may make sense to move some of the description of the driver from S6 to S2.

## Section 3: core language

* include example of checked semantics for both valid and invalid programs
* include example contract to highlight the streaming / reactive aspects of rely-guarantee contracts

## Section 4: abstract transition systems

* include example of abstract transition system, eg showing transition system of `count_when` program
* include example with contract instantiation to highlight abstraction semantics
* add subsection describing induction and k-induction

## Section 5: executable transition systems

* include example of translated executable transition system
* describe the structure of the generated C code, and highlight how it can be executed in a real-time (streaming) environment

## Section 6: Evaluation
Expand Kind2 evaluation:
* describe encoding issues and minor differences between properties in both systems
* include comparison of required proof size vs runnable code
* clarify wall-clock vs user-time in Figure 11
* expand Figure 11 to show Kind2 timeout on larger arrays (eg up to 128)

# Answers to questions

## Reviewer B

Relating to section 6 on evaluation, reviewer B asks:

> 1. Are Pipit and Kind2 verifying exactly the same properties? If not, why not?

Yes, the properties and proofs are almost the same, except for some encoding issues.
F* supports unbounded natural numbers, while Kind2 only supports integers or fixed-bounded integers, so we need some extra nonnegative conditions in some places.
The `next` function to find the next active trigger is naturally recursive, which requires a rather cumbersome encoding with a temporary array in the Kind2 version.
(Arrays are the only way to express loops / iteration within a single time-step in Kind2.)
We also encode the lemma `lemma_can_reach_next` in Kind2 as a function with a contract: the actual proof of the lemma is handled by the SMT solver in both cases, but explicitly instantiating the lemma with the right arguments is the important part.
This lemma encoding would have a slight runtime overhead if we were to compile and execute it, but it is sufficient for a verification-only example.

> 2.  The paper reports the total file size is about the same. However, it would be helpful to get a sense of how much of the Pipit is proofs vs runnable code.

After stripping whitespace and comments, the Pipit version is 103 lines of code, while the Kind2 version is 146 lines of code.
I'd like to add a disclaimer, however, that while Lustre is inherently a little bit more verbose, the style of Lustre used also favours shorter lines.

For actual runnable code, including type definitions and type signatures, Pipit has 23 lines, while Kind2 has 44 lines.
The majority of the remainder is the description of exactly what is required for a valid schedule (35 for Pipit, 45 Kind2), and the lemma and invariants (17 for Pipit, 47 for Kind2).

It's important to note that, in both cases, the proofs apply to an arbitrary valid schedule.
If we wanted to check for a concrete schedule, we wouldn't need the definition of what constitutes a valid schedule, and both programs would be a lot shorter.

For the paper, we will attempt to unify the whitespace style between the two versions before performing a similar analysis.

> The evaluation discusses two case studies. (One is already published.)

As a small remark, we should note a mistake in the bibliography: that citation (38) refers to an extended abstract that is not "officially published" in the workshop proceedings.
We will amend it to specify that it is an extended abstract.

# Proposed changes to include proof of entailment
In addition to the above changes to address reviewer comments, we would also like to expand the paper to include the proof of entailment and resolve the issues it exposed.
The changes do not affect the structure of the paper and do not diminish the claims at all.

## Section 3: core language

Figure 2:
* Figure 2 and its accompanying description should describe the proof of entailment as finished.

Checked semantics:
* The checked semantics rule for `let` needs to change.  The rule for `let x = e1 in e2` performs unfolding then checks `e2[x := e1]`; however, if the name `x` is not mentioned, then `e1` will be removed as dead code without being checked.  The amended rule checks e1 explicitly.

## Section 4: abstract transition system
In the translation of contracts to abstract transition systems, the rely clause for contracts is missing an implication.
In the submission, the translated assumptions for a contract lets us assume the contract's guarantee's assumptions hold.
However, these assumptions only hold if the contract's rely itself holds.
Concretely, in Figure 9, third-last line:
* `[| e_g |]_{rely}(i, f, s)` becomes `([| e_r|]_{value}(i, f, s) ==> [| e_g |]_{rely}(i, f, s))`

### 2024-03-21 2024pipit/section/30-Core.tex
entailment updates

@@
 The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
-One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
+The translation to abstract systems also generates proof obligations for the checked properties, contract definitions and contract instantiations in the core program; these proof obligations are verified to correspond to the original properties.
+% One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
 The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
@@
 We define the checked semantics of Pipit in \autoref{f:core-check}.
-The checked semantics mostly follows the structure of the dynamic semantics, additionally checking any properties and contracts as they are encountered.
+The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
+For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.
 
@@
 Rule {\sc ChkPrim} checks a primitive application by descending into the subexpressions.
+Similarly, rule {\sc ChkFby} descends into followed-by expressions.
 
-Rules $\textsc{ChkFby}_1$ and $\textsc{ChkFby}_S$ are derived from the structure of the big-step rules  $\textsc{Fby}_1$ and $\textsc{Fby}_S$.
-At an input stream of length one, $\textsc{ChkFby}_1$ asserts that all subproperties hold for the (non-existent) previous values in the stream.
-At subsequent parts of the stream, $\textsc{ChkFby}_S$ discards the most recent element of the stream history and checks the subexpression with the previous inputs.
-% One may imagine replacing both rules for $\xfby{v}{e}$ with a single rule that simply checks $e$ under the same streaming context.
-% However, such a rule would require infinite derivations for recursive streams.
+Rule {\sc ChkRec} checks a recursive-expression $\xrec{x}{e}$ by evaluating the overall expression to a stream of values $V$.
+The rule then extends the streaming environment $\Sigma$ with $x$ bound to the values from $V$; this extended environment is used to descend into the recursive expression.
 
-Rules {\sc ChkRec} and {\sc ChkLet} both perform the same unfolding as the corresponding big-step rules and check the resulting expression.
+Rule {\sc ChkLet} checks a let-expression $\xlet{x}{e}{e'}$ descends into both sub-expressions.
+To check the body $e'$, the rule first evaluates $e$ and extends the streaming environment.
 
@@
 Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
+First, we evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
 Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).

### 2024-03-21 2024pipit/section/50-Extraction.tex
copy
todos

@@
 
+\TODO{include example of translated executable transition system}
+
 This translation to transition systems is verified to preserve the original semantics.
@@
 For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
+
+\begin{theorem}[execution-equivalence]
+  For a well-typed causal expression $e$ and streaming history $\Sigma$, $e$ evaluates to stream $V$ $(\bigsteps{\Sigma}{e}{V})$ if-and-only-if repeated application of the transition system's step on $\Sigma$ also results in $V$.
+\end{theorem}
 
@@
 % We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
+
+\subsection{Generated C code}
+\TODO{describe the structure of the generated C code, and highlight how it can be executed in a real-time (streaming) environment}

### 2024-03-22 2024pipit/section/20-Motivation.tex
todos
minor copy

@@
 \label{s:motivation}
+
+\TODO{provide a better description of the TTCAN protocol and driver, as the current one is too terse. It may make sense to move some of the description of the driver from S6 to S2.}
 
@@
 To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
-The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
+The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
 
@@
   \> @let@ $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
-  \> \> @rely@ = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} );$ \\
+  \> \> @rely@ = $(\lambda \textit{time}.~ \textit{time} = \xfby{0}{(\textit{time} + 1)});$ \\
   \> \> @guar@ = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
@@
 % These kinds of self-dependencies and mutual-dependencies are fairly common in Lustre-style programs: for example, \emph{trigger_fetch}'s invocation of \emph{count_when}, whose input $\textit{inc}$ depends on the previous output $\textit{index}$.
+
+The key distinction between these streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
+In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one microsecond after the time at the previous tick of computation.
+Expressing such a rely in an imperative setting requires some extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

### 2024-03-22 2024pipit/section/30-Core.tex
todos
minor copy

@@
 \label{s:core}
+
+\TODO{include example contract to highlight the streaming / reactive aspects of rely-guarantee contracts}
 
@@
 % To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
+As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero upwards, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
@@
 \input{figures/core-check.tex}
+
+\TODO{include example of checked semantics for both valid and invalid programs}
 

### 2024-03-25 2024pipit/section/20-Motivation.tex
examples

@@
 
-When this function is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
+The key distinction between these streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
+In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one microsecond after the time at the previous tick of computation.
+Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.
+
+When \emph{trigger_fetch} is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
 In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.
@@
 % These kinds of self-dependencies and mutual-dependencies are fairly common in Lustre-style programs: for example, \emph{trigger_fetch}'s invocation of \emph{count_when}, whose input $\textit{inc}$ depends on the previous output $\textit{index}$.
-
-The key distinction between these streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
-In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one microsecond after the time at the previous tick of computation.
-Expressing such a rely in an imperative setting requires some extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

### 2024-03-25 2024pipit/section/30-Core.tex
examples

@@
 \label{s:core}
-
-\TODO{include example contract to highlight the streaming / reactive aspects of rely-guarantee contracts}
 
@@
 % One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
-The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
+The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
 The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.
@@
 
-For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.
+For example, the core expression $\sumrec$ computes the sum of values from a stream of integers $i$ by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
+If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
+$$
+\sumcontract
+$$
+To be considered a valid program, we would have to prove that the contract definition itself holds, similar to our earlier contract in \autoref{s:motivation:contract}.
+The unknown property status here allows us to defer the caller's proof that the input stream is always positive until we know more about the actual call-site.
 
+The remaining grammatical constructs of \autoref{f:core-grammar} describe streams, value environments, types and type environments.
 Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
 Types $\tau$ and type environments $\Gamma$ are standard.
+For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.
 
@@
 
-Rule {\sc Value} states that evaluating a value results in the value itself.
+At a high level, evaluation proceeds by unfolding the recursive streams until a concrete value is determined.
+For example, to evaluate the earlier sum example with input stream $i = [1; 2]$, we start with the judgment:
+$$
+\bigstep{\sumionetwo}{\sumrec}{v}
+$$
 
+First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
+To evaluate variables, we look for the variable in the current, or most recent, heap:
+$$
+\ruleIN{
+}{
+  \bigstep{\sumionetwo}{i}{2}
+}{Var}
+$$
+
+For delays, we discard the current heap and continue evaluation with the history prefix:
+$$
+\ruleIN{
+  \bigstep{\sgl{i \mapsto 1}}{\sumrec}{1}
+}{
+  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
+}{$\mbox{Fby}_S$}
+$$
+
+Returning to \autoref{f:core-bigstep}, rule {\sc Value} states that evaluating a value results in the value itself.
 Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.
+Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
+Rule {\sc Let} is standard.
 
-Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
-
-Rule $\textsc{Fby}_1$ evaluates a followed-by expression when the streaming history contains only a single element.
-Here, $\xfby{v}{e}$ evaluates to $v$, as there is no previous value of $e$ to use.
-
-Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
-In this case, $\xfby{v}{e}$ evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
+For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value or not.
+When there is no previous value -- the streaming history consists of just the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
+Otherwise, rule $\textsc{Fby}_S$ applies.
+In this case, the rule evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
 
@@
 For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding eventually terminates as each followed-by shortens the history.
-
-Rule {\sc Let} is standard.
 
@@
 Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
-% This ensures that the dynamic semantics correspond directly to the program we intend to execute.
+From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
+However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
+The deterministic and total nature of evaluation is key to our proofs and metatheory.
 
@@
 \label{s:core:checked}
-\input{figures/core-check.tex}
-
-\TODO{include example of checked semantics for both valid and invalid programs}
 
@@
 Contracts involve two proofs: one for the definition and one for the instantiation.
-To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
+To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), and assuming that the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee always holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
 Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
@@
 
-To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
-Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
-This sort of feedback is necessary for proving properties of mutually-dependent calls.
+For example, recall our earlier sum contract, which stated that the sum of strictly positive integers is itself positive:
+$$
+@let@~\text{sum}~i = \sumcontract
+$$
+To check the contract definition on a concrete input $i = [1; 2]$, we first evaluate the body:
+$$
+\bigsteps{\sumionetwo}{\sumrec}{[1; 3]}
+$$
+We then check that, assuming all inputs are positive, then all results are positive:
+$$
+\bigstepalways{\sumionetwo}{i > 0} \implies
+\bigstepalways{\sgl{i \mapsto 1, \sumvar \mapsto 1}; \sgl{i \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
+$$
+
+To prove that the contract definition is valid for all inputs, we perform a similar proof for arbitrary input streams ($\forall \Sigma$).
+We assume that the rely is true \emph{at all points} in the stream, and must show that the guarantee is true \emph{at all points}.
+Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
+In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
+At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
+This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.
+
+Contract instantiations often involve feedback loops, where the input to a contract depends, directly or indirectly, on previous output values of the contract itself.
+For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
+$$
+  % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
+  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
+$$
+
+To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
+To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
+We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
+We will describe this proof in more detail in \autoref{s:transition:ind}.
+
+% To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
+% Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
+% This sort of feedback is necessary for proving properties of mutually-dependent calls.
 This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
+For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
+The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
+However, there is no positive number whose negation is also positive, therefore the rely cannot imply the guarantee.
+This potential issue is resolved by outlawing such non-causal programs.
 
-We define the checked semantics of Pipit in \autoref{f:core-check}.
+\input{figures/core-check.tex}
+
+The checked semantics of Pipit is defined in \autoref{f:core-check}.
 The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
@@
 Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
-To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to be true.
+To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
 To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
@@
 % This rule must deal with the two different roles of a contract at once; in the next section, we will separate the two roles.
+
+We do not generally verify programs directly using the checked semantics.
+Instead, the checked semantics is intended as a specification describing what it means to be a valid program.
+To verify concrete programs, we generally translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).
+% The semantics for contracts is fairly subtle, but we believe it is simpler than the translation to transition systems.
+% Having a relatively simple high-level semantics of checks and contracts leaves room for future optimisations in translation to abstract transition systems.
+% Additionally, existing contract support in Kind2~\cite{champion2016kind2} has rather ad-hoc restrictions, such as disallowing calling other contracts inside contract relies and guarantees.
+Our semantics handles this case without any special treatment.
 
@@
 
+% XXX: the above uses substitution, but should use this bigsteps version. however, the above is much neater while below splits over the page.
+% they are equivalent here anyway, so why not use the neater one?
+% \begin{tabbing}
+%   @MM@\= @MMMMM@ \= @MMMMMMMMMMMMM@ \= \kill
+%   @let@ $\text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}: \text{prop}$ = \\
+%   \> $\forall \Sigma~V.$
+%   \> $ (
+%     \semcheck{\Sigma}{\PSValid}{(\erely, \ebody)}
+%     ~\wedge~
+%     \semcheck{\Sigma[x := V]}{\PSValid}{\eguar}
+%     ~\wedge~
+%     \bigstepalways{\Sigma}{\erely}
+%     ~\wedge~
+%     \bigsteps{\Sigma}{\ebody}{V}
+%   ) $ \\
+%   \> $\implies$
+%   \> $(
+%     \semcheck{\Sigma}{\PSUnknown}{(\erely, \ebody)}
+%     ~\wedge~
+%     \semcheck{\Sigma[x := V]}{\PSUnknown}{\eguar}
+%     ~\wedge~
+%     \bigstepalways{\Sigma}{\eguar}
+%     )$
+% \end{tabbing}
+
 After proving that the contract is valid for all inputs, we can bless the contract definition.
@@
 
-To ensure that recursive streams have a computational interpretation, we require that all recursive streams are guarded by a followed-by delay.
+To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
+This restriction checks that all recursive streams are guarded by a followed-by delay.
 We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
@@
 \begin{theorem}[bigstep-substitute]
-  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
+  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ $(\bigsteps{\Sigma}{e'}{V})$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
   The converse is also true.
@@
 
-The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
+The big-step semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
 An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.

### 2024-03-25 2024pipit/section/40-Transition.tex
todos
examples

@@
 
-Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics.
-The proof that the rely and guarantee propositions correspond to the checked semantics is future work.
+For example, recall again the sum contract:
+$$
+\sumcontractX{\PSUnknown}{\textit{ints}}
+$$
 
-% There is a slight difference with how followed-by delays are handled --- the checked semantics checks the delayed expression in the delayed context, while the transition system checks one step ahead ---
+To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
+The followed-by delay results in a local state variable called sum_fby, and the recursive stream results in a free variable called sum.
+We get the following transition system:
+
+  \begin{tabbing}
+  MM \= update: \= = \= update \= \kill
+  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
+  \> state   \> = (sum_fby: $\ZZ$); \\
+  \> free  \> = (sum: $\ZZ$); \\
+  \> init  \> = \{ sum_fby = 0 \}; \\
+  \> step  \> = $\lambda{} i~f~s.$ \{ \\
+  \> \> \> update \> = \{ sum_fby = f.sum \}; \\
+  \> \> \> value  \> = f.sum; \\
+  \> \> \> rely   \> = f.sum = (s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
+  \> \> \> guar   \> = f.sum > 0; \} \\
+  \} \\
+  \end{tabbing}
+
+The initial state of 0 corresponds to the initial value of the followed-by.
+In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containins f.sum, and $s$ refers to the state heap containing s.sum_fby.
+The result of the recursive stream is referred to as f.sum; in the rely / assumption of the step result, f.sum is restricted to have the correct sum.
+The step result's rely also includes the contract's rely, which allows proofs of system validity to assume that the input integer is positive.
+Finally, the step result's guarantee includes the contract's guarantee, which ensures that the proof of system validity actually proves that the contract holds.
+
+The translation for contract instantiations proceeds similarly, except that the contract itself is replaced by an arbitrary value in the free context.
+For example, recall the Fibonacci sequence:
+$
+  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
+$.
+
+This program does not require any input values, so we leave the input context empty.
+The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
+Similarly, the free context includes an entry for the recursive stream, and an entry for the contract itself:
+
+  \begin{tabbing}
+  MM \= update: \= = \= update \= \kill
+  @let@ fib_def: system () $\ZZ$ = \{ \\
+  \> state   \> = (fib_fby: $\ZZ$); \\
+  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
+  \> init  \> = \{ fib_fby = 1 \}; \\
+  \> step  \> = $\lambda{} i~f~s.$ \{ \\
+  \> \> \> update \> = \{ fib_fby = f.fib \}; \\
+  \> \> \> value  \> = f.fib; \\
+  \> \> \> rely   \> = f.fib = f.sum_contract $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
+  \> \> \> guar   \> = s.fib_fby > 0; \} \\
+  \} \\
+  \end{tabbing}
+
+As before, the step result's rely includes the assumption that the recursive stream's value (f.fib) agrees with its result (f.sum_contract).
+Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
+Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- that is, the followed-by's state variable must be positive.
+
+\subsection{Translation}
+
+We now describe the details of the translation before discussing the (k-)inductive proofs of system validity in \autoref{s:transition:ind}.
+Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics, and that the transition system's proof obligations correspond to the checked semantics.
 
@@
 At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
-The rely and guarantee differ from the checked semantics here: in the checked semantics, we check the subexpression on the previous inputs, but here we check the current subexpression.
-This means that a single step of the rely and guarantee do not exactly correspond to the checked semantics; however, we posit that they are equivalent for a rely and guarantee that has been proven to hold for any sequence of inputs.
 
@@
 
+\subsection{Proof obligations and induction}
+\label{s:transition:ind}
+\TODO{add subsection describing induction and k-induction}
+
+\TODO{describe entailment proof briefly}
+
 \subsection{Translation correctness proofs}

### 2024-03-26 2024pipit/section/40-Transition.tex
ok

@@
 Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
-Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- that is, the followed-by's state variable must be positive.
+Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- the followed-by's state variable is positive.
+
+Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
+This minor optimisation is sound, as the proof of system validity requires all steps to satisfy the rely and guarantee; we discuss this further in \autoref{s:transition:ind}.
 
@@
 
-We now describe the details of the translation before discussing the (k-)inductive proofs of system validity in \autoref{s:transition:ind}.
-Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics, and that the transition system's proof obligations correspond to the checked semantics.
+We now present the details of the translation.
+For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations correspond to the checked semantics.
 
@@
 
-In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
-The real semantics of the contract, however, requires the contract rely to be true \emph{at every step so far}.
-This difference is benign, as the inductive case of the proof of transition system validity assumes that both the rely and guarantee held at previous steps; if the rely also holds now, then it has held at every step so far.
-% to prove that a transition system satisfies its guarantee, we show that if the rely is true \emph{at every step so far}, then the guarantee is also true.
-% This simplification is essentially an application of the $\Box \Box p \iff \Box p$ equivalence of modal logic.
-% The mechanised proof of this simplification is future work.
-
 \subsection{Proof obligations and induction}
 \label{s:transition:ind}
-\TODO{add subsection describing induction and k-induction}
 
-\TODO{describe entailment proof briefly}
+To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
+A system satisfies its proof obligations if, for any sequence of steps that satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.
+
+Our definition of induction over a transition system's inputs is somewhat non-standard, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
+This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
+\begin{tabbing}
+  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
+  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
+  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
+  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
+  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
+\end{tabbing}
+
+For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, and assume that the first step satisfied the inductive property:
+\begin{tabbing}
+  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
+  @let@ $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
+  \> $\forall (i_0~ i_1: \textit{input}) (f_0~ f_1: \textit{sys}\text{.free}) (s_0: \textit{sys}\text{.state}).$ \\
+  \> @let@ $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ @in@ \\
+  \> @let@ $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ @in@ \\
+  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
+\end{tabbing}
+
+This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
+K-induction is a fairly standard invariant strengthening technique~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.
+
+To reason about system validity in general, we define a predicate \emph{system_holds_all} that formally defines a valid system as: for all sequences of inputs and their corresponding steps, if all of the steps' relies hold, then the guarantees also hold.
+Validity is implied by (k-)induction.
 
@@
 \end{theorem}
+
+Finally, we can show the main entailment result: that, if the proof obligations hold on the system, then the original program's proof obligations also hold.
+
+\begin{theorem}[translation-entailment]
+  For a well-typed causal expression $e$ and its translated system $s$, if the system holds $(\text{system_holds_all}~s)$, and the checked properties in $e$ hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSValid}{e})$, then
+  the unknown properties in $e$ also hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSUnknown}{e})$
+\end{theorem}
+
+The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
+Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them.

### 2024-03-26 2024pipit/section/50-Extraction.tex
ok

@@
 
-\TODO{include example of translated executable transition system}
+For example, for the \emph{sum} contract with body $\sumrecX{\textit{ints}}$, we generate an executable system that takes an input context containing integer variable \textit{ints}, with a single state variable for the followed-by, and returning an integer:
+
+  \begin{tabbing}
+  MM \= update: \= = \= @let (sum,y) @ \= \kill
+  @let@ sum_def: system (\textit{ints}: $\ZZ$) $\ZZ$ = \{ \\
+  \> state   \> = (sum_fby: $\ZZ$); \\
+  \> init  \> = \{ sum_fby = 0; \}; \\
+  \> step  \> = $\lambda{} i~s.$ \\
+  \> \> \> @let@ $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ @in@ \\
+  \> \> \> @let@ $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ @in@ \\
+  \> \> \> @let@ $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ @in@ \\
+  \> \> \> @let@ $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ @in@ \\
+  \> \> \> $(\textit{sum}_0, s_1)$
+  \end{tabbing}
+
+Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
+The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
+The last two bindings recompute the state, this time with the correct recursive value.
+This duplication of work is often removed by partial evaluation and dead-code-elimination, which are performed during code extraction.
 
@@
 
-The translation for recursive streams described above calls the step function of the sub-stream twice, which can duplicate work.
-The normalisation strategy used to partially-evaluate the translation inlines the two occurrences of the step function, and is often able to remove the duplicate work, but this removal is not guaranteed.
-Our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
+% The generated C code includes a struct containing the
+
+  \begin{tabbing}
+    @struct sum_state { int sum_fby; }@ \\
+    @void   sum_init(struct sum_state* state);@ \\
+    @int    sum_step(struct sum_state* state, int ints);@\\
+  \end{tabbing}
+
+Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
 In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
@@
 % We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
-
-\subsection{Generated C code}
-\TODO{describe the structure of the generated C code, and highlight how it can be executed in a real-time (streaming) environment}

### 2024-03-26 2024pipit/section/50-Extraction.tex
extraction, evaluation

@@
   MM \= update: \= = \= @let (sum,y) @ \= \kill
-  @let@ sum_def: system (\textit{ints}: $\ZZ$) $\ZZ$ = \{ \\
+  @let@ sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
   \> state   \> = (sum_fby: $\ZZ$); \\
@@
   \> \> \> @let@ $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ @in@ \\
-  \> \> \> $(\textit{sum}_0, s_1)$
+  \> \> \> $(\textit{sum}_0, s_1)$ \}
   \end{tabbing}
@@
 
-% The generated C code includes a struct containing the
+The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions; this interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular}:
+  \begin{tabbing}
+    @struct sum_state { uint32_t sum_fby; }@ \\
+    @void   sum_reset(struct sum_state* state);@ \\
+    @int    sum_step(struct sum_state* state, uint32_t ints);@
+  \end{tabbing}
 
-  \begin{tabbing}
-    @struct sum_state { int sum_fby; }@ \\
-    @void   sum_init(struct sum_state* state);@ \\
-    @int    sum_step(struct sum_state* state, int ints);@\\
-  \end{tabbing}
+The reset function takes the pointer to the state struct and sets it to its initial values.
+The step function takes the pointer to the state struct and the inputs, and returns the result integer.
+The state struct is updated in-place.
+The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.
 

### 2024-03-26 2024pipit/section/60-Evaluation.tex
todos
extraction, evaluation

@@
 We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
-The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
+The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $140\mu{}s$, while the average was $90\mu{}s$ with a standard deviation of $1.5\mu{}s$.
 Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
@@
 
-We were able to reliably verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, we ran into intermittent runtime errors.
-% Verification of a 4-size array took 11s of CPU time; an 8-size array took 13s; a 16-size array took 276s; and a 32-size array timed out after several hours and multiple runtime errors.
-For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
+We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
+For example, the specification-only function that finds the next trigger is naturally recursive.
+Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
+For example, given a hypothetical recursive definition of @next@ (left), the array-scan encoding is (right):
 
-We made a critical simplification in the Kind2 implementation, which was to modify the trigger-enabled set to be a single cycle index.
+\begin{minipage}{0.4\textwidth}
+\begin{verbatim}
+next(i) =
+  if enabled(triggers, i, c)
+  then i
+  else if i >= LENGTH - 1
+  then NO_NEXT_TRIGGER
+  else next (i + 1);
+\end{verbatim}
+\end{minipage}
+\begin{minipage}{0.6\textwidth}
+\begin{verbatim}
+next_array[i] =
+  if enabled(triggers, LENGTH - 1 - i, c)
+  then (LENGTH - 1 - i)
+  else if i <= 0
+  then NO_NEXT_TRIGGER
+  else next_array[i - 1];
+next(i) = next_array[LENGTH - 1 - i]
+
+\end{verbatim}
+\end{minipage}
+
+% Ignoring bounds checks, this roughly corresponds to @next(i) = if enabled(i) then i else next(i + 1)@.
+% However, Kind2 does not support recursive functions; instead, we were able to express the recursion using Kind2's array comprehension syntax, which allows one to define an array as a function of its index.
+% This encoding was further complicated by the fact that array comprehensions can only depend on values of the array with lower indices, not ones with a higher index: concretely, we cannot write @next[i] = if enabled(i) then i else next[i + 1]@; instead, we invert the indices: @next[i] = if enabled(LENGTH - 1 - i) then LENGTH - 1 - i else next[i - 1]@.
+
+We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
 In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
@@
 In our Pipit development, we can treat the definition of the cycle set abstractly.
-However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
+However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
 This limitation also makes the specification quite unwieldy, as functions must be manually inlined.
+
+\TODO{include comparison of required proof size vs runnable code}
 
@@
 Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+
+We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 32 elements; above that, our verification timed out after one hour.
+For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
 
@@
   & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
-  size & wall-clock & user-time & wall-clock & user-time & wall-clock & user-time \\
+  size & wall-clock & CPU-time & wall-clock & CPU-time & wall-clock & CPU-time \\
   \hline
 
-  1 & 1.51s&1.33s
-  & 1.61s&2.26s
-  & 5.28s&5.03s \\
-  2 & 1.50s&1.29s
-  & 1.68s&2.85s
-  & 5.28s&5.03s \\
-  4 & 1.56s&1.59s
-  & 2.10s&5.05s
-  & 5.28s&5.03s \\
-  8 & 1.80s&3.12s
-  & 4.29s&18.12s
-  & 5.28s&5.03s \\
-  16 & 3.41s&12.29s
-  & 16.41s&71.23s
-  & 5.28s&5.03s \\
-  32 & 11.24s&67.27s
-  & 941.64s&3853.16s
-  & 5.28s&5.03s \\
+ 1 & 1.48s&1.06s
+ & 1.57s&2.26s
+ & 5.25s&5.03s \\
+ 2 & 1.51s&1.26s
+ & 1.71s&2.93s
+ & 5.25s&5.03s \\
+ 4 & 1.57s&1.62s
+ & 2.08s&4.78s
+ & 5.25s&5.03s \\
+ 8 & 1.76s&3.07s
+ & 4.21s&16.98s
+ & 5.25s&5.03s \\
+ 16 & 3.36s&11.91s
+ & 13.82s&65.57s
+ & 5.25s&5.03s \\
+ 32 & 12.15s&62.38s
+ & 269.14s&1230.05s
+ & 5.25s&5.03s \\
+ 64 & 1701.01s&9096.99s
+ &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
+ 128 &  \multicolumn{2}{c|}{(timeout)}  &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
+
+% VERSION WITH SYSTEM TIME:
+%  1 & 1.48s&1.06s&0.46s
+%  & 1.57s&2.26s&0.84s
+%  & 5.25s&5.03s&0.22s \\
+%  2 & 1.51s&1.26s&0.40s
+%  & 1.71s&2.93s&0.98s
+%  & 5.25s&5.03s&0.22s \\
+%  4 & 1.57s&1.62s&0.44s
+%  & 2.08s&4.78s&1.03s
+%  & 5.25s&5.03s&0.22s \\
+%  8 & 1.76s&3.07s&0.51s
+%  & 4.21s&16.98s&1.67s
+%  & 5.25s&5.03s&0.22s \\
+%  16 & 3.36s&11.91s&0.78s
+%  & 13.82s&65.57s&3.71s
+%  & 5.25s&5.03s&0.22s \\
+%  32 & 12.15s&62.38s&2.43s
+%  & 269.14s&1230.05s&30.93s
+%  & 5.25s&5.03s&0.22s \\
+%  64 & 1701.01s&9096.99s&41.88s
+%  &  \multicolumn{2}{c|}{error}  & 5.25s&5.03s&0.22s \\
+%  128 &  \multicolumn{2}{c|}{error}  &  \multicolumn{2}{c|}{error}  & 5.25s&5.03s&0.22s \\
+
 \end{tabular}
 \caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
+The ``wall-clock time'' denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
 The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
-We were unable to verify arrays of size 64 with Kind2 as our verification timed out after three hours.}
+The time limit was one hour.}
 \label{f:evaluation:kind2-runtime}

### 2024-03-26 2024pipit/section/60-Evaluation.tex
evaluation

@@
 Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
-For example, given a hypothetical recursive definition of @next@ (left), the array-scan encoding is (right):
+Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
+Below we compare a hypothetical unsupported recursive definition of @next@ (left) with the array-scan encoding (right):
 
@@
 next(i) =
-  if enabled(triggers, i, c)
+  if enabled(triggers,i)
   then i
@@
   then NO_NEXT_TRIGGER
-  else next (i + 1);
+  else next(i + 1);
+
 \end{verbatim}
 \end{minipage}
-\begin{minipage}{0.6\textwidth}
+\begin{minipage}{0.55\textwidth}
 \begin{verbatim}
 next_array[i] =
-  if enabled(triggers, LENGTH - 1 - i, c)
-  then (LENGTH - 1 - i)
+  if enabled(triggers,LENGTH - 1 - i)
+  then LENGTH - 1 - i
   else if i <= 0
@@
   else next_array[i - 1];
-next(i) = next_array[LENGTH - 1 - i]
+next = next_array[LENGTH - 1 - index];
 
@@
 \end{minipage}
-
-% Ignoring bounds checks, this roughly corresponds to @next(i) = if enabled(i) then i else next(i + 1)@.
-% However, Kind2 does not support recursive functions; instead, we were able to express the recursion using Kind2's array comprehension syntax, which allows one to define an array as a function of its index.
-% This encoding was further complicated by the fact that array comprehensions can only depend on values of the array with lower indices, not ones with a higher index: concretely, we cannot write @next[i] = if enabled(i) then i else next[i + 1]@; instead, we invert the indices: @next[i] = if enabled(LENGTH - 1 - i) then LENGTH - 1 - i else next[i - 1]@.
 
@@
 
-\TODO{include comparison of required proof size vs runnable code}
-
 \autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
 We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
-Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+Ignoring whitespace and comments, the Pipit implementation has 26 lines of actual executable code, while the Kind2 code has 32.
+The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.
 
-We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 32 elements; above that, our verification timed out after one hour.
+We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
 For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.

### 2024-03-28 2024pipit/section/20-Motivation.tex
ttcan motivation frontload

@@
 
-\TODO{provide a better description of the TTCAN protocol and driver, as the current one is too terse. It may make sense to move some of the description of the driver from S6 to S2.}
+To introduce Pipit, we consider a \emph{time-triggered} network driver, which has a static schedule dictating the network traffic, and which all nodes on the network must adhere to.
+This driver is a simplification of the Time-Triggered Controller Area Network (TTCAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
 
-% Pure streaming languages like Lustre lie in an uneasy space somewhere between pure functions and imperative programs.
-% Like pure functions, pure streaming languages are equational and support reasoning by unfolding and rewriting.
-% Like imperative programs, they describe how state evolves over time.
-% Additionally, Lustre programs tend to be composed of many nested recursive loops.
-% When verifying both pure and imperative programs, recursion or loops generally require explicit loop invariants.
-% Requiring an explicit invariant for each recursive stream quickly becomes unwieldy for Lustre programs.
+At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
+Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
+Actions in the schedule may not be relevant to all nodes, so the node's \emph{node matrix} contains only the relevant actions.
+The node matrix is represented in-memory by a constant \emph{triggers array} containing the relevant triggers, sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
+Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
+Watch alerts are generally placed after an expected reference message and are used to signal an error if no reference message is received.
 
-To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered Controller Area Network (CAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
-% In this system, time is partitioned into a sequence of \emph{cycles}; at the start of each new cycle, the time is reset to zero.
-% Triggers also have a cycle offset and a repeat factor, which together determine the cycles in which the trigger is active.
+\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
+The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
+The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering a failure if the reference message is not received.
+\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.
+
+\begin{figure}
+  \begin{minipage}{0.5\textwidth}
+% \begin{table}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{r|ccccc}
+   & TM0 & TM1 & TM2 & $\cdots$ & TM9 \\
+  \hline
+  C0 & SEND A & SEND B & - & $\cdots$ & WATCH \\
+  C1 & SEND A & - & SEND C & $\cdots$ & WATCH
+\end{tabular}}
+% \end{table}
+\end{minipage}
+\begin{minipage}{0.49\textwidth}
+\scriptsize
+\begin{verbatim}
+ 0:{ time = 0; enabled = {C0,C1}; action = SEND(A); }
+ 1:{ time = 1; enabled = {C0};    action = SEND(B); }
+ 2:{ time = 2; enabled = {C1};    action = SEND(C); }
+ 3:{ time = 9; enabled = {C0,C1}; action = WATCH;   }
+\end{verbatim}
+\end{minipage}
+
+\caption{Left: node matrix; right: corresponding triggers array configuration}
+\label{f:tt-systemmatrix-ok}
+\end{figure}
+
+The network has strict timing requirements, which prohibit the driver from looping through the entire triggers array at each time-mark.
+Instead, the driver maintains an index that refers to the current trigger.
+At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
 
@@
 
-The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark.
-The driver maintains an index that refers to the current trigger.
-At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
-We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
-At each time step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
+We first implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
+At each time-mark, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
 
@@
 
-\subsection{The time-triggered system matrix}
+\subsection{Restrictions on the triggers array}
 
-The schedule of the time-triggered network is abstractly described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
-Each basic cycle is identified by its cycle index and each transmission column has an associated time-mark.
+Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
+To ensure that the schedule can be executed, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next-enabled trigger starts.
 
-\autoref{f:tt-systemmatrix-ok} (left) shows an example system matrix with cycles @C0@ and @C1@ and transmission columns at time-marks 0, 1 and 2.
-For this example, we assume that one message can be sent per clock cycle. % , though in practice a message could take on the order of ten microseconds
-To execute this system matrix, we synchronise the local time to zero at the start of basic cycle @C0@.
-After a basic cycle completes, the nodes on the network synchronise before execution continues to the next basic cycle.
+Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
+We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
+However, we could \emph{not} bring trigger 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
 
-\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding configuration for the triggers array.
-The enabled set denotes the basic cycles for which a trigger is active.
-
-The system has strict timing requirements which restrict how triggers can be defined.
-In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
-Additionally, the schedule must allow sufficient time for the driver to skip over the disabled triggers.
-Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2, as triggers 1 and 2 have distinct cycles.
-However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per tick, and it takes two steps to reach trigger 2 from the start of the array.
-
-\begin{figure}
-  \begin{minipage}{0.38\textwidth}
-\begin{tabular}{r|ccc}
-   & TM0 & TM1 & TM2 \\
-  \hline
-  C0 & MSG A & MSG B & - \\
-  C1 & MSG A & - & MSG C
-\end{tabular}
-\end{minipage}
-\begin{minipage}{0.6\textwidth}
-\small
-\begin{verbatim}
-  0: { time_mark = 0; enabled = {C0,C1}; msg = A; }
-  1: { time_mark = 1; enabled = {C0};    msg = B; }
-  2: { time_mark = 2; enabled = {C1};    msg = C; }
-\end{verbatim}
-\end{minipage}
-
-\caption{Left: system matrix; right: corresponding triggers array configuration}
-\label{f:tt-systemmatrix-ok}
-\end{figure}
-
-% The time-marks are strict deadlines on the order of microseconds, which prohibits the driver from looping through the whole array at each time step to find the next active trigger.
-% Instead, the driver maintains an index that refers to the current trigger; at each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
-% Multiple triggers can have the same time-mark, as long as they are active in different cycles; triggers that are active on the same cycle must have a sufficiently wide gap in their time-marks to allow the driver time to iterate through the inactive triggers between them and reach the next active trigger before it starts.
-% The implementation itself is simple, but formalising exactly what constitutes a ``sufficiently wide gap'' and proving that the schedule is never late requires care.
-
-We impose three restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
+We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
 
@@
 These properties are straightforward in a theorem prover, but are difficult to state in a model-checker with a limited specification language.
-
-% \begin{figure}
-%   \begin{minipage}{0.38\textwidth}
-% \begin{tabular}{r|c}
-%    & TM0 \\
-%   \hline
-%   C0 & MSG A \\
-%   C1 & MSG B
-% \end{tabular}
-% \end{minipage}
-% \begin{minipage}{0.6\textwidth}
-%   \small
-%   \begin{verbatim}
-%     { time_mark =  0; enabled = {C0}; msg = A; }
-%     { time_mark =  0; enabled = {C1}; msg = B; }
-%   \end{verbatim}
-%   \end{minipage}
-
-% \caption{Invalid trigger configuration}
-% \label{f:tt-systemmatrix-bad}
-% \end{figure}
 
@@
     \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} @ in@$ \\
-    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
+    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index})$; \\
     \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
@@
 
-We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 indicates that the lemma is being applied to one streaming argument (the index).
+We use the $\textit{pose}$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it.
 % The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
@@
 To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
-The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
+The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per time-mark --- then we guarantee that we never encounter a late trigger.
 

### 2024-03-28 2024pipit/figures/core-structure.tex
entailment updates
copy and tighten

--- 2024-01-16 2024pipit/figures/core-structure.tex
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/core-structure.pdf}
+  \includegraphics[width=\textwidth]{figures/core-structure-1920x450.pdf}
 \caption{Architecture of Pipit.
@@
 The white boxes and dashed arrows are trusted components.
-The labels next to the arrows denote properties of the translation.
-The refinement label ($\subseteq$) indicates a verified abstraction relation; the negated entailment ($\not\vdash$) indicates that the entailment relation --- contracts and checked properties proven on abstract system apply to the checked semantics --- is not yet verified; the equivalence ($=$) denotes a verified equivalence relation.}
+The labels denote verified properties of the translation: abstraction~($\subseteq$), entailment of proof obligations~($\vdash$), and equivalence~($=$).
+}
 \label{f:core:structure}

### 2024-03-28 2024pipit/section/01-Introduction.tex
copy and tighten

--- 2024-01-13 2024pipit/section/01-Introduction.tex
@@
 
-To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
+To be confident that our proofs hold on the real system, we need a single shared semantics for the compiler and the prover.
 In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
@@
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
-  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
+  \item we describe a \emph{checked semantics} for Pipit; after checking deferred properties, programs are \emph{blessed}, which marks their properties as valid (\autoref{s:core:checked});
   \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
@@
   \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
-  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver, which we have partially verified (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
 \end{itemize}

### 2024-03-28 2024pipit/section/20-Motivation.tex
copy and tighten

@@
 At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
-Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
-Actions in the schedule may not be relevant to all nodes, so the node's \emph{node matrix} contains only the relevant actions.
-The node matrix is represented in-memory by a constant \emph{triggers array} containing the relevant triggers, sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
+Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
+Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
+The node matrix is represented by a constant \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
 Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
-Watch alerts are generally placed after an expected reference message and are used to signal an error if no reference message is received.
+Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.
 
@@
 
-We first implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
-At each time-mark, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
+We implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
+At each step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the counter as-is.
 
@@
 We can now implement the trigger-fetch logic, which keeps track of the current trigger.
-The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
-We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
+We use the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
+We simplify our presentation here and only consider a constant cycle: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
 
@@
 To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
-The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per time-mark --- then we guarantee that we never encounter a late trigger.
+The contract we state is that if we are called once per time-mark then we guarantee that we never encounter a late trigger.
+% The contract we state is that if the environment ensures that the time progresses as expected --- that is, we are called once per time-mark --- then we guarantee that we never encounter a late trigger.
 
@@
 
-The key distinction between these streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
-In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one microsecond after the time at the previous tick of computation.
+The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
+In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous tick of computation.
 Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.

### 2024-03-28 2024pipit/section/30-Core.tex
ok
copy and tighten

@@
 % \pagebreak
-\section{Core language}
+\section{Pipit core language}
 \label{s:core}
@@
 The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
-The translation to abstract systems also generates proof obligations for the checked properties, contract definitions and contract instantiations in the core program; these proof obligations are verified to correspond to the original properties.
+The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
 % One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
@@
 
-Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
+Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
 % To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero upwards, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
+As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
@@
 $$
-To be considered a valid program, we would have to prove that the contract definition itself holds, similar to our earlier contract in \autoref{s:motivation:contract}.
-The unknown property status here allows us to defer the caller's proof that the input stream is always positive until we know more about the actual call-site.
+To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
+The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.
 
@@
 We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
-We will describe this proof in more detail in \autoref{s:transition:ind}.
+We describe the mechanisms of this proof in \autoref{s:transition}.
 

### 2024-03-28 2024pipit/section/30-Core.tex
copy and tighten

@@
 
-For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
-Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
+We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
 Primitives are pure, non-streaming functions.
@@
 
-Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.
+Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.
 
@@
 
-Rule \textsc{TCheck} states that statically checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.
+Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.
 
-Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
+Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
 The overall expression has result type $\tau$.
-Both rely and guarantee clauses must be boolean expressions.
-Additionally, the guarantee clause can refer to the result value as $x$.
+Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.
 
@@
 
-At a high level, evaluation proceeds by unfolding the recursive streams until a concrete value is determined.
-For example, to evaluate the earlier sum example with input stream $i = [1; 2]$, we start with the judgment:
+At a high level, evaluation unfolds recursive streams to determine a concrete value.
+For example, to evaluate the earlier sum example with input $i = [1; 2]$, we start with the judgment:
 $$
@@
 First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
-To evaluate variables, we look for the variable in the current, or most recent, heap:
+Evaluation of primitives is standard.
+For variables, we find the variable in the current heap:
 $$
@@
 
-Returning to \autoref{f:core-bigstep}, rule {\sc Value} states that evaluating a value results in the value itself.
-Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.
-Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
-Rule {\sc Let} is standard.
+Returning to \autoref{f:core-bigstep},
+rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
+Rules {\sc Value} and {\sc Let} are standard.
+Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.
+% rule {\sc Value} states that evaluating a value results in the value itself.
+% Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
+% Rule {\sc Let} is standard.
 
-For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value or not.
-When there is no previous value -- the streaming history consists of just the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
-Otherwise, rule $\textsc{Fby}_S$ applies.
-In this case, the rule evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
+For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
+When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
+Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.
 
@@
 
-Rule {\sc Check} states that check expressions always evaluate to unit.
-We do not perform a dynamic check that the property is true here; checking the truth of properties is dealt with in the checked semantics (\autoref{s:core:checked}).
+Rule {\sc Check} ignores the property when evaluating check expressions.
+We do not dynamically check the property here; this is done in the checked semantics (\autoref{s:core:checked}).
 
-Rule {\sc Contract} states that contracts evaluate by just evaluating their body.
-Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
+Similarly, rule {\sc Contract} ignores preconditions and postconditions when evaluating contracts.
 From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
@@
 In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
-We call these the \emph{checked} semantics.
-Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.
+We call these the \emph{checked} semantics; they are comparable to checking runtime assertions.
+% Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.
 
-The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties of $e$ with status $\PStatus$ hold.
+The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties and contracts of $e$ with status $\PStatus$ hold.
 The property status dictates which properties should be checked and which should be ignored.
 
-To show that an expression $e$'s unknown properties hold, we prove that for all streaming histories $\Sigma$, assuming the valid properties hold ($\semcheck{\Sigma}{\PSValid}{e}$), then the unknown properties ($\semcheck{\Sigma}{\PSUnknown}{e}$) hold.
-The assumption here means that we do not have to re-check properties after proving them once.
+We consider a program to be \emph{valid} if its checks hold for all histories ($\forall \Sigma.~ \semcheck{\Sigma}{\PSValid}{e}$).
+The checked semantics are a specification describing what it means to be a valid program.
+We do not generally verify programs directly using the checked semantics; instead, we translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).
 
-Contracts involve two proofs: one for the definition and one for the instantiation.
-To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), and assuming that the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee always holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
-Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
-The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
+% To show that an expression $e$'s unknown properties hold, we show that for all streaming histories $\Sigma$, the unknown properties hold ($\semcheck{\Sigma}{\PSUnknown}{e}$).
+% When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
+To check a property $(\xcheckP{\PStatus}{e})$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.
+
+Checking contracts is more involved.
+Conceptually, contracts involve two kinds of checks: one for the definition and one for the call-site.
+To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
+% The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
 
@@
 
-For example, recall our earlier sum contract, which stated that the sum of strictly positive integers is itself positive:
+For example, recall our earlier contract that the sum of strictly positive integers is positive:
 $$
@@
 
-To prove that the contract definition is valid for all inputs, we perform a similar proof for arbitrary input streams ($\forall \Sigma$).
-We assume that the rely is true \emph{at all points} in the stream, and must show that the guarantee is true \emph{at all points}.
+It is crucial that the rely is true \emph{at all points} in the stream.
 Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
@@
 
-Contract instantiations often involve feedback loops, where the input to a contract depends, directly or indirectly, on previous output values of the contract itself.
-For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
-$$
-  % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
-  @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
-$$
+To check a contract instance, we need to check whether the call-site satisfies the rely.
 
-To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
-To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
-We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
-We describe the mechanisms of this proof in \autoref{s:transition}.
+% Contract instantiations often involve feedback loops, where the input to a contract depends on previous output values of the contract itself.
+% For example, we can compute the Fibonacci sequence by instantiating the sum contract with its output fed back to it\footnote{The program shown here does not strictly fit within the core grammar, which does not support named functions; in the implementation we use the \fstar{} meta-language's top-level definitions for such definitions.}:
+% $$
+%   % @rec@~\fibvar.\, \sumcontractX{\PSUnknown}{(\xfby{1}{\fibvar})}
+%   @rec@~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
+% $$
+
+% To prove this instantiation valid, we need to show that the input $(\xfby{1}{\fibvar})$ is positive.
+% To do this, we perform induction on the input streams, allowing us to assume that the rely was previously true.
+% We then use the contract to show that, if the rely was true at the previous step, then the guarantee was true at the previous step -- which implies that the previous result is positive.
 
@@
 % This sort of feedback is necessary for proving properties of mutually-dependent calls.
-This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
-For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
-The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
-However, there is no positive number whose negation is also positive, therefore the rely cannot imply the guarantee.
-This potential issue is resolved by outlawing such non-causal programs.
+% This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
+% For example, the non-causal program $@rec@~\textit{nope}.\, \text{sum}~(-\textit{nope})$ depends instantaneously on itself.
+% The rely requires that the negated result is positive; the guarantee ensures that the result is positive.
+% However, there is no positive number whose negation is also positive, which suggests that the contract itself is inconsistent.
+% This potential issue is resolved by outlawing such non-causal programs.
 
@@
 The checked semantics of Pipit is defined in \autoref{f:core-check}.
-The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
-For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.
+% The checked semantics descends into the expression and checks the properties and contracts as they are encountered.
+% For binding forms such as let-expressions and recursive-expressions, the semantics evaluates the corresponding expression and extends the streaming history with the bound variable.
 
@@
 
-Rule {\sc ChkCheck} applies when checking property status $\PStatus$ of an expression $\xcheckP{\PStatus'}{e}$.
-If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we perform the actual check by evaluating the expression $e$ and requiring it to evaluate to a stream of trues.
-Otherwise, we do not need to evaluate the check-expression.
-In both cases, we descend into the expression and check its subexpressions, as they may have nested properties.
-Such nested properties are unlikely to be written directly by the user, but might occur after program transformations such as inlining.
+Rule {\sc ChkCheck} checks the properties marked $\PStatus$ in an expression $\xcheckP{\PStatus'}{e}$.
+If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we evaluate the expression $e$ and require it to be true at all steps.
+We then unconditionally descend into the subexpression to check any nested properties.
+Such nested properties are unlikely to be written directly by the user, but might occur after inlining.
 
 Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
-First, we evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
-Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
+This rule checks both the contract definition and the call-site.
+We evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
+Although the contract only has one property status, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition (assumed to be $\PSValid$).
 To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
@@
 
-We do not generally verify programs directly using the checked semantics.
-Instead, the checked semantics is intended as a specification describing what it means to be a valid program.
-To verify concrete programs, we generally translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).
 % The semantics for contracts is fairly subtle, but we believe it is simpler than the translation to transition systems.
@@
 Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
-Blessing an expression requires a proof that the checked semantics hold for all input streams:
+% Blessing an expression requires a proof that the unknown checks hold:
+Blessing an expression requires a proof that, for all input streams, assuming the valid checks hold, then the unknown checks hold:
 
@@
 $$
+
+We generally prove these properties via abstract transition systems (\autoref{s:transition}).
 

### 2024-03-28 2024pipit/section/40-Transition.tex
copy and tighten

@@
 
-For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: ``there exists some value that satisfies the specification''.
-Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
+For recursive streams and contract instantiations, which hide their implementation, the natural translation to a transition system would involve existentially quantifying a result that satisfies the specification.
+Unfortunately, using an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
 Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
@@
 To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
-The followed-by delay results in a local state variable called sum_fby, and the recursive stream results in a free variable called sum.
-We get the following transition system:
+The followed-by delay results in a local state variable called sum_fby, and we encode the existentially-quantified recursive stream as a free context variable called sum:
 
@@
   \> \> \> value  \> = f.sum; \\
-  \> \> \> rely   \> = f.sum = (s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
-  \> \> \> guar   \> = f.sum > 0; \} \\
-  \} \\
+  \> \> \> rely   \> = (f.sum = s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
+  \> \> \> guar   \> = f.sum > 0; \} \}
   \end{tabbing}
@@
 The initial state of 0 corresponds to the initial value of the followed-by.
-In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containins f.sum, and $s$ refers to the state heap containing s.sum_fby.
-The result of the recursive stream is referred to as f.sum; in the rely / assumption of the step result, f.sum is restricted to have the correct sum.
-The step result's rely also includes the contract's rely, which allows proofs of system validity to assume that the input integer is positive.
-Finally, the step result's guarantee includes the contract's guarantee, which ensures that the proof of system validity actually proves that the contract holds.
+In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containing the recursive stream f.sum, and $s$ refers to the state heap containing s.sum_fby.
+In the rely of the step result, f.sum is constrained to match the body of the recursive stream.
+The step result's rely also includes the contract's rely that the input integer is positive.
+Finally, the step result's guarantee includes the contract's guarantee that the contract holds.
 
-The translation for contract instantiations proceeds similarly, except that the contract itself is replaced by an arbitrary value in the free context.
-For example, recall the Fibonacci sequence:
+To verify the transition system, we prove inductively that if the rely always holds, then the guarantee holds; we discuss proofs of system validity further in \autoref{s:transition:ind}.
+
+The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
+For example, we can use the sum contract to implement the Fibonacci sequence with
 $
@@
 $.
-
 This program does not require any input values, so we leave the input context empty.
@@
   \> \> \> value  \> = f.fib; \\
-  \> \> \> rely   \> = f.fib = f.sum_contract $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
-  \> \> \> guar   \> = s.fib_fby > 0; \} \\
-  \} \\
+  \> \> \> rely   \> = (f.fib = f.sum_contract) \\
+  \> \> \>        \> $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
+  \> \> \> guar   \> = s.fib_fby > 0; \} \}
   \end{tabbing}
@@
 Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
-This minor optimisation is sound, as the proof of system validity requires all steps to satisfy the rely and guarantee; we discuss this further in \autoref{s:transition:ind}.
+This minor optimisation is sound, as we define system validity to require all steps to satisfy the rely.
 
@@
 
-Our definition of induction over a transition system's inputs is somewhat non-standard, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
+Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
 This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
@@
 This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
-K-induction is a fairly standard invariant strengthening technique~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.
+K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.
 
@@
 The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
+With this invariant, we can prove abstraction:
 

### 2024-03-28 2024pipit/section/50-Extraction.tex
copy and tighten

@@
 The last two bindings recompute the state, this time with the correct recursive value.
-This duplication of work is often removed by partial evaluation and dead-code-elimination, which are performed during code extraction.
+This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.
 
@@
 The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
-For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
+For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we use the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
 
@@
 The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
-We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
-This specialisation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.
+We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to partially-evaluate the application of the translation to a particular input program.
+This partial-evaluation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.
 
@@
 
-Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
-In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
+Unfortunately, our current approach is unsuitable for generating imperative array code, as our pure transition system only supports pure arrays.
+In the future, we intend to support efficient array computations and fix the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
 Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.

### 2024-03-28 2024pipit/section/60-Evaluation.tex
copy and tighten

@@
 
-To evaluate Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
-The CAN bus is commonly found in safety-critical automotive and industrial settings.
-The time-triggered network architecture defines a static schedule of network traffic.
-All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.
-
-At a high level, the schedule is described by a system matrix which consists of rows of basic cycles.
-Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
-Actions in the schedule may not be relevant to all nodes, so each node has its own local array containing the relevant triggers; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
-The trigger action for receiving an application-specific message checks that a particular message has been received since the trigger was last executed; depending on this, the driver increments or decrements a message-status-counter, which will in turn signal an error once the upper limit is reached.
-Reference messages start a new basic cycle and are used to synchronise the nodes.
-Watch alerts are generally placed after the expected end of the cycle and are used to signal an error if no reference message is received.
+To evaluate Pipit, we have implemented the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver~\cite{ISO11898_4}, described earlier in \autoref{s:motivation}.
+The CAN bus is common in safety-critical automotive and industrial settings.
+The time-triggered network architecture defines a static schedule of network traffic;
+by having all nodes on the network adhere to the schedule, the reliability of periodic messages is significantly increased~\cite{fuehrer2001time}.
 
 The TTCAN protocol can be implemented in two levels of increasing complexity.
-In the first level, reference messages contain the index of the newly-started cycle.
+In the first level, reference messages, which perform synchronisation between nodes, contain the index of the newly-started cycle.
 In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
@@
 We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
-In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
+In TTCAN proper, the enabled set is implemented as a cycle-offset and repeat-factor.
 Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
@@
 However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
-This limitation also makes the specification quite unwieldy, as functions must be manually inlined.
+This limitation also makes the specification quite unwieldy, as we must manually inline any functions in quantified formulas.
 
@@
 Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
-Ignoring whitespace and comments, the Pipit implementation has 26 lines of actual executable code, while the Kind2 code has 32.
+Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
 The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.
@@
   & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
-  size & wall-clock & CPU-time & wall-clock & CPU-time & wall-clock & CPU-time \\
+  size & wall-clock & CPU time & wall-clock & CPU time & wall-clock & CPU time \\
   \hline
@@
 \caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
-The ``wall-clock time'' denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
+The wall-clock time denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
 The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.

### 2024-03-28 2024pipit/section/70-Related.tex
copy and tighten

--- 2024-01-16 2024pipit/section/70-Related.tex
@@
 Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
-As described in \autoref{s:evaluation}, the Kind2 does include experimental array and quantifier support, but in our experiments was limited to verifying small arrays up to 8 or 16 triggers.
-Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
+As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
+Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
 
 Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
-It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
+It may be possible to encode the quantifiers as fixed-size loops in those that support arrays, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
 

### 2024-03-28 2024pipit/section/80-Conclusion.tex
copy and tighten

@@
 Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
-We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that the proof obligations generated by the abstract transition system match the checked semantics.
+We also introduced a checked semantics, which describes the semantics of checked properties and contracts; proof obligations generated by translation to abstract transition system are verified to correspond to these semantics.
 

### 2024-03-29 2024pipit/section/20-Motivation.tex
copy

@@
 Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
-The node matrix is represented by a constant \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
+The node matrix is represented in memory by a \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
 Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
@@
 
-The network has strict timing requirements, which prohibit the driver from looping through the entire triggers array at each time-mark.
+The network has strict timing requirements which prohibit the driver from looping through the entire triggers array at each time-mark.
 Instead, the driver maintains an index that refers to the current trigger.
@@
 Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
-To ensure that the schedule can be executed, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next-enabled trigger starts.
+To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.
 
@@
 We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
-However, we could \emph{not} bring trigger 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
+However, we could \emph{not} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
 

### 2024-03-29 2024pipit/section/30-Core.tex
copy

@@
 % To ensure that streams are productive,
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
+As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
@@
 
-Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
+Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
+Contracts involve two verification conditions.
 Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
@@
 
-For example, the core expression $\sumrec$ computes the sum of values from a stream of integers $i$ by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
+For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
 If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
@@
 
-At a high level, evaluation unfolds recursive streams to determine a concrete value.
-For example, to evaluate the earlier sum example with input $i = [1; 2]$, we start with the judgment:
+At a high level, evaluation unfolds recursive streams to determine a value.
+For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
 $$
@@
 
-First, we unfold the recursive stream to get $(\xfby{0}{\sumrec}) + i$.
+First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
 Evaluation of primitives is standard.
-For variables, we find the variable in the current heap:
+To evaluate variables, we look for the variable in the current heap:
 $$
@@
 }{
-  \bigstep{\sumionetwo}{i}{2}
+  \bigstep{\sumionetwo}{\textit{ints}}{2}
 }{Var}
@@
 \ruleIN{
-  \bigstep{\sgl{i \mapsto 1}}{\sumrec}{1}
+  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
 }{
@@
 % When performing this proof, we can assume that the valid properties already hold to avoid re-checking properties ($\semcheck{\Sigma}{\PSValid}{e}$).
-To check a property $(\xcheckP{\PStatus}{e})$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.
+To check a property $(\xcheckP{\PStatus}{e})$ in history $\Sigma$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.
 
@@
 $$
-@let@~\text{sum}~i = \sumcontract
+@let@~\text{sum}~\textit{i} = \sumcontractX{\PSUnknown}{i}
 $$
-To check the contract definition on a concrete input $i = [1; 2]$, we first evaluate the body:
+To check the contract definition on a concrete input $\textit{i} = [1; 2]$, we first evaluate the body:
 $$
-\bigsteps{\sumionetwo}{\sumrec}{[1; 3]}
+\bigsteps{\sumionetwo}{\sumrecX{i}}{[1; 3]}
 $$
@@
 $$
-\bigstepalways{\sumionetwo}{i > 0} \implies
-\bigstepalways{\sgl{i \mapsto 1, \sumvar \mapsto 1}; \sgl{i \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
+\bigstepalways{\sumionetwo}{\textit{i} > 0} \implies 
+\bigstepalways{\sgl{\textit{i} \mapsto 1, \sumvar \mapsto 1}; \sgl{\textit{i} \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
 $$
 
-It is crucial that the rely is true \emph{at all points} in the stream.
-Consider if we had instead used the input stream $i = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
+It is critical that the rely is true \emph{at all points} in the stream.
+Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
 In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
 At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
+The contract itself remains valid, as the input did not satisfy the rely at all steps.
 This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.
 
-To check a contract instance, we need to check whether the call-site satisfies the rely.
+Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.
 
@@
 
-We generally prove these properties via abstract transition systems (\autoref{s:transition}).
+We generally prove the required properties via abstract transition systems (\autoref{s:transition}).
 

### 2024-03-30 2024pipit/section/30-Core.tex
good

@@
 
-We generally prove the required properties via abstract transition systems (\autoref{s:transition}).
+We generally prove the required properties by first translating the program to an abstract transition system, as described in \autoref{s:transition}.
 
@@
 We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
-This check is stricter than necessary: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ does mention the recursive stream $x$ outside of the delay, but after inlining the let, it would be causal.
-We hope to relax this restriction somewhat in future work.
+This check ensures productivity of recursive streams, but can be too strict: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ mentions the recursive stream $x$ outside of the delay and is outlawed, but after inlining the let, it would be causal.
+We hope to relax this restriction in future work.
 

### 2024-03-30 2024pipit/section/40-Transition.tex
good

@@
 The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.
+% The type $\textit{heap}~c$, for some context c, refers to the type of heaps containing those values.
 
@@
 Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
-The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
-For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.
+The step-result additionally contains two propositions; one for the `rely', or assumptions about the execution environment, and another for the `guarantee', or obligations that the transition system must show.
+For the transition system corresponding to an expression $e$, these propositions are roughly analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.
 
@@
 $$
-\sumcontractX{\PSUnknown}{\textit{ints}}
+@let@~\text{sum}~\textit{ints}~=~\sumcontractX{\PSUnknown}{\textit{ints}}
 $$
@@
   \> step  \> = $\lambda{} i~f~s.$ \{ \\
-  \> \> \> update \> = \{ sum_fby = f.sum \}; \\
-  \> \> \> value  \> = f.sum; \\
-  \> \> \> rely   \> = (f.sum = s.sum_fby + i.ints) $\wedge$ i.ints > 0; \\
-  \> \> \> guar   \> = f.sum > 0; \} \}
+  \> \> \> update \> = \{ sum_fby = \textit{f}.sum \}; \\
+  \> \> \> value  \> = \textit{f}.sum; \\
+  \> \> \> rely   \> = (\textit{f}.sum = \textit{s}.sum_fby + \textit{i}.ints) $\wedge$ \textit{i}.ints > 0; \\
+  \> \> \> guar   \> = \textit{f}.sum > 0; \} \}
   \end{tabbing}
@@
 The initial state of 0 corresponds to the initial value of the followed-by.
-In the step function, argument $i$ refers to the input heap containing i.ints, $f$ refers to the free heap containing the recursive stream f.sum, and $s$ refers to the state heap containing s.sum_fby.
-In the rely of the step result, f.sum is constrained to match the body of the recursive stream.
-The step result's rely also includes the contract's rely that the input integer is positive.
-Finally, the step result's guarantee includes the contract's guarantee that the contract holds.
+In the step function, argument $i$ refers to the input heap containing \textit{i}.ints, $f$ refers to the free heap containing the recursive stream \textit{f}.sum, and $s$ refers to the state heap containing \textit{s}.sum_fby.
+In the rely of the step result, \textit{f}.sum is constrained to be the translated body of the recursive stream.
+The translated rely also includes the contract's rely that the input integer is positive.
+Finally, the translated guarantee includes the contract's guarantee that the output is positive.
 
@@
 The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
-Similarly, the free context includes an entry for the recursive stream, and an entry for the contract itself:
+Similarly, the free context includes an entry for the recursive stream, and an entry for the abstract, underspecified value of the contract:
 
@@
   \> step  \> = $\lambda{} i~f~s.$ \{ \\
-  \> \> \> update \> = \{ fib_fby = f.fib \}; \\
-  \> \> \> value  \> = f.fib; \\
-  \> \> \> rely   \> = (f.fib = f.sum_contract) \\
-  \> \> \>        \> $\wedge$ (s.fib_fby > 0 $\implies$ f.sum_contract > 0); \\
-  \> \> \> guar   \> = s.fib_fby > 0; \} \}
+  \> \> \> update \> = \{ fib_fby = \textit{f}.fib \}; \\
+  \> \> \> value  \> = \textit{f}.fib; \\
+  \> \> \> rely   \> = (\textit{f}.fib = \textit{f}.sum_contract) \\
+  \> \> \>        \> $\wedge$ (\textit{s}.fib_fby > 0 $\implies$ \textit{f}.sum_contract > 0); \\
+  \> \> \> guar   \> = \textit{s}.fib_fby > 0; \} \}
   \end{tabbing}
 
-As before, the step result's rely includes the assumption that the recursive stream's value (f.fib) agrees with its result (f.sum_contract).
-Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
-Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- the followed-by's state variable is positive.
+As before, the translated rely includes the assumption that the recursive stream's value (\textit{f}.fib) agrees with its body (\textit{f}.sum_contract).
+Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (\textit{s}.fib_fby) is positive, then its output (\textit{f}.sum_contract) is positive too.
+Finally, the translated guarantee encodes the obligation that the environment satisfies the \emph{contract's rely} -- the input to sum is positive.
 
@@
 We now present the details of the translation.
-For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations correspond to the checked semantics.
+For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations imply that the original expression satisfies the checked semantics.
 
@@
 \autoref{f:system-translation} defines the translation for expressions.
-Values and expressions have no internal state.
+Values and variables have no internal state.
 For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
@@
 
-To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{@fby@(e)}$ to $v$.
-At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
+To translate a followed-by $\xfby{v}{e}$, we initialise the followed-by's unique binder $x_{@fby@(e)}$ to the followed-by's default value $v$.
+At each step, we return the value in the local state \emph{before} updating the local state to the subexpression's new value.
 
@@
 To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
-A system satisfies its proof obligations if, for any sequence of steps that satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.
+A system satisfies its proof obligations if, for any sequence of steps that all satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.
 
 Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
-This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
+This means that the base case must take a single step from the initial state to be able to state the property that, if the step result's rely holds, then its guarantee holds:
 \begin{tabbing}
@@
   @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
-  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
+  \> $\forall (i: \text{heap}~\textit{input}) (f: \text{heap}~\textit{sys}\text{.free}).$ \\
   \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
@@
 
-For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, and assume that the first step satisfied the inductive property:
+For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, assuming that both steps satisfy the rely and the first step satisfied the inductive property:
 \begin{tabbing}
@@
   @let@ $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
-  \> $\forall (i_0~ i_1: \textit{input}) (f_0~ f_1: \textit{sys}\text{.free}) (s_0: \textit{sys}\text{.state}).$ \\
+  \> $\forall (i_0~ i_1: \text{heap}~\textit{input}) (f_0~ f_1: \text{heap}~\textit{sys}\text{.free}) (s_0: \text{heap}~\textit{sys}\text{.state}).$ \\
   \> @let@ $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ @in@ \\
@@
 
-This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
+This inductive scheme also generalises to \emph{k-induction}, which allows the inductive case to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
 K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.
@@
 We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
-The transition system itself is deterministic, but the free context provides the non-determinism; our theorem  statement existentially quantifies the free heap.
+The transition system itself is deterministic, but the free context provides the non-determinism which may occur from underspecified contracts; our theorem statement existentially quantifies the free heap.
 
@@
 Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
-The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
+The invariant for followed-by expressions asserts that the initial state of the followed-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
 With this invariant, we can prove abstraction:
@@
 
-Finally, we can show the main entailment result: that, if the proof obligations hold on the system, then the original program's proof obligations also hold.
+Finally, we can show the main entailment result that if the proof obligations hold on the system, then the original program is valid according to the checked semantics:
 
@@
 The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
-Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them.
+Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them, allowing for modular proofs.

### 2024-03-30 2024pipit/section/50-Extraction.tex
good

@@
 
-Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
+Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
 The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
-The last two bindings recompute the state, this time with the correct recursive value.
+The last two bindings recompute the state, this time with the correct recursive value used in the state.
 This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.
@@
 
-The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions; this interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular}:
+The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions:
   \begin{tabbing}
@@
 The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.
+This interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular} and other synchronous languages.
 

### 2024-03-30 2024pipit/section/60-Evaluation.tex
good

@@
 
-We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
+We were able to express equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
 For example, the specification-only function that finds the next trigger is naturally recursive.
@@
 Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
-Below we compare a hypothetical unsupported recursive definition of @next@ (left) with the array-scan encoding (right):
+\autoref{f:evaluation:recursion} illustrates this encoding with a simplified version of the next-trigger logic.
 
-\begin{minipage}{0.4\textwidth}
-\begin{verbatim}
-next(i) =
-  if enabled(triggers,i)
-  then i
-  else if i >= LENGTH - 1
-  then NO_NEXT_TRIGGER
-  else next(i + 1);
-
-\end{verbatim}
+\begin{figure}
+\begin{minipage}{0.45\textwidth}
+\begin{tabbing}
+  MM \= MM \= \kill
+@let rec@ next ($i$: int) ($c$: cycle): \\
+  \> Tot \> (option int) \\
+  \>     \> (decreases (count - $i$)) = \\
+\>  @if@ trigger_enabled $i$~$c$ \\
+\>  @then@ Some $i$ \\
+\>  @else if@ $i \ge \text{count} - 1$ \\
+\>  @then@ None \\
+\>  @else@ next $(i + 1)$~$c$
+\end{tabbing}
 \end{minipage}
@@
 \begin{verbatim}
+  function next(index:  int; c: cycle)
+       returns (result: int)
+  var next_array: int ^ COUNT;
+  let
     next_array[i] =
-  if enabled(triggers,LENGTH - 1 - i)
-  then LENGTH - 1 - i
+      if trigger_enabled(COUNT - 1 - i, c)
+      then COUNT - 1 - i
       else if i <= 0
@@
       else next_array[i - 1];
-next = next_array[LENGTH - 1 - index];
-
+    result =
+      next_array[COUNT - 1 - index];
+  tel
 \end{verbatim}
 \end{minipage}
+\caption{Left: next-trigger logic in \fstar{}; right: Kind2 encoding as array scan.
+In \fstar{}, the \emph{Tot $\tau$ (decreases \ldots)} syntax declares a total function with the given termination measure.
+In Kind2, the \tt{int\^{}COUNT} syntax declares an array of integers of length @COUNT@, while the @next_array[i]@ declaration defines the elements of the array as a function of the index @i@.
+}
+\label{f:evaluation:recursion}
+\end{figure}
 
@@
 We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
-Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
+Both Kind2 and Pipit developments of the trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
 Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
@@
 We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
-For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
+For the simplified trigger-fetch mechanism, we were able to verify up to 64 triggers.
+For reference, hardware implementations of TTCAN such as M_TTCAN support up to 64 triggers~\cite{bosch2019mttcan}.
 

### 2024-03-30 2024pipit/section/70-Related.tex
good

@@
 Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
-Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
+Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays, records, or a foreign-function interface.
 Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.
@@
 As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
-Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
+Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration, which is unlikely to result in acceptable performance.
 
@@
 In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
-If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not expand the trusted computing base.
+If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not necessarily expand the trusted computing base --- that is, we could augment our end-to-end verified workflow with \emph{unverified but validated} invariant generation.
 

### 2024-03-30 2024pipit/section/20-Motivation.tex
doublegood

@@
 The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
-The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering a failure if the reference message is not received.
+The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering an error if the reference message is not received.
 \autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.
@@
 
-Our driver may fall behind when trying to run certain schedules, as the driver only processes one trigger per time-mark.
+Our driver may fall behind when trying to execute certain schedules, as the driver only processes one trigger per time-mark.
 To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.
@@
 We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
-However, we could \emph{not} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
+However, we \emph{cannot} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
 
-We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
+We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index, so that it is reachable in time from the start of the array.
 
@@
 The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
-In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous tick of computation.
-Expressing such a rely in an imperative setting would require extra encoding, as imperative specification languages do not generally have an innate notion of discrete time steps.
+In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous \emph{tick} of computation.
+Expressing such a rely in an imperative setting would require extra encoding, as preconditions in imperative languages do not generally have an innate notion of the previous value with respect to a global shared clock.
 

### 2024-03-30 2024pipit/section/30-Core.tex
doublegood

@@
 The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.
+% The unknown property status here allows us to defer the caller's proof that the input stream is always positive until we have a concrete definition for the \emph{ints} input stream.
 
@@
 $$
-\bigstep{\sumionetwo}{\sumrec}{v}
+\bigstep{\sumintsonetwo}{\sumrec}{v}
 $$
@@
 Evaluation of primitives is standard.
-To evaluate variables, we look for the variable in the current heap:
+To evaluate variables, we look for the variable in the current (rightmost) heap:
 $$
@@
 }{
-  \bigstep{\sumionetwo}{\textit{ints}}{2}
+  \bigstep{\sumintsonetwo}{\textit{ints}}{2}
 }{Var}
@@
 }{
-  \bigstep{\sumionetwo}{\xfby{0}{\sumrec}}{1}
+  \bigstep{\sumintsonetwo}{\xfby{0}{\sumrec}}{1}
 }{$\mbox{Fby}_S$}
@@
 Checking contracts is more involved.
-Conceptually, contracts involve two kinds of checks: one for the definition and one for the call-site.
-To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
+For whole-program correctness, it would suffice to check that a contract's rely and guarantee both hold.
+However, the purpose of contracts is to enable modular reasoning about parts of the program: we need to be able to check contracts independently of their context.
+Conceptually, then, contracts involve two kinds of checks: one for the definition and one for the call-site.
+To check a contract definition, we check that the body satisfies the guarantee for all \emph{valid} contexts -- that is, those where the rely holds.
+Then, to check a contract instance, we just need to check that the call-site satisfies the rely.
+% To check that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid for some streaming history $\Sigma$, we want to check that the body satisfies the guarantee
+% check if, assuming the rely holds for the history ($\bigstepalways{\Sigma}{\erely}$), and assuming the body evaluates to some stream $V$ ($\bigsteps{\Sigma}{\ebody}{V}$), then the guarantee also holds ($\bigstepalways{\Sigma[x \mapsto V]}{\eguar}$).
 % The fact that the checked semantics refers to a particular $\Sigma$ is significant: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
@@
 It is critical that the rely is true \emph{at all points} in the stream.
-Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; here, the rely is false at the first step, but is instantaneously true at the second step.
+Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; the rely is false at the first step, but is instantaneously true at the second step.
 In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
 At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
-The contract itself remains valid, as the input did not satisfy the rely at all steps.
-This behaviour is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.
+The contract itself remains valid, however, as the assumption is invalid: the input did not satisfy the rely at all steps.
+% The \emph{at all steps} is a key difference between streaming contracts and imperative pre-post contracts, which consider their preconditions in isolation.
 
-Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.
+% Checking a contract instance is simple, in comparison: we just check that the call-site satisfies the rely at all steps.
 

### 2024-03-30 2024pipit/section/50-Extraction.tex
doublegood

@@
 This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
-To get the correct state, we call the step function again, this time with $e$ bound to $v$.
+To get the correct state, we call the step function again, this time with $x$ bound to the correct value, $v$.
 
@@
 The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
-The last two bindings recompute the state, this time with the correct recursive value used in the state.
+The last two bindings recompute the state, this time with the correct recursive value $\textit{sum}_0$ used in the state.
 This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

### 2024-04-02 2024pipit/section/60-Evaluation.tex
eval

@@
 In \fstar{}, the \emph{Tot $\tau$ (decreases \ldots)} syntax declares a total function with the given termination measure.
-In Kind2, the \tt{int\^{}COUNT} syntax declares an array of integers of length @COUNT@, while the @next_array[i]@ declaration defines the elements of the array as a function of the index @i@.
+In Kind2, the \tt{int\^{}COUNT} syntax denotes the type of an array of integers of length @COUNT@, while the @next_array[i]@ declaration defines the elements of the array as a function of the index @i@.
 }

### 2024-04-03 2024pipit/expand-20240403.tex
diff

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate,anonymous]{lipics-v2021}

\hideLIPIcs

\bibliographystyle{plainurl}

\ifx\authoranonymous\relax
\newcommand{\GITHUBURL}{(anonymised)}
\newcommand{\COFFEEURL}{(anonymised)}
\else
\newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
\newcommand{\COFFEEURL}{\url{https://youtu.be/6IybbQFPOl8}}
\fi

\usepackage{mathtools}
\usepackage{mathpartir}

\usepackage{stmaryrd}

\usepackage{wasysym}
\usepackage{underscore}

\usepackage{style/utils}
\usepackage{style/code}
\usepackage{style/proof}
\usepackage{style/keywords}
\usepackage{style/judgements}

\title{Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems}

\author{Amos Robinson}{Australian National University, Canberra, Australia}{amos.robinson@anu.edu.au}{https://orcid.org/0009-0004-4837-4981}{}

\author{Alex Potanin}{Australian National University, Canberra, Australia}{alex.potanin@anu.edu.au}{https://orcid.org/0000-0002-4242-2725}{}

\authorrunning{A. Robinson and A. Potanin}
\Copyright{Amos Robinson and Alex Potanin}
\ccsdesc[500]{Computer systems organization~Real-time languages}
\ccsdesc[500]{Theory of computation~Program verification}
\ccsdesc[500]{Software and its engineering~Specialized application languages}
\keywords{Lustre, streaming, reactive, verification}

\begin{document}

\maketitle

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate executable code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  The executable code is deterministic and total and preserves the semantics of the original program.
\end{abstract}

% \makeatactive

\section{Introduction}
Safety-critical control systems, such as the anti-lock braking systems that are present in most cars today, need to be correct and execute in real-time.
One approach, favoured by parts of the aerospace industry, is to implement the controllers in a high-level language such as Lustre~\cite{caspi1995functional} or Scade~\cite{colaco2017scade}, and verify that the implementations satisfy the high-level specification using a model-checker, such as Kind2~\cite{champion2016kind2}.
These model-checkers can prove many interesting safety properties automatically, but do not provide many options for manual proofs when the automated proof techniques fail.
Additionally, the semantics used by the model-checker may not match the semantics of the compiled code, in which case properties proved do not necessarily hold on the real system.
This mismatch may occur even when the compiler has been verified to be correct, as in the case of Vélus~\cite{bourke2017formally}.
For example, in Vélus, integer division rounds towards zero, matching the semantics of C; however, integer division in Kind2 rounds to negative infinity, matching SMT-lib~\cite{BarFT2016SMTLIB,kind2023intdiv}.

To be confident that our proofs hold on the real system, we need a single shared semantics for the compiler and the prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

In this paper, we make the following contributions:

\begin{itemize}
  \item we motivate the need to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit; after checking deferred properties, programs are \emph{blessed}, which marks their properties as valid (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:core:causality}, \autoref{s:transition:proof});
  \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
\end{itemize}

\section{Pipit for time-triggered networks}
\label{s:motivation}

To introduce Pipit, we consider a \emph{time-triggered} network driver, which has a static schedule dictating the network traffic, and which all nodes on the network must adhere to.
This driver is a simplification of the Time-Triggered Controller Area Network (TTCAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.

At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
The node matrix is represented in memory by a \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.

\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering an error if the reference message is not received.
\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.

\begin{figure}
  \begin{minipage}{0.5\textwidth}
\resizebox{\textwidth}{!}{
\begin{tabular}{r|ccccc}
   & TM0 & TM1 & TM2 & $\cdots$ & TM9 \\
  \hline
  C0 & SEND A & SEND B & - & $\cdots$ & WATCH \\
  C1 & SEND A & - & SEND C & $\cdots$ & WATCH
\end{tabular}}
\end{minipage}
\begin{minipage}{0.49\textwidth}
\scriptsize
\begin{verbatim}
 0:{ time = 0; enabled = {C0,C1}; action = SEND(A); }
 1:{ time = 1; enabled = {C0};    action = SEND(B); }
 2:{ time = 2; enabled = {C1};    action = SEND(C); }
 3:{ time = 9; enabled = {C0,C1}; action = WATCH;   }
\end{verbatim}
\end{minipage}

\caption{Left: node matrix; right: corresponding triggers array configuration}
\label{f:tt-systemmatrix-ok}
\end{figure}

The network has strict timing requirements which prohibit the driver from looping through the entire triggers array at each time-mark.
Instead, the driver maintains an index that refers to the current trigger.
At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.

\subsection{Deferring and proving properties}

We implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
At each step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the counter as-is.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \kill
  \tt{let} count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\NN$ = \\
    \> $\tt{rec}~\rawbind{\textit{count}}{}$ \\
    \> \> $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$; \\
    \> \> \tt{let} $\textit{count'}$ \tt{=} $(0~\tt{fby}~\textit{count}) + (\tt{if}~\textit{inc}~\tt{then}~1~\tt{else}~0)$ \tt{in} \\
    \> \> \tt{if } $\textit{count'} \ge \textit{max}$ \tt{then} $\textit{max}$  \tt{else} $\textit{count'}$
\end{tabbing}

The implementation of \emph{count_when} first defines a recursive stream, \emph{count}, which states an invariant about the count before defining the incremented stream \emph{count'}.
Inside \emph{count'}, the syntax $0~\tt{fby}~\textit{count}$ is read as ``the initial value of zero \emph{followed by} the previous count''.

The syntax $\xcheckP{\PSUnknown}{(0 \le \textit{count} \le \textit{max})}$ asserts that the count is within the range $[0, \textit{max}]$.
The subscript $\PSUnknown$ on the check is the \emph{property status}, which in this case denotes that the assertion has been stated, but it is not yet known whether it holds.
A property status of $\PSValid$, on the other hand, denotes that a property has been proved to hold.
These property statuses are used to defer checking properties until enough is known about the environment, and to avoid rechecking properties that have already been proven.
In practice, the user does not explicitly specify property statuses in the source language.
The stated property $(0 \le \textit{count} \le \textit{max})$ is a stream of booleans which must always be true.
Non-streaming operations such as $\le$ are implicitly lifted to streaming operations, and non-streaming values such as $0$ and $\textit{max}$ are implicitly lifted to constant streams.

We defer the proof of the property here because, at the point of stating the property inside the \tt{rec} combinator, we don't yet have a concrete definition for the count variable.
In this case, we could have instead deferred the \emph{statement} of the property by introducing a let-binding for the recursive count and putting the \tt{check} outside of the \tt{rec} combinator.
However, it is not always possible to defer property statements: for example, when calling other streaming functions that have their own preconditions, it may not be possible to move the function call outside of its enclosing \tt{rec}.

Pipit is an embedded domain-specific language.
The program above is really syntactic sugar for an \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable.
We will discuss the details of the core language in \autoref{s:core}, but for now we focus on the source program with some minor embedding details omitted.

To actually prove the property above, we use the meta-language \fstar{}'s tactics to translate the program into a transition system and prove the property inductively on the system.
Finally, we \emph{bless} the expression, which marks the properties as valid ($[\PSUnknown := \PSValid]$).
Blessing is an intensional operation that traverses the expression and updates the internal metadata, but does not affect the runtime semantics.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \kill
  \tt{let} $\mbox{count_when}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\BB$ $\to$ stream $\NN$ = \\
    \> \tt{let} $\textit{system} = \mbox{System.translate}_1 (\mbox{count_when}~\textit{max})$ \tt{in} \\
    \> \tt{assert} (System.inductive_check $\textit{system}$) \tt{by} (pipit_simplify ()); \\
    \> $\tt{bless}_1$ $(\mbox{count_when}~\textit{max})$
\end{tabbing}

The subscript 1 in the translation to transition system and blessing operations refers to the fact that the stream function has one stream parameter.
The \emph{pipit_simplify} tactic in the assertion performs normalisation-by-evaluation to simplify away the translation to a first-order transition system; \fstar{}'s proof-by-SMT can then solve the inductive check directly.

Callers of \emph{count_when} can now use the validated variant without needing to re-prove the count-range property.
In a dedicated model-checker such as Kind2 \cite{champion2016kind2} or Lesar \cite{raymond2008synchronous}, this kind of bookkeeping would all be performed under-the-hood.
By embedding Pipit in a general-purpose theorem prover, we move some of the bookkeeping burden onto the user; however, we have increased confidence that the compiled code matches the verified code and, as we shall see, we also have access to a rich specification language.

\subsection{Restrictions on the triggers array}

Our driver may fall behind when trying to execute certain schedules, as the driver only processes one trigger per time-mark.
To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.

Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
However, we \emph{cannot} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.

We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index, so that it is reachable in time from the start of the array.

With these restrictions in place, we prove a lemma \emph{lemma_can_reach_next}, which states that for all valid cycle indices and trigger indices, if the current trigger is enabled in the current cycle and there is another enabled trigger scheduled to occur somewhere in the array after the current one, then there is an adequate time-gap to allow the driver to skip over any disabled triggers in-between.
These properties are straightforward in a theorem prover, but are difficult to state in a model-checker with a limited specification language.

\subsection{Instantiating lemmas and defining contracts}
\label{s:motivation:contract}

We can now implement the trigger-fetch logic, which keeps track of the current trigger.
We use the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
We simplify our presentation here and only consider a constant cycle: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{let index} \= \kill
  \tt{let} trigger_fetch ($\textit{cycle}$: $\NN$) ($\textit{time}$: stream $\NN$): stream $\NN$ = \\
    \> $\tt{rec } \textit{index}.$ \\
    \> \> $\tt{let } \textit{inc} = \text{false} \tt{ fby } ((\text{time_mark}~\textit{index}) \le \textit{time} ~\vee~ \neg (\text{enabled}~\textit{index}~\textit{cycle})) \tt{ in}$\\
    \> \> $\tt{let } \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} \tt{ in}$ \\
    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index})$; \\
    \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
    \> \> $\textit{index}$
\end{tabbing}

The \emph{trigger_fetch} function takes a static cycle index and a stream denoting the current time.
The increment flag and the index are mutually dependent --- the increment flag depends on the previous value of the index, while the index depends on the current value of the increment flag --- so we introduce a recursive stream for the index.
We allow the index to go one past the end of the array to denote that there are no more triggers.

We use the $\textit{pose}$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it.
We then state an invariant as a deferred property.
Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if we are called once per time-mark then we guarantee that we never encounter a late trigger.

\begin{tabbing}
  \tt{MM}\= \tt{MM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{trigger_fetch}_{\PSValid}$ ($\textit{cycle}$: $\NN$): $\stream \NN \to \stream \NN$ = \\
  \> \tt{let} $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
  \> \> \tt{rely} = $(\lambda \textit{time}.~ \textit{time} = \xfby{0}{(\textit{time} + 1)});$ \\
  \> \> \tt{guar} = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
  \> \> \> $\implies (\text{time_mark}~\textit{index}) \ge \textit{time});$ \\
  \> \> \tt{body} = $(\lambda \textit{time}.~ \text{trigger_fetch}~\textit{cycle}~\textit{time} );$ \\
  \> \} \tt{in} \\
  \> \tt{assert} $(\text{Contract.inductive_check}~\textit{contract})$ \tt{by} (pipit_simplify ()); \\
  \> $\text{Contract.stream_of_contract}_1~\textit{contract}$
\end{tabbing}

In the implementation of the validated variant of \emph{trigger_fetch}, we first construct the contract from streaming functions.
The $\text{Contract.contract_of_stream}_1$ combinator describes a contract with one input (the time stream), and takes stream transformers for each of the rely, guarantee and body.
The combinator transforms the surface syntax into core expressions.
The assertion $(\text{Contract.inductive_check}~\textit{contract})$ then translates the expressions into a transition system, and checks that if the rely always holds then the guarantee always holds, and that the as-yet-unchecked subproperties hold.
Finally, $\text{Contract.stream_of_contract}_1$ blesses the core expression and converts it back to a stream transformer, so it can be easily used by other parts of the program.

The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous \emph{tick} of computation.
Expressing such a rely in an imperative setting would require extra encoding, as preconditions in imperative languages do not generally have an innate notion of the previous value with respect to a global shared clock.

When \emph{trigger_fetch} is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.

\section{Pipit core language}
\label{s:core}

\begin{figure}
  \includegraphics[width=\textwidth]{figures/core-structure-1920x450.pdf}
\caption{Architecture of Pipit. The gray boxes and solid arrows are defined in this paper. The white boxes and dashed arrows are trusted components. The labels denote verified properties of the translation: abstraction~($\subseteq$), entailment of proof obligations~($\vdash$), and equivalence~($=$).
}
\label{f:core:structure}
\end{figure}

We now introduce the core Pipit language.
Note that this form differs slightly from the surface syntax presented earlier in \autoref{s:motivation}, which used the syntax of the metalanguage \fstar{}, as well as including proofs in \fstar{} itself.

\autoref{f:core:structure} shows the high-level architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

\autoref{f:core-grammar} defines the grammar of Pipit.
The expression form $e$ includes standard syntax for values ($v$), variables ($x$) and primitive applications ($p(\ov{e})$).
Most of the expression forms were introduced informally in \autoref{s:motivation} and correspond to the clock-free expressions of Lustre~\cite{caspi1995functional}.

\begin{figure}
  \[
  \begin{array}{lrlr}
    e, e' & := & v ~|~ x ~|~ p(\ov{e}) & \mbox{(values, variables and operations)} \\
          & | & \xfby{v}{e} ~|~ \xrec{x}{e[x]} & \mbox{(delayed and recursive streams)} \\ & | & \xlet{x}{e}{e'[x]} & \mbox{(let-expressions)}\\
          & | & \xcheckP{\PStatus}{e_{\text{prop}}} & \mbox{(checked properties)} \\
          & | & \xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}} & \mbox{(rely-guarantee contracts)} \\
\\
    v & := & n \in \mathbb{N} ~|~ b \in \mathbb{B} ~|~ r \in \mathbb{R} ~|~ \hdots  & \mbox{(values)} \\
    p & := & (+) ~|~ (-) ~|~ (\times) ~|~ \tt{if-then-else} ~|~ \hdots & \mbox{(primitives)} \\
    \\
    \PStatus & := & \PSValid ~|~ \PSUnknown & \mbox{(property statuses: valid or unknown)}\\
    \\
    V & := & \cdot ~|~ V;v & \mbox{(streams of values)} \\
    \sigma & := & \sgl{\ov{x \mapsto v}} & \mbox{(heaps)} \\
    \Sigma & := & \cdot ~|~ \Sigma;\sigma & \mbox{(streaming history environments)} \\
\tau, \tau' & := & \mathbb{N} ~|~ \mathbb{B} ~|~ \tau \times \tau ~|~ \hdots & \mbox{(value types)} \\
    \Gamma & := & \cdot ~|~ x : \tau, \Gamma & \mbox{(type environments)}  \\
    \end{array}
  \]
  \caption{Core grammar: expressions $e$, values $v$, primitive operations $p$, and property statuses $\PStatus$.}
  \label{f:core-grammar}
\end{figure} 
The expression syntax for delayed streams ($\xfby{v}{e}$) denotes the previous value of the stream $e$, with an initial value of $v$ when there is no previous value.

Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax by combining the bindings together into a record or tuple.

Checked properties and contracts are annotated with their property status $\PStatus$, which can either be valid ($\PSValid$) or unknown ($\PSUnknown$).
For checked properies $\xcheckP{\PStatus}{e}$, the property status denotes whether the property has been proved to be valid.

Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
Contracts involve two verification conditions.
Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
Secondly, when a contract is \emph{instantiated}, the caller must prove that the environment satisfies the precondition: that is, $\erely$ is always true.
Conceptually, then, a contract could have two property statuses: one for the definition and one for the instantiation.
However, in practice, it is not useful to defer the proof of a contract definition --- one could achieve a similar effect by replacing the contract with its implementation.
For this reason, we only annotate contracts with one property status, which denotes whether the instantiation has been proved to satisfy the precondition.

For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
$$
\sumcontract
$$
To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.

The remaining grammatical constructs of \autoref{f:core-grammar} describe streams, value environments, types and type environments.
Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
Types $\tau$ and type environments $\Gamma$ are standard.
For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.

\begin{figure}
  \begin{mathpar}
    \boxed{\typing{\Gamma}{e}{\tau}}
\end{mathpar}

  \begin{mathpar}
    \ruleIN{
      \mtypingval{v}{\tau}
    }{
      \typing{\Gamma}{v}{\tau}
    }{TValue}

    \ruleAx{\typing{\Gamma, x: \tau, \Gamma'}{x}{\tau}}{TVar}

    \ruleIN{
      \mtypingprim{p}{(\tau_1 \times \dots \times \tau_n) \to \tau'}
      \qquad
      \typing{\Gamma}{e_1}{\tau_1}
      \quad \hdots \quad
      \typing{\Gamma}{e_n}{\tau_n}
    }{\typing{\Gamma}{p(\ov{e})}{\tau'}}{TPrim}

    \ruleIN{
      \mtypingval{v}{\tau}
      \qquad
      \typing{\Gamma}{e'}{\tau}
    }{
      \typing{\Gamma}{\xfby{v}{e'}}{\tau}
    }{TFby}

\ruleIN{
      \typing{\Gamma, x : \tau}{e}{\tau}
    }{
      \typing{\Gamma}{\xrec{x}{e[x]}}{\tau}
    }{TRec}

    \ruleIN{
      \typing{\Gamma}{e}{\tau}
      \qquad
      \typing{\Gamma, x : \tau}{e'}{\tau'}
    }{
      \typing{\Gamma}{\xlet{x}{e}{e'[x]}}{\tau'}
    }{TLet}

    \ruleIN{
      \typing{\Gamma}{e}{\mathbb{B}}
    }{
      \typing{\Gamma}{\xcheckP{\PStatus}{e}}{\tt{unit}}
    }{TCheck}
  \and
    \ruleIN{
      \typing{\Gamma}{\erely}{\mathbb{B}}
      \qquad
      \typing{\Gamma}{\ebody}{\tau}
      \qquad
      \typing{\Gamma, x: \tau}{\eguar}{\mathbb{B}}
    }{
      \typing{\Gamma}{\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}}}{\tau}
    }{TContract}
  \end{mathpar}

  \caption{Typing rules for Pipit; the judgment $\typing{\Gamma}{e}{\tau}$ denotes that expression $e$ describes a \emph{stream} of values of type $\tau$. Auxiliary functions are used for values and primitive operations.}\label{f:core-typing}
\end{figure} 
We define the typing judgments for Pipit in \autoref{f:core-typing}.
Most of the typing rules are standard for an unclocked Lustre.
The typing judgment $\typing{\Gamma}{e}{\tau}$ denotes that, in an environment of streams $\Gamma$, expression $e$ denotes a stream of type $\tau$.
This core typing judgment differs from the surface syntax used in \autoref{s:motivation}, which used an explicit stream type; for the core language, we instead assume that everything is a stream.

We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
Primitives are pure, non-streaming functions.

Rules \textsc{TValue}, \textsc{TVar}, \textsc{TPrim} and \textsc{TLet} are standard.

Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.

Rule \textsc{TRec} states that a recursive stream $\xrec{x}{e}$ has the recursive stream bound inside $e$.
The recursion must also be guarded, in that any recursive references to $x$ are delayed, but this requirement is performed as a separate syntactic check described in \autoref{s:core:causality}.

Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.

Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
The overall expression has result type $\tau$.
Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.

\subsection{Dynamic semantics}
\label{s:core:dynamic}

\begin{figure}[t]
  \begin{mathpar}
    \boxed{\bigstep{\Sigma}{e}{v}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\bigstep{\Sigma; \sigma}{x}{\sigma(x)}}{Var}
    \quad
    \ruleAx{\bigstep{\Sigma}{v}{v}}{Value}

    \ruleIN{
      \bigstep{\Sigma}{e'[x := e]}{v}
    }{
      \bigstep{\Sigma}{\xlet{x}{e}{e'[x]}}{v}
    }{Let}

    \ruleIN{
      \bigstep{\Sigma}{e_1}{v_1} \quad \hdots \quad
      \bigstep{\Sigma}{e_n}{v_n}
    }{\bigstep{\Sigma}{p(\ov{e})}{\text{prim-sem}(p, \ov{v})}}{Prim}

  \ruleAx{\bigstep{\sigma}{\xfby{v}{e'}}{v}}{$\mbox{Fby}_1$}
    \quad
    \ruleIN{\text{length}(\Sigma) > 0 \and \bigstep{\Sigma}{e'}{v'}}{\bigstep{\Sigma; \sigma}{\xfby{v}{e'}}{v'}}{$\mbox{Fby}_S$}

    \ruleIN{
      \bigstep{\Sigma}{e[x := \xrec{x}{e}]}{v}
    }{
      \bigstep{\Sigma}{\xrec{x}{e[x]}}{v}
    }{Rec}

    \ruleIN{
}{
      \bigstep{\Sigma}{\xcheckP{\PStatus}{e}}{()}
    }{Check}

    \ruleIN{
      \bigstep{\Sigma}{\ebody}{v}
    }{
      \bigstep{\Sigma}{\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}}}{v}
    }{Contract}
  \end{mathpar}

  \begin{mathpar}
    \boxed{\bigsteps{\Sigma}{e}{V}}

    \boxed{\bigstepalways{\Sigma}{e}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\bigsteps{\cdot}{e}{\cdot}}{$\mbox{Steps}_0$}

    \ruleIN{
      \bigstep{\Sigma}{e}{V}
      \and
      \bigstep{\Sigma; \sigma}{e}{v}
    }{\bigstep{\Sigma; \sigma}{e}{V; v}}{$\mbox{Steps}_S$}

    \ruleIN{\bigsteps{\Sigma}{e}{\true; \hdots}}{\bigstepalways{\Sigma}{e}}{Always}
  \end{mathpar}

  \caption{Dynamic semantics for Pipit; the judgment form $\bigstep{\Sigma}{e}{v}$ denotes that evaluating expression $e$ under streaming history $\Sigma$ results in value $v$.}\label{f:core-bigstep}
\end{figure}

The dynamic semantics of Pipit are defined in \autoref{f:core-bigstep}.
We present our semantics in a big-step form.
This differs somewhat from traditional \emph{reactive} semantics of Lustre~\cite{caspi1995functional}.
Our big-step semantics emphasises the equational nature of Pipit, as it is substitution-based and syntax-directed, while the reactive semantics emphasises the finite-state streaming execution of the system.
We use transition systems for reasoning about the finite-state execution (\autoref{s:transition}), which is fairly standard~\cite{brun2023equation,champion2016kind2,raymond2008synchronous}.
Previous work on the {\sc W-calculus}~\cite{gallego2021w} for linear digital-signal-processing filters makes a similar distinction and provides a non-streaming semantics for reasoning about programs and a streaming semantics for executing programs.

The judgment form $\bigstep{\Sigma}{e}{v}$ denotes that expression $e$ evaluates to value $v$ under streaming history $\Sigma$.
The streaming history is a stream of heaps; in practice, we only evaluate expressions with a non-empty streaming history.

At a high level, evaluation unfolds recursive streams to determine a value.
For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
$$
\bigstep{\sumintsonetwo}{\sumrec}{v}
$$

First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
Evaluation of primitives is standard.
To evaluate variables, we look for the variable in the current (rightmost) heap:
$$
\ruleIN{
}{
  \bigstep{\sumintsonetwo}{\textit{ints}}{2}
}{Var}
$$

For delays, we discard the current heap and continue evaluation with the history prefix:
$$
\ruleIN{
  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
}{
  \bigstep{\sumintsonetwo}{\xfby{0}{\sumrec}}{1}
}{$\mbox{Fby}_S$}
$$

Returning to \autoref{f:core-bigstep},
rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
Rules {\sc Value} and {\sc Let} are standard.
Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.

For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.

Rule {\sc Rec} evaluates a recursive stream $\xrec{x}{e}$ by unfolding the recursion one step.
For causal expressions (\autoref{s:core:causality}), where each recursive occurrence of $x$ is guarded by a followed-by, this unfolding eventually terminates as each followed-by shortens the history.

Rule {\sc Check} ignores the property when evaluating check expressions.
We do not dynamically check the property here; this is done in the checked semantics (\autoref{s:core:checked}).

Similarly, rule {\sc Contract} ignores preconditions and postconditions when evaluating contracts.
From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
The deterministic and total nature of evaluation is key to our proofs and metatheory.

We also define two auxiliary judgment forms: $\bigsteps{\Sigma}{e}{V}$ and $\bigstepalways{\Sigma}{e}$.

Judgment form $\bigsteps{\Sigma}{e}{V}$ denotes that, under history $\Sigma$, expression $e$ evaluates to the \emph{stream} $V$.
This judgment performs iterated application of single-value evaluation.

Judgment form $\bigstepalways{\Sigma}{e}$ denotes that a boolean expression $e$ evaluates to the stream of trues under history $\Sigma$.
Informally, it can be read as ``$e$ is always true in history $\Sigma$''.

\subsection{Checked semantics}
\label{s:core:checked}

In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
We call these the \emph{checked} semantics; they are comparable to checking runtime assertions.

The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties and contracts of $e$ with status $\PStatus$ hold.
The property status dictates which properties should be checked and which should be ignored.

We consider a program to be \emph{valid} if its checks hold for all histories ($\forall \Sigma.~ \semcheck{\Sigma}{\PSValid}{e}$).
The checked semantics are a specification describing what it means to be a valid program.
We do not generally verify programs directly using the checked semantics; instead, we translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).

To check a property $(\xcheckP{\PStatus}{e})$ in history $\Sigma$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.

Checking contracts is more involved.
For whole-program correctness, it would suffice to check that a contract's rely and guarantee both hold.
However, the purpose of contracts is to enable modular reasoning about parts of the program: we need to be able to check contracts independently of their context.
Conceptually, then, contracts involve two kinds of checks: one for the definition and one for the call-site.
To check a contract definition, we check that the body satisfies the guarantee for all \emph{valid} contexts -- that is, those where the rely holds.
Then, to check a contract instance, we just need to check that the call-site satisfies the rely.

For example, recall our earlier contract that the sum of strictly positive integers is positive:
$$
\tt{let}~\text{sum}~\textit{i} = \sumcontractX{\PSUnknown}{i}
$$
To check the contract definition on a concrete input $\textit{i} = [1; 2]$, we first evaluate the body:
$$
\bigsteps{\sumionetwo}{\sumrecX{i}}{[1; 3]}
$$
We then check that, assuming all inputs are positive, then all results are positive:
$$
\bigstepalways{\sumionetwo}{\textit{i} > 0} \implies 
\bigstepalways{\sgl{\textit{i} \mapsto 1, \sumvar \mapsto 1}; \sgl{\textit{i} \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
$$

It is critical that the rely is true \emph{at all points} in the stream.
Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; the rely is false at the first step, but is instantaneously true at the second step.
In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
The contract itself remains valid, however, as the assumption is invalid: the input did not satisfy the rely at all steps.

\begin{figure}[t]
  \begin{mathpar}
    \boxed{\semcheck{\Sigma}{\PStatus}{e}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\semcheck{\Sigma}{\PStatus}{v}}{ChkValue}
    \quad
    \ruleAx{\semcheck{\Sigma}{\PStatus}{x}}{ChkVar}

    \ruleIN{
      \semcheck{\Sigma}{\PStatus}{e_1} \quad \hdots \quad
      \semcheck{\Sigma}{\PStatus}{e_n}
    }{\semcheck{\Sigma}{\PStatus}{p(\ov{e})}}{ChkPrim}

    \ruleIN{\semcheck{\Sigma}{\PStatus}{e'}}{\semcheck{\Sigma}{\PStatus}{\xfby{v}{e'}}}{ChkFby}

    \ruleIN{
      \bigsteps{\Sigma}{\xrec{x}{e}}{V}
      \and
      \semcheck{\Sigma[x \mapsto V]}{\PStatus}{e}
    }{
      \semcheck{\Sigma}{\PStatus}{\xrec{x}{e[x]}}
    }{ChkRec}

    \ruleIN{
      \semcheck{\Sigma}{\PStatus}{e}
      \and
      \bigsteps{\Sigma}{e}{V}
      \and
      \semcheck{\Sigma[x \mapsto V]}{\PStatus}{e'}
    }{
      \semcheck{\Sigma}{\PStatus}{\xlet{x}{e}{e'[x]}}
    }{ChkLet}

    \ruleIN{
      (\PStatus = \PStatus' \implies \bigstepalways{\Sigma}{e})
      \and
      \semcheck{\Sigma}{\PStatus}{e}
    }{
      \semcheck{\Sigma}{\PStatus}{\xcheckP{\PStatus'}{e}}
    }{ChkCheck}

    \inferrule{
      \bigsteps{\Sigma}{\ebody}{V}
      \\\\
      (\PStatus = \PStatus' \implies \bigstepalways{\Sigma}{\erely})
      \\\\
      (\PStatus = \PSValid \implies \bigstepalways{\Sigma}{\erely} \implies \bigstepalways{\Sigma[x \mapsto V]}{\eguar})
      \\\\
      \semcheck{\Sigma}{\PStatus}{\erely}
      \\\\
      (\bigstepalways{\Sigma}{\erely} \implies \semcheck{\Sigma}{\PStatus}{\ebody} ~\wedge~ \semcheck{\Sigma[x \mapsto V]}{\PStatus}{\eguar})
    }{
      \semcheck{\Sigma}{\PStatus}{\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}}}
    }(\textsc{ChkContract})

\end{mathpar}

  \caption{Checked semantics for Pipit; the judgment form $\semcheck{\Sigma}{\PStatus}{e}$ denotes that evaluating expression $e$ under streaming history $\Sigma$ satisfies the checks and rely-guarantee contract requirements that are labelled with property status $\PStatus$.}\label{f:core-check}
\end{figure}

The checked semantics of Pipit is defined in \autoref{f:core-check}.

Rules {\sc ChkValue} and {\sc ChkVar} state that values and variables are always valid.

Rule {\sc ChkPrim} checks a primitive application by descending into the subexpressions.
Similarly, rule {\sc ChkFby} descends into followed-by expressions.

Rule {\sc ChkRec} checks a recursive-expression $\xrec{x}{e}$ by evaluating the overall expression to a stream of values $V$.
The rule then extends the streaming environment $\Sigma$ with $x$ bound to the values from $V$; this extended environment is used to descend into the recursive expression.

Rule {\sc ChkLet} checks a let-expression $\xlet{x}{e}{e'}$ descends into both sub-expressions.
To check the body $e'$, the rule first evaluates $e$ and extends the streaming environment.

Finally, the heavy lifting is performed by rules {\sc ChkCheck} and {\sc ChkContract}.

Rule {\sc ChkCheck} checks the properties marked $\PStatus$ in an expression $\xcheckP{\PStatus'}{e}$.
If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we evaluate the expression $e$ and require it to be true at all steps.
We then unconditionally descend into the subexpression to check any nested properties.
Such nested properties are unlikely to be written directly by the user, but might occur after inlining.

Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
This rule checks both the contract definition and the call-site.
We evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
Although the contract only has one property status, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition (assumed to be $\PSValid$).
To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
We also descend into the subexpressions to check them; when checking the body and guarantee, we can assume that the rely holds.

\subsubsection{Blessing expressions and contracts}
\label{s:core:blessing}

Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
Blessing an expression requires a proof that, for all input streams, assuming the valid checks hold, then the unknown checks hold:

$$
\ruleIN{
  \forall \Sigma.~
  \semcheck{\Sigma}{\PSValid}{e}
  \implies
  \semcheck{\Sigma}{\PSUnknown}{e}
}{\text{bless}~e}{BlessExpression}
$$

We generally prove the required properties by first translating the program to an abstract transition system, as described in \autoref{s:transition}.

Blessing is different for contract definitions, as we need to separate the definition of the contract from the instantiation.
To check that a contract definition is valid, we show that if the rely clause is always true for a particular input, then the body satisfies the guarantee for the same inputs.
We also assume that the valid properties in the rely, body and guarantee hold, and show the corresponding unknown properties:

\begin{tabbing}
  \tt{MM}\= \tt{MMMM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}: \text{prop}$ = \\
  \> $\forall \Sigma.$
  \> $ (
    \semcheck{\Sigma}{\PSValid}{(\erely, \ebody, \eguar[x := \ebody])}
    ~\wedge~
    \bigstepalways{\Sigma}{\erely}
  ) $ \\
  \> $\implies$
  \> $(
    \semcheck{\Sigma}{\PSUnknown}{(\erely, \ebody, \eguar[x := \ebody])}
    ~\wedge~
    \bigstepalways{\Sigma}{\eguar[x := \ebody]}
    )$
\end{tabbing}

After proving that the contract is valid for all inputs, we can bless the contract definition.
Blessing the contract definition blesses the subexpressions for the rely, body and guarantee, but leaves the contract's \emph{instantiation} property status as unknown:
$$
\inferrule{
  \text{contract_valid}~\{ \erely \} ~\ebody~ \{ \eguar \}
}{\text{bless_contract}~\{\erely\}~\ebody~\{ \eguar\}}(\textsc{BlessContract})
$$

\subsection{Causality and metatheory}
\label{s:core:causality}

To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
This restriction checks that all recursive streams are guarded by a followed-by delay.
We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
This check ensures productivity of recursive streams, but can be too strict: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ mentions the recursive stream $x$ outside of the delay and is outlawed, but after inlining the let, it would be causal.
We hope to relax this restriction in future work.

The causality restriction gives us some important properties about the metatheory.
The most important property is that the dynamic semantics form a total function: given a streaming history and a causal expression, we can evaluate the expression to a value.
These properties are mechanised in \fstar{}.

\begin{theorem}[bigstep-is-total]
  For any non-empty streaming history $\Sigma$ and causal expression $e$, there exists some value $v$ such that $e$ evaluates to $v$ $(\bigstep{\Sigma}{e}{v})$.
\end{theorem}

The relationship between substitution and the streaming history is also important.
In general, we have a substitution property that states that evaluating a substituted expression $e[x := e']$ under some context $\Sigma$ is equivalent to evaluating $e'$ and adding it to the context $\Sigma$:

\begin{theorem}[bigstep-substitute]
  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ $(\bigsteps{\Sigma}{e'}{V})$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
  The converse is also true.
\end{theorem}

The big-step semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
The above substitution theorem can be used to show that, for causal expressions, these two semantics are equivalent.
We can additionally show that, when evaluating $e$ with $x \mapsto V$, the most recent value in $V$ does not affect the result.
This fact can be used to ``seed'' evaluation by starting with an arbitrary value:
\begin{theorem}[bigstep-rec-causal]
  For a streaming history $\Sigma; \sigma$ and a causal recursive expression $\xrec{x}{e}$, if $(\bigstep{\Sigma; \sigma}{e}{v})$, then updating $\sigma[x]$ with any value $v'$ results in the same value: $(\bigstep{\Sigma; \sigma[x \mapsto v']}{e}{v})$.
\end{theorem}

\section{Abstract transition systems}
\label{s:transition}

\begin{figure}
  \begin{tabbing}
  MM \= update: \= \kill
  \tt{type} system (\textit{input}: $\Gamma$) (\textit{result}: $\tau$) = \{ \\
  \> state:  \> $\Gamma$; \\
  \> free: \> $\Gamma$; \\
  \> init: \> heap \textit{state}; \\
  \> step: \> heap \textit{input} $\to$ heap \textit{free} $\to$ heap \textit{state} $\to$ step_result \textit{state} \textit{result}; \\
  \} \\
  \\
  \tt{type} step_result (\textit{state}: $\Gamma$) (\textit{result}: $\tau$) = \{ \\
  \> update:  \> heap \textit{state}; \\
  \> value: \> \textit{result}; \\
  \> rely: \> \tt{prop}; \\
  \> guar: \> \tt{prop}; \\
  \}
  \end{tabbing}
  \caption{Abstract transition system type definitions}
  \label{f:system-types}
\end{figure} 
To prove properties about Pipit programs, we translate to an \emph{abstract} transition system, so-called because it abstracts away the implementation details of contract instantiations.
For extraction we also translate to \emph{executable} transition systems, which we discuss in \autoref{s:extraction}.

\autoref{f:system-types} shows the types of transition systems.
A transition system is parameterised by its input context and the result type.
It also contains two internal contexts: firstly, the state context describes the private state required to execute the machine; secondly, the free context contains any extra input values that the transition system would like to existentially quantify over.
The free context is used to allow the system to ask for arbitrary values from the environment, when it would not otherwise be able to return a concrete value.

For recursive streams and contract instantiations, which hide their implementation, the natural translation to a transition system would involve existentially quantifying a result that satisfies the specification.
Unfortunately, using an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
Instead, the free context provides the step function with a fresh unconstrained value of the desired type, which the step function can then constrain.

Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
The step-result additionally contains two propositions; one for the `rely', or assumptions about the execution environment, and another for the `guarantee', or obligations that the transition system must show.
For the transition system corresponding to an expression $e$, these propositions are roughly analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.

For example, recall again the sum contract:
$$
\tt{let}~\text{sum}~\textit{ints}~=~\sumcontractX{\PSUnknown}{\textit{ints}}
$$

To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
The followed-by delay results in a local state variable called sum_fby, and we encode the existentially-quantified recursive stream as a free context variable called sum:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  \tt{let} sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> free  \> = (sum: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ sum_fby = \textit{f}.sum \}; \\
  \> \> \> value  \> = \textit{f}.sum; \\
  \> \> \> rely   \> = (\textit{f}.sum = \textit{s}.sum_fby + \textit{i}.ints) $\wedge$ \textit{i}.ints > 0; \\
  \> \> \> guar   \> = \textit{f}.sum > 0; \} \}
  \end{tabbing}

The initial state of 0 corresponds to the initial value of the followed-by.
In the step function, argument $i$ refers to the input heap containing \textit{i}.ints, $f$ refers to the free heap containing the recursive stream \textit{f}.sum, and $s$ refers to the state heap containing \textit{s}.sum_fby.
In the rely of the step result, \textit{f}.sum is constrained to be the translated body of the recursive stream.
The translated rely also includes the contract's rely that the input integer is positive.
Finally, the translated guarantee includes the contract's guarantee that the output is positive.

To verify the transition system, we prove inductively that if the rely always holds, then the guarantee holds; we discuss proofs of system validity further in \autoref{s:transition:ind}.

The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
For example, we can use the sum contract to implement the Fibonacci sequence with
$
  \tt{rec}~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
$.
This program does not require any input values, so we leave the input context empty.
The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
Similarly, the free context includes an entry for the recursive stream, and an entry for the abstract, underspecified value of the contract:

  \begin{tabbing}
  MM \= update: \= = \= update \= \kill
  \tt{let} fib_def: system () $\ZZ$ = \{ \\
  \> state   \> = (fib_fby: $\ZZ$); \\
  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
  \> init  \> = \{ fib_fby = 1 \}; \\
  \> step  \> = $\lambda{} i~f~s.$ \{ \\
  \> \> \> update \> = \{ fib_fby = \textit{f}.fib \}; \\
  \> \> \> value  \> = \textit{f}.fib; \\
  \> \> \> rely   \> = (\textit{f}.fib = \textit{f}.sum_contract) \\
  \> \> \>        \> $\wedge$ (\textit{s}.fib_fby > 0 $\implies$ \textit{f}.sum_contract > 0); \\
  \> \> \> guar   \> = \textit{s}.fib_fby > 0; \} \}
  \end{tabbing}

As before, the translated rely includes the assumption that the recursive stream's value (\textit{f}.fib) agrees with its body (\textit{f}.sum_contract).
Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (\textit{s}.fib_fby) is positive, then its output (\textit{f}.sum_contract) is positive too.
Finally, the translated guarantee encodes the obligation that the environment satisfies the \emph{contract's rely} -- the input to sum is positive.

Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
This minor optimisation is sound, as we define system validity to require all steps to satisfy the rely.

\subsection{Translation}

We now present the details of the translation.
For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations imply that the original expression satisfies the checked semantics.

\newcommand{\sysinit}[1]{\systrans{#1}_{\text{init}}}
\newcommand{\sysvalue}[1]{\systrans{#1}_{\text{value}}}
\newcommand{\sysupdate}[1]{\systrans{#1}_{\text{update}}}
\newcommand{\sysrely}[1]{\systrans{#1}_{\text{rely}}}
\newcommand{\sysguar}[1]{\systrans{#1}_{\text{guar}}}
\newcommand{\xctr}{\xcontractP{\PStatus}{e_r}{e_b}{\rawbind{x}{e_g}}}

\newcommand{\sysstate}[1]{\systrans{#1}_{\text{state}}}
\newcommand{\sysoracle}[1]{\systrans{#1}_{\text{free}}}

\begin{figure}
  \small
  \[
  \begin{array}{rrlr}
    \sysstate{v} & = & \cdot \\
    \sysstate{x} & = & \cdot \\
    \sysstate{p(\ov{e})} & = & \bigcup_i \sysstate{e_i} \\
    \sysstate{\xfby{v}{e}} & = & x_{\tt{fby}(e)}: \tau, \sysstate{e} & \text{(fresh $x_{\tt{fby}(e)}$)} \\
    \sysstate{\xrec{x}{e}} & = & \sysstate{e} \\
    \sysstate{\xlet{x}{e}{e'}} & = & \sysstate{e} \cup \sysstate{e'} \\
    \sysstate{\xcheckP{\PStatus}{e}} & = & \sysstate{e} \\
    \sysstate{\xctr} & = & \sysstate{e_r} \cup \sysstate{e_b} \\
    \\
    \sysoracle{v} & = & \cdot \\
    \sysoracle{x} & = & \cdot \\
    \sysoracle{p(\ov{e})} & = & \bigcup_i \sysoracle{e_i} \\
    \sysoracle{\xfby{v}{e}} & = & \sysoracle{e} \\
    \sysoracle{\xrec{x}{e}} & = & x: \tau, \sysoracle{e} \\
    \sysoracle{\xlet{x}{e}{e'}} & = & \sysoracle{e} \cup \sysstate{e'} \\
    \sysoracle{\xcheckP{\PStatus}{e}} & = & \sysoracle{e} \\
    \sysoracle{\xctr} & = & x: \tau, \sysoracle{e_r} \cup \sysstate{e_b} \\
  \end{array}
\]
\caption{Transition system typing contexts of expressions; for an expression $e$, $\sysstate{e} : \Gamma$ and $\sysoracle{e} : \Gamma$ describe the heaps used to store the expression's internal state and extra inputs.}
\label{f:system-translation-contexts}
\end{figure}

\begin{figure}
  \small
  \[
  \begin{array}{lrl}
    \sysinit{v} & = & () \\
    \sysvalue{v}(i, f, s) & = & v \\
\\
    \sysinit{x} & = & () \\
    \sysvalue{x}(i, f, s) & = & (i \cup f).x \\
\\
    \sysinit{p(\ov{e})} & = & \bigcup_i \sysinit{e_i} \\
    \sysvalue{p(\ov{e})}(i, f, s) & = & \text{prim-sem}(p, \ov{\sysvalue{e}(i, f, s)}) \\
    \sysupdate{p(\ov{e})}(i, f, s) & = & \bigcup_i \sysupdate{e_i}(i, f, s) \\
    \sysrely{p(\ov{e})}(i, f, s) & = & \bigwedge_i \sysrely{e_i}(i, f, s) \\
    \sysguar{p(\ov{e})}(i, f, s) & = & \bigwedge_i \sysguar{e_i}(i, f, s) \\
    \\
    \sysinit{\xfby{v}{e}} & = & \sysinit{e} \cup \{ x_{\tt{fby}(e)} \mapsto v \} \\
    \sysvalue{\xfby{v}{e}}(i, f, s) & = & s.x_{\tt{fby}(e)} \\
    \sysupdate{\xfby{v}{e}}(i, f, s) & = & \sysupdate{e}(i, f, s) \cup \{x_{\tt{fby}(e)} \mapsto \sysvalue{e}(i, f, s)\}\\
    \sysrely{\xfby{v}{e}}(i, f, s) & = & \sysrely{e}(i, f, s) \\
    \sysguar{\xfby{v}{e}}(i, f, s) & = & \sysguar{e}(i, f, s) \\
    \\
    \sysinit{\xrec{x}{e}} & = & \sysinit{e} \\
    \sysvalue{\xrec{x}{e}}(i, f, s) & = & f.x \\
    \sysupdate{\xrec{x}{e}}(i, f, s) & = & \sysupdate{e}(i, f, s)\\
    \sysrely{\xrec{x}{e}}(i, f, s) & = & \sysrely{e}(i, f, s) \\
          & \wedge & f.x = \sysvalue{e}(i, f, s) \\
    \sysguar{\xrec{x}{e}}(i, f, s) & = & \sysguar{e}(i, f, s) \\
    \\
    \sysinit{\xlet{x}{e}{e'}} & = & \sysinit{e} \cup \sysinit{e'} \\
    \sysvalue{\xlet{x}{e}{e'}}(i, f, s) & = & \sysvalue{e'}(i \cup \{ x \mapsto \sysvalue{e}(i, f, s)\}, f, s) \\
    \sysupdate{\xlet{x}{e}{e'}}(i, f, s) & = & \sysupdate{e'}(i \cup \{ x \mapsto \sysvalue{e}(i, f, s)\}, f, s) \\
      & \cup & \sysupdate{e}(i, f, s) \\
    \sysrely{\xlet{x}{e}{e'}}(i, f, s) & = & \sysrely{e'}(i \cup \{ x \mapsto \sysvalue{e}(i, f, s)\}, f, s) \\
      & \wedge & \sysrely{e}(i, f, s) \\
    \sysguar{\xlet{x}{e}{e'}}(i, f, s) & = & \sysguar{e'}(i \cup \{ x \mapsto \sysvalue{e}(i, f, s)\}, f, s) \\
      & \wedge & \sysguar{e}(i, f, s) \\
    \\
    \sysinit{\xcheckP{\PStatus}{e}} & = & \sysinit{e} \\
    \sysvalue{\xcheckP{\PStatus}{e}}(i, f, s) & = & () \\
    \sysupdate{\xcheckP{\PStatus}{e}}(i, f, s) & = & \sysupdate{e}(i, f, s) \\
    \sysrely{\xcheckP{\PStatus}{e}}(i, f, s) & = & (\PStatus = \PSValid \implies \sysvalue{e}(i, f, s)) \wedge \sysrely{e}(i, f, s) \\
    \sysguar{\xcheckP{\PStatus}{e}}(i, f, s) & = & (\PStatus = \PSUnknown \implies \sysvalue{e}(i, f, s)) \wedge \sysguar{e}(i, f, s) \\
    \\
    \sysinit{\xctr} & = & \sysinit{e_r} \cup \sysinit{e_g} \\
    \sysvalue{\xctr}(i, f, s) & = & f.x \\
    \sysupdate{\xctr}(i, f, s) & = & \sysupdate{e_r}(i, f, s) \cup \sysupdate{e_g}(i, f, s) \\
    \sysrely{\xctr}(i, f, s) & = & (\sysvalue{e_r}(i, f, s) \implies \sysvalue{e_g}(i, f, s)) \\
                            & \wedge & (\PStatus = \PSValid \implies \sysvalue{e_r}(i, f, s)) \\
                            & \wedge & \sysrely{e_r}(i, f, s) \\
                            & \wedge & (\sysvalue{e_r}(i, f, s) \implies \sysrely{e_g}(i, f, s) \\
    \sysguar{\xctr}(i, f, s) & = & (\PStatus = \PSUnknown \implies \sysvalue{e_r}(i, f, s)) \\
    & \wedge & \sysguar{e_r}(i, f, s) \wedge \sysguar{e_g}(i, f, s) \\
\end{array}
  \]
  \caption{Transition system semantics; for an expression $\Gamma \vdash e: \tau$, $\sysinit{e} : \text{heap~}\sysstate{e}$ is the initial state. For each field of the step-result type, we define a translation function that takes the input, free and state heaps: for example, we define the value-result of a step with type $\sysvalue{e}: \text{heap~}\Gamma \to \text{heap~}\sysoracle{e} \to \text{heap~}\sysstate{e} \to \tau$.}
  \label{f:system-translation}
\end{figure} 
\autoref{f:system-translation-contexts} defines the internal state and free contexts required for an expression.
For most expression forms, the state and free contexts are defined by taking the union of the contexts of subexpressions.
Followed-by delays introduce a local state variable $x_{\tt{fby}(e)}$ in which to store the most recent stream value.
We generate a fresh variable here, although the implementation uses de Bruijn indices.
Recursive streams and contracts both introduce new bindings into the free context; we assume that their binders $x$ are unique.

\autoref{f:system-translation} defines the translation for expressions.
Values and variables have no internal state.
For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
We omit the rely and guarantee definitions here; both are trivially true.

To translate primitives, we union together the initial states of the subexpressions; updating the state is similar.
For the rely and guarantee definitions, we take the conjunction: we can assume that all subexpressions rely clauses hold, and must show that all guarantees hold.

To translate a followed-by $\xfby{v}{e}$, we initialise the followed-by's unique binder $x_{\tt{fby}(e)}$ to the followed-by's default value $v$.
At each step, we return the value in the local state \emph{before} updating the local state to the subexpression's new value.

To translate a recursive expression $\xrec{x}{e}$ of type $\tau$, we require an arbitrary value $x: \tau$ in the free heap.
The rely proposition constrains the free variable $x$ to be the result of evaluating $e$ with the binding for $x$ passed along, thus closing the recursive loop.

To translate let-expressions $\xlet{x}{e}{e'}$, we extend the input heap with the value of $e$ before evaluating $e'$.
The presentation here duplicates the computation of the value of $e$, but the actual implementation introduces a single binding.

To translate a check property, we inspect the property status.
If the property is known to be valid, then we can assume the property is true in the rely clause.
Otherwise, we include the property as an obligation in the guarantee clause.
In either case, we also include the subexpression's rely and guarantee clauses.

Finally, to translate contract instantiations, we use the contract's rely and guarantee and ignore the body.
As with recursive expressions, we require an arbitrary value $x: \tau$ in the free heap.
The translation's rely allows us to assume that the contract definition holds: that is, the contract's rely implies the contract's guarantee.
If the contract instantiation is known to be valid, we can also assume that the contract's rely holds.
Otherwise, we include the contract's rely as an obligation by putting it in the translation's guarantee.

\subsection{Proof obligations and induction}
\label{s:transition:ind}

To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
A system satisfies its proof obligations if, for any sequence of steps that all satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.

Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step result's rely holds, then its guarantee holds:
\begin{tabbing}
  \tt{MM}\= \tt{MMMM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \text{heap}~\textit{input}) (f: \text{heap}~\textit{sys}\text{.free}).$ \\
  \> \tt{let} $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ \tt{in} \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, assuming that both steps satisfy the rely and the first step satisfied the inductive property:
\begin{tabbing}
  \tt{MM}\= \tt{MMMM} \= \tt{MMMMMMMMMMMMM} \= \kill
  \tt{let} $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i_0~ i_1: \text{heap}~\textit{input}) (f_0~ f_1: \text{heap}~\textit{sys}\text{.free}) (s_0: \text{heap}~\textit{sys}\text{.state}).$ \\
  \> \tt{let} $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ \tt{in} \\
  \> \tt{let} $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ \tt{in} \\
  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
\end{tabbing}

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive case to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

To reason about system validity in general, we define a predicate \emph{system_holds_all} that formally defines a valid system as: for all sequences of inputs and their corresponding steps, if all of the steps' relies hold, then the guarantees also hold.
Validity is implied by (k-)induction.

\subsection{Translation correctness proofs}
\label{s:transition:proof}

We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
The transition system itself is deterministic, but the free context provides the non-determinism which may occur from underspecified contracts; our theorem statement existentially quantifies the free heap.

\begin{figure}[t]
  \begin{mathpar}
    \boxed{\sysinv{\Sigma}{e}{s}}
  \end{mathpar}
  \begin{mathpar}
    \ruleAx{\sysinv{\Sigma}{v}{s}}{IValue}

    \ruleAx{\sysinv{\Sigma}{x}{s}}{IVar}

    \ruleIN{
      \sysinv{\Sigma}{e_1}{s}
      \quad \hdots \quad
      \sysinv{\Sigma}{e_n}{s}
    }{\sysinv{\Sigma}{p(\ov{e})}{s}}{IPrim}

    \ruleIN{
      s.x_{\tt{fby}(e')} = v
      \qquad
      \sysinv{\cdot}{e'}{s}
    }{
      \sysinv{\cdot}{\xfby{v}{e'}}{s}
    }{$\mbox{IFby}_0$}

    \ruleIN{
      \bigstep{\Sigma; \sigma}{e'}{s.x_{\tt{fby}(e')}}
      \qquad
      \sysinv{\Sigma; \sigma}{e'}{s}
    }{
      \sysinv{\Sigma; \sigma}{\xfby{v}{e'}}{s}
    }{$\mbox{IFby}_S$}

    \ruleIN{
      \bigsteps{\Sigma}{\xrec{x}{e}}{V}
      \qquad
      \sysinv{\Sigma[x \mapsto V]}{e}{s}
    }{
      \sysinv{\Sigma}{\xrec{x}{e[x]}}{s}
    }{IRec}

    \ruleIN{
      \bigsteps{\Sigma}{e}{V}
      \qquad
      \sysinv{\Sigma}{e}{s}
      \qquad
      \sysinv{\Sigma[x \mapsto V]}{e'}{s}
    }{
      \sysinv{\Sigma}{\xlet{x}{e}{e'[x]}}{s}
    }{ILet}

    \ruleIN{
      \sysinv{\Sigma}{e}{s}
    }{
      \sysinv{\Sigma}{\xcheckP{\PStatus}{e}}{s}
    }{ICheck}

    \ruleIN{
      \bigsteps{\Sigma}{\ebody}{V}
      \qquad
      \sysinv{\Sigma}{\erely}{s}
      \qquad
      \sysinv{\Sigma[x \mapsto V]}{\eguar}{s}
    }{
      \sysinv{\Sigma}{\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}{\eguar[x]}}}{s}
    }{IContract}
  \end{mathpar}

  \caption{Transition system state invariant}
  \label{f:system-invariant}
\end{figure} 
The results presented here rely heavily on the totality and substitution metaproperties described in \autoref{s:core:causality}.
\autoref{f:system-invariant} defines the invariant for the abstraction proof; the judgment form $\sysinv{\Sigma}{e}{s}$ checks that $s$ is a valid state heap.
We use the invariant to state that, if executing the transition system for $e$ on the entire streaming history $\Sigma$ results in state heap $s$, then $s$ is a valid state.

As most expressions do not modify the state heap, the invariant for most expressions simply descends into the subexpressions.
Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
The invariant for followed-by expressions asserts that the initial state of the followed-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
With this invariant, we can prove abstraction:

\begin{theorem}[translation-abstraction]
  For a well-typed causal expression $e$ and streaming history $\Sigma$, if $e$ evaluates to stream $V$ $(\bigsteps{\Sigma}{e}{V})$, then there exists a sequence of free heaps $\Sigma_{F}$ such that repeated application of the transition system's step results in $V$.
\end{theorem}

Finally, we can show the main entailment result that if the proof obligations hold on the system, then the original program is valid according to the checked semantics:

\begin{theorem}[translation-entailment]
  For a well-typed causal expression $e$ and its translated system $s$, if the system holds $(\text{system_holds_all}~s)$, and the checked properties in $e$ hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSValid}{e})$, then
  the unknown properties in $e$ also hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSUnknown}{e})$
\end{theorem}

The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them, allowing for modular proofs.

\section{Extraction}
\label{s:extraction}

Pipit can generate executable code which is suitable for real-time execution on embedded devices.
The code extraction uses a variation of the abstract transition system described in \autoref{s:transition}, with two main differences to ensure that the result is executable without relying on the environment to provide values for the free context.
Contracts are straightforward to execute by using the body of the contract rather than abstracting over the implementation.

To execute recursive expressions $\xrec{x}{e} : \tau$, we require an arbitrary value of type $\tau$ to seed the fixpoint, as described in \autoref{s:core:causality}.
We first call the step function to evaluate $e$ with $x$ bound to $\bot_\tau$.
This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
To get the correct state, we call the step function again, this time with $x$ bound to the correct value, $v$.

For example, for the \emph{sum} contract with body $\sumrecX{\textit{ints}}$, we generate an executable system that takes an input context containing integer variable \textit{ints}, with a single state variable for the followed-by, and returning an integer:

  \begin{tabbing}
  MM \= update: \= = \= \tt{let (sum,y) } \= \kill
  \tt{let} sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0; \}; \\
  \> step  \> = $\lambda{} i~s.$ \\
  \> \> \> \tt{let} $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ \tt{in} \\
  \> \> \> \tt{let} $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ \tt{in} \\
  \> \> \> \tt{let} $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ \tt{in} \\
  \> \> \> \tt{let} $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ \tt{in} \\
  \> \> \> $(\textit{sum}_0, s_1)$ \}
  \end{tabbing}

Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value $\textit{sum}_0$ used in the state.
This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.

This translation to transition systems is verified to preserve the original semantics.
The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we use the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).

\begin{theorem}[execution-equivalence]
  For a well-typed causal expression $e$ and streaming history $\Sigma$, $e$ evaluates to stream $V$ $(\bigsteps{\Sigma}{e}{V})$ if-and-only-if repeated application of the transition system's step on $\Sigma$ also results in $V$.
\end{theorem}

To extract the program, we use a \emph{hybrid embedding} as described in \cite{ho2022noise}, which is similar to staged-compilation.
The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to partially-evaluate the application of the translation to a particular input program.
This partial-evaluation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.

The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions:
  \begin{tabbing}
    \tt{struct sum_state { uint32_t sum_fby; }} \\
    \tt{void   sum_reset(struct sum_state* state);} \\
    \tt{int    sum_step(struct sum_state* state, uint32_t ints);}
  \end{tabbing}

The reset function takes the pointer to the state struct and sets it to its initial values.
The step function takes the pointer to the state struct and the inputs, and returns the result integer.
The state struct is updated in-place.
The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.
This interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular} and other synchronous languages.

Unfortunately, our current approach is unsuitable for generating imperative array code, as our pure transition system only supports pure arrays.
In the future, we intend to support efficient array computations and fix the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.

\section{Evaluation}
\label{s:evaluation}

To evaluate Pipit, we have implemented the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver~\cite{ISO11898_4}, described earlier in \autoref{s:motivation}.
The CAN bus is common in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic;
by having all nodes on the network adhere to the schedule, the reliability of periodic messages is significantly increased~\cite{fuehrer2001time}.

The TTCAN protocol can be implemented in two levels of increasing complexity.
In the first level, reference messages, which perform synchronisation between nodes, contain the index of the newly-started cycle.
In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
We implement the first level as it is more amenable to software implementation~\cite{hartwich2002integration}.

The implementation defines a streaming function that takes a stream describing the current time, the state of the hardware, and any received messages.
It returns a stream of commands to be performed, such as sending a particular reference message.
The implementation defines a pure streaming function.
To actually interact with the hardware we assume a small hardware-interop layer that reads from the hardware registers and translates the commands to hardware-register writes, but we have not yet implemented this.
We package the driver's inputs into a record for convenience:

\begin{tabbing}
  MM \= bus_status: \= \kill
  \tt{type} driver_input = \{ \\
    \> local_time: \> network_time_unit; \\
    \> mode_cmd: \> option mode; \\
    \> tx_status: \> tx_status; \\
    \> bus_status: \> bus_status; \\
    \> rx_ref: \> option ref_message; \\
    \> rx_app: \> option app_message_index; \\
    \}
\end{tabbing}

Here, the local-time field denotes the time-since-boot in \emph{network time units}, which are based on the bitrate of the underlying network bus.
The mode-command is an optional field which indicates requests from the application to enter configuration or execution mode.
The transmission-status describes the status of the last transmission request and may be none, success, or various error conditions.
The bus-status describes whether the bus is currently idle, busy, or in an error state.
The two receive fields denote messages received from the bus; for application-specific messages the time-triggered logic only needs the message identifier.

The driver-logic returns a stream of commands for the hardware-interop layer to perform:

\begin{tabbing}
  MM \= enable_acks: \= \kill
  \tt{type} commands = \{ \\
  \> enable_acks: \> bool; \\
  \> tx_ref: \>       option ref_message; \\
  \> tx_app: \> option app_message_index; \\
  \> tx_delay: \>     network_time_unit; \\
 \}
\end{tabbing}

The enable-acknowledgements field denotes whether the hardware should respond to messages from other nodes with an acknowledgement bit; in the case of a severe error acknowledgements are disabled, as the node must not write to the bus at all.
The transmit fields denote whether to send a reference message or an application-specific message.
For application-specific messages, the hardware-interop layer maintains the transmission buffers containing the actual message payload.
To meet the schedule as closely as possible, the driver anticipates the next transmission and includes a transmission delay to tell the hardware exactly when to send the next message.

\subsection{Runtime}

The implementation includes an extension of the trigger-fetch logic described in \autoref{s:motivation}, as well as state machines for tracking node synchronisation, master status and fault handling.
We generate real-time C code as described in \autoref{s:extraction}.
We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $140\mu{}s$, while the average was $90\mu{}s$ with a standard deviation of $1.5\mu{}s$.
Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
For a 125Kbit/s bus, a slot size of approximately 1,500$\mu{}s$ is required to achieve utilisation above 85 per cent.
For the maximum CAN bus rate of 1Mbit/s, the required slot size is $184\mu{}s$.
Further evaluation is required to ensure that the complete runtime including the hardware-interop layer is sufficient for full-speed CAN.

Our code generation can be improved in a few ways.
A common optimisation in Lustre is to fuse consecutive if-statements with the same condition~\cite{bourke2017formally}; such an optimisation seems useful here, as our treatment of optional values introduces repeated unpacking and repacking.
Some form of array fusion~\cite{robinson2017machine} may also be useful for removing redundant array operations.
Our current extraction generates a transition-system with a step function which returns a tuple of the updated state and result.
Composing these step functions together results in repeated boxing and unboxing of this tuple; we currently rely on the \fstar{} normaliser to remove this boxing.
In the future, we plan to build on the current proofs to implement a more-sophisticated encoding that introduces less overhead.

\subsection{Verification}

We have verified a simplified trigger-fetch mechanism, as presented earlier (\autoref{s:motivation}).
For comparison, we implemented the same logic in the Kind2 model-checker~\cite{champion2016kind2}.
The restrictions placed on the triggers array --- that triggers are sorted by time-mark, that there must be an adequate time-gap between a trigger and its next-enabled, and that a trigger's time-mark must be greater-than-or-equal-to its index --- are naturally expressed with quantifiers.
The Kind2 model-checker includes experimental array and quantifier support~\cite{kind2userdoc}.
Due to the experimental nature of these features, we had to work around some limitations: for example, the use of arrays and quantifiers disables IC3-based invariant generation; quantified variables cannot be used in function calls; and the use of top-level constant arrays caused runtime errors that rendered most properties invalid~\cite{kind2024toparray}.

We were able to express equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
\autoref{f:evaluation:recursion} illustrates this encoding with a simplified version of the next-trigger logic.

\begin{figure}
\begin{minipage}{0.45\textwidth}
\begin{tabbing}
  MM \= MM \= \kill
\tt{let rec} next ($i$: int) ($c$: cycle): \\
  \> Tot \> (option int) \\
  \>     \> (decreases (count - $i$)) = \\
\>  \tt{if} trigger_enabled $i$~$c$ \\
\>  \tt{then} Some $i$ \\
\>  \tt{else if} $i \ge \text{count} - 1$ \\
\>  \tt{then} None \\
\>  \tt{else} next $(i + 1)$~$c$
\end{tabbing}
\end{minipage}
\begin{minipage}{0.55\textwidth}
\begin{verbatim}
  function next(index:  int; c: cycle)
       returns (result: int)
  var next_array: int ^ COUNT;
  let
    next_array[i] =
      if trigger_enabled(COUNT - 1 - i, c)
      then COUNT - 1 - i
      else if i <= 0
      then NO_NEXT_TRIGGER
      else next_array[i - 1];
    result =
      next_array[COUNT - 1 - index];
  tel
\end{verbatim}
\end{minipage}
\caption{Left: next-trigger logic in \fstar{}; right: Kind2 encoding as array scan. In \fstar{}, the \emph{Tot $\tau$ (decreases \ldots)} syntax declares a total function with the given termination measure.
In Kind2, the \tt{int\^{}COUNT} syntax denotes the type of an array of integers of length \tt{COUNT}, while the \tt{next_array[i]} declaration defines the elements of the array as a function of the index \tt{i}.
}
\label{f:evaluation:recursion}
\end{figure}

We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
In TTCAN proper, the enabled set is implemented as a cycle-offset and repeat-factor.
Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
In our Pipit development, we can treat the definition of the cycle set abstractly.
However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
This limitation also makes the specification quite unwieldy, as we must manually inline any functions in quantified formulas.

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
For the simplified trigger-fetch mechanism, we were able to verify up to 64 triggers.
For reference, hardware implementations of TTCAN such as M_TTCAN support up to 64 triggers~\cite{bosch2019mttcan}.

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU time & wall-clock & CPU time & wall-clock & CPU time \\
  \hline

 1 & 1.48s&1.06s
 & 1.57s&2.26s
 & 5.25s&5.03s \\
 2 & 1.51s&1.26s
 & 1.71s&2.93s
 & 5.25s&5.03s \\
 4 & 1.57s&1.62s
 & 2.08s&4.78s
 & 5.25s&5.03s \\
 8 & 1.76s&3.07s
 & 4.21s&16.98s
 & 5.25s&5.03s \\
 16 & 3.36s&11.91s
 & 13.82s&65.57s
 & 5.25s&5.03s \\
 32 & 12.15s&62.38s
 & 269.14s&1230.05s
 & 5.25s&5.03s \\
 64 & 1701.01s&9096.99s
 &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
 128 &  \multicolumn{2}{c|}{(timeout)}  &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\

\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
The wall-clock time denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
The time limit was one hour.}
\label{f:evaluation:kind2-runtime}
\end{figure}

We plan to verify the remainder of the TTCAN implementation and publish it separately.
Prior work formalising TTCAN has variously modeled the protocol itself~\cite{saha2007finite, pan2014modeling,li2018formal},
instances of the protocol~\cite{guo2020model},
and abstract models of TTCAN implementations~\cite{leen2006modeling}, but we are unaware of any prior work that has verified an \emph{executable} implementation of TTCAN.

Separately, Pipit has also been used to implement and verify a real-time controller for a coffee machine reservoir control system~\cite{robinson2023pipit}.
The reservoir has a float switch to sense the water level and a solenoid to allow the intake of water.
The specification includes a simple model of the water reservoir and shows that the reservoir does not exceed the maximum level under different failure-mode assumptions.

\section{Related work}
\label{s:related-work}

Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays, records, or a foreign-function interface.
Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.

Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration, which is unlikely to result in acceptable performance.

Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
It may be possible to encode the quantifiers as fixed-size loops in those that support arrays, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.

These model-checkers have definite usability advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples and implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, even when the problem can be expressed, these model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not necessarily expand the trusted computing base --- that is, we could augment our end-to-end verified workflow with \emph{unverified but validated} invariant generation.

Recent work has also introduced a form of refinement types for Lustre~\cite{chen2022synchronous}.
Rather than using transition systems, this work generates self-contained verification conditions based on the types of streams.
Such a type-based approach promises to allow abstraction of the implementation details.
However, for general-purpose functions such as \emph{count_when} from \autoref{s:motivation}, it is not clear how to give it a specification that actually \emph{abstracts} the implementation: a simple specification that the result is within some range would hide too much and be insufficient for verifying the rest of the system.
For such functions, the best specification is likely to include a re-statement of the implementation itself.

The embedded language Copilot generates real-time C code for runtime monitoring~\cite{laurent2015assuring}.
Recent work has used translation validation to show that the generated C code matches the high-level semantics~\cite{scott2023trustworthy}.
Copilot supports model-checking via Kind2; however, the model-checking has a limited specification language and does not support contracts.

Early work embedding a denotational semantics of Lucid Synchrone in an interactive theorem prover focussed on the semantics itself, rather than proving programs~\cite{boulme2001clocked}.
There is ongoing work to construct a denotational semantics of Vélus for program verification~\cite{bourke2022towards}.
We believe that the hybrid SMT approach of \fstar{} will allow for a better mixture of automated proofs with manual proofs.
Compared to Vélus alone, the trusted computing base of Pipit is larger: we depend on all of \fstar{}, \lowstar{}'s unverified C code extraction and the Z3 SMT solver; in comparison, Vélus' C code generation is verified and does not depend on any SMT solver.

The deferred aspect of our proofs is similar to the deferred proofs of verification conditions for imperative programs, such as \cite{oconnor2019deferring}.
However, such verification conditions are \emph{syntactically} deferred so that the verification condition can be proved later; in our case, the verification conditions are \emph{semantically} deferred, so that more knowledge of the enclosing program can be exploited in the proof.
In imperative programs, this sort of extra knowledge is generally provided explicitly as loop invariants, and non-looping statements have their weakest precondition computed automatically.
In Lustre-style reactive languages such as ours, programs tend to be composed of many nested recursive streams, which perform a similar function to loops.
Explicitly specifying an invariant for each recursive stream would be cumbersome; deferring the proof allows such invariants to be implicit.

\section{Conclusion}

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that, by embedding pure \fstar{} functions for array operations, Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the semantics of checked properties and contracts; proof obligations generated by translation to abstract transition system are verified to correspond to these semantics.

In the future, we intend to verify the remainder of the TTCAN driver logic.
We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation; they may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.

We are interested in further pursuing the intersection of model-checking with interactive theorem proving.
A smart-contract called Djed \cite{zahnentferner2023djed} currently uses a mixture of Kind2 \cite{champion2016kind2} and manual Isabelle/HOL proofs to show that the contract is well-behaved.
In future work, we would like to further investigate whether Pipit's integration of streaming proofs with \fstar{}'s automated proof system would be able to provide similar proofs, without introducing any semantic gap between the two systems.

\bibliography{Main}

\end{document}

### 2024-04-03 2024pipit/expand-ecoop24-submit0.tex
diff

--- 2024-04-03 2024pipit/expand-20240403.tex
@@
 
-To be confident that our proofs hold on the real system, we need a single shared semantics for the compiler and the prover.
+To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
 In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
@@
   \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
-  \item we describe a \emph{checked semantics} for Pipit; after checking deferred properties, programs are \emph{blessed}, which marks their properties as valid (\autoref{s:core:checked});
+  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
   \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions rather than relations; composing functions results in simpler transition systems (\autoref{s:transition});
@@
   \item similarly, we offer a mechanised proof that the executable transition system preserves the original semantics (\autoref{s:extraction});
-  \item finally, we evaluate Pipit by implementing the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver and verifying an abstract model of a key component (\autoref{s:evaluation}).
+  \item finally, we evaluate Pipit by implementing the high-level logic of a time-triggered Controller Area Network (CAN) bus driver and verifying an abstract model of a key operation (\autoref{s:evaluation}).
 \end{itemize}
@@
 
-To introduce Pipit, we consider a \emph{time-triggered} network driver, which has a static schedule dictating the network traffic, and which all nodes on the network must adhere to.
-This driver is a simplification of the Time-Triggered Controller Area Network (TTCAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
-
-At a high level, the network schedule is described by a \emph{system matrix} which consists of rows of \emph{basic cycles}.
-Each basic cycle consists of a sequence of actions to be performed at specific time-marks.
-Actions in the schedule may not be relevant to all nodes; the node's \emph{node matrix} contains only the relevant actions.
-The node matrix is represented in memory by a \emph{triggers array} containing triggers sorted by their time-marks; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
-Reference messages start a new basic cycle; a subset of nodes, designated as leaders, send reference messages to synchronise the network.
-Watch alerts are generally placed after an expected reference message to signal an error if no reference message is received.
-
-\autoref{f:tt-systemmatrix-ok} (left) shows an example node matrix for a non-leader node.
-The matrix consists of two basic cycles C0 and C1 with messages sent at time-marks 0, 1 and 2.
-The node expects to receive a reference message at time-mark 7; the watch at time-mark 9 allows a grace period before triggering an error if the reference message is not received.
-\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding triggers array.
-
-\begin{figure}
-  \begin{minipage}{0.5\textwidth}
-\resizebox{\textwidth}{!}{
-\begin{tabular}{r|ccccc}
-   & TM0 & TM1 & TM2 & $\cdots$ & TM9 \\
-  \hline
-  C0 & SEND A & SEND B & - & $\cdots$ & WATCH \\
-  C1 & SEND A & - & SEND C & $\cdots$ & WATCH
-\end{tabular}}
-\end{minipage}
-\begin{minipage}{0.49\textwidth}
-\scriptsize
-\begin{verbatim}
- 0:{ time = 0; enabled = {C0,C1}; action = SEND(A); }
- 1:{ time = 1; enabled = {C0};    action = SEND(B); }
- 2:{ time = 2; enabled = {C1};    action = SEND(C); }
- 3:{ time = 9; enabled = {C0,C1}; action = WATCH;   }
-\end{verbatim}
-\end{minipage}
-
-\caption{Left: node matrix; right: corresponding triggers array configuration}
-\label{f:tt-systemmatrix-ok}
-\end{figure}
-
-The network has strict timing requirements which prohibit the driver from looping through the entire triggers array at each time-mark.
-Instead, the driver maintains an index that refers to the current trigger.
-At each time-mark, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
+To introduce Pipit, we consider a driver with a static schedule of \emph{triggers}, or actions to be performed at a particular time; this driver is a simplification of the time-triggered Controller Area Network (CAN) bus specification \cite{fuehrer2001time} which we will discuss further in \autoref{s:evaluation}.
 
@@
 
-We implement a streaming function \emph{count_when} to maintain the index into the triggers array; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
-At each step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the counter as-is.
+The schedule of our time-triggered driver is determined by a constant array of triggers, sorted by their associated time-mark.
+The driver maintains an index that refers to the current trigger.
+At each instant in time, the driver checks if the current trigger has expired or is inactive, and if so, it increments the index.
+We first implement a streaming function \emph{count_when} to maintain the index; the function takes a constant natural number \emph{max} and a stream of booleans \emph{inc}.
+At each time step, \emph{count_when} checks whether the current increment flag is true; if so, it increments the previous counter, saturating at the maximum; otherwise, it leaves the previous counter as-is.
 
@@
 
-\subsection{Restrictions on the triggers array}
-
-Our driver may fall behind when trying to execute certain schedules, as the driver only processes one trigger per time-mark.
-To ensure that the schedule can be executed on time, the triggers array must allow sufficient time for the driver to skip over any disabled triggers before the next enabled trigger starts.
-
-Recall our concrete triggers array from \autoref{f:tt-systemmatrix-ok}, which contained trigger 1 (SEND B at time-mark 1 on cycle C0), and trigger 2 (SEND C at time-mark 2 on cycle C1).
-We could postpone trigger 1 to send B at time-mark 2, as the corresponding cell in the node matrix is empty.
-However, we \emph{cannot} bring the trigger at index 2 forward to send message C at time-mark 1, as it takes two steps to reach trigger 2 from the start of the array.
-
-We impose three restrictions on \emph{valid} triggers arrays: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index, so that it is reachable in time from the start of the array.
+\subsection{The time-triggered system matrix}
+
+The schedule of the time-triggered network is abstractly described by a \emph{system matrix}, consisting of rows of \emph{basic cycles}, columns of \emph{transmission columns}, and cells of optional messages.
+Each basic cycle is identified by its cycle index and each transmission column has an associated time-mark.
+
+\autoref{f:tt-systemmatrix-ok} (left) shows an example system matrix with cycles \tt{C0} and \tt{C1} and transmission columns at time-marks 0, 1 and 2.
+For this example, we assume that one message can be sent per clock cycle. To execute this system matrix, we synchronise the local time to zero at the start of basic cycle \tt{C0}.
+After a basic cycle completes, the nodes on the network synchronise before execution continues to the next basic cycle.
+
+\autoref{f:tt-systemmatrix-ok} (right) shows the corresponding configuration for the triggers array.
+The enabled set denotes the basic cycles for which a trigger is active.
+
+The system has strict timing requirements which restrict how triggers can be defined.
+In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
+Additionally, the schedule must allow sufficient time for the driver to skip over the disabled triggers.
+Concretely, we could postpone trigger 1 to send message \tt{B} at time-mark 2, as triggers 1 and 2 have distinct cycles.
+However, we could not bring forward trigger 2 to send message \tt{C} at time-mark 1: the driver can only process one trigger per tick, and it takes two steps to reach trigger 2 from the start of the array.
+
+\begin{figure}
+  \begin{minipage}{0.38\textwidth}
+\begin{tabular}{r|ccc}
+   & TM0 & TM1 & TM2 \\
+  \hline
+  C0 & MSG A & MSG B & - \\
+  C1 & MSG A & - & MSG C
+\end{tabular}
+\end{minipage}
+\begin{minipage}{0.6\textwidth}
+\small
+\begin{verbatim}
+  0: { time_mark = 0; enabled = {C0,C1}; msg = A; }
+  1: { time_mark = 1; enabled = {C0};    msg = B; }
+  2: { time_mark = 2; enabled = {C1};    msg = C; }
+\end{verbatim}
+\end{minipage}
+
+\caption{Left: system matrix; right: corresponding triggers array configuration}
+\label{f:tt-systemmatrix-ok}
+\end{figure}
+
+We impose three restrictions on the triggers array: the time-marks must be sorted; there must be an adequate time-gap between any two triggers that are enabled on the same cycle index; and each trigger's time-mark must be greater-than-or-equal to its index.
 
@@
 We can now implement the trigger-fetch logic, which keeps track of the current trigger.
-We use the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
-We simplify our presentation here and only consider a constant cycle: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
+The trigger-fetch logic uses the \emph{count_when} streaming function to define the index of the current trigger; we tell \emph{count_when} to increment the index whenever the previous index has expired or is inactive in the current basic cycle.
+We simplify our presentation here and only consider a single cycle in isolation: the real system presented in \autoref{s:evaluation} has some extra complexity such as resetting the index, incrementing the cycle index at the start of a new cycle, and using machine integers.
 
@@
     \> \> $\tt{let } \textit{index} = \mbox{count_when}_{\PSValid}~\text{trigger_count}~\textit{inc} \tt{ in}$ \\
-    \> \> $\text{pose}~(\text{lemma_can_reach_next}~\textit{cycle}~\textit{index})$; \\
+    \> \> $\text{pose}_1~(\text{lemma_can_reach_next}~\textit{cycle})~\textit{index}$; \\
     \> \> $\xcheckP{\PSUnknown}{(\text{can_reach_next_active}~\textit{cycle}~\textit{time}~\textit{index})}$; \\
@@
 
-We use the $\textit{pose}$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it.
+We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 indicates that the lemma is being applied to one streaming argument (the index).
 We then state an invariant as a deferred property.
@@
 To help compose this function with the rest of the system, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
-The contract we state is that if we are called once per time-mark then we guarantee that we never encounter a late trigger.
+The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.
 
@@
   \> \tt{let} $\textit{contract} = \text{Contract.contract_of_stream}_1$ \{ \\
-  \> \> \tt{rely} = $(\lambda \textit{time}.~ \textit{time} = \xfby{0}{(\textit{time} + 1)});$ \\
+  \> \> \tt{rely} = $(\lambda \textit{time}.~ \text{time_no_skips}~\textit{time} );$ \\
   \> \> \tt{guar} = $(\lambda \textit{time index}.~ (\text{index_valid}~\textit{index} \wedge \text{enabled}~\textit{index}~\textit{cycle})$ \\
@@
 
-The key distinction between our streaming rely-guarantee contracts and imperative pre-post contracts is that the rely and guarantee are both \emph{streams} of booleans, rather than instantaneous predicates.
-In this case, the rely $(\textit{time} = \xfby{0}{(\textit{time} + 1)})$ checks that the current time is exactly one time-mark after the time at the previous \emph{tick} of computation.
-Expressing such a rely in an imperative setting would require extra encoding, as preconditions in imperative languages do not generally have an innate notion of the previous value with respect to a global shared clock.
-
-When \emph{trigger_fetch} is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
+When this function is used in other parts of the program, the caller must ensure that the environment satisfies the rely clause.
 In the core language, this is tracked by another deferred property status attached to the contract; we will discuss this further in \autoref{s:core}.
 
-\section{Pipit core language}
+\section{Core language}
 \label{s:core}
@@
 \begin{figure}
-  \includegraphics[width=\textwidth]{figures/core-structure-1920x450.pdf}
+  \includegraphics[width=\textwidth]{figures/core-structure.pdf}
 \caption{Architecture of Pipit.
@@
 The white boxes and dashed arrows are trusted components.
-The labels denote verified properties of the translation: abstraction~($\subseteq$), entailment of proof obligations~($\vdash$), and equivalence~($=$).
+The labels next to the arrows denote properties of the translation.
 }
@@
 The translation to abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
-The translation to abstract systems also generates proof obligations, which are verified to correspond to the proof obligations on the original program.
-The translation to executable transition systems is proven to be semantics-preserving, as is the subsequent translation to \lowstar{}.
+One property that still remains to be proven is that the proof obligations on the abstract transition system entail the original proof obligations (\autoref{s:transition}); this pending proof is denoted as negated entailment in the figure ($\not\vdash$).
+The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
 The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.
@@
   \]
-  \caption{Core grammar: expressions $e$, values $v$, primitive operations $p$, and property statuses $\PStatus$.}
+  \caption{Pipit core language grammar, which contains expressions $e$, values $v$, primitive operations $p$, and property statuses $\PStatus$.}
   \label{f:core-grammar}
@@
 
-Recursive streams are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
-As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined and counts from zero up, but the stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
+Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
+As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
 This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
@@
 
-Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ allow modular reasoning by replacing the implementation with an abstract specification.
-Contracts involve two verification conditions.
+Contracts $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ involve two verification conditions.
 Firstly, when a contract is \emph{defined}, the definer must prove that the body satisfies the contract: roughly, if $\erely$ is always true, then $\eguar[x := \ebody]$ is always true.
@@
 
-For example, the core expression $\sumrec$ computes the sum of values from a stream of integers \emph{ints} by defining a recursive stream $\sumvar$, which is delayed and given an initial value of zero.
-If we were to use this sum in a context that required a strictly positive integer, we could give it a contract that states that if the input stream is always positive, then the resulting sum is also positive:
-$$
-\sumcontract
-$$
-To be considered a valid program, we must prove that the contract definition itself holds, as with our earlier contract (\autoref{s:motivation:contract}).
-The unknown property status here allows us to defer the caller's proof that the input stream is always positive until the contract is used.
-
-The remaining grammatical constructs of \autoref{f:core-grammar} describe streams, value environments, types and type environments.
+For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.
+
 Streams $V$ are represented as a sequence of values; streaming history environments $\Sigma$ are streams of heaps.
 Types $\tau$ and type environments $\Gamma$ are standard.
-For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.
 
@@
   \caption{Typing rules for Pipit; the judgment $\typing{\Gamma}{e}{\tau}$ denotes that expression $e$ describes a \emph{stream} of values of type $\tau$.
-Auxiliary functions are used for values and primitive operations.}\label{f:core-typing}
+Two auxiliary functions are used for values and primitive operations; their definitions are standard and are omitted.}\label{f:core-typing}
 \end{figure} 
@@
 
-We use an auxiliary function $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$; for primitives $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ denotes that $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
+For values, we use an auxiliary definition $\mtypingval{v}{\tau}$ to denote that value $v$ has type $\tau$.
+Likewise, for primitives we use $\mtypingprim{p}{(\tau_1 \times \cdots \hdots \times \tau_n) \to \tau'}$ to denote that primitive $p$ takes arguments of type $\tau_i$ and returns a result of type $\tau'$.
 Primitives are pure, non-streaming functions.
@@
 
-Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types.
+Rule \textsc{TFby} states that expression $\xfby{v}{e}$ requires both $v$ and $e$ to have equal types; the result is the same type.
 
@@
 
-Rule \textsc{TCheck} states that checked properties $\xcheckP{\PStatus}{e}$ require a boolean property $e$.
-
-Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of type $\tau$.
+Rule \textsc{TCheck} states that statically checking a property $\xcheckP{\PStatus}{e}$ requires a boolean property $e$ and returns unit.
+
+Finally, rule \textsc{TContract} applies for a contract $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ with a body expression of some type $\tau$.
 The overall expression has result type $\tau$.
-Both rely and guarantee must be boolean expressions, and the guarantee can refer to the result as $x$.
+Both rely and guarantee clauses must be boolean expressions.
+Additionally, the guarantee clause can refer to the result value as $x$.
 
@@
 
-\begin{figure}[t]
+\begin{figure}
   \begin{mathpar}
@@
   \end{mathpar}
+
   \begin{mathpar}
+    \ruleAx{\bigstep{\Sigma}{v}{v}}{Value}
+    \quad
     \ruleAx{\bigstep{\Sigma; \sigma}{x}{\sigma(x)}}{Var}
-    \quad
-    \ruleAx{\bigstep{\Sigma}{v}{v}}{Value}
-
-    \ruleIN{
-      \bigstep{\Sigma}{e'[x := e]}{v}
-    }{
-      \bigstep{\Sigma}{\xlet{x}{e}{e'[x]}}{v}
-    }{Let}
 
@@
     \ruleIN{
+      \bigstep{\Sigma}{e'[x := e]}{v}
+    }{
+      \bigstep{\Sigma}{\xlet{x}{e}{e'[x]}}{v}
+    }{Let}
+
+    \ruleIN{
 }{
@@
   \end{mathpar}
+
   \begin{mathpar}
@@
 
-At a high level, evaluation unfolds recursive streams to determine a value.
-For example, to evaluate the earlier sum example with input $\textit{ints} = [1; 2]$, we start with the judgment:
-$$
-\bigstep{\sumintsonetwo}{\sumrec}{v}
-$$
-
-First, we unfold the recursive stream one step to get $(\xfby{0}{\sumrec}) + \textit{ints}$.
-Evaluation of primitives is standard.
-To evaluate variables, we look for the variable in the current (rightmost) heap:
-$$
-\ruleIN{
-}{
-  \bigstep{\sumintsonetwo}{\textit{ints}}{2}
-}{Var}
-$$
-
-For delays, we discard the current heap and continue evaluation with the history prefix:
-$$
-\ruleIN{
-  \bigstep{\sgl{\textit{ints} \mapsto 1}}{\sumrec}{1}
-}{
-  \bigstep{\sumintsonetwo}{\xfby{0}{\sumrec}}{1}
-}{$\mbox{Fby}_S$}
-$$
-
-Returning to \autoref{f:core-bigstep},
-rule {\sc Var} evalutes a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap.
-Rules {\sc Value} and {\sc Let} are standard.
-Rule {\sc Prim} evaluates a primitive $p$ applied to many arguments $e_1$ to $e_n$ by evaluating each argument separately; we then apply the primitive with prim-sem metafunction.
-
-For delay expressions $\xfby{v}{e}$, we have two cases depending on whether there is a previous value.
-When there is no previous value -- the streaming history only contains the current heap -- rule $\textsc{Fby}_1$ evaluates to the default value $v$.
-Otherwise, rule $\textsc{Fby}_S$ applies; we evaluate the previous value of $e$ by discarding the most recent entry from the streaming history.
+Rule {\sc Value} states that evaluating a value results in the value itself.
+
+Rule {\sc Var} states that to evalute a variable $x$ under some non-empty stream history $\Sigma; \sigma$, where $\sigma$ is the most recent heap, we look up the variable in $\sigma$.
+
+Rule {\sc Prim} states that to evaluate a primitive $p$ applied to many arguments $e_1$ to $e_n$, we evaluate each argument separately; we then apply the primitive with prim-sem metafunction.
+
+Rule $\textsc{Fby}_1$ evaluates a followed-by expression when the streaming history contains only a single element.
+Here, $\xfby{v}{e}$ evaluates to $v$, as there is no previous value of $e$ to use.
+
+Rule $\textsc{Fby}_S$ evaluates a followed-by expression when the streaming history contains multiple entries.
+In this case, $\xfby{v}{e}$ evaluates the previous value of $e$ by discarding the most recent entry from the streaming history.
 
@@
 
-Rule {\sc Check} ignores the property when evaluating check expressions.
-We do not dynamically check the property here; this is done in the checked semantics (\autoref{s:core:checked}).
-
-Similarly, rule {\sc Contract} ignores preconditions and postconditions when evaluating contracts.
-From an abstraction perspective, it would be valid to return an arbitrary value that satisfies the contract.
-However, such an abstraction would make evaluation non-deterministic and, for contracts with unsatisfiable postconditions, non-total.
-The deterministic and total nature of evaluation is key to our proofs and metatheory.
+Rule {\sc Let} is standard.
+
+Rule {\sc Check} states that check expressions always evaluate to unit.
+We do not perform a dynamic check that the property is true here; checking the truth of properties is dealt with in the checked semantics (\autoref{s:core:checked}).
+
+Rule {\sc Contract} states that contracts evaluate by just evaluating their body.
+Like with checks, we do not perform a dynamic check that the precondition and postcondition hold.
 
@@
 
-In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
-We call these the \emph{checked} semantics; they are comparable to checking runtime assertions.
-
-The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties and contracts of $e$ with status $\PStatus$ hold.
-The property status dictates which properties should be checked and which should be ignored.
-
-We consider a program to be \emph{valid} if its checks hold for all histories ($\forall \Sigma.~ \semcheck{\Sigma}{\PSValid}{e}$).
-The checked semantics are a specification describing what it means to be a valid program.
-We do not generally verify programs directly using the checked semantics; instead, we translate to an abstract transition system and construct the proofs there (\autoref{s:transition}).
-
-To check a property $(\xcheckP{\PStatus}{e})$ in history $\Sigma$, we check that $e$ is always true $(\bigstepalways{\Sigma}{e})$.
-
-Checking contracts is more involved.
-For whole-program correctness, it would suffice to check that a contract's rely and guarantee both hold.
-However, the purpose of contracts is to enable modular reasoning about parts of the program: we need to be able to check contracts independently of their context.
-Conceptually, then, contracts involve two kinds of checks: one for the definition and one for the call-site.
-To check a contract definition, we check that the body satisfies the guarantee for all \emph{valid} contexts -- that is, those where the rely holds.
-Then, to check a contract instance, we just need to check that the call-site satisfies the rely.
-
-For example, recall our earlier contract that the sum of strictly positive integers is positive:
-$$
-\tt{let}~\text{sum}~\textit{i} = \sumcontractX{\PSUnknown}{i}
-$$
-To check the contract definition on a concrete input $\textit{i} = [1; 2]$, we first evaluate the body:
-$$
-\bigsteps{\sumionetwo}{\sumrecX{i}}{[1; 3]}
-$$
-We then check that, assuming all inputs are positive, then all results are positive:
-$$
-\bigstepalways{\sumionetwo}{\textit{i} > 0} \implies 
-\bigstepalways{\sgl{\textit{i} \mapsto 1, \sumvar \mapsto 1}; \sgl{\textit{i} \mapsto 2, \sumvar \mapsto 3}}{sum > 0}
-$$
-
-It is critical that the rely is true \emph{at all points} in the stream.
-Consider if we had instead used the input stream $\textit{i} = [-10; 1]$; the rely is false at the first step, but is instantaneously true at the second step.
-In this case, the sum is $-10$ at the first step, and $-9$ at the second step.
-At both steps the output is negative and the guarantee is false, even though the rely becomes true at the second step.
-The contract itself remains valid, however, as the assumption is invalid: the input did not satisfy the rely at all steps.
-
-\begin{figure}[t]
+\begin{figure}
   \begin{mathpar}
@@
 
-    \ruleIN{\semcheck{\Sigma}{\PStatus}{e'}}{\semcheck{\Sigma}{\PStatus}{\xfby{v}{e'}}}{ChkFby}
-
-    \ruleIN{
-      \bigsteps{\Sigma}{\xrec{x}{e}}{V}
-      \and
-      \semcheck{\Sigma[x \mapsto V]}{\PStatus}{e}
+  \ruleAx{\semcheck{\sigma}{\PStatus}{\xfby{v}{e'}}}{$\mbox{ChkFby}_1$}
+
+    \ruleIN{\text{length}(\Sigma) > 0 \and \semcheck{\Sigma}{\PStatus}{e'}}{\semcheck{\Sigma; \sigma}{\PStatus}{\xfby{v}{e'}}}{$\mbox{ChkFby}_S$}
+
+    \ruleIN{
+      \semcheck{\Sigma}{\PStatus}{e[x := \xrec{x}{e}]}
     }{
@@
     \ruleIN{
-      \semcheck{\Sigma}{\PStatus}{e}
-      \and
-      \bigsteps{\Sigma}{e}{V}
-      \and
-      \semcheck{\Sigma[x \mapsto V]}{\PStatus}{e'}
+      \semcheck{\Sigma}{\PStatus}{e'[x := e]}
     }{
@@
     \inferrule{
-      \bigsteps{\Sigma}{\ebody}{V}
-      \\\\
       (\PStatus = \PStatus' \implies \bigstepalways{\Sigma}{\erely})
       \\\\
-      (\PStatus = \PSValid \implies \bigstepalways{\Sigma}{\erely} \implies \bigstepalways{\Sigma[x \mapsto V]}{\eguar})
+      (\PStatus = \PSValid \implies \bigstepalways{\Sigma}{\erely} \implies \bigstepalways{\Sigma}{\eguar[x := \ebody]})
       \\\\
@@
       \\\\
-      (\bigstepalways{\Sigma}{\erely} \implies \semcheck{\Sigma}{\PStatus}{\ebody} ~\wedge~ \semcheck{\Sigma[x \mapsto V]}{\PStatus}{\eguar})
+      (\bigstepalways{\Sigma}{\erely} \implies \semcheck{\Sigma}{\PStatus}{\ebody} ~\wedge~ \semcheck{\Sigma}{\PStatus}{\eguar[x := \ebody]})
     }{
@@
 
-The checked semantics of Pipit is defined in \autoref{f:core-check}.
+In addition to the big-step semantics above, we also define a judgment form for checking that the properties and contracts of a program hold for a particular streaming history.
+We call these the \emph{checked} semantics.
+Unlike an axiomatic semantics, the checked semantics operate on a concrete set of input streams.
+
+The checked semantics have the judgment form $\semcheck{\Sigma}{\PStatus}{e}$, which denotes that under streaming history $\Sigma$, the properties of $e$ with status $\PStatus$ hold.
+The property status dictates which properties should be checked and which should be ignored.
+
+To show that an expression $e$'s unknown properties hold, we prove that for all streaming histories $\Sigma$, assuming the valid properties hold ($\semcheck{\Sigma}{\PSValid}{e}$), then the unknown properties ($\semcheck{\Sigma}{\PSUnknown}{e}$) hold.
+The assumption here means that we do not have to re-check properties after proving them once.
+
+Contracts involve two proofs: one for the definition and one for the instantiation.
+To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
+Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
+The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
+
+To prove that a contract \emph{instantiation} (a call-site) is valid, we show that, under the calling environment, the rely clause is always true.
+Crucially, the proof can also use the fact that, if the rely is always true, then the guarantee is always true.
+This sort of feedback is necessary for proving properties of mutually-dependent calls.
+This circular dependency is well-founded as our causality check ensures that recursive streams are guarded by delays (\autoref{s:core:causality}).
+
+We define the checked semantics of Pipit in \autoref{f:core-check}.
+The checked semantics mostly follows the structure of the dynamic semantics, additionally checking any properties and contracts as they are encountered.
 
@@
 Rule {\sc ChkPrim} checks a primitive application by descending into the subexpressions.
-Similarly, rule {\sc ChkFby} descends into followed-by expressions.
-
-Rule {\sc ChkRec} checks a recursive-expression $\xrec{x}{e}$ by evaluating the overall expression to a stream of values $V$.
-The rule then extends the streaming environment $\Sigma$ with $x$ bound to the values from $V$; this extended environment is used to descend into the recursive expression.
-
-Rule {\sc ChkLet} checks a let-expression $\xlet{x}{e}{e'}$ descends into both sub-expressions.
-To check the body $e'$, the rule first evaluates $e$ and extends the streaming environment.
+
+Rules $\textsc{ChkFby}_1$ and $\textsc{ChkFby}_S$ are derived from the structure of the big-step rules  $\textsc{Fby}_1$ and $\textsc{Fby}_S$.
+At an input stream of length one, $\textsc{ChkFby}_1$ asserts that all subproperties hold for the (non-existent) previous values in the stream.
+At subsequent parts of the stream, $\textsc{ChkFby}_S$ discards the most recent element of the stream history and checks the subexpression with the previous inputs.
+
+Rules {\sc ChkRec} and {\sc ChkLet} both perform the same unfolding as the corresponding big-step rules and check the resulting expression.
 
@@
 
-Rule {\sc ChkCheck} checks the properties marked $\PStatus$ in an expression $\xcheckP{\PStatus'}{e}$.
-If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we evaluate the expression $e$ and require it to be true at all steps.
-We then unconditionally descend into the subexpression to check any nested properties.
-Such nested properties are unlikely to be written directly by the user, but might occur after inlining.
+Rule {\sc ChkCheck} applies when checking property status $\PStatus$ of an expression $\xcheckP{\PStatus'}{e}$.
+If the check-expression has the same status as what we are checking ($\PStatus = \PStatus'$), then we perform the actual check by evaluating the expression $e$ and requiring it to evaluate to a stream of trues.
+Otherwise, we do not need to evaluate the check-expression.
+In both cases, we descend into the expression and check its subexpressions, as they may have nested properties.
+Such nested properties are unlikely to be written directly by the user, but might occur after program transformations such as inlining.
 
 Rule {\sc ChkContract} applies when checking property status $\PStatus$ of a contract with expression $\xcontractP{\PStatus'}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$.
-This rule checks both the contract definition and the call-site.
-We evaluate the body to a stream $V$; these values are used to check that the body satisfies guarantee.
-Although the contract only has one property status, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition (assumed to be $\PSValid$).
-To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to hold.
+Although we only include one property status on the contract, conceptually there are two distinct properties: one for the caller ($\PStatus'$) and one for the definition itself (assumed to be $\PSValid$).
+To check the caller property when $\PStatus = \PStatus'$, we evaluate the rely $\erely$ and require it to be true.
 To check the definition property when $\PStatus = \PSValid$, we assume that the rely holds, and check that the body satisfies the guarantee.
@@
 Blessing is a meta-operation that replaces the property statuses in an expression so that all checks and contracts are marked as valid ($\PSValid$).
-Blessing an expression requires a proof that, for all input streams, assuming the valid checks hold, then the unknown checks hold:
+Blessing an expression requires a proof that the checked semantics hold for all input streams:
 
@@
 $$
-
-We generally prove the required properties by first translating the program to an abstract transition system, as described in \autoref{s:transition}.
 
@@
 
-To ensure that recursive streams have a computational interpretation, we implement a causality restriction, similar to standard Lustre~\cite{caspi1995functional}.
-This restriction checks that all recursive streams are guarded by a followed-by delay.
+To ensure that recursive streams have a computational interpretation, we require that all recursive streams are guarded by a followed-by delay.
 We implement this as a simple syntactic check: each $\xrec{x}{e}$ can only mention $x$ inside a followed-by.
-This check ensures productivity of recursive streams, but can be too strict: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ mentions the recursive stream $x$ outside of the delay and is outlawed, but after inlining the let, it would be causal.
-We hope to relax this restriction in future work.
+This check is stricter than necessary: for example, the expression $\xrec{x}{(\xlet{x'}{x + 1}{\xfby{0}{x'}})}$ does mention the recursive stream $x$ outside of the delay, but after inlining the let, it would be causal.
+We hope to relax this restriction somewhat in future work.
 
@@
 \begin{theorem}[bigstep-substitute]
-  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ $(\bigsteps{\Sigma}{e'}{V})$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
+  For a streaming history $\Sigma$ and causal expressions $e$ and $e'$, if $e[x := e']$ evaluates to a value $v$ $(\bigstep{\Sigma}{e}{v})$, then we can evaluate $e'$ to some stream $V$ and extend the streaming history to evaluate $e$ to the original value $(\bigstep{\Sigma[x \mapsto V]}{e}{v})$.
   The converse is also true.
@@
 
-The big-step semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
+The semantics in \autoref{f:core-bigstep} for a recursive expression $\xrec{x}{e}$ performs one step of recursion by substituting $x$ for the recursive expression.
 An alternative non-syntax-directed semantics would be to have the environment outside the semantics supply a stream $V$ such that if we extend the streaming history with $x \mapsto V$, then $e$ evaluates to $V$ itself.
@@
   MM \= update: \= \kill
-  \tt{type} system (\textit{input}: $\Gamma$) (\textit{result}: $\tau$) = \{ \\
+  \tt{type} system (input: $\Gamma$) (result: $\tau$) = \{ \\
   \> state:  \> $\Gamma$; \\
   \> free: \> $\Gamma$; \\
-  \> init: \> heap \textit{state}; \\
-  \> step: \> heap \textit{input} $\to$ heap \textit{free} $\to$ heap \textit{state} $\to$ step_result \textit{state} \textit{result}; \\
+  \> init: \> heap state; \\
+  \> step: \> heap input $\to$ heap free $\to$ heap state $\to$ step_result state result; \\
   \} \\
   \\
-  \tt{type} step_result (\textit{state}: $\Gamma$) (\textit{result}: $\tau$) = \{ \\
-  \> update:  \> heap \textit{state}; \\
-  \> value: \> \textit{result}; \\
+  \tt{type} step_result (state: $\Gamma$) (result: $\tau$) = \{ \\
+  \> update:  \> heap state; \\
+  \> value: \> result; \\
   \> rely: \> \tt{prop}; \\
@@
 
-For recursive streams and contract instantiations, which hide their implementation, the natural translation to a transition system would involve existentially quantifying a result that satisfies the specification.
-Unfortunately, using an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
+For contract instantiations, which abstract over the implementation, the natural translation to a transition system would involve an existential quantifier: ``there exists some value that satisfies the specification''.
+Unfortunately, such an existential quantifier requires a step \emph{relation} rather than a step \emph{function}.
 Using a step relation complicates the resulting transition system, as other operations such as primitive application must also introduce existential quantifiers; such quantifiers block simplifications such as partial-evaluation and result in a more complex transition system.
@@
 Back to \autoref{f:system-types}, the step-result contains the updated state for the transition system, as well as the result value.
-The step-result additionally contains two propositions; one for the `rely', or assumptions about the execution environment, and another for the `guarantee', or obligations that the transition system must show.
-For the transition system corresponding to an expression $e$, these propositions are roughly analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.
-
-For example, recall again the sum contract:
-$$
-\tt{let}~\text{sum}~\textit{ints}~=~\sumcontractX{\PSUnknown}{\textit{ints}}
-$$
-
-To verify the contract definition, we first translate it to an abstract transition system whose input environment contains an integer \emph{ints}, and whose result type is also an integer.
-The followed-by delay results in a local state variable called sum_fby, and we encode the existentially-quantified recursive stream as a free context variable called sum:
-
-  \begin{tabbing}
-  MM \= update: \= = \= update \= \kill
-  \tt{let} sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
-  \> state   \> = (sum_fby: $\ZZ$); \\
-  \> free  \> = (sum: $\ZZ$); \\
-  \> init  \> = \{ sum_fby = 0 \}; \\
-  \> step  \> = $\lambda{} i~f~s.$ \{ \\
-  \> \> \> update \> = \{ sum_fby = \textit{f}.sum \}; \\
-  \> \> \> value  \> = \textit{f}.sum; \\
-  \> \> \> rely   \> = (\textit{f}.sum = \textit{s}.sum_fby + \textit{i}.ints) $\wedge$ \textit{i}.ints > 0; \\
-  \> \> \> guar   \> = \textit{f}.sum > 0; \} \}
-  \end{tabbing}
-
-The initial state of 0 corresponds to the initial value of the followed-by.
-In the step function, argument $i$ refers to the input heap containing \textit{i}.ints, $f$ refers to the free heap containing the recursive stream \textit{f}.sum, and $s$ refers to the state heap containing \textit{s}.sum_fby.
-In the rely of the step result, \textit{f}.sum is constrained to be the translated body of the recursive stream.
-The translated rely also includes the contract's rely that the input integer is positive.
-Finally, the translated guarantee includes the contract's guarantee that the output is positive.
-
-To verify the transition system, we prove inductively that if the rely always holds, then the guarantee holds; we discuss proofs of system validity further in \autoref{s:transition:ind}.
-
-The translation for contract instantiations is similar, except that the contract body is replaced by an arbitrary value from the free context.
-For example, we can use the sum contract to implement the Fibonacci sequence with
-$
-  \tt{rec}~\fibvar.\, \text{sum}~(\xfby{1}{\fibvar})
-$.
-This program does not require any input values, so we leave the input context empty.
-The state context includes an entry for the $\xfby{1}{\fibvar}$ followed-by expression, but does not include the followed-by expressions inside the contract definition.
-Similarly, the free context includes an entry for the recursive stream, and an entry for the abstract, underspecified value of the contract:
-
-  \begin{tabbing}
-  MM \= update: \= = \= update \= \kill
-  \tt{let} fib_def: system () $\ZZ$ = \{ \\
-  \> state   \> = (fib_fby: $\ZZ$); \\
-  \> free  \> = (fib: $\ZZ$; sum_contract: $\ZZ$); \\
-  \> init  \> = \{ fib_fby = 1 \}; \\
-  \> step  \> = $\lambda{} i~f~s.$ \{ \\
-  \> \> \> update \> = \{ fib_fby = \textit{f}.fib \}; \\
-  \> \> \> value  \> = \textit{f}.fib; \\
-  \> \> \> rely   \> = (\textit{f}.fib = \textit{f}.sum_contract) \\
-  \> \> \>        \> $\wedge$ (\textit{s}.fib_fby > 0 $\implies$ \textit{f}.sum_contract > 0); \\
-  \> \> \> guar   \> = \textit{s}.fib_fby > 0; \} \}
-  \end{tabbing}
-
-As before, the translated rely includes the assumption that the recursive stream's value (\textit{f}.fib) agrees with its body (\textit{f}.sum_contract).
-Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (\textit{s}.fib_fby) is positive, then its output (\textit{f}.sum_contract) is positive too.
-Finally, the translated guarantee encodes the obligation that the environment satisfies the \emph{contract's rely} -- the input to sum is positive.
-
-Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
-This minor optimisation is sound, as we define system validity to require all steps to satisfy the rely.
-
-\subsection{Translation}
-
-We now present the details of the translation.
-For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations imply that the original expression satisfies the checked semantics.
+The step-result additionally contains two propositions for the `rely', or assumptions about the execution environment, and `guarantee', or obligations that the transition system must show.
+For the transition system corresponding to an expression $e$, these propositions are analogous to the known checked semantics $\semcheck{\Sigma}{\PSValid}{e}$ and unknown checks $\semcheck{\Sigma}{\PSUnknown}{e}$ respectively.
+
+Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics.
+The proof that the rely and guarantee propositions correspond to the checked semantics is future work.
 
@@
                             & \wedge & (\PStatus = \PSValid \implies \sysvalue{e_r}(i, f, s)) \\
-                            & \wedge & \sysrely{e_r}(i, f, s) \\
-                            & \wedge & (\sysvalue{e_r}(i, f, s) \implies \sysrely{e_g}(i, f, s) \\
+                            & \wedge & \sysrely{e_r}(i, f, s) \wedge \sysrely{e_g}(i, f, s) \\
     \sysguar{\xctr}(i, f, s) & = & (\PStatus = \PSUnknown \implies \sysvalue{e_r}(i, f, s)) \\
@@
 Followed-by delays introduce a local state variable $x_{\tt{fby}(e)}$ in which to store the most recent stream value.
-We generate a fresh variable here, although the implementation uses de Bruijn indices.
+We generate a fresh variable here, though the implementation uses de Bruijn indices.
 Recursive streams and contracts both introduce new bindings into the free context; we assume that their binders $x$ are unique.
@@
 \autoref{f:system-translation} defines the translation for expressions.
-Values and variables have no internal state.
+Values and expressions have no internal state.
 For variables, we look for the variable binding in either of the input or free heaps; bindings are unique and cannot occur in both.
@@
 
-To translate a followed-by $\xfby{v}{e}$, we initialise the followed-by's unique binder $x_{\tt{fby}(e)}$ to the followed-by's default value $v$.
-At each step, we return the value in the local state \emph{before} updating the local state to the subexpression's new value.
+To translate a followed-by $\xfby{v}{e}$, we initialise the follow-by's unique binder $x_{\tt{fby}(e)}$ to $v$.
+At each step, we return the value in the local state, before updating the local state to the subexpression's new value.
+The rely and guarantee differ from the checked semantics here: in the checked semantics, we check the subexpression on the previous inputs, but here we check the current subexpression.
+This means that a single step of the rely and guarantee do not exactly correspond to the checked semantics; however, we posit that they are equivalent for a rely and guarantee that has been proven to hold for any sequence of inputs.
 
@@
 
-\subsection{Proof obligations and induction}
-\label{s:transition:ind}
-
-To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
-A system satisfies its proof obligations if, for any sequence of steps that all satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.
-
-Inductive proofs on Lustre programs generally use a non-standard definition of induction, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
-This means that the base case must take a single step from the initial state to be able to state the property that, if the step result's rely holds, then its guarantee holds:
-\begin{tabbing}
-  \tt{MM}\= \tt{MMMM} \= \tt{MMMMMMMMMMMMM} \= \kill
-  \tt{let} $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
-  \> $\forall (i: \text{heap}~\textit{input}) (f: \text{heap}~\textit{sys}\text{.free}).$ \\
-  \> \tt{let} $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ \tt{in} \\
-  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
-\end{tabbing}
-
-For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, assuming that both steps satisfy the rely and the first step satisfied the inductive property:
-\begin{tabbing}
-  \tt{MM}\= \tt{MMMM} \= \tt{MMMMMMMMMMMMM} \= \kill
-  \tt{let} $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
-  \> $\forall (i_0~ i_1: \text{heap}~\textit{input}) (f_0~ f_1: \text{heap}~\textit{sys}\text{.free}) (s_0: \text{heap}~\textit{sys}\text{.state}).$ \\
-  \> \tt{let} $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ \tt{in} \\
-  \> \tt{let} $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ \tt{in} \\
-  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
-\end{tabbing}
-
-This inductive scheme also generalises to \emph{k-induction}, which allows the inductive case to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
-K-induction is a fairly standard invariant strengthening technique; intuitively, it allows the proof to use more context of the history of execution~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.
-
-To reason about system validity in general, we define a predicate \emph{system_holds_all} that formally defines a valid system as: for all sequences of inputs and their corresponding steps, if all of the steps' relies hold, then the guarantees also hold.
-Validity is implied by (k-)induction.
+In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
+The real semantics of the contract, however, requires the contract rely to be true \emph{at every step so far}.
+This difference is benign, as the inductive case of the proof of transition system validity assumes that both the rely and guarantee held at previous steps; if the rely also holds now, then it has held at every step so far.
 
@@
 We prove that the transition system is an abstraction of the dynamic semantics: that is, if the expression evaluates to $v$ under some context, then there exists some execution of the transition system that also results in $v$.
-The transition system itself is deterministic, but the free context provides the non-determinism which may occur from underspecified contracts; our theorem statement existentially quantifies the free heap.
-
-\begin{figure}[t]
+The transition system itself is deterministic, but the free context provides the non-determinism; our theorem  statement existentially quantifies the free heap.
+
+\begin{figure}
   \begin{mathpar}
@@
     \ruleIN{
-      s.x_{\tt{fby}(e')} = v
+      s.x_{\tt{fby}} = v
       \qquad
@@
     \ruleIN{
-      \bigstep{\Sigma; \sigma}{e'}{s.x_{\tt{fby}(e')}}
+      \bigstep{\Sigma; \sigma}{e'}{s.x_{\tt{fby}}}
       \qquad
@@
 Where new bindings are added, we use the dynamic semantics to extend the context with the new values.
-The invariant for followed-by expressions asserts that the initial state of the followed-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
-With this invariant, we can prove abstraction:
+The invariant for follow-by expressions asserts that the initial state of the follow-by is the default value; on subsequent steps, the state corresponds to the dynamic semantics.
 
@@
 
-Finally, we can show the main entailment result that if the proof obligations hold on the system, then the original program is valid according to the checked semantics:
-
-\begin{theorem}[translation-entailment]
-  For a well-typed causal expression $e$ and its translated system $s$, if the system holds $(\text{system_holds_all}~s)$, and the checked properties in $e$ hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSValid}{e})$, then
-  the unknown properties in $e$ also hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSUnknown}{e})$
-\end{theorem}
-
-The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
-Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them, allowing for modular proofs.
-
 \section{Extraction}
@@
 This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
-To get the correct state, we call the step function again, this time with $x$ bound to the correct value, $v$.
-
-For example, for the \emph{sum} contract with body $\sumrecX{\textit{ints}}$, we generate an executable system that takes an input context containing integer variable \textit{ints}, with a single state variable for the followed-by, and returning an integer:
-
-  \begin{tabbing}
-  MM \= update: \= = \= \tt{let (sum,y) } \= \kill
-  \tt{let} sum_def: system (ints: $\ZZ$) $\ZZ$ = \{ \\
-  \> state   \> = (sum_fby: $\ZZ$); \\
-  \> init  \> = \{ sum_fby = 0; \}; \\
-  \> step  \> = $\lambda{} i~s.$ \\
-  \> \> \> \tt{let} $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ \tt{in} \\
-  \> \> \> \tt{let} $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ \tt{in} \\
-  \> \> \> \tt{let} $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ \tt{in} \\
-  \> \> \> \tt{let} $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ \tt{in} \\
-  \> \> \> $(\textit{sum}_0, s_1)$ \}
-  \end{tabbing}
-
-Here, the step function takes heaps of the input and state contexts, and returns a pair of the result value and the updated state.
-The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
-The last two bindings recompute the state, this time with the correct recursive value $\textit{sum}_0$ used in the state.
-This duplication of work can often be removed by the partial evaluation and dead-code-elimination which we perform during code extraction.
+To get the correct state, we call the step function again, this time with $e$ bound to $v$.
 
@@
 The invariant is very similar to that of \autoref{s:transition:proof}, except that the invariant descends into the implementations of contracts.
-For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we use the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
+For the abstract systems we only showed abstraction; to prove that executable systems are equivalent to the original semantics, we rely on the fact that the original semantics and transition systems are both deterministic and total (\autoref{s:core:causality}).
 
@@
 The hybrid embedding involves a deep embedding of the Pipit core language, while the translation to executable transition systems produces a shallow embedding.
-We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to partially-evaluate the application of the translation to a particular input program.
-This partial-evaluation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.
-
-The generated C code for \emph{sum}\footnote{This interface is for a variant of the sum contract with 32-bit integers instead of unbounded integers.} includes a struct type to hold the state information, as well as reset and step functions:
-  \begin{tabbing}
-    \tt{struct sum_state { uint32_t sum_fby; }} \\
-    \tt{void   sum_reset(struct sum_state* state);} \\
-    \tt{int    sum_step(struct sum_state* state, uint32_t ints);}
-  \end{tabbing}
-
-The reset function takes the pointer to the state struct and sets it to its initial values.
-The step function takes the pointer to the state struct and the inputs, and returns the result integer.
-The state struct is updated in-place.
-The implementations of these functions avoid dynamic (heap) allocation and are suitable for embedded systems.
-This interface is standard for Lustre compilers~\cite{bourke2017formally,gerard2012modular} and other synchronous languages.
-
-Unfortunately, our current approach is unsuitable for generating imperative array code, as our pure transition system only supports pure arrays.
-In the future, we intend to support efficient array computations and fix the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
+We use the \fstar{} host language's normalisation-by-evaluation and tactic support~\cite{martinez2019meta} to specialise the application of the translation to a particular input program.
+This specialisation results in a concrete transition system that fits in the \lowstar{} subset of \fstar{}, which can then be extracted to statically-allocated C code~\cite{protzenko2017verified}.
+
+The translation for recursive streams described above calls the step function of the sub-stream twice, which can duplicate work.
+The normalisation strategy used to partially-evaluate the translation inlines the two occurrences of the step function, and is often able to remove the duplicate work, but this removal is not guaranteed.
+Our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
+In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
 Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
@@
 
-To evaluate Pipit, we have implemented the high-level logic of a Time-Triggered Controller Area Network (TTCAN) bus driver~\cite{ISO11898_4}, described earlier in \autoref{s:motivation}.
-The CAN bus is common in safety-critical automotive and industrial settings.
-The time-triggered network architecture defines a static schedule of network traffic;
-by having all nodes on the network adhere to the schedule, the reliability of periodic messages is significantly increased~\cite{fuehrer2001time}.
+To evaluate Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
+The CAN bus is commonly found in safety-critical automotive and industrial settings.
+The time-triggered network architecture defines a static schedule of network traffic.
+All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.
+
+At a high level, the schedule is described by a system matrix which consists of rows of basic cycles.
+Each basic cycle consists of a sequence of actions to be performed at particular time-marks.
+Actions in the schedule may not be relevant to all nodes, so each node has its own local array containing the relevant triggers; trigger actions include sending and receiving application-specific messages, sending reference messages, and triggering `watch' alerts.
+The trigger action for receiving an application-specific message checks that a particular message has been received since the trigger was last executed; depending on this, the driver increments or decrements a message-status-counter, which will in turn signal an error once the upper limit is reached.
+Reference messages start a new basic cycle and are used to synchronise the nodes.
+Watch alerts are generally placed after the expected end of the cycle and are used to signal an error if no reference message is received.
 
 The TTCAN protocol can be implemented in two levels of increasing complexity.
-In the first level, reference messages, which perform synchronisation between nodes, contain the index of the newly-started cycle.
+In the first level, reference messages contain the index of the newly-started cycle.
 In the second level, the reference messages also contain the value of a global fractional clock and whether any gaps have occurred in the global clock, which allows other nodes to calibrate their own clocks.
@@
 We evaluated the generated C code by executing with randomised inputs and measuring the worst-case-execution-time on a Raspberry Pi Pico (RP2040) microcontroller.
-The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $140\mu{}s$, while the average was $90\mu{}s$ with a standard deviation of $1.5\mu{}s$.
+The runtime of the driver logic is fairly stable: over 5,000 executions, the measured worst-case execution time was $114\mu{}s$, while the average was $107\mu{}s$ with a standard deviation of $2.3\mu{}s$.
 Earlier work on fault-tolerant TTCAN~\cite{short2007fault} describes the required slot sizes --- the minimum time between triggers --- to achieve bus utilisation at different bus rates.
@@
 
-We were able to express equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
-For example, the specification-only function that finds the next trigger is naturally recursive.
-Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
-Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
-\autoref{f:evaluation:recursion} illustrates this encoding with a simplified version of the next-trigger logic.
-
-\begin{figure}
-\begin{minipage}{0.45\textwidth}
-\begin{tabbing}
-  MM \= MM \= \kill
-\tt{let rec} next ($i$: int) ($c$: cycle): \\
-  \> Tot \> (option int) \\
-  \>     \> (decreases (count - $i$)) = \\
-\>  \tt{if} trigger_enabled $i$~$c$ \\
-\>  \tt{then} Some $i$ \\
-\>  \tt{else if} $i \ge \text{count} - 1$ \\
-\>  \tt{then} None \\
-\>  \tt{else} next $(i + 1)$~$c$
-\end{tabbing}
-\end{minipage}
-\begin{minipage}{0.55\textwidth}
-\begin{verbatim}
-  function next(index:  int; c: cycle)
-       returns (result: int)
-  var next_array: int ^ COUNT;
-  let
-    next_array[i] =
-      if trigger_enabled(COUNT - 1 - i, c)
-      then COUNT - 1 - i
-      else if i <= 0
-      then NO_NEXT_TRIGGER
-      else next_array[i - 1];
-    result =
-      next_array[COUNT - 1 - index];
-  tel
-\end{verbatim}
-\end{minipage}
-\caption{Left: next-trigger logic in \fstar{}; right: Kind2 encoding as array scan.
-In \fstar{}, the \emph{Tot $\tau$ (decreases \ldots)} syntax declares a total function with the given termination measure.
-In Kind2, the \tt{int\^{}COUNT} syntax denotes the type of an array of integers of length \tt{COUNT}, while the \tt{next_array[i]} declaration defines the elements of the array as a function of the index \tt{i}.
-}
-\label{f:evaluation:recursion}
-\end{figure}
-
-We compare against two Kind2 implementations: one corresponds closely to the Pipit development, while the other includes a critical simplification to modify the trigger-enabled set to be a single cycle index.
-In TTCAN proper, the enabled set is implemented as a cycle-offset and repeat-factor.
+We were able to reliably verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 16 elements; above that, we ran into intermittent runtime errors.
+For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
+
+We made a critical simplification in the Kind2 implementation, which was to modify the trigger-enabled set to be a single cycle index.
+In the specification, the enabled set is implemented as a cycle-offset and repeat-factor.
 Checking if a trigger is enabled in the current cycle requires nonlinear arithmetic, which is difficult for SMT solvers.
 In our Pipit development, we can treat the definition of the cycle set abstractly.
-However, in the Kind2 development, quantified formulas cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
-This limitation also makes the specification quite unwieldy, as we must manually inline any functions in quantified formulas.
+However, in the Kind2 development, quantifiers cannot contain function calls, which means that we cannot hide the implementation of the enabled-set check by providing an abstract contract.
+This limitation also makes the specification quite unwieldy, as functions must be manually inlined.
 
@@
 We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
-Both Kind2 and Pipit developments of the trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
-Ignoring whitespace and comments, the Pipit implementation of trigger-fetch has 26 lines of actual executable code, while the Kind2 code has 32.
-The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.
-
-We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
-For the simplified trigger-fetch mechanism, we were able to verify up to 64 triggers.
-For reference, hardware implementations of TTCAN such as M_TTCAN support up to 64 triggers~\cite{bosch2019mttcan}.
+Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
 
@@
   & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
-  size & wall-clock & CPU time & wall-clock & CPU time & wall-clock & CPU time \\
+  size & wall-clock & user-time & wall-clock & user-time & wall-clock & user-time \\
   \hline
 
- 1 & 1.48s&1.06s
- & 1.57s&2.26s
- & 5.25s&5.03s \\
- 2 & 1.51s&1.26s
- & 1.71s&2.93s
- & 5.25s&5.03s \\
- 4 & 1.57s&1.62s
- & 2.08s&4.78s
- & 5.25s&5.03s \\
- 8 & 1.76s&3.07s
- & 4.21s&16.98s
- & 5.25s&5.03s \\
- 16 & 3.36s&11.91s
- & 13.82s&65.57s
- & 5.25s&5.03s \\
- 32 & 12.15s&62.38s
- & 269.14s&1230.05s
- & 5.25s&5.03s \\
- 64 & 1701.01s&9096.99s
- &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
- 128 &  \multicolumn{2}{c|}{(timeout)}  &  \multicolumn{2}{c|}{(timeout)}  & 5.25s&5.03s \\
-
+  1 & 1.51s&1.33s
+  & 1.61s&2.26s
+  & 5.28s&5.03s \\
+  2 & 1.50s&1.29s
+  & 1.68s&2.85s
+  & 5.28s&5.03s \\
+  4 & 1.56s&1.59s
+  & 2.10s&5.05s
+  & 5.28s&5.03s \\
+  8 & 1.80s&3.12s
+  & 4.29s&18.12s
+  & 5.28s&5.03s \\
+  16 & 3.41s&12.29s
+  & 16.41s&71.23s
+  & 5.28s&5.03s \\
+  32 & 11.24s&67.27s
+  & 941.64s&3853.16s
+  & 5.28s&5.03s \\
 \end{tabular}
 \caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.
-The wall-clock time denotes the elapsed time that an engineer must spend waiting for the result; the CPU time denotes the total time spent computing by all of the CPU cores.
 The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.
-The time limit was one hour.}
+We were unable to verify arrays of size 64 with Kind2 as our verification timed out after three hours.}
 \label{f:evaluation:kind2-runtime}
@@
 Using existing Lustre tools to verify \emph{and} execute the time-triggered CAN driver from \autoref{s:motivation} is nontrivial.
-Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays, records, or a foreign-function interface.
+Compiling the triggers array with an unverified compiler such as Lustre~V6~\cite{jahier2016lustre} or Heptagon~\cite{gerard2012modular} is straightforward; however, the verified Lustre compiler Vélus~\cite{bourke2023verified} does not support arrays or a foreign-function interface.
 Recent work on translation validation for LustreC~\cite{brun2023equation} also does not yet support arrays.
@@
 Verifying the time-triggered CAN driver is trickier, as the restrictions placed on the triggers array --- that triggers are sorted by time-mark, there must be an adequate time-gap between a trigger and its next-enabled, and a trigger's time-mark must be greater-than-or-equal-to its index --- naturally require quantifiers.
-As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify the full logic for arrays up to the 64 triggers, which is the size supported by hardware implementations of TTCAN.
-Additionally, due to the limitations that require the constant triggers array to be passed as an argument, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration, which is unlikely to result in acceptable performance.
+As described in \autoref{s:evaluation}, Kind2 does include experimental array and quantifier support, but in our experiments was unable to verify arrays up to the 64 triggers supported by hardware implementations.
+Additionally, due to the limitations on top-level array definitions, compiling the program with Lustre~V6 would result in the entire triggers array being copied to the stack each iteration.
 
 Other model-checkers for Lustre such as Lesar~\cite{raymond2008synchronous}, JKind~\cite{gacek2018jkind} and the original Kind~\cite{hagen2008scaling} do not support quantifiers.
-It may be possible to encode the quantifiers as fixed-size loops in those that support arrays, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
+It may be possible to encode the quantifiers as fixed-size loops, but ensuring that these loops do not affect the execution or runtime complexity of the generated code does not appear to be straightforward.
 
@@
 In the future, we would like to investigate integrating Pipit with a model-checker via an unverified extraction: such an extraction may allow some of the usability benefits such as counterexamples and invariant generation.
-If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not necessarily expand the trusted computing base --- that is, we could augment our end-to-end verified workflow with \emph{unverified but validated} invariant generation.
+If this integration were used solely for debugging and suggesting candidate invariants, then such a change would not expand the trusted computing base.
 
@@
 Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
-We also introduced a checked semantics, which describes the semantics of checked properties and contracts; proof obligations generated by translation to abstract transition system are verified to correspond to these semantics.
+We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that the proof obligations generated by the abstract transition system match the checked semantics.
 
